### 2017-02-11 syntax-icfp17.tex
abstract

%% For double-blind review submission
\documentclass[acmlarge,review,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission
%\documentclass[acmlarge,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission
%\documentclass[acmlarge]{acmart}\settopmatter{}

%% Note: Authors migrating a paper from PACMPL format to traditional
%% SIGPLAN proceedings format should change 'acmlarge' to
%% 'sigplan,10pt'.

%% Some recommended packages.
\usepackage{booktabs}   %% For formal tables:
                        %% http://ctan.org/pkg/booktabs
\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
                        %% http://ctan.org/pkg/subcaption

%% Cyrus packages
\usepackage{microtype}

%% Listings
\usepackage{listings}
\lstset{tabsize=2, 
basicstyle=\ttfamily\fontsize{7.5pt}{1em}\selectfont, 
commentstyle=\itshape\ttfamily\color{gray}, 
stringstyle=\ttfamily\color{red},
mathescape=false,escapechar=!,
numbers=left, numberstyle=\scriptsize\color{mygray}, language=ML,moredelim=[il][\sffamily]{?},showspaces=false,showstringspaces=false,xleftmargin=0pt, numbersep=3pt, morekeywords=[1]{tyfam,opfam,let,fn,val,def,casetype,objtype,metadata,of,*,datatype,new,toast,syntax,module,where,import,for,ana,syn,opcon,tycon,metasignature,metamodule,metasig,metamod,static,at,by,tycase,mod,macro,match,pattern,in,patterns,expressions,implicit,forall,exptsm,pattsm},deletekeywords={double},classoffset=0,
aboveskip=2pt,belowskip=0pt,
moredelim=**[is][\color{red}]{SSTR}{ESTR},
moredelim=**[is][\color{OliveGreen}]{SHTML}{EHTML},
moredelim=**[is][\color{purple}]{SCSS}{ECSS},
moredelim=**[is][\color{brown}]{SSQL}{ESQL},
moredelim=**[is][\color{orange}]{SCOLOR}{ECOLOR},
moredelim=**[is][\color{magenta}]{SPCT}{EPCT}, 
moredelim=**[is][\color{gray}]{SNAT}{ENAT}, 
moredelim=**[is][\color{blue}]{SURL}{EURL},
moredelim=**[is][\color{SeaGreen}]{SQT}{EQT},
moredelim=**[is][\color{Periwinkle}]{SGRM}{EGRM},
moredelim=**[is][\color{YellowGreen}]{SID}{EID},
moredelim=**[is][\color{Sepia}]{SUS}{EUS},
deletestring=[d]{"},
}
\newcommand{\liv}[1]{\lstinline{#1}}
\newcommand{\li}[1]{\lstinline{#1}}

\makeatletter\if@ACM@journal\makeatother
%% Journal information (used by PACMPL format)
%% Supplied to authors by publisher for camera-ready submission
\acmJournal{PACMPL}
\acmVolume{1}
\acmNumber{1}
\acmArticle{1}
\acmYear{2017}
\acmMonth{1}
\acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}
\else\makeatother
%% Conference information (used by SIGPLAN proceedings format)
%% Supplied to authors by publisher for camera-ready submission
\acmConference[PL'17]{ACM SIGPLAN Conference on Programming Languages}{January 01--03, 2017}{New York, NY, USA}
\acmYear{2017}
\acmISBN{978-x-xxxx-xxxx-x/YY/MM}
\acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}
\fi

%% Copyright information
%% Supplied to authors (based on authors' rights management selection;
%% see authors.acm.org) by publisher for camera-ready submission
\setcopyright{none}             %% For review submission
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\copyrightyear{2017}           %% If different from \acmYear

%% Bibliography style
\bibliographystyle{ACM-Reference-Format}
%% Citation style
%% Note: author/year citations are required for papers published as an
%% issue of PACMPL.
\citestyle{acmauthoryear}   %% For author/year citations

\begin{document}

%% Title information
\title[Reasonably Programmable Syntax]{Reasonably Programmable Syntax}         %% [Short Title] is optional;
                                        %% when present, will be used in
                                        %% header instead of Full Title.
% \titlenote{with title note}             %% \titlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'
% \subtitle{Subtitle}                     %% \subtitle is optional
% \subtitlenote{with subtitle note}       %% \subtitlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'

%% Author information
%% Contents and number of authors suppressed with 'anonymous'.
%% Each author should be introduced by \author, followed by
%% \authornote (optional), \orcid (optional), \affiliation, and
%% \email.
%% An author may have multiple affiliations and/or emails; repeat the
%% appropriate command.
%% Many elements are not rendered, but should be provided for metadata
%% extraction tools.

%% Author with single affiliation.
\author{First1 Last1}
\authornote{with author1 note}          %% \authornote is optional;
                                        %% can be repeated if necessary
\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
  \position{Position1}
  \department{Department1}              %% \department is recommended
  \institution{Institution1}            %% \institution is required
  \streetaddress{Street1 Address1}
  \city{City1}
  \state{State1}
  \postcode{Post-Code1}
  \country{Country1}
}
\email{first1.last1@inst1.edu}          %% \email is recommended

%% Author with two affiliations and emails.
\author{First2 Last2}
\authornote{with author2 note}          %% \authornote is optional;
                                        %% can be repeated if necessary
\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
  \position{Position2a}
  \department{Department2a}             %% \department is recommended
  \institution{Institution2a}           %% \institution is required
  \streetaddress{Street2a Address2a}
  \city{City2a}
  \state{State2a}
  \postcode{Post-Code2a}
  \country{Country2a}
}
\email{first2.last2@inst2a.com}         %% \email is recommended
\affiliation{
  \position{Position2b}
  \department{Department2b}             %% \department is recommended
  \institution{Institution2b}           %% \institution is required
  \streetaddress{Street3b Address2b}
  \city{City2b}
  \state{State2b}
  \postcode{Post-Code2b}
  \country{Country2b}
}
\email{first2.last2@inst2b.org}         %% \email is recommended

%% Paper note
%% The \thanks command may be used to create a "paper note" ---
%% similar to a title note or an author note, but not explicitly
%% associated with a particular element.  It will appear immediately
%% above the permission/copyright statement.
% \thanks{with paper note}                %% \thanks is optional
                                        %% can be repeated if necesary
                                        %% contents suppressed with 'anonymous'

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
Programmers, 
like ``pen-and-paper'' mathematicians, often seek to define ``syntactic sugar'' that  lowers the syntactic cost of common idioms. 
Unfortunately, the most direct mechanisms available to programmers, e.g. \li{camlp4} \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, do not support modular reasoning about determinism (i.e. separately defined forms can conflict syntactically with one another), and they obscure the type and binding structure of the program. We argue therefore that these mechanisms are unreasonable for programming ``in the large''.

This paper formally introduces \emph{parametric typed syntax macros} (pTSMs), which give library providers programmatic control over both the parsing and expansion of expressions and patterns of \emph{generalized literal form} at a specified type or parameterized family of types. Expansion is \emph{strictly hygienic}, meaning that expansions can refer to external bindings only via spliced terms or explicit module parameters, and expansions do not reveal internal bindings to spliced terms or to the remainder of the program. 
% Partial parameter application lowers the syntactic cost of this strict style. 
This design ensures that clients are able to reason about the binding structure of the program while holding the expansion of the program abstract. The system needs only convey to clients a \emph{segmentation} of each literal that gives the locations of the spliced terms. The system can determine this {segmentation} automatically and convey it to the client via syntax highlighting, or by presenting the client with a context-free grammar (without revealing the semantic actions associated with each production.) 
We argue that this mechanism occupies a ``sweet spot'' in the design space, in that it captures many common syntactic idioms while avoiding the problem of syntactic conflicts by construction and supplying clients with clear abstract reasoning principles.

% \emph{Typed syntax macros (TSMs)}, proposed in a recent short paper by Omar et al. \cite{sac15}, give library providers programmatic control over the parsing and expansion of only terms of {(generalized) literal form}. This appears to occupy a ``sweet spot'' in that it captures many common syntactic idioms while avoiding the problem of conflict. TSMs also maintain a reasonable type and binding discipline.% In particular, clients can use any combination of TSMs in a program without needing to consider conflicts between them, and the language validates each expansion that a TSM generates to maintain 1) a \emph{type discipline} (meaning clients can determine the type of an unexpanded expression without examining its expansion directly); 
% %and 2) a \emph{hygienic binding discipline}.
% %, meaning that the expansion cannot make any assumptions about bindings at the application site, nor  introduce ``hidden bindings'' into subterms.

% TSMs have only been described minimally -- it is not clear how they should be adapted for integration into languages like ML that support {pattern matching}, {parameterized datatypes}, {modules} and abstract types. Moreover, the prior work makes several simplifying assumptions related to binding that are impractically restrictive.% bno mechanism for binding values for use across TSM definitions has been described and the hygiene mechanism makes giving the expansions that they generate access to ``helper functions'' awkward.

% This paper gives a complete account of TSMs that addresses these deficiencies. In particular, we 1) integrate TSMs with pattern matching; 2) introduce a distinct static phase of evaluation, which gives TSM definitions access to libraries; and 3) introduce type and module parameters, which serve two purposes: they allow for TSMs that operate uniformly at a parameterized family of types (rather than only at a single type), and they give expansions explicit, hygienic access to libraries. Support for partial application  of parameters lowers the syntactic cost of this explicit approach.

% Put succinctly, we design a programming language in the ML tradition with a \emph{reasonably} programmable syntax.
\end{abstract}

%% 2012 ACM Computing Classification System (CSS) concepts
%% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
% \begin{CCSXML}
% <ccs2012>
% <concept>
% <concept_id>10011007.10011006.10011008</concept_id>
% <concept_desc>Software and its engineering~General programming languages</concept_desc>
% <concept_significance>500</concept_significance>
% </concept>
% <concept>
% <concept_id>10003456.10003457.10003521.10003525</concept_id>
% <concept_desc>Social and professional topics~History of programming languages</concept_desc>
% <concept_significance>300</concept_significance>
% </concept>
% </ccs2012>
% \end{CCSXML}

% \ccsdesc[500]{Software and its engineering~General programming languages}
% \ccsdesc[300]{Social and professional topics~History of programming languages}
%% End of generated code

%% Keywords
%% comma separated list
% \keywords{keyword1, keyword2, keyword3}  %% \keywords is optional

%% \maketitle
%% Note: \maketitle command must come after title commands, author
%% commands, abstract environment, Computing Classification System
%% environment and commands, and keywords command.
\maketitle

\section{Introduction}

Text of paper \ldots

%% Acknowledgments
\begin{acks}                            %% acks environment is optional
                                        %% contents suppressed with 'anonymous'
  %% Commands \grantsponsor{<sponsorID>}{<name>}{<url>} and
  %% \grantnum[<url>]{<sponsorID>}{<number>} should be used to
  %% acknowledge financial support and will be used by metadata
  %% extraction tools.
  This material is based upon work supported by the
  \grantsponsor{GS100000001}{National Science
    Foundation}{http://dx.doi.org/10.13039/100000001} under Grant
  No.~\grantnum{GS100000001}{nnnnnnn} and Grant
  No.~\grantnum{GS100000001}{mmmmmmm}.  Any opinions, findings, and
  conclusions or recommendations expressed in this material are those
  of the author and do not necessarily reflect the views of the
  National Science Foundation.
\end{acks}

%% Bibliography
\bibliography{../../papers/research}

%% Appendix
\appendix
\section{Appendix}

Text of appendix \ldots

\end{document}

### 2017-02-17 syntax-icfp17.tex
outline and beginning of intro from thesis

@@
 \usepackage{microtype}
+\usepackage{todonotes}
 
 %% Listings
+\definecolor{mygray}{rgb}{0.75,0.75,0.75}
 \usepackage{listings}
 \lstset{tabsize=2, 
-basicstyle=\ttfamily\fontsize{7.5pt}{1em}\selectfont, 
+basicstyle=\ttfamily\fontsize{10pt}{1em}\selectfont, 
 commentstyle=\itshape\ttfamily\color{gray}, 
 stringstyle=\ttfamily\color{red},
-mathescape=false,escapechar=!,
+mathescape=false,escapechar=~,
 numbers=left, numberstyle=\scriptsize\color{mygray}, language=ML,moredelim=[il][\sffamily]{?},showspaces=false,showstringspaces=false,xleftmargin=0pt, numbersep=3pt, morekeywords=[1]{tyfam,opfam,let,fn,val,def,casetype,objtype,metadata,of,*,datatype,new,toast,syntax,module,where,import,for,ana,syn,opcon,tycon,metasignature,metamodule,metasig,metamod,static,at,by,tycase,mod,macro,match,pattern,in,patterns,expressions,implicit,forall,exptsm,pattsm},deletekeywords={double},classoffset=0,
-aboveskip=2pt,belowskip=0pt,
+aboveskip=4pt,belowskip=2pt,
 moredelim=**[is][\color{red}]{SSTR}{ESTR},
@@
 Programmers, 
-like ``pen-and-paper'' mathematicians, often seek to define ``syntactic sugar'' that  lowers the syntactic cost of common idioms. 
-Unfortunately, the most direct mechanisms available to programmers, e.g.
+like ``pen-and-paper'' mathematicians, often seek to define new ``syntactic sugar'' that  lowers the syntactic cost of common idioms. 
+Unfortunately, the most direct mechanisms of syntactic control available to programmers, e.g.
 \li{camlp4} \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, do not support modular reasoning about determinism (i.e.
@@
 We argue therefore that these mechanisms are unreasonable for programming ``in the large''.
+Typed term-rewriting macro systems (e.g.
+as available in Scala) are more reasonable, but give programmers limited syntactic control.
 
 This paper formally introduces \emph{parametric typed syntax macros} (pTSMs), which give library providers programmatic control over both the parsing and expansion of expressions and patterns of \emph{generalized literal form} at a specified type or parameterized family of types.
-Expansion is \emph{strictly hygienic}, meaning that expansions can refer to external bindings only via spliced terms or explicit module parameters, and expansions do not reveal internal bindings to spliced terms or to the remainder of the program.
+Expansion is \emph{strictly hygienic}, meaning that 1) expansions can refer to external bindings only via spliced terms or supplied module parameters; and 2)  expansions do not reveal internal bindings to spliced terms or to the remainder of the program.
 % Partial parameter application lowers the syntactic cost of this strict style. 
-This design ensures that clients are able to reason about the binding structure of the program while holding the expansion of the program abstract.
-The system needs only convey to clients a \emph{segmentation} of each literal that gives the locations of the spliced terms.
-The system can determine this {segmentation} automatically and convey it to the client via syntax highlighting, or by presenting the client with a context-free grammar (without revealing the semantic actions associated with each production.)
+This design ensures that clients are able to reason about the binding structure of the program while holding the expansion of the program abstract -- the system needs only convey to clients the \emph{segmentation} of each literal, which gives the locations of the spliced terms.
+%The system can determine this {segmentation} automatically and convey it to the client via syntax highlighting, or by presenting the client with a context-free grammar (without revealing the semantic actions associated with each production.)
 We argue that this mechanism occupies a ``sweet spot'' in the design space, in that it captures many common syntactic idioms while avoiding the problem of syntactic conflicts by construction and supplying clients with clear abstract reasoning principles.
@@
 \section{Introduction}
+Programming languages commonly build in ``syntactic sugar'' that decreases the syntactic cost of working with certain standard library constructs.   
+For example, Standard ML and many other functional languages build in syntactic sugar for constructing and pattern matching on lists.
 
-Text of paper \ldots
+Third-party library providers are, justifiably, envious of the privileged status of standard library constructs -- they would like to be able to define syntactic sugar for constructing and pattern matching on values of the types that they have defined.
+For example, 1) the provider of a ``collections'' library might like to define syntactic sugar for finite sets and dictionaries; 2) the provider of a ``web programming'' library might like syntactic sugar for HTML and JSON values; and 3) the provider of a ``chemistry'' library might like syntactic sugar for chemical structures based on the SMILES standard \cite{anderson1987smiles}.
+This has led to efforts to eliminate derived forms  specific to standard library constructs from language definitions in favor of mechanisms that give more syntactic control to library providers.
+
+\subsection{Syntax Dialects are Unreasonable}
+One approach available to library providers is to use a syntax definition system like Camlp4 \cite{ocaml-manual}, Copper \cite{conf/gpce/WykS07} or SugarJ/Sugar* \cite{erdweg2011sugarj,erdweg2013framework} to construct a \emph{syntax dialect}, i.e.
+a new syntax definition that extends the original syntax definition with new derived forms.
+%library-specific (a.k.a. ``domain-specific'') 
+For example, Ur/Web extends Ur's textual syntax with derived forms for SQL queries, XHTML elements and other constructs defined in a  web programming library \cite{conf/popl/Chlipala15,conf/pldi/Chlipala10}.
+The example below demonstrates how an XHTML expression containing a ``spliced'' string expression can be expressed in Ur/Web.
+
+\begin{lstlisting}[numbers=none]
+  val p = SURL<xml><p>Hello, {[EURLjoin " " [first, last]SURL]}!</p></xml>EURL
+\end{lstlisting}
+The desugaring of this derived form (not shown) is substantially more verbose and, for programmers familiar with the standardized syntax for XHTML \cite{xhtml}, substantially more obscure.
+% Such dialects are sometimes qualitatively taxonomized as amongst the ``domain-specific language'' for this reason \cite{fowler2010domain}.
+%Syntactic cost is often assessed qualitatively \cite{green1996usability}, though quantitative metrics can be defined.
+% Syntax definition systems , which we will discuss in Sec.
+\ref{sec:syntax-dialects}, have simplified the task of defining ``library-specific'' (a.k.a.
+``domain-specific'') syntax dialects like Ur/Web, and have thereby contributed to their ongoing proliferation.
+
+% Many have argued that a proliferation of syntax dialects constructed using these tools is harmless or even desirable, because programmers can simply choose the right syntax dialect for each job at hand \cite{journals/stp/Ward94}.
+However, we argue that in fact
+This ``dialect-oriented approach'' is difficult to reconcile with the best practices of ``programming in the large''  \cite{DeRemer76}, i.e.
+developing large programs ``consisting of many small programs (modules), possibly written by different people'' whose interactions are mediated by a reasonable type and binding discipline.
+The problems that tend to arise are summarized below.
+%a more systematic treatment will follow in  Sec.
+\ref{sec:syntax-dialects}.
+
+\paragraph{Problem 1: Syntactic Conflicts}
+The first problem with the dialect-oriented approach is that clients  cannot always combine different syntax dialects when they want to use derived forms that they define together.
+This is problematic because client programs  cannot be expected to fall cleanly into a single preconceived ``problem domain'' -- large programs use many libraries \cite{DBLP:conf/sac/LammelPS11}.
+In some cases, this is simply because the dialects have been defined using different, incompatible syntax definition systems.
+In other cases, the problem is that the dialects define common forms, but assign them different desugarings.
+
+For example, consider two syntax dialects defined under a system like Camlp4: $\mathcal{D}_1$ defines derived forms for sets, and $\mathcal{D}_2$ defines derived forms for dictionaries, both delimited by \verb~{<~ and \verb~>}~.\footnote{In OCaml, simple curly braces are already reserved by the language for record types and values.} Though each dialect defines a deterministic grammar, i.e.
+$\mathrm{det}(\mathcal{D}_1)$ and $\mathrm{det}(\mathcal{D}_2)$, when the grammars are na\"ively combined by Camlp4, we do not have that $\mathrm{det}(\mathcal{D}_1 \cup \mathcal{D}_2)$ (i.e.
+syntactic ambiguities arise under the combined dialect.) In particular, \verb~{<>}~ can be parsed as either the empty set or the empty dictionary.%(Indeed, this ambiguity arose when Python added set literals to its grammar.) %A recent version of Python added derived forms for mutable sets.
+Due to a conflict with dictionary syntax, however, there is no derived form for the empty set.)
+ %A third syntax dialect might come along that uses the same forms that $\mathcal{D}_2$ defines, but for ordered finite maps.
+
+Schwerdfeger and Van Wyk have developed a modular grammar analysis, implemented in Copper \cite{conf/gpce/WykS07}, that ``nearly'' guarantees that determinism is conserved when syntax dialects (of a certain restricted class) are combined \cite{conf/pldi/SchwerdfegerW09}, the caveat being that the constituent dialects must prefix all newly introduced forms with starting tokens drawn from disjoint sets.\todo{more on this?}
+
+\paragraph{Problem 2: Abstract Reasoning About Derived Forms}
+
+Even putting aside the difficulties of conservatively combining syntax dialects, there are questions about how \emph{reasonable}  sprinkling library-specific derived forms throughout a large software system might be.
+For example, consider the perspective of a programmer attempting to comprehend (i.e.
+reason about) the program fragment in Figure \ref{fig:K-dialect}, which is drawn under a syntax dialect constructed by combining a number of dialects of Standard ML's textual syntax.
+
+\begin{figure}[h]
+\begin{lstlisting}
+    val a = get_a()
+    val w = get_w()
+    val x = read_data(a)
+    val y = {|(!R)@&{&/x!/:2_!x}'!R}|}
+\end{lstlisting}
+\caption{An example of a potentially confusing program fragment.}
+\label{fig:K-dialect}
+\end{figure}
+
+If the programmer happens to be familiar with the (intentionally terse) syntax of the stack-based database query processing language K, then Line 4 might pose few difficulties.
+If the programmer does not recognize this syntax, however, there are no simple, definitive protocols for answering questions like:\todo{label and expand upon these}
+
+\begin{enumerate}
+\item Which constituent dialect defined the derived form that appears on Line 4?
+\item Are the characters \li{x} and \li{R} inside this derived form parsed as ``spliced'' expressions \li{x} and \li{R} (of variable form), or parsed in some other way peculiar to this form?
+\item If \li{x} is the spliced expression \li{x}, does it refer to the binding on the previous line? Or was that binding shadowed by an unseen binding in the desugaring of Line 4?
+\item If \li{w} is renamed, could that possibly break the program, or change its meaning?
+In other words, does the desugaring  assume that some variable identified as \li{w} is in scope (even though \li{w} does not appear directly in the text of Line 4)?
+\item What type does \li{y} have?
+\end{enumerate}
+In short, syntax dialects do not come with useful principles of \emph{syntactic abstraction}: if the desugaring of the program is held abstract, programmers can no longer reason about types and binding (i.e.
+answer questions like those above) in the usual disciplined manner.
+This is burdensome at all scales, but particularly when programming in the large, where it is common to encounter a program fragment drawn by another programmer, or drawn  long ago.
+Forcing the programmer to examine the desugaring of the drawing in order to reason about types and binding defeats the ultimate purpose of using syntactic sugar -- lowering cognitive cost.
+
+%In other words, encountering an unfamiliar derived form has made it difficult for the programmer to maintain the usual \emph{type discipline} and \emph{binding discipline}.
+%Compelling the programmer to examine the desugaring directly defeat the purpose of defining the derived form -- decreasing cognitive cost.
+Indeed, it substantially increases cognitive cost.
+
+% In contrast, when a programmer encounters, for example, a function call like the call to \li{read_data} on Line 3, the analagous questions can be answered by following clear protocols that become ``cognitive reflexes'' after sufficient experience with the language, even if the programmer has no experience with the library defining \li{read_data}:
+% \begin{enumerate}
+% \item The language's syntax definition determines that \li{read_data(a)} is an expression of function application form.
+% \item Similarly, \li{read_data} and \li{a} are definitively expressions of variable form.
+% \item The variable \li{a} can only refer to the binding of \li{a} on Line 1.
+% \item The variable \li{w} can be renamed without knowing anything about the values that \li{read_data} and \li{a} stand for.
+% \item The type of \li{x} can be determined to be \li{B} by determining that the type of \li{read_data} is \li{A -> B} for some \li{A} and \li{B}, and checking that \li{a} has type \li{A}.
+Nothing else needs to be known about the values that \li{read_data} and \li{a} stand for.
+In Reynolds' words \cite{B304}:
+% \begin{quote}
+% \emph{Type structure is a syntactic discipline for enforcing levels of abstraction.}
+% \end{quote}
+% \end{enumerate}
+
+\subsection{Term-Rewriting Macros Give Limited Syntactic Control}
+
+\subsection{Contributions}
+
+\section{TSMs in VerseML}
+\subsection{Simple TSMs}
+
+\subsection{Parametric TSMs}
+
+\section{Simple TSMs, Formally}
+
+\section{Parametric TSMs, Formally}
+
+\section{Related Work}
+
+\section{Discussion \& Conclusion}
 

### 2017-02-28 syntax-icfp17.tex
ICFP 2017 submission

--- 2017-02-11 syntax-icfp17.tex
@@
+\PassOptionsToPackage{svgnames,dvipsnames,svgnames}{xcolor}
 %% For double-blind review submission
@@
 \usepackage{microtype}
+\usepackage{todonotes}
+\usepackage{mdframed}
+\usepackage{colortab}
+\usepackage{mathpartir}
+\usepackage{enumitem}
+\usepackage{bbm}
+
+% \newtheorem{theorem}{Theorem}[chapter]
+% \newtheorem{lemma}[theorem]{Lemma}
+% \newtheorem{corollary}[theorem]{Corollary}
+% \newtheorem{definition}[theorem]{Definition}
+\newtheorem{assumption}[theorem]{Assumption}
+\newtheorem{condition}[theorem]{Condition}
 
 %% Listings
+\definecolor{mygray}{rgb}{0.75,0.75,0.75}
 \usepackage{listings}
 \lstset{tabsize=2, 
-basicstyle=\ttfamily\fontsize{7.5pt}{1em}\selectfont, 
+basicstyle=\ttfamily\fontsize{10pt}{1em}\selectfont, 
 commentstyle=\itshape\ttfamily\color{gray}, 
 stringstyle=\ttfamily\color{red},
-mathescape=false,escapechar=!,
-numbers=left, numberstyle=\scriptsize\color{mygray}, language=ML,moredelim=[il][\sffamily]{?},showspaces=false,showstringspaces=false,xleftmargin=0pt, numbersep=3pt, morekeywords=[1]{tyfam,opfam,let,fn,val,def,casetype,objtype,metadata,of,*,datatype,new,toast,syntax,module,where,import,for,ana,syn,opcon,tycon,metasignature,metamodule,metasig,metamod,static,at,by,tycase,mod,macro,match,pattern,in,patterns,expressions,implicit,forall,exptsm,pattsm},deletekeywords={double},classoffset=0,
-aboveskip=2pt,belowskip=0pt,
+mathescape=false,escapechar=~,
+numbers=left, numberstyle=\scriptsize\color{mygray}, language=ML,moredelim=[il][\sffamily]{?},showspaces=false,showstringspaces=false,xleftmargin=0pt, numbersep=3pt, morekeywords=[1]{spliced,tyfam,opfam,let,fn,val,def,casetype,objtype,metadata,of,*,datatype,new,toast,syntax,module,where,import,for,ana,syn,opcon,tycon,metasignature,metamodule,metasig,metamod,static,at,by,tycase,mod,macro,match,pattern,in,patterns,expressions,implicit,forall,exptsm,pattsm},deletekeywords={double},classoffset=0,
+aboveskip=4pt,belowskip=2pt,
 moredelim=**[is][\color{red}]{SSTR}{ESTR},
@@
 moredelim=**[is][\color{gray}]{SNAT}{ENAT}, 
-moredelim=**[is][\color{blue}]{SURL}{EURL},
+moredelim=**[is][\color{OliveGreen}]{SURL}{EURL},
 moredelim=**[is][\color{SeaGreen}]{SQT}{EQT},
@@
 \citestyle{acmauthoryear}   %% For author/year citations
+
+\input{../../thesis/macros}
 
@@
 Programmers, 
-like ``pen-and-paper'' mathematicians, often seek to define ``syntactic sugar'' that  lowers the syntactic cost of common idioms. 
-Unfortunately, the most direct mechanisms available to programmers, e.g.
+like pen-and-paper mathematicians, often wish to define  ``syntactic sugar'' to  lower the syntactic cost of common idioms. 
+Unfortunately, the most direct mechanisms of syntactic control available to programmers, e.g.
 \li{camlp4} \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, do not support modular reasoning about determinism (i.e.
-separately defined forms can conflict syntactically with one another), and they obscure the type and binding structure of the program.
-We argue therefore that these mechanisms are unreasonable for programming ``in the large''.
+separately defined forms can conflict syntactically with one another.) Moreover, they obscure the type and binding structure of the program.
+Typed hygienic term-rewriting macros (e.g.
+as available in Scala) are more reasonable, but they give programmers only limited syntactic control.
 
-This paper formally introduces \emph{parametric typed syntax macros} (pTSMs), which give library providers programmatic control over both the parsing and expansion of expressions and patterns of \emph{generalized literal form} at a specified type or parameterized family of types.
-Expansion is \emph{strictly hygienic}, meaning that expansions can refer to external bindings only via spliced terms or explicit module parameters, and expansions do not reveal internal bindings to spliced terms or to the remainder of the program.
+This paper formally introduces \emph{simple} and \emph{parametric typed syntax macros} (TSMs), which give library providers programmatic control over both the parsing and expansion of expressions and patterns of \emph{generalized literal form} at a specified type or parameterized family of types.
+Expansion is {strictly hygienic}, meaning that 1) expansions can refer to external bindings only via spliced terms or supplied module parameters; and 2)  expansions do not reveal internal bindings to spliced terms or to the remainder of the program.
 % Partial parameter application lowers the syntactic cost of this strict style. 
-This design ensures that clients are able to reason about the binding structure of the program while holding the expansion of the program abstract.
-The system needs only convey to clients a \emph{segmentation} of each literal that gives the locations of the spliced terms.
-The system can determine this {segmentation} automatically and convey it to the client via syntax highlighting, or by presenting the client with a context-free grammar (without revealing the semantic actions associated with each production.)
+This design ensures that clients are able to reason about the binding structure of the program while holding the expansion of the program abstract -- the system needs only convey to clients the \emph{segmentation} of each literal, which gives the locations of the spliced terms.
+%The system can determine this {segmentation} automatically and convey it to the client via syntax highlighting, or by presenting the client with a context-free grammar (without revealing the semantic actions associated with each production.)
 We argue that this mechanism occupies a ``sweet spot'' in the design space, in that it captures many common syntactic idioms while avoiding the problem of syntactic conflicts by construction and supplying clients with clear abstract reasoning principles.
+We establish these reasoning principles formally.
 
@@
 
-\section{Introduction}
+\newcommand{\introSec}{Introduction}
+\section{\protect\introSec}
+\label{sec:intro}
+When defining the textual syntax of a programming language, it is common practice to include ``syntactic sugar'' specific to certain standard library constructs.   
+For example, Standard ML and other functional languages include sugar for constructing and pattern matching on lists, as defined in the standard library \cite{mthm97-for-dart}.
 
-Text of paper \ldots
+Third-party library providers are, quite justifiably, envious of the privileged status of these otherwise seemingly ordinary standard library definitions.
+After all, it is not difficult to find other examples of situationally useful library-specific syntactic sugar \cite{TSLs}.
+For example, 1) clients of a ``collections'' library might like syntactic sugar for finite sets and dictionaries; 2) clients of a ``web programming'' library might like syntactic sugar for HTML and JSON values; 3) a compiler writer might like syntactic sugar for the terms of the object language or various intermediate languages of interest; and 4) clients of a ``chemistry'' library might like syntactic sugar for chemical structures based on the SMILES standard \cite{anderson1987smiles}.
+This has motivated a long line of research into approaches that delegate more syntactic control to library providers.
+
+\subsection{Syntax Dialects are Unreasonable}\label{sec:syntax-dialects-intro}
+One well-worn approach is for library providers to use a syntax definition system to construct a library-specific \emph{syntax dialect}, i.e.
+a new syntax definition that extends the syntax definition given in the language definition with new derived forms (i.e.
+forms given meaning by expansion, rather than directly.) For example, Ur/Web defines a dialect of Ur's textual syntax with derived forms for SQL queries, XHTML elements and some other constructs defined in a  web programming library \cite{conf/popl/Chlipala15,conf/pldi/Chlipala10}.
+Figure \ref{fig:urweb} demonstrates how an XHTML expression containing ``spliced'' string expressions (Line 1) and XHTML expressions (Line 2) can be expressed in Ur/Web.
+The desugarings of these derived forms, not shown, are substantially more verbose and, for programmers familiar with the standardized syntax for XHTML \cite{xhtml}, substantially more obscure.
+
+\begin{figure}
+\begin{lstlisting}
+  fun heading first_name = SURL<xml><h1>Hello, {[EURLfirst_nameSURL]}!</h1></xml>EURL
+  val body = SURL<xml><body>{EURLheading "World"SURL} ...</body></xml>EURL
+\end{lstlisting}
+\vspace{-6px}
+\caption{Derived XHTML forms with support for splicing at two different types are built in to Ur/Web.}
+\vspace{-6px}
+\label{fig:urweb}
+\end{figure}
+ % Such dialects are sometimes qualitatively taxonomized as amongst the ``domain-specific language'' for this reason \cite{fowler2010domain}.
+%Syntactic cost is often assessed qualitatively \cite{green1996usability}, though quantitative metrics can be defined.
+% Syntax definition systems , which we will discuss in Sec.
+\ref{sec:syntax-dialects}, have simplified the task of defining ``library-specific'' (a.k.a.
+``domain-specific'') syntax dialects like Ur/Web, and have thereby contributed to their ongoing proliferation.
+
+% Many have argued that a proliferation of syntax dialects constructed using these tools is harmless or even desirable, because programmers can simply choose the right syntax dialect for each job at hand \cite{journals/stp/Ward94}.
+However, we argue that in fact
+There are hundreds of syntax definition systems of various design, e.g.
+grammar-based systems like Camlp4 \cite{ocaml-manual}, Copper \cite{conf/gpce/WykS07} or SugarJ/Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, parser combinator systems \cite{Hutton1992d} and somewhat less general systems built in to certain programming systems, like the system of fixity directives in Standard ML \cite{mthm97-for-dart} or more elaborate mixfix systems \cite{5134,DBLP:conf/gpce/TahaJ03} built in to proof assistants like Agda \cite{DBLP:conf/ifl/DanielssonN08} and Coq \cite{Coq:manual}.
+%library-specific (a.k.a. ``domain-specific'') 
+Unfortunately, this dialect-oriented approach is difficult to reconcile with the best practices of ``programming in the large'', i.e.
+developing large programs ``consisting of many small programs (modules), possibly written by different people''   \cite{DeRemer76} whose interactions are mediated by a reasonable type and binding discipline.
+The problems that tend to arise are summarized below.%a more systematic treatment will follow in  Sec.
+\ref{sec:syntax-dialects}.
+
+\paragraph{Problem 1: Incompatibility and Syntactic Conflict}
+The first problem with the dialect-oriented approach is that clients  cannot always combine different syntax dialects when they want to use derived forms that they define together.
+This is a significant problem because large programs use many separately defined libraries \cite{DBLP:conf/sac/LammelPS11} and often do not fall cleanly into a single preconceived ``problem domain''.
+
+In some cases, the dialects in question cannot be combined simply because they have been defined under incompatible syntax definition systems.
+In other cases, there is no dialect combination operation available that ensures that the combined dialect is an extension of both constituent dialects.
+And in other cases, the problem is one of \emph{conflict}: two dialects might define the same form, but determine different desugarings.
+
+For example, consider two syntax dialects defined under a system like Camlp4: $\mathcal{D}_1$ defines derived forms for sets, and $\mathcal{D}_2$ defines derived forms for dictionaries, both delimited by \verb~{<~ and \verb~>}~.\footnote{In OCaml, simple curly braces are already reserved by the language for record types and values.} Each dialect is deterministic (i.e.
+there are no possible syntactic conflicts), $\mathrm{det}(\mathcal{D}_1)$ and $\mathrm{det}(\mathcal{D}_2)$, but when the grammars are combined by Camlp4, it is not the case that $\mathrm{det}(\mathcal{D}_1 \cup \mathcal{D}_2)$.
+In particular, \verb~{<>}~ can desugar to the empty set and the empty dictionary.%(Indeed, this ambiguity arose when Python added set literals to its grammar.) %A recent version of Python added derived forms for mutable sets.
+Due to a conflict with dictionary syntax, however, there is no derived form for the empty set.)
+ %A third syntax dialect might come along that uses the same forms that $\mathcal{D}_2$ defines, but for ordered finite maps.
+
+\citet{conf/pldi/SchwerdfegerW09} have developed a modular grammar analysis, implemented in Copper, that ``nearly'' guarantees that determinism is conserved when syntax dialects (of a certain restricted class) are combined, the caveat being that the constituent dialects must prefix all newly introduced forms with marking tokens drawn from disjoint sets.
+To be confident that the marking tokens used are disjoint, providers must base them on the domain name system (or some other coordinating entity.) Because the mechanism operates at the level of the context-free grammar, it is difficult for the client to define scoped abbreviations for these verbose marking tokens (as one might do when, for example, deeply nested modules are used for namespacing in ML.)
+
+\paragraph{Problem 2: Abstract Reasoning About Derived Forms}
+
+Even putting aside the problem of syntactic conflict, there are questions about how \emph{reasonable}  sprinkling library-specific derived forms throughout a large software system might be. 
+For example, consider the perspective of a programmer attempting to comprehend (i.e.
+reason about) the program fragment in Figure \ref{fig:K-dialect}, which is written in a syntax dialect constructed by combining various library-specific dialects of Standard ML's syntax.
+
+\begin{figure}
+\begin{lstlisting}
+  val w = compute_w ()
+  val x = compute_x ()
+  val y = {|(!R)@&{&/x!/:2_!x}'!R}|}
+\end{lstlisting}
+\vspace{-8px}
+\caption{An example of unreasonable program text.}
+\vspace{-8px}
+\label{fig:K-dialect}
+\end{figure}
+
+If the programmer happens to be familiar with the (intentionally terse) syntax of the stack-based database query processing language K \cite{Whitney:2001:LOR:376284.375783}, then Line 3 might pose few difficulties.
+If the programmer does not recognize this syntax, however, there are no simple, definitive protocols for answering questions like:
+
+\begin{enumerate}
+\item \textbf{(Responsibility)} Which constituent dialect defined the derived form that appears on Line 3?
+\item \textbf{(Segmentation)} Are the characters \li{x} and \li{R} on Line 3 parsed as spliced expressions \li{x} and \li{R} (i.e.
+expressions of variable form), or parsed in some other way peculiar to this form?
+\item \textbf{(Capture)} If \li{x} is in fact a spliced expression, does it refer to the binding of \li{x} on Line 2? Or might it capture an unseen binding introduced in the desugaring of Line 3?
+\item \textbf{(Context Dependence)} If \li{w}, on Line 1, is renamed, could that possibly break the program, or change its meaning?
+In other words, might the desugaring of Line 3 assume that some variable identified as \li{w} is in scope (even though \li{w} is not mentioned in the text of Line 3)?
+\item \textbf{(Typing)} What type does \li{y} have?
+\end{enumerate}
+
+In short, syntax dialects do not come with reasonable principles of \emph{syntactic abstraction}: if the desugaring of the program text is held abstract, programmers can no longer reason about types and binding (i.e.
+answer questions like those above) in the usual disciplined manner.
+This is burdensome at all scales, but particularly when programming in the large, where it is common to encounter unfamiliar libraries.
+Forcing the programmer to examine the desugaring of the program text in order to reason about types and binding defeats the ultimate purpose of syntactic sugar -- decreasing cognitive cost \cite{Green89}.
+
+%In other words, encountering an unfamiliar derived form has made it difficult for the programmer to maintain the usual \emph{type discipline} and \emph{binding discipline}.
+%Compelling the programmer to examine the desugaring directly defeat the purpose of defining the derived form -- decreasing cognitive cost.
+Indeed, it substantially increases cognitive cost.
+
+% In contrast, when a programmer encounters, for example, a function call like the call to \li{read_data} on Line 3, the analagous questions can be answered by following clear protocols that become ``cognitive reflexes'' after sufficient experience with the language, even if the programmer has no experience with the library defining \li{read_data}:
+% \begin{enumerate}
+% \item The language's syntax definition determines that \li{read_data(a)} is an expression of function application form.
+% \item Similarly, \li{read_data} and \li{a} are definitively expressions of variable form.
+% \item The variable \li{a} can only refer to the binding of \li{a} on Line 1.
+% \item The variable \li{w} can be renamed without knowing anything about the values that \li{read_data} and \li{a} stand for.
+% \item The type of \li{x} can be determined to be \li{B} by determining that the type of \li{read_data} is \li{A -> B} for some \li{A} and \li{B}, and checking that \li{a} has type \li{A}.
+Nothing else needs to be known about the values that \li{read_data} and \li{a} stand for.
+In Reynolds' words \cite{B304}:
+% \begin{quote}
+% \emph{Type structure is a syntactic discipline for enforcing levels of abstraction.}
+% \end{quote}
+% \end{enumerate}
+
+\subsection{Term-Rewriting Systems Offer Limited Syntactic Control}
+An alternative approach that a library provider might consider when seeking to control syntactic cost is to leave the context-free syntax of the language unmodified and instead contextually repurpose select existing syntactic forms using a \emph{term rewriting system}.
+
+Na\"ive term rewriting systems suffer from problems analagous to those that plague syntax definition systems.
+In particular, there can be conflicts because separately defined rewriting rules might attempt to rewrite the same term differently.
+It can also be difficult to determine which rewriting rule, if any, is responsible for any particular term.
+Certain conventions have emerged, e.g.
+OCaml's textual syntax now includes \emph{preprocessor extension (ppx) points} used to identify terms that some presumably corresponding term rewriting preprocessor must rewrite \cite{ocaml-manual}, but a correspondence is not enforced.
+In any case, it is difficult to reason abstractly about types and binding given a program subject to a large number of context-free rewriting rules.
+
+Modern \emph{term-rewriting macro systems}, however, have made some progress. In particular:
+\begin{enumerate}[nolistsep]
+\item Macro systems require that the client explicitly apply the intended rewriting (implemented by a macro) to the term that is to be rewritten, thereby addressing the problems of conflict and determining responsibility.
+However, it is often unclear whether a given macro transforms or reinterprets the form of a given argument or sub-term thereof, as opposed to treating it parametrically by inserting it unmodified into the generated expansion.
+This is closely related to the problem of determining a {segmentation} discussed above.
+\item \emph{Hygienic} macro systems prevent variable {capture} and enforce {context-independence} \cite{Kohlbecker86a,DBLP:conf/popl/Adams15,DBLP:conf/popl/ClingerR91,DBLP:journals/lisp/DybvigHB92}.
+\item Although much of the research on macro systems has been for languages in the Lisp tradition \cite{mccarthy1978history}, some progress has also been made on reasoning about types with the design of \emph{typed macro systems} where annotations constrain the macro arguments and the generated expansions.
+Examples include Scala's macro system \cite{ScalaMacros2013}, formal calculi studied by \citet{DBLP:conf/esop/HermanW08} and \emph{staging macro systems} like MetaML \cite{Sheard:1999:UMS} and MacroML \cite{ganz2001macros} (which, it is worth noting, do not give the macro access to the syntax tree of arguments.)
+\end{enumerate}
+
+Perhaps the biggest problem, then, is that term-rewriting systems offer library providers only limited syntactic control.
+In other words many syntactic forms of interest will not particularly resemble those available in the host language.
+For example, consider the  XHTML and K examples above.
+In both cases, the syntactic conventions are quite distinct from those of ML-like languages (and, for that matter, languages that use S-expression.)
+
+It is tempting in these situations to consider repurposing string literal forms.
+For example, we might wish to apply a macro \li{html!} (following Rust's convention of using a post-fix \li{!} to distinguish macro names from variables) to rewrite string literals containing Ur/Web-style XHTML syntax as follows:
+\begin{lstlisting}[numbers=none]
+  fun heading first_name = html! 'SSTR<h1>Hello, {[first_name]}!</p>ESTR'
+  val body = html! 'SSTR<body>{heading "World"} ...</body>ESTR'
+\end{lstlisting}
+
+The problem here is that there is no way to extract the spliced expressions from the supplied string literal forms while satisfying the hygiene conditions, because variables that come from these spliced terms (here, \li{first_name} and \li{heading}) are indistinguishable from variables that inappropriately appear free in the generated expansion.
+In addition, the problem of segmentation becomes even more pernicious: to a human or tool unaware of Ur/Web's syntax, it is not immediately apparent which particular subsequences of the string literals supplied to \li{html!} are segmented out as spliced expressions.
+Reader macros have essentially the same problem  \cite{DBLP:journals/jfp/FlattCDF12}.
+
+  Some languages build in \emph{string splicing} (a.k.a.
+\emph{string interpolation}) forms, e.g.
+Scala \cite{odersky2008programming}, or similar but more general \emph{fragmentary quotation forms} \cite{conf/icfp/Slind91}, e.g.
+the SML/NJ dialect of ML.
+These designate particular delimiters to allow escape out into the expression language.
+The problem with using these (dynamically, or in conjunction with term rewriting macros) as vehicles to introduce syntactic sugar at various types is 1) there is no ``one-size-fits-all'' escape delimiter, and 2) reasoning about typing is problematic because every escaped term is checked against the same type.
+In the example above, we spliced in two different types of expressions using two different delimiters.
+In general, e.g.
+when defining syntax for a programming language with many sorts of terms, the most appropriate choice of delimiters might depend on where each spliced term appears.
+
+% These existing forms normally have other meanings, so this can be confusing \cite{pane1996usability}.
+
+\subsection{Contributions}
+This work introduces a system of \textbf{typed syntax macros (TSMs)} that gives library providers substantially more syntactic control than existing typed term-rewriting macro systems while maintaining the ability to reason abstractly about types, binding and segmentation.% abstract reasoning principles.
+% comparable to the level of control they have when defining a syntax dialect.
+
+% Clients apply TSMs to \emph{generalized literal forms}. 
+Our take on the running example is shown in Fig. \ref{fig:first-tsm-example}: we apply a TSM named 
+\li{$html} 
+ to \emph{generalized literal forms} delimited by backticks. TSM names are prefixed by \li{$} 
+ to clearly them from variables.
+The semantics delegates control over the parsing and expansion of each literal body to the applied TSM during a semantic phase called \emph{typed expansion}, which generalizes the usual typing phase.
+\begin{figure}[ht!]
+\vspace{-10px}
+\begin{lstlisting}
+  fun heading first_name = $html `SURL<h1>Hello, {[EURLfirst_nameSURL]}!</h1>EURL`
+  val body = $html `SURL<body>{EURLheading "World!"SURL} ...</body>EURL`
+\end{lstlisting}
+\vspace{-10px}
+\caption{Two examples of a TSM being applied to a generalized literal form.
+%In each case, literal body, between backticks, is initially left unparsed according to the language's context-free syntax.
+%The applied TSM determines a segmentation and expansion during the typed expansion phase, which generalizes the usual typing phase.
+We assume strings are primitive.}
+\vspace{-10px}
+\label{fig:first-tsm-example}
+\end{figure}
+
+Generalized literal forms, which first arose in related work by \citet{TSLs} that we will detail in Sec.
+\ref{sec:related-work}, subsume a variety of common syntactic forms because the context-free syntax of the language only defines which outer delimiters are available (in this paper, we will mostly use backticks, but in general, there can be many choices.) \emph{Literal bodies}, i.e.
+the characters between the delimiters, are otherwise syntactically unconstrained.
+Choices regarding internal delimitation of spliced terms is not imposed by the language definition.
+Notice, however, that Figure \ref{fig:first-tsm-example} reveals the locations of the spliced expressions by coloring them black.
+We have designed our system so that a figure like this, which presents a \emph{segmentation} of each literal body into spliced terms (in black) and characters parsed in some other way by the applied TSM (in green)\footnote{In general, spliced expressions might themselves apply TSMs, in which case the convention is to use a distinct color for unspliced segments at each depth.
+For example, if strings were not primitive, we might write \li{`SURL<body>\{EURLheading ($str "SSTRWorld!ESTR")SURL\} ...</body>EURL`}.}, can always be  generated.
+
+% Because the context-free syntax is never extended, syntactic conflicts are not a concern.
+
+%As such, the semantics can take the type and binding structure of the surrounding program into account when validating the expansion that the TSM programmatically generates to ensure that clients can answer critical questions related to types and binding, like those enumerated in Section \ref{sec:abs-reasoning-intro}.
+Clients need not have knowledge of the implementation of the TSM or of the generated expansion, i.e.
+there are useful principles of syntactic abstraction.
+
+% The primary technical challenge has to do with the fact that the applied TSM needs to be able to splice terms out of the literal body for inclusion in the expansion. For example,
+
+ % We design our mechanism such that these locations can easily be determined from the output of the TSM.
+This is essential for our hygiene mechanism, and it is also useful in that this information can be presented to the user (e.g.
+as shown in Figure \ref{fig:first-tsm-example-marked}).
+%As such, we must develop a mechanism where 1) the positions of spliced subterms can be determined without examining the macro implementation (e.g.
+so that they can be presented to the user differently by an editor or pretty-printer, ;  and 2) the hygiene mechanism must give only portions of the expansion that correspond to these spliced subterms access to the application site context.
+
+%In order to reason about types and binding, client programmers need only have knowledge of 1) the segmentation (e.g.
+by examining a figure like this presented by a code editor or pretty-printer) and 2) a type annotation on the definition of the applied TSM.
+No other details about the applied TSM's implementation or the expansion that it generates need to be revealed to the client programmer.
+In other words,
+TSMs come equipped with useful principles of syntactic abstraction.
+We will more precisely characterize these abstract reasoning principles as we proceed.
+For now, to develop some intuitions, consider Figure \ref{fig:K-tsm-example}, which uses TSMs to express the ``unreasonable'' example from Figure \ref{fig:K-dialect}.
+Without examining the expansion of Line 3, we can reason as follows:
+
+\begin{enumerate}
+\item \textbf{(Responsibility)} The applied TSM, \li{$kquery}, is solely responsible for typed expansion of the literal body. 
+\item \textbf{(Segmentation)} By examining the segmentation, we know that the two instances of \li{x} on Line 3 are parsed as spliced expressions, whereas \li{R} is parsed in some other way peculiar to this form.
+\item \textbf{(Capture)} The system prevents capture, so the spliced expression \li{x} must refer to the binding of \li{x} on Line 2 -- it cannot capture an unseen binding introduced in the expansion of Line 3.
+\item \textbf{(Context Dependence)} The system enforces context independence, so the expansion of Line 3 cannot  rely on the fact that, for example, \li{w} is in scope.
+\item \textbf{(Typing)} An explicit type annotation on the definition of \li{$kquery} determines the type that every expansion it generates will have.
+We will see an example of a TSM definition in the next section.
+
+Moreover, each segment in the segmentation also comes paired with the type it is expected to have.
+This information is usually not necessary to reason about typing, but it can be conveyed to the programmer upon request by the program editor if desired.
+%TSM definitions follow the usual scoping rules, so it is easy to ``jump to the definition'' of \li{$kquery}.
+\end{enumerate}
+
+\begin{figure}
+\begin{lstlisting}[numbers=none,xleftmargin=0px]
+  val w = compute_w ()
+  val x = compute_x ()
+  val y = $kquery `SURL(!R)@&{&/EURLxSURL!/:2_!EURLxSURL}'!R}EURL`
+\end{lstlisting}
+\vspace{-5px}
+\caption{TSMs make examples like the one from Fig. \ref{fig:K-dialect} more reasonable.}
+\vspace{-5px}
+\label{fig:K-tsm-example}
+\end{figure}
+
+% \begin{figure}[h]
+% \begin{lstlisting}
+% PElement Nil Seq(
+%   TextNode "Hello, ", 
+%   Seq(TextNode (join(" ", Cons(first, Cons(second, Nil)))), 
+%   TextNode "!"))
+% \end{lstlisting}
+% \caption{The desugaring.}
+% \end{figure}
+
+% There is also no ambiguity with regard to which TSM has control over each form, and searching for the definition of a TSM is no more difficult than searching for any other binding, i.e.
+there are well-defined scoping rules.
+
+% In other words, TSMs maintain a useful notion of syntactic abstraction.
+%More specifically, TSMs maintain a \emph{hygienic binding discipline}, meaning that questions Questions 4 and 5 above were concerned with are disallowed entirely.
+% We will, of course, make this notion more technically precise as we continue.
+
+The remainder of this document is organized as follows.
+\begin{itemize}
+\item Sec. \ref{sec:setsms} begins by going through the mechanics of defining a {simple expression TSM} and describes the process of typed expansion and expansion validation by example.
+\item Sec. \ref{sec:sptsms} does the same for a {simple pattern TSM}, i.e. a TSM that generates patterns that match values of a specified type. The hygiene condition for pattern TSMs is interesting.
+\item Sec. \ref{sec:setsms-formally} defines a calculus of simple expression and pattern TSMs and formally establishes the reasoning principles implied above.
+
+% In the remaining chapters, we enrich the language developed in Chapter \ref{chap:uetsms} with advanced features based on those found in full-scale general-purpose languages like ML and Scala, and enhance our TSM mechanism with these new features.
+
+% We next develop a more sophisticated account of TSMs in Chapters \ref{chap:ptsms} and \ref{chap:static-eval}, with the aim of working out details necessary to integrate TSMs into full-scale general-purpose languages like ML or Scala.
+%These two chapters constitute Part \ref{part:parametric-tsms} of our contributions.
+\item Sec.
+\ref{sec:ptsms} adds parameterized types and an ML-style module system and introduces \emph{parametric TSMs} (pTSMs), i.e.
+TSMs that take type and module parameters.
+Parameters serve two purposes:
+\begin{enumerate}
+\item They allow for TSMs that operate not just at a single type, but over a type- and module-parameterized family of types. For example, rather than defining a TSM % \li{$strlist} 
+for string-valued dictionaries and another TSM 
+%\li{$intlist} 
+for integer-valued dictionaries, or over different implementations of the dictionary signature, we can define a single parametric TSM 
+%\li{$list} 
+that operates uniformly across a type- and module-parameterized family of dictionary types.
+\item Parameters also give expansions access to terms other than those explicitly spliced in to the literal body.
+This is particularly convenient for working with abstract types \emph{a la} ML.
+Partial parameter application in TSM abbreviations lowers the cost of this explicit parameter-passing discipline.
+\end{enumerate}
+This section culminates with a calculus of ML-style modules that has support for parametric TSMs.
+\item Sec.
+\ref{sec:static-eval} considers the topic of term evaluation during the typed expansion phase, i.e.
+\emph{static evaluation}, in a bit more depth, and also gives examples of TSMs that are useful for defining other TSMs, e.g.
+TSMs that implement parser generators and quasiquotation.
+\item Sec.
+\ref{sec:related-work} discusses additional related work, including closely related work by \citet{TSLs,sac15} on type-specific languages.
+We conclude with a brief discussion of present limitations and future directions.
+\end{itemize}
+
+Due to space limitations, many of the standard or straightforward technical details of the calculi defined in Sec.
+\ref{sec:setsms-formally} through Sec.
+\ref{sec:static-eval} are relegated to the supplemental material.
+For convenience, the examples that we give are written in a full-scale functional language under development called VerseML.
+% For the present purposes, we can consider VerseML as essentially the calculus defined in the supplement extended with various conveniences that are commonly found in other ML-like languages and, notionally, orthogonal to TSMs.
+%VerseML is, as its name suggests, a conceptual descendent of ML.
+It diverges from other dialects of ML that have a similar type structure in that it has a bidirectional type system \cite{Pierce:2000:LTI:345099.345100} (like, for example, Scala \cite{OdeZenZen01}) for reasons that have to do with the mechanism of TSM implicits described in Chapters \ref{chap:tsls} and \ref{chap:ptsms}.
+%The reason we will not follow Standard ML \cite{mthm97-for-dart} in giving a complete formal definition of VerseML in this work is both to emphasize that the primitives we introduce are ``insensitive'' to the details of the underlying type structure of the language (so TSMs can be considered for inclusion in a variety of languages, not only dialects of ML), and to avoid distracting the reader (and the author) with definitions that are already well-understood in the literature and that are orthogonal to those that are the focus of this work.
+% We will not formally define these features mainly to avoid unnecessarily complicating our presentation with details that are not essential to the ideas presented herein. As such, 
+All examples written in VerseML should be understood to be informal motivating material for the subsequent formal material.
+%We anticipate that future full-scale language specifications will be able to combine the ideas  in the proposed work without trouble.
+%The purpose of the work being proposed is to serve as a reference for those interested in the new constructs we introduce, not to serve as a language specification.
+%We will give a brief overview of these languages are organized in Sec. \ref{sec:VerseML}.
+
+% \begin{figure}[h]
+% \begin{lstlisting}[numbers=none,xleftmargin=0px]
+% let syntax $strlist = $list string in 
+% $html `SURL<p>Hello, {[EURLjoin ($str ' ') ($strlist [firstSURL,EURL last])SURL]}</p>EURL`
+% \end{lstlisting}
+% \caption{The example from Figure \ref{fig:first-tsm-example-marked}, expressed using a parametric TSM.}
+% \label{fig:first-ptsm-example-marked}
+% \end{figure}
+
+% In Secs.
+\ref{chap:uetsms} and \ref{chap:uptsms}, we assume for the sake of technical simplicity that each TSM definition is self-contained, needing no access to libraries or to other TSMs.
+This is an impractical assumption in practice.
+We relax this assumption in In
+
+% %\item \textbf{Type-specific languages}, or \textbf{TSLs}. TSLs, described 
+% In Chapter \ref{chap:tsls}, we develop a mechanism of \emph{TSM implicits} that allows library clients to contextually designate, for any type, a privileged TSM at that type.
+The semantics applies this privileged TSM implicitly to unadorned literal forms that appear where a term of the associated type is expected.
+For example, if we designate
+% %\li{$str} 
+% as the privileged TSM at the \li{string} type and 
+% %\li{$strlist}
+% as the privileged TSM at the \li{list(string)} type, we can express the example from Figure \ref{fig:first-tsm-example-marked} instead as shown in Figure \ref{fig:first-tsm-example-implicit} (assuming \li{join} has type \li{string -> list(string) -> string}.)
+% \begin{figure}[h]
+% \begin{lstlisting}[numbers=none]
+% $html`SURL<p>Hello, {[EURLjoin ' ' [firstSURL,EURL last]SURL]}</p>EURL`
+% \end{lstlisting}
+% \caption{The example from Figure \ref{fig:first-tsm-example-marked} drawn to take advantage of TSM implicits.}
+% \label{fig:first-tsm-example-implicit}
+% \end{figure}
+
+% \noindent This approach is competitive in cost with library-specific syntax dialects (e.g.
+compare Figure \ref{fig:first-tsm-example-implicit} to Figure \ref{fig:urweb}), while maintaining the abstract reasoning principles characteristic of our approach.
+%\item \textbf{Metamodules}, introduced in Sec.
+\ref{sec:metamodules}, reduce the need to primitively build in the type structure of constructs like records (and variants thereof),  labeled sums and other interesting constructs that we will introduce later by giving library providers programmatic ``hooks'' directly into the semantics, which are specified as a \emph{type-directed translation semantics} targeting a small \emph{typed internal language} (introduced in Sec.
+\ref{sec:VerseML}).
+%For example, a library provider can implement the type structure of records with a metamodule that:
+%\begin{enumerate}
+%\item introduces a type constructor, \lstinline{record}, parameterized by finite mappings from labels to types, and defines, programmatically, a translation to unary and binary product types (which are built in to the internal language); and
+%\item introduces operators used to work with records, minimally record introduction and elimination (but perhaps also various functional update operators), and directly implements the logic governing their typechecking and translation to the IL (which builds in only nullary and binary products).
+%\end{enumerate}
+%We will see direct analogies between ML-style modules (which our mechanisms also support) and metamodules later.
+%\end{enumerate}
+
+% As vehicles for this work, we will define a small programming language in each of the three parts just mentioned, each building conceptually upon the previous language.
+All of our formal contributions are relative to these small languages.
+
+%TSMs, like other macro systems, perform \emph{static code generation} (also sometimes called \emph{static} or \emph{compile-time metaprogramming}), meaning that the relevant rules in the static semantics of the language call for the evaluation of \emph{static functions} that generate term encodings.
+Static functions are functions that are evaluated statically, i.e.
+during typing.
+%Library providers write these static functions using the VerseML \emph{static language} (SL).
+%Maintaining a separation between the static (or ``compile-time'') phase and the dynamic (or ``run-time'') phase is an important facet of VerseML's design.
+% static code generation.
+%We will  also introduce a simple variant of each of these primitives that leverages VerseML's support for local type inference to further reduce syntactic cost in certain common situations.
+
+\newcommand{\seTSMsSec}{Simple Expression TSMs (seTSMs)}
+\section{\protect\seTSMsSec}
+\label{sec:setsms}
+
+We will begin in this section by introducing \emph{simple expression TSMs} (seTSMs), i.e. TSMs where expansions are expressions of a single specified type.
+
+% \section{Simple Expression TSMs By Example}\label{sec:tsms-by-example}
+% We begin in this section with a ``tutorial-style'' introduction to seTSMs in VerseML.
+%In particular, we will define an seTSM for constructing values of the recursive labeled sum type \li{rx} that was defined in Figure \ref{fig:datatype-rx}.
+% Sec.
+\ref{sec:tsms-minimal-formalism} then formally defines a reduced dialect of VerseML called $\miniVerseUE$.
+This will serve as a ``conceptually minimal'' core calculus of TSMs, in the style of the simply typed lambda calculus.
+%We conclude in Sec.
+\ref{sec:uetsms-discussion}
+
+% \subsection{TSM Application}\label{sec:uetsms-usage}
+% The following VerseML expression, drawn textually, is of \emph{TSM application} form. Here, a TSM named \li{$rx} is applied to the \emph{generalized literal form} \li{/SURLA|T|G|CEURL/}:
+% \begin{lstlisting}[numbers=none,mathescape=|]
+% $rx /SURLA|T|G|CEURL/
+% \end{lstlisting}
+% Generalized literal forms are left unparsed according to the context-free syntax of VerseML.
+Several other outer delimiters are also available, as summarized in Figure \ref{fig:literal-forms}.
+The client is free to choose any of these for use with any TSM, as long as the \emph{literal body} (shown in green above) satisfies the requirements stated in Figure \ref{fig:literal-forms}.
+For example, we could have equivalently written the example above as \li{$rx `SURLA|T|G|CEURL`}.
+(In fact, this would have been convenient if we had wanted to express a regex containing forward slashes but not backticks.)
+
+% It is only during the subsequent \emph{typed expansion} phase that the applied TSM parses the {body} of the literal form to generate a \emph{proto-expansion}.
+The language then \emph{validates} this proto-expansion according to criteria that we will describe in Sec.
+\ref{sec:uetsms-validation}.
+If proto-expansion validation succeeds, the language generates the \emph{final expansion} (or more concisely, simply the \emph{expansion}) of the TSM application.
+The behavior of the program is determined by its expansion.
+
+% For example, the expansion of the TSM application above is equivalent to the following expression when the regex value constructors \li{Or} and \li{Str} are in scope:
+% \begin{lstlisting}[numbers=none]
+% Or(Str "SSTRAESTR", Or(Str "SSTRTESTR", Or(Str "SSTRGESTR", Str "SSTRCESTR")))
+% \end{lstlisting}
+% To avoid the assumption that the variables \li{Or} and \li{Str} are in scope at the TSM application site, the expansion actually uses the explicit \li{fold} and \li{inj} operators, as described in Sec.
+\ref{sec:lists}.
+In fact, the proto-expansion validation process enforces this notion of context independence -- we will return to proto-expansion validation below.
+(We will show how TSM parameters can reduce the awkwardness of this requirement in Chapter \ref{chap:ptsms}.)
+% %The constructors above are those of the type \li{Rx} that was defined in Figure \ref{fig:datatype-rx}.
+
+% % A number of literal forms, ,  are available in VerseML's concrete syntax.
+Any literal form can be used with any TSM,  TSMs have access only to the literal bodies.
+Because TSMs do not extend the concrete syntax of the language directly, there cannot be syntactic conflicts between TSMs.
+
+%  %The form does not directly determine the expansion.
+
+% \begin{figure}
+% \begin{lstlisting}
+% 'SURLbody cannot contain an apostropheEURL'
+% `SURLbody cannot contain a backtickEURL`
+% [SURLbody cannot contain unmatched square bracketsEURL]
+% {|SURLbody cannot contain unmatched barred curly bracesEURL|}
+% /SURLbody cannot contain a forward slashEURL/
+% \SURLbody cannot contain a backslashEURL\
+% \end{lstlisting}
+% %SURL<tag>body includes enclosing tags</tag>EURL
+% \caption[Available Generalized Literal Forms]{Generalized literal forms available for use in VerseML's textual syntax.
+The characters in green indicate the literal bodies and describe how the literal body is constrained by the form shown on that line.
+The Wyvern language defines additional forms, including whitespace-delimited forms \cite{TSLs} and multipart forms \cite{sac15}, but for simplicity we leave these out of VerseML.}
+% \label{fig:literal-forms}
+% \end{figure}
+\subsection{TSM Definitions}\label{sec:uetsms-definition}
+%The original expression, above, is statically rewritten to this expression.
+The definition of the seTSM named \lstinline{$html} that was shown being applied in Fig. \ref{fig:first-tsm-example} takes the following form:
+\begin{lstlisting}[mathescape=|]
+  syntax $html at html by 
+    static fn(b : body) -> parse_result(proto_expr) => 
+      (* Ur/Web-style HTML syntax parser here *)
+  in ... end 
+\end{lstlisting}
+
+Every seTSM definition consists of a TSM name, here \li{$html}, a \emph{type annotation}, here \lstinline{at html}, and a \emph{parse function} between \li{by} and \li{in}.
+The TSM definition is in scope between \li{in} and \li{end}.
+In practice, if an \li{in} clause is not provided, the definition is in scope until the end of the enclosing declaration (e.g.
+the enclosing function or module.) We will consider how TSM definitions are exported from libraries in Sec \ref{sec:static-eval}.
+
+% All TSM names must begin with the dollar symbol (\li{$}), which distinguishes them from variables.
+This is inspired by the Rust macro system, which uses post-fix exclamation points (\li{!}) to distinguish macro identifiers \cite{Rust/Macros}.
+
+The {parse function} is a \emph{static function} delegated responsibility over parsing the literal bodies of the literal forms to which the TSM is applied.
+Static functions, marked by the \li{static} keyword, are applied during the typed expansion process, so they cannot refer to the surrounding variable bindings (because those variables stand for dynamic values.) For now, we will simply assume that static functions are closed, and that they do not themselves make use of TSMs (we will eliminate these impractical limitations in Sec.
+\ref{sec:static-eval}.)
+
+The parse function must have type \li{body -> parse_result(proto_expr)}.
+The input type, \lstinline{body}, classifies encodings of literal {bodies}.
+In VerseML, literal bodies are sequences of characters, so it suffices to define \li{body} as an abbreviation for the \li{string} type, as shown in Figure \ref{fig:indexrange-and-parseresult}.
+The return type is a labeled sum type, defined by applying the type function \li{parse_result} defined in Figure \ref{fig:indexrange-and-parseresult}, that distinguishes between parse errors and successful parses.
+%\footnote{\li{parse_result} is defined as a type function because in Chapter \ref{chap:uptsms}, we will introduce pattern TSMs, which generate patterns rather than expressions.}
+Let us consider these two possibilities in turn.
+
+\paragraph{Parse Errors} If the parse function determines that the literal body is not well-formed (according to whatever syntax definition that it implements), it returns:
+\begin{lstlisting}[numbers=none]
+  ParseError {msg=~$e_\text{msg}$~, loc=~$e_\text{loc}$~}
+\end{lstlisting}
+where $e_\text{msg}$ is an error message and $e_\text{loc}$ is a value of type \li{segment}, defined in Figure \ref{fig:indexrange-and-parseresult}, that designates a segment of the literal body as the location of the error \cite{DBLP:journals/jsc/DeursenKT93}.
+This information is for use by VerseML compilers when reporting the error to the programmer (but it otherwise has no semantic significance.) We might alternatively have used an exception to signal a parse error, but sum types are easier to formalize.
+
+% In languages where the surface syntax is not textual, the definition of \li{loc} would need to designate
+
+\paragraph{Successful Parses} If parsing succeeds, the parse function returns 
+\begin{lstlisting}[numbers=none]
+  Success ~$\ecand$~
+\end{lstlisting} 
+where $\ecand$ is called the \emph{encoding of the proto-expansion}.
+For expression TSMs, proto-expansions are \emph{proto-expressions}, which are encoded as values of the recursive sum type \lstinline{proto_expr} outlined in Figure \ref{fig:candidate-exp-verseml}.
+Most of the variants of \li{proto_expr} are individually uninteresting -- they encode VerseML's various expression forms (just as in a self-hosted compiler, e.g. SML/NJ's Visible Compiler library.) 
+Expressions can mention types, so we also need the type \li{proto_typ} also outlined in Figure \ref{fig:candidate-exp-verseml}.
+More sophisticated encoding techniques for terms would cause no fundamental problems.
+It is only the \li{SplicedE} and \li{SplicedT} variants of these types that are interesting.
+%As we enrich our language in later chapters, we will need to define more encodings like these, for other sorts of trees.
+
+% The definitions of \li{proto_typ} and \li{proto_expr} are recursive labeled sum types to simplify our exposition, but we could have chosen alternative encodings, e.g.
+based on abstract binding trees \cite{pfpl}, with only minor modifications to our semantics.
+Indeed, when we formally define seTSMs in Sec.
+\ref{sec:miniVerseU}, we abstract over the particular encoding.
+
+% We will show a complete encoding when we describe our reduced formal system $\miniVerseUE$ in Sec. \ref{sec:tsms-minimal-formalism}. 
+% ; the remaining constructors (some of which are elided for concision) encode the abstract syntax of VerseML expressions and types.
+% It is extended with one additional form used to handled spliced subexpressions,
+
+%Notice that the types just described are those that one would expect to find in a typical parser.
+
+%One would find types analagous to those just described in any parser, so for concision, we elide the details of \li{$rx}'s parse function.
+%The parse function must treat the TSM parameters parametrically, i.e.
+it does not have access to any values in the supplied module parameter.
+Only the expansion the parse function generates can refer to module parameters.
+%For example, the following definition is ill-sorted:
+%\begin{lstlisting}[numbers=none]
+%syntax pattern_bad[Q : PATTERN] at Q.t {
+%  static fn (body : Body) : Exp => 
+%    if Q.flag then (* ... *) else (* ... *)
+%}
+%\end{lstlisting}%So the parse function parses the body of the delimited form to generate an encoding of the elaboration.
+
+\begin{figure}
+\begin{lstlisting}[numbers=none]
+type body = string
+type segment = {startIdx : nat, endIdx : nat} (* inclusive *)
+type parse_result('a) = ParseError of { msg : string, loc : segment }
+                      | Success of 'a
+type var_t = string (* ... or some other encoding of variables ... *)
+type proto_typ = TyVar of var_t 
+               | Arrow of proto_typ * proto_typ 
+               | (* ... *) 
+               | SplicedT of segment
+type proto_expr = Var of var_t 
+                | Fn of var_t * proto_typ * proto_expr
+                | Ap of proto_expr * proto_expr
+                | (* ... *) 
+                | SplicedE of segment * proto_typ
+
+\end{lstlisting}
+\vspace{-5px}
+\caption[Definitions of various types used by TSM definitions.]{Definitions of various types used by TSM definitions. We assume that these type definitions are available ambiently.}
+\label{fig:indexrange-and-parseresult}
+\label{fig:candidate-exp-verseml}
+\vspace{-5px}
+\end{figure}
+
+\subsection{Splicing}\label{sec:splicing-and-hygiene}
+When a parse function determines that a segment of the literal body should be taken as a spliced expression, it does not directly insert the syntax tree of that expression into the encoding of the expansion.
+Instead,
+the TSM must refer to the spliced expression \emph{by its relative location} using the \li{SplicedE} variant of \li{proto_expr}, which takes a value of type \li{segment} that indicates the zero-indexed location of the spliced expression relative to the start of the provided literal body.
+The \li{SplicedE} variant also requires a value of type \li{proto_typ}, which indicates the type that the spliced expression is expected to have.
+Types can be spliced out by using the \li{SplicedT} variant of \li{proto_typ} analagously.
+
+For example, consider again the TSM application on Line 1 of Fig. \ref{fig:first-tsm-example}, reproduced below:
+\begin{lstlisting}[numbers=none]
+  $html `SURL<h1>Hello, {[EURLfirst_nameSURL]}</h1>EURL`
+\end{lstlisting}
+Let us assume that \li{html} abbreviates a recursive labeled sum type having at least the cases
+\begin{lstlisting}[numbers=none]
+  H1Element of list(html_attr) * list(html) | TextNode of string | ...
+\end{lstlisting}
+The proto-expansion generated by \li{$html} for the supplied literal body, if written in a hypothetical textual syntax where references to spliced expressions are written \li{spliced<startIdx; endIndex; ty>}, is:
+\begin{lstlisting}[numbers=none]
+  H1Element(Nil, Cons(TextNode "Hello, ", Cons(
+                      TextNode spliced<13; 22; string>, Nil)))
+\end{lstlisting}
+Here, \li{spliced<13; 22; html>} refers to the spliced expression \li{first_name}, which must be of type \li{string}, by location, rather than directly.
+
+Requiring that the TSM refer to spliced expressions indirectly in this manner prevents it from ``forging'' a spliced expression (i.e.
+claiming that an expression is a spliced expression when it does not appear textually in the literal body.)% Our semantics distinguishes spliced expressions when validating the proto-expansion, which is important for reasons of hygiene that we will detail shortly.
+
+% The parse function can similarly extract \emph{spliced types} from a literal body using the \li{SplicedT} variant of \li{proto_typ}.
+%In particular, the parse function must provide the index range of spliced subexpressions to the \li{Spliced} constructor of the type \li{MarkedExp}.
+%Only subexpressions that actually appear in the body of the literal form can be marked as spliced subexpressions.
+
+%For example, had the  would not be a valid expansion, because the  that are not inside spliced subexpressions:
+%\begin{lstlisting}[numbers=none]
+%Q.Seq(Q.Str(name), Q.Seq(Q.Str ": ", ssn))
+%\end{lstlisting}
+
+% \subsection{Splice Summaries and Segmentations}
+The \emph{splice summary} of a proto-expression is the finite set of references to spliced terms contained within.
+For example, the summary of the proto-expression above is the finite set containing only \li{spliced<13; 22; html>}.% Notice that no information about  the spliced terms appear is communicated by the splice summary.
+
+The \emph{segmentation} of a proto-pression is the finite set of segments extracted from the splice summary, where a segment is simply a pair of natural numbers.
+For example, the segmentation of the proto-expansion above is $\{(13, 22)\}$.
+The semantics checks that all of the segments in the segmentation are 1) in bounds relative to the literal body; and 2) non-overlapping. 
+This resolves the problem of \textbf{Segmentation} described in Sec. \ref{sec:intro}, i.e. every literal body in a well-typed program has a well-defined segmentation.
+
+A program editor or pretty-printer can communicate the segmentation information to the programmer, e.g. by coloring non-spliced segments green as is our convention in this document. 
+A program editor or pretty-printer can also communicate the type of each spliced term, as indicated in the splice summary, to the programmer upon request (for example, the Emacs and Vim packages for working with OCaml defer to the Merlin tool when the programmer requests the type of an expression.)
+
+\subsection{Proto-Expansion Validation}\label{sec:uetsms-validation}
+Three concerns described in Sec.
+\ref{sec:intro} remain: those related to reasoning abstractly about \textbf{Capture}, \textbf{Context Dependence} and \textbf{Typing}.
+Addressing these concerns is the purpose of the \emph{proto-expansion validation} process.
+The result of proto-expansion validation is the \emph{final expansion}, which is simply the proto-expansion with  references to spliced terms replaced with their corresponding final expansions.
+
+\subsubsection{Capture}
+Proto-expansion validation ensures that spliced terms have access \emph{only} to the bindings that appear at the application site -- spliced terms cannot capture the bindings that appear in the proto-expansion.
+For example, consider the following application site:
+\begin{lstlisting}[numbers=none]
+  let tmp = (* ... application site temporary ... *) in 
+  $html /SURL<h1>{EURLf(tmp)SURL}</h1>EURL/
+\end{lstlisting}
+Now consider the scenario where the proto-expansion generated by \li{$html} has the following form:
+\begin{lstlisting}[numbers=none]
+  let tmp = (* ... expansion-internal temporary ... *) in 
+  H1Element(tmp, spliced<5; 10; html>)
+\end{lstlisting}
+Na\"ively, the binding of the variable \li{tmp} in the proto-expansion could shadow the application-site binding of \li{tmp} in the final expansion.
+
+To address this problem, proto-expansion validation is capture-avoiding.
+In other words, it discharges the requirement that capture not occur by implicitly alpha-varying the bindings in the proto-expansion as needed.
+For example, the final expansion of the example above might take the following form:
+\begin{lstlisting}[numbers=none]
+  let tmp = (* ... application site temporary ... *) in 
+  let tmp' = (* ... expansion-internal temporary ... *) in 
+  H1Element(tmp', f(tmp))
+\end{lstlisting}
+Notice that the expansion-internal binding of \li{tmp} has been alpha-varied to \li{tmp'} to avoid shadowing the application site binding of \li{tmp}.
+As such, the reference to \li{tmp} in the spliced expression refers, as intended, to the application site binding of \li{tmp}.
+
+For TSM providers, the benefit of this mechanism is that they can name the variables used internally within expansions freely, without worrying about whether their chosen identifiers might shadow those that a client might have used at the application site.
+There is no need for a mechanism that generates ``fresh variables''.
+
+TSM clients can, in turn, reliably reason about binding within every spliced expression without examining the expansion that the spliced expression appears within.
+
+The trade-off is that this prevents library providers from intentionally introducing bindings into spliced terms.
+For example, Haskell's derived form for monadic commands (i.e.
+\li{do}-notation) supports binding the result of executing a command to a variable that is then available in the subsequent commands in a command sequence.
+In VerseML, this cannot be expressed in the same way.
+Values can be communicated from the expansion to a spliced expression only via function arguments.
+%We will show an alternative formulation of Haskell's syntax for monadic commands that uses VerseML's anonymous function syntax to bind variables in Sec. \ref{sec:application-monadic-commands}. 
+We will return to this issue in Sec. \ref{sec:discussion}.
+
+\subsubsection{Context Dependence}\label{sec:context-dependence}
+%The prohibition on shadowing ensures only that variables that appear in spliced terms do not refer to bindings that appear in the surrounding expansion. 
+The proto-expansion validation process also ensures that variables that appear in the proto-expansion do not refer to bindings that appear either at the TSM definition or the application site.
+In other words, expansions must be completely \emph{context independent} -- they can make no assumptions about the surrounding context.
+
+A minimal example of a ``broken'' TSM that does not generate context-independent proto-expansions is below:
+\begin{lstlisting}[numbers=none]
+  syntax $broken at rx by
+    static fn(_) => Success (Var "SSTRxESTR")
+  end
+\end{lstlisting}
+The proto-expansion that this TSM generates (for any given literal body) refers to a variable \li{x} that it does not itself bind.
+If proto-expansion validation permitted such a proto-expansion, it would be well-typed only under those application site typing contexts where \li{x} is bound.
+This ``hidden assumption'' makes reasoning about binding and renaming difficult, so this proto-expansion is deemed invalid (even when \li{$broken} is applied where \li{x} is coincidentally bound.)
+
+Of course, this prohibition does not extend into the spliced terms in a proto-expansion -- spliced terms appear at the application site, so they can justifiably refer to application site bindings.
+Indeed, we saw examples of spliced terms that referred to variables bound at the application site above.
+Because proto-expansions refer to spliced terms indirectly, enforcing context independence is straightforward -- we need only that the proto-expansion itself be closed, without considering the spliced terms.% In the next section, we will formalize this intuition.
+% The TSM provider can only refer to them opaquely.
+
+One subtlety is that we assumed in the examples above that constructors like \li{H1Element} were available to the proto-expansion.
+In most dialects of ML, where datatype constructors are injected into the context as variables when the datatype is defined, this would violate context independence.
+As such, we will have to assume for now (and in our formalism in Sec.
+\ref{sec:setsms-formally}) that explicit injections into labeled sum types are available.
+Once we introduce module parameters in Sec.
+\ref{sec:ptsms}, we will be able to pass in a module exporting the datatype constructors as a parameter and partially apply that parameter to hide this detail from clients.
+
+% This prohibition on context dependence explains why the expansion generated by the TSM application in Sec.
+\ref{sec:uetsms-usage} cannot make use of the regex value constructors, e.g.
+\li{Str} and \li{Or}, directly.
+(In Chapter \ref{chap:ptsms}, we will relax this restriction to allow proto-expansions to access explicit parameters.)
+
+% Collectively, we refer to the prohibition on capture and the prohibition on context dependence as \emph{hygiene properties}, by conceptual analogy to corresponding properties in term-rewriting macro systems (see Sec.
+\ref{sec:macro-systems}.) The novelty here comes from the fact that spliced terms are being extracted from an initially unparsed sequence of characters.
+% In the examples in Sec.
+\ref{sec:uetsms-usage} and Sec.
+\ref{sec:splicing-and-hygiene}, the expansion used constructors associated with the \li{Rx} type, e.g.
+\li{Seq} and \li{Str}.
+This might appear to violate our prohibition on context-dependent expansions.
+This is not the case only because in VerseML, constructor labels are not variables or scoped symbols.
+Syntactically, they must begin with a capital letter (like Haskell's datatype constructors).
+Different labeled sum types can use common constructor labels without conflict because the type the term is being checked against -- e.g.
+\li{Rx}, due to the type ascription on \li{$rx} -- determines which type of value will be constructed.
+For dialects of ML where datatype definitions do introduce new variables or scoped symbols, we need parameterized TSMs.
+We will return to this topic in Chapter \ref{chap:ptsms}.
+% Indeed, we used the label \li{Spliced} for two different recursive labeled sum types in Figure \ref{fig:candidate-exp-verseml}.
+
+\subsubsection{Typing}
+Finally, proto-expansion validation maintains a reasonable \emph{typing discipline} by:
+\begin{enumerate}
+\item checking to ensure that the generated expansion is of the type specified in the TSM's type annotation; %For example, the type annotation on \li{$rx} is \li{at rx}, so proto-expansion validation ensures that the final expansion is of type \li{rx}.
+\item checking that the each spliced type is valid at the application site; 
+\item checking that the type annotation on each spliced expression is valid at the application site; and 
+\item checking each spliced expression against the provided type annotation.
+\end{enumerate}
+
+ % This addresses the problem of reasoning abstractly about \textbf{Typing} described in Sec. \ref{sec:intro}, i.e.:
+ % \begin{enumerate}
+ %   \item determining the type of an expansion requires examining only the type annotation on the TSM definition (much as determining the type of a function application requires examining only the function's type); and
+ % \item determining the type that a spliced expression must have requires only the information in the splice summary (rather than complete knowledge of the proto-expansion).
+ % \end{enumerate}
+
+% The language \emph{validates} proto-expansions before a final expansion is generated.
+One aspect of proto-expansion validation is checking  the proto-expansion against the type annotation specified by the TSM, e.g.
+the type \li{Rx} in the example above.
+This maintains a \emph{type discipline}: if a programmer sees a TSM being applied when examining a well-typed program, they need only look up the TSM's type annotation to determine the type of the generated expansion.
+Determining the type does not require examining the expansion directly.
+
+% \subsection{Hygiene}
+% The spliced subexpressions that the proto-expansion refers to (by their position within the literal body, cf.
+above) must be parsed, typed and expanded during the proto-expansion validation process (otherwise, the language would not be able to check the type of the proto-expansion).
+To maintain a useful \emph{binding discipline}, i.e.
+to allow programmers to reason also about variable binding without examining expansions directly, the validation process maintains two additional properties related to spliced subexpressions: \textbf{context independent expansion} and \textbf{expansion independent splicing}.
+These are collectively referred to as the \emph{hygiene properties} (because they are conceptually related to the concept of hygiene in term rewriting macro systems, cf.
+Sec.
+\ref{sec:term-rewriting}.)
+
+% \paragraph{Context Independent Expansion}
+
+% \paragraph{Expansion Independent Splicing} 
+% %These properties suffice to ensure that programmers and tools can freely rename a variable without changing the meaning of the program.
+The only information that is necessary to perform such a \emph{rename refactoring} is the locations of spliced subexpressions within all the literal forms for which the variable being renamed is in scope; the expansions need not otherwise be examined.
+It would be straightforward to develop a tool and/or editor plugin to indicate the locations of spliced subexpressions to the user, like we do in this document (by coloring spliced subexpressions black).
+We discuss tool support as future work in Sec.
+\ref{sec:interaction-with-tools}.
+
+% \subsubsection{Final Expansion}
+
+% For example, the final expansion of the body of \li{lookup_rx} is equivalent to the following, under an environment where the regex value constructors are available:
+% \begin{lstlisting}[numbers=none]
+% Seq(Str(name), Seq(Str "SSTR: ESTR", ssn))
+% \end{lstlisting}
+% (Again, due to the prohibition on context dependent expansions, the final expansion actually involves explicit \li{fold} and \li{inj} operators.)
+
+\newcommand{\spTSMsSec}{Simple Pattern TSMs (spTSMs)}
+\section{\protect\spTSMsSec}
+\label{sec:sptsms}
+Let us now briefly consider the topic of TSMs that expand to \emph{patterns}, rather than expressions.
+For example, we can pattern match on a value \li{x : html} by applying a pattern TSM \li{$html} as follows:
+\begin{lstlisting}[numbers=none]
+  match x with 
+  | $html `SURL<h1>{EURLySURL}</h1>EURL` -> y (* y : list(html) *)
+  | _ -> (* ... *)
+  end
+\end{lstlisting}
+
+Pattern TSM definitions look much like expression TSM definitions:
+\begin{lstlisting}
+  syntax $html at html for patterns by 
+    static fn(b : body) : parse_result(proto_pat) => 
+      (* HTML pattern parser here *)
+  in (* ... *) end
+\end{lstlisting}
+The \emph{sort qualifier}, \li{for patterns}, indicates that this is a pattern TSM definition.
+Expression TSM definitions can also include a sort qualifier, \li{for expressions}, but this is the default if omitted.
+Note that we used the same name, \li{$html}, for this pattern TSM as for the expression TSM defined in the previous section.
+This is possible because patterns and expressions are distinct sorts of terms.
+It does not make sense to apply an expression TSM in pattern position, and \emph{vice versa}.\footnote{The fact that some patterns look like expressions in the textual syntax is immaterial to this fundamental semantic distinction.
+Additionally, many expression-level constructs, e.g.
+lambdas, do not correspond even syntactically to patterns.}
+The return type of the parse function is \li{parse_result(proto_pat)}, rather than \li{parse_result(proto_expr)}.
+The type \li{proto_pat}, defined in part in Figure \ref{fig:CEPat}, classifies encodings of \emph{proto-patterns}.
+
+\subsection{Splicing}
+The constructor \li{SplicedP} serves much like \li{SplicedE} to allow a proto-pattern to refer indirectly to spliced pattern segments by their location within the literal body.
+For example, the proto-pattern generated for the example at the top of this section would be written concretely as follows:
+\begin{lstlisting}[numbers=none]
+  H1Element (_, spliced<5; 5; list(html)>)
+\end{lstlisting}
+
+\begin{figure}
+\begin{lstlisting}[numbers=none]
+type proto_pat = (* no variable pattern form! *) 
+                 Wild
+               | (* ... other standard pattern forms ... *)
+               | SplicedP of segment * proto_typ
+\end{lstlisting}
+\vspace{-5px}
+\caption[Abbreviated definition of \li{proto_pat} in VerseML]{Abbreviated definition of \li{proto_pat}.}
+\label{fig:CEPat}
+\end{figure}
+
+\subsection{Proto-Pattern Validation}
+Proto-pattern validation serves, like proto-expression validation, to maintain the ability to reason abstractly about binding and typing.
+
+\subsubsection{Binding} To maintain a reasonable abstract binding discipline, i.e.
+to allow clients to reason about variable binding without examining pattern TSM expansions directly, variable patterns can appear only within spliced patterns.
+Enforcing this restriction is straightforward: we simply have not defined a variant of the \li{proto_pat} type that encodes variable patterns (though wildcards are allowed.)  This prohibition ensures that no variables other than those visible to the client in a spliced pattern are captured by the corresponding branch of the match expression.
+
+In our simple language, patterns have no way to refer to surrounding bindings (they only induce bindings).
+However, the type annotations on references to spliced patterns could refer to type variables, so we also need to enforce context independence in the manner discussed in the previous discussion.
+
+Patterns cannot bind variables for use in sub-patterns, so we do not face the problem of capture.
+%However, in languages with more complex patterns, e.g.
+those that support boolean expressions as guards, we would need to follow the approach described in the Sec.
+\ref{sec:uetsms-validation} to enforce context independence.
+
+\subsubsection{Typing} To maintain a reasonable abstract typing discipline, proto-pattern validation checks:
+\begin{enumerate}
+  \item that the final expansion is a pattern that matches values of the type specified in the TSM type annotation;
+  \item that each spliced pattern matches values of the type indicated in the splice summary; and 
+  \item that each of these types are themselves well-formed types at the application site.
+\end{enumerate}
+\newcommand{\seTSMsFormallySec}{Simple TSMs, Formally}
+\section{\protect\seTSMsFormallySec}
+\label{sec:setsms-formally}
+\begin{figure}[t!]
+\begin{minipage}{\textwidth}
+$\arraycolsep=3pt\begin{array}{llcl}
+\mathsf{UTyp} & \utau & ::= & 
+\ut ~\vert~ 
+\parr{\utau}{\utau} ~\vert~
+\forallt{\ut}{\utau} ~\vert~
+\rect{\ut}{\utau} ~\vert~
+\prodt{\mapschema{\utau}{i}{\labelset}} ~\vert~
+\sumt{\mapschema{\utau}{i}{\labelset}}\\
+\mathsf{UExp} & \ue & ::= & 
+\ux ~\vert~
+% \asc{\ue}{\utau} ~\vert~
+% \letsyn{\ux}{\ue}{\ue} & \text{value binding}\\
+\lam{\ux}{\utau}{\ue} ~\vert~
+\ap{\ue}{\ue} ~\vert~
+\Lam{\ut}{\ue} ~\vert~
+\App{\ue}{\utau} ~\vert~
+\fold{\ue} ~\vert~
+% \unfold{\ue} ~\vert~
+\tpl{\mapschema{\ue}{i}{\labelset}} ~\vert~
+\inj{\ell}{\ue} ~\vert~
+\matchwith{\ue}{\seqschemaX{\urv}}
+ \\
+& & \vert & \uesyntaxq{\tsmv}{\utau}{e}{\ue} ~\vert~ \utsmap{\tsmv}{b} \\   
+& & \vert & \usyntaxup{\tsmv}{\utau}{e}{\ue} \\
+\mathsf{URule} & \urv & ::= & 
+%& \aumatchrule{\upv}{\ue} 
+\matchrule{\upv}{\ue} \\
+\mathsf{UPat} & \upv & ::= & 
+%& \ux 
+\ux ~\vert~
+\wildp ~\vert~ 
+\foldp{\upv} ~\vert~
+\tplp{\mapschema{\upv}{i}{\labelset}} ~\vert~
+\injp{\ell}{\upv} ~\vert~
+\utsmap{\tsmv}{b}
+
+% \LCC  &  & 
+% %& \lightgray 
+% & \color{Yellow} & \color{Yellow} \\
+% &&
+% %& \audefuetsm{\utau}{e}{\tsmv}{\ue} 
+% & \uesyntax{\tsmv}{\utau}{e}{\ue} & \text{seTSM definition}\\ 
+% &&
+% %& \autsmap{b}{\tsmv} 
+% & \utsmap{\tsmv}{b} & \text{seTSM application}\ECC
+\end{array}$
+\end{minipage}
+\vspace{-10px}
+\caption[Syntax of the $\miniVersePat$ unexpanded language (UL)]{Syntax of the $\miniVersePat$ unexpanded language (UL).
+Metavariable $\ut$ ranges over type identifiers, $\ux$ over expression identifiers, $\ell$ over labels, $\labelset$ over finite sets of labels, $\tsmv$ over TSM names and $b$ over literal bodies.
+We write $\mapschema{\utau}{i}{\labelset}$ for a finite mapping of each label $i$ in $\labelset$ to some unexpanded type $\utau_i$, and similarly for other sorts.
+We write $\seqschemaX{\urv}$ for a finite sequence of $n$ unexpanded rules.
+}
+\vspace{-5px}
+\label{fig:U-unexpanded-terms}
+\end{figure}
+\begin{figure}
+%\hspace{-5px}
+\begin{minipage}{\textwidth}
+$\arraycolsep=3pt\begin{array}{llcl}
+\mathsf{Typ} & \tau & ::= & t ~\vert~ \aparr{\tau}{\tau} ~\vert~ \aall{t}{\tau} ~\vert~ \arec{t}{\tau} ~\vert~ \aprod{\labelset}{\mapschema{\tau}{i}{\labelset}} ~\vert~ \asum{\labelset}{\mapschema{\tau}{i}{\labelset}}\\
+\mathsf{Exp} & e & ::= & x ~\vert~ \aelam{\tau}{x}{e} ~\vert~ \aeap{e}{e} ~\vert~ \aetlam{t}{e} ~\vert~ \aetap{e}{\tau} ~\vert~ \aefold{e} ~\vert~ \aetpl{\labelset}{\mapschema{e}{i}{\labelset}} ~\vert~ \aein{\ell}{e} \\
+& & \vert & \aematchwith{n}{e}{\seqschemaX{r}}\\
+\mathsf{Rule} & r & ::= & \aematchrule{p}{e}\\
+\mathsf{Pat} & p & ::= & x  ~\vert~ \aewildp ~\vert~ \aefoldp{p} ~\vert~ \aetplp{\labelset}{\mapschema{p}{i}{\labelset}} ~\vert~ \aeinjp{\ell}{p}
+\end{array}$
+\end{minipage}
+\vspace{-10px}
+\caption[Syntax of the XL of $\miniVersePat$]{Syntax of the $\miniVersePat$ (XL).
+XL terms are \emph{abstract binding trees} (ABTs) identified up to alpha-equivalence, so we follow the syntactic conventions of \citet{pfple1}.
+Metavariable $x$ ranges over variables and $t$ over type variables.
+%When using stylized forms, the label set is omitted when it can be inferred, e.g.
+the labeled product type $\prodt{\finmap{\mapitem{\ell_1}{e_1}, \mapitem{\ell_2}{e_2}}}$ leaves the label set $\{\ell_1, \ell_2\}$ implicit.
+% When we use the stylized forms, we assume that the reader can infer suppressed indices and arguments from the surrounding context.
+}
+\vspace{-4px}
+\label{fig:U-expanded-terms}
+\end{figure}
+
+Before continuing on to consider parametric TSMs, let us develop a calculus of simple expression and pattern TSMs called $\miniVersePat$.
+It is organized as an \emph{unexpanded language}, or \emph{UL}, defined by typed expansion to an \emph{expanded language}, or \emph{XL}.
+Figures \ref{fig:U-unexpanded-terms}  and \ref{fig:U-expanded-terms} summarize the syntax of the UL and  the XL, respectively.
+Programs are written as unexpanded expressions but evaluate as well-typed expanded expressions.
+We will start with a brief overview of our XL before turning in the remainder of the section on the UL and the typed expansion process.
+
+\subsection{Expanded Language (XL)}
+The {XL} of $\miniVersePat$ forms a standard pure functional language with partial function types, quantification over types, recursive types, labeled product types and labeled sum types. 
+ The reader is directed to \emph{PFPL} \cite{pfple1} for a detailed introductory account of these standard constructs.
+We will only tersely summarize the statics and dynamics of the XL below.
+The particularities of the XL are not critical to the ideas we will introduce below.
+
+\subsubsection{Statics of the Expanded Language}
+The \emph{statics of the XL} is defined by hypothetical judgements of the following form:
+\[\begin{array}{ll}
+% \textbf{Judgement Form} & \textbf{Description}\\
+\istypeU{\Delta}{\tau} & \text{$\tau$ is a well-formed type}\\
+%\isctxU{\Delta}{\Gamma} & \text{$\Gamma$ is a well-formed typing context assuming $\Delta$}\\
+\hastypeU{\Delta}{\Gamma}{e}{\tau} & \text{$e$ is assigned type $\tau$}\\
+\ruleType{\Delta}{\Gamma}{r}{\tau}{\tau'} & \text{$r$ takes values of type $\tau$ to values of type $\tau'$}\\
+\patType{\pctx}{p}{\tau} & \text{$p$ matches values of type $\tau$ and generates hypotheses $\pctx$} 
+\end{array}\]
+\emph{Type formation contexts}, $\Delta$, are finite sets of hypotheses of the form $\Dhyp{t}$.
+%Empty finite sets are written $\emptyset$, or omitted entirely within judgements, and non-empty finite sets are written as comma-separated finite sequences identified up to exchange and contraction.
+We write $\Delta, \Dhyp{t}$ when $\Dhyp{t} \notin \Delta$ for $\Delta$ extended with the hypothesis $\Dhyp{t}$.
+%Finite sets are written as finite sequences identified up to exchange.% We write $\Dcons{\Delta}{\Delta'}$ for the union of $\Delta$ and $\Delta'$.
+\emph{Typing contexts}, $\Gamma$, are finite functions that map each variable $x \in \domof{\Gamma}$, where $\domof{\Gamma}$ is a finite set of variables, to the hypothesis $\Ghyp{x}{\tau}$, for some $\tau$.
+The judgements above are inductively defined in the supplemental material and validate standard lemmas, also given in the supplement: Weakening, Substitution and Decomposition.
+%Empty typing contexts are written $\emptyset$, or omitted entirely within judgements, and non-empty typing contexts are written as finite sequences of hypotheses identified up to exchange and contraction.
+We write $\Gamma, \Ghyp{x}{\tau}$, when $x \notin \domof{\Gamma}$, for the extension of $\Gamma$ with a mapping from $x$ to $\Ghyp{x}{\tau}$, and $\Gcons{\Gamma}{\Gamma'}$ when $\domof{\Gamma} \cap \domof{\Gamma'} = \emptyset$ for the typing context mapping each $x \in \domof{\Gamma} \cup \domof{\Gamma'}$ to $x : \tau$ if $x : \tau \in \Gamma$ or $x : \tau \in \Gamma'$.
+% We write $\isctxU{\Delta}{\Gamma}$ if every type in $\Gamma$ is well-formed relative to $\Delta$.
+% \begin{definition}[Typing Context Formation] \label{def:isctxU}
+% $\isctxU{\Delta}{\Gamma}$ iff for each hypothesis $x : \tau \in \Gamma$, we have $\istypeU{\Delta}{\tau}$.
+% \end{definition}
+
+% \begin{subequations}\label{rules:istypeU}
+% \begin{equation*}\label{rule:istypeU-var}
+% \inferrule{ }{\istypeU{\Delta, \Dhyp{t}}{t}}
+% \end{equation*}
+% \begin{equation*}\label{rule:istypeU-parr}
+% \inferrule{
+%   \istypeU{\Delta}{\tau_1}\\
+%   \istypeU{\Delta}{\tau_2}
+% }{\istypeU{\Delta}{\aparr{\tau_1}{\tau_2}}}
+% \end{equation*}
+% \begin{equation*}\label{rule:istypeU-all}
+%   \inferrule{
+%     \istypeU{\Delta, \Dhyp{t}}{\tau}
+%   }{
+%     \istypeU{\Delta}{\aall{t}{\tau}}
+%   }
+% \end{equation*}
+% \begin{equation*}\label{rule:istypeU-rec}
+%   \inferrule{
+%     \istypeU{\Delta, \Dhyp{t}}{\tau}
+%   }{
+%     \istypeU{\Delta}{\arec{t}{\tau}}
+%   }
+% \end{equation*}
+% \begin{equation*}\label{rule:istypeU-prod}
+%   \inferrule{
+%     \{\istypeU{\Delta}{\tau_i}\}_{i \in \labelset}
+%   }{
+%     \istypeU{\Delta}{\aprod{\labelset}{\mapschema{\tau}{i}{\labelset}}}
+%   }
+% \end{equation*}
+% \begin{equation*}\label{rule:istypeU-sum}
+%   \inferrule{
+%     \{\istypeU{\Delta}{\tau_i}\}_{i \in \labelset}
+%   }{
+%     \istypeU{\Delta}{\asum{\labelset}{\mapschema{\tau}{i}{\labelset}}}
+%   }
+% \end{equation*}
+% \end{subequations}
+% Premises of the form $\{{J}_i\}_{i \in \labelset}$ mean that for each $i \in \labelset$, the judgement ${J}_i$ must hold.
+
+% \begin{subequations}\label{rules:hastypeU}
+% \begin{equation*}\label{rule:hastypeU-var}
+%   \inferrule{ }{
+%     \hastypeU{\Delta}{\Gamma, \Ghyp{x}{\tau}}{x}{\tau}
+%   }
+% \end{equation*}
+% \begin{equation*}\label{rule:hastypeU-lam}
+%   \inferrule{
+%     \istypeU{\Delta}{\tau}\\
+%     \hastypeU{\Delta}{\Gamma, \Ghyp{x}{\tau}}{e}{\tau'}
+%   }{
+%     \hastypeU{\Delta}{\Gamma}{\aelam{\tau}{x}{e}}{\aparr{\tau}{\tau'}}
+%   }
+% \end{equation*}
+% \begin{equation*}\label{rule:hastypeU-ap}
+%   \inferrule{
+%     \hastypeU{\Delta}{\Gamma}{e_1}{\aparr{\tau}{\tau'}}\\
+%     \hastypeU{\Delta}{\Gamma}{e_2}{\tau}
+%   }{
+%     \hastypeU{\Delta}{\Gamma}{\aeap{e_1}{e_2}}{\tau'}
+%   }
+% \end{equation*}
+% \begin{equation*}\label{rule:hastypeU-tlam}
+%   \inferrule{
+%     \hastypeU{\Delta, \Dhyp{t}}{\Gamma}{e}{\tau}
+%   }{
+%     \hastypeU{\Delta}{\Gamma}{\aetlam{t}{e}}{\aall{t}{\tau}}
+%   }
+% \end{equation*}
+% \begin{equation*}\label{rule:hastypeU-tap}
+%   \inferrule{
+%     \hastypeU{\Delta}{\Gamma}{e}{\aall{t}{\tau}}\\
+%     \istypeU{\Delta}{\tau'}
+%   }{
+%     \hastypeU{\Delta}{\Gamma}{\aetap{e}{\tau'}}{[\tau'/t]\tau}
+%   }
+% \end{equation*}
+% \begin{equation*}\label{rule:hastypeU-fold}
+%   \inferrule{\
+%     \istypeU{\Delta, \Dhyp{t}}{\tau}\\
+%     \hastypeU{\Delta}{\Gamma}{e}{[\arec{t}{\tau}/t]\tau}
+%   }{
+%     \hastypeU{\Delta}{\Gamma}{\aefold{e}}{\arec{t}{\tau}}
+%   }
+% \end{equation*}
+% \begin{equation*}\label{rule:hastypeU-unfold}
+%   \inferrule{
+%     \hastypeU{\Delta}{\Gamma}{e}{\arec{t}{\tau}}
+%   }{
+%     \hastypeU{\Delta}{\Gamma}{\aeunfold{e}}{[\arec{t}{\tau}/t]\tau}
+%   }
+% \end{equation*}
+% \begin{equation*}\label{rule:hastypeU-tpl}
+%   \inferrule{
+%     \{\hastypeU{\Delta}{\Gamma}{e_i}{\tau_i}\}_{i \in \labelset}
+%   }{
+%     \hastypeU{\Delta}{\Gamma}{\aetpl{\labelset}{\mapschema{e}{i}{\labelset}}}{\aprod{\labelset}{\mapschema{\tau}{i}{\labelset}}}
+%   }
+% \end{equation*}
+% \begin{equation*}\label{rule:hastypeU-pr}
+%   \inferrule{
+%     \hastypeU{\Delta}{\Gamma}{e}{\aprod{\labelset, \ell}{\mapschema{\tau}{i}{\labelset}; \ell \hookrightarrow \tau}}
+%   }{
+%     \hastypeU{\Delta}{\Gamma}{\aepr{\ell}{e}}{\tau}
+%   }
+% \end{equation*}
+% \begin{equation*}\label{rule:hastypeU-in}
+%   \inferrule{
+%     \{\istypeU{\Delta}{\tau_i}\}_{i \in \labelset}\\
+%     \istypeU{\Delta}{\tau}\\
+%     \hastypeU{\Delta}{\Gamma}{e}{\tau}
+%   }{
+%     \hastypeU{\Delta}{\Gamma}{\aein{\labelset, \ell}{\ell}{\mapschema{\tau}{i}{\labelset}; \ell \hookrightarrow \tau}{e}}{\asum{\labelset, \ell}{\mapschema{\tau}{i}{\labelset}; \ell \hookrightarrow \tau}}
+%   }
+% \end{equation*}
+% \begin{equation*}\label{rule:hastypeU-case}
+%   \inferrule{
+%     \hastypeU{\Delta}{\Gamma}{e}{\asum{\labelset}{\mapschema{\tau}{i}{\labelset}}}\\
+%     \istypeU{\Delta}{\tau}\\
+%     \{\hastypeU{\Delta}{\Gamma, x_i : \tau_i}{e_i}{\tau}\}_{i \in \labelset}
+%   }{
+%     \hastypeU{\Delta}{\Gamma}{\aecase{\labelset}{e}{\mapschemab{x}{e}{i}{\labelset}}}{\tau}
+%   }
+% \end{equation*}
+% \end{subequations}
+
+%Rules (\ref{rules:istypeU}) and (\ref{rules:hastypeU}) are syntax-directed, so we assume an inversion lemma for each rule as needed without stating it separately. 
+% The following standard lemmas also hold.
+
+\subsubsection{Structural Dynamics}\label{sec:dynamics-U}
+The \emph{structural dynamics}, a.k.a.
+the \emph{structural operational semantics} \cite{DBLP:journals/jlp/Plotkin04a}, of $\miniVersePat$ is defined as a transition system by judgements of the following form:
+\[\begin{array}{ll}
+% \textbf{Judgement Form} & \textbf{Description}\\
+\stepsU{e}{e'} & \text{$e$ transitions to $e'$}\\
+\isvalU{e} & \text{$e$ is a value}
+\end{array}\]
+We also define auxiliary judgements for \emph{iterated transition}, $\multistepU{e}{e'}$, and \emph{evaluation} to a value, $\evalU{e}{e'}$.
+Additional auxiliary judgements, not shown, are needed to define the dynamics of pattern matching, but they do not appear directly in our subsequent developments, so we omit them.
+We assume an eager dynamics for simplicity and again assume that standard lemmas, stated in the supplement, hold: Canonical Forms, Progress and Preservation.
+
+\subsection{Syntax of the Unexpanded Language}\label{sec:syntax-U}
+Unexpanded types and expressions are simple inductive structures.
+They are \textbf{not} abstract binding trees -- we do \textbf{not} define notions of renaming, alpha-equivalence or substitution for unexpanded terms.
+This is because unexpanded expressions remain ``partially parsed'' due to the presence of literal bodies, $b$, from which spliced terms might be extracted during typed expansion.
+In fact, unexpanded types and expressions do not involve variables at all, but rather \emph{type identifiers}, $\ut$, and \emph{expression identifiers}, $\ux$.
+Identifiers are given meaning by expansion to variables during typed expansion, as we will see below.
+This distinction between identifiers and variables is technically crucial.
+%We \textbf{cannot} adopt the usual definitions of $\alpha$-renaming of identifiers, because unexpanded types and expressions are still in a ``partially parsed'' state -- the literal bodies, $b$, within an unexpanded expression might contain spliced subterms that are ``surfaced'' by a TSM only during typed expansion, as we will detail below.
+%identifiers are given meaning by expansion to variables.
+%In other words, unexpanded expressions are not abstract binding trees, nor sequences of characters, but a ``transitional'' structure with some characteristics of each of these.
+%For this reason, we will need to handle generating fresh variables explicitly at binding sites in our semantics.
+%To do so, we distinguish \emph{type identifiers}, $\ut$, and \emph{expression identifiers}, $\ux$, from type variables, $t$, and expression variables, $x$.
+identifiers will be given meaning by expansion to variables (which, in turn, are given meaning by substitution, as described above).
+
+% There are only two unexpanded expression forms, highlighted in gray in Figure \ref{fig:U-unexpanded-terms}, that do not correspond to expanded expression forms -- the seTSM definition form and the seTSM application form.
+%These are the ``interesting'' forms.
+% These are the ``interesting'' forms.
+% Let us define this correspondence by the metafunction $\Uof{e}$:
+%\[
+%\begin{split}
+%\Uof{x} & = x\\
+%\Uof{\aelam{\tau}{x}{e}} & = \aulam{\tau}{x}{\Uof{e}}\\
+%\Uof{\aeap{e_1}{e_2}} & = \auap{\Uof{e_1}}{\Uof{e_2}}
+%\end{split}
+%\] and so on for the remaining expanded expression forms.
+
+There is also a corresponding context-free textual syntax for the UL. 
+Giving a complete definition of the context-free textual syntax as, e.g., a context-free grammar, is not critical to our purposes here. 
+%Our paper on Wyvern defines a textual syntax for a similar system \cite{TSLs}. 
+Instead, we need only posit the existence of partial metafunctions $\parseUTypF{b}$, $\parseUExpF{b}$ and $\parseUPatF{b}$  that go from character sequences, $b$, to unexpanded types, expressions and patterns, respectively.
+% \begingroup
+% \def\thetheorem{\ref{condition:textual-representability-SES}}
+\begin{condition}[Textual Representability] ~
+\begin{enumerate}[nolistsep]
+\item For each $\utau$, there exists $b$ such that $\parseUTyp{b}{\utau}$. 
+\item For each $\ue$, there exists $b$ such that $\parseUExp{b}{\ue}$.
+\item For each $\upv$, there exists $b$ such that $\parseUPat{b}{\upv}$.
+\end{enumerate}
+\end{condition}
+% \endgroup
+
+\subsection{Typed Expansion}\label{sec:typed-expansion-U}
+Unexpanded terms are checked and expanded simultaneously according to the \emph{typed expansion judgements}:
+\[\begin{array}{ll}
+% \textbf{Judgement Form} & \textbf{Description}\\
+\expandsTU{\uDelta}{\utau}{\tau} & \text{$\utau$ has well-formed expansion $\tau$}\\
+\expandsUPX{\ue}{e}{\tau} & \text{$\ue$ has expansion $e$ of type $\tau$}\\
+\ruleExpands{\uDelta}{\uGamma}{\uPsi}{\uPhi}{\urv}{r}{\tau}{\tau'} & \text{$\urv$ has expansion $r$ taking values of type $\tau$ to values of type $\tau'$}\\
+\patExpands{\upctx}{\uPhi}{\upv}{p}{\tau} & \text{$\upv$ has expansion $p$ matching against $\tau$ generating hypotheses $\upctx$}
+\end{array}\]
+%\newcommand{\gray}[1]{{\color{gray} #1}}
+
+These judgements are inductively defined in the supplement. We will reproduce the interesting rules below.
+% \begingroup 
+% \def\thetheorem{\ref{thm:typed-expansion-short-U}}
+
+%These rules validate the following theorem, which establishes that typed expansion produces an expansion of the assigned type. 
+%\begin{theorem}[Typed Expression Expansion] If $\expandsU{\uDD{\uD}{\Delta}}{\uGG{\uG}{\Gamma}}{\uPsi}{\ue}{e}{\tau}$ and $\uetsmenv{\Delta}{\uPsi}$ then $\hastypeU{\Delta}{\Gamma}{e}{\tau}$.\end{theorem}
+%\begin{proof} This is the first part of Theorem \ref{thm:typed-expansion-U}, defined and proven below.\end{proof}
+
+\paragraph{Common Forms} Most of the unexpanded forms in Figure \ref{fig:U-unexpanded-terms}  mirror the expanded forms.
+We refer to these as the \emph{common forms}.
+The mapping from expanded forms to common unexpanded forms is defined explicitly in the supplement.
+The typed expansion rules that handle these common forms mirror the corresponding typing rules. For example, the rules for variables and lambdas are reproduced below:
+%Each of these rules is based on the corresponding typing rule, i.e.
+Rules (\ref{rule:hastypeU-var}) through (\ref{rule:hastypeU-case}), respectively.
+For example, the following typed expansion rules are based on the typing rules (\ref{rule:hastypeU-var}), (\ref{rule:hastypeU-lam}) and (\ref{rule:hastypeU-ap}), respectively:% for unexpanded expressions of variable, function and application form, respectively:
+\begin{mathpar}
+  \inferrule[ee-id]{ }{\expandsUP{\uDelta}{\uGamma, \uGhyp{\ux}{x}{\tau}}{\uPsi}{\uPhi}{\ux}{x}{\tau}}
+
+  \inferrule[ee-lam]{
+    \expandsTU{\uDelta}{\utau}{\tau}\\
+    \expandsUP{\uDelta}{\uGamma, \uGhyp{\ux}{x}{\tau}}{\uPsi}{\uPhi}{\ue}{e}{\tau'}
+  }{\expandsUPX{\lam{\ux}{\utau}{\ue}}{\aelam{\tau}{x}{e}}{\aparr{\tau}{\tau'}}}
+\end{mathpar}
+
+The only subtlety has to do with the relationship between identifiers in the UL and variables in the XL. To understand this, we must first describe in detail how unexpanded contexts work.
+
+\emph{Unexpanded typing contexts}, $\uGamma$, are pairs of the form $\uGG{\uG}{\Gamma}$, where $\uG$ is an \emph{expression identifier expansion context}, and $\Gamma$ is a standard typing context.
+An expression identifier expansion context, $\uG$, is a finite function that maps each expression identifier $\ux \in \domof{\uG}$ to the hypothesis $\vExpands{\ux}{x}$, for some expression variable, $x$, called its expansion.
+We write $\ctxUpdate{\uG}{\ux}{x}$ for the expression identifier expansion context that maps $\ux$ to $\vExpands{\ux}{x}$ and defers to $\uG$ for all other expression identifiers (i.e.
+the previous mapping is \textbf{updated}.) Note the distinction between update and extension (which requires that the new identifier is not already in the domain.) %We write $\uGammaOK{\uGamma}$ when $\uGamma=\uGG{\uG}{\Gamma}$ and each expression variable in $\uG$ is assigned a type by $\Gamma$.
+%\begin{definition} $\uGammaOK{\uGG{\uG}{\Gamma}}$ iff for each $\vExpands{\ux}{x} \in \uG$, we have $\Ghyp{x}{\tau} \in \Gamma$ for some $\tau$.\end{definition}
+%\noindent 
+We define $\uGamma, \uGhyp{\ux}{x}{\tau}$ when $\uGamma = \uGG{\uG}{\Gamma}$ as an abbreviation of $\uGG{\ctxUpdate{\uG}{\ux}{x}}{\Gamma, \Ghyp{x}{\tau}}$.
+
+To develop an intuition for why the update operation is necessary, it is instructive to inspect the derivation of the expansion of the unexpanded expression $\lam{\ux}{\utau}{\lam{\ux}{\utau}{\ux}}$ to $\aelam{\tau}{x_1}{\aelam{\tau}{x_2}{x_2}}$ assuming $\expandsTU{\uDelta}{\utau}{\tau}$:
+\begin{mathpar}
+\inferrule{
+  \inferrule{ }{\expandsTU{\uDelta}{\utau}{\tau}}\\
+  \inferrule{
+    \inferrule{ }{\expandsTU{\uDelta}{\utau}{\tau}}\\
+    \inferrule{ }{
+      \expandsUP{\uDelta}{\uGG{\vExpands{\ux}{x_2}}{x_1 : \tau, x_2 : \tau}}{\uPsi}{\uPhi}{\ux}{x_2}{\tau}
+    }~\textsc{ee-id}
+  }{
+    \expandsUP{\uDelta}{\uGG{\vExpands{\ux}{x_1}}{x_1 : \tau}}{\uPsi}{\uPhi}{\lam{\ux}{\utau}{\ux}}{\aelam{\tau}{x_2}{x_2}}{\aparr{\tau}{\tau}}
+    % \expandsU{\uDelta}{\uGG}
+  }~\textsc{ee-lam}
+}{
+  \expandsUP{\uDelta}{\uGG{\emptyset}{\emptyset}}{\uPsi}{\uPhi}{\lam{\ux}{\utau}{\lam{\ux}{\utau}{\ux}}}{\aelam{\tau}{x_1}{\aelam{\tau}{x_2}{x_2}}}{\aparr{\tau}{\aparr{\tau}{\tau}}}
+}~\textsc{ee-lam}
+\end{mathpar}
+Notice that when Rule \textsc{ee-lam} is applied, the type identifier expansion context is updated but the typing context is extended with a (necessarily fresh) variable, first $x_1$ then $x_2$.
+Without this mechanism, expansions for unexpanded terms with shadowing, like this minimal example, would not exist, because we cannot implicitly alpha-vary the unexpanded term to sidestep this problem in the usual manner.
+
+\emph{Unexpanded type formation contexts}, $\uDelta$, consist of a \emph{type identifier expansion context}, $\uD$, paired with a standard type formation context, $\Delta$, and operate analagously (see supplement.)
+% of the form $\uDD{\uD}{\Delta}$, i.e.
+they .
+We similarly define $\uDelta, \uDhyp{\ut}{t}$ when $\uDelta=\uDD{\uD}{\Delta}$ as an abbreviation of $\uDD{\ctxUpdate{\uD}{\ut}{t}}{\Delta, \Dhyp{t}}$.%type identifier expansion context is always extended/updated together with
+
+\paragraph{Well-Typed Expansion} Before we continue, let us state an important invariant: that typed expansion produces a well-typed expression.
+\begin{theorem}[Typed Expression Expansion]\label{thm:typed-expansion-short-U}
+If $\expandsUP{\uDD{\uD}{\Delta}\hspace{-3px}}{\uGG{\uG}{\Gamma}\hspace{-3px}}{\uPsi}{\uPhi}{\ue}{e}{\tau}$ then $\hastypeU{\Delta}{\Gamma}{e}{\tau}$.
+\end{theorem}
+For the typed expansion rules governing common forms, like the two example rules above, the typed expansion rules mirror the corresponding typing rules so it is easy to see that this invariant holds.
+The details are in the supplement.
+The  rules of particular interest are the rules governing TSM definitions and TSM application, which are covered in the next two subsections.
+% \endgroup
+% \subsubsection{Type Expansion}
+% \emph{unexpanded type formation contexts}, $\udelta$, are of the form $\udd{\ud}{\delta}$, i.e.
+they consist of a \emph{type identifier expansion context}, $\ud$, paired with a standard type formation context, $\delta$.
+
+% A \emph{type identifier expansion context}, $\uD$, is a finite function that maps each type identifier $\ut \in \domof{\uD}$ to the hypothesis $\vExpands{\ut}{t}$, for some type variable $t$.
+We write $\ctxUpdate{\uD}{\ut}{t}$ for the type identifier expansion context that maps $\ut$ to $\vExpands{\ut}{t}$ and defers to $\uD$ for all other type identifiers (i.e.
+the previous mapping is \emph{updated}.) We define $\uDelta, \uDhyp{\ut}{t}$ when $\uDelta=\uDD{\uD}{\Delta}$ as an abbreviation of $\uDD{\ctxUpdate{\uD}{\ut}{t}}{\Delta, \Dhyp{t}}$.%type identifier expansion context is always extended/updated together with
+
+% The \emph{type expansion judgement}, $\expandsTU{\uDelta}{\utau}{\tau}$, is inductively defined by the rules given in the supplement. The first three of these rules are reproduced below:
+% % \begin{subequations}%\label{rules:expandsTU}
+% \begin{mathpar}
+% \inferrule[te-id]{ }{\expandsTU{\uDelta, \uDhyp{\ut}{t}}{\ut}{t}}
+
+% \inferrule[te-parr]{
+%   \expandsTU{\uDelta}{\utau_1}{\tau_1}\\
+%   \expandsTU{\uDelta}{\utau_2}{\tau_2}
+% }{\expandsTU{\uDelta}{\parr{\utau_1}{\utau_2}}{\aparr{\tau_1}{\tau_2}}}
+
+% \inferrule[te-all]{
+%     \expandsTU{\uDelta, \uDhyp{\ut}{t}}{\utau}{\tau}
+%   }{
+%     \expandsTU{\uDelta}{\forallt{\ut}{\utau}}{\aall{t}{\tau}}
+% }
+% \end{mathpar}
+% % \begin{equation*}\label{rule:expandsTU-rec}
+% %   \inferrule{
+% %     \expandsTU{\uDelta, \uDhyp{\ut}{t}}{\utau}{\tau}
+% %   }{
+% %     \expandsTU{\uDelta}{\aurec{\ut}{\utau}}{\arec{t}{\tau}}
+% %   }
+% % \end{equation*}
+% % \begin{equation*}\label{rule:expandsTU-prod}
+% %   \inferrule{
+% %     \{\expandsTU{\uDelta}{\utau_i}{\tau_i}\}_{i \in \labelset}
+% %   }{
+% %     \expandsTU{\uDelta}{\auprod{\labelset}{\mapschema{\utau}{i}{\labelset}}}{\aprod{\labelset}{\mapschema{\tau}{i}{\labelset}}}
+% %   }
+% % \end{equation*}
+% % \begin{equation*}\label{rule:expandsTU-sum}
+% %   \inferrule{
+% %     \{\expandsTU{\uDelta}{\utau_i}{\tau_i}\}_{i \in \labelset}
+% %   }{
+% %     \expandsTU{\uDelta}{\ausum{\labelset}{\mapschema{\utau}{i}{\labelset}}}{\asum{\labelset}{\mapschema{\tau}{i}{\labelset}}}
+% %   }
+% % \end{equation*}
+% % \end{subequations}
+% %We write $\uDeltaOK{\uDelta}$ when $\uDelta=\uDD{\uD}{\Delta}$ and each type variable in $\uD$ also appears in $\Delta$.
+% %\begin{definition}\label{def:uDeltaOK} $\uDeltaOK{\uDD{\uD}{\Delta}}$ iff for each $\vExpands{\ut}{t} \in \uD$, we have $\Dhyp{t} \in \Delta$.\end{definition}
+
+% The Type Expansion Lemma establishes that the expansion of an unexpanded type is a well-formed type.
+
+% % \begingroup
+% % \def\thetheorem{\ref{lemma:type-expansion-U}}
+% \begin{lemma}[Type Expansion] If $\expandsTU{\uDD{\uD}{\Delta}}{\utau}{\tau}$ then $\istypeU{\Delta}{\tau}$.\end{lemma}
+% % \begin{proof} By rule induction over Rules (\ref{rules:expandsTU}).
+In each case, we apply the IH to or over each premise, then apply the corresponding type formation rule in Rules (\ref{rules:istypeU}).
+\end{proof}
+% % \endgroup
+% \begin{subequations}\label{rules:expandsU}
+
+% The rules for the remaining expressions of common form are entirely straightforward, mirroring the corresponding typing rules, i.e.
+Rules (\ref{rules:hastypeU}).
+%In particular, observe that, in each of these rules, the unexpanded and expanded expression forms in the conclusion correspond, and each premise corresponds to a premise of the corresponding typing rule.
+%Type formation premises in the typing rule give rise to  type expansion premises in the corresponding typed expansion rule, and each typed expression expansion premise in each rule above corresponds to a typing premise in the corresponding typing rule.
+% The type assigned in the conclusion of each rule above is identical to the type assigned in the conclusion of the corresponding typing rule.
+The seTSM context, $\uPsi$, passes opaquely through these rules (we will define seTSM contexts below.) As such, the corresponding cases in the proof of Theorem \ref{thm:typed-expansion-short-U} are by application of the induction hypothesis and the  corresponding typing rule.
+%Rules (\ref{rules:expandsTU}) could similarly have been generated by mechanically transforming Rules (\ref{rules:istypeU}).
+
+% We can express this scheme more precisely with the rule transformation given in Appendix \ref{appendix:SES-uexps}. For each rule in Rules (\ref{rules:istypeU}) and Rules (\ref{rules:hastypeU}),
+% \begin{mathpar}
+% \refstepcounter{equation}
+% % \label{rule:expandsU-tlam}
+% % \refstepcounter{equation}
+% % \label{rule:expandsU-tap}
+% % \refstepcounter{equation}
+% \label{rule:expandsU-fold}
+% \refstepcounter{equation}
+% \label{rule:expandsU-unfold}
+% \refstepcounter{equation}
+% \label{rule:expandsU-tpl}
+% \refstepcounter{equation}
+% \label{rule:expandsU-pr}
+% \refstepcounter{equation}
+% \label{rule:expandsU-in}
+% \refstepcounter{equation}
+% \label{rule:expandsU-case}
+% \inferrule{J_1\\ \cdots \\ J_k}{J}
+% \end{mathpar}
+% the corresponding typed expansion rule is 
+% \begin{mathpar}
+% \inferrule{
+%   \Uof{J_1} \\
+%   \cdots\\
+%   \Uof{J_k}
+% }{
+%   \Uof{J}
+% }
+% \end{mathpar}
+% where
+% \[\begin{split}
+% \Uof{\istypeU{\Delta}{\tau}} & = \expandsTU{\Uof{\Delta}}{\Uof{\tau}}{\tau} \\
+% \Uof{\hastypeU{\Gamma}{\Delta}{e}{\tau}} & = \expandsU{\Uof{\Gamma}}{\Uof{\Delta}}{\uPsi}{\Uof{e}}{e}{\tau}\\
+% \Uof{\{J_i\}_{i \in \labelset}} & = \{\Uof{J_i}\}_{i \in \labelset}
+% \end{split}\]
+% and where:
+% \begin{itemize}
+% \item $\Uof{\tau}$ is defined as follows:
+%   \begin{itemize}
+%   \item When $\tau$ is of definite form, $\Uof{\tau}$ is defined as in Sec. \ref{sec:syntax-U}.
+%   \item When $\tau$ is of indefinite form, $\Uof{\tau}$ is a uniquely corresponding metavariable of sort $\mathsf{UTyp}$ also of indefinite form.
+For example, in Rule (\ref{rule:istypeU-parr}), $\tau_1$ and $\tau_2$ are of indefinite form, i.e.
+they match arbitrary types.
+The rule transformation simply ``hats'' them, i.e.
+$\Uof{\tau_1}=\utau_1$ and $\Uof{\tau_2}=\utau_2$.
+%   \end{itemize}
+% \item $\Uof{e}$ is defined as follows
+% \begin{itemize}
+% \item When $e$ is of definite form, $\Uof{e}$ is defined as in Sec. \ref{sec:syntax-U}. 
+% \item When $e$ is of indefinite form, $\Uof{e}$ is a uniquely corresponding metavariable of sort $\mathsf{UExp}$ also of indefinite form. For example, $\Uof{e_1}=\ue_1$ and $\Uof{e_2}=\ue_2$.
+% \end{itemize}
+% \item $\Uof{\Delta}$ is defined as follows:
+%   \begin{itemize} 
+%   \item When $\Delta$ is of definite form, $\Uof{\Delta}$ is defined as above.
+%   \item When $\Delta$ is of indefinite form, $\Uof{\Delta}$ is a uniquely corresponding metavariable ranging over unexpanded type formation contexts. For example, $\Uof{\Delta} = \uDelta$.
+%   \end{itemize}
+% \item $\Uof{\Gamma}$ is defined as follows:
+%   \begin{itemize}
+%   \item When $\Gamma$ is of definite form, $\Uof{\Gamma}$ produces the corresponding unexpanded typing context as follows:
+% \begin{align*}
+% \Uof{\emptyset} & = \uGG{\emptyset}{\emptyset}\\
+% \Uof{\Gamma, \Ghyp{x}{\tau}} & = \Uof{\Gamma}, \uGhyp{\identifierof{x}}{x}{\tau}
+% \end{align*}
+%   \item When $\Gamma$ is of indefinite form, $\Uof{\Gamma}$ is a uniquely corresponding metavariable ranging over unexpanded typing contexts. For example, $\Uof{\Gamma} = \uGamma$.
+% \end{itemize}
+% \end{itemize}
+
+% It is instructive to use this rule transformation to generate Rules (\ref{rules:expandsTU}) and Rules (\ref{rule:expandsU-var}) through (\ref{rule:expandsU-tap}) above.
+We omit the remaining rules, i.e.
+Rules (\ref*{rule:expandsU-fold}) through (\ref*{rule:expandsU-case}).
+By instead defining these rules solely by the rule transformation just described, we avoid having to write down a number of rules that are of limited marginal interest.
+Moreover, this demonstrates the general technique for generating typed expansion rules for unexpanded types and expressions of common form, so our exposition is somewhat ``robust'' to changes to the inner core.
+%o that when the inner core changes,  typed expansion rules  our exposition somewhat robust to changes to the inner core (though not to changes to the judgement forms in the statics of the inner core).% Even if changes to the judgement forms in the statics of the inner core are needed (e.g.
+the addition of a symbol context), it is easy to see would correspond to changes in the generic specification above.
+% \begin{subequations}\label{rules:expandsU}
+% \begin{equation*}\label{rule:expandsU-var}
+%   \inferrule{ }{\expandsU{\Delta}{\Gamma, x : \tau}{\uPsi}{x}{x}{\tau}}
+% \end{equation*}
+% \begin{equation*}\label{rule:expandsU-lam}
+%   \inferrule{
+%     \istypeU{\Delta}{\tau}\\
+%     \expandsU{\Delta}{\Gamma, x : \tau}{\uPsi}{\ue}{e}{\tau'}
+%   }{\expandsUX{\aulam{\tau}{x}{\ue}}{\aelam{\tau}{x}{e}}{\aparr{\tau}{\tau'}}}
+% \end{equation*}
+% \begin{equation*}\label{rule:expandsU-ap}
+%   \inferrule{
+%     \expandsUX{\ue_1}{e_1}{\aparr{\tau}{\tau'}}\\
+%     \expandsUX{\ue_2}{e_2}{\tau}
+%   }{
+%     \expandsUX{\auap{\ue_1}{\ue_2}}{\aeap{e_1}{e_2}}{\tau'}
+%   }
+% \end{equation*}
+% \begin{equation*}\label{rule:expandsU-tlam}
+%   \inferrule{
+%     \expandsU{\Delta, \Dhyp{t}}{\Gamma}{\uPsi}{\ue}{e}{\tau}
+%   }{
+%     \expandsUX{\autlam{t}{\ue}}{\aetlam{t}{e}}{\aall{t}{\tau}}
+%   }
+% \end{equation*}
+% \begin{equation*}\label{rule:expandsU-tap}
+%   \inferrule{
+%     \expandsUX{\ue}{e}{\aall{t}{\tau}}\\
+%     \istypeU{\Delta}{\tau'}
+%   }{
+%     \expandsUX{\autap{\ue}{\tau'}}{\aetap{e}{\tau'}}{[\tau'/t]\tau}
+%   }
+% \end{equation*}
+% \begin{equation*}\label{rule:expandsU-fold}
+%   \inferrule{
+%     \istypeU{\Delta, \Dhyp{t}}{\tau}\\
+%     \expandsUX{\ue}{e}{[\arec{t}{\tau}/t]\tau}
+%   }{
+%     \expandsUX{\aufold{t}{\tau}{\ue}}{\aefold{e}}{\arec{t}{\tau}}
+%   }
+% \end{equation*}
+% \begin{equation*}\label{rule:expandsU-unfold}
+%   \inferrule{
+%     \expandsUX{\ue}{e}{\arec{t}{\tau}}
+%   }{
+%     \expandsUX{\auunfold{\ue}}{\aeunfold{e}}{[\arec{t}{\tau}/t]\tau}
+%   }
+% \end{equation*}
+% \begin{equation*}\label{rule:expandsU-tpl}
+%   \inferrule{
+%     \{\expandsUX{\ue_i}{e_i}{\tau_i}\}_{i \in \labelset}
+%   }{
+%     \expandsUX{\autpl{\labelset}{\mapschema{\ue}{i}{\labelset}}}{\aetpl{\labelset}{\mapschema{e}{i}{\labelset}}}{\aprod{\labelset}{\mapschema{\tau}{i}{\labelset}}}
+%   }
+% \end{equation*}
+% \begin{equation*}\label{rule:expandsU-pr}
+%   \inferrule{
+%     \expandsUX{\ue}{e}{\aprod{\labelset, \ell}{\mapschema{\tau}{i}{\labelset}; \mapitem{\ell}{\tau}}}
+%   }{
+%     \expandsUX{\aupr{\ell}{\ue}}{\aepr{\ell}{e}}{\tau}
+%   }
+% \end{equation*}
+% \begin{equation*}\label{rule:expandsU-in}
+%   \inferrule{
+%     \{\istypeU{\Delta}{\tau_i}\}_{i \in \labelset}\\
+%     \istypeU{\Delta}{\tau}\\
+%     \expandsUX{\ue}{e}{\tau}
+%   }{
+%     \left\{\shortstack{$\Delta~\Gamma \vdash_\uPsi \auin{\labelset, \ell}{\ell}{\mapschema{\tau}{i}{\labelset}; \mapitem{\ell}{\tau}}{\ue}$\\$\leadsto$\\$\aein{\labelset, \ell}{\ell}{\mapschema{\tau}{i}{\labelset}; \mapitem{\ell}{\tau}}{e} : \asum{\labelset, \ell}{\mapschema{\tau}{i}{\labelset}; \mapitem{\ell}{\tau}}$\vspace{-1.2em}}\right\}
+%   }
+% \end{equation*}
+% \begin{equation*}\label{rule:expandsU-case}
+%   \inferrule{
+%     \expandsUX{\ue}{e}{\asum{\labelset}{\mapschema{\tau}{i}{\labelset}}}\\
+%     \{\expandsU{\Delta}{\Gamma, \Ghyp{x_i}{\tau_i}}{\uPsi}{\ue_i}{e_i}{\tau}\}_{i \in \labelset}
+%   }{
+%     \expandsUX{\aucase{\labelset}{\ue}{\mapschemab{x}{\ue}{i}{\labelset}}}{\aecase{\labelset}{e}{\mapschemab{x}{e}{i}{\labelset}}}{\tau}
+%   }
+% \end{equation*}
+% \end{subequations}
+
+% \begin{equation*}\label{rule:expandsU-syntax}
+% \inferrule{
+%   \istypeU{\Delta}{\tau}\\
+%   \expandsU{\emptyset}{\emptyset}{\emptyset}{\ueparse}{\eparse}{\aparr{\tBody}{\tParseResultExp}}\\\\
+%   a \notin \domof{\uPsi}\\
+%   \expandsU{\Delta}{\Gamma}{\uPsi, \xuetsmbnd{\tsmv}{\tau}{\eparse}}{\ue}{e}{\tau'}
+% }{
+%   \expandsUX{\audefuetsm{\tau}{\ueparse}{\tsmv}{\ue}}{e}{\tau'}
+% }
+% \end{equation*}
+% \begin{equation*}\label{rule:expandsU-tsmap}
+% \inferrule{
+%   \encodeBody{b}{\ebody}\\
+%   \evalU{\ap{\eparse}{\ebody}}{\inj{\lbltxt{SuccessE}}{\ecand}}\\
+%   \decodeCondE{\ecand}{\ce}\\\\
+%   \cvalidE{\emptyset}{\emptyset}{\esceneU{\Delta}{\Gamma}{\uPsi, \xuetsmbnd{\tsmv}{\tau}{\eparse}}{b}}{\ce}{e}{\tau}
+% }{
+%   \expandsU{\Delta}{\Gamma}{\uPsi, \xuetsmbnd{\tsmv}{\tau}{\eparse}}{\autsmap{b}{\tsmv}}{e}{\tau}
+% }
+% \end{equation*}
+%\end{subequations}
+
+%Notice that each form of expanded expression (Figure \ref{fig:U-expanded-terms}) corresponds to a form of unexpanded expression (Figure \ref{fig:U-unexpanded-terms}).
+For each typing rule in Rules (\ref{rules:hastypeU}), there is a corresponding typed expansion rule -- Rules (\ref{rule:expandsU-var}) through (\ref{rule:expandsU-case}) -- where the unexpanded and expanded forms correspond.
+The premises also correspond -- if a typing judgement appears as a premise of a typing rule, then the corresponding premise in the corresponding typed expansion rule is the corresponding typed expansion judgement.
+The seTSM context is not extended or inspected by these rules (it is only ``threaded through'' them opaquely).
+
+%There are two unexpanded expression forms that do not correspond to an expanded expression form: the seTSM definition form, and the seTSM application form.
+The rules governing these two forms interact with the seTSM context, and are the topics of the next two subsections, respectively.
+
+\subsection{TSM Definitions}\label{sec:U-uetsm-definition}
+An unexpanded expression of seTSM definition form, $\uesyntaxq{\tsmv}{\utau}{\eparse}{\ue}$, 
+%The operational form corresponding to this stylized form is \[\audefuetsm{\utau}{\eparse}{\tsmv}{\ue}\]
+ defines an {seTSM} identified as $\tsmv$ with \emph{unexpanded type annotation} $\utau$ and \emph{parse function} $\eparse$ for use within $\ue$. The rule below defines typed expansion of this form:
+% \begin{subequations}[resume]
+% \begin{equation*}\label{rule:expandsU-syntax}
+% \inferrule{
+%   \istypeU{\Delta}{\tau}\\
+%   \expandsU{\emptyset}{\emptyset}{\emptyset}{\ueparse}{\eparse}{\aparr{\tBody}{\tParseResultExp}}\\\\
+%   \expandsU{\Delta}{\Gamma}{\uPsi, \xuetsmbnd{\tsmv}{\tau}{\eparse}}{\ue}{e}{\tau'}
+% }{
+%   \expandsUX{\audefuetsm{\tau}{\ueparse}{\tsmv}{\ue}}{e}{\tau'}
+% }
+% \end{equation*}
+\begin{mathpar}
+\inferrule[ee-def-setsm]{
+  \expandsTU{\uDelta}{\utau}{\tau}\\
+  \hastypeU{\emptyset}{\emptyset}{\eparse}{\aparr{\tBody}{\tParseResultExp}}\\\\
+  \evalU{\eparse}{\eparse'}\\
+  \expandsUP{\uDelta}{\uGamma}{\uPsi, \uShyp{\tsmv}{a}{\tau}{\eparse'}}{\uPhi}{\ue}{e}{\tau'}
+}{
+  \expandsUPX{\uesyntaxq{\tsmv}{\utau}{\eparse}{\ue}}{e}{\tau'}
+}
+\end{mathpar}
+% \end{subequations}
+The premises of this rule can be understood as follows, in order:
+\begin{enumerate}
+\item The first premise expands the unexpanded type annotation.
+
+\item The second premise checks that $\eparse$ is a closed expanded function\footnote{In Sec.
+\ref{sec:static-eval}, we add the machinery necessary for parse functions that are neither closed nor yet expanded.} of type $\aparr{\tBody}{\tParseResultExp}$.
+
+The type abbreviated $\tBody$ classifies encodings of literal bodies, $b$.
+The mapping from literal bodies to values of type $\tBody$ is defined by the \emph{body encoding judgement} $\encodeBody{b}{\ebody}$.
+An inverse mapping is defined   by the \emph{body decoding judgement} $\decodeBody{\ebody}{b}$.
+Rather than defining $\tBody$ explicitly, and these judgements inductively against that definition (which would be tedious and uninteresting), it suffices to take as a condition that there is an isomorphism between literal bodies and values of type $\tBody$ mediated by these judgements (see supplement.)
+
+The return type of the parse function, $\tParseResultExp$, abbreviates a labeled sum type that distinguishes parse errors from successful parses:
+\begin{align*}
+L_\mathtt{SE} & \defeq \lbltxt{ParseError}, \lbltxt{SuccessE}\\
+\tParseResultExp & \defeq \asum{\labelset_\mathtt{SE}}{
+  \mapitem{\lbltxt{ParseError}}{\prodt{}}, 
+  \mapitem{\lbltxt{SuccessE}}{\tCEExp}
+}
+\end{align*} %[\mapitem{\lbltxt{ParseError}}{\prodt{}}, \mapitem{\lbltxt{SuccessE}}{\tCEExp}]
+% \]
+
+The type abbreviated $\tCEExp$ classifies encodings of \emph{proto-expressions}, $\ce$ (pronounced ``grave $e$''.) The syntax of proto-expressions, defined in Figure \ref{fig:U-candidate-terms}, will be described when we describe proto-expansion validation in Sec.
+\ref{sec:ce-syntax-U}.
+The mapping from proto-expressions to values of type $\tCEExp$ is defined by the \emph{proto-expression encoding judgement}, $\encodeCondE{\ce}{e}$.
+An inverse mapping is defined by the \emph{proto-expression decoding judgement}, $\decodeCondE{e}{\ce}$.
+Again, rather than picking a particular definition of $\tCEExp$ and defining the judgements above inductively against it, we take as a condition that there is an isomorphism between values of type $\tCEExp$ and proto-expressions mediated by these judgements (see supplement.)
+\item The third premise of Rule \textsc{ee-def-setsm} evaluates the parse function to a value.
+This is not semantically necessary, but it is the choice one would expect to make assuming an eagerly evaluated language.
+\item The final premise of Rule \textsc{ee-def-setsm} extends the seTSM context, $\uPsi$, with the newly determined {seTSM definition}, and proceeds to assign a type, $\tau'$, and expansion, $e$, to $\ue$.
+The conclusion of the rule then assigns this type and expansion to the seTSM definition as a whole.% i.e.
+TSMs define behavior that is relevant during typed expansion, but not during evaluation.
+
+\emph{seTSM contexts}, $\uPsi$, are of the form $\uAS{\uA}{\Psi}$, where $\uA$ is a \emph{TSM identifier expansion context} and $\Psi$ is an \emph{seTSM definition context}.
+
+A {TSM identifier expansion context}, $\uA$, is a finite function mapping each TSM identifier $\tsmv \in \domof{\uA}$ to the \emph{TSM identifier expansion}, $\vExpands{\tsmv}{a}$, for some \emph{TSM name}, $a$.
+We distinguish TSM identifiers, $\tsmv$, from TSM names, $a$, for much the same reason that we distinguish type and expression identifiers from type and expression variables: in order to allow a TSM definition to shadow a previously defined TSM definition without relying on an implicit identification convention.
+
+An {seTSM definition context}, $\Psi$, is a finite function mapping each TSM name $a \in \domof{\Psi}$ to an \emph{expanded seTSM definition}, $\xuetsmbnd{a}{\tau}{\eparse}$, where $\tau$ is the seTSM's type annotation, and $\eparse$ is its parse function.
+We define $\uPsi, \uShyp{\tsmv}{a}{\tau}{\eparse}$, when $\uPsi=\uAS{\uA}{\Psi}$, as an abbreviation of $\uAS{\ctxUpdate{\uA}{\tsmv}{a}}{\Psi, \xuetsmbnd{a}{\tau}{\eparse}}$.
+
+%Moreover, this distinction will be crucial in the semantics of TSM abbreviations in Chapter \ref{chap:ptsms}.
+
+\end{enumerate}
+
+The spTSM definition form, $\usyntaxup{\tsmv}{\utau}{\eparse}{\ue}$, operates analagously.
+The rule is reproduced below, and the details, which mirror those for seTSMs, are in the supplement.
+Notice that the spTSM context, $\uPhi$, rather than the $\uPsi$ is updated.
+This allows expression and pattern TSMs to share an identifier.
+\begin{mathpar}
+\inferrule[ee-def-sptsm]{
+  \expandsTU{\uDelta}{\utau}{\tau}\\
+  \hastypeU{\emptyset}{\emptyset}{\eparse}{\aparr{\tBody}{\tParseResultPat}}\\\\
+  \evalU{\eparse}{\eparse'}\\
+  \expandsUP{\uDelta}{\uGamma}{\uPsi}{\uPhi, \uPhyp{\tsmv}{a}{\tau}{\eparse'}}{\ue}{e}{\tau'}
+}{
+  \expandsUPX{\usyntaxup{\tsmv}{\utau}{\eparse}{\ue}}{e}{\tau'}
+}
+\end{mathpar}
+
+% \[\begin{array}{ll}
+% \textbf{Judgement Form} & \textbf{Description}\\
+% \uetsmenv{\Delta}{\uPsi} & \text{$\uPsi$ is well-formed assuming $\Delta$}\end{array}\]
+% This judgement is inductively defined by the following rules:
+% \begin{subequations}[intermezzo]\label{rules:uetsmenv-U}
+% \begin{equation*}\label{rule:uetsmenv-empty}
+% \inferrule{ }{\uetsmenv{\Delta}{\emptyset}}
+% \end{equation*}
+% \begin{equation*}\label{rule:uetsmenv-ext}
+% \inferrule{
+%   \uetsmenv{\Delta}{\uPsi}\\
+%   \istypeU{\Delta}{\tau}\\
+%   \hastypeU{\emptyset}{\emptyset}{\eparse}{\aparr{\tBody}{\tParseResultExp}}
+% }{
+%   \uetsmenv{\Delta}{\uPsi, \xuetsmbnd{\tsmv}{\tau}{\eparse}}
+% }
+% \end{equation*}
+% \end{subequations}
+
+\subsection{TSM Application}\label{sec:U-uetsm-application}
+The unexpanded expression form for applying an seTSM named $\tsmv$ to a literal form with literal body $b$ is $\utsmap{\tsmv}{b}$. 
+% This form uses forward slashes\todo{use backticks} to delimit the literal body, but other generalized literal forms could also be included as derived forms in the textual syntax.
+% (we omit them for simplicity).
+%The corresponding operational form is $\autsmap{b}{\tsmv}$.
+%i.e.
+for each literal body $b$, the operator $\texttt{uapuetsm}[b]$ is indexed by the TSM name $\tsmv$ and takes no arguments.
+%\footnote{This is in following the conventions in \emph{PFPL} \cite{pfpl}, where operators parameters allow for the use of metatheoretic objects that are not syntax trees or binding trees, e.g.
+$\mathsf{str}[s]$ and $\mathsf{num}[n]$.} This operator is indexed by the TSM name $\tsmv$ and takes no arguments.
+The typed expansion rule governing seTSM application is below:
+% \begin{subequations}[resume]
+% \begin{equation*}\label{rule:expandsU-tsmap}
+% \inferrule{
+%   \encodeBody{b}{\ebody}\\
+%   \evalU{\ap{\eparse}{\ebody}}{\inj{\lbltxt{SuccessE}}{\ecand}}\\
+%   \decodeCondE{\ecand}{\ce}\\\\
+%   \cvalidE{\emptyset}{\emptyset}{\esceneU{\Delta}{\Gamma}{\uPsi, \xuetsmbnd{\tsmv}{\tau}{\eparse}}{b}}{\ce}{e}{\tau}
+% }{
+%   \expandsU{\Delta}{\Gamma}{\uPsi, \xuetsmbnd{\tsmv}{\tau}{\eparse}}{\autsmap{b}{\tsmv}}{e}{\tau}
+% }
+% \end{equation*}
+\begin{mathpar}
+\inferrule[ee-ap-setsm]{
+  \uPsi = \uPsi', \uShyp{\tsmv}{a}{\tau}{\eparse}\\\\
+  \encodeBody{b}{\ebody}\\
+  \evalU{\ap{\eparse}{\ebody}}{\aein{\mathtt{SuccessE}}{\ecand}}\\
+  \decodeCondE{\ecand}{\ce}\\\\
+  \segOK{\segof{\ce}}{b}\\
+  \cvalidE{\emptyset}{\emptyset}{\esceneUP{\uDelta}{\uGamma}{\uPsi}{\uPhi}{b}}{\ce}{e}{\tau}
+}{
+  \expandsUP{\uDelta}{\uGamma}{\uPsi}{\uPhi}{\utsmap{\tsmv}{b}}{e}{\tau}
+}
+\end{mathpar}
+The premises of this rule can be understood as follows, in order:
+\begin{enumerate}
+\item The first premise ensures that $\tsmv$ has been defined and extracts the type annotation and parse function.
+\item The second premise determines the encoding of the literal body, $\ebody$, which is a value of type $\tBody$.
+\item The third premise applies the parse function $\eparse$ to the encoding of the literal body.
+If parsing succeeds, i.e.
+a value of the form $\aein{\mathtt{SuccessE}}{\ecand}$ results from evaluation, then $\ecand$ will be a value of type $\tCEExp$ (assuming a well-formed seTSM context, by application of the Preservation assumption.) We call $\ecand$ the \emph{encoding of the proto-expansion}.
+If the parse function produces a value labeled $\lbltxt{ParseError}$, then typed expansion fails.
+Formally, no rule is necessary to handle this case.
+
+\item The fourth premise decodes the encoding of the proto-expansion to produce the {proto-expansion} itself.
+
+\item The fifth premise determines the segmentation, $\segof{\ce}$, and ensures that it is valid with respect to $b$ via the predicate $\segOK{\psi}{b}$, which  checks that each segment in the finite set of segments $\psi$ has non-negative length and is within bounds of $b$, and that the segments in $\psi$ do not overlap.
+
+\item The final premise \emph{validates} the proto-expansion and simultaneously generates the \emph{final expansion}, $e$, which appears in the conclusion of the rule.
+The proto-expression validation judgement is discussed next.
+\end{enumerate}
+
+The typed pattern expansion rule governing spTSM application is analagous:
+\begin{mathpar}
+\inferrule[pe-ap-sptsm]{
+  \uPhi = \uPhi', \uPhyp{\tsmv}{a}{\tau}{\eparse}\\\\
+  \encodeBody{b}{\ebody}\\
+  \evalU{\ap{\eparse}{\ebody}}{\aein{\mathtt{SuccessP}}{\ecand}}\\
+  \decodeCEPat{\ecand}{\cpv}\\\\
+  \segOK{\segof{\cpv}}{b}\\
+  \cvalidP{\upctx}{\pscene{\uDelta}{\uPhi}{b}}{\cpv}{p}{\tau}
+}{
+  \patExpands{\upctx}{\uPhi}{\utsmap{\tsmv}{b}}{p}{\tau}
+}
+\end{mathpar}
+
+% \subsection{Syntax of Proto-Expansions}\label{sec:ce-syntax-U}
+
+\subsection{Proto-Expansion Validation}\label{sec:ce-validation-U}\label{sec:ce-syntax-U}
+
+\begin{figure}[h]
+\vspace{-10px}
+\begin{minipage}{\textwidth}
+$\arraycolsep=3pt\begin{array}{llcl}
+\mathsf{PrTyp} & \ctau & ::= & t ~\vert~
+\aceparr{\ctau}{\ctau} ~\vert~
+\aceall{t}{\ctau} ~\vert~
+\acerec{t}{\ctau} ~\vert~
+\aceprod{\labelset}{\mapschema{\ctau}{i}{\labelset}} ~\vert~
+\acesum{\labelset}{\mapschema{\ctau}{i}{\labelset}} \\
+& & \vert & \acesplicedt{m}{n}\\
+\mathsf{PrExp} & \ce & ::= & x ~\vert~
+% \aceasc{\ctau}{\ce} ~\vert~
+% \aceletsyn{x}{\ce}{\ce} ~\vert~
+\acelam{\ctau}{x}{\ce} ~\vert~
+\aceap{\ce}{\ce} ~\vert~
+\acetlam{t}{\ce} ~\vert~
+\acetap{\ce}{\ctau} ~\vert~
+\acefold{\ce} \\
+& & \vert & \acetpl{\labelset}{\mapschema{\ce}{i}{\labelset}} ~\vert~
+\acein{\ell}{\ce} ~\vert~
+\acematchwith{n}{\ce}{\seqschemaX{\crv}} ~\vert~
+\acesplicede{m}{n}{\ctau}\\
+\mathsf{PrRule} & \crv & ::= & \acematchrule{p}{\ce}\\
+\mathsf{PrPat} & \cpv & ::= & \acewildp ~\vert~
+\acefoldp{p} ~\vert~
+\acetplp{\labelset}{\mapschema{\cpv}{i}{\labelset}} ~\vert~
+\aceinjp{\ell}{\cpv} ~\vert~
+\acesplicedp{m}{n}{\ctau} 
+\end{array}$
+\end{minipage}
+\vspace{-5px}
+\caption[Syntax of $\miniVersePat$ proto-types and proto-expressions]{Syntax of $\miniVersePat$ proto-expansions. Proto-expansion terms are ABTs identified up to alpha-equivalence.}
+\label{fig:U-candidate-terms}
+\vspace{-10px}
+\end{figure}
+
+The \emph{proto-expansion validation judgements} validate proto-expansions and simultaneously generate their final expansions:% are types and expanded expressions, respectively.
+\[\begin{array}{ll}
+\cvalidT{\Delta}{\tscenev}{\ctau}{\tau} & \text{$\ctau$ has well-formed expansion $\tau$}\\
+\cvalidE{\Delta}{\Gamma}{\escenev}{\ce}{e}{\tau} & \text{$\ce$ has expansion $e$ of type $\tau$}\\
+\cvalidR{\Delta}{\Gamma}{\escenev}{\crv}{r}{\tau}{\tau'} & \text{$\crv$ has expansion $r$ taking values of type $\tau$ to values of type $\tau'$}\\
+\cvalidP{\upctx}{\pscenev}{\cpv}{p}{\tau} & \text{$\cpv$ has expansion $p$ matching against $\tau$ generating assumptions $\upctx$}
+\end{array}\]
+The purpose of the \emph{splicing scenes} $\tscenev$, $\escenev$ and $\pscenev$ is to ``remember'', during proto-expansion validation, the contexts and the literal body from the TSM application site (cf.
+Rules \textsc{ee-ap-setsm} and \textsc{pe-ap-sptsm}) for when validation encounters spliced terms.
+\emph{Type splicing scenes}, $\tscenev$, are of the form $\tsceneUP{\uDelta}{b}$.
+\emph{Expression splicing scenes}, $\escenev$, are of the form $\esceneUP{\uDelta}{\uGamma}{\uPsi}{\uPhi}{b}$.
+\emph{Pattern splicing scenes}, $\pscenev$, are of the form $\pscene{\uDelta}{\uPhi}{b}$.
+
+\subsubsection{Common Forms} Most of the proto-expansion forms mirror corresponding expanded forms.
+The rules governing proto-expansion validation for these common forms correspondingly mirror the typing rules.
+They are given in the supplement.
+Splicing scenes pass opaquely through these rules, i.e.
+none of these rules can access the application site contexts.
+This maintains context independence.
+
+Notice that proto-rules, $\crv$, involve expanded patterns, $p$, not proto-patterns, $\cpv$.
+The reason is that proto-rules appear in proto-expressions, which are generated by expression TSMs.
+Proto-patterns, in contrast, arise only from pattern TSMs.
+There a variable proto-pattern form.
+
+\subsubsection{References to Spliced Terms} The only interesting forms are the references to spliced unexpanded types, expressions and patterns.
+Let us first consider the rule for references to spliced unexpanded types:
+\begin{mathpar}
+  \inferrule[ptv-spliced]{
+    \parseUTyp{\bsubseq{b}{m}{n}}{\utau}\\
+    \expandsTU{\uDD{\uD}{\Delta_\text{app}}}{\utau}{\tau}\\
+    \Delta \cap \Delta_\text{app} = \emptyset
+  }{
+    \cvalidT{\Delta}{\tsceneU{\uDD{\uD}{\Delta_\text{app}}}{b}}{\acesplicedt{m}{n}}{\tau}
+  }
+\end{mathpar}
+This first premise of this rule parses out the requested segment of the literal body, $b$, to produce an unexpanded type, $\utau$.
+It then invokes type expansion to ensure that this unexpanded type is well-formed \emph{under the application site context}, $\uDD{\uD}{\Delta_\text{app}}$, but \emph{not} the expansion-local type formation context, $\Delta$.
+The final premise requires that the application site type formation context is disjoint from the expansion-local type formation context.
+Because proto-expansions are identified up to alpha-equivalence, we can always discharge the final premise by alpha-varying the proto-expansion.
+Collectively, this ensures that type variable capture does not occur (we will formally state this property in the next section.)
+
+The rule for references to spliced unexpanded expressions is fundamentally analagous:
+\begin{mathpar}
+\inferrule[pev-spliced]{
+  \parseUExp{\bsubseq{b}{m}{n}}{\ue}\\
+  \cvalidT{\emptyset}{\tsceneUP{\uDelta}{b}}{\ctau}{\tau}\\
+  \expandsUP{\uDD{\uD}{\Delta_\text{app}}}{\uGG{\uG}{\Gamma_\text{app}}}{\uPsi}{\uPhi}{\ue}{e}{\tau}\\\\
+  \Delta \cap \Delta_\text{app} = \emptyset\\
+  \domof{\Gamma} \cap \domof{\Gamma_\text{app}} = \emptyset
+}{
+  \cvalidE{\Delta}{\Gamma}{\esceneUP{\uDD{\uD}{\Delta_\text{app}}}{\uGG{\uG}{\Gamma_\text{app}}}{\uPsi}{\uPhi}{b}}{\acesplicede{m}{n}{\ctau}}{e}{\tau}
+}
+\end{mathpar}
+We first splice out the requested segment.
+The second premise expands the type annotation under an empty context, because the type annotation must be meaningful at the application site (so, independent of $\Delta$ and $\Gamma$) and not itself make any assumptions about the application site context.
+Spliced types can appear in the annotation.
+The third premise performs typed expansion of the spliced unexpanded expression under the application site contexts, but not the expansion-local contexts.
+The final two premises ensure that these contexts are disjoint, again to force capture avoidance.
+
+The rule for references to spliced unexpanded patterns is entirely analagous:
+\begin{mathpar}
+\inferrule[ppv-spliced]{
+  \parseUPat{\bsubseq{b}{m}{n}}{\upv}\\
+  \cvalidT{\emptyset}{\tsceneUP{\uDelta}{b}}{\ctau}{\tau}\\
+  \patExpands{\upctx}{\uPhi}{\upv}{p}{\tau}
+}{
+  \cvalidP{\upctx}{\pscene{\uDelta}{\uPhi}{b}}{\acesplicedp{m}{n}{\ctau}}{p}{\tau}
+}
+\end{mathpar}
+
+\subsection{Metatheory}\label{sec:s-metatheory}
+\subsubsection{Typed Expression Expansion} Let us now return to Theorem \ref{thm:typed-expansion-short-U}, the typed expression expansion theorem which was mentioned at the end of Sec.
+\ref{sec:typed-expansion-U}.
+As it turns out, in order to prove this theorem, we must  prove the following stronger theorem, because the proto-expression validation judgement is defined mutually inductively with the typed expansion judgement (due to the three rules just described.)
+
+% \begingroup
+% \def\thetheorem{\ref{thm:typed-expansion-full-U}}
+\begin{theorem}[Typed Expression Expansion (Strong)] ~
+\begin{enumerate}[nolistsep]
+\item If $\expandsU{\uDD{\uD}{\Delta}}{\uGG{\uG}{\Gamma}}{\uAS{\uA}{\Psi}}{\ue}{e}{\tau}$ then $\hastypeU{\Delta}{\Gamma}{e}{\tau}$.
+\item If $\cvalidE{\Delta}{\Gamma}{\esceneU{\uDD{\uD}{\Delta_\text{app}}}{\uGG{\uG}{\Gamma_\text{app}}}{\uAS{\uA}{\Psi}}{b}}{\ce}{e}{\tau}$ and $\Delta \cap \Delta_\text{app} = \emptyset$ and $\domof{\Gamma} \cap \domof{\Gamma_\text{app}} = \emptyset$ then $\hastypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\Gcons{\Gamma}{\Gamma_\text{app}}}{e}{\tau}$.
+\end{enumerate}
+\end{theorem}
+% \endgroup
+The additional second clause simply states that the final expansion produced by proto-expression validation is well-typed under the combined application site and expansion-internal context (spliced terms are distinguished only in the proto-expansion.) Such combined contexts can only be formed if the constituents are disjoint.
+
+The proof proceeds by mutual rule induction and appeal to simple lemmas about type expansion and proto-type validation (see supplement).
+The proof is straightforward but for one issue: it is not immediately clear that the mutual induction is well-founded, because the case in the proof of part 2 for Rule \textsc{pev-spliced} invokes part 1 of the induction hypothesis on a term that is not a sub-term of the conclusion, but rather parsed out of the literal body, $b$.
+To establish that the mutual induction is well-founded, then, we need to explicitly establish a decreasing metric.
+The intuition is that parsing a term of out a literal body cannot produce a bigger term than the term that contained that very literal body.
+More specifically, {the sum of the lengths of the literal bodies that appear in the term strictly decreases each time you perform a nested TSM application} because some portion of the term has to be consumed by the TSM name and the delimiters.
+The details are given in the supplemental material.
+
+\subsubsection{Typed Pattern Expansion} A similar argument is needed to prove Typed Pattern Expansion:
+\begin{theorem}[Typed Pattern Expansion (Strong)] ~
+\begin{enumerate}[nolistsep]
+  \item If $\pExpandsSP{\uDD{\uD}{\Delta}}{\uAS{\uA}{\Phi}}{\upv}{p}{\tau}{\uGG{\uG}{\pctx}}$ then $\patType{\pctx}{p}{\tau}$.
+  \item If $\cvalidP{\uGG{\uG}{\pctx}}{\pscene{\uDD{\uD}{\Delta}}{\uAP{\uA}{\Phi}}{b}}{\cpv}{p}{\tau}$ then $\patType{\pctx}{p}{\tau}$.
+\end{enumerate}
+\end{theorem}
+
+\subsubsection{seTSM Reasoning Principles} 
+The following theorem summarizes the abstract reasoning principles that programmers can rely on when applying an seTSM. Informal descriptions of important clauses are given inline. 
+% \begingroup
+% \def\thetheorem{\ref{thm:tsc-SES}}
+\begin{theorem}[seTSM Abstract Reasoning Principles]
+If $\expandsU{\uDD{\uD}{\Delta}}{\uGG{\uG}{\Gamma}}{\uPsi}{\utsmap{\tsmv}{b}}{e}{\tau}$ then:
+\begin{enumerate}[nolistsep]
+\item (\textbf{Typing 1}) $\uPsi = \uPsi', \uShyp{\tsmv}{a}{\tau}{\eparse}$ and $\hastypeU{\Delta}{\Gamma}{e}{\tau}$
+  \begin{quote}
+     The type of the expansion is consistent with the type annotation on the applied seTSM definition.
+  \end{quote}
+\item $\encodeBody{b}{\ebody}$
+\item $\evalU{\ap{\eparse}{\ebody}}{\aein{\lbltxt{SuccessE}}{\ecand}}$
+\item $\decodeCondE{\ecand}{\ce}$
+\item (\textbf{Segmentation}) $\segOK{\segof{\ce}}{b}$
+        \begin{quote}
+        The segmentation determined by the proto-expansion actually segments the literal body (i.e. each segment is in-bounds and the segments are non-overlapping.)
+        \end{quote}
+\item $\summaryOf{\ce} = \sseq{\acesplicedt{m'_i}{n'_i}}{\nty} \cup \sseq{\acesplicede{m_i}{n_i}{\ctau_i}}{\nexp}$
+\item \textbf{(Typing 2)} $\sseq{
+      \expandsTU{\uDD{\uD}{\Delta}}
+      {
+        \parseUTypF{\bsubseq{b}{m'_i}{n'_i}}
+      }{\tau'_i}
+    }{\nty}$ and $\sseq{\istypeU{\Delta}{\tau'_i}}{\nty}$
+      \begin{quote}
+        Each spliced type has a well-formed expansion at the application site.
+      \end{quote}
+\item \textbf{(Typing 3)} $\sseq{
+  \cvalidT{\emptyset}{
+    \tsceneUP
+      {\uDD
+        {\uD}{\Delta}
+      }{b}
+  }{
+    \ctau_i
+  }{\tau_i}
+}{\nexp}$ and $\sseq{\istypeU{\Delta}{\tau_i}}{\nexp}$
+    \begin{quote}
+      Each type annotation on a reference to a spliced expression has a well-formed expansion at the application site.
+    \end{quote}
+\item \textbf{(Typing 4)} $\sseq{
+  \expandsU
+    {\uDD{\uD}{\Delta}}
+    {\uGG{\uG}{\Gamma}}
+    {\uPsi}
+    {\parseUExpF{\bsubseq{b}{m_i}{n_i}}}
+    {e_i}
+    {\tau_i}
+}{\nexp}$ and $\sseq{\hastypeU{\Delta}{\Gamma}{e_i}{\tau_i}}{\nexp}$
+    \begin{quote}
+      Each spliced expression has a well-typed expansion consistent with its type annotation.
+    \end{quote}
+\item (\textbf{Capture Avoidance}) $e = [\sseq{\tau'_i/t_i}{\nty}, \sseq{e_i/x_i}{\nexp}]e'$ for some $\sseq{t_i}{\nty}$ and $\sseq{x_i}{\nexp}$ and $e'$
+    \begin{quote}
+      The final expansion can be decomposed into a  term with variables in place of each spliced type or expression.
+The expansions of these spliced types and expressions can be substituted into this term in the standard capture avoiding manner.
+    \end{quote}
+\item (\textbf{Context Independence}) $\mathsf{fv}(e') \subset \sseq{t_i}{\nty} \cup \sseq{x_i}{\nexp}$
+    \begin{quote}
+      The aforementioned decomposed term makes no mention of bindings in the application site context, i.e. the only free variables are those corresponding to the spliced terms.
+    \end{quote}
+  % $\hastypeU
+  % {\sseq{\Dhyp{t_i}}{\nty}}
+  % {\sseq{x_i : \tau_i}{\nexp}}
+  % {e'}{\tau}$
+\end{enumerate}
+\end{theorem}
+The proof, which involves auxiliary lemmas about the decomposition of proto-types and proto-expressions, is given in the supplement.
+
+\subsubsection{spTSM Reasoning Principles} Finally, the following theorem summarizes the abstract reasoning principles available to programmers when applying an spTSM.
+Most of the labeled clauses are analagous to those described above, so we omit their descriptions.
+% \begingroup
+% \def\thetheorem{\ref{thm:spTSM-Typing-Segmentation}}
+\begin{theorem}[spTSM Abstract Reasoning Principles]
+% \label{thm:spTSM-Typing-Segmentation}
+If $\patExpands{\upctx}{\uPhi}{\utsmap{\tsmv}{b}}{p}{\tau}$ where $\uDelta=\uDD{\uD}{\Delta}$ and $\uGamma=\uGG{\uG}{\Gamma}$ then all of the following hold:
+\begin{enumerate}
+        \item (\textbf{Typing 1}) $\uPhi=\uPhi', \uPhyp{\tsmv}{a}{\tau}{\eparse}$ and $\patType{\pctx}{p}{\tau}$
+        \item $\encodeBody{b}{\ebody}$
+        \item $\evalU{\eparse(\ebody)}{\aein{\mathtt{SuccessP}}{\ecand}}$
+        \item $\decodeCEPat{\ecand}{\cpv}$
+        \item (\textbf{Segmentation}) $\segOK{\segof{\cpv}}{b}$
+        \item $\summaryOf{\cpv} = \sseq{\acesplicedt{n'_i}{m'_i}}{\nty} \cup \sseq{\acesplicedp{m_i}{n_i}{\ctau_i}}{\npat}$
+        \item (\textbf{Typing 2}) $\sseq{
+              \expandsTU{\uDelta}
+              {
+                \parseUTypF{\bsubseq{b}{m'_i}{n'_i}}
+              }{\tau'_i}
+            }{\nty}$ and $\sseq{\istypeU{\Delta}{\tau'_i}}{\nty}$
+        \item (\textbf{Typing 3}) $\sseq{
+          \cvalidT{\emptyset}{
+            \tsceneUP
+              {\uDelta}{b}
+          }{
+            \ctau_i
+          }{\tau_i}
+        }{\npat}$ and $\sseq{\istypeU{\Delta}{\tau_i}}{\npat}$
+        \item (\textbf{Typing 4}) $\sseq{
+          \patExpands
+            {\upctx_i}
+            {\uPhi}
+            {\parseUPatF{\bsubseq{b}{m_i}{n_i}}}
+            {p_i}
+            {\tau_i}
+        }{\npat}$ and $\sseq{\patType{\pctx_i}{p_i}{\tau_i}}{\npat}$
+      \item (\textbf{No Hidden Bindings}) $\upctx = \biguplus_{0 \leq i < \npat} \upctx_i$
+        \begin{quote}
+          The hypotheses generated by the TSM application are exactly those generated by the spliced patterns.
+        \end{quote}
+\end{enumerate}
+\end{theorem}
+% \begin{proof} 
+The proof, in the supplement, relies on an auxiliary lemma about decomposing proto-patterns.
+
+% We write $\tsfrom{\escenev}$ for the type splicing scene constructed by dropping unnecessary contexts from $\escenev$:
+% \[\tsfrom{\esceneUP{\uDelta}{\uGamma}{\uPsi}{\uPhi}{b}} = \tsceneUP{\uDelta}{b}\]
+
+% Figure \ref{fig:U-candidate-terms} defines the syntax of proto-types, $\ctau$, and proto-expressions, $\ce$. Proto-types and -expressions
+
+% Each expanded form maps onto a proto-expansion form. We refer to these as the \emph{common proto-expansion forms}. The mapping is given explicitly in Appendix \ref{appendix:proto-expansions-SES}.
+
+% There are two ``interesting'' proto-expansion forms, highlighted in yellow in Figure \ref{fig:U-candidate-terms}: a proto-type form for \emph{references to spliced unexpanded types}, $\acesplicedt{m}{n}$, and a proto-expression form for \emph{references to spliced unexpanded expressions}, $\acesplicede{m}{n}{\ctau}$, where $m$ and $n$ are natural numbers.%TSM utilize these to splice types and unexpanded expressions out of literal bodies.
+
+% \subsubsection{Proto-Type Validation}\label{sec:SE-proto-type-validation}
+% The \emph{proto-type validation judgement}, $\cvalidT{\Delta}{\tscenev}{\ctau}{\tau}$, is inductively defined by Rules (\ref{rules:cvalidT-U}).
+
+% \paragraph{Common Forms} Rules (\ref{rule:cvalidT-U-tvar}) through (\ref{rule:cvalidT-U-sum}) validate proto-types of common form. The first three of these are reproduced below.
+% %Each of these rules is defined based on the corresponding type formation rule, i.e.
+Rules (\ref{rule:istypeU-var}) through (\ref{rule:istypeU-sum}), respectively.
+For example, the following proto-types validation rules are based on type formation rules (\ref{rule:istypeU-var}), (\ref{rule:istypeU-parr}) and (\ref{rule:istypeU-all}), respectively:
+% % \begin{subequations}%\label{rules:cvalidT-U}
+% \begin{equation*}\tag{\ref{rule:cvalidT-U-tvar}}
+% \inferrule{ }{
+%   \cvalidT{\Delta, \Dhyp{t}}{\tscenev}{t}{t}
+% }
+% \end{equation*}
+% \begin{equation*}\tag{\ref{rule:cvalidT-U-parr}}
+%   \inferrule{
+%     \cvalidT{\Delta}{\tscenev}{\ctau_1}{\tau_1}\\
+%     \cvalidT{\Delta}{\tscenev}{\ctau_2}{\tau_2}
+%   }{
+%     \cvalidT{\Delta}{\tscenev}{\aceparr{\ctau_1}{\ctau_2}}{\aparr{\tau_1}{\tau_2}}
+%   }
+% \end{equation*}
+% \begin{equation*}\tag{\ref{rule:cvalidT-U-all}}
+%   \inferrule {
+%     \cvalidT{\Delta, \Dhyp{t}}{\tscenev}{\ctau}{\tau}
+%   }{
+%     \cvalidT{\Delta}{\tscenev}{\aceall{t}{\ctau}}{\aall{t}{\tau}}
+%   }
+% \end{equation*}
+% % \begin{equation*}\label{rule:cvalidT-U-rec}
+% %   \inferrule{
+% %     \cvalidT{\Delta, \Dhyp{t}}{\tscenev}{\ctau}{\tau}
+% %   }{
+% %     \cvalidT{\Delta}{\tscenev}{\acerec{t}{\ctau}}{\arec{t}{\tau}}
+% %   }
+% % \end{equation*}
+% % \begin{equation*}\label{rule:cvalidT-U-prod}
+% %   \inferrule{
+% %     \{\cvalidT{\Delta}{\tscenev}{\ctau_i}{\tau_i}\}_{i \in \labelset}
+% %   }{
+% %     \cvalidT{\Delta}{\tscenev}{\aceprod{\labelset}{\mapschema{\ctau}{i}{\labelset}}}{\aprod{\labelset}{\mapschema{\tau}{i}{\labelset}}}
+% %   }
+% % \end{equation*}
+% % \begin{equation*}\label{rule:cvalidT-U-sum}
+% %   \inferrule{
+% %     \{\cvalidT{\Delta}{\tscenev}{\ctau_i}{\tau_i}\}_{i \in \labelset}
+% %   }{
+% %     \cvalidT{\Delta}{\tscenev}{\acesum{\labelset}{\mapschema{\ctau}{i}{\labelset}}}{\asum{\labelset}{\mapschema{\tau}{i}{\labelset}}}
+% %   }
+% % \end{equation*}
+
+% These rules, like the rules for common unexpanded type forms,  mirror the corresponding type formation rules, i.e.
+Rules (\ref{rules:istypeU}).
+The type splicing scene, $\tscenev$, passes opaquely through these rules.
+% % We can express this scheme more precisely with the following rule transformation. For each rule in Rules (\ref{rules:istypeU}), 
+% % \begin{mathpar}
+% % % \refstepcounter{equation}
+% % % \label{rule:cvalidT-U-rec}
+% % % \refstepcounter{equation}
+% % % \label{rule:cvalidT-U-prod}
+% % % \refstepcounter{equation}
+% % % \label{rule:cvalidT-U-sum}
+% % % \inferrule{J_1\\\cdots\\J_k}{J}
+% % \end{mathpar}
+% % the corresponding proto-types validation rule is
+% % \begin{mathpar}
+% % \inferrule{
+% %   \VTypof{J_1}\\
+% %   \cdots\\
+% %   \VTypof{J_k}
+% % }{
+% %   \VTypof{J}
+% % }
+% % \end{mathpar}
+% % where 
+% % \[\begin{split}
+% % \VTypof{\istypeU{\Delta}{\tau}} & = \cvalidT{\Delta}{\tscenev}{\VTypof{\tau}}{\tau}\\
+% % \VTypof{\{J_i\}_{i \in \labelset}} & = \{\VTypof{J_i}\}_{i \in \labelset}
+% % \end{split}\]
+% % and where $\VTypof{\tau}$, when $\tau$ is a metapattern of sort $\mathsf{Typ}$, is a metapattern of sort $\mathsf{CETyp}$ defined as follows:
+% % \begin{itemize}
+% % \item When $\tau$ is of definite form, $\VTypof{\tau}$ is defined as follows:
+% % \begin{align*}
+% % \VTypof{t} & = t\\
+% % \VTypof{\aparr{\tau_1}{\tau_2}} & = \aceparr{\VTypof{\tau_1}}{\VTypof{\tau_2}}\\
+% % \VTypof{\aall{t}{\tau}} & = \aceall{t}{\VTypof{\tau}}\\
+% % \VTypof{\arec{t}{\tau}} & = \acerec{t}{\VTypof{\tau}}\\
+% % \VTypof{\aprod{\labelset}{\mapschema{\tau}{i}{\labelset}}} & = \aceprod{\labelset}{\mapschemax{\VTypofv}{\tau}{i}{\labelset}}\\
+% % \VTypof{\asum{\labelset}{\mapschema{\tau}{i}{\labelset}}} & = \acesum{\labelset}{\mapschemax{\VTypofv}{\tau}{i}{\labelset}}
+% % \end{align*}
+% % \item When $\tau$ is of indefinite form, $\VTypof{\tau}$ is a uniquely corresponding metapattern also of indefinite form. For example, $\VTypof{\tau_1}=\ctau_1$ and $\VTypof{\tau_2}=\ctau_2$.
+% % \end{itemize}
+
+% % It is instructive to use this rule transformation to generate Rules (\ref{rule:cvalidT-U-tvar}) through (\ref{rule:cvalidT-U-all}) above.
+We omit the remaining rules, i.e.
+Rules (\ref*{rule:cvalidT-U-rec}) through (\ref*{rule:cvalidT-U-sum}).
+
+% Notice that in Rule (\ref{rule:cvalidT-U-tvar}), only type variables tracked by $\Delta$, the expansion's local type validation context, are well-formed.
+Type variables tracked by the application site unexpanded type formation context, which is a component of the type splicing scene, $\tscenev$, are not validated.
+%Indeed, $\tscenev$ passes opaquely through the rules above.
+%This achieves \emph{context-independent expansion} as described in Sec.
+\ref{sec:splicing-and-hygiene} for type variables -- seTSMs cannot impose ``hidden constraints'' on the application site unexpanded type formation context, because the type variables bound at the application site are simply not directly available to proto-types.
+
+% \paragraph{References to Spliced Types} The only proto-type form that does not correspond to a type form is $\acesplicedt{m}{n}$, which is a \emph{reference to a spliced unexpanded type}, i.e.
+it indicates that an unexpanded type should be parsed out from the literal body, which appears in the type splicing scene $\tscenev$, beginning at position $m$ and ending at position $n$, where $m$ and $n$ are natural numbers.
+Rule (\ref{rule:cvalidT-U-splicedt}) governs this form:
+% \begin{equation*}\tag{\ref{rule:cvalidT-U-splicedt}}
+%   \inferrule{
+%     \parseUTyp{\bsubseq{b}{m}{n}}{\utau}\\
+%     \expandsTU{\uDD{\uD}{\Delta_\text{app}}}{\utau}{\tau}\\
+%     \Delta \cap \Delta_\text{app} = \emptyset
+%   }{
+%     \cvalidT{\Delta}{\tsceneU{\uDD{\uD}{\Delta_\text{app}}}{b}}{\acesplicedt{m}{n}}{\tau}
+%   }
+% \end{equation*}
+% The first premise of this rule extracts the indicated subsequence of $b$ using the partial metafunction $\bsubseq{b}{m}{n}$ and parses it using the partial metafunction $\mathsf{parseUTyp}(b)$, characterized in Sec.
+\ref{sec:syntax-U}, to produce the spliced unexpanded type itself, $\utau$.
+
+% The second premise of Rule (\ref{rule:cvalidT-U-splicedt}) performs type expansion of $\utau$ under the application site unexpanded type formation context, $\uDD{\uD}{\Delta_\text{app}}$, which is a component of the type splicing scene.
+The hypotheses in the expansion's local type formation context, $\Delta$, are not made available to $\tau$.
+%This enforces the injunction on shadowing as described in Sec.
+\ref{sec:splicing-and-hygiene} for type variables that appear in proto-types.
+
+% The third premise of Rule (\ref{rule:cvalidT-U-splicedt}) imposes the constraint that the proto-expansion's type formation context, $\Delta$, be disjoint from the application site type formation context, $\Delta_\text{app}$.
+This premise can always be discharged by $\alpha$-varying the proto-expansion that the reference to the spliced type appears within.
+
+% Together, these two premises enforce the injunction on type variable capture as described in Sec.
+\ref{sec:uetsms-validation} -- the TSM provider can choose type variable names freely within a proto-expansion.
+We will consider this formally in Sec.
+\ref{sec:SE-metatheory} below.
+%, because the language prevents them from shadowing type variables at the application site (by $\alpha$-varying the proto-expansion as needed.)%Such a change in bound variable names is possible again because variables bound by the seTSM provider in a proto-expansion cannot ``leak into'' spliced terms because the hypotheses in $\Delta$ are not made available to the spliced type, $\tau$.
+
+% Rules (\ref{rules:cvalidT-U}) validate the following lemma, which establishes that the final expansion of a valid proto-type is a well-formed type under the combined type formation context.
+% \begingroup
+% \def\thetheorem{\ref{lemma:candidate-expansion-type-validation}}
+% \begin{lemma}[Proto-Expansion Type Validation]
+% If $\cvalidT{\Delta}{\tsceneU{\uDD{\uD}{\Delta_\text{app}}}{b}}{\ctau}{\tau}$ and $\Delta \cap \Delta_\text{app}=\emptyset$ then $\istypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\tau}$.
+% \end{lemma}
+% \endgroup
+
+% \subsubsection{Proto-Expression Validation}
+% The \emph{proto-expression validation judgement}, $\cvalidE{\Delta}{\Gamma}{\escenev}{\ce}{e}{\tau}$, is defined mutually inductively with the typed expansion judgement by Rules (\ref{rules:cvalidE-U}) as follows.% This is necessary because a typed expansion judgement appears as a premise in Rule (\ref{rule:cvalidE-U-splicede}) below, and a proto-expression validation judgement appears as a premise in Rule (\ref{rule:expandsU-tsmap}) above.
+
+% \paragraph{Common Forms} Rules (\ref{rule:cvalidE-U-var}) through (\ref{rule:cvalidE-U-case}) validate proto-expressions of common form, as well as ascriptions and let binding.
+The first five of these rules are reproduced below:
+% %For each expanded expression form defined in Figure \ref{fig:U-expanded-terms}, Figure \ref{fig:U-candidate-terms} defines a corresponding proto-expression form.
+The validation rules for proto-expressions of these forms are each based on the corresponding typing rule in Rules (\ref{rules:hastypeU}).
+For example, the validation rules for proto-expressions of variable, function and function application form  are based on Rules (\ref{rule:hastypeU-var}) through (\ref{rule:hastypeU-ap}), respectively:
+% %\begin{subequations}%\label{rules:cvalidE-U}
+% \begin{equation*}\tag{\ref{rule:cvalidE-U-var}}
+% \inferrule{ }{
+%   \cvalidE{\Delta}{\Gamma, \Ghyp{x}{\tau}}{\escenev}{x}{x}{\tau}
+% }
+% \end{equation*}
+% \begin{equation*}\tag{\ref{rule:cvalidE-U-asc}}
+% \inferrule{
+%   \cvalidT{\Delta}{\tsfrom{\escenev}}{\ctau}{\tau}\\
+%   \cvalidE{\Delta}{\Gamma}{\escenev}{\ce}{e}{\tau}
+% }{
+%   \cvalidE{\Delta}{\Gamma}{\escenev}{\aceasc{\ctau}{\ce}}{e}{\tau}
+% }
+% \end{equation*}
+% \begin{equation*}\tag{\ref{rule:cvalidE-U-letsyn}}
+%   \inferrule{
+%     \cvalidE{\Delta}{\Gamma}{\escenev}{\ce_1}{e_1}{\tau_1}\\
+%     \cvalidE{\Delta}{\Gamma, x : \tau_1}{\ce_2}{e_2}{\tau_2}
+%   }{
+%     \cvalidE{\Delta}{\Gamma}{\escenev}{\aceletsyn{x}{\ce_1}{\ce_2}}{
+%       \aeap{\aelam{\tau_1}{x}{e_2}}{e_1}
+%     }{\tau_2}
+%   }
+% \end{equation*}
+% \begin{equation*}\tag{\ref{rule:cvalidE-U-lam}}
+% \inferrule{
+%   \cvalidT{\Delta}{\tsfrom{\escenev}}{\ctau}{\tau}\\
+%   \cvalidE{\Delta}{\Gamma, \Ghyp{x}{\tau}}{\escenev}{\ce}{e}{\tau'}
+% }{
+%   \cvalidE{\Delta}{\Gamma}{\escenev}{\acelam{\ctau}{x}{\ce}}{\aelam{\tau}{x}{e}}{\aparr{\tau}{\tau'}}
+% }
+% \end{equation*}
+% \begin{equation*}\tag{\ref{rule:cvalidE-U-ap}}
+%   \inferrule{
+%     \cvalidE{\Delta}{\Gamma}{\escenev}{\ce_1}{e_1}{\aparr{\tau}{\tau'}}\\
+%     \cvalidE{\Delta}{\Gamma}{\escenev}{\ce_2}{e_2}{\tau}
+%   }{
+%     \cvalidE{\Delta}{\Gamma}{\escenev}{\aceap{\ce_1}{\ce_2}}{\aeap{e_1}{e_2}}{\tau'}
+%   }
+% \end{equation*}
+% Once again, the rules for common forms mirror the typing rules, i.e. Rules (\ref{rules:hastypeU}). The expression splicing scene, $\escenev$, passes opaquely through these rules.
+
+% Notice that in Rule (\ref{rule:cvalidE-U-var}), only variables tracked by the proto-expansion typing context, $\Gamma$, are validated.
+Variables  in the application site unexpanded typing context, which appears within the expression splicing scene $\escenev$, are not validated.
+This achieves \emph{context independence} as described in Sec.
+\ref{sec:uetsms-validation} -- seTSMs cannot impose ``hidden constraints'' on the application site unexpanded typing context, because the variable bindings at the application site are not directly available to proto-expansions.
+We will consider this formally in Sec.
+\ref{sec:SE-metatheory} below.
+
+% \paragraph{References to Spliced Unexpanded Expressions} The only proto-expression form that does not correspond to an expanded expression form is $\acesplicede{m}{n}{\ctau}$, which is a \emph{reference to a spliced unexpanded expression}, i.e.
+it indicates that an unexpanded expression should be parsed out from the literal body beginning at position $m$ and ending at position $n$.
+Rule (\ref{rule:cvalidE-U-splicede}) governs this form:
+% \begin{equation*}\tag{\ref{rule:cvalidE-U-splicede}}
+% \inferrule{
+%   \cvalidT{\emptyset}{\tsfrom{\escenev}}{\ctau}{\tau}\\
+%   \escenev=\esceneU{\uDD{\uD}{\Delta_\text{app}}}{\uGG{\uG}{\Gamma_\text{app}}}{\uPsi}{b}\\
+%   \parseUExp{\bsubseq{b}{m}{n}}{\ue}\\
+%   \expandsU{\uDD{\uD}{\Delta_\text{app}}}{\uGG{\uG}{\Gamma_\text{app}}}{\uPsi}{\ue}{e}{\tau}\\\\
+%   \Delta \cap \Delta_\text{app} = \emptyset\\
+%   \domof{\Gamma} \cap \domof{\Gamma_\text{app}} = \emptyset
+% }{
+%   \cvalidE{\Delta}{\Gamma}{\escenev}{\acesplicede{m}{n}{\ctau}}{e}{\tau}
+% }
+% \end{equation*}
+% % \begin{equation*}\label{rule:cvalidE-U-splicede}
+% % \inferrule{
+% %   \parseUExp{\bsubseq{b}{m}{n}}{\ue}\\\\
+% %   \expandsU{\Delta_\text{app}}{\Gamma_\text{app}}{\uPsi}{\ue}{e}{\tau}\\
+% %   \Delta \cap \Delta_\text{app} = \emptyset\\
+% %   \domof{\Gamma} \cap \domof{\Gamma_\text{app}} = \emptyset
+% % }{
+% %   \cvalidE{\Delta}{\Gamma}{\esceneU{\Delta_\text{app}}{\Gamma_\text{app}}{\uPsi}{b}}{\splicede{m}{n}}{e}{\tau}
+% % }
+% % \end{equation*}
+
+% The first premise of this rule validates and expands the type annotation. This type must be context independent.
+
+% The second premise of this rule serves simply to reveal the components of the expression splicing scene.
+
+% The third premise of this rule extracts the indicated subsequence of $b$ using the partial metafunction $\bsubseq{b}{m}{n}$ and parses it using the partial metafunction $\mathsf{parseUExp}(b)$, characterized in Sec.
+\ref{sec:syntax-U}, to produce the referenced spliced unexpanded expression, $\ue$.
+
+% The fourth premise of Rule (\ref{rule:cvalidE-U-splicede}) performs typed expansion of $\ue$ assuming the application site contexts that appear in the expression splicing scene.
+Notice that the hypotheses in $\Delta$ and $\Gamma$ are not made available to $\ue$.
+
+% The fifth premise of Rule (\ref{rule:cvalidE-U-splicede}) imposes the constraint that the proto-expansion's type formation context, $\Delta$, be disjoint from the application site type formation context, $\Delta_\text{app}$.
+Similarly, the sixth premise requires that the proto-expansion's typing context, $\Gamma$, be disjoint from the application site typing context, $\Gamma_\text{app}$.
+These two premises can always be discharged by $\alpha$-varying the proto-expression that the reference to the spliced unexpanded expression appears within.
+% Together, these premises enforce the prohibition on capture as described in Sec.
+\ref{sec:uetsms-validation} -- the TSM provider can choose variable names freely within a proto-expansion, because the language prevents them from shadowing those at the application site.
+% %\end{subequations}
+% % \begin{subequations}\label{rules:cvalidE-U}
+% % \begin{equation*}\label{rule:cvalidE-U-var}
+% % \inferrule{ }{
+% %   \cvalidE{\Delta}{\Gamma, \Ghyp{x}{\tau}}{\escenev}{x}{x}{\tau}
+% % }
+% % \end{equation*}
+% % \begin{equation*}\label{rule:cvalidE-U-lam}
+% % \inferrule{
+% %   \cvalidT{\Delta}{\tsfrom{\escenev}}{\ctau}{\tau}\\
+% %   \cvalidE{\Delta}{\Gamma, \Ghyp{x}{\tau}}{\escenev}{\ce}{e}{\tau'}
+% % }{
+% %   \cvalidE{\Delta}{\Gamma}{\escenev}{\acelam{\ctau}{x}{\ce}}{\aelam{\tau}{x}{e}}{\aparr{\tau}{\tau'}}
+% % }
+% % \end{equation*}
+% % \begin{equation*}\label{rule:cvalidE-U-ap}
+% %   \inferrule{
+% %     \cvalidE{\Delta}{\Gamma}{\escenev}{\ce_1}{e_1}{\aparr{\tau}{\tau'}}\\
+% %     \cvalidE{\Delta}{\Gamma}{\escenev}{\ce_2}{e_2}{\tau}
+% %   }{
+% %     \cvalidE{\Delta}{\Gamma}{\escenev}{\aceap{\ce_1}{\ce_2}}{\aeap{e_1}{e_2}}{\tau'}
+% %   }
+% % \end{equation*}
+% % \begin{equation*}\label{rule:cvalidE-U-tlam}
+% %   \inferrule{
+% %     \cvalidE{\Delta, \Dhyp{t}}{\Gamma}{\escenev}{\ce}{e}{\tau}
+% %   }{
+% %     \cvalidEX{\acetlam{t}{\ce}}{\aetlam{t}{e}}{\aall{t}{\tau}}
+% %   }
+% % \end{equation*}
+% % \begin{equation*}\label{rule:cvalidE-U-tap}
+% %   \inferrule{
+% %     \cvalidEX{\ce}{e}{\aall{t}{\tau}}\\
+% %     \cvalidT{\Delta}{\tsfrom{\escenev}}{\ctau'}{\tau'}
+% %   }{
+% %     \cvalidEX{\acetap{\ce}{\ctau'}}{\aetap{e}{\tau'}}{[\tau'/t]\tau}
+% %   }
+% % \end{equation*}
+% % \begin{equation*}\label{rule:cvalidE-U-fold}
+% %   \inferrule{
+% %     \cvalidT{\Delta, \Dhyp{t}}{\escenev}{\ctau}{\tau}\\
+% %     \cvalidEX{\ce}{e}{[\arec{t}{\tau}/t]\tau}
+% %   }{
+% %     \cvalidEX{\acefold{t}{\ctau}{\ce}}{\aefold{e}}{\arec{t}{\tau}}
+% %   }
+% % \end{equation*}
+% % \begin{equation*}\label{rule:cvalidE-U-unfold}
+% %   \inferrule{
+% %     \cvalidEX{\ce}{e}{\arec{t}{\tau}}
+% %   }{
+% %     \cvalidEX{\aceunfold{\ce}}{\aeunfold{e}}{[\arec{t}{\tau}/t]\tau}
+% %   }
+% % \end{equation*}
+% % \begin{equation*}\label{rule:cvalidE-U-tpl}
+% %   \inferrule{
+% %     \{\cvalidEX{\ce_i}{e_i}{\tau_i}\}_{i \in \labelset}
+% %   }{
+% %     \cvalidEX{\acetpl{\labelset}{\mapschema{\ce}{i}{\labelset}}}{\aetpl{\labelset}{\mapschema{e}{i}{\labelset}}}{\aprod{\labelset}{\mapschema{\tau}{i}{\labelset}}}
+% %   }
+% % \end{equation*}
+% % \begin{equation*}\label{rule:cvalidE-U-pr}
+% %   \inferrule{
+% %     \cvalidEX{\ce}{e}{\aprod{\labelset, \ell}{\mapschema{\tau}{i}{\labelset}; \mapitem{\ell}{\tau}}}
+% %   }{
+% %     \cvalidEX{\acepr{\ell}{\ce}}{\aepr{\ell}{e}}{\tau}
+% %   }
+% % \end{equation*}
+% % \begin{equation*}\label{rule:cvalidE-U-in}
+% %   \inferrule{
+% %     \{\cvalidT{\Delta}{\tsfrom{\escenev}}{\ctau_i}{\tau_i}\}_{i \in \labelset}\\
+% %     \cvalidT{\Delta}{\tsfrom{\escenev}}{\ctau}{\tau}\\
+% %     \cvalidEX{\ce}{e}{\tau}
+% %   }{
+% %     \left\{\shortstack{$\Delta~\Gamma \vdash_\uPsi \acein{\labelset, \ell}{\ell}{\mapschema{\ctau}{i}{\labelset}; \mapitem{\ell}{\ctau}}{\ce}$\\$\leadsto$\\$\aein{\labelset, \ell}{\ell}{\mapschema{\tau}{i}{\labelset}; \mapitem{\ell}{\tau}}{e} : \asum{\labelset, \ell}{\mapschema{\tau}{i}{\labelset}; \mapitem{\ell}{\tau}}$\vspace{-1.2em}}\right\}
+% %   }
+% % \end{equation*}
+% % \begin{equation*}\label{rule:cvalidE-U-case}
+% %   \inferrule{
+% %     \cvalidEX{\ce}{e}{\asum{\labelset}{\mapschema{\tau}{i}{\labelset}}}\\
+% %     \{\cvalidE{\Delta}{\Gamma, \Ghyp{x_i}{\tau_i}}{\escenev}{\ue_i}{e_i}{\tau}\}_{i \in \labelset}
+% %   }{
+% %     \cvalidEX{\acecase{\labelset}{\ce}{\mapschemab{x}{\ce}{i}{\labelset}}}{\aecase{\labelset}{e}{\mapschemab{x}{e}{i}{\labelset}}}{\tau}
+% %   }
+% % \end{equation*}
+% % \begin{equation*}\label{rule:cvalidE-U-splicede}
+% % \inferrule{
+% %   \parseUExp{\bsubseq{b}{m}{n}}{\ue}\\\\
+% %   \Delta \cap \Delta_\text{app} = \emptyset\\
+% %   \domof{\Gamma} \cap \domof{\Gamma_\text{app}} = \emptyset\\
+% %   \expandsU{\Delta_\text{app}}{\Gamma_\text{app}}{\uPsi}{\ue}{e}{\tau}
+% % }{
+% %   \cvalidE{\Delta}{\Gamma}{\esceneU{\Delta_\text{app}}{\Gamma_\text{app}}{\uPsi}{b}}{\acesplicede{m}{n}}{e}{\tau}
+% % }
+% % \end{equation*}
+% % \end{subequations}
+
+% % Each form of expanded expression, $e$, corresponds to a form of proto-expression, $\ce$ (compare Figure \ref{fig:U-expanded-terms} and Figure \ref{fig:U-candidate-terms}).
+For each typing rule in Rules \ref{rules:hastypeU}, there is a corresponding proto-expression validation rule -- Rules (\ref{rule:cvalidE-U-var}) to (\ref{rule:cvalidE-U-case}) -- where the proto-expression and expanded expression correspond.
+The premises also correspond.
+
+% %Candidate expansions cannot themselves define or apply TSMs.
+This simplifies our metatheory, though it can be inconvenient at times for TSM providers.
+We discuss adding the ability to use TSMs within proto-expansions in Sec.
+\ref{sec:tsms-in-expansions}.
+
+% \subsection{Metatheory}\label{sec:SE-metatheory}
+% \subsubsection{Typed Expansion}
+% Let us now consider Theorem \ref{thm:typed-expansion-short-U}, which was mentioned at the beginnning of Sec. \ref{sec:typed-expansion-U} and is reproduced below:
+% \begingroup
+% \def\thetheorem{\ref{thm:typed-expansion-short-U}}
+% \begin{theorem}[Typed Expression Expansion] \hspace{-3px}If $\expandsU{\uDD{\uD}{\Delta}\hspace{-3px}}{\uGG{\uG}{\Gamma}\hspace{-3px}}{\uPsi}{\ue}{e}{\tau}$ then $\hastypeU{\Delta}{\Gamma}{e}{\tau}$.
+% \end{theorem}
+% \endgroup
+
+%  To prove this theorem, we must  prove the following stronger theorem, because the proto-expression validation judgement is defined mutually inductively with the typed expansion judgement:
+
+% \begingroup
+% \def\thetheorem{\ref{thm:typed-expansion-full-U}}
+% \begin{theorem}[Typed Expansion (Full)] ~
+% \begin{enumerate}
+% \item If $\expandsU{\uDD{\uD}{\Delta}}{\uGG{\uG}{\Gamma}}{\uAS{\uA}{\Psi}}{\ue}{e}{\tau}$ then $\hastypeU{\Delta}{\Gamma}{e}{\tau}$.
+% \item If $\cvalidE{\Delta}{\Gamma}{\esceneU{\uDD{\uD}{\Delta_\text{app}}}{\uGG{\uG}{\Gamma_\text{app}}}{\uAS{\uA}{\Psi}}{b}}{\ce}{e}{\tau}$ and $\Delta \cap \Delta_\text{app} = \emptyset$ and $\domof{\Gamma} \cap \domof{\Gamma_\text{app}} = \emptyset$ then $\hastypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\Gcons{\Gamma}{\Gamma_\text{app}}}{e}{\tau}$.
+% \end{enumerate}
+% \end{theorem}
+% \endgroup
+% \begin{proof}
+% By mutual rule induction over Rules (\ref{rules:expandsU}) and Rules (\ref{rules:cvalidE-U}).
+The full proof is given in Appendix \ref{appendix:SES-typed-expression-expansion-metatheory}.
+We will reproduce the interesting cases below.
+
+% The proof of part 1 proceeds by inducting over the typed expansion assumption.
+The only interesting cases are those related to seTSM definition and application, reproduced below.
+In the following cases, let $\uDelta=\uDD{\uD}{\Delta}$ and $\uGamma=\uGG{\uG}{\Gamma}$ and $\uPsi=\uAS{\uA}{\Psi}$.
+
+% \begin{byCases}
+% \item[\text{(\ref{rule:expandsU-syntax})}] We have 
+% \begin{pfsteps}
+%   \item \ue=\uesyntax{\tsmv}{\utau'}{\eparse}{\ue'} \BY{assumption}
+%   \item \expandsTU{\uDelta}{\utau'}{\tau'} \BY{assumption} \pflabel{expandsTU}
+%  \item \hastypeU{\emptyset}{\emptyset}{\eparse}{\aparr{\tBody}{\tParseResultExp}} \BY{assumption}\pflabel{eparse}
+%   \item \expandsU{\uDelta}{\uGamma}{\uPsi, \uShyp{\tsmv}{a}{\tau'}{\eparse}}{\ue'}{e}{\tau} \BY{assumption}\pflabel{expandsU}
+% %  \item \uetsmenv{\Delta}{\Psi} \BY{assumption}\pflabel{uetsmenv1}
+%  \item \istypeU{\Delta}{\tau'} \BY{Lemma \ref{lemma:type-expansion-U} to \pfref{expandsTU}} \pflabel{istype}
+% %  \item \uetsmenv{\Delta}{\Psi, \xuetsmbnd{\tsmv}{\tau'}{\eparse}} \BY{Definition \ref{def:seTSM-def-ctx-formation} on \pfref{uetsmenv1}, \pfref{istype} and \pfref{eparse}}\pflabel{uetsmenv3}
+%   \item \hastypeU{\Delta}{\Gamma}{e}{\tau} \BY{IH, part 1(a) on \pfref{expandsU}}
+% \end{pfsteps}
+% \resetpfcounter
+
+% \item[\text{(\ref{rule:expandsU-tsmap})}] We have 
+% \begin{pfsteps}
+%   \item \ue=\utsmap{\tsmv}{b} \BY{assumption}
+%   \item \uA = \uA', \vExpands{\tsmv}{a} \BY{assumption}
+%   \item \Psi=\Psi', \xuetsmbnd{a}{\tau}{\eparse} \BY{assumption}
+%   \item \encodeBody{b}{\ebody} \BY{assumption}
+%   \item \evalU{\eparse(\ebody)}{\aein{\lbltxt{SuccessE}}{\ecand}} \BY{assumption}
+%   \item \decodeCondE{\ecand}{\ce} \BY{assumption}
+%   \item \cvalidE{\emptyset}{\emptyset}{\esceneU{\uDelta}{\uGamma}{\uPsi}{b}}{\ce}{e}{\tau} \BY{assumption}\pflabel{cvalidE}
+% %  \item \uetsmenv{\Delta}{\Psi} \BY{assumption} \pflabel{uetsmenv}
+%   \item \emptyset \cap \Delta = \emptyset \BY{finite set intersection} \pflabel{delta-cap}
+%   \item {\emptyset} \cap \domof{\Gamma} = \emptyset \BY{finite set intersection} \pflabel{gamma-cap}
+%   \item \hastypeU{\emptyset \cup \Delta}{\emptyset \cup \Gamma}{e}{\tau} \BY{IH, part 2 on \pfref{cvalidE}, \pfref{delta-cap}, and \pfref{gamma-cap}} \pflabel{penultimate}
+%   \item \hastypeU{\Delta}{\Gamma}{e}{\tau} \BY{finite set and finite function identity over \pfref{penultimate}}
+% \end{pfsteps}
+% \resetpfcounter
+% \end{byCases}
+
+% The proof of part 2 proceeds by induction over the proto-expression validation assumption.
+The only interesting case governs references to spliced expressions.
+In the following cases, let $\uDelta_\text{app}=\uDD{\uD}{\Delta_\text{app}}$ and $\uGamma_\text{app}=\uGG{\uG}{\Gamma_\text{app}}$ and $\uPsi = \uAS{\uA}{\Psi}$.
+% \begin{byCases}
+% % \item[\text{(\ref{rule:cvalidE-U-var})}] ~
+% % \begin{pfsteps*}
+% %   \item $\ce=x$ \BY{assumption}
+% %   \item $e=x$ \BY{assumption}
+% %   \item $\Gamma=\Gamma', \Ghyp{x}{\tau}$ \BY{assumption}
+% %   \item $\hastypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\Gamma', \Ghyp{x}{\tau}}{x}{\tau}$ \BY{Rule (\ref{rule:hastypeU-var})} \pflabel{hastypeU}
+% %   \item $\hastypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\Gcons{\Gamma', \Ghyp{x}{\tau}}{\Gamma_\text{app}}}{x}{\tau}$ \BY{Lemma \ref{lemma:weakening-U} over $\Gamma_\text{app}$ to \pfref{hastypeU}}
+% % \end{pfsteps*}
+% % \resetpfcounter
+
+% % \item[\text{(\ref{rule:cvalidE-U-lam})}] ~
+% % \begin{pfsteps*}
+% %   \item $\ce=\acelam{\ctau_1}{x}{\ce'}$ \BY{assumption}
+% %   \item $e=\aelam{\tau_1}{x}{e'}$ \BY{assumption}
+% %   \item $\tau=\aparr{\tau_1}{\tau_2}$ \BY{assumption}
+% %   \item $\cvalidT{\Delta}{\tsceneU{\uDelta_\text{app}}{b}}{\ctau_1}{\tau_1}$ \BY{assumption} \pflabel{cvalidT}
+% %   \item $\cvalidE{\Delta}{\Gamma, \Ghyp{x}{\tau_1}}{\esceneU{\uDelta_\text{app}}{\uGamma_\text{app}}{\uPsi}{b}}{\ce'}{e'}{\tau_2}$ \BY{assumption} \pflabel{cvalidE}
+% % %  \item $\uetsmenv{\Delta_\text{app}}{\Psi}$ \BY{assumption} \pflabel{uetsmenv}
+% %   \item $\Delta \cap \Delta_\text{app}=\emptyset$ \BY{assumption} \pflabel{delta-disjoint}
+% %   \item $\domof{\Gamma} \cap \domof{\Gamma_\text{app}}=\emptyset$ \BY{assumption} \pflabel{gamma-disjoint}
+% %   \item $x \notin \domof{\Gamma_\text{app}}$ \BY{identification convention} \pflabel{x-fresh}
+% %   \item $\domof{\Gamma, x : \tau_1} \cap \domof{\Gamma_\text{app}}=\emptyset$ \BY{\pfref{gamma-disjoint} and \pfref{x-fresh}} \pflabel{gamma-disjoint2}
+% %   \item $\istypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\tau_1}$ \BY{Lemma \ref{lemma:candidate-expansion-type-validation} on \pfref{cvalidT} and \pfref{delta-disjoint}} \pflabel{istype}
+% %   \item $\hastypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\Gcons{\Gamma, \Ghyp{x}{\tau_1}}{\Gamma_\text{app}}}{e'}{\tau_2}$ \BY{IH, part 2 on \pfref{cvalidE}, \pfref{delta-disjoint} and \pfref{gamma-disjoint2}} \pflabel{hastype1}
+% %   \item $\hastypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\Gcons{\Gamma}{\Gamma_\text{app}}, \Ghyp{x}{\tau_1}}{e'}{\tau_2}$ \BY{exchange over $\Gamma_\text{app}$ on \pfref{hastype1}} \pflabel{hastype2}
+% %   \item $\hastypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\Gcons{\Gamma}{\Gamma_\text{app}}}{\aelam{\tau_1}{x}{e'}}{\aparr{\tau_1}{\tau_2}}$ \BY{Rule (\ref{rule:hastypeU-lam}) on \pfref{istype} and \pfref{hastype2}}
+% % \end{pfsteps*}
+% % \resetpfcounter
+
+% % \item[\text{(\ref{rule:cvalidE-U-ap})}] ~
+% % \begin{pfsteps*}
+% %   \item $\ce=\aceap{\ce_1}{\ce_2}$ \BY{assumption}
+% %   \item $e=\aeap{e_1}{e_2}$ \BY{assumption}
+% %   \item $\cvalidE{\Delta}{\Gamma}{\esceneU{\uDelta_\text{app}}{\uGamma_\text{app}}{\uPsi}{b}}{\ce_1}{e_1}{\aparr{\tau_2}{\tau}}$ \BY{assumption} \pflabel{cvalidE1}
+% %   \item $\cvalidE{\Delta}{\Gamma}{\esceneU{\uDelta_\text{app}}{\uGamma_\text{app}}{\uPsi}{b}}{\ce_2}{e_2}{\tau_2}$ \BY{assumption} \pflabel{cvalidE2}
+% % %  \item $\uetsmenv{\Delta_\text{app}}{\Psi}$ \BY{assumption} \pflabel{uetsmenv}
+% %   \item $\Delta \cap \Delta_\text{app}=\emptyset$ \BY{assumption} \pflabel{delta-disjoint}
+% %   \item $\domof{\Gamma} \cap \domof{\Gamma_\text{app}}=\emptyset$ \BY{assumption} \pflabel{gamma-disjoint}
+% %   \item $\hastypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\Gcons{\Gamma}{\Gamma_\text{app}}}{e_1}{\aparr{\tau_2}{\tau}}$ \BY{IH, part 2 on \pfref{cvalidE1}, \pfref{delta-disjoint} and \pfref{gamma-disjoint}} \pflabel{hastypeU1}
+% %   \item $\hastypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\Gcons{\Gamma}{\Gamma_\text{app}}}{e_2}{\tau_2}$ \BY{IH, part 2 on \pfref{cvalidE2}, \pfref{delta-disjoint} and \pfref{gamma-disjoint}} \pflabel{hastypeU2}
+% %   \item $\hastypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\Gcons{\Gamma}{\Gamma_\text{app}}}{\aeap{e_1}{e_2}}{\tau}$ \BY{Rule (\ref{rule:hastypeU-ap}) on \pfref{hastypeU1} and \pfref{hastypeU2}}
+% % \end{pfsteps*}
+% % \resetpfcounter
+
+% % \item[\text{(\ref{rule:cvalidE-U-tlam})}] ~
+% % \begin{pfsteps}
+% %   \item \ce=\acetlam{t}{\ce'} \BY{assumption}
+% %   \item e = \aetlam{t}{e'} \BY{assumption}
+% %   \item \tau = \aall{t}{\tau'}\BY{assumption}
+% %   \item \cvalidE{\Delta, \Dhyp{t}}{\Gamma}{\esceneU{\uDelta_\text{app}}{\uGamma_\text{app}}{\uPsi}{b}}{\ce'}{e'}{\tau'} \BY{assumption} \pflabel{cvalidE}
+% % %  \item \uetsmenv{\Delta_\text{app}}{\Psi} \BY{assumption} \pflabel{uetsmenv}
+% %   \item \Delta \cap \Delta_\text{app}=\emptyset \BY{assumption} \pflabel{delta-disjoint}
+% %   \item \domof{\Gamma} \cap \domof{\Gamma_\text{app}}=\emptyset \BY{assumption} \pflabel{gamma-disjoint}
+% %   \item \Dhyp{t} \notin \Delta_\text{app} \BY{identification convention}\pflabel{t-fresh}
+% %   \item \Delta, \Dhyp{t} \cap \Delta_\text{app} = \emptyset \BY{\pfref{delta-disjoint} and \pfref{t-fresh}}\pflabel{delta-disjoint2}
+% %   \item \hastypeU{\Dcons{\Delta, \Dhyp{t}}{\Delta_\text{app}}}{\Gcons{\Gamma}{\Gamma_\text{app}}}{e'}{\tau'} \BY{IH, part 2 on \pfref{cvalidE}, \pfref{delta-disjoint2} and \pfref{gamma-disjoint}}\pflabel{hastype1}
+% %   \item \hastypeU{\Dcons{\Delta}{\Delta_\text{app}, \Dhyp{t}}}{\Gcons{\Gamma}{\Gamma_\text{app}}}{e'}{\tau'} \BY{exchange over $\Delta_\text{app}$ on \pfref{hastype1}}\pflabel{hastype2}
+% %   \item \hastypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\Gcons{\Gamma}{\Gamma_\text{app}}}{\aetlam{t}{e'}}{\aall{t}{\tau'}} \BY{Rule (\ref{rule:hastypeU-tlam}) on \pfref{hastype2}}
+% % \end{pfsteps}
+% % \resetpfcounter
+
+% % \item[{\text{(\ref{rule:cvalidE-U-tap})}}~\textbf{through}~{\text{(\ref{rule:cvalidE-U-case})}}] These cases follow analagously, i.e.
+we apply the IH, part 2 to all proto-expression validation judgements, Lemma \ref{lemma:candidate-expansion-type-validation} to all proto-type validation judgements, the identification convention to ensure that extended contexts remain disjoint, weakening and exchange as needed, and the corresponding typing rule in Rules (\ref{rule:hastypeU-tap}) through (\ref{rule:hastypeU-case}).
+% % \\
+
+% \item[\text{(\ref{rule:cvalidE-U-splicede})}] ~
+% \begin{pfsteps*}
+%   \item $\ce=\acesplicede{m}{n}{\ctau}$ \BY{assumption}
+%   \item $  \escenev=\esceneU{\uDD{\uD}{\Delta_\text{app}}}{\uGG{\uG}{\Gamma_\text{app}}}{\uPsi}{b}$ \BY{assumption}
+%   \item   $\cvalidT{\emptyset}{\tsfrom{\escenev}}{\ctau}{\tau}$ \BY{assumption}
+%   \item $\parseUExp{\bsubseq{b}{m}{n}}{\ue}$ \BY{assumption}
+%   \item $\expandsU{\uDelta_\text{app}}{\uGamma_\text{app}}{\uPsi}{\ue}{e}{\tau}$ \BY{assumption} \pflabel{expands}
+% %  \item $\uetsmenv{\Delta_\text{app}}{\Psi}$ \BY{assumption} \pflabel{uetsmenv}
+%   \item $\Delta \cap \Delta_\text{app}=\emptyset$ \BY{assumption} \pflabel{delta-disjoint}
+%   \item $\domof{\Gamma} \cap \domof{\Gamma_\text{app}}=\emptyset$ \BY{assumption} \pflabel{gamma-disjoint}
+%   \item $\hastypeU{\Delta_\text{app}}{\Gamma_\text{app}}{e}{\tau}$ \BY{IH, part 1 on \pfref{expands}} \pflabel{hastype}
+%   \item $\hastypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\Gcons{\Gamma}{\Gamma_\text{app}}}{e}{\tau}$ \BY{Lemma \ref{lemma:weakening-U} over $\Delta$ and $\Gamma$ and exchange on \pfref{hastype}}
+% \end{pfsteps*}
+% \resetpfcounter
+% \end{byCases}
+
+% The mutual induction can be shown to be well-founded by showing that the following numeric metric on the judgements that we induct over is decreasing:
+% \begin{align*}
+% \sizeof{\expandsU{\uDelta}{\uGamma}{\uPsi}{\ue}{e}{\tau}} & = \sizeof{\ue}\\
+% \sizeof{\cvalidE{\Delta}{\Gamma}{\esceneU{\uDelta_\text{app}}{\uGamma_\text{app}}{\uPsi}{b}}{\ce}{e}{\tau}} & = \sizeof{b}
+% \end{align*}
+% where $\sizeof{b}$ is the length of $b$ and $\sizeof{\ue}$ is the sum of the lengths of the literal bodies in $\ue$ (see Appendix \ref{appendix:SES-body-lengths}.)
+
+% The only case in the proof of part 1 that invokes part 2 is Case (\ref{rule:expandsU-tsmap}). There, we have that the metric remains stable: \begin{align*}
+%  & \sizeof{\expandsU{\uDelta}{\uGamma}{\uPsi}{\utsmap{\tsmv}{b}}{e}{\tau}}\\
+% =& \sizeof{\cvalidE{\emptyset}{\emptyset}{\esceneU{\uDelta}{\uGamma}{\uPsi}{b}}{\ce}{e}{\tau}}\\
+% =&\sizeof{b}\end{align*}
+
+% The only case in the proof of part 2 that invokes part 1 is Case (\ref{rule:cvalidE-U-splicede}).
+There, we have that $\parseUExp{\bsubseq{b}{m}{n}}{\ue}$ and the IH is applied to the judgement $\expandsU{\uDelta_\text{app}}{\uGamma_\text{app}}{\uPsi}{\ue}{e}{\tau}$ where $\uDelta_\text{app}=\uDD{\uD}{\Delta_\text{app}}$ and $\uGamma_\text{app}=\uGG{\uG}{\Gamma_\text{app}}$ and $\uPsi=\uAS{\uA}{\Psi}$.
+Because the metric is stable when passing from part 1 to part 2, we must have that it is strictly decreasing in the other direction:
+% \[\sizeof{\expandsU{\uDelta_\text{app}}{\uGamma_\text{app}}{\uPsi}{\ue}{e}{\tau}} < \sizeof{\cvalidE{\Delta}{\Gamma}{\esceneU{\uDelta_\text{app}}{\uGamma_\text{app}}{\uPsi}{b}}{\acesplicede{m}{n}{\ctau}}{e}{\tau}}\]
+% i.e. by the definitions above, 
+% \[\sizeof{\ue} < \sizeof{b}\]
+
+% This is established by appeal to the following two conditions.
+The first condition states that an unexpanded expression constructed by parsing a textual sequence $b$ is strictly smaller, as measured by the metric defined above, than the length of $b$, because some characters must necessarily be used to invoke a TSM and delimit each literal body.
+% \begingroup
+% \def\thetheorem{\ref{condition:body-parsing}}
+% \begin{condition}[Expression Parsing Monotonicity] If $\parseUExp{b}{\ue}$ then $\sizeof{\ue} < \sizeof{b}$.\end{condition}
+% \endgroup
+% The second condition simply states that subsequences of $b$ are no longer than $b$.
+% \begingroup
+% \def\thetheorem{\ref{condition:body-subsequences}}
+% \begin{condition}[Body Subsequencing] If $\bsubseq{b}{m}{n}=b'$ then $\sizeof{b'} \leq \sizeof{b}$. \end{condition}
+% \endgroup
+
+% Combining these two conditions, we have that $\sizeof{\ue} < \sizeof{b}$ as needed.
+% \end{proof}
+
+% % We need to define the following theorem about proto-expression validation mutually with Theorem \ref{thm:typed-expansion-U}. 
+% % \begin{theorem}[Proto-Expansion Expression Validation]\label{thm:candidate-expansion-validation-U}
+% % If $\cvalidE{\Delta}{\Gamma}{\esceneU{\Delta_\text{app}}{\Gamma_\text{app}}{\uPsi}{b}}{\ce}{e}{\tau}$ and $\uetsmenv{\Delta_\text{app}}{\uPsi}$ then $\hastypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\Gcons{\Gamma}{\Gamma_\text{app}}}{e}{\tau}$.
+% % \end{theorem}
+% % \begin{proof} By rule induction over Rules (\ref{rules:cvalidE-U}).
+% % \begin{byCases}
+% % \item[\text{(\ref{rule:cvalidE-U-var})}] ~
+% % \begin{pfsteps*}
+% %   \item $\ce=x$ \BY{assumption}
+% %   \item $e=x$ \BY{assumption}
+% %   \item $\Gamma=\Gamma', \Ghyp{x}{\tau}$ \BY{assumption}
+% %   \item $\hastypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\Gamma', \Ghyp{x}{\tau}}{x}{\tau}$ \BY{Rule (\ref{rule:hastypeU-var})} \pflabel{hastypeU}
+% %   \item $\hastypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\Gcons{\Gamma', \Ghyp{x}{\tau}}{\Gamma_\text{app}}}{x}{\tau}$ \BY{Lemma \ref{lemma:weakening-U} over $\Gamma_\text{app}$ to \pfref{hastypeU}}
+% % \end{pfsteps*}
+% % \resetpfcounter
+
+% % \item[\text{(\ref{rule:cvalidE-U-lam})}] ~
+% % \begin{pfsteps*}
+% %   \item $\ce=\acelam{\ctau_1}{x}{\ce'}$ \BY{assumption}
+% %   \item $e=\aelam{\tau_1}{x}{e'}$ \BY{assumption}
+% %   \item $\tau=\aparr{\tau_1}{\tau_2}$ \BY{assumption}
+% %   \item $\cvalidT{\Delta}{\esceneU{\Delta_\text{app}}{\Gamma_\text{app}}{\uPsi}{b}}{\ctau_1}{\tau_1}$ \BY{assumption} \pflabel{cvalidT}
+% %   \item $\cvalidE{\Delta}{\Gamma, \Ghyp{x}{\tau_1}}{\esceneU{\Delta_\text{app}}{\Gamma_\text{app}}{\uPsi}{b}}{\ce'}{e'}{\tau_2}$ \BY{assumption} \pflabel{cvalidE}
+% %   \item $\uetsmenv{\Delta_\text{app}}{\uPsi}$ \BY{assumption} \pflabel{uetsmenv}
+% %   \item $\istypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\tau_1}$ \BY{Lemma \ref{lemma:candidate-expansion-type-validation} on \pfref{cvalidT}} \pflabel{istype}
+% %   \item $\hastypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\Gcons{\Gamma, \Ghyp{x}{\tau_1}}{\Gamma_\text{app}}}{e'}{\tau_2}$ \BY{IH on \pfref{cvalidE} and \pfref{uetsmenv}} \pflabel{hastype1}
+% %   \item $\hastypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\Gcons{\Gamma}{\Gamma_\text{app}}, \Ghyp{x}{\tau_1}}{e'}{\tau_2}$ \BY{exchange over $\Gamma_\text{app}$ on \pfref{hastype1}} \pflabel{hastype2}
+% %   \item $\hastypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\Gcons{\Gamma}{\Gamma_\text{app}}}{\aelam{\tau_1}{x}{e'}}{\aparr{\tau_1}{\tau_2}}$ \BY{Rule (\ref{rule:hastypeU-lam}) on \pfref{istype} and \pfref{hastype2}}
+% % \end{pfsteps*}
+% % \resetpfcounter
+
+% % \item[\text{(\ref{rule:cvalidE-U-ap})}] ~
+% % \begin{pfsteps*}
+% %   \item $\ce=\aceap{\ce_1}{\ce_2}$ \BY{assumption}
+% %   \item $e=\aeap{e_1}{e_2}$ \BY{assumption}
+% %   \item $\cvalidE{\Delta}{\Gamma}{\esceneU{\Delta_\text{app}}{\Gamma_\text{app}}{\uPsi}{b}}{\ce_1}{e_1}{\aparr{\tau_1}{\tau}}$ \BY{assumption} \pflabel{cvalidE1}
+% %   \item $\cvalidE{\Delta}{\Gamma}{\esceneU{\Delta_\text{app}}{\Gamma_\text{app}}{\uPsi}{b}}{\ce_2}{e_2}{\tau_1}$ \BY{assumption} \pflabel{cvalidE2}
+% %   \item $\uetsmenv{\Delta_\text{app}}{\uPsi}$ \BY{assumption} \pflabel{uetsmenv}
+% %   \item $\hastypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\Gcons{\Gamma}{\Gamma_\text{app}}}{e_1}{\aparr{\tau_1}{\tau}}$ \BY{IH on \pfref{cvalidE1} and \pfref{uetsmenv}} \pflabel{hastypeU1}
+% %   \item $\hastypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\Gcons{\Gamma}{\Gamma_\text{app}}}{e_2}{\tau_1}$ \BY{IH on \pfref{cvalidE2} and \pfref{uetsmenv}} \pflabel{hastypeU2}
+% %   \item $\hastypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\Gcons{\Gamma}{\Gamma_\text{app}}}{\aeap{e_1}{e_2}}{\tau}$ \BY{Rule (\ref{rule:hastypeU-ap}) on \pfref{hastypeU1} and \pfref{hastypeU2}}
+% % \end{pfsteps*}
+% % \resetpfcounter
+
+% % \item[\VExpof{\text{\ref{rule:hastypeU-tlam}}}~\text{through}~\VExpof{\text{\ref{rule:hastypeU-case}}}] These cases follow analagously, i.e.
+we apply the IH to all proto-expression validation premises, Lemma \ref{lemma:candidate-expansion-type-validation} to all proto-types validation premises, weakening and exchange as needed, and then apply the corresponding typing rule.
+% % \\
+
+% % \item[\text{(\ref{rule:cvalidE-U-splicede})}] ~
+% % \begin{pfsteps*}
+% %   \item $\ce=\acesplicede{m}{n}$ \BY{assumption}
+% %   \item $\parseUExp{\bsubseq{b}{m}{n}}{\ue}$ \BY{assumption}
+% %   \item $\expandsU{\Delta_\text{app}}{\Gamma_\text{app}}{\uPsi}{\ue}{e}{\tau}$ \BY{assumption} \pflabel{expands}
+% %   \item $\uetsmenv{\Delta_\text{app}}{\uPsi}$ \BY{assumption} \pflabel{uetsmenv}
+% %   \item $\hastypeU{\Delta_\text{app}}{\Gamma_\text{app}}{e}{\tau}$ \BY{Theorem \ref{thm:typed-expansion-U} on \pfref{expands} and \pfref{uetsmenv}} \pflabel{hastype}
+% %   \item $\hastypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\Gcons{\Gamma}{\Gamma_\text{app}}}{e}{\tau}$ \BY{Lemma \ref{lemma:weakening-U} on \pfref{hastype}}
+% % \end{pfsteps*}
+% % \resetpfcounter
+% % \end{byCases}
+% % \end{proof}
+
+% %\qed
+
+% \subsubsection{Abstract Reasoning Principles}\label{sec:uetsms-reasoning-principles}
+% The following theorem summarizes the abstract reasoning principles that programmers can rely on when applying an seTSM. In particular, the programmer can be sure that:
+% \begin{enumerate} 
+% \item \textbf{Segmentation}: The segmentation determined by the proto-expansion actually segments the literal body (i.e. each segment is in-bounds and the segments are non-overlapping.)
+% \item \textbf{Typing 1}: The type of the expansion is consistent with the type annotation on the seTSM definition.
+% \item \textbf{Typing 2}: Each spliced type has a well-formed expansion at the application site.
+% \item \textbf{Typing 3}: Each type annotation on a reference to a spliced expression has a well-formed expansion at the application site.
+% \item \textbf{Typing 4}: Each spliced expression has a well-typed expansion consistent with its type annotation.
+% \item \textbf{Capture Avoidance}: The final expansion can be decomposed into a  term with variables in place of each spliced type or expression.
+The expansions of these spliced types and expressions can be substituted into this term in the standard capture avoiding manner.
+% \item \textbf{Context Independence}: The decomposed term is indeed well-typed independent of the application site contexts.
+
+% \end{enumerate}
+
+% \begingroup
+% \def\thetheorem{\ref{thm:tsc-SES}}
+% \begin{theorem}[seTSM Abstract Reasoning Principles]
+% If $\expandsU{\uDD{\uD}{\Delta}}{\uGG{\uG}{\Gamma}}{\uPsi}{\utsmap{\tsmv}{b}}{e}{\tau}$ then:
+% \begin{enumerate}
+% \item (\textbf{Typing 1}) $\uPsi = \uPsi', \uShyp{\tsmv}{a}{\tau}{\eparse}$ and $\hastypeU{\Delta}{\Gamma}{e}{\tau}$
+% \item $\encodeBody{b}{\ebody}$
+% \item $\evalU{\ap{\eparse}{\ebody}}{\aein{\lbltxt{SuccessE}}{\ecand}}$
+% \item $\decodeCondE{\ecand}{\ce}$
+% \item (\textbf{Segmentation}) $\segOK{\segof{\ce}}{b}$
+% \item $\summaryOf{\ce} = \sseq{\acesplicedt{m'_i}{n'_i}}{\nty} \cup \sseq{\acesplicede{m_i}{n_i}{\ctau_i}}{\nexp}$
+% \item \textbf{(Typing 2)} $\sseq{
+%       \expandsTU{\uDD{\uD}{\Delta}}
+%       {
+%         \parseUTypF{\bsubseq{b}{m'_i}{n'_i}}
+%       }{\tau'_i}
+%     }{\nty}$ and $\sseq{\istypeU{\Delta}{\tau'_i}}{\nty}$
+% \item \textbf{(Typing 3)} $\sseq{
+%   \cvalidT{\emptyset}{
+%     \tsceneUP
+%       {\uDD
+%         {\uD}{\Delta}
+%       }{b}
+%   }{
+%     \ctau_i
+%   }{\tau_i}
+% }{\nexp}$ and $\sseq{\istypeU{\Delta}{\tau_i}}{\nexp}$
+% \item \textbf{(Typing 4)} $\sseq{
+%   \expandsU
+%     {\uDD{\uD}{\Delta}}
+%     {\uGG{\uG}{\Gamma}}
+%     {\uPsi}
+%     {\parseUExpF{\bsubseq{b}{m_i}{n_i}}}
+%     {e_i}
+%     {\tau_i}
+% }{\nexp}$ and $\sseq{\hastypeU{\Delta}{\Gamma}{e_i}{\tau_i}}{\nexp}$
+% \item (\textbf{Capture Avoidance}) $e = [\sseq{\tau'_i/t_i}{\nty}, \sseq{e_i/x_i}{\nexp}]e'$ for some $\sseq{t_i}{\nty}$ and $\sseq{x_i}{\nexp}$ and $e'$
+% \item (\textbf{Context Independence}) $\mathsf{fv}(e') \subset \sseq{t_i}{\nty} \cup \sseq{x_i}{\nexp}$
+%   % $\hastypeU
+%   % {\sseq{\Dhyp{t_i}}{\nty}}
+%   % {\sseq{x_i : \tau_i}{\nexp}}
+%   % {e'}{\tau}$
+% \end{enumerate}
+% \end{theorem}
+% \begin{proof} The proof, which involves auxiliary lemmas about the decomposition of proto-types and proto-expressions, is given in Appendix \ref{appendix:SES-reasoning-principles}.
+% \end{proof}
+% \endgroup
+
+\newcommand{\pTSMsSec}{Parametric TSMs (pTSMs)}
+\section{\protect\pTSMsSec}
+\label{sec:ptsms}
+Simple TSMs operate at a single specified type and the proto-expansions they generate must be closed.
+This simplifies matters formally, but it is not convenient in practice.
+This section introduces \emph{parametric TSMs}.
+Parametric TSMs can be defined over a type- and module-parameterized family of types, and the proto-expansions they generate can refer to supplied type and module parameters.
+
+\subsection{Parametric TSMs By Example}\label{sec:ptsms-by-example}
+
+Consider the following ML signature specifying an abstract data type \cite{liskov1974programming,harper1997programming} of string-keyed polymorphic dictionaries:
+\begin{lstlisting}[numbers=none]
+  signature DICT = sig
+    type t('a)
+    val empty : t('a)
+    val extend : t('a) -> string * 'a -> t('a)
+    (* ... *)
+  end
+\end{lstlisting}
+Let us now define a TSM that is parametric over all implementations of this signature, \li{D : DICT}, and also over all choices of type \li{'a}:
+\begin{lstlisting}[numbers=none]
+  syntax $dict (D : DICT) (type 'a) at D.t('a) by static
+    fn(b : body) -> parse_result(proto_expr) => (* ... *)
+  end
+\end{lstlisting}
+For example, assuming some implementation \li{Dict : DICT} and some values \li{v1 : int} and \li{v2 : int} (and assuming that VerseML offers a generalized literal form delimited by \li{\{|} and \li{|\}}), we can apply \li{$dict} as follows:
+\begin{lstlisting}[numbers=none]
+  $dict Dict int {| "key1"SURL => EURLv1SURL; EURL"key2"SURL => EURLv2 |}
+\end{lstlisting}
+Notice that the segmentation immediately reveals which punctuation is particular to this TSM and where the spliced key and value expressions appear.
+Because the context-free syntax of unexpanded terms is never modified, it is possible to reason modularly about syntactic determinism (i.e.
+we can reason above that \li{=>} does not appear in the follow set of unexpanded terms, so there can never be an ambiguity about where a key expression ends.)
+
+If we will use the \li{Dict} implementation ubiquitously, we can abbreviate the partial application of \li{$dict} to \li{Dict}, resulting in a TSM that is parametric over only the type \li{'a}, as follows:
+\begin{lstlisting}[numbers=none]
+  let syntax $Dict = $dict Dict in 
+  $Dict int {| "key1"SURL => EURLv1SURL; EURL"key2"SURL => EURLv2 |}
+\end{lstlisting}
+The proto-expansion generated for the TSM application above might be:
+\begin{lstlisting}[numbers=none]
+  D.extend (D.extend D.empty (spliced<1; 6; string>, spliced<11; 12; 'a>)) 
+    (spliced<15; 20; string>, spliced<25; 26; 'a>)
+\end{lstlisting}
+The generated proto-expansion must truly be parametric, i.e.
+it must be valid for all modules \li{D : DICT} and types \li{'a}.
+It is only after proto-expansion validation that we substitute the actual parameters, here \li{Dict} for \li{D} and \li{int} for \li{'a}, into the final expansion.
+ADditionally, substitution occurs on the type annotations on references to spliced terms before recursively expanding those terms (otherwise, the expressions \li{v1} and \li{v2} above would need to be well-typed for all types, \li{'a}.)
+
+Module parameters are also useful for giving the expansion access to helper functions in a context-independent manner.
+For example, in Sec.
+\ref{sec:context-dependence} we discussed the problem of applying datatype constructors like \li{H1Element}, because they are introduced into the context as variables in most ML-like languages.
+With module parameters, it is possible to sidestep this problem by passing in a module containing the datatype constructors.
+In practice, the language might implicitly construct and open such a module within the expansion when the type annotation on the TSM is such a datatype.
+
+It is important to note that module parameters are not accessible by parse functions directly, because the applied module parameters will not have been dynamically instantiated when typed expansion occurs.
+Parameters are accessible only from within the expansion that the parse function generates.
+We will discuss a distinct mechanism for providing helper functions to parse functions in Sec.
+\ref{sec:static-eval}.
+
+\subsection{Parametric TSMs, Formally}
+% \begin{figure}[p]
+% \[\begin{array}{llcl}
+% \mathsf{UMType} & \urho & ::= 
+% %& \autype{\utau} 
+% & \utau & \text{type annotation}\\
+% % &&
+% %& \aualltypes{\ut}{\urho} 
+% % & \alltypes{\ut}{\urho} & \text{type parameterization}\\
+% &&
+% %& \auallmods{\usigma}{\uX}{\urho} 
+% & \allmods{\uX}{\usigma}{\urho} & \text{module parameterization}\\
+% \mathsf{UMExp} & \uepsilon & ::= 
+% %& \abindref{\tsmv} 
+% & \tsmv & \text{TSM identifier reference}\\
+% % &&
+% %& \auabstype{\ut}{\uepsilon} 
+% % & \abstype{\ut}{\uepsilon} & \text{type abstraction}\\
+% &&
+% %& \auabsmod{\usigma}{\uX}{\uepsilon} 
+% & \absmod{\uX}{\usigma}{\uepsilon} & \text{module abstraction}\\
+% % &&
+% %& \auaptype{\utau}{\uepsilon} 
+% % & \aptype{\uepsilon}{\utau} & \text{type application}\\
+% &&
+% %& \auapmod{\uM}{\uepsilon} 
+% & \apmod{\uepsilon}{\uX} & \text{module application}\ECC
+% \end{array}
+% \]
+% \caption{Syntax of unexpanded TSM types and expressions.}
+% \label{fig:P-macro-expressions-types-u}
+% \end{figure}
+
+We will now outline $\miniVerseParam$, a calculus that extends $\miniVersePat$ with support for parametric TSMs.
+This calculus is organized, like $\miniVersePat$, as an unexpanded language (UL) defined by typed expansion to an expanded language (XL).
+There is not enough space to describe $\miniVerseParam$ with the same level of detail as in Sec.
+\ref{sec:setsms-formally}, so we highlight only the most important concepts below.
+
+The XL consists of 1) module expressions, $M$, classified by signatures, $\sigma$; 2) constructions, $c$, classified by kinds, $\kappa$; and 3) expressions classified by types, which are constructions of kind $\akty$ (we use metavariables $\tau$ instead of $c$ for types.) Metavariables $X$ ranges over module variables and $u$ or $t$ over construction variables.
+The module and construction languages are based closely on those defined by \citet{pfple1}, which in turn are based on early work by \citet{MacQueen:1984:MSM:800055.802036,DBLP:conf/popl/MacQueen86}, subsequent work on the phase splitting interpretation of modules \cite{harper1989higher} and on using dependent singleton kinds to track type identity \cite{stone2006extensional,DBLP:conf/lfmtp/Crary09}, and finally on formal developments by \citet{dreyer2005understanding} and \citet{conf/popl/LeeCH07}.
+A complete account of these developments is unfortunately beyond the scope of this paper.
+The expression language extends the language of $\miniVersePat$ only to allow projection out of modules.
+
+The main conceptual difference between $\miniVersePat$ and $\miniVerseParam$ is that $\miniVerseParam$ introduces the notion of unexpanded and expanded TSM expressions and types, as shown in Figure \ref{fig:P-macro-expressions-types}.
+
+\begin{figure}[h]
+\begin{minipage}{0.36\textwidth}
+$\begin{array}{llcl}
+\mathsf{UMType} & \urho & ::= & \utau ~\vert~ \allmods{\uX}{\usigma}{\urho}\\
+\mathsf{UMExp} & \uepsilon & ::= & \tsmv ~\vert~ \absmod{\uX}{\usigma}{\uepsilon}~\vert~ \apmod{\uepsilon}{\uX}
+\end{array}$
+\end{minipage}
+\begin{minipage}{0.6\textwidth}
+$\begin{array}{llcl}
+\mathsf{MType} & \rho & ::= & \aetype{\tau} ~\vert~ \aeallmods{\sigma}{X}{\rho} \\
+\mathsf{MExp} & \epsilon & ::= & \adefref{a} ~\vert~ \aeabsmod{\sigma}{X}{\epsilon} ~\vert~\aeapmod{X}{\epsilon} 
+\end{array}$
+\end{minipage}
+\vspace{-6px}
+\caption{Syntax of unexpanded and expanded TSM types and expressions in $\miniVerseParam$}
+\label{fig:P-macro-expressions-types}
+\vspace{-6px}
+\end{figure}
+This is necessary to express abstraction over parameters and application of parameters.
+For simplicity, we formalize only module parameters.
+Type parameters can be realized as module parameters having exactly one abstract type member.
+
+The rule governing expression TSM application, reproduced below, touches all of the main ideas in $\miniVerseParam$. %It might be useful to compare this rule to Rule \textsc{ee-ap-sptsm}.
+\begin{mathpar}
+\inferrule[ee-ap-petsm]{
+  \uOmega = \uOmegaEx{\uD}{\uG}{\uMctx}{\Omega_\text{app}}\\
+  \uPsi=\uAS{\uA}{\Psi}\\\\
+  \tsmexpExpandsExp{\uOmega}{\uPsi}{\uepsilon}{\epsilon}{\aetype{\tau_\text{final}}}\\
+  \tsmexpEvalsExp{\Omega_\text{app}}{\Psi}{\epsilon}{\epsilon_\text{normal}}\\\\
+  \tsmdefof{\epsilon_\text{normal}}=a\\
+  \Psi = \Psi', \petsmdefn{a}{\rho}{\eparse}\\\\
+  \encodeBody{b}{\ebody}\\
+  \evalU{\ap{\eparse}{\ebody}}{\aein{\mathtt{SuccessE}}{e_\text{pproto}}}\\
+  \decodePCEExp{e_\text{pproto}}{\pce}\\\\
+  \prepce{\Omega_\text{app}}{\Psi}{\pce}{\ce}{\epsilon_\text{normal}}{\aetype{\tau_\text{proto}}}{\omega}{\Omega_\text{params}}\\\\
+  \segOK{\segof{\ce}}{b}\\
+  \cvalidEP{\Omega_\text{params}}{\esceneP{\omega : \OParams}{\uOmega}{\uPsi}{\uPhi}{b}}{\ce}{e}{\tau_\text{proto}}
+}{
+  \expandsP{\uOmega}{\uPsi}{\uPhi}{\utsmap{\uepsilon}{b}}{[\omega]e}{[\omega]\tau_\text{proto}}
+}
+\end{mathpar}
+The first two premises simply deconstruct the (unified) unexpanded context $\uOmega$ (which tracks the expansion of expression, constructor and module identifiers, as $\uDelta$ and $\uGamma$ did in $\miniVersePat$) and peTSM context, $\uPsi$.
+Next, we expand $\uepsilon$ according to straightforward unexpanded peTSM expression expansion rules.
+The resulting TSM expression, $\epsilon$, must be defined at a type (i.e.
+no quantification must remain.)
+
+The fourth premise performs \emph{peTSM expression normalization}.
+Normalization, $\tsmexpEvalsExp{\Omega_\text{app}}{\Psi}{\epsilon}{\epsilon_\text{normal}}$, is defined in terms of a simple structural dynamics with two stepping rules:
+% \begin{equation*}\tag{\ref{rule:tsmexpEvalsExp}}
+% \inferrule{
+%   \tsmexpMultistepsExp{\Omega}{\Psi}{\epsilon}{\epsilon'}\\
+%   \tsmexpNormalExp{\Omega}{\Psi}{\epsilon'}
+% }{
+%   \tsmexpEvalsExp{\Omega}{\Psi}{\epsilon}{\epsilon'}
+% }
+% \end{equation*}
+% where the multistep judgement, $\tsmexpMultistepsExp{\Omega}{\Psi}{\epsilon}{\epsilon'}$, is defined as the reflexive, transitive closure of the stepping judgement defined by the following rules:
+% \begin{equation*}\tag{\ref{rule:tsmexpStepsExp-aptype-1}}
+% \inferrule{
+%   \tsmexpStepsExp{\Omega}{\Psi}{\epsilon}{\epsilon'}
+% }{
+%   \tsmexpStepsExp{\Omega}{\Psi}{\aeaptype{\tau}{\epsilon}}{\aeaptype{\tau}{\epsilon'}}
+% }
+% \end{equation*}
+% \begin{equation*}\tag{\ref{rule:tsmexpStepsExp-aptype-2}}
+% \inferrule{ }{
+%   \tsmexpStepsExp{\Omega}{\Psi}{\aeaptype{\tau}{\aeabstype{t}{\epsilon}}}{[\tau/t]\epsilon}
+% }
+% \end{equation*}
+\begin{mathpar}
+\inferrule[eps-dyn-apmod-subst-e]{ }{
+  \tsmexpStepsExp{\Omega}{\Psi}{\aeapmod{X}{\aeabsmod{\sigma}{X'}{\epsilon}}}{[X/X']\epsilon}
+}
+
+\inferrule[eps-dyn-apmod-steps-e]{
+  \tsmexpStepsExp{\Omega}{\Psi}{\epsilon}{\epsilon'}
+}{
+  \tsmexpStepsExp{\Omega}{\Psi}{\aeapmod{X}{\epsilon}}{\aeapmod{X}{\epsilon'}}
+}
+\end{mathpar}
+% The peTSM expression normal forms are defined as follows:
+% \begin{equation*}\tag{\ref{rule:tsmexpNormalExp-defref}}
+% \inferrule{ }{
+%   \tsmexpNormalExp{\Omega}{\Psi, \petsmdefn{a}{\rho}{\eparse}}{\adefref{a}}
+% }
+% \end{equation*}
+% % \begin{equation*}\tag{\ref{rule:tsmexpNormalExp-abstype}}
+% % \inferrule{ }{
+% %   \tsmexpNormalExp{\Omega}{\Psi}{\aeabstype{t}{\epsilon}}
+% % }
+% % \end{equation*}
+% \begin{equation*}\tag{\ref{rule:tsmexpNormalExp-absmod}}
+% \inferrule{ }{
+%   \tsmexpNormalExp{\Omega}{\Psi}{\aeabsmod{\sigma}{X}{\epsilon}}
+% }
+% \end{equation*}
+% % \begin{equation*}\tag{\ref{rule:tsmexpNormalExp-aptype}}
+% % \inferrule{
+% %   \epsilon \neq \aeabstype{t}{\epsilon'}\\
+% %   \tsmexpNormalExp{\Omega}{\Psi}{\epsilon}
+% % }{
+% %   \tsmexpNormalExp{\Omega}{\Psi}{\aeaptype{\tau}{\epsilon}}
+% % }
+% % \end{equation*}
+% \begin{equation*}\tag{\ref{rule:tsmexpNormalExp-apmod}}
+% \inferrule{
+%   \epsilon \neq \aeabsmod{\sigma}{X'}{\epsilon'}\\
+%   \tsmexpNormalExp{\Omega}{\Psi}{\epsilon}
+% }{
+%   \tsmexpNormalExp{\Omega}{\Psi}{\aeapmod{X}{\epsilon}}
+% }
+% \end{equation*}
+Normalization eliminates parameters introduced in higher-order abbreviations, leaving only those parameter applications specified by the original TSM definition.
+Normal forms and progress and preservation theorems are established in the supplement.
+
+The third row of premises looks up the applied TSM's definition by first invoking a simple metafunction to extract the name, then looking up this name within the peTSM definition context, $\Psi$.
+% \begin{align}
+% \tsmdefof{\adefref{a}} & = a \tag{\ref{eqn:tsmdefof-adefref}}\\
+% % \tsmdefof{\aeabstype{t}{\epsilon}} & = \tsmdefof{\epsilon} \tag{\ref{eqn:tsmdefof-abstype}}\\
+% \tsmdefof{\aeabsmod{\sigma}{X}{\epsilon}} & = \tsmdefof{\epsilon} \tag{\ref{eqn:tsmdefof-absmod}}\\
+% % \tsmdefof{\aeaptype{\tau}{\epsilon}} & = \tsmdefof{\epsilon} \tag{\ref{eqn:tsmdefof-aptype}}\\
+% \tsmdefof{\aeapmod{X}{\epsilon}} & = \tsmdefof{\epsilon} \tag{\ref{eqn:tsmdefof-apmod}}
+% \end{align}
+
+The fourth row of premises 1) encodes the body as a value of the type $\tBody$; 2) applies the parse function; and 3) decodes the result, producing a \emph{parameterized proto-expression}, $\pce$.
+Parameterized proto-expressions, $\pce$, are ABTs that serve simply to introduce the parameter bindings into an underlying proto-expression, $\ce$.
+The syntax of parameterized proto-expressions is given below.
+\[\begin{array}{llcl}
+% \textbf{Sort} & & & \textbf{Operational Form} & \textbf{Stylized Form} & \textbf{Description}\\
+% \LCC \color{Yellow}&\color{Yellow}&\color{Yellow}& \color{Yellow} & \color{Yellow} & \color{Yellow}\\
+\mathsf{PPrExp} & \pce & ::= & \apceexp{\ce} ~\vert~ \apcebindmod{X}{\pce}
+\end{array}\]
+There must be one binder in $\pce$ for each TSM parameter specified by $\tsmdefof{\epsilon_\text{normal}}$.
+(VerseML can insert these binders automatically as a convenience, but we consider only the underlying mechanism in this core calculus.)
+
+The judgement on the fifth row of Rule \textsc{ee-ap-petsm} then \emph{deparameterizes} $\pce$ by peeling away these binders to produce 1) the underlying proto-expression, $\ce$, with the variables that stand for the parameters free; 2) a corresponding deparameterized type, $\tau_\text{proto}$, that uses the same free variables to stand for the parameters; 3) a \emph{substitution}, $\omega$, that pairs the applied parameters from $\epsilon_\text{normal}$ with the corresponding variables generated when peeling away the binders in $\pce$; and 4) a corresponding \emph{parameter context}, $\Omega_\text{params}$, that tracks the signatures of these variables.
+The two rules governing the proto-expression deparameterization judgement are below:
+\begin{mathpar}
+\inferrule{ }{
+  \prepce{\Omega_\text{app}}{\Psi, \petsmdefn{a}{\rho}{\eparse}}{\apceexp{\ce}}{\ce}{\adefref{a}}{\rho}{\emptyset}{\emptyset}
+}
+
+\inferrule{
+  \prepce{\Omega_\text{app}}{\Psi}{\pce}{\ce}{\epsilon}{\aeallmods{\sigma}{X}{\rho}}{\omega}{\Omega}\\
+  X \notin \domof{\Omega_\text{app}}
+}{
+  \prepce{\Omega_\text{app}}{\Psi}{\apcebindmod{X}{\pce}}{\ce}{\aeapmod{X'}{\epsilon}}{\rho}{(\omega, X'/X)}{(\Omega, X : \sigma)}
+}
+\end{mathpar}
+This judgement can be pronounced ``when applying peTSM $\epsilon$, $\pce$ has deparameterization $\ce$ leaving $\rho$ with parameter substitution $\omega$''. 
+Notice based on the second rule that every module binding in $\pce$ must pair with a corresponding module parameter application.
+Moreover, the variables standing for parameters must not appear in $\Omega_\text{app}$, i.e.
+$\domof{\Omega_\text{params}}$ must be disjoint from $\domof{\Omega_\text{app}}$ (this requirement can always be discharged by alpha-variation.)
+
+The final row of premises checks that the segmentation of $\ce$ is valid and  performs proto-expansion validation under the parameter context, $\Omega_\text{param}$ (rather than the empty context, as was the case in $\miniVersePat$.) The conclusion of the rule applies the parameter substitution, $\omega$, to the resulting expression and the deparameterized type.
+
+Proto-expansion validation operates conceptually as in $\miniVersePat$.
+The only subtlety has to do with the type annotations on references to spliced terms.
+As described at the end of Sec.
+\ref{sec:ptsms-by-example}, these annotations might refer to the parameters, so the parameter substitution, $\omega$, which is tracked by the splicing scene, must be applied to the type annotation before proceeding recursively to expand the referenced unexpanded term.
+However, the spliced term itself must treat parameters parametrically, so the substitution is not applied in the conclusion of the rule:
+\begin{mathpar}\label{rule:cvalidE-P-splicede}
+\inferrule[p-pev-spliced]{
+  \parseUExp{\bsubseq{b}{m}{n}}{\ue}\\
+    \cvalidC{\OParams}{\csceneP{\omega : \OParams}{\uOmega}{b}}{\ctau}{\tau}{\akty}\\
+  \expandsP{\uOmega}{\uPsi}{\uPhi}{\ue}{e}{[\omega]\tau}\\\\
+  \uOmega=\uOmegaEx{\uD}{\uG}{\uMctx}{\Omega_\text{app}}\\
+  \domof{\Omega} \cap \domof{\Omega_\text{app}} = \emptyset
+}{
+  \cvalidEP{\Omega}{\esceneP{\omega : \OParams}{\uOmega}{\uPsi}{\uPhi}{b}}{\acesplicede{m}{n}{\ctau}}{e}{\tau}
+}\vspace{-5px}
+\end{mathpar}
+(This is only sensible because we maintain the invariant that $\Omega$ is always an extension of $\Omega_\text{params}$.)
+
+The calculus enjoys metatheoretic properties analagous to those described in Sec.
+\ref{sec:s-metatheory}, modified to account for the presence of modules, kinds and parameterization.
+The following theorem establishes the abstract reasoning principles available when applying a parametric expression TSM.
+\vspace{-3px}
+\begin{theorem}[peTSM Reasoning Principles]
+If $\expandsP{\uOmega}{\uPsi}{\uPhi}{\utsmap{\uepsilon}{b}}{e}{\tau}$ then:
+\begin{enumerate}[nolistsep]
+  \item $\uOmega=\uOmegaEx{\uD}{\uG}{\uMctx}{\Omega_\text{app}}$
+  \item $\uPsi=\uAS{\uA}{\Psi}$
+  \item (\textbf{Typing 1}) $\tsmexpExpandsExp{\uOmega}{\uPsi}{\uepsilon}{\epsilon}{\aetype{\tau}}$ and $\hastypeP{\Omega_\text{app}}{e}{\tau}$
+  \item $\tsmexpEvalsExp{\Omega_\text{app}}{\Psi}{\epsilon}{\epsilon_\text{normal}}$
+  \item $\tsmdefof{\epsilon_\text{normal}}=a$
+  \item $\Psi = \Psi', \petsmdefn{a}{\rho}{\eparse}$
+  \item $\encodeBody{b}{\ebody}$
+    \item $\evalU{\ap{\eparse}{\ebody}}{\aein{\mathtt{SuccessE}}{e_\text{pproto}}}$
+  \item $\decodePCEExp{e_\text{pproto}}{\pce}$
+  \item $\prepce{\Omega_\text{app}}{\Psi}{\pce}{\ce}{\epsilon_\text{normal}}{\aetype{\tau_\text{proto}}}{\omega}{\Omega_\text{params}}$
+  \item (\textbf{Segmentation}) $\segOK{\segof{\ce}}{b}$
+  \item $\cvalidEP{\Omega_\text{params}}{\esceneP{\omega : \OParams}{\uOmega}{\uPsi}{\uPhi}{b}}{\ce}{e'}{\tau_\text{proto}}$
+  \item $e = [\omega]e'$
+  \item $\tau = [\omega]\tau_\text{proto}$
+  \item $
+    \summaryOf{\ce} = \sseq{\acesplicedk{m_i}{n_i}}{\nkind} \cup \sseq{\acesplicedc{m'_i}{n'_i}{\cekappa'_i}}{\ncon} \cup 
+               \sseq{\acesplicede{m''_i}{n''_i}{\ctau_i}}{\nexp}
+    $
+  \item (\textbf{Kinding 1}) $\sseq{\kExpands{\uOmega}{\parseUKindF{\bsubseq{b}{m_i}{n_i}}}{\kappa_i}}{\nkind}$ and $\sseq{\iskind{\Omega_\text{app}}{\kappa_i}}{\nkind}$
+  \item (\textbf{Kinding 2}) $\sseq{\cvalidK{\OParams}{\csceneP{\omega : \OParams}{\uOmega}{b}}{\cekappa'_i}{\kappa'_i}}{\ncon}$ and $\sseq{\iskind{\Omega_\text{app}}{[\omega]\kappa'_i}}{\ncon}$
+  \item (\textbf{Kinding 3}) $\sseq{\cExpands{\uOmega}{\parseUConF{\bsubseq{b}{m'_i}{n'_i}}}{c_i}{[\omega]\kappa'_i}}{\ncon}$ and $\sseq{\haskind{\Omega_\text{app}}{c_i}{[\omega]\kappa'_i}}{\ncon}$
+  \item (\textbf{Kinding 4}) $\sseq{\cvalidC{\OParams}{\csceneP{\omega : \OParams}{\uOmega}{b}}{\ctau_i}{\tau_i}{\akty}}{\nexp}$ and $\sseq{\haskind{\Omega_\text{app}}{[\omega]\tau_i}{\akty}}{\nexp}$
+  \item (\textbf{Typing 2}) $\sseq{\expandsP{\uOmega}{\uPsi}{\uPhi}{\parseUExpF{\bsubseq{b}{m''_i}{n''_i}}}{e_i}{[\omega]\tau_i}}{\nexp}$ and $\sseq{\hastypeP{\Omega_\text{app}}{e_i}{[\omega]\tau_i}}{\nexp}$
+  \item (\textbf{Capture Avoidance}) $e = [\sseq{\kappa_i/k_i}{\nkind}, \sseq{c_i/u_i}{\ncon}, \sseq{e_i/x_i}{\nexp}, \omega]e''$ for some $e''$ and fresh $\sseq{k_i}{\nkind}$ and fresh $\sseq{u_i}{\ncon}$ and fresh $\sseq{x_i}{\nexp}$
+  \item (\textbf{Context Independence}) $\mathsf{fv}(e'') \subset \sseq{k_i}{\nkind} \cup \sseq{u_i}{\ncon} \cup \sseq{x_i}{\nexp} \cup \domof{\OParams}$
+  % $\hastypeP{\sseq{\Khyp{k_i}}{\nkind} \cup \sseq{u_i :: [\omega]\kappa'_i}{\ncon} \cup \sseq{x_i : [\omega]\tau_i}{\nexp}}{[\omega]e''}{\tau}$\todo{maybe restate this in terms of free variables of e'' here and elsewhere, because context isn't technically well-formed here?}
+\end{enumerate}
+\end{theorem}
+We leave out the details of parametric pattern TSMs (ppTSMs), but they should again follow analagously.
+
+\newcommand{\staticEvalSec}{Static Evaluation}
+\section{\protect\staticEvalSec}
+\label{sec:static-eval}
+
+Thus far, we have assumed that the parse functions in TSM definitions are closed expanded expressions.
+This assumption simplifies matters formally, but in practice it is quite difficult, because this means that TSM definitions cannot themselves access any libraries nor use TSMs internally.
+To address this problem, we need to introduce the concept of a \emph{static environment}.
+
+\subsection{The Static Environment}
+Figure \ref{fig:static-module-example} shows an example of a module, \li{ParserCombos}, that defines a number of parser combinators following \citet{Hutton1992d}.
+The \li{static} qualifier indicates that this module is bound for use only within similarly qualified values, including in particular the parse functions of the subsequent TSM definitions.
+\begin{figure}[h]
+\vspace{-10px}
+\begin{lstlisting}
+  static module ParserCombos = struct 
+    type parser('c, 't) = list('c) -> list('t * list('c))
+    val alt : parser('c, 't) -> parser('c, 't) -> parser('c, 't)
+    (* ... *)
+  end
+  syntax $a at t by static fn(b) => (* OK *) ParserCombos.alt (* ... *) end 
+  static val y = (* OK *) ParserCombos.alt (* ... *)
+  val z = (* neither ParserCombos nor y are available here *)
+\end{lstlisting}
+\vspace{-5px}
+\caption{Binding a static module for use within parse functions.}
+\label{fig:static-module-example}
+\vspace{-10px}
+\end{figure}
+
+The values that arise during the static evaluation (in particular, of parse functions) do not themselves persist from ``compile-time'' to ``run-time'', so we do not need a full staged computation system, e.g.
+as described by \citet{Taha99multi-stageprogramming:}.
+Instead, a sequence of static bindings operates like a read-evaluate-print loop (REPL) scoped according to the program structure, in that each static expression is evaluated immediately and the evaluated values are tracked by a \emph{static environment}.
+The static environment is discarded at the end of typed expansion.
+
+A language designer might choose to restrict the external effects available to static terms in some way, e.g.
+to ensure deterministic builds.
+It might also be helpful to restrict mutable state shared between TSMs to prevent undesirable TSM application order dependencies.
+On the other hand, these features, if used for the purposes of caching in a disciplined manner, might speed up typed expansion.
+These are orthogonal design decisions.
+
+\subsection{Applying TSMs Within TSM Definitions}\label{sec:tsms-for-tsms}
+TSMs and TSM abbreviations can also be qualified with the \li{static} keyword, which marks them for use within subsequent static expressions and patterns.
+Let us consider some examples of relevance to TSM providers.
+
+\subsubsection{Quasiquotation}
+TSMs must construct values of type \li{proto_expr} or \li{proto_pat}.
+Constructing values of these types explicitly can have high syntactic cost.
+To decrease this cost, we can define TSMs that provide support for \emph{quasiquotation syntax} similar to that built in to languages like Lisp \cite{Bawd99a} and Scala \cite{shabalin2013quasiquotes}.
+The following TSM defines quasi-quotation for encodings of proto-expressions:
+\begin{lstlisting}[numbers=none]
+  static syntax $proto_expr at proto_expr by static fn(b) => (* ... *) end
+\end{lstlisting}
+For example, \li{$proto_expr `SQTg x ^EQTx`} might have expansion \li{App(App(Var "g", Var "x"), x)}.
+Notice that prefixing a variable (or parenthesized expression) with \li{^} serves to splice in its value, which here must be of type \li{proto_expr} (though in other syntactic positions, it might be \li{proto_typ}.) This is also known as \emph{anti-quotation}.
+
+% A similar approach can be taken for working with encodings of terms of other languages (e.g. when writing an interpretter or compiler in VerseML.)
+
+\subsubsection{Parser Generators}
+Abstractly, a grammar-based parser generator is a module matching the signature \li{PARSEGEN} defined below:
+\begin{lstlisting}
+  signature PARSEGEN = sig 
+    type grammar('a)
+    (* ... operations on grammars ... *)
+    val generate : grammar('a) -> (body -> parse_result('a))
+  end
+\end{lstlisting}
+
+Rather than constructing a grammar (equipped with semantic actions) using the associated operations (whose specifications are elided in \li{PARSEGEN}), we wish to use a syntax for context-free grammars that follows standard conventions.
+We can do so by defining a static parametric TSM \li{$grammar}:
+\begin{lstlisting}[numbers=none]
+  static syntax $grammar (P : PARSEGEN) 'a at P.grammar('a) by (* ... *) end
+\end{lstlisting}
+To support splicing, we need non-terminals that recognize unexpanded terms and produce the corresponding splice reference, rather than the AST itself.
+This would require that the parser generator keep track of location information (as most production-grade parser generators already do for error reporting.) For spliced expressions, this non-terminal would need to be a family of non-terminals indexed by a value of type \li{proto_typ}.
+
+A grammar containing such non-terminals can serve as a \emph{summary specification} -- a human can simply be take this grammar as a specification of what the splice summary will be for every recognized string, rather than relying on an editor to communicate this information.
+The associated semantic actions can be held abstract as long as the system performs a simple check to ensure that the generated proto-expansion does in fact mention each spliced expression from the corresponding production.
+
+% \begin{figure}[h!]
+% \vspace{-5px}
+% \begin{lstlisting}[deletekeywords={as}]
+% syntax $rx(R : RX) at R.t by static 
+%   P.generate ($grammar P proto_expr {|SHTML #\label{line:rx_parse_fn_start}#
+%     start <- ""
+%       EHTMLfn () => $proto_expr `SCSSR.EmptyECSS`SHTML
+%     start <- "(" start ")"
+%       EHTMLfn e => eSHTML
+%     token str_tok #\label{line:str_tok_start}#
+%       EHTMLRU.parse "SSTR[^(@$]+ESTR" (* cannot use $rx within its own def *)SHTML #\label{line:str_tok_end}#
+%     start <- str_tok
+%       EHTMLfn s => $proto_expr `SCSSR.Str %(ECSSstr_to_proto_lit sSCSS)ECSS`SHTML
+%     start <- start start
+%       EHTMLfn e1 e2 => $proto_expr `SCSSR.Seq (%ECSSe1SCSS, %ECSSe2SCSS)ECSS`SHTML
+%     start <- start "|" start 
+%       EHTMLfn e1 e2 => $proto_expr `SCSSR.Or (%ECSSe1SCSS, %ECSSe2SCSS)ECSS`SHTML
+%     start <- start "*"
+%       EHTMLfn e => $proto_expr `SCSSR.Star %ECSSe`SHTML
+
+%     using EHTMLspliced_uexp ($proto_typ `SCSSR.tECSS`) SHTML as spliced_rx #\label{line:splicede_using}#
+%     start <- "${" spliced_rx "}" #\label{line:splicing-start}#
+%       EHTMLfn e => eSHTML
+
+%     using EHTMLspliced_uexp ($proto_typ `SCSSstringECSS`) SHTML as spliced_str
+%     start <- "@{" spliced_str "}"
+%       EHTMLfn e => $proto_expr `SCSSR.Str %(ECSSeSCSS)ECSS`SHTML #\label{line:splicing-end}#
+%   EHTML|})
+% end #\label{line:rx_parse_fn_end}#
+% \end{lstlisting}
+% \vspace{-12px}
+% \caption{A grammar-based definition of \texttt{\$rx}.}
+% \vspace{-15px}
+% \label{fig:rx-grammar-based}
+% \end{figure}
+
+\subsection{Static Evaluation, Formally}
+It is not difficult to extend $\miniVerseParam$ to account for static evaluation.
+Static environments, $\Sigma$, take the form $\staticenv{\omega}{\uOmega}{\uPsi}{\uPhi}$, where $\omega$ is a substitution.
+Each binding form is annotated with a \emph{phase}, $\phi$, either $\staticphase$ or $\standardphase$.
+The rules for binding forms annotated with $\standardphase$ are essentially unchanged, differing only in that $\Sigma$ passes through opaquely.
+The rules for binding forms annotated with $\staticphase$ are based on the corresponding $\standardphase$ phase rules, differing only in that 1) they operate on $\Sigma$ and 2) evaluation occurs immediately.
+Finally, the forms for TSM definition are modified so that the parse function is now an unexpanded, rather than an expanded expression.
+The substitution $\omega$ is applied to the parse function after it is expanded.
+The full details are defined as a minor patch of $\miniVerseParam$ called $\miniVersePH$ in the supplement.
+
+\subsection{Library Management}
+In the examples above, and in our formal treatment, we explicitly qualified various definitions with the \li{static} keyword to make them available within static values.
+In practice, we would like to be able to use libraries within both static values and standard values as needed without duplicating code.
+This can be accomplished either by the package manager (e.g.
+SML/NJ's CM \cite{blume:smlnj-cm}, extended with phase annotations) or by allowing one to explicitly lower a new instance of a module imported into the standard phase for use also in  the static phase, as in a recent proposal for modular staging macros in OCaml \cite{Ocaml/macros}.
+We will leave these details as design decisions for the reader to consider.
+
+TSMs definitions can be exported by packages, but they cannot be exported from within modules because that would require that they also appear in signatures, and that, in turn, would complicate reasoning about signature equivalence, since TSM definitions contain arbitrary parse functions.
+It would also bring in confusion about whether the generated expansions can use knowledge about type identity available  within the module enclosing the TSM definition.
+That said, it should be possible to export TSM \emph{abbreviations} from modules, since they refer to TSM definitions only through symbolic names.
+We have not yet formalized this intuition.
+
+% For example, a language-external library manager for VerseML similar to SML/NJ's CM \cite{blume:smlnj-cm} could support a \li{static} qualifier on imported libraries, which would place the definitions exported by the imported library into the static phase of the library being defined.
+In particular, a library definition in such a compilation manager might look like this:
+% \begin{lstlisting}[numbers=none,morekeywords={Library,is}]
+% Library 
+%   (* ... exported module, signature and TSM names ... *)
+% is 
+%   (* ... files defining those exports ... *)
+
+%   (* imports: *)
+%   static parsegen.cm 
+% \end{lstlisting}
+
+% A similar approach could be taken for languages the incorporate library management directly into the syntax of programs, e.g. Scala \cite{odersky2008programming}:
+% \begin{lstlisting}[numbers=none]
+% static import edu.cmu.comar.parsegen
+% \end{lstlisting}
+
+% \begin{equation}\label{rule:mExpandsPH-syntaxpe-standard}
+% \inferrule{
+%   \tsmtyExpands{\uOmega}{\urho}{\rho}\\
+%   \Sigma = \staticenv{\omega}{\uOmega_S}{\uPsi_S}{\uPhi_S}\\\\
+%   \expandsP{\uOmega_S}{\uPsi_S}{\uPhi_S}{\ueparse}{\eparse}{\aparr{\tBody}{\tParseResultPCEExp}}\\\\
+%   \evalU{[\omega]\eparse}{\eparse'}\\
+%   \mExpandsPH{\uOmega}{\uAS{\uA \uplus \mapitem{\tsmv}{\adefref{a}}}{\Psi, \petsmdefn{a}{\rho}{\eparse'}}}{\uPhi}{\uM}{M}{\sigma}{\Sigma}
+% }{
+%   \mExpandsPH{\uOmega}{\uAS{\uA}{\Psi}}{\uPhi}{\defpetsmH{\standardphase}{\tsmv}{\urho}{\ueparse}{\uM}}{M}{\sigma}{\Sigma}
+% }
+% \end{equation}
+% \begin{equation}\label{rule:mExpandsPH-syntaxpe-static}
+% \inferrule{
+%   \tsmtyExpands{\uOmega}{\urho}{\rho}\\
+%   \Sigma = \staticenv{\omega}{\uOmega_S}{\uPsi_S}{\uPhi_S}\\
+%   \uPsi_S = \uAS{\uA_S}{\Psi_S}\\\\
+%   \expandsP{\uOmega_S}{\uPsi_S}{\uPhi_S}{\ueparse}{\eparse}{\aparr{\tBody}{\tParseResultPCEExp}}\\\\
+%   \evalU{[\omega]\eparse}{\eparse'}\\
+%   \mExpandsPH{\uOmega}{\uPsi}{\uPhi}{\uM}{M}{\sigma}{\staticenv{\omega}{\uOmega_S}{\uAS{\uA_S \uplus \mapitem{\tsmv}{\adefref{a}}}{\Psi_S, \petsmdefn{a}{\rho}{\eparse'}}}{\uPhi_S}}
+% }{
+%   \mExpandsPH{\uOmega}{\uPsi}{\uPhi}{\defpetsmH{\staticphase}{\tsmv}{\urho}{\ueparse}{\uM}}{M}{\sigma}{\Sigma}
+% }
+% \end{equation}
+
+\newcommand{\discussionSec}{Additional Related Work \& Concluding Discussion}
+\section{\protect\discussionSec}
+\label{sec:related-work}
+\label{sec:discussion}
+\label{sec:conclusion}
+\vspace{-3px}
+The most closely related work is that of \citet{TSLs} on \emph{type-specific languages} (TSLs), which introduced generalized literal forms and gave a bidirectionally typed protocol for type-directed parse function invocation.
+This work differs in several ways, and the differences we note below nicely summarize our specific contributions.
+
+Most obviously, TSMs are explicitly applied.
+Different TSMs can therefore be defined at the same type without conflict.
+Moreover, TSMs are not associated directly with generative type definitions, like TSLs,  but rather can operate at any type.
+In a subsequent short paper, \citet{sac15} proposed explicit application of simple expression TSMs in a bidirectional typed setting \cite{Pierce:2000:LTI:345099.345100}, but did not provide any formal details.
+We do not assume a bidirectionally typed language.
+Context independence implies that ML-style type inference can be performed using only the spliced terms and their types (because the remainder of an expansion cannot mention the very variables whose types are being inferred, nor constrain them further beyond what the annotations require.) We leave additional inference-related topics, e.g.
+TSM parameter inference, as future work.
+
+Another distinction is that the metatheory presented by \citet{TSLs} establishes only that generated expansions are well-typed.
+It does not establish the remaining abstract reasoning principles that have been a major focus of this paper.
+In particular, there is no formal notion of hygiene (though it is discussed informally), and the mechanism does not guarantee that a valid segmentation will exist, nor associate types with segments.
+
+Finally, the prior work on TSLs has not considered pattern matching, parameterized types, modules, parametric expansions or static evaluation.
+All of these are important for integration into an ML-like functional language.
+
+Another particularly compelling stream of related work is that by \citet{conf/icfp/LorenzenE13,conf/popl/LorenzenE16}, which describes SoundExt, a grammar-based syntax extension system where extension providers can equip their new forms with derived typing rules.
+The system then attempts to automatically verify that the expansion logic (expressed using a rewrite system, rather than an arbitrary function) is sound relative to these derived rules.
+
+Our work differs in several ways.
+First, we leave the context-free syntax fixed, so different TSMs cannot conflict.
+Second, SoundExt does not enforce hygiene, i.e.
+expansions might depend on the context and intentionally induce capture.
+Relatedly, TSMs do not support type-dependent expansion, as in \cite{conf/popl/LorenzenE16}.
+Another important distinction is that our system relies on (proto-)expansion validation, rather than verification.
+In other words, we do not prove properties about the parse function itself, but rather ``test'' each expansion.
+The trade-off is that our system does not require that parsing logic be written in a restricted manner, nor require a fully mechanized language definition and theorem proving machinery.
+Finally, it would be difficult to define syntax extensions in SoundExt that operate over module-parameterized families of types.
+Parameters can be approximated using splicing, but there is no clear notion of ``partial application''.
+
+% Other related work was already described in Sec. \ref{sec:intro}.\todo{reader macros}
+
+One important avenue has to do with automated refactoring.
+The unexpanded language does not come equipped with standard notions of renaming and substitution.
+Given a splice summary, it should be possible to ``backpatch'' refactorings into literal bodies, but the details are left as future work.
+
+%Another topic of interest has to do with intentional capture. 
+Haskell's \li{do}-notation for types with monadic structure cannot be expressed directly using TSMs because it introduces bindings into sub-terms.
+This might be an acceptable trade-off (indeed, that is our present position.) Alternatively, it might be worth exploring a design where spliced identifiers (but not arbitrary identifiers) can be designated in the splice summary as bound within spliced terms.
+
+At several points in the paper, we discussed editor integration.
+However, some important questions having to do with incremental parsing, error recovery and error reporting remain to be considered.
+Proto-expansions are context-independent and parametric, so they are likely to amenable to caching.
+
+Another interesting question is what this mechanism would look like in the setting of a projectional structure editor, i.e.
+one where the syntax is not textual but rather tree-shaped and projected in an interative manner to the programmer.
+This paper taken together with work by \citet{Omar:2012:ACC:2337223.2337324,DBLP:conf/popl/OmarVHAH17} can serve to guide such an effort.
+
+To conclude, TSMs give substantial syntactic control to library providers while leaving programmers with strong abstract reasoning principles.
+We believe TSMs therefore occupy a ``sweet spot'' in the  design space.
+%This paper serves to guide those who hope to incorporate TSMs into their future language designs.
 
@@
 
-%% Appendix
-\appendix
-\section{Appendix}
-
-Text of appendix \ldots
-
 \end{document}

### 2017-04-18 icfp17-rebuttal.txt
added ICFP17 reviews and draft rebuttal

Thank you for the thoughtful reviews. We were pleased to see that no significant technical problems were found during the review process, and that the reviewers found the technical approach to be thorough. The main doubts that seem to remain relate to (1) usefulness and (2) novelty.

On usefulness, Review B states:

"Although assessing the usefulness of a language design is difficult, reporting practical and substantial uses of the extension mechanism would go far in providing evidence that the language is useful."

We agree! So much so that we made just such a case for usefulness in our closely related work a few years ago (Omar et al, ECOOP 2014). That paper gave more sophisticated examples and even gathered empirical data from open source projects. The present paper is intended to be a follow-on "theory paper" that contributes more detailed formal results and develops various functional variants of the mechanism as was detailed in Sec. 7. Such a paper should be well within the scope of ICFP.

With regard to novelty, the first sentence of Sec. 1.3 ("Contributions") makes explicit the specific claims of our paper:

"This work introduces a system of typed syntax macros (TSMs) that gives library providers substantially more syntactic control than existing typed term-rewriting macro systems while maintaining the ability to reason abstractly about types, binding and segmentation."

In the remainder of the rebuttal, we will argue, in view of these specific claims, that the paper does in fact address all of the major concerns about novelty raised by the referees. We humbly ask that the scores assigned to our paper be reconsidered in this light.

== Review A ==

> "reminiscent of MacroML"

MacroML implements staging macros (stated on pg3). These are quite distinct from term-rewriting macros and TSMs in that they do not give the macro access to the syntax of the arguments at all (whether parsed or unparsed.) They cannot therefore be used as a mechanism for syntactic control -- staging macro invocation syntactically coincides with function application. Staging macros instead serve as a mechanism for staged evaluation, primarily for reasons of performance. This is unrelated to our stated goals.

> "reminiscent of TemplateHaskell in checking well-formedness after an extension's free-form generation of ASTs"

While Template Haskell does check that each expansion is well-typed (as any macro system in a statically typed language eventually must), there is no way for clients to reason abstractly about *what that type is* because there are no annotations on Template Haskell macros.

Furthermore, Template Haskell does not maintain any sort of hygienic binding discipline.

Being that Template Haskell achieves none of our central goals, and that there are a number of other languages that have broadly similar support for compile-time code generation, we neglected to mention it in Sec. 1.2 alongside, e.g. Scala's macro system, which is more disciplined. However, due to Template Haskell's popularity in the community, we agree that it would be worth mentioning explicitly.

> "The formalism to check hygiene seems novel, although with precedents in systems that synthesize hygiene (such as Lee et al., ECOOP'12)."

The hygiene mechanism in Lee et al (2012) is quite different, e.g. it sends queries at run-time to a compiler and reads back the error messages to determine which variables are free in an encoding of a term. No theoretical results whatsoever are established in that work.

> "The paper doesn't describe this parsing approach in one particular section. The description is partly at the bottom of page 4, and partly at the bottom of page 18."

Section 2 seems to be exactly the "one particular section" that is sought: it gives a thorough description of how typed expansion operates from start to finish.

One source of confusion may be that the paper only tersely describes how, before typed expansion, the initial context-free parsing of generalized literal forms works (see the paragraph immediately under Fig. 3, and Sec. 2's discussion of the body type.) There is nothing sophisticated about this step -- generalized literal forms are parsed like raw string literals. Our prior paper showed examples of various other delimiters and gave a context free grammar. Condition 4.1 (pg12) explains how we formally handle this first step.

The assertion that the description of the approach is "partly at the bottom of page 18" is incorrect: the bottom of Page 18 is the beginning of Sec. 5, which describes an *extension* of the core mechanism to handle parameterized families of types and module parameters. Nothing in Sec. 5 is necessary to understand prior sections, or the main conceptual points of the paper.

> "It's not immediately clear whether the segmentation should through colors (in figures 3 and 4 and footnote 2)
can be inferred without running the extension's implementation, but my current understanding is that it can't. Spelling out this design in once place and specifying the quoting convention (which isn't described, as far as I can tell) would improve the presentation."

The splice summary/segmentation/antiquotation mechanism, and how it relates to the colorings shown, is in fact spelled out in Sec. 2.2 (in particular, the final 3 paragraphs).

> "Along similar lines, the introduction provides considerable motivation for the parsing choice, but it's tangled with motivation for types, so drawing out the parsing strategy more explicitly could help."

By our reckoning, we gave little to no "motivation for types" in Sec. 1. Types are brought up briefly in the fifth point of "Problem 2" in Sec. 1.1, and in the third point in Sec. 1.2. Given that our research is focused on languages in the ML tradition, and our technical contributions do ultimately relate to types, we cannot eliminate statements about types from the introduction completely. We could perhaps mention that you can recover an "untyped" variant of TSMs by applying the usual type theorist's trick of defining just one type that classifies every well-scoped expression. The concerns about responsibility, segmentation, context independence and capture would still be relevant in that setting.

> "The paper's claim of a "sweet spot" seems to be supported only by the authors' opinions on how a language should behave."

The phrase "sweet spot" (which appears exclusively in quotes, and only at the end of the abstract and conclusion) is informal. That said, the phrase was chosen because TSMs sit between the well-studied extremes:

1. TSMs offer less syntactic control than syntax dialects, but stronger abstract reasoning principles (Sec. 1.1). The term "stronger" here is used in the technical sense -- clients can hold more information abstract when reasoning about types, binding and segmentation. The only opinion here is that abstract reasoning is itself a useful thing, but that is certainly a widely accepted belief within the ICFP community.

2. TSMs offer more syntactic control than term-rewriting macros and offer comparable reasoning principles (Sec. 1.2). Again, this is an objective claim. TSMs have full control over parsing the literal body, whereas term-rewriting macros are given an already-parsed term. The reasoning principles are theorems.

== Review B ==

> "The work is well presented and technically sound; my primary concern is on usability particularly on composable syntaxes and complex situations where two syntaxes might use the same symbols for starting/internal workings and the conflict might be confusing to users of both."

TSMs can be applied within spliced terms without conflict (e.g. see the footnote on page 4, the discussion about final expansion in Sec. 2, and also many more complex examples in Omar et al, 2014).

It is not clear to us what is meant here by "two syntaxes might use the same symbols for starting/internal workings". This concern is perhaps related to the discussion in the paper on "Responsibility" -- in which case note that the applied TSM unambiguously identifies which syntax is in use within each literal form. Of course, different TSMs might implement similar looking syntactic forms, but any semantic confusion that results can be resolved by carefully following the reasoning principles detailed in the paper. Ultimately, one cannot enforce "good taste" by purely technical means.

== Review C ==

> "However, the system presented is not novel in the ways the authors suggest. Most significantly, it is very similar to [...] the camlp4 metaprogramming system"

camlp4 was cited in Sec. 1.1 and suffers from *all* of the problems described therein (conflict + no way to reason about responsibility, segmentation, capture, context dependence or typing.)

> "Considering GHC quasiquotes, they also feature a single fixed delimiter, arbitrary parsers, explicit invocation, and typing. Because Template Haskell, on which quasiquotation is built, is more powerful than the system presented here, they do not have context independence or capture properties, but a system without those particular features would would have the relevant properties."

GHC quasiquotes address only the problems of conflict and responsibility. The use of fixed outer delimiters to handle conflict is explicitly disclaimed as a novel contribution at the top of page 4, where we discuss strings and the work of Slind 1991; also Omar et al, 2014. The fact that responsibility is handled by explicit invocation is also clearly attributed to existing macro systems in Sec. 1.2 (point 1).

The fact that unhygienic macro systems, like Template Haskell / GHC quasiquotes, can arbitrarily parse terms out of string literal bodies is explicitly acknowledged in Sec. 1.2 (second paragraph on p4). The problem, to reiterate, is (1) that no segmentation principles are available and (2) that this approach is incompatible with hygiene constraints that are necessary to "reason abstractly about ... binding". If you added standard hygiene constraints to Template Haskell, it would no longer be able to support arbitrary parsers for the reasons discussed on page 4. With regard to typing, see the response to Review A above.

TSMs are novel not for each feature individually, but because you need not make trade-offs between these features.

> "Segmentation is, in the present system, an out-of-band communication to the editor system, and could be so adapted for any extensible syntax system."

To be able to extract an accurate segmentation from an expansion, the origin of each spliced term in the expansion must be retained. A standard representation of ASTs (without the Spliced constructors discussed in Sec. 2, which are novel) would not retain this information because an arbitrary function sits between the input string and the output AST.

> "Additionally, the idea of directly generating an AST, rather than a term to be further expanded, and using this to build DSLs, goes back at least to Shriram Krishnamurthi's thesis (see the concept of "micros").

Direct AST generation emerged as the "default" solution in our mechanism in that adding support for TSM application within proto-expansions would have required significant additional machinery. Krishnamurthi's work on micros is focused on term-to-term translation from the source language to an IR, and macro expansion in the generated translation does eventually occur.

> "The approach to hygiene, which uses an auxiliary environment which maps symbols (here called identifiers) to variables, is familiar from the literature on hygenic macros in the Scheme community (this is known as a "renaming" in that literature; see for example [Dybvig 91]). That this approach enforces hygiene is proved for a significantly more expressive macro system by [Herman and Wand]."

Our mechanism for identifier expansion does not operate quite like both Dybvig91 and Herman and Wand (2008, 2010) describe -- in our work, renaming is *implicitly* necessary to discharge disjointness premises in the typed expansion and proto-expansion validation rules we give (see Harper, 2012, Chapters 1-2 on abstract binding trees, and Gabbay and Pitts 2001). Relying on disjointness premises and pushing the act of renaming into the more general ABT mechanism via an implicit identification convention a la Harper significantly simplifies our formalism relative to approaches that require explicitly specifying how unique names are generated during expansion. It also allows us to state the capture avoidance conditions (in Theorems 4.5 and 5.1) in terms of the standard capture-avoiding substitution operation on ABTs. To our knowledge, this ABT-based approach has not been explored before.

It is not clear what is meant here by "more expressive macro system". Herman and Wand's system, like other hygienic term-rewriting systems, does not support parsing terms out from string arguments, as already discussed. There are some similarities between our spliced segment references and the "tree locations" used in their work (which originate in Gorn, 1967), in that tree locations identify sub-terms by the location they appear in the argument list, rather than more generally in our system by their syntactic location in the literal body.

> "The parameterization is a somewhat odd feature. In the paper, it seems to mostly address the overly-strong hygiene restriction, by allowing the macro to generate references to identifiers from the module signature that is parameterized over. However, this seems a round-about way of recovering what is essentially lexical scope: macros should be able to reference names in their environment, and those references should be stable when expanded. The parameterization is independently useful, however, as the DICT syntax example shows."

This sort of lexical scope is not technically possible for macros that operate on non-primitive AST representations, like proto_expr, because variable references in the expansion are merely pieces of data of type var_t (which could very well just be a synonym for string or nat if using a de Bruijn-style encoding.) These values could result from arbitrary computations. Parameterization solves this problem without needing to privilege term representations primitively. It also makes it clear which bindings are going to be accessed by the computed expansion without examining the implementation directly, which is in following with our goal of making it easy to reason abstractly about binding.

"In this setting, the paper follows [Culpepper, Owens, Flatt; GPCE 2005] very closely (see also Owens' thesis)."

The work by Culpepper, Owens and Flatt, and by Owens in his thesis, on packaging macros into components/units is more closely related to the brief discussion in Sec. 6.4 (second paragraph) than to Sec. 5, and we will cite it there. Sec. 5 would correspond to parameterizing macros over units of a particular signature, which was not considered in their prior work.

> "The claims of the "Syntax Dialects are Unreasonable" section seems to ignore the long history of success of the Lisp community in this area."

It is not clear what specific efforts in the Lisp community this is referring to. Lisp has had notable success with term-rewriting macros, as discussed in Sec. 1.2. However, a Lisp macro given a string literal would have exactly the same problems as the Rust example described in that section. There have been some efforts to support arbitrary (non-S-expression) syntax, e.g. with reader macros / Racket's #lang mechanism, but these amount to syntax definition systems that suffer from some or all of the problems described in Sec. 1.1.

> "p3:44 the systems listed here are not all "typed" in the same sense. Herman and Wand's types, for example, describe the scoping of the resulting program, not its type."

Yes, you're right. That point wasn't stated as clearly as it could have been.

> "Would splices be more simply represented by providing an opaque type of strings with a "substring" operation?"

One way to guarantee that the splices are all in-bounds would indeed be to package together body and proto_exp as abstract types and allow the Spliced constructors to arise internally only via a checked substring operation on the provided body. A related mechanism was used in Omar et al, 2014. However, this would not guarantee that segments are non-overlapping so a check by the semantics would still be needed. Simply defining body as a string and including the bounds checks in along with the other checks is semantically simpler. Clients who favor the abstract approach can choose to use it, since parse functions are ML code.

> "p7:41 this type should be `string`, not `html`"

Yes, thanks.

> "The discussion of `do` notation on p8 is confused. `do` notation, and generation of binding forms more generally, is not prohibited here by capture-avoidance, but because (I infer, the full AST is not provided) there's no way to take a splice and put it in the binding position of a `lam` AST."

This possibility is mentioned at p24:37. Shape types ala Herman and Wand would be related to this mechanism. Absent such a mechanism, it is in fact capture avoidance that prevents capture by spliced terms of variables bound in the expansion.

* "p10:32 how would you address capture in a system which did feature binding in patterns that could be subsequently referenced?"

This would be addressed exactly as it is with expression TSMs.

> "p24:34 This "backpatching" is done in many Lisp IDEs, see DrRacket
  for an example [Florence et al, Language Workbench 2016]"

I will mention that in the future work, thanks.

### 2017-04-18 icfp17-rebuttal.txt
word wrap
tweaks to pp1

@@
-Thank you for the thoughtful reviews.
-We were pleased to see that no significant technical problems were found during the review process, and that the reviewers found the technical approach to be thorough.
-The main doubts that seem to remain relate to (1) usefulness and (2) novelty.
+Thank you for the thoughtful reviews. We were pleased to see that no
+significant technical problems were found during the review process. The main
+doubts that seem to remain relate to (1) usefulness and (2) novelty.
 
-On usefulness, Review B states:
+With regard to usefulness, Review B states:
 
-"Although assessing the usefulness of a language design is difficult, reporting practical and substantial uses of the extension mechanism would go far in providing evidence that the language is useful."
+"Although assessing the usefulness of a language design is difficult, reporting
+practical and substantial uses of the extension mechanism would go far in
+providing evidence that the language is useful."
 
-We agree!
-So much so that we made just such a case for usefulness in our closely related work a few years ago (Omar et al, ECOOP 2014).
-That paper gave more sophisticated examples and even gathered empirical data from open source projects.
-The present paper is intended to be a follow-on "theory paper" that contributes more detailed formal results and develops various functional variants of the mechanism as was detailed in Sec.
-7.
-Such a paper should be well within the scope of ICFP.
+We agree! So much so that we made just such a case for usefulness in our
+closely related work a few years ago (Omar et al, ECOOP 2014). That paper gave
+more sophisticated examples and even gathered empirical data from open source
+projects. The present paper is intended to be a follow-on "theory paper" that
+contributes more detailed formal results and develops various functional
+variants of the mechanism as was detailed in Sec. 7. Such a paper should be
+well within the scope of ICFP.
 
-With regard to novelty, the first sentence of Sec. 1.3 ("Contributions") makes explicit the specific claims of our paper:
+With regard to novelty, the first sentence of Sec. 1.3 ("Contributions") makes
+explicit the specific claims of our paper:
 
-"This work introduces a system of typed syntax macros (TSMs) that gives library providers substantially more syntactic control than existing typed term-rewriting macro systems while maintaining the ability to reason abstractly about types, binding and segmentation."
+"This work introduces a system of typed syntax macros (TSMs) that gives library
+providers substantially more syntactic control than existing typed
+term-rewriting macro systems while maintaining the ability to reason abstractly
+about types, binding and segmentation."
 
-In the remainder of the rebuttal, we will argue, in view of these specific claims, that the paper does in fact address all of the major concerns about novelty raised by the referees.
-We humbly ask that the scores assigned to our paper be reconsidered in this light.
+In the remainder of the rebuttal, we will argue, in view of these specific
+claims, that the paper does in fact address all of the major concerns about
+novelty raised by the referees. We humbly ask that the scores assigned to our
+paper be reconsidered in this light.
 
@@
 
-MacroML implements staging macros (stated on pg3).
-These are quite distinct from term-rewriting macros and TSMs in that they do not give the macro access to the syntax of the arguments at all (whether parsed or unparsed.) They cannot therefore be used as a mechanism for syntactic control -- staging macro invocation syntactically coincides with function application.
-Staging macros instead serve as a mechanism for staged evaluation, primarily for reasons of performance.
-This is unrelated to our stated goals.
+MacroML implements staging macros (stated on pg3). These are quite distinct
+from term-rewriting macros and TSMs in that they do not give the macro access
+to the syntax of the arguments at all (whether parsed or unparsed.) They cannot
+therefore be used as a mechanism for syntactic control -- staging macro
+invocation syntactically coincides with function application. Staging macros
+instead serve as a mechanism for staged evaluation, primarily for reasons of
+performance. This is unrelated to our stated goals.
 
-> "reminiscent of TemplateHaskell in checking well-formedness after an extension's free-form generation of ASTs"
+> "reminiscent of TemplateHaskell in checking well-formedness after an
+extension's free-form generation of ASTs"
 
-While Template Haskell does check that each expansion is well-typed (as any macro system in a statically typed language eventually must), there is no way for clients to reason abstractly about *what that type is* because there are no annotations on Template Haskell macros.
+While Template Haskell does check that each expansion is well-typed (as any
+macro system in a statically typed language eventually must), there is no way
+for clients to reason abstractly about *what that type is* because there are no
+	annotations on Template Haskell macros.
 
-Furthermore, Template Haskell does not maintain any sort of hygienic binding discipline.
+Furthermore, Template Haskell does not maintain any sort of hygienic binding
+discipline.
 
-Being that Template Haskell achieves none of our central goals, and that there are a number of other languages that have broadly similar support for compile-time code generation, we neglected to mention it in Sec.
-1.2 alongside, e.g.
-Scala's macro system, which is more disciplined.
-However, due to Template Haskell's popularity in the community, we agree that it would be worth mentioning explicitly.
+Being that Template Haskell achieves none of our central goals, and that there
+are a number of other languages that have broadly similar support for
+compile-time code generation, we neglected to mention it in Sec. 1.2 alongside,
+e.g. Scala's macro system, which is more disciplined. However, due to Template
+Haskell's popularity in the community, we agree that it would be worth
+mentioning explicitly.
 
-> "The formalism to check hygiene seems novel, although with precedents in systems that synthesize hygiene (such as Lee et al., ECOOP'12)."
+> "The formalism to check hygiene seems novel, although with precedents in
+systems that synthesize hygiene (such as Lee et al., ECOOP'12)."
 
-The hygiene mechanism in Lee et al (2012) is quite different, e.g.
-it sends queries at run-time to a compiler and reads back the error messages to determine which variables are free in an encoding of a term.
-No theoretical results whatsoever are established in that work.
+The hygiene mechanism in Lee et al (2012) is quite different, e.g. it sends
+queries at run-time to a compiler and reads back the error messages to
+determine which variables are free in an encoding of a term. No theoretical
+results whatsoever are established in that work.
 
-> "The paper doesn't describe this parsing approach in one particular section. The description is partly at the bottom of page 4, and partly at the bottom of page 18."
+> "The paper doesn't describe this parsing approach in one particular section.
+The description is partly at the bottom of page 4, and partly at the bottom of
+page 18."
 
-Section 2 seems to be exactly the "one particular section" that is sought: it gives a thorough description of how typed expansion operates from start to finish.
+Section 2 seems to be exactly the "one particular section" that is sought: it
+gives a thorough description of how typed expansion operates from start to
+finish.
 
-One source of confusion may be that the paper only tersely describes how, before typed expansion, the initial context-free parsing of generalized literal forms works (see the paragraph immediately under Fig.
-3, and Sec.
-2's discussion of the body type.) There is nothing sophisticated about this step -- generalized literal forms are parsed like raw string literals.
-Our prior paper showed examples of various other delimiters and gave a context free grammar.
+One source of confusion may be that the paper only tersely describes how,
+before typed expansion, the initial context-free parsing of generalized literal
+forms works (see the paragraph immediately under Fig. 3, and Sec. 2's
+discussion of the body type.) There is nothing sophisticated about this step --
+generalized literal forms are parsed like raw string literals. Our prior paper
+showed examples of various other delimiters and gave a context free grammar.
 Condition 4.1 (pg12) explains how we formally handle this first step.
 
-The assertion that the description of the approach is "partly at the bottom of page 18" is incorrect: the bottom of Page 18 is the beginning of Sec.
-5, which describes an *extension* of the core mechanism to handle parameterized families of types and module parameters.
-Nothing in Sec.
-5 is necessary to understand prior sections, or the main conceptual points of the paper.
+The assertion that the description of the approach is "partly at the bottom of
+page 18" is incorrect: the bottom of Page 18 is the beginning of Sec. 5, which
+describes an *extension* of the core mechanism to handle parameterized families
+of types and module parameters. Nothing in Sec. 5 is necessary to understand
+prior sections, or the main conceptual points of the paper.
 
-> "It's not immediately clear whether the segmentation should through colors (in figures 3 and 4 and footnote 2)
-can be inferred without running the extension's implementation, but my current understanding is that it can't.
-Spelling out this design in once place and specifying the quoting convention (which isn't described, as far as I can tell) would improve the presentation."
+> "It's not immediately clear whether the segmentation should through colors
+(in figures 3 and 4 and footnote 2) can be inferred without running the
+extension's implementation, but my current understanding is that it can't.
+Spelling out this design in once place and specifying the quoting convention
+(which isn't described, as far as I can tell) would improve the presentation."
 
-The splice summary/segmentation/antiquotation mechanism, and how it relates to the colorings shown, is in fact spelled out in Sec. 2.2 (in particular, the final 3 paragraphs).
+The splice summary/segmentation/antiquotation mechanism, and how it relates to
+the colorings shown, is in fact spelled out in Sec. 2.2 (in particular, the
+final 3 paragraphs).
 
-> "Along similar lines, the introduction provides considerable motivation for the parsing choice, but it's tangled with motivation for types, so drawing out the parsing strategy more explicitly could help."
+> "Along similar lines, the introduction provides considerable motivation for
+the parsing choice, but it's tangled with motivation for types, so drawing out
+the parsing strategy more explicitly could help."
 
-By our reckoning, we gave little to no "motivation for types" in Sec.
-1.
-Types are brought up briefly in the fifth point of "Problem 2" in Sec.
-1.1, and in the third point in Sec.
-1.2.
-Given that our research is focused on languages in the ML tradition, and our technical contributions do ultimately relate to types, we cannot eliminate statements about types from the introduction completely.
-We could perhaps mention that you can recover an "untyped" variant of TSMs by applying the usual type theorist's trick of defining just one type that classifies every well-scoped expression.
-The concerns about responsibility, segmentation, context independence and capture would still be relevant in that setting.
+By our reckoning, we gave little to no "motivation for types" in Sec. 1. Types
+are brought up briefly in the fifth point of "Problem 2" in Sec. 1.1, and in
+the third point in Sec. 1.2. Given that our research is focused on languages in
+the ML tradition, and our technical contributions do ultimately relate to
+types, we cannot eliminate statements about types from the introduction
+completely. We could perhaps mention that you can recover an "untyped" variant
+of TSMs by applying the usual type theorist's trick of defining just one type
+that classifies every well-scoped expression. The concerns about
+responsibility, segmentation, context independence and capture would still be
+relevant in that setting.
 
-> "The paper's claim of a "sweet spot" seems to be supported only by the authors' opinions on how a language should behave."
+> "The paper's claim of a "sweet spot" seems to be supported only by the
+authors' opinions on how a language should behave."
 
-The phrase "sweet spot" (which appears exclusively in quotes, and only at the end of the abstract and conclusion) is informal.
-That said, the phrase was chosen because TSMs sit between the well-studied extremes:
+The phrase "sweet spot" (which appears exclusively in quotes, and only at the
+end of the abstract and conclusion) is informal. That said, the phrase was
+chosen because TSMs sit between the well-studied extremes:
 
-1.
-TSMs offer less syntactic control than syntax dialects, but stronger abstract reasoning principles (Sec.
-1.1).
-The term "stronger" here is used in the technical sense -- clients can hold more information abstract when reasoning about types, binding and segmentation.
-The only opinion here is that abstract reasoning is itself a useful thing, but that is certainly a widely accepted belief within the ICFP community.
+1. TSMs offer less syntactic control than syntax dialects, but stronger
+abstract reasoning principles (Sec. 1.1). The term "stronger" here is used in
+the technical sense -- clients can hold more information abstract when
+reasoning about types, binding and segmentation. The only opinion here is that
+abstract reasoning is itself a useful thing, but that is certainly a widely
+accepted belief within the ICFP community.
 
-2.
-TSMs offer more syntactic control than term-rewriting macros and offer comparable reasoning principles (Sec.
-1.2).
-Again, this is an objective claim.
-TSMs have full control over parsing the literal body, whereas term-rewriting macros are given an already-parsed term.
-The reasoning principles are theorems.
+2. TSMs offer more syntactic control than term-rewriting macros and offer
+comparable reasoning principles (Sec. 1.2). Again, this is an objective claim.
+TSMs have full control over parsing the literal body, whereas term-rewriting
+macros are given an already-parsed term. The reasoning principles are theorems.
 
@@
 
-> "The work is well presented and technically sound; my primary concern is on usability particularly on composable syntaxes and complex situations where two syntaxes might use the same symbols for starting/internal workings and the conflict might be confusing to users of both."
+> "The work is well presented and technically sound; my primary concern is on
+usability particularly on composable syntaxes and complex situations where two
+syntaxes might use the same symbols for starting/internal workings and the
+conflict might be confusing to users of both."
 
-TSMs can be applied within spliced terms without conflict (e.g. see the footnote on page 4, the discussion about final expansion in Sec. 2, and also many more complex examples in Omar et al, 2014).
+TSMs can be applied within spliced terms without conflict (e.g. see the
+footnote on page 4, the discussion about final expansion in Sec. 2, and also
+many more complex examples in Omar et al, 2014).
 
-It is not clear to us what is meant here by "two syntaxes might use the same symbols for starting/internal workings".
-This concern is perhaps related to the discussion in the paper on "Responsibility" -- in which case note that the applied TSM unambiguously identifies which syntax is in use within each literal form.
-Of course, different TSMs might implement similar looking syntactic forms, but any semantic confusion that results can be resolved by carefully following the reasoning principles detailed in the paper.
-Ultimately, one cannot enforce "good taste" by purely technical means.
+It is not clear to us what is meant here by "two syntaxes might use the same
+symbols for starting/internal workings". This concern is perhaps related to the
+discussion in the paper on "Responsibility" -- in which case note that the
+applied TSM unambiguously identifies which syntax is in use within each literal
+form. Of course, different TSMs might implement similar looking syntactic
+forms, but any semantic confusion that results can be resolved by carefully
+following the reasoning principles detailed in the paper. Ultimately, one
+cannot enforce "good taste" by purely technical means.
 
@@
 
-> "However, the system presented is not novel in the ways the authors suggest. Most significantly, it is very similar to [...] the camlp4 metaprogramming system"
+> "However, the system presented is not novel in the ways the authors suggest.
+Most significantly, it is very similar to [...] the camlp4 metaprogramming
+system"
 
-camlp4 was cited in Sec. 1.1 and suffers from *all* of the problems described therein (conflict + no way to reason about responsibility, segmentation, capture, context dependence or typing.)
+camlp4 was cited in Sec. 1.1 and suffers from *all* of the problems described
+therein (conflict + no way to reason about responsibility, segmentation,
+capture, context dependence or typing.)
 
-> "Considering GHC quasiquotes, they also feature a single fixed delimiter, arbitrary parsers, explicit invocation, and typing.
-Because Template Haskell, on which quasiquotation is built, is more powerful than the system presented here, they do not have context independence or capture properties, but a system without those particular features would would have the relevant properties."
+> "Considering GHC quasiquotes, they also feature a single fixed delimiter,
+arbitrary parsers, explicit invocation, and typing. Because Template Haskell,
+on which quasiquotation is built, is more powerful than the system presented
+here, they do not have context independence or capture properties, but a system
+without those particular features would would have the relevant properties."
 
-GHC quasiquotes address only the problems of conflict and responsibility.
-The use of fixed outer delimiters to handle conflict is explicitly disclaimed as a novel contribution at the top of page 4, where we discuss strings and the work of Slind 1991; also Omar et al, 2014.
-The fact that responsibility is handled by explicit invocation is also clearly attributed to existing macro systems in Sec.
-1.2 (point 1).
+GHC quasiquotes address only the problems of conflict and responsibility. The
+use of fixed outer delimiters to handle conflict is explicitly disclaimed as a
+novel contribution at the top of page 4, where we discuss strings and the work
+of Slind 1991; also Omar et al, 2014. The fact that responsibility is handled
+by explicit invocation is also clearly attributed to existing macro systems in
+Sec. 1.2 (point 1).
 
-The fact that unhygienic macro systems, like Template Haskell / GHC quasiquotes, can arbitrarily parse terms out of string literal bodies is explicitly acknowledged in Sec.
-1.2 (second paragraph on p4).
-The problem, to reiterate, is (1) that no segmentation principles are available and (2) that this approach is incompatible with hygiene constraints that are necessary to "reason abstractly about ...
-binding".
-If you added standard hygiene constraints to Template Haskell, it would no longer be able to support arbitrary parsers for the reasons discussed on page 4.
-With regard to typing, see the response to Review A above.
+The fact that unhygienic macro systems, like Template Haskell / GHC
+quasiquotes, can arbitrarily parse terms out of string literal bodies is
+explicitly acknowledged in Sec. 1.2 (second paragraph on p4). The problem, to
+reiterate, is (1) that no segmentation principles are available and (2) that
+this approach is incompatible with hygiene constraints that are necessary to
+"reason abstractly about ... binding". If you added standard hygiene
+constraints to Template Haskell, it would no longer be able to support
+arbitrary parsers for the reasons discussed on page 4. With regard to typing,
+see the response to Review A above.
 
-TSMs are novel not for each feature individually, but because you need not make trade-offs between these features.
+TSMs are novel not for each feature individually, but because you need not make
+trade-offs between these features.
 
-> "Segmentation is, in the present system, an out-of-band communication to the editor system, and could be so adapted for any extensible syntax system."
+> "Segmentation is, in the present system, an out-of-band communication to the
+editor system, and could be so adapted for any extensible syntax system."
 
-To be able to extract an accurate segmentation from an expansion, the origin of each spliced term in the expansion must be retained.
-A standard representation of ASTs (without the Spliced constructors discussed in Sec.
-2, which are novel) would not retain this information because an arbitrary function sits between the input string and the output AST.
+To be able to extract an accurate segmentation from an expansion, the origin of
+each spliced term in the expansion must be retained. A standard representation
+of ASTs (without the Spliced constructors discussed in Sec. 2, which are novel)
+would not retain this information because an arbitrary function sits between
+the input string and the output AST.
 
-> "Additionally, the idea of directly generating an AST, rather than a term to be further expanded, and using this to build DSLs, goes back at least to Shriram Krishnamurthi's thesis (see the concept of "micros").
+> "Additionally, the idea of directly generating an AST, rather than a term to
+be further expanded, and using this to build DSLs, goes back at least to
+Shriram Krishnamurthi's thesis (see the concept of "micros").
 
-Direct AST generation emerged as the "default" solution in our mechanism in that adding support for TSM application within proto-expansions would have required significant additional machinery.
-Krishnamurthi's work on micros is focused on term-to-term translation from the source language to an IR, and macro expansion in the generated translation does eventually occur.
+Direct AST generation emerged as the "default" solution in our mechanism in
+that adding support for TSM application within proto-expansions would have
+required significant additional machinery. Krishnamurthi's work on micros is
+focused on term-to-term translation from the source language to an IR, and
+macro expansion in the generated translation does eventually occur.
 
-> "The approach to hygiene, which uses an auxiliary environment which maps symbols (here called identifiers) to variables, is familiar from the literature on hygenic macros in the Scheme community (this is known as a "renaming" in that literature; see for example [Dybvig 91]).
-That this approach enforces hygiene is proved for a significantly more expressive macro system by [Herman and Wand]."
+> "The approach to hygiene, which uses an auxiliary environment which maps
+symbols (here called identifiers) to variables, is familiar from the literature
+on hygenic macros in the Scheme community (this is known as a "renaming" in
+that literature; see for example [Dybvig 91]). That this approach enforces
+hygiene is proved for a significantly more expressive macro system by [Herman
+and Wand]."
 
-Our mechanism for identifier expansion does not operate quite like both Dybvig91 and Herman and Wand (2008, 2010) describe -- in our work, renaming is *implicitly* necessary to discharge disjointness premises in the typed expansion and proto-expansion validation rules we give (see Harper, 2012, Chapters 1-2 on abstract binding trees, and Gabbay and Pitts 2001).
-Relying on disjointness premises and pushing the act of renaming into the more general ABT mechanism via an implicit identification convention a la Harper significantly simplifies our formalism relative to approaches that require explicitly specifying how unique names are generated during expansion.
-It also allows us to state the capture avoidance conditions (in Theorems 4.5 and 5.1) in terms of the standard capture-avoiding substitution operation on ABTs.
-To our knowledge, this ABT-based approach has not been explored before.
+Our mechanism for identifier expansion does not operate quite like both
+Dybvig91 and Herman and Wand (2008, 2010) describe -- in our work, renaming is
+*implicitly* necessary to discharge disjointness premises in the typed
+expansion and proto-expansion validation rules we give (see Harper, 2012,
+Chapters 1-2 on abstract binding trees, and Gabbay and Pitts 2001). Relying on
+disjointness premises and pushing the act of renaming into the more general ABT
+mechanism via an implicit identification convention a la Harper significantly
+simplifies our formalism relative to approaches that require explicitly
+specifying how unique names are generated during expansion. It also allows us
+to state the capture avoidance conditions (in Theorems 4.5 and 5.1) in terms of
+the standard capture-avoiding substitution operation on ABTs. To our knowledge,
+this ABT-based approach has not been explored before.
 
-It is not clear what is meant here by "more expressive macro system".
-Herman and Wand's system, like other hygienic term-rewriting systems, does not support parsing terms out from string arguments, as already discussed.
-There are some similarities between our spliced segment references and the "tree locations" used in their work (which originate in Gorn, 1967), in that tree locations identify sub-terms by the location they appear in the argument list, rather than more generally in our system by their syntactic location in the literal body.
+It is not clear what is meant here by "more expressive macro system". Herman
+and Wand's system, like other hygienic term-rewriting systems, does not support
+parsing terms out from string arguments, as already discussed. There are some
+similarities between our spliced segment references and the "tree locations"
+used in their work (which originate in Gorn, 1967), in that tree locations
+identify sub-terms by the location they appear in the argument list, rather
+than more generally in our system by their syntactic location in the literal
+body.
 
-> "The parameterization is a somewhat odd feature.
-In the paper, it seems to mostly address the overly-strong hygiene restriction, by allowing the macro to generate references to identifiers from the module signature that is parameterized over.
-However, this seems a round-about way of recovering what is essentially lexical scope: macros should be able to reference names in their environment, and those references should be stable when expanded.
-The parameterization is independently useful, however, as the DICT syntax example shows."
+> "The parameterization is a somewhat odd feature. In the paper, it seems to
+mostly address the overly-strong hygiene restriction, by allowing the macro to
+generate references to identifiers from the module signature that is
+parameterized over. However, this seems a round-about way of recovering what is
+essentially lexical scope: macros should be able to reference names in their
+environment, and those references should be stable when expanded. The
+parameterization is independently useful, however, as the DICT syntax example
+shows."
 
-This sort of lexical scope is not technically possible for macros that operate on non-primitive AST representations, like proto_expr, because variable references in the expansion are merely pieces of data of type var_t (which could very well just be a synonym for string or nat if using a de Bruijn-style encoding.) These values could result from arbitrary computations.
-Parameterization solves this problem without needing to privilege term representations primitively.
-It also makes it clear which bindings are going to be accessed by the computed expansion without examining the implementation directly, which is in following with our goal of making it easy to reason abstractly about binding.
+This sort of lexical scope is not technically possible for macros that operate
+on non-primitive AST representations, like proto_expr, because variable
+references in the expansion are merely pieces of data of type var_t (which
+could very well just be a synonym for string or nat if using a de Bruijn-style
+encoding.) These values could result from arbitrary computations.
+Parameterization solves this problem without needing to privilege term
+representations primitively. It also makes it clear which bindings are going to
+be accessed by the computed expansion without examining the implementation
+directly, which is in following with our goal of making it easy to reason
+abstractly about binding.
 
-"In this setting, the paper follows [Culpepper, Owens, Flatt; GPCE 2005] very closely (see also Owens' thesis)."
+"In this setting, the paper follows [Culpepper, Owens, Flatt; GPCE 2005] very
+closely (see also Owens' thesis)."
 
-The work by Culpepper, Owens and Flatt, and by Owens in his thesis, on packaging macros into components/units is more closely related to the brief discussion in Sec.
-6.4 (second paragraph) than to Sec.
-5, and we will cite it there.
-Sec.
-5 would correspond to parameterizing macros over units of a particular signature, which was not considered in their prior work.
+The work by Culpepper, Owens and Flatt, and by Owens in his thesis, on
+packaging macros into components/units is more closely related to the brief
+discussion in Sec. 6.4 (second paragraph) than to Sec. 5, and we will cite it
+there. Sec. 5 would correspond to parameterizing macros over units of a
+particular signature, which was not considered in their prior work.
 
-> "The claims of the "Syntax Dialects are Unreasonable" section seems to ignore the long history of success of the Lisp community in this area."
+> "The claims of the "Syntax Dialects are Unreasonable" section seems to ignore
+the long history of success of the Lisp community in this area."
 
-It is not clear what specific efforts in the Lisp community this is referring to.
-Lisp has had notable success with term-rewriting macros, as discussed in Sec.
-1.2.
-However, a Lisp macro given a string literal would have exactly the same problems as the Rust example described in that section.
-There have been some efforts to support arbitrary (non-S-expression) syntax, e.g.
-with reader macros / Racket's #lang mechanism, but these amount to syntax definition systems that suffer from some or all of the problems described in Sec.
-1.1.
+It is not clear what specific efforts in the Lisp community this is referring
+to. Lisp has had notable success with term-rewriting macros, as discussed in
+Sec. 1.2. However, a Lisp macro given a string literal would have exactly the
+same problems as the Rust example described in that section. There have been
+some efforts to support arbitrary (non-S-expression) syntax, e.g. with reader
+macros / Racket's #lang mechanism, but these amount to syntax definition
+systems that suffer from some or all of the problems described in Sec. 1.1.
 
-> "p3:44 the systems listed here are not all "typed" in the same sense. Herman and Wand's types, for example, describe the scoping of the resulting program, not its type."
+> "p3:44 the systems listed here are not all "typed" in the same sense. Herman
+and Wand's types, for example, describe the scoping of the resulting program,
+not its type."
 
@@
 
-> "Would splices be more simply represented by providing an opaque type of strings with a "substring" operation?"
+> "Would splices be more simply represented by providing an opaque type of
+strings with a "substring" operation?"
 
-One way to guarantee that the splices are all in-bounds would indeed be to package together body and proto_exp as abstract types and allow the Spliced constructors to arise internally only via a checked substring operation on the provided body.
-A related mechanism was used in Omar et al, 2014.
-However, this would not guarantee that segments are non-overlapping so a check by the semantics would still be needed.
-Simply defining body as a string and including the bounds checks in along with the other checks is semantically simpler.
-Clients who favor the abstract approach can choose to use it, since parse functions are ML code.
+One way to guarantee that the splices are all in-bounds would indeed be to
+package together body and proto_exp as abstract types and allow the Spliced
+constructors to arise internally only via a checked substring operation on the
+provided body. A related mechanism was used in Omar et al, 2014. However, this
+would not guarantee that segments are non-overlapping so a check by the
+semantics would still be needed. Simply defining body as a string and including
+the bounds checks in along with the other checks is semantically simpler.
+Clients who favor the abstract approach can choose to use it, since parse
+functions are ML code.
 
@@
 
-> "The discussion of `do` notation on p8 is confused.
-`do` notation, and generation of binding forms more generally, is not prohibited here by capture-avoidance, but because (I infer, the full AST is not provided) there's no way to take a splice and put it in the binding position of a `lam` AST."
+> "The discussion of `do` notation on p8 is confused. `do` notation, and
+generation of binding forms more generally, is not prohibited here by
+capture-avoidance, but because (I infer, the full AST is not provided) there's
+no way to take a splice and put it in the binding position of a `lam` AST."
 
-This possibility is mentioned at p24:37.
-Shape types ala Herman and Wand would be related to this mechanism.
-Absent such a mechanism, it is in fact capture avoidance that prevents capture by spliced terms of variables bound in the expansion.
+This possibility is mentioned at p24:37. Shape types ala Herman and Wand would
+be related to this mechanism. Absent such a mechanism, it is in fact capture
+avoidance that prevents capture by spliced terms of variables bound in the
+expansion.
 
-* "p10:32 how would you address capture in a system which did feature binding in patterns that could be subsequently referenced?"
+* "p10:32 how would you address capture in a system which did feature binding
+  in patterns that could be subsequently referenced?"
 
@@
 
-> "p24:34 This "backpatching" is done in many Lisp IDEs, see DrRacket
-  for an example [Florence et al, Language Workbench 2016]"
+> "p24:34 This "backpatching" is done in many Lisp IDEs, see DrRacket for an
+example [Florence et al, Language Workbench 2016]"
 

### 2017-04-18 icfp17-rebuttal.txt
indent
tweak

@@
 
-Being that Template Haskell achieves none of our central goals, and that there
+Being that Template Haskell achieves neither of these central goals, and that there
 are a number of other languages that have broadly similar support for

### 2017-04-18 icfp17-rebuttal.txt
fwd ref
though

@@
 Haskell's popularity in the community, we agree that it would be worth
-mentioning explicitly.
+mentioning explicitly. We say a bit more about Template Haskell in response to 
+Review C below.
 
@@
 before typed expansion, the initial context-free parsing of generalized literal
-forms works (see the paragraph immediately under Fig. 3, and Sec. 2's
+forms works (though see the paragraph immediately under Fig. 3, and Sec. 2's
 discussion of the body type.) There is nothing sophisticated about this step --

### 2017-04-18 icfp17-rebuttal.txt
xtreme

@@
 page 18" is incorrect: the bottom of Page 18 is the beginning of Sec. 5, which
-describes an *extension* of the core mechanism to handle parameterized families
+describes an extension of the core mechanism to handle parameterized families
 of types and module parameters. Nothing in Sec. 5 is necessary to understand
@@
 The splice summary/segmentation/antiquotation mechanism, and how it relates to
-the colorings shown, is in fact spelled out in Sec. 2.2 (in particular, the
+the colorings shown, is in fact spelled out in Sec. 2.2 (in particular, in the
 final 3 paragraphs).
@@
 end of the abstract and conclusion) is informal. That said, the phrase was
-chosen because TSMs sit between the well-studied extremes:
-
-1. TSMs offer less syntactic control than syntax dialects, but stronger
-abstract reasoning principles (Sec. 1.1). The term "stronger" here is used in
-the technical sense -- clients can hold more information abstract when
-reasoning about types, binding and segmentation. The only opinion here is that
-abstract reasoning is itself a useful thing, but that is certainly a widely
-accepted belief within the ICFP community.
-
-2. TSMs offer more syntactic control than term-rewriting macros and offer
-comparable reasoning principles (Sec. 1.2). Again, this is an objective claim.
-TSMs have full control over parsing the literal body, whereas term-rewriting
-macros are given an already-parsed term. The reasoning principles are theorems.
+chosen because TSMs sit between the well-studied extremes as described in Sec.
+1.
 

### 2017-04-18 icfp17-rebuttal.txt
the
and

@@
 end of the abstract and conclusion) is informal. That said, the phrase was
-chosen because TSMs sit between the well-studied extremes as described in Sec.
-1.
+chosen because TSMs sit between the well-studied extremes described in Sec. 1.
 
@@
 
-Direct AST generation emerged as the "default" solution in our mechanism in
+Direct AST generation emerged as the "default" solution in our development in
 that adding support for TSM application within proto-expansions would have
 required significant additional machinery. Krishnamurthi's work on micros is
-focused on term-to-term translation from the source language to an IR, and
-macro expansion in the generated translation does eventually occur.
+focused on term-to-term translation from the source language to an IR encoding, 
+and macro expansion in the generated translation does eventually occur.
 

### 2017-04-18 icfp17-rebuttal.txt
thats
providers
tweaks

@@
 Thank you for the thoughtful reviews. We were pleased to see that no
-significant technical problems were found during the review process. The main
+technical problems were found during the review process. The main 
 doubts that seem to remain relate to (1) usefulness and (2) novelty.
 
-With regard to usefulness, Review B states:
+With regard to usefulness, Review A states:
 
-"Although assessing the usefulness of a language design is difficult, reporting
+> "Although assessing the usefulness of a language design is difficult, reporting
 practical and substantial uses of the extension mechanism would go far in
@@
 
-We agree! So much so that we made just such a case for usefulness in our
+We agree! And we made just such a case for usefulness in our
 closely related work a few years ago (Omar et al, ECOOP 2014). That paper gave
@@
 contributes more detailed formal results and develops various functional
-variants of the mechanism as was detailed in Sec. 7. Such a paper should be
-well within the scope of ICFP.
+variants of the mechanism as summarized in Sec. 7.
 
@@
 
-"This work introduces a system of typed syntax macros (TSMs) that gives library
+> "This work introduces a system of typed syntax macros (TSMs) that gives library
 providers substantially more syntactic control than existing typed
@@
 invocation syntactically coincides with function application. Staging macros
-instead serve as a mechanism for staged evaluation, primarily for reasons of
-performance. This is unrelated to our stated goals.
+are motivated by concerns about performance, rather than syntactic cost.
 
@@
 end of the abstract and conclusion) is informal. That said, the phrase was
-chosen because TSMs sit between the well-studied extremes described in Sec. 1.
+chosen because TSMs sit between the problematic extremes described in Sec. 1.
 
@@
 footnote on page 4, the discussion about final expansion in Sec. 2, and also
-many more complex examples in Omar et al, 2014).
+several more complex examples in Omar et al, 2014).
 
@@
 
-"In this setting, the paper follows [Culpepper, Owens, Flatt; GPCE 2005] very
+> "In this setting, the paper follows [Culpepper, Owens, Flatt; GPCE 2005] very
 closely (see also Owens' thesis)."
@@
 
-Yes, you're right. That point wasn't stated as clearly as it could have been.
+Yes, that's right. That point wasn't stated as clearly as it could have been.
 
@@
 the bounds checks in along with the other checks is semantically simpler.
-Clients who favor the abstract approach can choose to use it, since parse
+Providers who favor the abstract approach can choose to use it, since parse
 functions are ML code.
@@
 
-* "p10:32 how would you address capture in a system which did feature binding
+> "p10:32 how would you address capture in a system which did feature binding
   in patterns that could be subsequently referenced?"

### 2017-04-19 icfp17-rebuttal.txt
tweaks to rebuttal

@@
-Thank you for the thoughtful reviews. We were pleased to see that no
-technical problems were found during the review process. The main 
-doubts that seem to remain relate to (1) usefulness and (2) novelty.
+Thank you for the thoughtful reviews. We are encouraged by the fact that 
+the technical approach was considered sound. The main concerns expressed 
+seem to be regarding the (1) usefulness and (2) novelty of our approach,
 
@@
 
-> "Although assessing the usefulness of a language design is difficult, reporting
-practical and substantial uses of the extension mechanism would go far in
-providing evidence that the language is useful."
+> "Although assessing the usefulness of a language design is difficult,
+reporting practical and substantial uses of the extension mechanism would go
+far in providing evidence that the language is useful."
 
-We agree! And we made just such a case for usefulness in our
-closely related work a few years ago (Omar et al, ECOOP 2014). That paper gave
-more sophisticated examples and even gathered empirical data from open source
-projects. The present paper is intended to be a follow-on "theory paper" that
-contributes more detailed formal results and develops various functional
-variants of the mechanism as summarized in Sec. 7.
+In fact we did just that in our prior work at ECOOP 2014, which we referred to
+as the "most closely related work" and discussed in Sec. 1 and at length in
+Sec. 7. That paper gave many realistic examples, described two implementations
+(one that today has 400+ stars on GitHub) and summarized some empirical data
+gathered from open source projects. We drew brief attention to these aspects of
+the prior work in the second paragraph of Sec. 1. We will clarify in the final
+version that this paper defers questions about usefulness to our prior work so
+as to be able to focus on semantic questions.
 
-With regard to novelty, the first sentence of Sec. 1.3 ("Contributions") makes
-explicit the specific claims of our paper:
+With regard to novelty, note that the first sentence of Sec. 1.3
+("Contributions"), makes explicit the specific claims of our paper:
 
-> "This work introduces a system of typed syntax macros (TSMs) that gives library
-providers substantially more syntactic control than existing typed
+> "This work introduces a system of typed syntax macros (TSMs) that gives
+library providers substantially more syntactic control than existing typed
 term-rewriting macro systems while maintaining the ability to reason abstractly
@@
 
-In the remainder of the rebuttal, we will argue, in view of these specific
-claims, that the paper does in fact address all of the major concerns about
-novelty raised by the referees. We humbly ask that the scores assigned to our
-paper be reconsidered in this light.
+In the remainder of the rebuttal, we will argue, in view of this statement of
+our specific claims, that the paper does in fact address all of the major
+concerns about novelty raised by the referees. We also respond to a few other
+concerns, e.g. about the organization of the sections. We humbly ask that the
+scores assigned to our paper be reconsidered in light of these responses.
 
@@
 
-MacroML implements staging macros (stated on pg3). These are quite distinct
+MacroML implements staging macros (as stated on pg3). These are quite distinct
 from term-rewriting macros and TSMs in that they do not give the macro access
@@
 invocation syntactically coincides with function application. Staging macros
-are motivated by concerns about performance, rather than syntactic cost.
+are motivated mainly by concerns about performance, rather than syntactic cost.
 
@@
 for clients to reason abstractly about *what that type is* because there are no
-annotations on Template Haskell macros.
+	type annotations on Template Haskell macros.
 
-Furthermore, Template Haskell does not maintain any sort of hygienic binding
-discipline.
+Furthermore, Template Haskell does not maintain a hygienic binding discipline.
 
-Being that Template Haskell achieves neither of these central goals, and that there
-are a number of other languages that have broadly similar support for
-compile-time code generation, we neglected to mention it in Sec. 1.2 alongside,
-e.g. Scala's macro system, which is more disciplined. However, due to Template
-Haskell's popularity in the community, we agree that it would be worth
-mentioning explicitly. We say a bit more about Template Haskell in response to 
-Review C below.
+Being that Template Haskell achieves neither of these central goals, we
+neglected to mention it in Sec. 1.2 alongside, e.g. Scala's macro system and
+Rust's macro system, both of which are more disciplined. However, due to
+Template Haskell's prominence in the community, we agree that it would be worth
+mentioning explicitly. We say a bit more about Template Haskell and GHC in
+response to Review C below.
 
@@
 
-The hygiene mechanism in Lee et al (2012) is quite different, e.g. it sends
+The hygiene mechanism in Lee et al (2012) is interesting but formally ad hoc, e.g. it sends
 queries at run-time to a compiler and reads back the error messages to
 determine which variables are free in an encoding of a term. No theoretical
-results whatsoever are established in that work.
+results are established in that work.
 
@@
 
-Section 2 seems to be exactly the "one particular section" that is sought: it
-gives a thorough description of how typed expansion operates from start to
-finish.
+Section 2 describes how typed expansion works from start to finish, in the
+style of a detailed tutorial.
 
-One source of confusion may be that the paper only tersely describes how,
-before typed expansion, the initial context-free parsing of generalized literal
-forms works (though see the paragraph immediately under Fig. 3, and Sec. 2's
-discussion of the body type.) There is nothing sophisticated about this step --
-generalized literal forms are parsed like raw string literals. Our prior paper
-showed examples of various other delimiters and gave a context free grammar.
-Condition 4.1 (pg12) explains how we formally handle this first step.
+One source of confusion may be that the paper only tersely describes the step
+just before typed expansion -- context-free parsing. As stated in the paragraph
+immediately under Fig. 3, and as suggested by the definition of the body type
+in Sec. 2, there is nothing sophisticated about this step -- generalized
+literal forms are parsed like raw string literals. Condition 4.1 (pg12)
+explains how we formally handle this first step. The context-free syntax for
+various generalized literal forms was given in prior work (we would be happy to
+reproduce the grammar in the supplement.)
 
@@
 of types and module parameters. Nothing in Sec. 5 is necessary to understand
-prior sections, or the main conceptual points of the paper.
+prior sections, or the high-level conceptual points of the paper.
 
@@
 
-The splice summary/segmentation/antiquotation mechanism, and how it relates to
-the colorings shown, is in fact spelled out in Sec. 2.2 (in particular, in the
-final 3 paragraphs).
+The splice summary/segmentation mechanism, and how it relates to the coloring
+convention, is spelled out in Sec. 2.2 (in particular, in the final 3
+paragraphs). A segmentation cannot be produced if typed expansion fails
+(error recovery is mentioned as future work in Sec. 7).
 
@@
 
-By our reckoning, we gave little to no "motivation for types" in Sec. 1. Types
-are brought up briefly in the fifth point of "Problem 2" in Sec. 1.1, and in
-the third point in Sec. 1.2. Given that our research is focused on languages in
-the ML tradition, and our technical contributions do ultimately relate to
-types, we cannot eliminate statements about types from the introduction
-completely. We could perhaps mention that you can recover an "untyped" variant
-of TSMs by applying the usual type theorist's trick of defining just one type
-that classifies every well-scoped expression. The concerns about
-responsibility, segmentation, context independence and capture would still be
-relevant in that setting.
+[[[[This is a perplexing comment that I don't know how to respond to.]]]
+
+Readers who see no need to reason statically about types can apply the 
+usual type theorist's trick of reducing our system to one where there is 
+just one type that classifies every well-scoped expression. The concerns 
+about responsibility, segmentation, context independence and capture 
+would still be relevant in that setting.
 
@@
 
-The phrase "sweet spot" (which appears exclusively in quotes, and only at the
-end of the abstract and conclusion) is informal. That said, the phrase was
-chosen because TSMs sit between the problematic extremes described in Sec. 1.
+The informal phrase "sweet spot" appears appropriately in quotes, and only at
+the end of the abstract and conclusion. The phrase was chosen because TSMs
+avoid the problematic extremes of both term-rewriting macros and general syntax
+definition systems, as described in Sec. 1. It is perhaps "just an opinion"
+that being able to reason abstractly is a good thing, but this is not a 
+controversial position in the ICFP community.
 

### 2017-04-19 icfp17-rebuttal.txt
tweaks

@@
 Thank you for the thoughtful reviews. We are encouraged by the fact that 
-the technical approach was considered sound. The main concerns expressed 
-seem to be regarding the (1) usefulness and (2) novelty of our approach,
+no non-trivial technical mistakes were found. The main concerns expressed 
+seem to be regarding the (1) usefulness and (2) novelty of our approach.
 

### 2017-04-21 icfp17-rebuttal.txt
finalize rebuttal

@@
-Thank you for the thoughtful reviews. We are encouraged by the fact that 
-no non-trivial technical mistakes were found. The main concerns expressed 
-seem to be regarding the (1) usefulness and (2) novelty of our approach.
+Thank you for the thoughtful reviews. The main concerns expressed seem to be
+regarding the (1) usefulness and (2) novelty of our approach.
 
@@
 
-In fact we did just that in our prior work at ECOOP 2014, which we referred to
-as the "most closely related work" and discussed in Sec. 1 and at length in
-Sec. 7. That paper gave many realistic examples, described two implementations
-(one that today has 400+ stars on GitHub) and summarized some empirical data
-gathered from open source projects. We drew brief attention to these aspects of
-the prior work in the second paragraph of Sec. 1. We will clarify in the final
-version that this paper defers questions about usefulness to our prior work so
-as to be able to focus on semantic questions.
+We did just that in our prior work at ECOOP 2014, which we referred to as the
+"most closely related work" and discussed in Sec. 1 and Sec. 7.  That paper
+gave many realistic examples, described two implementations (one that today has
+400+ stars on GitHub) and summarized some empirical data gathered from open
+source projects. We drew brief attention to these aspects of the prior work in
+the second paragraph of Sec. 1, and we would be happy to expand on the text in
+the final version to make clear that usefulness concerns were substantially
+confronted in prior work.
+
+This paper was written as a "theory paper" that focuses on important semantic
+questions and introduces refinements of the mechanism from ECOOP 2014, as
+summarized in Sec. 7.
 
 With regard to novelty, note that the first sentence of Sec. 1.3
-("Contributions"), makes explicit the specific claims of our paper:
+("Contributions") makes explicit the specific claims of our paper:
 
@@
 In the remainder of the rebuttal, we will argue, in view of this statement of
-our specific claims, that the paper does in fact address all of the major
-concerns about novelty raised by the referees. We also respond to a few other
-concerns, e.g. about the organization of the sections. We humbly ask that the
-scores assigned to our paper be reconsidered in light of these responses.
+our specific claims, that the paper does in fact address all of the concerns
+about novelty raised by the referees. We also respond to a few other concerns,
+e.g. about the organization of the sections. We humbly ask that the scores
+assigned to our paper be reconsidered.
 
@@
 to the syntax of the arguments at all (whether parsed or unparsed.) They cannot
-therefore be used as a mechanism for syntactic control -- staging macro
-invocation syntactically coincides with function application. Staging macros
-are motivated mainly by concerns about performance, rather than syntactic cost.
+therefore be used as a mechanism for syntactic control. Staging macros are
+motivated mainly by concerns about performance, rather than syntactic cost.
 
@@
 
-Being that Template Haskell achieves neither of these central goals, we
-neglected to mention it in Sec. 1.2 alongside, e.g. Scala's macro system and
-Rust's macro system, both of which are more disciplined. However, due to
-Template Haskell's prominence in the community, we agree that it would be worth
-mentioning explicitly. We say a bit more about Template Haskell and GHC in
-response to Review C below.
+Because Template Haskell achieves neither of these central goals, we neglected
+to mention it in Sec. 1.2 alongside, e.g. Scala's macro system and Rust's macro
+system, both of which are more disciplined. However, due to Template Haskell's
+prominence in the community, we agree that it would be worth mentioning
+explicitly. We say a bit more about Template Haskell and GHC in response to
+Review C below.
 
@@
 
-The hygiene mechanism in Lee et al (2012) is interesting but formally ad hoc, e.g. it sends
-queries at run-time to a compiler and reads back the error messages to
-determine which variables are free in an encoding of a term. No theoretical
-results are established in that work.
+The hygiene mechanism in Lee et al (2012) is interesting but formally ad hoc,
+e.g. it sends queries at run-time to a compiler and reads back the error
+messages to determine which variables are free in an encoding of a term. No
+theoretical results are established in that work.
 
@@
 
-Section 2 describes how typed expansion works from start to finish, in the
-style of a detailed tutorial.
+Section 2 describes how typed expansion works from start to finish.
 
@@
 explains how we formally handle this first step. The context-free syntax for
-various generalized literal forms was given in prior work (we would be happy to
-reproduce the grammar in the supplement.)
+various generalized literal forms was given in our prior work (we would be
+happy to reproduce the grammar in the supplement.)
 
@@
 of types and module parameters. Nothing in Sec. 5 is necessary to understand
-prior sections, or the high-level conceptual points of the paper.
+prior sections, or the high-level points of the paper.
 
@@
 The splice summary/segmentation mechanism, and how it relates to the coloring
-convention, is spelled out in Sec. 2.2 (in particular, in the final 3
-paragraphs). A segmentation cannot be produced if typed expansion fails
-(error recovery is mentioned as future work in Sec. 7).
+convention, is spelled out in Sec. 2.2. A segmentation cannot be produced if
+typed expansion fails (error recovery is mentioned as future work in Sec. 7).
 
@@
 
-[[[[This is a perplexing comment that I don't know how to respond to.]]]
-
-Readers who see no need to reason statically about types can apply the 
-usual type theorist's trick of reducing our system to one where there is 
-just one type that classifies every well-scoped expression. The concerns 
-about responsibility, segmentation, context independence and capture 
-would still be relevant in that setting.
+Our approach is unabashedly rooted in the type-theoretic/ML tradition, yes, but
+we did not make any particular attempt to advocate for types in the
+introduction. Readers who see no need to reason statically about types can
+apply the usual type theorist's trick of reducing our system to one where there
+is just one type that classifies every well-scoped expression. The concerns
+about responsibility, segmentation, context independence and capture would
+still be relevant in that setting.
 
@@
 avoid the problematic extremes of both term-rewriting macros and general syntax
-definition systems, as described in Sec. 1. It is perhaps "just an opinion"
-that being able to reason abstractly is a good thing, but this is not a 
-controversial position in the ICFP community.
+definition systems, as described in Sec. 1. It is perhaps just an opinion
+that being able to reason abstractly about programs is a good thing, but this
+is not a controversial position in the ICFP community.
 
@@
 GHC quasiquotes address only the problems of conflict and responsibility. The
-use of fixed outer delimiters to handle conflict is explicitly disclaimed as a
-novel contribution at the top of page 4, where we discuss strings and the work
-of Slind 1991; also Omar et al, 2014. The fact that responsibility is handled
-by explicit invocation is also clearly attributed to existing macro systems in
-Sec. 1.2 (point 1).
+use of fixed outer delimiters to handle conflict is not claimed as a novel
+contribution and clearly attributed to prior work at the top of page 4, where
+we discuss strings and the work of Slind 1991; also Omar et al, 2014.  The fact
+that responsibility is handled by explicit invocation is also clearly
+attributed to existing macro systems in Sec. 1.2 (point 1).
 
@@
 TSMs are novel not for each feature individually, but because you need not make
-trade-offs between these features.
+trade-offs between the mentioned reasoning principles.
 
@@
 
-Our mechanism for identifier expansion does not operate quite like both
-Dybvig91 and Herman and Wand (2008, 2010) describe -- in our work, renaming is
+Our mechanism for identifier expansion is not specified in the same way as
+Dybvig91 and Herman and Wand (2008, 2010) -- in our formalism, renaming is
 *implicitly* necessary to discharge disjointness premises in the typed
@@
 macros / Racket's #lang mechanism, but these amount to syntax definition
-systems that suffer from some or all of the problems described in Sec. 1.1.
+systems that suffer from the problems described in Sec. 1.1.
 
@@
 
-This would be addressed exactly as it is with expression TSMs.
+The approach taken for expression TSMs should work in this situation too.
 
@@
 
-I will mention that in the future work, thanks.
+We will mention that in the future work, thanks.

### 2017-04-27 icfp17-rebuttal.txt
new review and rebuttal

@@
 macro system in a statically typed language eventually must), there is no way
-for clients to reason abstractly about *what that type is* because there are no
-	type annotations on Template Haskell macros.
+for clients to reason abstractly about *what that type is* because there are no type annotations on Template Haskell macros.
 
@@
 We will mention that in the future work, thanks.
+
+----
+
+> "Since these systems have quite different purposes, comparing them produces
+nothing meaningful."
+
+We organized our comparison by technical mechanism (syntax definition vs.
+term-rewriting), not by "purpose".
+A substantially more detailed comparison to related work, spanning 20 pages, is available in my PhD thesis [1] but obviously, this level of detail would not be appropriate for a conference paper.
+
+> "there are two quite distinct endeavours to consider"
+
+Comparing mechanisms of syntactic control in terms of reasoning principles, rather than simpler "expressability" concerns, is the major emphasis of our paper.
+From that perspective, the two proposed classes of examples are not distinct.
+
+In the paper, we did distinguish examples, like do-notation, that introduce bindings into sub-terms.
+TSMs as defined in the paper do not support these examples, though a simple modification could be made to add support.
+But we suggest that even if such examples were possible, it may not be worth it for the increased reasoning complexity that clients would have to contend with (because each spliced term would be associated not just with a type, but also with a list of bindings -- that is much harder to convey using, e.g., an editor.)
+
+> "For example, can TSMs handle the SQL binding constructs such as 'AS'?"
+
+Encodings of SQL queries, including those with 'AS' clauses, can be expressed using TSMs.
+If one were to somehow express the binding structure of SQL 'AS' clauses using the binding structure of the host language (a HOAS-style approach), it would not be possible -- see above.
+
+> "the only way to determine the parsing is to expand the syntax and find
+the 'spliced' sections, and a reader looking at a fragment of unexpanded
+source can't tell how to parse identifiers"
+
+The paper clearly states that the locations of spliced segments is communicated using secondary notation by the editor / pretty-printer.
+
+> "Hygiene (capture and context dependence) is a significant part of this
+work, and I'd like to see a deeper comparison between your design and
+existing work on hygiene (e.g.. Dave Herman's "A Theory of Typed Hygienic
+Macros")."
+
+We state on p4 that existing hygiene mechanisms prevent splicing terms out of string literals. Our mechanism does not have that problem.
+
+Technically, our approach to specifying hygiene is also different in that it uses an ABT-generic capture-avoiding substitution mechanism, rather than an explicit renaming approach.
+The rebuttal to the other reviews details this distinction.
+
+> "I'm troubled by the lack of complete examples."
+
+The rebuttal to the other reviews addresses this concern -- this paper is a "theory paper" that serves as a companion to our prior paper (ECOOP 2014), which did give several more examples.
+We did not have enough space to give a complete parser for a non-trivial syntax.
+
+> "Since TSMs generate untyped parse trees that are type-checked on expansion
+then I would be reluctant to describe them as "typed"."
+
+We call them "typed" because type annotations serve to allow clients to reason abstractly about types.
+
+The verification-validation distinction, which we also discussed, is important, but is not equivalent to "typed vs untyped".
+
+> "A short English summary alongside each reasoning principle (Theorems 4.5,
+4.6, 5.1) would make them much more accessible to the casual reader."
+
+There is in fact an English summary alongside each reasoning principle in Theorem 4.5.
+The correspond clauses would have nearly identical descriptions for the other theorems, so we did not duplicate them for space reasons.
+
+[1] http://www.cs.cmu.edu/~comar/omar-thesis.pdf

### 2017-05-02 icfp17/syntax-icfp17.tex
moved failed ICFP17 submission into a directory and copied it into popl18 for next try. made some tweaks to the abstract

--- 2017-02-28 syntax-icfp17.tex
@@
 
-In short, syntax dialects do not come with reasonable principles of \emph{syntactic abstraction}: if the desugaring of the program text is held abstract, programmers can no longer reason about types and binding (i.e.
+In short, syntax dialects do not come equipped with reasonable principles of \emph{syntactic abstraction}: if the desugaring of the program text is held abstract, programmers can no longer reason about types and binding (i.e.
 answer questions like those above) in the usual disciplined manner.
@@
 \item Macro systems require that the client explicitly apply the intended rewriting (implemented by a macro) to the term that is to be rewritten, thereby addressing the problems of conflict and determining responsibility.
-However, it is often unclear whether a given macro transforms or reinterprets the form of a given argument or sub-term thereof, as opposed to treating it parametrically by inserting it unmodified into the generated expansion.
+However, it is often unclear whether a given macro  is repurposing the form of a given argument or sub-term thereof, as opposed to treating it parametrically by inserting it unmodified into the generated expansion.
 This is closely related to the problem of determining a {segmentation} discussed above.
@@
 
-The problem here is that there is no way to extract the spliced expressions from the supplied string literal forms while satisfying the hygiene conditions, because variables that come from these spliced terms (here, \li{first_name} and \li{heading}) are indistinguishable from variables that inappropriately appear free in the generated expansion.
+The problem here is that there is no way to extract the spliced expressions from the supplied string literal forms while satisfying the hygiene conditions, because variables that come from these spliced terms (here, \li{first_name} and \li{heading}) are indistinguishable from variables that inappropriately appear free relative to the expansion.
 In addition, the problem of segmentation becomes even more pernicious: to a human or tool unaware of Ur/Web's syntax, it is not immediately apparent which particular subsequences of the string literals supplied to \li{html!} are segmented out as spliced expressions.
@@
 \begin{enumerate}
-\item They allow for TSMs that operate not just at a single type, but over a type- and module-parameterized family of types. For example, rather than defining a TSM % \li{$strlist} 
-for string-valued dictionaries and another TSM 
-%\li{$intlist} 
-for integer-valued dictionaries, or over different implementations of the dictionary signature, we can define a single parametric TSM 
-%\li{$list} 
-that operates uniformly across a type- and module-parameterized family of dictionary types.
+\item They allow for TSMs that operate not just at a single type, but over a type- and module-parameterized family of types.
+For example, we consider a parametric TSM for dictionaries that operates uniformly over different implementations of the dictionary signature, and also over different choices of the co-domain type.
 \item Parameters also give expansions access to terms other than those explicitly spliced in to the literal body.
@@
 \end{enumerate}
-This section culminates with a calculus of ML-style modules that has support for parametric TSMs.
+This section culminates with a calculus that extends the calculus of Sec. \ref{sec:setsms-formally} with ML-style modules and parametric TSMs.
 \item Sec.
 \ref{sec:static-eval} considers the topic of term evaluation during the typed expansion phase, i.e.
-\emph{static evaluation}, in a bit more depth, and also gives examples of TSMs that are useful for defining other TSMs, e.g.
+\emph{static evaluation}, in  more detail, and also gives examples of TSMs that are useful for defining other TSMs, e.g.
 TSMs that implement parser generators and quasiquotation.
@@
 \end{lstlisting}
-Here, \li{spliced<13; 22; html>} refers to the spliced expression \li{first_name}, which must be of type \li{string}, by location, rather than directly.
-
-Requiring that the TSM refer to spliced expressions indirectly in this manner prevents it from ``forging'' a spliced expression (i.e.
-claiming that an expression is a spliced expression when it does not appear textually in the literal body.)% Our semantics distinguishes spliced expressions when validating the proto-expansion, which is important for reasons of hygiene that we will detail shortly.
+Here, \li{spliced<13; 22; string>} refers to the spliced expression \li{first_name} by location, rather than directly, and indicates that it must be of type \li{string}.
+
+Requiring that TSMs refer to spliced expressions indirectly in this manner prevents them from ``forging'' spliced terms (i.e.
+claiming that an term is a spliced term when it does not in fact appear textually in the literal body.)% Our semantics distinguishes spliced expressions when validating the proto-expansion, which is important for reasons of hygiene that we will detail shortly.
 
@@
 The \emph{splice summary} of a proto-expression is the finite set of references to spliced terms contained within.
-For example, the summary of the proto-expression above is the finite set containing only \li{spliced<13; 22; html>}.% Notice that no information about  the spliced terms appear is communicated by the splice summary.
-
-The \emph{segmentation} of a proto-pression is the finite set of segments extracted from the splice summary, where a segment is simply a pair of natural numbers.
+For example, the summary of the proto-expression above is the finite set containing only \li{spliced<13; 22; string>}.% Notice that no information about  the spliced terms appear is communicated by the splice summary.
+
+The \emph{segmentation} of a proto-expression is the finite set of segments extracted from the splice summary, where a segment is simply a pair of natural numbers.
 For example, the segmentation of the proto-expansion above is $\{(13, 22)\}$.
@@
           \patExpands
-            {\upctx_i}
+            {\uGG{\uG_i}{\pctx_i}}
             {\uPhi}
@@
         }{\npat}$  and $\sseq{\patType{\pctx_i}{p_i}{\tau_i}}{\npat}$
-      \item (\textbf{No Hidden Bindings}) $\upctx = \biguplus_{0 \leq i < \npat} \upctx_i$
+      \item (\textbf{No Hidden Bindings}) $\uG = \biguplus_{0 \leq i < \npat} \uG_i$ and $\Gamma = \bigcup_{0 \leq i < \npat} \pctx_i$
         \begin{quote}

### 2017-05-02 popl18/syntax-popl18.tex
moved failed ICFP17 submission into a directory and copied it into popl18 for next try. made some tweaks to the abstract

--- 2017-05-02 icfp17/syntax-icfp17.tex
@@
 
-\input{../../thesis/macros}
+\input{../../../thesis/macros}
 
@@
 Programmers, 
-like pen-and-paper mathematicians, often wish to define  ``syntactic sugar'' to  lower the syntactic cost of common idioms. 
+like pen-and-paper mathematicians, often want to define new ``syntactic sugar'' to  lower the syntactic cost of common idioms. 
 Unfortunately, the most direct mechanisms of syntactic control available to programmers, e.g.
-\li{camlp4} \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, do not support modular reasoning about determinism (i.e.
-separately defined forms can conflict syntactically with one another.) Moreover, they obscure the type and binding structure of the program.
-Typed hygienic term-rewriting macros (e.g.
-as available in Scala) are more reasonable, but they give programmers only limited syntactic control.
-
-This paper formally introduces \emph{simple} and \emph{parametric typed syntax macros} (TSMs), which give library providers programmatic control over both the parsing and expansion of expressions and patterns of \emph{generalized literal form} at a specified type or parameterized family of types.
+\li{camlp4} \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, do not support modular reasoning about syntactic determinism -- separately defined syntax extensions can conflict syntactically with one another.
+Moreover, clients cannot reason abstractly, i.e.
+given only the program text, about types and binding.
+Instead, they must reason transparently, i.e.
+about the underlying expansion.
+To address these problems, this paper  formally introduces \emph{simple} and \emph{parametric typed literal macros} (TLMs), which give library providers programmatic control over both the parsing and expansion of expressions and patterns of \emph{generalized literal form} \cite{TSLs} at a specified type or parameterized family of types.
 Expansion is {strictly hygienic}, meaning that 1) expansions can refer to external bindings only via spliced terms or supplied module parameters; and 2)  expansions do not reveal internal bindings to spliced terms or to the remainder of the program.
 % Partial parameter application lowers the syntactic cost of this strict style. 
-This design ensures that clients are able to reason about the binding structure of the program while holding the expansion of the program abstract -- the system needs only convey to clients the \emph{segmentation} of each literal, which gives the locations of the spliced terms.
+TLMs offer substantially more syntactic control than existing hygienic term-rewriting macro systems while ensuring that clients are able to reason modularly about syntactic determinism and abstractly about the type and binding structure of the program.
+The system needs only to convey to clients (e.g.
+via secondary notation) the \emph{segmentation} of each literal, which gives the locations of the spliced terms.
 %The system can determine this {segmentation} automatically and convey it to the client via syntax highlighting, or by presenting the client with a context-free grammar (without revealing the semantic actions associated with each production.)
-We argue that this mechanism occupies a ``sweet spot'' in the design space, in that it captures many common syntactic idioms while avoiding the problem of syntactic conflicts by construction and supplying clients with clear abstract reasoning principles.
-We establish these reasoning principles formally.
+% This mechanism captures many common syntactic idioms while avoiding the problem of syntactic conflicts by construction and supplying clients with clear abstract reasoning principles. 
+We specify TLMs and establish these abstract reasoning principles formally using an ML-style calculus of expressions, patterns and modules.
 
@@
 
-Third-party library providers are, quite justifiably, envious of the privileged status of these otherwise seemingly ordinary standard library definitions.
+Third-party library providers are, quite justifiably, envious of the privileged status of these otherwise ordinary standard library definitions.
 After all, it is not difficult to find other examples of situationally useful library-specific syntactic sugar \cite{TSLs}.
@@
 %% Bibliography
-\bibliography{../../papers/research}
+\bibliography{../../../papers/research}
 

### 2017-05-18 popl18/syntax-popl18.tex
fixes from final thesis version

@@
 
-% \emph{Typed syntax macros (TSMs)}, proposed in a recent short paper by Omar et al.
+% \emph{Typed syntax macros (TLMs)}, proposed in a recent short paper by Omar et al.
 \cite{sac15}, give library providers programmatic control over the parsing and expansion of only terms of {(generalized) literal form}.
 This appears to occupy a ``sweet spot'' in that it captures many common syntactic idioms while avoiding the problem of conflict.
-TSMs also maintain a reasonable type and binding discipline.% In particular, clients can use any combination of TSMs in a program without needing to consider conflicts between them, and the language validates each expansion that a TSM generates to maintain 1) a \emph{type discipline} (meaning clients can determine the type of an unexpanded expression without examining its expansion directly);
+TLMs also maintain a reasonable type and binding discipline.% In particular, clients can use any combination of TLMs in a program without needing to consider conflicts between them, and the language validates each expansion that a TLM generates to maintain 1) a \emph{type discipline} (meaning clients can determine the type of an unexpanded expression without examining its expansion directly);
 % %and 2) a \emph{hygienic binding discipline}.
@@
 
-% TSMs have only been described minimally -- it is not clear how they should be adapted for integration into languages like ML that support {pattern matching}, {parameterized datatypes}, {modules} and abstract types.
-Moreover, the prior work makes several simplifying assumptions related to binding that are impractically restrictive.% bno mechanism for binding values for use across TSM definitions has been described and the hygiene mechanism makes giving the expansions that they generate access to ``helper functions'' awkward.
-
-% This paper gives a complete account of TSMs that addresses these deficiencies.
-In particular, we 1) integrate TSMs with pattern matching; 2) introduce a distinct static phase of evaluation, which gives TSM definitions access to libraries; and 3) introduce type and module parameters, which serve two purposes: they allow for TSMs that operate uniformly at a parameterized family of types (rather than only at a single type), and they give expansions explicit, hygienic access to libraries.
+% TLMs have only been described minimally -- it is not clear how they should be adapted for integration into languages like ML that support {pattern matching}, {parameterized datatypes}, {modules} and abstract types.
+Moreover, the prior work makes several simplifying assumptions related to binding that are impractically restrictive.% bno mechanism for binding values for use across TLM definitions has been described and the hygiene mechanism makes giving the expansions that they generate access to ``helper functions'' awkward.
+
+% This paper gives a complete account of TLMs that addresses these deficiencies.
+In particular, we 1) integrate TLMs with pattern matching; 2) introduce a distinct static phase of evaluation, which gives TLM definitions access to libraries; and 3) introduce type and module parameters, which serve two purposes: they allow for TLMs that operate uniformly at a parameterized family of types (rather than only at a single type), and they give expansions explicit, hygienic access to libraries.
 Support for partial application  of parameters lowers the syntactic cost of this explicit approach.
@@
 \subsection{Contributions}
-This work introduces a system of \textbf{typed syntax macros (TSMs)} that gives library providers substantially more syntactic control than existing typed term-rewriting macro systems while maintaining the ability to reason abstractly about types, binding and segmentation.% abstract reasoning principles.
+This work introduces a system of \textbf{typed literal macros (TLMs)} that gives library providers substantially more syntactic control than existing typed term-rewriting macro systems while maintaining the ability to reason abstractly about types, binding and segmentation.% abstract reasoning principles.
 % comparable to the level of control they have when defining a syntax dialect.
 
-% Clients apply TSMs to \emph{generalized literal forms}. 
-Our take on the running example is shown in Fig. \ref{fig:first-tsm-example}: we apply a TSM named 
+% Clients apply TLMs to \emph{generalized literal forms}. 
+Our take on the running example is shown in Fig. \ref{fig:first-tsm-example}: we apply a TLM named 
 \li{$html} 
- to \emph{generalized literal forms} delimited by backticks. TSM names are prefixed by \li{$} 
+ to \emph{generalized literal forms} delimited by backticks. TLM names are prefixed by \li{$} 
  to clearly them from variables.
-The semantics delegates control over the parsing and expansion of each literal body to the applied TSM during a semantic phase called \emph{typed expansion}, which generalizes the usual typing phase.
+The semantics delegates control over the parsing and expansion of each literal body to the applied TLM during a semantic phase called \emph{typed expansion}, which generalizes the usual typing phase.
 \begin{figure}[ht!]
@@
 \vspace{-10px}
-\caption{Two examples of a TSM being applied to a generalized literal form.
+\caption{Two examples of a TLM being applied to a generalized literal form.
 %In each case, literal body, between backticks, is initially left unparsed according to the language's context-free syntax.
-%The applied TSM determines a segmentation and expansion during the typed expansion phase, which generalizes the usual typing phase.
+%The applied TLM determines a segmentation and expansion during the typed expansion phase, which generalizes the usual typing phase.
 We assume strings are primitive.}
@@
 Notice, however, that Figure \ref{fig:first-tsm-example} reveals the locations of the spliced expressions by coloring them black.
-We have designed our system so that a figure like this, which presents a \emph{segmentation} of each literal body into spliced terms (in black) and characters parsed in some other way by the applied TSM (in green)\footnote{In general, spliced expressions might themselves apply TSMs, in which case the convention is to use a distinct color for unspliced segments at each depth.
+We have designed our system so that a figure like this, which presents a \emph{segmentation} of each literal body into spliced terms (in black) and characters parsed in some other way by the applied TLM (in green)\footnote{In general, spliced expressions might themselves apply TLMs, in which case the convention is to use a distinct color for unspliced segments at each depth.
 For example, if strings were not primitive, we might write \li{`SURL<body>\{EURLheading ($str "SSTRWorld!ESTR")SURL\} ...</body>EURL`}.}, can always be  generated.
@@
 
-%As such, the semantics can take the type and binding structure of the surrounding program into account when validating the expansion that the TSM programmatically generates to ensure that clients can answer critical questions related to types and binding, like those enumerated in Section \ref{sec:abs-reasoning-intro}.
-Clients need not have knowledge of the implementation of the TSM or of the generated expansion, i.e.
+%As such, the semantics can take the type and binding structure of the surrounding program into account when validating the expansion that the TLM programmatically generates to ensure that clients can answer critical questions related to types and binding, like those enumerated in Section \ref{sec:abs-reasoning-intro}.
+Clients need not have knowledge of the implementation of the TLM or of the generated expansion, i.e.
 there are useful principles of syntactic abstraction.
 
-% The primary technical challenge has to do with the fact that the applied TSM needs to be able to splice terms out of the literal body for inclusion in the expansion. For example,
-
- % We design our mechanism such that these locations can easily be determined from the output of the TSM.
+% The primary technical challenge has to do with the fact that the applied TLM needs to be able to splice terms out of the literal body for inclusion in the expansion. For example,
+
+ % We design our mechanism such that these locations can easily be determined from the output of the TLM.
 This is essential for our hygiene mechanism, and it is also useful in that this information can be presented to the user (e.g.
@@
 %In order to reason about types and binding, client programmers need only have knowledge of 1) the segmentation (e.g.
-by examining a figure like this presented by a code editor or pretty-printer) and 2) a type annotation on the definition of the applied TSM.
-No other details about the applied TSM's implementation or the expansion that it generates need to be revealed to the client programmer.
+by examining a figure like this presented by a code editor or pretty-printer) and 2) a type annotation on the definition of the applied TLM.
+No other details about the applied TLM's implementation or the expansion that it generates need to be revealed to the client programmer.
 In other words,
-TSMs come equipped with useful principles of syntactic abstraction.
+TLMs come equipped with useful principles of syntactic abstraction.
 We will more precisely characterize these abstract reasoning principles as we proceed.
-For now, to develop some intuitions, consider Figure \ref{fig:K-tsm-example}, which uses TSMs to express the ``unreasonable'' example from Figure \ref{fig:K-dialect}.
+For now, to develop some intuitions, consider Figure \ref{fig:K-tsm-example}, which uses TLMs to express the ``unreasonable'' example from Figure \ref{fig:K-dialect}.
 Without examining the expansion of Line 3, we can reason as follows:
@@
 \begin{enumerate}
-\item \textbf{(Responsibility)} The applied TSM, \li{$kquery}, is solely responsible for typed expansion of the literal body. 
+\item \textbf{(Responsibility)} The applied TLM, \li{$kquery}, is solely responsible for typed expansion of the literal body. 
 \item \textbf{(Segmentation)} By examining the segmentation, we know that the two instances of \li{x} on Line 3 are parsed as spliced expressions, whereas \li{R} is parsed in some other way peculiar to this form.
@@
 \item \textbf{(Typing)} An explicit type annotation on the definition of \li{$kquery} determines the type that every expansion it generates will have.
-We will see an example of a TSM definition in the next section.
+We will see an example of a TLM definition in the next section.
 
@@
 This information is usually not necessary to reason about typing, but it can be conveyed to the programmer upon request by the program editor if desired.
-%TSM definitions follow the usual scoping rules, so it is easy to ``jump to the definition'' of \li{$kquery}.
+%TLM definitions follow the usual scoping rules, so it is easy to ``jump to the definition'' of \li{$kquery}.
 \end{enumerate}
@@
 \vspace{-5px}
-\caption{TSMs make examples like the one from Fig. \ref{fig:K-dialect} more reasonable.}
+\caption{TLMs make examples like the one from Fig. \ref{fig:K-dialect} more reasonable.}
 \vspace{-5px}
@@
 
-% There is also no ambiguity with regard to which TSM has control over each form, and searching for the definition of a TSM is no more difficult than searching for any other binding, i.e.
+% There is also no ambiguity with regard to which TLM has control over each form, and searching for the definition of a TLM is no more difficult than searching for any other binding, i.e.
 there are well-defined scoping rules.
 
-% In other words, TSMs maintain a useful notion of syntactic abstraction.
-%More specifically, TSMs maintain a \emph{hygienic binding discipline}, meaning that questions Questions 4 and 5 above were concerned with are disallowed entirely.
+% In other words, TLMs maintain a useful notion of syntactic abstraction.
+%More specifically, TLMs maintain a \emph{hygienic binding discipline}, meaning that questions Questions 4 and 5 above were concerned with are disallowed entirely.
 % We will, of course, make this notion more technically precise as we continue.
@@
 \begin{itemize}
-\item Sec. \ref{sec:setsms} begins by going through the mechanics of defining a {simple expression TSM} and describes the process of typed expansion and expansion validation by example.
-\item Sec. \ref{sec:sptsms} does the same for a {simple pattern TSM}, i.e. a TSM that generates patterns that match values of a specified type. The hygiene condition for pattern TSMs is interesting.
-\item Sec. \ref{sec:setsms-formally} defines a calculus of simple expression and pattern TSMs and formally establishes the reasoning principles implied above.
-
-% In the remaining chapters, we enrich the language developed in Chapter \ref{chap:uetsms} with advanced features based on those found in full-scale general-purpose languages like ML and Scala, and enhance our TSM mechanism with these new features.
-
-% We next develop a more sophisticated account of TSMs in Chapters \ref{chap:ptsms} and \ref{chap:static-eval}, with the aim of working out details necessary to integrate TSMs into full-scale general-purpose languages like ML or Scala.
+\item Sec. \ref{sec:setsms} begins by going through the mechanics of defining a {simple expression TLM} and describes the process of typed expansion and expansion validation by example.
+\item Sec. \ref{sec:sptsms} does the same for a {simple pattern TLM}, i.e. a TLM that generates patterns that match values of a specified type. The hygiene condition for pattern TLMs is interesting.
+\item Sec. \ref{sec:setsms-formally} defines a calculus of simple expression and pattern TLMs and formally establishes the reasoning principles implied above.
+
+% In the remaining chapters, we enrich the language developed in Chapter \ref{chap:uetsms} with advanced features based on those found in full-scale general-purpose languages like ML and Scala, and enhance our TLM mechanism with these new features.
+
+% We next develop a more sophisticated account of TLMs in Chapters \ref{chap:ptsms} and \ref{chap:static-eval}, with the aim of working out details necessary to integrate TLMs into full-scale general-purpose languages like ML or Scala.
 %These two chapters constitute Part \ref{part:parametric-tsms} of our contributions.
 \item Sec.
-\ref{sec:ptsms} adds parameterized types and an ML-style module system and introduces \emph{parametric TSMs} (pTSMs), i.e.
-TSMs that take type and module parameters.
+\ref{sec:ptsms} adds parameterized types and an ML-style module system and introduces \emph{parametric TLMs} (pTLMs), i.e.
+TLMs that take type and module parameters.
 Parameters serve two purposes:
 \begin{enumerate}
-\item They allow for TSMs that operate not just at a single type, but over a type- and module-parameterized family of types.
-For example, we consider a parametric TSM for dictionaries that operates uniformly over different implementations of the dictionary signature, and also over different choices of the co-domain type.
+\item They allow for TLMs that operate not just at a single type, but over a type- and module-parameterized family of types.
+For example, we consider a parametric TLM for dictionaries that operates uniformly over different implementations of the dictionary signature, and also over different choices of the co-domain type.
 \item Parameters also give expansions access to terms other than those explicitly spliced in to the literal body.
 This is particularly convenient for working with abstract types \emph{a la} ML.
-Partial parameter application in TSM abbreviations lowers the cost of this explicit parameter-passing discipline.
+Partial parameter application in TLM abbreviations lowers the cost of this explicit parameter-passing discipline.
 \end{enumerate}
-This section culminates with a calculus that extends the calculus of Sec. \ref{sec:setsms-formally} with ML-style modules and parametric TSMs.
+This section culminates with a calculus that extends the calculus of Sec. \ref{sec:setsms-formally} with ML-style modules and parametric TLMs.
 \item Sec.
 \ref{sec:static-eval} considers the topic of term evaluation during the typed expansion phase, i.e.
-\emph{static evaluation}, in  more detail, and also gives examples of TSMs that are useful for defining other TSMs, e.g.
-TSMs that implement parser generators and quasiquotation.
+\emph{static evaluation}, in  more detail, and also gives examples of TLMs that are useful for defining other TLMs, e.g.
+TLMs that implement parser generators and quasiquotation.
 \item Sec.
@@
 For convenience, the examples that we give are written in a full-scale functional language under development called VerseML.
-% For the present purposes, we can consider VerseML as essentially the calculus defined in the supplement extended with various conveniences that are commonly found in other ML-like languages and, notionally, orthogonal to TSMs.
+% For the present purposes, we can consider VerseML as essentially the calculus defined in the supplement extended with various conveniences that are commonly found in other ML-like languages and, notionally, orthogonal to TLMs.
 %VerseML is, as its name suggests, a conceptual descendent of ML.
-It diverges from other dialects of ML that have a similar type structure in that it has a bidirectional type system \cite{Pierce:2000:LTI:345099.345100} (like, for example, Scala \cite{OdeZenZen01}) for reasons that have to do with the mechanism of TSM implicits described in Chapters \ref{chap:tsls} and \ref{chap:ptsms}.
-%The reason we will not follow Standard ML \cite{mthm97-for-dart} in giving a complete formal definition of VerseML in this work is both to emphasize that the primitives we introduce are ``insensitive'' to the details of the underlying type structure of the language (so TSMs can be considered for inclusion in a variety of languages, not only dialects of ML), and to avoid distracting the reader (and the author) with definitions that are already well-understood in the literature and that are orthogonal to those that are the focus of this work.
+It diverges from other dialects of ML that have a similar type structure in that it has a bidirectional type system \cite{Pierce:2000:LTI:345099.345100} (like, for example, Scala \cite{OdeZenZen01}) for reasons that have to do with the mechanism of TLM implicits described in Chapters \ref{chap:tsls} and \ref{chap:ptsms}.
+%The reason we will not follow Standard ML \cite{mthm97-for-dart} in giving a complete formal definition of VerseML in this work is both to emphasize that the primitives we introduce are ``insensitive'' to the details of the underlying type structure of the language (so TLMs can be considered for inclusion in a variety of languages, not only dialects of ML), and to avoid distracting the reader (and the author) with definitions that are already well-understood in the literature and that are orthogonal to those that are the focus of this work.
 % We will not formally define these features mainly to avoid unnecessarily complicating our presentation with details that are not essential to the ideas presented herein. As such, 
@@
 % \end{lstlisting}
-% \caption{The example from Figure \ref{fig:first-tsm-example-marked}, expressed using a parametric TSM.}
+% \caption{The example from Figure \ref{fig:first-tsm-example-marked}, expressed using a parametric TLM.}
 % \label{fig:first-ptsm-example-marked}
@@
 % In Secs.
-\ref{chap:uetsms} and \ref{chap:uptsms}, we assume for the sake of technical simplicity that each TSM definition is self-contained, needing no access to libraries or to other TSMs.
+\ref{chap:uetsms} and \ref{chap:uptsms}, we assume for the sake of technical simplicity that each TLM definition is self-contained, needing no access to libraries or to other TLMs.
 This is an impractical assumption in practice.
@@
 % %\item \textbf{Type-specific languages}, or \textbf{TSLs}. TSLs, described 
-% In Chapter \ref{chap:tsls}, we develop a mechanism of \emph{TSM implicits} that allows library clients to contextually designate, for any type, a privileged TSM at that type.
-The semantics applies this privileged TSM implicitly to unadorned literal forms that appear where a term of the associated type is expected.
+% In Chapter \ref{chap:tsls}, we develop a mechanism of \emph{TLM implicits} that allows library clients to contextually designate, for any type, a privileged TLM at that type.
+The semantics applies this privileged TLM implicitly to unadorned literal forms that appear where a term of the associated type is expected.
 For example, if we designate
 % %\li{$str} 
-% as the privileged TSM at the \li{string} type and 
+% as the privileged TLM at the \li{string} type and 
 % %\li{$strlist}
-% as the privileged TSM at the \li{list(string)} type, we can express the example from Figure \ref{fig:first-tsm-example-marked} instead as shown in Figure \ref{fig:first-tsm-example-implicit} (assuming \li{join} has type \li{string -> list(string) -> string}.)
+% as the privileged TLM at the \li{list(string)} type, we can express the example from Figure \ref{fig:first-tsm-example-marked} instead as shown in Figure \ref{fig:first-tsm-example-implicit} (assuming \li{join} has type \li{string -> list(string) -> string}.)
 % \begin{figure}[h]
@@
 % \end{lstlisting}
-% \caption{The example from Figure \ref{fig:first-tsm-example-marked} drawn to take advantage of TSM implicits.}
+% \caption{The example from Figure \ref{fig:first-tsm-example-marked} drawn to take advantage of TLM implicits.}
 % \label{fig:first-tsm-example-implicit}
@@
 
-%TSMs, like other macro systems, perform \emph{static code generation} (also sometimes called \emph{static} or \emph{compile-time metaprogramming}), meaning that the relevant rules in the static semantics of the language call for the evaluation of \emph{static functions} that generate term encodings.
+%TLMs, like other macro systems, perform \emph{static code generation} (also sometimes called \emph{static} or \emph{compile-time metaprogramming}), meaning that the relevant rules in the static semantics of the language call for the evaluation of \emph{static functions} that generate term encodings.
 Static functions are functions that are evaluated statically, i.e.
@@
 
-\newcommand{\seTSMsSec}{Simple Expression TSMs (seTSMs)}
-\section{\protect\seTSMsSec}
+\newcommand{\seTLMsSec}{Simple Expression TLMs (seTLMs)}
+\section{\protect\seTLMsSec}
 \label{sec:setsms}
 
-We will begin in this section by introducing \emph{simple expression TSMs} (seTSMs), i.e. TSMs where expansions are expressions of a single specified type.
-
-% \section{Simple Expression TSMs By Example}\label{sec:tsms-by-example}
-% We begin in this section with a ``tutorial-style'' introduction to seTSMs in VerseML.
-%In particular, we will define an seTSM for constructing values of the recursive labeled sum type \li{rx} that was defined in Figure \ref{fig:datatype-rx}.
+We will begin in this section by introducing \emph{simple expression TLMs} (seTLMs), i.e. TLMs where expansions are expressions of a single specified type.
+
+% \section{Simple Expression TLMs By Example}\label{sec:tsms-by-example}
+% We begin in this section with a ``tutorial-style'' introduction to seTLMs in VerseML.
+%In particular, we will define an seTLM for constructing values of the recursive labeled sum type \li{rx} that was defined in Figure \ref{fig:datatype-rx}.
 % Sec.
 \ref{sec:tsms-minimal-formalism} then formally defines a reduced dialect of VerseML called $\miniVerseUE$.
-This will serve as a ``conceptually minimal'' core calculus of TSMs, in the style of the simply typed lambda calculus.
+This will serve as a ``conceptually minimal'' core calculus of TLMs, in the style of the simply typed lambda calculus.
 %We conclude in Sec.
@@
 
-% \subsection{TSM Application}\label{sec:uetsms-usage}
-% The following VerseML expression, drawn textually, is of \emph{TSM application} form. Here, a TSM named \li{$rx} is applied to the \emph{generalized literal form} \li{/SURLA|T|G|CEURL/}:
+% \subsection{TLM Application}\label{sec:uetsms-usage}
+% The following VerseML expression, drawn textually, is of \emph{TLM application} form. Here, a TLM named \li{$rx} is applied to the \emph{generalized literal form} \li{/SURLA|T|G|CEURL/}:
 % \begin{lstlisting}[numbers=none,mathescape=|]
@@
 Several other outer delimiters are also available, as summarized in Figure \ref{fig:literal-forms}.
-The client is free to choose any of these for use with any TSM, as long as the \emph{literal body} (shown in green above) satisfies the requirements stated in Figure \ref{fig:literal-forms}.
+The client is free to choose any of these for use with any TLM, as long as the \emph{literal body} (shown in green above) satisfies the requirements stated in Figure \ref{fig:literal-forms}.
 For example, we could have equivalently written the example above as \li{$rx `SURLA|T|G|CEURL`}.
@@
 
-% It is only during the subsequent \emph{typed expansion} phase that the applied TSM parses the {body} of the literal form to generate a \emph{proto-expansion}.
+% It is only during the subsequent \emph{typed expansion} phase that the applied TLM parses the {body} of the literal form to generate a \emph{proto-expansion}.
 The language then \emph{validates} this proto-expansion according to criteria that we will describe in Sec.
 \ref{sec:uetsms-validation}.
-If proto-expansion validation succeeds, the language generates the \emph{final expansion} (or more concisely, simply the \emph{expansion}) of the TSM application.
+If proto-expansion validation succeeds, the language generates the \emph{final expansion} (or more concisely, simply the \emph{expansion}) of the TLM application.
 The behavior of the program is determined by its expansion.
 
-% For example, the expansion of the TSM application above is equivalent to the following expression when the regex value constructors \li{Or} and \li{Str} are in scope:
+% For example, the expansion of the TLM application above is equivalent to the following expression when the regex value constructors \li{Or} and \li{Str} are in scope:
 % \begin{lstlisting}[numbers=none]
@@
 % \end{lstlisting}
-% To avoid the assumption that the variables \li{Or} and \li{Str} are in scope at the TSM application site, the expansion actually uses the explicit \li{fold} and \li{inj} operators, as described in Sec.
+% To avoid the assumption that the variables \li{Or} and \li{Str} are in scope at the TLM application site, the expansion actually uses the explicit \li{fold} and \li{inj} operators, as described in Sec.
 \ref{sec:lists}.
 In fact, the proto-expansion validation process enforces this notion of context independence -- we will return to proto-expansion validation below.
-(We will show how TSM parameters can reduce the awkwardness of this requirement in Chapter \ref{chap:ptsms}.)
+(We will show how TLM parameters can reduce the awkwardness of this requirement in Chapter \ref{chap:ptsms}.)
 % %The constructors above are those of the type \li{Rx} that was defined in Figure \ref{fig:datatype-rx}.
@@
 % % A number of literal forms, ,  are available in VerseML's concrete syntax.
-Any literal form can be used with any TSM,  TSMs have access only to the literal bodies.
-Because TSMs do not extend the concrete syntax of the language directly, there cannot be syntactic conflicts between TSMs.
+Any literal form can be used with any TLM,  TLMs have access only to the literal bodies.
+Because TLMs do not extend the concrete syntax of the language directly, there cannot be syntactic conflicts between TLMs.
 
@@
 % \end{figure}
-\subsection{TSM Definitions}\label{sec:uetsms-definition}
+\subsection{TLM Definitions}\label{sec:uetsms-definition}
 %The original expression, above, is statically rewritten to this expression.
-The definition of the seTSM named \lstinline{$html} that was shown being applied in Fig. \ref{fig:first-tsm-example} takes the following form:
+The definition of the seTLM named \lstinline{$html} that was shown being applied in Fig. \ref{fig:first-tsm-example} takes the following form:
 \begin{lstlisting}[mathescape=|]
@@
 
-Every seTSM definition consists of a TSM name, here \li{$html}, a \emph{type annotation}, here \lstinline{at html}, and a \emph{parse function} between \li{by} and \li{in}.
-The TSM definition is in scope between \li{in} and \li{end}.
+Every seTLM definition consists of a TLM name, here \li{$html}, a \emph{type annotation}, here \lstinline{at html}, and a \emph{parse function} between \li{by} and \li{in}.
+The TLM definition is in scope between \li{in} and \li{end}.
 In practice, if an \li{in} clause is not provided, the definition is in scope until the end of the enclosing declaration (e.g.
-the enclosing function or module.) We will consider how TSM definitions are exported from libraries in Sec \ref{sec:static-eval}.
-
-% All TSM names must begin with the dollar symbol (\li{$}), which distinguishes them from variables.
+the enclosing function or module.) We will consider how TLM definitions are exported from libraries in Sec \ref{sec:static-eval}.
+
+% All TLM names must begin with the dollar symbol (\li{$}), which distinguishes them from variables.
 This is inspired by the Rust macro system, which uses post-fix exclamation points (\li{!}) to distinguish macro identifiers \cite{Rust/Macros}.
 
-The {parse function} is a \emph{static function} delegated responsibility over parsing the literal bodies of the literal forms to which the TSM is applied.
-Static functions, marked by the \li{static} keyword, are applied during the typed expansion process, so they cannot refer to the surrounding variable bindings (because those variables stand for dynamic values.) For now, we will simply assume that static functions are closed, and that they do not themselves make use of TSMs (we will eliminate these impractical limitations in Sec.
+The {parse function} is a \emph{static function} delegated responsibility over parsing the literal bodies of the literal forms to which the TLM is applied.
+Static functions, marked by the \li{static} keyword, are applied during the typed expansion process, so they cannot refer to the surrounding variable bindings (because those variables stand for dynamic values.) For now, we will simply assume that static functions are closed, and that they do not themselves make use of TLMs (we will eliminate these impractical limitations in Sec.
 \ref{sec:static-eval}.)
@@
 The return type is a labeled sum type, defined by applying the type function \li{parse_result} defined in Figure \ref{fig:indexrange-and-parseresult}, that distinguishes between parse errors and successful parses.
-%\footnote{\li{parse_result} is defined as a type function because in Chapter \ref{chap:uptsms}, we will introduce pattern TSMs, which generate patterns rather than expressions.}
+%\footnote{\li{parse_result} is defined as a type function because in Chapter \ref{chap:uptsms}, we will introduce pattern TLMs, which generate patterns rather than expressions.}
 Let us consider these two possibilities in turn.
@@
 where $\ecand$ is called the \emph{encoding of the proto-expansion}.
-For expression TSMs, proto-expansions are \emph{proto-expressions}, which are encoded as values of the recursive sum type \lstinline{proto_expr} outlined in Figure \ref{fig:candidate-exp-verseml}.
+For expression TLMs, proto-expansions are \emph{proto-expressions}, which are encoded as values of the recursive sum type \lstinline{proto_expr} outlined in Figure \ref{fig:candidate-exp-verseml}.
 Most of the variants of \li{proto_expr} are individually uninteresting -- they encode VerseML's various expression forms (just as in a self-hosted compiler, e.g. SML/NJ's Visible Compiler library.) 
@@
 based on abstract binding trees \cite{pfpl}, with only minor modifications to our semantics.
-Indeed, when we formally define seTSMs in Sec.
+Indeed, when we formally define seTLMs in Sec.
 \ref{sec:miniVerseU}, we abstract over the particular encoding.
@@
 %One would find types analagous to those just described in any parser, so for concision, we elide the details of \li{$rx}'s parse function.
-%The parse function must treat the TSM parameters parametrically, i.e.
+%The parse function must treat the TLM parameters parametrically, i.e.
 it does not have access to any values in the supplied module parameter.
@@
 \vspace{-5px}
-\caption[Definitions of various types used by TSM definitions.]{Definitions of various types used by TSM definitions. We assume that these type definitions are available ambiently.}
+\caption[Definitions of various types used by TLM definitions.]{Definitions of various types used by TLM definitions. We assume that these type definitions are available ambiently.}
 \label{fig:indexrange-and-parseresult}
@@
 Instead,
-the TSM must refer to the spliced expression \emph{by its relative location} using the \li{SplicedE} variant of \li{proto_expr}, which takes a value of type \li{segment} that indicates the zero-indexed location of the spliced expression relative to the start of the provided literal body.
+the TLM must refer to the spliced expression \emph{by its relative location} using the \li{SplicedE} variant of \li{proto_expr}, which takes a value of type \li{segment} that indicates the zero-indexed location of the spliced expression relative to the start of the provided literal body.
 The \li{SplicedE} variant also requires a value of type \li{proto_typ}, which indicates the type that the spliced expression is expected to have.
@@
 
-For example, consider again the TSM application on Line 1 of Fig. \ref{fig:first-tsm-example}, reproduced below:
+For example, consider again the TLM application on Line 1 of Fig. \ref{fig:first-tsm-example}, reproduced below:
 \begin{lstlisting}[numbers=none]
@@
 
-Requiring that TSMs refer to spliced expressions indirectly in this manner prevents them from ``forging'' spliced terms (i.e.
+Requiring that TLMs refer to spliced expressions indirectly in this manner prevents them from ``forging'' spliced terms (i.e.
 claiming that an term is a spliced term when it does not in fact appear textually in the literal body.)% Our semantics distinguishes spliced expressions when validating the proto-expansion, which is important for reasons of hygiene that we will detail shortly.
@@
 
-For TSM providers, the benefit of this mechanism is that they can name the variables used internally within expansions freely, without worrying about whether their chosen identifiers might shadow those that a client might have used at the application site.
+For TLM providers, the benefit of this mechanism is that they can name the variables used internally within expansions freely, without worrying about whether their chosen identifiers might shadow those that a client might have used at the application site.
 There is no need for a mechanism that generates ``fresh variables''.
 
-TSM clients can, in turn, reliably reason about binding within every spliced expression without examining the expansion that the spliced expression appears within.
+TLM clients can, in turn, reliably reason about binding within every spliced expression without examining the expansion that the spliced expression appears within.
 
@@
 %The prohibition on shadowing ensures only that variables that appear in spliced terms do not refer to bindings that appear in the surrounding expansion. 
-The proto-expansion validation process also ensures that variables that appear in the proto-expansion do not refer to bindings that appear either at the TSM definition or the application site.
+The proto-expansion validation process also ensures that variables that appear in the proto-expansion do not refer to bindings that appear either at the TLM definition or the application site.
 In other words, expansions must be completely \emph{context independent} -- they can make no assumptions about the surrounding context.
 
-A minimal example of a ``broken'' TSM that does not generate context-independent proto-expansions is below:
+A minimal example of a ``broken'' TLM that does not generate context-independent proto-expansions is below:
 \begin{lstlisting}[numbers=none]
@@
 \end{lstlisting}
-The proto-expansion that this TSM generates (for any given literal body) refers to a variable \li{x} that it does not itself bind.
+The proto-expansion that this TLM generates (for any given literal body) refers to a variable \li{x} that it does not itself bind.
 If proto-expansion validation permitted such a proto-expansion, it would be well-typed only under those application site typing contexts where \li{x} is bound.
@@
 Because proto-expansions refer to spliced terms indirectly, enforcing context independence is straightforward -- we need only that the proto-expansion itself be closed, without considering the spliced terms.% In the next section, we will formalize this intuition.
-% The TSM provider can only refer to them opaquely.
+% The TLM provider can only refer to them opaquely.
 
@@
 
-% This prohibition on context dependence explains why the expansion generated by the TSM application in Sec.
+% This prohibition on context dependence explains why the expansion generated by the TLM application in Sec.
 \ref{sec:uetsms-usage} cannot make use of the regex value constructors, e.g.
@@
 \li{Rx}, due to the type ascription on \li{$rx} -- determines which type of value will be constructed.
-For dialects of ML where datatype definitions do introduce new variables or scoped symbols, we need parameterized TSMs.
+For dialects of ML where datatype definitions do introduce new variables or scoped symbols, we need parameterized TLMs.
 We will return to this topic in Chapter \ref{chap:ptsms}.
@@
 \begin{enumerate}
-\item checking to ensure that the generated expansion is of the type specified in the TSM's type annotation; %For example, the type annotation on \li{$rx} is \li{at rx}, so proto-expansion validation ensures that the final expansion is of type \li{rx}.
+\item checking to ensure that the generated expansion is of the type specified in the TLM's type annotation; %For example, the type annotation on \li{$rx} is \li{at rx}, so proto-expansion validation ensures that the final expansion is of type \li{rx}.
 \item checking that the each spliced type is valid at the application site; 
@@
  % \begin{enumerate}
- %   \item determining the type of an expansion requires examining only the type annotation on the TSM definition (much as determining the type of a function application requires examining only the function's type); and
+ %   \item determining the type of an expansion requires examining only the type annotation on the TLM definition (much as determining the type of a function application requires examining only the function's type); and
  % \item determining the type that a spliced expression must have requires only the information in the splice summary (rather than complete knowledge of the proto-expansion).
@@
 % The language \emph{validates} proto-expansions before a final expansion is generated.
-One aspect of proto-expansion validation is checking  the proto-expansion against the type annotation specified by the TSM, e.g.
+One aspect of proto-expansion validation is checking  the proto-expansion against the type annotation specified by the TLM, e.g.
 the type \li{Rx} in the example above.
-This maintains a \emph{type discipline}: if a programmer sees a TSM being applied when examining a well-typed program, they need only look up the TSM's type annotation to determine the type of the generated expansion.
+This maintains a \emph{type discipline}: if a programmer sees a TLM being applied when examining a well-typed program, they need only look up the TLM's type annotation to determine the type of the generated expansion.
 Determining the type does not require examining the expansion directly.
@@
 
-\newcommand{\spTSMsSec}{Simple Pattern TSMs (spTSMs)}
-\section{\protect\spTSMsSec}
+\newcommand{\spTLMsSec}{Simple Pattern TLMs (spTLMs)}
+\section{\protect\spTLMsSec}
 \label{sec:sptsms}
-Let us now briefly consider the topic of TSMs that expand to \emph{patterns}, rather than expressions.
-For example, we can pattern match on a value \li{x : html} by applying a pattern TSM \li{$html} as follows:
+Let us now briefly consider the topic of TLMs that expand to \emph{patterns}, rather than expressions.
+For example, we can pattern match on a value \li{x : html} by applying a pattern TLM \li{$html} as follows:
 \begin{lstlisting}[numbers=none]
@@
 
-Pattern TSM definitions look much like expression TSM definitions:
+Pattern TLM definitions look much like expression TLM definitions:
 \begin{lstlisting}
@@
 \end{lstlisting}
-The \emph{sort qualifier}, \li{for patterns}, indicates that this is a pattern TSM definition.
-Expression TSM definitions can also include a sort qualifier, \li{for expressions}, but this is the default if omitted.
-Note that we used the same name, \li{$html}, for this pattern TSM as for the expression TSM defined in the previous section.
+The \emph{sort qualifier}, \li{for patterns}, indicates that this is a pattern TLM definition.
+Expression TLM definitions can also include a sort qualifier, \li{for expressions}, but this is the default if omitted.
+Note that we used the same name, \li{$html}, for this pattern TLM as for the expression TLM defined in the previous section.
 This is possible because patterns and expressions are distinct sorts of terms.
-It does not make sense to apply an expression TSM in pattern position, and \emph{vice versa}.\footnote{The fact that some patterns look like expressions in the textual syntax is immaterial to this fundamental semantic distinction.
+It does not make sense to apply an expression TLM in pattern position, and \emph{vice versa}.\footnote{The fact that some patterns look like expressions in the textual syntax is immaterial to this fundamental semantic distinction.
 Additionally, many expression-level constructs, e.g.
@@
 \subsubsection{Binding} To maintain a reasonable abstract binding discipline, i.e.
-to allow clients to reason about variable binding without examining pattern TSM expansions directly, variable patterns can appear only within spliced patterns.
+to allow clients to reason about variable binding without examining pattern TLM expansions directly, variable patterns can appear only within spliced patterns.
 Enforcing this restriction is straightforward: we simply have not defined a variant of the \li{proto_pat} type that encodes variable patterns (though wildcards are allowed.)  This prohibition ensures that no variables other than those visible to the client in a spliced pattern are captured by the corresponding branch of the match expression.
@@
 \begin{enumerate}
-  \item that the final expansion is a pattern that matches values of the type specified in the TSM type annotation;
+  \item that the final expansion is a pattern that matches values of the type specified in the TLM type annotation;
   \item that each spliced pattern matches values of the type indicated in the splice summary; and 
@@
 \end{enumerate}
-\newcommand{\seTSMsFormallySec}{Simple TSMs, Formally}
-\section{\protect\seTSMsFormallySec}
+\newcommand{\seTLMsFormallySec}{Simple TLMs, Formally}
+\section{\protect\seTLMsFormallySec}
 \label{sec:setsms-formally}
@@
 % %& \audefuetsm{\utau}{e}{\tsmv}{\ue} 
-% & \uesyntax{\tsmv}{\utau}{e}{\ue} & \text{seTSM definition}\\ 
+% & \uesyntax{\tsmv}{\utau}{e}{\ue} & \text{seTLM definition}\\ 
 % &&
 % %& \autsmap{b}{\tsmv} 
-% & \utsmap{\tsmv}{b} & \text{seTSM application}\ECC
+% & \utsmap{\tsmv}{b} & \text{seTLM application}\ECC
 \end{array}$
@@
 \caption[Syntax of the $\miniVersePat$ unexpanded language (UL)]{Syntax of the $\miniVersePat$ unexpanded language (UL).
-Metavariable $\ut$ ranges over type identifiers, $\ux$ over expression identifiers, $\ell$ over labels, $\labelset$ over finite sets of labels, $\tsmv$ over TSM names and $b$ over literal bodies.
+Metavariable $\ut$ ranges over type identifiers, $\ux$ over expression identifiers, $\ell$ over labels, $\labelset$ over finite sets of labels, $\tsmv$ over TLM names and $b$ over literal bodies.
 We write $\mapschema{\utau}{i}{\labelset}$ for a finite mapping of each label $i$ in $\labelset$ to some unexpanded type $\utau_i$, and similarly for other sorts.
@@
 
-Before continuing on to consider parametric TSMs, let us develop a calculus of simple expression and pattern TSMs called $\miniVersePat$.
+Before continuing on to consider parametric TLMs, let us develop a calculus of simple expression and pattern TLMs called $\miniVersePat$.
 It is organized as an \emph{unexpanded language}, or \emph{UL}, defined by typed expansion to an \emph{expanded language}, or \emph{XL}.
@@
 This distinction between identifiers and variables is technically crucial.
-%We \textbf{cannot} adopt the usual definitions of $\alpha$-renaming of identifiers, because unexpanded types and expressions are still in a ``partially parsed'' state -- the literal bodies, $b$, within an unexpanded expression might contain spliced subterms that are ``surfaced'' by a TSM only during typed expansion, as we will detail below.
+%We \textbf{cannot} adopt the usual definitions of $\alpha$-renaming of identifiers, because unexpanded types and expressions are still in a ``partially parsed'' state -- the literal bodies, $b$, within an unexpanded expression might contain spliced subterms that are ``surfaced'' by a TLM only during typed expansion, as we will detail below.
 %identifiers are given meaning by expansion to variables.
@@
 
-% There are only two unexpanded expression forms, highlighted in gray in Figure \ref{fig:U-unexpanded-terms}, that do not correspond to expanded expression forms -- the seTSM definition form and the seTSM application form.
+% There are only two unexpanded expression forms, highlighted in gray in Figure \ref{fig:U-unexpanded-terms}, that do not correspond to expanded expression forms -- the seTLM definition form and the seTLM application form.
 %These are the ``interesting'' forms.
@@
 The details are in the supplement.
-The  rules of particular interest are the rules governing TSM definitions and TSM application, which are covered in the next two subsections.
+The  rules of particular interest are the rules governing TLM definitions and TLM application, which are covered in the next two subsections.
 % \endgroup
@@
 % The type assigned in the conclusion of each rule above is identical to the type assigned in the conclusion of the corresponding typing rule.
-The seTSM context, $\uPsi$, passes opaquely through these rules (we will define seTSM contexts below.) As such, the corresponding cases in the proof of Theorem \ref{thm:typed-expansion-short-U} are by application of the induction hypothesis and the  corresponding typing rule.
+The seTLM context, $\uPsi$, passes opaquely through these rules (we will define seTLM contexts below.) As such, the corresponding cases in the proof of Theorem \ref{thm:typed-expansion-short-U} are by application of the induction hypothesis and the  corresponding typing rule.
 %Rules (\ref{rules:expandsTU}) could similarly have been generated by mechanically transforming Rules (\ref{rules:istypeU}).
@@
 The premises also correspond -- if a typing judgement appears as a premise of a typing rule, then the corresponding premise in the corresponding typed expansion rule is the corresponding typed expansion judgement.
-The seTSM context is not extended or inspected by these rules (it is only ``threaded through'' them opaquely).
-
-%There are two unexpanded expression forms that do not correspond to an expanded expression form: the seTSM definition form, and the seTSM application form.
-The rules governing these two forms interact with the seTSM context, and are the topics of the next two subsections, respectively.
-
-\subsection{TSM Definitions}\label{sec:U-uetsm-definition}
-An unexpanded expression of seTSM definition form, $\uesyntaxq{\tsmv}{\utau}{\eparse}{\ue}$, 
+The seTLM context is not extended or inspected by these rules (it is only ``threaded through'' them opaquely).
+
+%There are two unexpanded expression forms that do not correspond to an expanded expression form: the seTLM definition form, and the seTLM application form.
+The rules governing these two forms interact with the seTLM context, and are the topics of the next two subsections, respectively.
+
+\subsection{TLM Definitions}\label{sec:U-uetsm-definition}
+An unexpanded expression of seTLM definition form, $\uesyntaxq{\tsmv}{\utau}{\eparse}{\ue}$, 
 %The operational form corresponding to this stylized form is \[\audefuetsm{\utau}{\eparse}{\tsmv}{\ue}\]
- defines an {seTSM} identified as $\tsmv$ with \emph{unexpanded type annotation} $\utau$ and \emph{parse function} $\eparse$ for use within $\ue$. The rule below defines typed expansion of this form:
+ defines an {seTLM} identified as $\tsmv$ with \emph{unexpanded type annotation} $\utau$ and \emph{parse function} $\eparse$ for use within $\ue$. The rule below defines typed expansion of this form:
 % \begin{subequations}[resume]
@@
 This is not semantically necessary, but it is the choice one would expect to make assuming an eagerly evaluated language.
-\item The final premise of Rule \textsc{ee-def-setsm} extends the seTSM context, $\uPsi$, with the newly determined {seTSM definition}, and proceeds to assign a type, $\tau'$, and expansion, $e$, to $\ue$.
-The conclusion of the rule then assigns this type and expansion to the seTSM definition as a whole.% i.e.
-TSMs define behavior that is relevant during typed expansion, but not during evaluation.
-
-\emph{seTSM contexts}, $\uPsi$, are of the form $\uAS{\uA}{\Psi}$, where $\uA$ is a \emph{TSM identifier expansion context} and $\Psi$ is an \emph{seTSM definition context}.
-
-A {TSM identifier expansion context}, $\uA$, is a finite function mapping each TSM identifier $\tsmv \in \domof{\uA}$ to the \emph{TSM identifier expansion}, $\vExpands{\tsmv}{a}$, for some \emph{TSM name}, $a$.
-We distinguish TSM identifiers, $\tsmv$, from TSM names, $a$, for much the same reason that we distinguish type and expression identifiers from type and expression variables: in order to allow a TSM definition to shadow a previously defined TSM definition without relying on an implicit identification convention.
-
-An {seTSM definition context}, $\Psi$, is a finite function mapping each TSM name $a \in \domof{\Psi}$ to an \emph{expanded seTSM definition}, $\xuetsmbnd{a}{\tau}{\eparse}$, where $\tau$ is the seTSM's type annotation, and $\eparse$ is its parse function.
+\item The final premise of Rule \textsc{ee-def-setsm} extends the seTLM context, $\uPsi$, with the newly determined {seTLM definition}, and proceeds to assign a type, $\tau'$, and expansion, $e$, to $\ue$.
+The conclusion of the rule then assigns this type and expansion to the seTLM definition as a whole.% i.e.
+TLMs define behavior that is relevant during typed expansion, but not during evaluation.
+
+\emph{seTLM contexts}, $\uPsi$, are of the form $\uAS{\uA}{\Psi}$, where $\uA$ is a \emph{TLM identifier expansion context} and $\Psi$ is an \emph{seTLM definition context}.
+
+A {TLM identifier expansion context}, $\uA$, is a finite function mapping each TLM identifier $\tsmv \in \domof{\uA}$ to the \emph{TLM identifier expansion}, $\vExpands{\tsmv}{a}$, for some \emph{TLM name}, $a$.
+We distinguish TLM identifiers, $\tsmv$, from TLM names, $a$, for much the same reason that we distinguish type and expression identifiers from type and expression variables: in order to allow a TLM definition to shadow a previously defined TLM definition without relying on an implicit identification convention.
+
+An {seTLM definition context}, $\Psi$, is a finite function mapping each TLM name $a \in \domof{\Psi}$ to an \emph{expanded seTLM definition}, $\xuetsmbnd{a}{\tau}{\eparse}$, where $\tau$ is the seTLM's type annotation, and $\eparse$ is its parse function.
 We define $\uPsi, \uShyp{\tsmv}{a}{\tau}{\eparse}$, when $\uPsi=\uAS{\uA}{\Psi}$, as an abbreviation of $\uAS{\ctxUpdate{\uA}{\tsmv}{a}}{\Psi, \xuetsmbnd{a}{\tau}{\eparse}}$.
 
-%Moreover, this distinction will be crucial in the semantics of TSM abbreviations in Chapter \ref{chap:ptsms}.
+%Moreover, this distinction will be crucial in the semantics of TLM abbreviations in Chapter \ref{chap:ptsms}.
 
@@
 
-The spTSM definition form, $\usyntaxup{\tsmv}{\utau}{\eparse}{\ue}$, operates analagously.
-The rule is reproduced below, and the details, which mirror those for seTSMs, are in the supplement.
-Notice that the spTSM context, $\uPhi$, rather than the $\uPsi$ is updated.
-This allows expression and pattern TSMs to share an identifier.
+The spTLM definition form, $\usyntaxup{\tsmv}{\utau}{\eparse}{\ue}$, operates analagously.
+The rule is reproduced below, and the details, which mirror those for seTLMs, are in the supplement.
+Notice that the spTLM context, $\uPhi$, rather than the $\uPsi$ is updated.
+This allows expression and pattern TLMs to share an identifier.
 \begin{mathpar}
@@
 
-\subsection{TSM Application}\label{sec:U-uetsm-application}
-The unexpanded expression form for applying an seTSM named $\tsmv$ to a literal form with literal body $b$ is $\utsmap{\tsmv}{b}$. 
+\subsection{TLM Application}\label{sec:U-uetsm-application}
+The unexpanded expression form for applying an seTLM named $\tsmv$ to a literal form with literal body $b$ is $\utsmap{\tsmv}{b}$. 
 % This form uses forward slashes\todo{use backticks} to delimit the literal body, but other generalized literal forms could also be included as derived forms in the textual syntax.
@@
 %i.e.
-for each literal body $b$, the operator $\texttt{uapuetsm}[b]$ is indexed by the TSM name $\tsmv$ and takes no arguments.
+for each literal body $b$, the operator $\texttt{uapuetsm}[b]$ is indexed by the TLM name $\tsmv$ and takes no arguments.
 %\footnote{This is in following the conventions in \emph{PFPL} \cite{pfpl}, where operators parameters allow for the use of metatheoretic objects that are not syntax trees or binding trees, e.g.
-$\mathsf{str}[s]$ and $\mathsf{num}[n]$.} This operator is indexed by the TSM name $\tsmv$ and takes no arguments.
-The typed expansion rule governing seTSM application is below:
+$\mathsf{str}[s]$ and $\mathsf{num}[n]$.} This operator is indexed by the TLM name $\tsmv$ and takes no arguments.
+The typed expansion rule governing seTLM application is below:
 % \begin{subequations}[resume]
@@
 If parsing succeeds, i.e.
-a value of the form $\aein{\mathtt{SuccessE}}{\ecand}$ results from evaluation, then $\ecand$ will be a value of type $\tCEExp$ (assuming a well-formed seTSM context, by application of the Preservation assumption.) We call $\ecand$ the \emph{encoding of the proto-expansion}.
+a value of the form $\aein{\mathtt{SuccessE}}{\ecand}$ results from evaluation, then $\ecand$ will be a value of type $\tCEExp$ (assuming a well-formed seTLM context, by application of the Preservation assumption.) We call $\ecand$ the \emph{encoding of the proto-expansion}.
 If the parse function produces a value labeled $\lbltxt{ParseError}$, then typed expansion fails.
@@
 
-The typed pattern expansion rule governing spTSM application is analagous:
+The typed pattern expansion rule governing spTLM application is analagous:
 \begin{mathpar}
@@
 \end{array}\]
-The purpose of the \emph{splicing scenes} $\tscenev$, $\escenev$ and $\pscenev$ is to ``remember'', during proto-expansion validation, the contexts and the literal body from the TSM application site (cf.
+The purpose of the \emph{splicing scenes} $\tscenev$, $\escenev$ and $\pscenev$ is to ``remember'', during proto-expansion validation, the contexts and the literal body from the TLM application site (cf.
 Rules \textsc{ee-ap-setsm} and \textsc{pe-ap-sptsm}) for when validation encounters spliced terms.
@@
 Notice that proto-rules, $\crv$, involve expanded patterns, $p$, not proto-patterns, $\cpv$.
-The reason is that proto-rules appear in proto-expressions, which are generated by expression TSMs.
-Proto-patterns, in contrast, arise only from pattern TSMs.
+The reason is that proto-rules appear in proto-expressions, which are generated by expression TLMs.
+Proto-patterns, in contrast, arise only from pattern TLMs.
 There a variable proto-pattern form.
@@
 The intuition is that parsing a term of out a literal body cannot produce a bigger term than the term that contained that very literal body.
-More specifically, {the sum of the lengths of the literal bodies that appear in the term strictly decreases each time you perform a nested TSM application} because some portion of the term has to be consumed by the TSM name and the delimiters.
+More specifically, {the sum of the lengths of the literal bodies that appear in the term strictly decreases each time you perform a nested TLM application} because some portion of the term has to be consumed by the TLM name and the delimiters.
 The details are given in the supplemental material.
@@
 
-\subsubsection{seTSM Reasoning Principles} 
-The following theorem summarizes the abstract reasoning principles that programmers can rely on when applying an seTSM. Informal descriptions of important clauses are given inline. 
+\subsubsection{seTLM Reasoning Principles} 
+The following theorem summarizes the abstract reasoning principles that programmers can rely on when applying an seTLM. Informal descriptions of important clauses are given inline. 
 % \begingroup
 % \def\thetheorem{\ref{thm:tsc-SES}}
-\begin{theorem}[seTSM Abstract Reasoning Principles]
+\begin{theorem}[seTLM Abstract Reasoning Principles]
 If $\expandsU{\uDD{\uD}{\Delta}}{\uGG{\uG}{\Gamma}}{\uPsi}{\utsmap{\tsmv}{b}}{e}{\tau}$ then:
@@
   \begin{quote}
-     The type of the expansion is consistent with the type annotation on the applied seTSM definition.
+     The type of the expansion is consistent with the type annotation on the applied seTLM definition.
   \end{quote}
@@
         \end{quote}
-\item $\summaryOf{\ce} = \sseq{\acesplicedt{m'_i}{n'_i}}{\nty} \cup \sseq{\acesplicede{m_i}{n_i}{\ctau_i}}{\nexp}$
+\item $\segof{\ce} = \sseq{\acesplicedt{m'_i}{n'_i}}{\nty} \cup \sseq{\acesplicede{m_i}{n_i}{\ctau_i}}{\nexp}$
 \item \textbf{(Typing 2)} $\sseq{
@@
 
-\subsubsection{spTSM Reasoning Principles} Finally, the following theorem summarizes the abstract reasoning principles available to programmers when applying an spTSM.
+\subsubsection{spTLM Reasoning Principles} Finally, the following theorem summarizes the abstract reasoning principles available to programmers when applying an spTLM.
 Most of the labeled clauses are analagous to those described above, so we omit their descriptions.
 % \begingroup
-% \def\thetheorem{\ref{thm:spTSM-Typing-Segmentation}}
-\begin{theorem}[spTSM Abstract Reasoning Principles]
-% \label{thm:spTSM-Typing-Segmentation}
+% \def\thetheorem{\ref{thm:spTLM-Typing-Segmentation}}
+\begin{theorem}[spTLM Abstract Reasoning Principles]
+% \label{thm:spTLM-Typing-Segmentation}
 If $\patExpands{\upctx}{\uPhi}{\utsmap{\tsmv}{b}}{p}{\tau}$ where $\uDelta=\uDD{\uD}{\Delta}$ and $\uGamma=\uGG{\uG}{\Gamma}$ then all of the following hold:
@@
         \item (\textbf{Segmentation}) $\segOK{\segof{\cpv}}{b}$
-        \item $\summaryOf{\cpv} = \sseq{\acesplicedt{n'_i}{m'_i}}{\nty} \cup \sseq{\acesplicedp{m_i}{n_i}{\ctau_i}}{\npat}$
+        \item $\segof{\cpv} = \sseq{\acesplicedt{n'_i}{m'_i}}{\nty} \cup \sseq{\acesplicedp{m_i}{n_i}{\ctau_i}}{\npat}$
         \item (\textbf{Typing 2}) $\sseq{
@@
         \begin{quote}
-          The hypotheses generated by the TSM application are exactly those generated by the spliced patterns.
+          The hypotheses generated by the TLM application are exactly those generated by the spliced patterns.
         \end{quote}
@@
 
-% There are two ``interesting'' proto-expansion forms, highlighted in yellow in Figure \ref{fig:U-candidate-terms}: a proto-type form for \emph{references to spliced unexpanded types}, $\acesplicedt{m}{n}$, and a proto-expression form for \emph{references to spliced unexpanded expressions}, $\acesplicede{m}{n}{\ctau}$, where $m$ and $n$ are natural numbers.%TSM utilize these to splice types and unexpanded expressions out of literal bodies.
+% There are two ``interesting'' proto-expansion forms, highlighted in yellow in Figure \ref{fig:U-candidate-terms}: a proto-type form for \emph{references to spliced unexpanded types}, $\acesplicedt{m}{n}$, and a proto-expression form for \emph{references to spliced unexpanded expressions}, $\acesplicede{m}{n}{\ctau}$, where $m$ and $n$ are natural numbers.%TLM utilize these to splice types and unexpanded expressions out of literal bodies.
 
@@
 %This achieves \emph{context-independent expansion} as described in Sec.
-\ref{sec:splicing-and-hygiene} for type variables -- seTSMs cannot impose ``hidden constraints'' on the application site unexpanded type formation context, because the type variables bound at the application site are simply not directly available to proto-types.
+\ref{sec:splicing-and-hygiene} for type variables -- seTLMs cannot impose ``hidden constraints'' on the application site unexpanded type formation context, because the type variables bound at the application site are simply not directly available to proto-types.
 
@@
 % Together, these two premises enforce the injunction on type variable capture as described in Sec.
-\ref{sec:uetsms-validation} -- the TSM provider can choose type variable names freely within a proto-expansion.
+\ref{sec:uetsms-validation} -- the TLM provider can choose type variable names freely within a proto-expansion.
 We will consider this formally in Sec.
 \ref{sec:SE-metatheory} below.
-%, because the language prevents them from shadowing type variables at the application site (by $\alpha$-varying the proto-expansion as needed.)%Such a change in bound variable names is possible again because variables bound by the seTSM provider in a proto-expansion cannot ``leak into'' spliced terms because the hypotheses in $\Delta$ are not made available to the spliced type, $\tau$.
+%, because the language prevents them from shadowing type variables at the application site (by $\alpha$-varying the proto-expansion as needed.)%Such a change in bound variable names is possible again because variables bound by the seTLM provider in a proto-expansion cannot ``leak into'' spliced terms because the hypotheses in $\Delta$ are not made available to the spliced type, $\tau$.
 
@@
 This achieves \emph{context independence} as described in Sec.
-\ref{sec:uetsms-validation} -- seTSMs cannot impose ``hidden constraints'' on the application site unexpanded typing context, because the variable bindings at the application site are not directly available to proto-expansions.
+\ref{sec:uetsms-validation} -- seTLMs cannot impose ``hidden constraints'' on the application site unexpanded typing context, because the variable bindings at the application site are not directly available to proto-expansions.
 We will consider this formally in Sec.
@@
 % Together, these premises enforce the prohibition on capture as described in Sec.
-\ref{sec:uetsms-validation} -- the TSM provider can choose variable names freely within a proto-expansion, because the language prevents them from shadowing those at the application site.
+\ref{sec:uetsms-validation} -- the TLM provider can choose variable names freely within a proto-expansion, because the language prevents them from shadowing those at the application site.
 % %\end{subequations}
@@
 
-% %Candidate expansions cannot themselves define or apply TSMs.
-This simplifies our metatheory, though it can be inconvenient at times for TSM providers.
-We discuss adding the ability to use TSMs within proto-expansions in Sec.
+% %Candidate expansions cannot themselves define or apply TLMs.
+This simplifies our metatheory, though it can be inconvenient at times for TLM providers.
+We discuss adding the ability to use TLMs within proto-expansions in Sec.
 \ref{sec:tsms-in-expansions}.
@@
 % The proof of part 1 proceeds by inducting over the typed expansion assumption.
-The only interesting cases are those related to seTSM definition and application, reproduced below.
+The only interesting cases are those related to seTLM definition and application, reproduced below.
 In the following cases, let $\uDelta=\uDD{\uD}{\Delta}$ and $\uGamma=\uGG{\uG}{\Gamma}$ and $\uPsi=\uAS{\uA}{\Psi}$.
@@
 %  \item \istypeU{\Delta}{\tau'} \BY{Lemma \ref{lemma:type-expansion-U} to \pfref{expandsTU}} \pflabel{istype}
-% %  \item \uetsmenv{\Delta}{\Psi, \xuetsmbnd{\tsmv}{\tau'}{\eparse}} \BY{Definition \ref{def:seTSM-def-ctx-formation} on \pfref{uetsmenv1}, \pfref{istype} and \pfref{eparse}}\pflabel{uetsmenv3}
+% %  \item \uetsmenv{\Delta}{\Psi, \xuetsmbnd{\tsmv}{\tau'}{\eparse}} \BY{Definition \ref{def:seTLM-def-ctx-formation} on \pfref{uetsmenv1}, \pfref{istype} and \pfref{eparse}}\pflabel{uetsmenv3}
 %   \item \hastypeU{\Delta}{\Gamma}{e}{\tau} \BY{IH, part 1(a) on \pfref{expandsU}}
@@
 % This is established by appeal to the following two conditions.
-The first condition states that an unexpanded expression constructed by parsing a textual sequence $b$ is strictly smaller, as measured by the metric defined above, than the length of $b$, because some characters must necessarily be used to invoke a TSM and delimit each literal body.
+The first condition states that an unexpanded expression constructed by parsing a textual sequence $b$ is strictly smaller, as measured by the metric defined above, than the length of $b$, because some characters must necessarily be used to invoke a TLM and delimit each literal body.
 % \begingroup
@@
 % \subsubsection{Abstract Reasoning Principles}\label{sec:uetsms-reasoning-principles}
-% The following theorem summarizes the abstract reasoning principles that programmers can rely on when applying an seTSM. In particular, the programmer can be sure that:
+% The following theorem summarizes the abstract reasoning principles that programmers can rely on when applying an seTLM. In particular, the programmer can be sure that:
 % \begin{enumerate} 
 % \item \textbf{Segmentation}: The segmentation determined by the proto-expansion actually segments the literal body (i.e. each segment is in-bounds and the segments are non-overlapping.)
-% \item \textbf{Typing 1}: The type of the expansion is consistent with the type annotation on the seTSM definition.
+% \item \textbf{Typing 1}: The type of the expansion is consistent with the type annotation on the seTLM definition.
 % \item \textbf{Typing 2}: Each spliced type has a well-formed expansion at the application site.
@@
 % \def\thetheorem{\ref{thm:tsc-SES}}
-% \begin{theorem}[seTSM Abstract Reasoning Principles]
+% \begin{theorem}[seTLM Abstract Reasoning Principles]
 % If $\expandsU{\uDD{\uD}{\Delta}}{\uGG{\uG}{\Gamma}}{\uPsi}{\utsmap{\tsmv}{b}}{e}{\tau}$ then:
@@
 % \item (\textbf{Segmentation}) $\segOK{\segof{\ce}}{b}$
-% \item $\summaryOf{\ce} = \sseq{\acesplicedt{m'_i}{n'_i}}{\nty} \cup \sseq{\acesplicede{m_i}{n_i}{\ctau_i}}{\nexp}$
+% \item $\segof{\ce} = \sseq{\acesplicedt{m'_i}{n'_i}}{\nty} \cup \sseq{\acesplicede{m_i}{n_i}{\ctau_i}}{\nexp}$
 % \item \textbf{(Typing 2)} $\sseq{
@@
 
-\newcommand{\pTSMsSec}{Parametric TSMs (pTSMs)}
-\section{\protect\pTSMsSec}
+\newcommand{\pTLMsSec}{Parametric TLMs (pTLMs)}
+\section{\protect\pTLMsSec}
 \label{sec:ptsms}
-Simple TSMs operate at a single specified type and the proto-expansions they generate must be closed.
+Simple TLMs operate at a single specified type and the proto-expansions they generate must be closed.
 This simplifies matters formally, but it is not convenient in practice.
-This section introduces \emph{parametric TSMs}.
-Parametric TSMs can be defined over a type- and module-parameterized family of types, and the proto-expansions they generate can refer to supplied type and module parameters.
-
-\subsection{Parametric TSMs By Example}\label{sec:ptsms-by-example}
+This section introduces \emph{parametric TLMs}.
+Parametric TLMs can be defined over a type- and module-parameterized family of types, and the proto-expansions they generate can refer to supplied type and module parameters.
+
+\subsection{Parametric TLMs By Example}\label{sec:ptsms-by-example}
 
@@
 \end{lstlisting}
-Let us now define a TSM that is parametric over all implementations of this signature, \li{D : DICT}, and also over all choices of type \li{'a}:
+Let us now define a TLM that is parametric over all implementations of this signature, \li{D : DICT}, and also over all choices of type \li{'a}:
 \begin{lstlisting}[numbers=none]
@@
 \end{lstlisting}
-Notice that the segmentation immediately reveals which punctuation is particular to this TSM and where the spliced key and value expressions appear.
+Notice that the segmentation immediately reveals which punctuation is particular to this TLM and where the spliced key and value expressions appear.
 Because the context-free syntax of unexpanded terms is never modified, it is possible to reason modularly about syntactic determinism (i.e.
@@
 
-If we will use the \li{Dict} implementation ubiquitously, we can abbreviate the partial application of \li{$dict} to \li{Dict}, resulting in a TSM that is parametric over only the type \li{'a}, as follows:
+If we will use the \li{Dict} implementation ubiquitously, we can abbreviate the partial application of \li{$dict} to \li{Dict}, resulting in a TLM that is parametric over only the type \li{'a}, as follows:
 \begin{lstlisting}[numbers=none]
@@
 \end{lstlisting}
-The proto-expansion generated for the TSM application above might be:
+The proto-expansion generated for the TLM application above might be:
 \begin{lstlisting}[numbers=none]
@@
 With module parameters, it is possible to sidestep this problem by passing in a module containing the datatype constructors.
-In practice, the language might implicitly construct and open such a module within the expansion when the type annotation on the TSM is such a datatype.
+In practice, the language might implicitly construct and open such a module within the expansion when the type annotation on the TLM is such a datatype.
 
@@
 
-\subsection{Parametric TSMs, Formally}
+\subsection{Parametric TLMs, Formally}
 % \begin{figure}[p]
@@
 % %& \abindref{\tsmv} 
-% & \tsmv & \text{TSM identifier reference}\\
+% & \tsmv & \text{TLM identifier reference}\\
 % % &&
@@
 % \]
-% \caption{Syntax of unexpanded TSM types and expressions.}
+% \caption{Syntax of unexpanded TLM types and expressions.}
 % \label{fig:P-macro-expressions-types-u}
@@
 
-We will now outline $\miniVerseParam$, a calculus that extends $\miniVersePat$ with support for parametric TSMs.
+We will now outline $\miniVerseParam$, a calculus that extends $\miniVersePat$ with support for parametric TLMs.
 This calculus is organized, like $\miniVersePat$, as an unexpanded language (UL) defined by typed expansion to an expanded language (XL).
@@
 
-The main conceptual difference between $\miniVersePat$ and $\miniVerseParam$ is that $\miniVerseParam$ introduces the notion of unexpanded and expanded TSM expressions and types, as shown in Figure \ref{fig:P-macro-expressions-types}.
+The main conceptual difference between $\miniVersePat$ and $\miniVerseParam$ is that $\miniVerseParam$ introduces the notion of unexpanded and expanded TLM expressions and types, as shown in Figure \ref{fig:P-macro-expressions-types}.
 
@@
 \vspace{-6px}
-\caption{Syntax of unexpanded and expanded TSM types and expressions in $\miniVerseParam$}
+\caption{Syntax of unexpanded and expanded TLM types and expressions in $\miniVerseParam$}
 \label{fig:P-macro-expressions-types}
@@
 
-The rule governing expression TSM application, reproduced below, touches all of the main ideas in $\miniVerseParam$. %It might be useful to compare this rule to Rule \textsc{ee-ap-sptsm}.
+The rule governing expression TLM application, reproduced below, touches all of the main ideas in $\miniVerseParam$. %It might be useful to compare this rule to Rule \textsc{ee-ap-sptsm}.
 \begin{mathpar}
@@
 \end{mathpar}
-The first two premises simply deconstruct the (unified) unexpanded context $\uOmega$ (which tracks the expansion of expression, constructor and module identifiers, as $\uDelta$ and $\uGamma$ did in $\miniVersePat$) and peTSM context, $\uPsi$.
-Next, we expand $\uepsilon$ according to straightforward unexpanded peTSM expression expansion rules.
-The resulting TSM expression, $\epsilon$, must be defined at a type (i.e.
+The first two premises simply deconstruct the (unified) unexpanded context $\uOmega$ (which tracks the expansion of expression, constructor and module identifiers, as $\uDelta$ and $\uGamma$ did in $\miniVersePat$) and peTLM context, $\uPsi$.
+Next, we expand $\uepsilon$ according to straightforward unexpanded peTLM expression expansion rules.
+The resulting TLM expression, $\epsilon$, must be defined at a type (i.e.
 no quantification must remain.)
 
-The fourth premise performs \emph{peTSM expression normalization}.
+The fourth premise performs \emph{peTLM expression normalization}.
 Normalization, $\tsmexpEvalsExp{\Omega_\text{app}}{\Psi}{\epsilon}{\epsilon_\text{normal}}$, is defined in terms of a simple structural dynamics with two stepping rules:
@@
 \end{mathpar}
-% The peTSM expression normal forms are defined as follows:
+% The peTLM expression normal forms are defined as follows:
 % \begin{equation*}\tag{\ref{rule:tsmexpNormalExp-defref}}
@@
 % \end{equation*}
-Normalization eliminates parameters introduced in higher-order abbreviations, leaving only those parameter applications specified by the original TSM definition.
+Normalization eliminates parameters introduced in higher-order abbreviations, leaving only those parameter applications specified by the original TLM definition.
 Normal forms and progress and preservation theorems are established in the supplement.
 
-The third row of premises looks up the applied TSM's definition by first invoking a simple metafunction to extract the name, then looking up this name within the peTSM definition context, $\Psi$.
+The third row of premises looks up the applied TLM's definition by first invoking a simple metafunction to extract the name, then looking up this name within the peTLM definition context, $\Psi$.
 % \begin{align}
@@
 \end{array}\]
-There must be one binder in $\pce$ for each TSM parameter specified by $\tsmdefof{\epsilon_\text{normal}}$.
+There must be one binder in $\pce$ for each TLM parameter specified by $\tsmdefof{\epsilon_\text{normal}}$.
 (VerseML can insert these binders automatically as a convenience, but we consider only the underlying mechanism in this core calculus.)
@@
 \end{mathpar}
-This judgement can be pronounced ``when applying peTSM $\epsilon$, $\pce$ has deparameterization $\ce$ leaving $\rho$ with parameter substitution $\omega$''. 
+This judgement can be pronounced ``when applying peTLM $\epsilon$, $\pce$ has deparameterization $\ce$ leaving $\rho$ with parameter substitution $\omega$''. 
 Notice based on the second rule that every module binding in $\pce$ must pair with a corresponding module parameter application.
@@
 \ref{sec:s-metatheory}, modified to account for the presence of modules, kinds and parameterization.
-The following theorem establishes the abstract reasoning principles available when applying a parametric expression TSM.
+The following theorem establishes the abstract reasoning principles available when applying a parametric expression TLM.
 \vspace{-3px}
-\begin{theorem}[peTSM Reasoning Principles]
+\begin{theorem}[peTLM Reasoning Principles]
 If $\expandsP{\uOmega}{\uPsi}{\uPhi}{\utsmap{\uepsilon}{b}}{e}{\tau}$ then:
@@
   \item $
-    \summaryOf{\ce} = \sseq{\acesplicedk{m_i}{n_i}}{\nkind} \cup \sseq{\acesplicedc{m'_i}{n'_i}{\cekappa'_i}}{\ncon} \cup 
+    \segof{\ce} = \sseq{\acesplicedk{m_i}{n_i}}{\nkind} \cup \sseq{\acesplicedc{m'_i}{n'_i}{\cekappa'_i}}{\ncon} \cup 
                \sseq{\acesplicede{m''_i}{n''_i}{\ctau_i}}{\nexp}
@@
 \end{theorem}
-We leave out the details of parametric pattern TSMs (ppTSMs), but they should again follow analagously.
+We leave out the details of parametric pattern TLMs (ppTLMs), but they should again follow analagously.
 
@@
 
-Thus far, we have assumed that the parse functions in TSM definitions are closed expanded expressions.
-This assumption simplifies matters formally, but in practice it is quite difficult, because this means that TSM definitions cannot themselves access any libraries nor use TSMs internally.
+Thus far, we have assumed that the parse functions in TLM definitions are closed expanded expressions.
+This assumption simplifies matters formally, but in practice it is quite difficult, because this means that TLM definitions cannot themselves access any libraries nor use TLMs internally.
 To address this problem, we need to introduce the concept of a \emph{static environment}.
@@
 Figure \ref{fig:static-module-example} shows an example of a module, \li{ParserCombos}, that defines a number of parser combinators following \citet{Hutton1992d}.
-The \li{static} qualifier indicates that this module is bound for use only within similarly qualified values, including in particular the parse functions of the subsequent TSM definitions.
+The \li{static} qualifier indicates that this module is bound for use only within similarly qualified values, including in particular the parse functions of the subsequent TLM definitions.
 \begin{figure}[h]
@@
 to ensure deterministic builds.
-It might also be helpful to restrict mutable state shared between TSMs to prevent undesirable TSM application order dependencies.
+It might also be helpful to restrict mutable state shared between TLMs to prevent undesirable TLM application order dependencies.
 On the other hand, these features, if used for the purposes of caching in a disciplined manner, might speed up typed expansion.
@@
 
-\subsection{Applying TSMs Within TSM Definitions}\label{sec:tsms-for-tsms}
-TSMs and TSM abbreviations can also be qualified with the \li{static} keyword, which marks them for use within subsequent static expressions and patterns.
-Let us consider some examples of relevance to TSM providers.
+\subsection{Applying TLMs Within TLM Definitions}\label{sec:tsms-for-tsms}
+TLMs and TLM abbreviations can also be qualified with the \li{static} keyword, which marks them for use within subsequent static expressions and patterns.
+Let us consider some examples of relevance to TLM providers.
 
 \subsubsection{Quasiquotation}
-TSMs must construct values of type \li{proto_expr} or \li{proto_pat}.
+TLMs must construct values of type \li{proto_expr} or \li{proto_pat}.
 Constructing values of these types explicitly can have high syntactic cost.
-To decrease this cost, we can define TSMs that provide support for \emph{quasiquotation syntax} similar to that built in to languages like Lisp \cite{Bawd99a} and Scala \cite{shabalin2013quasiquotes}.
-The following TSM defines quasi-quotation for encodings of proto-expressions:
+To decrease this cost, we can define TLMs that provide support for \emph{quasiquotation syntax} similar to that built in to languages like Lisp \cite{Bawd99a} and Scala \cite{shabalin2013quasiquotes}.
+The following TLM defines quasi-quotation for encodings of proto-expressions:
 \begin{lstlisting}[numbers=none]
@@
 Rather than constructing a grammar (equipped with semantic actions) using the associated operations (whose specifications are elided in \li{PARSEGEN}), we wish to use a syntax for context-free grammars that follows standard conventions.
-We can do so by defining a static parametric TSM \li{$grammar}:
+We can do so by defining a static parametric TLM \li{$grammar}:
 \begin{lstlisting}[numbers=none]
@@
 The rules for binding forms annotated with $\staticphase$ are based on the corresponding $\standardphase$ phase rules, differing only in that 1) they operate on $\Sigma$ and 2) evaluation occurs immediately.
-Finally, the forms for TSM definition are modified so that the parse function is now an unexpanded, rather than an expanded expression.
+Finally, the forms for TLM definition are modified so that the parse function is now an unexpanded, rather than an expanded expression.
 The substitution $\omega$ is applied to the parse function after it is expanded.
@@
 
-TSMs definitions can be exported by packages, but they cannot be exported from within modules because that would require that they also appear in signatures, and that, in turn, would complicate reasoning about signature equivalence, since TSM definitions contain arbitrary parse functions.
-It would also bring in confusion about whether the generated expansions can use knowledge about type identity available  within the module enclosing the TSM definition.
-That said, it should be possible to export TSM \emph{abbreviations} from modules, since they refer to TSM definitions only through symbolic names.
+TLMs definitions can be exported by packages, but they cannot be exported from within modules because that would require that they also appear in signatures, and that, in turn, would complicate reasoning about signature equivalence, since TLM definitions contain arbitrary parse functions.
+It would also bring in confusion about whether the generated expansions can use knowledge about type identity available  within the module enclosing the TLM definition.
+That said, it should be possible to export TLM \emph{abbreviations} from modules, since they refer to TLM definitions only through symbolic names.
 We have not yet formalized this intuition.
@@
 % Library 
-%   (* ... exported module, signature and TSM names ... *)
+%   (* ... exported module, signature and TLM names ... *)
 % is 
@@
 
-Most obviously, TSMs are explicitly applied.
-Different TSMs can therefore be defined at the same type without conflict.
-Moreover, TSMs are not associated directly with generative type definitions, like TSLs,  but rather can operate at any type.
-In a subsequent short paper, \citet{sac15} proposed explicit application of simple expression TSMs in a bidirectional typed setting \cite{Pierce:2000:LTI:345099.345100}, but did not provide any formal details.
+Most obviously, TLMs are explicitly applied.
+Different TLMs can therefore be defined at the same type without conflict.
+Moreover, TLMs are not associated directly with generative type definitions, like TSLs,  but rather can operate at any type.
+In a subsequent short paper, \citet{sac15} proposed explicit application of simple expression TLMs in a bidirectional typed setting \cite{Pierce:2000:LTI:345099.345100}, but did not provide any formal details.
 We do not assume a bidirectionally typed language.
 Context independence implies that ML-style type inference can be performed using only the spliced terms and their types (because the remainder of an expansion cannot mention the very variables whose types are being inferred, nor constrain them further beyond what the annotations require.) We leave additional inference-related topics, e.g.
-TSM parameter inference, as future work.
+TLM parameter inference, as future work.
 
@@
 Our work differs in several ways.
-First, we leave the context-free syntax fixed, so different TSMs cannot conflict.
+First, we leave the context-free syntax fixed, so different TLMs cannot conflict.
 Second, SoundExt does not enforce hygiene, i.e.
 expansions might depend on the context and intentionally induce capture.
-Relatedly, TSMs do not support type-dependent expansion, as in \cite{conf/popl/LorenzenE16}.
+Relatedly, TLMs do not support type-dependent expansion, as in \cite{conf/popl/LorenzenE16}.
 Another important distinction is that our system relies on (proto-)expansion validation, rather than verification.
@@
 %Another topic of interest has to do with intentional capture. 
-Haskell's \li{do}-notation for types with monadic structure cannot be expressed directly using TSMs because it introduces bindings into sub-terms.
+Haskell's \li{do}-notation for types with monadic structure cannot be expressed directly using TLMs because it introduces bindings into sub-terms.
 This might be an acceptable trade-off (indeed, that is our present position.) Alternatively, it might be worth exploring a design where spliced identifiers (but not arbitrary identifiers) can be designated in the splice summary as bound within spliced terms.
@@
 
-To conclude, TSMs give substantial syntactic control to library providers while leaving programmers with strong abstract reasoning principles.
-We believe TSMs therefore occupy a ``sweet spot'' in the  design space.
-%This paper serves to guide those who hope to incorporate TSMs into their future language designs.
+To conclude, TLMs give substantial syntactic control to library providers while leaving programmers with strong abstract reasoning principles.
+We believe TLMs therefore occupy a ``sweet spot'' in the  design space.
+%This paper serves to guide those who hope to incorporate TLMs into their future language designs.
 

### 2017-05-24 popl18/syntax-popl18.tex
updated abstract

@@
 \begin{abstract}
-Programmers, 
-like pen-and-paper mathematicians, often want to define new ``syntactic sugar'' to  lower the syntactic cost of common idioms. 
-Unfortunately, the most direct mechanisms of syntactic control available to programmers, e.g.
-\li{camlp4} \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, do not support modular reasoning about syntactic determinism -- separately defined syntax extensions can conflict syntactically with one another.
-Moreover, clients cannot reason abstractly, i.e.
-given only the program text, about types and binding.
-Instead, they must reason transparently, i.e.
+%Many languages define convenient list literal forms.
+This paper questions this common practice: why should an otherwise ordinary data structure be privileged in the definition of a general-purpose programming language?
+When designing the surface syntax of a programming language, it is common practice to include literal forms for a select few standard data structures, e.g.
+lists.
+This practice neglects an unbounded assortment of other data structures for which literal forms might also be useful (consider, for example, sets, maps, matrices, regular expressions, HTML trees, URLs, SQL queries and chemical structures.) A more parsimonious approach would be to allow library providers to define new literal forms in a decentralized manner, but the mechanisms of  available to programmers today, e.g.
+\li{camlp4} \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, are, to put it pithily, \emph{unreasonable}.
+In particular,  they do not support modular reasoning about syntactic determinism, so separately defined literal forms can and do conflict syntactically with one another.
+Moreover, clients cannot reason abstractly about types and binding when they encounter an unfamiliar literal form (like they can when they encounter an unfamiliar function being applied.) Instead, they must reason transparently, i.e.
 about the underlying expansion.
-To address these problems, this paper  formally introduces \emph{simple} and \emph{parametric typed literal macros} (TLMs), which give library providers programmatic control over both the parsing and expansion of expressions and patterns of \emph{generalized literal form} \cite{TSLs} at a specified type or parameterized family of types.
-Expansion is {strictly hygienic}, meaning that 1) expansions can refer to external bindings only via spliced terms or supplied module parameters; and 2)  expansions do not reveal internal bindings to spliced terms or to the remainder of the program.
+This increases cognitive cost, and therefore defeats much of the purpose of syntactic sugar.
+%This makes it difficult to program ``in the large'', and these mechanisms have not been widely adopted.% Hygienic, typed term-rewriting systems, e.g.
+the Scala macro system \cite{ScalaMacros2013}, are only somewhat more reasonable, and do not offer direct syntactic control.
+
+This paper formally details a more reasonable alternative: \emph{typed literal macros} (TLMs).
+TLMs give library providers full programmatic control over both the parsing and expansion of expressions and patterns of \emph{generalized literal form} \cite{TSLs} at a specified type or parameterized family of types.
+Expansion is {strictly hygienic}, meaning 1) that expansions cannot make any assumptions about the application site context; and 2) that expansions do not reveal internal bindings to spliced terms or to the remainder of the program.
 % Partial parameter application lowers the syntactic cost of this strict style. 
-TLMs offer substantially more syntactic control than existing hygienic term-rewriting macro systems while ensuring that clients are able to reason modularly about syntactic determinism and abstractly about the type and binding structure of the program.
+This design ensures that clients are able to reason modularly about syntactic determinism and abstractly about the type and binding structure of the program.
 The system needs only to convey to clients (e.g.
-via secondary notation) the \emph{segmentation} of each literal, which gives the locations of the spliced terms.
+via secondary notation) the \emph{segmentation} of each literal, which gives the locations of the spliced segments.
 %The system can determine this {segmentation} automatically and convey it to the client via syntax highlighting, or by presenting the client with a context-free grammar (without revealing the semantic actions associated with each production.)
 % This mechanism captures many common syntactic idioms while avoiding the problem of syntactic conflicts by construction and supplying clients with clear abstract reasoning principles. 
-We specify TLMs and establish these abstract reasoning principles formally using an ML-style calculus of expressions, patterns and modules.
-
-% \emph{Typed syntax macros (TLMs)}, proposed in a recent short paper by Omar et al.
+We establish these abstract reasoning principles formally with a calculus of expressions, patterns and ML-style modules.
+This is the first detailed type-theoretic account of a typed, hygienic term-rewriting macro system for an ML-like language.
+% This calculus is the first such formalization of a typed macro system.
+
+% \emph{Typed syntax matcros (TLMs)}, proposed in a recent short paper by Omar et al.
 \cite{sac15}, give library providers programmatic control over the parsing and expansion of only terms of {(generalized) literal form}.
@@
 \label{sec:intro}
-When defining the textual syntax of a programming language, it is common practice to include ``syntactic sugar'' specific to certain standard library constructs.   
+When defining the surface syntax of a programming language, it is common practice to include a few ``literal forms'' specific to certain standard data structures.   
 For example, Standard ML and other functional languages include sugar for constructing and pattern matching on lists, as defined in the standard library \cite{mthm97-for-dart}.
+For example, programmers can write \li{[x, y, z]} rather than \li{Cons(x, Cons(y, Cons(z, Nil)))}.
 

### 2017-05-28 popl18/syntax-popl18.tex
tweaks to intro

@@
 This paper questions this common practice: why should an otherwise ordinary data structure be privileged in the definition of a general-purpose programming language?
-When designing the surface syntax of a programming language, it is common practice to include literal forms for a select few standard data structures, e.g.
+When designing the surface syntax of a programming language, it is common practice to build in literal forms for a select few standard data structures, e.g.
 lists.
-This practice neglects an unbounded assortment of other data structures for which literal forms might also be useful (consider, for example, sets, maps, matrices, regular expressions, HTML trees, URLs, SQL queries and chemical structures.) A more parsimonious approach would be to allow library providers to define new literal forms in a decentralized manner, but the mechanisms of  available to programmers today, e.g.
-\li{camlp4} \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, are, to put it pithily, \emph{unreasonable}.
+The problem is that this practice neglects an unbounded assortment of other data structures for which literal forms might also be useful (consider, for example, sets, maps, matrices, regular expressions, HTML trees, URLs, SQL queries and chemical structures.) A less \emph{ad hoc} design would be one that allows library providers to define new literal forms in a decentralized manner.
+However, the mechanisms available to programmers today, e.g.
+\li{camlp4} \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, are \emph{unreasonable}.
 In particular, they do not support modular reasoning about syntactic determinism, so separately defined literal forms can and do conflict syntactically with one another.
@@
 about the underlying expansion.
-This increases cognitive cost, and therefore defeats much of the purpose of syntactic sugar.
 %This makes it difficult to program ``in the large'', and these mechanisms have not been widely adopted.% Hygienic, typed term-rewriting systems, e.g.
@@
 We establish these abstract reasoning principles formally with a calculus of expressions, patterns and ML-style modules.
-This is the first detailed type-theoretic account of a typed, hygienic term-rewriting macro system for an ML-like language.
+This is the first detailed type-theoretic account of a typed, hygienic syntactic macro system for an ML-like language.
 % This calculus is the first such formalization of a typed macro system.
@@
 % Put succinctly, we design a programming language in the ML tradition with a \emph{reasonably} programmable syntax.
+\vspace{-10px}
 \end{abstract}
@@
 \label{sec:intro}
-When defining the surface syntax of a programming language, it is common practice to include a few ``literal forms'' specific to certain standard data structures.   
-For example, Standard ML and other functional languages include sugar for constructing and pattern matching on lists, as defined in the standard library \cite{mthm97-for-dart}.
-For example, programmers can write \li{[x, y, z]} rather than \li{Cons(x, Cons(y, Cons(z, Nil)))}.
-
-Third-party library providers are, quite justifiably, envious of the privileged status of these otherwise ordinary standard library definitions.
-After all, it is not difficult to find other examples of situationally useful library-specific syntactic sugar \cite{TSLs}.
-For example, 1) clients of a ``collections'' library might like syntactic sugar for finite sets and dictionaries; 2) clients of a ``web programming'' library might like syntactic sugar for HTML and JSON values; 3) a compiler writer might like syntactic sugar for the terms of the object language or various intermediate languages of interest; and 4) clients of a ``chemistry'' library might like syntactic sugar for chemical structures based on the SMILES standard \cite{anderson1987smiles}.
-This has motivated a long line of research into approaches that delegate more syntactic control to library providers.
+
+The surface syntax of a programming language serves as the user interface between human programmers and the language's semantics.
+As such, language designers often include various \emph{literal forms} in the surface syntax that are designed to decrease the cognitive cost of constructing and pattern matching over values of selected  types.
+
+For example, in Standard ML (SML), the list expression literal \li{[x, y, z]} desugars to a term  semantically equivalent to \li{Cons(x, Cons(y, Cons(z, Nil)))} where \li{Nil} and \li{Cons} stand for the constructors of the standard \li{list} datatype \cite{mthm97-for-dart,harper1997programming}.\footnote{In SML, list literals work even in contexts where the list expression constructors have been shadowed by other bindings, i.e.
+the expansion of literal forms is \emph{context independent}.
+We will return to the concept of context independence throughout this work.} List pattern literals are analagous.
+Another example is Ur/Web, which defines a dialect of Ur's textual syntax with expression and pattern literal forms for XHTML elements \cite{conf/popl/Chlipala15,conf/pldi/Chlipala10}.
+Figure \ref{fig:urweb} gives examles of XHTML literals containing ``spliced'' string expressions (Line 1) and XHTML expressions (Line 2).
+The desugarings of these literals, not shown, are substantially more verbose and, for programmers familiar with the standardized syntax for XHTML \cite{xhtml}, substantially more obscure.
+
+Third-party library providers are quite justifiably envious of the privileged status of semantically ordinary types like those just discussed that have been equipped with convenient literal forms by a language designer.
+After all, it is not difficult to find other examples of types for which programmers have invented specialized syntax.
+For example, 1) clients of a ``collections'' library might like vector, matrix, finite set and finite map literals following standard mathematical conventions; 2) clients of a ``web programming'' library in SML might like Ur/Web style XHTML literals; 3) Ur/Web programmers might further like CSS literals, which Ur/Web does not provide; 4) a compiler writer might like ``quotation'' literals for the terms of the object language and various intermediate languages of interest; and 5) clients of a ``chemistry'' library might like chemical structure literals based on the SMILES standard \cite{anderson1987smiles}.
+Research by \citet{Bravenboer:2007:PIA:1289971.1289975} suggests that literal forms for structured encodings of SQL queries and other query types can reduce the temptation to use unstructured string encodings (a.k.a.
+``stringly-typed programming'') and therefore reduce the prevalence of string injection attacks, which are amongst the most common and serious security vulnerabilities on the web today \cite{owasp2017}.
+\citet{TSLs} sampled string literals from open source projects and found that at least 15\% of them were parseable according to some readily apparent type-specific grammar (e.g.
+for SQL queries, regexes, file system paths, URLs, IP addresses and others.) Taken together, there appears to be a clear need for mechanisms that allow library providers to introduce new type-specific literal forms into a general-purpose language in a decentralized manner.
+
+Unfortunately, the mechanisms available today are, to put it pithily, \emph{unreasonable}.
+The next two subsections flesh out this claim and thereby motivate the specific contribution of this paper, introduced in Sec.
+\ref{sec:contributions}: a decidedly more reasonable mechanism for defining new literal forms within an ML- or Scala-like general-purpose language (i.e.
+a language that has a rich static type system, support for pattern matching and an ML-style module system or Scala-style object system with support for type members.)
 
 \subsection{Syntax Dialects are Unreasonable}\label{sec:syntax-dialects-intro}
-One well-worn approach is for library providers to use a syntax definition system to construct a library-specific \emph{syntax dialect}, i.e.
-a new syntax definition that extends the syntax definition given in the language definition with new derived forms (i.e.
-forms given meaning by expansion, rather than directly.) For example, Ur/Web defines a dialect of Ur's textual syntax with derived forms for SQL queries, XHTML elements and some other constructs defined in a  web programming library \cite{conf/popl/Chlipala15,conf/pldi/Chlipala10}.
-Figure \ref{fig:urweb} demonstrates how an XHTML expression containing ``spliced'' string expressions (Line 1) and XHTML expressions (Line 2) can be expressed in Ur/Web.
-The desugarings of these derived forms, not shown, are substantially more verbose and, for programmers familiar with the standardized syntax for XHTML \cite{xhtml}, substantially more obscure.
+One well-worn approach is for library providers to use a syntax definition system to construct a library-specific \emph{syntax dialect}: a new syntax definition that extends the syntax definition given in the language definition with new derived forms (i.e.
+forms given meaning by expansion, rather than directly.)
 
@@
 \vspace{-6px}
-\caption{Derived XHTML forms with support for splicing at two different types are built in to Ur/Web.}
+\caption{XHTML literals with support for splicing at two different types are built in to Ur/Web.}
 \vspace{-6px}
@@
 
-\subsection{Contributions}
+\subsection{Contributions}\label{sec:contributions}
 This work introduces a system of \textbf{typed literal macros (TLMs)} that gives library providers substantially more syntactic control than existing typed term-rewriting macro systems while maintaining the ability to reason abstractly about types, binding and segmentation.% abstract reasoning principles.

### 2017-05-30 popl18/syntax-popl18.tex
tweaks to sec 1.0 and 1.1

@@
 about the underlying expansion.
+This increases cognitive cost, defeating much of the purpose of convenient literal forms.
 %This makes it difficult to program ``in the large'', and these mechanisms have not been widely adopted.% Hygienic, typed term-rewriting systems, e.g.
@@
 TLMs give library providers full programmatic control over both the parsing and expansion of expressions and patterns of \emph{generalized literal form} \cite{TSLs} at a specified type or parameterized family of types.
-Expansion is {strictly hygienic}, meaning 1) that expansions cannot make any assumptions about the application site context; and 2) that expansions do not reveal internal bindings to spliced terms or to the remainder of the program.
+Expansion is {strictly hygienic}, meaning that 1) expansions cannot make any assumptions about the application site context; and 2)  expansions do not reveal internal bindings to spliced terms or to the remainder of the program.
 % Partial parameter application lowers the syntactic cost of this strict style. 
@@
 We will return to the concept of context independence throughout this work.} List pattern literals are analagous.
-Another example is Ur/Web, which defines a dialect of Ur's textual syntax with expression and pattern literal forms for XHTML elements \cite{conf/popl/Chlipala15,conf/pldi/Chlipala10}.
+Another example is Ur/Web, which extends Ur's textual syntax with expression and pattern literal forms for XHTML elements \cite{conf/popl/Chlipala15,conf/pldi/Chlipala10}.
 Figure \ref{fig:urweb} gives examles of XHTML literals containing ``spliced'' string expressions (Line 1) and XHTML expressions (Line 2).
-The desugarings of these literals, not shown, are substantially more verbose and, for programmers familiar with the standardized syntax for XHTML \cite{xhtml}, substantially more obscure.
-
-Third-party library providers are quite justifiably envious of the privileged status of semantically ordinary types like those just discussed that have been equipped with convenient literal forms by a language designer.
+The desugarings of these literals, not shown, are substantially more verbose and obscure.
+
+Third-party library providers are quite justifiably envious of the privileged status of semantically ordinary types that have been equipped with convenient literal forms by the language designer.
 After all, it is not difficult to find other examples of types for which programmers have invented specialized syntax.
-For example, 1) clients of a ``collections'' library might like vector, matrix, finite set and finite map literals following standard mathematical conventions; 2) clients of a ``web programming'' library in SML might like Ur/Web style XHTML literals; 3) Ur/Web programmers might further like CSS literals, which Ur/Web does not provide; 4) a compiler writer might like ``quotation'' literals for the terms of the object language and various intermediate languages of interest; and 5) clients of a ``chemistry'' library might like chemical structure literals based on the SMILES standard \cite{anderson1987smiles}.
-Research by \citet{Bravenboer:2007:PIA:1289971.1289975} suggests that literal forms for structured encodings of SQL queries and other query types can reduce the temptation to use unstructured string encodings (a.k.a.
-``stringly-typed programming'') and therefore reduce the prevalence of string injection attacks, which are amongst the most common and serious security vulnerabilities on the web today \cite{owasp2017}.
+For example, 1) clients of a ``collections'' library might like not just list literals, but also vector, matrix, finite set and finite map literals following standard mathematical conventions; 2) clients of a ``web programming'' library in SML might like Ur/Web style XHTML literals; 3) Ur/Web programmers might further like CSS literals, which Ur/Web does not provide; 4) a compiler writer might like ``quotation'' literals for the terms of the object language and various intermediate languages of interest; and 5) clients of a ``chemistry'' library might like chemical structure literals based on the SMILES standard \cite{anderson1987smiles}.
+Research by \citet{Bravenboer:2007:PIA:1289971.1289975} suggests that literal forms for structured encodings of SQL queries and other query types can reduce the temptation to use unstructured string encodings (``stringly-typed programming'') and therefore reduce the prevalence of string injection attacks, which are amongst the most common and serious security vulnerabilities on the web today \cite{owasp2017}.
 \citet{TSLs} sampled string literals from open source projects and found that at least 15\% of them were parseable according to some readily apparent type-specific grammar (e.g.
@@
 The next two subsections flesh out this claim and thereby motivate the specific contribution of this paper, introduced in Sec.
-\ref{sec:contributions}: a decidedly more reasonable mechanism for defining new literal forms within an ML- or Scala-like general-purpose language (i.e.
+\ref{sec:contributions}: a decidedly more reasonable mechanism for defining new literals within an ML- or Scala-like general-purpose language (i.e.
 a language that has a rich static type system, support for pattern matching and an ML-style module system or Scala-style object system with support for type members.)
@@
 \subsection{Syntax Dialects are Unreasonable}\label{sec:syntax-dialects-intro}
-One well-worn approach is for library providers to use a syntax definition system to construct a library-specific \emph{syntax dialect}: a new syntax definition that extends the syntax definition given in the language definition with new derived forms (i.e.
-forms given meaning by expansion, rather than directly.)
+One approach available to library providers is to use a syntax definition system to construct a library-specific \emph{syntax dialect}: a new syntax definition that extends the syntax definition given in the language definition with new forms, including literal forms.
+Ur/Web's syntax (Fig.
+\ref{fig:urweb}) is one example of a syntax dialect \cite{conf/popl/Chlipala15}.
 
@@
 \vspace{-6px}
-\caption{XHTML literals with support for splicing at two different types are built in to Ur/Web.}
+\caption{XHTML literals with support for splicing at two different types are built primitively into Ur/Web.}
 \vspace{-6px}
@@
 However, we argue that in fact
-There are hundreds of syntax definition systems of various design, e.g.
-grammar-based systems like Camlp4 \cite{ocaml-manual}, Copper \cite{conf/gpce/WykS07} or SugarJ/Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, parser combinator systems \cite{Hutton1992d} and somewhat less general systems built in to certain programming systems, like the system of fixity directives in Standard ML \cite{mthm97-for-dart} or more elaborate mixfix systems \cite{5134,DBLP:conf/gpce/TahaJ03} built in to proof assistants like Agda \cite{DBLP:conf/ifl/DanielssonN08} and Coq \cite{Coq:manual}.
+There are hundreds of syntax definition systems of various design, some integrated deeply into languages and build systems, others operating as standalone preprocessors.
+Examples include grammar-based systems like Camlp4 \cite{ocaml-manual}\todo{better citation?}, Copper \cite{conf/gpce/WykS07} or SugarJ/Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, parser combinator systems \cite{Hutton1992d} and somewhat less general systems built in to certain programming systems, like the system of fixity directives in Standard ML \cite{mthm97-for-dart}\todo{better citation?}~ or more elaborate mixfix systems \cite{5134,DBLP:conf/gpce/TahaJ03} built in to proof assistants like Agda \cite{DBLP:conf/ifl/DanielssonN08} and Coq \cite{Coq:manual}.
 %library-specific (a.k.a. ``domain-specific'') 
-Unfortunately, this dialect-oriented approach is difficult to reconcile with the best practices of ``programming in the large'', i.e.
-developing large programs ``consisting of many small programs (modules), possibly written by different people''   \cite{DeRemer76} whose interactions are mediated by a reasonable type and binding discipline.
-The problems that tend to arise are summarized below.%a more systematic treatment will follow in  Sec.
+Unfortunately, the dialect-oriented approach is difficult to reconcile with the best practices of ``programming in the large'', i.e.
+developing large programs ``consisting of many small programs (modules), possibly written by different people'' \cite{DeRemer76}.
+The reasoning problems that arise are summarized below.%a more systematic treatment will follow in  Sec.
 \ref{sec:syntax-dialects}.
 
-\paragraph{Problem 1: Incompatibility and Syntactic Conflict}
-The first problem with the dialect-oriented approach is that clients  cannot always combine different syntax dialects when they want to use derived forms that they define together.
-This is a significant problem because large programs use many separately defined libraries \cite{DBLP:conf/sac/LammelPS11} and often do not fall cleanly into a single preconceived ``problem domain''.
+\paragraph{Problem 1: Modular Reasoning About Syntactic Determinism}
+The first problem with the dialect-oriented approach is that clients cannot always deterministically combine different syntax dialects when they want to use the new forms that they define together.
+This is a significant problem because large programs use many separately developed libraries \cite{DBLP:conf/sac/LammelPS11} and often do not fall cleanly into a single preconceived ``problem domain''.
 
 In some cases, the dialects in question cannot be combined simply because they have been defined under incompatible syntax definition systems.
-In other cases, there is no dialect combination operation available that ensures that the combined dialect is an extension of both constituent dialects.
+In other cases, there is no dialect combination operation available.
 And in other cases, the problem is one of \emph{conflict}: two dialects might define the same form, but determine different desugarings.
-
-For example, consider two syntax dialects defined under a system like Camlp4: $\mathcal{D}_1$ defines derived forms for sets, and $\mathcal{D}_2$ defines derived forms for dictionaries, both delimited by \verb~{<~ and \verb~>}~.\footnote{In OCaml, simple curly braces are already reserved by the language for record types and values.} Each dialect is deterministic (i.e.
-there are no possible syntactic conflicts), $\mathrm{det}(\mathcal{D}_1)$ and $\mathrm{det}(\mathcal{D}_2)$, but when the grammars are combined by Camlp4, it is not the case that $\mathrm{det}(\mathcal{D}_1 \cup \mathcal{D}_2)$.
-In particular, \verb~{<>}~ can desugar to the empty set and the empty dictionary.%(Indeed, this ambiguity arose when Python added set literals to its grammar.) %A recent version of Python added derived forms for mutable sets.
+For example, consider two syntax dialects defined under a system like \li{camlp4}: $\mathcal{D}_1$ defines literal forms for sets, and $\mathcal{D}_2$ defines literal forms for finite sets, both delimited by \verb~{<~ and \verb~>}~.
+Each dialect is deterministic (i.e.
+there are no possible syntactic conflicts), denoted $\mathrm{det}(\mathcal{D}_1)$ and $\mathrm{det}(\mathcal{D}_2)$, but when the grammars are combined by Camlp4, it is not the case that $\mathrm{det}(\mathcal{D}_1 \cup \mathcal{D}_2)$.
+In particular, \verb~{<>}~ can desugar to the empty set and the empty dictionary.
+Reasoning about determinism is therefore not modular.
+%(Indeed, this ambiguity arose when Python added set literals to its grammar.) %A recent version of Python added derived forms for mutable sets.
 Due to a conflict with dictionary syntax, however, there is no derived form for the empty set.)
@@
 
-\citet{conf/pldi/SchwerdfegerW09} have developed a modular grammar analysis, implemented in Copper, that ``nearly'' guarantees that determinism is conserved when syntax dialects (of a certain restricted class) are combined, the caveat being that the constituent dialects must prefix all newly introduced forms with marking tokens drawn from disjoint sets.
+\citet{conf/pldi/SchwerdfegerW09} have developed a modular grammar analysis, implemented in Copper, that guarantees that determinism is conserved when syntax dialects of a certain restricted class are combined, the caveat being that the constituent dialects must prefix all newly introduced forms with marking tokens drawn from disjoint sets.
 To be confident that the marking tokens used are disjoint, providers must base them on the domain name system (or some other coordinating entity.) Because the mechanism operates at the level of the context-free grammar, it is difficult for the client to define scoped abbreviations for these verbose marking tokens (as one might do when, for example, deeply nested modules are used for namespacing in ML.)
 
-\paragraph{Problem 2: Abstract Reasoning About Derived Forms}
-
-Even putting aside the problem of syntactic conflict, there are questions about how \emph{reasonable}  sprinkling library-specific derived forms throughout a large software system might be. 
-For example, consider the perspective of a programmer attempting to comprehend (i.e.
-reason about) the program fragment in Figure \ref{fig:K-dialect}, which is written in a syntax dialect constructed by combining various library-specific dialects of Standard ML's syntax.
+\paragraph{Problem 2: Abstract Reasoning About Literal Forms}
 
@@
 
+Even putting aside the problem of syntactic conflict, there are questions about just how reasonable sprinkling many possibly unfamiliar library-specific literal forms  throughout a program may be.
+For example, consider the perspective of a programmer attempting to comprehend (i.e.
+reason about) the program fragment in Figure \ref{fig:K-dialect}, which is written in a syntax dialect constructed by combining various library-specific dialects of Standard ML's syntax.
 If the programmer happens to be familiar with the (intentionally terse) syntax of the stack-based database query processing language K \cite{Whitney:2001:LOR:376284.375783}, then Line 3 might pose few difficulties.
@@
 \begin{enumerate}
-\item \textbf{(Responsibility)} Which constituent dialect defined the derived form that appears on Line 3?
+\item \textbf{(Responsibility)} Which constituent dialect defined the literal form that appears on Line 3?
 \item \textbf{(Segmentation)} Are the characters \li{x} and \li{R} on Line 3 parsed as spliced expressions \li{x} and \li{R} (i.e.
-expressions of variable form), or parsed in some other way peculiar to this form?
+expressions of variable form), or parsed in some other way peculiar to this literal form?
 \item \textbf{(Capture)} If \li{x} is in fact a spliced expression, does it refer to the binding of \li{x} on Line 2? Or might it capture an unseen binding introduced in the desugaring of Line 3?

### 2017-06-03 popl18/syntax-popl18.tex
trying to reorganize intro to get to the point sooner and push related work to the end

@@
 \lstset{tabsize=2, 
-basicstyle=\ttfamily\fontsize{10pt}{1em}\selectfont, 
+basicstyle=\ttfamily\fontsize{9pt}{1em}\selectfont, 
 commentstyle=\itshape\ttfamily\color{gray}, 
@@
 However, the mechanisms available to programmers today, e.g.
-\li{camlp4} \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, are \emph{unreasonable}.
+Camlp4 \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, are \emph{unreasonable}.
 In particular, they do not support modular reasoning about syntactic determinism, so separately defined literal forms can and do conflict syntactically with one another.
@@
 We establish these abstract reasoning principles formally with a calculus of expressions, patterns and ML-style modules.
-This is the first detailed type-theoretic account of a typed, hygienic syntactic macro system for an ML-like language.
+This is the first detailed type-theoretic account of a typed, hygienic term-rewriting macro system for an ML-like\todo{or Scala-like?} language.
 % This calculus is the first such formalization of a typed macro system.
@@
 After all, it is not difficult to find other examples of types for which programmers have invented specialized syntax.
-For example, 1) clients of a ``collections'' library might like not just list literals, but also vector, matrix, finite set and finite map literals following standard mathematical conventions; 2) clients of a ``web programming'' library in SML might like Ur/Web style XHTML literals; 3) Ur/Web programmers might further like CSS literals, which Ur/Web does not provide; 4) a compiler writer might like ``quotation'' literals for the terms of the object language and various intermediate languages of interest; and 5) clients of a ``chemistry'' library might like chemical structure literals based on the SMILES standard \cite{anderson1987smiles}.
-Research by \citet{Bravenboer:2007:PIA:1289971.1289975} suggests that literal forms for structured encodings of SQL queries and other query types can reduce the temptation to use unstructured string encodings (``stringly-typed programming'') and therefore reduce the prevalence of string injection attacks, which are amongst the most common and serious security vulnerabilities on the web today \cite{owasp2017}.
-\citet{TSLs} sampled string literals from open source projects and found that at least 15\% of them were parseable according to some readily apparent type-specific grammar (e.g.
-for SQL queries, regexes, file system paths, URLs, IP addresses and others.) Taken together, there appears to be a clear need for mechanisms that allow library providers to introduce new type-specific literal forms into a general-purpose language in a decentralized manner.
-
-Unfortunately, the mechanisms available today are, to put it pithily, \emph{unreasonable}.
-The next two subsections flesh out this claim and thereby motivate the specific contribution of this paper, introduced in Sec.
-\ref{sec:contributions}: a decidedly more reasonable mechanism for defining new literals within an ML- or Scala-like general-purpose language (i.e.
-a language that has a rich static type system, support for pattern matching and an ML-style module system or Scala-style object system with support for type members.)
-
-\subsection{Syntax Dialects are Unreasonable}\label{sec:syntax-dialects-intro}
-One approach available to library providers is to use a syntax definition system to construct a library-specific \emph{syntax dialect}: a new syntax definition that extends the syntax definition given in the language definition with new forms, including literal forms.
-Ur/Web's syntax (Fig.
-\ref{fig:urweb}) is one example of a syntax dialect \cite{conf/popl/Chlipala15}.
+For example, 1) clients of a ``collections'' library might like not just list literals, but also vector, matrix, finite set and finite map literals following standard mathematical conventions; 2) clients of a ``web programming'' library in SML might like Ur/Web style XHTML literals; 3) Ur programmers might further like CSS literals, which Ur/Web does not provide; 4) a compiler writer might like ``quotation'' literals for the terms of the object language and various intermediate languages of interest; and 5) clients of a ``chemistry'' library might like chemical structure literals based on the SMILES standard \cite{anderson1987smiles}.
+Research by \citet{Bravenboer:2007:PIA:1289971.1289975} suggests that literal forms for structured encodings of SQL queries and other query types can reduce the temptation to use string encodings (``stringly-typed programming'') and therefore reduce the prevalence of string injection attacks, which are amongst the most common and serious security vulnerabilities on the web today \cite{owasp2017}.
+\citet{TSLs} sampled strings from open source projects and found that at least 15\% of them were parseable according to some readily apparent type-specific grammar (e.g.
+for SQL queries, regexes, file system paths, URLs and others.)
 
@@
 \vspace{-6px}
-\caption{XHTML literals with support for splicing at two different types are built primitively into Ur/Web.}
+\caption{XHTML literals with support for splicing at two different types are built primitively into Ur/Web \cite{conf/popl/Chlipala15}.}
 \vspace{-6px}
@@
 \end{figure}
- % Such dialects are sometimes qualitatively taxonomized as amongst the ``domain-specific language'' for this reason \cite{fowler2010domain}.
-%Syntactic cost is often assessed qualitatively \cite{green1996usability}, though quantitative metrics can be defined.
-% Syntax definition systems , which we will discuss in Sec.
-\ref{sec:syntax-dialects}, have simplified the task of defining ``library-specific'' (a.k.a.
-``domain-specific'') syntax dialects like Ur/Web, and have thereby contributed to their ongoing proliferation.
-
-% Many have argued that a proliferation of syntax dialects constructed using these tools is harmless or even desirable, because programmers can simply choose the right syntax dialect for each job at hand \cite{journals/stp/Ward94}.
-However, we argue that in fact
-There are hundreds of syntax definition systems of various design, some integrated deeply into languages and build systems, others operating as standalone preprocessors.
-Examples include grammar-based systems like Camlp4 \cite{ocaml-manual}\todo{better citation?}, Copper \cite{conf/gpce/WykS07} or SugarJ/Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, parser combinator systems \cite{Hutton1992d} and somewhat less general systems built in to certain programming systems, like the system of fixity directives in Standard ML \cite{mthm97-for-dart}\todo{better citation?}~ or more elaborate mixfix systems \cite{5134,DBLP:conf/gpce/TahaJ03} built in to proof assistants like Agda \cite{DBLP:conf/ifl/DanielssonN08} and Coq \cite{Coq:manual}.
-%library-specific (a.k.a. ``domain-specific'') 
-Unfortunately, the dialect-oriented approach is difficult to reconcile with the best practices of ``programming in the large'', i.e.
-developing large programs ``consisting of many small programs (modules), possibly written by different people'' \cite{DeRemer76}.
-The reasoning problems that arise are summarized below.%a more systematic treatment will follow in  Sec.
-\ref{sec:syntax-dialects}.
-
-\paragraph{Problem 1: Modular Reasoning About Syntactic Determinism}
-The first problem with the dialect-oriented approach is that clients cannot always deterministically combine different syntax dialects when they want to use the new forms that they define together.
-This is a significant problem because large programs use many separately developed libraries \cite{DBLP:conf/sac/LammelPS11} and often do not fall cleanly into a single preconceived ``problem domain''.
-
-In some cases, the dialects in question cannot be combined simply because they have been defined under incompatible syntax definition systems.
-In other cases, there is no dialect combination operation available.
-And in other cases, the problem is one of \emph{conflict}: two dialects might define the same form, but determine different desugarings.
-For example, consider two syntax dialects defined under a system like \li{camlp4}: $\mathcal{D}_1$ defines literal forms for sets, and $\mathcal{D}_2$ defines literal forms for finite sets, both delimited by \verb~{<~ and \verb~>}~.
-Each dialect is deterministic (i.e.
-there are no possible syntactic conflicts), denoted $\mathrm{det}(\mathcal{D}_1)$ and $\mathrm{det}(\mathcal{D}_2)$, but when the grammars are combined by Camlp4, it is not the case that $\mathrm{det}(\mathcal{D}_1 \cup \mathcal{D}_2)$.
-In particular, \verb~{<>}~ can desugar to the empty set and the empty dictionary.
-Reasoning about determinism is therefore not modular.
-%(Indeed, this ambiguity arose when Python added set literals to its grammar.) %A recent version of Python added derived forms for mutable sets.
-Due to a conflict with dictionary syntax, however, there is no derived form for the empty set.)
- %A third syntax dialect might come along that uses the same forms that $\mathcal{D}_2$ defines, but for ordered finite maps.
-
-\citet{conf/pldi/SchwerdfegerW09} have developed a modular grammar analysis, implemented in Copper, that guarantees that determinism is conserved when syntax dialects of a certain restricted class are combined, the caveat being that the constituent dialects must prefix all newly introduced forms with marking tokens drawn from disjoint sets.
-To be confident that the marking tokens used are disjoint, providers must base them on the domain name system (or some other coordinating entity.) Because the mechanism operates at the level of the context-free grammar, it is difficult for the client to define scoped abbreviations for these verbose marking tokens (as one might do when, for example, deeply nested modules are used for namespacing in ML.)
-
-\paragraph{Problem 2: Abstract Reasoning About Literal Forms}
-
-\begin{figure}
-\begin{lstlisting}
-  val w = compute_w ()
-  val x = compute_x ()
-  val y = {|(!R)@&{&/x!/:2_!x}'!R}|}
-\end{lstlisting}
-\vspace{-8px}
-\caption{An example of unreasonable program text.}
-\vspace{-8px}
-\label{fig:K-dialect}
-\end{figure}
-
-Even putting aside the problem of syntactic conflict, there are questions about just how reasonable sprinkling many possibly unfamiliar library-specific literal forms  throughout a program may be.
-For example, consider the perspective of a programmer attempting to comprehend (i.e.
-reason about) the program fragment in Figure \ref{fig:K-dialect}, which is written in a syntax dialect constructed by combining various library-specific dialects of Standard ML's syntax.
-If the programmer happens to be familiar with the (intentionally terse) syntax of the stack-based database query processing language K \cite{Whitney:2001:LOR:376284.375783}, then Line 3 might pose few difficulties.
-If the programmer does not recognize this syntax, however, there are no simple, definitive protocols for answering questions like:
-
-\begin{enumerate}
-\item \textbf{(Responsibility)} Which constituent dialect defined the literal form that appears on Line 3?
-\item \textbf{(Segmentation)} Are the characters \li{x} and \li{R} on Line 3 parsed as spliced expressions \li{x} and \li{R} (i.e.
-expressions of variable form), or parsed in some other way peculiar to this literal form?
-\item \textbf{(Capture)} If \li{x} is in fact a spliced expression, does it refer to the binding of \li{x} on Line 2? Or might it capture an unseen binding introduced in the desugaring of Line 3?
-\item \textbf{(Context Dependence)} If \li{w}, on Line 1, is renamed, could that possibly break the program, or change its meaning?
-In other words, might the desugaring of Line 3 assume that some variable identified as \li{w} is in scope (even though \li{w} is not mentioned in the text of Line 3)?
-\item \textbf{(Typing)} What type does \li{y} have?
-\end{enumerate}
-
-In short, syntax dialects do not come equipped with reasonable principles of \emph{syntactic abstraction}: if the desugaring of the program text is held abstract, programmers can no longer reason about types and binding (i.e.
-answer questions like those above) in the usual disciplined manner.
-This is burdensome at all scales, but particularly when programming in the large, where it is common to encounter unfamiliar libraries.
-Forcing the programmer to examine the desugaring of the program text in order to reason about types and binding defeats the ultimate purpose of syntactic sugar -- decreasing cognitive cost \cite{Green89}.
-
-%In other words, encountering an unfamiliar derived form has made it difficult for the programmer to maintain the usual \emph{type discipline} and \emph{binding discipline}.
-%Compelling the programmer to examine the desugaring directly defeat the purpose of defining the derived form -- decreasing cognitive cost.
-Indeed, it substantially increases cognitive cost.
-
-% In contrast, when a programmer encounters, for example, a function call like the call to \li{read_data} on Line 3, the analagous questions can be answered by following clear protocols that become ``cognitive reflexes'' after sufficient experience with the language, even if the programmer has no experience with the library defining \li{read_data}:
-% \begin{enumerate}
-% \item The language's syntax definition determines that \li{read_data(a)} is an expression of function application form.
-% \item Similarly, \li{read_data} and \li{a} are definitively expressions of variable form.
-% \item The variable \li{a} can only refer to the binding of \li{a} on Line 1.
-% \item The variable \li{w} can be renamed without knowing anything about the values that \li{read_data} and \li{a} stand for.
-% \item The type of \li{x} can be determined to be \li{B} by determining that the type of \li{read_data} is \li{A -> B} for some \li{A} and \li{B}, and checking that \li{a} has type \li{A}.
-Nothing else needs to be known about the values that \li{read_data} and \li{a} stand for.
-In Reynolds' words \cite{B304}:
-% \begin{quote}
-% \emph{Type structure is a syntactic discipline for enforcing levels of abstraction.}
-% \end{quote}
-% \end{enumerate}
-
-\subsection{Term-Rewriting Systems Offer Limited Syntactic Control}
-An alternative approach that a library provider might consider when seeking to control syntactic cost is to leave the context-free syntax of the language unmodified and instead contextually repurpose select existing syntactic forms using a \emph{term rewriting system}.
-
-Na\"ive term rewriting systems suffer from problems analagous to those that plague syntax definition systems.
-In particular, there can be conflicts because separately defined rewriting rules might attempt to rewrite the same term differently.
-It can also be difficult to determine which rewriting rule, if any, is responsible for any particular term.
-Certain conventions have emerged, e.g.
-OCaml's textual syntax now includes \emph{preprocessor extension (ppx) points} used to identify terms that some presumably corresponding term rewriting preprocessor must rewrite \cite{ocaml-manual}, but a correspondence is not enforced.
-In any case, it is difficult to reason abstractly about types and binding given a program subject to a large number of context-free rewriting rules.
-
-Modern \emph{term-rewriting macro systems}, however, have made some progress. In particular:
-\begin{enumerate}[nolistsep]
-\item Macro systems require that the client explicitly apply the intended rewriting (implemented by a macro) to the term that is to be rewritten, thereby addressing the problems of conflict and determining responsibility.
-However, it is often unclear whether a given macro  is repurposing the form of a given argument or sub-term thereof, as opposed to treating it parametrically by inserting it unmodified into the generated expansion.
-This is closely related to the problem of determining a {segmentation} discussed above.
-\item \emph{Hygienic} macro systems prevent variable {capture} and enforce {context-independence} \cite{Kohlbecker86a,DBLP:conf/popl/Adams15,DBLP:conf/popl/ClingerR91,DBLP:journals/lisp/DybvigHB92}.
-\item Although much of the research on macro systems has been for languages in the Lisp tradition \cite{mccarthy1978history}, some progress has also been made on reasoning about types with the design of \emph{typed macro systems} where annotations constrain the macro arguments and the generated expansions.
-Examples include Scala's macro system \cite{ScalaMacros2013}, formal calculi studied by \citet{DBLP:conf/esop/HermanW08} and \emph{staging macro systems} like MetaML \cite{Sheard:1999:UMS} and MacroML \cite{ganz2001macros} (which, it is worth noting, do not give the macro access to the syntax tree of arguments.)
-\end{enumerate}
-
-Perhaps the biggest problem, then, is that term-rewriting systems offer library providers only limited syntactic control.
-In other words many syntactic forms of interest will not particularly resemble those available in the host language.
-For example, consider the  XHTML and K examples above.
-In both cases, the syntactic conventions are quite distinct from those of ML-like languages (and, for that matter, languages that use S-expression.)
-
-It is tempting in these situations to consider repurposing string literal forms.
-For example, we might wish to apply a macro \li{html!} (following Rust's convention of using a post-fix \li{!} to distinguish macro names from variables) to rewrite string literals containing Ur/Web-style XHTML syntax as follows:
-\begin{lstlisting}[numbers=none]
-  fun heading first_name = html! 'SSTR<h1>Hello, {[first_name]}!</p>ESTR'
-  val body = html! 'SSTR<body>{heading "World"} ...</body>ESTR'
-\end{lstlisting}
-
-The problem here is that there is no way to extract the spliced expressions from the supplied string literal forms while satisfying the hygiene conditions, because variables that come from these spliced terms (here, \li{first_name} and \li{heading}) are indistinguishable from variables that inappropriately appear free relative to the expansion.
-In addition, the problem of segmentation becomes even more pernicious: to a human or tool unaware of Ur/Web's syntax, it is not immediately apparent which particular subsequences of the string literals supplied to \li{html!} are segmented out as spliced expressions.
-Reader macros have essentially the same problem  \cite{DBLP:journals/jfp/FlattCDF12}.
-
-  Some languages build in \emph{string splicing} (a.k.a.
-\emph{string interpolation}) forms, e.g.
-Scala \cite{odersky2008programming}, or similar but more general \emph{fragmentary quotation forms} \cite{conf/icfp/Slind91}, e.g.
-the SML/NJ dialect of ML.
-These designate particular delimiters to allow escape out into the expression language.
-The problem with using these (dynamically, or in conjunction with term rewriting macros) as vehicles to introduce syntactic sugar at various types is 1) there is no ``one-size-fits-all'' escape delimiter, and 2) reasoning about typing is problematic because every escaped term is checked against the same type.
-In the example above, we spliced in two different types of expressions using two different delimiters.
-In general, e.g.
-when defining syntax for a programming language with many sorts of terms, the most appropriate choice of delimiters might depend on where each spliced term appears.
+
+Given this profusion of possible literal forms, it is not sensible to ask the designers of general-purpose languages to keep up.
+Instead, there has been considerable interest in mechanisms that allow programmers to introduce new type-specific literal forms in a decentralized manner.
+Unfortunately, existing mechanisms ...
+
+{This paper addresses this established need by introducing a system of \emph{typed literal macros (TLMs)} that allows programmers to define new literal syntax of nearly arbitrary form within an ML- or Scala-like  language}, i.e.
+a language with a rich static type system (Sec.
+2), support for pattern matching (Sec.
+3) and an ML-style module system or Scala-style object system with support for type members and type abstraction (Sec.
+5).
+The examples in this paper are written in VerseML, a new dialect of ML that features TLMs.
+
+Compared to existing approaches, TLMs are more \emph{reasonable}.
+We justify this claim by (1) developing a collection of relevant reasoning principles based on the reasoning principles already available to client programmers when using an ML-like or Scala-like language without such a mechanism; (2) formally operationalizing these reasoning principles as metatheorems about a series of type-theoretic \emph{TLM calculi} of increasing sophistication (Sec.
+4-5); and (3) evaluating existing approaches -- including syntax definition systems like Camlp4 \cite{ocaml-manual}\todo{better citation}, SugarJ/Sugar* \cite{erdweg2011sugarj,erdweg2013framework} and Copper \cite{conf/pldi/SchwerdfegerW09,schwerdfeger2010context}, term rewriting macro systems as found in a variety of languages, and the system of type-specific languages (TSLs) and typed syntax macros (TSMs) described by \citet{TSLs,sac15} -- relative to these reasoning principles (Sec.
+6).
+We conclude by discussing limitations and future research directions in Sec.
+7\todo{sec refs}.
+A complete set of definitions and associated proofs are available in the  supplemental material.
+This is the first detailed type-theoretic accout of a typed, hygienic term-rewriting macro system for an ML- or Scala-like language.\todo{refine and justify this claim}
 
@@
 
+\section{Reasoning Criteria and Existing Approaches}
+\subsection{Library-Specific Syntax Dialects}\label{sec:syntax-dialects-intro}
+One approach available to library providers is to use a syntax definition system to construct a library-specific \emph{syntax dialect}: a new syntax definition that extends the syntax definition given in the language definition with new forms, including literal forms.
+Ur/Web's syntax (Fig.
+1) is a library-specific dialect of Ur's syntax.
+
+ % Such dialects are sometimes qualitatively taxonomized as amongst the ``domain-specific language'' for this reason \cite{fowler2010domain}.
+%Syntactic cost is often assessed qualitatively \cite{green1996usability}, though quantitative metrics can be defined.
+% Syntax definition systems , which we will discuss in Sec.
+\ref{sec:syntax-dialects}, have simplified the task of defining ``library-specific'' (a.k.a.
+``domain-specific'') syntax dialects like Ur/Web, and have thereby contributed to their ongoing proliferation.
+
+% Many have argued that a proliferation of syntax dialects constructed using these tools is harmless or even desirable, because programmers can simply choose the right syntax dialect for each job at hand \cite{journals/stp/Ward94}.
+However, we argue that in fact
+There are hundreds of syntax definition systems of various design, some integrated deeply into language front-ends and build systems, others operating as standalone preprocessors.
+Examples include grammar-based systems like Camlp4 \cite{ocaml-manual}\todo{better citation?}, Copper \cite{conf/gpce/WykS07} or SugarJ/Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, parser combinator systems \cite{Hutton1992d} and less general operator-based systems, like the system of fixity directives in Standard ML \cite{mthm97-for-dart}\todo{better citation?}~ and the more elaborate mixfix systems \cite{5134,DBLP:conf/gpce/TahaJ03} built in to proof assistants like Agda \cite{DBLP:conf/ifl/DanielssonN08} and Coq \cite{Coq:manual}.
+%library-specific (a.k.a. ``domain-specific'')
+
+\paragraph{Criteria 1: Modular Reasoning About Syntactic Determinism}
+The first problem with this approach is that clients cannot always deterministically combine different library-specific syntax dialects when they want to use the new forms that they define together.
+This is a significant problem because large programs use many separately developed libraries \cite{DBLP:conf/sac/LammelPS11} and often do not fall cleanly into a single  ``problem domain''.
+
+In some cases, the dialects in question cannot be combined simply because they have been defined under incompatible syntax definition systems.
+In other cases, the problem is one of \emph{conflict}: two dialects might define the same form, but determine different desugarings.
+For example, consider two syntax dialects defined under a system like Camlp4: $\mathcal{D}_1$ defines literal forms for sets, and $\mathcal{D}_2$ defines literal forms for finite sets, both delimited by \verb~{<~ and \verb~>}~.
+Each dialect is deterministic (i.e.
+there are no possible syntactic ambiguities), denoted $\mathrm{det}(\mathcal{D}_1)$ and $\mathrm{det}(\mathcal{D}_2)$.
+However, when the grammars are combined by Camlp4, it is not the case that $\mathrm{det}(\mathcal{D}_1 \cup \mathcal{D}_2)$ because \verb~{<>}~ can desugar to the empty set or the empty dictionary.
+Reasoning about determinism is therefore not modular.
+%(Indeed, this ambiguity arose when Python added set literals to its grammar.) %A recent version of Python added derived forms for mutable sets.
+Due to a conflict with dictionary syntax, however, there is no derived form for the empty set.)
+ %A third syntax dialect might come along that uses the same forms that $\mathcal{D}_2$ defines, but for ordered finite maps.
+
+\citet{conf/pldi/SchwerdfegerW09} have developed a modular grammar analysis, implemented in Copper, that guarantees that determinism is conserved when syntax dialects of a certain restricted class are combined, the caveat being that the constituent dialects must prefix all newly introduced forms with marking tokens drawn from disjoint sets.
+To be confident that the marking tokens used are disjoint, providers must base them on the domain name system or some other coordinating entity.
+Because the mechanism operates at the level of the context-free grammar, it is difficult for the client to define scoped abbreviations for these verbose marking tokens (as one might do when, for example, deeply nested modules are used for namespacing in ML.)
+
+\paragraph{Criteria 2: Abstract Reasoning About Literal Forms}
+
+\begin{figure}
+\begin{lstlisting}
+  val w = compute_w ()
+  val x = compute_x ()
+  val y = {|(!R)@&{&/x!/:2_!x}'!R}|}
+\end{lstlisting}
+\vspace{-8px}
+\caption{An example of unreasonable program text.}
+\vspace{-8px}
+\label{fig:K-dialect}
+\end{figure}
+
+Even putting aside the problem of syntactic conflict, there are questions about just how reasonable sprinkling many possibly unfamiliar library-specific literal forms  throughout a program may be.
+For example, consider the perspective of a programmer attempting to comprehend (i.e.
+reason about) the program fragment in Figure \ref{fig:K-dialect}, which is written in a syntax dialect constructed by combining various library-specific dialects of ML's (or Scala's) syntax.
+If the programmer happens to be familiar with the (intentionally terse) syntax of the stack-based database query processing language K \cite{Whitney:2001:LOR:376284.375783}, then Line 3 might pose few difficulties.
+If the programmer does not recognize this syntax, however, there are no simple, definitive protocols for answering questions like:
+
+\begin{enumerate}
+\item \textbf{(Responsibility)} Which constituent dialect defined the literal form that appears on Line 3?
+\item \textbf{(Segmentation)} Are the characters \li{x} and \li{R} on Line 3 parsed as spliced expressions \li{x} and \li{R} (i.e.
+expressions of variable form), or parsed in some other way peculiar to this literal form?
+\item \textbf{(Capture)} If \li{x} is in fact a spliced expression, does it refer to the binding of \li{x} on Line 2? Or might it capture an unseen binding introduced in the desugaring of Line 3?
+\item \textbf{(Context Dependence)} If \li{w}, on Line 1, is renamed, could that possibly break Line 3, or change its meaning?
+In other words, might the desugaring of Line 3 assume that some variable identified as \li{w} is in scope (even though \li{w} is not mentioned in the text of Line 3)?
+\item \textbf{(Typing)} What type does \li{y} have?
+\end{enumerate}
+
+In short, syntax dialects do not come equipped with  principles of \emph{syntactic abstraction}: if the desugaring of the program text is held abstract, programmers can no longer reason about types, binding and segmentation (i.e.
+answer questions like those above) in the usual disciplined manner.
+This is burdensome at all scales, but particularly when programming in the large, where it is common to encounter unfamiliar library constructs.
+Forcing the programmer to examine the desugaring of the program text, or the logic that produced that desugaring, in order to answer these sorts of basic questions about segmentation, binding and typing can defeat the ultimate purpose of syntactic sugar -- decreasing cognitive cost \cite{Green89}.
+
+%In other words, encountering an unfamiliar derived form has made it difficult for the programmer to maintain the usual \emph{type discipline} and \emph{binding discipline}.
+%Compelling the programmer to examine the desugaring directly defeat the purpose of defining the derived form -- decreasing cognitive cost.
+Indeed, it substantially increases cognitive cost.
+
+% In contrast, when a programmer encounters, for example, a function call like the call to \li{read_data} on Line 3, the analagous questions can be answered by following clear protocols that become ``cognitive reflexes'' after sufficient experience with the language, even if the programmer has no experience with the library defining \li{read_data}:
+% \begin{enumerate}
+% \item The language's syntax definition determines that \li{read_data(a)} is an expression of function application form.
+% \item Similarly, \li{read_data} and \li{a} are definitively expressions of variable form.
+% \item The variable \li{a} can only refer to the binding of \li{a} on Line 1.
+% \item The variable \li{w} can be renamed without knowing anything about the values that \li{read_data} and \li{a} stand for.
+% \item The type of \li{x} can be determined to be \li{B} by determining that the type of \li{read_data} is \li{A -> B} for some \li{A} and \li{B}, and checking that \li{a} has type \li{A}.
+Nothing else needs to be known about the values that \li{read_data} and \li{a} stand for.
+In Reynolds' words \cite{B304}:
+% \begin{quote}
+% \emph{Type structure is a syntactic discipline for enforcing levels of abstraction.}
+% \end{quote}
+% \end{enumerate}
+
+\subsection{Another Existing Approach: Term Rewriting}
+An alternative approach (and the approach that our mechanism is rooted in) is to leave the context-free syntax of the language unmodified and instead contextually repurpose existing forms using a \emph{term rewriting system}.
+
+For example, OCaml's textual syntax now includes \emph{preprocessor extension (ppx) points} used to identify terms that some term rewriting preprocessor must rewrite \cite{ocaml-manual}.
+We could attempt to mark a string literal containing Ur/Web-style HTML syntax with a ppx annotation, \li{xml}, as follows:
+\begin{lstlisting}[numbers=none]
+  [%xml "SSTR<h1>Hello, {[first_name]}!</h1>ESTR"]
+\end{lstlisting}
+Again, there are problems reasoning modularly and abstractly about such constructions.
+More than one applied preprocessor might recognize this annotation (there are, in practice, many XML/HTML libraries), so non-determinism can arise, and it is difficult to reason about \textbf{Responsibility}.
+It is also difficult to reason abstractly about \textbf{Segmentation}, \textbf{Capture}, \textbf{Context Dependence} and \textbf{Typing} because the code that the preprocessor generates is arbitrary.
+For these reasons, users of this mechanism warn that they should be used sparingly.\todo{citation}
+
+Term-rewriting macro systems address these problems -- the applied macro explicitly identifies the rewriting that wil
+
+Template Haskell / Haskell quasiquotations / early macro systems / reader macros
+
+hygienic macro systems -- problem!
+
+typed macro systems, e.g. Scala. mention MacroML stuff, Herman and Wand 2008/2010 work
+
+string splicing / interpolation
+
+TSLs / TSMs
+
+Non-local rewriting systems suffer from problems analagous to those that plague syntax definition systems.
+In particular, there can be conflicts because separately defined rewriting rules might attempt to rewrite the same term differently.
+It can also be difficult to determine which rewriting rule, if any, is responsible for any particular term.
+Certain conventions have emerged, e.g.
+OCaml's textual syntax now includes \emph{preprocessor extension (ppx) points} used to identify terms that some presumably corresponding term rewriting preprocessor must rewrite \cite{ocaml-manual}, but a correspondence is not enforced.
+In any case, it is also difficult to reason abstractly about types and binding given a program subject to a large number of rewriting rules.
+
+Modern \emph{term-rewriting macro systems}, however, have made some progress. In particular:
+\begin{enumerate}[nolistsep]
+\item Macro systems require that the client explicitly apply the intended rewriting (implemented by a macro) to the term that is to be rewritten, thereby addressing the problems of conflict and determining responsibility.
+However, it is often unclear whether a given macro  is repurposing the form of a given argument or sub-term thereof, as opposed to treating it parametrically by inserting it unmodified into the generated expansion.
+This is closely related to the problem of determining a {segmentation} discussed above.
+\item \emph{Hygienic} macro systems prevent variable {capture} and enforce {context-independence} \cite{Kohlbecker86a,DBLP:conf/popl/Adams15,DBLP:conf/popl/ClingerR91,DBLP:journals/lisp/DybvigHB92}.
+\item Although much of the research on macro systems has been for languages in the Lisp tradition \cite{mccarthy1978history}, some progress has also been made on reasoning about types with the design of \emph{typed macro systems} where annotations constrain the macro arguments and the generated expansions.
+Examples include Scala's macro system \cite{ScalaMacros2013}, formal calculi studied by \citet{DBLP:conf/esop/HermanW08} and \emph{staging macro systems} like MetaML \cite{Sheard:1999:UMS} and MacroML \cite{ganz2001macros} (which, it is worth noting, do not give the macro access to the syntax tree of arguments.)
+\end{enumerate}
+
+Perhaps the biggest problem, then, is that term-rewriting systems offer library providers only limited syntactic control.
+In other words many syntactic forms of interest will not particularly resemble those available in the host language.
+For example, consider the  XHTML and K examples above.
+In both cases, the syntactic conventions are quite distinct from those of ML-like languages (and, for that matter, languages that use S-expression.)
+
+It is tempting in these situations to consider repurposing string literal forms.
+For example, we might wish to apply a macro \li{html!} (following Rust's convention of using a post-fix \li{!} to distinguish macro names from variables) to rewrite string literals containing Ur/Web-style XHTML syntax as follows:
+\begin{lstlisting}[numbers=none]
+  fun heading first_name = html! 'SSTR<h1>Hello, {[first_name]}!</p>ESTR'
+  val body = html! 'SSTR<body>{heading "World"} ...</body>ESTR'
+\end{lstlisting}
+
+The problem here is that there is no way to extract the spliced expressions from the supplied string literal forms while satisfying the hygiene conditions, because variables that come from these spliced terms (here, \li{first_name} and \li{heading}) are indistinguishable from variables that inappropriately appear free relative to the expansion.
+In addition, the problem of segmentation becomes even more pernicious: to a human or tool unaware of Ur/Web's syntax, it is not immediately apparent which particular subsequences of the string literals supplied to \li{html!} are segmented out as spliced expressions.
+Reader macros have essentially the same problem  \cite{DBLP:journals/jfp/FlattCDF12}.
+
+  Some languages build in \emph{string splicing} (a.k.a.
+\emph{string interpolation}) forms, e.g.
+Scala \cite{odersky2008programming}, or similar but more general \emph{fragmentary quotation forms} \cite{conf/icfp/Slind91}, e.g.
+the SML/NJ dialect of ML.
+These designate particular delimiters to allow escape out into the expression language.
+The problem with using these (dynamically, or in conjunction with term rewriting macros) as vehicles to introduce syntactic sugar at various types is 1) there is no ``one-size-fits-all'' escape delimiter, and 2) reasoning about typing is problematic because every escaped term is checked against the same type.
+In the example above, we spliced in two different types of expressions using two different delimiters.
+In general, e.g.
+when defining syntax for a programming language with many sorts of terms, the most appropriate choice of delimiters might depend on where each spliced term appears.
+
 \newcommand{\discussionSec}{Additional Related Work \& Concluding Discussion}

### 2017-06-05 popl18/syntax-popl18.tex
progress on major reorg of intro and related work

@@
 
+\vspace{-4px}
 \newcommand{\introSec}{Introduction}
@@
 \label{sec:intro}
-
-The surface syntax of a programming language serves as the user interface between human programmers and the language's semantics.
-As such, language designers often include various \emph{literal forms} in the surface syntax that are designed to decrease the cognitive cost of constructing and pattern matching over values of selected  types.
-
-For example, in Standard ML (SML), the list expression literal \li{[x, y, z]} desugars to a term  semantically equivalent to \li{Cons(x, Cons(y, Cons(z, Nil)))} where \li{Nil} and \li{Cons} stand for the constructors of the standard \li{list} datatype \cite{mthm97-for-dart,harper1997programming}.\footnote{In SML, list literals work even in contexts where the list expression constructors have been shadowed by other bindings, i.e.
+The surface syntax of a programming language serves as the interface between human programmers and the language's semantics.
+As such, language designers often include various \emph{literal forms} in the surface syntax that are designed to decrease the cognitive cost of constructing and pattern matching over values of certain types defined in the standard library.
+For example, in Standard ML (SML), the list expression literal \li{[x, y, z]} desugars to a term  semantically equivalent to \li{Cons(x, Cons(y, Cons(z, Nil)))} where \li{Nil} and \li{Cons} identify the constructors of the standard \li{list} datatype \cite{mthm97-for-dart,harper1997programming}.\footnote{In SML, list literals work even in contexts where the list expression constructors have been shadowed by other bindings, i.e.
 the expansion of literal forms is \emph{context independent}.
 We will return to the concept of context independence throughout this work.} List pattern literals are analagous.
-Another example is Ur/Web, which extends Ur's textual syntax with expression and pattern literal forms for XHTML elements \cite{conf/popl/Chlipala15,conf/pldi/Chlipala10}.
-Figure \ref{fig:urweb} gives examles of XHTML literals containing ``spliced'' string expressions (Line 1) and XHTML expressions (Line 2).
-The desugarings of these literals, not shown, are substantially more verbose and obscure.
-
-Third-party library providers are quite justifiably envious of the privileged status of semantically ordinary types that have been equipped with convenient literal forms by the language designer.
-After all, it is not difficult to find other examples of types for which programmers have invented specialized syntax.
-For example, 1) clients of a ``collections'' library might like not just list literals, but also vector, matrix, finite set and finite map literals following standard mathematical conventions; 2) clients of a ``web programming'' library in SML might like Ur/Web style XHTML literals; 3) Ur programmers might further like CSS literals, which Ur/Web does not provide; 4) a compiler writer might like ``quotation'' literals for the terms of the object language and various intermediate languages of interest; and 5) clients of a ``chemistry'' library might like chemical structure literals based on the SMILES standard \cite{anderson1987smiles}.
-Research by \citet{Bravenboer:2007:PIA:1289971.1289975} suggests that literal forms for structured encodings of SQL queries and other query types can reduce the temptation to use string encodings (``stringly-typed programming'') and therefore reduce the prevalence of string injection attacks, which are amongst the most common and serious security vulnerabilities on the web today \cite{owasp2017}.
-\citet{TSLs} sampled strings from open source projects and found that at least 15\% of them were parseable according to some readily apparent type-specific grammar (e.g.
-for SQL queries, regexes, file system paths, URLs and others.)
+Another example is Ur/Web, which extends Ur's textual syntax with expression and pattern literals for XHTML trees, as defined in Ur/Web's standard library \cite{conf/popl/Chlipala15,conf/pldi/Chlipala10}.
+Figure \ref{fig:urweb} shows examples of XHTML literals containing ``spliced'' string expressions (Line 1) and XHTML expressions (Line 2).
+%The desugarings of these literal expressions, not shown, are substantially more verbose and less readable to programmers familiar with XHTML.
+
+Third-party library providers are quite justifiably envious of the special status of these few otherwise ordinary user-defined types.
+After all, it is not difficult to find other examples of types for which mathematicians, scientists and programmers have invented specialized notations.
+For example, 1) clients of a ``collections'' library might like not just list literals, but also vector, matrix, finite set and finite map literals following standard mathematical conventions; 2) clients of a ``web programming'' library might like not just XHTML literals but also CSS literals (which Ur/Web does not provide); 4) a compiler writer might like ``quotation'' literals for the terms of the object language and various intermediate languages of interest; and 5) clients of a ``chemistry'' library might like chemical structure literals based on the SMILES standard \cite{anderson1987smiles}.
+\citet{Bravenboer:2007:PIA:1289971.1289975} suggests that literal forms for structured encodings of SQL queries and other query types can reduce the temptation to use string encodings and therefore reduce the prevalence of string injection attacks, which are amongst the most common and serious security vulnerabilities on the web today \cite{owasp2017}.
+Choosing strings over composite data structures, a.k.a.
+``stringly-typed programming'', is quite common in practice -- \citet{TSLs} sampled strings from open source projects and found that at least 15\% of them were parseable according to some readily apparent type-specific grammar, e.g.
+for SQL queries, regexes, file system paths, URLs and others.
+Literal forms at these types, with support for splicing, would lower the cost of using more compositional encodings.
+%Literal forms would  preferable to string encodings because they support compositional construction and decomposition by pattern matching via splicing.
 
 \begin{figure}
+\vspace{-2px}
 \begin{lstlisting}
@@
 \end{lstlisting}
-\vspace{-6px}
+\vspace{-10px}
 \caption{XHTML literals with support for splicing at two different types are built primitively into Ur/Web \cite{conf/popl/Chlipala15}.}
-\vspace{-6px}
+\vspace{-7px}
 \label{fig:urweb}
@@
 
-Given this profusion of possible literal forms, it is not sensible to ask the designers of general-purpose languages to keep up.
-Instead, there has been considerable interest in mechanisms that allow programmers to introduce new type-specific literal forms in a decentralized manner.
-Unfortunately, existing mechanisms ...
-
-{This paper addresses this established need by introducing a system of \emph{typed literal macros (TLMs)} that allows programmers to define new literal syntax of nearly arbitrary form within an ML- or Scala-like  language}, i.e.
+Language designers cannot, of course, hope to accomodate all such types with primitive literal forms.
+Instead, there has been considerable interest in mechanisms that allow programmers to introduce new literal forms in a decentralized manner.
+We will cover a number of existing mechanisms, including Camlp4 \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, in Sec.
+\ref{sec:existing-approaches}.
+The problem is that these existing mechanisms quickly become \emph{unreasonable} when ``programming in the large'' \cite{DeRemer76}.
+%, i.e. when developing large programs ``consisting of many small programs (modules), possibly written by different people''
+For example, consider the perspective of a programmer attempting to comprehend (i.e.
+reason about) the program text in Figure \ref{fig:K-dialect}, which is written in a dialect of ML's (or Scala's) syntax where a large number of syntax extensions defining user-defined literal forms, some unfamiliar to the programmer, have been imported.
+
+\begin{figure}[h]
+\vspace{-9px}
+\begin{lstlisting}
+  val w = compute_w ()
+  val x = compute_x ()
+  val y = {|(!R)@&{&/x!/:2_!x}'!R}|}
+\end{lstlisting}
+\vspace{-12px}
+\caption{An example of unreasonable program text.}
+\vspace{-8px}
+\label{fig:K-dialect}
+\end{figure}
+
+\noindent
+If the programmer happens to be familiar with the intentionally terse syntax of the stack-based database query language K \cite{Whitney:2001:LOR:376284.375783}, then the literal form on Line 3 might be convenient.
+If the programmer does not recognize this syntax, however, there are no simple, definitive protocols for answering questions like:
+
+\begin{enumerate}
+\item \textbf{(Responsibility)} Which syntax extension determines the desugaring of the literal form on Line 3?
+Might there be many possible desugarings (i.e.
+syntactic non-determinism) due to conflicts between extensions?
+\item \textbf{(Segmentation)} Are the characters \li{x} and \li{R} on Line 3 taken as spliced expressions (meaning that they appear directly in the desugaring), or are they parsed in some other way peculiar to this literal form?
+\item \textbf{(Capture)} If \li{x} is in fact a spliced expression, does it refer to the binding of \li{x} on Line 2, or might it capture an unseen binding introduced in the desugaring of Line 3?
+\item \textbf{(Context Dependence)} If \li{w}, on Line 1, is renamed, could that possibly break Line 3, or change its meaning?
+In other words, might the desugaring of Line 3 assume that some variable identified as \li{w} is in scope (even though \li{w} is not mentioned in the text of Line 3)?
+\item \textbf{(Typing)} What type does \li{y} have?
+\end{enumerate}
+
+In short, the problem is that library providers cannot reason modularly about syntactic determinism (because other extensions might define overlapping forms), and if the desugaring of the program text is held abstract, programmers can no longer reason about types, binding and segmentation (i.e.
+answer questions like those above.) This is burdensome at all scales, but particularly when programming in the large, where it is common to encounter a wide variety of libraries \cite{DBLP:conf/sac/LammelPS11}.
+Forcing the programmer to examine the desugaring of the program text, or the logic that produced that desugaring, in order to answer these sorts of basic questions about segmentation, binding and typing can defeat the ultimate purpose of syntactic sugar -- decreasing cognitive cost \cite{Green89}.
+Analagous problems do not arise when programming without syntax extensions in languages like ML and Scala -- the binding discipline of the language ensures that programmers can reason lexically about binding and renaming, and types mediate abstraction over function and module implementations.
+%From this perspective, it is unsurprising that the use of Camlp4 has been deprecated in the OCaml system.
+
+This paper introduces a more reasonable approach: typed literal macros (TLMs).
+TLMs give library providers full programmatic control over the parsing and expansion of \emph{generalized literal forms}, whose bodies are left initially unparsed according to the fixed context free syntax of the language (i.e.
+they are initially parsed like raw string literals.) The applied TLM parses the literal body and generates an expansion during a phase known as typed expansion, which generalizes the usual typing phase.
+TLMs come equipped with uniquely powerful abstract reasoning principles.
+We will more precisely characterize these abstract reasoning principles as we proceed.
+For now, to develop some intuitions, consider Figure \ref{fig:K-tsm-example}, which applies a TLM to express the ``unreasonable'' example from Figure \ref{fig:K-dialect}.
+Without examining the expansion of Line 3, the client programmer can reason as follows:
+\begin{enumerate}
+\item \textbf{(Responsibility)} The applied TLM, \li{$kquery}, is solely responsible for typed expansion of the literal body. 
+\item \textbf{(Segmentation)} By examining the segmentation, we know that the two instances of \li{x} on Line 3 are parsed as spliced expressions, whereas \li{R} is parsed in some other way peculiar to this form.
+\item \textbf{(Capture)} The system prevents capture, so the spliced expression \li{x} must refer to the binding of \li{x} on Line 2 -- it cannot capture an unseen binding introduced in the expansion of Line 3.
+\item \textbf{(Context Dependence)} The system enforces context independence, so the expansion of Line 3 cannot  rely on the fact that, for example, \li{w} is in scope.
+\item \textbf{(Typing)} An explicit type annotation on the definition of \li{$kquery} determines the type that every expansion it generates will have.
+We will see an example of a TLM definition in the next section.
+
+Moreover, each segment in the segmentation also comes paired with the type it is expected to have.
+This information is usually not necessary to reason about typing, but it can be conveyed to the programmer upon request by the program editor if desired.
+%TLM definitions follow the usual scoping rules, so it is easy to ``jump to the definition'' of \li{$kquery}.
+\end{enumerate}
+
+\begin{figure}
+\begin{lstlisting}[numbers=none,xleftmargin=0px]
+  val w = compute_w ()
+  val x = compute_x ()
+  val y = $kquery `SURL(!R)@&{&/EURLxSURL!/:2_!EURLxSURL}'!R}EURL`
+\end{lstlisting}
+\vspace{-5px}
+\caption{TLMs make examples like the one from Fig. \ref{fig:K-dialect} more reasonable.}
+\vspace{-5px}
+\label{fig:K-tsm-example}
+\end{figure}
+
+WE WILL CONSIDER A VARIETY OF TERM REWRITING SYSTEMS, INCLUDING PPX, EARLY MACROS, READER MACROS, TEMPLATE HASKELL, QUASIQUOTES, HYGIENIC MACROS, TYPED HYGIENIC MACROS ALA SCALA, TYPE-SPECIFIC LANGUAGES, IN SEC 6.
+FOR NOW, LET US JUMP IMMEDIATELY TO OUR SYSTEM OF TYPED LITERAL MACROS AND DEVELOP AN INTUITION FOR HOW THEY ADDRESS THE PROBLEMS RAISED ABOVE.
+THE XHTML EXAMPLE CAN BE EXPRESSED IN OUR SYSTEM AS FOLLOWS:
+
+BUT THAT DOESNT WORK BECAUSE OF HYGIENE! THERE ARE OTHER PROBLEMS AS WELL,
+
+For example, OCaml's textual syntax now includes \emph{preprocessor extension (ppx) points} used to identify terms that some term rewriting preprocessor must rewrite \cite{ocaml-manual}.
+We could attempt to mark a string literal containing Ur/Web-style HTML syntax with a ppx annotation, \li{xml}, as follows:
+\begin{lstlisting}[numbers=none]
+  [%xml "SSTR<h1>Hello, {[first_name]}!</h1>ESTR"]
+\end{lstlisting}
+Again, there are problems reasoning modularly and abstractly about such constructions.
+More than one applied preprocessor might recognize this annotation (there are, in practice, many XML/HTML libraries), so the problems of \textbf{Responsibility} come up.
+It is also impossible to reason abstractly about \textbf{Segmentation}, \textbf{Capture}, \textbf{Context Dependence} and \textbf{Typing} because the code that the preprocessor generates is arbitrary.
+For these reasons, users of this mechanism warn that they should be used sparingly.\todo{citation}
+
+Term-rewriting macro systems are language-integrated local term rewriting systems that require that the client explicitly apply the intended rewriting, implemented by a macro, to the term that is to be rewritten\todo{readeer macros}.
+This addresses the issue of \textbf{Responsibility}.
+However, na\"ive macro systems, like the earliest variants of the Lisp macro system \cite{Hart63a}, Template Haskell \cite{SheardPeytonJones:Haskell-02} and GHC's quasiquotation system \cite{mainland2007s}, do not allow clients to reason abstractly about the remaining issues, again because the expansion that they produce is arbitrary.
+
+\emph{Hygienic} macro systems prevent, or abstractly account for \cite{DBLP:conf/esop/HermanW08,Herman10:Theory}, \textbf{Capture}, and they enforce application-site \textbf{Context Independence} \cite{Kohlbecker86a,DBLP:conf/popl/Adams15,DBLP:conf/popl/ClingerR91,DBLP:journals/lisp/DybvigHB92}.
+However, this turns out to make it difficult to use hygienic macros to introduce new literal forms of arbitrary design with support for splicing.
+Consider again our running XHTML example, which we might try to realize by applying a hygienic macro, \li{xml!}, to a string literal:
+\begin{lstlisting}[numbers=none]
+  (xml! "SSTR<h1>Hello, {[first_name]}!</h1>ESTR")
+\end{lstlisting}
+THIS IS BAD BECAUSE FIRSTNAME IS BAD HYGIENE GRAH. MAYBE TALK ABOUT STRING SPLICING AND FRAGMENT SYSTEM STUFF HERE.
+
+PUTTING THAT ASIDE, LET US CONSIDER TYPING.
+LISP IS DUMB NO TYPES.
+HERMAN AND WAND USED TYPES ONLY TO ACCOUNT FOR BINDING RELATIONSHIPS.
+SCALA HAS A TYPED MACRO SYSTEM, BUT THERE IS NO FORMAL ACCOUNT OF IT.
+MACROML IS ONLY A STAGING MACRO SYSTEM, NO WAY TO PARSE LITERALS.
+(It is not enough that with Template Haskell / GHC quasiquotation, the generated expansion is typechecked -- to satisfy the \textbf{Typing} criterion, it must be possible to reason abstractly about \emph{what the type of the generated expansion is}.) MAYBE BRING UP TSLS AND TSMS HERE.
+
+TRANSITION TO THE CONTRIBUTIONS OF THIS PAPER: TYPED LITERAL MACROS.
+REWRITE GENERALIZED LITERAL FORMS, KEEP TRACK OF SPLICED SEGMENTS SO THE HYGIENE MECHANISM IS COMPATIBLE, SEGMENTATION IS ALWAYS POSSIBLE.
+TYPED, INTEGRATED INTO ML SCALA-LIKE LANGUAGE (SECTION HEADINGS).
+FORMAL DEFINITION.
+REASONING PRINCIPLES ARE THEOREMS.
+THATS HOW WE EVALUATE OUR WORK.
+(USEFULNESS / USABILITY CONCERNS NEED TO BE ADDRESSED SOMEWHERES)
+
+Although much of the research on macro systems has been for languages in the Lisp tradition \cite{mccarthy1978history}, some progress has also been made on reasoning about types with the design of \emph{typed macro systems} where annotations constrain the macro arguments and the generated expansions.
+Examples include Scala's macro system \cite{ScalaMacros2013}, formal calculi studied by \citet{DBLP:conf/esop/HermanW08} and \emph{staging macro systems} like MetaML \cite{Sheard:1999:UMS} and MacroML \cite{ganz2001macros} (which, it is worth noting, do not give the macro access to the syntax tree of arguments.)
+
+Perhaps the biggest problem, then, is that term-rewriting systems offer library providers only limited syntactic control.
+In other words many syntactic forms of interest will not particularly resemble those available in the host language.
+For example, consider the  XHTML and K examples above.
+In both cases, the syntactic conventions are quite distinct from those of ML-like languages (and, for that matter, languages that use S-expression.)
+
+It is tempting in these situations to consider repurposing string literal forms.
+For example, we might wish to apply a macro \li{html!} (following Rust's convention of using a post-fix \li{!} to distinguish macro names from variables) to rewrite string literals containing Ur/Web-style XHTML syntax as follows:
+\begin{lstlisting}[numbers=none]
+  fun heading first_name = html! 'SSTR<h1>Hello, {[first_name]}!</p>ESTR'
+  val body = html! 'SSTR<body>{heading "World"} ...</body>ESTR'
+\end{lstlisting}
+
+The problem here is that there is no way to extract the spliced expressions from the supplied string literal forms while satisfying the hygiene conditions, because variables that come from these spliced terms (here, \li{first_name} and \li{heading}) are indistinguishable from variables that inappropriately appear free relative to the expansion.
+In addition, the problem of segmentation becomes even more pernicious: to a human or tool unaware of Ur/Web's syntax, it is not immediately apparent which particular subsequences of the string literals supplied to \li{html!} are segmented out as spliced expressions.
+Reader macros have essentially the same problem  \cite{DBLP:journals/jfp/FlattCDF12}.
+
+  Some languages build in \emph{string splicing} (a.k.a.
+\emph{string interpolation}) forms, e.g.
+Scala \cite{odersky2008programming}, or similar but more general \emph{fragmentary quotation forms} \cite{conf/icfp/Slind91}, e.g.
+the SML/NJ dialect of ML.
+These designate particular delimiters to allow escape out into the expression language.
+The problem with using these (dynamically, or in conjunction with term rewriting macros) as vehicles to introduce syntactic sugar at various types is 1) there is no ``one-size-fits-all'' escape delimiter, and 2) reasoning about typing is problematic because every escaped term is checked against the same type.
+In the example above, we spliced in two different types of expressions using two different delimiters.
+In general, e.g.
+when defining syntax for a programming language with many sorts of terms, the most appropriate choice of delimiters might depend on where each spliced term appears.
+
+{This paper introducing a system of \emph{typed literal macros (TLMs)} that allows programmers to define new literal syntax of nearly arbitrary form within an ML- or Scala-like  language}, i.e.
 a language with a rich static type system (Sec.
@@
 In other words,
-TLMs come equipped with useful principles of syntactic abstraction.
-We will more precisely characterize these abstract reasoning principles as we proceed.
-For now, to develop some intuitions, consider Figure \ref{fig:K-tsm-example}, which uses TLMs to express the ``unreasonable'' example from Figure \ref{fig:K-dialect}.
-Without examining the expansion of Line 3, we can reason as follows:
-
-\begin{enumerate}
-\item \textbf{(Responsibility)} The applied TLM, \li{$kquery}, is solely responsible for typed expansion of the literal body. 
-\item \textbf{(Segmentation)} By examining the segmentation, we know that the two instances of \li{x} on Line 3 are parsed as spliced expressions, whereas \li{R} is parsed in some other way peculiar to this form.
-\item \textbf{(Capture)} The system prevents capture, so the spliced expression \li{x} must refer to the binding of \li{x} on Line 2 -- it cannot capture an unseen binding introduced in the expansion of Line 3.
-\item \textbf{(Context Dependence)} The system enforces context independence, so the expansion of Line 3 cannot  rely on the fact that, for example, \li{w} is in scope.
-\item \textbf{(Typing)} An explicit type annotation on the definition of \li{$kquery} determines the type that every expansion it generates will have.
-We will see an example of a TLM definition in the next section.
-
-Moreover, each segment in the segmentation also comes paired with the type it is expected to have.
-This information is usually not necessary to reason about typing, but it can be conveyed to the programmer upon request by the program editor if desired.
-%TLM definitions follow the usual scoping rules, so it is easy to ``jump to the definition'' of \li{$kquery}.
-\end{enumerate}
-
-\begin{figure}
-\begin{lstlisting}[numbers=none,xleftmargin=0px]
-  val w = compute_w ()
-  val x = compute_x ()
-  val y = $kquery `SURL(!R)@&{&/EURLxSURL!/:2_!EURLxSURL}'!R}EURL`
-\end{lstlisting}
-\vspace{-5px}
-\caption{TLMs make examples like the one from Fig. \ref{fig:K-dialect} more reasonable.}
-\vspace{-5px}
-\label{fig:K-tsm-example}
-\end{figure}
 
@@
 
-\section{Reasoning Criteria and Existing Approaches}
+\section{Reasoning Criteria and Existing Approaches}\label{sec:existing-approaches}
 \subsection{Library-Specific Syntax Dialects}\label{sec:syntax-dialects-intro}
@@
 
-\begin{figure}
-\begin{lstlisting}
-  val w = compute_w ()
-  val x = compute_x ()
-  val y = {|(!R)@&{&/x!/:2_!x}'!R}|}
-\end{lstlisting}
-\vspace{-8px}
-\caption{An example of unreasonable program text.}
-\vspace{-8px}
-\label{fig:K-dialect}
-\end{figure}
-
 Even putting aside the problem of syntactic conflict, there are questions about just how reasonable sprinkling many possibly unfamiliar library-specific literal forms  throughout a program may be.
-For example, consider the perspective of a programmer attempting to comprehend (i.e.
-reason about) the program fragment in Figure \ref{fig:K-dialect}, which is written in a syntax dialect constructed by combining various library-specific dialects of ML's (or Scala's) syntax.
-If the programmer happens to be familiar with the (intentionally terse) syntax of the stack-based database query processing language K \cite{Whitney:2001:LOR:376284.375783}, then Line 3 might pose few difficulties.
-If the programmer does not recognize this syntax, however, there are no simple, definitive protocols for answering questions like:
-
-\begin{enumerate}
-\item \textbf{(Responsibility)} Which constituent dialect defined the literal form that appears on Line 3?
-\item \textbf{(Segmentation)} Are the characters \li{x} and \li{R} on Line 3 parsed as spliced expressions \li{x} and \li{R} (i.e.
-expressions of variable form), or parsed in some other way peculiar to this literal form?
-\item \textbf{(Capture)} If \li{x} is in fact a spliced expression, does it refer to the binding of \li{x} on Line 2? Or might it capture an unseen binding introduced in the desugaring of Line 3?
-\item \textbf{(Context Dependence)} If \li{w}, on Line 1, is renamed, could that possibly break Line 3, or change its meaning?
-In other words, might the desugaring of Line 3 assume that some variable identified as \li{w} is in scope (even though \li{w} is not mentioned in the text of Line 3)?
-\item \textbf{(Typing)} What type does \li{y} have?
-\end{enumerate}
-
-In short, syntax dialects do not come equipped with  principles of \emph{syntactic abstraction}: if the desugaring of the program text is held abstract, programmers can no longer reason about types, binding and segmentation (i.e.
-answer questions like those above) in the usual disciplined manner.
-This is burdensome at all scales, but particularly when programming in the large, where it is common to encounter unfamiliar library constructs.
-Forcing the programmer to examine the desugaring of the program text, or the logic that produced that desugaring, in order to answer these sorts of basic questions about segmentation, binding and typing can defeat the ultimate purpose of syntactic sugar -- decreasing cognitive cost \cite{Green89}.
 
@@
 \subsection{Another Existing Approach: Term Rewriting}
-An alternative approach (and the approach that our mechanism is rooted in) is to leave the context-free syntax of the language unmodified and instead contextually repurpose existing forms using a \emph{term rewriting system}.
-
-For example, OCaml's textual syntax now includes \emph{preprocessor extension (ppx) points} used to identify terms that some term rewriting preprocessor must rewrite \cite{ocaml-manual}.
-We could attempt to mark a string literal containing Ur/Web-style HTML syntax with a ppx annotation, \li{xml}, as follows:
-\begin{lstlisting}[numbers=none]
-  [%xml "SSTR<h1>Hello, {[first_name]}!</h1>ESTR"]
-\end{lstlisting}
-Again, there are problems reasoning modularly and abstractly about such constructions.
-More than one applied preprocessor might recognize this annotation (there are, in practice, many XML/HTML libraries), so non-determinism can arise, and it is difficult to reason about \textbf{Responsibility}.
-It is also difficult to reason abstractly about \textbf{Segmentation}, \textbf{Capture}, \textbf{Context Dependence} and \textbf{Typing} because the code that the preprocessor generates is arbitrary.
-For these reasons, users of this mechanism warn that they should be used sparingly.\todo{citation}
-
-Term-rewriting macro systems address these problems -- the applied macro explicitly identifies the rewriting that wil
-
-Template Haskell / Haskell quasiquotations / early macro systems / reader macros
-
-hygienic macro systems -- problem!
-
-typed macro systems, e.g. Scala. mention MacroML stuff, Herman and Wand 2008/2010 work
-
-string splicing / interpolation
-
-TSLs / TSMs
-
-Non-local rewriting systems suffer from problems analagous to those that plague syntax definition systems.
-In particular, there can be conflicts because separately defined rewriting rules might attempt to rewrite the same term differently.
-It can also be difficult to determine which rewriting rule, if any, is responsible for any particular term.
-Certain conventions have emerged, e.g.
-OCaml's textual syntax now includes \emph{preprocessor extension (ppx) points} used to identify terms that some presumably corresponding term rewriting preprocessor must rewrite \cite{ocaml-manual}, but a correspondence is not enforced.
-In any case, it is also difficult to reason abstractly about types and binding given a program subject to a large number of rewriting rules.
-
-Modern \emph{term-rewriting macro systems}, however, have made some progress. In particular:
-\begin{enumerate}[nolistsep]
-\item Macro systems require that the client explicitly apply the intended rewriting (implemented by a macro) to the term that is to be rewritten, thereby addressing the problems of conflict and determining responsibility.
-However, it is often unclear whether a given macro  is repurposing the form of a given argument or sub-term thereof, as opposed to treating it parametrically by inserting it unmodified into the generated expansion.
-This is closely related to the problem of determining a {segmentation} discussed above.
-\item \emph{Hygienic} macro systems prevent variable {capture} and enforce {context-independence} \cite{Kohlbecker86a,DBLP:conf/popl/Adams15,DBLP:conf/popl/ClingerR91,DBLP:journals/lisp/DybvigHB92}.
-\item Although much of the research on macro systems has been for languages in the Lisp tradition \cite{mccarthy1978history}, some progress has also been made on reasoning about types with the design of \emph{typed macro systems} where annotations constrain the macro arguments and the generated expansions.
-Examples include Scala's macro system \cite{ScalaMacros2013}, formal calculi studied by \citet{DBLP:conf/esop/HermanW08} and \emph{staging macro systems} like MetaML \cite{Sheard:1999:UMS} and MacroML \cite{ganz2001macros} (which, it is worth noting, do not give the macro access to the syntax tree of arguments.)
-\end{enumerate}
-
-Perhaps the biggest problem, then, is that term-rewriting systems offer library providers only limited syntactic control.
-In other words many syntactic forms of interest will not particularly resemble those available in the host language.
-For example, consider the  XHTML and K examples above.
-In both cases, the syntactic conventions are quite distinct from those of ML-like languages (and, for that matter, languages that use S-expression.)
-
-It is tempting in these situations to consider repurposing string literal forms.
-For example, we might wish to apply a macro \li{html!} (following Rust's convention of using a post-fix \li{!} to distinguish macro names from variables) to rewrite string literals containing Ur/Web-style XHTML syntax as follows:
-\begin{lstlisting}[numbers=none]
-  fun heading first_name = html! 'SSTR<h1>Hello, {[first_name]}!</p>ESTR'
-  val body = html! 'SSTR<body>{heading "World"} ...</body>ESTR'
-\end{lstlisting}
-
-The problem here is that there is no way to extract the spliced expressions from the supplied string literal forms while satisfying the hygiene conditions, because variables that come from these spliced terms (here, \li{first_name} and \li{heading}) are indistinguishable from variables that inappropriately appear free relative to the expansion.
-In addition, the problem of segmentation becomes even more pernicious: to a human or tool unaware of Ur/Web's syntax, it is not immediately apparent which particular subsequences of the string literals supplied to \li{html!} are segmented out as spliced expressions.
-Reader macros have essentially the same problem  \cite{DBLP:journals/jfp/FlattCDF12}.
-
-  Some languages build in \emph{string splicing} (a.k.a.
-\emph{string interpolation}) forms, e.g.
-Scala \cite{odersky2008programming}, or similar but more general \emph{fragmentary quotation forms} \cite{conf/icfp/Slind91}, e.g.
-the SML/NJ dialect of ML.
-These designate particular delimiters to allow escape out into the expression language.
-The problem with using these (dynamically, or in conjunction with term rewriting macros) as vehicles to introduce syntactic sugar at various types is 1) there is no ``one-size-fits-all'' escape delimiter, and 2) reasoning about typing is problematic because every escaped term is checked against the same type.
-In the example above, we spliced in two different types of expressions using two different delimiters.
-In general, e.g.
-when defining syntax for a programming language with many sorts of terms, the most appropriate choice of delimiters might depend on where each spliced term appears.
 

### 2017-06-05 popl18/syntax-popl18.tex
finished reorganizing intro

@@
 \begin{figure}[h]
-\vspace{-9px}
+\vspace{-10px}
 \begin{lstlisting}
@@
 \end{lstlisting}
-\vspace{-12px}
+\vspace{-10px}
 \caption{An example of unreasonable program text.}
-\vspace{-8px}
+\vspace{-9px}
 \label{fig:K-dialect}
@@
 This paper introduces a more reasonable approach: typed literal macros (TLMs).
-TLMs give library providers full programmatic control over the parsing and expansion of \emph{generalized literal forms}, whose bodies are left initially unparsed according to the fixed context free syntax of the language (i.e.
+TLMs give library providers full programmatic control over the parsing and expansion of \emph{generalized literal forms}, which syntactically subsume other literal forms because their bodies are left initially unparsed according to the fixed context free syntax of the language (i.e.
 they are initially parsed like raw string literals.) The applied TLM parses the literal body and generates an expansion during a phase known as typed expansion, which generalizes the usual typing phase.
+For example, Fig.
+\ref{fig:K-tsm-example} applies a TLM to express the ``unreasonable'' example from Fig.
+\ref{fig:K-dialect}.
+
 TLMs come equipped with uniquely powerful abstract reasoning principles.
 We will more precisely characterize these abstract reasoning principles as we proceed.
-For now, to develop some intuitions, consider Figure \ref{fig:K-tsm-example}, which applies a TLM to express the ``unreasonable'' example from Figure \ref{fig:K-dialect}.
-Without examining the expansion of Line 3, the client programmer can reason as follows:
+For now, let us develop some informal intuitions -- the client programmer can reason about Line 3 as follows:
 \begin{enumerate}
 \item \textbf{(Responsibility)} The applied TLM, \li{$kquery}, is solely responsible for typed expansion of the literal body.
-\item \textbf{(Segmentation)} By examining the segmentation, we know that the two instances of \li{x} on Line 3 are parsed as spliced expressions, whereas \li{R} is parsed in some other way peculiar to this form.
-\item \textbf{(Capture)} The system prevents capture, so the spliced expression \li{x} must refer to the binding of \li{x} on Line 2 -- it cannot capture an unseen binding introduced in the expansion of Line 3.
+The context-free syntax of the language is fixed, so the TLM provider can modularly establish that the syntax the TLM implements is deterministic.
+\item \textbf{(Segmentation)} The output that a TLM generates is structured so that the system can produce from it an accurate \emph{segmentation} of the literal body that distinguishes spliced segments, which appear directly in the expansion, from segments that are parsed in some other way by the TLM.
+This segmentation can be communicated by a pretty printer or code editor using secondary notation, e.g.
+colors in this document.
+By examining the segmentation of Line 3, we know that the two instances of \li{x} on Line 3 are parsed as spliced expressions, whereas \li{R} is parsed in some other way.
+\item \textbf{(Capture)} The system avoids capture, so the spliced expression \li{x} must refer to the binding of \li{x} on Line 2.
+It cannot capture an unseen binding of \li{x} in the expansion of Line 3.
 \item \textbf{(Context Dependence)} The system enforces context independence, so the expansion of Line 3 cannot  rely on the fact that, for example, \li{w} is in scope.
-\item \textbf{(Typing)} An explicit type annotation on the definition of \li{$kquery} determines the type that every expansion it generates will have.
+The client is therefore free to rename \li{w}.
+\item \textbf{(Typing)} An explicit type annotation on the definition of \li{$kquery} determines the type that the expansion must have.
 We will see an example of a TLM definition in the next section.
-
-Moreover, each segment in the segmentation also comes paired with the type it is expected to have.
+Moreover, each spliced segment in the segmentation also comes paired with the type it is expected to have.
 This information is usually not necessary to reason about typing, but it can be conveyed to the programmer upon request by the program editor if desired.
@@
 
+%\footnote{In general, spliced expressions might themselves apply TLMs, in which case the convention is to use a distinct color for unspliced segments at each depth.
+For example, if strings were not primitive, we might write \li{`SURL<body>\{EURLheading ($str "SSTRWorld!ESTR")SURL\} ...</body>EURL`}.}
+
 \begin{figure}
+\vspace{-4px}
 \begin{lstlisting}[numbers=none,xleftmargin=0px]
@@
 \end{lstlisting}
-\vspace{-5px}
+\vspace{-8px}
 \caption{TLMs make examples like the one from Fig. \ref{fig:K-dialect} more reasonable.}
-\vspace{-5px}
+\vspace{-8px}
 \label{fig:K-tsm-example}
@@
 
-WE WILL CONSIDER A VARIETY OF TERM REWRITING SYSTEMS, INCLUDING PPX, EARLY MACROS, READER MACROS, TEMPLATE HASKELL, QUASIQUOTES, HYGIENIC MACROS, TYPED HYGIENIC MACROS ALA SCALA, TYPE-SPECIFIC LANGUAGES, IN SEC 6.
-FOR NOW, LET US JUMP IMMEDIATELY TO OUR SYSTEM OF TYPED LITERAL MACROS AND DEVELOP AN INTUITION FOR HOW THEY ADDRESS THE PROBLEMS RAISED ABOVE.
-THE XHTML EXAMPLE CAN BE EXPRESSED IN OUR SYSTEM AS FOLLOWS:
-
-BUT THAT DOESNT WORK BECAUSE OF HYGIENE! THERE ARE OTHER PROBLEMS AS WELL,
-
-For example, OCaml's textual syntax now includes \emph{preprocessor extension (ppx) points} used to identify terms that some term rewriting preprocessor must rewrite \cite{ocaml-manual}.
-We could attempt to mark a string literal containing Ur/Web-style HTML syntax with a ppx annotation, \li{xml}, as follows:
-\begin{lstlisting}[numbers=none]
-  [%xml "SSTR<h1>Hello, {[first_name]}!</h1>ESTR"]
-\end{lstlisting}
-Again, there are problems reasoning modularly and abstractly about such constructions.
-More than one applied preprocessor might recognize this annotation (there are, in practice, many XML/HTML libraries), so the problems of \textbf{Responsibility} come up.
-It is also impossible to reason abstractly about \textbf{Segmentation}, \textbf{Capture}, \textbf{Context Dependence} and \textbf{Typing} because the code that the preprocessor generates is arbitrary.
-For these reasons, users of this mechanism warn that they should be used sparingly.\todo{citation}
-
-Term-rewriting macro systems are language-integrated local term rewriting systems that require that the client explicitly apply the intended rewriting, implemented by a macro, to the term that is to be rewritten\todo{readeer macros}.
-This addresses the issue of \textbf{Responsibility}.
-However, na\"ive macro systems, like the earliest variants of the Lisp macro system \cite{Hart63a}, Template Haskell \cite{SheardPeytonJones:Haskell-02} and GHC's quasiquotation system \cite{mainland2007s}, do not allow clients to reason abstractly about the remaining issues, again because the expansion that they produce is arbitrary.
-
-\emph{Hygienic} macro systems prevent, or abstractly account for \cite{DBLP:conf/esop/HermanW08,Herman10:Theory}, \textbf{Capture}, and they enforce application-site \textbf{Context Independence} \cite{Kohlbecker86a,DBLP:conf/popl/Adams15,DBLP:conf/popl/ClingerR91,DBLP:journals/lisp/DybvigHB92}.
-However, this turns out to make it difficult to use hygienic macros to introduce new literal forms of arbitrary design with support for splicing.
-Consider again our running XHTML example, which we might try to realize by applying a hygienic macro, \li{xml!}, to a string literal:
-\begin{lstlisting}[numbers=none]
-  (xml! "SSTR<h1>Hello, {[first_name]}!</h1>ESTR")
-\end{lstlisting}
-THIS IS BAD BECAUSE FIRSTNAME IS BAD HYGIENE GRAH. MAYBE TALK ABOUT STRING SPLICING AND FRAGMENT SYSTEM STUFF HERE.
-
-PUTTING THAT ASIDE, LET US CONSIDER TYPING.
-LISP IS DUMB NO TYPES.
-HERMAN AND WAND USED TYPES ONLY TO ACCOUNT FOR BINDING RELATIONSHIPS.
-SCALA HAS A TYPED MACRO SYSTEM, BUT THERE IS NO FORMAL ACCOUNT OF IT.
-MACROML IS ONLY A STAGING MACRO SYSTEM, NO WAY TO PARSE LITERALS.
-(It is not enough that with Template Haskell / GHC quasiquotation, the generated expansion is typechecked -- to satisfy the \textbf{Typing} criterion, it must be possible to reason abstractly about \emph{what the type of the generated expansion is}.) MAYBE BRING UP TSLS AND TSMS HERE.
-
-TRANSITION TO THE CONTRIBUTIONS OF THIS PAPER: TYPED LITERAL MACROS.
-REWRITE GENERALIZED LITERAL FORMS, KEEP TRACK OF SPLICED SEGMENTS SO THE HYGIENE MECHANISM IS COMPATIBLE, SEGMENTATION IS ALWAYS POSSIBLE.
-TYPED, INTEGRATED INTO ML SCALA-LIKE LANGUAGE (SECTION HEADINGS).
-FORMAL DEFINITION.
-REASONING PRINCIPLES ARE THEOREMS.
-THATS HOW WE EVALUATE OUR WORK.
-(USEFULNESS / USABILITY CONCERNS NEED TO BE ADDRESSED SOMEWHERES)
-
-Although much of the research on macro systems has been for languages in the Lisp tradition \cite{mccarthy1978history}, some progress has also been made on reasoning about types with the design of \emph{typed macro systems} where annotations constrain the macro arguments and the generated expansions.
-Examples include Scala's macro system \cite{ScalaMacros2013}, formal calculi studied by \citet{DBLP:conf/esop/HermanW08} and \emph{staging macro systems} like MetaML \cite{Sheard:1999:UMS} and MacroML \cite{ganz2001macros} (which, it is worth noting, do not give the macro access to the syntax tree of arguments.)
-
-Perhaps the biggest problem, then, is that term-rewriting systems offer library providers only limited syntactic control.
-In other words many syntactic forms of interest will not particularly resemble those available in the host language.
-For example, consider the  XHTML and K examples above.
-In both cases, the syntactic conventions are quite distinct from those of ML-like languages (and, for that matter, languages that use S-expression.)
-
-It is tempting in these situations to consider repurposing string literal forms.
-For example, we might wish to apply a macro \li{html!} (following Rust's convention of using a post-fix \li{!} to distinguish macro names from variables) to rewrite string literals containing Ur/Web-style XHTML syntax as follows:
-\begin{lstlisting}[numbers=none]
-  fun heading first_name = html! 'SSTR<h1>Hello, {[first_name]}!</p>ESTR'
-  val body = html! 'SSTR<body>{heading "World"} ...</body>ESTR'
-\end{lstlisting}
-
-The problem here is that there is no way to extract the spliced expressions from the supplied string literal forms while satisfying the hygiene conditions, because variables that come from these spliced terms (here, \li{first_name} and \li{heading}) are indistinguishable from variables that inappropriately appear free relative to the expansion.
-In addition, the problem of segmentation becomes even more pernicious: to a human or tool unaware of Ur/Web's syntax, it is not immediately apparent which particular subsequences of the string literals supplied to \li{html!} are segmented out as spliced expressions.
-Reader macros have essentially the same problem  \cite{DBLP:journals/jfp/FlattCDF12}.
-
-  Some languages build in \emph{string splicing} (a.k.a.
-\emph{string interpolation}) forms, e.g.
-Scala \cite{odersky2008programming}, or similar but more general \emph{fragmentary quotation forms} \cite{conf/icfp/Slind91}, e.g.
-the SML/NJ dialect of ML.
-These designate particular delimiters to allow escape out into the expression language.
-The problem with using these (dynamically, or in conjunction with term rewriting macros) as vehicles to introduce syntactic sugar at various types is 1) there is no ``one-size-fits-all'' escape delimiter, and 2) reasoning about typing is problematic because every escaped term is checked against the same type.
-In the example above, we spliced in two different types of expressions using two different delimiters.
-In general, e.g.
-when defining syntax for a programming language with many sorts of terms, the most appropriate choice of delimiters might depend on where each spliced term appears.
-
-{This paper introducing a system of \emph{typed literal macros (TLMs)} that allows programmers to define new literal syntax of nearly arbitrary form within an ML- or Scala-like  language}, i.e.
-a language with a rich static type system (Sec.
-2), support for pattern matching (Sec.
-3) and an ML-style module system or Scala-style object system with support for type members and type abstraction (Sec.
-5).
-The examples in this paper are written in VerseML, a new dialect of ML that features TLMs.
-
-Compared to existing approaches, TLMs are more \emph{reasonable}.
-We justify this claim by (1) developing a collection of relevant reasoning principles based on the reasoning principles already available to client programmers when using an ML-like or Scala-like language without such a mechanism; (2) formally operationalizing these reasoning principles as metatheorems about a series of type-theoretic \emph{TLM calculi} of increasing sophistication (Sec.
-4-5); and (3) evaluating existing approaches -- including syntax definition systems like Camlp4 \cite{ocaml-manual}\todo{better citation}, SugarJ/Sugar* \cite{erdweg2011sugarj,erdweg2013framework} and Copper \cite{conf/pldi/SchwerdfegerW09,schwerdfeger2010context}, term rewriting macro systems as found in a variety of languages, and the system of type-specific languages (TSLs) and typed syntax macros (TSMs) described by \citet{TSLs,sac15} -- relative to these reasoning principles (Sec.
-6).
-We conclude by discussing limitations and future research directions in Sec.
-7\todo{sec refs}.
-A complete set of definitions and associated proofs are available in the  supplemental material.
-This is the first detailed type-theoretic accout of a typed, hygienic term-rewriting macro system for an ML- or Scala-like language.\todo{refine and justify this claim}
-
-% These existing forms normally have other meanings, so this can be confusing \cite{pane1996usability}.
-
-\subsection{Contributions}\label{sec:contributions}
-This work introduces a system of \textbf{typed literal macros (TLMs)} that gives library providers substantially more syntactic control than existing typed term-rewriting macro systems while maintaining the ability to reason abstractly about types, binding and segmentation.% abstract reasoning principles.
+The remainder of this document is organized as follows.
+\begin{itemize}
+\item Sec.
+\ref{sec:setsms} begins by going through the mechanics of defining a {simple expression TLM} for XHTML and describes initial parsing, typed expansion and expansion validation in more detail.
+\item Sec. \ref{sec:sptsms} does the same for a {simple pattern TLM}, i.e. a TLM that generates patterns. %The hygiene condition for pattern TLMs is interesting.
+\item Sec. \ref{sec:setsms-formally} defines a calculus of simple expression and pattern TLMs and formally establishes the reasoning principles implied above.
+
+% In the remaining chapters, we enrich the language developed in Chapter \ref{chap:uetsms} with advanced features based on those found in full-scale general-purpose languages like ML and Scala, and enhance our TLM mechanism with these new features.
+
+% We next develop a more sophisticated account of TLMs in Chapters \ref{chap:ptsms} and \ref{chap:static-eval}, with the aim of working out details necessary to integrate TLMs into full-scale general-purpose languages like ML or Scala.
+%These two chapters constitute Part \ref{part:parametric-tsms} of our contributions.
+\item Sec.
+\ref{sec:ptsms} adds parameterized types and an ML-style module system and introduces \emph{parametric TLMs} (pTLMs), i.e.
+TLMs that take type and module parameters.
+Parameters serve two purposes:
+\begin{enumerate}
+\item They allow for TLMs that operate not just at a single type, but over a type- and module-parameterized family of types.
+For example, we consider a parametric TLM for dictionaries that operates uniformly over different implementations of the abstract dictionary signature, and also over different choices of the co-domain type.
+\item Parameters also give expansions access to terms other than those explicitly spliced in. %This is particularly convenient for working with abstract types \emph{a la} ML. 
+Partial parameter application in TLM abbreviations lowers the cost of this parameter-passing discipline.
+\end{enumerate}
+This section culminates with a calculus that extends the calculus of Sec. \ref{sec:setsms-formally} with ML-style modules and parametric TLMs.
+\item Sec.
+\ref{sec:static-eval} considers the topic of term evaluation during the typed expansion phase, i.e.
+\emph{static evaluation}, in  more detail, and also gives examples of TLMs that are useful for defining other TLMs, e.g.
+TLMs that implement parser generators and quasiquotation.
+\item Sec.
+\ref{sec:related-work} compares TLMs to the related work, including closely related work by \citet{TSLs,sac15} on type-specific languages.
+\item Sec.
+\ref{sec:discussion} concludes with a summary of our contributions and a brief discussion of present limitations and future research directions.
+\end{itemize}
+
+Due to space limitations, many standard or straightforward technical details of the calculi defined in Sec.
+\ref{sec:setsms-formally} through Sec.
+\ref{sec:static-eval} are relegated to the supplemental material.
+For convenience, the examples that we give are written in a full-scale functional language under development called VerseML.
+% For the present purposes, we can consider VerseML as essentially the calculus defined in the supplement extended with various conveniences that are commonly found in other ML-like languages and, notionally, orthogonal to TLMs.
+%VerseML is, as its name suggests, a conceptual descendent of ML.
+It diverges from other dialects of ML that have a similar type structure in that it has a bidirectional type system \cite{Pierce:2000:LTI:345099.345100} (like, for example, Scala \cite{OdeZenZen01}) for reasons that have to do with the mechanism of TLM implicits described in Chapters \ref{chap:tsls} and \ref{chap:ptsms}.
+%The reason we will not follow Standard ML \cite{mthm97-for-dart} in giving a complete formal definition of VerseML in this work is both to emphasize that the primitives we introduce are ``insensitive'' to the details of the underlying type structure of the language (so TLMs can be considered for inclusion in a variety of languages, not only dialects of ML), and to avoid distracting the reader (and the author) with definitions that are already well-understood in the literature and that are orthogonal to those that are the focus of this work.
+% We will not formally define these features mainly to avoid unnecessarily complicating our presentation with details that are not essential to the ideas presented herein. As such, 
+All examples written in VerseML should be understood to be informal motivating material for the subsequent formal material.
+%We anticipate that future full-scale language specifications will be able to combine the ideas  in the proposed work without trouble.
+%The purpose of the work being proposed is to serve as a reference for those interested in the new constructs we introduce, not to serve as a language specification.
+%We will give a brief overview of these languages are organized in Sec. \ref{sec:VerseML}.
+
+% \subsection{Contributions}\label{sec:contributions}
+% This work introduces a system of \textbf{typed literal macros (TLMs)} that gives library providers substantially more syntactic control than existing typed term-rewriting macro systems while maintaining the ability to reason abstractly about types, binding and segmentation.% abstract reasoning principles.
 % comparable to the level of control they have when defining a syntax dialect.
@@
 % Clients apply TLMs to \emph{generalized literal forms}.
+
+% Because the context-free syntax is never extended, syntactic conflicts are not a concern.
+
+%As such, the semantics can take the type and binding structure of the surrounding program into account when validating the expansion that the TLM programmatically generates to ensure that clients can answer critical questions related to types and binding, like those enumerated in Section \ref{sec:abs-reasoning-intro}.
+Clients need not have knowledge of the implementation of the TLM or of the generated expansion, i.e.
+there are useful principles of syntactic abstraction.
+
+% The primary technical challenge has to do with the fact that the applied TLM needs to be able to splice terms out of the literal body for inclusion in the expansion. For example,
+
+ % We design our mechanism such that these locations can easily be determined from the output of the TLM.
+This is essential for our hygiene mechanism, and it is also useful in that this information can be presented to the user (e.g.
+as shown in Figure \ref{fig:first-tsm-example-marked}).
+%As such, we must develop a mechanism where 1) the positions of spliced subterms can be determined without examining the macro implementation (e.g.
+so that they can be presented to the user differently by an editor or pretty-printer, ;  and 2) the hygiene mechanism must give only portions of the expansion that correspond to these spliced subterms access to the application site context.
+
+%In order to reason about types and binding, client programmers need only have knowledge of 1) the segmentation (e.g.
+by examining a figure like this presented by a code editor or pretty-printer) and 2) a type annotation on the definition of the applied TLM.
+No other details about the applied TLM's implementation or the expansion that it generates need to be revealed to the client programmer.
+In other words,
+
+% \begin{figure}[h]
+% \begin{lstlisting}
+% PElement Nil Seq(
+%   TextNode "Hello, ", 
+%   Seq(TextNode (join(" ", Cons(first, Cons(second, Nil)))), 
+%   TextNode "!"))
+% \end{lstlisting}
+% \caption{The desugaring.}
+% \end{figure}
+
+% There is also no ambiguity with regard to which TLM has control over each form, and searching for the definition of a TLM is no more difficult than searching for any other binding, i.e.
+there are well-defined scoping rules.
+
+% In other words, TLMs maintain a useful notion of syntactic abstraction.
+%More specifically, TLMs maintain a \emph{hygienic binding discipline}, meaning that questions Questions 4 and 5 above were concerned with are disallowed entirely.
+% We will, of course, make this notion more technically precise as we continue.
+
+% \begin{figure}[h]
+% \begin{lstlisting}[numbers=none,xleftmargin=0px]
+% let syntax $strlist = $list string in 
+% $html `SURL<p>Hello, {[EURLjoin ($str ' ') ($strlist [firstSURL,EURL last])SURL]}</p>EURL`
+% \end{lstlisting}
+% \caption{The example from Figure \ref{fig:first-tsm-example-marked}, expressed using a parametric TLM.}
+% \label{fig:first-ptsm-example-marked}
+% \end{figure}
+
+% In Secs.
+\ref{chap:uetsms} and \ref{chap:uptsms}, we assume for the sake of technical simplicity that each TLM definition is self-contained, needing no access to libraries or to other TLMs.
+This is an impractical assumption in practice.
+We relax this assumption in In
+
+% %\item \textbf{Type-specific languages}, or \textbf{TSLs}. TSLs, described 
+% In Chapter \ref{chap:tsls}, we develop a mechanism of \emph{TLM implicits} that allows library clients to contextually designate, for any type, a privileged TLM at that type.
+The semantics applies this privileged TLM implicitly to unadorned literal forms that appear where a term of the associated type is expected.
+For example, if we designate
+% %\li{$str} 
+% as the privileged TLM at the \li{string} type and 
+% %\li{$strlist}
+% as the privileged TLM at the \li{list(string)} type, we can express the example from Figure \ref{fig:first-tsm-example-marked} instead as shown in Figure \ref{fig:first-tsm-example-implicit} (assuming \li{join} has type \li{string -> list(string) -> string}.)
+% \begin{figure}[h]
+% \begin{lstlisting}[numbers=none]
+% $html`SURL<p>Hello, {[EURLjoin ' ' [firstSURL,EURL last]SURL]}</p>EURL`
+% \end{lstlisting}
+% \caption{The example from Figure \ref{fig:first-tsm-example-marked} drawn to take advantage of TLM implicits.}
+% \label{fig:first-tsm-example-implicit}
+% \end{figure}
+
+% \noindent This approach is competitive in cost with library-specific syntax dialects (e.g.
+compare Figure \ref{fig:first-tsm-example-implicit} to Figure \ref{fig:urweb}), while maintaining the abstract reasoning principles characteristic of our approach.
+%\item \textbf{Metamodules}, introduced in Sec.
+\ref{sec:metamodules}, reduce the need to primitively build in the type structure of constructs like records (and variants thereof),  labeled sums and other interesting constructs that we will introduce later by giving library providers programmatic ``hooks'' directly into the semantics, which are specified as a \emph{type-directed translation semantics} targeting a small \emph{typed internal language} (introduced in Sec.
+\ref{sec:VerseML}).
+%For example, a library provider can implement the type structure of records with a metamodule that:
+%\begin{enumerate}
+%\item introduces a type constructor, \lstinline{record}, parameterized by finite mappings from labels to types, and defines, programmatically, a translation to unary and binary product types (which are built in to the internal language); and
+%\item introduces operators used to work with records, minimally record introduction and elimination (but perhaps also various functional update operators), and directly implements the logic governing their typechecking and translation to the IL (which builds in only nullary and binary products).
+%\end{enumerate}
+%We will see direct analogies between ML-style modules (which our mechanisms also support) and metamodules later.
+%\end{enumerate}
+
+% As vehicles for this work, we will define a small programming language in each of the three parts just mentioned, each building conceptually upon the previous language.
+All of our formal contributions are relative to these small languages.
+
+%TLMs, like other macro systems, perform \emph{static code generation} (also sometimes called \emph{static} or \emph{compile-time metaprogramming}), meaning that the relevant rules in the static semantics of the language call for the evaluation of \emph{static functions} that generate term encodings.
+Static functions are functions that are evaluated statically, i.e.
+during typing.
+%Library providers write these static functions using the VerseML \emph{static language} (SL).
+%Maintaining a separation between the static (or ``compile-time'') phase and the dynamic (or ``run-time'') phase is an important facet of VerseML's design.
+% static code generation.
+%We will  also introduce a simple variant of each of these primitives that leverages VerseML's support for local type inference to further reduce syntactic cost in certain common situations.
+
+\newcommand{\seTLMsSec}{Simple Expression TLMs (seTLMs)}
+\section{\protect\seTLMsSec}
+\label{sec:setsms}
+
+We will begin in this section by introducing \emph{simple expression TLMs} (seTLMs), i.e. TLMs where expansions are expressions of a single specified type.
+
 Our take on the running example is shown in Fig. \ref{fig:first-tsm-example}: we apply a TLM named 
@@
 For example, if strings were not primitive, we might write \li{`SURL<body>\{EURLheading ($str "SSTRWorld!ESTR")SURL\} ...</body>EURL`}.}, can always be  generated.
-
-% Because the context-free syntax is never extended, syntactic conflicts are not a concern.
-
-%As such, the semantics can take the type and binding structure of the surrounding program into account when validating the expansion that the TLM programmatically generates to ensure that clients can answer critical questions related to types and binding, like those enumerated in Section \ref{sec:abs-reasoning-intro}.
-Clients need not have knowledge of the implementation of the TLM or of the generated expansion, i.e.
-there are useful principles of syntactic abstraction.
-
-% The primary technical challenge has to do with the fact that the applied TLM needs to be able to splice terms out of the literal body for inclusion in the expansion. For example,
-
- % We design our mechanism such that these locations can easily be determined from the output of the TLM.
-This is essential for our hygiene mechanism, and it is also useful in that this information can be presented to the user (e.g.
-as shown in Figure \ref{fig:first-tsm-example-marked}).
-%As such, we must develop a mechanism where 1) the positions of spliced subterms can be determined without examining the macro implementation (e.g.
-so that they can be presented to the user differently by an editor or pretty-printer, ;  and 2) the hygiene mechanism must give only portions of the expansion that correspond to these spliced subterms access to the application site context.
-
-%In order to reason about types and binding, client programmers need only have knowledge of 1) the segmentation (e.g.
-by examining a figure like this presented by a code editor or pretty-printer) and 2) a type annotation on the definition of the applied TLM.
-No other details about the applied TLM's implementation or the expansion that it generates need to be revealed to the client programmer.
-In other words,
-
-% \begin{figure}[h]
-% \begin{lstlisting}
-% PElement Nil Seq(
-%   TextNode "Hello, ", 
-%   Seq(TextNode (join(" ", Cons(first, Cons(second, Nil)))), 
-%   TextNode "!"))
-% \end{lstlisting}
-% \caption{The desugaring.}
-% \end{figure}
-
-% There is also no ambiguity with regard to which TLM has control over each form, and searching for the definition of a TLM is no more difficult than searching for any other binding, i.e.
-there are well-defined scoping rules.
-
-% In other words, TLMs maintain a useful notion of syntactic abstraction.
-%More specifically, TLMs maintain a \emph{hygienic binding discipline}, meaning that questions Questions 4 and 5 above were concerned with are disallowed entirely.
-% We will, of course, make this notion more technically precise as we continue.
-
-The remainder of this document is organized as follows.
-\begin{itemize}
-\item Sec. \ref{sec:setsms} begins by going through the mechanics of defining a {simple expression TLM} and describes the process of typed expansion and expansion validation by example.
-\item Sec. \ref{sec:sptsms} does the same for a {simple pattern TLM}, i.e. a TLM that generates patterns that match values of a specified type. The hygiene condition for pattern TLMs is interesting.
-\item Sec. \ref{sec:setsms-formally} defines a calculus of simple expression and pattern TLMs and formally establishes the reasoning principles implied above.
-
-% In the remaining chapters, we enrich the language developed in Chapter \ref{chap:uetsms} with advanced features based on those found in full-scale general-purpose languages like ML and Scala, and enhance our TLM mechanism with these new features.
-
-% We next develop a more sophisticated account of TLMs in Chapters \ref{chap:ptsms} and \ref{chap:static-eval}, with the aim of working out details necessary to integrate TLMs into full-scale general-purpose languages like ML or Scala.
-%These two chapters constitute Part \ref{part:parametric-tsms} of our contributions.
-\item Sec.
-\ref{sec:ptsms} adds parameterized types and an ML-style module system and introduces \emph{parametric TLMs} (pTLMs), i.e.
-TLMs that take type and module parameters.
-Parameters serve two purposes:
-\begin{enumerate}
-\item They allow for TLMs that operate not just at a single type, but over a type- and module-parameterized family of types.
-For example, we consider a parametric TLM for dictionaries that operates uniformly over different implementations of the dictionary signature, and also over different choices of the co-domain type.
-\item Parameters also give expansions access to terms other than those explicitly spliced in to the literal body.
-This is particularly convenient for working with abstract types \emph{a la} ML.
-Partial parameter application in TLM abbreviations lowers the cost of this explicit parameter-passing discipline.
-\end{enumerate}
-This section culminates with a calculus that extends the calculus of Sec. \ref{sec:setsms-formally} with ML-style modules and parametric TLMs.
-\item Sec.
-\ref{sec:static-eval} considers the topic of term evaluation during the typed expansion phase, i.e.
-\emph{static evaluation}, in  more detail, and also gives examples of TLMs that are useful for defining other TLMs, e.g.
-TLMs that implement parser generators and quasiquotation.
-\item Sec.
-\ref{sec:related-work} discusses additional related work, including closely related work by \citet{TSLs,sac15} on type-specific languages.
-We conclude with a brief discussion of present limitations and future directions.
-\end{itemize}
-
-Due to space limitations, many of the standard or straightforward technical details of the calculi defined in Sec.
-\ref{sec:setsms-formally} through Sec.
-\ref{sec:static-eval} are relegated to the supplemental material.
-For convenience, the examples that we give are written in a full-scale functional language under development called VerseML.
-% For the present purposes, we can consider VerseML as essentially the calculus defined in the supplement extended with various conveniences that are commonly found in other ML-like languages and, notionally, orthogonal to TLMs.
-%VerseML is, as its name suggests, a conceptual descendent of ML.
-It diverges from other dialects of ML that have a similar type structure in that it has a bidirectional type system \cite{Pierce:2000:LTI:345099.345100} (like, for example, Scala \cite{OdeZenZen01}) for reasons that have to do with the mechanism of TLM implicits described in Chapters \ref{chap:tsls} and \ref{chap:ptsms}.
-%The reason we will not follow Standard ML \cite{mthm97-for-dart} in giving a complete formal definition of VerseML in this work is both to emphasize that the primitives we introduce are ``insensitive'' to the details of the underlying type structure of the language (so TLMs can be considered for inclusion in a variety of languages, not only dialects of ML), and to avoid distracting the reader (and the author) with definitions that are already well-understood in the literature and that are orthogonal to those that are the focus of this work.
-% We will not formally define these features mainly to avoid unnecessarily complicating our presentation with details that are not essential to the ideas presented herein. As such, 
-All examples written in VerseML should be understood to be informal motivating material for the subsequent formal material.
-%We anticipate that future full-scale language specifications will be able to combine the ideas  in the proposed work without trouble.
-%The purpose of the work being proposed is to serve as a reference for those interested in the new constructs we introduce, not to serve as a language specification.
-%We will give a brief overview of these languages are organized in Sec. \ref{sec:VerseML}.
-
-% \begin{figure}[h]
-% \begin{lstlisting}[numbers=none,xleftmargin=0px]
-% let syntax $strlist = $list string in 
-% $html `SURL<p>Hello, {[EURLjoin ($str ' ') ($strlist [firstSURL,EURL last])SURL]}</p>EURL`
-% \end{lstlisting}
-% \caption{The example from Figure \ref{fig:first-tsm-example-marked}, expressed using a parametric TLM.}
-% \label{fig:first-ptsm-example-marked}
-% \end{figure}
-
-% In Secs.
-\ref{chap:uetsms} and \ref{chap:uptsms}, we assume for the sake of technical simplicity that each TLM definition is self-contained, needing no access to libraries or to other TLMs.
-This is an impractical assumption in practice.
-We relax this assumption in In
-
-% %\item \textbf{Type-specific languages}, or \textbf{TSLs}. TSLs, described 
-% In Chapter \ref{chap:tsls}, we develop a mechanism of \emph{TLM implicits} that allows library clients to contextually designate, for any type, a privileged TLM at that type.
-The semantics applies this privileged TLM implicitly to unadorned literal forms that appear where a term of the associated type is expected.
-For example, if we designate
-% %\li{$str} 
-% as the privileged TLM at the \li{string} type and 
-% %\li{$strlist}
-% as the privileged TLM at the \li{list(string)} type, we can express the example from Figure \ref{fig:first-tsm-example-marked} instead as shown in Figure \ref{fig:first-tsm-example-implicit} (assuming \li{join} has type \li{string -> list(string) -> string}.)
-% \begin{figure}[h]
-% \begin{lstlisting}[numbers=none]
-% $html`SURL<p>Hello, {[EURLjoin ' ' [firstSURL,EURL last]SURL]}</p>EURL`
-% \end{lstlisting}
-% \caption{The example from Figure \ref{fig:first-tsm-example-marked} drawn to take advantage of TLM implicits.}
-% \label{fig:first-tsm-example-implicit}
-% \end{figure}
-
-% \noindent This approach is competitive in cost with library-specific syntax dialects (e.g.
-compare Figure \ref{fig:first-tsm-example-implicit} to Figure \ref{fig:urweb}), while maintaining the abstract reasoning principles characteristic of our approach.
-%\item \textbf{Metamodules}, introduced in Sec.
-\ref{sec:metamodules}, reduce the need to primitively build in the type structure of constructs like records (and variants thereof),  labeled sums and other interesting constructs that we will introduce later by giving library providers programmatic ``hooks'' directly into the semantics, which are specified as a \emph{type-directed translation semantics} targeting a small \emph{typed internal language} (introduced in Sec.
-\ref{sec:VerseML}).
-%For example, a library provider can implement the type structure of records with a metamodule that:
-%\begin{enumerate}
-%\item introduces a type constructor, \lstinline{record}, parameterized by finite mappings from labels to types, and defines, programmatically, a translation to unary and binary product types (which are built in to the internal language); and
-%\item introduces operators used to work with records, minimally record introduction and elimination (but perhaps also various functional update operators), and directly implements the logic governing their typechecking and translation to the IL (which builds in only nullary and binary products).
-%\end{enumerate}
-%We will see direct analogies between ML-style modules (which our mechanisms also support) and metamodules later.
-%\end{enumerate}
-
-% As vehicles for this work, we will define a small programming language in each of the three parts just mentioned, each building conceptually upon the previous language.
-All of our formal contributions are relative to these small languages.
-
-%TLMs, like other macro systems, perform \emph{static code generation} (also sometimes called \emph{static} or \emph{compile-time metaprogramming}), meaning that the relevant rules in the static semantics of the language call for the evaluation of \emph{static functions} that generate term encodings.
-Static functions are functions that are evaluated statically, i.e.
-during typing.
-%Library providers write these static functions using the VerseML \emph{static language} (SL).
-%Maintaining a separation between the static (or ``compile-time'') phase and the dynamic (or ``run-time'') phase is an important facet of VerseML's design.
-% static code generation.
-%We will  also introduce a simple variant of each of these primitives that leverages VerseML's support for local type inference to further reduce syntactic cost in certain common situations.
-
-\newcommand{\seTLMsSec}{Simple Expression TLMs (seTLMs)}
-\section{\protect\seTLMsSec}
-\label{sec:setsms}
-
-We will begin in this section by introducing \emph{simple expression TLMs} (seTLMs), i.e. TLMs where expansions are expressions of a single specified type.
 
@@
 \subsection{Another Existing Approach: Term Rewriting}
+For example, OCaml's textual syntax now includes \emph{preprocessor extension (ppx) points} used to identify terms that some term rewriting preprocessor must rewrite \cite{ocaml-manual}.
+We could attempt to mark a string literal containing Ur/Web-style HTML syntax with a ppx annotation, \li{xml}, as follows:
+\begin{lstlisting}[numbers=none]
+  [%xml "SSTR<h1>Hello, {[first_name]}!</h1>ESTR"]
+\end{lstlisting}
+Again, there are problems reasoning modularly and abstractly about such constructions.
+More than one applied preprocessor might recognize this annotation (there are, in practice, many XML/HTML libraries), so the problems of \textbf{Responsibility} come up.
+It is also impossible to reason abstractly about \textbf{Segmentation}, \textbf{Capture}, \textbf{Context Dependence} and \textbf{Typing} because the code that the preprocessor generates is arbitrary.
+For these reasons, users of this mechanism warn that they should be used sparingly.\todo{citation}
+
+Term-rewriting macro systems are language-integrated local term rewriting systems that require that the client explicitly apply the intended rewriting, implemented by a macro, to the term that is to be rewritten\todo{readeer macros}.
+This addresses the issue of \textbf{Responsibility}.
+However, na\"ive macro systems, like the earliest variants of the Lisp macro system \cite{Hart63a}, Template Haskell \cite{SheardPeytonJones:Haskell-02} and GHC's quasiquotation system \cite{mainland2007s}, do not allow clients to reason abstractly about the remaining issues, again because the expansion that they produce is arbitrary.
+
+\emph{Hygienic} macro systems prevent, or abstractly account for \cite{DBLP:conf/esop/HermanW08,Herman10:Theory}, \textbf{Capture}, and they enforce application-site \textbf{Context Independence} \cite{Kohlbecker86a,DBLP:conf/popl/Adams15,DBLP:conf/popl/ClingerR91,DBLP:journals/lisp/DybvigHB92}.
+However, this turns out to make it difficult to use hygienic macros to introduce new literal forms of arbitrary design with support for splicing.
+Consider again our running XHTML example, which we might try to realize by applying a hygienic macro, \li{xml!}, to a string literal:
+\begin{lstlisting}[numbers=none]
+  (xml! "SSTR<h1>Hello, {[first_name]}!</h1>ESTR")
+\end{lstlisting}
+THIS IS BAD BECAUSE FIRSTNAME IS BAD HYGIENE GRAH. MAYBE TALK ABOUT STRING SPLICING AND FRAGMENT SYSTEM STUFF HERE.
+
+PUTTING THAT ASIDE, LET US CONSIDER TYPING.
+LISP IS DUMB NO TYPES.
+HERMAN AND WAND USED TYPES ONLY TO ACCOUNT FOR BINDING RELATIONSHIPS.
+SCALA HAS A TYPED MACRO SYSTEM, BUT THERE IS NO FORMAL ACCOUNT OF IT.
+MACROML IS ONLY A STAGING MACRO SYSTEM, NO WAY TO PARSE LITERALS.
+(It is not enough that with Template Haskell / GHC quasiquotation, the generated expansion is typechecked -- to satisfy the \textbf{Typing} criterion, it must be possible to reason abstractly about \emph{what the type of the generated expansion is}.) MAYBE BRING UP TSLS AND TSMS HERE.
+
+TRANSITION TO THE CONTRIBUTIONS OF THIS PAPER: TYPED LITERAL MACROS.
+REWRITE GENERALIZED LITERAL FORMS, KEEP TRACK OF SPLICED SEGMENTS SO THE HYGIENE MECHANISM IS COMPATIBLE, SEGMENTATION IS ALWAYS POSSIBLE.
+TYPED, INTEGRATED INTO ML SCALA-LIKE LANGUAGE (SECTION HEADINGS).
+FORMAL DEFINITION.
+REASONING PRINCIPLES ARE THEOREMS.
+THATS HOW WE EVALUATE OUR WORK.
+(USEFULNESS / USABILITY CONCERNS NEED TO BE ADDRESSED SOMEWHERES)
+
+Although much of the research on macro systems has been for languages in the Lisp tradition \cite{mccarthy1978history}, some progress has also been made on reasoning about types with the design of \emph{typed macro systems} where annotations constrain the macro arguments and the generated expansions.
+Examples include Scala's macro system \cite{ScalaMacros2013}, formal calculi studied by \citet{DBLP:conf/esop/HermanW08} and \emph{staging macro systems} like MetaML \cite{Sheard:1999:UMS} and MacroML \cite{ganz2001macros} (which, it is worth noting, do not give the macro access to the syntax tree of arguments.)
+
+Perhaps the biggest problem, then, is that term-rewriting systems offer library providers only limited syntactic control.
+In other words many syntactic forms of interest will not particularly resemble those available in the host language.
+For example, consider the  XHTML and K examples above.
+In both cases, the syntactic conventions are quite distinct from those of ML-like languages (and, for that matter, languages that use S-expression.)
+
+It is tempting in these situations to consider repurposing string literal forms.
+For example, we might wish to apply a macro \li{html!} (following Rust's convention of using a post-fix \li{!} to distinguish macro names from variables) to rewrite string literals containing Ur/Web-style XHTML syntax as follows:
+\begin{lstlisting}[numbers=none]
+  fun heading first_name = html! 'SSTR<h1>Hello, {[first_name]}!</p>ESTR'
+  val body = html! 'SSTR<body>{heading "World"} ...</body>ESTR'
+\end{lstlisting}
+
+The problem here is that there is no way to extract the spliced expressions from the supplied string literal forms while satisfying the hygiene conditions, because variables that come from these spliced terms (here, \li{first_name} and \li{heading}) are indistinguishable from variables that inappropriately appear free relative to the expansion.
+In addition, the problem of segmentation becomes even more pernicious: to a human or tool unaware of Ur/Web's syntax, it is not immediately apparent which particular subsequences of the string literals supplied to \li{html!} are segmented out as spliced expressions.
+Reader macros have essentially the same problem  \cite{DBLP:journals/jfp/FlattCDF12}.
+
+  Some languages build in \emph{string splicing} (a.k.a.
+\emph{string interpolation}) forms, e.g.
+Scala \cite{odersky2008programming}, or similar but more general \emph{fragmentary quotation forms} \cite{conf/icfp/Slind91}, e.g.
+the SML/NJ dialect of ML.
+These designate particular delimiters to allow escape out into the expression language.
+The problem with using these (dynamically, or in conjunction with term rewriting macros) as vehicles to introduce syntactic sugar at various types is 1) there is no ``one-size-fits-all'' escape delimiter, and 2) reasoning about typing is problematic because every escaped term is checked against the same type.
+In the example above, we spliced in two different types of expressions using two different delimiters.
+In general, e.g.
+when defining syntax for a programming language with many sorts of terms, the most appropriate choice of delimiters might depend on where each spliced term appears.
+
+{This paper introducing a system of \emph{typed literal macros (TLMs)} that allows programmers to define new literal syntax of nearly arbitrary form within an ML- or Scala-like  language}, i.e.
+a language with a rich static type system (Sec.
+2), support for pattern matching (Sec.
+3) and an ML-style module system or Scala-style object system with support for type members and type abstraction (Sec.
+5).
+The examples in this paper are written in VerseML, a new dialect of ML that features TLMs.
+
+Compared to existing approaches, TLMs are more \emph{reasonable}.
+We justify this claim by (1) developing a collection of relevant reasoning principles based on the reasoning principles already available to client programmers when using an ML-like or Scala-like language without such a mechanism; (2) formally operationalizing these reasoning principles as metatheorems about a series of type-theoretic \emph{TLM calculi} of increasing sophistication (Sec.
+4-5); and (3) evaluating existing approaches -- including syntax definition systems like Camlp4 \cite{ocaml-manual}\todo{better citation}, SugarJ/Sugar* \cite{erdweg2011sugarj,erdweg2013framework} and Copper \cite{conf/pldi/SchwerdfegerW09,schwerdfeger2010context}, term rewriting macro systems as found in a variety of languages, and the system of type-specific languages (TSLs) and typed syntax macros (TSMs) described by \citet{TSLs,sac15} -- relative to these reasoning principles (Sec.
+6).
+We conclude by discussing limitations and future research directions in Sec.
+7\todo{sec refs}.
+A complete set of definitions and associated proofs are available in the  supplemental material.
+This is the first detailed type-theoretic accout of a typed, hygienic term-rewriting macro system for an ML- or Scala-like language.\todo{refine and justify this claim}
+
+% These existing forms normally have other meanings, so this can be confusing \cite{pane1996usability}.
 

### 2017-06-06 popl18/syntax-popl18.tex
switch to acmsmall template, collapsing abstract, reorganized intro

@@
 %% For double-blind review submission
-\documentclass[acmlarge,review,anonymous]{acmart}\settopmatter{printfolios=true}
+\documentclass[acmsmall,10pt,review,anonymous]{acmart}\settopmatter{printfolios=true}
 %% For single-blind review submission
@@
 This paper questions this common practice: why should an otherwise ordinary data structure be privileged in the definition of a general-purpose programming language?
-When designing the surface syntax of a programming language, it is common practice to build in literal forms for a select few standard data structures, e.g.
+When designing the surface syntax of a programming language, it is common practice to build in literal forms for a few standard data structures, e.g.
 lists.
-The problem is that this practice neglects an unbounded assortment of other data structures for which literal forms might also be useful (consider, for example, sets, maps, matrices, regular expressions, HTML trees, URLs, SQL queries and chemical structures.) A less \emph{ad hoc} design would be one that allows library providers to define new literal forms in a decentralized manner.
+The problem is that this practice neglects an unbounded variety of other data structures for which literal forms might also be useful (consider, for example,  maps, regexes, HTML trees, URLs, SQL queries and chemical structures.) %A less \emph{ad hoc} design would be one that allows library providers to define new literal forms in a decentralized manner.
 However, the mechanisms available to programmers today, e.g.
@@
 In particular, they do not support modular reasoning about syntactic determinism, so separately defined literal forms can and do conflict syntactically with one another.
-Moreover, clients cannot reason abstractly about types and binding when they encounter an unfamiliar literal form (like they can when they encounter an unfamiliar function being applied.) Instead, they must reason transparently, i.e.
+Moreover, clients cannot reason abstractly about types and binding when they encounter an unfamiliar literal form (like they can when they encounter an unfamiliar function being applied.)% Instead, they must reason transparently, i.e.
 about the underlying expansion.
@@
 the Scala macro system \cite{ScalaMacros2013}, are only somewhat more reasonable, and do not offer direct syntactic control.
-
-This paper formally details a more reasonable alternative: \emph{typed literal macros} (TLMs).
-TLMs give library providers full programmatic control over both the parsing and expansion of expressions and patterns of \emph{generalized literal form} \cite{TSLs} at a specified type or parameterized family of types.
-Expansion is {strictly hygienic}, meaning that 1) expansions cannot make any assumptions about the application site context; and 2)  expansions do not reveal internal bindings to spliced terms or to the remainder of the program.
+This paper introduces a system of \emph{typed literal macros (TLMs)} that allows library providers to define new literal forms in a decentralized manner.
+More specifically, TLMs give library providers full programmatic control over both the parsing and expansion of expressions and patterns of \emph{generalized literal form} \cite{TSLs} at a specified type or parameterized family of types.
+% The mechanism is {strictly hygienic}, meaning that 1) the expansion must be context-independent; and 2)  splicing is capture avoiding.
 % Partial parameter application lowers the syntactic cost of this strict style. 
-This design ensures that clients are able to reason modularly about syntactic determinism and abstractly about the type and binding structure of the program.
+Compared to existing approaches, including syntax extension systems and macro-based quotation rewriting systems, TLMs are uniquely \emph{reasonable}: TLM providers can reason modularly about syntactic determinism, and TLM clients can reason abstractly about the type and binding structure of the program.
 The system needs only to convey to clients (e.g.
-via secondary notation) the \emph{segmentation} of each literal, which gives the locations of the spliced segments.
+via secondary notation) the inferred \emph{segmentation} of each literal, which gives the locations of the spliced terms.
+The inferred segmentation also informs the hygiene mechanism (previously, only unhygienic macros could freely splice terms out of literal bodies.)
 %The system can determine this {segmentation} automatically and convey it to the client via syntax highlighting, or by presenting the client with a context-free grammar (without revealing the semantic actions associated with each production.)
@@
 We establish these abstract reasoning principles formally with a calculus of expressions, patterns and ML-style modules.
-This is the first detailed type-theoretic account of a typed, hygienic term-rewriting macro system for an ML-like\todo{or Scala-like?} language.
+This is the  first detailed type-theoretic account of a typed, hygienic term-rewriting macro system for an ML-like language.
 % This calculus is the first such formalization of a typed macro system.
@@
 
-\vspace{-4px}
+% \vspace{-4px}
 \newcommand{\introSec}{Introduction}
@@
 \label{sec:intro}
-The surface syntax of a programming language serves as the interface between human programmers and the language's semantics.
-As such, language designers often include various \emph{literal forms} in the surface syntax that are designed to decrease the cognitive cost of constructing and pattern matching over values of certain types defined in the standard library.
-For example, in Standard ML (SML), the list expression literal \li{[x, y, z]} desugars to a term  semantically equivalent to \li{Cons(x, Cons(y, Cons(z, Nil)))} where \li{Nil} and \li{Cons} identify the constructors of the standard \li{list} datatype \cite{mthm97-for-dart,harper1997programming}.\footnote{In SML, list literals work even in contexts where the list expression constructors have been shadowed by other bindings, i.e.
+The surface syntax of a programming language serves as the interface between human programmers and a semantics.
+As such, language designers often include in the surface syntax various \emph{literal forms}  that are designed to decrease the cognitive cost of constructing and pattern matching over values of certain otherwise ordinary types from the standard library.
+For example, in Standard ML (SML), the list expression literal \li{[x, y, z]} desugars to a term equivalent to \li{Cons(x, Cons(y, Cons(z, Nil)))} where \li{Nil} and \li{Cons} identify the constructors of the standard \li{list} datatype \cite{mthm97-for-dart,harper1997programming}.\footnote{In SML, list literals work even in contexts where the list expression constructors have been shadowed by other bindings, i.e.
 the expansion of literal forms is \emph{context independent}.
-We will return to the concept of context independence throughout this work.} List pattern literals are analagous.
+We will return to the concept of context independence again.} List pattern literals are analagous.
 Another example is Ur/Web, which extends Ur's textual syntax with expression and pattern literals for XHTML trees, as defined in Ur/Web's standard library \cite{conf/popl/Chlipala15,conf/pldi/Chlipala10}.
-Figure \ref{fig:urweb} shows examples of XHTML literals containing ``spliced'' string expressions (Line 1) and XHTML expressions (Line 2).
+Fig.
+\ref{fig:urweb} shows examples of XHTML literals containing ``spliced'' string expressions (Line 1) and XHTML expressions (Line 2).
 %The desugarings of these literal expressions, not shown, are substantially more verbose and less readable to programmers familiar with XHTML.
 
-Third-party library providers are quite justifiably envious of the special status of these few otherwise ordinary user-defined types.
+Third-party library providers are quite justifiably envious of the special status of these few otherwise ordinary standard library constructs.
 After all, it is not difficult to find other examples of types for which mathematicians, scientists and programmers have invented specialized notations.
-For example, 1) clients of a ``collections'' library might like not just list literals, but also vector, matrix, finite set and finite map literals following standard mathematical conventions; 2) clients of a ``web programming'' library might like not just XHTML literals but also CSS literals (which Ur/Web does not provide); 4) a compiler writer might like ``quotation'' literals for the terms of the object language and various intermediate languages of interest; and 5) clients of a ``chemistry'' library might like chemical structure literals based on the SMILES standard \cite{anderson1987smiles}.
+For example, 1) clients of a ``collections'' library might like not just list literals, but also matrix, finite set and finite map literals following standard mathematical conventions; 2) clients of a ``web programming'' library might like not just XHTML literals but also CSS literals (which Ur/Web does not provide); 4) a compiler writer might like ``quotation'' literals for the terms of the object language and various intermediate languages of interest; and 5) clients of a ``chemistry'' library might like chemical structure literals based on the SMILES standard \cite{anderson1987smiles}.
 \citet{Bravenboer:2007:PIA:1289971.1289975} suggests that literal forms for structured encodings of SQL queries and other query types can reduce the temptation to use string encodings and therefore reduce the prevalence of string injection attacks, which are amongst the most common and serious security vulnerabilities on the web today \cite{owasp2017}.
 Choosing strings over composite data structures, a.k.a.
-``stringly-typed programming'', is quite common in practice -- \citet{TSLs} sampled strings from open source projects and found that at least 15\% of them were parseable according to some readily apparent type-specific grammar, e.g.
+``stringly-typed programming'', is rather common in practice -- \citet{TSLs} sampled strings from open source projects and found that at least 15\% of them were parseable according to some readily apparent type-specific grammar, e.g.
 for SQL queries, regexes, file system paths, URLs and others.
-Literal forms at these types, with support for splicing, would lower the cost of using more compositional encodings.
-%Literal forms would  preferable to string encodings because they support compositional construction and decomposition by pattern matching via splicing.
+Literal forms at these types, with support for splicing, would lower the cost of compositional encodings.%Literal forms would  preferable to string encodings because they support compositional construction and decomposition by pattern matching via splicing.
 
@@
 \vspace{-10px}
-\caption{XHTML literals with support for splicing at two different types are built primitively into Ur/Web \cite{conf/popl/Chlipala15}.}
+\caption{XHTML literals with support for splicing at two different types are built primitively into Ur/Web.}
 \vspace{-7px}
@@
 Language designers cannot, of course, hope to accomodate all such types with primitive literal forms.
-Instead, there has been considerable interest in mechanisms that allow programmers to introduce new literal forms in a decentralized manner.
-We will cover a number of existing mechanisms, including Camlp4 \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, in Sec.
+Instead, there is considerable interest in mechanisms that allow programmers to define new literal forms in a decentralized manner.
+For example, we will consider a number of syntax extension mechanisms, including Camlp4 \cite{ocaml-manual} and SugarJ/Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, in Sec.
 \ref{sec:existing-approaches}.
-The problem is that these existing mechanisms quickly become \emph{unreasonable} when ``programming in the large'' \cite{DeRemer76}.
+The problem is that these mechanisms quickly become \emph{unreasonable} when ``programming in the large'' \cite{DeRemer76}.
 %, i.e. when developing large programs ``consisting of many small programs (modules), possibly written by different people''
 For example, consider the perspective of a programmer attempting to comprehend (i.e.
-reason about) the program text in Figure \ref{fig:K-dialect}, which is written in a dialect of ML's (or Scala's) syntax where a large number of syntax extensions defining user-defined literal forms, some unfamiliar to the programmer, have been imported.
+reason about) the program text in Fig.
+\ref{fig:K-dialect}, which is written in a dialect of ML's (or Scala's) syntax where a variety of syntax extensions defining user-defined literal forms, some unfamiliar to the programmer, are active.
 
@@
 \begin{lstlisting}
-  val w = compute_w ()
-  val x = compute_x ()
+  val w = compute_w()
+  val x = compute_x(w)
   val y = {|(!R)@&{&/x!/:2_!x}'!R}|}
@@
 If the programmer happens to be familiar with the intentionally terse syntax of the stack-based database query language K \cite{Whitney:2001:LOR:376284.375783}, then the literal form on Line 3 might be convenient.
-If the programmer does not recognize this syntax, however, there are no simple, definitive protocols for answering questions like:
+If the programmer does not recognize this rather obscure syntax, however, there are no simple, definitive protocols for answering questions like:
 
 \begin{enumerate}
-\item \textbf{(Responsibility)} Which syntax extension determines the desugaring of the literal form on Line 3?
-Might there be many possible desugarings (i.e.
-syntactic non-determinism) due to conflicts between extensions?
+\item \textbf{(Responsibility)} Which syntax extension determines the desugaring of the literal form on Line 3? Might there be conflicting desugarings (i.e. syntactic non-determinism)?
 \item \textbf{(Segmentation)} Are the characters \li{x} and \li{R} on Line 3 taken as spliced expressions (meaning that they appear directly in the desugaring), or are they parsed in some other way peculiar to this literal form?
-\item \textbf{(Capture)} If \li{x} is in fact a spliced expression, does it refer to the binding of \li{x} on Line 2, or might it capture an unseen binding introduced in the desugaring of Line 3?
+\item \textbf{(Capture)} If \li{x} is in fact a spliced expression, does it refer to the binding of \li{x} on Line 2, or might it capture an unseen binding of the same identifier used in the desugaring of Line 3?
 \item \textbf{(Context Dependence)} If \li{w}, on Line 1, is renamed, could that possibly break Line 3, or change its meaning?
@@
 
-In short, the problem is that library providers cannot reason modularly about syntactic determinism (because other extensions might define overlapping forms), and if the desugaring of the program text is held abstract, programmers can no longer reason about types, binding and segmentation (i.e.
+% In short, the problem is that library providers cannot reason modularly about syntactic determinism (because other extensions might define overlapping forms), and if the desugaring of the program text is held abstract, programmers can no longer reason about types, binding and segmentation (i.e.
 answer questions like those above.) This is burdensome at all scales, but particularly when programming in the large, where it is common to encounter a wide variety of libraries \cite{DBLP:conf/sac/LammelPS11}.
-Forcing the programmer to examine the desugaring of the program text, or the logic that produced that desugaring, in order to answer these sorts of basic questions about segmentation, binding and typing can defeat the ultimate purpose of syntactic sugar -- decreasing cognitive cost \cite{Green89}.
+Forcing the programmer to directly examine the desugaring of the program text, or the logic that produced that desugaring, in order to answer these sorts of basic questions  can defeat the purpose of syntactic sugar -- decreasing cognitive cost \cite{Green89}.
 Analagous problems do not arise when programming without syntax extensions in languages like ML and Scala -- the binding discipline of the language ensures that programmers can reason lexically about binding and renaming, and types mediate abstraction over function and module implementations.
@@
 
-This paper introduces a more reasonable approach: typed literal macros (TLMs).
-TLMs give library providers full programmatic control over the parsing and expansion of \emph{generalized literal forms}, which syntactically subsume other literal forms because their bodies are left initially unparsed according to the fixed context free syntax of the language (i.e.
+This paper details a more reasonable approach: typed literal macros (TLMs).
+TLMs give library providers full programmatic control over the parsing and expansion of \emph{generalized literal forms}, which syntactically subsume other literal forms because their bodies are left initially unparsed according to the fixed context-free syntax of the language (i.e.
 they are initially parsed like raw string literals.) The applied TLM parses the literal body and generates an expansion during a phase known as typed expansion, which generalizes the usual typing phase.
 For example, Fig.
-\ref{fig:K-tsm-example} applies a TLM to express the ``unreasonable'' example from Fig.
+\ref{fig:K-tsm-example} applies a TLM, \li{$kquery}, to a backtick-delimited generalized literal form to express the ``unreasonable'' example from Fig.
 \ref{fig:K-dialect}.
-
 TLMs come equipped with uniquely powerful abstract reasoning principles.
-We will more precisely characterize these abstract reasoning principles as we proceed.
-For now, let us develop some informal intuitions -- the client programmer can reason about Line 3 as follows:
+We will characterize these reasoning principles absolutely precisely as we proceed, but for now, let us develop some informal intuitions -- the client programmer can reason about Line 3 as follows:
 \begin{enumerate}
-\item \textbf{(Responsibility)} The applied TLM, \li{$kquery}, is solely responsible for typed expansion of the literal body.
+\item \textbf{(Responsibility)} The applied TLM, \li{$kquery}, is solely responsible for parsing and expansion of the literal body.
 The context-free syntax of the language is fixed, so the TLM provider can modularly establish that the syntax the TLM implements is deterministic.
-\item \textbf{(Segmentation)} The output that a TLM generates is structured so that the system can produce from it an accurate \emph{segmentation} of the literal body that distinguishes spliced segments, which appear directly in the expansion, from segments that are parsed in some other way by the TLM.
-This segmentation can be communicated by a pretty printer or code editor using secondary notation, e.g.
+\item \textbf{(Segmentation)} The output that a TLM generates is structured so that the system can infer from it an accurate \emph{segmentation} of the literal body that distinguishes spliced segments, which appear directly in the expansion, from segments that are parsed in some other way by the TLM.
+The segmentation can be communicated by a code editor using secondary notation, e.g.
 colors in this document.
-By examining the segmentation of Line 3, we know that the two instances of \li{x} on Line 3 are parsed as spliced expressions, whereas \li{R} is parsed in some other way.
-\item \textbf{(Capture)} The system avoids capture, so the spliced expression \li{x} must refer to the binding of \li{x} on Line 2.
-It cannot capture an unseen binding of \li{x} in the expansion of Line 3.
-\item \textbf{(Context Dependence)} The system enforces context independence, so the expansion of Line 3 cannot  rely on the fact that, for example, \li{w} is in scope.
-The client is therefore free to rename \li{w}.
+By examining the segmentation of Line 3, we know that the two instances of \li{x} on Line 3 are parsed as spliced expressions, whereas \li{R} and the other characters are parsed in some other way.
+\item \textbf{(Capture)} Splicing is capture-avoiding, so the spliced expression \li{x} must refer to the binding of \li{x} on Line 2.
+It cannot capture a coincidental binding of \li{x} in the expansion of Line 3.
+\item \textbf{(Context Dependence)} The system enforces context independence, so the expansion of Line 3 cannot  rely on the fact that, e.g., \li{w} is in scope.
+The client is free to rename \li{w}.
+(As it turns out, enforcing context independence while also allowing macros to parse literal bodies arbitrarily is not straightforward.
+We return to this point in Sec.
+\ref{sec:existing-approaches}.)
 \item \textbf{(Typing)} An explicit type annotation on the definition of \li{$kquery} determines the type that the expansion must have.
@@
 \vspace{-4px}
-\begin{lstlisting}[numbers=none,xleftmargin=0px]
-  val w = compute_w ()
-  val x = compute_x ()
+\begin{lstlisting}
+  val w = compute_w()
+  val x = compute_x(w)
   val y = $kquery `SURL(!R)@&{&/EURLxSURL!/:2_!EURLxSURL}'!R}EURL`
 \end{lstlisting}
-\vspace{-8px}
+\vspace{-10px}
 \caption{TLMs make examples like the one from Fig. \ref{fig:K-dialect} more reasonable.}
-\vspace{-8px}
+\vspace{-12px}
 \label{fig:K-tsm-example}
@@
 \begin{itemize}
-\item Sec.
-\ref{sec:setsms} begins by going through the mechanics of defining a {simple expression TLM} for XHTML and describes initial parsing, typed expansion and expansion validation in more detail.
-\item Sec. \ref{sec:sptsms} does the same for a {simple pattern TLM}, i.e. a TLM that generates patterns. %The hygiene condition for pattern TLMs is interesting.
-\item Sec. \ref{sec:setsms-formally} defines a calculus of simple expression and pattern TLMs and formally establishes the reasoning principles implied above.
+\item Sec. \ref{sec:setsms} begins by defining a {simple expression TLM} for XHTML and describes initial parsing of generalized literal forms, typed expansion and expansion validation in more detail.
+\item Sec. \ref{sec:sptsms} introduces {simple pattern TLMs}, i.e. TLMs that generate patterns. %The hygiene condition for pattern TLMs is interesting.
+\item Sec. \ref{sec:setsms-formally} defines a type-theoretic calculus of simple expression and pattern TLMs and formally establishes the reasoning principles implied above.
 
@@
 \item Sec.
-\ref{sec:ptsms} adds parameterized types and an ML-style module system and introduces \emph{parametric TLMs} (pTLMs), i.e.
+\ref{sec:ptsms} adds type functions and an ML-style module system and introduces \emph{parametric TLMs} (pTLMs), i.e.
 TLMs that take type and module parameters.
@@
 \begin{enumerate}
-\item They allow for TLMs that operate not just at a single type, but over a type- and module-parameterized family of types.
-For example, we consider a parametric TLM for dictionaries that operates uniformly over different implementations of the abstract dictionary signature, and also over different choices of the co-domain type.
+\item They allow a single TLM to operate not just at a single type, but over a type- and module-parameterized family of types.
+For example, we consider a parametric TLM for dictionaries that operates uniformly over different implementations of the abstract dictionary signature, and also over different choices of the range type.
 \item Parameters also give expansions access to terms other than those explicitly spliced in. %This is particularly convenient for working with abstract types \emph{a la} ML. 
-Partial parameter application in TLM abbreviations lowers the cost of this parameter-passing discipline.
+Partial parameter application is possible, which lowers the syntactic cost of the strict context independence discipline that the system enforces.
 \end{enumerate}
-This section culminates with a calculus that extends the calculus of Sec. \ref{sec:setsms-formally} with ML-style modules and parametric TLMs.
+This section culminates with a calculus that extends the simpler calculus of Sec. \ref{sec:setsms-formally} with type functions, ML-style modules and parametric TLMs.
 \item Sec.
@@
 \item Sec.
-\ref{sec:related-work} compares TLMs to the related work, including closely related work by \citet{TSLs,sac15} on type-specific languages.
-\item Sec.
-\ref{sec:discussion} concludes with a summary of our contributions and a brief discussion of present limitations and future research directions.
+\ref{sec:existing-approaches} compares TLMs to the related work, including closely related work by \citet{TSLs,sac15} on type-specific languages, and the long history of work on macro systems.
+\item Sec. \ref{sec:discussion} concludes with a summary of the specific contributions of this paper and a brief discussion of present limitations and future research directions.
 \end{itemize}
@@
 \ref{sec:static-eval} are relegated to the supplemental material.
-For convenience, the examples that we give are written in a full-scale functional language under development called VerseML.
+The examples that we give are written in an emerging dialect of ML called VerseML that supports TLMs as described herein.
+Our design should straightforwardly generalize to other languages with a broadly similar philosophy regarding types and binding (e.g.
+Haskell, Scala and others.) Languages that have a disciplined binding structure but that lack a rich static type structure, e.g.
+Racket and other Lisp derivatives, may also benefit from TLMs -- one can apply the results of this paper by deploying the usual type theorist's trick of casting such a language as a ``{unityped}'' language \cite{pfpl,scott1980lambda}.
 % For the present purposes, we can consider VerseML as essentially the calculus defined in the supplement extended with various conveniences that are commonly found in other ML-like languages and, notionally, orthogonal to TLMs.
@@
 % We will not formally define these features mainly to avoid unnecessarily complicating our presentation with details that are not essential to the ideas presented herein. As such, 
-All examples written in VerseML should be understood to be informal motivating material for the subsequent formal material.
+% All examples written in VerseML should be understood to be informal motivating material for the subsequent formal material.
 %We anticipate that future full-scale language specifications will be able to combine the ideas  in the proposed work without trouble.
@@
 \vspace{-5px}
-\caption[Definitions of various types used by TLM definitions.]{Definitions of various types used by TLM definitions. We assume that these type definitions are available ambiently.}
+\caption[Definitions of various types used by TLM definitions.]{Definitions of various types used by TLM definitions. We assume that these are available ambiently.}
 \label{fig:indexrange-and-parseresult}

### 2017-06-10 popl18/syntax-popl18.tex
another revision pass over sec 1

@@
 \label{sec:intro}
-The surface syntax of a programming language serves as the interface between human programmers and a semantics.
-As such, language designers often include in the surface syntax various \emph{literal forms}  that are designed to decrease the cognitive cost of constructing and pattern matching over values of certain otherwise ordinary types from the standard library.
-For example, in Standard ML (SML), the list expression literal \li{[x, y, z]} desugars to a term equivalent to \li{Cons(x, Cons(y, Cons(z, Nil)))} where \li{Nil} and \li{Cons} identify the constructors of the standard \li{list} datatype \cite{mthm97-for-dart,harper1997programming}.\footnote{In SML, list literals work even in contexts where the list expression constructors have been shadowed by other bindings, i.e.
+The surface syntax of a programming language is a user interface.
+As such, language designers often include in the surface syntax various \emph{literal forms}  that are designed to decrease the cognitive cost of constructing and pattern matching over values of certain otherwise ordinary types from the language's standard library.
+For example, in Standard ML (SML) and its cousins, the list expression literal \li{[x, y, z]} desugars to a term equivalent to \li{Cons(x, Cons(y, Cons(z, Nil)))} where \li{Nil} and \li{Cons} identify the constructors of the standard \li{list} datatype \cite{mthm97-for-dart,harper1997programming}.\footnote{In SML, list literals work even in contexts where the list expression constructors have been shadowed by other bindings, i.e.
 the expansion of literal forms is \emph{context independent}.
 We will return to the concept of context independence again.} List pattern literals are analagous.
-Another example is Ur/Web, which extends Ur's textual syntax with expression and pattern literals for XHTML trees, as defined in Ur/Web's standard library \cite{conf/popl/Chlipala15,conf/pldi/Chlipala10}.
+Another example is Ur/Web, which extends Ur's textual syntax with expression and pattern literals for encodings of XHTML data, as defined in Ur/Web's standard library \cite{conf/popl/Chlipala15,conf/pldi/Chlipala10}.
 Fig.
@@
 
-Third-party library providers are quite justifiably envious of the special status of these few otherwise ordinary standard library constructs.
+Third-party library providers are quite justifiably envious of the special syntactic status of these few semantically ordinary standard library constructs.
 After all, it is not difficult to find other examples of types for which mathematicians, scientists and programmers have invented specialized notations.
@@
 \citet{Bravenboer:2007:PIA:1289971.1289975} suggests that literal forms for structured encodings of SQL queries and other query types can reduce the temptation to use string encodings and therefore reduce the prevalence of string injection attacks, which are amongst the most common and serious security vulnerabilities on the web today \cite{owasp2017}.
-Choosing strings over composite data structures, a.k.a.
+Choosing strings over composite data structures for the sake of convenience, a.k.a.
 ``stringly-typed programming'', is rather common in practice -- \citet{TSLs} sampled strings from open source projects and found that at least 15\% of them were parseable according to some readily apparent type-specific grammar, e.g.
 for SQL queries, regexes, file system paths, URLs and others.
-Literal forms at these types, with support for splicing, would lower the cost of compositional encodings.%Literal forms would  preferable to string encodings because they support compositional construction and decomposition by pattern matching via splicing.
+Literal forms at these types, with support for splicing, would lower the cost of richer compositional encodings.%Literal forms would  preferable to string encodings because they support compositional construction and decomposition by pattern matching via splicing.
 
@@
 
-Language designers cannot, of course, hope to accomodate all such types with primitive literal forms.
-Instead, there is considerable interest in mechanisms that allow programmers to define new literal forms in a decentralized manner.
-For example, we will consider a number of syntax extension mechanisms, including Camlp4 \cite{ocaml-manual} and SugarJ/Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, in Sec.
+Language designers cannot, of course, hope to accomodate all such types with primitive literal forms, so there has been considerable interest in mechanisms that allow programmers to define new literal forms in a decentralized manner.
+For example, we describe a number of syntax extension mechanisms, including Camlp4 \cite{ocaml-manual} and SugarJ/Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, in Sec.
 \ref{sec:existing-approaches}.
@@
 %, i.e. when developing large programs ``consisting of many small programs (modules), possibly written by different people''
-For example, consider the perspective of a programmer attempting to comprehend (i.e.
+Consider the perspective of a programmer attempting to comprehend (i.e.
 reason about) the program text in Fig.
-\ref{fig:K-dialect}, which is written in a dialect of ML's (or Scala's) syntax where a variety of syntax extensions defining user-defined literal forms, some unfamiliar to the programmer, are active.
+\ref{fig:K-dialect}, which is written in a dialect of ML's (or Scala's) syntax where a large number of syntax extensions defining novel literal forms, some unfamiliar to the programmer, are active.
 
@@
 \noindent
-If the programmer happens to be familiar with the intentionally terse syntax of the stack-based database query language K \cite{Whitney:2001:LOR:376284.375783}, then the literal form on Line 3 might be convenient.
-If the programmer does not recognize this rather obscure syntax, however, there are no simple, definitive protocols for answering questions like:
+The literal form on Line 3 constructs an encoding of a query in the stack-based database query language K, using its terse textual syntax \cite{Whitney:2001:LOR:376284.375783}.
+Of course, the programmer reading this code may well be unfamiliar with K's query syntax.
+The problem then is that there are no simple, abstract protocols for answering questions like:
 
 \begin{enumerate}
-\item \textbf{(Responsibility)} Which syntax extension determines the desugaring of the literal form on Line 3? Might there be conflicting desugarings (i.e. syntactic non-determinism)?
-\item \textbf{(Segmentation)} Are the characters \li{x} and \li{R} on Line 3 taken as spliced expressions (meaning that they appear directly in the desugaring), or are they parsed in some other way peculiar to this literal form?
+\item \textbf{(Responsibility)} Which syntax extension determined the desugaring of the literal form on Line 3? Might there be many possible desugarings (i.e. syntactic non-determinism)?
+\item \textbf{(Segmentation)} Are the characters \li{x} and \li{R} on Line 3 taken as spliced expressions (meaning that they appear directly in the desugaring), or are they parsed in some other way peculiar to this literal form (e.g.
+as operators in the K language)?
 \item \textbf{(Capture)} If \li{x} is in fact a spliced expression, does it refer to the binding of \li{x} on Line 2, or might it capture an unseen binding of the same identifier used in the desugaring of Line 3?
-\item \textbf{(Context Dependence)} If \li{w}, on Line 1, is renamed, could that possibly break Line 3, or change its meaning?
-In other words, might the desugaring of Line 3 assume that some variable identified as \li{w} is in scope (even though \li{w} is not mentioned in the text of Line 3)?
+\item \textbf{(Context Dependence)} If \li{w}, on Line 1, is renamed, could that possibly break Line 3, or change its meaning because the desugaring of Line 3 assume that some variable identified as \li{w} is in scope (even though \li{w} is not mentioned in the text of Line 3)?
 \item \textbf{(Typing)} What type does \li{y} have?
@@
 answer questions like those above.) This is burdensome at all scales, but particularly when programming in the large, where it is common to encounter a wide variety of libraries \cite{DBLP:conf/sac/LammelPS11}.
-Forcing the programmer to directly examine the desugaring of the program text, or the logic that produced that desugaring, in order to answer these sorts of basic questions  can defeat the purpose of syntactic sugar -- decreasing cognitive cost \cite{Green89}.
-Analagous problems do not arise when programming without syntax extensions in languages like ML and Scala -- the binding discipline of the language ensures that programmers can reason lexically about binding and renaming, and types mediate abstraction over function and module implementations.
+Forcing the programmer to reason transparently, i.e.
+to examine the desugaring of the program text, or the logic that produced that desugaring, in order to answer these sorts of basic questions,  can defeat the purpose of syntactic sugar -- decreasing cognitive cost \cite{Green89}.
+Analagous problems do not arise when programming without syntax extensions in languages like ML and Scala -- the binding discipline of the language ensures that programmers can reason lexically about binding, and types mediate abstraction over function and module implementations.
 %From this perspective, it is unsurprising that the use of Camlp4 has been deprecated in the OCaml system.
 
-This paper details a more reasonable approach: typed literal macros (TLMs).
+This paper details a more reasonable mechanism: \textbf{typed literal macros (TLMs)}.
 TLMs give library providers full programmatic control over the parsing and expansion of \emph{generalized literal forms}, which syntactically subsume other literal forms because their bodies are left initially unparsed according to the fixed context-free syntax of the language (i.e.
-they are initially parsed like raw string literals.) The applied TLM parses the literal body and generates an expansion during a phase known as typed expansion, which generalizes the usual typing phase.
-For example, Fig.
-\ref{fig:K-tsm-example} applies a TLM, \li{$kquery}, to a backtick-delimited generalized literal form to express the ``unreasonable'' example from Fig.
+they are initially parsed like raw string literals.) For example, Fig.
+\ref{fig:K-tsm-example} applies a TLM, \li{$kquery}, to a backtick-delimited generalized literal form to express the ``unreasonable'' K query example from Fig.
 \ref{fig:K-dialect}.
+The applied TLM parses the literal body and generates an expansion during a semantic phase called typed expansion, which generalizes the usual typing phase.
 TLMs come equipped with uniquely powerful abstract reasoning principles.
-We will characterize these reasoning principles absolutely precisely as we proceed, but for now, let us develop some informal intuitions -- the client programmer can reason about Line 3 as follows:
+We will characterize these reasoning principles absolutely precisely as we proceed, but for now, let us develop some intuitions -- the programmer can reason about Line 3 as follows:
 \begin{enumerate}
@@
 The context-free syntax of the language is fixed, so the TLM provider can modularly establish that the syntax the TLM implements is deterministic.
-\item \textbf{(Segmentation)} The output that a TLM generates is structured so that the system can infer from it an accurate \emph{segmentation} of the literal body that distinguishes spliced segments, which appear directly in the expansion, from segments that are parsed in some other way by the TLM.
+\item \textbf{(Segmentation)} The output that a TLM generates is structured so that the system can infer from it an accurate \emph{segmentation} of the literal body that distinguishes spliced terms from segments that are parsed in some other way by the TLM.
 The segmentation can be communicated by a code editor using secondary notation, e.g.
 colors in this document.
-By examining the segmentation of Line 3, we know that the two instances of \li{x} on Line 3 are parsed as spliced expressions, whereas \li{R} and the other characters are parsed in some other way.
-\item \textbf{(Capture)} Splicing is capture-avoiding, so the spliced expression \li{x} must refer to the binding of \li{x} on Line 2.
-It cannot capture a coincidental binding of \li{x} in the expansion of Line 3.
-\item \textbf{(Context Dependence)} The system enforces context independence, so the expansion of Line 3 cannot  rely on the fact that, e.g., \li{w} is in scope.
-The client is free to rename \li{w}.
-(As it turns out, enforcing context independence while also allowing macros to parse literal bodies arbitrarily is not straightforward.
-We return to this point in Sec.
-\ref{sec:existing-approaches}.)
+By examining the segmentation of Line 3, we know that the two instances of \li{x} on Line 3 are parsed as spliced expressions (because they are in black), whereas the \li{R}'s are parsed in some other way (because they are in green.)
+\item \textbf{(Capture)} Splicing is capture-avoiding, so the spliced expressions \li{x} must refer to the binding of \li{x} on Line 2.
+They cannot capture a coincidental binding of \li{x} in the expansion.
+\item \textbf{(Context Dependence)} The system enforces context independence, so the expansion of Line 3 cannot  rely on the fact that, e.g., \li{w} is in scope. The client is free to rename \li{w}.
 \item \textbf{(Typing)} An explicit type annotation on the definition of \li{$kquery} determines the type that the expansion must have.
@@
 \begin{itemize}
-\item Sec. \ref{sec:setsms} begins by defining a {simple expression TLM} for XHTML and describes initial parsing of generalized literal forms, typed expansion and expansion validation in more detail.
+\item Sec. \ref{sec:setsms} defines a {simple expression TLM} for XHTML, inspired by Ur/Web, and describes the typed expansion process in more detail.
 \item Sec. \ref{sec:sptsms} introduces {simple pattern TLMs}, i.e. TLMs that generate patterns. %The hygiene condition for pattern TLMs is interesting.
@@
 \begin{enumerate}
-\item They allow a single TLM to operate not just at a single type, but over a type- and module-parameterized family of types.
-For example, we consider a parametric TLM for dictionaries that operates uniformly over different implementations of the abstract dictionary signature, and also over different choices of the range type.
-\item Parameters also give expansions access to terms other than those explicitly spliced in. %This is particularly convenient for working with abstract types \emph{a la} ML. 
-Partial parameter application is possible, which lowers the syntactic cost of the strict context independence discipline that the system enforces.
+\item They allow a TLM to operate not just at a single type, but over a type- and module-parameterized family of types.
+For example, we consider a parametric TLM for string-keyed dictionaries that operates uniformly over different implementations of the abstract dictionary signature, and also over different choices of the range type.
+\item Parameters also give expansions access to terms other than those explicitly spliced in to the literal body by the client.
+%This is particularly convenient for working with abstract types \emph{a la} ML.
+Partial parameter application is possible, which lowers the syntactic cost of the strict context independence discipline that TLMs enforce.
 \end{enumerate}
@@
 \ref{sec:existing-approaches} compares TLMs to the related work, including closely related work by \citet{TSLs,sac15} on type-specific languages, and the long history of work on macro systems.
-\item Sec. \ref{sec:discussion} concludes with a summary of the specific contributions of this paper and a brief discussion of present limitations and future research directions.
+\item Sec. \ref{sec:discussion} concludes with a summary of the specific contributions of this paper and a brief discussion of the present limitations and some future research directions.
 \end{itemize}
@@
 Haskell, Scala and others.) Languages that have a disciplined binding structure but that lack a rich static type structure, e.g.
-Racket and other Lisp derivatives, may also benefit from TLMs -- one can apply the results of this paper by deploying the usual type theorist's trick of casting such a language as a ``{unityped}'' language \cite{pfpl,scott1980lambda}.
+Racket and other Lisp derivatives, may also benefit from TLMs -- they can apply the results of this paper by deploying the usual type theorist's trick of viewing such a language as a ``{unityped}'' language \cite{pfpl,scott1980lambda}.
 % For the present purposes, we can consider VerseML as essentially the calculus defined in the supplement extended with various conveniences that are commonly found in other ML-like languages and, notionally, orthogonal to TLMs.

### 2017-06-12 popl18/syntax-popl18.tex
substantially revamped related work section

@@
 lists.
-The problem is that this practice neglects an unbounded variety of other data structures for which literal forms might also be useful (consider, for example,  maps, regexes, HTML trees, URLs, SQL queries and chemical structures.) %A less \emph{ad hoc} design would be one that allows library providers to define new literal forms in a decentralized manner.
+This practice neglects an unbounded variety of other data structures for which literal forms might also be useful (consider, for example,  maps, regexes, HTML data, SQL queries and chemical structures.) %A less \emph{ad hoc} design would be one that allows library providers to define new literal forms in a decentralized manner.
 However, the mechanisms available to programmers today, e.g.
@@
 the Scala macro system \cite{ScalaMacros2013}, are only somewhat more reasonable, and do not offer direct syntactic control.
-This paper introduces a system of \emph{typed literal macros (TLMs)} that allows library providers to define new literal forms in a decentralized manner.
-More specifically, TLMs give library providers full programmatic control over both the parsing and expansion of expressions and patterns of \emph{generalized literal form} \cite{TSLs} at a specified type or parameterized family of types.
+This paper introduces a system of \emph{typed literal macros (TLMs)} that gives library providers full programmatic control over both the parsing and expansion of expressions and patterns of \emph{generalized literal form} at any specified type or parameterized family of types.
 % The mechanism is {strictly hygienic}, meaning that 1) the expansion must be context-independent; and 2)  splicing is capture avoiding.
 % Partial parameter application lowers the syntactic cost of this strict style. 
-Compared to existing approaches, including syntax extension systems and macro-based quotation rewriting systems, TLMs are uniquely \emph{reasonable}: TLM providers can reason modularly about syntactic determinism, and TLM clients can reason abstractly about the type and binding structure of the program.
+Compared to existing approaches, including syntax extension systems and macro-based quotation systems, TLMs are uniquely \emph{reasonable}: TLM providers can reason modularly about syntactic determinism, and TLM clients can reason abstractly about the type and binding structure of the program.
 The system needs only to convey to clients (e.g.
 via secondary notation) the inferred \emph{segmentation} of each literal, which gives the locations of the spliced terms.
-The inferred segmentation also informs the hygiene mechanism (previously, only unhygienic macros could freely splice terms out of literal bodies.)
+The inferred segmentation also informs the hygiene mechanism (previously, only unhygienic macros could freely parse terms out of literal bodies.)
 %The system can determine this {segmentation} automatically and convey it to the client via syntax highlighting, or by presenting the client with a context-free grammar (without revealing the semantic actions associated with each production.)
@@
 
-Third-party library providers are quite justifiably envious of the special syntactic status of these few semantically ordinary standard library constructs.
+Third-party library providers are quite justifiably envious of the special syntactic status of these semantically ordinary standard library constructs.
 After all, it is not difficult to find other examples of types for which mathematicians, scientists and programmers have invented specialized notations.
@@
 
-Language designers cannot, of course, hope to accomodate all such types with primitive literal forms, so there has been considerable interest in mechanisms that allow programmers to define new literal forms in a decentralized manner.
-For example, we describe a number of syntax extension mechanisms, including Camlp4 \cite{ocaml-manual} and SugarJ/Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, in Sec.
+Language designers cannot, of course, hope to accomodate all such types with primitive literal forms \emph{a priori}, so there has been considerable interest in mechanisms that allow programmers to define new literal forms on their own.
+For example, we will describe a number of syntax extension mechanisms, including Camlp4 \cite{ocaml-manual} and SugarJ/Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, in Sec.
 \ref{sec:existing-approaches}.
@@
 The literal form on Line 3 constructs an encoding of a query in the stack-based database query language K, using its terse textual syntax \cite{Whitney:2001:LOR:376284.375783}.
-Of course, the programmer reading this code may well be unfamiliar with K's query syntax.
+Of course, the programmer reading this program text may well be unfamiliar with K's query syntax.
 The problem then is that there are no simple, abstract protocols for answering questions like:
@@
 
-This paper details a more reasonable mechanism: \textbf{typed literal macros (TLMs)}.
+This paper details a more reasonable mechanism: \emph{typed literal macros (TLMs)}.
 TLMs give library providers full programmatic control over the parsing and expansion of \emph{generalized literal forms}, which syntactically subsume other literal forms because their bodies are left initially unparsed according to the fixed context-free syntax of the language (i.e.
@@
 TLMs come equipped with uniquely powerful abstract reasoning principles.
-We will characterize these reasoning principles absolutely precisely as we proceed, but for now, let us develop some intuitions -- the programmer can reason about Line 3 as follows:
+We will characterize these reasoning principles precisely as we proceed, but for now, let us develop some intuitions -- the programmer can reason abstractly about Line 3 as follows:
 \begin{enumerate}
 \item \textbf{(Responsibility)} The applied TLM, \li{$kquery}, is solely responsible for parsing and expansion of the literal body.
-The context-free syntax of the language is fixed, so the TLM provider can modularly establish that the syntax the TLM implements is deterministic.
+The context-free syntax of the language is fixed, so the TLM provider can modularly establish that the syntax this TLM implements is deterministic.
 \item \textbf{(Segmentation)} The output that a TLM generates is structured so that the system can infer from it an accurate \emph{segmentation} of the literal body that distinguishes spliced terms from segments that are parsed in some other way by the TLM.
@@
 colors in this document.
-By examining the segmentation of Line 3, we know that the two instances of \li{x} on Line 3 are parsed as spliced expressions (because they are in black), whereas the \li{R}'s are parsed in some other way (because they are in green.)
+By examining Fig.
+\ref{fig:K-tsm-example}, we know that the two instances of \li{x} on Line 3 are parsed as spliced expressions (because they are in black), whereas the \li{R}'s must be parsed in some other way (because they are in green.)
 \item \textbf{(Capture)} Splicing is capture-avoiding, so the spliced expressions \li{x} must refer to the binding of \li{x} on Line 2.
@@
 The remainder of this document is organized as follows.
-\begin{itemize}
-\item Sec. \ref{sec:setsms} defines a {simple expression TLM} for XHTML, inspired by Ur/Web, and describes the typed expansion process in more detail.
-\item Sec. \ref{sec:sptsms} introduces {simple pattern TLMs}, i.e. TLMs that generate patterns. %The hygiene condition for pattern TLMs is interesting.
-\item Sec. \ref{sec:setsms-formally} defines a type-theoretic calculus of simple expression and pattern TLMs and formally establishes the reasoning principles implied above.
-
-% In the remaining chapters, we enrich the language developed in Chapter \ref{chap:uetsms} with advanced features based on those found in full-scale general-purpose languages like ML and Scala, and enhance our TLM mechanism with these new features.
-
-% We next develop a more sophisticated account of TLMs in Chapters \ref{chap:ptsms} and \ref{chap:static-eval}, with the aim of working out details necessary to integrate TLMs into full-scale general-purpose languages like ML or Scala.
-%These two chapters constitute Part \ref{part:parametric-tsms} of our contributions.
-\item Sec.
-\ref{sec:ptsms} adds type functions and an ML-style module system and introduces \emph{parametric TLMs} (pTLMs), i.e.
-TLMs that take type and module parameters.
-Parameters serve two purposes:
-\begin{enumerate}
-\item They allow a TLM to operate not just at a single type, but over a type- and module-parameterized family of types.
-For example, we consider a parametric TLM for string-keyed dictionaries that operates uniformly over different implementations of the abstract dictionary signature, and also over different choices of the range type.
+% \begin{itemize}
+% \item 
+Sec. \ref{sec:setsms} defines a {simple expression TLM} for XHTML, which implements Ur/Web's XHTML literal syntax, and describes the typed expansion process in more detail.
+% \item 
+Sec. \ref{sec:sptsms} briefly introduces {simple pattern TLMs}, i.e. TLMs that generate patterns. %The hygiene condition for pattern TLMs is interesting.
+% \item 
+Sec. \ref{sec:setsms-formally} defines a type-theoretic calculus of simple expression and pattern TLMs and formally establishes the reasoning principles implied above.
+% \item 
+Sec. \ref{sec:ptsms} adds type functions and an ML-style module system and introduces \emph{parametric TLMs} (pTLMs), i.e. TLMs that take type and module parameters. Parameters serve two purposes:
+\begin{enumerate}[noitemsep,nolistsep]
+\item They allow a TLM to operate not just at a single specified type, but over a type- and module-parameterized family of types.
+We describe a parametric TLM for string-keyed dictionaries that operates uniformly over different implementations of the abstract dictionary signature, and also over different choices of the range type.
 \item Parameters also give expansions access to terms other than those explicitly spliced in to the literal body by the client.
@@
 \end{enumerate}
-This section culminates with a calculus that extends the simpler calculus of Sec. \ref{sec:setsms-formally} with type functions, ML-style modules and parametric TLMs.
-\item Sec.
+This section culminates with a calculus that extends the calculus of Sec. \ref{sec:setsms-formally} with type functions, ML-style modules and parametric TLMs.
+% \item
+Sec.
 \ref{sec:static-eval} considers the topic of term evaluation during the typed expansion phase, i.e.
@@
 TLMs that implement parser generators and quasiquotation.
-\item Sec.
-\ref{sec:existing-approaches} compares TLMs to the related work, including closely related work by \citet{TSLs,sac15} on type-specific languages, and the long history of work on macro systems.
-\item Sec. \ref{sec:discussion} concludes with a summary of the specific contributions of this paper and a brief discussion of the present limitations and some future research directions.
-\end{itemize}
+% \item 
+Sec. \ref{sec:existing-approaches} compares TLMs to related work, including closely related work by \citet{TSLs,sac15} on type-specific languages, and the long history of work on macro systems.
+% \item 
+Sec. \ref{sec:discussion} concludes with a summary of the specific contributions of this paper and a brief discussion of the present limitations and some future research directions.
+% \end{itemize}
 
@@
 \ref{sec:static-eval} are relegated to the supplemental material.
-The examples that we give are written in an emerging dialect of ML called VerseML that supports TLMs as described herein.
-Our design should straightforwardly generalize to other languages with a broadly similar philosophy regarding types and binding (e.g.
-Haskell, Scala and others.) Languages that have a disciplined binding structure but that lack a rich static type structure, e.g.
-Racket and other Lisp derivatives, may also benefit from TLMs -- they can apply the results of this paper by deploying the usual type theorist's trick of viewing such a language as a ``{unityped}'' language \cite{pfpl,scott1980lambda}.
+The examples that we give are written in an emerging dialect of ML called VerseML that supports TLMs as described herein, but the essence of the mechanism is not ML-specific -- it should be straightforward to adapt it to other languages that take a similar approach to types and binding (e.g.
+Haskell, Scala and others.) When a design choice is ML-specific, we will make note of it\todo{do this}.
+Languages that have a disciplined binding structure but that lack a rich static type structure, e.g.
+Racket and other LISP derivatives, may also benefit from TLMs -- they can apply the results of this paper by deploying the usual type theorist's trick of viewing such a language as ``{unityped}'' \cite{pfpl,scott1980lambda}.
 % For the present purposes, we can consider VerseML as essentially the calculus defined in the supplement extended with various conveniences that are commonly found in other ML-like languages and, notionally, orthogonal to TLMs.
@@
 
-We will begin in this section by introducing \emph{simple expression TLMs} (seTLMs), i.e. TLMs where expansions are expressions of a single specified type.
+We will begin in this section by introducing \emph{simple expression TLMs} (seTLMs), i.e. TLMs where expansions are expressions of a single specified type, by example.
 
@@
 
-\section{Reasoning Criteria and Existing Approaches}\label{sec:existing-approaches}
-\subsection{Library-Specific Syntax Dialects}\label{sec:syntax-dialects-intro}
-One approach available to library providers is to use a syntax definition system to construct a library-specific \emph{syntax dialect}: a new syntax definition that extends the syntax definition given in the language definition with new forms, including literal forms.
-Ur/Web's syntax (Fig.
-1) is a library-specific dialect of Ur's syntax.
+\newcommand{\rwSec}{Related Work}
+\section{\protect\rwSec}
+\label{sec:existing-approaches}
+\subsection{Syntax Definition Systems}\label{sec:syntax-dialects-intro}
+One approach available to library providers seeking to introduce new literal forms is to use a syntax definition system to construct a library-specific \emph{syntax dialect}: a new syntax definition that extends the syntax definition given in the language definition with new forms, including literal forms.
+For example, Ur/Web's syntax (Fig.
+1) is a library-specific dialect of Ur's syntax \cite{conf/pldi/Chlipala10,conf/popl/Chlipala15}.
 
@@
 However, we argue that in fact
-There are hundreds of syntax definition systems of various design, some integrated deeply into language front-ends and build systems, others operating as standalone preprocessors.
-Examples include grammar-based systems like Camlp4 \cite{ocaml-manual}\todo{better citation?}, Copper \cite{conf/gpce/WykS07} or SugarJ/Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, parser combinator systems \cite{Hutton1992d} and less general operator-based systems, like the system of fixity directives in Standard ML \cite{mthm97-for-dart}\todo{better citation?}~ and the more elaborate mixfix systems \cite{5134,DBLP:conf/gpce/TahaJ03} built in to proof assistants like Agda \cite{DBLP:conf/ifl/DanielssonN08} and Coq \cite{Coq:manual}.
+There are hundreds of syntax definition systems of various design.
+Notable examples include grammar-oriented systems like Camlp4 \cite{ocaml-manual}\todo{better citation?}, Copper \cite{conf/gpce/WykS07,conf/pldi/SchwerdfegerW09}, SugarJ/Sugar*/SoundExt \cite{erdweg2011sugarj,erdweg2013framework,conf/icfp/LorenzenE13,conf/popl/LorenzenE16}, as well as various parser combinator systems \cite{Hutton1992d}.
+The parsers generated by these systems can be invoked to preprocess program text in various ways, e.g.
+by invoking them from within a build script, by using a preprocessor-aware build system (e.g.
+\li{ocamlbuild}), or via language-integrated preprocessing directives (e.g.
+Racket's \li{#lang} directive or its reader macros \cite{Flatt:2012:CLR:2063176.2063195}, or the import mechanism of SugarJ/Sugar*/SoundExt.)
 %library-specific (a.k.a. ``domain-specific'')
 
-\paragraph{Criteria 1: Modular Reasoning About Syntactic Determinism}
-The first problem with this approach is that clients cannot always deterministically combine different library-specific syntax dialects when they want to use the new forms that they define together.
+The first problem with using a syntax definition system to directly extend the context-free syntax of a language is that clients cannot always deterministically combine the resulting extended syntax dialects when they want to use the new forms that they define together, i.e.
+there can be syntactic conflicts.
 This is a significant problem because large programs use many separately developed libraries \cite{DBLP:conf/sac/LammelPS11} and often do not fall cleanly into a single ``problem domain''.
 
-In some cases, the dialects in question cannot be combined simply because they have been defined under incompatible syntax definition systems.
+One possible solution is found in Copper, which integrates a modular grammar analysis that guarantees that determinism is conserved when syntax dialects of a certain restricted class are combined \citet{conf/pldi/SchwerdfegerW09}.
+The caveat is that the constituent dialects must prefix all newly introduced forms with marking tokens drawn from disjoint sets.
+To be confident that the marking tokens used are disjoint, providers must base them on the domain name system or some other coordinating entity.
+Because the mechanism operates at the level of the context-free grammar, it is difficult for the client to define scoped abbreviations for these verbose marking tokens.
+
+TLMs sidestep these complexities entirely because the context-free syntax of the language is fixed, and composition is via splicing rather than direct combination.
+The TLM provider can therefore modularly establish that the syntax definition that the TLM implements (possibly using a syntax definition system internally, as discussed in Sec.
+\ref{sec:static-eval}) is deterministic.
+There is no need for verbose marking tokens, and programmers can define scoped TLM abbreviations as discussed in Sec.
+\ref{sec:ptsms}.
+
+% In some cases, the dialects in question cannot be combined simply because they have been defined under incompatible syntax definition systems.
 In other cases, the problem is one of \emph{conflict}: two dialects might define the same form, but determine different desugarings.
@@
 
-\citet{conf/pldi/SchwerdfegerW09} have developed a modular grammar analysis, implemented in Copper, that guarantees that determinism is conserved when syntax dialects of a certain restricted class are combined, the caveat being that the constituent dialects must prefix all newly introduced forms with marking tokens drawn from disjoint sets.
-To be confident that the marking tokens used are disjoint, providers must base them on the domain name system or some other coordinating entity.
-Because the mechanism operates at the level of the context-free grammar, it is difficult for the client to define scoped abbreviations for these verbose marking tokens (as one might do when, for example, deeply nested modules are used for namespacing in ML.)
-
-\paragraph{Criteria 2: Abstract Reasoning About Literal Forms}
-
 Even putting aside the problem of syntactic conflict, there are questions about just how reasonable sprinkling many possibly unfamiliar library-specific literal forms  throughout a program may be.
+The intuition was given by example in Sec.
+\ref{sec:intro}.
+To reiterate, more generally:
+\begin{enumerate}[noitemsep,nolistsep]
+\item \textbf{Responsibility}: Clients using a combined syntax dialect cannot easily determine which constituent extension is responsible for a given form, whereas TLM clients can follow the binding structure of the language in the usual manner.
+\item \textbf{Segmentation}: Clients of a syntax dialect cannot accurately determine which segments of the program text appear directly in the desugaring.
+In contrast, TLM clients can inspect the inferred segmentation (communicate via, e.g., secondary notation.)
+\item \textbf{Binding}: Clients of a context-free syntax dialect cannot be sure that the desugaring is context-independent and that spliced terms are capture avoiding, as TLM clients can.
+\item \textbf{Typing}: Clients of a syntax dialect cannot reason abstractly about types.
+In contrast, TLM clients clients can determine the type of any expansion by referring to the parameter and type declarations on the TLM definition, and nothing else.
+The inferred segmentation also gives types for each spliced expression or pattern.
+This information can be communicated upon request by the type inspection service of a program editor.
+\end{enumerate}
+
+The work of \citet{conf/icfp/LorenzenE13,conf/popl/LorenzenE16} introduces SoundExt, a grammar-based syntax extension system where extension providers can equip their new forms with derived typing rules.
+The system then attempts to automatically verify that the expansion logic (expressed using a rewrite system, rather than an arbitrary function) is sound relative to these derived rules.
+
+TLMs differ in several ways.
+First, as already discussed, we leave the context-free syntax fixed, so different TLMs cannot conflict.
+Second, SoundExt does not enforce hygiene, i.e.
+expansions might depend on the context and intentionally induce capture.
+Similarly, there is no abstract segmentation discipline.
+While the client can indirectly reason about binding (but not segmentation) by inspecting the derived typing rules, this is weaker than the strict hygiene and segmentation discipline that TLMs enforce.
+Another distinction is that TLMs do not support type-dependent expansions.
+The trade-off is that this allows expansions, and therefore segmentations, to be generated even when the program is ill-typed, as long as the static values are well-typed\todo{elaborate}.
+Another important distinction is that TLMs rely on proto-expansion validation, rather than verification as in SoundExt.
+The trade-off is that TLMs do not require that the expansion logic be written using a restricted rewriting system, nor does the system require a fully mechanized language definition and heavyweight theorem proving machinery.
+Finally, it would be difficult to define syntax extensions in SoundExt that operate over module-parameterized families of types.
+Parameters can be approximated using splicing, but there is no clear notion of ``partial application''.
 
@@
 
-\subsection{Another Existing Approach: Term Rewriting}
+\subsection{Term Rewriting Systems}
+Another approach -- and the approach that TLMs are rooted in -- is to leave the context-free syntax of the language fixed, and instead contextually rewrite existing literal forms.
+
 For example, OCaml's textual syntax now includes \emph{preprocessor extension (ppx) points} used to identify terms that some term rewriting preprocessor must rewrite \cite{ocaml-manual}.
-We could attempt to mark a string literal containing Ur/Web-style HTML syntax with a ppx annotation, \li{xml}, as follows:
+We could  mark a string literal containing Ur/Web-style HTML syntax with a ppx annotation, \li{xml}, as follows:
 \begin{lstlisting}[numbers=none]
@@
 \end{lstlisting}
-Again, there are problems reasoning modularly and abstractly about such constructions.
+There are problems reasoning modularly and abstractly about such constructions.
 More than one applied preprocessor might recognize this annotation (there are, in practice, many XML/HTML libraries), so the problems of \textbf{Responsibility} come up.
-It is also impossible to reason abstractly about \textbf{Segmentation}, \textbf{Capture}, \textbf{Context Dependence} and \textbf{Typing} because the code that the preprocessor generates is arbitrary.
-For these reasons, users of this mechanism warn that they should be used sparingly.\todo{citation}
-
-Term-rewriting macro systems are language-integrated local term rewriting systems that require that the client explicitly apply the intended rewriting, implemented by a macro, to the term that is to be rewritten\todo{readeer macros}.
+It is also impossible to reason abstractly about \textbf{Segmentation}, \textbf{Capture}, \textbf{Context Dependence} and \textbf{Typing} because the code that the preprocessor generates is unconstrained.% For these reasons, users of this mechanism warn that they should be used sparingly.\todo{citation}
+
+Term-rewriting macro systems are language-integrated local term rewriting systems that require that the client explicitly apply the intended rewriting, implemented by a macro, to the term that is to be rewritten.
 This addresses the issue of \textbf{Responsibility}.
-However, na\"ive macro systems, like the earliest variants of the Lisp macro system \cite{Hart63a}, Template Haskell \cite{SheardPeytonJones:Haskell-02} and GHC's quasiquotation system \cite{mainland2007s}, do not allow clients to reason abstractly about the remaining issues, again because the expansion that they produce is arbitrary.
+However, unhygienic, untyped macro systems, like the earliest variants of the Lisp macro system \cite{Hart63a}, Template Haskell \cite{SheardPeytonJones:Haskell-02} and GHC's quasiquotation system \cite{mainland2007s} (which is based on Template Haskell), do not allow clients to reason abstractly about the remaining issues, again because the expansion that they produce is unconstrained.
+(It is not enough that with Template Haskell / GHC quasiquotation, the generated expansion is typechecked -- to satisfy the \textbf{Typing} criterion, it must be possible to reason abstractly about \emph{what the type of the generated expansion is}.)
 
 \emph{Hygienic} macro systems prevent, or abstractly account for \cite{DBLP:conf/esop/HermanW08,Herman10:Theory}, \textbf{Capture}, and they enforce application-site \textbf{Context Independence} \cite{Kohlbecker86a,DBLP:conf/popl/Adams15,DBLP:conf/popl/ClingerR91,DBLP:journals/lisp/DybvigHB92}.
-However, this turns out to make it difficult to use hygienic macros to introduce new literal forms of arbitrary design with support for splicing.
-Consider again our running XHTML example, which we might try to realize by applying a hygienic macro, \li{xml!}, to a string literal:
+However, this turns out to make it impossible to repurpose string literal forms to introduce compositional literal forms at other types.
+Consider again our running XHTML example, which we might try to realize by applying a hygienic macro, \li{xml!}, to a string literal that the macro parses:
 \begin{lstlisting}[numbers=none]
@@
 \end{lstlisting}
-THIS IS BAD BECAUSE FIRSTNAME IS BAD HYGIENE GRAH. MAYBE TALK ABOUT STRING SPLICING AND FRAGMENT SYSTEM STUFF HERE.
-
-PUTTING THAT ASIDE, LET US CONSIDER TYPING.
-LISP IS DUMB NO TYPES.
-HERMAN AND WAND USED TYPES ONLY TO ACCOUNT FOR BINDING RELATIONSHIPS.
-SCALA HAS A TYPED MACRO SYSTEM, BUT THERE IS NO FORMAL ACCOUNT OF IT.
-MACROML IS ONLY A STAGING MACRO SYSTEM, NO WAY TO PARSE LITERALS.
-(It is not enough that with Template Haskell / GHC quasiquotation, the generated expansion is typechecked -- to satisfy the \textbf{Typing} criterion, it must be possible to reason abstractly about \emph{what the type of the generated expansion is}.) MAYBE BRING UP TSLS AND TSMS HERE.
-
-TRANSITION TO THE CONTRIBUTIONS OF THIS PAPER: TYPED LITERAL MACROS.
-REWRITE GENERALIZED LITERAL FORMS, KEEP TRACK OF SPLICED SEGMENTS SO THE HYGIENE MECHANISM IS COMPATIBLE, SEGMENTATION IS ALWAYS POSSIBLE.
-TYPED, INTEGRATED INTO ML SCALA-LIKE LANGUAGE (SECTION HEADINGS).
-FORMAL DEFINITION.
-REASONING PRINCIPLES ARE THEOREMS.
-THATS HOW WE EVALUATE OUR WORK.
-(USEFULNESS / USABILITY CONCERNS NEED TO BE ADDRESSED SOMEWHERES)
-
-Although much of the research on macro systems has been for languages in the Lisp tradition \cite{mccarthy1978history}, some progress has also been made on reasoning about types with the design of \emph{typed macro systems} where annotations constrain the macro arguments and the generated expansions.
-Examples include Scala's macro system \cite{ScalaMacros2013}, formal calculi studied by \citet{DBLP:conf/esop/HermanW08} and \emph{staging macro systems} like MetaML \cite{Sheard:1999:UMS} and MacroML \cite{ganz2001macros} (which, it is worth noting, do not give the macro access to the syntax tree of arguments.)
-
-Perhaps the biggest problem, then, is that term-rewriting systems offer library providers only limited syntactic control.
-In other words many syntactic forms of interest will not particularly resemble those available in the host language.
-For example, consider the  XHTML and K examples above.
-In both cases, the syntactic conventions are quite distinct from those of ML-like languages (and, for that matter, languages that use S-expression.)
-
-It is tempting in these situations to consider repurposing string literal forms.
-For example, we might wish to apply a macro \li{html!} (following Rust's convention of using a post-fix \li{!} to distinguish macro names from variables) to rewrite string literals containing Ur/Web-style XHTML syntax as follows:
-\begin{lstlisting}[numbers=none]
-  fun heading first_name = html! 'SSTR<h1>Hello, {[first_name]}!</p>ESTR'
-  val body = html! 'SSTR<body>{heading "World"} ...</body>ESTR'
-\end{lstlisting}
-
-The problem here is that there is no way to extract the spliced expressions from the supplied string literal forms while satisfying the hygiene conditions, because variables that come from these spliced terms (here, \li{first_name} and \li{heading}) are indistinguishable from variables that inappropriately appear free relative to the expansion.
-In addition, the problem of segmentation becomes even more pernicious: to a human or tool unaware of Ur/Web's syntax, it is not immediately apparent which particular subsequences of the string literals supplied to \li{html!} are segmented out as spliced expressions.
-Reader macros have essentially the same problem  \cite{DBLP:journals/jfp/FlattCDF12}.
-
-  Some languages build in \emph{string splicing} (a.k.a.
-\emph{string interpolation}) forms, e.g.
-Scala \cite{odersky2008programming}, or similar but more general \emph{fragmentary quotation forms} \cite{conf/icfp/Slind91}, e.g.
+The expansion of this macro will either fail the check for context independence because \li{first_name} will appear as a free variable, no different from any other, or else it will have incorrectly captured an expansion-local binding of the same identifier.
+The hygiene mechanism for TLMs addresses this problem by explicitly distinguishing spliced segments of the literal body in the proto-expansion.
+This also addresses the problem of \textbf{Segmentation} -- the segmentation abstractly communicates the fact that (only) \li{first_name} is a spliced expression of type \li{string}.
+
+Much of the research on macro systems has been for languages in the LISP tradition \cite{mccarthy1978history} that do not have rich static type structure.
+The formal macro calculus studied by \citet{DBLP:conf/esop/HermanW08} uses types to encode the binding structure of the generated expansion (in particular, to indicate which identifiers passed as arguments can be captured by which other arguments), but the expanded language itself does not have rich type structure (nor is this calculus is not capable of expressing new literal forms, for the reasons already discussed.) Research on typed \emph{staging macro systems} like MetaML \cite{Sheard:1999:UMS} and MacroML \cite{ganz2001macros} is also not directly applicable to the problem of defining new literal forms -- the syntax tree of the arguments cannot be inspected at all (staging macros are useful mainly for performance reasons.) The Scala macro system is a hygienic macro system which supports ``black-box'' term rewriting macros.
+These support reasoning abstractly about \textbf{Typing} because type annotations constrain the macro arguments and the generated expansions, though the precise reasoning principles available are unclear because the Scala macro system has not been formally specified.
+In any case, black-box macros also cannot be used to repurpose string literal forms for the reasons above.
+
+  Some languages, including Scala \cite{odersky2008programming}, build in \emph{string splicing} (a.k.a.
+\emph{string interpolation}) forms, or similar but more general \emph{fragmentary quotation forms} \cite{conf/icfp/Slind91}, e.g.
 the SML/NJ dialect of ML.
-These designate particular delimiters to allow escape out into the expression language.
-The problem with using these (dynamically, or in conjunction with term rewriting macros) as vehicles to introduce syntactic sugar at various types is 1) there is no ``one-size-fits-all'' escape delimiter, and 2) reasoning about typing is problematic because every escaped term is checked against the same type.
-In the example above, we spliced in two different types of expressions using two different delimiters.
-In general, e.g.
+These designate a particular delimiter to escape out into the expression language.
+The problem with using these (dynamically, or in conjunction with term rewriting macros) as vehicles to introduce literal forms at various other types is 1) there is no ``one-size-fits-all'' escape delimiter, and 2) typing is problematic because every escaped term is checked against the same type.
+In the XHTML example, we have splicing at two different types using two different delimiters.
+These forms are also not available in pattern position.%In general, e.g.
 when defining syntax for a programming language with many sorts of terms, the most appropriate choice of delimiters might depend on where each spliced term appears.
 
+This brings us to the most closely related work, that of \citet{TSLs} on \emph{type-specific languages} (TSLs).
+Like simple expression TLMs (Sec.
+\ref{sec:setsms}), TSLs allow library providers to programmatically control the parsing of expressions of generalized literal form.
+Parse functions are associated directly with nominal types and invoked according to a bidirectionally typed protocol.
+In contrast, TLMs are separately defined and explicitly applied.
+Accordingly, different TLMs can therefore be defined at the same type without conflict, and can operate at any type, not just at nominal (i.e.
+abstract) types.
+In a subsequent short paper, \citet{sac15} proposed explicit application of simple expression TLMs in a bidirectional typed setting \cite{Pierce:2000:LTI:345099.345100}, but this proposal did not give any formal details.
+With TLMs, it is not necessary for the language to be bidirectionally typed -- context independence implies that ML-style type inference can be performed using only the segmentation (because the remainder of an expansion cannot mention the very variables whose types are being inferred.) Another critical distinction is that the metatheory presented by \citet{TSLs} establishes only that generated expansions are well-typed.
+It does not establish the remaining abstract reasoning principles that have been a major focus of this paper.
+In particular, there is no formal notion of hygiene (though it is discussed informally), and the mechanism does not guarantee that a valid segmentation will exist, nor associate types with segments.
+Finally, this prior work did not consider pattern matching, type functions, ML-style modules, parameters or static evaluation.
+All of these are important for integration into an ML- or Scala-like language.
+
+% These existing forms normally have other meanings, so this can be confusing \cite{pane1996usability}.
+
+\newcommand{\discussionSec}{Discussion}
+\section{\protect\discussionSec}
+\label{sec:related-work}
+\label{sec:discussion}
+\label{sec:conclusion}
+\vspace{-3px}
 {This paper introducing a system of \emph{typed literal macros (TLMs)} that allows programmers to define new literal syntax of nearly arbitrary form within an ML- or Scala-like  language}, i.e.
@@
 This is the first detailed type-theoretic accout of a typed, hygienic term-rewriting macro system for an ML- or Scala-like language.\todo{refine and justify this claim}
-
-% These existing forms normally have other meanings, so this can be confusing \cite{pane1996usability}.
-
-\newcommand{\discussionSec}{Additional Related Work \& Concluding Discussion}
-\section{\protect\discussionSec}
-\label{sec:related-work}
-\label{sec:discussion}
-\label{sec:conclusion}
-\vspace{-3px}
-The most closely related work is that of \citet{TSLs} on \emph{type-specific languages} (TSLs), which introduced generalized literal forms and gave a bidirectionally typed protocol for type-directed parse function invocation.
-This work differs in several ways, and the differences we note below nicely summarize our specific contributions.
-
-Most obviously, TLMs are explicitly applied.
-Different TLMs can therefore be defined at the same type without conflict.
-Moreover, TLMs are not associated directly with generative type definitions, like TSLs,  but rather can operate at any type.
-In a subsequent short paper, \citet{sac15} proposed explicit application of simple expression TLMs in a bidirectional typed setting \cite{Pierce:2000:LTI:345099.345100}, but did not provide any formal details.
-We do not assume a bidirectionally typed language.
-Context independence implies that ML-style type inference can be performed using only the spliced terms and their types (because the remainder of an expansion cannot mention the very variables whose types are being inferred, nor constrain them further beyond what the annotations require.) We leave additional inference-related topics, e.g.
-TLM parameter inference, as future work.
-
-Another distinction is that the metatheory presented by \citet{TSLs} establishes only that generated expansions are well-typed.
-It does not establish the remaining abstract reasoning principles that have been a major focus of this paper.
-In particular, there is no formal notion of hygiene (though it is discussed informally), and the mechanism does not guarantee that a valid segmentation will exist, nor associate types with segments.
-
-Finally, the prior work on TSLs has not considered pattern matching, parameterized types, modules, parametric expansions or static evaluation.
-All of these are important for integration into an ML-like functional language.
-
-Another particularly compelling stream of related work is that by \citet{conf/icfp/LorenzenE13,conf/popl/LorenzenE16}, which describes SoundExt, a grammar-based syntax extension system where extension providers can equip their new forms with derived typing rules.
-The system then attempts to automatically verify that the expansion logic (expressed using a rewrite system, rather than an arbitrary function) is sound relative to these derived rules.
-
-Our work differs in several ways.
-First, we leave the context-free syntax fixed, so different TLMs cannot conflict.
-Second, SoundExt does not enforce hygiene, i.e.
-expansions might depend on the context and intentionally induce capture.
-Relatedly, TLMs do not support type-dependent expansion, as in \cite{conf/popl/LorenzenE16}.
-Another important distinction is that our system relies on (proto-)expansion validation, rather than verification.
-In other words, we do not prove properties about the parse function itself, but rather ``test'' each expansion.
-The trade-off is that our system does not require that parsing logic be written in a restricted manner, nor require a fully mechanized language definition and theorem proving machinery.
-Finally, it would be difficult to define syntax extensions in SoundExt that operate over module-parameterized families of types.
-Parameters can be approximated using splicing, but there is no clear notion of ``partial application''.
 

### 2017-06-13 popl18/syntax-popl18.tex
started revamping discussion sec

@@
                 | SplicedE of segment * proto_typ
-
 \end{lstlisting}
@@
 \label{sec:conclusion}
-\vspace{-3px}
-{This paper introducing a system of \emph{typed literal macros (TLMs)} that allows programmers to define new literal syntax of nearly arbitrary form within an ML- or Scala-like  language}, i.e.
-a language with a rich static type system (Sec.
-2), support for pattern matching (Sec.
-3) and an ML-style module system or Scala-style object system with support for type members and type abstraction (Sec.
-5).
-The examples in this paper are written in VerseML, a new dialect of ML that features TLMs.
-
-Compared to existing approaches, TLMs are more \emph{reasonable}.
-We justify this claim by (1) developing a collection of relevant reasoning principles based on the reasoning principles already available to client programmers when using an ML-like or Scala-like language without such a mechanism; (2) formally operationalizing these reasoning principles as metatheorems about a series of type-theoretic \emph{TLM calculi} of increasing sophistication (Sec.
-4-5); and (3) evaluating existing approaches -- including syntax definition systems like Camlp4 \cite{ocaml-manual}\todo{better citation}, SugarJ/Sugar* \cite{erdweg2011sugarj,erdweg2013framework} and Copper \cite{conf/pldi/SchwerdfegerW09,schwerdfeger2010context}, term rewriting macro systems as found in a variety of languages, and the system of type-specific languages (TSLs) and typed syntax macros (TSMs) described by \citet{TSLs,sac15} -- relative to these reasoning principles (Sec.
-6).
-We conclude by discussing limitations and future research directions in Sec.
-7\todo{sec refs}.
-A complete set of definitions and associated proofs are available in the  supplemental material.
-This is the first detailed type-theoretic accout of a typed, hygienic term-rewriting macro system for an ML- or Scala-like language.\todo{refine and justify this claim}
-
-% Other related work was already described in Sec. \ref{sec:intro}.\todo{reader macros}
+
+The importance of good notation as a ``tool for thought'' has long been recognized\todo{cite iverson}.
+A good notation, according to A.
+N.
+Whitehead, ``relieves the brain of [...] unnecessary work'' and ``sets it free to concentrate on more advanced problems''\todo{citation}.
+Indeed, advances in mathematics, science and software engineering have quite often been accompanied by new notation.
+Of course, this desire to ``relieve the brain of all unnecessary work'' has influenced not just the surface syntax but also the semantics of languages like ML and Scala, which maintain a strong type and binding discipline so that programmers, and their tools, can hold certain ``implementation details'' abstract when reasoning about program behavior.
+
+Previously, these  were in tension -- .
+
+Not all notation is good, and we make no empirical claims as to the value of any particular notation in this paper.
+TLMs leave essentially all choices about literal notation, and therefore the burden of establishing the value of any particular literal notation, to individual library providers, rather than to the language designer.
+This decentralization of control is justified by the fact that when the programmer encounters an unfamiliar or bad notation, it does not ruin their day.
+The programmer is in no worse a position...
+
+The broader empirical question of whether the capabilities that TLMs provide are of interest to programmers ``in the wild'' is well beyond the scope of a conference paper such as this.
+The intended audience for this paper is language designers who wish to decentralize control over literal forms but seek a detailed understanding, rooted in the first principles of type theory, of a reasonable mechanism by which to achieve that goal in the setting of an ML- or Scala-like language.
+This paper is the first such account of such a mechanism.
+More generally, this paper contributes the first detailed type-theoretic account of a typed, hygienic macro system for an ML-like language, i.e.
+one with a rich static type system, support for pattern matching and an ML-like module system with support for type functions and abstract type members.
+(To summarize Sec.
+\ref{sec:existing-approaches}, the closest existing typed, hygienic macro system -- the Scala macro system \cite{ScalaMacros2013} -- is not formally specified, the work of \cite{DBLP:conf/esop/HermanW08} uses types only to reason abstractly about binding, and the specification of TSLs in the work of \citet{TSLs} did not formalize hygiene nor handle these advanced language features.)
+
+It is trivial to observe that the literal forms already chosen to appear in various languages, including the examples in Sec. \ref{sec:intro},
 
@@
 
-Another interesting question is what this mechanism would look like in the setting of a projectional structure editor, i.e.
+Other directions for future work are given in the supplement...
+
+% Another interesting question is what this mechanism would look like in the setting of a projectional structure editor, i.e.
 one where the syntax is not textual but rather tree-shaped and projected in an interative manner to the programmer.
@@
 
-To conclude, TLMs give substantial syntactic control to library providers while leaving programmers with strong abstract reasoning principles.
+% To conclude, TLMs give substantial syntactic control to library providers while leaving programmers with strong abstract reasoning principles.
 We believe TLMs therefore occupy a ``sweet spot'' in the  design space.

### 2017-06-25 popl18/syntax-popl18.tex
revisions to introduction and abstract

@@
 lists.
-This practice neglects an unbounded variety of other data structures for which literal forms might also be useful (consider, for example,  maps, regexes, HTML data, SQL queries and chemical structures.) %A less \emph{ad hoc} design would be one that allows library providers to define new literal forms in a decentralized manner.
+This practice is \emph{ad hoc} -- there are an unbounded variety of other data structures for which literal forms might also be useful (consider, for example,  maps, regexes, HTML data, SQL queries, syntax trees and chemical structures.) %A less \emph{ad hoc} design would be one that allows library providers to define new literal forms in a decentralized manner.
 However, the mechanisms available to programmers today, e.g.
@@
 the Scala macro system \cite{ScalaMacros2013}, are only somewhat more reasonable, and do not offer direct syntactic control.
-This paper introduces a system of \emph{typed literal macros (TLMs)} that gives library providers full programmatic control over both the parsing and expansion of expressions and patterns of \emph{generalized literal form} at any specified type or parameterized family of types.
+This paper introduces a system of \emph{typed literal macros (TLMs)} that gives library providers full programmatic control over the parsing and expansion of expressions and patterns of \emph{generalized literal form} at \emph{any} specified type or parameterized family of types.
 % The mechanism is {strictly hygienic}, meaning that 1) the expansion must be context-independent; and 2)  splicing is capture avoiding.
@@
 Compared to existing approaches, including syntax extension systems and macro-based quotation systems, TLMs are uniquely \emph{reasonable}: TLM providers can reason modularly about syntactic determinism, and TLM clients can reason abstractly about the type and binding structure of the program.
-The system needs only to convey to clients (e.g.
-via secondary notation) the inferred \emph{segmentation} of each literal, which gives the locations of the spliced terms.
-The inferred segmentation also informs the hygiene mechanism (previously, only unhygienic macros could freely parse terms out of literal bodies.)
+The system only needs to convey to clients (via secondary notation) the inferred \emph{segmentation} of each literal, which gives the locations of the spliced terms.
+The inferred segmentation also informs the hygiene mechanism (previously, only unhygienic macros could freely parse terms out of string literal bodies.)
 %The system can determine this {segmentation} automatically and convey it to the client via syntax highlighting, or by presenting the client with a context-free grammar (without revealing the semantic actions associated with each production.)
@@
 We establish these abstract reasoning principles formally with a calculus of expressions, patterns and ML-style modules.
-This is the  first detailed type-theoretic account of a typed, hygienic term-rewriting macro system for an ML-like language.
+This is the  first detailed type-theoretic account of a typed, hygienic term-rewriting macro system of any design for an ML- or Scala-like language.
 % This calculus is the first such formalization of a typed macro system.
@@
 \label{sec:intro}
-The surface syntax of a programming language is a user interface.
-As such, language designers often include in the surface syntax various \emph{literal forms}  that are designed to decrease the cognitive cost of constructing and pattern matching over values of certain otherwise ordinary types from the language's standard library.
-For example, in Standard ML (SML) and its cousins, the list expression literal \li{[x, y, z]} desugars to a term equivalent to \li{Cons(x, Cons(y, Cons(z, Nil)))} where \li{Nil} and \li{Cons} identify the constructors of the standard \li{list} datatype \cite{mthm97-for-dart,harper1997programming}.\footnote{In SML, list literals work even in contexts where the list expression constructors have been shadowed by other bindings, i.e.
-the expansion of literal forms is \emph{context independent}.
+% The surface syntax of a programming language is its user interface. 
+When designing the surface syntax of a general-purpose programming language, it is common practice to build in a few \emph{literal forms} that lower the syntactic cost of constructing and pattern matching over values of certain standard types.
+For example, in Standard ML (SML) and its cousins, the list expression literal \li{[x, y, z]} is equivalent by desugaring to \li{Cons(x, Cons(y, Cons(z, Nil)))} where \li{Nil} and \li{Cons} identify the constructors of the standard \li{list} datatype \cite{mthm97-for-dart,harper1997programming}.\footnote{In SML, list literals work even in contexts where the list expression constructors have been shadowed by other bindings, i.e.
+the meaning of literal forms is \emph{context independent}.
 We will return to the concept of context independence again.} List pattern literals are analagous.
-Another example is Ur/Web, which extends Ur's textual syntax with expression and pattern literals for encodings of XHTML data, as defined in Ur/Web's standard library \cite{conf/popl/Chlipala15,conf/pldi/Chlipala10}.
+Another example is found in Ur/Web's surface syntax, which extends Ur's surface syntax with expression and pattern literal forms for encodings of XHTML data \cite{conf/popl/Chlipala15,conf/pldi/Chlipala10}.
 Fig.
-\ref{fig:urweb} shows examples of XHTML literals containing ``spliced'' string expressions (Line 1) and XHTML expressions (Line 2).
+\ref{fig:urweb} gives examples of XHTML literals that ``splice in'' a string expression (Line 1) and another XHTML expression (Line 2).
 %The desugarings of these literal expressions, not shown, are substantially more verbose and less readable to programmers familiar with XHTML.
 
-Third-party library providers are quite justifiably envious of the special syntactic status of these semantically ordinary standard library constructs.
-After all, it is not difficult to find other examples of types for which mathematicians, scientists and programmers have invented specialized notations.
-For example, 1) clients of a ``collections'' library might like not just list literals, but also matrix, finite set and finite map literals following standard mathematical conventions; 2) clients of a ``web programming'' library might like not just XHTML literals but also CSS literals (which Ur/Web does not provide); 4) a compiler writer might like ``quotation'' literals for the terms of the object language and various intermediate languages of interest; and 5) clients of a ``chemistry'' library might like chemical structure literals based on the SMILES standard \cite{anderson1987smiles}.
-\citet{Bravenboer:2007:PIA:1289971.1289975} suggests that literal forms for structured encodings of SQL queries and other query types can reduce the temptation to use string encodings and therefore reduce the prevalence of string injection attacks, which are amongst the most common and serious security vulnerabilities on the web today \cite{owasp2017}.
-Choosing strings over composite data structures for the sake of convenience, a.k.a.
-``stringly-typed programming'', is rather common in practice -- \citet{TSLs} sampled strings from open source projects and found that at least 15\% of them were parseable according to some readily apparent type-specific grammar, e.g.
-for SQL queries, regexes, file system paths, URLs and others.
-Literal forms at these types, with support for splicing, would lower the cost of richer compositional encodings.%Literal forms would  preferable to string encodings because they support compositional construction and decomposition by pattern matching via splicing.
+This design practice is decidedly \emph{ad hoc} -- it is easy to find other examples of  library constructs that are not privileged with literal forms by these languages, but for which mathematicians, scientists or programmers have invented and adopted specialized notation \cite{DBLP:journals/cacm/Iverson80,cajori1928history}.
+Consider, for example, that 1) clients of a ``collections'' library might like not just list literals, but also matrix, finite set and finite map literals following standard mathematical conventions; 2) clients of a ``web programming'' library might like not just XHTML literals but also CSS literals (which Ur/Web does not provide); 4) a compiler writer might like ``quotation'' literals for the terms of the object language and various intermediate languages of interest; and 5) clients of a ``chemistry'' library might like composite chemical structure literals based on the SMILES standard \cite{anderson1987smiles}.
+
+Although concerns about literal notation are easy to dismiss as superficial, the reality is that literal forms, or the absence thereof, have had a substantial influence on the quality of software that is written today.
+For example, \citet{Bravenboer:2007:PIA:1289971.1289975} finds that literal forms for structured encodings of queries, like the SQL query literals found in many languages \cite{LINQ}\todo{find citation}, reduce the temptation to use string encodings of queries and therefore reduce the prevalence of string injection attacks, which are amongst the most common and serious security vulnerabilities on the web today \cite{owasp2017}.
+More generally, evidence suggests that programmers frequently engage in ``stringly-typed programming'', i.e.
+they choose strings instead of composite data structures largely for reasons of syntactic convenience.
+In particular, \citet{TSLs} sampled strings from open source projects and found that about 15\% of them were parseable according to some readily apparent type-specific grammar, e.g.
+for SQL queries, regexes, file system paths, URLs and many other structures.
+Literal forms at these types, with support for splicing, would decrease the syntactic cost of composite encodings, which are easier than flat string encodings to manipulate programmatically and to reason about compositionally.
+%Literal forms would  preferable to string encodings because they support compositional construction and decomposition by pattern matching via splicing.
 
@@
 \caption{XHTML literals with support for splicing at two different types are built primitively into Ur/Web.}
-\vspace{-7px}
+\vspace{-8px}
 \label{fig:urweb}
@@
 
-Language designers cannot, of course, hope to accomodate all such types with primitive literal forms \emph{a priori}, so there has been considerable interest in mechanisms that allow programmers to define new literal forms on their own.
-For example, we will describe a number of syntax extension mechanisms, including Camlp4 \cite{ocaml-manual} and SugarJ/Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, in Sec.
-\ref{sec:existing-approaches}.
-The problem is that these mechanisms quickly become \emph{unreasonable} when ``programming in the large'' \cite{DeRemer76}.
+Of course, it would not be sensible to ask general-purpose language designers to accomodate all known notations with primitive literal forms.
+Instead, there has been considerable interest in mechanisms that allow library providers to define new literal forms on their own.
+For example, general syntax extension mechanisms, including Camlp4 \cite{ocaml-manual}, SugarJ/Sugar* \cite{erdweg2011sugarj,erdweg2013framework} and others that we will return to in Sec.
+\ref{sec:existing-approaches}, can be used to define new literal forms.
+The problem is that these mechanisms reveal themselves to be quite \emph{unreasonable} when ``programming in the large'' \cite{DeRemer76}.
 %, i.e. when developing large programs ``consisting of many small programs (modules), possibly written by different people''
-Consider the perspective of a programmer attempting to comprehend (i.e.
+Consider, for instance, the perspective of a programmer attempting to comprehend (i.e.
 reason about) the program text in Fig.
-\ref{fig:K-dialect}, which is written in a dialect of ML's (or Scala's) syntax where a large number of syntax extensions defining novel literal forms, some unfamiliar to the programmer, are active.
+\ref{fig:K-dialect}, which is written in a dialect of ML's (or Scala's) syntax where a large number of new literal forms, some unfamiliar to the programmer, have been defined using one of these mechanisms.
 
@@
 \caption{An example of unreasonable program text.}
-\vspace{-9px}
+\vspace{-10px}
 \label{fig:K-dialect}
@@
 \noindent
-The literal form on Line 3 constructs an encoding of a query in the stack-based database query language K, using its terse textual syntax \cite{Whitney:2001:LOR:376284.375783}.
+The literal form on Line 3 above constructs an encoding of a query in the stack-based database query language K, using its terse surface syntax \cite{Whitney:2001:LOR:376284.375783}.
 Of course, the programmer reading this program text may well be unfamiliar with K's query syntax.
-The problem then is that there are no simple, abstract protocols for answering questions like:
-
-\begin{enumerate}
+Unfortunately for the programmer, there are no simple, abstract protocols for answering questions like:
+
+\begin{enumerate}[nolistsep,noitemsep]
 \item \textbf{(Responsibility)} Which syntax extension determined the desugaring of the literal form on Line 3? Might there be many possible desugarings (i.e. syntactic non-determinism)?
@@
 \item \textbf{(Capture)} If \li{x} is in fact a spliced expression, does it refer to the binding of \li{x} on Line 2, or might it capture an unseen binding of the same identifier used in the desugaring of Line 3?
-\item \textbf{(Context Dependence)} If \li{w}, on Line 1, is renamed, could that possibly break Line 3, or change its meaning because the desugaring of Line 3 assume that some variable identified as \li{w} is in scope (even though \li{w} is not mentioned in the text of Line 3)?
+\item \textbf{(Context Dependence)} If \li{w}, on Lines 1-2, is renamed, could that possibly break Line 3 or change its meaning because the desugaring assumes that some variable identified as \li{w} is in scope?
+What other hidden assumptions might the desugaring be making?
 \item \textbf{(Typing)} What type does \li{y} have?
@@
 Forcing the programmer to reason transparently, i.e.
-to examine the desugaring of the program text, or the logic that produced that desugaring, in order to answer these sorts of basic questions,  can defeat the purpose of syntactic sugar -- decreasing cognitive cost \cite{Green89}.
+to examine the desugaring of the program text, or even the logic that produced that desugaring, defeats the ultimate purpose of syntactic sugar: decreasing cognitive cost \cite{Green89}.
 Analagous problems do not arise when programming without syntax extensions in languages like ML and Scala -- the binding discipline of the language ensures that programmers can reason lexically about binding, and types mediate abstraction over function and module implementations.
-%From this perspective, it is unsurprising that the use of Camlp4 has been deprecated in the OCaml system.
-
-This paper details a more reasonable mechanism: \emph{typed literal macros (TLMs)}.
-TLMs give library providers full programmatic control over the parsing and expansion of \emph{generalized literal forms}, which syntactically subsume other literal forms because their bodies are left initially unparsed according to the fixed context-free syntax of the language (i.e.
+% From this perspective, it is unsurprising that the use of Camlp4 has been deprecated in the OCaml system.
+
+This paper formally details a more reasonable mechanism for defining new literal forms: \emph{typed literal macros (TLMs)}.
+TLMs give library providers full programmatic control over the parsing and expansion of \emph{generalized literal forms} \cite{TSLs}, which syntactically subsume other literal forms because their bodies are left initially unparsed according to the fixed context-free syntax of the language (i.e.
 they are initially parsed like raw string literals.) For example, Fig.
@@
 \ref{fig:K-dialect}.
-The applied TLM parses the literal body and generates an expansion during a semantic phase called typed expansion, which generalizes the usual typing phase.
+The applied TLM parses the literal body and generates an expansion statically, during a semantic phase called typed expansion, which generalizes the usual typing phase.
 TLMs come equipped with uniquely powerful abstract reasoning principles.
-We will characterize these reasoning principles precisely as we proceed, but for now, let us develop some intuitions -- the programmer can reason abstractly about Line 3 as follows:
-\begin{enumerate}
+We will characterize these reasoning principles absolutely precisely as we proceed, but for now, let us develop some intuitions: the programmer can reason abstractly about Line 3 as follows:
+\begin{enumerate}[noitemsep,nolistsep]
 \item \textbf{(Responsibility)} The applied TLM, \li{$kquery}, is solely responsible for parsing and expansion of the literal body.
@@
 \item \textbf{(Typing)} An explicit type annotation on the definition of \li{$kquery} determines the type that the expansion must have.
-We will see an example of a TLM definition in the next section.
-Moreover, each spliced segment in the segmentation also comes paired with the type it is expected to have.
+(We will see an example of a TLM definition in the next section.) Moreover, each spliced segment in the segmentation also comes paired with the type it is expected to have.
 This information is usually not necessary to reason about typing, but it can be conveyed to the programmer upon request by the program editor if desired.
@@
 \begin{figure}
-\vspace{-4px}
+\vspace{-2px}
 \begin{lstlisting}
@@
 % \item 
-Sec. \ref{sec:setsms} defines a {simple expression TLM} for XHTML, which implements Ur/Web's XHTML literal syntax, and describes the typed expansion process in more detail.
+Sec. \ref{sec:setsms} defines a {simple expression TLM} that implements Ur/Web's XHTML literal syntax, and describes the typed expansion process in more detail.
 % \item 
@@
 % \item 
-Sec. \ref{sec:ptsms} adds type functions and an ML-style module system and introduces \emph{parametric TLMs} (pTLMs), i.e. TLMs that take type and module parameters. Parameters serve two purposes:
-\begin{enumerate}[noitemsep,nolistsep]
-\item They allow a TLM to operate not just at a single specified type, but over a type- and module-parameterized family of types.
+Sec.
+\ref{sec:ptsms} then adds type functions and an ML-style module system and introduces \emph{parametric TLMs} (pTLMs), i.e.
+TLMs that take type and module parameters.
+Parameters serve two purposes.
+First, they allow a TLM to operate not just at a single specified type, but over a type- and module-parameterized family of types.
 We describe a parametric TLM for string-keyed dictionaries that operates uniformly over different implementations of the abstract dictionary signature, and also over different choices of the range type.
-\item Parameters also give expansions access to terms other than those explicitly spliced in to the literal body by the client.
+Second, parameters give expansions access to terms other than those explicitly spliced in to the literal body by the client.
 %This is particularly convenient for working with abstract types \emph{a la} ML.
 Partial parameter application is possible, which lowers the syntactic cost of the strict context independence discipline that TLMs enforce.
-\end{enumerate}
-This section culminates with a calculus that extends the calculus of Sec. \ref{sec:setsms-formally} with type functions, ML-style modules and parametric TLMs.
+% Sec.
+\ref{sec:ptsms} culminates with a calculus that extends the calculus of Sec.
+\ref{sec:setsms-formally} with type functions, ML-style modules and parametric TLMs.
 % \item
@@
 % \item 
-Sec. \ref{sec:existing-approaches} compares TLMs to related work, including closely related work by \citet{TSLs,sac15} on type-specific languages, and the long history of work on macro systems.
+Sec. \ref{sec:existing-approaches} compares TLMs to related work, including recent related work by \citet{TSLs,sac15}, and the long history of work on macro systems.
 % \item 
-Sec. \ref{sec:discussion} concludes with a summary of the specific contributions of this paper and a brief discussion of the present limitations and some future research directions.
+Sec. \ref{sec:discussion} concludes with a summary of the specific contributions of this paper and a brief discussion of present limitations and future work.
 % \end{itemize}
 
-Due to space limitations, many standard or straightforward technical details of the calculi defined in Sec.
+Due to space limitations, many technical details of the calculi defined in Sec.
 \ref{sec:setsms-formally} through Sec.
-\ref{sec:static-eval} are relegated to the supplemental material.
-The examples that we give are written in an emerging dialect of ML called VerseML that supports TLMs as described herein, but the essence of the mechanism is not ML-specific -- it should be straightforward to adapt it to other languages that take a similar approach to types and binding (e.g.
+\ref{sec:static-eval}, as well as proofs, are relegated to the supplemental material.
+Examples are written in an emerging dialect of  ML called VerseML that is focused on making (Standard) ML more ergonomic.
+However, the essence of the mechanism is not ML-specific -- it should be straightforward to adapt it to other languages that take a similar approach to types and binding (e.g.
 Haskell, Scala and others.) When a design choice is ML-specific, we will make note of it\todo{do this}.
 Languages that have a disciplined binding structure but that lack a rich static type structure, e.g.
-Racket and other LISP derivatives, may also benefit from TLMs -- they can apply the results of this paper by deploying the usual type theorist's trick of viewing such a language as ``{unityped}'' \cite{pfpl,scott1980lambda}.
+Racket and other LISP derivatives, would also benefit from TLMs -- they can apply the results of this paper by deploying the usual type theorist's trick of viewing such a language as ``{unityped}'' \cite{pfpl,scott1980lambda}.
 % For the present purposes, we can consider VerseML as essentially the calculus defined in the supplement extended with various conveniences that are commonly found in other ML-like languages and, notionally, orthogonal to TLMs.
@@
 
-We will begin in this section by introducing \emph{simple expression TLMs} (seTLMs), i.e. TLMs where expansions are expressions of a single specified type, by example.
+We will begin in this section with another example that informally introduces \emph{simple expression TLMs} (seTLMs), i.e. TLMs for generating expressions of a single specified type.
 
@@
  to \emph{generalized literal forms} delimited by backticks. TLM names are prefixed by \li{$} 
- to clearly them from variables.
+ to clearly them from variable identifiers.
 The semantics delegates control over the parsing and expansion of each literal body to the applied TLM during a semantic phase called \emph{typed expansion}, which generalizes the usual typing phase.
@@
 
-The importance of good notation as a ``tool for thought'' has long been recognized\todo{cite iverson}.
-A good notation, according to A.
-N.
-Whitehead, ``relieves the brain of [...] unnecessary work'' and ``sets it free to concentrate on more advanced problems''\todo{citation}.
-Indeed, advances in mathematics, science and software engineering have quite often been accompanied by new notation.
-Of course, this desire to ``relieve the brain of all unnecessary work'' has influenced not just the surface syntax but also the semantics of languages like ML and Scala, which maintain a strong type and binding discipline so that programmers, and their tools, can hold certain ``implementation details'' abstract when reasoning about program behavior.
-
-Previously, these  were in tension -- .
-
-Not all notation is good, and we make no empirical claims as to the value of any particular notation in this paper.
-TLMs leave essentially all choices about literal notation, and therefore the burden of establishing the value of any particular literal notation, to individual library providers, rather than to the language designer.
-This decentralization of control is justified by the fact that when the programmer encounters an unfamiliar or bad notation, it does not ruin their day.
-The programmer is in no worse a position...
-
-The broader empirical question of whether the capabilities that TLMs provide are of interest to programmers ``in the wild'' is well beyond the scope of a conference paper such as this.
-The intended audience for this paper is language designers who wish to decentralize control over literal forms but seek a detailed understanding, rooted in the first principles of type theory, of a reasonable mechanism by which to achieve that goal in the setting of an ML- or Scala-like language.
-This paper is the first such account of such a mechanism.
-More generally, this paper contributes the first detailed type-theoretic account of a typed, hygienic macro system for an ML-like language, i.e.
-one with a rich static type system, support for pattern matching and an ML-like module system with support for type functions and abstract type members.
-(To summarize Sec.
+The importance of specialized notation as a ``tool for thought'' has long been recognized \cite{DBLP:journals/cacm/Iverson80}.
+According to Whitehead, a good notation ``relieves the brain of [...] unnecessary work'' and ``sets it free to concentrate on more advanced problems'' \cite{cajori1928history}.
+And indeed, advances in mathematics, science and programming have often been accompanied by new notation.
+
+Of course, this desire to ``relieve the brain of unnecessary work'' has motivated not only the syntax but also the semantics of languages like ML and Scala -- these languages maintain a strong type and binding discipline so that programmers, and their tools, can hold certain implementation details abstract when reasoning about program behavior.
+In the words of \citet{B304}, ``type structure is a syntactic discipline for enforcing levels of abstraction.''
+
+Previously, these two relief mechanisms were in tension -- mechanisms  that allowed programmers to express new notation would obscure the type and binding structure of program text, as discussed in Sec.
+\ref{sec:intro} and detailed in Sec.
+\ref{sec:existing-approaches}.
+TLMs resolve this tension for the broad class of literal forms that generalized literal forms subsume.
+This class includes all of the examples enumerated in Sec.
+\ref{sec:intro} (up to the choice of outermost delimiter), the additional examples described in Sec.
+\ref{sec:static-eval}, the more detailed example in the supplement\todo{do this}, and the many examples in the prior papers by \citet{TSLs,sac15}.
+
+Of course, not all possible literal notation will prove to be in good taste, and we make no empirical claims about particular notation in this paper.
+TLMs leave the burden of establishing the value of any particular literal notation to individual library providers, rather than to the language designer.
+The reasoning principles that TLMs provide, which are the specific contributions of this paper, allow clients to "reason around" poor literal designs, using principles analagous to those already familiar to programmers in languages like these.
+% We must leave a detailed empirical evaluation of the impact of particular TLMs on various quality attributes, like programmer productivity and program comprehensibility, as future work.
+The technical content of this paper, which formally establishes these reasoning principles, constitutes the first detailed type-theoretic account of a typed, hygienic macro system for an ML-like language, i.e.
+one with a rich static type system, support for pattern matching and an ML-like module system with type functions and abstract type members.
+Our own language integration efforts have been focused on an emerging ML dialect called VerseML, but Scala supports analagous features \cite{conf/oopsla/AminRO14} and would also be a suitable host for the TLM mechanism.
+We also plan to implement TLMs into \li{typy}, a typed functional language embedded into Python as a library \cite{gpce/Omar16}.%We have evaluated this claim by stating the available abstract reasoning principles formally and proving that they hold for the calculi that we have described.
+
+% The intended audience for this paper is language designers who want to decentralize control over literal forms but seek a detailed understanding, rooted in the first principles of type theory, of a reasonable mechanism by which to achieve that goal.
+
+%(To summarize Sec.
 \ref{sec:existing-approaches}, the closest existing typed, hygienic macro system -- the Scala macro system \cite{ScalaMacros2013} -- is not formally specified, the work of \cite{DBLP:conf/esop/HermanW08} uses types only to reason abstractly about binding, and the specification of TSLs in the work of \citet{TSLs} did not formalize hygiene nor handle these advanced language features.)
 
-It is trivial to observe that the literal forms already chosen to appear in various languages, including the examples in Sec. \ref{sec:intro},
-
-One important avenue has to do with automated refactoring.
+Haskell's \li{do}-notation at types with monadic structure cannot be expressed directly using TLMs because it needs to purposefully induce capture in spliced sub-terms, and TLMs prevent capture.
+This limitation could be resolved by allowing identifiers that appear in literal bodies to be explicitly designated as bound within spliced expressions in the segmentation.
+The prior work by \citet{DBLP:conf/esop/HermanW08} considered a similar approach in a setting where the locations of sub-terms is known ahead of time (building on the \emph{tree locations} of \citet{gorn1965explicit}, which are conceptually similar to our spliced segment locations.)  We leave as future work the details of this proposal and a consideration of whether this additional expressive power would justify the increased reasoning complexity for clients (it is presently the author's opinion that it would not.)
+
+A correct parse function never returns an encoding of a proto-expansion that fails validation. This invariant cannot be enforced by the simple ML-style type system we
+have been working with. Using a proof assistant, it would be possible to verify that
+a parse function generates only encodings of valid proto-expansions\todo{cite something here}. Alternatively, in a
+dependently typed setting, the type of the parse function itself could be enriched so as to
+enforce this invariant intrinsically. We leave the details of this approach as future work.
+
+%Another topic of interest has to do with intentional capture. 
+Another important direction has to do with automated refactoring.
 The unexpanded language does not come equipped with standard notions of renaming and substitution.
-Given a splice summary, it should be possible to ``backpatch'' refactorings into literal bodies, but the details are left as future work.
-
-%Another topic of interest has to do with intentional capture. 
-Haskell's \li{do}-notation for types with monadic structure cannot be expressed directly using TLMs because it introduces bindings into sub-terms.
-This might be an acceptable trade-off (indeed, that is our present position.) Alternatively, it might be worth exploring a design where spliced identifiers (but not arbitrary identifiers) can be designated in the splice summary as bound within spliced terms.
-
-At several points in the paper, we discussed editor integration.
-However, some important questions having to do with incremental parsing, error recovery and error reporting remain to be considered.
-Proto-expansions are context-independent and parametric, so they are likely to amenable to caching.
-
-Other directions for future work are given in the supplement...
+Given a splice summary, it should be possible to ``backpatch'' refactorings into literal bodies, but the details are left as future work\todo{cite some papers}.
+
+At several points in the paper, we allude to editor integration.
+However, some important questions having to do with TLM-specific syntax highlighting, incremental parsing and error recovery\todo{citation} remain to be considered.
+Another interesting direction would be to generalize TLMs to support non-textual notation in the setting of a structure editor.
+The semantic foundations for structure editors proposed recently by \citet{DBLP:conf/popl/OmarVHAH17} may guide such an effort.
+
+Other directions for future work are given in the supplement\todo{do this?}.
 

### 2017-07-01 popl18/syntax-popl18.tex
more tweaks to front matter

@@
 %% Title information
-\title[Reasonably Programmable Syntax]{Reasonably Programmable Syntax}         %% [Short Title] is optional;
+\title[Reasonably Programmable Literal Forms]{Reasonably Programmable Literal Forms}         %% [Short Title] is optional;
                                         %% when present, will be used in
@@
 This paper questions this common practice: why should an otherwise ordinary data structure be privileged in the definition of a general-purpose programming language?
-When designing the surface syntax of a programming language, it is common practice to build in literal forms for a few standard data structures, e.g.
-lists.
-This practice is \emph{ad hoc} -- there are an unbounded variety of other data structures for which literal forms might also be useful (consider, for example,  maps, regexes, HTML data, SQL queries, syntax trees and chemical structures.) %A less \emph{ad hoc} design would be one that allows library providers to define new literal forms in a decentralized manner.
+General-purpose programming languages typically provide literal notation for only a small number of standard data structures, like lists.
+This language design practice is decidedly \emph{ad hoc}: there are many more data structures for which literal notation might also be useful (consider, for example,  maps, regexes, HTML data, SQL queries, syntax trees and chemical structures.) Availability of literal notation is far from a frivolous concern: studies suggest that literal notation can have a non-trivial impact on software quality by removing the temptation to use string encodings, which are vectors for injection attacks and cause other difficulties.
+%A less \emph{ad hoc} design would be one that allows library providers to define new literal forms in a decentralized manner.
 However, the mechanisms available to programmers today, e.g.
@@
 the Scala macro system \cite{ScalaMacros2013}, are only somewhat more reasonable, and do not offer direct syntactic control.
-This paper introduces a system of \emph{typed literal macros (TLMs)} that gives library providers full programmatic control over the parsing and expansion of expressions and patterns of \emph{generalized literal form} at \emph{any} specified type or parameterized family of types.
+This paper details a mechanism for decentralizing control over literal notation: \emph{typed literal macros (TLMs)}.
+TLMs give library providers programmatic control over the parsing and expansion of expressions and patterns of \emph{generalized literal form} at {a} specified type or parameteric family of types.
 % The mechanism is {strictly hygienic}, meaning that 1) the expansion must be context-independent; and 2)  splicing is capture avoiding.
@@
 Compared to existing approaches, including syntax extension systems and macro-based quotation systems, TLMs are uniquely \emph{reasonable}: TLM providers can reason modularly about syntactic determinism, and TLM clients can reason abstractly about the type and binding structure of the program.
-The system only needs to convey to clients (via secondary notation) the inferred \emph{segmentation} of each literal, which gives the locations of the spliced terms.
-The inferred segmentation also informs the hygiene mechanism (previously, only unhygienic macros could freely parse terms out of string literal bodies.)
+The system only needs to convey to clients (via secondary notation) the inferred \emph{segmentation} of each literal, which gives the locations of spliced terms.
+Maintaining information about segmentation supports the novel TLM hygiene mechanism, i.e.
+the mechanism that ensures that clients can reason abstractly about binding (previously, only unhygienic macros could freely splice terms out of string literal bodies.)
 %The system can determine this {segmentation} automatically and convey it to the client via syntax highlighting, or by presenting the client with a context-free grammar (without revealing the semantic actions associated with each production.)
 % This mechanism captures many common syntactic idioms while avoiding the problem of syntactic conflicts by construction and supplying clients with clear abstract reasoning principles. 
-We establish these abstract reasoning principles formally with a calculus of expressions, patterns and ML-style modules.
-This is the  first detailed type-theoretic account of a typed, hygienic term-rewriting macro system of any design for an ML- or Scala-like language.
+This paper establishes these abstract reasoning principles formally with a calculus of expressions, patterns and ML-style modules.
+This calculus is the  first detailed type-theoretic account of a typed, hygienic term-rewriting macro system of any design for a language with these features of ML or Scala.
 % This calculus is the first such formalization of a typed macro system.
@@
 % The surface syntax of a programming language is its user interface. 
-When designing the surface syntax of a general-purpose programming language, it is common practice to build in a few \emph{literal forms} that lower the syntactic cost of constructing and pattern matching over values of certain standard types.
-For example, in Standard ML (SML) and its cousins, the list expression literal \li{[x, y, z]} is equivalent by desugaring to \li{Cons(x, Cons(y, Cons(z, Nil)))} where \li{Nil} and \li{Cons} identify the constructors of the standard \li{list} datatype \cite{mthm97-for-dart,harper1997programming}.\footnote{In SML, list literals work even in contexts where the list expression constructors have been shadowed by other bindings, i.e.
+When designing the surface syntax of a general-purpose programming language, it is common practice to build in a few \emph{literal forms} that decrease the syntactic cost of constructing and pattern matching over values of certain standard types.
+For example, in Standard ML (SML) and its cousins, a list expression literal like \li{[x, y, z]} is equivalent by desugaring to \li{Cons(x, Cons(y, Cons(z, Nil)))} where \li{Nil} and \li{Cons} identify the constructors of the standard \li{list} datatype \cite{mthm97-for-dart,harper1997programming}.\footnote{In SML, list literals work even in contexts where the list expression constructors have been shadowed by other bindings, i.e.
 the meaning of literal forms is \emph{context independent}.
-We will return to the concept of context independence again.} List pattern literals are analagous.
-Another example is found in Ur/Web's surface syntax, which extends Ur's surface syntax with expression and pattern literal forms for encodings of XHTML data \cite{conf/popl/Chlipala15,conf/pldi/Chlipala10}.
+We will return to the concept of context independence again.} Another example is found in Ur/Web's surface syntax, which extends Ur's surface syntax with expression and pattern literal forms for encodings of HTML data \cite{conf/popl/Chlipala15,conf/pldi/Chlipala10}.
 Fig.
-\ref{fig:urweb} gives examples of XHTML literals that ``splice in'' a string expression (Line 1) and another XHTML expression (Line 2).
+\ref{fig:urweb} shows two such HTML literals, one that ``splices in'' a string expression (Line 1) and the other an HTML expression (Line 2).
 %The desugarings of these literal expressions, not shown, are substantially more verbose and less readable to programmers familiar with XHTML.
 
-This design practice is decidedly \emph{ad hoc} -- it is easy to find other examples of  library constructs that are not privileged with literal forms by these languages, but for which mathematicians, scientists or programmers have invented and adopted specialized notation \cite{DBLP:journals/cacm/Iverson80,cajori1928history}.
-Consider, for example, that 1) clients of a ``collections'' library might like not just list literals, but also matrix, finite set and finite map literals following standard mathematical conventions; 2) clients of a ``web programming'' library might like not just XHTML literals but also CSS literals (which Ur/Web does not provide); 4) a compiler writer might like ``quotation'' literals for the terms of the object language and various intermediate languages of interest; and 5) clients of a ``chemistry'' library might like composite chemical structure literals based on the SMILES standard \cite{anderson1987smiles}.
-
-Although concerns about literal notation are easy to dismiss as superficial, the reality is that literal forms, or the absence thereof, have had a substantial influence on the quality of software that is written today.
-For example, \citet{Bravenboer:2007:PIA:1289971.1289975} finds that literal forms for structured encodings of queries, like the SQL query literals found in many languages \cite{LINQ}\todo{find citation}, reduce the temptation to use string encodings of queries and therefore reduce the prevalence of string injection attacks, which are amongst the most common and serious security vulnerabilities on the web today \cite{owasp2017}.
+This language design practice is decidedly \emph{ad hoc} -- it is easy to come up with other examples of data structures that are not privileged with literal forms in these languages, but for which mathematicians, scientists or programmers have elsewhere invented specialized notation \cite{DBLP:journals/cacm/Iverson80,cajori1928history}.
+Consider, for example, that 1) clients of a ``collections'' library might like not just list literals, but also matrix, set and map literals following standard mathematical conventions; 2) clients of a ``web programming'' library might like not just HTML literals but also CSS literals (which Ur/Web does not provide); 4) a compiler writer might like ``quotation'' literals for the terms of the object language and various intermediate languages of interest; and 5) clients of a ``chemistry'' library might like composite chemical structure literals based on the SMILES standard \cite{anderson1987smiles}.
+
+Although concerns about ``missing'' literal notation are easy to dismiss as superficial or frivolous, the reality is that literal forms, or the absence thereof, have had a substantial influence on the quality of software that is written today.
+For example, \citet{Bravenboer:2007:PIA:1289971.1289975} finds that literal forms for structured encodings of queries, like the SQL-based query literals now found in many languages \cite{LINQ}\todo{find citation}, reduce the temptation to use string encodings of queries and therefore reduce the prevalence of string injection attacks, which are amongst the most common and serious security vulnerabilities on the web today \cite{owasp2017}.
 More generally, evidence suggests that programmers frequently engage in ``stringly-typed programming'', i.e.
@@
 In particular, \citet{TSLs} sampled strings from open source projects and found that about 15\% of them were parseable according to some readily apparent type-specific grammar, e.g.
-for SQL queries, regexes, file system paths, URLs and many other structures.
-Literal forms at these types, with support for splicing, would decrease the syntactic cost of composite encodings, which are easier than flat string encodings to manipulate programmatically and to reason about compositionally.
+for SQL queries, regexes, file system paths, URLs and many other data structures.
+Literal forms at these types, with support for splicing, would decrease the syntactic cost of composite encodings, which are more amenable to programmatic manipulation and compositional reasoning than flat string encodings.
 %Literal forms would  preferable to string encodings because they support compositional construction and decomposition by pattern matching via splicing.
@@
 \vspace{-10px}
-\caption{XHTML literals with support for splicing at two different types are built primitively into Ur/Web.}
+\caption{HTML literals with support for splicing at two different types are built primitively into Ur/Web.}
 \vspace{-8px}
@@
 
-Of course, it would not be sensible to ask general-purpose language designers to accomodate all known notations with primitive literal forms.
-Instead, there has been considerable interest in mechanisms that allow library providers to define new literal forms on their own.
+Of course, it would not be sensible to ask general-purpose language designers to accomodate all known notations with built-in literal forms.
+Instead, there has been considerable interest in mechanisms that allow library providers to define new literal notation on their own.
 For example, general syntax extension mechanisms, including Camlp4 \cite{ocaml-manual}, SugarJ/Sugar* \cite{erdweg2011sugarj,erdweg2013framework} and others that we will return to in Sec.
 \ref{sec:existing-approaches}, can be used to define new literal forms.
-The problem is that these mechanisms reveal themselves to be quite \emph{unreasonable} when ``programming in the large'' \cite{DeRemer76}.
+The problem that specifically motivates this paper is that these mechanisms are quite \emph{unreasonable} when ``programming in the large'' \cite{DeRemer76}.
 %, i.e. when developing large programs ``consisting of many small programs (modules), possibly written by different people''
@@
 reason about) the program text in Fig.
-\ref{fig:K-dialect}, which is written in a dialect of ML's (or Scala's) syntax where a large number of new literal forms, some unfamiliar to the programmer, have been defined using one of these mechanisms.
+\ref{fig:K-dialect}, which is written in a dialect of ML's (or Scala's) syntax, constructed by one of these mechanisms, where a large number of new literal forms have been defined.
 
@@
 Forcing the programmer to reason transparently, i.e.
-to examine the desugaring of the program text, or even the logic that produced that desugaring, defeats the ultimate purpose of syntactic sugar: decreasing cognitive cost \cite{Green89}.
+to examine the desugaring of the program text, or even the logic that produced that desugaring, to answer basic questions like these defeats the ultimate purpose of syntactic sugar: decreasing cognitive cost \cite{Green89}.
 Analagous problems do not arise when programming without syntax extensions in languages like ML and Scala -- the binding discipline of the language ensures that programmers can reason lexically about binding, and types mediate abstraction over function and module implementations.
@@
 \ref{fig:K-tsm-example} applies a TLM, \li{$kquery}, to a backtick-delimited generalized literal form to express the ``unreasonable'' K query example from Fig.
-\ref{fig:K-dialect}.
+\ref{fig:K-dialect} just discussed.
 The applied TLM parses the literal body and generates an expansion statically, during a semantic phase called typed expansion, which generalizes the usual typing phase.
 TLMs come equipped with uniquely powerful abstract reasoning principles.
-We will characterize these reasoning principles absolutely precisely as we proceed, but for now, let us develop some intuitions: the programmer can reason abstractly about Line 3 as follows:
+We will characterize these reasoning principles precisely as we proceed, but for now, let us develop some intuitions: the programmer can reason abstractly about Line 3 as follows:
 \begin{enumerate}[noitemsep,nolistsep]
@@
 By examining Fig.
-\ref{fig:K-tsm-example}, we know that the two instances of \li{x} on Line 3 are parsed as spliced expressions (because they are in black), whereas the \li{R}'s must be parsed in some other way (because they are in green.)
-\item \textbf{(Capture)} Splicing is capture-avoiding, so the spliced expressions \li{x} must refer to the binding of \li{x} on Line 2.
-They cannot capture a coincidental binding of \li{x} in the expansion.
+\ref{fig:K-tsm-example}, we know that the two instances of \li{x} on Line 3 are parsed as spliced expressions (because they are in black), whereas the \li{R}'s must be parsed in some other way, e.g.
+as operators in the K query language (because they are in green.)
+\item \textbf{(Capture)} Splicing is capture-avoiding, so the spliced expression \li{x} must refer to the binding of \li{x} on Line 2.
+It cannot capture a coincidental binding of \li{x} in the expansion.
 \item \textbf{(Context Dependence)} The system enforces context independence, so the expansion of Line 3 cannot  rely on the fact that, e.g., \li{w} is in scope. The client is free to rename \li{w}.
 \item \textbf{(Typing)} An explicit type annotation on the definition of \li{$kquery} determines the type that the expansion must have.
-(We will see an example of a TLM definition in the next section.) Moreover, each spliced segment in the segmentation also comes paired with the type it is expected to have.
+We will see an example of a TLM definition in the next section.
+Moreover, each spliced segment in the segmentation also comes paired with the type it is expected to have.
 This information is usually not necessary to reason about typing, but it can be conveyed to the programmer upon request by the program editor if desired.
@@
 % \item 
-Sec. \ref{sec:setsms} defines a {simple expression TLM} that implements Ur/Web's XHTML literal syntax, and describes the typed expansion process in more detail.
+Sec.
+\ref{sec:setsms} defines another {simple expression TLM} -- one that implements Ur/Web-style HTML literal syntax (Fig.
+\ref{fig:urweb}) -- and uses it as a running example to informally detail how TLMs work from start to finish.
 % \item 
-Sec. \ref{sec:sptsms} briefly introduces {simple pattern TLMs}, i.e. TLMs that generate patterns. %The hygiene condition for pattern TLMs is interesting.
+Sec. \ref{sec:sptsms} briefly introduces {pattern TLMs}, i.e. TLMs that generate patterns. %The hygiene condition for pattern TLMs is interesting.
 % \item 
@@
 TLMs that take type and module parameters.
-Parameters serve two purposes.
-First, they allow a TLM to operate not just at a single specified type, but over a type- and module-parameterized family of types.
-We describe a parametric TLM for string-keyed dictionaries that operates uniformly over different implementations of the abstract dictionary signature, and also over different choices of the range type.
+Parameters serve two key purposes.
+First, they allow a TLM to operate not just at a single type, but over a type- and module-parameterized family of types.
+We give as an example a parametric TLM for string-keyed dictionaries that operates uniformly over different implementations of the abstract dictionary signature, and also over different choices of the range type.
 Second, parameters give expansions access to terms other than those explicitly spliced in to the literal body by the client.
 %This is particularly convenient for working with abstract types \emph{a la} ML.
-Partial parameter application is possible, which lowers the syntactic cost of the strict context independence discipline that TLMs enforce.
+Partial parameter application is possible.
+Together, this lowers the syntactic cost of the strict context independence discipline that TLMs enforce.
 % Sec.
@@
 % \item 
-Sec. \ref{sec:existing-approaches} compares TLMs to related work, including recent related work by \citet{TSLs,sac15}, and the long history of work on macro systems.
+Sec. \ref{sec:existing-approaches} compares TLMs to related work, including recent related work by \citet{TSLs,sac15}, and the long history of related work on macro systems.
 % \item 
-Sec. \ref{sec:discussion} concludes with a summary of the specific contributions of this paper and a brief discussion of present limitations and future work.
+Sec. \ref{sec:discussion} concludes with a summary of the specific contributions of this paper and a brief discussion of the limitations of TLMs, along with some suggestions for future work.
 % \end{itemize}
 
-Due to space limitations, many technical details of the calculi defined in Sec.
-\ref{sec:setsms-formally} through Sec.
-\ref{sec:static-eval}, as well as proofs, are relegated to the supplemental material.
-Examples are written in an emerging dialect of  ML called VerseML that is focused on making (Standard) ML more ergonomic.
-However, the essence of the mechanism is not ML-specific -- it should be straightforward to adapt it to other languages that take a similar approach to types and binding (e.g.
+Due to space limitations, certain technical details of the calculi defined in this paper, as well as proofs, are relegated to the supplemental material.
+The examples in this paper are written in an emerging dialect of  ML called VerseML that is focused on improving the ergonomics of ML.
+However, the essential ideas are not ML-specific -- it should be possible to adapt TLMs to other languages that take a similar approach to types and binding (e.g.
 Haskell, Scala and others.) When a design choice is ML-specific, we will make note of it\todo{do this}.
 Languages that have a disciplined binding structure but that lack a rich static type structure, e.g.
-Racket and other LISP derivatives, would also benefit from TLMs -- they can apply the results of this paper by deploying the usual type theorist's trick of viewing such a language as ``{unityped}'' \cite{pfpl,scott1980lambda}.
-% For the present purposes, we can consider VerseML as essentially the calculus defined in the supplement extended with various conveniences that are commonly found in other ML-like languages and, notionally, orthogonal to TLMs.
+Racket and other LISP derivatives, might also benefit from TLMs -- they can apply the results of this paper by deploying the usual trick of reducing the language to one that is ``{unityped}'' \cite{pfpl,scott1980lambda}.%  We say more about LISP macros in Sec.
+\ref{sec:existing-approaches}.% For the present purposes, we can consider VerseML as essentially the calculus defined in the supplement extended with various conveniences that are commonly found in other ML-like languages and, notionally, orthogonal to TLMs.
 %VerseML is, as its name suggests, a conceptual descendent of ML.
@@
 
-We will begin in this section with another example that informally introduces \emph{simple expression TLMs} (seTLMs), i.e. TLMs for generating expressions of a single specified type.
-
-Our take on the running example is shown in Fig. \ref{fig:first-tsm-example}: we apply a TLM named 
+This section informally details the TLM mechanism.
+It will suffice in this section to consider only \emph{simple expression TLMs (seTLMs)}, i.e.
+TLMs for abbreviating expressions of a single specified type.
+In particular, this section is organized around an seTLM named \li{$html} that implements Ur/Web-style HTML literals, which abbreviate expressions of the type \li{html} defined as follows:
+\begin{lstlisting}[numbers=none]
+  type html = BodyElement of list(html_attr) * list(html)
+            | H1Element of list(html_attr) * list(html) 
+            | (* ... *) | TextNode of string 
+\end{lstlisting}
+The type \li{html_attr} is straightforward and the type function \li{list} is standard, so neither are shown.
+In VerseML, \li{string}s are primitive (though this is not essential, see footnote \todo{reference}1 below.)
+
+Using this TLM, we can express the Ur/Web program from Fig. \ref{fig:urweb} as shown in Fig. \ref{fig:first-tsm-example}: on each line, we apply 
 \li{$html} 
- to \emph{generalized literal forms} delimited by backticks. TLM names are prefixed by \li{$} 
+ to a \emph{generalized literal form} delimited by backticks. TLM names are prefixed by \li{$} 
  to clearly them from variable identifiers.

### 2017-07-08 popl18/syntax-popl18.tex
POPL18 submission in, o my

@@
 \usepackage{bbm}
+\definecolor{light-gray}{gray}{0.95}
 
@@
 mathescape=false,escapechar=~,
-numbers=left, numberstyle=\scriptsize\color{mygray}, language=ML,moredelim=[il][\sffamily]{?},showspaces=false,showstringspaces=false,xleftmargin=0pt, numbersep=3pt, morekeywords=[1]{spliced,tyfam,opfam,let,fn,val,def,casetype,objtype,metadata,of,*,datatype,new,toast,syntax,module,where,import,for,ana,syn,opcon,tycon,metasignature,metamodule,metasig,metamod,static,at,by,tycase,mod,macro,match,pattern,in,patterns,expressions,implicit,forall,exptsm,pattsm},deletekeywords={double},classoffset=0,
+numbers=left, numberstyle=\scriptsize\color{mygray}, language=ML,moredelim=[il][\sffamily]{?},showspaces=false,showstringspaces=false,xleftmargin=0pt, numbersep=3pt, morekeywords=[1]{spliced,tyfam,opfam,let,fn,val,def,casetype,objtype,metadata,of,*,string,datatype,new,toast,syntax,module,switch,where,using,import,for,ana,syn,opcon,tycon,metasignature,metamodule,metasig,metamod,static,at,by,tycase,mod,macro,match,pattern,in,patterns,expressions,implicit,forall,exptsm,pattsm},deletekeywords={double},classoffset=0,
 aboveskip=4pt,belowskip=2pt,
@@
 }
+\lstset{morecomment=[n]{/*}{*/}}
+
 \newcommand{\liv}[1]{\lstinline{#1}}
@@
 
-\input{../../../thesis/macros}
+\input{macros}
 
@@
 General-purpose programming languages typically provide literal notation for only a small number of standard data structures, like lists.
-This language design practice is decidedly \emph{ad hoc}: there are many more data structures for which literal notation might also be useful (consider, for example,  maps, regexes, HTML data, SQL queries, syntax trees and chemical structures.) Availability of literal notation is far from a frivolous concern: studies suggest that literal notation can have a non-trivial impact on software quality by removing the temptation to use string encodings, which are vectors for injection attacks and cause other difficulties.
+This common design practice is decidedly \emph{ad hoc}: there are many other data structures for which literal notation might also be useful, e.g.
+finite maps, regexes, HTML data, SQL queries, syntax trees and chemical structures.
+Availability of literal notation is far from a superficial concern: studies suggest that literal notation can have a serious impact on software quality by subverting string encodings, which are vectors for injection attacks and cause other difficulties.
 %A less \emph{ad hoc} design would be one that allows library providers to define new literal forms in a decentralized manner.
@@
 the Scala macro system \cite{ScalaMacros2013}, are only somewhat more reasonable, and do not offer direct syntactic control.
-This paper details a mechanism for decentralizing control over literal notation: \emph{typed literal macros (TLMs)}.
-TLMs give library providers programmatic control over the parsing and expansion of expressions and patterns of \emph{generalized literal form} at {a} specified type or parameteric family of types.
+This paper develops, from type-theoretic first principles, a mechanism that decentralizes control over literal notation: \emph{typed literal macros (TLMs)}.
+TLMs give library providers programmatic control over the parsing and expansion of expressions and patterns of a flexible \emph{generalized literal form} at {a} specified type or parameteric family of types.
 % The mechanism is {strictly hygienic}, meaning that 1) the expansion must be context-independent; and 2)  splicing is capture avoiding.
@@
 The system only needs to convey to clients (via secondary notation) the inferred \emph{segmentation} of each literal, which gives the locations of spliced terms.
-Maintaining information about segmentation supports the novel TLM hygiene mechanism, i.e.
-the mechanism that ensures that clients can reason abstractly about binding (previously, only unhygienic macros could freely splice terms out of string literal bodies.)
+Maintaining information about segmentation is necessary for the novel TLM hygiene mechanism, i.e.
+the mechanism that ensures that clients can reason abstractly about binding (previously, only unhygienic macro systems supported splicing of terms out of string literal bodies.)
 %The system can determine this {segmentation} automatically and convey it to the client via syntax highlighting, or by presenting the client with a context-free grammar (without revealing the semantic actions associated with each production.)
 % This mechanism captures many common syntactic idioms while avoiding the problem of syntactic conflicts by construction and supplying clients with clear abstract reasoning principles. 
-This paper establishes these abstract reasoning principles formally with a calculus of expressions, patterns and ML-style modules.
-This calculus is the  first detailed type-theoretic account of a typed, hygienic term-rewriting macro system of any design for a language with these features of ML or Scala.
-% This calculus is the first such formalization of a typed macro system.
+This paper establishes these abstract reasoning principles formally with a calculus of typed expressions, patterns and modules.
+This calculus is the  first detailed type-theoretic account of a hygienic term-rewriting macro system, of any design, for a language with these essential features of ML.
+We are integrating TLMs into Reason, an emerging alternative front-end for OCaml.% This calculus is the first such formalization of a typed macro system.
 
@@
 % The surface syntax of a programming language is its user interface. 
-When designing the surface syntax of a general-purpose programming language, it is common practice to build in a few \emph{literal forms} that decrease the syntactic cost of constructing and pattern matching over values of certain standard types.
-For example, in Standard ML (SML) and its cousins, a list expression literal like \li{[x, y, z]} is equivalent by desugaring to \li{Cons(x, Cons(y, Cons(z, Nil)))} where \li{Nil} and \li{Cons} identify the constructors of the standard \li{list} datatype \cite{mthm97-for-dart,harper1997programming}.\footnote{In SML, list literals work even in contexts where the list expression constructors have been shadowed by other bindings, i.e.
+When designing the surface syntax of a general-purpose programming language, it is a common practice to build in \emph{literal forms} that decrease the syntactic cost of constructing and pattern matching over values of a select few standard types.
+For example, in Standard ML (SML) and its cousins, a list expression literal like \li{[x, y, z]} is equivalent by desugaring to \li{Cons(x, Cons(y, Cons(z, Nil)))} where \li{Nil} and \li{Cons} stand for the constructors of the standard \li{list} datatype \cite{mthm97-for-dart,harper1997programming}.\footnote{In SML, list literals work even in contexts where the list expression constructors have been shadowed by other bindings, i.e.
 the meaning of literal forms is \emph{context independent}.
@@
 
-This language design practice is decidedly \emph{ad hoc} -- it is easy to come up with other examples of data structures that are not privileged with literal forms in these languages, but for which mathematicians, scientists or programmers have elsewhere invented specialized notation \cite{DBLP:journals/cacm/Iverson80,cajori1928history}.
-Consider, for example, that 1) clients of a ``collections'' library might like not just list literals, but also matrix, set and map literals following standard mathematical conventions; 2) clients of a ``web programming'' library might like not just HTML literals but also CSS literals (which Ur/Web does not provide); 4) a compiler writer might like ``quotation'' literals for the terms of the object language and various intermediate languages of interest; and 5) clients of a ``chemistry'' library might like composite chemical structure literals based on the SMILES standard \cite{anderson1987smiles}.
-
-Although concerns about ``missing'' literal notation are easy to dismiss as superficial or frivolous, the reality is that literal forms, or the absence thereof, have had a substantial influence on the quality of software that is written today.
-For example, \citet{Bravenboer:2007:PIA:1289971.1289975} finds that literal forms for structured encodings of queries, like the SQL-based query literals now found in many languages \cite{LINQ}\todo{find citation}, reduce the temptation to use string encodings of queries and therefore reduce the prevalence of string injection attacks, which are amongst the most common and serious security vulnerabilities on the web today \cite{owasp2017}.
+This common language design practice is decidedly \emph{ad hoc} -- it is quite easy to come up with other examples of data structures that are not privileged with literal forms in these languages, but for which mathematicians, scientists or programmers have invented specialized notation \cite{DBLP:journals/cacm/Iverson80,cajori1928history}.
+Consider, for example, that 1) clients of a ``collections'' library might like not just list literals, but also matrix, set and map literals following standard mathematical conventions; 2) clients of a ``web programming'' library might like not just HTML literals but also CSS literals (which Ur/Web does not have); 4) a compiler writer might like ``quotation'' literals for the terms of the object language and various intermediate languages; and 5) clients of a ``chemistry'' library might like chemical structure literals based on the SMILES standard \cite{anderson1987smiles}.
+
+Although concerns about ``missing'' literal notation are easy to dismiss as superficial or frivolous, the reality is that literal forms, or the absence thereof, have a substantial influence on software quality.
+For example, \citet{Bravenboer:2007:PIA:1289971.1289975} finds that literal forms for structured encodings of queries, like the SQL-based query literals now found in many languages \cite{meijer2006linq}, reduce the temptation to use string encodings of queries and therefore reduce the prevalence of string injection attacks, which are amongst the most common and serious security vulnerabilities on web servers today \cite{owasp2017}.
 More generally, evidence suggests that programmers frequently engage in ``stringly-typed programming'', i.e.
@@
 for SQL queries, regexes, file system paths, URLs and many other data structures.
-Literal forms at these types, with support for splicing, would decrease the syntactic cost of composite encodings, which are more amenable to programmatic manipulation and compositional reasoning than flat string encodings.
+Literal forms, with support for splicing, would decrease the syntactic cost of composite encodings of such data, which are more amenable to programmatic manipulation and compositional reasoning than flat string encodings.
 %Literal forms would  preferable to string encodings because they support compositional construction and decomposition by pattern matching via splicing.
@@
 \ref{sec:existing-approaches}, can be used to define new literal forms.
-The problem that specifically motivates this paper is that these mechanisms are quite \emph{unreasonable} when ``programming in the large'' \cite{DeRemer76}.
+The problem that specifically motivates this paper is that these mechanisms can obscure program structure and are therefore \emph{unreasonable} for ``programming in the large'' \cite{DeRemer76}.
 %, i.e. when developing large programs ``consisting of many small programs (modules), possibly written by different people''
@@
 reason about) the program text in Fig.
-\ref{fig:K-dialect}, which is written in a dialect of ML's (or Scala's) syntax, constructed by one of these mechanisms, where a large number of new literal forms have been defined.
+\ref{fig:K-dialect}, which is written in a dialect of OCaml's syntax called Reason\footnote{\url{http://facebook.github.io/reason}} that has, hypothetically, been extended with a large number of new literal forms.
 
@@
 \begin{lstlisting}
-  val w = compute_w()
-  val x = compute_x(w)
-  val y = {|(!R)@&{&/x!/:2_!x}'!R}|}
+  let w = compute_w();
+  let x = compute_x(w);
+  let y = [|(!R)@&{&/x!/:2_!x}'!R}|];
 \end{lstlisting}
 \vspace{-10px}
-\caption{An example of unreasonable program text.}
+\caption{Syntax extension systems produce unreasonable program text.}
 \vspace{-10px}
@@
 \noindent
-The literal form on Line 3 above constructs an encoding of a query in the stack-based database query language K, using its terse surface syntax \cite{Whitney:2001:LOR:376284.375783}.
-Of course, the programmer reading this program text may well be unfamiliar with K's query syntax.
-Unfortunately for the programmer, there are no simple, abstract protocols for answering questions like:
+The literal form on Line 3 constructs an encoding of a query in the niche stack-based database query language K, using its terse surface syntax \cite{Whitney:2001:LOR:376284.375783}.
+Of course, the programmer reading this program text may be unfamiliar with K's query syntax.
+Unfortunately for this programmer, there are no simple, abstract protocols for answering questions like:
 
@@
 \item \textbf{(Segmentation)} Are the characters \li{x} and \li{R} on Line 3 taken as spliced expressions (meaning that they appear directly in the desugaring), or are they parsed in some other way peculiar to this literal form (e.g.
-as operators in the K language)?
-\item \textbf{(Capture)} If \li{x} is in fact a spliced expression, does it refer to the binding of \li{x} on Line 2, or might it capture an unseen binding of the same identifier used in the desugaring of Line 3?
-\item \textbf{(Context Dependence)} If \li{w}, on Lines 1-2, is renamed, could that possibly break Line 3 or change its meaning because the desugaring assumes that some variable identified as \li{w} is in scope?
-What other hidden assumptions might the desugaring be making?
+as operators in the K query language)?
+\item \textbf{(Capture)} If \li{x} is in fact a spliced expression, does it refer to the binding of \li{x} on Line 2, or might it capture an unseen binding of the same identifier in the desugaring of Line 3?
+\item \textbf{(Context Dependence)} If \li{w}, on Lines 1-2, is renamed, could that possibly break Line 3 or change its meaning because the desugaring assumes that some variable \li{w} is in scope?% What other hidden assumptions might the desugaring be making?
 \item \textbf{(Typing)} What type does \li{y} have?
@@
 they are initially parsed like raw string literals.) For example, Fig.
-\ref{fig:K-tsm-example} applies a TLM, \li{$kquery}, to a backtick-delimited generalized literal form to express the ``unreasonable'' K query example from Fig.
-\ref{fig:K-dialect} just discussed.
+\ref{fig:K-tsm-example} applies a TLM, \li{$kquery}, to a generalized literal form delimited by \li{[|} and \li{|]} to express the ``unreasonable'' K query example from Fig.
+\ref{fig:K-dialect}.
 The applied TLM parses the literal body and generates an expansion statically, during a semantic phase called typed expansion, which generalizes the usual typing phase.
 TLMs come equipped with uniquely powerful abstract reasoning principles.
-We will characterize these reasoning principles precisely as we proceed, but for now, let us develop some intuitions: the programmer can reason abstractly about Line 3 as follows:
+We will characterize these reasoning principles precisely as we proceed, but for now, let us develop the basic intuitions -- the programmer can reason abstractly about Line 3 as follows:
 \begin{enumerate}[noitemsep,nolistsep]
-\item \textbf{(Responsibility)} The applied TLM, \li{$kquery}, is solely responsible for parsing and expansion of the literal body.
-The context-free syntax of the language is fixed, so the TLM provider can modularly establish that the syntax this TLM implements is deterministic.
-\item \textbf{(Segmentation)} The output that a TLM generates is structured so that the system can infer from it an accurate \emph{segmentation} of the literal body that distinguishes spliced terms from segments that are parsed in some other way by the TLM.
+\item \textbf{(Responsibility)} The applied TLM, \li{$kquery}, parses and determines the expansion of the literal body.
+The context-free syntax of the language is fixed, so the TLM provider is able to modularly establish that the syntax this TLM implements is deterministic.
+\item \textbf{(Segmentation)} The output that the TLM generates is structured so that the system can infer from it an accurate \emph{segmentation} of the literal body that distinguishes spliced terms from segments that are parsed in some other way by the TLM.
 The segmentation can be communicated by a code editor using secondary notation, e.g.
 colors in this document.
-By examining Fig.
-\ref{fig:K-tsm-example}, we know that the two instances of \li{x} on Line 3 are parsed as spliced expressions (because they are in black), whereas the \li{R}'s must be parsed in some other way, e.g.
+So by examining Fig.
+\ref{fig:K-tsm-example}, the programmer knows that the two instances of \li{x} on Line 3 are parsed as spliced expressions (because they are in black), whereas the \li{R}'s must be parsed in some other way, e.g.
 as operators in the K query language (because they are in green.)
@@
 We will see an example of a TLM definition in the next section.
-Moreover, each spliced segment in the segmentation also comes paired with the type it is expected to have.
-This information is usually not necessary to reason about typing, but it can be conveyed to the programmer upon request by the program editor if desired.
+Moreover, each spliced segment in the segmentation also comes with a type annotation.
+%This information is usually not necessary to reason about typing, but it can be conveyed to the programmer upon request by the program editor if desired.
 %TLM definitions follow the usual scoping rules, so it is easy to ``jump to the definition'' of \li{$kquery}.
@@
 \begin{lstlisting}
-  val w = compute_w()
-  val x = compute_x(w)
-  val y = $kquery `SURL(!R)@&{&/EURLxSURL!/:2_!EURLxSURL}'!R}EURL`
+  let w = compute_w();
+  let x = compute_x(w);
+  let y = $kquery [|SURL(!R)@&{&/EURLxSURL!/:2_!EURLxSURL}'!R}EURL|];
 \end{lstlisting}
@@
 Sec.
-\ref{sec:setsms} defines another {simple expression TLM} -- one that implements Ur/Web-style HTML literal syntax (Fig.
-\ref{fig:urweb}) -- and uses it as a running example to informally detail how TLMs work from start to finish.
+\ref{sec:setsms} defines another {simple expression TLM} -- one that implements the Ur/Web-style HTML literal syntax from Fig.
+\ref{fig:urweb} -- and uses it as a running example to informally describe how TLMs work from start to finish.
 % \item 
-Sec. \ref{sec:sptsms} briefly introduces {pattern TLMs}, i.e. TLMs that generate patterns. %The hygiene condition for pattern TLMs is interesting.
+Sec. \ref{sec:sptsms} then briefly introduces {pattern TLMs}, i.e. TLMs that generate patterns. %The hygiene condition for pattern TLMs is interesting.
 % \item 
@@
 \ref{sec:ptsms} then adds type functions and an ML-style module system and introduces \emph{parametric TLMs} (pTLMs), i.e.
-TLMs that take type and module parameters.
-Parameters serve two key purposes.
+TLMs that can take type and module parameters.
+Parameters serve two important purposes.
 First, they allow a TLM to operate not just at a single type, but over a type- and module-parameterized family of types.
 We give as an example a parametric TLM for string-keyed dictionaries that operates uniformly over different implementations of the abstract dictionary signature, and also over different choices of the range type.
-Second, parameters give expansions access to terms other than those explicitly spliced in to the literal body by the client.
-%This is particularly convenient for working with abstract types \emph{a la} ML.
-Partial parameter application is possible.
-Together, this lowers the syntactic cost of the strict context independence discipline that TLMs enforce.
+Second, parameters lower the syntactic cost of the strict context independence discipline that TLMs enforce because they give expansions access to terms other than those explicitly spliced in to the literal body by the client, and because %This is particularly convenient for working with abstract types \emph{a la} ML.
+partial parameter application is possible.
 % Sec.
@@
 
-Due to space limitations, certain technical details of the calculi defined in this paper, as well as proofs, are relegated to the supplemental material.
-The examples in this paper are written in an emerging dialect of  ML called VerseML that is focused on improving the ergonomics of ML.
-However, the essential ideas are not ML-specific -- it should be possible to adapt TLMs to other languages that take a similar approach to types and binding (e.g.
-Haskell, Scala and others.) When a design choice is ML-specific, we will make note of it\todo{do this}.
-Languages that have a disciplined binding structure but that lack a rich static type structure, e.g.
-Racket and other LISP derivatives, might also benefit from TLMs -- they can apply the results of this paper by deploying the usual trick of reducing the language to one that is ``{unityped}'' \cite{pfpl,scott1980lambda}.%  We say more about LISP macros in Sec.
-\ref{sec:existing-approaches}.% For the present purposes, we can consider VerseML as essentially the calculus defined in the supplement extended with various conveniences that are commonly found in other ML-like languages and, notionally, orthogonal to TLMs.
-%VerseML is, as its name suggests, a conceptual descendent of ML.
+Due to space limitations, certain technical details of the calculi described in this paper, as well as detailed proofs, are given in the supplemental material.
+The examples in the remainder of the paper are written in an open source alternative syntactic front-end for OCaml called Reason, mentioned above, because we are working to incorporate TLMs into Reason.
+However, the essential ideas are not Reason-, OCaml- or ML-specific -- it should be possible to adapt TLMs to other languages that take a similar approach to types and binding (e.g.
+Haskell, Scala and others.) Languages that have a disciplined binding structure but that lack a rich static type structure, e.g.
+standard Racket, would also benefit from TLMs -- they can apply the results of this paper by deploying the usual trick of viewing the language as statically ``{unityped}'' \cite{pfpl,scott1980lambda}.%  We say more about LISP macros in Sec.
+\ref{sec:existing-approaches}.% For the present purposes, we can consider Reason as essentially the calculus defined in the supplement extended with various conveniences that are commonly found in other ML-like languages and, notionally, orthogonal to TLMs.
+%Reason is, as its name suggests, a conceptual descendent of ML.
 It diverges from other dialects of ML that have a similar type structure in that it has a bidirectional type system \cite{Pierce:2000:LTI:345099.345100} (like, for example, Scala \cite{OdeZenZen01}) for reasons that have to do with the mechanism of TLM implicits described in Chapters \ref{chap:tsls} and \ref{chap:ptsms}.
-%The reason we will not follow Standard ML \cite{mthm97-for-dart} in giving a complete formal definition of VerseML in this work is both to emphasize that the primitives we introduce are ``insensitive'' to the details of the underlying type structure of the language (so TLMs can be considered for inclusion in a variety of languages, not only dialects of ML), and to avoid distracting the reader (and the author) with definitions that are already well-understood in the literature and that are orthogonal to those that are the focus of this work.
+%The reason we will not follow Standard ML \cite{mthm97-for-dart} in giving a complete formal definition of Reason in this work is both to emphasize that the primitives we introduce are ``insensitive'' to the details of the underlying type structure of the language (so TLMs can be considered for inclusion in a variety of languages, not only dialects of ML), and to avoid distracting the reader (and the author) with definitions that are already well-understood in the literature and that are orthogonal to those that are the focus of this work.
 % We will not formally define these features mainly to avoid unnecessarily complicating our presentation with details that are not essential to the ideas presented herein. As such, 
-% All examples written in VerseML should be understood to be informal motivating material for the subsequent formal material.
+% All examples written in Reason should be understood to be informal motivating material for the subsequent formal material.
 %We anticipate that future full-scale language specifications will be able to combine the ideas  in the proposed work without trouble.
 %The purpose of the work being proposed is to serve as a reference for those interested in the new constructs we introduce, not to serve as a language specification.
-%We will give a brief overview of these languages are organized in Sec. \ref{sec:VerseML}.
+%We will give a brief overview of these languages are organized in Sec. \ref{sec:Reason}.
 
@@
 \ref{sec:metamodules}, reduce the need to primitively build in the type structure of constructs like records (and variants thereof),  labeled sums and other interesting constructs that we will introduce later by giving library providers programmatic ``hooks'' directly into the semantics, which are specified as a \emph{type-directed translation semantics} targeting a small \emph{typed internal language} (introduced in Sec.
-\ref{sec:VerseML}).
+\ref{sec:Reason}).
 %For example, a library provider can implement the type structure of records with a metamodule that:
@@
 during typing.
-%Library providers write these static functions using the VerseML \emph{static language} (SL).
-%Maintaining a separation between the static (or ``compile-time'') phase and the dynamic (or ``run-time'') phase is an important facet of VerseML's design.
+%Library providers write these static functions using the Reason \emph{static language} (SL).
+%Maintaining a separation between the static (or ``compile-time'') phase and the dynamic (or ``run-time'') phase is an important facet of Reason's design.
 % static code generation.
-%We will  also introduce a simple variant of each of these primitives that leverages VerseML's support for local type inference to further reduce syntactic cost in certain common situations.
-
+%We will  also introduce a simple variant of each of these primitives that leverages Reason's support for local type inference to further reduce syntactic cost in certain common situations.
+
+\vspace{-4px}
 \newcommand{\seTLMsSec}{Simple Expression TLMs (seTLMs)}
@@
 \label{sec:setsms}
-
-This section informally details the TLM mechanism.
+\vspace{-2px}
+We will start in this section by informally describing TLMs in Reason.
 It will suffice in this section to consider only \emph{simple expression TLMs (seTLMs)}, i.e.
 TLMs for abbreviating expressions of a single specified type.
-In particular, this section is organized around an seTLM named \li{$html} that implements Ur/Web-style HTML literals, which abbreviate expressions of the type \li{html} defined as follows:
-\begin{lstlisting}[numbers=none]
-  type html = BodyElement of list(html_attr) * list(html)
-            | H1Element of list(html_attr) * list(html) 
-            | (* ... *) | TextNode of string 
-\end{lstlisting}
-The type \li{html_attr} is straightforward and the type function \li{list} is standard, so neither are shown.
-In VerseML, \li{string}s are primitive (though this is not essential, see footnote \todo{reference}1 below.)
-
-Using this TLM, we can express the Ur/Web program from Fig. \ref{fig:urweb} as shown in Fig. \ref{fig:first-tsm-example}: on each line, we apply 
-\li{$html} 
- to a \emph{generalized literal form} delimited by backticks. TLM names are prefixed by \li{$} 
- to clearly them from variable identifiers.
-The semantics delegates control over the parsing and expansion of each literal body to the applied TLM during a semantic phase called \emph{typed expansion}, which generalizes the usual typing phase.
+In particular, we will consider an seTLM named \li{$html} that implements Ur/Web style HTML literal notation for expressions of the \li{html} type that is defined in Fig.
+\ref{fig:html-type-def}.
+
 \begin{figure}[ht!]
 \vspace{-10px}
-\begin{lstlisting}
-  fun heading first_name = $html `SURL<h1>Hello, {[EURLfirst_nameSURL]}!</h1>EURL`
-  val body = $html `SURL<body>{EURLheading "World!"SURL} ...</body>EURL`
+\begin{lstlisting}[numbers=none]
+  type html = BodyElement(list(html_attr), list(html))
+            | H1Element(list(html_attr), list(html))
+            | TextNode(string) | /* ... */;
 \end{lstlisting}
 \vspace{-10px}
-\caption{Two examples of a TLM being applied to a generalized literal form.
+\caption{The \li{html} type, which classifies encodings of HTML data.
+The type \li{html_attr} is also straightforward and the type function \li{list} is standard, so neither are shown.
+Strings are primitive.}
+\label{fig:html-type-def}
+\vspace{-10px}
+\end{figure}
+
+\vspace{-4px}
+\subsection{TLM Definition and Application}
+\vspace{-2px}
+The definition of the \lstinline{$html} TLM is outlined in Fig. \ref{fig:html-tlm-def}. TLM names are prefixed by \li{$} 
+ to distinguish them from variable identifiers.
+Every TLM definition also has a \emph{type annotation}, here \lstinline{at html}, and a \emph{parse function} between \li{by static \{} and \li{\}}.
+The TLM definition is in scope until the end of the enclosing scope.
+We will describe how TLM definitions are packaged into libraries in Sec \ref{sec:static-eval}.
+
+\begin{figure}[ht!]
+\vspace{-12px}
+\begin{lstlisting}[mathescape=|]
+  syntax $html at html by static {
+    fun(b : body) : parse_result(proto_expr) => 
+      /* ... Ur/Web-style HTML syntax parser ... */
+  };
+\end{lstlisting}
+\vspace{-15px}
+\caption{The definition of the seTLM named \li{$html}.}
+\label{fig:html-tlm-def}
+\vspace{-10px}
+\end{figure}
+
+Using the TLM defined in Fig. \ref{fig:html-tlm-def}, we can express the example Ur/Web program from Fig. \ref{fig:urweb} as shown in Fig. \ref{fig:first-tsm-example}. On both lines, we apply 
+\li{$html} 
+ to a \emph{generalized literal form} delimited by \li{[|} and \li{|]}.
+Generalized literal forms, which first arose in related work by \citet{TSLs} that we will detail in Sec.
+\ref{sec:existing-approaches}, subsume other literal forms because the context-free syntax of the language only specifies the outer delimiters (in this paper, we will use  \li{[|} and \li{|]}, but \citet{TSLs} formally specified several other choices, including layout-sensitive delimitation.) \emph{Literal bodies}, i.e.
+the characters between these delimiters, are constrained only in that occurrences of \li{[|} must be balanced by \li{|]}.
+%In particular, there is no pre-specified syntactic to splice out sub-terms.%The initial context-free parser therefore parses generalized literal forms in much the same way as parsers typically parse ``raw'' string literals (i.e.
+string literals where escape sequences like \li{\\n} have not yet been processed.)
+
+\begin{figure}[t!]
+\vspace{-2px}
+\begin{lstlisting}
+  fun heading first_name = $html [|SURL<h1>Hello, {[EURLfirst_nameSURL]}!</h1>EURL|]
+  val body = $html [|SURL<body>{EURLheading "World!"SURL} ...</body>EURL|]
+\end{lstlisting}
+\vspace{-10px}
+\caption{Two examples of the \li{$html} TLM being applied.
 %In each case, literal body, between backticks, is initially left unparsed according to the language's context-free syntax.
 %The applied TLM determines a segmentation and expansion during the typed expansion phase, which generalizes the usual typing phase.
-We assume strings are primitive.}
-\vspace{-10px}
+}
+\vspace{-12px}
 \label{fig:first-tsm-example}
@@
 
-Generalized literal forms, which first arose in related work by \citet{TSLs} that we will detail in Sec.
-\ref{sec:related-work}, subsume a variety of common syntactic forms because the context-free syntax of the language only defines which outer delimiters are available (in this paper, we will mostly use backticks, but in general, there can be many choices.) \emph{Literal bodies}, i.e.
-the characters between the delimiters, are otherwise syntactically unconstrained.
-Choices regarding internal delimitation of spliced terms is not imposed by the language definition.
-Notice, however, that Figure \ref{fig:first-tsm-example} reveals the locations of the spliced expressions by coloring them black.
-We have designed our system so that a figure like this, which presents a \emph{segmentation} of each literal body into spliced terms (in black) and characters parsed in some other way by the applied TLM (in green)\footnote{In general, spliced expressions might themselves apply TLMs, in which case the convention is to use a distinct color for unspliced segments at each depth.
-For example, if strings were not primitive, we might write \li{`SURL<body>\{EURLheading ($str "SSTRWorld!ESTR")SURL\} ...</body>EURL`}.}, can always be  generated.
+ The semantics delegates responsibility over parsing and expansion of each literal body to the applied TLM's parse function during a semantic phase called \emph{typed expansion}, which generalizes the usual typing phase.
+Because the {parse function}  is applied during the typed expansion process, rather than at run-time, we call it a \emph{static function}.
+Static functions  cannot refer to the surrounding variable bindings (because those variables stand for run-time values.) For now, we will assume that static functions are completely closed, and that they do not themselves make use of TLMs (we will eliminate these impractical limitations by introducing distinct static phase bindings in Sec.
+\ref{sec:static-eval}.)
+
+ The parse function must have type \li{body -> parse_result(proto_expr)}.
+The input type, \lstinline{body}, classifies encodings of literal {bodies}.
+Literal bodies are sequences of characters, so it suffices to define \li{body} as an abbreviation for \li{string}, as shown in Fig.
+\ref{fig:indexrange-and-parseresult}.
+The return type is a labeled sum type, defined by applying the type function \li{parse_result} defined in Fig.
+\ref{fig:indexrange-and-parseresult}, that distinguishes between parse errors and successful parses.\footnote{We might alternatively have used an exception to signal a parse error, but sum types are formally simpler.} Let us consider these two possibilities in turn.
+
+ If the parse function determines that the literal body is not well-formed according to whatever syntax definition that it implements, it must return \li{ParseError \{msg=}$e_\text{msg}$\li{, loc=}$e_\text{loc}$\li{\}}
+where $e_\text{msg}$ is a custom error message and $e_\text{loc}$ is a value of type \li{segment}, defined in Fig.
+\ref{fig:indexrange-and-parseresult}, that designates a segment of the literal body as the origin of the error \cite{DBLP:journals/jsc/DeursenKT93}.
+This information is reported to the programmer by the Reason compiler or editor services.
+
+If, on the other hand, parsing succeeds, the parse function returns \li{Success} ~$\ecand$, 
+where $\ecand$ is called the \emph{encoding of the proto-expansion}.
+For expression TLMs, the proto-expansions is a \emph{proto-expression}.
+These are encoded as values of the recursive sum type \lstinline{proto_expr} that is outlined in Fig.
+\ref{fig:candidate-exp-Reason}.\footnote{More sophisticated encoding techniques for syntax trees would cause no fundamental problems; we have kept it simple here for the sake of clarity.}
+Most of the constructors of \li{proto_expr} are individually uninteresting -- they encode Reason's various expression forms (just as in a self-hosted compiler, e.g.
+SML/NJ's Visible Compiler library.)
+Expressions can mention types, so we also need the type \li{proto_typ} also outlined in Fig.
+\ref{fig:candidate-exp-Reason}.
+It is only the \li{SplicedE} and \li{SplicedT} constructors that are novel.
+These are discussed next.
 
 % \section{Simple Expression TLMs By Example}\label{sec:tsms-by-example}
-% We begin in this section with a ``tutorial-style'' introduction to seTLMs in VerseML.
+% We begin in this section with a ``tutorial-style'' introduction to seTLMs in Reason.
 %In particular, we will define an seTLM for constructing values of the recursive labeled sum type \li{rx} that was defined in Figure \ref{fig:datatype-rx}.
 % Sec.
-\ref{sec:tsms-minimal-formalism} then formally defines a reduced dialect of VerseML called $\miniVerseUE$.
+\ref{sec:tsms-minimal-formalism} then formally defines a reduced dialect of Reason called $\miniVerseUE$.
 This will serve as a ``conceptually minimal'' core calculus of TLMs, in the style of the simply typed lambda calculus.
@@
 % \subsection{TLM Application}\label{sec:uetsms-usage}
-% The following VerseML expression, drawn textually, is of \emph{TLM application} form. Here, a TLM named \li{$rx} is applied to the \emph{generalized literal form} \li{/SURLA|T|G|CEURL/}:
+% The following Reason expression, drawn textually, is of \emph{TLM application} form. Here, a TLM named \li{$rx} is applied to the \emph{generalized literal form} \li{/SURLA|T|G|CEURL/}:
 % \begin{lstlisting}[numbers=none,mathescape=|]
@@
 % \end{lstlisting}
-% Generalized literal forms are left unparsed according to the context-free syntax of VerseML.
+% Generalized literal forms are left unparsed according to the context-free syntax of Reason.
 Several other outer delimiters are also available, as summarized in Figure \ref{fig:literal-forms}.
@@
 
-% % A number of literal forms, ,  are available in VerseML's concrete syntax.
+% % A number of literal forms, ,  are available in Reason's concrete syntax.
 Any literal form can be used with any TLM,  TLMs have access only to the literal bodies.
@@
 % %SURL<tag>body includes enclosing tags</tag>EURL
-% \caption[Available Generalized Literal Forms]{Generalized literal forms available for use in VerseML's textual syntax.
+% \caption[Available Generalized Literal Forms]{Generalized literal forms available for use in Reason's textual syntax.
 The characters in green indicate the literal bodies and describe how the literal body is constrained by the form shown on that line.
-The Wyvern language defines additional forms, including whitespace-delimited forms \cite{TSLs} and multipart forms \cite{sac15}, but for simplicity we leave these out of VerseML.}
+The Wyvern language defines additional forms, including whitespace-delimited forms \cite{TSLs} and multipart forms \cite{sac15}, but for simplicity we leave these out of Reason.}
 % \label{fig:literal-forms}
 % \end{figure}
-\subsection{TLM Definitions}\label{sec:uetsms-definition}
-%The original expression, above, is statically rewritten to this expression.
-The definition of the seTLM named \lstinline{$html} that was shown being applied in Fig. \ref{fig:first-tsm-example} takes the following form:
-\begin{lstlisting}[mathescape=|]
-  syntax $html at html by 
-    static fn(b : body) -> parse_result(proto_expr) => 
-      (* Ur/Web-style HTML syntax parser here *)
-  in ... end 
+
+\vspace{-4px}
+\subsection{Splicing}\label{sec:splicing-and-hygiene}
+\vspace{-2px}
+When the parse function determines that some segment of the literal body should be taken as a spliced expression, according to whatever syntactic criteria it deems suitable, it must not directly insert the syntax tree of that expression into the encoding of the expansion.
+Instead,
+the TLM must indirectly refer to the spliced expression using the \li{SplicedE} variant of \li{proto_expr}, which takes a value of type \li{segment} that indicates the zero-indexed location of the spliced expression relative to the start of the provided literal body.
+The \li{SplicedE} variant also requires a value of type \li{proto_typ}, which indicates the type that the spliced expression is expected to have.
+Types can be spliced out by using the \li{SplicedT} variant of \li{proto_typ} analagously.
+
+\begin{figure}
+\vspace{-2px}
+\begin{lstlisting}[numbers=none]
+type body = string;
+type segment = {startIdx : int, endIdx : int}; /* inclusive */
+type parse_result('a) = ParseError { msg : string, loc : segment }
+                      | Success('a);
+type var_t = string;
+type proto_typ = TyVar(var_t)
+               | Arrow(proto_typ, proto_typ)
+               | StringTy
+               | /* ... */ 
+               | SplicedT(segment);
+type proto_expr = Var(var_t)
+                | Fn(var_t, proto_typ, proto_expr)
+                | Ap(proto_expr, proto_expr)
+                | /* ... */ 
+                | SplicedE(segment, proto_typ);
 \end{lstlisting}
-
-Every seTLM definition consists of a TLM name, here \li{$html}, a \emph{type annotation}, here \lstinline{at html}, and a \emph{parse function} between \li{by} and \li{in}.
-The TLM definition is in scope between \li{in} and \li{end}.
-In practice, if an \li{in} clause is not provided, the definition is in scope until the end of the enclosing declaration (e.g.
-the enclosing function or module.) We will consider how TLM definitions are exported from libraries in Sec \ref{sec:static-eval}.
-
-% All TLM names must begin with the dollar symbol (\li{$}), which distinguishes them from variables.
-This is inspired by the Rust macro system, which uses post-fix exclamation points (\li{!}) to distinguish macro identifiers \cite{Rust/Macros}.
-
-The {parse function} is a \emph{static function} delegated responsibility over parsing the literal bodies of the literal forms to which the TLM is applied.
-Static functions, marked by the \li{static} keyword, are applied during the typed expansion process, so they cannot refer to the surrounding variable bindings (because those variables stand for dynamic values.) For now, we will simply assume that static functions are closed, and that they do not themselves make use of TLMs (we will eliminate these impractical limitations in Sec.
-\ref{sec:static-eval}.)
-
-The parse function must have type \li{body -> parse_result(proto_expr)}.
-The input type, \lstinline{body}, classifies encodings of literal {bodies}.
-In VerseML, literal bodies are sequences of characters, so it suffices to define \li{body} as an abbreviation for the \li{string} type, as shown in Figure \ref{fig:indexrange-and-parseresult}.
-The return type is a labeled sum type, defined by applying the type function \li{parse_result} defined in Figure \ref{fig:indexrange-and-parseresult}, that distinguishes between parse errors and successful parses.
-%\footnote{\li{parse_result} is defined as a type function because in Chapter \ref{chap:uptsms}, we will introduce pattern TLMs, which generate patterns rather than expressions.}
-Let us consider these two possibilities in turn.
-
-\paragraph{Parse Errors} If the parse function determines that the literal body is not well-formed (according to whatever syntax definition that it implements), it returns:
-\begin{lstlisting}[numbers=none]
-  ParseError {msg=~$e_\text{msg}$~, loc=~$e_\text{loc}$~}
-\end{lstlisting}
-where $e_\text{msg}$ is an error message and $e_\text{loc}$ is a value of type \li{segment}, defined in Figure \ref{fig:indexrange-and-parseresult}, that designates a segment of the literal body as the location of the error \cite{DBLP:journals/jsc/DeursenKT93}.
-This information is for use by VerseML compilers when reporting the error to the programmer (but it otherwise has no semantic significance.) We might alternatively have used an exception to signal a parse error, but sum types are easier to formalize.
-
-% In languages where the surface syntax is not textual, the definition of \li{loc} would need to designate
-
-\paragraph{Successful Parses} If parsing succeeds, the parse function returns 
-\begin{lstlisting}[numbers=none]
-  Success ~$\ecand$~
-\end{lstlisting} 
-where $\ecand$ is called the \emph{encoding of the proto-expansion}.
-For expression TLMs, proto-expansions are \emph{proto-expressions}, which are encoded as values of the recursive sum type \lstinline{proto_expr} outlined in Figure \ref{fig:candidate-exp-verseml}.
-Most of the variants of \li{proto_expr} are individually uninteresting -- they encode VerseML's various expression forms (just as in a self-hosted compiler, e.g. SML/NJ's Visible Compiler library.) 
-Expressions can mention types, so we also need the type \li{proto_typ} also outlined in Figure \ref{fig:candidate-exp-verseml}.
-More sophisticated encoding techniques for terms would cause no fundamental problems.
-It is only the \li{SplicedE} and \li{SplicedT} variants of these types that are interesting.
-%As we enrich our language in later chapters, we will need to define more encodings like these, for other sorts of trees.
-
-% The definitions of \li{proto_typ} and \li{proto_expr} are recursive labeled sum types to simplify our exposition, but we could have chosen alternative encodings, e.g.
-based on abstract binding trees \cite{pfpl}, with only minor modifications to our semantics.
-Indeed, when we formally define seTLMs in Sec.
-\ref{sec:miniVerseU}, we abstract over the particular encoding.
-
-% We will show a complete encoding when we describe our reduced formal system $\miniVerseUE$ in Sec. \ref{sec:tsms-minimal-formalism}. 
-% ; the remaining constructors (some of which are elided for concision) encode the abstract syntax of VerseML expressions and types.
-% It is extended with one additional form used to handled spliced subexpressions,
-
-%Notice that the types just described are those that one would expect to find in a typical parser.
-
-%One would find types analagous to those just described in any parser, so for concision, we elide the details of \li{$rx}'s parse function.
-%The parse function must treat the TLM parameters parametrically, i.e.
-it does not have access to any values in the supplied module parameter.
-Only the expansion the parse function generates can refer to module parameters.
-%For example, the following definition is ill-sorted:
-%\begin{lstlisting}[numbers=none]
-%syntax pattern_bad[Q : PATTERN] at Q.t {
-%  static fn (body : Body) : Exp => 
-%    if Q.flag then (* ... *) else (* ... *)
-%}
-%\end{lstlisting}%So the parse function parses the body of the delimited form to generate an encoding of the elaboration.
-
-\begin{figure}
-\begin{lstlisting}[numbers=none]
-type body = string
-type segment = {startIdx : nat, endIdx : nat} (* inclusive *)
-type parse_result('a) = ParseError of { msg : string, loc : segment }
-                      | Success of 'a
-type var_t = string (* ... or some other encoding of variables ... *)
-type proto_typ = TyVar of var_t 
-               | Arrow of proto_typ * proto_typ 
-               | (* ... *) 
-               | SplicedT of segment
-type proto_expr = Var of var_t 
-                | Fn of var_t * proto_typ * proto_expr
-                | Ap of proto_expr * proto_expr
-                | (* ... *) 
-                | SplicedE of segment * proto_typ
-\end{lstlisting}
-\vspace{-5px}
+\vspace{-10px}
 \caption[Definitions of various types used by TLM definitions.]{Definitions of various types used by TLM definitions. We assume that these are available ambiently.}
 \label{fig:indexrange-and-parseresult}
-\label{fig:candidate-exp-verseml}
-\vspace{-5px}
+\label{fig:candidate-exp-Reason}
+\vspace{-12px}
 \end{figure}
-
-\subsection{Splicing}\label{sec:splicing-and-hygiene}
-When a parse function determines that a segment of the literal body should be taken as a spliced expression, it does not directly insert the syntax tree of that expression into the encoding of the expansion.
-Instead,
-the TLM must refer to the spliced expression \emph{by its relative location} using the \li{SplicedE} variant of \li{proto_expr}, which takes a value of type \li{segment} that indicates the zero-indexed location of the spliced expression relative to the start of the provided literal body.
-The \li{SplicedE} variant also requires a value of type \li{proto_typ}, which indicates the type that the spliced expression is expected to have.
-Types can be spliced out by using the \li{SplicedT} variant of \li{proto_typ} analagously.
 
@@
 \begin{lstlisting}[numbers=none]
-  $html `SURL<h1>Hello, {[EURLfirst_nameSURL]}</h1>EURL`
+  $html [|SURL<h1>Hello, {[EURLfirst_nameSURL]}</h1>EURL|]
 \end{lstlisting}
-Let us assume that \li{html} abbreviates a recursive labeled sum type having at least the cases
-\begin{lstlisting}[numbers=none]
-  H1Element of list(html_attr) * list(html) | TextNode of string | ...
-\end{lstlisting}
-The proto-expansion generated by \li{$html} for the supplied literal body, if written in a hypothetical textual syntax where references to spliced expressions are written \li{spliced<startIdx; endIndex; ty>}, is:
+The proto-expansion generated by \li{$html}, if written in a textual syntax for proto-expressions where references to spliced expressions are written \li{spliced<startIdx; endIndex; ty>}, is:
 \begin{lstlisting}[numbers=none]
@@
 \end{lstlisting}
-Here, \li{spliced<13; 22; string>} refers to the spliced expression \li{first_name} by location, rather than directly, and indicates that it must be of type \li{string}.
-
-Requiring that TLMs refer to spliced expressions indirectly in this manner prevents them from ``forging'' spliced terms (i.e.
-claiming that an term is a spliced term when it does not in fact appear textually in the literal body.)% Our semantics distinguishes spliced expressions when validating the proto-expansion, which is important for reasons of hygiene that we will detail shortly.
+The proto-expression \li{spliced<13; 22; string>} is a reference to the spliced string expression \li{first_name} by its location relative to the start of the literal body being expanded.
+It corresponds to the following encoding:
+\li{SplicedE (\{startIdx: 13, endIdx: 22\}, StringTy)}.
+
+Requiring that TLMs refer to spliced expressions indirectly in this manner ensures that a TLM cannot ``forge'' spliced terms (i.e.
+claim that some sub-term of the expansion should be given the privileges of a spliced term, discussed below, when it does not in fact appear in the literal body.)% Our semantics distinguishes spliced expressions when validating the proto-expansion, which is important for reasons of hygiene that we will detail shortly.
 
@@
 % \subsection{Splice Summaries and Segmentations}
-The \emph{splice summary} of a proto-expression is the finite set of references to spliced terms contained within.
-For example, the summary of the proto-expression above is the finite set containing only \li{spliced<13; 22; string>}.% Notice that no information about  the spliced terms appear is communicated by the splice summary.
-
-The \emph{segmentation} of a proto-expression is the finite set of segments extracted from the splice summary, where a segment is simply a pair of natural numbers.
-For example, the segmentation of the proto-expansion above is $\{(13, 22)\}$.
-The semantics checks that all of the segments in the segmentation are 1) in bounds relative to the literal body; and 2) non-overlapping. 
-This resolves the problem of \textbf{Segmentation} described in Sec. \ref{sec:intro}, i.e. every literal body in a well-typed program has a well-defined segmentation.
-
-A program editor or pretty-printer can communicate the segmentation information to the programmer, e.g. by coloring non-spliced segments green as is our convention in this document. 
-A program editor or pretty-printer can also communicate the type of each spliced term, as indicated in the splice summary, to the programmer upon request (for example, the Emacs and Vim packages for working with OCaml defer to the Merlin tool when the programmer requests the type of an expression.)
-
+The \emph{segmentation} inferred from a proto-expansion is the finite set of references to spliced terms contained within.
+For example, the segmentation inferred from the proto-expression above is the finite set containing only \li{spliced<13; 22; string>}.
+% Notice that no information about  the spliced terms appear is communicated by the splice summary.
+The semantics checks that all of the locations in the segmentation are 1) in bounds relative to the literal body; and 2) non-overlapping. 
+This resolves the problem of \textbf{Segmentation} described in Sec.
+\ref{sec:intro}, i.e.
+every literal body in a well-typed program has a well-defined segmentation.
+A program editor or pretty-printer can communicate this location information to the programmer, e.g.
+by coloring non-spliced segments green as is our convention in this document.
+In general, spliced expressions might themselves apply TLMs, in which case the convention is to use a distinct color for unspliced segments at each depth.
+For example, if strings were not primitive but rather defined as sequences of characters, we might define a TLM \li{$str} to recover string literal notation and write \li{$html [|SURL<body>\{EURLheading ($str [|SSTRWorld!ESTR|])SURL\} ...</body>EURL|]}.
+
+A program editor or pretty-printer can also communicate the type of each spliced expression, as specified by the segmentation, to the programmer upon request (the editor packages for working with Reason defer to the Merlin tool when the programmer requests the type of an expression.) %When there is a type mismatch, the type annotation on the spliced segment also allows the error message to report the problem without revealing the full expansion.
+
+\vspace{-6px}
 \subsection{Proto-Expansion Validation}\label{sec:uetsms-validation}
-Three concerns described in Sec.
+Three important concerns described in Sec.
 \ref{sec:intro} remain: those related to reasoning abstractly about \textbf{Capture}, \textbf{Context Dependence} and \textbf{Typing}.
-Addressing these concerns is the purpose of the \emph{proto-expansion validation} process.
-The result of proto-expansion validation is the \emph{final expansion}, which is simply the proto-expansion with  references to spliced terms replaced with their corresponding final expansions.
+Addressing these concerns is the purpose of the \emph{proto-expansion validation} process, which occurs once a proto-expansion has been generated by the parse function.
+Proto-expansion validation results in the \emph{final expansion}, which is simply the proto-expansion with the references to spliced segments replaced with their own final expansions.
 
 \subsubsection{Capture}
-Proto-expansion validation ensures that spliced terms have access \emph{only} to the bindings that appear at the application site -- spliced terms cannot capture the bindings that appear in the proto-expansion.
+Proto-expansion validation ensures that spliced terms have access \emph{only} to the bindings at the application site -- spliced terms cannot capture the bindings in the proto-expansion.
 For example, consider the following application site:
 \begin{lstlisting}[numbers=none]
-  let tmp = (* ... application site temporary ... *) in 
-  $html /SURL<h1>{EURLf(tmp)SURL}</h1>EURL/
+  let tmp = /* ... application site temporary ... */;
+  $html [|SURL<h1>{EURLf(tmp)SURL}</h1>EURL|];
 \end{lstlisting}
@@
 \begin{lstlisting}[numbers=none]
-  let tmp = (* ... expansion-internal temporary ... *) in 
-  H1Element(tmp, spliced<5; 10; html>)
+  let tmp = /* ... expansion-internal temporary ... */;
+  H1Element(tmp, spliced<5; 10; html>);
 \end{lstlisting}
@@
 
-To address this problem, proto-expansion validation is capture-avoiding.
-In other words, it discharges the requirement that capture not occur by implicitly alpha-varying the bindings in the proto-expansion as needed.
+To address this problem, splicing is guaranteed to be capture-avoiding.
+When generating the final expansion, the system discharges the requirement that capture not occur by implicitly alpha-varying the bindings in the proto-expansion as needed.
 For example, the final expansion of the example above might take the following form:
 \begin{lstlisting}[numbers=none]
-  let tmp = (* ... application site temporary ... *) in 
-  let tmp' = (* ... expansion-internal temporary ... *) in 
-  H1Element(tmp', f(tmp))
+  let tmp = /* ... application site temporary ... */; 
+  let tmp' = /* ... expansion-internal temporary ... */;
+  H1Element(tmp', f(tmp));
 \end{lstlisting}
-Notice that the expansion-internal binding of \li{tmp} has been alpha-varied to \li{tmp'} to avoid shadowing the application site binding of \li{tmp}.
-As such, the reference to \li{tmp} in the spliced expression refers, as intended, to the application site binding of \li{tmp}.
-
-For TLM providers, the benefit of this mechanism is that they can name the variables used internally within expansions freely, without worrying about whether their chosen identifiers might shadow those that a client might have used at the application site.
-There is no need for a mechanism that generates ``fresh variables''.
-
-TLM clients can, in turn, reliably reason about binding within every spliced expression without examining the expansion that the spliced expression appears within.
-
-The trade-off is that this prevents library providers from intentionally introducing bindings into spliced terms.
-For example, Haskell's derived form for monadic commands (i.e.
-\li{do}-notation) supports binding the result of executing a command to a variable that is then available in the subsequent commands in a command sequence.
-In VerseML, this cannot be expressed in the same way.
-Values can be communicated from the expansion to a spliced expression only via function arguments.
-%We will show an alternative formulation of Haskell's syntax for monadic commands that uses VerseML's anonymous function syntax to bind variables in Sec. \ref{sec:application-monadic-commands}. 
-We will return to this issue in Sec. \ref{sec:discussion}.
-
+Notice that the expansion-internal binding of \li{tmp} has been alpha-varied to \li{tmp'}.
+The reference to \li{tmp} in the spliced expression then refers, as intended, to the application site binding.
+
+For TLM providers, the benefit of this mechanism is that they can name the variables used internally within expansions freely.
+There is no need to explicitly deploy a mechanism that generates ``fresh variables''.
+TLM clients can, in turn, can reason abstractly about capture, i.e.
+without examining the expansion that the spliced term appears within.
+
+This does prevent library providers from intentionally introducing bindings into spliced terms.
+For example, Haskell's literal notation for monadic commands, i.e.
+\li{do}-notation,  supports binding the result of executing a command to a variable that is then available in the subsequent commands in the command sequence.
+Using TLMs in Reason, this notation cannot be expressed -- values can be communicated from the expansion to a spliced expression only via function arguments.
+It would be possible to distinguish spliced identifiers, much as we do spliced terms, and explicitly make them available to spliced expressions by extending the \li{SplicedE} and \li{SplicedT} constructors to take finite sets of spliced identifiers (following the basic approach taken by \citet{DBLP:conf/esop/HermanW08}.) However, the opinion of the authors is that this would unjustifiably increase the reasoning burden on clients when they encounter an unfamiliar literal form: \emph{might this new literal form be doing obscure things with binding too?} Haskell-style infix notation for monadic commands, where bind is \li{e >>= f}, can be expressed using TLMs and splicing.
+%We will show an alternative formulation of Haskell's syntax for monadic commands that uses Reason's anonymous function syntax to bind variables in Sec. \ref{sec:application-monadic-commands}.
+
+\vspace{-6px}
 \subsubsection{Context Dependence}\label{sec:context-dependence}
@@
 The proto-expansion validation process also ensures that variables that appear in the proto-expansion do not refer to bindings that appear either at the TLM definition or the application site.
-In other words, expansions must be completely \emph{context independent} -- they can make no assumptions about the surrounding context.
-
-A minimal example of a ``broken'' TLM that does not generate context-independent proto-expansions is below:
+In other words, expansions must be completely \emph{context independent} -- they can make no assumptions about the surrounding context whatsoever.
+A minimal example of a ``broken'' TLM that never generates context-independent proto-expansions is below:
 \begin{lstlisting}[numbers=none]
-  syntax $broken at rx by
-    static fn(_) => Success (Var "SSTRxESTR")
-  end
+  syntax $broken at rx by static { fun(_) => Success (Var "SSTRxESTR") };
 \end{lstlisting}
@@
 Indeed, we saw examples of spliced terms that referred to variables bound at the application site above.
-Because proto-expansions refer to spliced terms indirectly, enforcing context independence is straightforward -- we need only that the proto-expansion itself be closed, without considering the spliced terms.% In the next section, we will formalize this intuition.
+Because proto-expansions refer to spliced terms indirectly, enforcing context independence is straightforward -- we need only that the proto-expansion itself be closed.% In the next section, we will formalize this intuition.
 % The TLM provider can only refer to them opaquely.
@@
 One subtlety is that we assumed in the examples above that constructors like \li{H1Element} were available to the proto-expansion.
-In most dialects of ML, where datatype constructors are injected into the context as variables when the datatype is defined, this would violate context independence.
-As such, we will have to assume for now (and in our formalism in Sec.
-\ref{sec:setsms-formally}) that explicit injections into labeled sum types are available.
+In OCaml and other dialects of ML where datatype constructors are injected into the context as variables when the datatype is defined, this would violate context independence.
+As such, we will have to assume for now and in our formalism in Sec.
+\ref{sec:setsms-formally} that explicit injections into labeled sum types are available.
 Once we introduce module parameters in Sec.
@@
 This might appear to violate our prohibition on context-dependent expansions.
-This is not the case only because in VerseML, constructor labels are not variables or scoped symbols.
+This is not the case only because in Reason, constructor labels are not variables or scoped symbols.
 Syntactically, they must begin with a capital letter (like Haskell's datatype constructors).
@@
 We will return to this topic in Chapter \ref{chap:ptsms}.
-% Indeed, we used the label \li{Spliced} for two different recursive labeled sum types in Figure \ref{fig:candidate-exp-verseml}.
+% Indeed, we used the label \li{Spliced} for two different recursive labeled sum types in Figure \ref{fig:candidate-exp-Reason}.
 
@@
 Finally, proto-expansion validation maintains a reasonable \emph{typing discipline} by:
-\begin{enumerate}
-\item checking to ensure that the generated expansion is of the type specified in the TLM's type annotation; %For example, the type annotation on \li{$rx} is \li{at rx}, so proto-expansion validation ensures that the final expansion is of type \li{rx}.
-\item checking that the each spliced type is valid at the application site; 
-\item checking that the type annotation on each spliced expression is valid at the application site; and 
-\item checking each spliced expression against the provided type annotation.
+\begin{enumerate}[noitemsep,nolistsep]
+\item checking that the expansion is of the type specified by the TLM's type annotation; %For example, the type annotation on \li{$rx} is \li{at rx}, so proto-expansion validation ensures that the final expansion is of type \li{rx}.
+\item checking that each spliced type is valid; 
+\item checking that the type annotation on each spliced expression is valid; and 
+\item checking each spliced expression against the specified type annotation.
 \end{enumerate}
-
+%The OCaml type system is not strong enough to allow us to express only contextually well-typed syntax trees. In The details are discussed as future work in Sec. \ref{sec:discussion}.
  % This addresses the problem of reasoning abstractly about \textbf{Typing} described in Sec. \ref{sec:intro}, i.e.:
@@
 
+\vspace{-4px}
 \newcommand{\spTLMsSec}{Simple Pattern TLMs (spTLMs)}
@@
 \label{sec:sptsms}
+\vspace{-2px}
+
 Let us now briefly consider the topic of TLMs that expand to \emph{patterns}, rather than expressions.
@@
 \begin{lstlisting}[numbers=none]
-  match x with 
-  | $html `SURL<h1>{EURLySURL}</h1>EURL` -> y (* y : list(html) *)
-  | _ -> (* ... *)
-  end
+  let children_of_h1 x => 
+    switch x {
+    | $html [|SURL<h1>{EURLcsSURL}</h1>EURL|] -> Some cs; /* cs : list(html) */
+    | _ -> None; };
 \end{lstlisting}
+Any list pattern, including one generated by another TLM application, can appear where \li{cs} appears in the example pattern above.
+For longer \li{switch} expressions, the shorthand \li{switch x using $html} applies \li{$html} to every rule where the outermost pattern is of generalized literal form.
 
@@
 \begin{lstlisting}
-  syntax $html at html for patterns by 
-    static fn(b : body) : parse_result(proto_pat) => 
-      (* HTML pattern parser here *)
-  in (* ... *) end
+  syntax $html at html for patterns by static {
+    fun(b : body) : parse_result(proto_pat) => 
+      /* ... Ur/Web-style HTML pattern parser here ...  */
+  };
 \end{lstlisting}
@@
 Expression TLM definitions can also include a sort qualifier, \li{for expressions}, but this is the default if omitted.
-Note that we used the same name, \li{$html}, for this pattern TLM as for the expression TLM defined in the previous section.
+Notice that we used the same name, \li{$html}, for this pattern TLM as for the expression TLM defined in the previous section.
 This is possible because patterns and expressions are distinct sorts of terms.
 It does not make sense to apply an expression TLM in pattern position, and \emph{vice versa}.\footnote{The fact that some patterns look like expressions in the textual syntax is immaterial to this fundamental semantic distinction.
-Additionally, many expression-level constructs, e.g.
+Many expression-level constructs, e.g.
 lambdas, do not correspond even syntactically to patterns.}
 The return type of the parse function is \li{parse_result(proto_pat)}, rather than \li{parse_result(proto_expr)}.
-The type \li{proto_pat}, defined in part in Figure \ref{fig:CEPat}, classifies encodings of \emph{proto-patterns}.
-
-\subsection{Splicing}
+The type \li{proto_pat}, outlined in Fig.
+\ref{fig:CEPat}, classifies encodings of \emph{proto-patterns}.
+
+\begin{figure}[h]
+\vspace{-10px}
+\begin{lstlisting}[numbers=none]
+  type proto_pat = /* IMPORTANT: no variable pattern form! */
+                 | Wild
+                 | /* ... other standard pattern forms ... */
+                 | SplicedP(segment, proto_typ);
+\end{lstlisting}
+\vspace{-10px}
+\caption[Abbreviated definition of \li{proto_pat} in Reason]{Abbreviated definition of \li{proto_pat}.}
+\label{fig:CEPat}
+\vspace{-10px}
+\end{figure}
+
+% \subsection{Splicing}
 The constructor \li{SplicedP} serves much like \li{SplicedE} to allow a proto-pattern to refer indirectly to spliced pattern segments by their location within the literal body.
@@
 \begin{lstlisting}[numbers=none]
-  H1Element (_, spliced<5; 5; list(html)>)
+  H1Element (_, spliced<6; 7; list(html)>)
 \end{lstlisting}
 
-\begin{figure}
-\begin{lstlisting}[numbers=none]
-type proto_pat = (* no variable pattern form! *) 
-                 Wild
-               | (* ... other standard pattern forms ... *)
-               | SplicedP of segment * proto_typ
-\end{lstlisting}
-\vspace{-5px}
-\caption[Abbreviated definition of \li{proto_pat} in VerseML]{Abbreviated definition of \li{proto_pat}.}
-\label{fig:CEPat}
-\end{figure}
-
-\subsection{Proto-Pattern Validation}
-Proto-pattern validation serves, like proto-expression validation, to maintain the ability to reason abstractly about binding and typing.
-
-\subsubsection{Binding} To maintain a reasonable abstract binding discipline, i.e.
+% \subsection{Proto-Pattern Validation}
+% Proto-pattern validation serves, like proto-expression validation, to maintain the ability to reason abstractly about binding and typing.
+
+To maintain a reasonable abstract binding discipline, i.e.
 to allow clients to reason about variable binding without examining pattern TLM expansions directly, variable patterns can appear only within spliced patterns.
@@
 
-In our simple language, patterns have no way to refer to surrounding bindings (they only induce bindings).
-However, the type annotations on references to spliced patterns could refer to type variables, so we also need to enforce context independence in the manner discussed in the previous discussion.
-
-Patterns cannot bind variables for use in sub-patterns, so we do not face the problem of capture.
-%However, in languages with more complex patterns, e.g.
-those that support boolean expressions as guards, we would need to follow the approach described in the Sec.
-\ref{sec:uetsms-validation} to enforce context independence.
-
-\subsubsection{Typing} To maintain a reasonable abstract typing discipline, proto-pattern validation checks:
-\begin{enumerate}
-  \item that the final expansion is a pattern that matches values of the type specified in the TLM type annotation;
-  \item that each spliced pattern matches values of the type indicated in the splice summary; and 
-  \item that each of these types are themselves well-formed types at the application site.
+Patterns have no way to refer to surrounding bindings (they only induce bindings in other expressions, e.g.
+in the corresponding branch of the \li{switch} expression.) However, type annotations on references to spliced patterns could refer to type variables, so we also need to enforce context independence in the manner discussed in the previous section.
+% In languages like OCaml, which support arbitrary boolean guard expressions, we would also need to enforce both context independence and capture avoidance as discussed in the previous section for spliced expressions that end up in a guard expression.
+In our formalism, we do not support guard expressions.
+
+% \subsubsection{Typing} 
+To maintain a reasonable abstract typing discipline, proto-pattern validation checks:
+\begin{enumerate}[noitemsep,nolistsep]
+  \item that the final expansion is a pattern that matches values of the type specified by the TLM's type annotation;
+  \item that each spliced pattern matches values of the type indicated in the segmentation; and 
+  \item that each of these types are themselves well-formed types.
 \end{enumerate}
+
+\vspace{-4px}
 \newcommand{\seTLMsFormallySec}{Simple TLMs, Formally}
@@
 \label{sec:setsms-formally}
+\vspace{-2px}
+
+Before continuing on to consider more general parametric TLMs, let us develop a calculus of simple expression and pattern TLMs called $\miniVersePat$.
+This calculus consists of an \emph{unexpanded language}, or \emph{UL}, defined by typed expansion to an \emph{expanded language}, or \emph{XL}.
+Figs.
+\ref{fig:U-unexpanded-terms}  and \ref{fig:U-expanded-terms} summarize the syntax of the UL and  the XL, respectively.
+Programs are written as unexpanded expressions but evaluate as well-typed expanded expressions.
+We will start with a brief overview of our XL before turning in the remainder of the section on the UL and the typed expansion process.
+
+\vspace{-4px}
+\subsection{Expanded Language (XL)}\label{sec:s-XL}
+\vspace{-2px}
+The {XL} of $\miniVersePat$ forms a standard pure functional language with partial function types, quantification over types, recursive types, labeled product types and labeled sum types. 
+ The reader is directed to \emph{PFPL} \cite{pfple1} for a detailed introductory account of these standard constructs.
+We will only tersely summarize the statics and dynamics of the XL below because the particularities of the XL are not critical to the ideas we will introduce below.
+
+% \subsubsection{Statics of the Expanded Language}
+The \emph{statics of the XL} is defined by hypothetical judgements of the following form:
+\[\begin{array}{ll}
+% \textbf{Judgement Form} & \textbf{Description}\\
+\istypeU{\Delta}{\tau} & \text{$\tau$ is a well-formed type}\\
+%\isctxU{\Delta}{\Gamma} & \text{$\Gamma$ is a well-formed typing context assuming $\Delta$}\\
+\hastypeU{\Delta}{\Gamma}{e}{\tau} & \text{$e$ is assigned type $\tau$}\\
+\ruleType{\Delta}{\Gamma}{r}{\tau}{\tau'} & \text{$r$ takes values of type $\tau$ to values of type $\tau'$}\\
+\patType{\pctx}{p}{\tau} & \text{$p$ matches values of type $\tau$ and generates hypotheses $\pctx$} 
+\end{array}\]
+\emph{Type formation contexts}, $\Delta$, are finite sets of hypotheses of the form $\Dhyp{t}$.
+%Empty finite sets are written $\emptyset$, or omitted entirely within judgements, and non-empty finite sets are written as comma-separated finite sequences identified up to exchange and contraction.
+We write $\Delta, \Dhyp{t}$ when $\Dhyp{t} \notin \Delta$ for $\Delta$ extended with the hypothesis $\Dhyp{t}$.
+%Finite sets are written as finite sequences identified up to exchange.% We write $\Dcons{\Delta}{\Delta'}$ for the union of $\Delta$ and $\Delta'$.
+\emph{Typing contexts}, $\Gamma$, are finite functions that map each variable $x \in \domof{\Gamma}$, where $\domof{\Gamma}$ is a finite set of variables, to the hypothesis $\Ghyp{x}{\tau}$, for some $\tau$.
+The judgements above are inductively defined in the supplemental material and validate standard lemmas, also given in the supplement.
+%Empty typing contexts are written $\emptyset$, or omitted entirely within judgements, and non-empty typing contexts are written as finite sequences of hypotheses identified up to exchange and contraction.
+We write $\Gamma, \Ghyp{x}{\tau}$, when $x \notin \domof{\Gamma}$, for the extension of $\Gamma$ with a mapping from $x$ to $\Ghyp{x}{\tau}$, and $\Gcons{\Gamma}{\Gamma'}$ when $\domof{\Gamma} \cap \domof{\Gamma'} = \emptyset$ for the typing context mapping each $x \in \domof{\Gamma} \cup \domof{\Gamma'}$ to $x : \tau$ if $x : \tau \in \Gamma$ or $x : \tau \in \Gamma'$.
+% We write $\isctxU{\Delta}{\Gamma}$ if every type in $\Gamma$ is well-formed relative to $\Delta$.
+% \begin{definition}[Typing Context Formation] \label{def:isctxU}
+% $\isctxU{\Delta}{\Gamma}$ iff for each hypothesis $x : \tau \in \Gamma$, we have $\istypeU{\Delta}{\tau}$.
+% \end{definition}
+
+% \begin{subequations}\label{rules:istypeU}
+% \begin{equation*}\label{rule:istypeU-var}
+% \inferrule{ }{\istypeU{\Delta, \Dhyp{t}}{t}}
+% \end{equation*}
+% \begin{equation*}\label{rule:istypeU-parr}
+% \inferrule{
+%   \istypeU{\Delta}{\tau_1}\\
+%   \istypeU{\Delta}{\tau_2}
+% }{\istypeU{\Delta}{\aparr{\tau_1}{\tau_2}}}
+% \end{equation*}
+% \begin{equation*}\label{rule:istypeU-all}
+%   \inferrule{
+%     \istypeU{\Delta, \Dhyp{t}}{\tau}
+%   }{
+%     \istypeU{\Delta}{\aall{t}{\tau}}
+%   }
+% \end{equation*}
+% \begin{equation*}\label{rule:istypeU-rec}
+%   \inferrule{
+%     \istypeU{\Delta, \Dhyp{t}}{\tau}
+%   }{
+%     \istypeU{\Delta}{\arec{t}{\tau}}
+%   }
+% \end{equation*}
+% \begin{equation*}\label{rule:istypeU-prod}
+%   \inferrule{
+%     \{\istypeU{\Delta}{\tau_i}\}_{i \in \labelset}
+%   }{
+%     \istypeU{\Delta}{\aprod{\labelset}{\mapschema{\tau}{i}{\labelset}}}
+%   }
+% \end{equation*}
+% \begin{equation*}\label{rule:istypeU-sum}
+%   \inferrule{
+%     \{\istypeU{\Delta}{\tau_i}\}_{i \in \labelset}
+%   }{
+%     \istypeU{\Delta}{\asum{\labelset}{\mapschema{\tau}{i}{\labelset}}}
+%   }
+% \end{equation*}
+% \end{subequations}
+% Premises of the form $\{{J}_i\}_{i \in \labelset}$ mean that for each $i \in \labelset$, the judgement ${J}_i$ must hold.
+
 \begin{figure}[t!]
 \begin{minipage}{\textwidth}
+\small
 $\arraycolsep=3pt\begin{array}{llcl}
@@
 }
-\vspace{-5px}
+\vspace{-6px}
 \label{fig:U-unexpanded-terms}
@@
 \begin{minipage}{\textwidth}
+\small
 $\arraycolsep=3pt\begin{array}{llcl}
 \mathsf{Typ} & \tau & ::= & t ~\vert~ \aparr{\tau}{\tau} ~\vert~ \aall{t}{\tau} ~\vert~ \arec{t}{\tau} ~\vert~ \aprod{\labelset}{\mapschema{\tau}{i}{\labelset}} ~\vert~ \asum{\labelset}{\mapschema{\tau}{i}{\labelset}}\\
-\mathsf{Exp} & e & ::= & x ~\vert~ \aelam{\tau}{x}{e} ~\vert~ \aeap{e}{e} ~\vert~ \aetlam{t}{e} ~\vert~ \aetap{e}{\tau} ~\vert~ \aefold{e} ~\vert~ \aetpl{\labelset}{\mapschema{e}{i}{\labelset}} ~\vert~ \aein{\ell}{e} \\
-& & \vert & \aematchwith{n}{e}{\seqschemaX{r}}\\
+\mathsf{Exp} & e & ::= & x ~\vert~ \aelam{\tau}{x}{e} ~\vert~ \aeap{e}{e} ~\vert~ \aetlam{t}{e} ~\vert~ \aetap{e}{\tau} ~\vert~ \aefold{e} ~\vert~ \aetpl{\labelset}{\mapschema{e}{i}{\labelset}}  \\
+& & \vert & \aein{\ell}{e} ~\vert~ \aematchwith{n}{e}{\seqschemaX{r}}\\
 \mathsf{Rule} & r & ::= & \aematchrule{p}{e}\\
@@
 \vspace{-10px}
-\caption[Syntax of the XL of $\miniVersePat$]{Syntax of the $\miniVersePat$ (XL).
+\caption[Syntax of the XL of $\miniVersePat$]{Syntax of the $\miniVersePat$ expanded language (XL).
 XL terms are \emph{abstract binding trees} (ABTs) identified up to alpha-equivalence, so we follow the syntactic conventions of \citet{pfple1}.
@@
 \end{figure}
-
-Before continuing on to consider parametric TLMs, let us develop a calculus of simple expression and pattern TLMs called $\miniVersePat$.
-It is organized as an \emph{unexpanded language}, or \emph{UL}, defined by typed expansion to an \emph{expanded language}, or \emph{XL}.
-Figures \ref{fig:U-unexpanded-terms}  and \ref{fig:U-expanded-terms} summarize the syntax of the UL and  the XL, respectively.
-Programs are written as unexpanded expressions but evaluate as well-typed expanded expressions.
-We will start with a brief overview of our XL before turning in the remainder of the section on the UL and the typed expansion process.
-
-\subsection{Expanded Language (XL)}
-The {XL} of $\miniVersePat$ forms a standard pure functional language with partial function types, quantification over types, recursive types, labeled product types and labeled sum types. 
- The reader is directed to \emph{PFPL} \cite{pfple1} for a detailed introductory account of these standard constructs.
-We will only tersely summarize the statics and dynamics of the XL below.
-The particularities of the XL are not critical to the ideas we will introduce below.
-
-\subsubsection{Statics of the Expanded Language}
-The \emph{statics of the XL} is defined by hypothetical judgements of the following form:
-\[\begin{array}{ll}
-% \textbf{Judgement Form} & \textbf{Description}\\
-\istypeU{\Delta}{\tau} & \text{$\tau$ is a well-formed type}\\
-%\isctxU{\Delta}{\Gamma} & \text{$\Gamma$ is a well-formed typing context assuming $\Delta$}\\
-\hastypeU{\Delta}{\Gamma}{e}{\tau} & \text{$e$ is assigned type $\tau$}\\
-\ruleType{\Delta}{\Gamma}{r}{\tau}{\tau'} & \text{$r$ takes values of type $\tau$ to values of type $\tau'$}\\
-\patType{\pctx}{p}{\tau} & \text{$p$ matches values of type $\tau$ and generates hypotheses $\pctx$} 
-\end{array}\]
-\emph{Type formation contexts}, $\Delta$, are finite sets of hypotheses of the form $\Dhyp{t}$.
-%Empty finite sets are written $\emptyset$, or omitted entirely within judgements, and non-empty finite sets are written as comma-separated finite sequences identified up to exchange and contraction.
-We write $\Delta, \Dhyp{t}$ when $\Dhyp{t} \notin \Delta$ for $\Delta$ extended with the hypothesis $\Dhyp{t}$.
-%Finite sets are written as finite sequences identified up to exchange.% We write $\Dcons{\Delta}{\Delta'}$ for the union of $\Delta$ and $\Delta'$.
-\emph{Typing contexts}, $\Gamma$, are finite functions that map each variable $x \in \domof{\Gamma}$, where $\domof{\Gamma}$ is a finite set of variables, to the hypothesis $\Ghyp{x}{\tau}$, for some $\tau$.
-The judgements above are inductively defined in the supplemental material and validate standard lemmas, also given in the supplement: Weakening, Substitution and Decomposition.
-%Empty typing contexts are written $\emptyset$, or omitted entirely within judgements, and non-empty typing contexts are written as finite sequences of hypotheses identified up to exchange and contraction.
-We write $\Gamma, \Ghyp{x}{\tau}$, when $x \notin \domof{\Gamma}$, for the extension of $\Gamma$ with a mapping from $x$ to $\Ghyp{x}{\tau}$, and $\Gcons{\Gamma}{\Gamma'}$ when $\domof{\Gamma} \cap \domof{\Gamma'} = \emptyset$ for the typing context mapping each $x \in \domof{\Gamma} \cup \domof{\Gamma'}$ to $x : \tau$ if $x : \tau \in \Gamma$ or $x : \tau \in \Gamma'$.
-% We write $\isctxU{\Delta}{\Gamma}$ if every type in $\Gamma$ is well-formed relative to $\Delta$.
-% \begin{definition}[Typing Context Formation] \label{def:isctxU}
-% $\isctxU{\Delta}{\Gamma}$ iff for each hypothesis $x : \tau \in \Gamma$, we have $\istypeU{\Delta}{\tau}$.
-% \end{definition}
-
-% \begin{subequations}\label{rules:istypeU}
-% \begin{equation*}\label{rule:istypeU-var}
-% \inferrule{ }{\istypeU{\Delta, \Dhyp{t}}{t}}
-% \end{equation*}
-% \begin{equation*}\label{rule:istypeU-parr}
-% \inferrule{
-%   \istypeU{\Delta}{\tau_1}\\
-%   \istypeU{\Delta}{\tau_2}
-% }{\istypeU{\Delta}{\aparr{\tau_1}{\tau_2}}}
-% \end{equation*}
-% \begin{equation*}\label{rule:istypeU-all}
-%   \inferrule{
-%     \istypeU{\Delta, \Dhyp{t}}{\tau}
-%   }{
-%     \istypeU{\Delta}{\aall{t}{\tau}}
-%   }
-% \end{equation*}
-% \begin{equation*}\label{rule:istypeU-rec}
-%   \inferrule{
-%     \istypeU{\Delta, \Dhyp{t}}{\tau}
-%   }{
-%     \istypeU{\Delta}{\arec{t}{\tau}}
-%   }
-% \end{equation*}
-% \begin{equation*}\label{rule:istypeU-prod}
-%   \inferrule{
-%     \{\istypeU{\Delta}{\tau_i}\}_{i \in \labelset}
-%   }{
-%     \istypeU{\Delta}{\aprod{\labelset}{\mapschema{\tau}{i}{\labelset}}}
-%   }
-% \end{equation*}
-% \begin{equation*}\label{rule:istypeU-sum}
-%   \inferrule{
-%     \{\istypeU{\Delta}{\tau_i}\}_{i \in \labelset}
-%   }{
-%     \istypeU{\Delta}{\asum{\labelset}{\mapschema{\tau}{i}{\labelset}}}
-%   }
-% \end{equation*}
-% \end{subequations}
-% Premises of the form $\{{J}_i\}_{i \in \labelset}$ mean that for each $i \in \labelset$, the judgement ${J}_i$ must hold.
-
 % \begin{subequations}\label{rules:hastypeU}
@@
 
-\subsubsection{Structural Dynamics}\label{sec:dynamics-U}
-The \emph{structural dynamics}, a.k.a.
-the \emph{structural operational semantics} \cite{DBLP:journals/jlp/Plotkin04a}, of $\miniVersePat$ is defined as a transition system by judgements of the following form:
-\[\begin{array}{ll}
-% \textbf{Judgement Form} & \textbf{Description}\\
-\stepsU{e}{e'} & \text{$e$ transitions to $e'$}\\
-\isvalU{e} & \text{$e$ is a value}
-\end{array}\]
-We also define auxiliary judgements for \emph{iterated transition}, $\multistepU{e}{e'}$, and \emph{evaluation} to a value, $\evalU{e}{e'}$.
-Additional auxiliary judgements, not shown, are needed to define the dynamics of pattern matching, but they do not appear directly in our subsequent developments, so we omit them.
-We assume an eager dynamics for simplicity and again assume that standard lemmas, stated in the supplement, hold: Canonical Forms, Progress and Preservation.
-
-\subsection{Syntax of the Unexpanded Language}\label{sec:syntax-U}
+% \vspace{-4px}
+% \subsubsection{Evaluation Semantics}\label{sec:dynamics-U}
+% \vspace{-2px}
+The \emph{evaluation semantics} of $\miniVersePat$ is organized around the judgements $\isvalU{e}$, which says that $e$ is a value, and $\evalU{e}{e'}$, which says that $e$ evaluates to the value $e'$.
+Additional  judgements, not shown, are needed to define the dynamics of pattern matching, but they do not appear directly in our subsequent developments, so we omit them.
+We assume an eager dynamics and the standard type safety theorem (see supplement.)
+
+\vspace{-4px}
+\subsection{Syntax of the Unexpanded Language}\label{sec:syntax-U}\label{sec:s-UL}
+\vspace{-2px}
 Unexpanded types and expressions are simple inductive structures.
-They are \textbf{not} abstract binding trees -- we do \textbf{not} define notions of renaming, alpha-equivalence or substitution for unexpanded terms.
+Unlike expanded types and expressions, they are \textbf{not} abstract binding trees -- we do \textbf{not} define the standard notions of renaming, alpha-equivalence or substitution for unexpanded terms.
 This is because unexpanded expressions remain ``partially parsed'' due to the presence of literal bodies, $b$, from which spliced terms might be extracted during typed expansion.
 In fact, unexpanded types and expressions do not involve variables at all, but rather \emph{type identifiers}, $\ut$, and \emph{expression identifiers}, $\ux$.
-Identifiers are given meaning by expansion to variables during typed expansion, as we will see below.
-This distinction between identifiers and variables is technically crucial.
+Identifiers are given meaning by expansion to variables during typed expansion, as we will see.
+This distinction between identifiers and variables is technically crucial to our developments, and we return to it below.
 %We \textbf{cannot} adopt the usual definitions of $\alpha$-renaming of identifiers, because unexpanded types and expressions are still in a ``partially parsed'' state -- the literal bodies, $b$, within an unexpanded expression might contain spliced subterms that are ``surfaced'' by a TLM only during typed expansion, as we will detail below.
@@
 %\] and so on for the remaining expanded expression forms.
+Most of the unexpanded forms in Figure \ref{fig:U-unexpanded-terms}  mirror the expanded forms.
+We refer to these as the \emph{common forms}.
+The mapping from expanded forms to common unexpanded forms is defined explicitly in the supplement.
 
@@
 %Our paper on Wyvern defines a textual syntax for a similar system \cite{TSLs}. 
-Instead, we need only posit the existence of partial metafunctions $\parseUTypF{b}$, $\parseUExpF{b}$ and $\parseUPatF{b}$  that go from character sequences, $b$, to unexpanded types, expressions and patterns, respectively.
+Instead, we need only posit partial metafunctions $\parseUTypF{b}$, $\parseUExpF{b}$ and $\parseUPatF{b}$  that go from character sequences, $b$, to unexpanded types, expressions and patterns (the supplement states the necessary condition in full.)
 % \begingroup
 % \def\thetheorem{\ref{condition:textual-representability-SES}}
-\begin{condition}[Textual Representability] ~
-\begin{enumerate}[nolistsep]
-\item For each $\utau$, there exists $b$ such that $\parseUTyp{b}{\utau}$. 
-\item For each $\ue$, there exists $b$ such that $\parseUExp{b}{\ue}$.
-\item For each $\upv$, there exists $b$ such that $\parseUPat{b}{\upv}$.
-\end{enumerate}
-\end{condition}
+% \begin{condition}[Textual Representability] ~
+% \begin{enumerate}[nolistsep]
+% \item For each $\utau$, there exists $b$ such that $\parseUTyp{b}{\utau}$. 
+% \item For each $\ue$, there exists $b$ such that $\parseUExp{b}{\ue}$.
+% \item For each $\upv$, there exists $b$ such that $\parseUPat{b}{\upv}$.
+% \end{enumerate}
+% \end{condition}
 % \endgroup
 
-\subsection{Typed Expansion}\label{sec:typed-expansion-U}
-Unexpanded terms are checked and expanded simultaneously according to the \emph{typed expansion judgements}:
-\[\begin{array}{ll}
+\subsection{Typed Expansion}\label{sec:typed-expansion-U}\label{sec:s-TE}
+Unexpanded terms are checked and expanded simultaneously according to the central judgements of our calculus, the \emph{typed expansion judgements}:
+\vspace{-3px}\[\begin{array}{ll}
 % \textbf{Judgement Form} & \textbf{Description}\\
@@
 \patExpands{\upctx}{\uPhi}{\upv}{p}{\tau} & \text{$\upv$ has expansion $p$ matching against $\tau$ generating hypotheses $\upctx$}
-\end{array}\]
+\end{array}\]\vspace{-3px}
 %\newcommand{\gray}[1]{{\color{gray} #1}}
 
-These judgements are inductively defined in the supplement. We will reproduce the interesting rules below.
+% These judgements are inductively defined in the supplement. 
 % \begingroup 
@@
 
-\paragraph{Common Forms} Most of the unexpanded forms in Figure \ref{fig:U-unexpanded-terms}  mirror the expanded forms.
-We refer to these as the \emph{common forms}.
-The mapping from expanded forms to common unexpanded forms is defined explicitly in the supplement.
-The typed expansion rules that handle these common forms mirror the corresponding typing rules. For example, the rules for variables and lambdas are reproduced below:
+The typed expansion rules that handle common forms mirror the corresponding typing rules.
+The \emph{expression TLM context}, $\uPsi$, and the \emph{pattern TLM context}, $\uPhi$, pass through these rules opaquely.
+For example, the rules for variables and lambdas are below, and the remainder are in the supplement:
 %Each of these rules is based on the corresponding typing rule, i.e.
@@
 For example, the following typed expansion rules are based on the typing rules (\ref{rule:hastypeU-var}), (\ref{rule:hastypeU-lam}) and (\ref{rule:hastypeU-ap}), respectively:% for unexpanded expressions of variable, function and application form, respectively:
-\begin{mathpar}
+{\small\begin{mathpar}
   \inferrule[ee-id]{ }{\expandsUP{\uDelta}{\uGamma, \uGhyp{\ux}{x}{\tau}}{\uPsi}{\uPhi}{\ux}{x}{\tau}}
@@
   }{\expandsUPX{\lam{\ux}{\utau}{\ue}}{\aelam{\tau}{x}{e}}{\aparr{\tau}{\tau'}}}
-\end{mathpar}
-
-The only subtlety has to do with the relationship between identifiers in the UL and variables in the XL. To understand this, we must first describe in detail how unexpanded contexts work.
-
+\end{mathpar}}
+
+The only subtlety here has to do with the relationship between identifiers, $\ux$, in the UL and variables, $x$, in the XL.
+To understand this, we must first describe in detail how unexpanded contexts work.
 \emph{Unexpanded typing contexts}, $\uGamma$, are pairs of the form $\uGG{\uG}{\Gamma}$, where $\uG$ is an \emph{expression identifier expansion context}, and $\Gamma$ is a standard typing context.
@@
 To develop an intuition for why the update operation is necessary, it is instructive to inspect the derivation of the expansion of the unexpanded expression $\lam{\ux}{\utau}{\lam{\ux}{\utau}{\ux}}$ to $\aelam{\tau}{x_1}{\aelam{\tau}{x_2}{x_2}}$ assuming $\expandsTU{\uDelta}{\utau}{\tau}$:
-\begin{mathpar}
+{\small\begin{mathpar}
 \inferrule{
@@
 }~\textsc{ee-lam}
-\end{mathpar}
+\end{mathpar}}
+\noindent
 Notice that when Rule \textsc{ee-lam} is applied, the type identifier expansion context is updated but the typing context is extended with a (necessarily fresh) variable, first $x_1$ then $x_2$.
@@
 
-\paragraph{Well-Typed Expansion} Before we continue, let us state an important invariant: that typed expansion produces a well-typed expression.
+Before we continue, let us state an important invariant: that typed expression expansion produces a well-typed expression.
 \begin{theorem}[Typed Expression Expansion]\label{thm:typed-expansion-short-U}
@@
 The details are in the supplement.
-The  rules of particular interest are the rules governing TLM definitions and TLM application, which are covered in the next two subsections.
+The  rules of particular interest are the rules governing TLM definitions and TLM application, which are the topic of the remainder of this section.
 % \endgroup
@@
 
-\subsection{TLM Definitions}\label{sec:U-uetsm-definition}
-An unexpanded expression of seTLM definition form, $\uesyntaxq{\tsmv}{\utau}{\eparse}{\ue}$, 
+\vspace{-4px}
+\subsection{TLM Definitions}\label{sec:U-uetsm-definition}\label{sec:s-TLM-def}
+\vspace{-2px}
+% An unexpanded expression of seTLM definition form, $\uesyntaxq{\tsmv}{\utau}{\eparse}{\ue}$, 
 %The operational form corresponding to this stylized form is \[\audefuetsm{\utau}{\eparse}{\tsmv}{\ue}\]
- defines an {seTLM} identified as $\tsmv$ with \emph{unexpanded type annotation} $\utau$ and \emph{parse function} $\eparse$ for use within $\ue$. The rule below defines typed expansion of this form:
+ % defines an {seTLM} identified as $\tsmv$ with \emph{unexpanded type annotation} $\utau$ and \emph{parse function} $\eparse$ for use within $\ue$. 
+ The rule below defines typed expansion of the seTLM definition form:
 % \begin{subequations}[resume]
@@
 % \end{equation*}
-\begin{mathpar}
+{\small\begin{mathpar}
 \inferrule[ee-def-setsm]{
@@
 }
-\end{mathpar}
+\end{mathpar}}
 % \end{subequations}
-The premises of this rule can be understood as follows, in order:
-\begin{enumerate}
-\item The first premise expands the unexpanded type annotation.
-
-\item The second premise checks that $\eparse$ is a closed expanded function\footnote{In Sec.
+\vspace{-3px}
+
+The first premise expands the unexpanded type annotation.
+The second premise checks that $\eparse$ is a closed expanded function\footnote{In Sec.
 \ref{sec:static-eval}, we add the machinery necessary for parse functions that are neither closed nor yet expanded.} of type $\aparr{\tBody}{\tParseResultExp}$.
@@
 The type abbreviated $\tBody$ classifies encodings of literal bodies, $b$.
-The mapping from literal bodies to values of type $\tBody$ is defined by the \emph{body encoding judgement} $\encodeBody{b}{\ebody}$.
-An inverse mapping is defined   by the \emph{body decoding judgement} $\decodeBody{\ebody}{b}$.
+The mapping from literal bodies, $b$, to values of type $\tBody$ is defined by the \emph{body encoding judgement} $\encodeBody{b}{\ebody}$.
+An inverse mapping can also be defined   by the \emph{body decoding judgement} $\decodeBody{\ebody}{b}$.
 Rather than defining $\tBody$ explicitly, and these judgements inductively against that definition (which would be tedious and uninteresting), it suffices to take as a condition that there is an isomorphism between literal bodies and values of type $\tBody$ mediated by these judgements (see supplement.)
 
-The return type of the parse function, $\tParseResultExp$, abbreviates a labeled sum type that distinguishes parse errors from successful parses:
-\begin{align*}
-L_\mathtt{SE} & \defeq \lbltxt{ParseError}, \lbltxt{SuccessE}\\
-\tParseResultExp & \defeq \asum{\labelset_\mathtt{SE}}{
+The return type, $\tParseResultExp$, abbreviates a labeled sum type that distinguishes parse errors from successful parses:
+{$
+% L_\mathtt{SE} & \defeq \lbltxt{ParseError}, \lbltxt{SuccessE}\\ \asumNL{
   \mapitem{\lbltxt{ParseError}}{\prodt{}}, 
   \mapitem{\lbltxt{SuccessE}}{\tCEExp}
-}
-\end{align*} %[\mapitem{\lbltxt{ParseError}}{\prodt{}}, \mapitem{\lbltxt{SuccessE}}{\tCEExp}]
+$}.
+% \end{align*}} %[\mapitem{\lbltxt{ParseError}}{\prodt{}}, \mapitem{\lbltxt{SuccessE}}{\tCEExp}]
 % \]
 
-The type abbreviated $\tCEExp$ classifies encodings of \emph{proto-expressions}, $\ce$ (pronounced ``grave $e$''.) The syntax of proto-expressions, defined in Figure \ref{fig:U-candidate-terms}, will be described when we describe proto-expansion validation in Sec.
+The type abbreviated $\tCEExp$ classifies encodings of \emph{proto-expressions}, $\ce$ (pronounced ``grave $e$''.) The syntax of proto-expressions, defined in Fig.
+\ref{fig:U-candidate-terms}, will be described when we describe proto-expansion validation in Sec.
 \ref{sec:ce-syntax-U}.
@@
 An inverse mapping is defined by the \emph{proto-expression decoding judgement}, $\decodeCondE{e}{\ce}$.
-Again, rather than picking a particular definition of $\tCEExp$ and defining the judgements above inductively against it, we take as a condition that there is an isomorphism between values of type $\tCEExp$ and proto-expressions mediated by these judgements (see supplement.)
-\item The third premise of Rule \textsc{ee-def-setsm} evaluates the parse function to a value.
-This is not semantically necessary, but it is the choice one would expect to make assuming an eagerly evaluated language.
-\item The final premise of Rule \textsc{ee-def-setsm} extends the seTLM context, $\uPsi$, with the newly determined {seTLM definition}, and proceeds to assign a type, $\tau'$, and expansion, $e$, to $\ue$.
-The conclusion of the rule then assigns this type and expansion to the seTLM definition as a whole.% i.e.
+Again, rather than picking a particular definition of $\tCEExp$ and defining the judgements above inductively against it, we take as a condition that there is an isomorphism between values of type $\tCEExp$ and closed proto-expressions mediated by these judgements (see supplement.)
+
+The third premise of Rule \textsc{ee-def-setsm} evaluates the parse function to a value. This is not semantically necessary, but it is the choice one would expect to make in an eager language.
+
+The final premise of Rule \textsc{ee-def-setsm} extends the expression TLM context, $\uPsi$, with the newly determined {seTLM definition}, and proceeds to assign a type, $\tau'$, and expansion, $e$, to $\ue$.
+The conclusion of the rule then assigns this type and expansion to the seTLM definition as a whole.
+% i.e.
 TLMs define behavior that is relevant during typed expansion, but not during evaluation.
-
-\emph{seTLM contexts}, $\uPsi$, are of the form $\uAS{\uA}{\Psi}$, where $\uA$ is a \emph{TLM identifier expansion context} and $\Psi$ is an \emph{seTLM definition context}.
+{Expression TLM contexts}, $\uPsi$, are of the form $\uAS{\uA}{\Psi}$, where $\uA$ is a \emph{TLM identifier expansion context} and $\Psi$ is an \emph{expression TLM definition context}.
 
@@
 
-An {seTLM definition context}, $\Psi$, is a finite function mapping each TLM name $a \in \domof{\Psi}$ to an \emph{expanded seTLM definition}, $\xuetsmbnd{a}{\tau}{\eparse}$, where $\tau$ is the seTLM's type annotation, and $\eparse$ is its parse function.
+An {expression TLM definition context}, $\Psi$, is a finite function mapping each TLM name $a \in \domof{\Psi}$ to an \emph{expanded seTLM definition}, $\xuetsmbnd{a}{\tau}{\eparse}$, where $\tau$ is the seTLM's type annotation, and $\eparse$ is its parse function.
 We define $\uPsi, \uShyp{\tsmv}{a}{\tau}{\eparse}$, when $\uPsi=\uAS{\uA}{\Psi}$, as an abbreviation of $\uAS{\ctxUpdate{\uA}{\tsmv}{a}}{\Psi, \xuetsmbnd{a}{\tau}{\eparse}}$.
@@
 
-\end{enumerate}
-
-The spTLM definition form, $\usyntaxup{\tsmv}{\utau}{\eparse}{\ue}$, operates analagously.
+% \end{enumerate}
+
+The spTLM definition form operates analagously.
 The rule is reproduced below, and the details, which mirror those for seTLMs, are in the supplement.
 Notice that the spTLM context, $\uPhi$, rather than the $\uPsi$ is updated.
-This allows expression and pattern TLMs to share an identifier.
-\begin{mathpar}
+This allows expression and pattern TLMs to use the same identifiers.
+{\vspace{-3px}\small\begin{mathpar}
 \inferrule[ee-def-sptsm]{
@@
 }
-\end{mathpar}
+\end{mathpar}}
+\vspace{-3px}
 
@@
 
-\subsection{TLM Application}\label{sec:U-uetsm-application}
+\vspace{-4px}
+\subsection{TLM Application}\label{sec:U-uetsm-application}\label{sec:s-TLM-ap}
+\vspace{-2px}
 The unexpanded expression form for applying an seTLM named $\tsmv$ to a literal form with literal body $b$ is $\utsmap{\tsmv}{b}$. 
@@
 % \end{equation*}
-\begin{mathpar}
+{\small\begin{mathpar}
 \inferrule[ee-ap-setsm]{
-  \uPsi = \uPsi', \uShyp{\tsmv}{a}{\tau}{\eparse}\\\\
   \encodeBody{b}{\ebody}\\
@@
 }{
-  \expandsUP{\uDelta}{\uGamma}{\uPsi}{\uPhi}{\utsmap{\tsmv}{b}}{e}{\tau}
+  \expandsUP{\uDelta}{\uGamma}{\uPsi', \uShyp{\tsmv}{a}{\tau}{\eparse}}{\uPhi}{\utsmap{\tsmv}{b}}{e}{\tau}
 }
-\end{mathpar}
-The premises of this rule can be understood as follows, in order:
-\begin{enumerate}
-\item The first premise ensures that $\tsmv$ has been defined and extracts the type annotation and parse function.
-\item The second premise determines the encoding of the literal body, $\ebody$, which is a value of type $\tBody$.
-\item The third premise applies the parse function $\eparse$ to the encoding of the literal body.
+\end{mathpar}}
+
+The first premise determines the encoding of the literal body, $\ebody$, which, as discussed above, is a value of type $\tBody$.
+
+The second premise applies the parse function $\eparse$ to the encoding of the literal body.
 If parsing succeeds, i.e.
-a value of the form $\aein{\mathtt{SuccessE}}{\ecand}$ results from evaluation, then $\ecand$ will be a value of type $\tCEExp$ (assuming a well-formed seTLM context, by application of the Preservation assumption.) We call $\ecand$ the \emph{encoding of the proto-expansion}.
-If the parse function produces a value labeled $\lbltxt{ParseError}$, then typed expansion fails.
-Formally, no rule is necessary to handle this case.
-
-\item The fourth premise decodes the encoding of the proto-expansion to produce the {proto-expansion} itself.
-
-\item The fifth premise determines the segmentation, $\segof{\ce}$, and ensures that it is valid with respect to $b$ via the predicate $\segOK{\psi}{b}$, which  checks that each segment in the finite set of segments $\psi$ has non-negative length and is within bounds of $b$, and that the segments in $\psi$ do not overlap.
-
-\item The final premise \emph{validates} the proto-expansion and simultaneously generates the \emph{final expansion}, $e$, which appears in the conclusion of the rule.
-The proto-expression validation judgement is discussed next.
-\end{enumerate}
+a value of the form $\aein{\mathtt{SuccessE}}{\ecand}$ results from evaluation, then $\ecand$ will be a value of type $\tCEExp$ (assuming a well-formed expression TLM context, by application of the Type Safety assumption.) We call $\ecand$ the \emph{encoding of the proto-expansion}.
+If the parse function produces a value labeled $\lbltxt{ParseError}$, then typed expansion fails and formally, no rule is necessary.
+
+The third premise decodes the encoding of the proto-expansion.
+
+The fourth premise determines the segmentation of the proto-expansion, $\segof{\ce}$, and ensures that it is valid with respect to $b$ via the predicate $\segOK{\psi}{b}$, which  checks that each segment in the finite set of segments $\psi$ has non-negative length and is within bounds of $b$, and that the segments in $\psi$ do not overlap.
+
+The final premise \emph{validates} the proto-expansion and simultaneously generates the \emph{final expansion}, $e$, which appears in the conclusion of the rule.
+The proto-expression validation judgement is defined in the next subsection.
 
 The typed pattern expansion rule governing spTLM application is analagous:
-\begin{mathpar}
+{\small\begin{mathpar}
 \inferrule[pe-ap-sptsm]{
-  \uPhi = \uPhi', \uPhyp{\tsmv}{a}{\tau}{\eparse}\\\\
   \encodeBody{b}{\ebody}\\
@@
 }{
-  \patExpands{\upctx}{\uPhi}{\utsmap{\tsmv}{b}}{p}{\tau}
+  \patExpands{\upctx}{\uPhi', \uPhyp{\tsmv}{a}{\tau}{\eparse}}{\utsmap{\tsmv}{b}}{p}{\tau}
 }
-\end{mathpar}
+\end{mathpar}}
 
@@
 
-\subsection{Proto-Expansion Validation}\label{sec:ce-validation-U}\label{sec:ce-syntax-U}
-
-\begin{figure}[h]
+\vspace{-4px}
+\subsection{Proto-Expansion Validation}\label{sec:ce-validation-U}\label{sec:ce-syntax-U}\label{sec:s-PEV}
+\vspace{-2px}
+
+\begin{figure}
 \vspace{-10px}
 \begin{minipage}{\textwidth}
+\small
 $\arraycolsep=3pt\begin{array}{llcl}
@@
 \acerec{t}{\ctau} ~\vert~
-\aceprod{\labelset}{\mapschema{\ctau}{i}{\labelset}} ~\vert~
-\acesum{\labelset}{\mapschema{\ctau}{i}{\labelset}} \\
-& & \vert & \acesplicedt{m}{n}\\
+\aceprod{\labelset}{\mapschema{\ctau}{i}{\labelset}} \\
+& & \vert & \acesum{\labelset}{\mapschema{\ctau}{i}{\labelset}} ~\vert~ \acesplicedt{m}{n}\\
 \mathsf{PrExp} & \ce & ::= & x ~\vert~
@@
 
-The \emph{proto-expansion validation judgements} validate proto-expansions and simultaneously generate their final expansions:% are types and expanded expressions, respectively.
+Finally, the \emph{proto-expansion validation judgements} validate the proto-expansions generated by TLMs and simultaneously generate their final expansions:% are types and expanded expressions, respectively.
 \[\begin{array}{ll}
@@
 \end{array}\]
-The purpose of the \emph{splicing scenes} $\tscenev$, $\escenev$ and $\pscenev$ is to ``remember'', during proto-expansion validation, the contexts and the literal body from the TLM application site (cf.
+The purpose of the \emph{splicing scenes} $\tscenev$, $\escenev$ and $\pscenev$ is to ``remember'' the contexts and literal body from the TLM application site (cf.
 Rules \textsc{ee-ap-setsm} and \textsc{pe-ap-sptsm}) for when validation encounters spliced terms.
@@
 none of these rules can access the application site contexts.
-This maintains context independence.
+This maintains context independence (defined formally below.)
 
@@
 Proto-patterns, in contrast, arise only from pattern TLMs.
-There a variable proto-pattern form.
+There is not a variable proto-pattern form.
 
@@
 Let us first consider the rule for references to spliced unexpanded types:
-\begin{mathpar}
+{\small\begin{mathpar}
   \inferrule[ptv-spliced]{
@@
   }
-\end{mathpar}
+\end{mathpar}}
+
+\noindent
 This first premise of this rule parses out the requested segment of the literal body, $b$, to produce an unexpanded type, $\utau$.
@@
 The final premise requires that the application site type formation context is disjoint from the expansion-local type formation context.
-Because proto-expansions are identified up to alpha-equivalence, we can always discharge the final premise by alpha-varying the proto-expansion.
+Because proto-expansions are ABTs identified up to alpha-equivalence, we can always discharge the final premise by alpha-varying the proto-expansion.
 Collectively, this ensures that type variable capture does not occur (we will formally state this property in the next section.)
@@
 The rule for references to spliced unexpanded expressions is fundamentally analagous:
-\begin{mathpar}
+{\small\begin{mathpar}
 \inferrule[pev-spliced]{
@@
 }
-\end{mathpar}
+\end{mathpar}}
+
+\noindent
 We first splice out the requested segment.
@@
 The rule for references to spliced unexpanded patterns is entirely analagous:
-\begin{mathpar}
+{\small\begin{mathpar}
 \inferrule[ppv-spliced]{
@@
 }
-\end{mathpar}
-
+\end{mathpar}}
+
+\vspace{-4px}
 \subsection{Metatheory}\label{sec:s-metatheory}
-\subsubsection{Typed Expression Expansion} Let us now return to Theorem \ref{thm:typed-expansion-short-U}, the typed expression expansion theorem which was mentioned at the end of Sec.
+\vspace{-2px}
+\subsubsection{Typed Expansion} Let us now return to Theorem \ref{thm:typed-expansion-short-U}, the typed expression expansion theorem that was mentioned at the end of Sec.
 \ref{sec:typed-expansion-U}.
@@
 % \endgroup
-The additional second clause simply states that the final expansion produced by proto-expression validation is well-typed under the combined application site and expansion-internal context (spliced terms are distinguished only in the proto-expansion.) Such combined contexts can only be formed if the constituents are disjoint.
+The additional second clause simply states that the final expansion produced by proto-expression validation is well-typed under the combined application site and expansion-internal context (because spliced terms are distinguished only in the proto-expansion, not in the final expansion.) Such combined contexts can only be formed if the constituents are disjoint.
 
@@
 The details are given in the supplemental material.
-
-\subsubsection{Typed Pattern Expansion} A similar argument is needed to prove Typed Pattern Expansion:
+A similar argument is needed to prove Typed Pattern Expansion:
 \begin{theorem}[Typed Pattern Expansion (Strong)] ~
@@
 
+\vspace{-8px}
 \subsubsection{seTLM Reasoning Principles} 
-The following theorem summarizes the abstract reasoning principles that programmers can rely on when applying an seTLM. Informal descriptions of important clauses are given inline. 
+The following theorem summarizes the abstract reasoning principles that programmers can rely on when applying an seTLM. Informal descriptions of the labeled clauses are given inline, in gray boxes. 
 % \begingroup
 % \def\thetheorem{\ref{thm:tsc-SES}}
-\begin{theorem}[seTLM Abstract Reasoning Principles]
+\vspace{-4px}
+\begin{theorem}[seTLM Reasoning Principles]\label{thm:setlm-reasoning}
 If $\expandsU{\uDD{\uD}{\Delta}}{\uGG{\uG}{\Gamma}}{\uPsi}{\utsmap{\tsmv}{b}}{e}{\tau}$ then:
@@
   \begin{quote}
+  \begin{grayparbox}
      The type of the expansion is consistent with the type annotation on the applied seTLM definition.
+  \end{grayparbox}
   \end{quote}
@@
         \begin{quote}
+          \begin{grayparbox}
         The segmentation determined by the proto-expansion actually segments the literal body (i.e. each segment is in-bounds and the segments are non-overlapping.)
+          \end{grayparbox}
 \end{quote}
@@
       \begin{quote}
-        Each spliced type has a well-formed expansion at the application site.
+        \begin{grayparbox}
+
+        Each spliced type has a well-formed expansion.
+        \end{grayparbox}
 \end{quote}
@@
     \begin{quote}
-      Each type annotation on a reference to a spliced expression has a well-formed expansion at the application site.
+      \begin{grayparbox}
+      Each type annotation on a reference to a spliced expression has a well-formed expansion.
+      \end{grayparbox}
 \end{quote}
@@
     \begin{quote}
-      Each spliced expression has a well-typed expansion consistent with its type annotation.
+      \begin{grayparbox}
+      Each spliced expression has a well-typed expansion consistent with the type annotation in the segmentation.
+      \end{grayparbox}
 \end{quote}
-\item (\textbf{Capture Avoidance}) $e = [\sseq{\tau'_i/t_i}{\nty}, \sseq{e_i/x_i}{\nexp}]e'$ for some $\sseq{t_i}{\nty}$ and $\sseq{x_i}{\nexp}$ and $e'$
+\item (\textbf{Capture Avoidance}) $e = [\sseq{\tau'_i/t_i}{\nty}, \sseq{e_i/x_i}{\nexp}]e'$ for some variables $\sseq{t_i}{\nty}$ and $\sseq{x_i}{\nexp}$, and $e'$
     \begin{quote}
+    \begin{grayparbox}
       The final expansion can be decomposed into a  term with variables in place of each spliced type or expression.
 The expansions of these spliced types and expressions can be substituted into this term in the standard capture avoiding manner.
+    \end{grayparbox}
     \end{quote}
@@
     \begin{quote}
-      The aforementioned decomposed term makes no mention of bindings in the application site context, i.e. the only free variables are those corresponding to the spliced terms.
+      \begin{grayparbox}
+
+      The decomposed term makes no mention of bindings in the application site context, i.e. the only free variables are those standing for spliced terms.
+      \end{grayparbox}
 \end{quote}
@@
 \end{theorem}
+\vspace{-4px}
+
 The proof, which involves auxiliary lemmas about the decomposition of proto-types and proto-expressions, is given in the supplement.
 
+Notice that we were able to state the hygiene properties (\textbf{Capture Avoidance} and \textbf{Context Independence}) without needing a notion of alpha-equivalence of source terms, as is the case in the typical formal accounts of hygiene \cite{Kohlbecker86a,DBLP:conf/popl/Adams15,DBLP:conf/popl/ClingerR91,DBLP:journals/lisp/DybvigHB92,DBLP:conf/esop/HermanW08,Herman10:Theory}.
+Instead, we used standard notions of capture avoiding substitution and free variables combined with the context disjointness conditions in the rules above.
+This is possible only because we keep track of spliced terms explicitly in the proto-expansion, rather than going straight to the final expansion.
+In fact, doing so is critical for TLMs -- there is no notion of alpha-conversion for partially parsed terms, so any notion of hygiene that relies on this notion would be inapplicable.
+
+\vspace{-5px}
 \subsubsection{spTLM Reasoning Principles} Finally, the following theorem summarizes the abstract reasoning principles available to programmers when applying an spTLM.
@@
 % \def\thetheorem{\ref{thm:spTLM-Typing-Segmentation}}
-\begin{theorem}[spTLM Abstract Reasoning Principles]
+\begin{theorem}[spTLM Reasoning Principles]
 % \label{thm:spTLM-Typing-Segmentation}
 If $\patExpands{\upctx}{\uPhi}{\utsmap{\tsmv}{b}}{p}{\tau}$ where $\uDelta=\uDD{\uD}{\Delta}$ and $\uGamma=\uGG{\uG}{\Gamma}$ then all of the following hold:
-\begin{enumerate}
+\begin{enumerate}[noitemsep,nolistsep]
         \item (\textbf{Typing 1}) $\uPhi=\uPhi', \uPhyp{\tsmv}{a}{\tau}{\eparse}$ and $\patType{\pctx}{p}{\tau}$
@@
         }{\npat}$  and $\sseq{\patType{\pctx_i}{p_i}{\tau_i}}{\npat}$
-      \item (\textbf{No Hidden Bindings}) $\uG = \biguplus_{0 \leq i < \npat} \uG_i$ and $\Gamma = \bigcup_{0 \leq i < \npat} \pctx_i$
+      \item (\textbf{Visibility}) $\uG = \biguplus_{0 \leq i < \npat} \uG_i$ and $\Gamma = \bigcup_{0 \leq i < \npat} \pctx_i$
         \begin{quote}
+          \begin{grayparbox}
           The hypotheses generated by the TLM application are exactly those generated by the spliced patterns.
+          \end{grayparbox}
         \end{quote}
@@
 \end{theorem}
+\vspace{-3px}
 % \begin{proof} 
-The proof, in the supplement, relies on an auxiliary lemma about decomposing proto-patterns.
-
+% The proof, in the supplement, relies on an auxiliary lemma about decomposing proto-patterns.
+
+\subsection{Retrospective} Let us step back and briefly review what we have accomplished so far.
+First, we defined an entirely standard expanded language (Sec.
+\ref{sec:s-XL}).
+Then we mirrored the forms in this language to form the common forms of the unexpanded language (Sec.
+\ref{sec:s-UL}), taking care to handle identifiers carefully (Sec.
+\ref{sec:s-TE}).
+We also added forms for defining TLMs (Sec.
+\ref{sec:s-TLM-def}).
+We then considered TLM application, which invokes a parse function to programmatically produce an encoding of a proto-expansion (Sec.
+\ref{sec:s-TLM-ap}).
+Each proto-expansion is then decoded, validated and inductively expanded  (Sec.
+\ref{sec:s-PEV}) to establish the abstract reasoning principles just described (Sec.
+\ref{sec:s-metatheory}).
 % We write $\tsfrom{\escenev}$ for the type splicing scene constructed by dropping unnecessary contexts from $\escenev$:
@@
 
+\vspace{-4px}
 \newcommand{\pTLMsSec}{Parametric TLMs (pTLMs)}
@@
 \label{sec:ptsms}
-Simple TLMs operate at a single specified type and the proto-expansions they generate must be closed.
-This simplifies matters formally, but it is not convenient in practice.
+\vspace{-2px}
+Simple TLMs operate at a single specified type and the proto-expansions they generate must be closed (cf.
+the final premise of Rule \textsc{ee-ap-setsm} in the Sec.
+\ref{sec:s-TLM-ap}.) While this suited our purposes in the previous section because we intend $\miniVersePat$ to communicate the essential concepts, it is not realistic for an ML-like language to impose these limitations.
 This section introduces \emph{parametric TLMs}.
@@
 
+\vspace{-4px}
 \subsection{Parametric TLMs By Example}\label{sec:ptsms-by-example}
-
-Consider the following ML signature specifying an abstract data type \cite{liskov1974programming,harper1997programming} of string-keyed polymorphic dictionaries:
+\vspace{-2px}
+
+Consider the following Reason module type (a.k.a.
+signature), which specifies an abstract data type \cite{liskov1974programming,harper1997programming} of string-keyed polymorphic immutable dictionaries:
 \begin{lstlisting}[numbers=none]
-  signature DICT = sig
-    type t('a)
-    val empty : t('a)
-    val extend : t('a) -> string * 'a -> t('a)
-    (* ... *)
-  end
+  module type DICT = {
+    type t('a); 
+    let empty : t('a); 
+    let extend : t('a) -> (string, 'a) -> t('a); 
+    /* ... */ 
+  };
 \end{lstlisting}
-Let us now define a TLM that is parametric over all implementations of this signature, \li{D : DICT}, and also over all choices of type \li{'a}:
+Let us now define a TLM that is parametric over all implementations of this signature, \li{D : DICT}, and also over all choices of the range type, \li{'a}:
 \begin{lstlisting}[numbers=none]
-  syntax $dict (D : DICT) (type 'a) at D.t('a) by static
-    fn(b : body) -> parse_result(proto_expr) => (* ... *)
-  end
+  syntax $dict (D : DICT) (type 'a) at D.t('a) by static {
+    fun(b : body) -> parse_result(proto_expr) => /* ... */
+  };
 \end{lstlisting}
-For example, assuming some implementation \li{Dict : DICT} and some values \li{v1 : int} and \li{v2 : int} (and assuming that VerseML offers a generalized literal form delimited by \li{\{|} and \li{|\}}), we can apply \li{$dict} as follows:
+Assuming some module that implements this signature, \li{HashDict : DICT}, and some values \li{v1 : int} and \li{v2 : int}, we can apply \li{$dict} as follows:
 \begin{lstlisting}[numbers=none]
-  $dict Dict int {| "key1"SURL => EURLv1SURL; EURL"key2"SURL => EURLv2 |}
+  $dict HashDict int [| "key1"SURL => EURLv1SURL; EURL"key2"SURL => EURLv2 |]
 \end{lstlisting}
 Notice that the segmentation immediately reveals which punctuation is particular to this TLM and where the spliced key and value expressions appear.
-Because the context-free syntax of unexpanded terms is never modified, it is possible to reason modularly about syntactic determinism (i.e.
-we can reason above that \li{=>} does not appear in the follow set of unexpanded terms, so there can never be an ambiguity about where a key expression ends.)
-
-If we will use the \li{Dict} implementation ubiquitously, we can abbreviate the partial application of \li{$dict} to \li{Dict}, resulting in a TLM that is parametric over only the type \li{'a}, as follows:
+Because the context-free syntax of unexpanded terms is never modified, it is possible to reason modularly about syntactic determinism, i.e.
+we can reason above that \li{=>} does not appear in the follow set of unexpanded terms \cite{conf/pldi/SchwerdfegerW09}, so there can never be an ambiguity about where a key expression ends.
+
+If we will use the \li{HashDict} implementation ubiquitously, we can abbreviate the partial application of \li{$dict} to \li{HashDict}, resulting in a TLM that is parametric over only the type \li{'a}, as follows:
 \begin{lstlisting}[numbers=none]
-  let syntax $Dict = $dict Dict in 
-  $Dict int {| "key1"SURL => EURLv1SURL; EURL"key2"SURL => EURLv2 |}
+  syntax $dict' = $dict HashDict; 
+  $dict' int [| "key1"SURL => EURLv1SURL; EURL"key2"SURL => EURLv2 |];
 \end{lstlisting}
@@
 it must be valid for all modules \li{D : DICT} and types \li{'a}.
-It is only after proto-expansion validation that we substitute the actual parameters, here \li{Dict} for \li{D} and \li{int} for \li{'a}, into the final expansion.
-ADditionally, substitution occurs on the type annotations on references to spliced terms before recursively expanding those terms (otherwise, the expressions \li{v1} and \li{v2} above would need to be well-typed for all types, \li{'a}.)
+It is only after proto-expansion validation that we substitute the actual parameters, here \li{HashDict} for \li{D} and \li{int} for \li{'a}, into the final expansion.
+Additionally, substitution occurs on the type annotations on references to spliced terms before recursively expanding those terms (otherwise, the expressions \li{v1} and \li{v2} above would need to be well-typed for all types, \li{'a}.)
 
@@
 With module parameters, it is possible to sidestep this problem by passing in a module containing the datatype constructors.
-In practice, the language might implicitly construct and open such a module within the expansion when the type annotation on the TLM is such a datatype.
-
-It is important to note that module parameters are not accessible by parse functions directly, because the applied module parameters will not have been dynamically instantiated when typed expansion occurs.
-Parameters are accessible only from within the expansion that the parse function generates.
+Because this will be common in practice, we provide the shorthand \li{syntax $a at T using t1, (*...*), tn, X1, (*...*), Xm by static \{ ...
+\}} for the parameterization of \li{$a} by $n$ type variable parameters and $m$ module variable parameters, followed by the partial application of the given parameters.
+Additionally, for the type variable parameters that refer to datatypes, the constructors are bound automatically in the expansion.
+This explicit form of capture is reminiscent of work on explicit capture for distributed functions by \citet{DBLP:conf/ecoop/MillerHO14}.
+Although a bit cumbersome, asking TLM providers to make their intent explicit allows them to avoid having to reason about the computations that the parse function performs on syntax tree representations when refactoring code.
+Otherwise, proto-expansion representations would need to be built primitively into the binding structure of the language.
+We want to leave the semantics of OCaml alone when incorporating TLMs into Reason, so this would not be feasible.
+
+It is important to note that module parameters are accessible by expansions, but not by parse functions directly, because the applied module parameters will not have been dynamically instantiated when typed expansion occurs.
 We will discuss a distinct mechanism for providing helper functions to parse functions in Sec.
@@
 
+\vspace{-4px}
 \subsection{Parametric TLMs, Formally}
+\vspace{-2px}
 % \begin{figure}[p]
@@
 
-We will now outline $\miniVerseParam$, a calculus that extends $\miniVersePat$ with support for parametric TLMs.
+We will now outline $\miniVerseParam$, a calculus that extends $\miniVersePat$ with parametric TLMs.
 This calculus is organized, like $\miniVersePat$, as an unexpanded language (UL) defined by typed expansion to an expanded language (XL).
@@
 \ref{sec:setsms-formally}, so we highlight only the most important concepts below.
-
-The XL consists of 1) module expressions, $M$, classified by signatures, $\sigma$; 2) constructions, $c$, classified by kinds, $\kappa$; and 3) expressions classified by types, which are constructions of kind $\akty$ (we use metavariables $\tau$ instead of $c$ for types.) Metavariables $X$ ranges over module variables and $u$ or $t$ over construction variables.
+The details are in the supplement.
+
+The XL consists of 1) module expressions, $M$, classified by signatures, $\sigma$; 2) constructions, $c$, classified by kinds, $\kappa$; and 3) expressions classified by types, which are constructions of kind $\akty$ (we use metavariables $\tau$ instead of $c$ for types by convention.) Metavariables $X$ ranges over module variables and $u$ or $t$ over construction variables.
 The module and construction languages are based closely on those defined by \citet{pfple1}, which in turn are based on early work by \citet{MacQueen:1984:MSM:800055.802036,DBLP:conf/popl/MacQueen86}, subsequent work on the phase splitting interpretation of modules \cite{harper1989higher} and on using dependent singleton kinds to track type identity \cite{stone2006extensional,DBLP:conf/lfmtp/Crary09}, and finally on formal developments by \citet{dreyer2005understanding} and \citet{conf/popl/LeeCH07}.
@@
 
-The main conceptual difference between $\miniVersePat$ and $\miniVerseParam$ is that $\miniVerseParam$ introduces the notion of unexpanded and expanded TLM expressions and types, as shown in Figure \ref{fig:P-macro-expressions-types}.
+The main conceptual difference between $\miniVersePat$ and $\miniVerseParam$ is that $\miniVerseParam$ introduces the notion of unexpanded and expanded TLM expressions and types, as shown in Fig.
+\ref{fig:P-macro-expressions-types}.
 
 \begin{figure}[h]
-\begin{minipage}{0.36\textwidth}
-$\begin{array}{llcl}
+\vspace{-6px}
+\small
+\begin{minipage}{0.38\textwidth}
+$\arraycolsep=3pt\begin{array}{llcl}
 \mathsf{UMType} & \urho & ::= & \utau ~\vert~ \allmods{\uX}{\usigma}{\urho}\\
@@
 \begin{minipage}{0.6\textwidth}
-$\begin{array}{llcl}
+$\arraycolsep=3pt\begin{array}{llcl}
 \mathsf{MType} & \rho & ::= & \aetype{\tau} ~\vert~ \aeallmods{\sigma}{X}{\rho} \\
@@
 \end{minipage}
-\vspace{-6px}
+\vspace{-10px}
 \caption{Syntax of unexpanded and expanded TLM types and expressions in $\miniVerseParam$}
 \label{fig:P-macro-expressions-types}
-\vspace{-6px}
+\vspace{-10px}
 \end{figure}
-This is necessary to express abstraction over parameters and application of parameters.
+The TLM type $\aeallmods{\sigma}{X}{\rho}$ classifies TLM expressions that have one module parameter matching $\sigma$.
 For simplicity, we formalize only module parameters.
-Type parameters can be realized as module parameters having exactly one abstract type member.
-
-The rule governing expression TLM application, reproduced below, touches all of the main ideas in $\miniVerseParam$. %It might be useful to compare this rule to Rule \textsc{ee-ap-sptsm}.
-\begin{mathpar}
+Type parameters can be expressed as module parameters having exactly one abstract type member.
+
+The rule governing expression TLM application, reproduced below, touches all of the main ideas in $\miniVerseParam$, so we will refer to it throughout the remainder of this section.%It might be useful to compare this rule to Rule \textsc{ee-ap-sptsm}.
+{\small\begin{mathpar}
 \inferrule[ee-ap-petsm]{
@@
 }
-\end{mathpar}
+\end{mathpar}}
+
 The first two premises simply deconstruct the (unified) unexpanded context $\uOmega$ (which tracks the expansion of expression, constructor and module identifiers, as $\uDelta$ and $\uGamma$ did in $\miniVersePat$) and peTLM context, $\uPsi$.
@@
 The resulting TLM expression, $\epsilon$, must be defined at a type (i.e.
-no quantification must remain.)
-
-The fourth premise performs \emph{peTLM expression normalization}.
-Normalization, $\tsmexpEvalsExp{\Omega_\text{app}}{\Psi}{\epsilon}{\epsilon_\text{normal}}$, is defined in terms of a simple structural dynamics with two stepping rules:
+no quantification over modules must remain once the literal body is encountered.)
+
+The fourth premise performs \emph{peTLM expression normalization}, $\tsmexpEvalsExp{\Omega_\text{app}}{\Psi}{\epsilon}{\epsilon_\text{normal}}$.
+This is defined in terms of a structural operational semantics \cite{DBLP:journals/jlp/Plotkin04a} with two stepping rules:
 % \begin{equation*}\tag{\ref{rule:tsmexpEvalsExp}}
@@
 % \end{equation*}
-\begin{mathpar}
+{\small\begin{mathpar}
 \inferrule[eps-dyn-apmod-subst-e]{ }{
@@
 }
-\end{mathpar}
+\end{mathpar}}
 % The peTLM expression normal forms are defined as follows:
@@
 % \end{equation*}
+\vspace{-5px}
+
 Normalization eliminates parameters introduced in higher-order abbreviations, leaving only those parameter applications specified by the original TLM definition.
@@
 
-The third row of premises looks up the applied TLM's definition by first invoking a simple metafunction to extract the name, then looking up this name within the peTLM definition context, $\Psi$.
+The third row of premises looks up the applied TLM's definition by invoking a simple metafunction to extract its name, $a$, then looking up $a$ within the peTLM definition context, $\Psi$.
 % \begin{align}
@@
 The syntax of parameterized proto-expressions is given below.
-\[\begin{array}{llcl}
+
+\vspace{-4px}{\small\[\begin{array}{llcl}
 % \textbf{Sort} & & & \textbf{Operational Form} & \textbf{Stylized Form} & \textbf{Description}\\
@@
 \mathsf{PPrExp} & \pce & ::= & \apceexp{\ce} ~\vert~ \apcebindmod{X}{\pce}
-\end{array}\]
-There must be one binder in $\pce$ for each TLM parameter specified by $\tsmdefof{\epsilon_\text{normal}}$.
-(VerseML can insert these binders automatically as a convenience, but we consider only the underlying mechanism in this core calculus.)
+\end{array}\]}%
+\vspace{-6px}%
+
+There must be one binder in $\pce$ for each TLM parameter specified by $a$. (In Reason, we can insert these binders automatically as a convenience.)
 
@@
 The two rules governing the proto-expression deparameterization judgement are below:
-\begin{mathpar}
+{\small\begin{mathpar}
 \inferrule{ }{
@@
 
-\inferrule{
+\vspace{-4px}\inferrule{
   \prepce{\Omega_\text{app}}{\Psi}{\pce}{\ce}{\epsilon}{\aeallmods{\sigma}{X}{\rho}}{\omega}{\Omega}\\
@@
 }
-\end{mathpar}
+\end{mathpar}}%
 This judgement can be pronounced ``when applying peTLM $\epsilon$, $\pce$ has deparameterization $\ce$ leaving $\rho$ with parameter substitution $\omega$''. 
@@
 \ref{sec:ptsms-by-example}, these annotations might refer to the parameters, so the parameter substitution, $\omega$, which is tracked by the splicing scene, must be applied to the type annotation before proceeding recursively to expand the referenced unexpanded term.
-However, the spliced term itself must treat parameters parametrically, so the substitution is not applied in the conclusion of the rule:
+However, the spliced term itself must treat parameters parametrically, so the substitution is not applied in the conclusion of the following rule:
 \begin{mathpar}\label{rule:cvalidE-P-splicede}
-\inferrule[p-pev-spliced]{
+\inferrule{
   \parseUExp{\bsubseq{b}{m}{n}}{\ue}\\
@@
   \cvalidEP{\Omega}{\esceneP{\omega : \OParams}{\uOmega}{\uPsi}{\uPhi}{b}}{\acesplicede{m}{n}{\ctau}}{e}{\tau}
-}\vspace{-5px}
+}
 \end{mathpar}
@@
 The following theorem establishes the abstract reasoning principles available when applying a parametric expression TLM.
+The clauses are directly analagous to those of Theorem \ref{thm:setlm-reasoning}, so for reasons of space we do not repeat the inline descriptions.
+The \textbf{Kinding} clauses can be understood by analogy to the \textbf{Typing} clauses.
+The details of parametric pattern TLMs (ppTLMs) are analagous (see supplement.)
 \vspace{-3px}
@@
 If $\expandsP{\uOmega}{\uPsi}{\uPhi}{\utsmap{\uepsilon}{b}}{e}{\tau}$ then:
-\begin{enumerate}[nolistsep]
+\begin{enumerate}[nolistsep,noitemsep]
   \item $\uOmega=\uOmegaEx{\uD}{\uG}{\uMctx}{\Omega_\text{app}}$
@@
   \item $
-    \segof{\ce} = \sseq{\acesplicedk{m_i}{n_i}}{\nkind} \cup \sseq{\acesplicedc{m'_i}{n'_i}{\cekappa'_i}}{\ncon} \cup 
-               \sseq{\acesplicede{m''_i}{n''_i}{\ctau_i}}{\nexp}
+    \segof{\ce} = \sseq{\acesplicedk{m_i}{n_i}}{\nkind} \cup \sseq{\acesplicedc{m'_i}{n'_i}{\cekappa'_i}}{\ncon} \cup $ \\
+     ~~~~$          \sseq{\acesplicede{m''_i}{n''_i}{\ctau_i}}{\nexp}
     $
-  \item (\textbf{Kinding 1}) $\sseq{\kExpands{\uOmega}{\parseUKindF{\bsubseq{b}{m_i}{n_i}}}{\kappa_i}}{\nkind}$ and $\sseq{\iskind{\Omega_\text{app}}{\kappa_i}}{\nkind}$
+  \item (\textbf{Kinding 1}) $\sseq{\kExpands{\uOmega}{\parseUKindF{\bsubseq{b}{m_i}{n_i}}}{\kappa_i}}{\nkind}$ and \\ ~~~~$\sseq{\iskind{\Omega_\text{app}}{\kappa_i}}{\nkind}$
   \item (\textbf{Kinding 2}) $\sseq{\cvalidK{\OParams}{\csceneP{\omega : \OParams}{\uOmega}{b}}{\cekappa'_i}{\kappa'_i}}{\ncon}$ and $\sseq{\iskind{\Omega_\text{app}}{[\omega]\kappa'_i}}{\ncon}$
@@
 \end{theorem}
-We leave out the details of parametric pattern TLMs (ppTLMs), but they should again follow analagously.
-
+
+\vspace{-6px}
 \newcommand{\staticEvalSec}{Static Evaluation}
@@
 \label{sec:static-eval}
-
-Thus far, we have assumed that the parse functions in TLM definitions are closed expanded expressions.
+\vspace{-2px}
+
+There is one more major impracticality that we can now address.
+So far, we have assumed that parse functions are closed expanded expressions.
 This assumption simplifies matters formally, but in practice it is quite difficult, because this means that TLM definitions cannot themselves access any libraries nor use TLMs internally.
-To address this problem, we need to introduce the concept of a \emph{static environment}.
-
+To address this problem, we introduce a \emph{static environment}.
+
+\vspace{-4px}
 \subsection{The Static Environment}
-Figure \ref{fig:static-module-example} shows an example of a module, \li{ParserCombos}, that defines a number of parser combinators following \citet{Hutton1992d}.
-The \li{static} qualifier indicates that this module is bound for use only within similarly qualified values, including in particular the parse functions of the subsequent TLM definitions.
+\vspace{-2px}
+Fig.
+\ref{fig:static-module-example} shows an example of a module, \li{ParserCombos}, that defines a number of parser combinators following \citet{Hutton1992d}.
+The \li{static} qualifier indicates that this module is bound for use only within similarly qualified values, including in particular parse functions of subsequent TLM definitions.
 \begin{figure}[h]
@@
 \begin{lstlisting}
-  static module ParserCombos = struct 
-    type parser('c, 't) = list('c) -> list('t * list('c))
-    val alt : parser('c, 't) -> parser('c, 't) -> parser('c, 't)
-    (* ... *)
-  end
-  syntax $a at t by static fn(b) => (* OK *) ParserCombos.alt (* ... *) end 
-  static val y = (* OK *) ParserCombos.alt (* ... *)
-  val z = (* neither ParserCombos nor y are available here *)
+  static module ParserCombos = { 
+    type parser('c, 't) = list('c) -> list('t, list('c));
+    let alt : parser('c, 't) -> parser('c, 't) -> parser('c, 't);
+    /* ... */
+  };
+  syntax $a at t by static { fun(b) => ParserCombos.alt /* OK ... */ };
+  static val y = ParserCombos.alt /* OK ... */;
+  val z = /* ... neither ParserCombos nor y are available here */;
 \end{lstlisting}
-\vspace{-5px}
-\caption{Binding a static module for use within parse functions.}
+\vspace{-10px}
+\caption{Binding static modules and values for use within parse functions.}
 \label{fig:static-module-example}
@@
 
-The values that arise during the static evaluation (in particular, of parse functions) do not themselves persist from ``compile-time'' to ``run-time'', so we do not need a full staged computation system, e.g.
+The values that arise when the static phase runs do not persist from ``compile-time'' to ``run-time'', so we do not need a full staged computation system, e.g.
 as described by \citet{Taha99multi-stageprogramming:}.
 Instead, a sequence of static bindings operates like a read-evaluate-print loop (REPL) scoped according to the program structure, in that each static expression is evaluated immediately and the evaluated values are tracked by a \emph{static environment}.
-The static environment is discarded at the end of typed expansion.
+The static environment is discarded after typed expansion.
 
@@
 It might also be helpful to restrict mutable state shared between TLMs to prevent undesirable TLM application order dependencies.
-On the other hand, these features, if used for the purposes of caching in a disciplined manner, might speed up typed expansion.
+On the other hand, these features, if used for the purposes of caching, might speed up typed expansion.
 These are orthogonal design decisions.
 
+\vspace{-6px}
 \subsection{Applying TLMs Within TLM Definitions}\label{sec:tsms-for-tsms}
@@
 
+\vspace{-6px}
 \subsubsection{Quasiquotation}
@@
 \begin{lstlisting}[numbers=none]
-  static syntax $proto_expr at proto_expr by static fn(b) => (* ... *) end
+  static syntax $proto_expr at proto_expr by static { /* ... */ };
 \end{lstlisting}
@@
 
-% A similar approach can be taken for working with encodings of terms of other languages (e.g. when writing an interpretter or compiler in VerseML.)
-
+% A similar approach can be taken for working with encodings of terms of other languages (e.g. when writing an interpretter or compiler in Reason.)
+
+\vspace{-6px}
 \subsubsection{Parser Generators}
@@
 \begin{lstlisting}
-  signature PARSEGEN = sig 
-    type grammar('a)
-    (* ... operations on grammars ... *)
-    val generate : grammar('a) -> (body -> parse_result('a))
-  end
+  module type PARSEGEN = { 
+    type grammar('a);
+    /* ... operations on grammars ... */
+    val generate : grammar('a) -> (body -> parse_result('a));
+  };
 \end{lstlisting}
 
-Rather than constructing a grammar (equipped with semantic actions) using the associated operations (whose specifications are elided in \li{PARSEGEN}), we wish to use a syntax for context-free grammars that follows standard conventions.
-We can do so by defining a static parametric TLM \li{$grammar}:
+Rather than constructing a grammar equipped with semantic actions using the associated operations (whose specifications are elided in \li{PARSEGEN}), we wish to use a syntax for context-free grammars that follows standard conventions.
+We can do so by defining a static parametric TLM:
 \begin{lstlisting}[numbers=none]
-  static syntax $grammar (P : PARSEGEN) 'a at P.grammar('a) by (* ... *) end
+  static syntax $grammar(P : PARSEGEN)(type 'a) at P.grammar('a) /*...*/
 \end{lstlisting}
-To support splicing, we need non-terminals that recognize unexpanded terms and produce the corresponding splice reference, rather than the AST itself.
-This would require that the parser generator keep track of location information (as most production-grade parser generators already do for error reporting.) For spliced expressions, this non-terminal would need to be a family of non-terminals indexed by a value of type \li{proto_typ}.
-
-A grammar containing such non-terminals can serve as a \emph{summary specification} -- a human can simply be take this grammar as a specification of what the splice summary will be for every recognized string, rather than relying on an editor to communicate this information.
-The associated semantic actions can be held abstract as long as the system performs a simple check to ensure that the generated proto-expansion does in fact mention each spliced expression from the corresponding production.
+To support splicing, we need non-terminals that recognize unexpanded terms and produce the corresponding splice references, rather than the AST itself.
+This requires that the parser generator keep track of location information (as most production-grade parser generators already do for error reporting.) A more detailed example of this mechanism being used to define a TLM for a modular encoding of regular expressions is given in the supplement.% For spliced expressions, this non-terminal would need to be a family of non-terminals indexed by a value of type \li{proto_typ}.
+
+A grammar containing such non-terminals can serve as a \emph{summary specification} -- a human can simply be take this grammar as a specification of what the segmentation will be for every recognized string, rather than relying on an editor to communicate this information.
+The associated semantic actions can be held abstract as long as the system performs a simple check to ensure that the proto-expansion does mention each spliced expression from the corresponding production.
 
@@
 %     token str_tok #\label{line:str_tok_start}#
-%       EHTMLRU.parse "SSTR[^(@$]+ESTR" (* cannot use $rx within its own def *)SHTML #\label{line:str_tok_end}#
+%       EHTMLRU.parse "SSTR[^(@$]+ESTR" /* cannot use $rx within its own def */SHTML #\label{line:str_tok_end}#
 %     start <- str_tok
@@
 
+\vspace{-6px}
 \subsection{Static Evaluation, Formally}
@@
 The substitution $\omega$ is applied to the parse function after it is expanded.
-The full details are defined as a minor patch of $\miniVerseParam$ called $\miniVersePH$ in the supplement.
-
+The full details are defined as a small patch of $\miniVerseParam$ called $\miniVersePH$ in the supplement.
+
+\vspace{-6px}
 \subsection{Library Management}
@@
 This can be accomplished either by the package manager (e.g.
-SML/NJ's CM \cite{blume:smlnj-cm}, extended with phase annotations) or by allowing one to explicitly lower a new instance of a module imported into the standard phase for use also in  the static phase, as in a recent proposal for modular staging macros in OCaml \cite{Ocaml/macros}.
-We will leave these details as design decisions for the reader to consider.
-
-TLMs definitions can be exported by packages, but they cannot be exported from within modules because that would require that they also appear in signatures, and that, in turn, would complicate reasoning about signature equivalence, since TLM definitions contain arbitrary parse functions.
-It would also bring in confusion about whether the generated expansions can use knowledge about type identity available  within the module enclosing the TLM definition.
+SML/NJ's CM \cite{blume:smlnj-cm}, extended with phase annotations) or by allowing one to explicitly lower an instance of a module define in the standard phase for use also in  the static phase, as in a recent proposal for modular staging macros in OCaml \cite{Ocaml/macros}.
+
+TLMs definitions can be exported from the top level of packages, but they cannot be exported from within ML-style modules because that would require that they also appear in signatures, and that, in turn, would complicate reasoning about signature equivalence, since TLM definitions contain arbitrary parse functions.
+%It would also bring in confusion about whether the generated expansions can use private knowledge about type identity. 
 That said, it should be possible to export TLM \emph{abbreviations} from modules, since they refer to TLM definitions only through symbolic names.
-We have not yet formalized this intuition.
-
-% For example, a language-external library manager for VerseML similar to SML/NJ's CM \cite{blume:smlnj-cm} could support a \li{static} qualifier on imported libraries, which would place the definitions exported by the imported library into the static phase of the library being defined.
+We have not yet formalized this intuition, but the work of \citet{culpepper2005syntactic,culpepper2007advanced} considered a closely related question: how should Typed Scheme's macros interact with its unit (i.e.
+package) system.
+
+% For example, a language-external library manager for Reason similar to SML/NJ's CM \cite{blume:smlnj-cm} could support a \li{static} qualifier on imported libraries, which would place the definitions exported by the imported library into the static phase of the library being defined.
 In particular, a library definition in such a compilation manager might look like this:
@@
 % Library 
-%   (* ... exported module, signature and TLM names ... *)
+%   /* ... exported module, signature and TLM names ... */
 % is 
-%   (* ... files defining those exports ... *)
-
-%   (* imports: *)
+%   /* ... files defining those exports ... */
+
+%   /* imports: */
 %   static parsegen.cm 
@@
 
+\vspace{-6px}
 \newcommand{\rwSec}{Related Work}
@@
 There are hundreds of syntax definition systems of various design.
-Notable examples include grammar-oriented systems like Camlp4 \cite{ocaml-manual}\todo{better citation?}, Copper \cite{conf/gpce/WykS07,conf/pldi/SchwerdfegerW09}, SugarJ/Sugar*/SoundExt \cite{erdweg2011sugarj,erdweg2013framework,conf/icfp/LorenzenE13,conf/popl/LorenzenE16}, as well as various parser combinator systems \cite{Hutton1992d}.
+Notable examples include grammar-oriented systems like Camlp4 \cite{ocaml-manual}, Copper \cite{conf/gpce/WykS07,conf/pldi/SchwerdfegerW09}, SugarJ/Sugar*/SoundExt \cite{erdweg2011sugarj,erdweg2013framework,conf/icfp/LorenzenE13,conf/popl/LorenzenE16}, as well as various parser combinator systems \cite{Hutton1992d}.
 The parsers generated by these systems can be invoked to preprocess program text in various ways, e.g.
@@
 \item \textbf{Segmentation}: Clients of a syntax dialect cannot accurately determine which segments of the program text appear directly in the desugaring.
-In contrast, TLM clients can inspect the inferred segmentation (communicate via, e.g., secondary notation.)
+In contrast, TLM clients can inspect the inferred segmentation (communicated via secondary notation.)
 \item \textbf{Binding}: Clients of a context-free syntax dialect cannot be sure that the desugaring is context-independent and that spliced terms are capture avoiding, as TLM clients can.
 \item \textbf{Typing}: Clients of a syntax dialect cannot reason abstractly about types.
-In contrast, TLM clients clients can determine the type of any expansion by referring to the parameter and type declarations on the TLM definition, and nothing else.
+In contrast, TLM clients can determine the type of any expansion by referring to the parameter and type declarations on the TLM definition, and nothing else.
 The inferred segmentation also gives types for each spliced expression or pattern.
-This information can be communicated upon request by the type inspection service of a program editor.
+%This information can be communicated upon request by the type inspection service of a program editor.
 \end{enumerate}
@@
 The system then attempts to automatically verify that the expansion logic (expressed using a rewrite system, rather than an arbitrary function) is sound relative to these derived rules.
-
 TLMs differ in several ways.
@@
 Another distinction is that TLMs do not support type-dependent expansions.
-The trade-off is that this allows expansions, and therefore segmentations, to be generated even when the program is ill-typed, as long as the static values are well-typed\todo{elaborate}.
-Another important distinction is that TLMs rely on proto-expansion validation, rather than verification as in SoundExt.
+The trade-off is that this allows expansions, and therefore segmentations, to be generated even when the program is ill-typed, as long as the static values are well-typed (this notion of \emph{untyped expansion} was not considered formally above, but it follows straightforwardly from the fact that all of the premises to the TLM application rules except the final proto-expansion validation premises are independent of the typing context.) Another important distinction is that TLMs rely on proto-expansion validation, rather than verification as in SoundExt.
 The trade-off is that TLMs do not require that the expansion logic be written using a restricted rewriting system, nor does the system require a fully mechanized language definition and heavyweight theorem proving machinery.
@@
 
+\vspace{-6px}
 \subsection{Term Rewriting Systems}
+\vspace{-2px}
 Another approach -- and the approach that TLMs are rooted in -- is to leave the context-free syntax of the language fixed, and instead contextually rewrite existing literal forms.
@@
 \end{lstlisting}
-The expansion of this macro will either fail the check for context independence because \li{first_name} will appear as a free variable, no different from any other, or else it will have incorrectly captured an expansion-local binding of the same identifier.
+The expansion of this macro will fail the check for context independence because \li{first_name} will appear as a free variable, no different from any other.
 The hygiene mechanism for TLMs addresses this problem by explicitly distinguishing spliced segments of the literal body in the proto-expansion.
@@
 Much of the research on macro systems has been for languages in the LISP tradition \cite{mccarthy1978history} that do not have rich static type structure.
-The formal macro calculus studied by \citet{DBLP:conf/esop/HermanW08} uses types to encode the binding structure of the generated expansion (in particular, to indicate which identifiers passed as arguments can be captured by which other arguments), but the expanded language itself does not have rich type structure (nor is this calculus is not capable of expressing new literal forms, for the reasons already discussed.) Research on typed \emph{staging macro systems} like MetaML \cite{Sheard:1999:UMS} and MacroML \cite{ganz2001macros} is also not directly applicable to the problem of defining new literal forms -- the syntax tree of the arguments cannot be inspected at all (staging macros are useful mainly for performance reasons.) The Scala macro system is a hygienic macro system which supports ``black-box'' term rewriting macros.
+The formal macro calculus studied by \citet{DBLP:conf/esop/HermanW08} uses types to encode the binding structure of the generated expansion (in particular, to indicate which identifiers passed as arguments can be captured by which other arguments), but the expanded language itself does not have rich type structure (nor is this calculus capable of expressing new literal forms, for the reasons already discussed.) Research on typed \emph{staging macro systems} like MetaML \cite{Sheard:1999:UMS} and MacroML \cite{ganz2001macros} is also not directly applicable to the problem of defining new literal forms -- the syntax tree of the arguments cannot be inspected at all (staging macros are useful mainly for performance reasons.) The Scala macro system is a hygienic macro system which supports ``black-box'' term rewriting macros.
 These support reasoning abstractly about \textbf{Typing} because type annotations constrain the macro arguments and the generated expansions, though the precise reasoning principles available are unclear because the Scala macro system has not been formally specified.
-In any case, black-box macros also cannot be used to repurpose string literal forms for the reasons above.
+In any case, black-box macros also cannot be used to repurpose string literal forms because they are hygienic.
 
@@
 These designate a particular delimiter to escape out into the expression language.
-The problem with using these (dynamically, or in conjunction with term rewriting macros) as vehicles to introduce literal forms at various other types is 1) there is no ``one-size-fits-all'' escape delimiter, and 2) typing is problematic because every escaped term is checked against the same type.
+The problem with using these together with macros as vehicles to introduce literal forms at various other types is 1) there is no ``one-size-fits-all'' escape delimiter, and 2) typing is problematic because every escaped term is checked against the same type.
 In the XHTML example, we have splicing at two different types using two different delimiters.
@@
 In contrast, TLMs are separately defined and explicitly applied.
-Accordingly, different TLMs can therefore be defined at the same type without conflict, and can operate at any type, not just at nominal (i.e.
+Accordingly, different TLMs can be defined at the same type without conflict, and can operate at any type, not just at nominal (i.e.
 abstract) types.
-In a subsequent short paper, \citet{sac15} proposed explicit application of simple expression TLMs in a bidirectional typed setting \cite{Pierce:2000:LTI:345099.345100}, but this proposal did not give any formal details.
-With TLMs, it is not necessary for the language to be bidirectionally typed -- context independence implies that ML-style type inference can be performed using only the segmentation (because the remainder of an expansion cannot mention the very variables whose types are being inferred.) Another critical distinction is that the metatheory presented by \citet{TSLs} establishes only that generated expansions are well-typed.
-It does not establish the remaining abstract reasoning principles that have been a major focus of this paper.
-In particular, there is no formal notion of hygiene (though it is discussed informally), and the mechanism does not guarantee that a valid segmentation will exist, nor associate types with segments.
+In a subsequent short paper, \citet{sac15} proposed explicit application of simple expression TLMs (there called \emph{typed syntax macros}) in a bidirectional typed setting \cite{Pierce:2000:LTI:345099.345100}, but this short proposal did not give any formal details.
+With TLMs, it is not necessary for the language to be bidirectionally typed -- context independence implies that ML-style type inference can be performed using only the segmentation (because the remainder of an expansion cannot mention the very variables whose types are being inferred.) It should be possible to implicitly invoke TLMs based on the expected type in situations where the inference engine had enough information, but we leave this as an interesting direction for future work.
+
+ Another critical distinction is that the metatheory presented by \citet{TSLs} establishes only that generated expansions are of the expected type (i.e.
+a variant of the Typed Expression Expansion theorem from Sec.
+\ref{sec:s-metatheory}.) It does not establish the remaining abstract reasoning principles that have been a major focus of this paper.
+In particular, there is no formal hygiene theorem (though it is discussed informally), and the mechanism does not guarantee that a valid segmentation will exist, nor associate types with segments.
 Finally, this prior work did not consider pattern matching, type functions, ML-style modules, parameters or static evaluation.
-All of these are important for integration into an ML- or Scala-like language.
+All of these are important for integration into an ML- or Scala-like language, and the focus of most of this paper.
 
@@
 
+\vspace{-6px}
 \newcommand{\discussionSec}{Discussion}
 \section{\protect\discussionSec}
-\label{sec:related-work}
 \label{sec:discussion}
 \label{sec:conclusion}
+\vspace{-2px}
 
 The importance of specialized notation as a ``tool for thought'' has long been recognized \cite{DBLP:journals/cacm/Iverson80}.
-According to Whitehead, a good notation ``relieves the brain of [...] unnecessary work'' and ``sets it free to concentrate on more advanced problems'' \cite{cajori1928history}.
-And indeed, advances in mathematics, science and programming have often been accompanied by new notation.
+According to Whitehead, a good notation ``relieves the brain of unnecessary work'' and ``sets it free to concentrate on more advanced problems'' \cite{cajori1928history}, and indeed, advances in mathematics, science and programming have often been accompanied by new notation.
 
@@
 
-Previously, these two relief mechanisms were in tension -- mechanisms  that allowed programmers to express new notation would obscure the type and binding structure of program text, as discussed in Sec.
-\ref{sec:intro} and detailed in Sec.
-\ref{sec:existing-approaches}.
+Previously, these two relief mechanisms were in tension -- mechanisms  that allowed programmers to express new notation would obscure the type and binding structure of the program text.
 TLMs resolve this tension for the broad class of literal forms that generalized literal forms subsume.
@@
 \ref{sec:intro} (up to the choice of outermost delimiter), the additional examples described in Sec.
-\ref{sec:static-eval}, the more detailed example in the supplement\todo{do this}, and the many examples in the prior papers by \citet{TSLs,sac15}.
+\ref{sec:static-eval}, the more detailed example in the supplement, and the many examples in the prior papers by \citet{TSLs,sac15}.
 
@@
 one with a rich static type system, support for pattern matching and an ML-like module system with type functions and abstract type members.
-Our own language integration efforts have been focused on an emerging ML dialect called VerseML, but Scala supports analagous features \cite{conf/oopsla/AminRO14} and would also be a suitable host for the TLM mechanism.
-We also plan to implement TLMs into \li{typy}, a typed functional language embedded into Python as a library \cite{gpce/Omar16}.%We have evaluated this claim by stating the available abstract reasoning principles formally and proving that they hold for the calculi that we have described.
+Our own language integration efforts have been focused on an emerging alternative front-end for OCaml called Reason, but Scala supports analagous features \cite{conf/oopsla/AminRO14} and would also be a suitable host for the TLM mechanism.
+The intended audience for this paper is language designers seeking a reasonable solution to the problem of user-defined literal notation.% We also plan to implement TLMs into \li{typy}, a typed functional language embedded into Python as a library \cite{gpce/Omar16}.%We have evaluated this claim by stating the available abstract reasoning principles formally and proving that they hold for the calculi that we have described.
 
@@
 
-Haskell's \li{do}-notation at types with monadic structure cannot be expressed directly using TLMs because it needs to purposefully induce capture in spliced sub-terms, and TLMs prevent capture.
+% Haskell's \li{do}-notation at types with monadic structure cannot be expressed directly using TLMs because it needs to purposefully induce capture in spliced sub-terms, and TLMs prevent capture.
 This limitation could be resolved by allowing identifiers that appear in literal bodies to be explicitly designated as bound within spliced expressions in the segmentation.
@@
 
-A correct parse function never returns an encoding of a proto-expansion that fails validation. This invariant cannot be enforced by the simple ML-style type system we
-have been working with. Using a proof assistant, it would be possible to verify that
-a parse function generates only encodings of valid proto-expansions\todo{cite something here}. Alternatively, in a
-dependently typed setting, the type of the parse function itself could be enriched so as to
-enforce this invariant intrinsically. We leave the details of this approach as future work.
+There are several interesting avenues for future work beyond those already discussed.
+A correct parse function never returns an encoding of a proto-expansion that fails validation given well-typed splices, but this invariant cannot be enforced by the ML type system.
+% Using a proof assistant, it would be possible to verify that
+% a parse function generates only encodings of valid proto-expansions. 
+Under a
+richer type system, the return type of the parse function itself could be refined so as to
+enforce this invariant intrinsically.
+This problem -- of typed first-class term representations -- has been studied in a variety of settings, e.g.
+in MetaML \cite{Sheard:1999:UMS} and in the modal logic tradition \cite{DBLP:conf/popl/DaviesP96}.
+We leave integration of these approaches into the TLM mechanism as future work.
+Our initial efforts aim to leave the semantics of OCaml unchanged, consistent with the philosophy of the Reason project.
 
 %Another topic of interest has to do with intentional capture. 
-Another important direction has to do with automated refactoring.
-The unexpanded language does not come equipped with standard notions of renaming and substitution.
-Given a splice summary, it should be possible to ``backpatch'' refactorings into literal bodies, but the details are left as future work\todo{cite some papers}.
+Another direction has to do with automated refactoring.
+The unexpanded language does not come with context-free notions of renaming and substitution.
+However, given a segmentation, it should be possible to ``backpatch'' refactorings into literal bodies.
+Recent work by \citet{wand2017inferring} on tracking bindings ``backwards'' from an expansion to the source program is likely relevant.
 
 At several points in the paper, we allude to editor integration.
-However, some important questions having to do with TLM-specific syntax highlighting, incremental parsing and error recovery\todo{citation} remain to be considered.
+However, several important questions having to do with TLM-specific syntax highlighting, incremental parsing and error recovery \cite{graham1979practical} remain to be considered.
 Another interesting direction would be to generalize TLMs to support non-textual notation in the setting of a structure editor.
@@
 
-Other directions for future work are given in the supplement\todo{do this?}.
+Let us conclude by considering the oft-uttered phrase ``syntax doesn't matter'', which is generally taken to mean that syntactic concerns are orthogonal to various more important semantic concerns.
+Our hope is that the reader will leave this paper with an understanding that this is a rather simplistic proclamation.
+Instead, let us proclaim that ``semantics matter for syntax too!'' % We hope that this work will lead to more ``reasonable'' variants of other syntactic conveniences, e.g.
+the ubiquitious ``deriving'' clauses on datatype definitions.
+
+% Other directions for future work are given in the supplement\todo{do this?}.
 
@@
 
+\clearpage
 %% Bibliography

### 2017-10-26 pldi18/syntax-pldi18.tex
ported POPL submission to PLDI format and started doing some edits

--- 2017-07-08 popl18/syntax-popl18.tex
@@
 %% For double-blind review submission
-\documentclass[acmsmall,10pt,review,anonymous]{acmart}\settopmatter{printfolios=true}
+\documentclass[sigplan,review,10pt,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
+% \documentclass[acmsmall,10pt,review,anonymous]{acmart}\settopmatter{printfolios=true}
 %% For single-blind review submission
@@
 % \newtheorem{definition}[theorem]{Definition}
-\newtheorem{assumption}[theorem]{Assumption}
-\newtheorem{condition}[theorem]{Condition}
+% \newtheorem{assumption}[theorem]{Assumption}
+% \newtheorem{condition}[theorem]{Condition}
+
+% \usepackage{titlesec}
+% \titlespacing{\paragraph}{0pt}{10pt}{10pt}
+
+\makeatletter
+\renewcommand{\paragraph}{%
+  \@startsection{paragraph}{4}%
+  {\z@}{0.25ex \@plus 1ex \@minus .2ex}{-1em}%
+  {\normalfont\normalsize\bfseries}%
+}
+\makeatother
 
@@
 mathescape=false,escapechar=~,
-numbers=left, numberstyle=\scriptsize\color{mygray}, language=ML,moredelim=[il][\sffamily]{?},showspaces=false,showstringspaces=false,xleftmargin=0pt, numbersep=3pt, morekeywords=[1]{spliced,tyfam,opfam,let,fn,val,def,casetype,objtype,metadata,of,*,string,datatype,new,toast,syntax,module,switch,where,using,import,for,ana,syn,opcon,tycon,metasignature,metamodule,metasig,metamod,static,at,by,tycase,mod,macro,match,pattern,in,patterns,expressions,implicit,forall,exptsm,pattsm},deletekeywords={double},classoffset=0,
+numbers=left, numberstyle=\scriptsize\color{mygray}, language=ML,moredelim=[il][\sffamily]{?},showspaces=false,showstringspaces=false,xleftmargin=0pt, numbersep=-6pt, morekeywords=[1]{spliced,tyfam,opfam,let,fn,val,def,casetype,objtype,metadata,of,*,string,datatype,new,toast,syntax,module,switch,where,using,import,for,ana,syn,opcon,tycon,metasignature,metamodule,metasig,metamod,static,at,by,tycase,mod,macro,match,pattern,in,patterns,expressions,implicit,forall,exptsm,pattsm},deletekeywords={double},classoffset=0, xleftmargin=-10pt,
 aboveskip=4pt,belowskip=2pt,
@@
 %% issue of PACMPL.
-\citestyle{acmauthoryear}   %% For author/year citations
+% \citestyle{acmauthoryear}   %% For author/year citations
 
@@
 %% Title information
-\title[Reasonably Programmable Literal Forms]{Reasonably Programmable Literal Forms}         %% [Short Title] is optional;
+\title[Reasonably Programmable Literal Forms]{Reasonably Programmable Literal Notation}         %% [Short Title] is optional;
                                         %% when present, will be used in
@@
 This paper questions this common practice: why should an otherwise ordinary data structure be privileged in the definition of a general-purpose programming language?
-General-purpose programming languages typically provide literal notation for only a small number of standard data structures, like lists.
-This common design practice is decidedly \emph{ad hoc}: there are many other data structures for which literal notation might also be useful, e.g.
-finite maps, regexes, HTML data, SQL queries, syntax trees and chemical structures.
-Availability of literal notation is far from a superficial concern: studies suggest that literal notation can have a serious impact on software quality by subverting string encodings, which are vectors for injection attacks and cause other difficulties.
+General-purpose programming languages typically provide literal notation for only a small number of data structures, like lists.
+This design practice is unsatisfying because there are many other data structures for which literal notation might be useful, e.g.
+finite maps, regexes, HTML data, SQL queries, syntax trees and encodings of chemical structures.
 %A less \emph{ad hoc} design would be one that allows library providers to define new literal forms in a decentralized manner.
@@
 the Scala macro system \cite{ScalaMacros2013}, are only somewhat more reasonable, and do not offer direct syntactic control.
-This paper develops, from type-theoretic first principles, a mechanism that decentralizes control over literal notation: \emph{typed literal macros (TLMs)}.
-TLMs give library providers programmatic control over the parsing and expansion of expressions and patterns of a flexible \emph{generalized literal form} at {a} specified type or parameteric family of types.
+This paper introduces \emph{typed literal macros (TLMs)}, which allow library providers to define new literal notation of nearly arbitrary form at a specified type or parameterized family of types. 
+% TLMs give library providers programmatic control over the parsing and expansion of expressions and patterns of a flexible \emph{generalized literal form} at {a} specified type or parameteric family of types.
 % The mechanism is {strictly hygienic}, meaning that 1) the expansion must be context-independent; and 2)  splicing is capture avoiding.
 % Partial parameter application lowers the syntactic cost of this strict style. 
-Compared to existing approaches, including syntax extension systems and macro-based quotation systems, TLMs are uniquely \emph{reasonable}: TLM providers can reason modularly about syntactic determinism, and TLM clients can reason abstractly about the type and binding structure of the program.
-The system only needs to convey to clients (via secondary notation) the inferred \emph{segmentation} of each literal, which gives the locations of spliced terms.
-Maintaining information about segmentation is necessary for the novel TLM hygiene mechanism, i.e.
+Compared to existing approaches, including direct grammar extension systems and prior  macro-based parsing systems, TLMs are uniquely \emph{reasonable}: TLM providers can reason modularly about syntactic ambiguity, and TLM clients can reason abstractly, i.e.
+without examining the underlying expansion, about types and binding.
+The system only needs to convey to clients via secondary notation the inferred \emph{segmentation} of each literal, which gives the locations of spliced subterms.
+%Maintaining information about segmentation is necessary for the novel TLM hygiene mechanism, i.e.
 the mechanism that ensures that clients can reason abstractly about binding (previously, only unhygienic macro systems supported splicing of terms out of string literal bodies.)
@@
 This paper establishes these abstract reasoning principles formally with a calculus of typed expressions, patterns and modules.
-This calculus is the  first detailed type-theoretic account of a hygienic term-rewriting macro system, of any design, for a language with these essential features of ML.
-We are integrating TLMs into Reason, an emerging alternative front-end for OCaml.% This calculus is the first such formalization of a typed macro system.
+This calculus is the first detailed type-theoretic account of a hygienic term-rewriting macro system, of any design, for a language with these essential features of ML, and the first literal macro system equipped with these essential reasoning principles.
+We are integrating TLMs into Reason, a popular alternative front-end for OCaml.% This calculus is the first such formalization of a typed macro system.
 
@@
 % The surface syntax of a programming language is its user interface. 
-When designing the surface syntax of a general-purpose programming language, it is a common practice to build in \emph{literal forms} that decrease the syntactic cost of constructing and pattern matching over values of a select few standard types.
+\paragraph{Motivation} When designing the textual syntax of a general-purpose programming language, it is a common practice to build in \emph{literal forms} that decrease the syntactic cost of constructing and pattern matching over values of a select few standard types.
 For example, in Standard ML (SML) and its cousins, a list expression literal like \li{[x, y, z]} is equivalent by desugaring to \li{Cons(x, Cons(y, Cons(z, Nil)))} where \li{Nil} and \li{Cons} stand for the constructors of the standard \li{list} datatype \cite{mthm97-for-dart,harper1997programming}.\footnote{In SML, list literals work even in contexts where the list expression constructors have been shadowed by other bindings, i.e.
 the meaning of literal forms is \emph{context independent}.
-We will return to the concept of context independence again.} Another example is found in Ur/Web's surface syntax, which extends Ur's surface syntax with expression and pattern literal forms for encodings of HTML data \cite{conf/popl/Chlipala15,conf/pldi/Chlipala10}.
-Fig.
-\ref{fig:urweb} shows two such HTML literals, one that ``splices in'' a string expression (Line 1) and the other an HTML expression (Line 2).
+We will return to the concept of context independence again.} Another example is found in Ur/Web's textual syntax, which extends Ur's syntax with expression and pattern literal forms for encodings of HTML data \cite{conf/popl/Chlipala15,conf/pldi/Chlipala10}.
+Figure \ref{fig:urweb} shows two such HTML literals, one that ``splices in'' a string expression (Line 1) and the other an HTML expression (Line 2).
 %The desugarings of these literal expressions, not shown, are substantially more verbose and less readable to programmers familiar with XHTML.
 
-This common language design practice is decidedly \emph{ad hoc} -- it is quite easy to come up with other examples of data structures that are not privileged with literal forms in these languages, but for which mathematicians, scientists or programmers have invented specialized notation \cite{DBLP:journals/cacm/Iverson80,cajori1928history}.
+This language design practice is decidedly \emph{ad hoc} -- it is quite easy to come up with other examples of data structures that are not privileged with literal forms in these languages, but for which mathematicians, scientists or programmers have invented specialized notation \cite{DBLP:journals/cacm/Iverson80,cajori1928history}.
 Consider, for example, that 1) clients of a ``collections'' library might like not just list literals, but also matrix, set and map literals following standard mathematical conventions; 2) clients of a ``web programming'' library might like not just HTML literals but also CSS literals (which Ur/Web does not have); 4) a compiler writer might like ``quotation'' literals for the terms of the object language and various intermediate languages; and 5) clients of a ``chemistry'' library might like chemical structure literals based on the SMILES standard \cite{anderson1987smiles}.
 
-Although concerns about ``missing'' literal notation are easy to dismiss as superficial or frivolous, the reality is that literal forms, or the absence thereof, have a substantial influence on software quality.
-For example, \citet{Bravenboer:2007:PIA:1289971.1289975} finds that literal forms for structured encodings of queries, like the SQL-based query literals now found in many languages \cite{meijer2006linq}, reduce the temptation to use string encodings of queries and therefore reduce the prevalence of string injection attacks, which are amongst the most common and serious security vulnerabilities on web servers today \cite{owasp2017}.
+Although concerns about the availability of literal notation are easy to dismiss as superficial or frivolous, the reality is that literal notation, or the absence thereof, can have a substantial influence on software quality.
+For example, \citet{Bravenboer:2007:PIA:1289971.1289975} finds that literal notation for structured encodings of queries, like the SQL-based query literals now found in many languages \cite{meijer2006linq}, reduce the temptation to use string encodings of queries and therefore reduce the prevalence of string injection attacks, which are amongst the most common and serious security vulnerabilities on web servers today \cite{owasp2017}.
 More generally, evidence suggests that programmers frequently engage in ``stringly-typed programming'', i.e.
-they choose strings instead of composite data structures largely for reasons of syntactic convenience.
+they choose strings instead of composite data structures largely for reasons of notational convenience.
 In particular, \citet{TSLs} sampled strings from open source projects and found that about 15\% of them were parseable according to some readily apparent type-specific grammar, e.g.
@@
 
-\begin{figure}
+\begin{figure*}
 \vspace{-2px}
@@
 \vspace{-10px}
-\caption{HTML literals with support for splicing at two different types are built primitively into Ur/Web.}
+\caption{HTML literals with support for splicing at two different types are built primitively into Ur/Web \cite{conf/popl/Chlipala15}.}
 \vspace{-8px}
 \label{fig:urweb}
-\end{figure}
+\end{figure*}
 
@@
 Instead, there has been considerable interest in mechanisms that allow library providers to define new literal notation on their own.
-For example, general syntax extension mechanisms, including Camlp4 \cite{ocaml-manual}, SugarJ/Sugar* \cite{erdweg2011sugarj,erdweg2013framework} and others that we will return to in Sec.
+For example, general syntax extension mechanisms, including Camlp4 \cite{ocaml-manual}, SugarJ/Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, macro-based systems like Omar et al's type-specific languages (TSLs) \cite{TSLs}, and others that we will return to in Sec.
 \ref{sec:existing-approaches}, can be used to define new literal forms.
-The problem that specifically motivates this paper is that these mechanisms can obscure program structure and are therefore \emph{unreasonable} for ``programming in the large'' \cite{DeRemer76}.
+
+\paragraph{Problem} The problem that specifically motivates this paper is that these mechanisms obscure the formal structure of the program and thus make it difficult for programmers to reason about the program text directly.
 %, i.e. when developing large programs ``consisting of many small programs (modules), possibly written by different people''
+
 Consider, for instance, the perspective of a programmer attempting to comprehend (i.e.
-reason about) the program text in Fig.
-\ref{fig:K-dialect}, which is written in a dialect of OCaml's syntax called Reason\footnote{\url{http://facebook.github.io/reason}} that has, hypothetically, been extended with a large number of new literal forms.
-
-\begin{figure}[h]
-\vspace{-10px}
+reason about) the program text in Figure \ref{fig:K-dialect}, which is written in a dialect of OCaml's syntax called Reason\footnote{\url{http://facebook.github.io/reason}} that has, hypothetically, been extended with some large number of new literal forms.
+
+\begin{figure}[t!]
+\vspace{-2px}
 \begin{lstlisting}
@@
 
-\noindent
 The literal form on Line 3 constructs an encoding of a query in the niche stack-based database query language K, using its terse surface syntax \cite{Whitney:2001:LOR:376284.375783}.
@@
 
+\paragraph{Approach}
+
+\begin{figure}
+\vspace{-2px}
+\begin{lstlisting}
+  let w = compute_w();
+  let x = compute_x(w);
+  let y = $kquery [|SURL(!R)@&{&/EURLxSURL!/:2_!EURLxSURL}'!R}EURL|];
+\end{lstlisting}
+\vspace{-10px}
+\caption{TLMs make examples like the one from Figure \ref{fig:K-dialect} more reasonable.}
+\vspace{-10px}
+\label{fig:K-tsm-example}
+\end{figure}
+
 This paper formally details a more reasonable mechanism for defining new literal forms: \emph{typed literal macros (TLMs)}.
 TLMs give library providers full programmatic control over the parsing and expansion of \emph{generalized literal forms} \cite{TSLs}, which syntactically subsume other literal forms because their bodies are left initially unparsed according to the fixed context-free syntax of the language (i.e.
-they are initially parsed like raw string literals.) For example, Fig.
-\ref{fig:K-tsm-example} applies a TLM, \li{$kquery}, to a generalized literal form delimited by \li{[|} and \li{|]} to express the ``unreasonable'' K query example from Fig.
-\ref{fig:K-dialect}.
+they are initially parsed like raw string literals.) For example, Figure \ref{fig:K-tsm-example} applies a TLM, \li{$kquery}, to a generalized literal form delimited by \li{[|} and \li{|]} to express the ``unreasonable'' K query example from Figure \ref{fig:K-dialect}.
 The applied TLM parses the literal body and generates an expansion statically, during a semantic phase called typed expansion, which generalizes the usual typing phase.
@@
 colors in this document.
-So by examining Fig.
-\ref{fig:K-tsm-example}, the programmer knows that the two instances of \li{x} on Line 3 are parsed as spliced expressions (because they are in black), whereas the \li{R}'s must be parsed in some other way, e.g.
+So by examining Figure \ref{fig:K-tsm-example}, the programmer knows that the two instances of \li{x} on Line 3 are parsed as spliced expressions (because they are in black), whereas the \li{R}'s must be parsed in some other way, e.g.
 as operators in the K query language (because they are in green.)
@@
 
-\begin{figure}
-\vspace{-2px}
-\begin{lstlisting}
-  let w = compute_w();
-  let x = compute_x(w);
-  let y = $kquery [|SURL(!R)@&{&/EURLxSURL!/:2_!EURLxSURL}'!R}EURL|];
-\end{lstlisting}
-\vspace{-10px}
-\caption{TLMs make examples like the one from Fig. \ref{fig:K-dialect} more reasonable.}
-\vspace{-12px}
-\label{fig:K-tsm-example}
-\end{figure}
-
 The remainder of this document is organized as follows.
@@
 Sec.
-\ref{sec:setsms} defines another {simple expression TLM} -- one that implements the Ur/Web-style HTML literal syntax from Fig.
-\ref{fig:urweb} -- and uses it as a running example to informally describe how TLMs work from start to finish.
+\ref{sec:setsms} defines another {simple expression TLM} -- one that implements the Ur/Web-style HTML literal syntax from Figure \ref{fig:urweb} -- and uses it as a running example to informally describe how TLMs work from start to finish.
 % \item 
@@
 TLMs for abbreviating expressions of a single specified type.
-In particular, we will consider an seTLM named \li{$html} that implements Ur/Web style HTML literal notation for expressions of the \li{html} type that is defined in Fig.
-\ref{fig:html-type-def}.
+In particular, we will consider an seTLM named \li{$html} that implements Ur/Web style HTML literal notation for expressions of the \li{html} type that is defined in Figure \ref{fig:html-type-def}.
 
@@
 \vspace{-2px}
-The definition of the \lstinline{$html} TLM is outlined in Fig. \ref{fig:html-tlm-def}. TLM names are prefixed by \li{$} 
+The definition of the \lstinline{$html} TLM is outlined in Figure \ref{fig:html-tlm-def}. TLM names are prefixed by \li{$} 
  to distinguish them from variable identifiers.
@@
 
-Using the TLM defined in Fig. \ref{fig:html-tlm-def}, we can express the example Ur/Web program from Fig. \ref{fig:urweb} as shown in Fig. \ref{fig:first-tsm-example}. On both lines, we apply 
+Using the TLM defined in Figure \ref{fig:html-tlm-def}, we can express the example Ur/Web program from Figure \ref{fig:urweb} as shown in Figure \ref{fig:first-tsm-example}. On both lines, we apply 
 \li{$html} 
@@
 The input type, \lstinline{body}, classifies encodings of literal {bodies}.
-Literal bodies are sequences of characters, so it suffices to define \li{body} as an abbreviation for \li{string}, as shown in Fig.
-\ref{fig:indexrange-and-parseresult}.
-The return type is a labeled sum type, defined by applying the type function \li{parse_result} defined in Fig.
-\ref{fig:indexrange-and-parseresult}, that distinguishes between parse errors and successful parses.\footnote{We might alternatively have used an exception to signal a parse error, but sum types are formally simpler.} Let us consider these two possibilities in turn.
+Literal bodies are sequences of characters, so it suffices to define \li{body} as an abbreviation for \li{string}, as shown in Figure \ref{fig:indexrange-and-parseresult}.
+The return type is a labeled sum type, defined by applying the type function \li{parse_result} defined in Figure \ref{fig:indexrange-and-parseresult}, that distinguishes between parse errors and successful parses.\footnote{We might alternatively have used an exception to signal a parse error, but sum types are formally simpler.} Let us consider these two possibilities in turn.
 
  If the parse function determines that the literal body is not well-formed according to whatever syntax definition that it implements, it must return \li{ParseError \{msg=}$e_\text{msg}$\li{, loc=}$e_\text{loc}$\li{\}}
-where $e_\text{msg}$ is a custom error message and $e_\text{loc}$ is a value of type \li{segment}, defined in Fig.
-\ref{fig:indexrange-and-parseresult}, that designates a segment of the literal body as the origin of the error \cite{DBLP:journals/jsc/DeursenKT93}.
+where $e_\text{msg}$ is a custom error message and $e_\text{loc}$ is a value of type \li{segment}, defined in Figure \ref{fig:indexrange-and-parseresult}, that designates a segment of the literal body as the origin of the error \cite{DBLP:journals/jsc/DeursenKT93}.
 This information is reported to the programmer by the Reason compiler or editor services.
@@
 For expression TLMs, the proto-expansions is a \emph{proto-expression}.
-These are encoded as values of the recursive sum type \lstinline{proto_expr} that is outlined in Fig.
-\ref{fig:candidate-exp-Reason}.\footnote{More sophisticated encoding techniques for syntax trees would cause no fundamental problems; we have kept it simple here for the sake of clarity.}
+These are encoded as values of the recursive sum type \lstinline{proto_expr} that is outlined in Figure \ref{fig:candidate-exp-Reason}.\footnote{More sophisticated encoding techniques for syntax trees would cause no fundamental problems; we have kept it simple here for the sake of clarity.}
 Most of the constructors of \li{proto_expr} are individually uninteresting -- they encode Reason's various expression forms (just as in a self-hosted compiler, e.g.
 SML/NJ's Visible Compiler library.)
-Expressions can mention types, so we also need the type \li{proto_typ} also outlined in Fig.
-\ref{fig:candidate-exp-Reason}.
+Expressions can mention types, so we also need the type \li{proto_typ} also outlined in Figure \ref{fig:candidate-exp-Reason}.
 It is only the \li{SplicedE} and \li{SplicedT} constructors that are novel.
@@
 
-For example, consider again the TLM application on Line 1 of Fig. \ref{fig:first-tsm-example}, reproduced below:
+For example, consider again the TLM application on Line 1 of Figure \ref{fig:first-tsm-example}, reproduced below:
 \begin{lstlisting}[numbers=none]
@@
 The return type of the parse function is \li{parse_result(proto_pat)}, rather than \li{parse_result(proto_expr)}.
-The type \li{proto_pat}, outlined in Fig.
-\ref{fig:CEPat}, classifies encodings of \emph{proto-patterns}.
+The type \li{proto_pat}, outlined in Figure \ref{fig:CEPat}, classifies encodings of \emph{proto-patterns}.
 
@@
 
-The type abbreviated $\tCEExp$ classifies encodings of \emph{proto-expressions}, $\ce$ (pronounced ``grave $e$''.) The syntax of proto-expressions, defined in Fig.
-\ref{fig:U-candidate-terms}, will be described when we describe proto-expansion validation in Sec.
+The type abbreviated $\tCEExp$ classifies encodings of \emph{proto-expressions}, $\ce$ (pronounced ``grave $e$''.) The syntax of proto-expressions, defined in Figure \ref{fig:U-candidate-terms}, will be described when we describe proto-expansion validation in Sec.
 \ref{sec:ce-syntax-U}.
@@
 
-The main conceptual difference between $\miniVersePat$ and $\miniVerseParam$ is that $\miniVerseParam$ introduces the notion of unexpanded and expanded TLM expressions and types, as shown in Fig.
-\ref{fig:P-macro-expressions-types}.
+The main conceptual difference between $\miniVersePat$ and $\miniVerseParam$ is that $\miniVerseParam$ introduces the notion of unexpanded and expanded TLM expressions and types, as shown in Figure \ref{fig:P-macro-expressions-types}.
 
@@
 \vspace{-2px}
-Fig.
-\ref{fig:static-module-example} shows an example of a module, \li{ParserCombos}, that defines a number of parser combinators following \citet{Hutton1992d}.
+Figure \ref{fig:static-module-example} shows an example of a module, \li{ParserCombos}, that defines a number of parser combinators following \citet{Hutton1992d}.
 The \li{static} qualifier indicates that this module is bound for use only within similarly qualified values, including in particular parse functions of subsequent TLM definitions.
@@
 One approach available to library providers seeking to introduce new literal forms is to use a syntax definition system to construct a library-specific \emph{syntax dialect}: a new syntax definition that extends the syntax definition given in the language definition with new forms, including literal forms.
-For example, Ur/Web's syntax (Fig.
-1) is a library-specific dialect of Ur's syntax \cite{conf/pldi/Chlipala10,conf/popl/Chlipala15}.
+For example, Ur/Web's syntax (Figure 1) is a library-specific dialect of Ur's syntax \cite{conf/pldi/Chlipala10,conf/popl/Chlipala15}.
 

### 2017-10-28 pldi18/syntax-pldi18.tex
intro revisions

@@
 \lstset{tabsize=2, 
-basicstyle=\ttfamily\fontsize{9pt}{1em}\selectfont, 
+basicstyle=\ttfamily\fontsize{8pt}{1em}\selectfont, 
 commentstyle=\itshape\ttfamily\color{gray},
@@
 %% Title information
-\title[Reasonably Programmable Literal Forms]{Reasonably Programmable Literal Notation}         %% [Short Title] is optional;
+\title[Reasonably Programmable Literal Notation]{Reasonably Programmable Literal Notation}         %% [Short Title] is optional;
                                         %% when present, will be used in
@@
 This paper questions this common practice: why should an otherwise ordinary data structure be privileged in the definition of a general-purpose programming language?
-General-purpose programming languages typically provide literal notation for only a small number of data structures, like lists.
+General-purpose programming languages typically define literal notation for only a small number of data structures, e.g.
+lists.
 This design practice is unsatisfying because there are many other data structures for which literal notation might be useful, e.g.
@@
 the Scala macro system \cite{ScalaMacros2013}, are only somewhat more reasonable, and do not offer direct syntactic control.
-This paper introduces \emph{typed literal macros (TLMs)}, which allow library providers to define new literal notation of nearly arbitrary form at a specified type or parameterized family of types. 
+This paper introduces \emph{typed literal macros (TLMs)}, which allow library providers to define new literal notation of nearly arbitrary form at any specified type or parameterized family of types. 
 % TLMs give library providers programmatic control over the parsing and expansion of expressions and patterns of a flexible \emph{generalized literal form} at {a} specified type or parameteric family of types.
@@
 This paper establishes these abstract reasoning principles formally with a calculus of typed expressions, patterns and modules.
-This calculus is the first detailed type-theoretic account of a hygienic term-rewriting macro system, of any design, for a language with these essential features of ML, and the first literal macro system equipped with these essential reasoning principles.
-We are integrating TLMs into Reason, a popular alternative front-end for OCaml.% This calculus is the first such formalization of a typed macro system.
+This calculus is the first detailed type-theoretic account of a hygienic macro system, of any design, for a language with these essential features of ML, and the first macro system with support for literal parsing that comes equipped with these essential reasoning principles.
+We are integrating TLMs into Reason, an increasingly popular alternative front-end for OCaml.% This calculus is the first such formalization of a typed macro system.
 
@@
 % The surface syntax of a programming language is its user interface. 
-\paragraph{Motivation} When designing the textual syntax of a general-purpose programming language, it is a common practice to build in \emph{literal forms} that decrease the syntactic cost of constructing and pattern matching over values of a select few standard types.
-For example, in Standard ML (SML) and its cousins, a list expression literal like \li{[x, y, z]} is equivalent by desugaring to \li{Cons(x, Cons(y, Cons(z, Nil)))} where \li{Nil} and \li{Cons} stand for the constructors of the standard \li{list} datatype \cite{mthm97-for-dart,harper1997programming}.\footnote{In SML, list literals work even in contexts where the list expression constructors have been shadowed by other bindings, i.e.
-the meaning of literal forms is \emph{context independent}.
-We will return to the concept of context independence again.} Another example is found in Ur/Web's textual syntax, which extends Ur's syntax with expression and pattern literal forms for encodings of HTML data \cite{conf/popl/Chlipala15,conf/pldi/Chlipala10}.
+When designing the surface syntax of a general-purpose programming language, it is common to build in \emph{literal notation} that decreases the syntactic cost of constructing and pattern matching over values of a few standard data structures.
+For example, in Standard ML (SML), the list literal \li{[x, y, z]} is equivalent by desugaring to \li{Cons(x,Cons(y,Cons(z,Nil)))} where \li{Nil} and \li{Cons} stand for the constructors of the standard \li{list} datatype \cite{mthm97-for-dart,harper1997programming}.
+%\footnote{In SML, list literals work even in contexts where the list expression constructors have been shadowed by other bindings, i.e. the meaning of literal forms is \emph{context independent}. 
+% We will return to this concept again later.} 
+Similarly, Ur/Web's surface syntax extends Ur's syntax with expression and pattern literals for encodings of XML and HTML data \cite{conf/popl/Chlipala15,conf/pldi/Chlipala10}.
 Figure \ref{fig:urweb} shows two such HTML literals, one that ``splices in'' a string expression (Line 1) and the other an HTML expression (Line 2).
@@
 This language design practice is decidedly \emph{ad hoc} -- it is quite easy to come up with other examples of data structures that are not privileged with literal forms in these languages, but for which mathematicians, scientists or programmers have invented specialized notation \cite{DBLP:journals/cacm/Iverson80,cajori1928history}.
-Consider, for example, that 1) clients of a ``collections'' library might like not just list literals, but also matrix, set and map literals following standard mathematical conventions; 2) clients of a ``web programming'' library might like not just HTML literals but also CSS literals (which Ur/Web does not have); 4) a compiler writer might like ``quotation'' literals for the terms of the object language and various intermediate languages; and 5) clients of a ``chemistry'' library might like chemical structure literals based on the SMILES standard \cite{anderson1987smiles}.
-
-Although concerns about the availability of literal notation are easy to dismiss as superficial or frivolous, the reality is that literal notation, or the absence thereof, can have a substantial influence on software quality.
+For example, 1) clients of a ``collections'' library might also want matrix, set and map literals following usual mathematical conventions; 2) clients of a ``web programming'' library might like not just HTML literals but also CSS literals (which Ur/Web lacks); 3) a compiler writer might like ``quotation'' literals for the terms of the object language and various intermediate languages; and 4) clients of a ``chemistry'' library might like chemical structure literals based on the SMILES standard \cite{anderson1987smiles}.
+
+Although requests for new literal notations are easy to dismiss as superficial, the reality is that literal notation, or the absence thereof, can have a substantial influence on software quality.
 For example, \citet{Bravenboer:2007:PIA:1289971.1289975} finds that literal notation for structured encodings of queries, like the SQL-based query literals now found in many languages \cite{meijer2006linq}, reduce the temptation to use string encodings of queries and therefore reduce the prevalence of string injection attacks, which are amongst the most common and serious security vulnerabilities on web servers today \cite{owasp2017}.
@@
 \caption{HTML literals with support for splicing at two different types are built primitively into Ur/Web \cite{conf/popl/Chlipala15}.}
-\vspace{-8px}
+\vspace{-10px}
 \label{fig:urweb}
@@
 Of course, it would not be sensible to ask general-purpose language designers to accomodate all known notations with built-in literal forms.
-Instead, there has been considerable interest in mechanisms that allow library providers to define new literal notation on their own.
-For example, general syntax extension mechanisms, including Camlp4 \cite{ocaml-manual}, SugarJ/Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, macro-based systems like Omar et al's type-specific languages (TSLs) \cite{TSLs}, and others that we will return to in Sec.
-\ref{sec:existing-approaches}, can be used to define new literal forms.
-
-\paragraph{Problem} The problem that specifically motivates this paper is that these mechanisms obscure the formal structure of the program and thus make it difficult for programmers to reason about the program text directly.
+Instead, there has been considerable interest for many years in mechanisms that allow library providers to define new literal notation on their own.
+For example, direct grammar extension systems like Camlp4 \cite{ocaml-manual} and SugarJ \cite{erdweg2011sugarj,erdweg2013framework}, macro-based systems like Omar et al's \emph{type-specific languages} (TSLs) \cite{TSLs}, and others that we will discuss in Sec.
+\ref{sec:existing-approaches} can all be used to define new literal notation.
+
+\paragraph{Problem} The problem that motivates this paper is that it is impossible to reason abstractly, i.e.
+by inspecting only the surface syntax, about such fundamental issues as types and variable binding when a program text is littered with  unfamiliar user-defined literal forms.
+Instead, programmers can only reason transparently, i.e.
+by inspecting the underlying expansion or the extension responsible for producing the expansion.
+In some existing systems, even determining which extension is responsible for a literal form is difficult.
 %, i.e. when developing large programs ``consisting of many small programs (modules), possibly written by different people''
-
-Consider, for instance, the perspective of a programmer attempting to comprehend (i.e.
-reason about) the program text in Figure \ref{fig:K-dialect}, which is written in a dialect of OCaml's syntax called Reason\footnote{\url{http://facebook.github.io/reason}} that has, hypothetically, been extended with some large number of new literal forms.
 
@@
 \caption{Syntax extension systems produce unreasonable program text.}
-\vspace{-10px}
+\vspace{-14px}
 \label{fig:K-dialect}
@@
 
-The literal form on Line 3 constructs an encoding of a query in the niche stack-based database query language K, using its terse surface syntax \cite{Whitney:2001:LOR:376284.375783}.
-Of course, the programmer reading this program text may be unfamiliar with K's query syntax.
-Unfortunately for this programmer, there are no simple, abstract protocols for answering questions like:
-
-\begin{enumerate}[nolistsep,noitemsep]
-\item \textbf{(Responsibility)} Which syntax extension determined the desugaring of the literal form on Line 3? Might there be many possible desugarings (i.e. syntactic non-determinism)?
-\item \textbf{(Segmentation)} Are the characters \li{x} and \li{R} on Line 3 taken as spliced expressions (meaning that they appear directly in the desugaring), or are they parsed in some other way peculiar to this literal form (e.g.
+Consider, for example, the perspective of a programmer attempting to comprehend, i.e.
+reason about, the program text in Figure \ref{fig:K-dialect}, which is written in a dialect of OCaml's syntax called Reason \cite{reason} that has, hypothetically, been extended with some large number of new literal forms by a grammar extension system, like Camlp4 \cite{ocaml-manual} or Sugar* \cite{erdweg2011sugarj,erdweg2013framework}.
+The literal form on Line 3 constructs an encoding of a query in the niche stack-based database query language K, using its intentionally terse notation \cite{Whitney:2001:LOR:376284.375783}.
+The problem is that a programmer unfamiliar with this notation cannot fall back on any abstract reasoning principles to answer questions like:
+
+\begin{enumerate}[nolistsep,noitemsep,leftmargin=10pt]
+\item \textbf{(Responsibility)} Which syntax extension determined the expansion of the literal form on Line 3?
+Might different extensions generate different expansions for this form (i.e.
+is there syntactic ambiguity)?
+\item \textbf{(Segmentation)} Are the characters \li{x} and \li{R} on Line 3 parsed as spliced expressions (meaning that they appear directly in the underlying expansion), or are they parsed in some other way peculiar to this literal form (e.g.
 as operators in the K query language)?
-\item \textbf{(Capture)} If \li{x} is in fact a spliced expression, does it refer to the binding of \li{x} on Line 2, or might it capture an unseen binding of the same identifier in the desugaring of Line 3?
-\item \textbf{(Context Dependence)} If \li{w}, on Lines 1-2, is renamed, could that possibly break Line 3 or change its meaning because the desugaring assumes that some variable \li{w} is in scope?% What other hidden assumptions might the desugaring be making?
+\item \textbf{(Capture)} If \li{x} is in fact a spliced expression, does it refer to the binding of \li{x} on Line 2, or might it capture an unseen binding of the same identifier in the expansion of Line 3?
+\item \textbf{(Context Dependence)} If \li{w}, on Lines 1-2, is renamed, could that possibly break Line 3 or change its meaning because the expansion assumes that \li{w} is in scope?% What other hidden assumptions might the desugaring be making?
 \item \textbf{(Typing)} What type does \li{y} have?
@@
 answer questions like those above.) This is burdensome at all scales, but particularly when programming in the large, where it is common to encounter a wide variety of libraries \cite{DBLP:conf/sac/LammelPS11}.
-Forcing the programmer to reason transparently, i.e.
-to examine the desugaring of the program text, or even the logic that produced that desugaring, to answer basic questions like these defeats the ultimate purpose of syntactic sugar: decreasing cognitive cost \cite{Green89}.
+Forcing the programmer to reason transparently to answer basic questions like these defeats the ultimate purpose of syntactic sugar: decreasing cognitive cost \cite{Green89}.
 Analagous problems do not arise when programming without syntax extensions in languages like ML and Scala -- the binding discipline of the language ensures that programmers can reason lexically about binding, and types mediate abstraction over function and module implementations.
@@
 \paragraph{Approach}
+
+This paper formally defines a more reasonable mechanism for defining new literal notation: \emph{typed literal macros (TLMs)}.
+TLMs give library providers  programmatic control over the parsing and expansion of \emph{generalized literal forms} \cite{TSLs}, which syntactically subsume other literal forms because their bodies are left initially unparsed according to the fixed context-free syntax of the language (i.e.
+they are initially parsed like raw string literals.)
 
@@
 \caption{TLMs make examples like the one from Figure \ref{fig:K-dialect} more reasonable.}
-\vspace{-10px}
+\vspace{-14px}
 \label{fig:K-tsm-example}
@@
 
-This paper formally details a more reasonable mechanism for defining new literal forms: \emph{typed literal macros (TLMs)}.
-TLMs give library providers full programmatic control over the parsing and expansion of \emph{generalized literal forms} \cite{TSLs}, which syntactically subsume other literal forms because their bodies are left initially unparsed according to the fixed context-free syntax of the language (i.e.
-they are initially parsed like raw string literals.) For example, Figure \ref{fig:K-tsm-example} applies a TLM, \li{$kquery}, to a generalized literal form delimited by \li{[|} and \li{|]} to express the ``unreasonable'' K query example from Figure \ref{fig:K-dialect}.
-The applied TLM parses the literal body and generates an expansion statically, during a semantic phase called typed expansion, which generalizes the usual typing phase.
-TLMs come equipped with uniquely powerful abstract reasoning principles.
-We will characterize these reasoning principles precisely as we proceed, but for now, let us develop the basic intuitions -- the programmer can reason abstractly about Line 3 as follows:
-\begin{enumerate}[noitemsep,nolistsep]
+For example, Figure \ref{fig:K-tsm-example} applies a TLM named \li{$kquery} to a generalized literal form delimited by \li{[|} and \li{|]} to express the K query example from Figure \ref{fig:K-dialect} just discussed.
+The applied TLM parses the literal body (i.e.
+the characters between the delimiters) and generates an expansion statically.
+During the typing phase, a validation step ensures that programmers can reason abstractly about the five issues just identified.
+We will characterize these reasoning principles precisely as we proceed, but for now, let us develop the basic intuition -- the programmer can reason abstractly about Line 3 as follows:
+\begin{enumerate}[noitemsep,nolistsep,leftmargin=10pt]
 \item \textbf{(Responsibility)} The applied TLM, \li{$kquery}, parses and determines the expansion of the literal body.
-The context-free syntax of the language is fixed, so the TLM provider is able to modularly establish that the syntax this TLM implements is deterministic.
-\item \textbf{(Segmentation)} The output that the TLM generates is structured so that the system can infer from it an accurate \emph{segmentation} of the literal body that distinguishes spliced terms from segments that are parsed in some other way by the TLM.
+The context-free syntax of the language is fixed, so the TLM provider is able to modularly establish that the syntax that this parser implements is unambiguous (even in situations where a spliced term itself applies a TLM, not shown here.)
+\item \textbf{(Segmentation)} The intermediate output that the TLM generates is structured so that the system can infer from it an accurate \emph{segmentation} of the literal body that distinguishes spliced terms from segments that are parsed in some other way by the TLM.
 The segmentation can be communicated by a code editor using secondary notation, e.g.
 colors in this document.
-So by examining Figure \ref{fig:K-tsm-example}, the programmer knows that the two instances of \li{x} on Line 3 are parsed as spliced expressions (because they are in black), whereas the \li{R}'s must be parsed in some other way, e.g.
-as operators in the K query language (because they are in green.)
+So by examining Figure \ref{fig:K-tsm-example}, the programmer knows that the two instances of \li{x} on Line 3 are parsed as spliced expressions (because they are black), whereas the \li{R}'s must be parsed in some other way, e.g.
+as operators in the K language (because they are  green.)
 \item \textbf{(Capture)} Splicing is capture-avoiding, so the spliced expression \li{x} must refer to the binding of \li{x} on Line 2.
@@
 \item \textbf{(Context Dependence)} The system enforces context independence, so the expansion of Line 3 cannot  rely on the fact that, e.g., \li{w} is in scope. The client is free to rename \li{w}.
-\item \textbf{(Typing)} An explicit type annotation on the definition of \li{$kquery} determines the type that the expansion must have.
+\item \textbf{(Typing)} A type annotation on the definition of \li{$kquery} (not shown) determines the type that the expansion must have.
 We will see an example of a TLM definition in the next section.
@@
 
-The remainder of this document is organized as follows.
+\paragraph{Outline} 
 % \begin{itemize}
 % \item 
-Sec.
-\ref{sec:setsms} defines another {simple expression TLM} -- one that implements the Ur/Web-style HTML literal syntax from Figure \ref{fig:urweb} -- and uses it as a running example to informally describe how TLMs work from start to finish.
+We begin in Sec.~\ref{sec:setsms} with a more detailed case study: a TLM that implements the HTML literal notation shown in Figure \ref{fig:urweb}.
+We use this case study as a running example to informally describe how TLMs operate, from start to finish.
 % \item 
-Sec. \ref{sec:sptsms} then briefly introduces {pattern TLMs}, i.e. TLMs that generate patterns. %The hygiene condition for pattern TLMs is interesting.
+Sec.~\ref{sec:sptsms} then introduces {pattern TLMs}, i.e. TLMs that generate patterns, rather than expressions. %The hygiene condition for pattern TLMs is interesting.
 % \item 
-Sec. \ref{sec:setsms-formally} defines a type-theoretic calculus of simple expression and pattern TLMs and formally establishes the reasoning principles implied above. 
+Sec.~\ref{sec:ptsms} adds type functions and an ML-style module system and introduces \emph{parametric TLMs} (pTLMs), i.e.
+TLMs that can take type and module parameters.
+This section is organized around a second case study: finite map literals.
+Sec.~\ref{sec:static-eval} considers the topic of term evaluation during the  expansion phase, i.e.
+\emph{static evaluation}, in  more detail.
+Sec \ref{sec:static-eval} also gives examples of TLMs that are useful for defining other TLMs, e.g.
+TLMs that implement parser generators and quasiquotation, and uses these in a third case study of the mechanism: IPv4 address literals.
+Sec.~\ref{sec:setsms-formally} defines a type-theoretic calculus of simple expression and pattern TLMs and formally establishes the reasoning principles implied above. 
 % \item 
-Sec.
-\ref{sec:ptsms} then adds type functions and an ML-style module system and introduces \emph{parametric TLMs} (pTLMs), i.e.
-TLMs that can take type and module parameters.
-Parameters serve two important purposes.
+Sec.~\ref{sec:existing-approaches} compares TLMs to related work using the rubric of five reasoning principles just discussed. 
+% \item 
+Sec.~\ref{sec:discussion} concludes with a summary of the specific contributions of this paper and a discussion of the limitations of TLMs, together with suggestions for future work.
+
+% Parameters serve two important purposes.
 First, they allow a TLM to operate not just at a single type, but over a type- and module-parameterized family of types.
@@
 Second, parameters lower the syntactic cost of the strict context independence discipline that TLMs enforce because they give expansions access to terms other than those explicitly spliced in to the literal body by the client, and because %This is particularly convenient for working with abstract types \emph{a la} ML.
-partial parameter application is possible.
+% partial parameter application is possible.
 % Sec.
@@
 % \item
-Sec.
-\ref{sec:static-eval} considers the topic of term evaluation during the typed expansion phase, i.e.
-\emph{static evaluation}, in  more detail, and also gives examples of TLMs that are useful for defining other TLMs, e.g.
-TLMs that implement parser generators and quasiquotation.
 % \item 
-Sec. \ref{sec:existing-approaches} compares TLMs to related work, including recent related work by \citet{TSLs,sac15}, and the long history of related work on macro systems.
-% \item 
-Sec. \ref{sec:discussion} concludes with a summary of the specific contributions of this paper and a brief discussion of the limitations of TLMs, along with some suggestions for future work.
 % \end{itemize}
 
-Due to space limitations, certain technical details of the calculi described in this paper, as well as detailed proofs, are given in the supplemental material.
-The examples in the remainder of the paper are written in an open source alternative syntactic front-end for OCaml called Reason, mentioned above, because we are working to incorporate TLMs into Reason.
-However, the essential ideas are not Reason-, OCaml- or ML-specific -- it should be possible to adapt TLMs to other languages that take a similar approach to types and binding (e.g.
-Haskell, Scala and others.) Languages that have a disciplined binding structure but that lack a rich static type structure, e.g.
-standard Racket, would also benefit from TLMs -- they can apply the results of this paper by deploying the usual trick of viewing the language as statically ``{unityped}'' \cite{pfpl,scott1980lambda}.%  We say more about LISP macros in Sec.
+Due to the space limitations, certain technical details and extensions to the calculi described in this paper, as well as detailed proofs, are given in the supplemental material.
+Some notes on implementation and a fourth case study, involving an encoding of regular expressions as an abstract data type, are also both available in the supplement.
+
+%  We say more about LISP macros in Sec.
 \ref{sec:existing-approaches}.% For the present purposes, we can consider Reason as essentially the calculus defined in the supplement extended with various conveniences that are commonly found in other ML-like languages and, notionally, orthogonal to TLMs.
@@
 
+The examples in this paper are written in an open source alternative syntactic front-end for OCaml called Reason because we are incorporating TLMs into Reason.
+However, the essential ideas are not Reason-, OCaml- or ML-specific -- it should be possible to adapt TLMs to other languages that take a similar approach to types and binding (e.g.
+Haskell, Scala and others.) Languages that have a disciplined binding structure but that lack a rich static type structure, e.g.
+standard Racket, would also benefit from TLMs -- they can apply these results by deploying the usual trick of viewing the language as statically ``{unityped}'' \cite{pfpl,scott1980lambda}.
+
 % The intended audience for this paper is language designers who want to decentralize control over literal forms but seek a detailed understanding, rooted in the first principles of type theory, of a reasonable mechanism by which to achieve that goal.

### 2017-10-28 pldi18/syntax-pldi18.tex
figure tweaks

@@
 mathescape=false,escapechar=~,
-numbers=left, numberstyle=\scriptsize\color{mygray}, language=ML,moredelim=[il][\sffamily]{?},showspaces=false,showstringspaces=false,xleftmargin=0pt, numbersep=-6pt, morekeywords=[1]{spliced,tyfam,opfam,let,fn,val,def,casetype,objtype,metadata,of,*,string,datatype,new,toast,syntax,module,switch,where,using,import,for,ana,syn,opcon,tycon,metasignature,metamodule,metasig,metamod,static,at,by,tycase,mod,macro,match,pattern,in,patterns,expressions,implicit,forall,exptsm,pattsm},deletekeywords={double},classoffset=0, xleftmargin=-10pt,
+numbers=left, numberstyle=\scriptsize\color{mygray}, language=ML,moredelim=[il][\sffamily]{?},showspaces=false,showstringspaces=false,xleftmargin=0pt, numbersep=-6pt, morekeywords=[1]{try,spliced,tyfam,opfam,let,fn,val,def,casetype,objtype,metadata,of,*,string,datatype,new,toast,syntax,module,switch,where,using,import,for,ana,syn,opcon,tycon,metasignature,metamodule,metasig,metamod,static,at,by,tycase,mod,macro,match,pattern,in,patterns,expressions,implicit,forall,exptsm,pattsm},deletekeywords={double},classoffset=0, xleftmargin=-10pt,
 aboveskip=4pt,belowskip=2pt,
@@
 For example, if strings were not primitive, we might write \li{`SURL<body>\{EURLheading ($str "SSTRWorld!ESTR")SURL\} ...</body>EURL`}.}
+\begin{figure*}[t]
+
+\begin{subfigure}[t]{0.48\textwidth}
+\begin{lstlisting}
+  type html_attrs = list(string, string)
+  type html = BodyElement(html_attrs, list(html))
+            | H1Element(html_attrs, list(html))
+            | TextNode(string) 
+            | /* ... */;
+\end{lstlisting}
+\vspace{-5px}
+\caption{The \li{html} type, which classifies encodings of HTML data.}
+\label{fig:html-type-def}
+\end{subfigure}
+\hfill
+\begin{subfigure}[t]{0.48\textwidth}
+\begin{lstlisting}[mathescape=|]
+  syntax $html at html by static {
+    fun(b : body) : parse_result(proto_expr) => 
+      try (parse_html b |> html_to_ast |> Success) {
+      | InvalidHTML msg loc => ParseError msg loc }
+  };
+\end{lstlisting}
+\vspace{-5px}
+\caption{The definition of the \li{$html} TLM.
+It calls into an existing HTML parser, then converts the result to an AST, extracting the spliced segments as it goes.
+Figure \ref{fig:candidate-exp-Reason} defines the  TLM-related types.}
+\label{fig:html-tlm-def}
+\end{subfigure}
+
+\begin{subfigure}[t]{\textwidth}
+\begin{lstlisting}
+  let heading first_name = $html [|SURL<h1>Hello, {[EURLfirst_nameSURL]}!</h1>EURL|]
+  let body = $html [|SURL<body>{EURLheading "World!"SURL} ...</body>EURL|]
+\end{lstlisting}
+\vspace{-5px}
+\caption{Two examples of the \li{$html} TLM being applied.
+Compare to Ur/Web's built in HTML literal notation from Figure \ref{fig:urweb}.
+%In each case, literal body, between backticks, is initially left unparsed according to the language's context-free syntax.
+%The applied TLM determines a segmentation and expansion during the typed expansion phase, which generalizes the usual typing phase.
+}
+\label{fig:first-tsm-example}
+\end{subfigure}
+\vspace{-8px}
+\caption{Case Study: HTML literals}
+\vspace{-5px}
+\end{figure*}
 
@@
 \vspace{-2px}
+
 We will start in this section by informally describing TLMs in Reason.
@@
 In particular, we will consider an seTLM named \li{$html} that implements Ur/Web style HTML literal notation for expressions of the \li{html} type that is defined in Figure \ref{fig:html-type-def}.
-
-\begin{figure}[ht!]
-\vspace{-10px}
-\begin{lstlisting}[numbers=none]
-  type html = BodyElement(list(html_attr), list(html))
-            | H1Element(list(html_attr), list(html))
-            | TextNode(string) | /* ... */;
-\end{lstlisting}
-\vspace{-10px}
-\caption{The \li{html} type, which classifies encodings of HTML data.
-The type \li{html_attr} is also straightforward and the type function \li{list} is standard, so neither are shown.
-Strings are primitive.}
-\label{fig:html-type-def}
-\vspace{-10px}
-\end{figure}
 
@@
 
-\begin{figure}[ht!]
-\vspace{-12px}
-\begin{lstlisting}[mathescape=|]
-  syntax $html at html by static {
-    fun(b : body) : parse_result(proto_expr) => 
-      /* ... Ur/Web-style HTML syntax parser ... */
-  };
-\end{lstlisting}
-\vspace{-15px}
-\caption{The definition of the seTLM named \li{$html}.}
-\label{fig:html-tlm-def}
-\vspace{-10px}
-\end{figure}
-
 Using the TLM defined in Figure \ref{fig:html-tlm-def}, we can express the example Ur/Web program from Figure \ref{fig:urweb} as shown in Figure \ref{fig:first-tsm-example}. On both lines, we apply 
@@
 string literals where escape sequences like \li{\\n} have not yet been processed.)
-
-\begin{figure}[t!]
-\vspace{-2px}
-\begin{lstlisting}
-  fun heading first_name = $html [|SURL<h1>Hello, {[EURLfirst_nameSURL]}!</h1>EURL|]
-  val body = $html [|SURL<body>{EURLheading "World!"SURL} ...</body>EURL|]
-\end{lstlisting}
-\vspace{-10px}
-\caption{Two examples of the \li{$html} TLM being applied.
-%In each case, literal body, between backticks, is initially left unparsed according to the language's context-free syntax.
-%The applied TLM determines a segmentation and expansion during the typed expansion phase, which generalizes the usual typing phase.
-}
-\vspace{-12px}
-\label{fig:first-tsm-example}
-\end{figure}
 
@@
 \vspace{-2px}
-\begin{lstlisting}[numbers=none]
+\begin{lstlisting}[numbers=none,xleftmargin=0pt]
 type body = string;
-type segment = {startIdx : int, endIdx : int}; /* inclusive */
-type parse_result('a) = ParseError { msg : string, loc : segment }
+type segment = {startIdx : int, endIdx : int};
+type parse_result('a) 
+  = ParseError { msg : string, loc : segment }
   | Success('a);
@@
 \vspace{-10px}
-\caption[Definitions of various types used by TLM definitions.]{Definitions of various types used by TLM definitions. We assume that these are available ambiently.}
+\caption[Definitions of various types used by TLM definitions.]{Definitions of various types available ambiently to TLM definitions.}
 \label{fig:indexrange-and-parseresult}

### 2017-10-29 pldi18/syntax-pldi18.tex
revised sec 2.1

@@
 \newcommand{\liv}[1]{\lstinline{#1}}
-\newcommand{\li}[1]{\lstinline{#1}}
+\newcommand{\li}[1]{\lstinline[basicstyle=\ttfamily\fontsize{9pt}{1em}\selectfont]{#1}}
 
@@
 \vspace{-5px}
-\caption{The definition of the \li{$html} TLM.
-It calls into an existing HTML parser, then converts the result to an AST, extracting the spliced segments as it goes.
-Figure \ref{fig:candidate-exp-Reason} defines the  TLM-related types.}
+\caption{The \li{$html} TLM definition. Figure \ref{fig:candidate-exp-Reason} defines   TLM-related types.}
 \label{fig:html-tlm-def}
@@
 Due to the space limitations, certain technical details and extensions to the calculi described in this paper, as well as detailed proofs, are given in the supplemental material.
-Some notes on implementation and a fourth case study, involving an encoding of regular expressions as an abstract data type, are also both available in the supplement.
+Some notes on implementation and a fourth case study, involving an encoding of regular expressions as an abstract data type, are also both given in the supplement.
 
@@
 \vspace{-4px}
-\newcommand{\seTLMsSec}{Simple Expression TLMs (seTLMs)}
+\newcommand{\seTLMsSec}{Expression TLMs}
 \section{\protect\seTLMsSec}
@@
 
-We will start in this section by informally describing TLMs in Reason.
-It will suffice in this section to consider only \emph{simple expression TLMs (seTLMs)}, i.e.
-TLMs for abbreviating expressions of a single specified type.
-In particular, we will consider an seTLM named \li{$html} that implements Ur/Web style HTML literal notation for expressions of the \li{html} type that is defined in Figure \ref{fig:html-type-def}.
+Let us begin by detailing the TLM mechanism by way of a case study: HTML literal notation for expressions of the \li{html} type defined in Figure \ref{fig:html-type-def}.
 
@@
 \vspace{-2px}
-The definition of the \lstinline{$html} TLM is outlined in Figure \ref{fig:html-tlm-def}. TLM names are prefixed by \li{$} 
- to distinguish them from variable identifiers.
-Every TLM definition also has a \emph{type annotation}, here \lstinline{at html}, and a \emph{parse function} between \li{by static \{} and \li{\}}.
-The TLM definition is in scope until the end of the enclosing scope.
+The definition of the \li{$html} TLM is given in Figure \ref{fig:html-tlm-def}. TLM names are prefixed by \li{$} 
+ to distinguish them from variables.
+Every TLM definition also provides a \emph{type annotation}, here \li{at html}, and a \emph{parse function} between \li{by static \{} and \li{\}}.
+% The TLM definition follows the same scoping rules as type definitions.
 We will describe how TLM definitions are packaged into libraries in Sec \ref{sec:static-eval}.
@@
 Generalized literal forms, which first arose in related work by \citet{TSLs} that we will detail in Sec.
-\ref{sec:existing-approaches}, subsume other literal forms because the context-free syntax of the language only specifies the outer delimiters (in this paper, we will use  \li{[|} and \li{|]}, but \citet{TSLs} formally specified several other choices, including layout-sensitive delimitation.) \emph{Literal bodies}, i.e.
+\ref{sec:existing-approaches}, subsume other literal forms because the context-free syntax of the language only specifies the outer delimiters.
+In this paper, we will use  \li{[|} and \li{|]}, but \citet{TSLs} formally specified several other choices, including layout-sensitive delimitation.
+\emph{Literal bodies}, i.e.
 the characters between these delimiters, are constrained only in that occurrences of \li{[|} must be balanced by \li{|]}.
@@
 
- The semantics delegates responsibility over parsing and expansion of each literal body to the applied TLM's parse function during a semantic phase called \emph{typed expansion}, which generalizes the usual typing phase.
-Because the {parse function}  is applied during the typed expansion process, rather than at run-time, we call it a \emph{static function}.
-Static functions  cannot refer to the surrounding variable bindings (because those variables stand for run-time values.) For now, we will assume that static functions are completely closed, and that they do not themselves make use of TLMs (we will eliminate these impractical limitations by introducing distinct static phase bindings in Sec.
-\ref{sec:static-eval}.)
-
- The parse function must have type \li{body -> parse_result(proto_expr)}.
-The input type, \lstinline{body}, classifies encodings of literal {bodies}.
-Literal bodies are sequences of characters, so it suffices to define \li{body} as an abbreviation for \li{string}, as shown in Figure \ref{fig:indexrange-and-parseresult}.
-The return type is a labeled sum type, defined by applying the type function \li{parse_result} defined in Figure \ref{fig:indexrange-and-parseresult}, that distinguishes between parse errors and successful parses.\footnote{We might alternatively have used an exception to signal a parse error, but sum types are formally simpler.} Let us consider these two possibilities in turn.
+ The system delegates responsibility over the parsing and expansion of each literal body to the applied TLM's parse function during a semantic phase called \emph{typed expansion}, which generalizes the typing phase.
+Because the parse function is applied during this phase, rather than at run-time, we call it a \emph{static function}.
+Static functions  cannot refer to the surrounding variable bindings (because those variables stand for run-time values.) Instead, there is a separate class of \emph{static bindings}.
+We will return to static bindings in Sec.
+\ref{sec:static-eval}.
+
+ The parse function must have the following type: 
+ \begin{lstlisting}[numbers=none]
+      body -> parse_result(proto_expr)
+ \end{lstlisting}
+ The input type, \li{body}, classifies encodings of literal {bodies}.
+Literal bodies are sequences of characters, so we define \li{body} as a synonym of \li{string}, as shown in \autoref{fig:indexrange-and-parseresult}.
+The return type is a sum type, defined by applying the type function \li{parse_result} defined in \autoref{fig:indexrange-and-parseresult}, that distinguishes between parse errors and successful parses.\footnote{We might alternatively have used an exception to signal a parse error, but sum types are formally simpler.} Let us consider these two possibilities in turn.
 
@@
 where $e_\text{msg}$ is a custom error message and $e_\text{loc}$ is a value of type \li{segment}, defined in Figure \ref{fig:indexrange-and-parseresult}, that designates a segment of the literal body as the origin of the error \cite{DBLP:journals/jsc/DeursenKT93}.
-This information is reported to the programmer by the Reason compiler or editor services.
-
-If, on the other hand, parsing succeeds, the parse function returns \li{Success} ~$\ecand$, 
+This information is for use in reporting the problem to the programmer.
+
+If instead parsing succeeds, the parse function returns \li{Success} ~$\ecand$, 
 where $\ecand$ is called the \emph{encoding of the proto-expansion}.
-For expression TLMs, the proto-expansions is a \emph{proto-expression}.
-These are encoded as values of the recursive sum type \lstinline{proto_expr} that is outlined in Figure \ref{fig:candidate-exp-Reason}.\footnote{More sophisticated encoding techniques for syntax trees would cause no fundamental problems; we have kept it simple here for the sake of clarity.}
-Most of the constructors of \li{proto_expr} are individually uninteresting -- they encode Reason's various expression forms (just as in a self-hosted compiler, e.g.
-SML/NJ's Visible Compiler library.)
+For expression TLMs, the proto-expansion is a \emph{proto-expression} and it is encoded as a value of the recursive datatype \li{proto_expr} that is outlined in Figure \ref{fig:candidate-exp-Reason}.
+Most of the constructors of \li{proto_expr} are individually uninteresting -- they encode OCaml's various expression forms. 
 Expressions can mention types, so we also need the type \li{proto_typ} also outlined in Figure \ref{fig:candidate-exp-Reason}.
@@
 type body = string;
-type segment = {startIdx : int, endIdx : int};
+type segment = {startIdx: int, endIdx: int};
 type parse_result('a) 
-  = ParseError { msg : string, loc : segment }
+  = ParseError {msg: string, loc: segment}
   | Success('a);
@@
 
-For example, consider again the TLM application on Line 1 of Figure \ref{fig:first-tsm-example}, reproduced below:
-\begin{lstlisting}[numbers=none]
-  $html [|SURL<h1>Hello, {[EURLfirst_nameSURL]}</h1>EURL|]
-\end{lstlisting}
+For example, consider again the TLM application on Line 1 of Figure \ref{fig:first-tsm-example}:
+\li{$html [|SURL<h1>Hello, {[EURLfirst_nameSURL]}</h1>EURL|]}.
 The proto-expansion generated by \li{$html}, if written in a textual syntax for proto-expressions where references to spliced expressions are written \li{spliced<startIdx; endIndex; ty>}, is:

### 2017-11-12 pldi18/syntax-pldi18.tex
work on intro

@@
 lists.
-This design practice is unsatisfying because there are many other data structures for which literal notation might be useful, e.g.
+This is unsatisfying because there are many other data structures for which literal notation might be useful, e.g.
 finite maps, regexes, HTML data, SQL queries, syntax trees and encodings of chemical structures.
@@
 This paper establishes these abstract reasoning principles formally with a calculus of typed expressions, patterns and modules. 
-This calculus is the first detailed type-theoretic account of a hygienic macro system, of any design, for a language with these essential features of ML, and the first macro system with support for literal parsing that comes equipped with these essential reasoning principles.
-We are integrating TLMs into Reason, an increasingly popular alternative front-end for OCaml.% This calculus is the first such formalization of a typed macro system.
+%This calculus is the first detailed type-theoretic account of a hygienic macro system, of any design, for a language with these essential features of ML, and the first macro system with support for literal parsing that comes equipped with these essential reasoning principles.
+We are integrating TLMs into Reason, an alternative front-end for OCaml.% This calculus is the first such formalization of a typed macro system.
 
@@
 % The surface syntax of a programming language is its user interface. 
-When designing the surface syntax of a general-purpose programming language, it is common to build in \emph{literal notation} that decreases the syntactic cost of constructing and pattern matching over values of a few standard data structures.
+When designing the surface syntax of a general-purpose programming language, it is common practice to define \emph{literal notation} that decreases the syntactic cost of constructing and pattern matching over values of a few data structures.
 For example, in Standard ML (SML), the list literal \li{[x, y, z]} is equivalent by desugaring to \li{Cons(x,Cons(y,Cons(z,Nil)))} where \li{Nil} and \li{Cons} stand for the constructors of the standard \li{list} datatype \cite{mthm97-for-dart,harper1997programming}.
@@
 
-This language design practice is decidedly \emph{ad hoc} -- it is quite easy to come up with other examples of data structures that are not privileged with literal forms in these languages, but for which mathematicians, scientists or programmers have invented specialized notation \cite{DBLP:journals/cacm/Iverson80,cajori1928history}.
+This language design practice is decidedly \emph{ad hoc}, in that it is easy to come up with other examples of data structures that are not privileged with literal forms in these languages, but for which mathematicians, scientists or programmers have invented specialized notation \cite{DBLP:journals/cacm/Iverson80,cajori1928history}.
 For example, 1) clients of a ``collections'' library might also want matrix, set and map literals following usual mathematical conventions; 2) clients of a ``web programming'' library might like not just HTML literals but also CSS literals (which Ur/Web lacks); 3) a compiler writer might like ``quotation'' literals for the terms of the object language and various intermediate languages; and 4) clients of a ``chemistry'' library might like chemical structure literals based on the SMILES standard \cite{anderson1987smiles}.
 
-Although requests for new literal notations are easy to dismiss as superficial, the reality is that literal notation, or the absence thereof, can have a substantial influence on software quality.
+Although requests for specialized literal notation are easy to dismiss as superficial, the reality is that literal notation, or the absence thereof, can have a substantial influence on software quality.
 For example, \citet{Bravenboer:2007:PIA:1289971.1289975} finds that literal notation for structured encodings of queries, like the SQL-based query literals now found in many languages \cite{meijer2006linq}, reduce the temptation to use string encodings of queries and therefore reduce the prevalence of string injection attacks, which are amongst the most common and serious security vulnerabilities on web servers today \cite{owasp2017}.
@@
 In particular, \citet{TSLs} sampled strings from open source projects and found that about 15\% of them were parseable according to some readily apparent type-specific grammar, e.g.
-for SQL queries, regexes, file system paths, URLs and many other data structures.
+for SQL queries, regexes, URLs and many other data structures.
 Literal forms, with support for splicing, would decrease the syntactic cost of composite encodings of such data, which are more amenable to programmatic manipulation and compositional reasoning than flat string encodings.
@@
 \paragraph{Problem} The problem that motivates this paper is that it is impossible to reason abstractly, i.e.
-by inspecting only the surface syntax, about such fundamental issues as types and variable binding when a program text is littered with  unfamiliar user-defined literal forms.
+by inspecting only the surface syntax, about such fundamental issues as types and variable binding when the program text is littered with  unfamiliar user-defined literal forms.
 Instead, programmers can only reason transparently, i.e.
@@
 \begin{figure}[t!]
-\vspace{-2px}
 \begin{lstlisting}
@@
 \caption{Syntax extension systems produce unreasonable program text.}
-\vspace{-14px}
+\vspace{-11px}
 \label{fig:K-dialect}
@@
 The literal form on Line 3 constructs an encoding of a query in the niche stack-based database query language K, using its intentionally terse notation \cite{Whitney:2001:LOR:376284.375783}.
-The problem is that a programmer unfamiliar with this notation cannot fall back on any abstract reasoning principles to answer questions like:
-
-\begin{enumerate}[nolistsep,noitemsep,leftmargin=10pt]
+The problem is that a programmer unfamiliar with this notation cannot fall back on abstract reasoning principles to answer questions like:
+
+\begin{enumerate}[leftmargin=10pt]
 \item \textbf{(Responsibility)} Which syntax extension determined the expansion of the literal form on Line 3?
@@
 
-\paragraph{Approach}
-
-This paper formally defines a more reasonable mechanism for defining new literal notation: \emph{typed literal macros (TLMs)}.
-TLMs give library providers  programmatic control over the parsing and expansion of \emph{generalized literal forms} \cite{TSLs}, which syntactically subsume other literal forms because their bodies are left initially unparsed according to the fixed context-free syntax of the language (i.e.
-they are initially parsed like raw string literals.)
+An approach that has made some headway against these problems is Omar et al.'s \emph{type-specific languages} (TSLs) \cite{TSLs}.
+In this approach, the language delegates control over the parsing and expansion of certain \emph{generalized literal forms} to a parse function associated with the type that the form is being checked against.
+Generalized literal forms syntactically subsume other literal forms because their bodies (i.e.
+the characters between the outer delimiters, here \li{[|} and \li{|]}) are left  unparsed by the fixed context-free syntax of the language (i.e.
+they are initially parsed like raw string literals.) With TSLs, Line 3 of our example would require a type annotation:
+\begin{lstlisting}[numbers=none]
+    let y : kquery = [|(!R)@&{&/x!/:2_!x}'!R}|]
+\end{lstlisting}
+The parse function associated with the \li{kquery} type would be invoked to parse and expand the literal body.
+
+Although TSLs address the problems of \textbf{Responsibility} and \textbf{Typing}, there is no guarantee that a valid segmentation exists, nor was the problem of reasoning about binding (i.e.
+capture and context dependence) adequately addressed in the prior work.
+Moreover, this prior work was introduced in the context of a very simple monomorphic, nominally typed language.
+There was no support for important features of modern languages in the ML tradition, such as structural types (e.g.
+arrow and tuple types), parameterized types, abstract types, modules, pattern notation and ML-style type inference.
+Finally, parse functions and the expansions they generate needed to be closed, i.e.
+they could not make use of any libraries at all.
+Taken together, we found that TSLs were not suitable as-is for integration into Reason, and other languages with a broadly similar type and binding discipline (e.g.
+Haskell, Scala, Rust and so on.)
+
+\paragraph{Contributions} The aim of this paper is to address all of these deficiencies, resulting in the first system for defining new literal notation that (1) provably maintains all five of the abstract reasoning principles just outlined; and (2) that is expressive enough for integration into a full-scale typed functional language (in this paper, the Reason front-end for OCaml.) We evaluate the first claim by developing a series of formal calculi equipped with proofs of these reasoning principles, and the second claim with several non-trivial case studies that appear throughout the paper.
+The empirical evaluation in the prior work on TSLs \cite{TSLs} supports the claim that a mechanism for defining new literal notation is useful, so our focus is on formal reasoning principles and integration with advanced language features in this paper.
 
 \begin{figure}
-\vspace{-2px}
 \begin{lstlisting}
@@
 \caption{TLMs make examples like the one from Figure \ref{fig:K-dialect} more reasonable.}
-\vspace{-14px}
+\vspace{-10px}
 \label{fig:K-tsm-example}
@@
 
-For example, Figure \ref{fig:K-tsm-example} applies a TLM named \li{$kquery} to a generalized literal form delimited by \li{[|} and \li{|]} to express the K query example from Figure \ref{fig:K-dialect} just discussed.
-The applied TLM parses the literal body (i.e.
-the characters between the delimiters) and generates an expansion statically.
+The most obvious point of departure between TSLs and \emph{typed literal macros} (TLMs), which are the topic of this paper, is that we separate the mechanism for defining new literal notation from the type definition mechanism of the language.
+TLMs are explicitly named and applied.
+For example, Figure \ref{fig:K-tsm-example} applies a TLM named \li{$kquery} to a generalized literal form delimited by \li{[|} and \li{|]} to express the K query example just discussed.
+The applied TLM parses the literal body and generates an expansion statically.
+
 During the typing phase, a validation step ensures that programmers can reason abstractly about the five issues just identified.
-We will characterize these reasoning principles precisely as we proceed, but for now, let us develop the basic intuition -- the programmer can reason abstractly about Line 3 as follows:
-\begin{enumerate}[noitemsep,nolistsep,leftmargin=10pt]
+We will characterize these reasoning principles precisely as we proceed, but for now, let us develop the basic intuition -- the programmer can reason abstractly as follows:
+\begin{enumerate}[nolistsep,leftmargin=10pt]
 \item \textbf{(Responsibility)} The applied TLM, \li{$kquery}, parses and determines the expansion of the literal body.
@@
 \vspace{-2px}
-When the parse function determines that some segment of the literal body should be taken as a spliced expression, according to whatever syntactic criteria it deems suitable, it must not directly insert the syntax tree of that expression into the encoding of the expansion.
-Instead,
-the TLM must indirectly refer to the spliced expression using the \li{SplicedE} variant of \li{proto_expr}, which takes a value of type \li{segment} that indicates the zero-indexed location of the spliced expression relative to the start of the provided literal body.
+When the parse function determines that some segment of the literal body is a spliced expression, according to whatever syntactic criteria it deems suitable, it can indirectly refer to it using the \li{SplicedE} variant of \li{proto_expr}, which takes a value of type \li{segment} that indicates the zero-indexed location of the spliced expression relative to the start of the provided literal body.
 The \li{SplicedE} variant also requires a value of type \li{proto_typ}, which indicates the type that the spliced expression is expected to have.
@@
 \li{$html [|SURL<h1>Hello, {[EURLfirst_nameSURL]}</h1>EURL|]}.
-The proto-expansion generated by \li{$html}, if written in a textual syntax for proto-expressions where references to spliced expressions are written \li{spliced<startIdx; endIndex; ty>}, is:
+The parse function of the \li{$html} TLM (Figure \ref{fig:html-tlm-def}, Lines 2-4) first sends the literal body through a pre-existing HTML parser, \li{parse_html}\todo{which one?}.
+It then passes the result to a function \li{html_to_ast}, not shown, which produces a value of type \li{proto_expr}.
+When this function encounters a text node containing matched \li{\{[} and \li{]\}}, then that segment is inserted as a spliced expression of type \li{string}.
+Similarly, text nodes containing matched curly braces produce a spliced expression of type \li{html}.
+We assume that the pre-existing HTML parser annotates each element or node with its position in the originally parsed text (as most production-grade HTML/XML parsers are capable of doing.) If written in a textual syntax for proto-expressions where references to spliced expressions are written \li{spliced<startIdx; endIndex; ty>}, the output of this function for the example above is:
 \begin{lstlisting}[numbers=none]
@@
 It corresponds to the following encoding:
-\li{SplicedE (\{startIdx: 13, endIdx: 22\}, StringTy)}.
+\begin{lstlisting}[numbers=none]
+    SplicedE ({startIdx=13, endIdx=22}, StringTy)
+\end{lstlisting}
 
@@
 The \emph{segmentation} inferred from a proto-expansion is the finite set of references to spliced terms contained within.
-For example, the segmentation inferred from the proto-expression above is the finite set containing only \li{spliced<13; 22; string>}.
+For example, the segmentation inferred from the proto-expression above contains only \li{spliced<13; 22; string>}.
 % Notice that no information about  the spliced terms appear is communicated by the splice summary.
-The semantics checks that all of the locations in the segmentation are 1) in bounds relative to the literal body; and 2) non-overlapping. 
+The system checks that all of the locations in the segmentation are 1) in bounds relative to the literal body; and 2) non-overlapping. 
 This resolves the problem of \textbf{Segmentation} described in Sec.
@@
 In general, spliced expressions might themselves apply TLMs, in which case the convention is to use a distinct color for unspliced segments at each depth.
-For example, if strings were not primitive but rather defined as sequences of characters, we might define a TLM \li{$str} to recover string literal notation and write \li{$html [|SURL<body>\{EURLheading ($str [|SSTRWorld!ESTR|])SURL\} ...</body>EURL|]}.
+For example, if strings were not primitive but rather defined as sequences of characters, we might define a TLM \li{$str} to recover string literal notation and write \li{$html [|SURL<body>\{EURLheading $str [|SSTRWorld!ESTR|]SURL\} ...</body>EURL|]}.
 

### 2018-02-19 pldi18/syntax-pldi18.tex
forgot to commit these

@@
 
+% \usepackage{titlesec}
+
+% \titlespacing*\section{0pt}{1pt plus 2pt minus 2pt}{2pt plus 2pt minus 2pt}
+% \titlespacing*\subsection{0pt}{1pt plus 2pt minus 2pt}{2pt plus 2pt minus 2pt}
+% \titlespacing*\subsubsection{0pt}{1pt plus 2pt minus 2pt}{2pt plus 2pt minus 2pt}
+% \titlespacing*\paragraph{0pt}{5pt plus 2pt minus 2pt}{2pt plus 2pt minus 2pt}
+
 \makeatletter
@@
 mathescape=false,escapechar=~,
-numbers=left, numberstyle=\scriptsize\color{mygray}, language=ML,moredelim=[il][\sffamily]{?},showspaces=false,showstringspaces=false,xleftmargin=0pt, numbersep=-6pt, morekeywords=[1]{try,spliced,tyfam,opfam,let,fn,val,def,casetype,objtype,metadata,of,*,string,datatype,new,toast,syntax,module,switch,where,using,import,for,ana,syn,opcon,tycon,metasignature,metamodule,metasig,metamod,static,at,by,tycase,mod,macro,match,pattern,in,patterns,expressions,implicit,forall,exptsm,pattsm},deletekeywords={double},classoffset=0, xleftmargin=-10pt,
+numbers=left, numberstyle=\scriptsize\color{mygray}, language=ML,moredelim=[il][\sffamily]{?},showspaces=false,showstringspaces=false,xleftmargin=0pt, numbersep=-6pt, morekeywords=[1]{try,spliced,tyfam,opfam,let,fn,val,def,casetype,objtype,metadata,of,*,string,datatype,new,toast,syntax,module,switch,where,using,import,for,ana,syn,opcon,tycon,metasignature,metamodule,metasig,metamod,static,at,by,tycase,mod,macro,match,pattern,in,patterns,expressions,implicit,forall,exptsm,pattsm},deletekeywords={double,structure,of},classoffset=0, xleftmargin=-10pt,
 aboveskip=4pt,belowskip=2pt,
@@
 without examining the underlying expansion, about types and binding.
-The system only needs to convey to clients via secondary notation the inferred \emph{segmentation} of each literal, which gives the locations of spliced subterms.
+The system only needs to convey to clients, via secondary notation, the inferred \emph{segmentation} of each literal, which gives the locations of spliced subterms.
 %Maintaining information about segmentation is necessary for the novel TLM hygiene mechanism, i.e.
@@
 %This calculus is the first detailed type-theoretic account of a hygienic macro system, of any design, for a language with these essential features of ML, and the first macro system with support for literal parsing that comes equipped with these essential reasoning principles.
-We are integrating TLMs into Reason, an alternative front-end for OCaml.% This calculus is the first such formalization of a typed macro system.
+We are integrating TLMs into Reason, an emerging alternative front-end for OCaml.% This calculus is the first such formalization of a typed macro system.
 
@@
 
-% \vspace{-4px}
 \newcommand{\introSec}{Introduction}
@@
 \label{sec:intro}
+\vspace{-3px}
 % The surface syntax of a programming language is its user interface. 
@@
 Similarly, Ur/Web's surface syntax extends Ur's syntax with expression and pattern literals for encodings of XML and HTML data \cite{conf/popl/Chlipala15,conf/pldi/Chlipala10}.
-Figure \ref{fig:urweb} shows two such HTML literals, one that ``splices in'' a string expression (Line 1) and the other an HTML expression (Line 2).
+Figure \ref{fig:urweb} shows two HTML literals, one that ``splices in'' a string expression (Line 1) and the other an HTML expression (Line 2).
 %The desugarings of these literal expressions, not shown, are substantially more verbose and less readable to programmers familiar with XHTML.
@@
 they choose strings instead of composite data structures largely for reasons of notational convenience.
-In particular, \citet{TSLs} sampled strings from open source projects and found that about 15\% of them were parseable according to some readily apparent type-specific grammar, e.g.
-for SQL queries, regexes, URLs and many other data structures.
-Literal forms, with support for splicing, would decrease the syntactic cost of composite encodings of such data, which are more amenable to programmatic manipulation and compositional reasoning than flat string encodings.
+In particular, \citet{TSLs} sampled strings from open source projects and found that about 15\% of them could be parsed by some readily apparent type-specific grammar.
+Literal notation, with support for splicing, would decrease the syntactic cost of composite encodings of such data, which are more amenable to programmatic manipulation and compositional reasoning than flat strings.
 %Literal forms would  preferable to string encodings because they support compositional construction and decomposition by pattern matching via splicing.
@@
 \begin{figure*}
-\vspace{-2px}
 \begin{lstlisting}
@@
 Instead, programmers can only reason transparently, i.e.
-by inspecting the underlying expansion or the extension responsible for producing the expansion.
+by inspecting the underlying expansion or the code responsible for producing the expansion.
 In some existing systems, even determining which extension is responsible for a literal form is difficult.
@@
 \caption{Syntax extension systems produce unreasonable program text.}
-\vspace{-11px}
+\vspace{-7px}
 \label{fig:K-dialect}
@@
 Consider, for example, the perspective of a programmer attempting to comprehend, i.e.
-reason about, the program text in Figure \ref{fig:K-dialect}, which is written in a dialect of OCaml's syntax called Reason \cite{reason} that has, hypothetically, been extended with some large number of new literal forms by a grammar extension system, like Camlp4 \cite{ocaml-manual} or Sugar* \cite{erdweg2011sugarj,erdweg2013framework}.
+reason about, the program text in Figure \ref{fig:K-dialect}, which is written in a dialect of OCaml's syntax called Reason \cite{reason-what}\footnote{This paper uses Reason 2.0 as 3.0 was released just before submission.} that has, hypothetically, been extended with some number of new literal forms by a grammar extension system like Camlp4 \cite{ocaml-manual} or Sugar*~\cite{erdweg2011sugarj,erdweg2013framework}.
 The literal form on Line 3 constructs an encoding of a query in the niche stack-based database query language K, using its intentionally terse notation \cite{Whitney:2001:LOR:376284.375783}.
-The problem is that a programmer unfamiliar with this notation cannot fall back on abstract reasoning principles to answer questions like:
-
-\begin{enumerate}[leftmargin=10pt]
+The problem is that a programmer unfamiliar with this notation cannot easily answer questions like the following:
+
+\begin{enumerate}[nolistsep,leftmargin=10pt]
 \item \textbf{(Responsibility)} Which syntax extension determined the expansion of the literal form on Line 3?
-Might different extensions generate different expansions for this form (i.e.
+Might activating a new extension generate a conflicting expansion for the same form (i.e.
 is there syntactic ambiguity)?
-\item \textbf{(Segmentation)} Are the characters \li{x} and \li{R} on Line 3 parsed as spliced expressions (meaning that they appear directly in the underlying expansion), or are they parsed in some other way peculiar to this literal form (e.g.
+\item \textbf{(Segmentation)} Are the characters \li{x}, \li{R} and \li{2} on Line 3 parsed as spliced expressions (meaning that they appear directly in the underlying expansion), or are they parsed in some other way peculiar to this literal form (e.g.
 as operators in the K query language)?
@@
 Forcing the programmer to reason transparently to answer basic questions like these defeats the ultimate purpose of syntactic sugar: decreasing cognitive cost \cite{Green89}.
-Analagous problems do not arise when programming without syntax extensions in languages like ML and Scala -- the binding discipline of the language ensures that programmers can reason lexically about binding, and types mediate abstraction over function and module implementations.
+Analagous problems do not arise when programming without syntax extensions in languages like ML and Scala --- programmers can reason lexically about where variables and other symbols are bound, and types mediate abstraction over function and module implementations \cite{B304}.
 % From this perspective, it is unsurprising that the use of Camlp4 has been deprecated in the OCaml system.
@@
 An approach that has made some headway against these problems is Omar et al.'s \emph{type-specific languages} (TSLs) \cite{TSLs}.
-In this approach, the language delegates control over the parsing and expansion of certain \emph{generalized literal forms} to a parse function associated with the type that the form is being checked against.
-Generalized literal forms syntactically subsume other literal forms because their bodies (i.e.
-the characters between the outer delimiters, here \li{[|} and \li{|]}) are left  unparsed by the fixed context-free syntax of the language (i.e.
+With this approach, the language delegates control over the parsing and expansion of certain \emph{generalized literal forms} to a parse function associated with the type that the form is being checked against.
+Generalized literal forms syntactically subsume other literal forms because their bodies---the characters between the outer delimiters, here \li{[|} and \li{|]}---are left  unparsed by the fixed context-free syntax of the language (i.e.
 they are initially parsed like raw string literals.) With TSLs, Line 3 of our example would require a type annotation:
@@
 
-Although TSLs address the problems of \textbf{Responsibility} and \textbf{Typing}, there is no guarantee that a valid segmentation exists, nor was the problem of reasoning about binding (i.e.
-capture and context dependence) adequately addressed in the prior work.
-Moreover, this prior work was introduced in the context of a very simple monomorphic, nominally typed language.
-There was no support for important features of modern languages in the ML tradition, such as structural types (e.g.
-arrow and tuple types), parameterized types, abstract types, modules, pattern notation and ML-style type inference.
-Finally, parse functions and the expansions they generate needed to be closed, i.e.
-they could not make use of any libraries at all.
-Taken together, we found that TSLs were not suitable as-is for integration into Reason, and other languages with a broadly similar type and binding discipline (e.g.
-Haskell, Scala, Rust and so on.)
-
-\paragraph{Contributions} The aim of this paper is to address all of these deficiencies, resulting in the first system for defining new literal notation that (1) provably maintains all five of the abstract reasoning principles just outlined; and (2) that is expressive enough for integration into a full-scale typed functional language (in this paper, the Reason front-end for OCaml.) We evaluate the first claim by developing a series of formal calculi equipped with proofs of these reasoning principles, and the second claim with several non-trivial case studies that appear throughout the paper.
-The empirical evaluation in the prior work on TSLs \cite{TSLs} supports the claim that a mechanism for defining new literal notation is useful, so our focus is on formal reasoning principles and integration with advanced language features in this paper.
+TSLs support reasoning abstractly about \textbf{Responsibility} and \textbf{Typing}, due to the type-driven dispatch mechanism, but the remaining reasoning principles were not fully and formally established in the prior work.
+Moreover, the prior work considered only a monomorphic, nominally-typed language with local type inference, leaving open a number of problems that came up as we considered integrating TSLs into Reason/OCaml.
+These included non-trivial problems related to support for structural types (e.g.
+arrow types), parameterized types, abstract types, modules, pattern notation and ML-style type inference.
+Another problem was that parse functions and expansions were completely restricted from accessing libraries.
+Taken together, we found that TSLs as described were not suitable for integration into Reason/OCaml.
+
+\paragraph{Contributions} The aim of this paper is to address all of these deficiencies, resulting in the first system for defining new literal notation that (1) provably maintains all five of the abstract reasoning principles just outlined; and (2) that is expressive enough for integration into Reason/OCaml and other full-scale statically typed languages.
+We evaluate the first claim by developing a series of formal calculi equipped with proofs of these reasoning principles, and the second claim with several non-trivial case studies that appear throughout the paper and involve all of the aforementioned language features.
+The empirical study in the prior work on TSLs \cite{TSLs} supports the claim that a macro-based mechanism for defining new literal notation is broadly useful, so this paper is targeted at language designers that seek a deeper understanding of the reasoning principles at play, particularly as they relate to features common to ML and other full-scale statically typed languages (e.g.
+Haskell, Scala, Rust, etc.)
 
@@
 \caption{TLMs make examples like the one from Figure \ref{fig:K-dialect} more reasonable.}
-\vspace{-10px}
+\vspace{-7px}
 \label{fig:K-tsm-example}
@@
 
-The most obvious point of departure between TSLs and \emph{typed literal macros} (TLMs), which are the topic of this paper, is that we separate the mechanism for defining new literal notation from the type definition mechanism of the language.
-TLMs are explicitly named and applied.
+The most obvious point of departure between TSLs and \emph{typed literal macros} (TLMs), which are the topic of this paper, is that we separate out the mechanism for defining new literal notation from the type definition mechanism of the language.
+TLMs are instead explicitly named and applied.
 For example, Figure \ref{fig:K-tsm-example} applies a TLM named \li{$kquery} to a generalized literal form delimited by \li{[|} and \li{|]} to express the K query example just discussed.
 The applied TLM parses the literal body and generates an expansion statically.
-
-During the typing phase, a validation step ensures that programmers can reason abstractly about the five issues just identified.
-We will characterize these reasoning principles precisely as we proceed, but for now, let us develop the basic intuition -- the programmer can reason abstractly as follows:
+During the typing phase, a validation step ensures that programmers can reason abstractly about types, binding and segmentation.
+Taken together, the programmer can reason as follows:
 \begin{enumerate}[nolistsep,leftmargin=10pt]
-\item \textbf{(Responsibility)} The applied TLM, \li{$kquery}, parses and determines the expansion of the literal body.
-The context-free syntax of the language is fixed, so the TLM provider is able to modularly establish that the syntax that this parser implements is unambiguous (even in situations where a spliced term itself applies a TLM, not shown here.)
+\item \textbf{(Responsibility)} The applied TLM, \li{$kquery}, is delegated responsibility for parsing the literal body.
+The context-free syntax of the language is fixed, so the TLM provider can modularly establish that the syntax that the TLM implements is unambiguous (even in situations where a spliced term itself applies a TLM, shown in a later example.)
 \item \textbf{(Segmentation)} The intermediate output that the TLM generates is structured so that the system can infer from it an accurate \emph{segmentation} of the literal body that distinguishes spliced terms from segments that are parsed in some other way by the TLM.
@@
 \caption{Case Study: HTML literals}
-\vspace{-5px}
+\vspace{-7px}
 \end{figure*}
 
+\vspace{-3px}
 \paragraph{Outline} 
@@
 % \item 
-We begin in Sec.~\ref{sec:setsms} with a more detailed case study: a TLM that implements the HTML literal notation shown in Figure \ref{fig:urweb}.
-We use this case study as a running example to informally describe how TLMs operate, from start to finish.
+We begin in Sec.~\ref{sec:setsms} by introducing TLMs with a detailed case study. 
 % \item 
-Sec.~\ref{sec:sptsms} then introduces {pattern TLMs}, i.e. TLMs that generate patterns, rather than expressions. %The hygiene condition for pattern TLMs is interesting.
+Sec.~\ref{sec:sptsms} then introduces {pattern TLMs}, i.e.
+TLMs that generate patterns, and describes the special hygiene conditions in pattern position.
+%The hygiene condition for pattern TLMs is interesting.
 % \item 
-Sec.~\ref{sec:ptsms} adds type functions and an ML-style module system and introduces \emph{parametric TLMs} (pTLMs), i.e.
-TLMs that can take type and module parameters.
-This section is organized around a second case study: finite map literals.
-Sec.~\ref{sec:static-eval} considers the topic of term evaluation during the  expansion phase, i.e.
-\emph{static evaluation}, in  more detail.
-Sec \ref{sec:static-eval} also gives examples of TLMs that are useful for defining other TLMs, e.g.
-TLMs that implement parser generators and quasiquotation, and uses these in a third case study of the mechanism: IPv4 address literals.
-Sec.~\ref{sec:setsms-formally} defines a type-theoretic calculus of simple expression and pattern TLMs and formally establishes the reasoning principles implied above. 
+Sec.~\ref{sec:ptsms} adds type functions and an ML-style module system and introduces \emph{parametric TLMs} (pTLMs), i.e. TLMs that can take type and module parameters, which solve several problems. 
+Sec.~\ref{sec:setsms-formally} presents a type-theoretic calculus of simple expression and pattern TLMs and formally establishes the reasoning principles implied above. 
 % \item 
@@
 % \item 
-Sec.~\ref{sec:discussion} concludes with a summary of the specific contributions of this paper and a discussion of the limitations of TLMs, together with suggestions for future work.
+Sec.~\ref{sec:discussion} concludes with a summary of the specific contributions of this paper and a discussion.
 
@@
 
-Due to the space limitations, certain technical details and extensions to the calculi described in this paper, as well as detailed proofs, are given in the supplemental material.
-Some notes on implementation and a fourth case study, involving an encoding of regular expressions as an abstract data type, are also both given in the supplement.
+Due to the space limitations, certain technical details and extensions to the calculi presented in this paper, as well as detailed proofs, are given in the supplemental material.
+Also available is an additional series of case studies: notation for an encoding of regular expressions as an abstract data type, implemented using TLMs for quasiquotation and grammar-based parser generators.
+%are also available in the supplement.
+Sec.~\ref{sec:static-eval} considers the topic of term evaluation during the  expansion phase, i.e.
+\emph{static evaluation}, in  more detail.
+Sec \ref{sec:static-eval} also gives examples of TLMs that are useful for defining other TLMs, e.g.
+TLMs that implement parser generators and quasiquotation.
 
@@
 
-\vspace{-4px}
+\vspace{-6px}
 \newcommand{\seTLMsSec}{Expression TLMs}
@@
 \label{sec:setsms}
-\vspace{-2px}
-
-Let us begin by detailing the TLM mechanism by way of a case study: HTML literal notation for expressions of the \li{html} type defined in Figure \ref{fig:html-type-def}.
-
-\vspace{-4px}
+% \vspace{-3px}
+
+We start in this section by describing the TLM mechanism in detail by way of a substantial case study: literal notation, like that built in to Ur/Web (see Figure \ref{fig:urweb}) and also, presently, Reason, for expressions of the type  \li{html} in Figure \ref{fig:html-type-def}.
+
+% \vspace{-8px}
 \subsection{TLM Definition and Application}
@@
  to distinguish them from variables.
-Every TLM definition also provides a \emph{type annotation}, here \li{at html}, and a \emph{parse function} between \li{by static \{} and \li{\}}.
+Every TLM definition also includes a \emph{type annotation}, here \li{at html}, and a \emph{parse function} between \li{by static \{} and \li{\}}.
 % The TLM definition follows the same scoping rules as type definitions.
@@
 
-Using the TLM defined in Figure \ref{fig:html-tlm-def}, we can express the example Ur/Web program from Figure \ref{fig:urweb} as shown in Figure \ref{fig:first-tsm-example}. On both lines, we apply 
+Using this TLM, we can express the Ur/Web example from Figure \ref{fig:urweb} as shown in Figure \ref{fig:first-tsm-example}. On both Lines 1 and 2, we apply 
 \li{$html} 
  to a \emph{generalized literal form} delimited by \li{[|} and \li{|]}.
-Generalized literal forms, which first arose in related work by \citet{TSLs} that we will detail in Sec.
-\ref{sec:existing-approaches}, subsume other literal forms because the context-free syntax of the language only specifies the outer delimiters.
+Generalized literal forms, which first arose in the prior work on TSLs \cite{TSLs}, syntactically subsume other literal forms because the context-free syntax of the language only specifies the outer delimiters.
 In this paper, we will use  \li{[|} and \li{|]}, but \citet{TSLs} formally specified several other choices, including layout-sensitive delimitation.
-\emph{Literal bodies}, i.e.
-the characters between these delimiters, are constrained only in that occurrences of \li{[|} must be balanced by \li{|]}.
+\emph{Literal bodies} are constrained only in that \li{[|} and \li{|]} must be balanced.
 %In particular, there is no pre-specified syntactic to splice out sub-terms.%The initial context-free parser therefore parses generalized literal forms in much the same way as parsers typically parse ``raw'' string literals (i.e.
@@
 
- The system delegates responsibility over the parsing and expansion of each literal body to the applied TLM's parse function during a semantic phase called \emph{typed expansion}, which generalizes the typing phase.
+ The system delegates responsibility over the parsing and expansion of each literal body to the applied TLM's parse function during a semantic phase called \emph{expansion}, which starts just before and continues into the typing phase.
+
  Because the parse function is applied during this phase, rather than at run-time, we call it a \emph{static function}.
-Static functions  cannot refer to the surrounding variable bindings (because those variables stand for run-time values.) Instead, there is a separate class of \emph{static bindings}.
-We will return to static bindings in Sec.
-\ref{sec:static-eval}.
-
- The parse function must have the following type: 
- \begin{lstlisting}[numbers=none]
-      body -> parse_result(proto_expr)
- \end{lstlisting}
- The input type, \li{body}, classifies encodings of literal {bodies}.
-Literal bodies are sequences of characters, so we define \li{body} as a synonym of \li{string}, as shown in \autoref{fig:indexrange-and-parseresult}.
-The return type is a sum type, defined by applying the type function \li{parse_result} defined in \autoref{fig:indexrange-and-parseresult}, that distinguishes between parse errors and successful parses.\footnote{We might alternatively have used an exception to signal a parse error, but sum types are formally simpler.} Let us consider these two possibilities in turn.
-
- If the parse function determines that the literal body is not well-formed according to whatever syntax definition that it implements, it must return \li{ParseError \{msg=}$e_\text{msg}$\li{, loc=}$e_\text{loc}$\li{\}}
+Static functions  cannot refer to the surrounding variable bindings because those variables stand for run-time values.
+Instead, there is are separate \emph{static bindings} marked by the \li{static} keyword that populate a \emph{static environment} that is discarded after expansion finishes.
+A more detailed account of static evaluation, both informal and informal, is available in the supplement.
+An alternative design that allows for the explicit lowering of standard-phase modules to the static phase has also been proposed for OCaml \cite{Ocaml/macros}.
+%\ificfp We are closely following this work in our implementation.
+\else \fi
+  % return to in Sec. \ref{sec:static-eval}.
+
+ The input type of the parse function, \li{body}, classifies encodings of literal {bodies}.
+Literal bodies are sequences of characters, so we define \li{body} as a synonym of \li{string} in \autoref{fig:indexrange-and-parseresult}.
+The return type is a sum type, defined by applying the parameterized type \li{parse_result} defined in \autoref{fig:indexrange-and-parseresult}, that distinguishes between parse errors and successful parses.
+Let us consider these two possibilities in turn.
+
+% \vspace{-3px}\paragraph{Parse Errors} 
+If the parse function determines that the literal body is not well-formed according to the syntax that it implements, it must return \li{ParseError \{msg=}$e_\text{msg}$\li{, loc=}$e_\text{loc}$\li{\}}
 where $e_\text{msg}$ is a custom error message and $e_\text{loc}$ is a value of type \li{segment}, defined in Figure \ref{fig:indexrange-and-parseresult}, that designates a segment of the literal body as the origin of the error \cite{DBLP:journals/jsc/DeursenKT93}.
-This information is for use in reporting the problem to the programmer.
-
+
+% \vspace{-3px}\paragraph{Success} 
 If instead parsing succeeds, the parse function returns \li{Success} ~$\ecand$, 
@@
 \vspace{-2px}
-When the parse function determines that some segment of the literal body is a spliced expression, according to whatever syntactic criteria it deems suitable, it can indirectly refer to it using the \li{SplicedE} variant of \li{proto_expr}, which takes a value of type \li{segment} that indicates the zero-indexed location of the spliced expression relative to the start of the provided literal body.
-The \li{SplicedE} variant also requires a value of type \li{proto_typ}, which indicates the type that the spliced expression is expected to have.
-Types can be spliced out by using the \li{SplicedT} variant of \li{proto_typ} analagously.
+When the parse function determines that some segment of the literal body is a spliced expression, according to whatever syntactic criteria it deems suitable, it can indirectly refer to it in the encoding it produces using the \li{SplicedE} constructor of \li{proto_expr}, which takes a value of type \li{segment} that indicates the zero-indexed location of the spliced expression relative to the start of the provided literal body.
+The \li{SplicedE} constructor also requires a value of type \li{proto_typ}, which indicates the type that the spliced expression is expected to have.
+Types can be spliced out by using the \li{SplicedT} constructor of \li{proto_typ} analagously.
 
@@
   | Success('a);
-type var_t = string;
-type proto_typ = TyVar(var_t)
-               | Arrow(proto_typ, proto_typ)
+type proto_typ = Arrow(proto_typ, proto_typ)
                | StringTy
@@
                | SplicedT(segment);
-type proto_expr = Var(var_t)
-                | Fn(var_t, proto_typ, proto_expr)
-                | Ap(proto_expr, proto_expr)
+type proto_expr = Tuple(list(proto_expr))
                 | /* ... */
@@
 \label{fig:candidate-exp-Reason}
-\vspace{-12px}
+\vspace{-14px}
 \end{figure}
 
-For example, consider again the TLM application on Line 1 of Figure \ref{fig:first-tsm-example}:
-\li{$html [|SURL<h1>Hello, {[EURLfirst_nameSURL]}</h1>EURL|]}.
-The parse function of the \li{$html} TLM (Figure \ref{fig:html-tlm-def}, Lines 2-4) first sends the literal body through a pre-existing HTML parser, \li{parse_html}\todo{which one?}.
-It then passes the result to a function \li{html_to_ast}, not shown, which produces a value of type \li{proto_expr}.
-When this function encounters a text node containing matched \li{\{[} and \li{]\}}, then that segment is inserted as a spliced expression of type \li{string}.
-Similarly, text nodes containing matched curly braces produce a spliced expression of type \li{html}.
-We assume that the pre-existing HTML parser annotates each element or node with its position in the originally parsed text (as most production-grade HTML/XML parsers are capable of doing.) If written in a textual syntax for proto-expressions where references to spliced expressions are written \li{spliced<startIdx; endIndex; ty>}, the output of this function for the example above is:
+For example, consider again the two TLM applications in Figure \ref{fig:first-tsm-example}.
+In each case, the parse function of the \li{$html} TLM (Figure \ref{fig:html-tlm-def}, Lines 2-4) first sends the literal body through an off-the-shelf HTML parser, \li{parse_html}.
+It then passes the result to a function \li{html_to_ast}, not shown, which produces the corresponding expression encoding of type \li{proto_expr}.
+When this function encounters an HTML text node containing matched \li{\{[} and \li{]\}}, then that segment is inserted as a spliced expression of type \li{string}, and similarly text nodes containing matched curly braces produce a spliced expression of type \li{html}.
+For instance,  the proto-expansion generated for first TLM application in \ref{fig:first-tsm-example}, pretty-printed, is:
 \begin{lstlisting}[numbers=none]
@@
 \end{lstlisting}
-The proto-expression \li{spliced<13; 22; string>} is a reference to the spliced string expression \li{first_name} by its location relative to the start of the literal body being expanded.
-It corresponds to the following encoding:
-\begin{lstlisting}[numbers=none]
-    SplicedE ({startIdx=13, endIdx=22}, StringTy)
-\end{lstlisting}
-
-Requiring that TLMs refer to spliced expressions indirectly in this manner ensures that a TLM cannot ``forge'' spliced terms (i.e.
-claim that some sub-term of the expansion should be given the privileges of a spliced term, discussed below, when it does not in fact appear in the literal body.)% Our semantics distinguishes spliced expressions when validating the proto-expansion, which is important for reasons of hygiene that we will detail shortly.
+Here, \li{spliced<13; 22; string>} is a reference to the spliced string expression \li{first_name} by its location relative to the start of the literal body being expanded (the off-the-shelf HTML parser provides the necessary baseline location information for use by \li{html_to_ast}).
+It corresponds to the encoding \li{SplicedE(\{startIdx=13, endIdx=22\}, StringTy)}.
+Requiring that TLMs refer to spliced expressions indirectly in this manner ensures that a TLM cannot ``forge'' spliced terms, i.e.
+claim that some sub-term of the expansion should be given the privileges of a spliced term, discussed in Sec.
+\ref{sec:uetsms-validation}, when it does not in fact appear in the literal body.% Our semantics distinguishes spliced expressions when validating the proto-expansion, which is important for reasons of hygiene that we will detail shortly.
 
@@
 \ref{sec:intro}, i.e.
-every literal body in a well-typed program has a well-defined segmentation.
-A program editor or pretty-printer can communicate this location information to the programmer, e.g.
+every literal body in a well-formed program has a well-defined segmentation.
+The TSL mechanism  did not maintain this reasoning principle \cite{TSLs}.
+A program editor or pretty-printer can communicate this segmentation information to the programmer, e.g.
 by coloring non-spliced segments green as is our convention in this document.
 In general, spliced expressions might themselves apply TLMs, in which case the convention is to use a distinct color for unspliced segments at each depth.
-For example, if strings were not primitive but rather defined as sequences of characters, we might define a TLM \li{$str} to recover string literal notation and write \li{$html [|SURL<body>\{EURLheading $str [|SSTRWorld!ESTR|]SURL\} ...</body>EURL|]}.
-
-A program editor or pretty-printer can also communicate the type of each spliced expression, as specified by the segmentation, to the programmer upon request (the editor packages for working with Reason defer to the Merlin tool when the programmer requests the type of an expression.) %When there is a type mismatch, the type annotation on the spliced segment also allows the error message to report the problem without revealing the full expansion.
+For example, consider a TLM \li{$smiles} for chemical structures \cite{anderson1987smiles} with support for splicing using curly braces:
+\vspace{-3px}
+\begin{lstlisting}[numbers=none]
+    $html [|SURLChemical structure of sucrose: {EURL
+        $smiles [|SCSS{ECSSm_glucoseSCSS}-O-{ECSSm_fructoseSCSS}ECSS|] 
+        |> SMILES.to_svg SURL}EURL|]
+\end{lstlisting}
+\vspace{-2px}
+%For example, if strings were not primitive but rather defined as sequences of characters, we might define a TLM \li{$str} to recover string literal notation and write \li{$html [|SURL<body>\{EURLheading $str [|SSTRWorld!ESTR|]SURL\} ...</body>EURL|]}.
+
+A program editor or pretty-printer can communicate the type of each spliced expression, also specified abstractly by the segmentation, upon request (for Reason, via Merlin \cite{Merlin}.) %When there is a type mismatch, the type annotation on the spliced segment also allows the error message to report the problem without revealing the full expansion.
 
@@
 \subsection{Proto-Expansion Validation}\label{sec:uetsms-validation}
+\vspace{-2px}
 Three important concerns described in Sec.
-\ref{sec:intro} remain: those related to reasoning abstractly about \textbf{Capture}, \textbf{Context Dependence} and \textbf{Typing}.
-Addressing these concerns is the purpose of the \emph{proto-expansion validation} process, which occurs once a proto-expansion has been generated by the parse function.
+\ref{sec:intro} remain: those related to reasoning abstractly about the \emph{hygiene properties}, i.e.
+\textbf{Capture} and \textbf{Context Dependence}, and \textbf{Typing}.
+Addressing these concerns is the purpose of the \emph{proto-expansion validation} process, which occurs during the typing phase.
 Proto-expansion validation results in the \emph{final expansion}, which is simply the proto-expansion with the references to spliced segments replaced with their own final expansions.
 
-\subsubsection{Capture}
-Proto-expansion validation ensures that spliced terms have access \emph{only} to the bindings at the application site -- spliced terms cannot capture the bindings in the proto-expansion.
+\subsubsection{Capture}\label{sec:capture}
+\vspace{-3px}
+
+Proto-expansion validation ensures that spliced terms have access \emph{only} to the bindings at the application site---spliced terms cannot capture bindings internal to the proto-expansion.
 For example, consider the following application site:
+\vspace{-3px}
 \begin{lstlisting}[numbers=none]
-  let tmp = /* ... application site temporary ... */;
+    let tmp = /* ... application site temp ... */;
     $html [|SURL<h1>{EURLf(tmp)SURL}</h1>EURL|];
 \end{lstlisting}
+\vspace{-2px}
 Now consider the scenario where the proto-expansion generated by \li{$html} has the following form:
+\vspace{-3px}
 \begin{lstlisting}[numbers=none]
-  let tmp = /* ... expansion-internal temporary ... */;
+    let tmp = /* ... expansion-internal temp ... */;
     H1Element(tmp, spliced<5; 10; html>);
 \end{lstlisting}
+\vspace{-2px}
 Na\"ively, the binding of the variable \li{tmp} in the proto-expansion could shadow the application-site binding of \li{tmp} in the final expansion. 
-
 To address this problem, splicing is guaranteed to be capture-avoiding.
 When generating the final expansion, the system discharges the requirement that capture not occur by implicitly alpha-varying the bindings in the proto-expansion as needed.
-For example, the final expansion of the example above might take the following form:
-\begin{lstlisting}[numbers=none]
-  let tmp = /* ... application site temporary ... */; 
-  let tmp' = /* ... expansion-internal temporary ... */;
-  H1Element(tmp', f(tmp));
-\end{lstlisting}
-Notice that the expansion-internal binding of \li{tmp} has been alpha-varied to \li{tmp'}.
+There is no need for TLM providers to explicitly deploy a mechanism that generates fresh variables (as in, e.g., prior \emph{reader macro} systems \cite{Flatt:2012:CLR:2063176.2063195}.
+% For example, the final expansion of the example above might take the following form:
+%\begin{lstlisting}[numbers=none]
+%  let tmp = /* ... application site temporary ... */; 
+%  let tmp' = /* ... expansion-internal temporary ... */;
+%  H1Element(tmp', f(tmp));
+%\end{lstlisting}
+%Notice that the expansion-internal binding of \li{tmp} has been alpha-varied to \li{tmp'}.
 The reference to \li{tmp} in the spliced expression then refers, as intended, to the application site binding.
 
-For TLM providers, the benefit of this mechanism is that they can name the variables used internally within expansions freely.
-There is no need to explicitly deploy a mechanism that generates ``fresh variables''.
-TLM clients can, in turn, can reason abstractly about capture, i.e.
-without examining the expansion that the spliced term appears within.
-
-This does prevent library providers from intentionally introducing bindings into spliced terms.
-For example, Haskell's literal notation for monadic commands, i.e.
-\li{do}-notation,  supports binding the result of executing a command to a variable that is then available in the subsequent commands in the command sequence.
-Using TLMs in Reason, this notation cannot be expressed -- values can be communicated from the expansion to a spliced expression only via function arguments.
-It would be possible to distinguish spliced identifiers, much as we do spliced terms, and explicitly make them available to spliced expressions by extending the \li{SplicedE} and \li{SplicedT} constructors to take finite sets of spliced identifiers (following the basic approach taken by \citet{DBLP:conf/esop/HermanW08}.) However, the opinion of the authors is that this would unjustifiably increase the reasoning burden on clients when they encounter an unfamiliar literal form: \emph{might this new literal form be doing obscure things with binding too?} Haskell-style infix notation for monadic commands, where bind is \li{e >>= f}, can be expressed using TLMs and splicing.
+% For TLM providers, the benefit of this mechanism is that they can name the variables used internally within expansions freely. TLM clients can, in turn, can reason abstractly about \textbf{Capture}.
+
+Capture avoidance does prevent library providers from intentionally introducing bindings into spliced terms.
+For example, Haskell's literal notation for monadic values, i.e.
+\li{do}-notation, cannot be expressed because it introduces a new binding syntax \cite{jones2003haskell}.
+
+In our initial designs, we were able to express this example because we distinguished spliced identifiers, much as we do spliced types and expressions.
+These could be explicitly bound inside spliced expressions by extending the \li{SplicedE} and \li{SplicedT} constructors to take finite sets of spliced identifiers.
+This is conceptually similar to the approach taken by \citet{DBLP:conf/esop/HermanW08} for giving (already parsed) sub-terms access to internal variables.
+However, we ultimately decided to remove this feature because it would disproportionately increase the reasoning burden on clients when they encounter an unfamiliar literal: \emph{might this obscure literal also be introducing bindings?} It is difficult for a program editor to display a set of ``hidden bindings'' to the programmer.
+Haskell-style infix notation for monadic values, where bind is \li{e >>= f}, can be expressed using TLMs without support for spliced identifiers.
+Reason 3.0 is adopting more concise function syntax, which will decrease the cost of this syntax.
 %We will show an alternative formulation of Haskell's syntax for monadic commands that uses Reason's anonymous function syntax to bind variables in Sec. \ref{sec:application-monadic-commands}.
 
-\vspace{-6px}
+\vspace{-7px}
 \subsubsection{Context Dependence}\label{sec:context-dependence}
+\vspace{-3px}
 %The prohibition on shadowing ensures only that variables that appear in spliced terms do not refer to bindings that appear in the surrounding expansion. 
@@
 In other words, expansions must be completely \emph{context independent} -- they can make no assumptions about the surrounding context whatsoever.
-A minimal example of a ``broken'' TLM that never generates context-independent proto-expansions is below:
+
+A minimal example of a ``broken'' TLM that does not generate context-independent proto-expansions is below:
+\vspace{-3px}
 \begin{lstlisting}[numbers=none]
-  syntax $broken at rx by static { fun(_) => Success (Var "SSTRxESTR") };
+    syntax $broken at t by static { 
+      fun(_) => Success (Var "SSTRxESTR") };
 \end{lstlisting}
-The proto-expansion that this TLM generates (for any given literal body) refers to a variable \li{x} that it does not itself bind.
+\vspace{-2px}
+The proto-expansion that this TLM generates (for any literal body) refers to a variable \li{x} that it does not itself bind.
 If proto-expansion validation permitted such a proto-expansion, it would be well-typed only under those application site typing contexts where \li{x} is bound.
-This ``hidden assumption'' makes reasoning about binding and renaming difficult, so this proto-expansion is deemed invalid (even when \li{$broken} is applied where \li{x} is coincidentally bound.)
+This ``hidden assumption'' makes reasoning about binding and renaming difficult.
 
 Of course, this prohibition does not extend into the spliced terms in a proto-expansion -- spliced terms appear at the application site, so they can justifiably refer to application site bindings.
-Indeed, we saw examples of spliced terms that referred to variables bound at the application site above.
-Because proto-expansions refer to spliced terms indirectly, enforcing context independence is straightforward -- we need only that the proto-expansion itself be closed.% In the next section, we will formalize this intuition.
+(like \li{first_name} in Fig.
+\ref{fig:first-tsm-example}.) Because proto-expansions refer to spliced terms indirectly, enforcing context independence is straightforward -- we need only that the proto-expansion itself be closed.% In the next section, we will formalize this intuition.
 % The TLM provider can only refer to them opaquely.
 
-One subtlety is that we assumed in the examples above that constructors like \li{H1Element} were available to the proto-expansion.
-In OCaml and other dialects of ML where datatype constructors are injected into the context as variables when the datatype is defined, this would violate context independence.
-As such, we will have to assume for now and in our formalism in Sec.
-\ref{sec:setsms-formally} that explicit injections into labeled sum types are available.
-Once we introduce module parameters in Sec.
-\ref{sec:ptsms}, we will be able to pass in a module exporting the datatype constructors as a parameter and partially apply that parameter to hide this detail from clients.
+Na\"ively, this restriction, also present in the prior work on TSLs \cite{TSLs}, is quite restrictive -- expansions cannot access any library functions.
+At best, they can require the client to ``pass in'' required library functions via splicing at every application.
+In Sec.
+\ref{sec:ptsms}, we will introduce module parameters and partial parameter application to neatly resolve this problem.
 
@@
 
-\subsubsection{Typing}
-Finally, proto-expansion validation maintains a reasonable \emph{typing discipline} by:
-\begin{enumerate}[noitemsep,nolistsep]
-\item checking that the expansion is of the type specified by the TLM's type annotation; %For example, the type annotation on \li{$rx} is \li{at rx}, so proto-expansion validation ensures that the final expansion is of type \li{rx}.
-\item checking that each spliced type is valid; 
-\item checking that the type annotation on each spliced expression is valid; and 
-\item checking each spliced expression against the specified type annotation.
-\end{enumerate}
+\vspace{-6px}
+\subsubsection{Typing}\label{sec:typing-e}
+\vspace{-3px}
+Finally, proto-expansion validation maintains a reasonable {typing discipline} by (1) checking that the expansion is of the type specified by the TLM's type annotation; (2) checking that each spliced type is valid; (3) checking that the type annotation on each spliced expression is valid; and (4) checking each spliced expression against the specified type annotation.
+Context independence implies that ML-style type inference can be performed using only the segmentation (because the remainder of an expansion cannot mention the very variables whose types are being inferred).
+In the prior work on TSLs, spliced terms did not have type annotations
 %The OCaml type system is not strong enough to allow us to express only contextually well-typed syntax trees. In The details are discussed as future work in Sec. \ref{sec:discussion}.
@@
 
-\vspace{-4px}
-\newcommand{\spTLMsSec}{Simple Pattern TLMs (spTLMs)}
+\vspace{-6px}
+\newcommand{\spTLMsSec}{Pattern TLMs}
 \section{\protect\spTLMsSec}
 \label{sec:sptsms}
+\vspace{-3px}
+
+Let us now briefly consider the topic of TLMs that generate \emph{patterns}, rather than expressions.
+Pattern literals are the dual to expression literals in that expression literals support construction whereas pattern literals support deconstruction \cite{moriconi2008inversion}.
+For example, we can pattern match on a value \li{x : html} by applying a pattern TLM \li{$html} as follows:
+\vspace{-3px}
+\begin{lstlisting}[numbers=none]
+    switch x {
+    | $html [|SURL<h1>{EURLcsSURL}</h1>EURL|] => /* cs:list(html) */
+    | _ -> None
+    };
+\end{lstlisting}
 \vspace{-2px}
-
-Let us now briefly consider the topic of TLMs that expand to \emph{patterns}, rather than expressions.
-For example, we can pattern match on a value \li{x : html} by applying a pattern TLM \li{$html} as follows:
-\begin{lstlisting}[numbers=none]
-  let children_of_h1 x => 
-    switch x {
-    | $html [|SURL<h1>{EURLcsSURL}</h1>EURL|] -> Some cs; /* cs : list(html) */
-    | _ -> None; };
-\end{lstlisting}
 Any list pattern, including one generated by another TLM application, can appear where \li{cs} appears in the example pattern above.
@@
 
-Pattern TLM definitions look much like expression TLM definitions:
-\begin{lstlisting}
-  syntax $html at html for patterns by static {
-    fun(b : body) : parse_result(proto_pat) => 
-      /* ... Ur/Web-style HTML pattern parser here ...  */
+Notice that we can use the same name, \li{$html}, for this pattern TLM as for the expression TLM defined in the previous section.
+It does not make sense to apply an expression TLM in pattern position  (many expression-level constructs, e.g.
+lambdas, do not correspond even syntactically to patterns), and \emph{vice versa}, so this is unambiguous.
+
+Pattern TLM definitions differ from expression TLM definitions in two ways: (1) a \emph{sort qualifier}, \li{for patterns}, distinguishes them from expression TLM definitions; (2) the return type of the parse function is \li{parse_result(proto_pat)}, rather than \li{parse_result(proto_expr)}.
+The type \li{proto_pat}, outlined below, classifies encodings of \emph{proto-patterns}.
+\vspace{-3px}
+\begin{lstlisting}[numbers=none]
+   type proto_pat = /* no variable pattern form! */
+                  | Wild
+                  | /* ... */
+                  | SplicedP(segment, proto_typ);
+\end{lstlisting}
+\vspace{-1px}
+
+% \subsection{Splicing}
+The constructor \li{SplicedP} operates much like \li{SplicedE} to allow a proto-pattern to refer indirectly to spliced patterns.% by their location within the literal body.
+
+%  For example, the proto-pattern generated for the example at the top of this section would be written concretely as follows:
+% \begin{lstlisting}[numbers=none]
+%   H1Element (_, spliced<6; 7; list(html)>)
+% \end{lstlisting}
+
+% \subsection{Proto-Pattern Validation}
+% Proto-pattern validation serves, like proto-expression validation, to maintain the ability to reason abstractly about binding and typing.
+
+To maintain the abstract binding discipline, variable patterns can appear only within spliced patterns.
+Enforcing this restriction is straightforward: we simply have not defined a variant of the \li{proto_pat} type that encodes variable patterns (wildcards are allowed.)  This restriction ensures that only variables visible to the client in a spliced pattern are bound in the corresponding branch expression.
+This is analagous to the capture avoidance principle for expression TLMs.
+
+Type annotations on references to spliced patterns could refer to type variables, so we also need to enforce context independence in the manner discussed in the previous section.
+%Pattern guards are part of pattern matching rules, not patterns themselves, in Reason/OCaml, but were they part of patterns, context independence would also need to be enforced for guard expressions.% In languages like OCaml, which support arbitrary boolean guard expressions, we would also need to enforce both context independence and capture avoidance as discussed in the previous section for spliced expressions that end up in a guard expression.
+In our formalism, we do not support guard expressions.
+
+% \subsubsection{Typing} 
+To maintain an abstract typing discipline, proto-pattern validation checks type annotations much as in Sec.
+\ref{sec:typing-e}.% (1) that the final expansion is a pattern that matches values of the type specified by the TLM's type annotation; (2) that each spliced pattern matches values of the type indicated in the segmentation; and (3) that each of these types are themselves well-formed types.
+
+\vspace{-7px}
+\newcommand{\pTLMsSec}{Parametric TLMs}
+\section{\protect\pTLMsSec}
+\label{sec:ptsms}
+\vspace{-2px}
+
+The simple TLMs in the previous sections operate only at one specified type, as did the TSLs in the prior work \cite{TSLs}.
+This is rather limiting.
+This section introduces \emph{parametric TLMs}, which can operate over a type- and module-parameterized family of types.
+They also neatly solve the problem discussed in Sec.
+\ref{sec:context-dependence} of giving expansions access to helper functions.
+%Moreover, as discussed at the end of Sec.
+\ref{sec:context-dependence}, the expansions that they generate have no access to libraries.
+This section introduces \emph{parametric TLMs} to neatly resolve both of these limitations (which also limited the prior work on TSL \cite{TSLs}.) Parametric TLMs can be defined over a type- and module-parameterized family of types, and the proto-expansions they generate can refer to supplied type and module parameters.
+Partial parameter application decreases the syntactic cost of this explicit parameter passing style.
+% \vspace{-4px}
+% \subsection{Parametric TLMs By Example}\label{sec:ptsms-by-example}
+% \vspace{-2px}
+
+Consider the following Reason/OCaml module type (a.k.a. signature), which specifies an abstract data type \cite{liskov1974programming,harper1997programming} of string-keyed polymorphic dictionaries:
+\vspace{-3px}
+\begin{lstlisting}[numbers=none]
+    module type DICT = {
+      type t('a); 
+      let empty : t('a); 
+      let extend : t('a) -> (string, 'a) -> t('a); 
+      /* ... */ 
     };
 \end{lstlisting}
-The \emph{sort qualifier}, \li{for patterns}, indicates that this is a pattern TLM definition.
-Expression TLM definitions can also include a sort qualifier, \li{for expressions}, but this is the default if omitted.
-Notice that we used the same name, \li{$html}, for this pattern TLM as for the expression TLM defined in the previous section.
-This is possible because patterns and expressions are distinct sorts of terms.
-It does not make sense to apply an expression TLM in pattern position, and \emph{vice versa}.\footnote{The fact that some patterns look like expressions in the textual syntax is immaterial to this fundamental semantic distinction.
-Many expression-level constructs, e.g.
-lambdas, do not correspond even syntactically to patterns.}
-The return type of the parse function is \li{parse_result(proto_pat)}, rather than \li{parse_result(proto_expr)}.
-The type \li{proto_pat}, outlined in Figure \ref{fig:CEPat}, classifies encodings of \emph{proto-patterns}.
-
-\begin{figure}[h]
-\vspace{-10px}
+\vspace{-2px}
+We can define a TLM that is parametric over implementations of this signature, \li{D : DICT}, and over choices of the codomain type, \li{'a} as follows:
+\vspace{-3px}
 \begin{lstlisting}[numbers=none]
-  type proto_pat = /* IMPORTANT: no variable pattern form! */
-                 | Wild
-                 | /* ... other standard pattern forms ... */
-                 | SplicedP(segment, proto_typ);
+    syntax $dict (D : DICT) (type 'a) at D.t('a) 
+    by static { fun(b) => /* ... */ };
 \end{lstlisting}
-\vspace{-10px}
-\caption[Abbreviated definition of \li{proto_pat} in Reason]{Abbreviated definition of \li{proto_pat}.}
-\label{fig:CEPat}
-\vspace{-10px}
-\end{figure}
-
-% \subsection{Splicing}
-The constructor \li{SplicedP} serves much like \li{SplicedE} to allow a proto-pattern to refer indirectly to spliced pattern segments by their location within the literal body.
-For example, the proto-pattern generated for the example at the top of this section would be written concretely as follows:
+\vspace{-2px}
+For example, given some module that implements this signature, \li{HashDict : DICT}, we can apply \li{$dict} as follows:
+\vspace{-3px}
 \begin{lstlisting}[numbers=none]
-  H1Element (_, spliced<6; 7; list(html)>)
+    $dict HashDict int [|"key1"SURL=>EURL10SURL; EURL"key2"SURL=>EURL15|]
 \end{lstlisting}
-
-% \subsection{Proto-Pattern Validation}
-% Proto-pattern validation serves, like proto-expression validation, to maintain the ability to reason abstractly about binding and typing.
-
-To maintain a reasonable abstract binding discipline, i.e.
-to allow clients to reason about variable binding without examining pattern TLM expansions directly, variable patterns can appear only within spliced patterns.
-Enforcing this restriction is straightforward: we simply have not defined a variant of the \li{proto_pat} type that encodes variable patterns (though wildcards are allowed.)  This prohibition ensures that no variables other than those visible to the client in a spliced pattern are captured by the corresponding branch of the match expression.
-
-Patterns have no way to refer to surrounding bindings (they only induce bindings in other expressions, e.g.
-in the corresponding branch of the \li{switch} expression.) However, type annotations on references to spliced patterns could refer to type variables, so we also need to enforce context independence in the manner discussed in the previous section.
-% In languages like OCaml, which support arbitrary boolean guard expressions, we would also need to enforce both context independence and capture avoidance as discussed in the previous section for spliced expressions that end up in a guard expression.
-In our formalism, we do not support guard expressions.
-
-% \subsubsection{Typing} 
-To maintain a reasonable abstract typing discipline, proto-pattern validation checks:
-\begin{enumerate}[noitemsep,nolistsep]
-  \item that the final expansion is a pattern that matches values of the type specified by the TLM's type annotation;
-  \item that each spliced pattern matches values of the type indicated in the segmentation; and 
-  \item that each of these types are themselves well-formed types.
-\end{enumerate}
-
-\vspace{-4px}
-\newcommand{\seTLMsFormallySec}{Simple TLMs, Formally}
-\section{\protect\seTLMsFormallySec}
-\label{sec:setsms-formally}
 \vspace{-2px}
-
-Before continuing on to consider more general parametric TLMs, let us develop a calculus of simple expression and pattern TLMs called $\miniVersePat$.
-This calculus consists of an \emph{unexpanded language}, or \emph{UL}, defined by typed expansion to an \emph{expanded language}, or \emph{XL}.
-Figs.
-\ref{fig:U-unexpanded-terms}  and \ref{fig:U-expanded-terms} summarize the syntax of the UL and  the XL, respectively.
-Programs are written as unexpanded expressions but evaluate as well-typed expanded expressions.
-We will start with a brief overview of our XL before turning in the remainder of the section on the UL and the typed expansion process.
-
-\vspace{-4px}
-\subsection{Expanded Language (XL)}\label{sec:s-XL}
+Notice that the segmentation immediately reveals which punctuation is particular to this TLM and where the spliced key and value expressions appear.
+Because the context-free syntax of unexpanded terms is never modified, it is possible to reason modularly about syntactic determinism, i.e.
+we can reason above that \li{=>} does not appear in the follow set of unexpanded expressions \cite{conf/pldi/SchwerdfegerW09}, so there can never be an ambiguity about where a key expression ends.
+
+The proto-expansion generated for the TLM application above, shown below, can refer to the parameters:
+\vspace{-3px}
+\begin{lstlisting}[numbers=none]
+    D.extend(D.extend D.empty (spliced<1;6;string>, 
+      spliced<11;12;'a>)) (spliced<15;20;string>, 
+      spliced<25;26;'a>)
+\end{lstlisting}
 \vspace{-2px}
-The {XL} of $\miniVersePat$ forms a standard pure functional language with partial function types, quantification over types, recursive types, labeled product types and labeled sum types. 
- The reader is directed to \emph{PFPL} \cite{pfple1} for a detailed introductory account of these standard constructs.
-We will only tersely summarize the statics and dynamics of the XL below because the particularities of the XL are not critical to the ideas we will introduce below.
-
-% \subsubsection{Statics of the Expanded Language}
-The \emph{statics of the XL} is defined by hypothetical judgements of the following form:
-\[\begin{array}{ll}
-% \textbf{Judgement Form} & \textbf{Description}\\
-\istypeU{\Delta}{\tau} & \text{$\tau$ is a well-formed type}\\
-%\isctxU{\Delta}{\Gamma} & \text{$\Gamma$ is a well-formed typing context assuming $\Delta$}\\
-\hastypeU{\Delta}{\Gamma}{e}{\tau} & \text{$e$ is assigned type $\tau$}\\
-\ruleType{\Delta}{\Gamma}{r}{\tau}{\tau'} & \text{$r$ takes values of type $\tau$ to values of type $\tau'$}\\
-\patType{\pctx}{p}{\tau} & \text{$p$ matches values of type $\tau$ and generates hypotheses $\pctx$} 
-\end{array}\]
-\emph{Type formation contexts}, $\Delta$, are finite sets of hypotheses of the form $\Dhyp{t}$.
-%Empty finite sets are written $\emptyset$, or omitted entirely within judgements, and non-empty finite sets are written as comma-separated finite sequences identified up to exchange and contraction.
-We write $\Delta, \Dhyp{t}$ when $\Dhyp{t} \notin \Delta$ for $\Delta$ extended with the hypothesis $\Dhyp{t}$.
-%Finite sets are written as finite sequences identified up to exchange.% We write $\Dcons{\Delta}{\Delta'}$ for the union of $\Delta$ and $\Delta'$.
-\emph{Typing contexts}, $\Gamma$, are finite functions that map each variable $x \in \domof{\Gamma}$, where $\domof{\Gamma}$ is a finite set of variables, to the hypothesis $\Ghyp{x}{\tau}$, for some $\tau$.
-The judgements above are inductively defined in the supplemental material and validate standard lemmas, also given in the supplement.
-%Empty typing contexts are written $\emptyset$, or omitted entirely within judgements, and non-empty typing contexts are written as finite sequences of hypotheses identified up to exchange and contraction.
-We write $\Gamma, \Ghyp{x}{\tau}$, when $x \notin \domof{\Gamma}$, for the extension of $\Gamma$ with a mapping from $x$ to $\Ghyp{x}{\tau}$, and $\Gcons{\Gamma}{\Gamma'}$ when $\domof{\Gamma} \cap \domof{\Gamma'} = \emptyset$ for the typing context mapping each $x \in \domof{\Gamma} \cup \domof{\Gamma'}$ to $x : \tau$ if $x : \tau \in \Gamma$ or $x : \tau \in \Gamma'$.
-% We write $\isctxU{\Delta}{\Gamma}$ if every type in $\Gamma$ is well-formed relative to $\Delta$.
-% \begin{definition}[Typing Context Formation] \label{def:isctxU}
-% $\isctxU{\Delta}{\Gamma}$ iff for each hypothesis $x : \tau \in \Gamma$, we have $\istypeU{\Delta}{\tau}$.
-% \end{definition}
-
-% \begin{subequations}\label{rules:istypeU}
-% \begin{equation*}\label{rule:istypeU-var}
-% \inferrule{ }{\istypeU{\Delta, \Dhyp{t}}{t}}
-% \end{equation*}
-% \begin{equation*}\label{rule:istypeU-parr}
-% \inferrule{
-%   \istypeU{\Delta}{\tau_1}\\
-%   \istypeU{\Delta}{\tau_2}
-% }{\istypeU{\Delta}{\aparr{\tau_1}{\tau_2}}}
-% \end{equation*}
-% \begin{equation*}\label{rule:istypeU-all}
-%   \inferrule{
-%     \istypeU{\Delta, \Dhyp{t}}{\tau}
-%   }{
-%     \istypeU{\Delta}{\aall{t}{\tau}}
-%   }
-% \end{equation*}
-% \begin{equation*}\label{rule:istypeU-rec}
-%   \inferrule{
-%     \istypeU{\Delta, \Dhyp{t}}{\tau}
-%   }{
-%     \istypeU{\Delta}{\arec{t}{\tau}}
-%   }
-% \end{equation*}
-% \begin{equation*}\label{rule:istypeU-prod}
-%   \inferrule{
-%     \{\istypeU{\Delta}{\tau_i}\}_{i \in \labelset}
-%   }{
-%     \istypeU{\Delta}{\aprod{\labelset}{\mapschema{\tau}{i}{\labelset}}}
-%   }
-% \end{equation*}
-% \begin{equation*}\label{rule:istypeU-sum}
-%   \inferrule{
-%     \{\istypeU{\Delta}{\tau_i}\}_{i \in \labelset}
-%   }{
-%     \istypeU{\Delta}{\asum{\labelset}{\mapschema{\tau}{i}{\labelset}}}
-%   }
-% \end{equation*}
-% \end{subequations}
-% Premises of the form $\{{J}_i\}_{i \in \labelset}$ mean that for each $i \in \labelset$, the judgement ${J}_i$ must hold.
+Validation checks that the proto-expansion is truly parametric, i.e.
+it must be valid for all modules \li{D : DICT} and types \li{'a}.
+It is only after validation that we substitute the actual parameters, here \li{HashDict} for \li{D} and \li{int} for \li{'a}, into the final expansion.
+Only the type annotations on references to spliced terms are subject to early parameter substitution (because they classify application site terms.)
+
+If we will use the \li{HashDict} implementation ubiquitously, we can abbreviate the partial application of \li{$dict} to \li{HashDict}, resulting in a TLM that is parametric over only the type \li{'a}:
+\vspace{-3px}
+\begin{lstlisting}[numbers=none]
+    syntax $dict' = $dict HashDict; 
+\end{lstlisting}
+\vspace{-2px}
+TLM abbreviations can themselves be parameterized to support partial application of parameters other than the last.
+
+In Sec.
+\ref{sec:context-dependence} we discussed the problem of the strict context independence discipline being too restrictive, in that it would seem to restrict expansions from referring to useful helpers bound at the TLM definition site.
+Module parameters address this problem -- the helper values, types and modules can be packaged into a module, passed in and partially applied to hide this detail from clients.
+Because this will be common in practice, we provide the following shorthand:
+\begin{lstlisting}[numbers=none]
+    syntax $a at t using X~${}_1$~=M~${}_1$~,...,X~${}_n$~=M~${}_n$~ by ...
+\end{lstlisting} 
+% This is the same as a definition that specifies $n$ type, expression or module parameters and immediately applies the given arguments, binding the result to the same TSM name.
+
+This explicit parameter passing discipline is reminiscent of work on explicit capture for distributed functions \cite{DBLP:conf/ecoop/MillerHO14}.
+By not implicitly giving expansions access to all definition-site bindings, we need not examine parse functions to reason about, e.g., renaming.
+Consequently, encodings of proto-terms can be values of standard datatypes (e.g.
+\li{proto_expr}) with variables represented simply as, e.g., strings, and quasiquotation notation can be expressing using TLMs (see supplement).
+A central design goal of Reason is to leave the OCaml semantics unchanged, so building quasiquotation in primitively, to integrate the free variables in term encodings into the overall binding discipline \cite{Bawd99a}, was in any case infeasible.
+
+%
 
@@
 \small
-$\arraycolsep=3pt\begin{array}{llcl}
+$\arraycolsep=2pt\begin{array}{llcl}
 \mathsf{UTyp} & \utau & ::= & 
@@
 \forallt{\ut}{\utau} ~\vert~
-\rect{\ut}{\utau} ~\vert~
+\rect{\ut}{\utau}\\
+& & \vert & 
 \prodt{\mapschema{\utau}{i}{\labelset}} ~\vert~
@@
 \App{\ue}{\utau} ~\vert~
-\fold{\ue} ~\vert~
+\fold{\ue} \\
 % \unfold{\ue} ~\vert~
-\tpl{\mapschema{\ue}{i}{\labelset}} ~\vert~
+&& \vert & \tpl{\mapschema{\ue}{i}{\labelset}} ~\vert~
 \inj{\ell}{\ue} ~\vert~
@@
  \\
-& & \vert & \uesyntaxq{\tsmv}{\utau}{e}{\ue} ~\vert~ \utsmap{\tsmv}{b} \\   
+& & \vert & \uesyntaxq{\tsmv}{\utau}{e}{\ue} \\   
 & & \vert & \usyntaxup{\tsmv}{\utau}{e}{\ue} \\
+&&\vert&  \utsmap{\tsmv}{b} \\
 \mathsf{URule} & \urv & ::= & 
@@
 }
-\vspace{-6px}
+\vspace{-10px}
 \label{fig:U-unexpanded-terms}
 \end{figure}
-\begin{figure}
-%\hspace{-5px}
-\begin{minipage}{\textwidth}
-\small
-$\arraycolsep=3pt\begin{array}{llcl}
-\mathsf{Typ} & \tau & ::= & t ~\vert~ \aparr{\tau}{\tau} ~\vert~ \aall{t}{\tau} ~\vert~ \arec{t}{\tau} ~\vert~ \aprod{\labelset}{\mapschema{\tau}{i}{\labelset}} ~\vert~ \asum{\labelset}{\mapschema{\tau}{i}{\labelset}}\\
-\mathsf{Exp} & e & ::= & x ~\vert~ \aelam{\tau}{x}{e} ~\vert~ \aeap{e}{e} ~\vert~ \aetlam{t}{e} ~\vert~ \aetap{e}{\tau} ~\vert~ \aefold{e} ~\vert~ \aetpl{\labelset}{\mapschema{e}{i}{\labelset}}  \\
-& & \vert & \aein{\ell}{e} ~\vert~ \aematchwith{n}{e}{\seqschemaX{r}}\\
-\mathsf{Rule} & r & ::= & \aematchrule{p}{e}\\
-\mathsf{Pat} & p & ::= & x  ~\vert~ \aewildp ~\vert~ \aefoldp{p} ~\vert~ \aetplp{\labelset}{\mapschema{p}{i}{\labelset}} ~\vert~ \aeinjp{\ell}{p}
-\end{array}$
-\end{minipage}
+\begin{comment}
+\vspace{-6px}
+\newcommand{\staticEvalSec}{Static Evaluation}
+\section{\protect\staticEvalSec}
+\label{sec:static-eval}
+\vspace{-2px}
+
+% It is important to note that module parameters are accessible by expansions, but not by parse functions directly, because the applied module parameters will not have been dynamically instantiated when typed expansion occurs.
+We will discuss a distinct mechanism for providing helper functions to parse functions in Sec.
+\ref{sec:static-eval}.
+
+There is one more major impracticality from the prior work on TSLs \cite{TSLs} that we will now briefly address: in the prior work, parse functions had no access to any libraries, nor any ability to themselves apply TLMs.
+To address this problem, we introduce a \emph{static environment}.
+
+\vspace{-4px}
+\subsection{The Static Environment}
+\vspace{-2px}
+Figure \ref{fig:static-module-example} shows an example of a module, \li{ParserCombos}, that defines a number of parser combinators following \citet{Hutton1992d}.
+The \li{static} qualifier indicates that this module is bound for use only within similarly qualified values, including in particular parse functions of subsequent TLM definitions.
+\begin{figure}[h]
 \vspace{-10px}
-\caption[Syntax of the XL of $\miniVersePat$]{Syntax of the $\miniVersePat$ expanded language (XL).
-XL terms are \emph{abstract binding trees} (ABTs) identified up to alpha-equivalence, so we follow the syntactic conventions of \citet{pfple1}.
-Metavariable $x$ ranges over variables and $t$ over type variables.
-%When using stylized forms, the label set is omitted when it can be inferred, e.g.
-the labeled product type $\prodt{\finmap{\mapitem{\ell_1}{e_1}, \mapitem{\ell_2}{e_2}}}$ leaves the label set $\{\ell_1, \ell_2\}$ implicit.
-% When we use the stylized forms, we assume that the reader can infer suppressed indices and arguments from the surrounding context.
-}
+\begin{lstlisting}[numbers=none]
+  static module ParserCombos = { 
+    type p('c, 't) = list('c) -> list('t, list('c));
+    let alt : p('c, 't) -> p('c, 't) -> p('c, 't);
+    /* ... */
+  };
+  syntax $a at t by static { fun(b) => 
+    ParserCombos.alt /* ... OK */ };
+  static let y = ParserCombos.alt /* ... OK */;
+  let z = /* ... ParserCombos nor y bound here */;
+\end{lstlisting}
+\vspace{-10px}
+\caption{Binding static modules and values for use within parse functions.}
+\label{fig:static-module-example}
+\vspace{-10px}
+\end{figure}
+
+The values that arise when the static phase runs do not persist from ``compile-time'' to ``run-time'', so we do not need a full staged computation system, e.g.
+as described by \citet{Taha99multi-stageprogramming:}.
+Instead, a sequence of static bindings operates like a read-evaluate-print loop (REPL) scoped according to the program structure, in that each static expression is evaluated immediately and the evaluated values are tracked by a \emph{static environment}, which is discarded after expansion finishes.
+We do not restrict the language features available to the static phase, though this is a worth considering in languages where that is more straightforward (e.g.
+to ensure deterministic builds.)
+
+% A language designer might choose to restrict the external effects available to static terms in some way, e.g.
+to ensure deterministic builds.
+It might also be helpful to restrict mutable state shared between TLMs to prevent undesirable TLM application order dependencies.
+On the other hand, these features, if used for the purposes of caching, might speed up typed expansion.
+These are orthogonal design decisions.
+
+\vspace{-6px}
+\subsection{Applying TLMs Within TLM Definitions}\label{sec:tsms-for-tsms}
+TLMs and TLM abbreviations can also be qualified with the \li{static} keyword, which marks them for use within subsequent static expressions and patterns.
+Let us consider some examples of relevance to TLM providers.
+
+\vspace{-6px}
+\subsubsection{Quasiquotation}\label{sec:quasiquotation}
+TLMs must construct values of type \li{proto_expr} or \li{proto_pat}.
+Constructing values of these types explicitly can have high syntactic cost.
+To decrease this cost, we can define TLMs that provide support for \emph{quasiquotation syntax} similar to that built in to languages like Lisp \cite{Bawd99a} and Scala \cite{shabalin2013quasiquotes}.
+The following TLM defines quasi-quotation for encodings of proto-expressions:
+\begin{lstlisting}[numbers=none]
+  static syntax $proto_expr at proto_expr by static { /* ... */ };
+\end{lstlisting}
+For example, \li{$proto_expr `SQTg x ^EQTx`} might have expansion \li{App(App(Var "g", Var "x"), x)}.
+Notice that prefixing a variable (or parenthesized expression) with \li{^} serves to splice in its value, which here must be of type \li{proto_expr} (though in other syntactic positions, it might be \li{proto_typ}.) This is also known as \emph{anti-quotation}.
+
+% A similar approach can be taken for working with encodings of terms of other languages (e.g. when writing an interpretter or compiler in Reason.)
+
+\vspace{-6px}
+\subsubsection{Parser Generators}
+Abstractly, a grammar-based parser generator is a module matching the signature \li{PARSEGEN} defined below:
+\begin{lstlisting}
+  module type PARSEGEN = { 
+    type grammar('a);
+    /* ... operations on grammars ... */
+    val generate : grammar('a) -> (body -> parse_result('a));
+  };
+\end{lstlisting}
+
+Rather than constructing a grammar equipped with semantic actions using the associated operations (whose specifications are elided in \li{PARSEGEN}), we wish to use a syntax for context-free grammars that follows standard conventions.
+We can do so by defining a static parametric TLM:
+\begin{lstlisting}[numbers=none]
+  static syntax $grammar(P : PARSEGEN)(type 'a) at P.grammar('a) /*...*/
+\end{lstlisting}
+To support splicing, we need non-terminals that recognize unexpanded terms and produce the corresponding splice references, rather than the AST itself.
+This requires that the parser generator keep track of location information (as most production-grade parser generators already do for error reporting.) A more detailed example of this mechanism being used to define a TLM for a modular encoding of regular expressions is given in the supplement.% For spliced expressions, this non-terminal would need to be a family of non-terminals indexed by a value of type \li{proto_typ}.
+
+A grammar containing such non-terminals can serve as a \emph{summary specification} -- a human can simply be take this grammar as a specification of what the segmentation will be for every recognized string, rather than relying on an editor to communicate this information.
+The associated semantic actions can be held abstract as long as the system performs a simple check to ensure that the proto-expansion does mention each spliced expression from the corresponding production.
+
+% \begin{figure}[h!]
+% \vspace{-5px}
+% \begin{lstlisting}[deletekeywords={as}]
+% syntax $rx(R : RX) at R.t by static 
+%   P.generate ($grammar P proto_expr {|SHTML #\label{line:rx_parse_fn_start}#
+%     start <- ""
+%       EHTMLfn () => $proto_expr `SCSSR.EmptyECSS`SHTML
+%     start <- "(" start ")"
+%       EHTMLfn e => eSHTML
+%     token str_tok #\label{line:str_tok_start}#
+%       EHTMLRU.parse "SSTR[^(@$]+ESTR" /* cannot use $rx within its own def */SHTML #\label{line:str_tok_end}#
+%     start <- str_tok
+%       EHTMLfn s => $proto_expr `SCSSR.Str %(ECSSstr_to_proto_lit sSCSS)ECSS`SHTML
+%     start <- start start
+%       EHTMLfn e1 e2 => $proto_expr `SCSSR.Seq (%ECSSe1SCSS, %ECSSe2SCSS)ECSS`SHTML
+%     start <- start "|" start 
+%       EHTMLfn e1 e2 => $proto_expr `SCSSR.Or (%ECSSe1SCSS, %ECSSe2SCSS)ECSS`SHTML
+%     start <- start "*"
+%       EHTMLfn e => $proto_expr `SCSSR.Star %ECSSe`SHTML
+
+%     using EHTMLspliced_uexp ($proto_typ `SCSSR.tECSS`) SHTML as spliced_rx #\label{line:splicede_using}#
+%     start <- "${" spliced_rx "}" #\label{line:splicing-start}#
+%       EHTMLfn e => eSHTML
+
+%     using EHTMLspliced_uexp ($proto_typ `SCSSstringECSS`) SHTML as spliced_str
+%     start <- "@{" spliced_str "}"
+%       EHTMLfn e => $proto_expr `SCSSR.Str %(ECSSeSCSS)ECSS`SHTML #\label{line:splicing-end}#
+%   EHTML|})
+% end #\label{line:rx_parse_fn_end}#
+% \end{lstlisting}
+% \vspace{-12px}
+% \caption{A grammar-based definition of \texttt{\$rx}.}
+% \vspace{-15px}
+% \label{fig:rx-grammar-based}
+% \end{figure}
+
+\vspace{-6px}
+\subsection{Static Evaluation, Formally}
+It is not difficult to extend $\miniVerseParam$ to account for static evaluation.
+Static environments, $\Sigma$, take the form $\staticenv{\omega}{\uOmega}{\uPsi}{\uPhi}$, where $\omega$ is a substitution.
+Each binding form is annotated with a \emph{phase}, $\phi$, either $\staticphase$ or $\standardphase$.
+The rules for binding forms annotated with $\standardphase$ are essentially unchanged, differing only in that $\Sigma$ passes through opaquely.
+The rules for binding forms annotated with $\staticphase$ are based on the corresponding $\standardphase$ phase rules, differing only in that 1) they operate on $\Sigma$ and 2) evaluation occurs immediately.
+Finally, the forms for TLM definition are modified so that the parse function is now an unexpanded, rather than an expanded expression.
+The substitution $\omega$ is applied to the parse function after it is expanded.
+The full details are defined as a small patch of $\miniVerseParam$ called $\miniVersePH$ in the supplement.
+
+\vspace{-6px}
+\subsection{Library Management}
+In the examples above, and in our formal treatment, we explicitly qualified various definitions with the \li{static} keyword to make them available within static values.
+In practice, we would like to be able to use libraries within both static values and standard values as needed without duplicating code.
+This can be accomplished either by the package manager (e.g.
+SML/NJ's CM \cite{blume:smlnj-cm}, extended with phase annotations) or by allowing one to explicitly lower an instance of a module define in the standard phase for use also in  the static phase, as in a recent proposal for modular staging macros in OCaml \cite{Ocaml/macros}.
+
+TLMs definitions can be exported from the top level of packages, but they cannot be exported from within ML-style modules because that would require that they also appear in signatures, and that, in turn, would complicate reasoning about signature equivalence, since TLM definitions contain arbitrary parse functions.
+%It would also bring in confusion about whether the generated expansions can use private knowledge about type identity. 
+That said, it should be possible to export TLM \emph{abbreviations} from modules, since they refer to TLM definitions only through symbolic names.
+We have not yet formalized this intuition, but the work of \citet{culpepper2005syntactic,culpepper2007advanced} considered a closely related question: how should Typed Scheme's macros interact with its unit (i.e.
+package) system.
+
+% For example, a language-external library manager for Reason similar to SML/NJ's CM \cite{blume:smlnj-cm} could support a \li{static} qualifier on imported libraries, which would place the definitions exported by the imported library into the static phase of the library being defined.
+In particular, a library definition in such a compilation manager might look like this:
+% \begin{lstlisting}[numbers=none,morekeywords={Library,is}]
+% Library 
+%   /* ... exported module, signature and TLM names ... */
+% is 
+%   /* ... files defining those exports ... */
+
+%   /* imports: */
+%   static parsegen.cm 
+% \end{lstlisting}
+
+% A similar approach could be taken for languages the incorporate library management directly into the syntax of programs, e.g. Scala \cite{odersky2008programming}:
+% \begin{lstlisting}[numbers=none]
+% static import edu.cmu.comar.parsegen
+% \end{lstlisting}
+
+% \begin{equation}\label{rule:mExpandsPH-syntaxpe-standard}
+% \inferrule{
+%   \tsmtyExpands{\uOmega}{\urho}{\rho}\\
+%   \Sigma = \staticenv{\omega}{\uOmega_S}{\uPsi_S}{\uPhi_S}\\\\
+%   \expandsP{\uOmega_S}{\uPsi_S}{\uPhi_S}{\ueparse}{\eparse}{\aparr{\tBody}{\tParseResultPCEExp}}\\\\
+%   \evalU{[\omega]\eparse}{\eparse'}\\
+%   \mExpandsPH{\uOmega}{\uAS{\uA \uplus \mapitem{\tsmv}{\adefref{a}}}{\Psi, \petsmdefn{a}{\rho}{\eparse'}}}{\uPhi}{\uM}{M}{\sigma}{\Sigma}
+% }{
+%   \mExpandsPH{\uOmega}{\uAS{\uA}{\Psi}}{\uPhi}{\defpetsmH{\standardphase}{\tsmv}{\urho}{\ueparse}{\uM}}{M}{\sigma}{\Sigma}
+% }
+% \end{equation}
+% \begin{equation}\label{rule:mExpandsPH-syntaxpe-static}
+% \inferrule{
+%   \tsmtyExpands{\uOmega}{\urho}{\rho}\\
+%   \Sigma = \staticenv{\omega}{\uOmega_S}{\uPsi_S}{\uPhi_S}\\
+%   \uPsi_S = \uAS{\uA_S}{\Psi_S}\\\\
+%   \expandsP{\uOmega_S}{\uPsi_S}{\uPhi_S}{\ueparse}{\eparse}{\aparr{\tBody}{\tParseResultPCEExp}}\\\\
+%   \evalU{[\omega]\eparse}{\eparse'}\\
+%   \mExpandsPH{\uOmega}{\uPsi}{\uPhi}{\uM}{M}{\sigma}{\staticenv{\omega}{\uOmega_S}{\uAS{\uA_S \uplus \mapitem{\tsmv}{\adefref{a}}}{\Psi_S, \petsmdefn{a}{\rho}{\eparse'}}}{\uPhi_S}}
+% }{
+%   \mExpandsPH{\uOmega}{\uPsi}{\uPhi}{\defpetsmH{\staticphase}{\tsmv}{\urho}{\ueparse}{\uM}}{M}{\sigma}{\Sigma}
+% }
+% \end{equation}
+\end{comment}
+
 \vspace{-4px}
-\label{fig:U-expanded-terms}
-\end{figure}
+\newcommand{\seTLMsFormallySec}{Simple TLMs, Formally}
+\section{\protect\seTLMsFormallySec}
+\label{sec:setsms-formally}
+\vspace{-2px}
+
+This section will present a calculus of simple expression and pattern TLMs called $\miniVersePat$.
+By the end of this section, we will have a theorem that encodes the five reasoning principles that were outlined informally in the previous sections.
+
+Because our focus is on these reasoning principles, and for reasons of space, we leave our full calculus of parametric TLMs to the supplement.
+The full calculus extends the simple calculus of this section with an ML module calculus based closely on the system defined by \citet{pfple1}, which in turn is based on early work by \citet{MacQueen:1984:MSM:800055.802036,DBLP:conf/popl/MacQueen86}, subsequent work on the phase splitting interpretation of modules \cite{harper1989higher} and on using dependent singleton kinds to track type identity \cite{stone2006extensional,DBLP:conf/lfmtp/Crary09}, and finally on formal developments by \citet{dreyer2005understanding} and \citet{conf/popl/LeeCH07}.
+These additional mechanisms are necessary only to formalize the advanced features of Sec.
+\ref{sec:ptsms}.
+Proofs are in the supplement for both the simple and full calculus.
+
+Both the simple and full calculus consist of an \emph{unexpanded language}, or \emph{UL}, defined by typed expansion to an \emph{expanded language}, or \emph{XL}.
+Figs.
+\ref{fig:U-unexpanded-terms}  and \ref{fig:U-expanded-terms} summarize the syntax of the UL and  the XL, respectively.
+% Programs are written as unexpanded expressions but evaluate as well-typed expanded expressions.
+We will start with a brief overview of our XL before turning in the remainder of the section on the UL.
+
+\vspace{-6px}
+\subsection{Expanded Language (XL)}\label{sec:s-XL}
+\vspace{-2px}
+The {XL} of $\miniVersePat$ forms a standard pure functional language with partial function types, quantification over types, recursive types, labeled product types and labeled sum types and support for pattern matching.
+ The reader is directed to \emph{PFPL} \cite{pfple1} for a detailed introductory account of these constructs.
+We will only tersely summarize the statics and dynamics of the XL below because the particularities are not critical.
+
+% \subsubsection{Statics of the Expanded Language}
+The \emph{statics of the XL} is organized around the type formation judgement, $\istypeU{\Delta}{\tau}$, the expression typing judgement, $\hastypeU{\Delta}{\Gamma}{e}{\tau}$, and the pattern typing judgement, $\patType{\pctx}{p}{\tau}$.
+In the latter, $\pctx$ is a typing context that tracks the typing hypotheses generated by $p$.
+These judgements are inductively defined in the supplemental material along with necessary auxiliary structures and  standard lemmas.
+%\emph{Type formation contexts}, $\Delta$, and \emph{typing contexts}, $\Gamma$ are defined in the supplement in the standard way \cite{pfpl},
+% \[\begin{array}{ll}
+% % \textbf{Judgement Form} & \textbf{Description}\\
+% \istypeU{\Delta}{\tau} & \text{$\tau$ is a well-formed type}\\
+% %\isctxU{\Delta}{\Gamma} & \text{$\Gamma$ is a well-formed typing context assuming $\Delta$}\\
+% \hastypeU{\Delta}{\Gamma}{e}{\tau} & \text{$e$ is assigned type $\tau$}\\
+% \ruleType{\Delta}{\Gamma}{r}{\tau}{\tau'} & \text{$r$ takes values of type $\tau$ to $\tau'$}\\
+% \patType{\pctx}{p}{\tau} & \text{$p$ matches values of type $\tau$}\\
+% & \text{and generates hypotheses $\pctx$} 
+% \end{array}\]
+% \emph{Type formation contexts}, $\Delta$, are finite sets of hypotheses of the form $\Dhyp{t}$.
+%Empty finite sets are written $\emptyset$, or omitted entirely within judgements, and non-empty finite sets are written as comma-separated finite sequences identified up to exchange and contraction.
+We write $\Delta, \Dhyp{t}$ when $\Dhyp{t} \notin \Delta$ for $\Delta$ extended with the hypothesis $\Dhyp{t}$.
+%Finite sets are written as finite sequences identified up to exchange.% We write $\Dcons{\Delta}{\Delta'}$ for the union of $\Delta$ and $\Delta'$.
+% \emph{Typing contexts}, $\Gamma$, are finite functions that map each variable $x \in \domof{\Gamma}$, where $\domof{\Gamma}$ is a finite set of variables, to the hypothesis $\Ghyp{x}{\tau}$, for some $\tau$.
+The judgements above are inductively defined in the supplemental material and validate standard lemmas, also given in the supplement.
+%Empty typing contexts are written $\emptyset$, or omitted entirely within judgements, and non-empty typing contexts are written as finite sequences of hypotheses identified up to exchange and contraction.
+We write $\Gamma, \Ghyp{x}{\tau}$, when $x \notin \domof{\Gamma}$, for the extension of $\Gamma$ with a mapping from $x$ to $\Ghyp{x}{\tau}$, and $\Gcons{\Gamma}{\Gamma'}$ when $\domof{\Gamma} \cap \domof{\Gamma'} = \emptyset$ for the typing context mapping each $x \in \domof{\Gamma} \cup \domof{\Gamma'}$ to $x : \tau$ if $x : \tau \in \Gamma$ or $x : \tau \in \Gamma'$.
+% We write $\isctxU{\Delta}{\Gamma}$ if every type in $\Gamma$ is well-formed relative to $\Delta$.
+% \begin{definition}[Typing Context Formation] \label{def:isctxU}
+% $\isctxU{\Delta}{\Gamma}$ iff for each hypothesis $x : \tau \in \Gamma$, we have $\istypeU{\Delta}{\tau}$.
+% \end{definition}
+
+% \begin{subequations}\label{rules:istypeU}
+% \begin{equation*}\label{rule:istypeU-var}
+% \inferrule{ }{\istypeU{\Delta, \Dhyp{t}}{t}}
+% \end{equation*}
+% \begin{equation*}\label{rule:istypeU-parr}
+% \inferrule{
+%   \istypeU{\Delta}{\tau_1}\\
+%   \istypeU{\Delta}{\tau_2}
+% }{\istypeU{\Delta}{\aparr{\tau_1}{\tau_2}}}
+% \end{equation*}
+% \begin{equation*}\label{rule:istypeU-all}
+%   \inferrule{
+%     \istypeU{\Delta, \Dhyp{t}}{\tau}
+%   }{
+%     \istypeU{\Delta}{\aall{t}{\tau}}
+%   }
+% \end{equation*}
+% \begin{equation*}\label{rule:istypeU-rec}
+%   \inferrule{
+%     \istypeU{\Delta, \Dhyp{t}}{\tau}
+%   }{
+%     \istypeU{\Delta}{\arec{t}{\tau}}
+%   }
+% \end{equation*}
+% \begin{equation*}\label{rule:istypeU-prod}
+%   \inferrule{
+%     \{\istypeU{\Delta}{\tau_i}\}_{i \in \labelset}
+%   }{
+%     \istypeU{\Delta}{\aprod{\labelset}{\mapschema{\tau}{i}{\labelset}}}
+%   }
+% \end{equation*}
+% \begin{equation*}\label{rule:istypeU-sum}
+%   \inferrule{
+%     \{\istypeU{\Delta}{\tau_i}\}_{i \in \labelset}
+%   }{
+%     \istypeU{\Delta}{\asum{\labelset}{\mapschema{\tau}{i}{\labelset}}}
+%   }
+% \end{equation*}
+% \end{subequations}
+% Premises of the form $\{{J}_i\}_{i \in \labelset}$ mean that for each $i \in \labelset$, the judgement ${J}_i$ must hold.
+
 % \begin{subequations}\label{rules:hastypeU}
@@
 % \vspace{-2px}
-The \emph{evaluation semantics} of $\miniVersePat$ is organized around the judgements $\isvalU{e}$, which says that $e$ is a value, and $\evalU{e}{e'}$, which says that $e$ evaluates to the value $e'$.
-Additional  judgements, not shown, are needed to define the dynamics of pattern matching, but they do not appear directly in our subsequent developments, so we omit them.
-We assume an eager dynamics and the standard type safety theorem (see supplement.)
-
-\vspace{-4px}
+The \emph{evaluation semantics} of $\miniVersePat$ is organized around the judgements $\isvalU{e}$, which says that $e$ is a value, and $\evalU{e}{e'}$, which says that $e$ evaluates to the value $e'$.% Additional  judgements, not shown, are needed to define the dynamics of pattern matching, but they do not appear directly in our subsequent developments, so we omit them.
+% We assume an eager dynamics.
+
+\vspace{-6px}
 \subsection{Syntax of the Unexpanded Language}\label{sec:syntax-U}\label{sec:s-UL}
@@
 Identifiers are given meaning by expansion to variables during typed expansion, as we will see.
-This distinction between identifiers and variables is technically crucial to our developments, and we return to it below.
+This distinction between identifiers and variables is technically crucial to our developments.
 %We \textbf{cannot} adopt the usual definitions of $\alpha$-renaming of identifiers, because unexpanded types and expressions are still in a ``partially parsed'' state -- the literal bodies, $b$, within an unexpanded expression might contain spliced subterms that are ``surfaced'' by a TLM only during typed expansion, as we will detail below.
@@
 We refer to these as the \emph{common forms}.
-The mapping from expanded forms to common unexpanded forms is defined explicitly in the supplement.
+% The mapping from expanded forms to common unexpanded forms is defined explicitly in the supplement.
 
@@
 %Our paper on Wyvern defines a textual syntax for a similar system \cite{TSLs}. 
-Instead, we need only posit partial metafunctions $\parseUTypF{b}$, $\parseUExpF{b}$ and $\parseUPatF{b}$  that go from character sequences, $b$, to unexpanded types, expressions and patterns (the supplement states the necessary condition in full.)
+Instead, we only posit partial metafunctions $\parseUTypF{b}$, $\parseUExpF{b}$ and $\parseUPatF{b}$  that go from character sequences, $b$, to unexpanded types, expressions and patterns (the supplement states the full condition.)
 % \begingroup
@@
 
-\subsection{Typed Expansion}\label{sec:typed-expansion-U}\label{sec:s-TE}
-Unexpanded terms are checked and expanded simultaneously according to the central judgements of our calculus, the \emph{typed expansion judgements}:
-\vspace{-3px}\[\begin{array}{ll}
-% \textbf{Judgement Form} & \textbf{Description}\\
-\expandsTU{\uDelta}{\utau}{\tau} & \text{$\utau$ has well-formed expansion $\tau$}\\
-\expandsUPX{\ue}{e}{\tau} & \text{$\ue$ has expansion $e$ of type $\tau$}\\
-\ruleExpands{\uDelta}{\uGamma}{\uPsi}{\uPhi}{\urv}{r}{\tau}{\tau'} & \text{$\urv$ has expansion $r$ taking values of type $\tau$ to values of type $\tau'$}\\
-\patExpands{\upctx}{\uPhi}{\upv}{p}{\tau} & \text{$\upv$ has expansion $p$ matching against $\tau$ generating hypotheses $\upctx$}
-\end{array}\]\vspace{-3px}
-%\newcommand{\gray}[1]{{\color{gray} #1}}
-
-% These judgements are inductively defined in the supplement. 
-% \begingroup 
-% \def\thetheorem{\ref{thm:typed-expansion-short-U}}
-
-%These rules validate the following theorem, which establishes that typed expansion produces an expansion of the assigned type. 
-%\begin{theorem}[Typed Expression Expansion] If $\expandsU{\uDD{\uD}{\Delta}}{\uGG{\uG}{\Gamma}}{\uPsi}{\ue}{e}{\tau}$ and $\uetsmenv{\Delta}{\uPsi}$ then $\hastypeU{\Delta}{\Gamma}{e}{\tau}$.\end{theorem}
-%\begin{proof} This is the first part of Theorem \ref{thm:typed-expansion-U}, defined and proven below.\end{proof}
-
-The typed expansion rules that handle common forms mirror the corresponding typing rules.
-The \emph{expression TLM context}, $\uPsi$, and the \emph{pattern TLM context}, $\uPhi$, pass through these rules opaquely.
-For example, the rules for variables and lambdas are below, and the remainder are in the supplement:
-%Each of these rules is based on the corresponding typing rule, i.e.
-Rules (\ref{rule:hastypeU-var}) through (\ref{rule:hastypeU-case}), respectively.
-For example, the following typed expansion rules are based on the typing rules (\ref{rule:hastypeU-var}), (\ref{rule:hastypeU-lam}) and (\ref{rule:hastypeU-ap}), respectively:% for unexpanded expressions of variable, function and application form, respectively:
-{\small\begin{mathpar}
-  \inferrule[ee-id]{ }{\expandsUP{\uDelta}{\uGamma, \uGhyp{\ux}{x}{\tau}}{\uPsi}{\uPhi}{\ux}{x}{\tau}}
-
-  \inferrule[ee-lam]{
-    \expandsTU{\uDelta}{\utau}{\tau}\\
-    \expandsUP{\uDelta}{\uGamma, \uGhyp{\ux}{x}{\tau}}{\uPsi}{\uPhi}{\ue}{e}{\tau'}
-  }{\expandsUPX{\lam{\ux}{\utau}{\ue}}{\aelam{\tau}{x}{e}}{\aparr{\tau}{\tau'}}}
-\end{mathpar}}
-
-The only subtlety here has to do with the relationship between identifiers, $\ux$, in the UL and variables, $x$, in the XL.
-To understand this, we must first describe in detail how unexpanded contexts work.
-\emph{Unexpanded typing contexts}, $\uGamma$, are pairs of the form $\uGG{\uG}{\Gamma}$, where $\uG$ is an \emph{expression identifier expansion context}, and $\Gamma$ is a standard typing context.
-An expression identifier expansion context, $\uG$, is a finite function that maps each expression identifier $\ux \in \domof{\uG}$ to the hypothesis $\vExpands{\ux}{x}$, for some expression variable, $x$, called its expansion.
-We write $\ctxUpdate{\uG}{\ux}{x}$ for the expression identifier expansion context that maps $\ux$ to $\vExpands{\ux}{x}$ and defers to $\uG$ for all other expression identifiers (i.e.
-the previous mapping is \textbf{updated}.) Note the distinction between update and extension (which requires that the new identifier is not already in the domain.) %We write $\uGammaOK{\uGamma}$ when $\uGamma=\uGG{\uG}{\Gamma}$ and each expression variable in $\uG$ is assigned a type by $\Gamma$.
-%\begin{definition} $\uGammaOK{\uGG{\uG}{\Gamma}}$ iff for each $\vExpands{\ux}{x} \in \uG$, we have $\Ghyp{x}{\tau} \in \Gamma$ for some $\tau$.\end{definition}
-%\noindent 
-We define $\uGamma, \uGhyp{\ux}{x}{\tau}$ when $\uGamma = \uGG{\uG}{\Gamma}$ as an abbreviation of $\uGG{\ctxUpdate{\uG}{\ux}{x}}{\Gamma, \Ghyp{x}{\tau}}$.
-
-To develop an intuition for why the update operation is necessary, it is instructive to inspect the derivation of the expansion of the unexpanded expression $\lam{\ux}{\utau}{\lam{\ux}{\utau}{\ux}}$ to $\aelam{\tau}{x_1}{\aelam{\tau}{x_2}{x_2}}$ assuming $\expandsTU{\uDelta}{\utau}{\tau}$:
+\begin{figure}
+%\hspace{-5px}
+\begin{minipage}{\textwidth}
+\small
+$\arraycolsep=2pt\begin{array}{llcl}
+\mathsf{Typ} & \tau & ::= & t ~\vert~ \aparr{\tau}{\tau} ~\vert~ \aall{t}{\tau} ~\vert~ \arec{t}{\tau} \\
+& & \vert &  \aprod{\labelset}{\mapschema{\tau}{i}{\labelset}} ~\vert~ \asum{\labelset}{\mapschema{\tau}{i}{\labelset}}\\
+\mathsf{Exp} & e & ::= & x ~\vert~ \aelam{\tau}{x}{e} ~\vert~ \aeap{e}{e} ~\vert~ \aetlam{t}{e} ~\vert~ \aetap{e}{\tau} \\
+& & \vert & \aefold{e} ~\vert~ \aetpl{\labelset}{\mapschema{e}{i}{\labelset}} ~\vert~  \aein{\ell}{e} \\
+& & \vert & \aematchwith{n}{e}{\seqschemaX{r}}\\
+\mathsf{Rule} & r & ::= & \aematchrule{p}{e}\\
+\mathsf{Pat} & p & ::= & x  ~\vert~ \aewildp ~\vert~ \aefoldp{p} ~\vert~ \aetplp{\labelset}{\mapschema{p}{i}{\labelset}}\\
+& & \vert & \aeinjp{\ell}{p}
+\end{array}$
+\end{minipage}
+\vspace{-10px}
+\caption[Syntax of the XL of $\miniVersePat$]{Syntax of the expanded language (XL).
+XL terms are \emph{abstract binding trees} (ABTs) identified up to alpha-equivalence, so we follow the syntactic conventions of \citet{pfple1}.
+Metavariable $x$ and $t$ ranges over variables.
+%When using stylized forms, the label set is omitted when it can be inferred, e.g.
+the labeled product type $\prodt{\finmap{\mapitem{\ell_1}{e_1}, \mapitem{\ell_2}{e_2}}}$ leaves the label set $\{\ell_1, \ell_2\}$ implicit.
+% When we use the stylized forms, we assume that the reader can infer suppressed indices and arguments from the surrounding context.
+}
+\vspace{-10px}
+\label{fig:U-expanded-terms}
+\end{figure}
+
+\begin{figure*}
 {\small\begin{mathpar}
@@
 \end{mathpar}}
-\noindent
+\vspace{-12px}
+\caption{An example expansion derviation demonstrating how identifiers and variables are separately tracked.}
+\label{fig:expansion-exmpl}
+\vspace{-10px}
+\end{figure*}
+
+\vspace{-6px}
+\subsection{Typed Expansion}\label{sec:typed-expansion-U}\label{sec:s-TE}
+\vspace{-3px}
+Unexpanded terms are checked and expanded simultaneously according to the central \emph{typed expansion judgements}:
+\vspace{-5px}\[\begin{array}{ll}
+% \textbf{Judgement Form} & \textbf{Description}\\
+\expandsTU{\uDelta}{\utau}{\tau} & \text{$\utau$ has well-formed expansion $\tau$}\\
+\expandsUPX{\ue}{e}{\tau} & \text{$\ue$ has expansion $e$ of type $\tau$}\\
+% \ruleExpands{\uDelta}{\uGamma}{\uPsi}{\uPhi}{\urv}{r}{\tau}{\tau'} & \text{$\urv$ has expansion $r$ taking values of type $\tau$ to values of type $\tau'$}\\
+\patExpands{\upctx}{\uPhi}{\upv}{p}{\tau} & \text{$\upv$ has expansion $p$ matching $\tau$}
+\end{array}\]\vspace{-5px}
+%\newcommand{\gray}[1]{{\color{gray} #1}}
+
+% These judgements are inductively defined in the supplement. 
+% \begingroup 
+% \def\thetheorem{\ref{thm:typed-expansion-short-U}}
+
+%These rules validate the following theorem, which establishes that typed expansion produces an expansion of the assigned type. 
+%\begin{theorem}[Typed Expression Expansion] If $\expandsU{\uDD{\uD}{\Delta}}{\uGG{\uG}{\Gamma}}{\uPsi}{\ue}{e}{\tau}$ and $\uetsmenv{\Delta}{\uPsi}$ then $\hastypeU{\Delta}{\Gamma}{e}{\tau}$.\end{theorem}
+%\begin{proof} This is the first part of Theorem \ref{thm:typed-expansion-U}, defined and proven below.\end{proof}
+
+The typed expansion rules that handle common forms mirror the corresponding typing rules.
+The \emph{expression TLM context}, $\uPsi$, and the \emph{pattern TLM context}, $\uPhi$, pass through these rules opaquely.
+For example, the rules for variables and lambdas are shown being applied in Fig.
+\ref{fig:expansion-exmpl}, discussed below.
+%Each of these rules is based on the corresponding typing rule, i.e.
+Rules (\ref{rule:hastypeU-var}) through (\ref{rule:hastypeU-case}), respectively.
+For example, the following typed expansion rules are based on the typing rules (\ref{rule:hastypeU-var}), (\ref{rule:hastypeU-lam}) and (\ref{rule:hastypeU-ap}), respectively:% for unexpanded expressions of variable, function and application form, respectively:
+% {\small\begin{mathpar}
+%   \inferrule[ee-id]{ }{\expandsUP{\uDelta}{\uGamma, \uGhyp{\ux}{x}{\tau}}{\uPsi}{\uPhi}{\ux}{x}{\tau}}
+
+%   \inferrule[ee-lam]{
+%     \expandsTU{\uDelta}{\utau}{\tau}\\
+%     \expandsUP{\uDelta}{\uGamma, \uGhyp{\ux}{x}{\tau}}{\uPsi}{\uPhi}{\ue}{e}{\tau'}
+%   }{\expandsUPX{\lam{\ux}{\utau}{\ue}}{\aelam{\tau}{x}{e}}{\aparr{\tau}{\tau'}}}
+% \end{mathpar}}
+
+The only subtlety related to common forms has to do with the relationship between identifiers, $\ux$, in the UL and variables, $x$, in the XL.
+To understand this, we must first describe in detail how unexpanded contexts work.
+\emph{Unexpanded typing contexts}, $\uGamma$, are pairs of the form $\uGG{\uG}{\Gamma}$, where $\uG$ maps each expression identifier $\ux \in \domof{\uG}$ to the hypothesis $\vExpands{\ux}{x}$, for some expression variable, $x$, called its expansion.
+The standard typing context, $\Gamma$, then tracks the type of $x$.
+We write $\ctxUpdate{\uG}{\ux}{x}$ for the expression identifier expansion context that maps $\ux$ to $\vExpands{\ux}{x}$ and defers to $\uG$ for all other expression identifiers (i.e.
+the previous mapping is \textbf{updated}.) Note the distinction between update and extension (which requires that the new identifier is not already in the domain.) %We write $\uGammaOK{\uGamma}$ when $\uGamma=\uGG{\uG}{\Gamma}$ and each expression variable in $\uG$ is assigned a type by $\Gamma$.
+%\begin{definition} $\uGammaOK{\uGG{\uG}{\Gamma}}$ iff for each $\vExpands{\ux}{x} \in \uG$, we have $\Ghyp{x}{\tau} \in \Gamma$ for some $\tau$.\end{definition}
+%\noindent 
+We define $\uGamma, \uGhyp{\ux}{x}{\tau}$ when $\uGamma = \uGG{\uG}{\Gamma}$ as an abbreviation of $\uGG{\ctxUpdate{\uG}{\ux}{x}}{\Gamma, \Ghyp{x}{\tau}}$.
+
+To develop an intuition for why the update operation is necessary, it is instructive to inspect in Fig.
+\ref{fig:expansion-exmpl} the derivation of the expansion of the unexpanded expression $\lam{\ux}{\utau}{\lam{\ux}{\utau}{\ux}}$ to $\aelam{\tau}{x_1}{\aelam{\tau}{x_2}{x_2}}$ assuming $\expandsTU{\uDelta}{\utau}{\tau}$.
 Notice that when Rule \textsc{ee-lam} is applied, the type identifier expansion context is updated but the typing context is extended with a (necessarily fresh) variable, first $x_1$ then $x_2$.
@@
 
-\emph{Unexpanded type formation contexts}, $\uDelta$, consist of a \emph{type identifier expansion context}, $\uD$, paired with a standard type formation context, $\Delta$, and operate analagously (see supplement.)
+% \emph{Unexpanded type formation contexts}, $\uDelta$, consist of a \emph{type identifier expansion context}, $\uD$, paired with a standard type formation context, $\Delta$, and operate analagously (see supplement.)
 % of the form $\uDD{\uD}{\Delta}$, i.e.
@@
 
-Before we continue, let us state an important invariant: that typed expression expansion produces a well-typed expression.
-\begin{theorem}[Typed Expression Expansion]\label{thm:typed-expansion-short-U}
-If $\expandsUP{\uDD{\uD}{\Delta}\hspace{-3px}}{\uGG{\uG}{\Gamma}\hspace{-3px}}{\uPsi}{\uPhi}{\ue}{e}{\tau}$ then $\hastypeU{\Delta}{\Gamma}{e}{\tau}$.
-\end{theorem}
-For the typed expansion rules governing common forms, like the two example rules above, the typed expansion rules mirror the corresponding typing rules so it is easy to see that this invariant holds.
-The details are in the supplement.
+% % Before we continue, let us state an important invariant: that Typed expansion produces a well-typed expression. 
+% \begin{theorem}[Typed Expression Expansion]\label{thm:typed-expansion-short-U} ~\\
+% If $\expandsUP{\uDD{\uD}{\Delta}\hspace{-3px}}{\uGG{\uG}{\Gamma}\hspace{-3px}}{\uPsi}{\uPhi}{\ue}{e}{\tau}$ then $\hastypeU{\Delta}{\Gamma}{e}{\tau}$.
+% \end{theorem}
+% For the typed expansion rules governing common forms, like the two example rules applied in Fig.
+\ref{fig:expansion-exmpl}, it is easy to see that this invariant holds.
 The  rules of particular interest are the rules governing TLM definitions and TLM application, which are the topic of the remainder of this section.
@@
 
-\vspace{-4px}
+\vspace{-6px}
 \subsection{TLM Definitions}\label{sec:U-uetsm-definition}\label{sec:s-TLM-def}
@@
  % defines an {seTLM} identified as $\tsmv$ with \emph{unexpanded type annotation} $\utau$ and \emph{parse function} $\eparse$ for use within $\ue$. 
- The rule below defines typed expansion of the seTLM definition form:
+Rule \textsc{ee-def-setsm} in Fig.
+\ref{fig:ee-def-setsm} governs simple expression TLM (seTLM) definitions.
+The first premise expands the unexpanded type annotation.
+The second premise checks that $\eparse$ is a closed expanded function of the given function type.
+(In the supplement, we add the machinery necessary for parse functions that are neither closed nor yet expanded.)
+
 % \begin{subequations}[resume]
@@
 % \end{equation*}
+\begin{figure*}
 {\small\begin{mathpar}
@@
 }
+~~~~~~~~~~~~\hspace{15px}
+\inferrule[ee-ap-setsm]{
+  \encodeBody{b}{\ebody}\\
+  \evalU{\ap{\eparse}{\ebody}}{\aein{\mathtt{SuccessE}}{\ecand}}\\\\
+  \decodeCondE{\ecand}{\ce}\\
+  \segOK{\segof{\ce}}{b}\\\\
+  \cvalidE{\emptyset}{\emptyset}{\esceneUP{\uDelta}{\uGamma}{\uPsi}{\uPhi}{b}}{\ce}{e}{\tau}
+}{
+  \expandsUP{\uDelta}{\uGamma}{\uPsi', \uShyp{\tsmv}{a}{\tau}{\eparse}}{\uPhi}{\utsmap{\tsmv}{b}}{e}{\tau}
+}
 \end{mathpar}}
+\vspace{-12px}
+\caption{The typed expansion rules for expression TLM definition and application.}
+\label{fig:ee-def-setsm}
+\vspace{-10px}
+\end{figure*}
 % \end{subequations}
-\vspace{-3px}
-
-The first premise expands the unexpanded type annotation.
-The second premise checks that $\eparse$ is a closed expanded function\footnote{In Sec.
-\ref{sec:static-eval}, we add the machinery necessary for parse functions that are neither closed nor yet expanded.} of type $\aparr{\tBody}{\tParseResultExp}$.
 
 The type abbreviated $\tBody$ classifies encodings of literal bodies, $b$.
-The mapping from literal bodies, $b$, to values of type $\tBody$ is defined by the \emph{body encoding judgement} $\encodeBody{b}{\ebody}$.
-An inverse mapping can also be defined   by the \emph{body decoding judgement} $\decodeBody{\ebody}{b}$.
-Rather than defining $\tBody$ explicitly, and these judgements inductively against that definition (which would be tedious and uninteresting), it suffices to take as a condition that there is an isomorphism between literal bodies and values of type $\tBody$ mediated by these judgements (see supplement.)
+Rather than defining $\tBody$ explicitly it suffices to take as a condition that there is an isomorphism between literal bodies and values of type $\tBody$ mediated in one direction by a judgement $\encodeBody{b}{\ebody}$ that will come up below.
 
@@
 An inverse mapping is defined by the \emph{proto-expression decoding judgement}, $\decodeCondE{e}{\ce}$.
-Again, rather than picking a particular definition of $\tCEExp$ and defining the judgements above inductively against it, we take as a condition that there is an isomorphism between values of type $\tCEExp$ and closed proto-expressions mediated by these judgements (see supplement.)
-
-The third premise of Rule \textsc{ee-def-setsm} evaluates the parse function to a value. This is not semantically necessary, but it is the choice one would expect to make in an eager language.
+Again, we need only take as a condition that there is an isomorphism between values of type $\tCEExp$ and closed proto-expressions mediated by these judgements (see supplement.)
+
+The third premise of Rule \textsc{ee-def-setsm} evaluates the parse function to a value. This is not necessary, but it is the choice one would expect to make in an eager language.
 
@@
 TLMs define behavior that is relevant during typed expansion, but not during evaluation.
+
 {Expression TLM contexts}, $\uPsi$, are of the form $\uAS{\uA}{\Psi}$, where $\uA$ is a \emph{TLM identifier expansion context} and $\Psi$ is an \emph{expression TLM definition context}.
-
-A {TLM identifier expansion context}, $\uA$, is a finite function mapping each TLM identifier $\tsmv \in \domof{\uA}$ to the \emph{TLM identifier expansion}, $\vExpands{\tsmv}{a}$, for some \emph{TLM name}, $a$.
 We distinguish TLM identifiers, $\tsmv$, from TLM names, $a$, for much the same reason that we distinguish type and expression identifiers from type and expression variables: in order to allow a TLM definition to shadow a previously defined TLM definition without relying on an implicit identification convention.
+
+% A {TLM identifier expansion context}, $\uA$, maps each TLM identifier $\tsmv \in \domof{\uA}$ to the \emph{TLM identifier expansion}, $\vExpands{\tsmv}{a}$, for some \emph{TLM name}, $a$.
 
@@
 
-The spTLM definition form operates analagously.
-The rule is reproduced below, and the details, which mirror those for seTLMs, are in the supplement.
-Notice that the spTLM context, $\uPhi$, rather than the $\uPsi$ is updated.
+The simple pattern TLM (spTLM) definition form operates analagously (see supplement), with the spTLM context, $\uPhi$, rather than the $\uPsi$ updated.
 This allows expression and pattern TLMs to use the same identifiers.
-{\vspace{-3px}\small\begin{mathpar}
-\inferrule[ee-def-sptsm]{
-  \expandsTU{\uDelta}{\utau}{\tau}\\
-  \hastypeU{\emptyset}{\emptyset}{\eparse}{\aparr{\tBody}{\tParseResultPat}}\\\\
-  \evalU{\eparse}{\eparse'}\\
-  \expandsUP{\uDelta}{\uGamma}{\uPsi}{\uPhi, \uPhyp{\tsmv}{a}{\tau}{\eparse'}}{\ue}{e}{\tau'}
-}{
-  \expandsUPX{\usyntaxup{\tsmv}{\utau}{\eparse}{\ue}}{e}{\tau'}
-}
-\end{mathpar}}
-\vspace{-3px}
+% {\vspace{-3px}\small\begin{mathpar}
+% \inferrule[ee-def-sptsm]{
+%   \expandsTU{\uDelta}{\utau}{\tau}\\
+%   \hastypeU{\emptyset}{\emptyset}{\eparse}{\aparr{\tBody}{\tParseResultPat}}\\\\
+%   \evalU{\eparse}{\eparse'}\\
+%   \expandsUP{\uDelta}{\uGamma}{\uPsi}{\uPhi, \uPhyp{\tsmv}{a}{\tau}{\eparse'}}{\ue}{e}{\tau'}
+% }{
+%   \expandsUPX{\usyntaxup{\tsmv}{\utau}{\eparse}{\ue}}{e}{\tau'}
+% }
+% \end{mathpar}}
+% \vspace{-3px}
 
@@
 $\mathsf{str}[s]$ and $\mathsf{num}[n]$.} This operator is indexed by the TLM name $\tsmv$ and takes no arguments.
-The typed expansion rule governing seTLM application is below:
+Rule \textsc{ee-ap-setsm} governing this form is shown in Fig. \ref{fig:ee-def-setsm}.
+
 % \begin{subequations}[resume]
@@
 % \end{equation*}
-{\small\begin{mathpar}
-\inferrule[ee-ap-setsm]{
-  \encodeBody{b}{\ebody}\\
-  \evalU{\ap{\eparse}{\ebody}}{\aein{\mathtt{SuccessE}}{\ecand}}\\
-  \decodeCondE{\ecand}{\ce}\\\\
-  \segOK{\segof{\ce}}{b}\\
-  \cvalidE{\emptyset}{\emptyset}{\esceneUP{\uDelta}{\uGamma}{\uPsi}{\uPhi}{b}}{\ce}{e}{\tau}
-}{
-  \expandsUP{\uDelta}{\uGamma}{\uPsi', \uShyp{\tsmv}{a}{\tau}{\eparse}}{\uPhi}{\utsmap{\tsmv}{b}}{e}{\tau}
-}
-\end{mathpar}}
-
-The first premise determines the encoding of the literal body, $\ebody$, which, as discussed above, is a value of type $\tBody$.
+% {\small\begin{mathpar}
+% \inferrule[ee-ap-setsm]{
+%   \encodeBody{b}{\ebody}\\
+%   \evalU{\ap{\eparse}{\ebody}}{\aein{\mathtt{SuccessE}}{\ecand}}\\
+%   \decodeCondE{\ecand}{\ce}\\\\
+%   \segOK{\segof{\ce}}{b}\\
+%   \cvalidE{\emptyset}{\emptyset}{\esceneUP{\uDelta}{\uGamma}{\uPsi}{\uPhi}{b}}{\ce}{e}{\tau}
+% }{
+%   \expandsUP{\uDelta}{\uGamma}{\uPsi', \uShyp{\tsmv}{a}{\tau}{\eparse}}{\uPhi}{\utsmap{\tsmv}{b}}{e}{\tau}
+% }
+% \end{mathpar}}
+
+The first premise encodes the literal body, $\ebody$, which, as described above, is a value of type $\tBody$.
 
@@
 If parsing succeeds, i.e.
-a value of the form $\aein{\mathtt{SuccessE}}{\ecand}$ results from evaluation, then $\ecand$ will be a value of type $\tCEExp$ (assuming a well-formed expression TLM context, by application of the Type Safety assumption.) We call $\ecand$ the \emph{encoding of the proto-expansion}.
+a value of the form $\aein{\mathtt{SuccessE}}{\ecand}$ results from evaluation, then $\ecand$ will be a value of type $\tCEExp$ (assuming a well-formed expression TLM context, by application of Type Safety.) We call $\ecand$ the \emph{encoding of the proto-expansion}.
 If the parse function produces a value labeled $\lbltxt{ParseError}$, then typed expansion fails and formally, no rule is necessary.
 
-The third premise decodes the encoding of the proto-expansion.
-
-The fourth premise determines the segmentation of the proto-expansion, $\segof{\ce}$, and ensures that it is valid with respect to $b$ via the predicate $\segOK{\psi}{b}$, which  checks that each segment in the finite set of segments $\psi$ has non-negative length and is within bounds of $b$, and that the segments in $\psi$ do not overlap.
+The third premise decodes the encoding of the proto-expansion using the judgement described in Sec. \ref{sec:U-uetsm-definition}.
+
+The fourth premise of Rule \textsc{ee-ap-setsm} determines the segmentation of the proto-expansion, $\segof{\ce}$, and ensures that it is valid with respect to $b$ via the predicate $\segOK{\psi}{b}$, which  checks that each segment in the finite set of segments $\psi$ has non-negative length and is within bounds of $b$, and that the segments in $\psi$ do not overlap.
 
@@
 
-The typed pattern expansion rule governing spTLM application is analagous:
-{\small\begin{mathpar}
-\inferrule[pe-ap-sptsm]{
-  \encodeBody{b}{\ebody}\\
-  \evalU{\ap{\eparse}{\ebody}}{\aein{\mathtt{SuccessP}}{\ecand}}\\
-  \decodeCEPat{\ecand}{\cpv}\\\\
-  \segOK{\segof{\cpv}}{b}\\
-  \cvalidP{\upctx}{\pscene{\uDelta}{\uPhi}{b}}{\cpv}{p}{\tau}
-}{
-  \patExpands{\upctx}{\uPhi', \uPhyp{\tsmv}{a}{\tau}{\eparse}}{\utsmap{\tsmv}{b}}{p}{\tau}
-}
-\end{mathpar}}
+The typed pattern expansion rule governing pattern TLM application is analagous (see supplement).
+% {\small\begin{mathpar}
+% \inferrule[pe-ap-sptsm]{
+%   \encodeBody{b}{\ebody}\\
+%   \evalU{\ap{\eparse}{\ebody}}{\aein{\mathtt{SuccessP}}{\ecand}}\\
+%   \decodeCEPat{\ecand}{\cpv}\\\\
+%   \segOK{\segof{\cpv}}{b}\\
+%   \cvalidP{\upctx}{\pscene{\uDelta}{\uPhi}{b}}{\cpv}{p}{\tau}
+% }{
+%   \patExpands{\upctx}{\uPhi', \uPhyp{\tsmv}{a}{\tau}{\eparse}}{\utsmap{\tsmv}{b}}{p}{\tau}
+% }
+% \end{mathpar}}
 
@@
 
-\vspace{-4px}
+\vspace{-6px}
 \subsection{Proto-Expansion Validation}\label{sec:ce-validation-U}\label{sec:ce-syntax-U}\label{sec:s-PEV}
@@
 \begin{figure}
-\vspace{-10px}
 \begin{minipage}{\textwidth}
@@
 \mathsf{PrTyp} & \ctau & ::= & t ~\vert~
-\aceparr{\ctau}{\ctau} ~\vert~
-\aceall{t}{\ctau} ~\vert~
-\acerec{t}{\ctau} ~\vert~
-\aceprod{\labelset}{\mapschema{\ctau}{i}{\labelset}} \\
-& & \vert & \acesum{\labelset}{\mapschema{\ctau}{i}{\labelset}} ~\vert~ \acesplicedt{m}{n}\\
+\aceparr{\ctau}{\ctau} ~\vert~ \cdots ~\vert~
+% \aceall{t}{\ctau} ~\vert~
+% \acerec{t}{\ctau} ~\vert~
+% \aceprod{\labelset}{\mapschema{\ctau}{i}{\labelset}} \\
+% & & \vert & 
+% \acesum{\labelset}{\mapschema{\ctau}{i}{\labelset}} ~\vert~ 
+\acesplicedt{m}{n}\\
 \mathsf{PrExp} & \ce & ::= & x ~\vert~
@@
 \aceap{\ce}{\ce} ~\vert~
-\acetlam{t}{\ce} ~\vert~
-\acetap{\ce}{\ctau} ~\vert~
-\acefold{\ce} \\
-& & \vert & \acetpl{\labelset}{\mapschema{\ce}{i}{\labelset}} ~\vert~
-\acein{\ell}{\ce} ~\vert~
+\cdots \\
+% \acetlam{t}{\ce} ~\vert~
+% \acetap{\ce}{\ctau} ~\vert~
+% \acefold{\ce} \\
+% & & \vert & \acetpl{\labelset}{\mapschema{\ce}{i}{\labelset}} ~\vert~
+% \acein{\ell}{\ce} ~\vert~
+& & \vert & 
 \acematchwith{n}{\ce}{\seqschemaX{\crv}} ~\vert~
@@
 \mathsf{PrPat} & \cpv & ::= & \acewildp ~\vert~
-\acefoldp{p} ~\vert~
-\acetplp{\labelset}{\mapschema{\cpv}{i}{\labelset}} ~\vert~
-\aceinjp{\ell}{\cpv} ~\vert~
+\cdots ~\vert~
+% \acefoldp{p} ~\vert~
+% \acetplp{\labelset}{\mapschema{\cpv}{i}{\labelset}} ~\vert~
+% \aceinjp{\ell}{\cpv} ~\vert~
 \acesplicedp{m}{n}{\ctau} 
@@
 \end{minipage}
-\vspace{-5px}
-\caption[Syntax of $\miniVersePat$ proto-types and proto-expressions]{Syntax of $\miniVersePat$ proto-expansions. Proto-expansion terms are ABTs identified up to alpha-equivalence.}
+\vspace{-10px}
+\caption[Syntax of proto-types and proto-expressions]{Syntax of proto-expansions. Proto-expansion terms are ABTs identified up to alpha-equivalence.}
 \label{fig:U-candidate-terms}
@@
 
-Finally, the \emph{proto-expansion validation judgements} validate the proto-expansions generated by TLMs and simultaneously generate their final expansions:% are types and expanded expressions, respectively.
-\[\begin{array}{ll}
+Finally, we arrive at the crucial \emph{proto-expansion validation judgements}, which validate the proto-expansions generated by TLMs and simultaneously generate their final expansions:% are types and expanded expressions, respectively.
+\vspace{-5px}\[\begin{array}{ll}
 \cvalidT{\Delta}{\tscenev}{\ctau}{\tau} & \text{$\ctau$ has well-formed expansion $\tau$}\\
 \cvalidE{\Delta}{\Gamma}{\escenev}{\ce}{e}{\tau} & \text{$\ce$ has expansion $e$ of type $\tau$}\\
-\cvalidR{\Delta}{\Gamma}{\escenev}{\crv}{r}{\tau}{\tau'} & \text{$\crv$ has expansion $r$ taking values of type $\tau$ to values of type $\tau'$}\\
-\cvalidP{\upctx}{\pscenev}{\cpv}{p}{\tau} & \text{$\cpv$ has expansion $p$ matching against $\tau$ generating assumptions $\upctx$}
+% \cvalidR{\Delta}{\Gamma}{\escenev}{\crv}{r}{\tau}{\tau'} & \text{$\crv$ has expansion $r$ taking values of type $\tau$ to values of type $\tau'$}\\
+\cvalidP{\upctx}{\pscenev}{\cpv}{p}{\tau} & \text{$\cpv$ has expansion $p$ matching $\tau$}
 \end{array}\]
+\vspace{-5px}
+
 The purpose of the \emph{splicing scenes} $\tscenev$, $\escenev$ and $\pscenev$ is to ``remember'' the contexts and literal body from the TLM application site (cf.
-Rules \textsc{ee-ap-setsm} and \textsc{pe-ap-sptsm}) for when validation encounters spliced terms.
-\emph{Type splicing scenes}, $\tscenev$, are of the form $\tsceneUP{\uDelta}{b}$.
-\emph{Expression splicing scenes}, $\escenev$, are of the form $\esceneUP{\uDelta}{\uGamma}{\uPsi}{\uPhi}{b}$.
-\emph{Pattern splicing scenes}, $\pscenev$, are of the form $\pscene{\uDelta}{\uPhi}{b}$.
-
-\subsubsection{Common Forms} Most of the proto-expansion forms mirror corresponding expanded forms.
-The rules governing proto-expansion validation for these common forms correspondingly mirror the typing rules.
-They are given in the supplement.
-Splicing scenes pass opaquely through these rules, i.e.
+the final premise of Rule \textsc{ee-ap-setsm} in Fig.
+\ref{fig:ee-def-setsm}) for when validation encounters spliced terms.
+For example, \emph{expression splicing scenes}, $\escenev$, are of the form $\esceneUP{\uDelta}{\uGamma}{\uPsi}{\uPhi}{b}$.
+
+\paragraph{Common Forms} Most of the proto-expansion forms, including all of those elided in Fig.
+\ref{fig:U-candidate-terms} mirror corresponding expanded forms.
+The rules governing proto-expansion validation for these common forms  in the supplement correspondingly mirror the typing rules.
+Splicing scenes---$\escenev$, $\tscenev$ and $\pscenev$--pass opaquely through these rules, i.e.
 none of these rules can access the application site contexts.
@@
 Proto-patterns, in contrast, arise only from pattern TLMs.
-There is not a variable proto-pattern form.
-
-\subsubsection{References to Spliced Terms} The only interesting forms are the references to spliced unexpanded types, expressions and patterns.
-Let us first consider the rule for references to spliced unexpanded types:
-{\small\begin{mathpar}
-  \inferrule[ptv-spliced]{
-    \parseUTyp{\bsubseq{b}{m}{n}}{\utau}\\
-    \expandsTU{\uDD{\uD}{\Delta_\text{app}}}{\utau}{\tau}\\
-    \Delta \cap \Delta_\text{app} = \emptyset
-  }{
-    \cvalidT{\Delta}{\tsceneU{\uDD{\uD}{\Delta_\text{app}}}{b}}{\acesplicedt{m}{n}}{\tau}
-  }
-\end{mathpar}}
-
-\noindent
-This first premise of this rule parses out the requested segment of the literal body, $b$, to produce an unexpanded type, $\utau$.
-It then invokes type expansion to ensure that this unexpanded type is well-formed \emph{under the application site context}, $\uDD{\uD}{\Delta_\text{app}}$, but \emph{not} the expansion-local type formation context, $\Delta$.
-The final premise requires that the application site type formation context is disjoint from the expansion-local type formation context.
-Because proto-expansions are ABTs identified up to alpha-equivalence, we can always discharge the final premise by alpha-varying the proto-expansion.
-Collectively, this ensures that type variable capture does not occur (we will formally state this property in the next section.)
-
-The rule for references to spliced unexpanded expressions is fundamentally analagous:
+There is not a variable proto-pattern form, for the reasons described in Sec.
+\ref{sec:sptsms}.
+
+\paragraph{References to Spliced Terms} The only interesting forms are the references to spliced unexpanded types, expressions and patterns.
+Let us consider the rule for references to spliced unexpanded expressions:
 {\small\begin{mathpar}
@@
 \end{mathpar}}
+\vspace{-5px}
+% {\small\begin{mathpar}
+%   \inferrule[ptv-spliced]{
+%     \parseUTyp{\bsubseq{b}{m}{n}}{\utau}\\
+%     \expandsTU{\uDD{\uD}{\Delta_\text{app}}}{\utau}{\tau}\\
+%     \Delta \cap \Delta_\text{app} = \emptyset
+%   }{
+%     \cvalidT{\Delta}{\tsceneU{\uDD{\uD}{\Delta_\text{app}}}{b}}{\acesplicedt{m}{n}}{\tau}
+%   }
+% \end{mathpar}}
 
 \noindent
-We first splice out the requested segment.
+This first premise of this rule parses out the requested segment of the literal body, $b$, to produce an unexpanded expression, $\ue$.
+The second premise performs proto-type expansion on the given type annotation, $\ctau$, producing a type, $\tau$.
+The third premise then invokes type expansion on $\ue$ \emph{under the application site contexts}, $\uDD{\uD}{\Delta_\text{app}}$ and $\uGG{\uG}{\Gamma_\text{app}}$, but \emph{not} the expansion-local contexts, $\Delta$ and $\Gamma$.
+The final premise requires that the application site contexts are disjoint from the expansion-local type formation context.
+Because proto-expansions are ABTs identified up to alpha-equivalence, we can always discharge the final premise by alpha-varying the proto-expansion.
+This serves to enforce capture avoidance.
+%(we will formally state this property in the next section.)
+
+The rule for references to spliced unexpanded types and patterns are fundamentally analagous (see supplement). 
+% \noindent
+% We first splice out the requested segment.
 The second premise expands the type annotation under an empty context, because the type annotation must be meaningful at the application site (so, independent of $\Delta$ and $\Gamma$) and not itself make any assumptions about the application site context.
@@
 
-The rule for references to spliced unexpanded patterns is entirely analagous:
-{\small\begin{mathpar}
-\inferrule[ppv-spliced]{
-  \parseUPat{\bsubseq{b}{m}{n}}{\upv}\\
-  \cvalidT{\emptyset}{\tsceneUP{\uDelta}{b}}{\ctau}{\tau}\\
-  \patExpands{\upctx}{\uPhi}{\upv}{p}{\tau}
-}{
-  \cvalidP{\upctx}{\pscene{\uDelta}{\uPhi}{b}}{\acesplicedp{m}{n}{\ctau}}{p}{\tau}
-}
-\end{mathpar}}
-
-\vspace{-4px}
+% The rule for references to spliced unexpanded patterns is entirely analagous (see supplement).
+% {\small\begin{mathpar}
+% \inferrule[ppv-spliced]{
+%   \parseUPat{\bsubseq{b}{m}{n}}{\upv}\\
+%   \cvalidT{\emptyset}{\tsceneUP{\uDelta}{b}}{\ctau}{\tau}\\
+%   \patExpands{\upctx}{\uPhi}{\upv}{p}{\tau}
+% }{
+%   \cvalidP{\upctx}{\pscene{\uDelta}{\uPhi}{b}}{\acesplicedp{m}{n}{\ctau}}{p}{\tau}
+% }
+% \end{mathpar}}
+
+\vspace{-6px}
 \subsection{Metatheory}\label{sec:s-metatheory}
-\vspace{-2px}
-\subsubsection{Typed Expansion} Let us now return to Theorem \ref{thm:typed-expansion-short-U}, the typed expression expansion theorem that was mentioned at the end of Sec.
-\ref{sec:typed-expansion-U}.
-As it turns out, in order to prove this theorem, we must  prove the following stronger theorem, because the proto-expression validation judgement is defined mutually inductively with the typed expansion judgement (due to the three rules just described.)
+\vspace{-1px}
+\subsubsection{Typed Expansion} The first property that we are interested in is simple: that typed expansion produces a well-typed expansion.
+As it turns out, in order to prove this theorem, we must  prove the following stronger theorem, because the proto-expression validation judgement is defined mutually inductively with the typed expansion judgement (due to splicing).
 
@@
 \end{theorem}
+\vspace{-5px}
 % \endgroup
-The additional second clause simply states that the final expansion produced by proto-expression validation is well-typed under the combined application site and expansion-internal context (because spliced terms are distinguished only in the proto-expansion, not in the final expansion.) Such combined contexts can only be formed if the constituents are disjoint.
+The additional second clause simply states that the final expansion produced by proto-expression validation is well-typed under the combined application site and expansion-internal context (because spliced terms are distinguished only in the proto-expansion, not in the final expansion.) The combined context can only be formed if these are disjoint.
 
@@
 The intuition is that parsing a term of out a literal body cannot produce a bigger term than the term that contained that very literal body.
-More specifically, {the sum of the lengths of the literal bodies that appear in the term strictly decreases each time you perform a nested TLM application} because some portion of the term has to be consumed by the TLM name and the delimiters.
+% More specifically, {the sum of the lengths of the literal bodies that appear in the term strictly decreases each time you perform a nested TLM application} because some portion of the term has to be consumed by the TLM name and the delimiters.
 The details are given in the supplemental material.
-A similar argument is needed to prove Typed Pattern Expansion:
-\begin{theorem}[Typed Pattern Expansion (Strong)] ~
-\begin{enumerate}[nolistsep]
-  \item If $\pExpandsSP{\uDD{\uD}{\Delta}}{\uAS{\uA}{\Phi}}{\upv}{p}{\tau}{\uGG{\uG}{\pctx}}$ then $\patType{\pctx}{p}{\tau}$.
-  \item If $\cvalidP{\uGG{\uG}{\pctx}}{\pscene{\uDD{\uD}{\Delta}}{\uAP{\uA}{\Phi}}{b}}{\cpv}{p}{\tau}$ then $\patType{\pctx}{p}{\tau}$.
-\end{enumerate}
-\end{theorem}
-
-\vspace{-8px}
+
+%  A similar argument is needed to prove Typed Pattern Expansion:
+% \begin{theorem}[Typed Pattern Expansion (Strong)] ~
+% \begin{enumerate}[nolistsep]
+%   \item If $\pExpandsSP{\uDD{\uD}{\Delta}}{\uAS{\uA}{\Phi}}{\upv}{p}{\tau}{\uGG{\uG}{\pctx}}$ then $\patType{\pctx}{p}{\tau}$.
+%   \item If $\cvalidP{\uGG{\uG}{\pctx}}{\pscene{\uDD{\uD}{\Delta}}{\uAP{\uA}{\Phi}}{b}}{\cpv}{p}{\tau}$ then $\patType{\pctx}{p}{\tau}$.
+% \end{enumerate}
+% \end{theorem}
+
+\vspace{-6px}
 \subsubsection{seTLM Reasoning Principles} 
-The following theorem summarizes the abstract reasoning principles that programmers can rely on when applying an seTLM. Informal descriptions of the labeled clauses are given inline, in gray boxes. 
+\vspace{-3px}
+The following theorem summarizes the abstract reasoning principles that programmers can rely on when applying a simple expression TLM.
+Informal descriptions of the labeled clauses are given inline, in gray boxes.
 % \begingroup
@@
 \vspace{-4px}
-\begin{theorem}[seTLM Reasoning Principles]\label{thm:setlm-reasoning}
+\begin{theorem}[seTLM Reasoning Principles]\label{thm:setlm-reasoning} ~\\
 If $\expandsU{\uDD{\uD}{\Delta}}{\uGG{\uG}{\Gamma}}{\uPsi}{\utsmap{\tsmv}{b}}{e}{\tau}$ then:
@@
 \item $\encodeBody{b}{\ebody}$
-\item $\evalU{\ap{\eparse}{\ebody}}{\aein{\lbltxt{SuccessE}}{\ecand}}$
+\item (\textbf{Responsibility}) $\evalU{\ap{\eparse}{\ebody}}{\aein{\lbltxt{SuccessE}}{\ecand}}$
+  \begin{quote}
+  \begin{grayparbox}
+  The parse function of the invoked TLM is responsible for the expansion.
+  \end{grayparbox}
+  \end{quote}
 \item $\decodeCondE{\ecand}{\ce}$
@@
 \end{quote}
-\item $\segof{\ce} = \sseq{\acesplicedt{m'_i}{n'_i}}{\nty} \cup \sseq{\acesplicede{m_i}{n_i}{\ctau_i}}{\nexp}$
+\item $\segof{\ce} = \sseq{\acesplicedt{m'_i}{n'_i}}{\nty} \cup$\\ $\quad\sseq{\acesplicede{m_i}{n_i}{\ctau_i}}{\nexp}$
 \item \textbf{(Typing 2)} $\sseq{
@@
 \end{quote}
-\item \textbf{(Typing 4)} $\sseq{
+\item \textbf{(Typing 4)} ~\\ $\sseq{
   \expandsU
@@
 \end{quote}
-\item (\textbf{Capture Avoidance}) $e = [\sseq{\tau'_i/t_i}{\nty}, \sseq{e_i/x_i}{\nexp}]e'$ for some variables $\sseq{t_i}{\nty}$ and $\sseq{x_i}{\nexp}$, and $e'$
+\item (\textbf{Capture Avoidance}) ~\\$e = [\sseq{\tau'_i/t_i}{\nty}, \sseq{e_i/x_i}{\nexp}]e'$ for some variables $\sseq{t_i}{\nty}$ and $\sseq{x_i}{\nexp}$, and $e'$
     \begin{quote}
@@
     \end{quote}
-\item (\textbf{Context Independence}) $\mathsf{fv}(e') \subset \sseq{t_i}{\nty} \cup \sseq{x_i}{\nexp}$
+\item (\textbf{Context Independence}) ~\\$\mathsf{fv}(e') \subset \sseq{t_i}{\nty} \cup \sseq{x_i}{\nexp}$
     \begin{quote}
@@
 
-The proof, which involves auxiliary lemmas about the decomposition of proto-types and proto-expressions, is given in the supplement.
-
-Notice that we were able to state the hygiene properties (\textbf{Capture Avoidance} and \textbf{Context Independence}) without needing a notion of alpha-equivalence of source terms, as is the case in the typical formal accounts of hygiene \cite{Kohlbecker86a,DBLP:conf/popl/Adams15,DBLP:conf/popl/ClingerR91,DBLP:journals/lisp/DybvigHB92,DBLP:conf/esop/HermanW08,Herman10:Theory}.
+% The proof, which involves auxiliary lemmas about the decomposition of proto-types and proto-expressions, is given in the supplement.
+
+Notice that we were able to state the hygiene properties (\textbf{Capture Avoidance} and \textbf{Context Independence}) without needing a notion of alpha-equivalence of source terms, as in typical formal accounts of hygiene \cite{Kohlbecker86a,DBLP:conf/popl/Adams15,DBLP:conf/popl/ClingerR91,DBLP:journals/lisp/DybvigHB92,DBLP:conf/esop/HermanW08,Herman10:Theory}.
 Instead, we used standard notions of capture avoiding substitution and free variables combined with the context disjointness conditions in the rules above.
 This is possible only because we keep track of spliced terms explicitly in the proto-expansion, rather than going straight to the final expansion.
-In fact, doing so is critical for TLMs -- there is no notion of alpha-conversion for partially parsed terms, so any notion of hygiene that relies on this notion would be inapplicable.
-
-\vspace{-5px}
-\subsubsection{spTLM Reasoning Principles} Finally, the following theorem summarizes the abstract reasoning principles available to programmers when applying an spTLM.
+%In fact, doing so is critical for TLMs -- there is no notion of alpha-conversion for partially parsed terms, so any notion of hygiene that relies on this notion would be inapplicable.
+
+The reasoning principles theorem for pattern TLMs is in the supplement.
+The key clause establishes that the hypotheses generated by the TLM application form are exactly the union of the hypothesis generated by the spliced patterns.
+% No additional hypotheses can be produced by the TLM.
+
+In the full calculus, which supports parametric TLMs, the context independence clause allows reference to the variables standing for parameters.
+
+% The following theorem summarizes the abstract reasoning principles available to programmers when applying a pattern TLM.
 Most of the labeled clauses are analagous to those described above, so we omit their descriptions.
-% \begingroup
-% \def\thetheorem{\ref{thm:spTLM-Typing-Segmentation}}
-\begin{theorem}[spTLM Reasoning Principles]
-% \label{thm:spTLM-Typing-Segmentation}
-If $\patExpands{\upctx}{\uPhi}{\utsmap{\tsmv}{b}}{p}{\tau}$ where $\uDelta=\uDD{\uD}{\Delta}$ and $\uGamma=\uGG{\uG}{\Gamma}$ then all of the following hold:
-\begin{enumerate}[noitemsep,nolistsep]
-        \item (\textbf{Typing 1}) $\uPhi=\uPhi', \uPhyp{\tsmv}{a}{\tau}{\eparse}$ and $\patType{\pctx}{p}{\tau}$
-        \item $\encodeBody{b}{\ebody}$
-        \item $\evalU{\eparse(\ebody)}{\aein{\mathtt{SuccessP}}{\ecand}}$
-        \item $\decodeCEPat{\ecand}{\cpv}$
-        \item (\textbf{Segmentation}) $\segOK{\segof{\cpv}}{b}$
-        \item $\segof{\cpv} = \sseq{\acesplicedt{n'_i}{m'_i}}{\nty} \cup \sseq{\acesplicedp{m_i}{n_i}{\ctau_i}}{\npat}$
-        \item (\textbf{Typing 2}) $\sseq{
-              \expandsTU{\uDelta}
-              {
-                \parseUTypF{\bsubseq{b}{m'_i}{n'_i}}
-              }{\tau'_i}
-            }{\nty}$ and $\sseq{\istypeU{\Delta}{\tau'_i}}{\nty}$
-        \item (\textbf{Typing 3}) $\sseq{
-          \cvalidT{\emptyset}{
-            \tsceneUP
-              {\uDelta}{b}
-          }{
-            \ctau_i
-          }{\tau_i}
-        }{\npat}$ and $\sseq{\istypeU{\Delta}{\tau_i}}{\npat}$
-        \item (\textbf{Typing 4}) $\sseq{
-          \patExpands
-            {\uGG{\uG_i}{\pctx_i}}
-            {\uPhi}
-            {\parseUPatF{\bsubseq{b}{m_i}{n_i}}}
-            {p_i}
-            {\tau_i}
-        }{\npat}$  and $\sseq{\patType{\pctx_i}{p_i}{\tau_i}}{\npat}$
-      \item (\textbf{Visibility}) $\uG = \biguplus_{0 \leq i < \npat} \uG_i$ and $\Gamma = \bigcup_{0 \leq i < \npat} \pctx_i$
-        \begin{quote}
-          \begin{grayparbox}
-          The hypotheses generated by the TLM application are exactly those generated by the spliced patterns.
-          \end{grayparbox}
-        \end{quote}
-\end{enumerate}
-\end{theorem}
-\vspace{-3px}
+% % \begingroup
+% % \def\thetheorem{\ref{thm:spTLM-Typing-Segmentation}}
+% \begin{theorem}[spTLM Reasoning Principles]
+% % \label{thm:spTLM-Typing-Segmentation}
+% If $\patExpands{\upctx}{\uPhi}{\utsmap{\tsmv}{b}}{p}{\tau}$ where $\uDelta=\uDD{\uD}{\Delta}$ and $\uGamma=\uGG{\uG}{\Gamma}$ then all of the following hold:
+% \begin{enumerate}[noitemsep,nolistsep]
+%         \item (\textbf{Typing 1}) $\uPhi=\uPhi', \uPhyp{\tsmv}{a}{\tau}{\eparse}$ and $\patType{\pctx}{p}{\tau}$
+%         \item $\encodeBody{b}{\ebody}$
+%         \item $\evalU{\eparse(\ebody)}{\aein{\mathtt{SuccessP}}{\ecand}}$
+%         \item $\decodeCEPat{\ecand}{\cpv}$
+%         \item (\textbf{Segmentation}) $\segOK{\segof{\cpv}}{b}$
+%         \item $\segof{\cpv} = \sseq{\acesplicedt{n'_i}{m'_i}}{\nty} \cup \sseq{\acesplicedp{m_i}{n_i}{\ctau_i}}{\npat}$
+%         \item (\textbf{Typing 2}) $\sseq{
+%               \expandsTU{\uDelta}
+%               {
+%                 \parseUTypF{\bsubseq{b}{m'_i}{n'_i}}
+%               }{\tau'_i}
+%             }{\nty}$ and $\sseq{\istypeU{\Delta}{\tau'_i}}{\nty}$
+%         \item (\textbf{Typing 3}) $\sseq{
+%           \cvalidT{\emptyset}{
+%             \tsceneUP
+%               {\uDelta}{b}
+%           }{
+%             \ctau_i
+%           }{\tau_i}
+%         }{\npat}$ and $\sseq{\istypeU{\Delta}{\tau_i}}{\npat}$
+%         \item (\textbf{Typing 4}) $\sseq{
+%           \patExpands
+%             {\uGG{\uG_i}{\pctx_i}}
+%             {\uPhi}
+%             {\parseUPatF{\bsubseq{b}{m_i}{n_i}}}
+%             {p_i}
+%             {\tau_i}
+%         }{\npat}$  and $\sseq{\patType{\pctx_i}{p_i}{\tau_i}}{\npat}$
+%       \item (\textbf{Visibility}) $\uG = \biguplus_{0 \leq i < \npat} \uG_i$ and $\Gamma = \bigcup_{0 \leq i < \npat} \pctx_i$
+%         \begin{quote}
+%           \begin{grayparbox}
+%           The hypotheses generated by the TLM application are exactly those generated by the spliced patterns.
+%           \end{grayparbox}
+%         \end{quote}
+% \end{enumerate}
+% \end{theorem}
+% \vspace{-3px}
 % \begin{proof} 
@@
 
-\subsection{Retrospective} Let us step back and briefly review what we have accomplished so far.
+% \subsection{Retrospective} Let us step back and briefly review what we have accomplished so far.
 First, we defined an entirely standard expanded language (Sec.
@@
 
-\vspace{-4px}
-\newcommand{\pTLMsSec}{Parametric TLMs (pTLMs)}
-\section{\protect\pTLMsSec}
-\label{sec:ptsms}
+\vspace{-6px}
+\newcommand{\rwSec}{Related Work}
+\section{\protect\rwSec}
+\label{sec:existing-approaches}
+\subsection{Syntax Definition Systems}\label{sec:syntax-dialects-intro}
+\vspace{-3px}
+One approach available to library providers seeking to introduce new literal forms is to use a syntax definition system to construct a library-specific \emph{syntax dialect}: a new syntax definition that extends the syntax definition given in the language definition with new forms, including literal forms.% For example, Ur/Web's syntax (Figure 1) is a library-specific dialect of Ur's syntax \cite{conf/pldi/Chlipala10,conf/popl/Chlipala15}.
+
+ % Such dialects are sometimes qualitatively taxonomized as amongst the ``domain-specific language'' for this reason \cite{fowler2010domain}.
+%Syntactic cost is often assessed qualitatively \cite{green1996usability}, though quantitative metrics can be defined.
+% Syntax definition systems , which we will discuss in Sec.
+\ref{sec:syntax-dialects}, have simplified the task of defining ``library-specific'' (a.k.a.
+``domain-specific'') syntax dialects like Ur/Web, and have thereby contributed to their ongoing proliferation.
+
+% Many have argued that a proliferation of syntax dialects constructed using these tools is harmless or even desirable, because programmers can simply choose the right syntax dialect for each job at hand \cite{journals/stp/Ward94}.
+However, we argue that in fact
+There are hundreds of syntax definition systems of various design.
+Notable examples include grammar-oriented systems like Camlp4 \cite{ocaml-manual}, Copper \cite{conf/gpce/WykS07,conf/pldi/SchwerdfegerW09} and Sugar*/SoundExt \cite{erdweg2011sugarj,erdweg2013framework,conf/icfp/LorenzenE13,conf/popl/LorenzenE16}, as well as parser combinator systems \cite{Hutton1992d}.
+The parsers generated by these systems can be invoked to preprocess program text in various ways, e.g.
+by invoking them from within a build script, by using a preprocessor-aware build system (e.g.
+\li{ocamlbuild}), or via language-integrated preprocessing directives, e.g.
+Racket's \li{#lang} directive or its reader macros \cite{Flatt:2012:CLR:2063176.2063195}, or the import mechanism of SugarJ \cite{erdweg2011sugarj}.% the directives of language-integrated ``mixfix'' systems \cite{wieland2009parsing,missura1997higher,5134} like Coq \cite{Coq:manual}.)
+%library-specific (a.k.a. ``domain-specific'')
+
+The problem was described by example in Sec.
+\ref{sec:intro}: these systems make it difficult to reason abstractly.
+Let us reitierate more generally, before discussing a few systems that are exceptional along some strict subset of these dimensions:
+\begin{enumerate}[noitemsep,nolistsep,leftmargin=10pt]
+\item \textbf{Responsibility}: Clients using a combined syntax dialect cannot easily determine which constituent extension is responsible for a given form, whereas TLMs have explicit names which follow the usual scoping conventions.
+
+Moreover, there can be syntactic conflicts because multiple extensions can claim responsibility for the same form.
+TLMs sidestep these complexities entirely because the context-free syntax of the language is fixed, and composition is via splicing rather than direct combination.
+\item \textbf{Segmentation}: Clients of a syntax dialect cannot accurately determine which segments of the program text appear directly in the desugaring.
+In contrast, TLM clients can inspect the inferred segmentation (communicated via secondary notation.)
+\item \textbf{Capture}: Unlike TLM clients, clients of a syntax dialect cannot be sure that spliced terms are capture avoiding. 
+\item \textbf{Context Dependence}: Similarly, clients cannot be sure that the desugaring is context independent.
+Indeed, without a method to pass in parameters (Sec.
+\ref{sec:ptsms}), achieving strict context independence would be impractical.
+\item \textbf{Typing}: Clients of a syntax dialect cannot reason abstractly about what type a desugaring has.
+In contrast, TLM clients can determine the type of any expansion by referring to the parameter and type declarations on the TLM definition, and nothing else.
+The inferred segmentation also gives types for each spliced expression or pattern.
+%This information can be communicated upon request by the type inspection service of a program editor.
+\end{enumerate}
+
+% The first problem with using a syntax definition systems to directly extend the context-free syntax of a language is that clients cannot always deterministically combine the resulting extended syntax dialects when they want to use the new forms that they define together, i.e.
+there can be syntactic conflicts.
+This is a significant problem because large programs use many separately developed libraries \cite{DBLP:conf/sac/LammelPS11}.
+
+An extensible syntax definition system that has confronted the problem of \textbf{Responsibility} (but not the other problems) is Copper \cite{conf/pldi/SchwerdfegerW09}.
+Copper integrates a modular grammar analysis that guarantees that determinism is conserved when extensions of a certain restricted class are combined.
+The caveat is that the constituent extensions must prefix all newly introduced forms with marking tokens drawn from disjoint sets.
+To be confident that the marking tokens used are disjoint, providers must base them on the domain name system or some other coordinating entity.
+Because the mechanism operates at the level of the context-free grammar, it is difficult for the client to define scoped abbreviations for these verbose marking tokens.
+TLMs can be abbreviated (Sec.
+\ref{sec:ptsms}).
+
+Some programming languages, notably including theorem provers like Coq \cite{Coq:manual} and Agda \cite{norell2007towards}, support ``mixfix'' notation directives \cite{wieland2009parsing,missura1997higher,5134}.
+Many of these systems enforce capture avoidance and application-site context independence \cite{5134,DBLP:conf/gpce/TahaJ03,Coq:manual,DBLP:conf/ifl/DanielssonN08}.
+The problem is that mixfix notation requires a fixed number of sub-trees, e.g.
+\li{if _ then _ else _}.
+Coq has some limited extensions for list-like literals \cite{Coq:manual}.
+These systems cannot express the example literal forms from this paper, because they can have any number of spliced terms.
+
+The work of \citet{conf/icfp/LorenzenE13,conf/popl/LorenzenE16} introduces SoundExt, a grammar-based syntax extension system where extension providers can equip their new forms with derived typing rules.
+The system then attempts to automatically verify that the expansion logic (expressed using a rewrite system, rather than an arbitrary function) is sound relative to these derived rules.
+TLMs differ in several ways.
+First, as already discussed, we leave the context-free syntax fixed, so different TLMs cannot conflict.
+Second, SoundExt does not enforce hygiene, i.e.
+expansions might depend on the context and intentionally induce capture.
+Similarly, there is no abstract segmentation discipline.
+A client can only indirectly reason about binding (but not segmentation) by inspecting the derived typing rules.
+Unlike TLMs, SoundExt supports type-dependent expansions \cite{conf/popl/LorenzenE16}.
+The trade-off is that TLMs can generate expansions, and therefore segmentations, even when the program is ill-typed.
+Another important distinction is that TLMs rely on proto-expansion validation, rather than verification as in SoundExt.
+The trade-off is that TLMs do not require that the expansion logic be written using a restricted rewriting system, nor does the system require a fully mechanized language definition.
+Finally, there is no clear notion of ``partial application'' in SoundExt or other syntax definition systems.
+
+%In other words, encountering an unfamiliar derived form has made it difficult for the programmer to maintain the usual \emph{type discipline} and \emph{binding discipline}.
+%Compelling the programmer to examine the desugaring directly defeat the purpose of defining the derived form -- decreasing cognitive cost.
+Indeed, it substantially increases cognitive cost.
+
+% In contrast, when a programmer encounters, for example, a function call like the call to \li{read_data} on Line 3, the analagous questions can be answered by following clear protocols that become ``cognitive reflexes'' after sufficient experience with the language, even if the programmer has no experience with the library defining \li{read_data}:
+% \begin{enumerate}
+% \item The language's syntax definition determines that \li{read_data(a)} is an expression of function application form.
+% \item Similarly, \li{read_data} and \li{a} are definitively expressions of variable form.
+% \item The variable \li{a} can only refer to the binding of \li{a} on Line 1.
+% \item The variable \li{w} can be renamed without knowing anything about the values that \li{read_data} and \li{a} stand for.
+% \item The type of \li{x} can be determined to be \li{B} by determining that the type of \li{read_data} is \li{A -> B} for some \li{A} and \li{B}, and checking that \li{a} has type \li{A}.
+Nothing else needs to be known about the values that \li{read_data} and \li{a} stand for.
+In Reynolds' words \cite{B304}:
+% \begin{quote}
+% \emph{Type structure is a syntactic discipline for enforcing levels of abstraction.}
+% \end{quote}
+% \end{enumerate}
+
+\vspace{-6px}
+\subsection{Term Rewriting Systems}
+\vspace{-3px}
+Another approach -- and the approach that TLMs are rooted in -- is to leave the context-free syntax of the language fixed, and instead contextually rewrite existing literal forms.
+
+OCaml's textual syntax now includes \emph{preprocessor extension (ppx) points} used to identify terms that some external term rewriting preprocessor must rewrite \cite{ocaml-manual}.
+We could  mark a string literal as follows:
+\vspace{-3px}
+\begin{lstlisting}[numbers=none]
+    [%xml "SSTR<h1>Hello, {[first_name]}!</h1>ESTR"]
+\end{lstlisting}
 \vspace{-2px}
-Simple TLMs operate at a single specified type and the proto-expansions they generate must be closed (cf.
-the final premise of Rule \textsc{ee-ap-setsm} in the Sec.
-\ref{sec:s-TLM-ap}.) While this suited our purposes in the previous section because we intend $\miniVersePat$ to communicate the essential concepts, it is not realistic for an ML-like language to impose these limitations.
-This section introduces \emph{parametric TLMs}.
-Parametric TLMs can be defined over a type- and module-parameterized family of types, and the proto-expansions they generate can refer to supplied type and module parameters.
-
-\vspace{-4px}
-\subsection{Parametric TLMs By Example}\label{sec:ptsms-by-example}
+More than one applied preprocessor might recognize this annotation (there are, in practice, many XML/HTML libraries), so the problems of \textbf{Responsibility} comes up.
+It is also impossible to reason abstractly about \textbf{Segmentation}, \textbf{Capture}, \textbf{Context Dependence} and \textbf{Typing} because the code that the preprocessor generates is unconstrained.% For these reasons, users of this mechanism warn that they should be used sparingly.\todo{citation}
+
+Term-rewriting macro systems are language-integrated local term rewriting systems that require that the client explicitly apply the intended rewriting, implemented by a macro, to the term that is to be rewritten.
+This addresses the issue of \textbf{Responsibility}.
+However, unhygienic, untyped macro systems, like the earliest variants of the Lisp macro system \cite{Hart63a}, Template Haskell \cite{SheardPeytonJones:Haskell-02} and GHC's quasiquotation system \cite{mainland2007s} (which is based on Template Haskell), do not allow clients to reason abstractly about the remaining issues, again because the expansion that they produce is unconstrained.
+(It is not enough that with Template Haskell / GHC quasiquotation, the generated expansion is typechecked -- to satisfy the \textbf{Typing} criterion, it must be possible to reason abstractly about \emph{what the type of the generated expansion is}.)
+
+\emph{Hygienic} macro systems prevent, or abstractly account for \cite{DBLP:conf/esop/HermanW08,Herman10:Theory}, \textbf{Capture}, and they enforce application-site \textbf{Context Independence} \cite{Kohlbecker86a,DBLP:conf/popl/Adams15,DBLP:conf/popl/ClingerR91,DBLP:journals/lisp/DybvigHB92}.
+The critical problem is that a standard hygiene discipline makes it impossible to repurpose string literal forms to introduce compositional literal forms at other types.
+Consider again our running XHTML example, which we might try to realize by applying a hygienic macro, \li{xml!}, to a string literal that the macro parses:
+\vspace{-3px}
+\begin{lstlisting}[numbers=none]
+    (xml! "SSTR<h1>Hello, {[first_name]}!</h1>ESTR")
+\end{lstlisting}
 \vspace{-2px}
-
-Consider the following Reason module type (a.k.a.
-signature), which specifies an abstract data type \cite{liskov1974programming,harper1997programming} of string-keyed polymorphic immutable dictionaries:
-\begin{lstlisting}[numbers=none]
-  module type DICT = {
-    type t('a); 
-    let empty : t('a); 
-    let extend : t('a) -> (string, 'a) -> t('a); 
-    /* ... */ 
-  };
-\end{lstlisting}
-Let us now define a TLM that is parametric over all implementations of this signature, \li{D : DICT}, and also over all choices of the range type, \li{'a}:
-\begin{lstlisting}[numbers=none]
-  syntax $dict (D : DICT) (type 'a) at D.t('a) by static {
-    fun(b : body) -> parse_result(proto_expr) => /* ... */
-  };
-\end{lstlisting}
-Assuming some module that implements this signature, \li{HashDict : DICT}, and some values \li{v1 : int} and \li{v2 : int}, we can apply \li{$dict} as follows:
-\begin{lstlisting}[numbers=none]
-  $dict HashDict int [| "key1"SURL => EURLv1SURL; EURL"key2"SURL => EURLv2 |]
-\end{lstlisting}
-Notice that the segmentation immediately reveals which punctuation is particular to this TLM and where the spliced key and value expressions appear.
-Because the context-free syntax of unexpanded terms is never modified, it is possible to reason modularly about syntactic determinism, i.e.
-we can reason above that \li{=>} does not appear in the follow set of unexpanded terms \cite{conf/pldi/SchwerdfegerW09}, so there can never be an ambiguity about where a key expression ends.
-
-If we will use the \li{HashDict} implementation ubiquitously, we can abbreviate the partial application of \li{$dict} to \li{HashDict}, resulting in a TLM that is parametric over only the type \li{'a}, as follows:
-\begin{lstlisting}[numbers=none]
-  syntax $dict' = $dict HashDict; 
-  $dict' int [| "key1"SURL => EURLv1SURL; EURL"key2"SURL => EURLv2 |];
-\end{lstlisting}
-The proto-expansion generated for the TLM application above might be:
-\begin{lstlisting}[numbers=none]
-  D.extend (D.extend D.empty (spliced<1; 6; string>, spliced<11; 12; 'a>)) 
-    (spliced<15; 20; string>, spliced<25; 26; 'a>)
-\end{lstlisting}
-The generated proto-expansion must truly be parametric, i.e.
-it must be valid for all modules \li{D : DICT} and types \li{'a}.
-It is only after proto-expansion validation that we substitute the actual parameters, here \li{HashDict} for \li{D} and \li{int} for \li{'a}, into the final expansion.
-Additionally, substitution occurs on the type annotations on references to spliced terms before recursively expanding those terms (otherwise, the expressions \li{v1} and \li{v2} above would need to be well-typed for all types, \li{'a}.)
-
-Module parameters are also useful for giving the expansion access to helper functions in a context-independent manner.
-For example, in Sec.
-\ref{sec:context-dependence} we discussed the problem of applying datatype constructors like \li{H1Element}, because they are introduced into the context as variables in most ML-like languages.
-With module parameters, it is possible to sidestep this problem by passing in a module containing the datatype constructors.
-Because this will be common in practice, we provide the shorthand \li{syntax $a at T using t1, (*...*), tn, X1, (*...*), Xm by static \{ ...
-\}} for the parameterization of \li{$a} by $n$ type variable parameters and $m$ module variable parameters, followed by the partial application of the given parameters.
-Additionally, for the type variable parameters that refer to datatypes, the constructors are bound automatically in the expansion.
-This explicit form of capture is reminiscent of work on explicit capture for distributed functions by \citet{DBLP:conf/ecoop/MillerHO14}.
-Although a bit cumbersome, asking TLM providers to make their intent explicit allows them to avoid having to reason about the computations that the parse function performs on syntax tree representations when refactoring code.
-Otherwise, proto-expansion representations would need to be built primitively into the binding structure of the language.
-We want to leave the semantics of OCaml alone when incorporating TLMs into Reason, so this would not be feasible.
-
-It is important to note that module parameters are accessible by expansions, but not by parse functions directly, because the applied module parameters will not have been dynamically instantiated when typed expansion occurs.
-We will discuss a distinct mechanism for providing helper functions to parse functions in Sec.
-\ref{sec:static-eval}.
-
+The expansion of this macro will fail the check for application-site context independence because \li{first_name} will appear as a free variable, in no way different from any other.
+The hygiene mechanism for TLMs addresses this problem by explicitly distinguishing spliced segments of the literal body in the proto-expansion.
+This also addresses the problem of \textbf{Segmentation} -- the segmentation abstractly communicates the fact that (only) \li{first_name} is a spliced \li{string} expression.
+
+Much of the research on macro systems has been for languages in the LISP tradition \cite{mccarthy1978history} that do not have rich static type structure.
+The formal macro calculus studied by \citet{DBLP:conf/esop/HermanW08} (which is not capable of expressing new literal forms, for the reasons just discussed) uses types only to encode the binding structure of the generated expansion (as discussed in Sec.
+\ref{sec:capture}).
+Research on typed \emph{staging macro systems} like MetaML \cite{Sheard:1999:UMS}, MetaOCaml \cite{DBLP:conf/flops/Kiselyov14} and MacroML \cite{ganz2001macros} is also not directly applicable to the problem of defining new literal forms -- the syntax tree of the arguments cannot be inspected at all (staging macros are used mainly for partial evaluation and performance-related reasons.)
+
+The Scala macro system is a hygienic macro system.
+Its ``black box'' macros support reasoning abstractly about \textbf{Typing} because type annotations constrain the macro arguments and the generated expansions, though the precise reasoning principles available are unclear because the Scala macro system has not been formally specified.
+The full calculus we have defined is the first detailed type-theoretic accounts of a typed, hygienic macro system of any design for an ML-like language, i.e.
+one with a rich static type system, support for pattern matching, type functions and ML-like modules.
+
+  Some languages, including Scala \cite{odersky2008programming}, build in \emph{string splicing} (a.k.a.
+\emph{string interpolation}) forms, or similar but more general \emph{fragmentary quotation forms} \cite{conf/icfp/Slind91}, e.g.
+SML/NJ.
+These designate a particular delimiter to escape out into the expression language.
+The problem with using these together with macros as vehicles to introduce literal forms at various other types is 1) there is no ``one-size-fits-all'' escape delimiter, and 2) typing is problematic because every escaped term is checked against the same type.
+In the HTML example, we have splicing at two different types using two different delimiters.
+These forms also cannot appear in patterns.%In general, e.g.
+when defining syntax for a programming language with many sorts of terms, the most appropriate choice of delimiters might depend on where each spliced term appears.
+
+This brings us back to the most closely related work, that of \citet{TSLs} on \emph{type-specific languages} (TSLs).
+Like simple expression TLMs (Sec.
+\ref{sec:setsms}), TSLs allow library providers to programmatically control the parsing of expressions of generalized literal form.
+With TSLs, parse functions are associated directly with nominal types and invoked according to a bidirectionally typed protocol.
+In contrast, TLMs are separately defined and explicitly applied.
+Accordingly, different TLMs can operate at the same type, and can operate at any type, including structural types.
+In a subsequent short paper, \citet{sac15} suggested explicit application of simple expression TLMs  also in a bidirectional typed setting \cite{Pierce:2000:LTI:345099.345100}, but this paper did not have any formal content.
+With TLMs, it is not necessary for the language to be bidirectionally typed (see Sec.
+\ref{sec:typing-e} on type inference).% It should be possible to implicitly invoke TLMs based on the expected type in situations where the inference engine had enough information, but we leave this as future work.
+
+Perhaps most importantly, the metatheory presented by \citet{TSLs} establishes only that generated expansions are of the expected type (i.e.
+a variant of the Typed Expression Expansion theorem from Sec.
+\ref{sec:s-metatheory}.) It does not establish the remaining abstract reasoning principles that have been the major focus of this paper.
+In particular, there is no formal hygiene theorem and indeed the formal system in the paper does not correctly handle substitution or capture avoidance, issues we emphasized because they were non-obvious in Sec.
+5.
+Moreover, the TLM does not guarantee that a valid segmentation will exist, nor associate types with segments.
+
+Finally, the prior work did not consider pattern matching, type functions, ML-style modules, parameters or static evaluation. This paper addresses all of these.
+
+% These existing forms normally have other meanings, so this can be confusing \cite{pane1996usability}.
+
+\vspace{-6px}
+\newcommand{\discussionSec}{Discussion}
+\section{\protect\discussionSec}
+\label{sec:discussion}
+\label{sec:conclusion}
+\vspace{-2px}
+
+The importance of specialized notation as a ``tool for thought'' has long been recognized \cite{DBLP:journals/cacm/Iverson80}.
+According to Whitehead, a good notation ``relieves the brain of unnecessary work'' and ``sets it free to concentrate on more advanced problems'' \cite{cajori1928history}, and indeed, advances in mathematics, science and programming have often been accompanied by new notation.
+
+Of course, this desire to ``relieve the brain of unnecessary work'' has motivated not only the syntax but also the semantics of languages like ML and Scala -- these languages maintain a strong type and binding discipline so that programmers, and their tools, can hold certain implementation details abstract when reasoning about program behavior.
+In the words of \citet{B304}, ``type structure is a syntactic discipline for enforcing levels of abstraction.''
+
+Previously, these two relief mechanisms were in tension---mechanisms  that allowed programmers to express new notation would obscure the type and binding structure of the program text.
+TLMs resolve this tension for the broad class of literal forms that generalized literal forms subsume.
+This class includes all of the examples enumerated in Sec.
+\ref{sec:intro} (up to the choice of outermost delimiter), the case studies detailed in this paper and in the supplement, and the examples collected from the empirical study by \citet{TSLs}.
+
+Of course, not all possible literal notation will prove to be in good taste.
+%TLMs leave the burden of establishing the value of any particular literal notation to individual library providers, rather than to the language designer.
+The reasoning principles that TLMs provide, which are the primary contributions of this paper, allow clients to "reason around" poor literal designs, using principles analagous to those already familiar to programmers in languages like ML and Scala.
+% We must leave a detailed empirical evaluation of the impact of particular TLMs on various quality attributes, like programmer productivity and program comprehensibility, as future work.
+% Our own implementation efforts have focused on an emerging alternative front-end for OCaml called Reason, but Scala supports analagous features \cite{conf/oopsla/AminRO14} and would also be a suitable host for the TLM mechanism.
+The intended audience for this paper is language designers seeking a reasonable solution to the problem of user-defined literal notation.% We also plan to implement TLMs into \li{typy}, a typed functional language embedded into Python as a library \cite{gpce/Omar16}.%We have evaluated this claim by stating the available abstract reasoning principles formally and proving that they hold for the calculi that we have described.
+
+% The examples in this paper are written in an open source alternative syntactic front-end for OCaml called Reason because we are incorporating TLMs into Reason.
+However, the essential ideas are not Reason-, OCaml- or ML-specific -- it should be possible to adapt TLMs to other languages that take a similar approach to types and binding (e.g.
+Haskell, Scala and others.) Languages that have a disciplined binding structure but that lack a rich static type structure, e.g.
+standard Racket, would also benefit from TLMs -- they can apply these results by deploying the usual trick of viewing the language as statically ``{unityped}'' \cite{pfpl,scott1980lambda}.
+
+% The intended audience for this paper is language designers who want to decentralize control over literal forms but seek a detailed understanding, rooted in the first principles of type theory, of a reasonable mechanism by which to achieve that goal.
+
+%(To summarize Sec.
+\ref{sec:existing-approaches}, the closest existing typed, hygienic macro system -- the Scala macro system \cite{ScalaMacros2013} -- is not formally specified, the work of \cite{DBLP:conf/esop/HermanW08} uses types only to reason abstractly about binding, and the specification of TSLs in the work of \citet{TSLs} did not formalize hygiene nor handle these advanced language features.)
+
+% Haskell's \li{do}-notation at types with monadic structure cannot be expressed directly using TLMs because it needs to purposefully induce capture in spliced sub-terms, and TLMs prevent capture.
+This limitation could be resolved by allowing identifiers that appear in literal bodies to be explicitly designated as bound within spliced expressions in the segmentation.
+The prior work by \citet{DBLP:conf/esop/HermanW08} considered a similar approach in a setting where the locations of sub-terms is known ahead of time (building on the \emph{tree locations} of \citet{gorn1965explicit}, which are conceptually similar to our spliced segment locations.)  We leave as future work the details of this proposal and a consideration of whether this additional expressive power would justify the increased reasoning complexity for clients (it is presently the author's opinion that it would not.)
+
+A correct parse function never returns an encoding of a proto-expansion that fails validation given well-typed splices, but this invariant cannot be enforced by the ML type system.
+% Using a proof assistant, it would be possible to verify that
+% a parse function generates only encodings of valid proto-expansions. 
+Under a
+richer type system, the return type of the parse function itself could be refined so as to
+enforce this invariant intrinsically.
+This problem -- of typed first-class typed term representations -- has been studied in a variety of settings, e.g.
+in MetaML \cite{Sheard:1999:UMS} and in the modal logic tradition \cite{DBLP:conf/popl/DaviesP96}.
+% We leave integration of these approaches into the TLM mechanism as future work.
+ Our present efforts aim to leave the semantics of OCaml unchanged.
+
+%Another topic of interest has to do with intentional capture. 
+Another direction has to do with automated refactoring.
+The unexpanded language does not come with context-free notions of renaming and substitution.
+However, given a segmentation, it should be possible to ``backpatch'' refactorings into literal bodies.
+Recent work by \citet{wand2017inferring} on tracking bindings ``backwards'' from an expansion to the source program is likely relevant.
+The challenge is that the TLM's splicing logic might not be invariant to refactorings.
+
+At several points in the paper, we allude to editor integration.
+However, several important questions having to do with TLM-specific syntax highlighting, incremental parsing and error recovery \cite{graham1979practical} remain to be considered, and indeed these are our biggest remaining implementation challenges.
+% Another interesting direction would be to generalize TLMs to support non-textual notation in the setting of a structure editor.
+The semantic foundations for structure editors proposed recently by \citet{DBLP:conf/popl/OmarVHAH17} may guide such an effort.
+
+% Let us conclude by considering the oft-uttered phrase ``syntax doesn't matter'', which is generally taken to mean that syntactic concerns are orthogonal to various more important semantic concerns.
+Our hope is that the reader will leave this paper with an understanding that this is a rather simplistic proclamation, because semantic concerns are relevant to notation design as well!
+% We hope that this work will lead to more ``reasonable'' variants of other syntactic conveniences, e.g.
+the ubiquitious ``deriving'' clauses on datatype definitions.
+
+% Other directions for future work are given in the supplement\todo{do this?}.
+
+% Another interesting question is what this mechanism would look like in the setting of a projectional structure editor, i.e.
+one where the syntax is not textual but rather tree-shaped and projected in an interative manner to the programmer.
+This paper taken together with work by \citet{Omar:2012:ACC:2337223.2337324,DBLP:conf/popl/OmarVHAH17} can serve to guide such an effort.
+
+% To conclude, TLMs give substantial syntactic control to library providers while leaving programmers with strong abstract reasoning principles.
+We believe TLMs therefore occupy a ``sweet spot'' in the  design space.
+%This paper serves to guide those who hope to incorporate TLMs into their future language designs.
+
+%% Acknowledgments
+\begin{acks}                            %% acks environment is optional
+                                        %% contents suppressed with 'anonymous'
+  %% Commands \grantsponsor{<sponsorID>}{<name>}{<url>} and
+  %% \grantnum[<url>]{<sponsorID>}{<number>} should be used to
+  %% acknowledge financial support and will be used by metadata
+  %% extraction tools.
+  This material is based upon work supported by the
+  \grantsponsor{GS100000001}{National Science
+    Foundation}{http://dx.doi.org/10.13039/100000001} under Grant
+  No.~\grantnum{GS100000001}{nnnnnnn} and Grant
+  No.~\grantnum{GS100000001}{mmmmmmm}.  Any opinions, findings, and
+  conclusions or recommendations expressed in this material are those
+  of the author and do not necessarily reflect the views of the
+  National Science Foundation.
+\end{acks}
+
+\clearpage
+%% Bibliography
+\bibliography{../../../papers/research}
+
+\begin{comment}\appendix
 \vspace{-4px}
@@
 \end{theorem}
-
-\vspace{-6px}
-\newcommand{\staticEvalSec}{Static Evaluation}
-\section{\protect\staticEvalSec}
-\label{sec:static-eval}
-\vspace{-2px}
-
-There is one more major impracticality that we can now address.
-So far, we have assumed that parse functions are closed expanded expressions.
-This assumption simplifies matters formally, but in practice it is quite difficult, because this means that TLM definitions cannot themselves access any libraries nor use TLMs internally.
-To address this problem, we introduce a \emph{static environment}.
-
-\vspace{-4px}
-\subsection{The Static Environment}
-\vspace{-2px}
-Figure \ref{fig:static-module-example} shows an example of a module, \li{ParserCombos}, that defines a number of parser combinators following \citet{Hutton1992d}.
-The \li{static} qualifier indicates that this module is bound for use only within similarly qualified values, including in particular parse functions of subsequent TLM definitions.
-\begin{figure}[h]
-\vspace{-10px}
-\begin{lstlisting}
-  static module ParserCombos = { 
-    type parser('c, 't) = list('c) -> list('t, list('c));
-    let alt : parser('c, 't) -> parser('c, 't) -> parser('c, 't);
-    /* ... */
-  };
-  syntax $a at t by static { fun(b) => ParserCombos.alt /* OK ... */ };
-  static val y = ParserCombos.alt /* OK ... */;
-  val z = /* ... neither ParserCombos nor y are available here */;
-\end{lstlisting}
-\vspace{-10px}
-\caption{Binding static modules and values for use within parse functions.}
-\label{fig:static-module-example}
-\vspace{-10px}
-\end{figure}
-
-The values that arise when the static phase runs do not persist from ``compile-time'' to ``run-time'', so we do not need a full staged computation system, e.g.
-as described by \citet{Taha99multi-stageprogramming:}.
-Instead, a sequence of static bindings operates like a read-evaluate-print loop (REPL) scoped according to the program structure, in that each static expression is evaluated immediately and the evaluated values are tracked by a \emph{static environment}.
-The static environment is discarded after typed expansion.
-
-A language designer might choose to restrict the external effects available to static terms in some way, e.g.
-to ensure deterministic builds.
-It might also be helpful to restrict mutable state shared between TLMs to prevent undesirable TLM application order dependencies.
-On the other hand, these features, if used for the purposes of caching, might speed up typed expansion.
-These are orthogonal design decisions.
-
-\vspace{-6px}
-\subsection{Applying TLMs Within TLM Definitions}\label{sec:tsms-for-tsms}
-TLMs and TLM abbreviations can also be qualified with the \li{static} keyword, which marks them for use within subsequent static expressions and patterns.
-Let us consider some examples of relevance to TLM providers.
-
-\vspace{-6px}
-\subsubsection{Quasiquotation}
-TLMs must construct values of type \li{proto_expr} or \li{proto_pat}.
-Constructing values of these types explicitly can have high syntactic cost.
-To decrease this cost, we can define TLMs that provide support for \emph{quasiquotation syntax} similar to that built in to languages like Lisp \cite{Bawd99a} and Scala \cite{shabalin2013quasiquotes}.
-The following TLM defines quasi-quotation for encodings of proto-expressions:
-\begin{lstlisting}[numbers=none]
-  static syntax $proto_expr at proto_expr by static { /* ... */ };
-\end{lstlisting}
-For example, \li{$proto_expr `SQTg x ^EQTx`} might have expansion \li{App(App(Var "g", Var "x"), x)}.
-Notice that prefixing a variable (or parenthesized expression) with \li{^} serves to splice in its value, which here must be of type \li{proto_expr} (though in other syntactic positions, it might be \li{proto_typ}.) This is also known as \emph{anti-quotation}.
-
-% A similar approach can be taken for working with encodings of terms of other languages (e.g. when writing an interpretter or compiler in Reason.)
-
-\vspace{-6px}
-\subsubsection{Parser Generators}
-Abstractly, a grammar-based parser generator is a module matching the signature \li{PARSEGEN} defined below:
-\begin{lstlisting}
-  module type PARSEGEN = { 
-    type grammar('a);
-    /* ... operations on grammars ... */
-    val generate : grammar('a) -> (body -> parse_result('a));
-  };
-\end{lstlisting}
-
-Rather than constructing a grammar equipped with semantic actions using the associated operations (whose specifications are elided in \li{PARSEGEN}), we wish to use a syntax for context-free grammars that follows standard conventions.
-We can do so by defining a static parametric TLM:
-\begin{lstlisting}[numbers=none]
-  static syntax $grammar(P : PARSEGEN)(type 'a) at P.grammar('a) /*...*/
-\end{lstlisting}
-To support splicing, we need non-terminals that recognize unexpanded terms and produce the corresponding splice references, rather than the AST itself.
-This requires that the parser generator keep track of location information (as most production-grade parser generators already do for error reporting.) A more detailed example of this mechanism being used to define a TLM for a modular encoding of regular expressions is given in the supplement.% For spliced expressions, this non-terminal would need to be a family of non-terminals indexed by a value of type \li{proto_typ}.
-
-A grammar containing such non-terminals can serve as a \emph{summary specification} -- a human can simply be take this grammar as a specification of what the segmentation will be for every recognized string, rather than relying on an editor to communicate this information.
-The associated semantic actions can be held abstract as long as the system performs a simple check to ensure that the proto-expansion does mention each spliced expression from the corresponding production.
-
-% \begin{figure}[h!]
-% \vspace{-5px}
-% \begin{lstlisting}[deletekeywords={as}]
-% syntax $rx(R : RX) at R.t by static 
-%   P.generate ($grammar P proto_expr {|SHTML #\label{line:rx_parse_fn_start}#
-%     start <- ""
-%       EHTMLfn () => $proto_expr `SCSSR.EmptyECSS`SHTML
-%     start <- "(" start ")"
-%       EHTMLfn e => eSHTML
-%     token str_tok #\label{line:str_tok_start}#
-%       EHTMLRU.parse "SSTR[^(@$]+ESTR" /* cannot use $rx within its own def */SHTML #\label{line:str_tok_end}#
-%     start <- str_tok
-%       EHTMLfn s => $proto_expr `SCSSR.Str %(ECSSstr_to_proto_lit sSCSS)ECSS`SHTML
-%     start <- start start
-%       EHTMLfn e1 e2 => $proto_expr `SCSSR.Seq (%ECSSe1SCSS, %ECSSe2SCSS)ECSS`SHTML
-%     start <- start "|" start 
-%       EHTMLfn e1 e2 => $proto_expr `SCSSR.Or (%ECSSe1SCSS, %ECSSe2SCSS)ECSS`SHTML
-%     start <- start "*"
-%       EHTMLfn e => $proto_expr `SCSSR.Star %ECSSe`SHTML
-
-%     using EHTMLspliced_uexp ($proto_typ `SCSSR.tECSS`) SHTML as spliced_rx #\label{line:splicede_using}#
-%     start <- "${" spliced_rx "}" #\label{line:splicing-start}#
-%       EHTMLfn e => eSHTML
-
-%     using EHTMLspliced_uexp ($proto_typ `SCSSstringECSS`) SHTML as spliced_str
-%     start <- "@{" spliced_str "}"
-%       EHTMLfn e => $proto_expr `SCSSR.Str %(ECSSeSCSS)ECSS`SHTML #\label{line:splicing-end}#
-%   EHTML|})
-% end #\label{line:rx_parse_fn_end}#
-% \end{lstlisting}
-% \vspace{-12px}
-% \caption{A grammar-based definition of \texttt{\$rx}.}
-% \vspace{-15px}
-% \label{fig:rx-grammar-based}
-% \end{figure}
-
-\vspace{-6px}
-\subsection{Static Evaluation, Formally}
-It is not difficult to extend $\miniVerseParam$ to account for static evaluation.
-Static environments, $\Sigma$, take the form $\staticenv{\omega}{\uOmega}{\uPsi}{\uPhi}$, where $\omega$ is a substitution.
-Each binding form is annotated with a \emph{phase}, $\phi$, either $\staticphase$ or $\standardphase$.
-The rules for binding forms annotated with $\standardphase$ are essentially unchanged, differing only in that $\Sigma$ passes through opaquely.
-The rules for binding forms annotated with $\staticphase$ are based on the corresponding $\standardphase$ phase rules, differing only in that 1) they operate on $\Sigma$ and 2) evaluation occurs immediately.
-Finally, the forms for TLM definition are modified so that the parse function is now an unexpanded, rather than an expanded expression.
-The substitution $\omega$ is applied to the parse function after it is expanded.
-The full details are defined as a small patch of $\miniVerseParam$ called $\miniVersePH$ in the supplement.
-
-\vspace{-6px}
-\subsection{Library Management}
-In the examples above, and in our formal treatment, we explicitly qualified various definitions with the \li{static} keyword to make them available within static values.
-In practice, we would like to be able to use libraries within both static values and standard values as needed without duplicating code.
-This can be accomplished either by the package manager (e.g.
-SML/NJ's CM \cite{blume:smlnj-cm}, extended with phase annotations) or by allowing one to explicitly lower an instance of a module define in the standard phase for use also in  the static phase, as in a recent proposal for modular staging macros in OCaml \cite{Ocaml/macros}.
-
-TLMs definitions can be exported from the top level of packages, but they cannot be exported from within ML-style modules because that would require that they also appear in signatures, and that, in turn, would complicate reasoning about signature equivalence, since TLM definitions contain arbitrary parse functions.
-%It would also bring in confusion about whether the generated expansions can use private knowledge about type identity. 
-That said, it should be possible to export TLM \emph{abbreviations} from modules, since they refer to TLM definitions only through symbolic names.
-We have not yet formalized this intuition, but the work of \citet{culpepper2005syntactic,culpepper2007advanced} considered a closely related question: how should Typed Scheme's macros interact with its unit (i.e.
-package) system.
-
-% For example, a language-external library manager for Reason similar to SML/NJ's CM \cite{blume:smlnj-cm} could support a \li{static} qualifier on imported libraries, which would place the definitions exported by the imported library into the static phase of the library being defined.
-In particular, a library definition in such a compilation manager might look like this:
-% \begin{lstlisting}[numbers=none,morekeywords={Library,is}]
-% Library 
-%   /* ... exported module, signature and TLM names ... */
-% is 
-%   /* ... files defining those exports ... */
-
-%   /* imports: */
-%   static parsegen.cm 
-% \end{lstlisting}
-
-% A similar approach could be taken for languages the incorporate library management directly into the syntax of programs, e.g. Scala \cite{odersky2008programming}:
-% \begin{lstlisting}[numbers=none]
-% static import edu.cmu.comar.parsegen
-% \end{lstlisting}
-
-% \begin{equation}\label{rule:mExpandsPH-syntaxpe-standard}
-% \inferrule{
-%   \tsmtyExpands{\uOmega}{\urho}{\rho}\\
-%   \Sigma = \staticenv{\omega}{\uOmega_S}{\uPsi_S}{\uPhi_S}\\\\
-%   \expandsP{\uOmega_S}{\uPsi_S}{\uPhi_S}{\ueparse}{\eparse}{\aparr{\tBody}{\tParseResultPCEExp}}\\\\
-%   \evalU{[\omega]\eparse}{\eparse'}\\
-%   \mExpandsPH{\uOmega}{\uAS{\uA \uplus \mapitem{\tsmv}{\adefref{a}}}{\Psi, \petsmdefn{a}{\rho}{\eparse'}}}{\uPhi}{\uM}{M}{\sigma}{\Sigma}
-% }{
-%   \mExpandsPH{\uOmega}{\uAS{\uA}{\Psi}}{\uPhi}{\defpetsmH{\standardphase}{\tsmv}{\urho}{\ueparse}{\uM}}{M}{\sigma}{\Sigma}
-% }
-% \end{equation}
-% \begin{equation}\label{rule:mExpandsPH-syntaxpe-static}
-% \inferrule{
-%   \tsmtyExpands{\uOmega}{\urho}{\rho}\\
-%   \Sigma = \staticenv{\omega}{\uOmega_S}{\uPsi_S}{\uPhi_S}\\
-%   \uPsi_S = \uAS{\uA_S}{\Psi_S}\\\\
-%   \expandsP{\uOmega_S}{\uPsi_S}{\uPhi_S}{\ueparse}{\eparse}{\aparr{\tBody}{\tParseResultPCEExp}}\\\\
-%   \evalU{[\omega]\eparse}{\eparse'}\\
-%   \mExpandsPH{\uOmega}{\uPsi}{\uPhi}{\uM}{M}{\sigma}{\staticenv{\omega}{\uOmega_S}{\uAS{\uA_S \uplus \mapitem{\tsmv}{\adefref{a}}}{\Psi_S, \petsmdefn{a}{\rho}{\eparse'}}}{\uPhi_S}}
-% }{
-%   \mExpandsPH{\uOmega}{\uPsi}{\uPhi}{\defpetsmH{\staticphase}{\tsmv}{\urho}{\ueparse}{\uM}}{M}{\sigma}{\Sigma}
-% }
-% \end{equation}
-
-\vspace{-6px}
-\newcommand{\rwSec}{Related Work}
-\section{\protect\rwSec}
-\label{sec:existing-approaches}
-\subsection{Syntax Definition Systems}\label{sec:syntax-dialects-intro}
-One approach available to library providers seeking to introduce new literal forms is to use a syntax definition system to construct a library-specific \emph{syntax dialect}: a new syntax definition that extends the syntax definition given in the language definition with new forms, including literal forms.
-For example, Ur/Web's syntax (Figure 1) is a library-specific dialect of Ur's syntax \cite{conf/pldi/Chlipala10,conf/popl/Chlipala15}.
-
- % Such dialects are sometimes qualitatively taxonomized as amongst the ``domain-specific language'' for this reason \cite{fowler2010domain}.
-%Syntactic cost is often assessed qualitatively \cite{green1996usability}, though quantitative metrics can be defined.
-% Syntax definition systems , which we will discuss in Sec.
-\ref{sec:syntax-dialects}, have simplified the task of defining ``library-specific'' (a.k.a.
-``domain-specific'') syntax dialects like Ur/Web, and have thereby contributed to their ongoing proliferation.
-
-% Many have argued that a proliferation of syntax dialects constructed using these tools is harmless or even desirable, because programmers can simply choose the right syntax dialect for each job at hand \cite{journals/stp/Ward94}.
-However, we argue that in fact
-There are hundreds of syntax definition systems of various design.
-Notable examples include grammar-oriented systems like Camlp4 \cite{ocaml-manual}, Copper \cite{conf/gpce/WykS07,conf/pldi/SchwerdfegerW09}, SugarJ/Sugar*/SoundExt \cite{erdweg2011sugarj,erdweg2013framework,conf/icfp/LorenzenE13,conf/popl/LorenzenE16}, as well as various parser combinator systems \cite{Hutton1992d}.
-The parsers generated by these systems can be invoked to preprocess program text in various ways, e.g.
-by invoking them from within a build script, by using a preprocessor-aware build system (e.g.
-\li{ocamlbuild}), or via language-integrated preprocessing directives (e.g.
-Racket's \li{#lang} directive or its reader macros \cite{Flatt:2012:CLR:2063176.2063195}, or the import mechanism of SugarJ/Sugar*/SoundExt.)
-%library-specific (a.k.a. ``domain-specific'')
-
-The first problem with using a syntax definition system to directly extend the context-free syntax of a language is that clients cannot always deterministically combine the resulting extended syntax dialects when they want to use the new forms that they define together, i.e.
-there can be syntactic conflicts.
-This is a significant problem because large programs use many separately developed libraries \cite{DBLP:conf/sac/LammelPS11} and often do not fall cleanly into a single ``problem domain''.
-
-One possible solution is found in Copper, which integrates a modular grammar analysis that guarantees that determinism is conserved when syntax dialects of a certain restricted class are combined \citet{conf/pldi/SchwerdfegerW09}.
-The caveat is that the constituent dialects must prefix all newly introduced forms with marking tokens drawn from disjoint sets.
-To be confident that the marking tokens used are disjoint, providers must base them on the domain name system or some other coordinating entity.
-Because the mechanism operates at the level of the context-free grammar, it is difficult for the client to define scoped abbreviations for these verbose marking tokens.
-
-TLMs sidestep these complexities entirely because the context-free syntax of the language is fixed, and composition is via splicing rather than direct combination.
-The TLM provider can therefore modularly establish that the syntax definition that the TLM implements (possibly using a syntax definition system internally, as discussed in Sec.
-\ref{sec:static-eval}) is deterministic.
-There is no need for verbose marking tokens, and programmers can define scoped TLM abbreviations as discussed in Sec.
-\ref{sec:ptsms}.
-
-% In some cases, the dialects in question cannot be combined simply because they have been defined under incompatible syntax definition systems.
-In other cases, the problem is one of \emph{conflict}: two dialects might define the same form, but determine different desugarings.
-For example, consider two syntax dialects defined under a system like Camlp4: $\mathcal{D}_1$ defines literal forms for sets, and $\mathcal{D}_2$ defines literal forms for finite sets, both delimited by \verb~{<~ and \verb~>}~.
-Each dialect is deterministic (i.e.
-there are no possible syntactic ambiguities), denoted $\mathrm{det}(\mathcal{D}_1)$ and $\mathrm{det}(\mathcal{D}_2)$.
-However, when the grammars are combined by Camlp4, it is not the case that $\mathrm{det}(\mathcal{D}_1 \cup \mathcal{D}_2)$ because \verb~{<>}~ can desugar to the empty set or the empty dictionary.
-Reasoning about determinism is therefore not modular.
-%(Indeed, this ambiguity arose when Python added set literals to its grammar.) %A recent version of Python added derived forms for mutable sets.
-Due to a conflict with dictionary syntax, however, there is no derived form for the empty set.)
- %A third syntax dialect might come along that uses the same forms that $\mathcal{D}_2$ defines, but for ordered finite maps.
-
-Even putting aside the problem of syntactic conflict, there are questions about just how reasonable sprinkling many possibly unfamiliar library-specific literal forms  throughout a program may be.
-The intuition was given by example in Sec.
-\ref{sec:intro}.
-To reiterate, more generally:
-\begin{enumerate}[noitemsep,nolistsep]
-\item \textbf{Responsibility}: Clients using a combined syntax dialect cannot easily determine which constituent extension is responsible for a given form, whereas TLM clients can follow the binding structure of the language in the usual manner.
-\item \textbf{Segmentation}: Clients of a syntax dialect cannot accurately determine which segments of the program text appear directly in the desugaring.
-In contrast, TLM clients can inspect the inferred segmentation (communicated via secondary notation.)
-\item \textbf{Binding}: Clients of a context-free syntax dialect cannot be sure that the desugaring is context-independent and that spliced terms are capture avoiding, as TLM clients can.
-\item \textbf{Typing}: Clients of a syntax dialect cannot reason abstractly about types.
-In contrast, TLM clients can determine the type of any expansion by referring to the parameter and type declarations on the TLM definition, and nothing else.
-The inferred segmentation also gives types for each spliced expression or pattern.
-%This information can be communicated upon request by the type inspection service of a program editor.
-\end{enumerate}
-
-The work of \citet{conf/icfp/LorenzenE13,conf/popl/LorenzenE16} introduces SoundExt, a grammar-based syntax extension system where extension providers can equip their new forms with derived typing rules.
-The system then attempts to automatically verify that the expansion logic (expressed using a rewrite system, rather than an arbitrary function) is sound relative to these derived rules.
-TLMs differ in several ways.
-First, as already discussed, we leave the context-free syntax fixed, so different TLMs cannot conflict.
-Second, SoundExt does not enforce hygiene, i.e.
-expansions might depend on the context and intentionally induce capture.
-Similarly, there is no abstract segmentation discipline.
-While the client can indirectly reason about binding (but not segmentation) by inspecting the derived typing rules, this is weaker than the strict hygiene and segmentation discipline that TLMs enforce.
-Another distinction is that TLMs do not support type-dependent expansions.
-The trade-off is that this allows expansions, and therefore segmentations, to be generated even when the program is ill-typed, as long as the static values are well-typed (this notion of \emph{untyped expansion} was not considered formally above, but it follows straightforwardly from the fact that all of the premises to the TLM application rules except the final proto-expansion validation premises are independent of the typing context.) Another important distinction is that TLMs rely on proto-expansion validation, rather than verification as in SoundExt.
-The trade-off is that TLMs do not require that the expansion logic be written using a restricted rewriting system, nor does the system require a fully mechanized language definition and heavyweight theorem proving machinery.
-Finally, it would be difficult to define syntax extensions in SoundExt that operate over module-parameterized families of types.
-Parameters can be approximated using splicing, but there is no clear notion of ``partial application''.
-
-%In other words, encountering an unfamiliar derived form has made it difficult for the programmer to maintain the usual \emph{type discipline} and \emph{binding discipline}.
-%Compelling the programmer to examine the desugaring directly defeat the purpose of defining the derived form -- decreasing cognitive cost.
-Indeed, it substantially increases cognitive cost.
-
-% In contrast, when a programmer encounters, for example, a function call like the call to \li{read_data} on Line 3, the analagous questions can be answered by following clear protocols that become ``cognitive reflexes'' after sufficient experience with the language, even if the programmer has no experience with the library defining \li{read_data}:
-% \begin{enumerate}
-% \item The language's syntax definition determines that \li{read_data(a)} is an expression of function application form.
-% \item Similarly, \li{read_data} and \li{a} are definitively expressions of variable form.
-% \item The variable \li{a} can only refer to the binding of \li{a} on Line 1.
-% \item The variable \li{w} can be renamed without knowing anything about the values that \li{read_data} and \li{a} stand for.
-% \item The type of \li{x} can be determined to be \li{B} by determining that the type of \li{read_data} is \li{A -> B} for some \li{A} and \li{B}, and checking that \li{a} has type \li{A}.
-Nothing else needs to be known about the values that \li{read_data} and \li{a} stand for.
-In Reynolds' words \cite{B304}:
-% \begin{quote}
-% \emph{Type structure is a syntactic discipline for enforcing levels of abstraction.}
-% \end{quote}
-% \end{enumerate}
-
-\vspace{-6px}
-\subsection{Term Rewriting Systems}
-\vspace{-2px}
-Another approach -- and the approach that TLMs are rooted in -- is to leave the context-free syntax of the language fixed, and instead contextually rewrite existing literal forms.
-
-For example, OCaml's textual syntax now includes \emph{preprocessor extension (ppx) points} used to identify terms that some term rewriting preprocessor must rewrite \cite{ocaml-manual}.
-We could  mark a string literal containing Ur/Web-style HTML syntax with a ppx annotation, \li{xml}, as follows:
-\begin{lstlisting}[numbers=none]
-  [%xml "SSTR<h1>Hello, {[first_name]}!</h1>ESTR"]
-\end{lstlisting}
-There are problems reasoning modularly and abstractly about such constructions.
-More than one applied preprocessor might recognize this annotation (there are, in practice, many XML/HTML libraries), so the problems of \textbf{Responsibility} come up.
-It is also impossible to reason abstractly about \textbf{Segmentation}, \textbf{Capture}, \textbf{Context Dependence} and \textbf{Typing} because the code that the preprocessor generates is unconstrained.% For these reasons, users of this mechanism warn that they should be used sparingly.\todo{citation}
-
-Term-rewriting macro systems are language-integrated local term rewriting systems that require that the client explicitly apply the intended rewriting, implemented by a macro, to the term that is to be rewritten.
-This addresses the issue of \textbf{Responsibility}.
-However, unhygienic, untyped macro systems, like the earliest variants of the Lisp macro system \cite{Hart63a}, Template Haskell \cite{SheardPeytonJones:Haskell-02} and GHC's quasiquotation system \cite{mainland2007s} (which is based on Template Haskell), do not allow clients to reason abstractly about the remaining issues, again because the expansion that they produce is unconstrained.
-(It is not enough that with Template Haskell / GHC quasiquotation, the generated expansion is typechecked -- to satisfy the \textbf{Typing} criterion, it must be possible to reason abstractly about \emph{what the type of the generated expansion is}.)
-
-\emph{Hygienic} macro systems prevent, or abstractly account for \cite{DBLP:conf/esop/HermanW08,Herman10:Theory}, \textbf{Capture}, and they enforce application-site \textbf{Context Independence} \cite{Kohlbecker86a,DBLP:conf/popl/Adams15,DBLP:conf/popl/ClingerR91,DBLP:journals/lisp/DybvigHB92}.
-However, this turns out to make it impossible to repurpose string literal forms to introduce compositional literal forms at other types.
-Consider again our running XHTML example, which we might try to realize by applying a hygienic macro, \li{xml!}, to a string literal that the macro parses:
-\begin{lstlisting}[numbers=none]
-  (xml! "SSTR<h1>Hello, {[first_name]}!</h1>ESTR")
-\end{lstlisting}
-The expansion of this macro will fail the check for context independence because \li{first_name} will appear as a free variable, no different from any other.
-The hygiene mechanism for TLMs addresses this problem by explicitly distinguishing spliced segments of the literal body in the proto-expansion.
-This also addresses the problem of \textbf{Segmentation} -- the segmentation abstractly communicates the fact that (only) \li{first_name} is a spliced expression of type \li{string}.
-
-Much of the research on macro systems has been for languages in the LISP tradition \cite{mccarthy1978history} that do not have rich static type structure.
-The formal macro calculus studied by \citet{DBLP:conf/esop/HermanW08} uses types to encode the binding structure of the generated expansion (in particular, to indicate which identifiers passed as arguments can be captured by which other arguments), but the expanded language itself does not have rich type structure (nor is this calculus capable of expressing new literal forms, for the reasons already discussed.) Research on typed \emph{staging macro systems} like MetaML \cite{Sheard:1999:UMS} and MacroML \cite{ganz2001macros} is also not directly applicable to the problem of defining new literal forms -- the syntax tree of the arguments cannot be inspected at all (staging macros are useful mainly for performance reasons.) The Scala macro system is a hygienic macro system which supports ``black-box'' term rewriting macros.
-These support reasoning abstractly about \textbf{Typing} because type annotations constrain the macro arguments and the generated expansions, though the precise reasoning principles available are unclear because the Scala macro system has not been formally specified.
-In any case, black-box macros also cannot be used to repurpose string literal forms because they are hygienic.
-
-  Some languages, including Scala \cite{odersky2008programming}, build in \emph{string splicing} (a.k.a.
-\emph{string interpolation}) forms, or similar but more general \emph{fragmentary quotation forms} \cite{conf/icfp/Slind91}, e.g.
-the SML/NJ dialect of ML.
-These designate a particular delimiter to escape out into the expression language.
-The problem with using these together with macros as vehicles to introduce literal forms at various other types is 1) there is no ``one-size-fits-all'' escape delimiter, and 2) typing is problematic because every escaped term is checked against the same type.
-In the XHTML example, we have splicing at two different types using two different delimiters.
-These forms are also not available in pattern position.%In general, e.g.
-when defining syntax for a programming language with many sorts of terms, the most appropriate choice of delimiters might depend on where each spliced term appears.
-
-This brings us to the most closely related work, that of \citet{TSLs} on \emph{type-specific languages} (TSLs).
-Like simple expression TLMs (Sec.
-\ref{sec:setsms}), TSLs allow library providers to programmatically control the parsing of expressions of generalized literal form.
-Parse functions are associated directly with nominal types and invoked according to a bidirectionally typed protocol.
-In contrast, TLMs are separately defined and explicitly applied.
-Accordingly, different TLMs can be defined at the same type without conflict, and can operate at any type, not just at nominal (i.e.
-abstract) types.
-In a subsequent short paper, \citet{sac15} proposed explicit application of simple expression TLMs (there called \emph{typed syntax macros}) in a bidirectional typed setting \cite{Pierce:2000:LTI:345099.345100}, but this short proposal did not give any formal details.
-With TLMs, it is not necessary for the language to be bidirectionally typed -- context independence implies that ML-style type inference can be performed using only the segmentation (because the remainder of an expansion cannot mention the very variables whose types are being inferred.) It should be possible to implicitly invoke TLMs based on the expected type in situations where the inference engine had enough information, but we leave this as an interesting direction for future work.
-
- Another critical distinction is that the metatheory presented by \citet{TSLs} establishes only that generated expansions are of the expected type (i.e.
-a variant of the Typed Expression Expansion theorem from Sec.
-\ref{sec:s-metatheory}.) It does not establish the remaining abstract reasoning principles that have been a major focus of this paper.
-In particular, there is no formal hygiene theorem (though it is discussed informally), and the mechanism does not guarantee that a valid segmentation will exist, nor associate types with segments.
-Finally, this prior work did not consider pattern matching, type functions, ML-style modules, parameters or static evaluation.
-All of these are important for integration into an ML- or Scala-like language, and the focus of most of this paper.
-
-% These existing forms normally have other meanings, so this can be confusing \cite{pane1996usability}.
-
-\vspace{-6px}
-\newcommand{\discussionSec}{Discussion}
-\section{\protect\discussionSec}
-\label{sec:discussion}
-\label{sec:conclusion}
-\vspace{-2px}
-
-The importance of specialized notation as a ``tool for thought'' has long been recognized \cite{DBLP:journals/cacm/Iverson80}.
-According to Whitehead, a good notation ``relieves the brain of unnecessary work'' and ``sets it free to concentrate on more advanced problems'' \cite{cajori1928history}, and indeed, advances in mathematics, science and programming have often been accompanied by new notation.
-
-Of course, this desire to ``relieve the brain of unnecessary work'' has motivated not only the syntax but also the semantics of languages like ML and Scala -- these languages maintain a strong type and binding discipline so that programmers, and their tools, can hold certain implementation details abstract when reasoning about program behavior.
-In the words of \citet{B304}, ``type structure is a syntactic discipline for enforcing levels of abstraction.''
-
-Previously, these two relief mechanisms were in tension -- mechanisms  that allowed programmers to express new notation would obscure the type and binding structure of the program text.
-TLMs resolve this tension for the broad class of literal forms that generalized literal forms subsume.
-This class includes all of the examples enumerated in Sec.
-\ref{sec:intro} (up to the choice of outermost delimiter), the additional examples described in Sec.
-\ref{sec:static-eval}, the more detailed example in the supplement, and the many examples in the prior papers by \citet{TSLs,sac15}.
-
-Of course, not all possible literal notation will prove to be in good taste, and we make no empirical claims about particular notation in this paper.
-TLMs leave the burden of establishing the value of any particular literal notation to individual library providers, rather than to the language designer.
-The reasoning principles that TLMs provide, which are the specific contributions of this paper, allow clients to "reason around" poor literal designs, using principles analagous to those already familiar to programmers in languages like these.
-% We must leave a detailed empirical evaluation of the impact of particular TLMs on various quality attributes, like programmer productivity and program comprehensibility, as future work.
-The technical content of this paper, which formally establishes these reasoning principles, constitutes the first detailed type-theoretic account of a typed, hygienic macro system for an ML-like language, i.e.
-one with a rich static type system, support for pattern matching and an ML-like module system with type functions and abstract type members.
-Our own language integration efforts have been focused on an emerging alternative front-end for OCaml called Reason, but Scala supports analagous features \cite{conf/oopsla/AminRO14} and would also be a suitable host for the TLM mechanism.
-The intended audience for this paper is language designers seeking a reasonable solution to the problem of user-defined literal notation.% We also plan to implement TLMs into \li{typy}, a typed functional language embedded into Python as a library \cite{gpce/Omar16}.%We have evaluated this claim by stating the available abstract reasoning principles formally and proving that they hold for the calculi that we have described.
-
-The examples in this paper are written in an open source alternative syntactic front-end for OCaml called Reason because we are incorporating TLMs into Reason.
-However, the essential ideas are not Reason-, OCaml- or ML-specific -- it should be possible to adapt TLMs to other languages that take a similar approach to types and binding (e.g.
-Haskell, Scala and others.) Languages that have a disciplined binding structure but that lack a rich static type structure, e.g.
-standard Racket, would also benefit from TLMs -- they can apply these results by deploying the usual trick of viewing the language as statically ``{unityped}'' \cite{pfpl,scott1980lambda}.
-
-% The intended audience for this paper is language designers who want to decentralize control over literal forms but seek a detailed understanding, rooted in the first principles of type theory, of a reasonable mechanism by which to achieve that goal.
-
-%(To summarize Sec.
-\ref{sec:existing-approaches}, the closest existing typed, hygienic macro system -- the Scala macro system \cite{ScalaMacros2013} -- is not formally specified, the work of \cite{DBLP:conf/esop/HermanW08} uses types only to reason abstractly about binding, and the specification of TSLs in the work of \citet{TSLs} did not formalize hygiene nor handle these advanced language features.)
-
-% Haskell's \li{do}-notation at types with monadic structure cannot be expressed directly using TLMs because it needs to purposefully induce capture in spliced sub-terms, and TLMs prevent capture.
-This limitation could be resolved by allowing identifiers that appear in literal bodies to be explicitly designated as bound within spliced expressions in the segmentation.
-The prior work by \citet{DBLP:conf/esop/HermanW08} considered a similar approach in a setting where the locations of sub-terms is known ahead of time (building on the \emph{tree locations} of \citet{gorn1965explicit}, which are conceptually similar to our spliced segment locations.)  We leave as future work the details of this proposal and a consideration of whether this additional expressive power would justify the increased reasoning complexity for clients (it is presently the author's opinion that it would not.)
-
-There are several interesting avenues for future work beyond those already discussed.
-A correct parse function never returns an encoding of a proto-expansion that fails validation given well-typed splices, but this invariant cannot be enforced by the ML type system.
-% Using a proof assistant, it would be possible to verify that
-% a parse function generates only encodings of valid proto-expansions. 
-Under a
-richer type system, the return type of the parse function itself could be refined so as to
-enforce this invariant intrinsically.
-This problem -- of typed first-class term representations -- has been studied in a variety of settings, e.g.
-in MetaML \cite{Sheard:1999:UMS} and in the modal logic tradition \cite{DBLP:conf/popl/DaviesP96}.
-We leave integration of these approaches into the TLM mechanism as future work.
-Our initial efforts aim to leave the semantics of OCaml unchanged, consistent with the philosophy of the Reason project.
-
-%Another topic of interest has to do with intentional capture. 
-Another direction has to do with automated refactoring.
-The unexpanded language does not come with context-free notions of renaming and substitution.
-However, given a segmentation, it should be possible to ``backpatch'' refactorings into literal bodies.
-Recent work by \citet{wand2017inferring} on tracking bindings ``backwards'' from an expansion to the source program is likely relevant.
-
-At several points in the paper, we allude to editor integration.
-However, several important questions having to do with TLM-specific syntax highlighting, incremental parsing and error recovery \cite{graham1979practical} remain to be considered.
-Another interesting direction would be to generalize TLMs to support non-textual notation in the setting of a structure editor.
-The semantic foundations for structure editors proposed recently by \citet{DBLP:conf/popl/OmarVHAH17} may guide such an effort.
-
-Let us conclude by considering the oft-uttered phrase ``syntax doesn't matter'', which is generally taken to mean that syntactic concerns are orthogonal to various more important semantic concerns.
-Our hope is that the reader will leave this paper with an understanding that this is a rather simplistic proclamation.
-Instead, let us proclaim that ``semantics matter for syntax too!'' % We hope that this work will lead to more ``reasonable'' variants of other syntactic conveniences, e.g.
-the ubiquitious ``deriving'' clauses on datatype definitions.
-
-% Other directions for future work are given in the supplement\todo{do this?}.
-
-% Another interesting question is what this mechanism would look like in the setting of a projectional structure editor, i.e.
-one where the syntax is not textual but rather tree-shaped and projected in an interative manner to the programmer.
-This paper taken together with work by \citet{Omar:2012:ACC:2337223.2337324,DBLP:conf/popl/OmarVHAH17} can serve to guide such an effort.
-
-% To conclude, TLMs give substantial syntactic control to library providers while leaving programmers with strong abstract reasoning principles.
-We believe TLMs therefore occupy a ``sweet spot'' in the  design space.
-%This paper serves to guide those who hope to incorporate TLMs into their future language designs.
-
-%% Acknowledgments
-\begin{acks}                            %% acks environment is optional
-                                        %% contents suppressed with 'anonymous'
-  %% Commands \grantsponsor{<sponsorID>}{<name>}{<url>} and
-  %% \grantnum[<url>]{<sponsorID>}{<number>} should be used to
-  %% acknowledge financial support and will be used by metadata
-  %% extraction tools.
-  This material is based upon work supported by the
-  \grantsponsor{GS100000001}{National Science
-    Foundation}{http://dx.doi.org/10.13039/100000001} under Grant
-  No.~\grantnum{GS100000001}{nnnnnnn} and Grant
-  No.~\grantnum{GS100000001}{mmmmmmm}.  Any opinions, findings, and
-  conclusions or recommendations expressed in this material are those
-  of the author and do not necessarily reflect the views of the
-  National Science Foundation.
-\end{acks}
-
-\clearpage
-%% Bibliography
-\bibliography{../../../papers/research}
+\end{comment}
 

### 2018-02-19 pldi18/rebuttal.txt
added reviews and rebuttal from PLDI 18

Thank you for the detailed reviews! We respond to the main questions and concerns below, then address additional questions beyond 700 words.

RA asked for more detailed examples in order to assess ease-of-use from the TLM provider’s perspective. The paper spoke to this perspective by pointing out that TLM parse functions are quite ordinary -- they take strings and produce ASTs represented as ordinary values -- so they can be generated using a parser generator of the provider’s choice, e.g. lex/yacc or a parser combinator library. Ease-of-use from the provider’s perspective is substantially a function of this choice. Including, e.g., a yacc grammar in the paper body would have consumed a large amount of space without conveying information relevant to the claims of the paper, which revolve around reasoning principles from the TLM client’s perspective.

The paper does provide a complete example -- Figure 4 shows the full definition of $html and two example applications. The only omission was a straightforward helper function that was instead detailed in the text. This example also spoke to this question of ease-of-use by demonstrating how providers can repurpose off-the-shelf parsers.

RB and RC expressed a more general desire for more examples. The prior work on TSLs presented a variety of examples and an empirical study that taxonomized additional possibilities, so in Sec1 we explicitly deferred to that paper in this regard. The case studies in the present paper were chosen to support our specific claims in consideration of the page limit. We will include additional examples in the final supplement.

We use “literal notation” to mean “notation for terms within a specified parametric type family” and will make this explicit in the final version.

RA said “I don't agree with the (implicit) argument in the paper that expansions
examined by a tool (e.g., IDE) are more abstract than expansions examined by a programmer”. The tool needs only examine the proto-expansion at the lowest level. It can then pass along only the type annotation and segmentation (a finite set) to other services, and ultimately to the programmer. The service/programmer can then reason about, e.g., the type of expression that is expected at a splice site. By analogy, a function’s type determines the type of its argument without requiring that the programmer inspect how that argument is used inside the function.

RA questioned the progression through simple TLMs before arriving at parametric TLMs. Simple TLMs are certainly not themselves practical (further discussion below the fold). Instead, they serve a pedagogical purpose: they allow us to tackle all five reasoning principles without bringing in the full complexity of the ML module system. Moreover, the impracticalities revealed by considering simple TLMs nicely motivate the introduction of parametric TLMs, which is why that section may seem concise relative to its practical importance.

RC says “all that TLMs do in this respect is to tag every “$” keyword with a nominal type”. The type need not be nominal (e.g. structural types like arrows are supported). Moreover, each spliced segment also has a type.

RC suggested that strict capture avoidance discipline we settled on prevents the expression of, e.g., regex backreferences. This is not the case -- TLMs can send values into spliced functions by application. Reason3 introduces concise function notation -- x => e -- that decreases the cost of this technique. The characteristic argument of this paper is that simpler reasoning principles justify these small syntactic costs.

RD wonders why we define the mechanism using arbitrary parse functions. Note that the reasoning principles we formulate would be equally relevant to designs that privilege a particular syntax definition formalism. Regarding AST representation, the ML type system is not powerful enough to express an intrinsically typed AST in the parametric setting, but if it were, the necessary invariants are exactly those maintained by proto-expansion validation.

RA and RB asked about inferring TLM parameters. It should be possible to infer those type parameters that appear in the TLM’s type annotation by unification at type-constrained  application sites. However, it would violate the Typing criterion to use the expansion itself to infer the type parameters. This might be acceptable if the editor reveals inferred parameters upon request, but we leave that to future work on editor design.

--- (additional responses below go beyond 700 words) --

RA brought up properties that clients cannot reason about abstractly, e.g. non-termination. However, we claim only to support the sort of abstract reasoning otherwise possible in the language, i.e. reasoning supported by the type system.

Several reviewers asked why we used “maps \hat{x} to \hat{x} ~> x” and similar instead of “maps \hat{x} to x”. This was an admittedly awkward way to show what the mapping connective looks like (“~>”). We will make this less awkward in the final version.

RA said “it is not clear why it is necessary to have $\hat{\Gamma} = \langle \mathcal{G}; \Gamma \rangle$ [...]” and RC had related questions. The point is that \hat{x} might be shadowed, which will cause it to be remapped, but we still need to track the “old” mapping (assumed unique by alpha-renaming in the XL) in \Gamma because it may have appeared elsewhere in the expansion. Figure 8 and the discussion around it gives such an example.

RA says that “one would need to change the variables in $\Delta$ and $\Gamma$ --- but they have already been picked”. Right -- an implementation strategy that would pre-empt the need to “backtrack” is one that generates fresh variables relative to \Delta/\Gamma_{app} each time \Delta/\Gamma is extended. The formalism is not meant to uniquely determine an implementation strategy, only to constrain it.

RC asks “is it not redundant then to have separate Typ and UTyp grammars”. The two grammars differ in that Typ contains variables and UTyp contains identifiers. The distinction is important because types can appear within expressions, including spliced expressions.

RC comments that “there is a reason we don’t write ASTs in our formalisms”. We used ABT notation deliberately because it is important to distinguish between terms that are and are not identified up to alpha variation. Moreover, UL terms are those that are exposed to the programmer, so we used conventional idioms, while XL terms are only ever manipulated as ABTs by parse functions, so we used ABT notation.

RC said that “markup is inconsistent between splicing using {[ ]} and just { }, within an HTML literal”. Those are different splicing forms -- the first for string splicing and the second for HTML splicing (see Figure 1 and accompanying text).

RD says “you couldn’t define useful pattern expansions using guards”. In OCaml, guards are not part of the syntax of patterns (they can appear only at the outermost level of a match rule). If guards were part of patterns, then the mechanisms already considered for expression TLMs would become relevant to the guard portion of a pattern expansion as well.

RC asked about the “fold” operation and RD asks “what is the `fold` pattern?” These are the introductory form for recursive types. Similarly, the “inj[\ell]” operation is the introductory form for sum types. We based our terminology on Harper’s PFPL, but the same or similar notation is common in other textbooks (e.g. Pierce) and in the literature.

RA says “I think that the paper drops some non-trivial information in the simplification and abbreviation of the proto_typ, proto_expr, and proto_pat types.” Yes, these datatypes were simplified as acknowledged in the text. In practice, we use the existing OCaml AST, with a special variable “__tlm_spliced” that takes a pair of number literals to represent spliced segments. We will mention this implementation detail in the final version.

The $grammar TLM in the supplement, and related definitions, were shown only to outline how a parser generator could be formulated as a mode of use of TLMs. A full parser generator implementation would span thousands of lines of code. There is certainly no requirement that the programmer use this method, and we expect in practice that existing parser generators will be used to generate parse functions.

### 2018-02-20 icfp18/syntax-icfp18.tex
initial port to ICFP18 format and some edits to the intro that preview my plan for the submission

--- 2018-02-19 pldi18/syntax-pldi18.tex
@@
 %% For double-blind review submission
-\documentclass[sigplan,review,10pt,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
+\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
+% \documentclass[sigplan,review,10pt,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
 % \documentclass[acmsmall,10pt,review,anonymous]{acmart}\settopmatter{printfolios=true}
@@
 \usepackage{bbm}
+\usepackage{wasysym}
 \definecolor{light-gray}{gray}{0.95}
@@
 stringstyle=\ttfamily\color{red},
-mathescape=false,escapechar=~,
-numbers=left, numberstyle=\scriptsize\color{mygray}, language=ML,moredelim=[il][\sffamily]{?},showspaces=false,showstringspaces=false,xleftmargin=0pt, numbersep=-6pt, morekeywords=[1]{try,spliced,tyfam,opfam,let,fn,val,def,casetype,objtype,metadata,of,*,string,datatype,new,toast,syntax,module,switch,where,using,import,for,ana,syn,opcon,tycon,metasignature,metamodule,metasig,metamod,static,at,by,tycase,mod,macro,match,pattern,in,patterns,expressions,implicit,forall,exptsm,pattsm},deletekeywords={double,structure,of},classoffset=0, xleftmargin=-10pt,
+mathescape=false,escapeinside={(~}{~)},
+numbers=left, numberstyle=\scriptsize\color{mygray}, language=ML,moredelim=[il][\sffamily]{?},showspaces=false,showstringspaces=false,xleftmargin=0pt, numbersep=-6pt, morekeywords=[1]{try,spliced,tyfam,opfam,let,fn,val,def,casetype,objtype,metadata,of,*,string,datatype,new,toast,notation,module,switch,where,using,import,for,ana,syn,opcon,tycon,metasignature,metamodule,metasig,metamod,static,at,by,tycase,mod,macro,match,pattern,in,patterns,expressions,implicit,forall,exptsm,pattsm},deletekeywords={double,structure,of},classoffset=0, xleftmargin=0pt,
 aboveskip=4pt,belowskip=2pt,
@@
 \input{macros}
+
+\setlength{\abovecaptionskip}{4pt plus 3pt minus 2pt} % Chosen fairly arbitrarily
+\setlength{\belowcaptionskip}{-2pt plus 3pt minus 2pt} % Chosen fairly arbitrarily
 
@@
 This paper questions this common practice: why should an otherwise ordinary data structure be privileged in the definition of a general-purpose programming language?
-General-purpose programming languages typically define literal notation for only a small number of data structures, e.g.
+General-purpose programming languages typically define literal notation for only a small number of common data structures, e.g.
 lists.
 This is unsatisfying because there are many other data structures for which literal notation might be useful, e.g.
-finite maps, regexes, HTML data, SQL queries, syntax trees and encodings of chemical structures.
+hash tables, regular expressions, HTML data, SQL queries, syntax trees and chemical structures.
+There may also be different implementations of each of these data structures, e.g.
+with different performance characteristics, that could all benefit from common literal notation.
 %A less \emph{ad hoc} design would be one that allows library providers to define new literal forms in a decentralized manner.
@@
 the Scala macro system \cite{ScalaMacros2013}, are only somewhat more reasonable, and do not offer direct syntactic control.
-This paper introduces \emph{typed literal macros (TLMs)}, which allow library providers to define new literal notation of nearly arbitrary form at any specified type or parameterized family of types. 
+This paper introduces \emph{typed literal macros (TLMs)}, which allow library providers to define new literal notation of nearly arbitrary design at any specified type or parameterized family of types.
 % TLMs give library providers programmatic control over the parsing and expansion of expressions and patterns of a flexible \emph{generalized literal form} at {a} specified type or parameteric family of types.
@@
 % Partial parameter application lowers the syntactic cost of this strict style. 
-Compared to existing approaches, including direct grammar extension systems and prior  macro-based parsing systems, TLMs are uniquely \emph{reasonable}: TLM providers can reason modularly about syntactic ambiguity, and TLM clients can reason abstractly, i.e.
+Compared to existing approaches, TLMs are uniquely \emph{reasonable}: TLM providers can reason modularly about syntactic ambiguity, and TLM clients can reason abstractly, i.e.
 without examining the underlying expansion, about types and binding.
-The system only needs to convey to clients, via secondary notation, the inferred \emph{segmentation} of each literal, which gives the locations of spliced subterms.
+The system only needs to convey to clients, via secondary notation, the inferred \emph{segmentation} of each literal, which gives the locations and types of spliced subterms.
 %Maintaining information about segmentation is necessary for the novel TLM hygiene mechanism, i.e.
@@
 % This mechanism captures many common syntactic idioms while avoiding the problem of syntactic conflicts by construction and supplying clients with clear abstract reasoning principles. 
-This paper establishes these abstract reasoning principles formally with a calculus of typed expressions, patterns and modules. 
-%This calculus is the first detailed type-theoretic account of a hygienic macro system, of any design, for a language with these essential features of ML, and the first macro system with support for literal parsing that comes equipped with these essential reasoning principles.
+This paper establishes these abstract reasoning principles formally with a calculus of typed expressions, pattern matching and ML-style modules. 
+This calculus is the first detailed type-theoretic account of a hygienic macro system, of any design, for a language with these essential features of ML. 
 We are integrating TLMs into Reason, an emerging alternative front-end for OCaml.% This calculus is the first such formalization of a typed macro system.
@@
 % Put succinctly, we design a programming language in the ML tradition with a \emph{reasonably} programmable syntax.
-\vspace{-10px}
 \end{abstract}
@@
 \label{sec:intro}
-\vspace{-3px}
 % The surface syntax of a programming language is its user interface. 
-When designing the surface syntax of a general-purpose programming language, it is common practice to define \emph{literal notation} that decreases the syntactic cost of constructing and pattern matching over values of a few data structures.
-For example, in Standard ML (SML), the list literal \li{[x, y, z]} is equivalent by desugaring to \li{Cons(x,Cons(y,Cons(z,Nil)))} where \li{Nil} and \li{Cons} stand for the constructors of the standard \li{list} datatype \cite{mthm97-for-dart,harper1997programming}.
+When designing the surface syntax of a general-purpose programming language, it is common practice to define \emph{literal notation} that decreases the syntactic cost of constructing and pattern matching over values of a select few data structures.
+For example, Standard ML supports list literals like \li{[x1, x2, x3]} in both expressions and patterns \cite{harper1997programming,mthm97-for-dart}, and many other general-purpose languages support substantially similar list literal notation.
 %\footnote{In SML, list literals work even in contexts where the list expression constructors have been shadowed by other bindings, i.e. the meaning of literal forms is \emph{context independent}. 
 % We will return to this concept again later.} 
-Similarly, Ur/Web's surface syntax extends Ur's syntax with expression and pattern literals for encodings of XML and HTML data \cite{conf/popl/Chlipala15,conf/pldi/Chlipala10}.
-Figure \ref{fig:urweb} shows two HTML literals, one that ``splices in'' a string expression (Line 1) and the other an HTML expression (Line 2).
+Another example is Ur/Web, which extends the surface syntax of Ur (an ML-like language \cite{conf/pldi/Chlipala10}) with expression and pattern literals for encodings of XML and HTML data \cite{conf/popl/Chlipala15}.
+The example in Fig.
+\ref{fig:urweb} below shows two HTML literals, one that ``splices in'' a string expression delimited by \li{\{[} and \li{]\}} (Line 1) and the other an HTML expression delimited by \li{\{} and \li{\}} (Line 2).
 %The desugarings of these literal expressions, not shown, are substantially more verbose and less readable to programmers familiar with XHTML.
 
-This language design practice is decidedly \emph{ad hoc}, in that it is easy to come up with other examples of data structures that are not privileged with literal forms in these languages, but for which mathematicians, scientists or programmers have invented specialized notation \cite{DBLP:journals/cacm/Iverson80,cajori1928history}.
-For example, 1) clients of a ``collections'' library might also want matrix, set and map literals following usual mathematical conventions; 2) clients of a ``web programming'' library might like not just HTML literals but also CSS literals (which Ur/Web lacks); 3) a compiler writer might like ``quotation'' literals for the terms of the object language and various intermediate languages; and 4) clients of a ``chemistry'' library might like chemical structure literals based on the SMILES standard \cite{anderson1987smiles}.
-
-Although requests for specialized literal notation are easy to dismiss as superficial, the reality is that literal notation, or the absence thereof, can have a substantial influence on software quality.
-For example, \citet{Bravenboer:2007:PIA:1289971.1289975} finds that literal notation for structured encodings of queries, like the SQL-based query literals now found in many languages \cite{meijer2006linq}, reduce the temptation to use string encodings of queries and therefore reduce the prevalence of string injection attacks, which are amongst the most common and serious security vulnerabilities on web servers today \cite{owasp2017}.
-More generally, evidence suggests that programmers frequently engage in ``stringly-typed programming'', i.e.
-they choose strings instead of composite data structures largely for reasons of notational convenience.
-In particular, \citet{TSLs} sampled strings from open source projects and found that about 15\% of them could be parsed by some readily apparent type-specific grammar.
-Literal notation, with support for splicing, would decrease the syntactic cost of composite encodings of such data, which are more amenable to programmatic manipulation and compositional reasoning than flat strings.
-%Literal forms would  preferable to string encodings because they support compositional construction and decomposition by pattern matching via splicing.
-
-\begin{figure*}
+\begin{figure}[h]
+\vspace{-10px}
 \begin{lstlisting}
@@
 \end{lstlisting}
-\vspace{-10px}
+\vspace{-4px}
 \caption{HTML literals with support for splicing at two different types are built primitively into Ur/Web \cite{conf/popl/Chlipala15}.}
-\vspace{-10px}
 \label{fig:urweb}
-\end{figure*}
+\vspace{-8px}
+\end{figure}
+
+This design practice, where the language designer privileges certain user-defined types with built-in literal notation, is decidedly \emph{ad hoc} in that it is easy to come up with other examples of data structures for which mathematicians, scientists or programmers have invented specialized notation \cite{DBLP:journals/cacm/Iverson80,cajori1928history}.
+For example, 1) clients of a ``collections'' library might want not just list literals, but also literal notation for matrices, finite sets, maps, hash tables and so on, following usual mathematical conventions; 2) clients of a ``web programming'' library might like not just HTML literals but also CSS literals (which Ur/Web lacks); 3) a compiler author might like ``quotation'' literals for the terms of the object language and various intermediate languages of interest; and 4) clients of a ``chemistry'' library might like chemical structure literals based on the SMILES standard \cite{anderson1987smiles}.
+
+Although requests for specialized literal notation are easy to dismiss as superficial, the reality is that literal notation, or the absence thereof, can have a substantial influence on software quality.
+For example, \citet{Bravenboer:2007:PIA:1289971.1289975} finds that literal notation for structured encodings of queries, like the SQL-based query literals now found in many languages \cite{meijer2006linq}, reduce the temptation to use string encodings of queries and therefore reduce the risk of string injection attacks, which are amongst the most common and serious security vulnerabilities in web applications \cite{owasp2017}.
+More generally, evidence suggests that programmers frequently engage in ``stringly-typed programming'', i.e.
+they choose strings instead of composite data structures largely for reasons of notational convenience.
+In particular, \citet{TSLs} sampled strings from open source projects and found that about 15\% of them could be parsed by some readily apparent type-specific grammar.
+Literal notation, with support for splicing, would decrease the syntactic cost of composite encodings of such data, which are more amenable to programmatic manipulation and compositional reasoning than strings.
+%Literal forms would  preferable to string encodings because they support compositional construction and decomposition by pattern matching via splicing.
 
 Of course, it would not be sensible to ask general-purpose language designers to accomodate all known notations with built-in literal forms.
-Instead, there has been considerable interest for many years in mechanisms that allow library providers to define new literal notation on their own.
-For example, direct grammar extension systems like Camlp4 \cite{ocaml-manual} and SugarJ \cite{erdweg2011sugarj,erdweg2013framework}, macro-based systems like Omar et al's \emph{type-specific languages} (TSLs) \cite{TSLs}, and others that we will discuss in Sec.
-\ref{sec:existing-approaches} can all be used to define new literal notation.
+Instead, there has been considerable and persistent interest in mechanisms that allow library providers to define new literal notation on their own.
+For example, direct grammar extension systems like Camlp4 \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, macro-based systems like Omar et al's \emph{type-specific languages} (TSLs) \cite{TSLs}, and others that we will discuss in Sec.~\ref{sec:existing-approaches} can all be used to define new literal notation (and, in some cases, other forms of new notation, such as new infix operators and type declaration forms, which are beyond the scope of this paper)\todo{normalize parens-period thing}.
 
 \paragraph{Problem} The problem that motivates this paper is that it is impossible to reason abstractly, i.e.
-by inspecting only the surface syntax, about such fundamental issues as types and variable binding when the program text is littered with  unfamiliar user-defined literal forms.
-Instead, programmers can only reason transparently, i.e.
-by inspecting the underlying expansion or the code responsible for producing the expansion.
-In some existing systems, even determining which extension is responsible for a literal form is difficult.
-%, i.e. when developing large programs ``consisting of many small programs (modules), possibly written by different people''
+by inspecting only the program as presented to the programmer, about such fundamental issues as types and variable binding when the program text is littered with  user-defined literal forms.
+Instead, programmers and editor services can only reason transparently, i.e.
+by inspecting the underlying expansion or the extension responsible for producing that expansion.
+In some existing systems, even determining which syntax extension is responsible for a literal form is difficult.
+%, i.e. when developing large programs ``consisting of many small programs o(modules), possibly written by different people''
+\newcommand*\Suppressnumber{%
+  \lst@AddToHook{OnNewLine}{%
+    \let\thelstnumber\relax%
+     \advance\c@lstnumber-\@ne\relax%
+    }%
+}
+
+\newcommand*\Reactivatenumber{%
+  \lst@AddToHook{OnNewLine}{%
+   \let\thelstnumber\origthelstnumber%
+   \advance\c@lstnumber\@ne\relax}%
+}
 
 \begin{figure}[t!]
-\begin{lstlisting}
+\begin{subfigure}[t]{0.48\textwidth}
+\begin{lstlisting}[xleftmargin=-2pt, morekeywords={context,free,syntax}]
+  context-free syntax 
+    (* ...direct mods to the CFG... *)
+  (* ... *)
   let w = compute_w();
   let x = compute_x(w);
-  let y = [|(!R)@&{&/x!/:2_!x}'!R}|];
+  let y = `(!R)@&{&/x!/:2_!x}'!R`;
 \end{lstlisting}
+\vspace{-5px}
+\caption{It is difficult to reason abstractly about programs that use unfamiliar grammar extensions (see the five reasoning criteria in the text).}
+\label{fig:K-dialect}
+\end{subfigure}
+\hfill
+\begin{subfigure}[t]{0.51\textwidth}
+\begin{lstlisting}[xleftmargin=-2pt]
+  notation $kq at KQuery.t by KQueryParser 
+    using module Helpers as Helpers;
+  (* ... *)
+  let w = compute_w();
+  let x = compute_x(w);
+  let y = $kq `(SURL(!R)@&{&/EURLxSURL!/:2_!EURLxSURL}'!REURL)`;
+\end{lstlisting}
+\vspace{-5px}
+\caption{TLMs make examples like these more reasonable by leaving the grammar fixed and making the segmentation and the type and binding structure explicit.}
+\label{fig:K-tsm-example}
+\end{subfigure}
+\caption{Literal notation for encodings of K queries}
 \vspace{-10px}
-\caption{Syntax extension systems produce unreasonable program text.}
-\vspace{-7px}
-\label{fig:K-dialect}
 \end{figure}
@@
 Consider, for example, the perspective of a programmer attempting to comprehend, i.e.
-reason about, the program text in Figure \ref{fig:K-dialect}, which is written in a dialect of OCaml's syntax called Reason \cite{reason-what}\footnote{This paper uses Reason 2.0 as 3.0 was released just before submission.} that has, hypothetically, been extended with some number of new literal forms by a grammar extension system like Camlp4 \cite{ocaml-manual} or Sugar*~\cite{erdweg2011sugarj,erdweg2013framework}.
-The literal form on Line 3 constructs an encoding of a query in the niche stack-based database query language K, using its intentionally terse notation \cite{Whitney:2001:LOR:376284.375783}.
+reason about, the program text in Figure \ref{fig:K-dialect}, which is written in a dialect of OCaml's surface syntax called Reason \cite{reason-what}\todo{convert to Reason 3}{} that has, hypothetically, been extended with some number of new literal forms by a grammar extension system like Camlp4 \cite{ocaml-manual} or Sugar*~\cite{erdweg2011sugarj,erdweg2013framework} (Line 1 suggestively uses the syntax for introducing a grammar extension from SugarHaskell \cite{erdweg2012layout}).
+The literal form on Line 6 constructs an encoding of a query in the niche stack-based database query language K, using its intentionally terse notation \cite{Whitney:2001:LOR:376284.375783}.
 The problem is that a programmer unfamiliar with this notation cannot easily answer questions like the following:
 
-\begin{enumerate}[nolistsep,leftmargin=10pt]
-\item \textbf{(Responsibility)} Which syntax extension determined the expansion of the literal form on Line 3?
+\begin{enumerate}[leftmargin=12px]
+\item[\Circle] \textbf{(Responsibility)} Which syntax extension determined the expansion of the literal form on Line 6?
 Might activating a new extension generate a conflicting expansion for the same form (i.e.
-is there syntactic ambiguity)?
-\item \textbf{(Segmentation)} Are the characters \li{x}, \li{R} and \li{2} on Line 3 parsed as spliced expressions (meaning that they appear directly in the underlying expansion), or are they parsed in some other way peculiar to this literal form (e.g.
+might there be a syntactic ambiguity)?
+\item[\Circle] \textbf{(Segmentation)} Are the characters \li{x}, \li{R} and \li{2} on Line 6 parsed as spliced expressions (meaning that they appear directly in the underlying expansion), or are they parsed in some other way peculiar to this literal form (e.g.
 as operators in the K query language)?
-\item \textbf{(Capture)} If \li{x} is in fact a spliced expression, does it refer to the binding of \li{x} on Line 2, or might it capture an unseen binding of the same identifier in the expansion of Line 3?
-\item \textbf{(Context Dependence)} If \li{w}, on Lines 1-2, is renamed, could that possibly break Line 3 or change its meaning because the expansion assumes that \li{w} is in scope?% What other hidden assumptions might the desugaring be making?
-\item \textbf{(Typing)} What type does \li{y} have?
+\item[\Circle] \textbf{(Capture)} If \li{x} is in fact a spliced expression, does it refer to the binding of \li{x} on Line 5, or might it capture an unseen binding of the same identifier in the expansion of Line 6?
+\item[\Circle] \textbf{(Context Dependence)} If \li{w}, on Lines 4-5, is renamed, could that possibly break Line 6 or change its meaning because the expansion assumes that \li{w} is in scope?% What other hidden assumptions might the desugaring be making?
+\item[\Circle] \textbf{(Typing)} What type does \li{y} have? What type is each spliced expression expected to have?
 \end{enumerate}
@@
 Analagous problems do not arise when programming without syntax extensions in languages like ML and Scala --- programmers can reason lexically about where variables and other symbols are bound, and types mediate abstraction over function and module implementations \cite{B304}.
+Ideally, the programmer would be able to abstract in some analagous manner over the implementation of a notation.
 % From this perspective, it is unsurprising that the use of Camlp4 has been deprecated in the OCaml system.
 
-An approach that has made some headway against these problems is Omar et al.'s \emph{type-specific languages} (TSLs) \cite{TSLs}.
-With this approach, the language delegates control over the parsing and expansion of certain \emph{generalized literal forms} to a parse function associated with the type that the form is being checked against.
-Generalized literal forms syntactically subsume other literal forms because their bodies---the characters between the outer delimiters, here \li{[|} and \li{|]}---are left  unparsed by the fixed context-free syntax of the language (i.e.
-they are initially parsed like raw string literals.) With TSLs, Line 3 of our example would require a type annotation:
+An approach that has made some headway toward this ideal is Omar et al.'s \emph{type-specific languages} (TSLs) \cite{TSLs}.
+With this approach, the language delegates control over the parsing and expansion of certain \emph{generalized literal forms} to a parser associated with the type that the form is being checked against.
+With TSLs, Line 3 of our example would require a type annotation:
 \begin{lstlisting}[numbers=none]
-    let y : kquery = [|(!R)@&{&/x!/:2_!x}'!R}|]
+  let y : KQuery.t = `(!R)@&{&/x!/:2_!x}'!R`
 \end{lstlisting}
-The parse function associated with the \li{kquery} type would be invoked to parse and expand the literal body.
-
-TSLs support reasoning abstractly about \textbf{Responsibility} and \textbf{Typing}, due to the type-driven dispatch mechanism, but the remaining reasoning principles were not fully and formally established in the prior work.
-Moreover, the prior work considered only a monomorphic, nominally-typed language with local type inference, leaving open a number of problems that came up as we considered integrating TSLs into Reason/OCaml.
-These included non-trivial problems related to support for structural types (e.g.
-arrow types), parameterized types, abstract types, modules, pattern notation and ML-style type inference.
-Another problem was that parse functions and expansions were completely restricted from accessing libraries.
-Taken together, we found that TSLs as described were not suitable for integration into Reason/OCaml.
-
-\paragraph{Contributions} The aim of this paper is to address all of these deficiencies, resulting in the first system for defining new literal notation that (1) provably maintains all five of the abstract reasoning principles just outlined; and (2) that is expressive enough for integration into Reason/OCaml and other full-scale statically typed languages.
-We evaluate the first claim by developing a series of formal calculi equipped with proofs of these reasoning principles, and the second claim with several non-trivial case studies that appear throughout the paper and involve all of the aforementioned language features.
-The empirical study in the prior work on TSLs \cite{TSLs} supports the claim that a macro-based mechanism for defining new literal notation is broadly useful, so this paper is targeted at language designers that seek a deeper understanding of the reasoning principles at play, particularly as they relate to features common to ML and other full-scale statically typed languages (e.g.
+The parser associated with the type \li{KQuery.t} would be statically invoked to parse and expand the literal body, i.e.
+the characters between the outer delimiters, here backticks.
+Generalized literal forms syntactically subsume other literal forms because their bodies are left unparsed by the fixed context-free syntax of the language (i.e.
+they are initially parsed like raw string literals.) TSLs are closely related to the mechanism proposed in this paper, so let us analyze TSLs from the perspective of the five reasoning criteria just outlined.
+\begin{enumerate}[leftmargin=12px]
+  \item[\CIRCLE] \textbf{(Responsibility)} The type-directed dispatch mechanism allows the client to easily determine which parser is responsible for each literal form.
+The context-free syntax of the language remains fixed and composition is mediated by splicing, so clients do not need to worry about different TSLs conflicting syntactically with one another.
+  \item[\Circle] \textbf{(Segmentation)} The TSL can splice expressions out of the literal body, but there is no way to associate spliced terms with segments of the literal body, nor is there a guarantee that spliced expressions are non-overlapping.
+  \item[\Circle] \textbf{(Capture)} The TSL mechanism as described does not guarantee that splicing is capture-avoiding.
+  \item[\LEFTcircle] \textbf{(Context Independence)} The TSL mechanism as described requires that the expansion be completely closed.
+Although no theorem about context independence was established, this appears to maintain context independence.
+However, this comes at a significant cost -- expansions cannot use any libraries whatsoever.
+  \item[\LEFTcircle] \textbf{(Typing)} Each parser is associated with a type, and the generated expansion must necessarily be of the expected type.
+However, there is no easy way to determine the type expected for each spliced expression.
+In addition, the prior work on TSLs considered only a monomorphic, nominally-typed language with only local type inference, leaving open a number of problems that came up as we considered integrating TSLs into Reason/OCaml.
+These included non-trivial problems related to support for:
+\begin{enumerate}[leftmargin=15px]
+  \item \textbf{Structural Types}: There is no way to define literal notation for structural types, e.g. tuple and arrow types.
+  \item \textbf{Parameterized Types}: There is no way to define literal notation over a type-parameterized family of types, e.g. for all types \li{list('a)}.
+  \item \textbf{Abstract Types}: There is no way to define literal notation over a module-parameterized family of abstract types, e.g.
+for every abstract type defined by a module implementing the \li{QUEUE} signature.
+  \item \textbf{ML-style Type Inference}: The type-directed dispatch mechanism relied on strictly local type inference, specified as a bidirectional type system \cite{Pierce:2000:LTI:345099.345100}.
+It is not clear that this is suitable for languages with ML-style constraint-based type inference.
+  \item \textbf{Pattern Literals}: TSLs support only expression literals.
+No consideration was given to literals in pattern position, which are dual to expression literals in languages with support for ML-style pattern matching.
+\end{enumerate}
+\end{enumerate}
+Taken together, we found that no existing approach for defining literal notation was suitable as-is for integration into Reason/OCaml.
+
+\paragraph{Contributions} The aim of this paper is to introduce \emph{typed literal macros} (TLMs): a new system for defining literal notation that addresses all of these deficiencies.
+The result is the first system for defining new literal notation that (1) provably maintains all five of the abstract reasoning principles just outlined; and (2) that is expressive enough for integration into Reason/OCaml and other full-scale statically typed languages.
+We evaluate the first claim by developing a series of formal calculi equipped with proofs of these reasoning principles, and the second claim with a number of non-trivial examples that appear throughout the paper and involve the language features marked \textbf{(a)-(e)} above.
+In describing these examples, we will also demonstrate that the literal notation itself can be defined using off-the-shelf parser generators, so the burden on notation providers is comparable to that of existing systems.
+
+% The empirical study in the prior work on TSLs \cite{TSLs} supports the claim that a macro-based mechanism for defining new literal notation is broadly useful, so This paper is targeted at language designers that seek a deeper understanding of the reasoning principles at play, particularly as they relate to features common to ML and other full-scale statically typed languages (e.g.
 Haskell, Scala, Rust, etc.)
 
-\begin{figure}
-\begin{lstlisting}
-  let w = compute_w();
-  let x = compute_x(w);
-  let y = $kquery [|SURL(!R)@&{&/EURLxSURL!/:2_!EURLxSURL}'!R}EURL|];
-\end{lstlisting}
-\vspace{-10px}
-\caption{TLMs make examples like the one from Figure \ref{fig:K-dialect} more reasonable.}
-\vspace{-7px}
-\label{fig:K-tsm-example}
-\end{figure}
-
-The most obvious point of departure between TSLs and \emph{typed literal macros} (TLMs), which are the topic of this paper, is that we separate out the mechanism for defining new literal notation from the type definition mechanism of the language.
+The most obvious point of departure between TSLs, described above, and TLMs, which are the topic of this paper, is that we separate out the mechanism for defining new literal notation from the type definition mechanism of the language.
 TLMs are instead explicitly named and applied.
-For example, Figure \ref{fig:K-tsm-example} applies a TLM named \li{$kquery} to a generalized literal form delimited by \li{[|} and \li{|]} to express the K query example just discussed.
+For example, Figure \ref{fig:K-tsm-example} applies a TLM named \li{$kq} to a generalized literal form delimited by \li{`(} and \li{)`} to express the K query example just discussed.
 The applied TLM parses the literal body and generates an expansion statically.
@@
 \begin{enumerate}[nolistsep,leftmargin=10pt]
-\item \textbf{(Responsibility)} The applied TLM, \li{$kquery}, is delegated responsibility for parsing the literal body.
-The context-free syntax of the language is fixed, so the TLM provider can modularly establish that the syntax that the TLM implements is unambiguous (even in situations where a spliced term itself applies a TLM, shown in a later example.)
-\item \textbf{(Segmentation)} The intermediate output that the TLM generates is structured so that the system can infer from it an accurate \emph{segmentation} of the literal body that distinguishes spliced terms from segments that are parsed in some other way by the TLM.
+\item[\CIRCLE] \textbf{(Responsibility)} The applied TLM, \li{$kq}, is delegated responsibility for parsing the literal body.
+As with TSLs, the context-free syntax of the language is fixed, so the TLM provider can modularly establish that the syntax that the TLM implements is unambiguous (even in situations where a spliced term itself applies a TLM, shown in a later example.)
+\item[\CIRCLE] \textbf{(Segmentation)} The intermediate output that the TLM generates is structured so that the system can infer from it an accurate \emph{segmentation} of the literal body that distinguishes spliced terms from segments that are parsed in some other way by the TLM.
 The segmentation can be communicated by a code editor using secondary notation, e.g.
@@
 as operators in the K language (because they are  green.)
-\item \textbf{(Capture)} Splicing is capture-avoiding, so the spliced expression \li{x} must refer to the binding of \li{x} on Line 2.
+\item[\CIRCLE] \textbf{(Capture)} Splicing is capture-avoiding, so the spliced expression \li{x} must refer to the binding of \li{x} on Line 2.
 It cannot capture a coincidental binding of \li{x} in the expansion.
-\item \textbf{(Context Dependence)} The system enforces context independence, so the expansion of Line 3 cannot  rely on the fact that, e.g., \li{w} is in scope. The client is free to rename \li{w}.
-\item \textbf{(Typing)} A type annotation on the definition of \li{$kquery} (not shown) determines the type that the expansion must have.
+\item[\CIRCLE] \textbf{(Context Dependence)} The system enforces context independence, so the expansion of Line 3 cannot  rely on the fact that, e.g., \li{w} is in scope.
+The client is free to rename \li{w}.
+The library dependencies that the generated expansion might have are explicit (here, the expansion uses the module \li{Helpers}), and the system ensures that this dependency is available even if the identifier \li{Helpers} has been shadowed at the use site.
+\item[\CIRCLE] \textbf{(Typing)} The type annotation on the definition of \li{$kq} determines the type that the expansion must have, here \li{KQuery.t}.
 We will see an example of a TLM definition in the next section.
@@
 %This information is usually not necessary to reason about typing, but it can be conveyed to the programmer upon request by the program editor if desired.
-%TLM definitions follow the usual scoping rules, so it is easy to ``jump to the definition'' of \li{$kquery}.
+%TLM definitions follow the usual scoping rules, so it is easy to ``jump to the definition'' of \li{$kq}.
 \end{enumerate}
@@
 \end{lstlisting}
-\vspace{-5px}
 \caption{The \li{html} type, which classifies encodings of HTML data.}
@@
 \end{lstlisting}
-\vspace{-5px}
 \caption{The \li{$html} TLM definition. Figure \ref{fig:candidate-exp-Reason} defines   TLM-related types.}
@@
 \end{lstlisting}
-\vspace{-5px}
 \caption{Two examples of the \li{$html} TLM being applied.
@@
 \end{subfigure}
-\vspace{-8px}
 \caption{Case Study: HTML literals}
-\vspace{-7px}
 \end{figure*}
 
-\vspace{-3px}
 \paragraph{Outline} 
@@
 
-\vspace{-6px}
 \newcommand{\seTLMsSec}{Expression TLMs}
@@
 \label{sec:setsms}
-% \vspace{-3px}
 
@@
 
-% \vspace{-8px}
 \subsection{TLM Definition and Application}
-\vspace{-2px}
 The definition of the \li{$html} TLM is given in Figure \ref{fig:html-tlm-def}. TLM names are prefixed by \li{$} 
@@
 
-\vspace{-4px}
 \subsection{Splicing}\label{sec:splicing-and-hygiene}
-\vspace{-2px}
 When the parse function determines that some segment of the literal body is a spliced expression, according to whatever syntactic criteria it deems suitable, it can indirectly refer to it in the encoding it produces using the \li{SplicedE} constructor of \li{proto_expr}, which takes a value of type \li{segment} that indicates the zero-indexed location of the spliced expression relative to the start of the provided literal body.
@@
 \begin{figure}
-\vspace{-2px}
 \begin{lstlisting}[numbers=none,xleftmargin=0pt]
@@
 \end{lstlisting}
-\vspace{-10px}
 \caption[Definitions of various types used by TLM definitions.]{Definitions of various types available ambiently to TLM definitions.}
@@
 \label{fig:candidate-exp-Reason}
-\vspace{-14px}
 \end{figure}
@@
 For example, consider a TLM \li{$smiles} for chemical structures \cite{anderson1987smiles} with support for splicing using curly braces:
-\vspace{-3px}
 \begin{lstlisting}[numbers=none]
@@
 \end{lstlisting}
-\vspace{-2px}
 %For example, if strings were not primitive but rather defined as sequences of characters, we might define a TLM \li{$str} to recover string literal notation and write \li{$html [|SURL<body>\{EURLheading $str [|SSTRWorld!ESTR|]SURL\} ...</body>EURL|]}.
@@
 
-\vspace{-6px}
 \subsection{Proto-Expansion Validation}\label{sec:uetsms-validation}
-\vspace{-2px}
 Three important concerns described in Sec.
@@
 \subsubsection{Capture}\label{sec:capture}
-\vspace{-3px}
 
@@
 For example, consider the following application site:
-\vspace{-3px}
 \begin{lstlisting}[numbers=none]
@@
 \end{lstlisting}
-\vspace{-2px}
 Now consider the scenario where the proto-expansion generated by \li{$html} has the following form:
-\vspace{-3px}
 \begin{lstlisting}[numbers=none]
@@
 \end{lstlisting}
-\vspace{-2px}
 Na\"ively, the binding of the variable \li{tmp} in the proto-expansion could shadow the application-site binding of \li{tmp} in the final expansion. 
@@
 
-\vspace{-7px}
 \subsubsection{Context Dependence}\label{sec:context-dependence}
-\vspace{-3px}
 %The prohibition on shadowing ensures only that variables that appear in spliced terms do not refer to bindings that appear in the surrounding expansion. 
@@
 A minimal example of a ``broken'' TLM that does not generate context-independent proto-expansions is below:
-\vspace{-3px}
 \begin{lstlisting}[numbers=none]
@@
 \end{lstlisting}
-\vspace{-2px}
 The proto-expansion that this TLM generates (for any literal body) refers to a variable \li{x} that it does not itself bind.
@@
 
-\vspace{-6px}
 \subsubsection{Typing}\label{sec:typing-e}
-\vspace{-3px}
 Finally, proto-expansion validation maintains a reasonable {typing discipline} by (1) checking that the expansion is of the type specified by the TLM's type annotation; (2) checking that each spliced type is valid; (3) checking that the type annotation on each spliced expression is valid; and (4) checking each spliced expression against the specified type annotation.
@@
 
-\vspace{-6px}
 \newcommand{\spTLMsSec}{Pattern TLMs}
@@
 \label{sec:sptsms}
-\vspace{-3px}
 
@@
 For example, we can pattern match on a value \li{x : html} by applying a pattern TLM \li{$html} as follows:
-\vspace{-3px}
 \begin{lstlisting}[numbers=none]
@@
 \end{lstlisting}
-\vspace{-2px}
 Any list pattern, including one generated by another TLM application, can appear where \li{cs} appears in the example pattern above.
@@
 The type \li{proto_pat}, outlined below, classifies encodings of \emph{proto-patterns}.
-\vspace{-3px}
 \begin{lstlisting}[numbers=none]
@@
 \end{lstlisting}
-\vspace{-1px}
 
@@
 
-\vspace{-7px}
 \newcommand{\pTLMsSec}{Parametric TLMs}
@@
 \label{sec:ptsms}
-\vspace{-2px}
 
@@
 Partial parameter application decreases the syntactic cost of this explicit parameter passing style.
-% \vspace{-4px}
 % \subsection{Parametric TLMs By Example}\label{sec:ptsms-by-example}
-% \vspace{-2px}
 
 Consider the following Reason/OCaml module type (a.k.a. signature), which specifies an abstract data type \cite{liskov1974programming,harper1997programming} of string-keyed polymorphic dictionaries:
-\vspace{-3px}
 \begin{lstlisting}[numbers=none]
@@
 \end{lstlisting}
-\vspace{-2px}
 We can define a TLM that is parametric over implementations of this signature, \li{D : DICT}, and over choices of the codomain type, \li{'a} as follows:
-\vspace{-3px}
 \begin{lstlisting}[numbers=none]
@@
 \end{lstlisting}
-\vspace{-2px}
 For example, given some module that implements this signature, \li{HashDict : DICT}, we can apply \li{$dict} as follows:
-\vspace{-3px}
 \begin{lstlisting}[numbers=none]
@@
 \end{lstlisting}
-\vspace{-2px}
 Notice that the segmentation immediately reveals which punctuation is particular to this TLM and where the spliced key and value expressions appear.
@@
 The proto-expansion generated for the TLM application above, shown below, can refer to the parameters:
-\vspace{-3px}
 \begin{lstlisting}[numbers=none]
@@
 \end{lstlisting}
-\vspace{-2px}
 Validation checks that the proto-expansion is truly parametric, i.e.
@@
 If we will use the \li{HashDict} implementation ubiquitously, we can abbreviate the partial application of \li{$dict} to \li{HashDict}, resulting in a TLM that is parametric over only the type \li{'a}:
-\vspace{-3px}
 \begin{lstlisting}[numbers=none]
@@
 \end{lstlisting}
-\vspace{-2px}
 TLM abbreviations can themselves be parameterized to support partial application of parameters other than the last.
@@
 \end{minipage}
-\vspace{-10px}
 \caption[Syntax of the $\miniVersePat$ unexpanded language (UL)]{Syntax of the $\miniVersePat$ unexpanded language (UL).
@@
 }
-\vspace{-10px}
 \label{fig:U-unexpanded-terms}
@@
 \begin{comment}
-\vspace{-6px}
 \newcommand{\staticEvalSec}{Static Evaluation}
@@
 \label{sec:static-eval}
-\vspace{-2px}
 
@@
 
-\vspace{-4px}
 \subsection{The Static Environment}
-\vspace{-2px}
 Figure \ref{fig:static-module-example} shows an example of a module, \li{ParserCombos}, that defines a number of parser combinators following \citet{Hutton1992d}.
@@
 \begin{figure}[h]
-\vspace{-10px}
 \begin{lstlisting}[numbers=none]
@@
 \end{lstlisting}
-\vspace{-10px}
 \caption{Binding static modules and values for use within parse functions.}
 \label{fig:static-module-example}
-\vspace{-10px}
 \end{figure}
@@
 
-\vspace{-6px}
 \subsection{Applying TLMs Within TLM Definitions}\label{sec:tsms-for-tsms}
@@
 
-\vspace{-6px}
 \subsubsection{Quasiquotation}\label{sec:quasiquotation}
@@
 
-\vspace{-6px}
 \subsubsection{Parser Generators}
@@
 % \begin{figure}[h!]
-% \vspace{-5px}
 % \begin{lstlisting}[deletekeywords={as}]
@@
 % \end{lstlisting}
-% \vspace{-12px}
 % \caption{A grammar-based definition of \texttt{\$rx}.}
-% \vspace{-15px}
 % \label{fig:rx-grammar-based}
@@
 
-\vspace{-6px}
 \subsection{Static Evaluation, Formally}
@@
 
-\vspace{-6px}
 \subsection{Library Management}
@@
 
-\vspace{-4px}
 \newcommand{\seTLMsFormallySec}{Simple TLMs, Formally}
@@
 \label{sec:setsms-formally}
-\vspace{-2px}
 
@@
 
-\vspace{-6px}
 \subsection{Expanded Language (XL)}\label{sec:s-XL}
-\vspace{-2px}
 The {XL} of $\miniVersePat$ forms a standard pure functional language with partial function types, quantification over types, recursive types, labeled product types and labeled sum types and support for pattern matching.
@@
 
-% \vspace{-4px}
 % \subsubsection{Evaluation Semantics}\label{sec:dynamics-U}
-% \vspace{-2px}
 The \emph{evaluation semantics} of $\miniVersePat$ is organized around the judgements $\isvalU{e}$, which says that $e$ is a value, and $\evalU{e}{e'}$, which says that $e$ evaluates to the value $e'$.% Additional  judgements, not shown, are needed to define the dynamics of pattern matching, but they do not appear directly in our subsequent developments, so we omit them.
@@
 
-\vspace{-6px}
 \subsection{Syntax of the Unexpanded Language}\label{sec:syntax-U}\label{sec:s-UL}
-\vspace{-2px}
 Unexpanded types and expressions are simple inductive structures.
@@
 \end{minipage}
-\vspace{-10px}
 \caption[Syntax of the XL of $\miniVersePat$]{Syntax of the expanded language (XL).
@@
 }
-\vspace{-10px}
 \label{fig:U-expanded-terms}
@@
 \end{mathpar}}
-\vspace{-12px}
 \caption{An example expansion derviation demonstrating how identifiers and variables are separately tracked.}
 \label{fig:expansion-exmpl}
-\vspace{-10px}
 \end{figure*}
 
-\vspace{-6px}
 \subsection{Typed Expansion}\label{sec:typed-expansion-U}\label{sec:s-TE}
-\vspace{-3px}
 Unexpanded terms are checked and expanded simultaneously according to the central \emph{typed expansion judgements}:
-\vspace{-5px}\[\begin{array}{ll}
+\[\begin{array}{ll}
 % \textbf{Judgement Form} & \textbf{Description}\\
@@
 \patExpands{\upctx}{\uPhi}{\upv}{p}{\tau} & \text{$\upv$ has expansion $p$ matching $\tau$}
-\end{array}\]\vspace{-5px}
+\end{array}\]
 %\newcommand{\gray}[1]{{\color{gray} #1}}
@@
 
-\vspace{-6px}
 \subsection{TLM Definitions}\label{sec:U-uetsm-definition}\label{sec:s-TLM-def}
-\vspace{-2px}
 % An unexpanded expression of seTLM definition form, $\uesyntaxq{\tsmv}{\utau}{\eparse}{\ue}$, 
@@
 \end{mathpar}}
-\vspace{-12px}
 \caption{The typed expansion rules for expression TLM definition and application.}
 \label{fig:ee-def-setsm}
-\vspace{-10px}
 \end{figure*}
@@
 
-\vspace{-4px}
 \subsection{TLM Application}\label{sec:U-uetsm-application}\label{sec:s-TLM-ap}
-\vspace{-2px}
 The unexpanded expression form for applying an seTLM named $\tsmv$ to a literal form with literal body $b$ is $\utsmap{\tsmv}{b}$. 
@@
 
-\vspace{-6px}
 \subsection{Proto-Expansion Validation}\label{sec:ce-validation-U}\label{sec:ce-syntax-U}\label{sec:s-PEV}
-\vspace{-2px}
 
@@
 \end{minipage}
-\vspace{-10px}
 \caption[Syntax of proto-types and proto-expressions]{Syntax of proto-expansions. Proto-expansion terms are ABTs identified up to alpha-equivalence.}
 \label{fig:U-candidate-terms}
-\vspace{-10px}
 \end{figure}
@@
 Finally, we arrive at the crucial \emph{proto-expansion validation judgements}, which validate the proto-expansions generated by TLMs and simultaneously generate their final expansions:% are types and expanded expressions, respectively.
-\vspace{-5px}\[\begin{array}{ll}
+\[\begin{array}{ll}
 \cvalidT{\Delta}{\tscenev}{\ctau}{\tau} & \text{$\ctau$ has well-formed expansion $\tau$}\\
@@
 \end{array}\]
-\vspace{-5px}
 
@@
 \end{mathpar}}
-\vspace{-5px}
+
 % {\small\begin{mathpar}
@@
 
-\vspace{-6px}
 \subsection{Metatheory}\label{sec:s-metatheory}
-\vspace{-1px}
+
 \subsubsection{Typed Expansion} The first property that we are interested in is simple: that typed expansion produces a well-typed expansion.
@@
 \end{theorem}
-\vspace{-5px}
+
 % \endgroup
@@
 
-\vspace{-6px}
 \subsubsection{seTLM Reasoning Principles}
-\vspace{-3px}
+
 The following theorem summarizes the abstract reasoning principles that programmers can rely on when applying a simple expression TLM.
@@
 % \def\thetheorem{\ref{thm:tsc-SES}}
-\vspace{-4px}
+
 \begin{theorem}[seTLM Reasoning Principles]\label{thm:setlm-reasoning} ~\\
@@
 \end{theorem}
-\vspace{-4px}
 
@@
 % \end{theorem}
-% \vspace{-3px}
+% 
 % \begin{proof} 
@@
 
-\vspace{-6px}
 \newcommand{\rwSec}{Related Work}
@@
 \subsection{Syntax Definition Systems}\label{sec:syntax-dialects-intro}
-\vspace{-3px}
 One approach available to library providers seeking to introduce new literal forms is to use a syntax definition system to construct a library-specific \emph{syntax dialect}: a new syntax definition that extends the syntax definition given in the language definition with new forms, including literal forms.% For example, Ur/Web's syntax (Figure 1) is a library-specific dialect of Ur's syntax \cite{conf/pldi/Chlipala10,conf/popl/Chlipala15}.
@@
 
-\vspace{-6px}
 \subsection{Term Rewriting Systems}
-\vspace{-3px}
 Another approach -- and the approach that TLMs are rooted in -- is to leave the context-free syntax of the language fixed, and instead contextually rewrite existing literal forms.
@@
 We could  mark a string literal as follows:
-\vspace{-3px}
 \begin{lstlisting}[numbers=none]
@@
 \end{lstlisting}
-\vspace{-2px}
 More than one applied preprocessor might recognize this annotation (there are, in practice, many XML/HTML libraries), so the problems of \textbf{Responsibility} comes up.
@@
 Consider again our running XHTML example, which we might try to realize by applying a hygienic macro, \li{xml!}, to a string literal that the macro parses:
-\vspace{-3px}
 \begin{lstlisting}[numbers=none]
@@
 \end{lstlisting}
-\vspace{-2px}
 The expansion of this macro will fail the check for application-site context independence because \li{first_name} will appear as a free variable, in no way different from any other.
@@
 
-\vspace{-6px}
 \newcommand{\discussionSec}{Discussion}
@@
 \label{sec:conclusion}
-\vspace{-2px}
 
@@
 \begin{comment}\appendix
-\vspace{-4px}
 \subsection{Parametric TLMs, Formally}

### 2018-02-20 icfp18/syntax-icfp18.tex
more progress on intro

@@
 
+\vspace{-5px}
 \newcommand{\introSec}{Introduction}
@@
 
-Of course, it would not be sensible to ask general-purpose language designers to accomodate all known notations with built-in literal forms.
-Instead, there has been considerable and persistent interest in mechanisms that allow library providers to define new literal notation on their own.
+Of course, it would not be scalable to ask general-purpose language designers to build in support for all known notations \emph{a priori}.
+Instead, there has been persistent interest in mechanisms that allow library providers to define new literal notation on their own.
 For example, direct grammar extension systems like Camlp4 \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, macro-based systems like Omar et al's \emph{type-specific languages} (TSLs) \cite{TSLs}, and others that we will discuss in Sec.~\ref{sec:existing-approaches} can all be used to define new literal notation (and, in some cases, other forms of new notation, such as new infix operators and type declaration forms, which are beyond the scope of this paper)\todo{normalize parens-period thing}.
@@
   context-free syntax 
-    (* ...direct mods to the CFG... *)
-  (* ... *)
+    /* ...direct mods to the CFG... */
+  /* ... */
   let w = compute_w();
   let x = compute_x(w);
-  let y = `(!R)@&{&/x!/:2_!x}'!R`;
+  let y = `((!R)@&{&/x!/:2_!x}'!R)`;
 \end{lstlisting}
@@
     using module Helpers as Helpers;
-  (* ... *)
+  /* ... */
   let w = compute_w();
@@
 \end{subfigure}
-\caption{Literal notation for encodings of K queries}
+\vspace{3px}
+\caption{Two of the possible ways to introduce literal notation for encodings of K queries}
 \vspace{-10px}
@@
 
-Consider, for example, the perspective of a programmer attempting to comprehend, i.e.
-reason about, the program text in Figure \ref{fig:K-dialect}, which is written in a dialect of OCaml's surface syntax called Reason \cite{reason-what}\todo{convert to Reason 3}{} that has, hypothetically, been extended with some number of new literal forms by a grammar extension system like Camlp4 \cite{ocaml-manual} or Sugar*~\cite{erdweg2011sugarj,erdweg2013framework} (Line 1 suggestively uses the syntax for introducing a grammar extension from SugarHaskell \cite{erdweg2012layout}).
+\todo{show expr <- thing}Consider, for example, the perspective of a programmer attempting to comprehend, i.e.
+reason about, the program text in Figure \ref{fig:K-dialect}, which is written in a dialect of OCaml's surface syntax called Reason \cite{reason-what}\todo{convert to Reason 3}{} that has, hypothetically, been extended with some number of new literal forms by a grammar extension system like Camlp4 \cite{ocaml-manual} or Sugar*~\cite{erdweg2011sugarj,erdweg2013framework} (Line 1 suggestively borrows the syntax for introducing a grammar extension from SugarHaskell \cite{erdweg2012layout}).
 The literal form on Line 6 constructs an encoding of a query in the niche stack-based database query language K, using its intentionally terse notation \cite{Whitney:2001:LOR:376284.375783}.
@@
 
-\begin{enumerate}[leftmargin=12px]
+\begin{enumerate}[nolistsep,noitemsep,leftmargin=12px]
 \item[\Circle] \textbf{(Responsibility)} Which syntax extension determined the expansion of the literal form on Line 6?
-Might activating a new extension generate a conflicting expansion for the same form (i.e.
-might there be a syntactic ambiguity)?
+Might activating a new extension generate a conflicting expansion for the same form?
 \item[\Circle] \textbf{(Segmentation)} Are the characters \li{x}, \li{R} and \li{2} on Line 6 parsed as spliced expressions (meaning that they appear directly in the underlying expansion), or are they parsed in some other way peculiar to this literal form (e.g.
@@
 An approach that has made some headway toward this ideal is Omar et al.'s \emph{type-specific languages} (TSLs) \cite{TSLs}.
-With this approach, the language delegates control over the parsing and expansion of certain \emph{generalized literal forms} to a parser associated with the type that the form is being checked against.
-With TSLs, Line 3 of our example would require a type annotation:
-\begin{lstlisting}[numbers=none]
+Under this approach, the language delegates control over the parsing and expansion of certain \emph{generalized literal forms} to a parser associated with the type that the form is being checked against.
+So with TSLs, Line 3 of our example would require a type annotation:
+\begin{lstlisting}[numbers=none,basicstyle=\ttfamily\fontsize{9pt}{1em}\selectfont]
   let y : KQuery.t = `(!R)@&{&/x!/:2_!x}'!R`
@@
 The parser associated with the type \li{KQuery.t} would be statically invoked to parse and expand the literal body, i.e.
-the characters between the outer delimiters, here backticks.
-Generalized literal forms syntactically subsume other literal forms because their bodies are left unparsed by the fixed context-free syntax of the language (i.e.
-they are initially parsed like raw string literals.) TSLs are closely related to the mechanism proposed in this paper, so let us analyze TSLs from the perspective of the five reasoning criteria just outlined.
+the characters between the outer delimiters, here \li{`(} and \li{)`}, which are left unparsed by the fixed context-free syntax of the language (i.e.
+generalized literal forms are initially parsed like raw string literals.) TSLs are closely related to the mechanism proposed in this paper, so let us analyze TSLs from the perspective of the five reasoning criteria just outlined.
 \begin{enumerate}[leftmargin=12px]
-  \item[\CIRCLE] \textbf{(Responsibility)} The type-directed dispatch mechanism allows the client to easily determine which parser is responsible for each literal form.
+  \item[\CIRCLE] \textbf{(Responsibility)} The type-directed dispatch mechanism allows the client to easily determine which parser is responsible for each literal form -- the parser defined when the type was defined.
 The context-free syntax of the language remains fixed and composition is mediated by splicing, so clients do not need to worry about different TSLs conflicting syntactically with one another.
-  \item[\Circle] \textbf{(Segmentation)} The TSL can splice expressions out of the literal body, but there is no way to associate spliced terms with segments of the literal body, nor is there a guarantee that spliced expressions are non-overlapping.
-  \item[\Circle] \textbf{(Capture)} The TSL mechanism as described does not guarantee that splicing is capture-avoiding.
-  \item[\LEFTcircle] \textbf{(Context Independence)} The TSL mechanism as described requires that the expansion be completely closed.
-Although no theorem about context independence was established, this appears to maintain context independence.
-However, this comes at a significant cost -- expansions cannot use any libraries whatsoever.
-  \item[\LEFTcircle] \textbf{(Typing)} Each parser is associated with a type, and the generated expansion must necessarily be of the expected type.
+  \item[\Circle] \textbf{(Segmentation)} The TSL can splice expressions out of the literal body, but there is no clear way to associate each spliced term with some particular segment (i.e.
+subsequence) of the literal body, nor is there a guarantee that spliced expressions are non-overlapping.
+  \item[\Circle] \textbf{(Capture)} The TSL mechanism as described does not guarantee that splicing is capture-avoiding. (We will return to say more about this problem later.)
+  \item[\LEFTcircle] \textbf{(Context Independence)} The TSL mechanism as described requires that the expansion be completely closed, so it is trivially context independent.
+However, this comes at a significant expressive cost: the generated expansions cannot make use of any libraries whatsoever.
+  \item[\LEFTcircle] \textbf{(Typing)} Each parser is associated with a type, and the generated expansion must necessarily be of that type, so abstract reasoning about the type of the expansion is possible.
 However, there is no easy way to determine the type expected for each spliced expression.
-In addition, the prior work on TSLs considered only a monomorphic, nominally-typed language with only local type inference, leaving open a number of problems that came up as we considered integrating TSLs into Reason/OCaml.
-These included non-trivial problems related to support for:
+In addition, the prior work on TSLs considered only a monomorphic, nominally-typed language with only local type inference, leaving open a number of problems that came up as we considered integrating TSLs into Reason, which is a syntactic layer over all of OCaml \cite{ocaml-manual}:
 \begin{enumerate}[leftmargin=15px]
-  \item \textbf{Structural Types}: There is no way to define literal notation for structural types, e.g. tuple and arrow types.
-  \item \textbf{Parameterized Types}: There is no way to define literal notation over a type-parameterized family of types, e.g. for all types \li{list('a)}.
+  \item \textbf{Structural Types}: There is no way to define literal notation at structural types, e.g. tuple and arrow types.
+  \item \textbf{Parameterized Types}: There is no way to define literal notation over a type-parameterized family of types, e.g. at all types \li{list('a)}.
   \item \textbf{Abstract Types}: There is no way to define literal notation over a module-parameterized family of abstract types, e.g.
-for every abstract type defined by a module implementing the \li{QUEUE} signature.
-  \item \textbf{ML-style Type Inference}: The type-directed dispatch mechanism relied on strictly local type inference, specified as a bidirectional type system \cite{Pierce:2000:LTI:345099.345100}.
+at every abstract type defined by a module implementing the \li{QUEUE} signature.
+Parameterized and abstract types are ubiquitous in ML-family languages.% Programmers in ML-family languages commonly define various implementations of abstract data types in this way.
+  \item \textbf{ML-Style Type Inference}: The type-directed dispatch mechanism is specified by a bidirectional type system, which supports only local type inference \cite{Pierce:2000:LTI:345099.345100}.
 It is not clear that this is suitable for languages with ML-style constraint-based type inference.
   \item \textbf{Pattern Literals}: TSLs support only expression literals.
-No consideration was given to literals in pattern position, which are dual to expression literals in languages with support for ML-style pattern matching.
+No consideration was given to the problem of defining new literal notation in pattern position.
+Pattern matching is ubiquitous in ML-family languages.
+%Pattern literals are dual to expression literals in that expression literals support value construction, while pattern literals support value deconstruction.
 \end{enumerate}
 \end{enumerate}
-Taken together, we found that no existing approach for defining literal notation was suitable as-is for integration into Reason/OCaml.
-
-\paragraph{Contributions} The aim of this paper is to introduce \emph{typed literal macros} (TLMs): a new system for defining literal notation that addresses all of these deficiencies.
-The result is the first system for defining new literal notation that (1) provably maintains all five of the abstract reasoning principles just outlined; and (2) that is expressive enough for integration into Reason/OCaml and other full-scale statically typed languages.
-We evaluate the first claim by developing a series of formal calculi equipped with proofs of these reasoning principles, and the second claim with a number of non-trivial examples that appear throughout the paper and involve the language features marked \textbf{(a)-(e)} above.
-In describing these examples, we will also demonstrate that the literal notation itself can be defined using off-the-shelf parser generators, so the burden on notation providers is comparable to that of existing systems.
+We consider other existing approaches against these criteria in Sec.~\ref{sec:existing-approaches}.
+All together, we found that no existing approach was suitable as-is for integration into Reason/OCaml.
+
+\paragraph{Contributions} This paper introduces \emph{typed literal macros} (TLMs): a new system for defining literal notation that addresses all of these deficiencies.
+The result is the first system for defining new literal notation that (1) provably maintains all five of the abstract reasoning principles just outlined; and (2) is expressive enough for integration into Reason/OCaml and other full-scale statically typed languages.
+We evaluate the first claim by developing a series of formal calculi equipped with proofs of these abstract reasoning principles, and the second claim with a number of non-trivial examples that appear throughout the paper and involve the language features marked {(a)-(e)} above.
+In describing these examples, we also demonstrate that the literal parsers can be defined using off-the-shelf parser generators, so the burden on notation providers is comparable to that of existing systems for integrating new literal notation into a programming language.
 
@@
 For example, Figure \ref{fig:K-tsm-example} applies a TLM named \li{$kq} to a generalized literal form delimited by \li{`(} and \li{)`} to express the K query example just discussed.
-The applied TLM parses the literal body and generates an expansion statically.
-During the typing phase, a validation step ensures that programmers can reason abstractly about types, binding and segmentation.
-Taken together, the programmer can reason as follows:
-\begin{enumerate}[nolistsep,leftmargin=10pt]
-\item[\CIRCLE] \textbf{(Responsibility)} The applied TLM, \li{$kq}, is delegated responsibility for parsing the literal body.
-As with TSLs, the context-free syntax of the language is fixed, so the TLM provider can modularly establish that the syntax that the TLM implements is unambiguous (even in situations where a spliced term itself applies a TLM, shown in a later example.)
+The programmer can reason abstractly as follows:
+\begin{enumerate}[leftmargin=12pt]
+\item[\CIRCLE] \textbf{(Responsibility)} The parser named by applied TLM, here \li{KQueryParser} (Line 1), is exclusively responsible for parsing the literal body and generates an expansion statically.
+Due to this explicit dispatch mechanism, there can be any number of TLMs defined at a given type.
+As with TSLs, the context-free syntax of the language is fixed, so the TLM provider can modularly establish that the notation that the TLM implements is unambiguous (even in situations where a spliced term itself applies another TLM, shown in a later example.)
 \item[\CIRCLE] \textbf{(Segmentation)} The intermediate output that the TLM generates is structured so that the system can infer from it an accurate \emph{segmentation} of the literal body that distinguishes spliced terms from segments that are parsed in some other way by the TLM.
@@
 It cannot capture a coincidental binding of \li{x} in the expansion.
-\item[\CIRCLE] \textbf{(Context Dependence)} The system enforces context independence, so the expansion of Line 3 cannot  rely on the fact that, e.g., \li{w} is in scope.
+\item[\CIRCLE] \textbf{(Context Independence)} The system enforces context independence, so the expansion of Line 3 cannot  rely on the fact that, e.g., \li{w} is in scope.
 The client is free to rename \li{w}.
-The library dependencies that the generated expansion might have are explicit (here, the expansion uses the module \li{Helpers}), and the system ensures that this dependency is available even if the identifier \li{Helpers} has been shadowed at the use site.
-\item[\CIRCLE] \textbf{(Typing)} The type annotation on the definition of \li{$kq} determines the type that the expansion must have, here \li{KQuery.t}.
-We will see an example of a TLM definition in the next section.
-Moreover, each spliced segment in the segmentation also comes with a type annotation.
+The external dependencies of the generated expansion are explicit -- here, the expansion uses the module \li{Helpers} -- and the system ensures that this dependency is properly bound even if the identifier \li{Helpers} has been shadowed at the application site.
+\item[\CIRCLE] \textbf{(Typing)} The type annotation on the definition of \li{$kq} (Line 1) determines the type that the expansion is validated against, here \li{KQuery.t} (not shown).
+Moreover, each spliced segment in the inferred segmentation also has  a type annotation.
+This, together with the context independence property, ensures that type inference can be performed abstractly, i.e.
+given only the inferred segmentation rather than the full expansion\todo{say something formally about this?}.
+This segment-specific type information can also be communicated to the programmer upon request via the Merlin editor service.
+We will see examples of TLMs that make use of more advanced features of the OCaml type system in subsequent examples.
 %This information is usually not necessary to reason about typing, but it can be conveyed to the programmer upon request by the program editor if desired.

### 2018-02-25 icfp18/syntax-icfp18.tex
more progress on intro

@@
 mathescape=false,escapeinside={(~}{~)},
-numbers=left, numberstyle=\scriptsize\color{mygray}, language=ML,moredelim=[il][\sffamily]{?},showspaces=false,showstringspaces=false,xleftmargin=0pt, numbersep=-6pt, morekeywords=[1]{try,spliced,tyfam,opfam,let,fn,val,def,casetype,objtype,metadata,of,*,string,datatype,new,toast,notation,module,switch,where,using,import,for,ana,syn,opcon,tycon,metasignature,metamodule,metasig,metamod,static,at,by,tycase,mod,macro,match,pattern,in,patterns,expressions,implicit,forall,exptsm,pattsm},deletekeywords={double,structure,of},classoffset=0, xleftmargin=0pt,
+numbers=left, numberstyle=\scriptsize\color{mygray}, language=ML,moredelim=[il][\sffamily]{?},showspaces=false,showstringspaces=false,xleftmargin=0pt, numbersep=-6pt, morekeywords=[1]{try,spliced,tyfam,opfam,let,fn,val,def,casetype,objtype,metadata,of,*,string,lexer,parser,datatype,new,toast,notation,module,switch,where,expansions,require,import,for,ana,syn,opcon,tycon,metasignature,metamodule,metasig,metamod,static,at,by,tycase,mod,macro,match,pattern,in,patterns,expressions,implicit,forall,exptsm,pattsm},deletekeywords={double,structure,of},classoffset=0, xleftmargin=0pt,
 aboveskip=4pt,belowskip=2pt,
@@
 Instead, there has been persistent interest in mechanisms that allow library providers to define new literal notation on their own.
-For example, direct grammar extension systems like Camlp4 \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, macro-based systems like Omar et al's \emph{type-specific languages} (TSLs) \cite{TSLs}, and others that we will discuss in Sec.~\ref{sec:existing-approaches} can all be used to define new literal notation (and, in some cases, other forms of new notation, such as new infix operators and type declaration forms, which are beyond the scope of this paper)\todo{normalize parens-period thing}.
-
-\paragraph{Problem} The problem that motivates this paper is that it is impossible to reason abstractly, i.e.
+For example, direct grammar extension systems like Camlp4 \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, macro-based systems like Omar et al's \emph{type-specific languages} (TSLs) \cite{TSLs}, and others that we will discuss in Sec.~\ref{sec:existing-approaches} can all be used to define new literal notation (and, in some cases, other forms of new notation, such as new infix expression forms and type declaration forms, which are beyond the scope of this paper)\todo{normalize parens-period thing}.
+
+\paragraph{Problem} The problem that motivates this paper is that existing syntax extension systems make it difficult or impossible to reason abstractly, i.e.
 by inspecting only the program as presented to the programmer, about such fundamental issues as types and variable binding when the program text is littered with  user-defined literal forms.
 Instead, programmers and editor services can only reason transparently, i.e.
-by inspecting the underlying expansion or the extension responsible for producing that expansion.
-In some existing systems, even determining which syntax extension is responsible for a literal form is difficult.
+by inspecting the underlying expansion or the implementation details of the extensions that may have been responsible for producing that expansion.% In some existing systems, even determining which syntax extension is responsible for a literal form is difficult.
 %, i.e. when developing large programs ``consisting of many small programs o(modules), possibly written by different people''
-\newcommand*\Suppressnumber{%
-  \lst@AddToHook{OnNewLine}{%
-    \let\thelstnumber\relax%
-     \advance\c@lstnumber-\@ne\relax%
-    }%
-}
-
-\newcommand*\Reactivatenumber{%
-  \lst@AddToHook{OnNewLine}{%
-   \let\thelstnumber\origthelstnumber%
-   \advance\c@lstnumber\@ne\relax}%
-}
-
 \begin{figure}[t!]
 \begin{subfigure}[t]{0.48\textwidth}
-\begin{lstlisting}[xleftmargin=-2pt, morekeywords={context,free,syntax}]
-  context-free syntax 
-    /* ...direct mods to the CFG... */
-  /* ... */
+\fcolorbox{gray!20}{gray!20}{\makebox[0.96\textwidth][l]{\textsc{existing grammar extension systems}}}
+\begin{lstlisting}[xleftmargin=-2pt, morekeywords={EXTEND}]
+  EXTEND /* loaded by, e.g., camlp4 */
+    expr:
+      |  "`(" q = kquery ")`" -> q
+    kquery: 
+      | /* ... kquery grammar ... */
+  /* ...more extensions defined... */
   let w = compute_w();
@@
 \begin{subfigure}[t]{0.51\textwidth}
+\fcolorbox{gray!20}{gray!20}{\makebox[\textwidth][l]{\textsc{\textbf{this paper}: typed literal macros}}}
 \begin{lstlisting}[xleftmargin=-2pt]
-  notation $kq at KQuery.t by KQueryParser 
-    using module Helpers as Helpers;
-  /* ... */
+  notation $kq at KQuery.t 
+    lexer  KQueryLexer
+    parser KQueryParser.start
+    expansions require
+      module KQuery as KQuery;
+  /* ...more notations defined... */
   let w = compute_w();
@@
 \vspace{-5px}
-\caption{TLMs make examples like these more reasonable by leaving the grammar fixed and making the segmentation and the type and binding structure explicit.}
+\caption{TLMs make examples like these more reasonable by leaving the base grammar fixed and making the type and binding structure and the segmentation explicit.}
 \label{fig:K-tsm-example}
@@
 
-\todo{show expr <- thing}Consider, for example, the perspective of a programmer attempting to comprehend, i.e.
-reason about, the program text in Figure \ref{fig:K-dialect}, which is written in a dialect of OCaml's surface syntax called Reason \cite{reason-what}\todo{convert to Reason 3}{} that has, hypothetically, been extended with some number of new literal forms by a grammar extension system like Camlp4 \cite{ocaml-manual} or Sugar*~\cite{erdweg2011sugarj,erdweg2013framework} (Line 1 suggestively borrows the syntax for introducing a grammar extension from SugarHaskell \cite{erdweg2012layout}).
-The literal form on Line 6 constructs an encoding of a query in the niche stack-based database query language K, using its intentionally terse notation \cite{Whitney:2001:LOR:376284.375783}.
-The problem is that a programmer unfamiliar with this notation cannot easily answer questions like the following:
+Consider, for example, the perspective of a programmer attempting to comprehend, i.e.
+reason about, the program text in Figure \ref{fig:K-dialect}, which is written in a dialect of OCaml's surface syntax called Reason \cite{reason-what}\todo{convert to Reason 3}{} that has, hypothetically, been extended with some number of new literal forms by a grammar extension system like Camlp4 \cite{ocaml-manual} or Sugar*~\cite{erdweg2011sugarj,erdweg2013framework} -- Lines 1-5 use the grammar extension syntax from Camlp4, but see SugarHaskell for another design \cite{erdweg2012layout}.
+The literal form on Line 9 constructs an encoding of a query in the niche stack-based database query language K, using its intentionally terse notation \cite{Whitney:2001:LOR:376284.375783}.
+The problem is that a programmer unfamiliar with this notation cannot abstractly answer questions like the following:
 
 \begin{enumerate}[nolistsep,noitemsep,leftmargin=12px]
-\item[\Circle] \textbf{(Responsibility)} Which syntax extension determined the expansion of the literal form on Line 6?
+\item[\Circle] \textbf{Responsibility}: Which syntax extension determined the expansion of the literal form on Line 9?
 Might activating a new extension generate a conflicting expansion for the same form?
-\item[\Circle] \textbf{(Segmentation)} Are the characters \li{x}, \li{R} and \li{2} on Line 6 parsed as spliced expressions (meaning that they appear directly in the underlying expansion), or are they parsed in some other way peculiar to this literal form (e.g.
+\item[\Circle] \textbf{Segmentation}: Are the characters \li{x}, \li{R} and \li{2} on Line 6 parsed as spliced expressions (meaning that they appear directly in the underlying expansion), or are they parsed in some other way peculiar to this literal form (e.g.
 as operators in the K query language)?
-\item[\Circle] \textbf{(Capture)} If \li{x} is in fact a spliced expression, does it refer to the binding of \li{x} on Line 5, or might it capture an unseen binding of the same identifier in the expansion of Line 6?
-\item[\Circle] \textbf{(Context Dependence)} If \li{w}, on Lines 4-5, is renamed, could that possibly break Line 6 or change its meaning because the expansion assumes that \li{w} is in scope?% What other hidden assumptions might the desugaring be making?
-\item[\Circle] \textbf{(Typing)} What type does \li{y} have? What type is each spliced expression expected to have?
+\item[\Circle] \textbf{Capture}: If \li{x} is in fact a spliced expression, does it refer to the binding of \li{x} on Line 5, or might it capture an unseen binding of the same identifier in the expansion of Line 6?
+\item[\Circle] \textbf{Context Dependence}: If \li{w}, on Lines 4-5, is renamed, could that possibly break Line 6 or change its meaning because the expansion assumes that \li{w} is in scope?% What other hidden assumptions might the desugaring be making?
+\item[\Circle] \textbf{Typing}: What type does \li{y} have? What type is each spliced expression expected to have?
 \end{enumerate}
@@
 Analagous problems do not arise when programming without syntax extensions in languages like ML and Scala --- programmers can reason lexically about where variables and other symbols are bound, and types mediate abstraction over function and module implementations \cite{B304}.
-Ideally, the programmer would be able to abstract in some analagous manner over the implementation of a notation.
+Ideally, the programmer would be able to abstract in some analagous manner over the implementation of a new literal notation.
 % From this perspective, it is unsurprising that the use of Camlp4 has been deprecated in the OCaml system.
@@
 An approach that has made some headway toward this ideal is Omar et al.'s \emph{type-specific languages} (TSLs) \cite{TSLs}.
-Under this approach, the language delegates control over the parsing and expansion of certain \emph{generalized literal forms} to a parser associated with the type that the form is being checked against.
-So with TSLs, Line 3 of our example would require a type annotation:
+With this approach, the language delegates control over the parsing and expansion of certain \emph{generalized literal forms} to a parser associated with the type that the form is being checked against.
+So with TSLs, Line 9 of our example would require a type annotation:
 \begin{lstlisting}[numbers=none,basicstyle=\ttfamily\fontsize{9pt}{1em}\selectfont]
@@
 \begin{enumerate}[leftmargin=12px]
-  \item[\CIRCLE] \textbf{(Responsibility)} The type-directed dispatch mechanism allows the client to easily determine which parser is responsible for each literal form -- the parser defined when the type was defined.
+  \item[\CIRCLE] \textbf{Responsibility}: The type-directed dispatch mechanism allows the client to easily determine which parser is responsible for each literal form -- the parser defined when the type was defined.
 The context-free syntax of the language remains fixed and composition is mediated by splicing, so clients do not need to worry about different TSLs conflicting syntactically with one another.
-  \item[\Circle] \textbf{(Segmentation)} The TSL can splice expressions out of the literal body, but there is no clear way to associate each spliced term with some particular segment (i.e.
+  \item[\Circle] \textbf{Segmentation}: The TSL can splice expressions out of the literal body, but there is no clear way to associate each spliced term with some particular segment (i.e.
 subsequence) of the literal body, nor is there a guarantee that spliced expressions are non-overlapping.
-  \item[\Circle] \textbf{(Capture)} The TSL mechanism as described does not guarantee that splicing is capture-avoiding. (We will return to say more about this problem later.)
-  \item[\LEFTcircle] \textbf{(Context Independence)} The TSL mechanism as described requires that the expansion be completely closed, so it is trivially context independent.
+  \item[\Circle] \textbf{Capture}: The TSL mechanism as described does not guarantee that splicing is capture-avoiding. (We will return to say more about this problem later.)
+  \item[\LEFTcircle] \textbf{Context Dependence}: The TSL mechanism as described requires that the expansion be completely closed, so it is trivially context independent.
 However, this comes at a significant expressive cost: the generated expansions cannot make use of any libraries whatsoever.
-  \item[\LEFTcircle] \textbf{(Typing)} Each parser is associated with a type, and the generated expansion must necessarily be of that type, so abstract reasoning about the type of the expansion is possible.
+  \item[\LEFTcircle] \textbf{Typing}: Each parser is associated with a type, and the generated expansion must necessarily be of that type, so abstract reasoning about the type of the expansion is possible.
 However, there is no easy way to determine the type expected for each spliced expression.
@@
 It cannot capture a coincidental binding of \li{x} in the expansion.
-\item[\CIRCLE] \textbf{(Context Independence)} The system enforces context independence, so the expansion of Line 3 cannot  rely on the fact that, e.g., \li{w} is in scope.
+\item[\CIRCLE] \textbf{(Context Dependence)} The system enforces context independence, so the expansion of Line 3 cannot  rely on the fact that, e.g., \li{w} is in scope.
 The client is free to rename \li{w}.

### 2018-02-25 icfp18/syntax-icfp18.tex
first four pages are in decent shape and page breaks are in pleasant locations _twitch_

@@
 \begin{abstract}
+\vspace{-5px}
 %Many languages define convenient list literal forms.
@@
 \label{sec:intro}
+\vspace{-2px}
 % The surface syntax of a programming language is its user interface. 
 When designing the surface syntax of a general-purpose programming language, it is common practice to define \emph{literal notation} that decreases the syntactic cost of constructing and pattern matching over values of a select few data structures.
-For example, Standard ML supports list literals like \li{[x1, x2, x3]} in both expressions and patterns \cite{harper1997programming,mthm97-for-dart}, and many other general-purpose languages support substantially similar list literal notation.
+For example, Standard ML supports list literals like \li{[x1, x2, x3]} in both expression and pattern position \cite{harper1997programming,mthm97-for-dart}.
+Many other languages have similar list literal notation.
 %\footnote{In SML, list literals work even in contexts where the list expression constructors have been shadowed by other bindings, i.e. the meaning of literal forms is \emph{context independent}. 
 % We will return to this concept again later.} 
-Another example is Ur/Web, which extends the surface syntax of Ur (an ML-like language \cite{conf/pldi/Chlipala10}) with expression and pattern literals for encodings of XML and HTML data \cite{conf/popl/Chlipala15}.
+Another example is Ur/Web, which extends the surface syntax of Ur (an ML-like language \cite{conf/pldi/Chlipala10}) with expression and pattern literals for encodings of XML and HTML data~\cite{conf/popl/Chlipala15}.
 The example in Fig.
@@
 \begin{figure}[h]
-\vspace{-10px}
+\vspace{-11px}
 \begin{lstlisting}
@@
 \end{lstlisting}
-\vspace{-4px}
+\vspace{-5px}
 \caption{HTML literals with support for splicing at two different types are built primitively into Ur/Web \cite{conf/popl/Chlipala15}.}
 \label{fig:urweb}
-\vspace{-8px}
+\vspace{-9px}
 \end{figure}
 
-This design practice, where the language designer privileges certain user-defined types with built-in literal notation, is decidedly \emph{ad hoc} in that it is easy to come up with other examples of data structures for which mathematicians, scientists or programmers have invented specialized notation \cite{DBLP:journals/cacm/Iverson80,cajori1928history}.
+This design practice, where the language designer privileges certain user-defined types with built-in literal notation, is decidedly \emph{ad hoc} in that it is easy to come up with other examples of data structures for which mathematicians, scientists or programmers have invented specialized notation \cite{DBLP:journals/cacm/Iverson80,cajori1928history,TSLs}.
 For example, 1) clients of a ``collections'' library might want not just list literals, but also literal notation for matrices, finite sets, maps, hash tables and so on, following usual mathematical conventions; 2) clients of a ``web programming'' library might like not just HTML literals but also CSS literals (which Ur/Web lacks); 3) a compiler author might like ``quotation'' literals for the terms of the object language and various intermediate languages of interest; and 4) clients of a ``chemistry'' library might like chemical structure literals based on the SMILES standard \cite{anderson1987smiles}.
@@
 Although requests for specialized literal notation are easy to dismiss as superficial, the reality is that literal notation, or the absence thereof, can have a substantial influence on software quality.
-For example, \citet{Bravenboer:2007:PIA:1289971.1289975} finds that literal notation for structured encodings of queries, like the SQL-based query literals now found in many languages \cite{meijer2006linq}, reduce the temptation to use string encodings of queries and therefore reduce the risk of string injection attacks, which are amongst the most common and serious security vulnerabilities in web applications \cite{owasp2017}.
-More generally, evidence suggests that programmers frequently engage in ``stringly-typed programming'', i.e.
+For example, \citet{Bravenboer:2007:PIA:1289971.1289975} finds that literal notation for structured encodings of queries, like the SQL-based query literals now found in many languages \cite{meijer2006linq}, reduce the temptation to use string encodings of queries and therefore reduce the risk of catastrophic string injection attacks~\cite{owasp2017}.
+More generally, evidence suggests that programmers frequently resort to ``stringly-typed programming'', i.e.
 they choose strings instead of composite data structures largely for reasons of notational convenience.
-In particular, \citet{TSLs} sampled strings from open source projects and found that about 15\% of them could be parsed by some readily apparent type-specific grammar.
-Literal notation, with support for splicing, would decrease the syntactic cost of composite encodings of such data, which are more amenable to programmatic manipulation and compositional reasoning than strings.
+In particular, \citet{TSLs} sampled strings from open source projects and found that at least 15\% of them could be parsed by some readily apparent type-specific grammar.
+%, e.g. for URLs, paths, regular expressions and so on
+Literal notation, with support for splicing, would decrease the syntactic cost of composite encodings of this data, which are more amenable to programmatic manipulation and compositional reasoning.
 %Literal forms would  preferable to string encodings because they support compositional construction and decomposition by pattern matching via splicing.
@@
 Instead, there has been persistent interest in mechanisms that allow library providers to define new literal notation on their own.
-For example, direct grammar extension systems like Camlp4 \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, macro-based systems like Omar et al's \emph{type-specific languages} (TSLs) \cite{TSLs}, and others that we will discuss in Sec.~\ref{sec:existing-approaches} can all be used to define new literal notation (and, in some cases, other forms of new notation, such as new infix expression forms and type declaration forms, which are beyond the scope of this paper)\todo{normalize parens-period thing}.
-
-\paragraph{Problem} The problem that motivates this paper is that existing syntax extension systems make it difficult or impossible to reason abstractly, i.e.
-by inspecting only the program as presented to the programmer, about such fundamental issues as types and variable binding when the program text is littered with  user-defined literal forms.
+For example, direct grammar extension systems like Camlp4 \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, macro-based systems like Omar et al's \emph{type-specific languages} (TSLs) \cite{TSLs}, and other systems that we will discuss in Sec.~\ref{sec:existing-approaches} can all be used to define new literal notation (and, in some cases, other forms of new notation, such as new infix expression forms or type declaration forms, which are beyond the scope of this paper).
+
+\paragraph{Problem} The problem that specifically motivates this paper is that existing syntax extension systems  make it difficult or impossible to reason abstractly about such fundamental issues as types and variable binding when the program text is littered with  user-defined literal forms.
 Instead, programmers and editor services can only reason transparently, i.e.
-by inspecting the underlying expansion or the implementation details of the extensions that may have been responsible for producing that expansion.% In some existing systems, even determining which syntax extension is responsible for a literal form is difficult.
+by inspecting the underlying expansion or the implementation details of the extensions responsible for producing the expansion.% In some existing systems, even determining which syntax extension is responsible for a literal form is difficult.
 %, i.e. when developing large programs ``consisting of many small programs o(modules), possibly written by different people''
@@
 \begin{subfigure}[t]{0.48\textwidth}
-\fcolorbox{gray!20}{gray!20}{\makebox[0.96\textwidth][l]{\textsc{existing grammar extension systems}}}
+\fcolorbox{gray!20}{gray!20}{\makebox[0.97\textwidth][l]{\textsc{existing grammar extension systems}}}
+\vspace{-12px}
 \begin{lstlisting}[xleftmargin=-2pt, morekeywords={EXTEND}]
@@
     kquery: 
-      | /* ... kquery grammar ... */
+      | /* ...K query grammar... */
   /* ...more extensions defined... */
@@
 \fcolorbox{gray!20}{gray!20}{\makebox[\textwidth][l]{\textsc{\textbf{this paper}: typed literal macros}}}
+\vspace{-12px}
 \begin{lstlisting}[xleftmargin=-2pt]
@@
 \end{subfigure}
-\vspace{3px}
+% \vspace{4px}
 \caption{Two of the possible ways to introduce literal notation for encodings of K queries}
@@
 Consider, for example, the perspective of a programmer attempting to comprehend, i.e.
-reason about, the program text in Figure \ref{fig:K-dialect}, which is written in a dialect of OCaml's surface syntax called Reason \cite{reason-what}\todo{convert to Reason 3}{} that has, hypothetically, been extended with some number of new literal forms by a grammar extension system like Camlp4 \cite{ocaml-manual} or Sugar*~\cite{erdweg2011sugarj,erdweg2013framework} -- Lines 1-5 use the grammar extension syntax from Camlp4, but see SugarHaskell for another design \cite{erdweg2012layout}.
-The literal form on Line 9 constructs an encoding of a query in the niche stack-based database query language K, using its intentionally terse notation \cite{Whitney:2001:LOR:376284.375783}.
-The problem is that a programmer unfamiliar with this notation cannot abstractly answer questions like the following:
-
-\begin{enumerate}[nolistsep,noitemsep,leftmargin=12px]
-\item[\Circle] \textbf{Responsibility}: Which syntax extension determined the expansion of the literal form on Line 9?
+reason about, the program text in Figure \ref{fig:K-dialect}, which is written in a dialect of OCaml's surface syntax called Reason \cite{reason-what} that has, hypothetically, been extended with some number of new literal forms by a grammar extension system -- Lines 1-5 show the grammar extension syntax from Camlp4 \cite{ocaml-manual}, but Sugar*/SugarHaskell is  similar \cite{erdweg2011sugarj,erdweg2013framework,erdweg2012layout}.
+The literal form on Line 9 then constructs an encoding of a query in the niche stack-based database query language K, using its intentionally terse notation~\cite{Whitney:2001:LOR:376284.375783}.
+The problem is that a programmer looking at the program as presented, and unfamiliar with (i.e.
+holding abstract) the details elided on Lines 5-6, cannot easily answer questions like the following:
+
+\begin{enumerate}[leftmargin=12px]
+\item[\Circle] \textbf{Responsibility}: Which syntax extension even determines the expansion of the literal on Line~9?
 Might activating a new extension generate a conflicting expansion for the same form?
-\item[\Circle] \textbf{Segmentation}: Are the characters \li{x}, \li{R} and \li{2} on Line 6 parsed as spliced expressions (meaning that they appear directly in the underlying expansion), or are they parsed in some other way peculiar to this literal form (e.g.
+\item[\Circle] \textbf{Segmentation}: Are the characters \li{x}, \li{R} and \li{2} on Line 9 parsed as spliced expressions (meaning that they appear directly in the underlying expansion), or are they parsed in some other way peculiar to this literal form (e.g.
 as operators in the K query language)?
-\item[\Circle] \textbf{Capture}: If \li{x} is in fact a spliced expression, does it refer to the binding of \li{x} on Line 5, or might it capture an unseen binding of the same identifier in the expansion of Line 6?
-\item[\Circle] \textbf{Context Dependence}: If \li{w}, on Lines 4-5, is renamed, could that possibly break Line 6 or change its meaning because the expansion assumes that \li{w} is in scope?% What other hidden assumptions might the desugaring be making?
+\item[\Circle] \textbf{Capture}: If \li{x} is in fact a spliced expression, does it refer to the binding of \li{x} on Line 5, or might it capture an unseen binding of the same identifier in the expansion of Line 9?
+\item[\Circle] \textbf{Context Dependence}: If \li{w}, on Lines 7-8, is renamed, could that break Line 9  because the expansion assumes that \li{w} is in scope?
+What other hidden assumptions is the expansion making?
 \item[\Circle] \textbf{Typing}: What type does \li{y} have? What type is each spliced expression expected to have?
@@
 Forcing the programmer to reason transparently to answer basic questions like these defeats the ultimate purpose of syntactic sugar: decreasing cognitive cost \cite{Green89}.
-Analagous problems do not arise when programming without syntax extensions in languages like ML and Scala --- programmers can reason lexically about where variables and other symbols are bound, and types mediate abstraction over function and module implementations \cite{B304}.
-Ideally, the programmer would be able to abstract in some analagous manner over the implementation of a new literal notation.
+Analagous problems do not arise when programming without syntax extensions in languages like ML --- programmers can reason lexically about where variables and other symbols are bound, and types mediate abstraction over function and module implementations \cite{B304}.
+Ideally, the programmer would be able to abstract in some analagous manner over the implementation of an unfamiliar notation.
 % From this perspective, it is unsurprising that the use of Camlp4 has been deprecated in the OCaml system.
@@
 An approach that has made some headway toward this ideal is Omar et al.'s \emph{type-specific languages} (TSLs) \cite{TSLs}.
-With this approach, the language delegates control over the parsing and expansion of certain \emph{generalized literal forms} to a parser associated with the type that the form is being checked against.
-So with TSLs, Line 9 of our example would require a type annotation:
+A language that supports TSLs delegates control over the parsing and expansion of certain \emph{generalized literal forms} to a parser associated with the type that the form is being checked against.
+So in a language with TSLs, our example requires a type annotation on \li{y}:
 \begin{lstlisting}[numbers=none,basicstyle=\ttfamily\fontsize{9pt}{1em}\selectfont]
-  let y : KQuery.t = `(!R)@&{&/x!/:2_!x}'!R`
+  let y : KQuery.t = `((!R)@&{&/x!/:2_!x}'!R)`
 \end{lstlisting}
-The parser associated with the type \li{KQuery.t} would be statically invoked to parse and expand the literal body, i.e.
-the characters between the outer delimiters, here \li{`(} and \li{)`}, which are left unparsed by the fixed context-free syntax of the language (i.e.
-generalized literal forms are initially parsed like raw string literals.) TSLs are closely related to the mechanism proposed in this paper, so let us analyze TSLs from the perspective of the five reasoning criteria just outlined.
+The parser associated with the type \li{KQuery.t} is statically invoked to lex, parse and expand the literal body, i.e.
+the sequence of characters between the outer delimiters, here \li{`(} and \li{)`}.
+The literal body is constrained by the context-free syntax of the language only in that nested delimiters must be matched (like comments in OCaml).
+TSLs are closely related to the mechanism proposed in this paper, so let us analyze TSLs from the perspective of the five reasoning criteria just outlined.
 \begin{enumerate}[leftmargin=12px]
-  \item[\CIRCLE] \textbf{Responsibility}: The type-directed dispatch mechanism allows the client to easily determine which parser is responsible for each literal form -- the parser defined when the type was defined.
-The context-free syntax of the language remains fixed and composition is mediated by splicing, so clients do not need to worry about different TSLs conflicting syntactically with one another.
-  \item[\Circle] \textbf{Segmentation}: The TSL can splice expressions out of the literal body, but there is no clear way to associate each spliced term with some particular segment (i.e.
-subsequence) of the literal body, nor is there a guarantee that spliced expressions are non-overlapping.
-  \item[\Circle] \textbf{Capture}: The TSL mechanism as described does not guarantee that splicing is capture-avoiding. (We will return to say more about this problem later.)
+  \item[\LEFTcircle] \textbf{Responsibility}: The type-directed dispatch mechanism allows the client to easily determine which parser is responsible for each literal form: the parser that was associated with the type when it was defined.
+Clients do not need to worry about different TSLs conflicting syntactically because the context-free syntax of the language remains fixed and composition is mediated by splicing.
+The main limitation here is that it is impossible to define literal notation after a type has been defined, and it is also impossible to define multiple notations at a single type.
+  \item[\LEFTcircle] \textbf{Segmentation}: The parser can splice expressions out of the literal body, but there is no clear way to associate each spliced term with some particular segment (i.e.
+subsequence) of the literal body, nor is there a guarantee that spliced terms are non-overlapping.
+  \item[\LEFTcircle] \textbf{Capture}: The TSL mechanism as described enforces complete capture avoidance, though the formal specification given in the paper did not adequately enforce this constraint.
   \item[\LEFTcircle] \textbf{Context Dependence}: The TSL mechanism as described requires that the expansion be completely closed, so it is trivially context independent.
@@
 However, there is no easy way to determine the type expected for each spliced expression.
-In addition, the prior work on TSLs considered only a monomorphic, nominally-typed language with only local type inference, leaving open a number of problems that came up as we considered integrating TSLs into Reason, which is a syntactic layer over all of OCaml \cite{ocaml-manual}:
-\begin{enumerate}[leftmargin=15px]
-  \item \textbf{Structural Types}: There is no way to define literal notation at structural types, e.g. tuple and arrow types.
-  \item \textbf{Parameterized Types}: There is no way to define literal notation over a type-parameterized family of types, e.g. at all types \li{list('a)}.
-  \item \textbf{Abstract Types}: There is no way to define literal notation over a module-parameterized family of abstract types, e.g.
+In addition, the prior work considered only a monomorphic, nominally-typed language with local type inference, leaving open a number of problems that came up as we considered integrating TSLs into Reason, which is a syntactic layer over all of OCaml's semantics \cite{ocaml-manual}:
+\begin{enumerate}[leftmargin=15px,nolistsep,noitemsep]
+  \item \textbf{Structural Types}: There is no way to define literal notation at structural types, e.g. tuple and arrow types, because there is no ``definition site'' for such types.
+  \item \textbf{Parameterized \& Abstract Types}: There is no way to define literal notation over a type-parameterized family of types, e.g.
+at all types \li{list('a)}.
+Similarly, there is no way to define literal notation over a module-parameterized family of abstract types, e.g.
 at every abstract type defined by a module implementing the \li{QUEUE} signature.
-Parameterized and abstract types are ubiquitous in ML-family languages.% Programmers in ML-family languages commonly define various implementations of abstract data types in this way.
-  \item \textbf{ML-Style Type Inference}: The type-directed dispatch mechanism is specified by a bidirectional type system, which supports only local type inference \cite{Pierce:2000:LTI:345099.345100}.
-It is not clear that this is suitable for languages with ML-style constraint-based type inference.
-  \item \textbf{Pattern Literals}: TSLs support only expression literals.
-No consideration was given to the problem of defining new literal notation in pattern position.
-Pattern matching is ubiquitous in ML-family languages.
+Parameterized and abstract type families are ubiquitous in ML-family languages.% Programmers in ML-family languages commonly define various implementations of abstract data types in this way.
+  \item \textbf{Pattern Literals}: Pattern matching is ubiquitous in ML-family languages but the prior work on TSLs considered only expression literals.
 %Pattern literals are dual to expression literals in that expression literals support value construction, while pattern literals support value deconstruction.
+  \item \textbf{ML-Style Type Inference}: It is not clear that a type-directed dispatch scheme can be cleanly  reconciled with ML-style type inference.
 \end{enumerate}
@@
 We consider other existing approaches against these criteria in Sec.~\ref{sec:existing-approaches}.
-All together, we found that no existing approach was suitable as-is for integration into Reason/OCaml.
-
-\paragraph{Contributions} This paper introduces \emph{typed literal macros} (TLMs): a new system for defining literal notation that addresses all of these deficiencies.
-The result is the first system for defining new literal notation that (1) provably maintains all five of the abstract reasoning principles just outlined; and (2) is expressive enough for integration into Reason/OCaml and other full-scale statically typed languages.
-We evaluate the first claim by developing a series of formal calculi equipped with proofs of these abstract reasoning principles, and the second claim with a number of non-trivial examples that appear throughout the paper and involve the language features marked {(a)-(e)} above.
-In describing these examples, we also demonstrate that the literal parsers can be defined using off-the-shelf parser generators, so the burden on notation providers is comparable to that of existing systems for integrating new literal notation into a programming language.
+Taken together, we found that no existing approach was reasonable and expressive enough for integration into Reason/OCaml.
+
+\paragraph{Contributions} This paper introduces \emph{typed literal macros} (TLMs): the first system for defining new literal notation that (1) provably maintains all five of the abstract reasoning principles just outlined; and (2) is semantically expressive enough for integration into Reason/OCaml and other full-scale statically typed functional languages.
+We evaluate these claims with (1) a series of increasingly ML-like calculi that are equipped with proofs of these abstract reasoning principles; and (2) a number of non-trivial examples that appear throughout the paper and involve the language features mentioned above.
+In describing these examples, we also demonstrate that literal parsing logic can be defined using off-the-shelf parser generators, so the burden on notation providers is comparable to that of existing systems.
+% From the client's perspective, which is the focus of our contributions, TLMs are uniquely reasonable.
+Unique to our approach in this paper is the focus, both formal and informal, on maintaining client-side abstract reasoning principles.
 
@@
 TLMs are instead explicitly named and applied.
-For example, Figure \ref{fig:K-tsm-example} applies a TLM named \li{$kq} to a generalized literal form delimited by \li{`(} and \li{)`} to express the K query example just discussed.
+For example, Line 9 of Figure \ref{fig:K-tsm-example} applies a TLM named \li{$kq} to a generalized literal form delimited by \li{`(} and \li{)`} to express the K query example.
 The programmer can reason abstractly as follows:
 \begin{enumerate}[leftmargin=12pt]
-\item[\CIRCLE] \textbf{(Responsibility)} The parser named by applied TLM, here \li{KQueryParser} (Line 1), is exclusively responsible for parsing the literal body and generates an expansion statically.
+\item[\CIRCLE] \textbf{Responsibility}: The lexer and parser specified by the applied TLM on Lines 2-3 is exclusively responsible for lexing, parsing and expanding the literal body.
+We will give more details on constructing a TLM lexer and parser in the next section.
 Due to this explicit dispatch mechanism, there can be any number of TLMs defined at a given type.
-As with TSLs, the context-free syntax of the language is fixed, so the TLM provider can modularly establish that the notation that the TLM implements is unambiguous (even in situations where a spliced term itself applies another TLM, shown in a later example.)
-\item[\CIRCLE] \textbf{(Segmentation)} The intermediate output that the TLM generates is structured so that the system can infer from it an accurate \emph{segmentation} of the literal body that distinguishes spliced terms from segments that are parsed in some other way by the TLM.
-The segmentation can be communicated by a code editor using secondary notation, e.g.
+As with TSLs, the context-free syntax of the language is fixed, so the TLM provider can modularly establish that the notation that the TLM implements is unambiguous (even in situations where a spliced term itself applies another TLM, which we will show in a later example).
+\item[\CIRCLE] \textbf{Segmentation}: The intermediate output that the TLM generates is structured so that the system can infer from it an accurate \emph{segmentation} of the literal body that distinguishes spliced terms from segments that are parsed in some other way by the TLM.
+The segmentation is all that needs to be communicated to editor services downstream of the expander, and ultimately to the programmer using secondary notation, e.g.
 colors in this document.
 So by examining Figure \ref{fig:K-tsm-example}, the programmer knows that the two instances of \li{x} on Line 3 are parsed as spliced expressions (because they are black), whereas the \li{R}'s must be parsed in some other way, e.g.
-as operators in the K language (because they are  green.)
-\item[\CIRCLE] \textbf{(Capture)} Splicing is capture-avoiding, so the spliced expression \li{x} must refer to the binding of \li{x} on Line 2.
+as operators in the K language (because they are green).
+\item[\CIRCLE] \textbf{Capture}: Splicing is capture-avoiding, so the spliced expression \li{x} must refer to the binding of \li{x} on Line 2.
 It cannot capture a coincidental binding of \li{x} in the expansion.
-\item[\CIRCLE] \textbf{(Context Dependence)} The system enforces context independence, so the expansion of Line 3 cannot  rely on the fact that, e.g., \li{w} is in scope.
+\item[\CIRCLE] \textbf{Context Dependence}: The system enforces context independence, so the expansion of Line 3 cannot  rely on the fact that, e.g., \li{w} is in scope.
 The client is free to rename \li{w}.
-The external dependencies of the generated expansion are explicit -- here, the expansion uses the module \li{Helpers} -- and the system ensures that this dependency is properly bound even if the identifier \li{Helpers} has been shadowed at the application site.
-\item[\CIRCLE] \textbf{(Typing)} The type annotation on the definition of \li{$kq} (Line 1) determines the type that the expansion is validated against, here \li{KQuery.t} (not shown).
+The external dependencies of the generated expansion are explicit -- here, Lines 4-5 specify that the expansion requires the module \li{KQuery} -- and the system ensures that this dependency is properly bound even if the identifier \li{KQuery} has been shadowed at the application site.
+\item[\CIRCLE] \textbf{Typing}: The type annotation on the definition of \li{$kq} (Line 1) determines the type that the expansion is validated against, here \li{KQuery.t} (not shown).
 Moreover, each spliced segment in the inferred segmentation also has  a type annotation.
-This, together with the context independence property, ensures that type inference can be performed abstractly, i.e.
-given only the inferred segmentation rather than the full expansion\todo{say something formally about this?}.
-This segment-specific type information can also be communicated to the programmer upon request via the Merlin editor service.
-We will see examples of TLMs that make use of more advanced features of the OCaml type system in subsequent examples.
+This, together with the context independence property, ensures that type inference at the TLM application site can be performed (by editor services or in the programmer's mind) abstractly, i.e.
+given only the inferred segmentation rather than the full expansion.
+We will see examples of TLMs that make use of more advanced features of the OCaml type system -- notably, pattern matching, parameterized types and modules -- later.
 %This information is usually not necessary to reason about typing, but it can be conveyed to the programmer upon request by the program editor if desired.

### 2018-02-26 icfp18/syntax-icfp18.tex
initial pass over intro done

@@
 The literal body is constrained by the context-free syntax of the language only in that nested delimiters must be matched (like comments in OCaml).
-TSLs are closely related to the mechanism proposed in this paper, so let us analyze TSLs from the perspective of the five reasoning criteria just outlined.
-\begin{enumerate}[leftmargin=12px]
+TSLs are closely related to the mechanism that we will propose, so let us analyze TSLs from the perspective of the five reasoning criteria just outlined.
+\begin{enumerate}[leftmargin=12px,nolistsep,noitemsep]
   \item[\LEFTcircle] \textbf{Responsibility}: The type-directed dispatch mechanism allows the client to easily determine which parser is responsible for each literal form: the parser that was associated with the type when it was defined.
@@
 subsequence) of the literal body, nor is there a guarantee that spliced terms are non-overlapping.
+As such, there is no way to indicate to the programmer where base language expressions are located within a literal form.
   \item[\LEFTcircle] \textbf{Capture}: The TSL mechanism as described enforces complete capture avoidance, though the formal specification given in the paper did not adequately enforce this constraint.
@@
 However, this comes at a significant expressive cost: the generated expansions cannot make use of any libraries whatsoever.
-  \item[\LEFTcircle] \textbf{Typing}: Each parser is associated with a type, and the generated expansion must necessarily be of that type, so abstract reasoning about the type of the expansion is possible.
+  \item[\LEFTcircle] \textbf{Typing}: The expansion must necessarily be of the associated type, so abstract reasoning about the type of the expansion is straightforward.
 However, there is no easy way to determine the type expected for each spliced expression.
-In addition, the prior work considered only a monomorphic, nominally-typed language with local type inference, leaving open a number of problems that came up as we considered integrating TSLs into Reason, which is a syntactic layer over all of OCaml's semantics \cite{ocaml-manual}:
+In addition, the prior work considered only a monomorphic, nominally-typed language with local type inference, leaving open a number of problems that came up as we considered integrating TSLs into Reason/OCaml:
 \begin{enumerate}[leftmargin=15px,nolistsep,noitemsep]
@@
 \end{enumerate}
+
 We consider other existing approaches against these criteria in Sec.~\ref{sec:existing-approaches}.
@@
 % \item 
-Sec.~\ref{sec:ptsms} adds type functions and an ML-style module system and introduces \emph{parametric TLMs} (pTLMs), i.e. TLMs that can take type and module parameters, which solve several problems. 
-Sec.~\ref{sec:setsms-formally} presents a type-theoretic calculus of simple expression and pattern TLMs and formally establishes the reasoning principles implied above. 
+Sec.~\ref{sec:ptsms}  introduces \emph{parametric TLMs} (pTLMs), i.e.
+TLMs that can take type and module parameters, which allow us to define literal notation over a type- or module-parameterized family of types.
+Sec.~\ref{sec:ptsms} also reveals that the \li{expansions require} clause is syntactic sugar for parameterization followed immediately by partial parameter application.
+Having introduced the basic machinery, we continue in Sec.~\ref{sec:more-examples} with additional examples that demonstrate both the expressive power of this approach and certain alternative parser implementation strategies that notation providers might consider.
+Sec.~\ref{sec:setsms-formally} presents a type-theoretic calculus of simple expression and pattern TLMs and formally establishes the reasoning principles implied above, in the formally simplest setting.
+Sec.~\ref{sec:ptlms-formally} adds type functions and an ML-style module system to this calculus, and gives the most general variant of the reasoning principles theorem. 
 % \item 
-Sec.~\ref{sec:existing-approaches} compares TLMs to related work using the rubric of five reasoning principles just discussed. 
+Sec.~\ref{sec:existing-approaches} compares TLMs to related work, guided by the rubric of five reasoning principles just discussed. 
 % \item 
-Sec.~\ref{sec:discussion} concludes with a summary of the specific contributions of this paper and a discussion.
+Sec.~\ref{sec:discussion} concludes with a summary of the contributions of this paper and a discussion of future work.
 
@@
 
-Due to the space limitations, certain technical details and extensions to the calculi presented in this paper, as well as detailed proofs, are given in the supplemental material.
-Also available is an additional series of case studies: notation for an encoding of regular expressions as an abstract data type, implemented using TLMs for quasiquotation and grammar-based parser generators.
+Due to the space limitations, certain technical details and extensions to the calculi presented in this paper, as well as detailed proofs, are available in the supplemental material.
+% Also available is an additional series of case studies: notation for an encoding of regular expressions as an abstract data type, implemented using TLMs for quasiquotation and grammar-based parser generators.
 %are also available in the supplement.
@@
 
+\newcommand{\moreExamplesSec}{Additional Examples}
+\section{\protect\moreExamplesSec}
+\label{sec:more-examples}
+
+\subsection{HTML}
+
+\subsection{Quotation}
+
+\subsection{More Examples, Briefly}
 %
@@
 In the full calculus, which supports parametric TLMs, the context independence clause allows reference to the variables standing for parameters.
+
+\newcommand{\pTLMsFormallySec}{Parametric TLMs, Formally}
+\section{\protect\pTLMsFormallySec}
+\label{sec:ptlms-formally}
+
+% \begin{figure}[p]
+% \[\begin{array}{llcl}
+% \mathsf{UMType} & \urho & ::= 
+% %& \autype{\utau} 
+% & \utau & \text{type annotation}\\
+% % &&
+% %& \aualltypes{\ut}{\urho} 
+% % & \alltypes{\ut}{\urho} & \text{type parameterization}\\
+% &&
+% %& \auallmods{\usigma}{\uX}{\urho} 
+% & \allmods{\uX}{\usigma}{\urho} & \text{module parameterization}\\
+% \mathsf{UMExp} & \uepsilon & ::= 
+% %& \abindref{\tsmv} 
+% & \tsmv & \text{TLM identifier reference}\\
+% % &&
+% %& \auabstype{\ut}{\uepsilon} 
+% % & \abstype{\ut}{\uepsilon} & \text{type abstraction}\\
+% &&
+% %& \auabsmod{\usigma}{\uX}{\uepsilon} 
+% & \absmod{\uX}{\usigma}{\uepsilon} & \text{module abstraction}\\
+% % &&
+% %& \auaptype{\utau}{\uepsilon} 
+% % & \aptype{\uepsilon}{\utau} & \text{type application}\\
+% &&
+% %& \auapmod{\uM}{\uepsilon} 
+% & \apmod{\uepsilon}{\uX} & \text{module application}\ECC
+% \end{array}
+% \]
+% \caption{Syntax of unexpanded TLM types and expressions.}
+% \label{fig:P-macro-expressions-types-u}
+% \end{figure}
+
+We will now outline $\miniVerseParam$, a calculus that extends $\miniVersePat$ with parametric TLMs.
+This calculus is organized, like $\miniVersePat$, as an unexpanded language (UL) defined by typed expansion to an expanded language (XL).
+There is not enough space to describe $\miniVerseParam$ with the same level of detail as in Sec.
+\ref{sec:setsms-formally}, so we highlight only the most important concepts below.
+The details are in the supplement.
+
+The XL consists of 1) module expressions, $M$, classified by signatures, $\sigma$; 2) constructions, $c$, classified by kinds, $\kappa$; and 3) expressions classified by types, which are constructions of kind $\akty$ (we use metavariables $\tau$ instead of $c$ for types by convention.) Metavariables $X$ ranges over module variables and $u$ or $t$ over construction variables.
+The module and construction languages are based closely on those defined by \citet{pfple1}, which in turn are based on early work by \citet{MacQueen:1984:MSM:800055.802036,DBLP:conf/popl/MacQueen86}, subsequent work on the phase splitting interpretation of modules \cite{harper1989higher} and on using dependent singleton kinds to track type identity \cite{stone2006extensional,DBLP:conf/lfmtp/Crary09}, and finally on formal developments by \citet{dreyer2005understanding} and \citet{conf/popl/LeeCH07}.
+A complete account of these developments is unfortunately beyond the scope of this paper.
+The expression language extends the language of $\miniVersePat$ only to allow projection out of modules.
+
+The main conceptual difference between $\miniVersePat$ and $\miniVerseParam$ is that $\miniVerseParam$ introduces the notion of unexpanded and expanded TLM expressions and types, as shown in Fig.
+\ref{fig:P-macro-expressions-types}.
+
+\begin{figure}[h]
+\small
+\begin{minipage}{0.38\textwidth}
+$\arraycolsep=3pt\begin{array}{llcl}
+\mathsf{UMType} & \urho & ::= & \utau ~\vert~ \allmods{\uX}{\usigma}{\urho}\\
+\mathsf{UMExp} & \uepsilon & ::= & \tsmv ~\vert~ \absmod{\uX}{\usigma}{\uepsilon}~\vert~ \apmod{\uepsilon}{\uX}
+\end{array}$
+\end{minipage}
+\begin{minipage}{0.6\textwidth}
+$\arraycolsep=3pt\begin{array}{llcl}
+\mathsf{MType} & \rho & ::= & \aetype{\tau} ~\vert~ \aeallmods{\sigma}{X}{\rho} \\
+\mathsf{MExp} & \epsilon & ::= & \adefref{a} ~\vert~ \aeabsmod{\sigma}{X}{\epsilon} ~\vert~\aeapmod{X}{\epsilon} 
+\end{array}$
+\end{minipage}
+\caption{Syntax of unexpanded and expanded TLM types and expressions in $\miniVerseParam$}
+\label{fig:P-macro-expressions-types}
+\end{figure}
+The TLM type $\aeallmods{\sigma}{X}{\rho}$ classifies TLM expressions that have one module parameter matching $\sigma$.
+For simplicity, we formalize only module parameters.
+Type parameters can be expressed as module parameters having exactly one abstract type member.
+
+The rule governing expression TLM application, reproduced below, touches all of the main ideas in $\miniVerseParam$, so we will refer to it throughout the remainder of this section.%It might be useful to compare this rule to Rule \textsc{ee-ap-sptsm}.
+{\small\begin{mathpar}
+\inferrule[ee-ap-petsm]{
+  \uOmega = \uOmegaEx{\uD}{\uG}{\uMctx}{\Omega_\text{app}}\\
+  \uPsi=\uAS{\uA}{\Psi}\\\\
+  \tsmexpExpandsExp{\uOmega}{\uPsi}{\uepsilon}{\epsilon}{\aetype{\tau_\text{final}}}\\
+  \tsmexpEvalsExp{\Omega_\text{app}}{\Psi}{\epsilon}{\epsilon_\text{normal}}\\\\
+  \tsmdefof{\epsilon_\text{normal}}=a\\
+  \Psi = \Psi', \petsmdefn{a}{\rho}{\eparse}\\\\
+  \encodeBody{b}{\ebody}\\
+  \evalU{\ap{\eparse}{\ebody}}{\aein{\mathtt{SuccessE}}{e_\text{pproto}}}\\
+  \decodePCEExp{e_\text{pproto}}{\pce}\\\\
+  \prepce{\Omega_\text{app}}{\Psi}{\pce}{\ce}{\epsilon_\text{normal}}{\aetype{\tau_\text{proto}}}{\omega}{\Omega_\text{params}}\\\\
+  \segOK{\segof{\ce}}{b}\\
+  \cvalidEP{\Omega_\text{params}}{\esceneP{\omega : \OParams}{\uOmega}{\uPsi}{\uPhi}{b}}{\ce}{e}{\tau_\text{proto}}
+}{
+  \expandsP{\uOmega}{\uPsi}{\uPhi}{\utsmap{\uepsilon}{b}}{[\omega]e}{[\omega]\tau_\text{proto}}
+}
+\end{mathpar}}
+
+The first two premises simply deconstruct the (unified) unexpanded context $\uOmega$ (which tracks the expansion of expression, constructor and module identifiers, as $\uDelta$ and $\uGamma$ did in $\miniVersePat$) and peTLM context, $\uPsi$.
+Next, we expand $\uepsilon$ according to straightforward unexpanded peTLM expression expansion rules.
+The resulting TLM expression, $\epsilon$, must be defined at a type (i.e.
+no quantification over modules must remain once the literal body is encountered.)
+
+The fourth premise performs \emph{peTLM expression normalization}, $\tsmexpEvalsExp{\Omega_\text{app}}{\Psi}{\epsilon}{\epsilon_\text{normal}}$.
+This is defined in terms of a structural operational semantics \cite{DBLP:journals/jlp/Plotkin04a} with two stepping rules:
+% \begin{equation*}\tag{\ref{rule:tsmexpEvalsExp}}
+% \inferrule{
+%   \tsmexpMultistepsExp{\Omega}{\Psi}{\epsilon}{\epsilon'}\\
+%   \tsmexpNormalExp{\Omega}{\Psi}{\epsilon'}
+% }{
+%   \tsmexpEvalsExp{\Omega}{\Psi}{\epsilon}{\epsilon'}
+% }
+% \end{equation*}
+% where the multistep judgement, $\tsmexpMultistepsExp{\Omega}{\Psi}{\epsilon}{\epsilon'}$, is defined as the reflexive, transitive closure of the stepping judgement defined by the following rules:
+% \begin{equation*}\tag{\ref{rule:tsmexpStepsExp-aptype-1}}
+% \inferrule{
+%   \tsmexpStepsExp{\Omega}{\Psi}{\epsilon}{\epsilon'}
+% }{
+%   \tsmexpStepsExp{\Omega}{\Psi}{\aeaptype{\tau}{\epsilon}}{\aeaptype{\tau}{\epsilon'}}
+% }
+% \end{equation*}
+% \begin{equation*}\tag{\ref{rule:tsmexpStepsExp-aptype-2}}
+% \inferrule{ }{
+%   \tsmexpStepsExp{\Omega}{\Psi}{\aeaptype{\tau}{\aeabstype{t}{\epsilon}}}{[\tau/t]\epsilon}
+% }
+% \end{equation*}
+{\small\begin{mathpar}
+\inferrule[eps-dyn-apmod-subst-e]{ }{
+  \tsmexpStepsExp{\Omega}{\Psi}{\aeapmod{X}{\aeabsmod{\sigma}{X'}{\epsilon}}}{[X/X']\epsilon}
+}
+
+\inferrule[eps-dyn-apmod-steps-e]{
+  \tsmexpStepsExp{\Omega}{\Psi}{\epsilon}{\epsilon'}
+}{
+  \tsmexpStepsExp{\Omega}{\Psi}{\aeapmod{X}{\epsilon}}{\aeapmod{X}{\epsilon'}}
+}
+\end{mathpar}}
+% The peTLM expression normal forms are defined as follows:
+% \begin{equation*}\tag{\ref{rule:tsmexpNormalExp-defref}}
+% \inferrule{ }{
+%   \tsmexpNormalExp{\Omega}{\Psi, \petsmdefn{a}{\rho}{\eparse}}{\adefref{a}}
+% }
+% \end{equation*}
+% % \begin{equation*}\tag{\ref{rule:tsmexpNormalExp-abstype}}
+% % \inferrule{ }{
+% %   \tsmexpNormalExp{\Omega}{\Psi}{\aeabstype{t}{\epsilon}}
+% % }
+% % \end{equation*}
+% \begin{equation*}\tag{\ref{rule:tsmexpNormalExp-absmod}}
+% \inferrule{ }{
+%   \tsmexpNormalExp{\Omega}{\Psi}{\aeabsmod{\sigma}{X}{\epsilon}}
+% }
+% \end{equation*}
+% % \begin{equation*}\tag{\ref{rule:tsmexpNormalExp-aptype}}
+% % \inferrule{
+% %   \epsilon \neq \aeabstype{t}{\epsilon'}\\
+% %   \tsmexpNormalExp{\Omega}{\Psi}{\epsilon}
+% % }{
+% %   \tsmexpNormalExp{\Omega}{\Psi}{\aeaptype{\tau}{\epsilon}}
+% % }
+% % \end{equation*}
+% \begin{equation*}\tag{\ref{rule:tsmexpNormalExp-apmod}}
+% \inferrule{
+%   \epsilon \neq \aeabsmod{\sigma}{X'}{\epsilon'}\\
+%   \tsmexpNormalExp{\Omega}{\Psi}{\epsilon}
+% }{
+%   \tsmexpNormalExp{\Omega}{\Psi}{\aeapmod{X}{\epsilon}}
+% }
+% \end{equation*}
+\vspace{-5px}
+
+Normalization eliminates parameters introduced in higher-order abbreviations, leaving only those parameter applications specified by the original TLM definition.
+Normal forms and progress and preservation theorems are established in the supplement.
+
+The third row of premises looks up the applied TLM's definition by invoking a simple metafunction to extract its name, $a$, then looking up $a$ within the peTLM definition context, $\Psi$.
+% \begin{align}
+% \tsmdefof{\adefref{a}} & = a \tag{\ref{eqn:tsmdefof-adefref}}\\
+% % \tsmdefof{\aeabstype{t}{\epsilon}} & = \tsmdefof{\epsilon} \tag{\ref{eqn:tsmdefof-abstype}}\\
+% \tsmdefof{\aeabsmod{\sigma}{X}{\epsilon}} & = \tsmdefof{\epsilon} \tag{\ref{eqn:tsmdefof-absmod}}\\
+% % \tsmdefof{\aeaptype{\tau}{\epsilon}} & = \tsmdefof{\epsilon} \tag{\ref{eqn:tsmdefof-aptype}}\\
+% \tsmdefof{\aeapmod{X}{\epsilon}} & = \tsmdefof{\epsilon} \tag{\ref{eqn:tsmdefof-apmod}}
+% \end{align}
+
+The fourth row of premises 1) encodes the body as a value of the type $\tBody$; 2) applies the parse function; and 3) decodes the result, producing a \emph{parameterized proto-expression}, $\pce$.
+Parameterized proto-expressions, $\pce$, are ABTs that serve simply to introduce the parameter bindings into an underlying proto-expression, $\ce$.
+The syntax of parameterized proto-expressions is given below.
+
+\vspace{-4px}{\small\[\begin{array}{llcl}
+% \textbf{Sort} & & & \textbf{Operational Form} & \textbf{Stylized Form} & \textbf{Description}\\
+% \LCC \color{Yellow}&\color{Yellow}&\color{Yellow}& \color{Yellow} & \color{Yellow} & \color{Yellow}\\
+\mathsf{PPrExp} & \pce & ::= & \apceexp{\ce} ~\vert~ \apcebindmod{X}{\pce}
+\end{array}\]}%
+\vspace{-6px}%
+
+There must be one binder in $\pce$ for each TLM parameter specified by $a$. (In Reason, we can insert these binders automatically as a convenience.)
+
+The judgement on the fifth row of Rule \textsc{ee-ap-petsm} then \emph{deparameterizes} $\pce$ by peeling away these binders to produce 1) the underlying proto-expression, $\ce$, with the variables that stand for the parameters free; 2) a corresponding deparameterized type, $\tau_\text{proto}$, that uses the same free variables to stand for the parameters; 3) a \emph{substitution}, $\omega$, that pairs the applied parameters from $\epsilon_\text{normal}$ with the corresponding variables generated when peeling away the binders in $\pce$; and 4) a corresponding \emph{parameter context}, $\Omega_\text{params}$, that tracks the signatures of these variables.
+The two rules governing the proto-expression deparameterization judgement are below:
+{\small\begin{mathpar}
+\inferrule{ }{
+  \prepce{\Omega_\text{app}}{\Psi, \petsmdefn{a}{\rho}{\eparse}}{\apceexp{\ce}}{\ce}{\adefref{a}}{\rho}{\emptyset}{\emptyset}
+}
+
+\vspace{-4px}\inferrule{
+  \prepce{\Omega_\text{app}}{\Psi}{\pce}{\ce}{\epsilon}{\aeallmods{\sigma}{X}{\rho}}{\omega}{\Omega}\\
+  X \notin \domof{\Omega_\text{app}}
+}{
+  \prepce{\Omega_\text{app}}{\Psi}{\apcebindmod{X}{\pce}}{\ce}{\aeapmod{X'}{\epsilon}}{\rho}{(\omega, X'/X)}{(\Omega, X : \sigma)}
+}
+\end{mathpar}}%
+This judgement can be pronounced ``when applying peTLM $\epsilon$, $\pce$ has deparameterization $\ce$ leaving $\rho$ with parameter substitution $\omega$''. 
+Notice based on the second rule that every module binding in $\pce$ must pair with a corresponding module parameter application.
+Moreover, the variables standing for parameters must not appear in $\Omega_\text{app}$, i.e.
+$\domof{\Omega_\text{params}}$ must be disjoint from $\domof{\Omega_\text{app}}$ (this requirement can always be discharged by alpha-variation.)
+
+The final row of premises checks that the segmentation of $\ce$ is valid and  performs proto-expansion validation under the parameter context, $\Omega_\text{param}$ (rather than the empty context, as was the case in $\miniVersePat$.) The conclusion of the rule applies the parameter substitution, $\omega$, to the resulting expression and the deparameterized type.
+
+Proto-expansion validation operates conceptually as in $\miniVersePat$.
+The only subtlety has to do with the type annotations on references to spliced terms.
+As described at the end of Sec.
+\ref{sec:ptsms-by-example}, these annotations might refer to the parameters, so the parameter substitution, $\omega$, which is tracked by the splicing scene, must be applied to the type annotation before proceeding recursively to expand the referenced unexpanded term.
+However, the spliced term itself must treat parameters parametrically, so the substitution is not applied in the conclusion of the following rule:
+\begin{mathpar}\label{rule:cvalidE-P-splicede}
+\inferrule{
+  \parseUExp{\bsubseq{b}{m}{n}}{\ue}\\
+    \cvalidC{\OParams}{\csceneP{\omega : \OParams}{\uOmega}{b}}{\ctau}{\tau}{\akty}\\
+  \expandsP{\uOmega}{\uPsi}{\uPhi}{\ue}{e}{[\omega]\tau}\\\\
+  \uOmega=\uOmegaEx{\uD}{\uG}{\uMctx}{\Omega_\text{app}}\\
+  \domof{\Omega} \cap \domof{\Omega_\text{app}} = \emptyset
+}{
+  \cvalidEP{\Omega}{\esceneP{\omega : \OParams}{\uOmega}{\uPsi}{\uPhi}{b}}{\acesplicede{m}{n}{\ctau}}{e}{\tau}
+}
+\end{mathpar}
+(This is only sensible because we maintain the invariant that $\Omega$ is always an extension of $\Omega_\text{params}$.)
+
+The calculus enjoys metatheoretic properties analagous to those described in Sec.
+\ref{sec:s-metatheory}, modified to account for the presence of modules, kinds and parameterization.
+The following theorem establishes the abstract reasoning principles available when applying a parametric expression TLM.
+The clauses are directly analagous to those of Theorem \ref{thm:setlm-reasoning}, so for reasons of space we do not repeat the inline descriptions.
+The \textbf{Kinding} clauses can be understood by analogy to the \textbf{Typing} clauses.
+The details of parametric pattern TLMs (ppTLMs) are analagous (see supplement.)
+\vspace{-3px}
+\begin{theorem}[peTLM Reasoning Principles]
+If $\expandsP{\uOmega}{\uPsi}{\uPhi}{\utsmap{\uepsilon}{b}}{e}{\tau}$ then:
+\begin{enumerate}[nolistsep,noitemsep]
+  \item $\uOmega=\uOmegaEx{\uD}{\uG}{\uMctx}{\Omega_\text{app}}$
+  \item $\uPsi=\uAS{\uA}{\Psi}$
+  \item (\textbf{Typing 1}) $\tsmexpExpandsExp{\uOmega}{\uPsi}{\uepsilon}{\epsilon}{\aetype{\tau}}$ and $\hastypeP{\Omega_\text{app}}{e}{\tau}$
+  \item $\tsmexpEvalsExp{\Omega_\text{app}}{\Psi}{\epsilon}{\epsilon_\text{normal}}$
+  \item $\tsmdefof{\epsilon_\text{normal}}=a$
+  \item $\Psi = \Psi', \petsmdefn{a}{\rho}{\eparse}$
+  \item $\encodeBody{b}{\ebody}$
+    \item $\evalU{\ap{\eparse}{\ebody}}{\aein{\mathtt{SuccessE}}{e_\text{pproto}}}$
+  \item $\decodePCEExp{e_\text{pproto}}{\pce}$
+  \item $\prepce{\Omega_\text{app}}{\Psi}{\pce}{\ce}{\epsilon_\text{normal}}{\aetype{\tau_\text{proto}}}{\omega}{\Omega_\text{params}}$
+  \item (\textbf{Segmentation}) $\segOK{\segof{\ce}}{b}$
+  \item $\cvalidEP{\Omega_\text{params}}{\esceneP{\omega : \OParams}{\uOmega}{\uPsi}{\uPhi}{b}}{\ce}{e'}{\tau_\text{proto}}$
+  \item $e = [\omega]e'$
+  \item $\tau = [\omega]\tau_\text{proto}$
+  \item $
+    \segof{\ce} = \sseq{\acesplicedk{m_i}{n_i}}{\nkind} \cup \sseq{\acesplicedc{m'_i}{n'_i}{\cekappa'_i}}{\ncon} \cup $ \\
+     ~~~~$          \sseq{\acesplicede{m''_i}{n''_i}{\ctau_i}}{\nexp}
+    $
+  \item (\textbf{Kinding 1}) $\sseq{\kExpands{\uOmega}{\parseUKindF{\bsubseq{b}{m_i}{n_i}}}{\kappa_i}}{\nkind}$ and \\ ~~~~$\sseq{\iskind{\Omega_\text{app}}{\kappa_i}}{\nkind}$
+  \item (\textbf{Kinding 2}) $\sseq{\cvalidK{\OParams}{\csceneP{\omega : \OParams}{\uOmega}{b}}{\cekappa'_i}{\kappa'_i}}{\ncon}$ and $\sseq{\iskind{\Omega_\text{app}}{[\omega]\kappa'_i}}{\ncon}$
+  \item (\textbf{Kinding 3}) $\sseq{\cExpands{\uOmega}{\parseUConF{\bsubseq{b}{m'_i}{n'_i}}}{c_i}{[\omega]\kappa'_i}}{\ncon}$ and $\sseq{\haskind{\Omega_\text{app}}{c_i}{[\omega]\kappa'_i}}{\ncon}$
+  \item (\textbf{Kinding 4}) $\sseq{\cvalidC{\OParams}{\csceneP{\omega : \OParams}{\uOmega}{b}}{\ctau_i}{\tau_i}{\akty}}{\nexp}$ and $\sseq{\haskind{\Omega_\text{app}}{[\omega]\tau_i}{\akty}}{\nexp}$
+  \item (\textbf{Typing 2}) $\sseq{\expandsP{\uOmega}{\uPsi}{\uPhi}{\parseUExpF{\bsubseq{b}{m''_i}{n''_i}}}{e_i}{[\omega]\tau_i}}{\nexp}$ and $\sseq{\hastypeP{\Omega_\text{app}}{e_i}{[\omega]\tau_i}}{\nexp}$
+  \item (\textbf{Capture Avoidance}) $e = [\sseq{\kappa_i/k_i}{\nkind}, \sseq{c_i/u_i}{\ncon}, \sseq{e_i/x_i}{\nexp}, \omega]e''$ for some $e''$ and fresh $\sseq{k_i}{\nkind}$ and fresh $\sseq{u_i}{\ncon}$ and fresh $\sseq{x_i}{\nexp}$
+  \item (\textbf{Context Independence}) $\mathsf{fv}(e'') \subset \sseq{k_i}{\nkind} \cup \sseq{u_i}{\ncon} \cup \sseq{x_i}{\nexp} \cup \domof{\OParams}$
+  % $\hastypeP{\sseq{\Khyp{k_i}}{\nkind} \cup \sseq{u_i :: [\omega]\kappa'_i}{\ncon} \cup \sseq{x_i : [\omega]\tau_i}{\nexp}}{[\omega]e''}{\tau}$\todo{maybe restate this in terms of free variables of e'' here and elsewhere, because context isn't technically well-formed here?}
+\end{enumerate}
+\end{theorem}
+
+\vspace{-6px}
 

### 2018-02-26 icfp18/syntax-icfp18.tex
new figure 3 -- regexes

@@
 \paragraph{Contributions} This paper introduces \emph{typed literal macros} (TLMs): the first system for defining new literal notation that (1) provably maintains all five of the abstract reasoning principles just outlined; and (2) is semantically expressive enough for integration into Reason/OCaml and other full-scale statically typed functional languages.
-We evaluate these claims with (1) a series of increasingly ML-like calculi that are equipped with proofs of these abstract reasoning principles; and (2) a number of non-trivial examples that appear throughout the paper and involve the language features mentioned above.
+We evaluate these claims with (1) a series of increasingly ML-like calculi equipped with proofs of these abstract reasoning principles; and (2) a number of non-trivial examples that appear throughout the paper and involve the language features mentioned above.
 In describing these examples, we also demonstrate that literal parsing logic can be defined using off-the-shelf parser generators, so the burden on notation providers is comparable to that of existing systems.
@@
 For example, if strings were not primitive, we might write \li{`SURL<body>\{EURLheading ($str "SSTRWorld!ESTR")SURL\} ...</body>EURL`}.}
+
+\begin{figure}[t]
+\begin{subfigure}[t]{0.45\textwidth}
+\begin{lstlisting}[mathescape=~]
+  module Regex = {
+    type t = AnyChar | Str(string)
+           | Seq(t, t) | Or(t, t) 
+           | Star(t);
+  };
+\end{lstlisting}
+\vspace{-5px}
+\caption{The \li{Regex} module, which defines the recursive datatype \li{Regex.t}.}
+\label{fig:Regex-module-def}
+\end{subfigure}
+\hfill
+\begin{subfigure}[t]{0.53\textwidth}
+\begin{lstlisting}[mathescape=|]
+  notation $regex at Regex.t
+    lexer   RegexLexer
+    parser  RegexParser.start
+    expansions require
+      module Regex as Regex;
+\end{lstlisting}
+\vspace{-5px}
+\caption{The definition of the \li{$regex} TLM. Figure \ref{fig:candidate-exp-Reason} defines \li{RegexLexer} and \li{RegexParser}.}
+\label{fig:html-tlm-def}
+\end{subfigure}
+
+\begin{subfigure}[t]{\textwidth}
+\vspace{4px}
+\begin{lstlisting}
+  let any_base = $regex `(SURLA|T|G|CEURL)`;
+  let bisA = $regex `(SURLGC$(EURLany_baseSURL)GCEURL)`;
+  let gene_restriction_template(gene_sequence : string) => 
+    $regex `(SURL$(EURLbisASURL)$(EURLany_baseSURL)*$$(EURLgene_sequenceSURL)$(EURLany_baseSURL)*$(EURLbisASURL)EURL)`
+\end{lstlisting}
+\vspace{-2px}
+\caption{Examples of the \li{$regex} TLM being applied in a bioinformatics application.
+%In each case, literal body, between backticks, is initially left unparsed according to the language's context-free syntax.
+%The applied TLM determines a segmentation and expansion during the typed expansion phase, which generalizes the usual typing phase.
+}
+\label{fig:first-tsm-example}
+\end{subfigure}
+\vspace{5px}
+\caption{Case Study: POSIX-style regex literal notation, with support for string and regex splicing.}
+\end{figure}
+
+\paragraph{Outline} 
+% \begin{itemize}
+% \item 
+We begin in Sec.~\ref{sec:setsms} by introducing TLMs with a detailed case study. 
+% \item 
+Sec.~\ref{sec:sptsms} then introduces {pattern TLMs}, i.e.
+TLMs that generate patterns, and describes the special hygiene conditions in pattern position.
+%The hygiene condition for pattern TLMs is interesting.
+% \item 
+Sec.~\ref{sec:ptsms}  introduces \emph{parametric TLMs} (pTLMs), i.e.
+TLMs that can take type and module parameters, which allow us to define literal notation over a type- or module-parameterized family of types.
+Sec.~\ref{sec:ptsms} also reveals that the \li{expansions require} clause is syntactic sugar for parameterization followed immediately by partial parameter application.
+Having introduced the basic machinery, we continue in Sec.~\ref{sec:more-examples} with additional examples that demonstrate both the expressive power of this approach and alternative parser implementation strategies that notation providers can consider.
+Sec.~\ref{sec:setsms-formally} presents a type-theoretic calculus of simple expression and pattern TLMs and formally establishes the reasoning principles implied above, in the formally simplest setting.
+Sec.~\ref{sec:ptlms-formally} adds type functions and an ML-style module system to this calculus, and gives the most general variant of the reasoning principles theorem. 
+% \item 
+Sec.~\ref{sec:existing-approaches} compares TLMs to related work, guided by the rubric of five reasoning principles just discussed. 
+% \item 
+Sec.~\ref{sec:discussion} concludes with a summary of the contributions of this paper and a discussion of future work.
+Due to the space limitations, certain technical details and extensions to the calculi presented in this paper, as well as detailed proofs, are available in the anonymized supplemental material.
+% Also available is an additional series of case studies: notation for an encoding of regular expressions as an abstract data type, implemented using TLMs for quasiquotation and grammar-based parser generators.
+%are also available in the supplement.
+Sec.~\ref{sec:static-eval} considers the topic of term evaluation during the  expansion phase, i.e.
+\emph{static evaluation}, in  more detail.
+Sec \ref{sec:static-eval} also gives examples of TLMs that are useful for defining other TLMs, e.g.
+TLMs that implement parser generators and quasiquotation.
+
+%  We say more about LISP macros in Sec.
+\ref{sec:existing-approaches}.% For the present purposes, we can consider Reason as essentially the calculus defined in the supplement extended with various conveniences that are commonly found in other ML-like languages and, notionally, orthogonal to TLMs.
+%Reason is, as its name suggests, a conceptual descendent of ML.
+It diverges from other dialects of ML that have a similar type structure in that it has a bidirectional type system \cite{Pierce:2000:LTI:345099.345100} (like, for example, Scala \cite{OdeZenZen01}) for reasons that have to do with the mechanism of TLM implicits described in Chapters \ref{chap:tsls} and \ref{chap:ptsms}.
+%The reason we will not follow Standard ML \cite{mthm97-for-dart} in giving a complete formal definition of Reason in this work is both to emphasize that the primitives we introduce are ``insensitive'' to the details of the underlying type structure of the language (so TLMs can be considered for inclusion in a variety of languages, not only dialects of ML), and to avoid distracting the reader (and the author) with definitions that are already well-understood in the literature and that are orthogonal to those that are the focus of this work.
+% We will not formally define these features mainly to avoid unnecessarily complicating our presentation with details that are not essential to the ideas presented herein. As such, 
+% All examples written in Reason should be understood to be informal motivating material for the subsequent formal material.
+%We anticipate that future full-scale language specifications will be able to combine the ideas  in the proposed work without trouble.
+%The purpose of the work being proposed is to serve as a reference for those interested in the new constructs we introduce, not to serve as a language specification.
+%We will give a brief overview of these languages are organized in Sec. \ref{sec:Reason}.
+
+% \subsection{Contributions}\label{sec:contributions}
+% This work introduces a system of \textbf{typed literal macros (TLMs)} that gives library providers substantially more syntactic control than existing typed term-rewriting macro systems while maintaining the ability to reason abstractly about types, binding and segmentation.% abstract reasoning principles.
+% comparable to the level of control they have when defining a syntax dialect.
+
+% Clients apply TLMs to \emph{generalized literal forms}.
+
+% Because the context-free syntax is never extended, syntactic conflicts are not a concern.
+
+%As such, the semantics can take the type and binding structure of the surrounding program into account when validating the expansion that the TLM programmatically generates to ensure that clients can answer critical questions related to types and binding, like those enumerated in Section \ref{sec:abs-reasoning-intro}.
+Clients need not have knowledge of the implementation of the TLM or of the generated expansion, i.e.
+there are useful principles of syntactic abstraction.
+
+% The primary technical challenge has to do with the fact that the applied TLM needs to be able to splice terms out of the literal body for inclusion in the expansion. For example,
+
+ % We design our mechanism such that these locations can easily be determined from the output of the TLM.
+This is essential for our hygiene mechanism, and it is also useful in that this information can be presented to the user (e.g.
+as shown in Figure \ref{fig:first-tsm-example-marked}).
+%As such, we must develop a mechanism where 1) the positions of spliced subterms can be determined without examining the macro implementation (e.g.
+so that they can be presented to the user differently by an editor or pretty-printer, ;  and 2) the hygiene mechanism must give only portions of the expansion that correspond to these spliced subterms access to the application site context.
+
+%In order to reason about types and binding, client programmers need only have knowledge of 1) the segmentation (e.g.
+by examining a figure like this presented by a code editor or pretty-printer) and 2) a type annotation on the definition of the applied TLM.
+No other details about the applied TLM's implementation or the expansion that it generates need to be revealed to the client programmer.
+In other words,
+
+% \begin{figure}[h]
+% \begin{lstlisting}
+% PElement Nil Seq(
+%   TextNode "Hello, ", 
+%   Seq(TextNode (join(" ", Cons(first, Cons(second, Nil)))), 
+%   TextNode "!"))
+% \end{lstlisting}
+% \caption{The desugaring.}
+% \end{figure}
+
+% There is also no ambiguity with regard to which TLM has control over each form, and searching for the definition of a TLM is no more difficult than searching for any other binding, i.e.
+there are well-defined scoping rules.
+
+% In other words, TLMs maintain a useful notion of syntactic abstraction.
+%More specifically, TLMs maintain a \emph{hygienic binding discipline}, meaning that questions Questions 4 and 5 above were concerned with are disallowed entirely.
+% We will, of course, make this notion more technically precise as we continue.
+
+% \begin{figure}[h]
+% \begin{lstlisting}[numbers=none,xleftmargin=0px]
+% let syntax $strlist = $list string in 
+% $html `SURL<p>Hello, {[EURLjoin ($str ' ') ($strlist [firstSURL,EURL last])SURL]}</p>EURL`
+% \end{lstlisting} 
+% \caption{The example from Figure \ref{fig:first-tsm-example-marked}, expressed using a parametric TLM.}
+% \label{fig:first-ptsm-example-marked}
+% \end{figure}
+
+% In Secs.
+\ref{chap:uetsms} and \ref{chap:uptsms}, we assume for the sake of technical simplicity that each TLM definition is self-contained, needing no access to libraries or to other TLMs.
+This is an impractical assumption in practice.
+We relax this assumption in In
+
+% %\item \textbf{Type-specific languages}, or \textbf{TSLs}. TSLs, described 
+% In Chapter \ref{chap:tsls}, we develop a mechanism of \emph{TLM implicits} that allows library clients to contextually designate, for any type, a privileged TLM at that type.
+The semantics applies this privileged TLM implicitly to unadorned literal forms that appear where a term of the associated type is expected.
+For example, if we designate
+% %\li{$str} 
+% as the privileged TLM at the \li{string} type and 
+% %\li{$strlist}
+% as the privileged TLM at the \li{list(string)} type, we can express the example from Figure \ref{fig:first-tsm-example-marked} instead as shown in Figure \ref{fig:first-tsm-example-implicit} (assuming \li{join} has type \li{string -> list(string) -> string}.)
+% \begin{figure}[h]
+% \begin{lstlisting}[numbers=none]
+% $html`SURL<p>Hello, {[EURLjoin ' ' [firstSURL,EURL last]SURL]}</p>EURL`
+% \end{lstlisting}
+% \caption{The example from Figure \ref{fig:first-tsm-example-marked} drawn to take advantage of TLM implicits.}
+% \label{fig:first-tsm-example-implicit}
+% \end{figure}
+
+% \noindent This approach is competitive in cost with library-specific syntax dialects (e.g.
+compare Figure \ref{fig:first-tsm-example-implicit} to Figure \ref{fig:urweb}), while maintaining the abstract reasoning principles characteristic of our approach.
+%\item \textbf{Metamodules}, introduced in Sec.
+\ref{sec:metamodules}, reduce the need to primitively build in the type structure of constructs like records (and variants thereof),  labeled sums and other interesting constructs that we will introduce later by giving library providers programmatic ``hooks'' directly into the semantics, which are specified as a \emph{type-directed translation semantics} targeting a small \emph{typed internal language} (introduced in Sec.
+\ref{sec:Reason}).
+%For example, a library provider can implement the type structure of records with a metamodule that:
+%\begin{enumerate}
+%\item introduces a type constructor, \lstinline{record}, parameterized by finite mappings from labels to types, and defines, programmatically, a translation to unary and binary product types (which are built in to the internal language); and
+%\item introduces operators used to work with records, minimally record introduction and elimination (but perhaps also various functional update operators), and directly implements the logic governing their typechecking and translation to the IL (which builds in only nullary and binary products).
+%\end{enumerate}
+%We will see direct analogies between ML-style modules (which our mechanisms also support) and metamodules later.
+%\end{enumerate}
+
+% As vehicles for this work, we will define a small programming language in each of the three parts just mentioned, each building conceptually upon the previous language.
+All of our formal contributions are relative to these small languages.
+
+%TLMs, like other macro systems, perform \emph{static code generation} (also sometimes called \emph{static} or \emph{compile-time metaprogramming}), meaning that the relevant rules in the static semantics of the language call for the evaluation of \emph{static functions} that generate term encodings.
+Static functions are functions that are evaluated statically, i.e.
+during typing.
+%Library providers write these static functions using the Reason \emph{static language} (SL).
+%Maintaining a separation between the static (or ``compile-time'') phase and the dynamic (or ``run-time'') phase is an important facet of Reason's design.
+% static code generation.
+%We will  also introduce a simple variant of each of these primitives that leverages Reason's support for local type inference to further reduce syntactic cost in certain common situations.
+
+\newcommand{\seTLMsSec}{Expression TLMs}
+\section{\protect\seTLMsSec}
+\label{sec:setsms}
+
+We start in this section by describing the TLM mechanism in detail by way of a substantial case study: literal notation, like that built in to Ur/Web (see Figure \ref{fig:urweb}) and also, presently, Reason, for expressions of the type  \li{html} in Figure \ref{fig:html-type-def}.
+
+\subsection{TLM Definition and Application}
+The definition of the \li{$html} TLM is given in Figure \ref{fig:html-tlm-def}. TLM names are prefixed by \li{$} 
+ to distinguish them from variables.
+Every TLM definition also includes a \emph{type annotation}, here \li{at html}, and a \emph{parse function} between \li{by static \{} and \li{\}}.
+% The TLM definition follows the same scoping rules as type definitions.
+We will describe how TLM definitions are packaged into libraries in Sec \ref{sec:static-eval}.
+
+Using this TLM, we can express the Ur/Web example from Figure \ref{fig:urweb} as shown in Figure \ref{fig:first-tsm-example}. On both Lines 1 and 2, we apply 
+\li{$html} 
+ to a \emph{generalized literal form} delimited by \li{[|} and \li{|]}.
+Generalized literal forms, which first arose in the prior work on TSLs \cite{TSLs}, syntactically subsume other literal forms because the context-free syntax of the language only specifies the outer delimiters.
+In this paper, we will use  \li{[|} and \li{|]}, but \citet{TSLs} formally specified several other choices, including layout-sensitive delimitation.
+\emph{Literal bodies} are constrained only in that \li{[|} and \li{|]} must be balanced.
+%In particular, there is no pre-specified syntactic to splice out sub-terms.%The initial context-free parser therefore parses generalized literal forms in much the same way as parsers typically parse ``raw'' string literals (i.e.
+string literals where escape sequences like \li{\\n} have not yet been processed.)
+
+ The system delegates responsibility over the parsing and expansion of each literal body to the applied TLM's parse function during a semantic phase called \emph{expansion}, which starts just before and continues into the typing phase.
+
+ Because the parse function is applied during this phase, rather than at run-time, we call it a \emph{static function}.
+Static functions  cannot refer to the surrounding variable bindings because those variables stand for run-time values.
+Instead, there is are separate \emph{static bindings} marked by the \li{static} keyword that populate a \emph{static environment} that is discarded after expansion finishes.
+A more detailed account of static evaluation, both informal and informal, is available in the supplement.
+An alternative design that allows for the explicit lowering of standard-phase modules to the static phase has also been proposed for OCaml \cite{Ocaml/macros}.
+%\ificfp We are closely following this work in our implementation.
+\else \fi
+  % return to in Sec. \ref{sec:static-eval}.
+
+ The input type of the parse function, \li{body}, classifies encodings of literal {bodies}.
+Literal bodies are sequences of characters, so we define \li{body} as a synonym of \li{string} in \autoref{fig:indexrange-and-parseresult}.
+The return type is a sum type, defined by applying the parameterized type \li{parse_result} defined in \autoref{fig:indexrange-and-parseresult}, that distinguishes between parse errors and successful parses.
+Let us consider these two possibilities in turn.
+
+% \vspace{-3px}\paragraph{Parse Errors} 
+If the parse function determines that the literal body is not well-formed according to the syntax that it implements, it must return \li{ParseError \{msg=}$e_\text{msg}$\li{, loc=}$e_\text{loc}$\li{\}}
+where $e_\text{msg}$ is a custom error message and $e_\text{loc}$ is a value of type \li{segment}, defined in Figure \ref{fig:indexrange-and-parseresult}, that designates a segment of the literal body as the origin of the error \cite{DBLP:journals/jsc/DeursenKT93}.
+
+% \vspace{-3px}\paragraph{Success} 
+If instead parsing succeeds, the parse function returns \li{Success} ~$\ecand$, 
+where $\ecand$ is called the \emph{encoding of the proto-expansion}.
+For expression TLMs, the proto-expansion is a \emph{proto-expression} and it is encoded as a value of the recursive datatype \li{proto_expr} that is outlined in Figure \ref{fig:candidate-exp-Reason}.
+Most of the constructors of \li{proto_expr} are individually uninteresting -- they encode OCaml's various expression forms. 
+Expressions can mention types, so we also need the type \li{proto_typ} also outlined in Figure \ref{fig:candidate-exp-Reason}.
+It is only the \li{SplicedE} and \li{SplicedT} constructors that are novel.
+These are discussed next.
+
+% \section{Simple Expression TLMs By Example}\label{sec:tsms-by-example}
+% We begin in this section with a ``tutorial-style'' introduction to seTLMs in Reason.
+%In particular, we will define an seTLM for constructing values of the recursive labeled sum type \li{rx} that was defined in Figure \ref{fig:datatype-rx}.
+% Sec.
+\ref{sec:tsms-minimal-formalism} then formally defines a reduced dialect of Reason called $\miniVerseUE$.
+This will serve as a ``conceptually minimal'' core calculus of TLMs, in the style of the simply typed lambda calculus.
+%We conclude in Sec.
+\ref{sec:uetsms-discussion}
+
+% \subsection{TLM Application}\label{sec:uetsms-usage}
+% The following Reason expression, drawn textually, is of \emph{TLM application} form. Here, a TLM named \li{$rx} is applied to the \emph{generalized literal form} \li{/SURLA|T|G|CEURL/}:
+% \begin{lstlisting}[numbers=none,mathescape=|]
+% $rx /SURLA|T|G|CEURL/
+% \end{lstlisting}
+% Generalized literal forms are left unparsed according to the context-free syntax of Reason.
+Several other outer delimiters are also available, as summarized in Figure \ref{fig:literal-forms}.
+The client is free to choose any of these for use with any TLM, as long as the \emph{literal body} (shown in green above) satisfies the requirements stated in Figure \ref{fig:literal-forms}.
+For example, we could have equivalently written the example above as \li{$rx `SURLA|T|G|CEURL`}.
+(In fact, this would have been convenient if we had wanted to express a regex containing forward slashes but not backticks.)
+
+% It is only during the subsequent \emph{typed expansion} phase that the applied TLM parses the {body} of the literal form to generate a \emph{proto-expansion}.
+The language then \emph{validates} this proto-expansion according to criteria that we will describe in Sec.
+\ref{sec:uetsms-validation}.
+If proto-expansion validation succeeds, the language generates the \emph{final expansion} (or more concisely, simply the \emph{expansion}) of the TLM application.
+The behavior of the program is determined by its expansion.
+
+% For example, the expansion of the TLM application above is equivalent to the following expression when the regex value constructors \li{Or} and \li{Str} are in scope:
+% \begin{lstlisting}[numbers=none]
+% Or(Str "SSTRAESTR", Or(Str "SSTRTESTR", Or(Str "SSTRGESTR", Str "SSTRCESTR")))
+% \end{lstlisting}
+% To avoid the assumption that the variables \li{Or} and \li{Str} are in scope at the TLM application site, the expansion actually uses the explicit \li{fold} and \li{inj} operators, as described in Sec.
+\ref{sec:lists}.
+In fact, the proto-expansion validation process enforces this notion of context independence -- we will return to proto-expansion validation below.
+(We will show how TLM parameters can reduce the awkwardness of this requirement in Chapter \ref{chap:ptsms}.)
+% %The constructors above are those of the type \li{Rx} that was defined in Figure \ref{fig:datatype-rx}.
+
+% % A number of literal forms, ,  are available in Reason's concrete syntax.
+Any literal form can be used with any TLM,  TLMs have access only to the literal bodies.
+Because TLMs do not extend the concrete syntax of the language directly, there cannot be syntactic conflicts between TLMs.
+
+%  %The form does not directly determine the expansion.
+
+% \begin{figure}
+% \begin{lstlisting}
+% 'SURLbody cannot contain an apostropheEURL'
+% `SURLbody cannot contain a backtickEURL`
+% [SURLbody cannot contain unmatched square bracketsEURL]
+% {|SURLbody cannot contain unmatched barred curly bracesEURL|}
+% /SURLbody cannot contain a forward slashEURL/
+% \SURLbody cannot contain a backslashEURL\
+% \end{lstlisting}
+% %SURL<tag>body includes enclosing tags</tag>EURL
+% \caption[Available Generalized Literal Forms]{Generalized literal forms available for use in Reason's textual syntax.
+The characters in green indicate the literal bodies and describe how the literal body is constrained by the form shown on that line.
+The Wyvern language defines additional forms, including whitespace-delimited forms \cite{TSLs} and multipart forms \cite{sac15}, but for simplicity we leave these out of Reason.}
+% \label{fig:literal-forms}
+% \end{figure}
+
+\subsection{Splicing}\label{sec:splicing-and-hygiene}
+When the parse function determines that some segment of the literal body is a spliced expression, according to whatever syntactic criteria it deems suitable, it can indirectly refer to it in the encoding it produces using the \li{SplicedE} constructor of \li{proto_expr}, which takes a value of type \li{segment} that indicates the zero-indexed location of the spliced expression relative to the start of the provided literal body.
+The \li{SplicedE} constructor also requires a value of type \li{proto_typ}, which indicates the type that the spliced expression is expected to have.
+Types can be spliced out by using the \li{SplicedT} constructor of \li{proto_typ} analagously.
+
+\begin{figure}
+\begin{lstlisting}[numbers=none,xleftmargin=0pt]
+type body = string;
+type segment = {startIdx: int, endIdx: int};
+type parse_result('a) 
+  = ParseError {msg: string, loc: segment}
+  | Success('a);
+type proto_typ = Arrow(proto_typ, proto_typ)
+               | StringTy
+               | /* ... */ 
+               | SplicedT(segment);
+type proto_expr = Tuple(list(proto_expr))
+                | /* ... */
+                | SplicedE(segment, proto_typ);
+\end{lstlisting}
+\caption[Definitions of various types used by TLM definitions.]{Definitions of various types available ambiently to TLM definitions.}
+\label{fig:indexrange-and-parseresult}
+\label{fig:candidate-exp-Reason}
+\end{figure}
+
+For example, consider again the two TLM applications in Figure \ref{fig:first-tsm-example}.
+In each case, the parse function of the \li{$html} TLM (Figure \ref{fig:html-tlm-def}, Lines 2-4) first sends the literal body through an off-the-shelf HTML parser, \li{parse_html}.
+It then passes the result to a function \li{html_to_ast}, not shown, which produces the corresponding expression encoding of type \li{proto_expr}.
+When this function encounters an HTML text node containing matched \li{\{[} and \li{]\}}, then that segment is inserted as a spliced expression of type \li{string}, and similarly text nodes containing matched curly braces produce a spliced expression of type \li{html}.
+For instance,  the proto-expansion generated for first TLM application in \ref{fig:first-tsm-example}, pretty-printed, is:
+\begin{lstlisting}[numbers=none]
+    H1Element(Nil, Cons(TextNode "Hello, ", Cons(
+      TextNode spliced<13; 22; string>, Nil)))
+\end{lstlisting}
+Here, \li{spliced<13; 22; string>} is a reference to the spliced string expression \li{first_name} by its location relative to the start of the literal body being expanded (the off-the-shelf HTML parser provides the necessary baseline location information for use by \li{html_to_ast}).
+It corresponds to the encoding \li{SplicedE(\{startIdx=13, endIdx=22\}, StringTy)}.
+Requiring that TLMs refer to spliced expressions indirectly in this manner ensures that a TLM cannot ``forge'' spliced terms, i.e.
+claim that some sub-term of the expansion should be given the privileges of a spliced term, discussed in Sec.
+\ref{sec:uetsms-validation}, when it does not in fact appear in the literal body.% Our semantics distinguishes spliced expressions when validating the proto-expansion, which is important for reasons of hygiene that we will detail shortly.
+
+% The parse function can similarly extract \emph{spliced types} from a literal body using the \li{SplicedT} variant of \li{proto_typ}.
+%In particular, the parse function must provide the index range of spliced subexpressions to the \li{Spliced} constructor of the type \li{MarkedExp}.
+%Only subexpressions that actually appear in the body of the literal form can be marked as spliced subexpressions.
+
+%For example, had the  would not be a valid expansion, because the  that are not inside spliced subexpressions:
+%\begin{lstlisting}[numbers=none]
+%Q.Seq(Q.Str(name), Q.Seq(Q.Str ": ", ssn))
+%\end{lstlisting}
+
+% \subsection{Splice Summaries and Segmentations}
+The \emph{segmentation} inferred from a proto-expansion is the finite set of references to spliced terms contained within.
+For example, the segmentation inferred from the proto-expression above contains only \li{spliced<13; 22; string>}.
+% Notice that no information about  the spliced terms appear is communicated by the splice summary.
+The system checks that all of the locations in the segmentation are 1) in bounds relative to the literal body; and 2) non-overlapping. 
+This resolves the problem of \textbf{Segmentation} described in Sec.
+\ref{sec:intro}, i.e.
+every literal body in a well-formed program has a well-defined segmentation.
+The TSL mechanism  did not maintain this reasoning principle \cite{TSLs}.
+A program editor or pretty-printer can communicate this segmentation information to the programmer, e.g.
+by coloring non-spliced segments green as is our convention in this document.
+In general, spliced expressions might themselves apply TLMs, in which case the convention is to use a distinct color for unspliced segments at each depth.
+For example, consider a TLM \li{$smiles} for chemical structures \cite{anderson1987smiles} with support for splicing using curly braces:
+\begin{lstlisting}[numbers=none]
+    $html [|SURLChemical structure of sucrose: {EURL
+        $smiles [|SCSS{ECSSm_glucoseSCSS}-O-{ECSSm_fructoseSCSS}ECSS|] 
+        |> SMILES.to_svg SURL}EURL|]
+\end{lstlisting}
+%For example, if strings were not primitive but rather defined as sequences of characters, we might define a TLM \li{$str} to recover string literal notation and write \li{$html [|SURL<body>\{EURLheading $str [|SSTRWorld!ESTR|]SURL\} ...</body>EURL|]}.
+
+A program editor or pretty-printer can communicate the type of each spliced expression, also specified abstractly by the segmentation, upon request (for Reason, via Merlin \cite{Merlin}.) %When there is a type mismatch, the type annotation on the spliced segment also allows the error message to report the problem without revealing the full expansion.
+
+\subsection{Proto-Expansion Validation}\label{sec:uetsms-validation}
+Three important concerns described in Sec.
+\ref{sec:intro} remain: those related to reasoning abstractly about the \emph{hygiene properties}, i.e.
+\textbf{Capture} and \textbf{Context Dependence}, and \textbf{Typing}.
+Addressing these concerns is the purpose of the \emph{proto-expansion validation} process, which occurs during the typing phase.
+Proto-expansion validation results in the \emph{final expansion}, which is simply the proto-expansion with the references to spliced segments replaced with their own final expansions.
+
+\subsubsection{Capture}\label{sec:capture}
+
+Proto-expansion validation ensures that spliced terms have access \emph{only} to the bindings at the application site---spliced terms cannot capture bindings internal to the proto-expansion.
+For example, consider the following application site:
+\begin{lstlisting}[numbers=none]
+    let tmp = /* ... application site temp ... */;
+    $html [|SURL<h1>{EURLf(tmp)SURL}</h1>EURL|];
+\end{lstlisting}
+Now consider the scenario where the proto-expansion generated by \li{$html} has the following form:
+\begin{lstlisting}[numbers=none]
+    let tmp = /* ... expansion-internal temp ... */;
+    H1Element(tmp, spliced<5; 10; html>);
+\end{lstlisting}
+Na\"ively, the binding of the variable \li{tmp} in the proto-expansion could shadow the application-site binding of \li{tmp} in the final expansion. 
+To address this problem, splicing is guaranteed to be capture-avoiding.
+When generating the final expansion, the system discharges the requirement that capture not occur by implicitly alpha-varying the bindings in the proto-expansion as needed.
+There is no need for TLM providers to explicitly deploy a mechanism that generates fresh variables (as in, e.g., prior \emph{reader macro} systems \cite{Flatt:2012:CLR:2063176.2063195}.
+% For example, the final expansion of the example above might take the following form:
+%\begin{lstlisting}[numbers=none]
+%  let tmp = /* ... application site temporary ... */; 
+%  let tmp' = /* ... expansion-internal temporary ... */;
+%  H1Element(tmp', f(tmp));
+%\end{lstlisting}
+%Notice that the expansion-internal binding of \li{tmp} has been alpha-varied to \li{tmp'}.
+The reference to \li{tmp} in the spliced expression then refers, as intended, to the application site binding.
+
+% For TLM providers, the benefit of this mechanism is that they can name the variables used internally within expansions freely. TLM clients can, in turn, can reason abstractly about \textbf{Capture}.
+
+Capture avoidance does prevent library providers from intentionally introducing bindings into spliced terms.
+For example, Haskell's literal notation for monadic values, i.e.
+\li{do}-notation, cannot be expressed because it introduces a new binding syntax \cite{jones2003haskell}.
+
+In our initial designs, we were able to express this example because we distinguished spliced identifiers, much as we do spliced types and expressions.
+These could be explicitly bound inside spliced expressions by extending the \li{SplicedE} and \li{SplicedT} constructors to take finite sets of spliced identifiers.
+This is conceptually similar to the approach taken by \citet{DBLP:conf/esop/HermanW08} for giving (already parsed) sub-terms access to internal variables.
+However, we ultimately decided to remove this feature because it would disproportionately increase the reasoning burden on clients when they encounter an unfamiliar literal: \emph{might this obscure literal also be introducing bindings?} It is difficult for a program editor to display a set of ``hidden bindings'' to the programmer.
+Haskell-style infix notation for monadic values, where bind is \li{e >>= f}, can be expressed using TLMs without support for spliced identifiers.
+Reason 3.0 is adopting more concise function syntax, which will decrease the cost of this syntax.
+%We will show an alternative formulation of Haskell's syntax for monadic commands that uses Reason's anonymous function syntax to bind variables in Sec. \ref{sec:application-monadic-commands}.
+
+\subsubsection{Context Dependence}\label{sec:context-dependence}
+%The prohibition on shadowing ensures only that variables that appear in spliced terms do not refer to bindings that appear in the surrounding expansion. 
+The proto-expansion validation process also ensures that variables that appear in the proto-expansion do not refer to bindings that appear either at the TLM definition or the application site.
+In other words, expansions must be completely \emph{context independent} -- they can make no assumptions about the surrounding context whatsoever.
+
+A minimal example of a ``broken'' TLM that does not generate context-independent proto-expansions is below:
+\begin{lstlisting}[numbers=none]
+    syntax $broken at t by static { 
+      fun(_) => Success (Var "SSTRxESTR") };
+\end{lstlisting}
+The proto-expansion that this TLM generates (for any literal body) refers to a variable \li{x} that it does not itself bind.
+If proto-expansion validation permitted such a proto-expansion, it would be well-typed only under those application site typing contexts where \li{x} is bound.
+This ``hidden assumption'' makes reasoning about binding and renaming difficult.
+
+Of course, this prohibition does not extend into the spliced terms in a proto-expansion -- spliced terms appear at the application site, so they can justifiably refer to application site bindings.
+(like \li{first_name} in Fig.
+\ref{fig:first-tsm-example}.) Because proto-expansions refer to spliced terms indirectly, enforcing context independence is straightforward -- we need only that the proto-expansion itself be closed.% In the next section, we will formalize this intuition.
+% The TLM provider can only refer to them opaquely.
+
+Na\"ively, this restriction, also present in the prior work on TSLs \cite{TSLs}, is quite restrictive -- expansions cannot access any library functions.
+At best, they can require the client to ``pass in'' required library functions via splicing at every application.
+In Sec.
+\ref{sec:ptsms}, we will introduce module parameters and partial parameter application to neatly resolve this problem.
+
+% This prohibition on context dependence explains why the expansion generated by the TLM application in Sec.
+\ref{sec:uetsms-usage} cannot make use of the regex value constructors, e.g.
+\li{Str} and \li{Or}, directly.
+(In Chapter \ref{chap:ptsms}, we will relax this restriction to allow proto-expansions to access explicit parameters.)
+
+% Collectively, we refer to the prohibition on capture and the prohibition on context dependence as \emph{hygiene properties}, by conceptual analogy to corresponding properties in term-rewriting macro systems (see Sec.
+\ref{sec:macro-systems}.) The novelty here comes from the fact that spliced terms are being extracted from an initially unparsed sequence of characters.
+% In the examples in Sec.
+\ref{sec:uetsms-usage} and Sec.
+\ref{sec:splicing-and-hygiene}, the expansion used constructors associated with the \li{Rx} type, e.g.
+\li{Seq} and \li{Str}.
+This might appear to violate our prohibition on context-dependent expansions.
+This is not the case only because in Reason, constructor labels are not variables or scoped symbols.
+Syntactically, they must begin with a capital letter (like Haskell's datatype constructors).
+Different labeled sum types can use common constructor labels without conflict because the type the term is being checked against -- e.g.
+\li{Rx}, due to the type ascription on \li{$rx} -- determines which type of value will be constructed.
+For dialects of ML where datatype definitions do introduce new variables or scoped symbols, we need parameterized TLMs.
+We will return to this topic in Chapter \ref{chap:ptsms}.
+% Indeed, we used the label \li{Spliced} for two different recursive labeled sum types in Figure \ref{fig:candidate-exp-Reason}.
+
+\subsubsection{Typing}\label{sec:typing-e}
+Finally, proto-expansion validation maintains a reasonable {typing discipline} by (1) checking that the expansion is of the type specified by the TLM's type annotation; (2) checking that each spliced type is valid; (3) checking that the type annotation on each spliced expression is valid; and (4) checking each spliced expression against the specified type annotation.
+Context independence implies that ML-style type inference can be performed using only the segmentation (because the remainder of an expansion cannot mention the very variables whose types are being inferred).
+In the prior work on TSLs, spliced terms did not have type annotations
+%The OCaml type system is not strong enough to allow us to express only contextually well-typed syntax trees. In The details are discussed as future work in Sec. \ref{sec:discussion}.
+ % This addresses the problem of reasoning abstractly about \textbf{Typing} described in Sec. \ref{sec:intro}, i.e.:
+ % \begin{enumerate}
+ %   \item determining the type of an expansion requires examining only the type annotation on the TLM definition (much as determining the type of a function application requires examining only the function's type); and
+ % \item determining the type that a spliced expression must have requires only the information in the splice summary (rather than complete knowledge of the proto-expansion).
+ % \end{enumerate}
+
+% The language \emph{validates} proto-expansions before a final expansion is generated.
+One aspect of proto-expansion validation is checking  the proto-expansion against the type annotation specified by the TLM, e.g.
+the type \li{Rx} in the example above.
+This maintains a \emph{type discipline}: if a programmer sees a TLM being applied when examining a well-typed program, they need only look up the TLM's type annotation to determine the type of the generated expansion.
+Determining the type does not require examining the expansion directly.
+
+% \subsection{Hygiene}
+% The spliced subexpressions that the proto-expansion refers to (by their position within the literal body, cf.
+above) must be parsed, typed and expanded during the proto-expansion validation process (otherwise, the language would not be able to check the type of the proto-expansion).
+To maintain a useful \emph{binding discipline}, i.e.
+to allow programmers to reason also about variable binding without examining expansions directly, the validation process maintains two additional properties related to spliced subexpressions: \textbf{context independent expansion} and \textbf{expansion independent splicing}.
+These are collectively referred to as the \emph{hygiene properties} (because they are conceptually related to the concept of hygiene in term rewriting macro systems, cf.
+Sec.
+\ref{sec:term-rewriting}.)
+
+% \paragraph{Context Independent Expansion}
+
+% \paragraph{Expansion Independent Splicing} 
+% %These properties suffice to ensure that programmers and tools can freely rename a variable without changing the meaning of the program.
+The only information that is necessary to perform such a \emph{rename refactoring} is the locations of spliced subexpressions within all the literal forms for which the variable being renamed is in scope; the expansions need not otherwise be examined.
+It would be straightforward to develop a tool and/or editor plugin to indicate the locations of spliced subexpressions to the user, like we do in this document (by coloring spliced subexpressions black).
+We discuss tool support as future work in Sec.
+\ref{sec:interaction-with-tools}.
+
+% \subsubsection{Final Expansion}
+
+% For example, the final expansion of the body of \li{lookup_rx} is equivalent to the following, under an environment where the regex value constructors are available:
+% \begin{lstlisting}[numbers=none]
+% Seq(Str(name), Seq(Str "SSTR: ESTR", ssn))
+% \end{lstlisting}
+% (Again, due to the prohibition on context dependent expansions, the final expansion actually involves explicit \li{fold} and \li{inj} operators.)
+
+\newcommand{\spTLMsSec}{Pattern TLMs}
+\section{\protect\spTLMsSec}
+\label{sec:sptsms}
+
+Let us now briefly consider the topic of TLMs that generate \emph{patterns}, rather than expressions.
+Pattern literals are the dual to expression literals in that expression literals support construction whereas pattern literals support deconstruction \cite{moriconi2008inversion}.
+For example, we can pattern match on a value \li{x : html} by applying a pattern TLM \li{$html} as follows:
+\begin{lstlisting}[numbers=none]
+    switch x {
+    | $html [|SURL<h1>{EURLcsSURL}</h1>EURL|] => /* cs:list(html) */
+    | _ -> None
+    };
+\end{lstlisting}
+Any list pattern, including one generated by another TLM application, can appear where \li{cs} appears in the example pattern above.
+For longer \li{switch} expressions, the shorthand \li{switch x using $html} applies \li{$html} to every rule where the outermost pattern is of generalized literal form.
+
+Notice that we can use the same name, \li{$html}, for this pattern TLM as for the expression TLM defined in the previous section.
+It does not make sense to apply an expression TLM in pattern position  (many expression-level constructs, e.g.
+lambdas, do not correspond even syntactically to patterns), and \emph{vice versa}, so this is unambiguous.
+
+Pattern TLM definitions differ from expression TLM definitions in two ways: (1) a \emph{sort qualifier}, \li{for patterns}, distinguishes them from expression TLM definitions; (2) the return type of the parse function is \li{parse_result(proto_pat)}, rather than \li{parse_result(proto_expr)}.
+The type \li{proto_pat}, outlined below, classifies encodings of \emph{proto-patterns}.
+\begin{lstlisting}[numbers=none]
+   type proto_pat = /* no variable pattern form! */
+                  | Wild
+                  | /* ... */
+                  | SplicedP(segment, proto_typ);
+\end{lstlisting}
+
+% \subsection{Splicing}
+The constructor \li{SplicedP} operates much like \li{SplicedE} to allow a proto-pattern to refer indirectly to spliced patterns.% by their location within the literal body.
+
+%  For example, the proto-pattern generated for the example at the top of this section would be written concretely as follows:
+% \begin{lstlisting}[numbers=none]
+%   H1Element (_, spliced<6; 7; list(html)>)
+% \end{lstlisting}
+
+% \subsection{Proto-Pattern Validation}
+% Proto-pattern validation serves, like proto-expression validation, to maintain the ability to reason abstractly about binding and typing.
+
+To maintain the abstract binding discipline, variable patterns can appear only within spliced patterns.
+Enforcing this restriction is straightforward: we simply have not defined a variant of the \li{proto_pat} type that encodes variable patterns (wildcards are allowed.)  This restriction ensures that only variables visible to the client in a spliced pattern are bound in the corresponding branch expression.
+This is analagous to the capture avoidance principle for expression TLMs.
+
+Type annotations on references to spliced patterns could refer to type variables, so we also need to enforce context independence in the manner discussed in the previous section.
+%Pattern guards are part of pattern matching rules, not patterns themselves, in Reason/OCaml, but were they part of patterns, context independence would also need to be enforced for guard expressions.% In languages like OCaml, which support arbitrary boolean guard expressions, we would also need to enforce both context independence and capture avoidance as discussed in the previous section for spliced expressions that end up in a guard expression.
+In our formalism, we do not support guard expressions.
+
+% \subsubsection{Typing} 
+To maintain an abstract typing discipline, proto-pattern validation checks type annotations much as in Sec.
+\ref{sec:typing-e}.% (1) that the final expansion is a pattern that matches values of the type specified by the TLM's type annotation; (2) that each spliced pattern matches values of the type indicated in the segmentation; and (3) that each of these types are themselves well-formed types.
+
+\newcommand{\pTLMsSec}{Parametric TLMs}
+\section{\protect\pTLMsSec}
+\label{sec:ptsms}
+
+The simple TLMs in the previous sections operate only at one specified type, as did the TSLs in the prior work \cite{TSLs}.
+This is rather limiting.
+This section introduces \emph{parametric TLMs}, which can operate over a type- and module-parameterized family of types.
+They also neatly solve the problem discussed in Sec.
+\ref{sec:context-dependence} of giving expansions access to helper functions.
+%Moreover, as discussed at the end of Sec.
+\ref{sec:context-dependence}, the expansions that they generate have no access to libraries.
+This section introduces \emph{parametric TLMs} to neatly resolve both of these limitations (which also limited the prior work on TSL \cite{TSLs}.) Parametric TLMs can be defined over a type- and module-parameterized family of types, and the proto-expansions they generate can refer to supplied type and module parameters.
+Partial parameter application decreases the syntactic cost of this explicit parameter passing style.
+% \subsection{Parametric TLMs By Example}\label{sec:ptsms-by-example}
+
+Consider the following Reason/OCaml module type (a.k.a. signature), which specifies an abstract data type \cite{liskov1974programming,harper1997programming} of string-keyed polymorphic dictionaries:
+\begin{lstlisting}[numbers=none]
+    module type DICT = {
+      type t('a); 
+      let empty : t('a); 
+      let extend : t('a) -> (string, 'a) -> t('a); 
+      /* ... */ 
+    };
+\end{lstlisting}
+We can define a TLM that is parametric over implementations of this signature, \li{D : DICT}, and over choices of the codomain type, \li{'a} as follows:
+\begin{lstlisting}[numbers=none]
+    syntax $dict (D : DICT) (type 'a) at D.t('a) 
+    by static { fun(b) => /* ... */ };
+\end{lstlisting}
+For example, given some module that implements this signature, \li{HashDict : DICT}, we can apply \li{$dict} as follows:
+\begin{lstlisting}[numbers=none]
+    $dict HashDict int [|"key1"SURL=>EURL10SURL; EURL"key2"SURL=>EURL15|]
+\end{lstlisting}
+Notice that the segmentation immediately reveals which punctuation is particular to this TLM and where the spliced key and value expressions appear.
+Because the context-free syntax of unexpanded terms is never modified, it is possible to reason modularly about syntactic determinism, i.e.
+we can reason above that \li{=>} does not appear in the follow set of unexpanded expressions \cite{conf/pldi/SchwerdfegerW09}, so there can never be an ambiguity about where a key expression ends.
+
+The proto-expansion generated for the TLM application above, shown below, can refer to the parameters:
+\begin{lstlisting}[numbers=none]
+    D.extend(D.extend D.empty (spliced<1;6;string>, 
+      spliced<11;12;'a>)) (spliced<15;20;string>, 
+      spliced<25;26;'a>)
+\end{lstlisting}
+Validation checks that the proto-expansion is truly parametric, i.e.
+it must be valid for all modules \li{D : DICT} and types \li{'a}.
+It is only after validation that we substitute the actual parameters, here \li{HashDict} for \li{D} and \li{int} for \li{'a}, into the final expansion.
+Only the type annotations on references to spliced terms are subject to early parameter substitution (because they classify application site terms.)
+
+If we will use the \li{HashDict} implementation ubiquitously, we can abbreviate the partial application of \li{$dict} to \li{HashDict}, resulting in a TLM that is parametric over only the type \li{'a}:
+\begin{lstlisting}[numbers=none]
+    syntax $dict' = $dict HashDict; 
+\end{lstlisting}
+TLM abbreviations can themselves be parameterized to support partial application of parameters other than the last.
+
+In Sec.
+\ref{sec:context-dependence} we discussed the problem of the strict context independence discipline being too restrictive, in that it would seem to restrict expansions from referring to useful helpers bound at the TLM definition site.
+Module parameters address this problem -- the helper values, types and modules can be packaged into a module, passed in and partially applied to hide this detail from clients.
+Because this will be common in practice, we provide the following shorthand:
+\begin{lstlisting}[numbers=none]
+    syntax $a at t using X~${}_1$~=M~${}_1$~,...,X~${}_n$~=M~${}_n$~ by ...
+\end{lstlisting} 
+% This is the same as a definition that specifies $n$ type, expression or module parameters and immediately applies the given arguments, binding the result to the same TSM name.
+
+This explicit parameter passing discipline is reminiscent of work on explicit capture for distributed functions \cite{DBLP:conf/ecoop/MillerHO14}.
+By not implicitly giving expansions access to all definition-site bindings, we need not examine parse functions to reason about, e.g., renaming.
+Consequently, encodings of proto-terms can be values of standard datatypes (e.g.
+\li{proto_expr}) with variables represented simply as, e.g., strings, and quasiquotation notation can be expressing using TLMs (see supplement).
+A central design goal of Reason is to leave the OCaml semantics unchanged, so building quasiquotation in primitively, to integrate the free variables in term encodings into the overall binding discipline \cite{Bawd99a}, was in any case infeasible.
+
+\newcommand{\moreExamplesSec}{Additional Examples}
+\section{\protect\moreExamplesSec}
+\label{sec:more-examples}
+
+\subsection{HTML}
 \begin{figure*}[t]
@@
 \end{figure*}
-
-\paragraph{Outline} 
-% \begin{itemize}
-% \item 
-We begin in Sec.~\ref{sec:setsms} by introducing TLMs with a detailed case study. 
-% \item 
-Sec.~\ref{sec:sptsms} then introduces {pattern TLMs}, i.e.
-TLMs that generate patterns, and describes the special hygiene conditions in pattern position.
-%The hygiene condition for pattern TLMs is interesting.
-% \item 
-Sec.~\ref{sec:ptsms}  introduces \emph{parametric TLMs} (pTLMs), i.e.
-TLMs that can take type and module parameters, which allow us to define literal notation over a type- or module-parameterized family of types.
-Sec.~\ref{sec:ptsms} also reveals that the \li{expansions require} clause is syntactic sugar for parameterization followed immediately by partial parameter application.
-Having introduced the basic machinery, we continue in Sec.~\ref{sec:more-examples} with additional examples that demonstrate both the expressive power of this approach and certain alternative parser implementation strategies that notation providers might consider.
-Sec.~\ref{sec:setsms-formally} presents a type-theoretic calculus of simple expression and pattern TLMs and formally establishes the reasoning principles implied above, in the formally simplest setting.
-Sec.~\ref{sec:ptlms-formally} adds type functions and an ML-style module system to this calculus, and gives the most general variant of the reasoning principles theorem. 
-% \item 
-Sec.~\ref{sec:existing-approaches} compares TLMs to related work, guided by the rubric of five reasoning principles just discussed. 
-% \item 
-Sec.~\ref{sec:discussion} concludes with a summary of the contributions of this paper and a discussion of future work.
-
-% Parameters serve two important purposes.
-First, they allow a TLM to operate not just at a single type, but over a type- and module-parameterized family of types.
-We give as an example a parametric TLM for string-keyed dictionaries that operates uniformly over different implementations of the abstract dictionary signature, and also over different choices of the range type.
-Second, parameters lower the syntactic cost of the strict context independence discipline that TLMs enforce because they give expansions access to terms other than those explicitly spliced in to the literal body by the client, and because %This is particularly convenient for working with abstract types \emph{a la} ML.
-% partial parameter application is possible.
-% Sec.
-\ref{sec:ptsms} culminates with a calculus that extends the calculus of Sec.
-\ref{sec:setsms-formally} with type functions, ML-style modules and parametric TLMs.
-% \item
-% \item 
-% \end{itemize}
-
-Due to the space limitations, certain technical details and extensions to the calculi presented in this paper, as well as detailed proofs, are available in the supplemental material.
-% Also available is an additional series of case studies: notation for an encoding of regular expressions as an abstract data type, implemented using TLMs for quasiquotation and grammar-based parser generators.
-%are also available in the supplement.
-Sec.~\ref{sec:static-eval} considers the topic of term evaluation during the  expansion phase, i.e.
-\emph{static evaluation}, in  more detail.
-Sec \ref{sec:static-eval} also gives examples of TLMs that are useful for defining other TLMs, e.g.
-TLMs that implement parser generators and quasiquotation.
-
-%  We say more about LISP macros in Sec.
-\ref{sec:existing-approaches}.% For the present purposes, we can consider Reason as essentially the calculus defined in the supplement extended with various conveniences that are commonly found in other ML-like languages and, notionally, orthogonal to TLMs.
-%Reason is, as its name suggests, a conceptual descendent of ML.
-It diverges from other dialects of ML that have a similar type structure in that it has a bidirectional type system \cite{Pierce:2000:LTI:345099.345100} (like, for example, Scala \cite{OdeZenZen01}) for reasons that have to do with the mechanism of TLM implicits described in Chapters \ref{chap:tsls} and \ref{chap:ptsms}.
-%The reason we will not follow Standard ML \cite{mthm97-for-dart} in giving a complete formal definition of Reason in this work is both to emphasize that the primitives we introduce are ``insensitive'' to the details of the underlying type structure of the language (so TLMs can be considered for inclusion in a variety of languages, not only dialects of ML), and to avoid distracting the reader (and the author) with definitions that are already well-understood in the literature and that are orthogonal to those that are the focus of this work.
-% We will not formally define these features mainly to avoid unnecessarily complicating our presentation with details that are not essential to the ideas presented herein. As such, 
-% All examples written in Reason should be understood to be informal motivating material for the subsequent formal material.
-%We anticipate that future full-scale language specifications will be able to combine the ideas  in the proposed work without trouble.
-%The purpose of the work being proposed is to serve as a reference for those interested in the new constructs we introduce, not to serve as a language specification.
-%We will give a brief overview of these languages are organized in Sec. \ref{sec:Reason}.
-
-% \subsection{Contributions}\label{sec:contributions}
-% This work introduces a system of \textbf{typed literal macros (TLMs)} that gives library providers substantially more syntactic control than existing typed term-rewriting macro systems while maintaining the ability to reason abstractly about types, binding and segmentation.% abstract reasoning principles.
-% comparable to the level of control they have when defining a syntax dialect.
-
-% Clients apply TLMs to \emph{generalized literal forms}.
-
-% Because the context-free syntax is never extended, syntactic conflicts are not a concern.
-
-%As such, the semantics can take the type and binding structure of the surrounding program into account when validating the expansion that the TLM programmatically generates to ensure that clients can answer critical questions related to types and binding, like those enumerated in Section \ref{sec:abs-reasoning-intro}.
-Clients need not have knowledge of the implementation of the TLM or of the generated expansion, i.e.
-there are useful principles of syntactic abstraction.
-
-% The primary technical challenge has to do with the fact that the applied TLM needs to be able to splice terms out of the literal body for inclusion in the expansion. For example,
-
- % We design our mechanism such that these locations can easily be determined from the output of the TLM.
-This is essential for our hygiene mechanism, and it is also useful in that this information can be presented to the user (e.g.
-as shown in Figure \ref{fig:first-tsm-example-marked}).
-%As such, we must develop a mechanism where 1) the positions of spliced subterms can be determined without examining the macro implementation (e.g.
-so that they can be presented to the user differently by an editor or pretty-printer, ;  and 2) the hygiene mechanism must give only portions of the expansion that correspond to these spliced subterms access to the application site context.
-
-%In order to reason about types and binding, client programmers need only have knowledge of 1) the segmentation (e.g.
-by examining a figure like this presented by a code editor or pretty-printer) and 2) a type annotation on the definition of the applied TLM.
-No other details about the applied TLM's implementation or the expansion that it generates need to be revealed to the client programmer.
-In other words,
-
-% \begin{figure}[h]
-% \begin{lstlisting}
-% PElement Nil Seq(
-%   TextNode "Hello, ", 
-%   Seq(TextNode (join(" ", Cons(first, Cons(second, Nil)))), 
-%   TextNode "!"))
-% \end{lstlisting}
-% \caption{The desugaring.}
-% \end{figure}
-
-% There is also no ambiguity with regard to which TLM has control over each form, and searching for the definition of a TLM is no more difficult than searching for any other binding, i.e.
-there are well-defined scoping rules.
-
-% In other words, TLMs maintain a useful notion of syntactic abstraction.
-%More specifically, TLMs maintain a \emph{hygienic binding discipline}, meaning that questions Questions 4 and 5 above were concerned with are disallowed entirely.
-% We will, of course, make this notion more technically precise as we continue.
-
-% \begin{figure}[h]
-% \begin{lstlisting}[numbers=none,xleftmargin=0px]
-% let syntax $strlist = $list string in 
-% $html `SURL<p>Hello, {[EURLjoin ($str ' ') ($strlist [firstSURL,EURL last])SURL]}</p>EURL`
-% \end{lstlisting} 
-% \caption{The example from Figure \ref{fig:first-tsm-example-marked}, expressed using a parametric TLM.}
-% \label{fig:first-ptsm-example-marked}
-% \end{figure}
-
-% In Secs.
-\ref{chap:uetsms} and \ref{chap:uptsms}, we assume for the sake of technical simplicity that each TLM definition is self-contained, needing no access to libraries or to other TLMs.
-This is an impractical assumption in practice.
-We relax this assumption in In
-
-% %\item \textbf{Type-specific languages}, or \textbf{TSLs}. TSLs, described 
-% In Chapter \ref{chap:tsls}, we develop a mechanism of \emph{TLM implicits} that allows library clients to contextually designate, for any type, a privileged TLM at that type.
-The semantics applies this privileged TLM implicitly to unadorned literal forms that appear where a term of the associated type is expected.
-For example, if we designate
-% %\li{$str} 
-% as the privileged TLM at the \li{string} type and 
-% %\li{$strlist}
-% as the privileged TLM at the \li{list(string)} type, we can express the example from Figure \ref{fig:first-tsm-example-marked} instead as shown in Figure \ref{fig:first-tsm-example-implicit} (assuming \li{join} has type \li{string -> list(string) -> string}.)
-% \begin{figure}[h]
-% \begin{lstlisting}[numbers=none]
-% $html`SURL<p>Hello, {[EURLjoin ' ' [firstSURL,EURL last]SURL]}</p>EURL`
-% \end{lstlisting}
-% \caption{The example from Figure \ref{fig:first-tsm-example-marked} drawn to take advantage of TLM implicits.}
-% \label{fig:first-tsm-example-implicit}
-% \end{figure}
-
-% \noindent This approach is competitive in cost with library-specific syntax dialects (e.g.
-compare Figure \ref{fig:first-tsm-example-implicit} to Figure \ref{fig:urweb}), while maintaining the abstract reasoning principles characteristic of our approach.
-%\item \textbf{Metamodules}, introduced in Sec.
-\ref{sec:metamodules}, reduce the need to primitively build in the type structure of constructs like records (and variants thereof),  labeled sums and other interesting constructs that we will introduce later by giving library providers programmatic ``hooks'' directly into the semantics, which are specified as a \emph{type-directed translation semantics} targeting a small \emph{typed internal language} (introduced in Sec.
-\ref{sec:Reason}).
-%For example, a library provider can implement the type structure of records with a metamodule that:
-%\begin{enumerate}
-%\item introduces a type constructor, \lstinline{record}, parameterized by finite mappings from labels to types, and defines, programmatically, a translation to unary and binary product types (which are built in to the internal language); and
-%\item introduces operators used to work with records, minimally record introduction and elimination (but perhaps also various functional update operators), and directly implements the logic governing their typechecking and translation to the IL (which builds in only nullary and binary products).
-%\end{enumerate}
-%We will see direct analogies between ML-style modules (which our mechanisms also support) and metamodules later.
-%\end{enumerate}
-
-% As vehicles for this work, we will define a small programming language in each of the three parts just mentioned, each building conceptually upon the previous language.
-All of our formal contributions are relative to these small languages.
-
-%TLMs, like other macro systems, perform \emph{static code generation} (also sometimes called \emph{static} or \emph{compile-time metaprogramming}), meaning that the relevant rules in the static semantics of the language call for the evaluation of \emph{static functions} that generate term encodings.
-Static functions are functions that are evaluated statically, i.e.
-during typing.
-%Library providers write these static functions using the Reason \emph{static language} (SL).
-%Maintaining a separation between the static (or ``compile-time'') phase and the dynamic (or ``run-time'') phase is an important facet of Reason's design.
-% static code generation.
-%We will  also introduce a simple variant of each of these primitives that leverages Reason's support for local type inference to further reduce syntactic cost in certain common situations.
-
-\newcommand{\seTLMsSec}{Expression TLMs}
-\section{\protect\seTLMsSec}
-\label{sec:setsms}
-
-We start in this section by describing the TLM mechanism in detail by way of a substantial case study: literal notation, like that built in to Ur/Web (see Figure \ref{fig:urweb}) and also, presently, Reason, for expressions of the type  \li{html} in Figure \ref{fig:html-type-def}.
-
-\subsection{TLM Definition and Application}
-The definition of the \li{$html} TLM is given in Figure \ref{fig:html-tlm-def}. TLM names are prefixed by \li{$} 
- to distinguish them from variables.
-Every TLM definition also includes a \emph{type annotation}, here \li{at html}, and a \emph{parse function} between \li{by static \{} and \li{\}}.
-% The TLM definition follows the same scoping rules as type definitions.
-We will describe how TLM definitions are packaged into libraries in Sec \ref{sec:static-eval}.
-
-Using this TLM, we can express the Ur/Web example from Figure \ref{fig:urweb} as shown in Figure \ref{fig:first-tsm-example}. On both Lines 1 and 2, we apply 
-\li{$html} 
- to a \emph{generalized literal form} delimited by \li{[|} and \li{|]}.
-Generalized literal forms, which first arose in the prior work on TSLs \cite{TSLs}, syntactically subsume other literal forms because the context-free syntax of the language only specifies the outer delimiters.
-In this paper, we will use  \li{[|} and \li{|]}, but \citet{TSLs} formally specified several other choices, including layout-sensitive delimitation.
-\emph{Literal bodies} are constrained only in that \li{[|} and \li{|]} must be balanced.
-%In particular, there is no pre-specified syntactic to splice out sub-terms.%The initial context-free parser therefore parses generalized literal forms in much the same way as parsers typically parse ``raw'' string literals (i.e.
-string literals where escape sequences like \li{\\n} have not yet been processed.)
-
- The system delegates responsibility over the parsing and expansion of each literal body to the applied TLM's parse function during a semantic phase called \emph{expansion}, which starts just before and continues into the typing phase.
-
- Because the parse function is applied during this phase, rather than at run-time, we call it a \emph{static function}.
-Static functions  cannot refer to the surrounding variable bindings because those variables stand for run-time values.
-Instead, there is are separate \emph{static bindings} marked by the \li{static} keyword that populate a \emph{static environment} that is discarded after expansion finishes.
-A more detailed account of static evaluation, both informal and informal, is available in the supplement.
-An alternative design that allows for the explicit lowering of standard-phase modules to the static phase has also been proposed for OCaml \cite{Ocaml/macros}.
-%\ificfp We are closely following this work in our implementation.
-\else \fi
-  % return to in Sec. \ref{sec:static-eval}.
-
- The input type of the parse function, \li{body}, classifies encodings of literal {bodies}.
-Literal bodies are sequences of characters, so we define \li{body} as a synonym of \li{string} in \autoref{fig:indexrange-and-parseresult}.
-The return type is a sum type, defined by applying the parameterized type \li{parse_result} defined in \autoref{fig:indexrange-and-parseresult}, that distinguishes between parse errors and successful parses.
-Let us consider these two possibilities in turn.
-
-% \vspace{-3px}\paragraph{Parse Errors} 
-If the parse function determines that the literal body is not well-formed according to the syntax that it implements, it must return \li{ParseError \{msg=}$e_\text{msg}$\li{, loc=}$e_\text{loc}$\li{\}}
-where $e_\text{msg}$ is a custom error message and $e_\text{loc}$ is a value of type \li{segment}, defined in Figure \ref{fig:indexrange-and-parseresult}, that designates a segment of the literal body as the origin of the error \cite{DBLP:journals/jsc/DeursenKT93}.
-
-% \vspace{-3px}\paragraph{Success} 
-If instead parsing succeeds, the parse function returns \li{Success} ~$\ecand$, 
-where $\ecand$ is called the \emph{encoding of the proto-expansion}.
-For expression TLMs, the proto-expansion is a \emph{proto-expression} and it is encoded as a value of the recursive datatype \li{proto_expr} that is outlined in Figure \ref{fig:candidate-exp-Reason}.
-Most of the constructors of \li{proto_expr} are individually uninteresting -- they encode OCaml's various expression forms. 
-Expressions can mention types, so we also need the type \li{proto_typ} also outlined in Figure \ref{fig:candidate-exp-Reason}.
-It is only the \li{SplicedE} and \li{SplicedT} constructors that are novel.
-These are discussed next.
-
-% \section{Simple Expression TLMs By Example}\label{sec:tsms-by-example}
-% We begin in this section with a ``tutorial-style'' introduction to seTLMs in Reason.
-%In particular, we will define an seTLM for constructing values of the recursive labeled sum type \li{rx} that was defined in Figure \ref{fig:datatype-rx}.
-% Sec.
-\ref{sec:tsms-minimal-formalism} then formally defines a reduced dialect of Reason called $\miniVerseUE$.
-This will serve as a ``conceptually minimal'' core calculus of TLMs, in the style of the simply typed lambda calculus.
-%We conclude in Sec.
-\ref{sec:uetsms-discussion}
-
-% \subsection{TLM Application}\label{sec:uetsms-usage}
-% The following Reason expression, drawn textually, is of \emph{TLM application} form. Here, a TLM named \li{$rx} is applied to the \emph{generalized literal form} \li{/SURLA|T|G|CEURL/}:
-% \begin{lstlisting}[numbers=none,mathescape=|]
-% $rx /SURLA|T|G|CEURL/
-% \end{lstlisting}
-% Generalized literal forms are left unparsed according to the context-free syntax of Reason.
-Several other outer delimiters are also available, as summarized in Figure \ref{fig:literal-forms}.
-The client is free to choose any of these for use with any TLM, as long as the \emph{literal body} (shown in green above) satisfies the requirements stated in Figure \ref{fig:literal-forms}.
-For example, we could have equivalently written the example above as \li{$rx `SURLA|T|G|CEURL`}.
-(In fact, this would have been convenient if we had wanted to express a regex containing forward slashes but not backticks.)
-
-% It is only during the subsequent \emph{typed expansion} phase that the applied TLM parses the {body} of the literal form to generate a \emph{proto-expansion}.
-The language then \emph{validates} this proto-expansion according to criteria that we will describe in Sec.
-\ref{sec:uetsms-validation}.
-If proto-expansion validation succeeds, the language generates the \emph{final expansion} (or more concisely, simply the \emph{expansion}) of the TLM application.
-The behavior of the program is determined by its expansion.
-
-% For example, the expansion of the TLM application above is equivalent to the following expression when the regex value constructors \li{Or} and \li{Str} are in scope:
-% \begin{lstlisting}[numbers=none]
-% Or(Str "SSTRAESTR", Or(Str "SSTRTESTR", Or(Str "SSTRGESTR", Str "SSTRCESTR")))
-% \end{lstlisting}
-% To avoid the assumption that the variables \li{Or} and \li{Str} are in scope at the TLM application site, the expansion actually uses the explicit \li{fold} and \li{inj} operators, as described in Sec.
-\ref{sec:lists}.
-In fact, the proto-expansion validation process enforces this notion of context independence -- we will return to proto-expansion validation below.
-(We will show how TLM parameters can reduce the awkwardness of this requirement in Chapter \ref{chap:ptsms}.)
-% %The constructors above are those of the type \li{Rx} that was defined in Figure \ref{fig:datatype-rx}.
-
-% % A number of literal forms, ,  are available in Reason's concrete syntax.
-Any literal form can be used with any TLM,  TLMs have access only to the literal bodies.
-Because TLMs do not extend the concrete syntax of the language directly, there cannot be syntactic conflicts between TLMs.
-
-%  %The form does not directly determine the expansion.
-
-% \begin{figure}
-% \begin{lstlisting}
-% 'SURLbody cannot contain an apostropheEURL'
-% `SURLbody cannot contain a backtickEURL`
-% [SURLbody cannot contain unmatched square bracketsEURL]
-% {|SURLbody cannot contain unmatched barred curly bracesEURL|}
-% /SURLbody cannot contain a forward slashEURL/
-% \SURLbody cannot contain a backslashEURL\
-% \end{lstlisting}
-% %SURL<tag>body includes enclosing tags</tag>EURL
-% \caption[Available Generalized Literal Forms]{Generalized literal forms available for use in Reason's textual syntax.
-The characters in green indicate the literal bodies and describe how the literal body is constrained by the form shown on that line.
-The Wyvern language defines additional forms, including whitespace-delimited forms \cite{TSLs} and multipart forms \cite{sac15}, but for simplicity we leave these out of Reason.}
-% \label{fig:literal-forms}
-% \end{figure}
-
-\subsection{Splicing}\label{sec:splicing-and-hygiene}
-When the parse function determines that some segment of the literal body is a spliced expression, according to whatever syntactic criteria it deems suitable, it can indirectly refer to it in the encoding it produces using the \li{SplicedE} constructor of \li{proto_expr}, which takes a value of type \li{segment} that indicates the zero-indexed location of the spliced expression relative to the start of the provided literal body.
-The \li{SplicedE} constructor also requires a value of type \li{proto_typ}, which indicates the type that the spliced expression is expected to have.
-Types can be spliced out by using the \li{SplicedT} constructor of \li{proto_typ} analagously.
-
-\begin{figure}
-\begin{lstlisting}[numbers=none,xleftmargin=0pt]
-type body = string;
-type segment = {startIdx: int, endIdx: int};
-type parse_result('a) 
-  = ParseError {msg: string, loc: segment}
-  | Success('a);
-type proto_typ = Arrow(proto_typ, proto_typ)
-               | StringTy
-               | /* ... */ 
-               | SplicedT(segment);
-type proto_expr = Tuple(list(proto_expr))
-                | /* ... */
-                | SplicedE(segment, proto_typ);
-\end{lstlisting}
-\caption[Definitions of various types used by TLM definitions.]{Definitions of various types available ambiently to TLM definitions.}
-\label{fig:indexrange-and-parseresult}
-\label{fig:candidate-exp-Reason}
-\end{figure}
-
-For example, consider again the two TLM applications in Figure \ref{fig:first-tsm-example}.
-In each case, the parse function of the \li{$html} TLM (Figure \ref{fig:html-tlm-def}, Lines 2-4) first sends the literal body through an off-the-shelf HTML parser, \li{parse_html}.
-It then passes the result to a function \li{html_to_ast}, not shown, which produces the corresponding expression encoding of type \li{proto_expr}.
-When this function encounters an HTML text node containing matched \li{\{[} and \li{]\}}, then that segment is inserted as a spliced expression of type \li{string}, and similarly text nodes containing matched curly braces produce a spliced expression of type \li{html}.
-For instance,  the proto-expansion generated for first TLM application in \ref{fig:first-tsm-example}, pretty-printed, is:
-\begin{lstlisting}[numbers=none]
-    H1Element(Nil, Cons(TextNode "Hello, ", Cons(
-      TextNode spliced<13; 22; string>, Nil)))
-\end{lstlisting}
-Here, \li{spliced<13; 22; string>} is a reference to the spliced string expression \li{first_name} by its location relative to the start of the literal body being expanded (the off-the-shelf HTML parser provides the necessary baseline location information for use by \li{html_to_ast}).
-It corresponds to the encoding \li{SplicedE(\{startIdx=13, endIdx=22\}, StringTy)}.
-Requiring that TLMs refer to spliced expressions indirectly in this manner ensures that a TLM cannot ``forge'' spliced terms, i.e.
-claim that some sub-term of the expansion should be given the privileges of a spliced term, discussed in Sec.
-\ref{sec:uetsms-validation}, when it does not in fact appear in the literal body.% Our semantics distinguishes spliced expressions when validating the proto-expansion, which is important for reasons of hygiene that we will detail shortly.
-
-% The parse function can similarly extract \emph{spliced types} from a literal body using the \li{SplicedT} variant of \li{proto_typ}.
-%In particular, the parse function must provide the index range of spliced subexpressions to the \li{Spliced} constructor of the type \li{MarkedExp}.
-%Only subexpressions that actually appear in the body of the literal form can be marked as spliced subexpressions.
-
-%For example, had the  would not be a valid expansion, because the  that are not inside spliced subexpressions:
-%\begin{lstlisting}[numbers=none]
-%Q.Seq(Q.Str(name), Q.Seq(Q.Str ": ", ssn))
-%\end{lstlisting}
-
-% \subsection{Splice Summaries and Segmentations}
-The \emph{segmentation} inferred from a proto-expansion is the finite set of references to spliced terms contained within.
-For example, the segmentation inferred from the proto-expression above contains only \li{spliced<13; 22; string>}.
-% Notice that no information about  the spliced terms appear is communicated by the splice summary.
-The system checks that all of the locations in the segmentation are 1) in bounds relative to the literal body; and 2) non-overlapping. 
-This resolves the problem of \textbf{Segmentation} described in Sec.
-\ref{sec:intro}, i.e.
-every literal body in a well-formed program has a well-defined segmentation.
-The TSL mechanism  did not maintain this reasoning principle \cite{TSLs}.
-A program editor or pretty-printer can communicate this segmentation information to the programmer, e.g.
-by coloring non-spliced segments green as is our convention in this document.
-In general, spliced expressions might themselves apply TLMs, in which case the convention is to use a distinct color for unspliced segments at each depth.
-For example, consider a TLM \li{$smiles} for chemical structures \cite{anderson1987smiles} with support for splicing using curly braces:
-\begin{lstlisting}[numbers=none]
-    $html [|SURLChemical structure of sucrose: {EURL
-        $smiles [|SCSS{ECSSm_glucoseSCSS}-O-{ECSSm_fructoseSCSS}ECSS|] 
-        |> SMILES.to_svg SURL}EURL|]
-\end{lstlisting}
-%For example, if strings were not primitive but rather defined as sequences of characters, we might define a TLM \li{$str} to recover string literal notation and write \li{$html [|SURL<body>\{EURLheading $str [|SSTRWorld!ESTR|]SURL\} ...</body>EURL|]}.
-
-A program editor or pretty-printer can communicate the type of each spliced expression, also specified abstractly by the segmentation, upon request (for Reason, via Merlin \cite{Merlin}.) %When there is a type mismatch, the type annotation on the spliced segment also allows the error message to report the problem without revealing the full expansion.
-
-\subsection{Proto-Expansion Validation}\label{sec:uetsms-validation}
-Three important concerns described in Sec.
-\ref{sec:intro} remain: those related to reasoning abstractly about the \emph{hygiene properties}, i.e.
-\textbf{Capture} and \textbf{Context Dependence}, and \textbf{Typing}.
-Addressing these concerns is the purpose of the \emph{proto-expansion validation} process, which occurs during the typing phase.
-Proto-expansion validation results in the \emph{final expansion}, which is simply the proto-expansion with the references to spliced segments replaced with their own final expansions.
-
-\subsubsection{Capture}\label{sec:capture}
-
-Proto-expansion validation ensures that spliced terms have access \emph{only} to the bindings at the application site---spliced terms cannot capture bindings internal to the proto-expansion.
-For example, consider the following application site:
-\begin{lstlisting}[numbers=none]
-    let tmp = /* ... application site temp ... */;
-    $html [|SURL<h1>{EURLf(tmp)SURL}</h1>EURL|];
-\end{lstlisting}
-Now consider the scenario where the proto-expansion generated by \li{$html} has the following form:
-\begin{lstlisting}[numbers=none]
-    let tmp = /* ... expansion-internal temp ... */;
-    H1Element(tmp, spliced<5; 10; html>);
-\end{lstlisting}
-Na\"ively, the binding of the variable \li{tmp} in the proto-expansion could shadow the application-site binding of \li{tmp} in the final expansion. 
-To address this problem, splicing is guaranteed to be capture-avoiding.
-When generating the final expansion, the system discharges the requirement that capture not occur by implicitly alpha-varying the bindings in the proto-expansion as needed.
-There is no need for TLM providers to explicitly deploy a mechanism that generates fresh variables (as in, e.g., prior \emph{reader macro} systems \cite{Flatt:2012:CLR:2063176.2063195}.
-% For example, the final expansion of the example above might take the following form:
-%\begin{lstlisting}[numbers=none]
-%  let tmp = /* ... application site temporary ... */; 
-%  let tmp' = /* ... expansion-internal temporary ... */;
-%  H1Element(tmp', f(tmp));
-%\end{lstlisting}
-%Notice that the expansion-internal binding of \li{tmp} has been alpha-varied to \li{tmp'}.
-The reference to \li{tmp} in the spliced expression then refers, as intended, to the application site binding.
-
-% For TLM providers, the benefit of this mechanism is that they can name the variables used internally within expansions freely. TLM clients can, in turn, can reason abstractly about \textbf{Capture}.
-
-Capture avoidance does prevent library providers from intentionally introducing bindings into spliced terms.
-For example, Haskell's literal notation for monadic values, i.e.
-\li{do}-notation, cannot be expressed because it introduces a new binding syntax \cite{jones2003haskell}.
-
-In our initial designs, we were able to express this example because we distinguished spliced identifiers, much as we do spliced types and expressions.
-These could be explicitly bound inside spliced expressions by extending the \li{SplicedE} and \li{SplicedT} constructors to take finite sets of spliced identifiers.
-This is conceptually similar to the approach taken by \citet{DBLP:conf/esop/HermanW08} for giving (already parsed) sub-terms access to internal variables.
-However, we ultimately decided to remove this feature because it would disproportionately increase the reasoning burden on clients when they encounter an unfamiliar literal: \emph{might this obscure literal also be introducing bindings?} It is difficult for a program editor to display a set of ``hidden bindings'' to the programmer.
-Haskell-style infix notation for monadic values, where bind is \li{e >>= f}, can be expressed using TLMs without support for spliced identifiers.
-Reason 3.0 is adopting more concise function syntax, which will decrease the cost of this syntax.
-%We will show an alternative formulation of Haskell's syntax for monadic commands that uses Reason's anonymous function syntax to bind variables in Sec. \ref{sec:application-monadic-commands}.
-
-\subsubsection{Context Dependence}\label{sec:context-dependence}
-%The prohibition on shadowing ensures only that variables that appear in spliced terms do not refer to bindings that appear in the surrounding expansion. 
-The proto-expansion validation process also ensures that variables that appear in the proto-expansion do not refer to bindings that appear either at the TLM definition or the application site.
-In other words, expansions must be completely \emph{context independent} -- they can make no assumptions about the surrounding context whatsoever.
-
-A minimal example of a ``broken'' TLM that does not generate context-independent proto-expansions is below:
-\begin{lstlisting}[numbers=none]
-    syntax $broken at t by static { 
-      fun(_) => Success (Var "SSTRxESTR") };
-\end{lstlisting}
-The proto-expansion that this TLM generates (for any literal body) refers to a variable \li{x} that it does not itself bind.
-If proto-expansion validation permitted such a proto-expansion, it would be well-typed only under those application site typing contexts where \li{x} is bound.
-This ``hidden assumption'' makes reasoning about binding and renaming difficult.
-
-Of course, this prohibition does not extend into the spliced terms in a proto-expansion -- spliced terms appear at the application site, so they can justifiably refer to application site bindings.
-(like \li{first_name} in Fig.
-\ref{fig:first-tsm-example}.) Because proto-expansions refer to spliced terms indirectly, enforcing context independence is straightforward -- we need only that the proto-expansion itself be closed.% In the next section, we will formalize this intuition.
-% The TLM provider can only refer to them opaquely.
-
-Na\"ively, this restriction, also present in the prior work on TSLs \cite{TSLs}, is quite restrictive -- expansions cannot access any library functions.
-At best, they can require the client to ``pass in'' required library functions via splicing at every application.
-In Sec.
-\ref{sec:ptsms}, we will introduce module parameters and partial parameter application to neatly resolve this problem.
-
-% This prohibition on context dependence explains why the expansion generated by the TLM application in Sec.
-\ref{sec:uetsms-usage} cannot make use of the regex value constructors, e.g.
-\li{Str} and \li{Or}, directly.
-(In Chapter \ref{chap:ptsms}, we will relax this restriction to allow proto-expansions to access explicit parameters.)
-
-% Collectively, we refer to the prohibition on capture and the prohibition on context dependence as \emph{hygiene properties}, by conceptual analogy to corresponding properties in term-rewriting macro systems (see Sec.
-\ref{sec:macro-systems}.) The novelty here comes from the fact that spliced terms are being extracted from an initially unparsed sequence of characters.
-% In the examples in Sec.
-\ref{sec:uetsms-usage} and Sec.
-\ref{sec:splicing-and-hygiene}, the expansion used constructors associated with the \li{Rx} type, e.g.
-\li{Seq} and \li{Str}.
-This might appear to violate our prohibition on context-dependent expansions.
-This is not the case only because in Reason, constructor labels are not variables or scoped symbols.
-Syntactically, they must begin with a capital letter (like Haskell's datatype constructors).
-Different labeled sum types can use common constructor labels without conflict because the type the term is being checked against -- e.g.
-\li{Rx}, due to the type ascription on \li{$rx} -- determines which type of value will be constructed.
-For dialects of ML where datatype definitions do introduce new variables or scoped symbols, we need parameterized TLMs.
-We will return to this topic in Chapter \ref{chap:ptsms}.
-% Indeed, we used the label \li{Spliced} for two different recursive labeled sum types in Figure \ref{fig:candidate-exp-Reason}.
-
-\subsubsection{Typing}\label{sec:typing-e}
-Finally, proto-expansion validation maintains a reasonable {typing discipline} by (1) checking that the expansion is of the type specified by the TLM's type annotation; (2) checking that each spliced type is valid; (3) checking that the type annotation on each spliced expression is valid; and (4) checking each spliced expression against the specified type annotation.
-Context independence implies that ML-style type inference can be performed using only the segmentation (because the remainder of an expansion cannot mention the very variables whose types are being inferred).
-In the prior work on TSLs, spliced terms did not have type annotations
-%The OCaml type system is not strong enough to allow us to express only contextually well-typed syntax trees. In The details are discussed as future work in Sec. \ref{sec:discussion}.
- % This addresses the problem of reasoning abstractly about \textbf{Typing} described in Sec. \ref{sec:intro}, i.e.:
- % \begin{enumerate}
- %   \item determining the type of an expansion requires examining only the type annotation on the TLM definition (much as determining the type of a function application requires examining only the function's type); and
- % \item determining the type that a spliced expression must have requires only the information in the splice summary (rather than complete knowledge of the proto-expansion).
- % \end{enumerate}
-
-% The language \emph{validates} proto-expansions before a final expansion is generated.
-One aspect of proto-expansion validation is checking  the proto-expansion against the type annotation specified by the TLM, e.g.
-the type \li{Rx} in the example above.
-This maintains a \emph{type discipline}: if a programmer sees a TLM being applied when examining a well-typed program, they need only look up the TLM's type annotation to determine the type of the generated expansion.
-Determining the type does not require examining the expansion directly.
-
-% \subsection{Hygiene}
-% The spliced subexpressions that the proto-expansion refers to (by their position within the literal body, cf.
-above) must be parsed, typed and expanded during the proto-expansion validation process (otherwise, the language would not be able to check the type of the proto-expansion).
-To maintain a useful \emph{binding discipline}, i.e.
-to allow programmers to reason also about variable binding without examining expansions directly, the validation process maintains two additional properties related to spliced subexpressions: \textbf{context independent expansion} and \textbf{expansion independent splicing}.
-These are collectively referred to as the \emph{hygiene properties} (because they are conceptually related to the concept of hygiene in term rewriting macro systems, cf.
-Sec.
-\ref{sec:term-rewriting}.)
-
-% \paragraph{Context Independent Expansion}
-
-% \paragraph{Expansion Independent Splicing} 
-% %These properties suffice to ensure that programmers and tools can freely rename a variable without changing the meaning of the program.
-The only information that is necessary to perform such a \emph{rename refactoring} is the locations of spliced subexpressions within all the literal forms for which the variable being renamed is in scope; the expansions need not otherwise be examined.
-It would be straightforward to develop a tool and/or editor plugin to indicate the locations of spliced subexpressions to the user, like we do in this document (by coloring spliced subexpressions black).
-We discuss tool support as future work in Sec.
-\ref{sec:interaction-with-tools}.
-
-% \subsubsection{Final Expansion}
-
-% For example, the final expansion of the body of \li{lookup_rx} is equivalent to the following, under an environment where the regex value constructors are available:
-% \begin{lstlisting}[numbers=none]
-% Seq(Str(name), Seq(Str "SSTR: ESTR", ssn))
-% \end{lstlisting}
-% (Again, due to the prohibition on context dependent expansions, the final expansion actually involves explicit \li{fold} and \li{inj} operators.)
-
-\newcommand{\spTLMsSec}{Pattern TLMs}
-\section{\protect\spTLMsSec}
-\label{sec:sptsms}
-
-Let us now briefly consider the topic of TLMs that generate \emph{patterns}, rather than expressions.
-Pattern literals are the dual to expression literals in that expression literals support construction whereas pattern literals support deconstruction \cite{moriconi2008inversion}.
-For example, we can pattern match on a value \li{x : html} by applying a pattern TLM \li{$html} as follows:
-\begin{lstlisting}[numbers=none]
-    switch x {
-    | $html [|SURL<h1>{EURLcsSURL}</h1>EURL|] => /* cs:list(html) */
-    | _ -> None
-    };
-\end{lstlisting}
-Any list pattern, including one generated by another TLM application, can appear where \li{cs} appears in the example pattern above.
-For longer \li{switch} expressions, the shorthand \li{switch x using $html} applies \li{$html} to every rule where the outermost pattern is of generalized literal form.
-
-Notice that we can use the same name, \li{$html}, for this pattern TLM as for the expression TLM defined in the previous section.
-It does not make sense to apply an expression TLM in pattern position  (many expression-level constructs, e.g.
-lambdas, do not correspond even syntactically to patterns), and \emph{vice versa}, so this is unambiguous.
-
-Pattern TLM definitions differ from expression TLM definitions in two ways: (1) a \emph{sort qualifier}, \li{for patterns}, distinguishes them from expression TLM definitions; (2) the return type of the parse function is \li{parse_result(proto_pat)}, rather than \li{parse_result(proto_expr)}.
-The type \li{proto_pat}, outlined below, classifies encodings of \emph{proto-patterns}.
-\begin{lstlisting}[numbers=none]
-   type proto_pat = /* no variable pattern form! */
-                  | Wild
-                  | /* ... */
-                  | SplicedP(segment, proto_typ);
-\end{lstlisting}
-
-% \subsection{Splicing}
-The constructor \li{SplicedP} operates much like \li{SplicedE} to allow a proto-pattern to refer indirectly to spliced patterns.% by their location within the literal body.
-
-%  For example, the proto-pattern generated for the example at the top of this section would be written concretely as follows:
-% \begin{lstlisting}[numbers=none]
-%   H1Element (_, spliced<6; 7; list(html)>)
-% \end{lstlisting}
-
-% \subsection{Proto-Pattern Validation}
-% Proto-pattern validation serves, like proto-expression validation, to maintain the ability to reason abstractly about binding and typing.
-
-To maintain the abstract binding discipline, variable patterns can appear only within spliced patterns.
-Enforcing this restriction is straightforward: we simply have not defined a variant of the \li{proto_pat} type that encodes variable patterns (wildcards are allowed.)  This restriction ensures that only variables visible to the client in a spliced pattern are bound in the corresponding branch expression.
-This is analagous to the capture avoidance principle for expression TLMs.
-
-Type annotations on references to spliced patterns could refer to type variables, so we also need to enforce context independence in the manner discussed in the previous section.
-%Pattern guards are part of pattern matching rules, not patterns themselves, in Reason/OCaml, but were they part of patterns, context independence would also need to be enforced for guard expressions.% In languages like OCaml, which support arbitrary boolean guard expressions, we would also need to enforce both context independence and capture avoidance as discussed in the previous section for spliced expressions that end up in a guard expression.
-In our formalism, we do not support guard expressions.
-
-% \subsubsection{Typing} 
-To maintain an abstract typing discipline, proto-pattern validation checks type annotations much as in Sec.
-\ref{sec:typing-e}.% (1) that the final expansion is a pattern that matches values of the type specified by the TLM's type annotation; (2) that each spliced pattern matches values of the type indicated in the segmentation; and (3) that each of these types are themselves well-formed types.
-
-\newcommand{\pTLMsSec}{Parametric TLMs}
-\section{\protect\pTLMsSec}
-\label{sec:ptsms}
-
-The simple TLMs in the previous sections operate only at one specified type, as did the TSLs in the prior work \cite{TSLs}.
-This is rather limiting.
-This section introduces \emph{parametric TLMs}, which can operate over a type- and module-parameterized family of types.
-They also neatly solve the problem discussed in Sec.
-\ref{sec:context-dependence} of giving expansions access to helper functions.
-%Moreover, as discussed at the end of Sec.
-\ref{sec:context-dependence}, the expansions that they generate have no access to libraries.
-This section introduces \emph{parametric TLMs} to neatly resolve both of these limitations (which also limited the prior work on TSL \cite{TSLs}.) Parametric TLMs can be defined over a type- and module-parameterized family of types, and the proto-expansions they generate can refer to supplied type and module parameters.
-Partial parameter application decreases the syntactic cost of this explicit parameter passing style.
-% \subsection{Parametric TLMs By Example}\label{sec:ptsms-by-example}
-
-Consider the following Reason/OCaml module type (a.k.a. signature), which specifies an abstract data type \cite{liskov1974programming,harper1997programming} of string-keyed polymorphic dictionaries:
-\begin{lstlisting}[numbers=none]
-    module type DICT = {
-      type t('a); 
-      let empty : t('a); 
-      let extend : t('a) -> (string, 'a) -> t('a); 
-      /* ... */ 
-    };
-\end{lstlisting}
-We can define a TLM that is parametric over implementations of this signature, \li{D : DICT}, and over choices of the codomain type, \li{'a} as follows:
-\begin{lstlisting}[numbers=none]
-    syntax $dict (D : DICT) (type 'a) at D.t('a) 
-    by static { fun(b) => /* ... */ };
-\end{lstlisting}
-For example, given some module that implements this signature, \li{HashDict : DICT}, we can apply \li{$dict} as follows:
-\begin{lstlisting}[numbers=none]
-    $dict HashDict int [|"key1"SURL=>EURL10SURL; EURL"key2"SURL=>EURL15|]
-\end{lstlisting}
-Notice that the segmentation immediately reveals which punctuation is particular to this TLM and where the spliced key and value expressions appear.
-Because the context-free syntax of unexpanded terms is never modified, it is possible to reason modularly about syntactic determinism, i.e.
-we can reason above that \li{=>} does not appear in the follow set of unexpanded expressions \cite{conf/pldi/SchwerdfegerW09}, so there can never be an ambiguity about where a key expression ends.
-
-The proto-expansion generated for the TLM application above, shown below, can refer to the parameters:
-\begin{lstlisting}[numbers=none]
-    D.extend(D.extend D.empty (spliced<1;6;string>, 
-      spliced<11;12;'a>)) (spliced<15;20;string>, 
-      spliced<25;26;'a>)
-\end{lstlisting}
-Validation checks that the proto-expansion is truly parametric, i.e.
-it must be valid for all modules \li{D : DICT} and types \li{'a}.
-It is only after validation that we substitute the actual parameters, here \li{HashDict} for \li{D} and \li{int} for \li{'a}, into the final expansion.
-Only the type annotations on references to spliced terms are subject to early parameter substitution (because they classify application site terms.)
-
-If we will use the \li{HashDict} implementation ubiquitously, we can abbreviate the partial application of \li{$dict} to \li{HashDict}, resulting in a TLM that is parametric over only the type \li{'a}:
-\begin{lstlisting}[numbers=none]
-    syntax $dict' = $dict HashDict; 
-\end{lstlisting}
-TLM abbreviations can themselves be parameterized to support partial application of parameters other than the last.
-
-In Sec.
-\ref{sec:context-dependence} we discussed the problem of the strict context independence discipline being too restrictive, in that it would seem to restrict expansions from referring to useful helpers bound at the TLM definition site.
-Module parameters address this problem -- the helper values, types and modules can be packaged into a module, passed in and partially applied to hide this detail from clients.
-Because this will be common in practice, we provide the following shorthand:
-\begin{lstlisting}[numbers=none]
-    syntax $a at t using X~${}_1$~=M~${}_1$~,...,X~${}_n$~=M~${}_n$~ by ...
-\end{lstlisting} 
-% This is the same as a definition that specifies $n$ type, expression or module parameters and immediately applies the given arguments, binding the result to the same TSM name.
-
-This explicit parameter passing discipline is reminiscent of work on explicit capture for distributed functions \cite{DBLP:conf/ecoop/MillerHO14}.
-By not implicitly giving expansions access to all definition-site bindings, we need not examine parse functions to reason about, e.g., renaming.
-Consequently, encodings of proto-terms can be values of standard datatypes (e.g.
-\li{proto_expr}) with variables represented simply as, e.g., strings, and quasiquotation notation can be expressing using TLMs (see supplement).
-A central design goal of Reason is to leave the OCaml semantics unchanged, so building quasiquotation in primitively, to integrate the free variables in term encodings into the overall binding discipline \cite{Bawd99a}, was in any case infeasible.
-
-\newcommand{\moreExamplesSec}{Additional Examples}
-\section{\protect\moreExamplesSec}
-\label{sec:more-examples}
-
-\subsection{HTML}
 

### 2018-03-06 icfp18/syntax-icfp18.tex
tweaks to intro

@@
 When designing the surface syntax of a general-purpose programming language, it is common practice to define \emph{literal notation} that decreases the syntactic cost of constructing and pattern matching over values of a select few data structures.
-For example, Standard ML supports list literals like \li{[x1, x2, x3]} in both expression and pattern position \cite{harper1997programming,mthm97-for-dart}.
-Many other languages have similar list literal notation.
+For example, many languages in the ML family support list literals like \li{[x1, x2, x3]} in both expression and pattern position \cite{harper1997programming,mthm97-for-dart}.
 %\footnote{In SML, list literals work even in contexts where the list expression constructors have been shadowed by other bindings, i.e. the meaning of literal forms is \emph{context independent}. 
 % We will return to this concept again later.} 
-Another example is Ur/Web, which extends the surface syntax of Ur (an ML-like language \cite{conf/pldi/Chlipala10}) with expression and pattern literals for encodings of XML and HTML data~\cite{conf/popl/Chlipala15}.
+While lists are common across problem domains, other literal notation is more specialized.
+For example, Ur/Web extends the surface syntax of Ur (an ML-like language \cite{conf/pldi/Chlipala10}) with expression and pattern literals for encodings of XML and HTML data~\cite{conf/popl/Chlipala15}.
 The example in Fig.
@@
 \begin{figure}[h]
-\vspace{-11px}
+\vspace{-12px}
 \begin{lstlisting}
@@
 \end{lstlisting}
-\vspace{-5px}
+\vspace{-4px}
 \caption{HTML literals with support for splicing at two different types are built primitively into Ur/Web \cite{conf/popl/Chlipala15}.}
@@
 
-This design practice, where the language designer privileges certain user-defined types with built-in literal notation, is decidedly \emph{ad hoc} in that it is easy to come up with other examples of data structures for which mathematicians, scientists or programmers have invented specialized notation \cite{DBLP:journals/cacm/Iverson80,cajori1928history,TSLs}.
-For example, 1) clients of a ``collections'' library might want not just list literals, but also literal notation for matrices, finite sets, maps, hash tables and so on, following usual mathematical conventions; 2) clients of a ``web programming'' library might like not just HTML literals but also CSS literals (which Ur/Web lacks); 3) a compiler author might like ``quotation'' literals for the terms of the object language and various intermediate languages of interest; and 4) clients of a ``chemistry'' library might like chemical structure literals based on the SMILES standard \cite{anderson1987smiles}.
+This design practice, where the language designer privileges certain library constructs with built-in literal notation, is decidedly \emph{ad hoc} in that it is easy to come up with other examples of data structures for which mathematicians, scientists or programmers have invented specialized notation \cite{DBLP:journals/cacm/Iverson80,cajori1928history,TSLs}.
+For example, 1) clients of a ``collections'' library might want not just list literals, but also literal notation for matrices, finite sets, maps and so on; 2) clients of a ``web programming'' library might like CSS literals (which Ur/Web lacks); 3) a compiler author might like ``quotation'' literals for the terms of the object language and various intermediate languages of interest; and 4) clients of a ``chemistry'' library might like chemical structure literals based on the SMILES standard \cite{anderson1987smiles}.
 
@@
 %, e.g. for URLs, paths, regular expressions and so on
-Literal notation, with support for splicing, would decrease the syntactic cost of composite encodings of this data, which are more amenable to programmatic manipulation and compositional reasoning.
+Literal notation, with support for splicing, would decrease the syntactic cost of composite encodings of this data, which is more amenable to programmatic manipulation and compositional reasoning.
 %Literal forms would  preferable to string encodings because they support compositional construction and decomposition by pattern matching via splicing.
@@
 Instead, there has been persistent interest in mechanisms that allow library providers to define new literal notation on their own.
-For example, direct grammar extension systems like Camlp4 \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, macro-based systems like Omar et al's \emph{type-specific languages} (TSLs) \cite{TSLs}, and other systems that we will discuss in Sec.~\ref{sec:existing-approaches} can all be used to define new literal notation (and, in some cases, other forms of new notation, such as new infix expression forms or type declaration forms, which are beyond the scope of this paper).
-
-\paragraph{Problem} The problem that specifically motivates this paper is that existing syntax extension systems  make it difficult or impossible to reason abstractly about such fundamental issues as types and variable binding when the program text is littered with  user-defined literal forms.
+For example, direct grammar extension systems like Camlp4 \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, macro-based systems like Omar et al's \emph{type-specific languages} (TSLs) \cite{TSLs}, and other systems that we will discuss in Sec.~\ref{sec:existing-approaches} can all be used to define new literal notation (and, in some cases, other forms of new notation, such as new infix expression forms or type declaration forms, which we leave beyond the scope of this paper).
+
+\paragraph{Problem} The problem that specifically motivates this paper is that these existing syntax extension systems  make it difficult or impossible to reason abstractly about such fundamental issues as types and variable binding.
 Instead, programmers and editor services can only reason transparently, i.e.
-by inspecting the underlying expansion or the implementation details of the extensions responsible for producing the expansion.% In some existing systems, even determining which syntax extension is responsible for a literal form is difficult.
+by inspecting the underlying expansion or the implementation details of the collection of extensions responsible for producing the expansion.% In some existing systems, even determining which syntax extension is responsible for a literal form is difficult.
 %, i.e. when developing large programs ``consisting of many small programs o(modules), possibly written by different people''
@@
 
-Consider, for example, the perspective of a programmer attempting to comprehend, i.e.
-reason about, the program text in Figure \ref{fig:K-dialect}, which is written in a dialect of OCaml's surface syntax called Reason \cite{reason-what} that has, hypothetically, been extended with some number of new literal forms by a grammar extension system -- Lines 1-5 show the grammar extension syntax from Camlp4 \cite{ocaml-manual}, but Sugar*/SugarHaskell is  similar \cite{erdweg2011sugarj,erdweg2013framework,erdweg2012layout}.
-The literal form on Line 9 then constructs an encoding of a query in the niche stack-based database query language K, using its intentionally terse notation~\cite{Whitney:2001:LOR:376284.375783}.
+Consider, for example, the perspective of a programmer attempting to comprehend the program text in Figure \ref{fig:K-dialect}, which is written in a dialect of OCaml's surface syntax called Reason \cite{reason-what} that has, hypothetically, been extended with some number of new literal forms by a grammar extension system -- Lines 1-5 show the grammar extension syntax from Camlp4 \cite{ocaml-manual}, but Sugar*/SugarHaskell is  similar \cite{erdweg2011sugarj,erdweg2013framework,erdweg2012layout}.
+The literal form on Line 9 constructs an encoding of a query in the niche stack-based database query language K, using its intentionally terse notation~\cite{Whitney:2001:LOR:376284.375783}.
 The problem is that a programmer looking at the program as presented, and unfamiliar with (i.e.
@@
 A language that supports TSLs delegates control over the parsing and expansion of certain \emph{generalized literal forms} to a parser associated with the type that the form is being checked against.
-So in a language with TSLs, our example requires a type annotation on \li{y}:
+So if we add TSLs to our language, our example then requires a type annotation on \li{y}:
 \begin{lstlisting}[numbers=none,basicstyle=\ttfamily\fontsize{9pt}{1em}\selectfont]
@@
 As such, there is no way to indicate to the programmer where base language expressions are located within a literal form.
-  \item[\LEFTcircle] \textbf{Capture}: The TSL mechanism as described enforces complete capture avoidance, though the formal specification given in the paper did not adequately enforce this constraint.
+  \item[\LEFTcircle] \textbf{Capture}: The TSL mechanism as described enforces complete capture avoidance (though the formal specification given in the paper did not adequately enforce this constraint).
   \item[\LEFTcircle] \textbf{Context Dependence}: The TSL mechanism as described requires that the expansion be completely closed, so it is trivially context independent.
@@
 We evaluate these claims with (1) a series of increasingly ML-like calculi equipped with proofs of these abstract reasoning principles; and (2) a number of non-trivial examples that appear throughout the paper and involve the language features mentioned above.
-In describing these examples, we also demonstrate that literal parsing logic can be defined using off-the-shelf parser generators, so the burden on notation providers is comparable to that of existing systems.
+In describing these examples, we also demonstrate that literal parsing logic can be defined using standard, unmodified parser generators, so the burden on notation providers is comparable to that of existing systems despite the stronger client-side reasoning principles.
 % From the client's perspective, which is the focus of our contributions, TLMs are uniquely reasonable.
-Unique to our approach in this paper is the focus, both formal and informal, on maintaining client-side abstract reasoning principles.
+
+%Unique to our approach in this paper is the focus, both formal and informal, on maintaining client-side abstract reasoning principles.
 
@@
 
-The most obvious point of departure between TSLs, described above, and TLMs, which are the topic of this paper, is that we separate out the mechanism for defining new literal notation from the type definition mechanism of the language.
+The most obvious point of departure between TSLs, described above, and TLMs is that we separate out the mechanism for defining new literal notation from the type definition mechanism of the language.
 TLMs are instead explicitly named and applied.
@@
 The client is free to rename \li{w}.
-The external dependencies of the generated expansion are explicit -- here, Lines 4-5 specify that the expansion requires the module \li{KQuery} -- and the system ensures that this dependency is properly bound even if the identifier \li{KQuery} has been shadowed at the application site.
+The external dependencies of the generated expansion are explicit -- here, Lines 4-5 specify that the expansion requires the module \li{KQuery} -- and the system ensures that this dependency is properly bound as specified even if the identifier \li{KQuery} has been shadowed at the application site.
 \item[\CIRCLE] \textbf{Typing}: The type annotation on the definition of \li{$kq} (Line 1) determines the type that the expansion is validated against, here \li{KQuery.t} (not shown).
@@
   let bisA = $regex `(SURLGC$(EURLany_baseSURL)GCEURL)`;
-  let gene_restriction_template(gene_sequence : string) => 
+  let gene_restriction_template(gene_sequence) => 
     $regex `(SURL$(EURLbisASURL)$(EURLany_baseSURL)*$$(EURLgene_sequenceSURL)$(EURLany_baseSURL)*$(EURLbisASURL)EURL)`
@@
 % \item 
-We begin in Sec.~\ref{sec:setsms} by introducing TLMs with a detailed case study. 
+We begin in Sec.~\ref{sec:setsms} by introducing expression TLMs with a more detailed case study. 
 % \item 
 Sec.~\ref{sec:sptsms} then introduces {pattern TLMs}, i.e.
-TLMs that generate patterns, and describes the special hygiene conditions in pattern position.
+TLMs that generate patterns, and describes the special reasoning conditions in pattern position.
 %The hygiene condition for pattern TLMs is interesting.
@@
 TLMs that can take type and module parameters, which allow us to define literal notation over a type- or module-parameterized family of types.
-Sec.~\ref{sec:ptsms} also reveals that the \li{expansions require} clause is syntactic sugar for parameterization followed immediately by partial parameter application.
-Having introduced the basic machinery, we continue in Sec.~\ref{sec:more-examples} with additional examples that demonstrate both the expressive power of this approach and alternative parser implementation strategies that notation providers can consider.
+This section also reveals that the \li{expansions require} clause is syntactic sugar for parameterization followed immediately by partial parameter application.
+Having introduced the basic machinery, we continue in Sec.~\ref{sec:more-examples} with additional examples that demonstrate both the expressive power of this approach and various parser implementation strategies that notation providers can consider.
 Sec.~\ref{sec:setsms-formally} presents a type-theoretic calculus of simple expression and pattern TLMs and formally establishes the reasoning principles implied above, in the formally simplest setting.
@@
 Sec.~\ref{sec:discussion} concludes with a summary of the contributions of this paper and a discussion of future work.
-Due to the space limitations, certain technical details and extensions to the calculi presented in this paper, as well as detailed proofs, are available in the anonymized supplemental material.
+Due to the space limitations, certain technical details, proofs and extensions to the calculi presented in this paper are available in the anonymized supplemental material.
 % Also available is an additional series of case studies: notation for an encoding of regular expressions as an abstract data type, implemented using TLMs for quasiquotation and grammar-based parser generators.

### 2018-03-06 icfp18/syntax-icfp18.tex
first 5 pages done...

@@
 numbers=left, numberstyle=\scriptsize\color{mygray}, language=ML,moredelim=[il][\sffamily]{?},showspaces=false,showstringspaces=false,xleftmargin=0pt, numbersep=-6pt, morekeywords=[1]{try,spliced,tyfam,opfam,let,fn,val,def,casetype,objtype,metadata,of,*,string,lexer,parser,datatype,new,toast,notation,module,switch,where,expansions,require,import,for,ana,syn,opcon,tycon,metasignature,metamodule,metasig,metamod,static,at,by,tycase,mod,macro,match,pattern,in,patterns,expressions,implicit,forall,exptsm,pattsm},deletekeywords={double,structure,of},classoffset=0, xleftmargin=0pt,
-aboveskip=4pt,belowskip=2pt,
+aboveskip=3pt,belowskip=2pt,
 moredelim=**[is][\color{red}]{SSTR}{ESTR},
@@
 the sequence of characters between the outer delimiters, here \li{`(} and \li{)`}.
-The literal body is constrained by the context-free syntax of the language only in that nested delimiters must be matched (like comments in OCaml).
+The literal body is constrained by the context-free grammar of the language only in that nested delimiters must be matched (like comments in OCaml).
 TSLs are closely related to the mechanism that we will propose, so let us analyze TSLs from the perspective of the five reasoning criteria just outlined.
-\begin{enumerate}[leftmargin=12px,nolistsep,noitemsep]
+\begin{enumerate}[leftmargin=12px]
   \item[\LEFTcircle] \textbf{Responsibility}: The type-directed dispatch mechanism allows the client to easily determine which parser is responsible for each literal form: the parser that was associated with the type when it was defined.
-Clients do not need to worry about different TSLs conflicting syntactically because the context-free syntax of the language remains fixed and composition is mediated by splicing.
+Clients do not need to worry about different TSLs conflicting syntactically because the context-free grammar of the language remains fixed and composition is mediated by splicing.
 The main limitation here is that it is impossible to define literal notation after a type has been defined, and it is also impossible to define multiple notations at a single type.
@@
 We evaluate these claims with (1) a series of increasingly ML-like calculi equipped with proofs of these abstract reasoning principles; and (2) a number of non-trivial examples that appear throughout the paper and involve the language features mentioned above.
-In describing these examples, we also demonstrate that literal parsing logic can be defined using standard, unmodified parser generators, so the burden on notation providers is comparable to that of existing systems despite the stronger client-side reasoning principles.
+In describing these examples, we also demonstrate that literal parsing logic can be defined using standard, unmodified parser generators, so the burden on notation providers is comparable to that of existing systems despite these stronger client-side reasoning principles.
 % From the client's perspective, which is the focus of our contributions, TLMs are uniquely reasonable.
@@
 
-The most obvious point of departure between TSLs, described above, and TLMs is that we separate out the mechanism for defining new literal notation from the type definition mechanism of the language.
+TLMs, like TSLs, operate on generalized literal forms.
+With TLMs, however, the mechanism for defining new literal notation is not tied up with the type definition mechanism of the language.
 TLMs are instead explicitly named and applied.
@@
 Due to this explicit dispatch mechanism, there can be any number of TLMs defined at a given type.
-As with TSLs, the context-free syntax of the language is fixed, so the TLM provider can modularly establish that the notation that the TLM implements is unambiguous (even in situations where a spliced term itself applies another TLM, which we will show in a later example).
+As with TSLs, the context-free grammar of the language is fixed, so the TLM provider can modularly establish that the notation that the TLM implements is unambiguous (even in situations where a spliced term itself applies another TLM, which we will show in a later example).
 \item[\CIRCLE] \textbf{Segmentation}: The intermediate output that the TLM generates is structured so that the system can infer from it an accurate \emph{segmentation} of the literal body that distinguishes spliced terms from segments that are parsed in some other way by the TLM.
@@
 It cannot capture a coincidental binding of \li{x} in the expansion.
+We will say more about capture later on.
 \item[\CIRCLE] \textbf{Context Dependence}: The system enforces context independence, so the expansion of Line 3 cannot  rely on the fact that, e.g., \li{w} is in scope.
 The client is free to rename \li{w}.
-The external dependencies of the generated expansion are explicit -- here, Lines 4-5 specify that the expansion requires the module \li{KQuery} -- and the system ensures that this dependency is properly bound as specified even if the identifier \li{KQuery} has been shadowed at the application site.
+The external dependencies of the generated expansion are explicit -- here, Lines 4-5 specify that expansions generated by \li{$kq} require the module \li{KQuery} -- and the system ensures that this dependency is  bound as specified even if the identifier \li{KQuery} has been shadowed at the application site.
 \item[\CIRCLE] \textbf{Typing}: The type annotation on the definition of \li{$kq} (Line 1) determines the type that the expansion is validated against, here \li{KQuery.t} (not shown).
@@
 \caption{The definition of the \li{$regex} TLM. Figure \ref{fig:candidate-exp-Reason} defines \li{RegexLexer} and \li{RegexParser}.}
-\label{fig:html-tlm-def}
+\label{fig:regex-tlm-def}
 \end{subfigure}
@@
 \end{lstlisting}
-\vspace{-2px}
+\vspace{-5px}
 \caption{Examples of the \li{$regex} TLM being applied in a bioinformatics application.
@@
 }
-\label{fig:first-tsm-example}
+\label{fig:first-tlm-example}
 \end{subfigure}
-\vspace{5px}
+\vspace{3px}
 \caption{Case Study: POSIX-style regex literal notation, with support for string and regex splicing.}
+\vspace{-3px}
 \end{figure}
@@
 % \item 
-We begin in Sec.~\ref{sec:setsms} by introducing expression TLMs with a more detailed case study. 
+Sec.~\ref{sec:setsms} introduces expression TLMs with a detailed case study. 
 % \item 
@@
 % \item 
-Sec.~\ref{sec:ptsms}  introduces \emph{parametric TLMs} (pTLMs), i.e.
+Sec.~\ref{sec:ptsms} then introduces the more general \emph{parametric TLMs}, i.e.
 TLMs that can take type and module parameters, which allow us to define literal notation over a type- or module-parameterized family of types.
-This section also reveals that the \li{expansions require} clause is syntactic sugar for parameterization followed immediately by partial parameter application.
-Having introduced the basic machinery, we continue in Sec.~\ref{sec:more-examples} with additional examples that demonstrate both the expressive power of this approach and various parser implementation strategies that notation providers can consider.
-Sec.~\ref{sec:setsms-formally} presents a type-theoretic calculus of simple expression and pattern TLMs and formally establishes the reasoning principles implied above, in the formally simplest setting.
+This section also reveals that the \li{expansions require} clause is but syntactic sugar for parameterization followed immediately by partial parameter application.
+Having introduced the basic machinery, we continue in Sec.~\ref{sec:more-examples} with examples that further demonstrate the expressive power of this approach and various parser implementation strategies that notation providers can consider.
+Sec.~\ref{sec:setsms-formally} presents a type-theoretic calculus of simple expression and pattern TLMs and formally establishes the reasoning principles implied above, in the simplest setting. 
 Sec.~\ref{sec:ptlms-formally} adds type functions and an ML-style module system to this calculus, and gives the most general variant of the reasoning principles theorem. 
@@
 Sec.~\ref{sec:discussion} concludes with a summary of the contributions of this paper and a discussion of future work.
-Due to the space limitations, certain technical details, proofs and extensions to the calculi presented in this paper are available in the anonymized supplemental material.
+Certain technical details, proofs and extensions to the calculi presented in this paper are given in the supplement.
 % Also available is an additional series of case studies: notation for an encoding of regular expressions as an abstract data type, implemented using TLMs for quasiquotation and grammar-based parser generators.
@@
 
+\vspace{-3px}
 \newcommand{\seTLMsSec}{Expression TLMs}
@@
 \label{sec:setsms}
-
-We start in this section by describing the TLM mechanism in detail by way of a substantial case study: literal notation, like that built in to Ur/Web (see Figure \ref{fig:urweb}) and also, presently, Reason, for expressions of the type  \li{html} in Figure \ref{fig:html-type-def}.
-
+\vspace{-2px}
+
+Consider the recursive datatype \li{Regex.t} defined by the module in Fig.~\ref{fig:Regex-module-def}, which encodes \emph{regular expressions} (regexes) into Reason \cite{Thompson:1968:PTR:363347.363387}.
+Regular expressions are  common in fields like bioinformatics, where they are used to express patterns in DNA sequences.
+For example, we can construct a regex that matches the strings \li{"A"}, \li{"T"}, \li{"G"} or \li{"C"}, which represent the four bases in DNA, as follows:
+\begin{lstlisting}[numbers=none]
+  let any_base = Regex.(Or(Str "A", Or(Str "T", Or(Str "G", Or(Str "C")))))
+\end{lstlisting}
+In Reason, the notation \li{Regex.(}$e$\li{)} locally opens the module \li{Regex} for use within the expression $e$, so we do not need to qualify the constructors \li{Regex.Or} and \li{Regex.Str}.
+Even allowing for this shorthand, however, this notation for constructing regexes is rather unwieldy.
+Instead, we would like to be able to use the common POSIX-style notation for constructing values of type \li{Regex.t} \cite{STD95954}.
+
+% One approach would be to simply define a function \li{parse_rx : string -> Regex.t} that parses a string at run-time to produce a regex.
+This approach works for simple expressions like the one above -- we can write \li{parse_rx "A|T|G|C"}.
+However, this approach breaks down when one needs to construct regexes compositionally.
+For example, the DNA pattern that the BisI restriction enzyme recognizes is \li{AT}$x$\li{GC}, where $x$ is any of the four DNA bases.
+However, one cannot pass \li{any_base} to the string concatenation operator, because \li{any_base} is already of type \li{Regex.t}.
+These kinds of situations cause programmers to engage in \emph{stringly-typed programming}, as discussed in Sec.~\ref{sec:intro} -- they simply use strings to encode regexes throughout their application, rather than structured encodings like \li{Regex.t}.
+This can be catastrophic if the programmer is not extremely careful with string concatenation because strings that happen to contain special characters recognized by the POSIX notation can be misinterpreted because they are not distinguished in any way from strings that intentionally use POSIX notation.
+When some of these strings come from untrusted sources, e.g.
+web forms as is common in bioinformatics applications, the result can be a breach in privacy.
+
+% Moreover, there already exists a broadly adopted shorthand notational standard -- the POSIX standard for regular expressions \cite{STD95954}.
+For the purposes of this section, we will stick to the subset of the POSIX standard that applies to the encoding in Fig.~\ref{fig:Regex-module-def}, but there is nothing that would fundamentally prevent us from implementing the full POSIX standard using TLMs.
+
+\vspace{-3px}
 \subsection{TLM Definition and Application}
-The definition of the \li{$html} TLM is given in Figure \ref{fig:html-tlm-def}. TLM names are prefixed by \li{$} 
+Figure \ref{fig:regex-tlm-def} defines a TLM named \li{$regex} that supports a POSIX-style literal notation, extended with splice forms, for values of type \li{Regex.t}.
+Line 1 of Fig.
+\ref{fig:first-tlm-example} applies this TLM to construct the regex \li{any_base} just described.
+Line 2 of Fig.
+\ref{fig:first-tlm-example} then applies \li{$regex} again, using its regex splicing notation to construct a regex \li{bisA} matching the DNA sequences recognized by the BisA restriction enzyme, where the middle base can be any base.
+Finally, Lines 3-4 of Fig.
+\ref{fig:first-tlm-example} define a function that constructs a more complex regex from these two regexes and a given gene sequence represented as a string.
+
+Notice that regexes are spliced in using \li{$(}$e$\li{)}, where $e$ is any expression of type \li{Regex.t}, while strings are spliced in using \li{$$(}$e${\li)}, where $e$ is any expression of type \li{string}.
+These choices are made entire
+
+TLM names are prefixed by \li{$} 
  to distinguish them from variables.
@@
 Instead, there is are separate \emph{static bindings} marked by the \li{static} keyword that populate a \emph{static environment} that is discarded after expansion finishes.
-A more detailed account of static evaluation, both informal and informal, is available in the supplement.
+A more detailed account of static evaluation, both informal and informal, is given in the supplement.
 An alternative design that allows for the explicit lowering of standard-phase modules to the static phase has also been proposed for OCaml \cite{Ocaml/macros}.
@@
 \subsection{HTML}
+We start in this section by describing the TLM mechanism in detail by way of a substantial case study: literal notation, like that built in to Ur/Web (see Figure \ref{fig:urweb}) and also, presently, Reason, for expressions of the type  \li{html} in Figure \ref{fig:html-type-def}.
+
 \begin{figure*}[t]

### 2018-03-06 icfp18/syntax-icfp18.tex
first 6 pages done...

@@
 \vspace{-5px}
-\caption{The definition of the \li{$regex} TLM. Figure \ref{fig:candidate-exp-Reason} defines \li{RegexLexer} and \li{RegexParser}.}
+\caption{The definition of the \li{$regex} TLM. Figure \ref{fig:lexer-and-parser} defines \li{RegexLexer} and \li{RegexParser}.}
 \label{fig:regex-tlm-def}
@@
 
+Let us consider the second of these TLM applications more closely:
+\begin{lstlisting}[numbers=none]
+    $regex `(GC$(any_base)GC)`
+\end{lstlisting}
+According to the context-free grammar of the language, this form is simply a leaf of the unexpanded syntax tree, like a string literal would be. TLM names are prefixed by \li{$} 
+ to distinguish them from variables.
+There are no lexical constraints on the body of the generalized literal form except that any occurrences of \li{`(} must be balanced by \li{)`}, much like nested comments in OCaml, Reason and other languages.
+Generalized literal forms, which first arose in the prior work on TSLs \cite{TSLs}, therefore syntactically subsume other literal forms.
+The prior work specified several other choices of outer delimitation, including layout-sensitive delimitation, but for the purposes of this work, we will use \li{`(} and \li{)`} exclusively.
+The fact that the context-free grammar of the base language is fixed ensures that notation providers can reason modularly about syntactic ambiguity in their own grammars.
+
+Notice that in the unexpanded expression above, we did not distinguish the spliced expression from the rest of the literal.
+This is to emphasize that it is only during the subsequent \emph{expansion} phase that the body of the generalized literal form is lexed, parsed and expanded to produce an expanded syntax tree where spliced expressions, like \li{any_base}, have been revealed.
+Responsibility for lexing, parsing and expanding the literal body is delegated to the lexer and parser specified by the applied TLM by the clauses \li{lexer} and \li{parser}, respectively.
+
+\vspace{-4px}
+\subsection{Abstract Reasoning Principles} 
+Rather than looking immediately at Fig.
+\ref{fig:lexer-and-parser}, which defines the lexer and parser specified by \li{$regex}, let us first consider the perspective of a client programmer who does not want to delve into the details of lexing and parsing for each piece of literal notation that they encounter.
+What can this programmer deduce just by reasoning from the information presented in Figures \ref{fig:Regex-module-def} through \ref{fig:first-tlm-example}?
+
+ TLM definitions follow the same scoping rules as other definitions, so the client programmer can follow the binding structure of the language in the usual manner (manually or assisted by a ``go to definition'' editor service) to determine which TLM is uniquely responsible for each literal form.
+In Sec.
+\ref{sec:ptsms}, we will see an example of TLM synonyms (as a trivial case of partial parameter application).
+
+Having arrived at the definition of \li{$regex}, the client can immediately determine the type of the expansion being generated at each application site because it is specified explicitly by the clause \li{at Regex.t} on Line 1 of Fig.
+\ref{fig:regex-tlm-def}.
+This is analagous to the return type of a function.
+The identity of \li{Regex.t} is determined relative to the TLM definition site, not at each application site (so the module \li{Regex} need not be in scope at the application site, or it can be shadowed by a different module).
+
+The TLM mechanism enforces an similarly strong context independence property on generated expansions by requiring that the TLM explicitly specify the modules that the expansions it generates might internally depend on.
+In this case, Lines 4-5 of Fig.
+\ref{fig:regex-tlm-def} specify that generated expansion might use the module \li{Regex}, again as it is bound at the definition site, using the module variable \li{Regex} internally.
+These happen to coincide in this case, but in general, the dependency can be an arbitrary module path while internally it must be bound to a module variable, for example:
+\begin{lstlisting}[numbers=none]
+  expansions require module Core.List as L
+\end{lstlisting}
+The module variable \li{L} will be bound internally to \li{Core.List} in all expansions generated by this TLM, including those where \li{L} or \li{Core.List} are not bound, or otherwise bound.
+All other bindings, whether at the TLM definition site or the TLM application site, are not internally available to the expansion.
+This allows client programmers to freely rename and shadow variables as they are used to, without needing to inspect each parser to determine whether it requires certain modules be bound to particular names at each use site (as is common with, e.g., syntax extensions based on \li{camlp4} or Sugar*, or when using unhygienic term rewriting approaches like OCaml's PPX system or Template Haskell \cite{SheardPeytonJones:Haskell-02}, as further discussed in Sec.
+\ref{sec:existing-approaches}).
+
+Enforcing this context independence property is actually quite subtle because TLM parsers can extract spliced expressions from the literal body.
+Na\"ively checking that the final expansion is closed except for the explicitly named dependencies would inappropriately enforce the same context independence property on client-side spliced expressions.
+For example, consider the final expansion of the example from Line 2 of Fig.
+\ref{fig:first-tlm-example} above:
+\begin{lstlisting}[numbers=none]
+  Regex.Seq(Regex.Str "GC", Regex.Seq(any_base, Regex.Str "GC"))
+\end{lstlisting}
+In this term, both \li{Regex} and \li{any_base} are free variables.
+The origin of the sub-expression \li{any_base} -- i.e.
+that it is a spliced expression in the literal body at the application site, and not based on an inappropriate assumption that the parser is making about the application site -- is not apparent from the final expansion.
+As we will discuss more extensively in Sec.
+\ref{sec:existing-approaches}, this is why existing hygienic macro systems cannot be used to repurpose string literals for literal notation at other types.
+
+To address this problem,
+
+When a TLM definition appears inside a module, it must also appear in the signature with the same specification, up to the usual notions of type and module path equivalence in the type annotation and module dependencies (like datatype definitions in ML).
+Note that even when a TLM needs only a single helper function, it must be placed into a named module.
+This is to avoid needing to reason about the equivalence of arbitrary expressions when determining whether two signatures are compatible.
+
 Notice that regexes are spliced in using \li{$(}$e$\li{)}, where $e$ is any expression of type \li{Regex.t}, while strings are spliced in using \li{$$(}$e${\li)}, where $e$ is any expression of type \li{string}.
@@
 
-TLM names are prefixed by \li{$} 
- to distinguish them from variables.
  Every TLM definition also includes a \emph{type annotation}, here \li{at html}, and a \emph{parse function} between \li{by static \{} and \li{\}}.

### 2018-03-06 icfp18/syntax-icfp18.tex
finished thru p7

@@
 commentstyle=\itshape\ttfamily\color{gray},
-stringstyle=\ttfamily\color{red},
+stringstyle=\ttfamily\color{purple},
 mathescape=false,escapeinside={(~}{~)},
@@
 moredelim=**[is][\color{Sepia}]{SUS}{EUS},
-deletestring=[d]{"},
+% deletestring=[d]{"},
 }
@@
 \begin{subfigure}[t]{0.45\textwidth}
+\vspace{-1px}
 \begin{lstlisting}[mathescape=~]
@@
 \begin{subfigure}[t]{0.53\textwidth}
+\vspace{-1px}
 \begin{lstlisting}[mathescape=|]
@@
   let bisA = $regex `(SURLGC$(EURLany_baseSURL)GCEURL)`;
-  let gene_restriction_template(gene_sequence) => 
-    $regex `(SURL$(EURLbisASURL)$(EURLany_baseSURL)*$$(EURLgene_sequenceSURL)$(EURLany_baseSURL)*$(EURLbisASURL)EURL)`
+  let gene_restriction_template(gene) => 
+    $regex `(SURL$(EURLbisASURL)$(EURLany_baseSURL)*$$(EURLgeneSURL)$(EURLany_baseSURL)*$(EURLbisASURL)EURL)`
 \end{lstlisting}
@@
 \caption{Case Study: POSIX-style regex literal notation, with support for string and regex splicing.}
-\vspace{-3px}
+\vspace{-5px}
 \end{figure}
@@
 Having introduced the basic machinery, we continue in Sec.~\ref{sec:more-examples} with examples that further demonstrate the expressive power of this approach and various parser implementation strategies that notation providers can consider.
-Sec.~\ref{sec:setsms-formally} presents a type-theoretic calculus of simple expression and pattern TLMs and formally establishes the reasoning principles implied above, in the simplest setting. 
-Sec.~\ref{sec:ptlms-formally} adds type functions and an ML-style module system to this calculus, and gives the most general variant of the reasoning principles theorem. 
+This notably includes our take on the example from  Fig.~\ref{fig:urweb} of Ur/Web-style HTML literals.
+Next, Sec.~\ref{sec:setsms-formally} defines a type-theoretic calculus of simple expression and pattern TLMs and formally establishes the reasoning principles implied above in their simplest form. 
+Sec.~\ref{sec:ptlms-formally} adds type functions and an ML-style module system to this calculus, and gives a more general variant of the reasoning principles theorem. 
 % \item 
@@
 \subsection{TLM Definition and Application}
-Figure \ref{fig:regex-tlm-def} defines a TLM named \li{$regex} that supports a POSIX-style literal notation, extended with splice forms, for values of type \li{Regex.t}.
+Figure \ref{fig:regex-tlm-def} defines a TLM named \li{$regex} for values of type \li{Regex.t} that supports a POSIX-style literal notation extended with splice forms for regexes and strings.
 Line 1 of Fig.
@@
 Line 2 of Fig.
-\ref{fig:first-tlm-example} then applies \li{$regex} again, using its regex splicing notation to construct a regex \li{bisA} matching the DNA sequences recognized by the BisA restriction enzyme, where the middle base can be any base.
+\ref{fig:first-tlm-example} then applies \li{$regex} again, using its regex splice form to construct a regex \li{bisA} matching the DNA sequences recognized by the BisA restriction enzyme, where the middle base can be any base.
 Finally, Lines 3-4 of Fig.
-\ref{fig:first-tlm-example} define a function that constructs a more complex regex from these two regexes and a given gene sequence represented as a string.
-
-Let us consider the second of these TLM applications more closely:
+\ref{fig:first-tlm-example} define a function that constructs a more complex regex from these two regexes and a given gene sequence represented as a string, which is spliced in using a string splice form delimited by \li{$$(} and \li{)}.
+
+Let us consider the second of these three TLM applications more closely:
 \begin{lstlisting}[numbers=none]
@@
 
- TLM definitions follow the same scoping rules as other definitions, so the client programmer can follow the binding structure of the language in the usual manner (manually or assisted by a ``go to definition'' editor service) to determine which TLM is uniquely responsible for each literal form.
+% TLM definitions follow well-defined scoping rules, so the client programmer can follow the binding structure of the language in the usual manner (manually or assisted by a ``go to definition'' editor service) to determine which TLM is uniquely responsible for each literal form.
 In Sec.
@@
 
-Having arrived at the definition of \li{$regex}, the client can immediately determine the type of the expansion being generated at each application site because it is specified explicitly by the clause \li{at Regex.t} on Line 1 of Fig.
+From the definition of the \li{$regex} TLM, the client can immediately determine the type of the expansion being generated at each application site because it is specified explicitly by the clause \li{at Regex.t} on Line 1 of Fig.
 \ref{fig:regex-tlm-def}.
@@
 \begin{lstlisting}[numbers=none]
-  expansions require module Core.List as L
+  expansions require 
+    module Regex as Regex
+    module Core.List as L
 \end{lstlisting}
@@
 
-Enforcing this context independence property is actually quite subtle because TLM parsers can extract spliced expressions from the literal body.
-Na\"ively checking that the final expansion is closed except for the explicitly named dependencies would inappropriately enforce the same context independence property on client-side spliced expressions.
+Enforcing this strong context independence condition is quite subtle because TLM parsers need to be able to extract spliced expressions from the literal body and place them in the final expansion.
+Na\"ively checking that the final expansion is closed except for the explicitly named dependencies would inappropriately enforce context independence on application site spliced expressions, which should certainly not be prevented from referring to variables in the application site context.
 For example, consider the final expansion of the example from Line 2 of Fig.
@@
 \end{lstlisting}
-In this term, both \li{Regex} and \li{any_base} are free variables.
-The origin of the sub-expression \li{any_base} -- i.e.
-that it is a spliced expression in the literal body at the application site, and not based on an inappropriate assumption that the parser is making about the application site -- is not apparent from the final expansion.
+In this term, both \li{Regex} and \li{any_base} are free variables. There is nothing to indicate which free variables came from the literal body via an intentional act of splicing by the TLM. 
+%\footnote{An exhaustive search would not be feasible because final expansions are abstract syntax trees.
+There is not a unique mapping back to concrete syntax, e.g.
+because whitespace can be inserted arbitrarily.}
 As we will discuss more extensively in Sec.
-\ref{sec:existing-approaches}, this is why existing hygienic macro systems cannot be used to repurpose string literals for literal notation at other types.
-
-To address this problem,
+\ref{sec:existing-approaches}, this is why traditional hygienic term-rewriting macro systems, like those available in various Lisp-family languages \cite{mccarthy1978history} and in Scala \cite{ScalaMacros2013}, cannot be used to repurpose string literals for literal notation at other types.
+
+To address this problem, the TLM parser does not generate the final expansion directly.
+Instead, the parser generates a \emph{proto-expansion} that refers to spliced expressions indirectly by location relative to the start of the provided literal body.
+For example, the proto-expansion generated by \li{$regex} for the example above can be expressed as follows:
+\begin{lstlisting}[numbers=none]
+  Regex.Seq(Regex.Str "GC", Regex.Seq(spliced<4; 11; Regex.t>, Regex.Str "GC"))
+\end{lstlisting}
+
+Here, \li{spliced<4; 11; Regex.t>} is a reference to the spliced expression \li{any_base} by its location relative to the start of the literal body being expanded.
+The context independence condition can be enforced straightforwardly on the proto-expansion -- the only free variable in the proto-expansion is \li{Regex}, which is an explicitly listed dependency.
+% Requiring that TLMs refer to spliced expressions indirectly in this manner ensures that a TLM cannot ``forge'' spliced terms, i.e.
+mark some sub-term as being spliced when it does not in fact appear at some specified segment of the literal body.
+
+The set of splice references in the proto-expansion generated for a literal body is called the \emph{segmentation} of that literal body.
+The system checks that the segmentation does indeed segment the literal body, i.e.
+that the segments are non-overlapping.
+For the purposes of abstract reasoning, only the segmentation needs to be reported to downstream editor services, and ultimately to the programmer.
+The rest of the proto-expansion can be held abstract.
+In this paper, spliced segment locations are reported using colors, with segments that are not part of spliced segment shown in color (green, so far) while spliced segments start out black.
+When TLM applications are nested, a distinct color can be used at each depth.
+For example, in Sec.
+\ref{sec:more-examples}, we will describe a TLM \li{$html} for HTML notation in the style of Ur/Web (see Fig.~\ref{fig:urweb}).
+We might apply it, together with another TLM \li{$smiles} for chemical structures described using the SMILES standard \cite{anderson1987smiles}, as follows:
+\begin{lstlisting}[numbers=none]
+    $html `(SURL<div>Chemical structure of sucrose: {EURL
+      Smiles.to_svg
+        $smiles `(SCOLOR{ECOLORmono_glucoseSCOLOR}-O-{ECOLORmono_fructoseSCOLOR}ECOLOR)` 
+    SURL}</div>EURL)`
+\end{lstlisting}
+
+Each splice reference in the segmentation gives not just the location of the spliced segment but also a type annotation.
+Returning to the regex example above, the specified type annotation is \li{Regex.t}.
+This type annotation, which must also be context independent, constrains the spliced expression.
+Consequently, type inference can be performed using only the segmentation together with the type annotation on the applied TLM's definition.
+For example, consider the function \li{gene_restriction_template} on Lines 3-4 of Fig.
+\ref{fig:first-tlm-example}.
+The return type of this function can be determined to be \li{Regex.t} because of the type annotation on the \li{$regex} TLM.
+The type of the argument, \li{gene}, can be inferred to be \li{string} because the segmentation specifies the type \li{string} for the spliced segment where it appears.
+Context independence implies that \li{gene} cannot appear elsewhere in the expansion, and so no further typing constraints could possibly be collected from examining the portions of the expansion held abstract.
+Spliced segment types can be communicated directly to the programmer upon request by an editor service, e.g.
+\li{merlin} for OCaml/Reason \cite{Merlin}.
+
+---------------------- {\color{red} stuff below is notes / not yet revised}
 

### 2018-03-07 icfp18/syntax-icfp18.tex
restructured introduction

@@
 This is unsatisfying because there are many other data structures for which literal notation might be useful, e.g.
-hash tables, regular expressions, HTML data, SQL queries, syntax trees and chemical structures.
-There may also be different implementations of each of these data structures, e.g.
-with different performance characteristics, that could all benefit from common literal notation.
+finite maps, regular expressions, HTML data, SQL queries, syntax trees and chemical structures.
+There may also be different implementations of each of these data structures, perhaps with different performance characteristics, that could all benefit from common literal notation.
 %A less \emph{ad hoc} design would be one that allows library providers to define new literal forms in a decentralized manner.
@@
 % The surface syntax of a programming language is its user interface. 
-When designing the surface syntax of a general-purpose programming language, it is common practice to define \emph{literal notation} that decreases the syntactic cost of constructing and pattern matching over values of a select few data structures.
+When designing the surface syntax of a general-purpose programming language, it is common practice to define shorthand \emph{literal notation}, i.e.
+notation that decreases the syntactic cost of constructing and pattern matching over values of a particular data structure or parameterized family of data structures.
 For example, many languages in the ML family support list literals like \li{[x1, x2, x3]} in both expression and pattern position \cite{harper1997programming,mthm97-for-dart}.
@@
 The example in Fig.
-\ref{fig:urweb} below shows two HTML literals, one that ``splices in'' a string expression delimited by \li{\{[} and \li{]\}} (Line 1) and the other an HTML expression delimited by \li{\{} and \li{\}} (Line 2).
+\ref{fig:urweb} shows two HTML literals, one that ``splices in'' a string expression delimited by \li{\{[} and \li{]\}} and the other an HTML expression delimited by \li{\{} and \li{\}}.
 %The desugarings of these literal expressions, not shown, are substantially more verbose and less readable to programmers familiar with XHTML.
@@
 \begin{figure}[h]
-\vspace{-12px}
+\vspace{-9px}
 \begin{lstlisting}
@@
 \label{fig:urweb}
-\vspace{-9px}
+\vspace{-7px}
 \end{figure}
 
-This design practice, where the language designer privileges certain library constructs with built-in literal notation, is decidedly \emph{ad hoc} in that it is easy to come up with other examples of data structures for which mathematicians, scientists or programmers have invented specialized notation \cite{DBLP:journals/cacm/Iverson80,cajori1928history,TSLs}.
-For example, 1) clients of a ``collections'' library might want not just list literals, but also literal notation for matrices, finite sets, maps and so on; 2) clients of a ``web programming'' library might like CSS literals (which Ur/Web lacks); 3) a compiler author might like ``quotation'' literals for the terms of the object language and various intermediate languages of interest; and 4) clients of a ``chemistry'' library might like chemical structure literals based on the SMILES standard \cite{anderson1987smiles}.
+This design practice, where the language designer privileges certain library constructs with built-in literal notation, is \emph{ad hoc} in that it is easy to come up with other examples of data structures for which mathematicians, scientists or programmers have invented specialized notation \cite{DBLP:journals/cacm/Iverson80,cajori1928history,TSLs}.
+For example, 1) clients of a ``collections'' library might want not just list literals, but also literal notation for matrices, finite sets, maps and so on; 2) clients of a ``web programming'' library might want CSS literals (which Ur/Web lacks); 3) a compiler author might want ``quotation'' literals for the terms of the object language and various intermediate languages of interest; and 4) clients of a ``chemistry'' library might want chemical structure literals based on the SMILES standard \cite{anderson1987smiles}.
 
@@
 they choose strings instead of composite data structures largely for reasons of notational convenience.
-In particular, \citet{TSLs} sampled strings from open source projects and found that at least 15\% of them could be parsed by some readily apparent type-specific grammar.
-%, e.g. for URLs, paths, regular expressions and so on
-Literal notation, with support for splicing, would decrease the syntactic cost of composite encodings of this data, which is more amenable to programmatic manipulation and compositional reasoning.
+In particular, \citet{TSLs} sampled strings from open source projects and found that at least 15\% of them could be parsed by some readily apparent type-specific grammar, e.g.
+for URLs, paths, regular expressions and many others.
+Literal notation, with support for splicing, would decrease the syntactic cost of composite encodings, which are more amenable to programmatic manipulation and compositional reasoning.
 %Literal forms would  preferable to string encodings because they support compositional construction and decomposition by pattern matching via splicing.
@@
 Instead, there has been persistent interest in mechanisms that allow library providers to define new literal notation on their own.
-For example, direct grammar extension systems like Camlp4 \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, macro-based systems like Omar et al's \emph{type-specific languages} (TSLs) \cite{TSLs}, and other systems that we will discuss in Sec.~\ref{sec:existing-approaches} can all be used to define new literal notation (and, in some cases, other forms of new notation, such as new infix expression forms or type declaration forms, which we leave beyond the scope of this paper).
-
-\paragraph{Problem} The problem that specifically motivates this paper is that these existing syntax extension systems  make it difficult or impossible to reason abstractly about such fundamental issues as types and variable binding.
+For example, direct grammar extension systems like Camlp4 \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, macro-based systems like Template Haskell~\cite{SheardPeytonJones:Haskell-02,mainland2007s} and the \emph{type-specific languages} (TSLs) of \citet{TSLs}, and other systems that we will discuss below can all be used to define new literal notation (and, in some cases, other forms of new notation, such as new infix expression forms, control flow operations or type declaration forms, which we leave beyond the scope of this paper).
+
+\paragraph{Problem} The problem that specifically motivates this paper is that these existing systems  make it difficult or impossible to reason abstractly about such fundamental issues as types and variable binding.
 Instead, programmers and editor services can only reason transparently, i.e.
@@
 \vspace{-5px}
-\caption{It is difficult to reason abstractly about programs that use unfamiliar grammar extensions (see the five reasoning criteria in the text).}
+\caption{It is difficult to reason abstractly about programs that use unfamiliar grammar extensions (see the six reasoning criteria in the text).}
 \label{fig:K-dialect}
@@
 \end{subfigure}
-% \vspace{4px}
+\vspace{4px}
 \caption{Two of the possible ways to introduce literal notation for encodings of K queries}
@@
 
-Consider, for example, the perspective of a programmer attempting to comprehend the program text in Figure \ref{fig:K-dialect}, which is written in a dialect of OCaml's surface syntax called Reason \cite{reason-what} that has, hypothetically, been extended with some number of new literal forms by a grammar extension system -- Lines 1-5 show the grammar extension syntax from Camlp4 \cite{ocaml-manual}, but Sugar*/SugarHaskell is  similar \cite{erdweg2011sugarj,erdweg2013framework,erdweg2012layout}.
-The literal form on Line 9 constructs an encoding of a query in the niche stack-based database query language K, using its intentionally terse notation~\cite{Whitney:2001:LOR:376284.375783}.
+Consider, for example, the perspective of a programmer attempting to comprehend the program text in Figure \ref{fig:K-dialect}, which is written in a dialect of OCaml's surface syntax called Reason \cite{reason-what} that has, hypothetically, been extended with some number of new literal forms by a grammar extension system --- Lines 1-5 show the grammar extension syntax from Camlp4 \cite{ocaml-manual}, but Sugar*/SugarHaskell is  similar \cite{erdweg2011sugarj,erdweg2013framework,erdweg2012layout}.
+Line 9 uses one of the active syntax extensions to construct an encoding of a query in the niche database query language K, using its intentionally terse notation~\cite{Whitney:2001:LOR:376284.375783}.
 The problem is that a programmer looking at the program as presented, and unfamiliar with (i.e.
@@
 
-\begin{enumerate}[leftmargin=12px]
-\item[\Circle] \textbf{Responsibility}: Which syntax extension even determines the expansion of the literal on Line~9?
-Might activating a new extension generate a conflicting expansion for the same form?
-\item[\Circle] \textbf{Segmentation}: Are the characters \li{x}, \li{R} and \li{2} on Line 9 parsed as spliced expressions (meaning that they appear directly in the underlying expansion), or are they parsed in some other way peculiar to this literal form (e.g.
+\begin{enumerate}[leftmargin=15px]
+\item \textbf{Responsibility}: Which syntax extension determined the expansion of the literal on Line~9? Might activating a new extension generate a conflicting expansion for the same literal?
+\item \textbf{Expansion Typing}: What type does \li{y} have?
+\item \textbf{Context Dependence}: Which bindings does the expansion of Line~9 invisibly depend on?
+If I shadow or remove a module or other binding, could that break or change the meaning of Line~9 because its expansion depended invisibly on the original binding?
+\item \textbf{Segmentation}: Are the characters \li{x}, \li{R} and \li{2} on Line~9 parsed as spliced terms, meaning that they appear directly in the underlying expansion, or are they parsed in some other way peculiar to this literal notation (e.g.
 as operators in the K query language)?
-\item[\Circle] \textbf{Capture}: If \li{x} is in fact a spliced expression, does it refer to the binding of \li{x} on Line 5, or might it capture an unseen binding of the same identifier in the expansion of Line 9?
-\item[\Circle] \textbf{Context Dependence}: If \li{w}, on Lines 7-8, is renamed, could that break Line 9  because the expansion assumes that \li{w} is in scope?
-What other hidden assumptions is the expansion making?
-\item[\Circle] \textbf{Typing}: What type does \li{y} have? What type is each spliced expression expected to have?
+\item \textbf{Segment Typing}: What type is each spliced term expected to have?
+How can I infer a type for a variable that appears in a spliced term without looking at where it ends up in the expansion?
+\item \textbf{Capture}: If \li{x} is in fact a spliced term, does it refer to the binding of \li{x} on Line 5, or might it capture an invisible binding of the same identifier in the expansion of Line 9?
 \end{enumerate}
@@
 
-An approach that has made some headway toward this ideal is Omar et al.'s \emph{type-specific languages} (TSLs) \cite{TSLs}.
-A language that supports TSLs delegates control over the parsing and expansion of certain \emph{generalized literal forms} to a parser associated with the type that the form is being checked against.
-So if we add TSLs to our language, our example then requires a type annotation on \li{y}:
-\begin{lstlisting}[numbers=none,basicstyle=\ttfamily\fontsize{9pt}{1em}\selectfont]
-  let y : KQuery.t = `((!R)@&{&/x!/:2_!x}'!R)`
-\end{lstlisting}
-The parser associated with the type \li{KQuery.t} is statically invoked to lex, parse and expand the literal body, i.e.
-the sequence of characters between the outer delimiters, here \li{`(} and \li{)`}.
-The literal body is constrained by the context-free grammar of the language only in that nested delimiters must be matched (like comments in OCaml).
-TSLs are closely related to the mechanism that we will propose, so let us analyze TSLs from the perspective of the five reasoning criteria just outlined.
-\begin{enumerate}[leftmargin=12px]
-  \item[\LEFTcircle] \textbf{Responsibility}: The type-directed dispatch mechanism allows the client to easily determine which parser is responsible for each literal form: the parser that was associated with the type when it was defined.
-Clients do not need to worry about different TSLs conflicting syntactically because the context-free grammar of the language remains fixed and composition is mediated by splicing.
-The main limitation here is that it is impossible to define literal notation after a type has been defined, and it is also impossible to define multiple notations at a single type.
-  \item[\LEFTcircle] \textbf{Segmentation}: The parser can splice expressions out of the literal body, but there is no clear way to associate each spliced term with some particular segment (i.e.
-subsequence) of the literal body, nor is there a guarantee that spliced terms are non-overlapping.
-As such, there is no way to indicate to the programmer where base language expressions are located within a literal form.
-  \item[\LEFTcircle] \textbf{Capture}: The TSL mechanism as described enforces complete capture avoidance (though the formal specification given in the paper did not adequately enforce this constraint).
-  \item[\LEFTcircle] \textbf{Context Dependence}: The TSL mechanism as described requires that the expansion be completely closed, so it is trivially context independent.
-However, this comes at a significant expressive cost: the generated expansions cannot make use of any libraries whatsoever.
-  \item[\LEFTcircle] \textbf{Typing}: The expansion must necessarily be of the associated type, so abstract reasoning about the type of the expansion is straightforward.
-However, there is no easy way to determine the type expected for each spliced expression.
-In addition, the prior work considered only a monomorphic, nominally-typed language with local type inference, leaving open a number of problems that came up as we considered integrating TSLs into Reason/OCaml:
-\begin{enumerate}[leftmargin=15px,nolistsep,noitemsep]
-  \item \textbf{Structural Types}: There is no way to define literal notation at structural types, e.g. tuple and arrow types, because there is no ``definition site'' for such types.
-  \item \textbf{Parameterized \& Abstract Types}: There is no way to define literal notation over a type-parameterized family of types, e.g.
-at all types \li{list('a)}.
-Similarly, there is no way to define literal notation over a module-parameterized family of abstract types, e.g.
-at every abstract type defined by a module implementing the \li{QUEUE} signature.
-Parameterized and abstract type families are ubiquitous in ML-family languages.% Programmers in ML-family languages commonly define various implementations of abstract data types in this way.
-  \item \textbf{Pattern Literals}: Pattern matching is ubiquitous in ML-family languages but the prior work on TSLs considered only expression literals.
-%Pattern literals are dual to expression literals in that expression literals support value construction, while pattern literals support value deconstruction.
-  \item \textbf{ML-Style Type Inference}: It is not clear that a type-directed dispatch scheme can be cleanly  reconciled with ML-style type inference.
-\end{enumerate}
-\end{enumerate}
-
-We consider other existing approaches against these criteria in Sec.~\ref{sec:existing-approaches}.
-Taken together, we found that no existing approach was reasonable and expressive enough for integration into Reason/OCaml.
-
-\paragraph{Contributions} This paper introduces \emph{typed literal macros} (TLMs): the first system for defining new literal notation that (1) provably maintains all five of the abstract reasoning principles just outlined; and (2) is semantically expressive enough for integration into Reason/OCaml and other full-scale statically typed functional languages.
-We evaluate these claims with (1) a series of increasingly ML-like calculi equipped with proofs of these abstract reasoning principles; and (2) a number of non-trivial examples that appear throughout the paper and involve the language features mentioned above.
+Given these issues, we concluded that direct grammar extension systems were not ideally suited for integration into Reason.
+We also evaluated various approaches that are based not on direct grammar extension but on term rewriting over a fixed grammar.
+We give a full account of this evaluation in Sec.~\ref{sec:existing-approaches}.
+Briefly, we found that:
+\begin{itemize}[leftmargin=15px]
+\item Unhygienic approaches, like OCaml's preprocessor extension point (PPX) rewriters \cite{ocaml-manual} and Template Haskell \cite{SheardPeytonJones:Haskell-02,mainland2007s} allow us to define new literal notation with support for splicing by repurposing existing string literal forms.
+They also partially or completely solve the problem of reasoning about \textbf{Responsibility} by using explicit annotations on terms that are being rewritten.
+However, they do not satisfy the remaining five reasoning criteria.
+\item Hygienic staging macro systems, like MetaOCaml and related systems \cite{DBLP:conf/flops/Kiselyov14,ganz2001macros,Sheard:1999:UMS} do not give the expander access to the syntax trees of arguments, i.e.
+the rewriting is parametric in the arguments, so they are not suitable for the problem of defining new literal notation.
+\item Hygienic term rewriting macro systems, like those in various Lisp-family languages \cite{DBLP:conf/esop/HermanW08,Herman10:Theory,Kohlbecker86a,DBLP:conf/popl/Adams15,DBLP:conf/popl/ClingerR91,DBLP:journals/lisp/DybvigHB92,mccarthy1978history,Flatt:2012:CLR:2063176.2063195} and Scala \cite{ScalaMacros2013}, do not allow us to repurpose string literal forms to define literal forms at other types because the hygiene discipline does not allow us to splice terms out of literal bodies via parsing.
+Only existing composite forms can therefore be repurposed.
+\item An approach that came close was Omar et al.'s \emph{type-specific languages} (TSLs) \cite{TSLs}, but it fell short with regard to the six reasoning principles just discussed in ways that we will return to and in any case, the mechanism was designed for a simple monomorphic, nominally-typed language and relied critically on a particular local type inference scheme.
+It is not suitable for a language with an ML-like semantics, meaning a language with support for structural types (like tuple types and arrow types in ML), parameterized types, abstract types, modules, pattern matching and non-local type inference.
+\end{itemize}
+
+% Based on this evaluation, we concluded that the existing approaches were not ideally suited for integration into Reason, an increasingly popular open source project that seeks to provide a more clear and practical surface syntax for the OCaml programming language.
+
+\paragraph{Contributions} This paper introduces \emph{typed literal macros} (TLMs): the first system for defining new literal notation, of nearly arbitrary design, that (1) provides the ability to reason abstractly about all six of the topics just outlined; and (2) is semantically expressive enough for integration into Reason/OCaml and other full-scale statically typed functional languages.
+We evaluate these claims with (1) a series of increasingly ML-like calculi equipped with proofs of the claimed abstract reasoning principles; and (2) a number of non-trivial examples that appear throughout the paper and involve the language features mentioned above.
 In describing these examples, we also demonstrate that literal parsing logic can be defined using standard, unmodified parser generators, so the burden on notation providers is comparable to that of existing systems despite these stronger client-side reasoning principles.
@@
 
-TLMs, like TSLs, operate on generalized literal forms.
-With TLMs, however, the mechanism for defining new literal notation is not tied up with the type definition mechanism of the language.
-TLMs are instead explicitly named and applied.
-For example, Line 9 of Figure \ref{fig:K-tsm-example} applies a TLM named \li{$kq} to a generalized literal form delimited by \li{`(} and \li{)`} to express the K query example.
-The programmer can reason abstractly as follows:
-\begin{enumerate}[leftmargin=12pt]
-\item[\CIRCLE] \textbf{Responsibility}: The lexer and parser specified by the applied TLM on Lines 2-3 is exclusively responsible for lexing, parsing and expanding the literal body.
-We will give more details on constructing a TLM lexer and parser in the next section.
-Due to this explicit dispatch mechanism, there can be any number of TLMs defined at a given type.
-As with TSLs, the context-free grammar of the language is fixed, so the TLM provider can modularly establish that the notation that the TLM implements is unambiguous (even in situations where a spliced term itself applies another TLM, which we will show in a later example).
-\item[\CIRCLE] \textbf{Segmentation}: The intermediate output that the TLM generates is structured so that the system can infer from it an accurate \emph{segmentation} of the literal body that distinguishes spliced terms from segments that are parsed in some other way by the TLM.
+\paragraph{Brief Overview} To finish up our discussion of the K query example above, let us give a brief overview of the TLM-based solution outlined in Figure \ref{fig:K-tsm-example}.
+On lines 1-5, we define a TLM named \li{$kq}.
+On Line 9, we apply this TLM.
+We can reason abstractly about this program as follows:
+\begin{enumerate}[leftmargin=15pt]
+\item \textbf{Responsibility}: The lexer and parser specified by the applied TLM on Lines 2-3 are exclusively responsible for lexing, parsing and expanding the body of the generalized literal form, i.e.
+the characters between \li{`(} and \li{)`}.
+We will give more details on generalized literal forms, and on constructing a TLM lexer and parser in the next section.
+For now, it suffices to say that our design goal is to provide a mechanism where the programmer does not normally need to look up the definitions of \li{KQueryLexer} and \li{KQueryParser} to reason about types and binding.
+% To briefly summarize, our system uses Menhir, which is the most popular parser generator in the OCaml ecosystem and integrated into its standard build tools.
+%the context-free grammar of the language is fixed, so the TLM provider can modularly establish that the notation that the TLM implements is unambiguous (even in situations where a spliced term itself applies another TLM, which we will show in a later example).
+\item \textbf{Expansion Typing}: The type annotation on the definition of \li{$kq} (Line 1) determines the type that the expansion must have, here \li{KQuery.t}.
+\item \textbf{Context Dependence}: The external dependencies of the generated expansion are explicitly specified.
+Here, Lines 4-5 specify that expansions generated by \li{$kq} require the module \li{KQuery}.
+The system ensures that this dependency is  bound as specified even if the identifier \li{KQuery} has been shadowed at the application site.
+By making dependencies explicit, we can be sure that the expansion does not rely on the fact that, e.g., \li{w} is in scope at the application site.
+\item \textbf{Segmentation}: The intermediate output that the TLM generates is structured so that the system can infer from it an accurate \emph{segmentation} of the literal body that distinguishes spliced terms, which appear directly in the expansion, from segments that are parsed in some other way by the TLM.
 The segmentation is all that needs to be communicated to editor services downstream of the expander, and ultimately to the programmer using secondary notation, e.g.
 colors in this document.
-So by examining Figure \ref{fig:K-tsm-example}, the programmer knows that the two instances of \li{x} on Line 3 are parsed as spliced expressions (because they are black), whereas the \li{R}'s must be parsed in some other way, e.g.
-as operators in the K language (because they are green).
-\item[\CIRCLE] \textbf{Capture}: Splicing is capture-avoiding, so the spliced expression \li{x} must refer to the binding of \li{x} on Line 2.
-It cannot capture a coincidental binding of \li{x} in the expansion.
-We will say more about capture later on.
-\item[\CIRCLE] \textbf{Context Dependence}: The system enforces context independence, so the expansion of Line 3 cannot  rely on the fact that, e.g., \li{w} is in scope.
-The client is free to rename \li{w}.
-The external dependencies of the generated expansion are explicit -- here, Lines 4-5 specify that expansions generated by \li{$kq} require the module \li{KQuery} -- and the system ensures that this dependency is  bound as specified even if the identifier \li{KQuery} has been shadowed at the application site.
-\item[\CIRCLE] \textbf{Typing}: The type annotation on the definition of \li{$kq} (Line 1) determines the type that the expansion is validated against, here \li{KQuery.t} (not shown).
-Moreover, each spliced segment in the inferred segmentation also has  a type annotation.
+So by examining Fig.
+\ref{fig:K-tsm-example}, the programmer knows that the two instances of \li{x} on Line 3 are parsed as spliced expressions (because they are black), whereas the \li{R}'s must be parsed in some other way, e.g.
+as operators of the K language (because they are green).
+\item \textbf{Segment Typing}: Each spliced segment in the inferred segmentation also has  a type annotation.
 This, together with the context independence property, ensures that type inference at the TLM application site can be performed (by editor services or in the programmer's mind) abstractly, i.e.
-given only the inferred segmentation rather than the full expansion.
-We will see examples of TLMs that make use of more advanced features of the OCaml type system -- notably, pattern matching, parameterized types and modules -- later.
+by requesting just the type annotations on the spliced segments.
 %This information is usually not necessary to reason about typing, but it can be conveyed to the programmer upon request by the program editor if desired.
 %TLM definitions follow the usual scoping rules, so it is easy to ``jump to the definition'' of \li{$kq}.
+\item \textbf{Capture}: Splicing is guaranteed to be capture-avoiding, so the spliced expression \li{x} must refer to the binding of \li{x} on Line 2.
+It cannot have captured a coincidental binding of \li{x} in the expansion.
+% We will say more about capture later on.
 \end{enumerate}
+
+% We will see examples of TLMs that make use of the more advanced features of the OCaml type system -- notably, pattern matching, parameterized types and modules -- later.
 
@@
 
+\paragraph{Paper Outline} 
+% \begin{itemize}
+% \item 
+Sec.~\ref{sec:setsms} introduces expression TLMs with a substantially more detailed case study. 
+% \item 
+Sec.~\ref{sec:sptsms} then introduces {pattern TLMs} and describes the special reasoning conditions in pattern position. %The hygiene condition for pattern TLMs is interesting.
+% \item 
+Sec.~\ref{sec:ptsms} introduces the more general {parametric TLMs}, i.e.
+TLMs that can take type and module parameters, which allow us to define literal notation over a type- or module-parameterized family of types.
+This section also reveals that the \li{expansions require} clause can be understood as parameterization followed immediately by partial parameter application.
+Having introduced the basic machinery, we continue in Sec.~\ref{sec:more-examples} with examples that further demonstrate the expressive power of this approach and various parser implementation strategies that notation providers can consider.
+Notably, this section gives our take on the example from  Fig.~\ref{fig:urweb} of Ur/Web-style HTML literals as well as a TLM that implements quasiquotation for Reason language terms, which is useful for writing other TLMs.
+Next, Sec.~\ref{sec:setsms-formally} defines a type-theoretic calculus of simple expression and pattern TLMs and formally establishes the reasoning principles implied above in their simplest form. 
+Sec.~\ref{sec:ptlms-formally} adds type functions and an ML-style module system to this calculus, and gives a more general variant of the reasoning principles theorem. 
+% \item 
+Sec.~\ref{sec:existing-approaches} compares TLMs to related work, guided by the rubric of  reasoning principles just discussed. 
+% \item 
+Sec.~\ref{sec:discussion} concludes with a summary of the contributions of this paper and a discussion of limitations and future work.
+Certain technical details, proofs and extensions to the calculi presented in this paper are given in the supplement.
+% Also available is an additional series of case studies: notation for an encoding of regular expressions as an abstract data type, implemented using TLMs for quasiquotation and grammar-based parser generators.
+%are also available in the supplement.
+Sec.~\ref{sec:static-eval} considers the topic of term evaluation during the  expansion phase, i.e.
+\emph{static evaluation}, in  more detail.
+Sec \ref{sec:static-eval} also gives examples of TLMs that are useful for defining other TLMs, e.g.
+TLMs that implement parser generators and quasiquotation.
+
+%  We say more about LISP macros in Sec.
+\ref{sec:existing-approaches}.% For the present purposes, we can consider Reason as essentially the calculus defined in the supplement extended with various conveniences that are commonly found in other ML-like languages and, notionally, orthogonal to TLMs.
+%Reason is, as its name suggests, a conceptual descendent of ML.
+It diverges from other dialects of ML that have a similar type structure in that it has a bidirectional type system \cite{Pierce:2000:LTI:345099.345100} (like, for example, Scala \cite{OdeZenZen01}) for reasons that have to do with the mechanism of TLM implicits described in Chapters \ref{chap:tsls} and \ref{chap:ptsms}.
+%The reason we will not follow Standard ML \cite{mthm97-for-dart} in giving a complete formal definition of Reason in this work is both to emphasize that the primitives we introduce are ``insensitive'' to the details of the underlying type structure of the language (so TLMs can be considered for inclusion in a variety of languages, not only dialects of ML), and to avoid distracting the reader (and the author) with definitions that are already well-understood in the literature and that are orthogonal to those that are the focus of this work.
+% We will not formally define these features mainly to avoid unnecessarily complicating our presentation with details that are not essential to the ideas presented herein. As such, 
+% All examples written in Reason should be understood to be informal motivating material for the subsequent formal material.
+%We anticipate that future full-scale language specifications will be able to combine the ideas  in the proposed work without trouble.
+%The purpose of the work being proposed is to serve as a reference for those interested in the new constructs we introduce, not to serve as a language specification.
+%We will give a brief overview of these languages are organized in Sec. \ref{sec:Reason}.
+
+% \subsection{Contributions}\label{sec:contributions}
+% This work introduces a system of \textbf{typed literal macros (TLMs)} that gives library providers substantially more syntactic control than existing typed term-rewriting macro systems while maintaining the ability to reason abstractly about types, binding and segmentation.% abstract reasoning principles.
+% comparable to the level of control they have when defining a syntax dialect.
+
+% Clients apply TLMs to \emph{generalized literal forms}.
+
+% Because the context-free syntax is never extended, syntactic conflicts are not a concern.
+
+%As such, the semantics can take the type and binding structure of the surrounding program into account when validating the expansion that the TLM programmatically generates to ensure that clients can answer critical questions related to types and binding, like those enumerated in Section \ref{sec:abs-reasoning-intro}.
+Clients need not have knowledge of the implementation of the TLM or of the generated expansion, i.e.
+there are useful principles of syntactic abstraction.
+
+% The primary technical challenge has to do with the fact that the applied TLM needs to be able to splice terms out of the literal body for inclusion in the expansion. For example,
+
+ % We design our mechanism such that these locations can easily be determined from the output of the TLM.
+This is essential for our hygiene mechanism, and it is also useful in that this information can be presented to the user (e.g.
+as shown in Figure \ref{fig:first-tsm-example-marked}).
+%As such, we must develop a mechanism where 1) the positions of spliced subterms can be determined without examining the macro implementation (e.g.
+so that they can be presented to the user differently by an editor or pretty-printer, ;  and 2) the hygiene mechanism must give only portions of the expansion that correspond to these spliced subterms access to the application site context.
+
+%In order to reason about types and binding, client programmers need only have knowledge of 1) the segmentation (e.g.
+by examining a figure like this presented by a code editor or pretty-printer) and 2) a type annotation on the definition of the applied TLM.
+No other details about the applied TLM's implementation or the expansion that it generates need to be revealed to the client programmer.
+In other words,
+
+% \begin{figure}[h]
+% \begin{lstlisting}
+% PElement Nil Seq(
+%   TextNode "Hello, ", 
+%   Seq(TextNode (join(" ", Cons(first, Cons(second, Nil)))), 
+%   TextNode "!"))
+% \end{lstlisting}
+% \caption{The desugaring.}
+% \end{figure}
+
+% There is also no ambiguity with regard to which TLM has control over each form, and searching for the definition of a TLM is no more difficult than searching for any other binding, i.e.
+there are well-defined scoping rules.
+
+% In other words, TLMs maintain a useful notion of syntactic abstraction.
+%More specifically, TLMs maintain a \emph{hygienic binding discipline}, meaning that questions Questions 4 and 5 above were concerned with are disallowed entirely.
+% We will, of course, make this notion more technically precise as we continue.
+
+% \begin{figure}[h]
+% \begin{lstlisting}[numbers=none,xleftmargin=0px]
+% let syntax $strlist = $list string in 
+% $html `SURL<p>Hello, {[EURLjoin ($str ' ') ($strlist [firstSURL,EURL last])SURL]}</p>EURL`
+% \end{lstlisting} 
+% \caption{The example from Figure \ref{fig:first-tsm-example-marked}, expressed using a parametric TLM.}
+% \label{fig:first-ptsm-example-marked}
+% \end{figure}
+
+% In Secs.
+\ref{chap:uetsms} and \ref{chap:uptsms}, we assume for the sake of technical simplicity that each TLM definition is self-contained, needing no access to libraries or to other TLMs.
+This is an impractical assumption in practice.
+We relax this assumption in In
+
+% %\item \textbf{Type-specific languages}, or \textbf{TSLs}. TSLs, described 
+% In Chapter \ref{chap:tsls}, we develop a mechanism of \emph{TLM implicits} that allows library clients to contextually designate, for any type, a privileged TLM at that type.
+The semantics applies this privileged TLM implicitly to unadorned literal forms that appear where a term of the associated type is expected.
+For example, if we designate
+% %\li{$str} 
+% as the privileged TLM at the \li{string} type and 
+% %\li{$strlist}
+% as the privileged TLM at the \li{list(string)} type, we can express the example from Figure \ref{fig:first-tsm-example-marked} instead as shown in Figure \ref{fig:first-tsm-example-implicit} (assuming \li{join} has type \li{string -> list(string) -> string}.)
+% \begin{figure}[h]
+% \begin{lstlisting}[numbers=none]
+% $html`SURL<p>Hello, {[EURLjoin ' ' [firstSURL,EURL last]SURL]}</p>EURL`
+% \end{lstlisting}
+% \caption{The example from Figure \ref{fig:first-tsm-example-marked} drawn to take advantage of TLM implicits.}
+% \label{fig:first-tsm-example-implicit}
+% \end{figure}
+
+% \noindent This approach is competitive in cost with library-specific syntax dialects (e.g.
+compare Figure \ref{fig:first-tsm-example-implicit} to Figure \ref{fig:urweb}), while maintaining the abstract reasoning principles characteristic of our approach.
+%\item \textbf{Metamodules}, introduced in Sec.
+\ref{sec:metamodules}, reduce the need to primitively build in the type structure of constructs like records (and variants thereof),  labeled sums and other interesting constructs that we will introduce later by giving library providers programmatic ``hooks'' directly into the semantics, which are specified as a \emph{type-directed translation semantics} targeting a small \emph{typed internal language} (introduced in Sec.
+\ref{sec:Reason}).
+%For example, a library provider can implement the type structure of records with a metamodule that:
+%\begin{enumerate}
+%\item introduces a type constructor, \lstinline{record}, parameterized by finite mappings from labels to types, and defines, programmatically, a translation to unary and binary product types (which are built in to the internal language); and
+%\item introduces operators used to work with records, minimally record introduction and elimination (but perhaps also various functional update operators), and directly implements the logic governing their typechecking and translation to the IL (which builds in only nullary and binary products).
+%\end{enumerate}
+%We will see direct analogies between ML-style modules (which our mechanisms also support) and metamodules later.
+%\end{enumerate}
+
+% As vehicles for this work, we will define a small programming language in each of the three parts just mentioned, each building conceptually upon the previous language.
+All of our formal contributions are relative to these small languages.
+
+%TLMs, like other macro systems, perform \emph{static code generation} (also sometimes called \emph{static} or \emph{compile-time metaprogramming}), meaning that the relevant rules in the static semantics of the language call for the evaluation of \emph{static functions} that generate term encodings.
+Static functions are functions that are evaluated statically, i.e.
+during typing.
+%Library providers write these static functions using the Reason \emph{static language} (SL).
+%Maintaining a separation between the static (or ``compile-time'') phase and the dynamic (or ``run-time'') phase is an important facet of Reason's design.
+% static code generation.
+%We will  also introduce a simple variant of each of these primitives that leverages Reason's support for local type inference to further reduce syntactic cost in certain common situations.
+
+\vspace{-3px}
+\newcommand{\seTLMsSec}{Expression TLMs}
+\section{\protect\seTLMsSec}
+\label{sec:setsms}
+\vspace{-2px}
+
+Consider the recursive datatype \li{Regex.t} defined by the module in Fig.~\ref{fig:Regex-module-def}, which encodes \emph{regular expressions} (regexes) into Reason \cite{Thompson:1968:PTR:363347.363387}.
+Regular expressions are  common in fields like bioinformatics, where they are used to express patterns in DNA sequences.
+For example, we can construct a regex that matches the strings \li{"A"}, \li{"T"}, \li{"G"} or \li{"C"}, which represent the four bases in DNA, as follows:
+\begin{lstlisting}[numbers=none]
+  let any_base = Regex.(Or(Str "A", Or(Str "T", Or(Str "G", Or(Str "C")))))
+\end{lstlisting}
+In Reason, the notation \li{Regex.(}$e$\li{)} locally opens the module \li{Regex} for use within the expression $e$, so we do not need to qualify the constructors \li{Regex.Or} and \li{Regex.Str}.
+Even allowing for this shorthand, however, this notation for constructing regexes is rather unwieldy.
+Instead, we would like to be able to use the common POSIX-style notation for constructing values of type \li{Regex.t} \cite{STD95954}.
+
+% One approach would be to simply define a function \li{parse_rx : string -> Regex.t} that parses a string at run-time to produce a regex.
+This approach works for simple expressions like the one above -- we can write \li{parse_rx "A|T|G|C"}.
+However, this approach breaks down when one needs to construct regexes compositionally.
+For example, the DNA pattern that the BisI restriction enzyme recognizes is \li{AT}$x$\li{GC}, where $x$ is any of the four DNA bases.
+However, one cannot pass \li{any_base} to the string concatenation operator, because \li{any_base} is already of type \li{Regex.t}.
+These kinds of situations cause programmers to engage in \emph{stringly-typed programming}, as discussed in Sec.~\ref{sec:intro} -- they simply use strings to encode regexes throughout their application, rather than structured encodings like \li{Regex.t}.
+This can be catastrophic if the programmer is not extremely careful with string concatenation because strings that happen to contain special characters recognized by the POSIX notation can be misinterpreted because they are not distinguished in any way from strings that intentionally use POSIX notation.
+When some of these strings come from untrusted sources, e.g.
+web forms as is common in bioinformatics applications, the result can be a breach in privacy.
+
+% Moreover, there already exists a broadly adopted shorthand notational standard -- the POSIX standard for regular expressions \cite{STD95954}.
+For the purposes of this section, we will stick to the subset of the POSIX standard that applies to the encoding in Fig.~\ref{fig:Regex-module-def}, but there is nothing that would fundamentally prevent us from implementing the full POSIX standard using TLMs.
 \begin{figure}[t]
@@
 \end{figure}
-
-\paragraph{Outline} 
-% \begin{itemize}
-% \item 
-Sec.~\ref{sec:setsms} introduces expression TLMs with a detailed case study. 
-% \item 
-Sec.~\ref{sec:sptsms} then introduces {pattern TLMs}, i.e.
-TLMs that generate patterns, and describes the special reasoning conditions in pattern position.
-%The hygiene condition for pattern TLMs is interesting.
-% \item 
-Sec.~\ref{sec:ptsms} then introduces the more general \emph{parametric TLMs}, i.e.
-TLMs that can take type and module parameters, which allow us to define literal notation over a type- or module-parameterized family of types.
-This section also reveals that the \li{expansions require} clause is but syntactic sugar for parameterization followed immediately by partial parameter application.
-Having introduced the basic machinery, we continue in Sec.~\ref{sec:more-examples} with examples that further demonstrate the expressive power of this approach and various parser implementation strategies that notation providers can consider.
-This notably includes our take on the example from  Fig.~\ref{fig:urweb} of Ur/Web-style HTML literals.
-Next, Sec.~\ref{sec:setsms-formally} defines a type-theoretic calculus of simple expression and pattern TLMs and formally establishes the reasoning principles implied above in their simplest form. 
-Sec.~\ref{sec:ptlms-formally} adds type functions and an ML-style module system to this calculus, and gives a more general variant of the reasoning principles theorem. 
-% \item 
-Sec.~\ref{sec:existing-approaches} compares TLMs to related work, guided by the rubric of five reasoning principles just discussed. 
-% \item 
-Sec.~\ref{sec:discussion} concludes with a summary of the contributions of this paper and a discussion of future work.
-Certain technical details, proofs and extensions to the calculi presented in this paper are given in the supplement.
-% Also available is an additional series of case studies: notation for an encoding of regular expressions as an abstract data type, implemented using TLMs for quasiquotation and grammar-based parser generators.
-%are also available in the supplement.
-Sec.~\ref{sec:static-eval} considers the topic of term evaluation during the  expansion phase, i.e.
-\emph{static evaluation}, in  more detail.
-Sec \ref{sec:static-eval} also gives examples of TLMs that are useful for defining other TLMs, e.g.
-TLMs that implement parser generators and quasiquotation.
-
-%  We say more about LISP macros in Sec.
-\ref{sec:existing-approaches}.% For the present purposes, we can consider Reason as essentially the calculus defined in the supplement extended with various conveniences that are commonly found in other ML-like languages and, notionally, orthogonal to TLMs.
-%Reason is, as its name suggests, a conceptual descendent of ML.
-It diverges from other dialects of ML that have a similar type structure in that it has a bidirectional type system \cite{Pierce:2000:LTI:345099.345100} (like, for example, Scala \cite{OdeZenZen01}) for reasons that have to do with the mechanism of TLM implicits described in Chapters \ref{chap:tsls} and \ref{chap:ptsms}.
-%The reason we will not follow Standard ML \cite{mthm97-for-dart} in giving a complete formal definition of Reason in this work is both to emphasize that the primitives we introduce are ``insensitive'' to the details of the underlying type structure of the language (so TLMs can be considered for inclusion in a variety of languages, not only dialects of ML), and to avoid distracting the reader (and the author) with definitions that are already well-understood in the literature and that are orthogonal to those that are the focus of this work.
-% We will not formally define these features mainly to avoid unnecessarily complicating our presentation with details that are not essential to the ideas presented herein. As such, 
-% All examples written in Reason should be understood to be informal motivating material for the subsequent formal material.
-%We anticipate that future full-scale language specifications will be able to combine the ideas  in the proposed work without trouble.
-%The purpose of the work being proposed is to serve as a reference for those interested in the new constructs we introduce, not to serve as a language specification.
-%We will give a brief overview of these languages are organized in Sec. \ref{sec:Reason}.
-
-% \subsection{Contributions}\label{sec:contributions}
-% This work introduces a system of \textbf{typed literal macros (TLMs)} that gives library providers substantially more syntactic control than existing typed term-rewriting macro systems while maintaining the ability to reason abstractly about types, binding and segmentation.% abstract reasoning principles.
-% comparable to the level of control they have when defining a syntax dialect.
-
-% Clients apply TLMs to \emph{generalized literal forms}.
-
-% Because the context-free syntax is never extended, syntactic conflicts are not a concern.
-
-%As such, the semantics can take the type and binding structure of the surrounding program into account when validating the expansion that the TLM programmatically generates to ensure that clients can answer critical questions related to types and binding, like those enumerated in Section \ref{sec:abs-reasoning-intro}.
-Clients need not have knowledge of the implementation of the TLM or of the generated expansion, i.e.
-there are useful principles of syntactic abstraction.
-
-% The primary technical challenge has to do with the fact that the applied TLM needs to be able to splice terms out of the literal body for inclusion in the expansion. For example,
-
- % We design our mechanism such that these locations can easily be determined from the output of the TLM.
-This is essential for our hygiene mechanism, and it is also useful in that this information can be presented to the user (e.g.
-as shown in Figure \ref{fig:first-tsm-example-marked}).
-%As such, we must develop a mechanism where 1) the positions of spliced subterms can be determined without examining the macro implementation (e.g.
-so that they can be presented to the user differently by an editor or pretty-printer, ;  and 2) the hygiene mechanism must give only portions of the expansion that correspond to these spliced subterms access to the application site context.
-
-%In order to reason about types and binding, client programmers need only have knowledge of 1) the segmentation (e.g.
-by examining a figure like this presented by a code editor or pretty-printer) and 2) a type annotation on the definition of the applied TLM.
-No other details about the applied TLM's implementation or the expansion that it generates need to be revealed to the client programmer.
-In other words,
-
-% \begin{figure}[h]
-% \begin{lstlisting}
-% PElement Nil Seq(
-%   TextNode "Hello, ", 
-%   Seq(TextNode (join(" ", Cons(first, Cons(second, Nil)))), 
-%   TextNode "!"))
-% \end{lstlisting}
-% \caption{The desugaring.}
-% \end{figure}
-
-% There is also no ambiguity with regard to which TLM has control over each form, and searching for the definition of a TLM is no more difficult than searching for any other binding, i.e.
-there are well-defined scoping rules.
-
-% In other words, TLMs maintain a useful notion of syntactic abstraction.
-%More specifically, TLMs maintain a \emph{hygienic binding discipline}, meaning that questions Questions 4 and 5 above were concerned with are disallowed entirely.
-% We will, of course, make this notion more technically precise as we continue.
-
-% \begin{figure}[h]
-% \begin{lstlisting}[numbers=none,xleftmargin=0px]
-% let syntax $strlist = $list string in 
-% $html `SURL<p>Hello, {[EURLjoin ($str ' ') ($strlist [firstSURL,EURL last])SURL]}</p>EURL`
-% \end{lstlisting} 
-% \caption{The example from Figure \ref{fig:first-tsm-example-marked}, expressed using a parametric TLM.}
-% \label{fig:first-ptsm-example-marked}
-% \end{figure}
-
-% In Secs.
-\ref{chap:uetsms} and \ref{chap:uptsms}, we assume for the sake of technical simplicity that each TLM definition is self-contained, needing no access to libraries or to other TLMs.
-This is an impractical assumption in practice.
-We relax this assumption in In
-
-% %\item \textbf{Type-specific languages}, or \textbf{TSLs}. TSLs, described 
-% In Chapter \ref{chap:tsls}, we develop a mechanism of \emph{TLM implicits} that allows library clients to contextually designate, for any type, a privileged TLM at that type.
-The semantics applies this privileged TLM implicitly to unadorned literal forms that appear where a term of the associated type is expected.
-For example, if we designate
-% %\li{$str} 
-% as the privileged TLM at the \li{string} type and 
-% %\li{$strlist}
-% as the privileged TLM at the \li{list(string)} type, we can express the example from Figure \ref{fig:first-tsm-example-marked} instead as shown in Figure \ref{fig:first-tsm-example-implicit} (assuming \li{join} has type \li{string -> list(string) -> string}.)
-% \begin{figure}[h]
-% \begin{lstlisting}[numbers=none]
-% $html`SURL<p>Hello, {[EURLjoin ' ' [firstSURL,EURL last]SURL]}</p>EURL`
-% \end{lstlisting}
-% \caption{The example from Figure \ref{fig:first-tsm-example-marked} drawn to take advantage of TLM implicits.}
-% \label{fig:first-tsm-example-implicit}
-% \end{figure}
-
-% \noindent This approach is competitive in cost with library-specific syntax dialects (e.g.
-compare Figure \ref{fig:first-tsm-example-implicit} to Figure \ref{fig:urweb}), while maintaining the abstract reasoning principles characteristic of our approach.
-%\item \textbf{Metamodules}, introduced in Sec.
-\ref{sec:metamodules}, reduce the need to primitively build in the type structure of constructs like records (and variants thereof),  labeled sums and other interesting constructs that we will introduce later by giving library providers programmatic ``hooks'' directly into the semantics, which are specified as a \emph{type-directed translation semantics} targeting a small \emph{typed internal language} (introduced in Sec.
-\ref{sec:Reason}).
-%For example, a library provider can implement the type structure of records with a metamodule that:
-%\begin{enumerate}
-%\item introduces a type constructor, \lstinline{record}, parameterized by finite mappings from labels to types, and defines, programmatically, a translation to unary and binary product types (which are built in to the internal language); and
-%\item introduces operators used to work with records, minimally record introduction and elimination (but perhaps also various functional update operators), and directly implements the logic governing their typechecking and translation to the IL (which builds in only nullary and binary products).
-%\end{enumerate}
-%We will see direct analogies between ML-style modules (which our mechanisms also support) and metamodules later.
-%\end{enumerate}
-
-% As vehicles for this work, we will define a small programming language in each of the three parts just mentioned, each building conceptually upon the previous language.
-All of our formal contributions are relative to these small languages.
-
-%TLMs, like other macro systems, perform \emph{static code generation} (also sometimes called \emph{static} or \emph{compile-time metaprogramming}), meaning that the relevant rules in the static semantics of the language call for the evaluation of \emph{static functions} that generate term encodings.
-Static functions are functions that are evaluated statically, i.e.
-during typing.
-%Library providers write these static functions using the Reason \emph{static language} (SL).
-%Maintaining a separation between the static (or ``compile-time'') phase and the dynamic (or ``run-time'') phase is an important facet of Reason's design.
-% static code generation.
-%We will  also introduce a simple variant of each of these primitives that leverages Reason's support for local type inference to further reduce syntactic cost in certain common situations.
-
-\vspace{-3px}
-\newcommand{\seTLMsSec}{Expression TLMs}
-\section{\protect\seTLMsSec}
-\label{sec:setsms}
-\vspace{-2px}
-
-Consider the recursive datatype \li{Regex.t} defined by the module in Fig.~\ref{fig:Regex-module-def}, which encodes \emph{regular expressions} (regexes) into Reason \cite{Thompson:1968:PTR:363347.363387}.
-Regular expressions are  common in fields like bioinformatics, where they are used to express patterns in DNA sequences.
-For example, we can construct a regex that matches the strings \li{"A"}, \li{"T"}, \li{"G"} or \li{"C"}, which represent the four bases in DNA, as follows:
-\begin{lstlisting}[numbers=none]
-  let any_base = Regex.(Or(Str "A", Or(Str "T", Or(Str "G", Or(Str "C")))))
-\end{lstlisting}
-In Reason, the notation \li{Regex.(}$e$\li{)} locally opens the module \li{Regex} for use within the expression $e$, so we do not need to qualify the constructors \li{Regex.Or} and \li{Regex.Str}.
-Even allowing for this shorthand, however, this notation for constructing regexes is rather unwieldy.
-Instead, we would like to be able to use the common POSIX-style notation for constructing values of type \li{Regex.t} \cite{STD95954}.
-
-% One approach would be to simply define a function \li{parse_rx : string -> Regex.t} that parses a string at run-time to produce a regex.
-This approach works for simple expressions like the one above -- we can write \li{parse_rx "A|T|G|C"}.
-However, this approach breaks down when one needs to construct regexes compositionally.
-For example, the DNA pattern that the BisI restriction enzyme recognizes is \li{AT}$x$\li{GC}, where $x$ is any of the four DNA bases.
-However, one cannot pass \li{any_base} to the string concatenation operator, because \li{any_base} is already of type \li{Regex.t}.
-These kinds of situations cause programmers to engage in \emph{stringly-typed programming}, as discussed in Sec.~\ref{sec:intro} -- they simply use strings to encode regexes throughout their application, rather than structured encodings like \li{Regex.t}.
-This can be catastrophic if the programmer is not extremely careful with string concatenation because strings that happen to contain special characters recognized by the POSIX notation can be misinterpreted because they are not distinguished in any way from strings that intentionally use POSIX notation.
-When some of these strings come from untrusted sources, e.g.
-web forms as is common in bioinformatics applications, the result can be a breach in privacy.
-
-% Moreover, there already exists a broadly adopted shorthand notational standard -- the POSIX standard for regular expressions \cite{STD95954}.
-For the purposes of this section, we will stick to the subset of the POSIX standard that applies to the encoding in Fig.~\ref{fig:Regex-module-def}, but there is nothing that would fundamentally prevent us from implementing the full POSIX standard using TLMs.
 
@@
 This section will present a calculus of simple expression and pattern TLMs called $\miniVersePat$.
-By the end of this section, we will have a theorem that encodes the five reasoning principles that were outlined informally in the previous sections.
+By the end of this section, we will have a theorem that encodes the six reasoning principles that were outlined informally in the previous sections.
 
@@
 \label{sec:existing-approaches}
+
+Unhygienic term rewriting systems like OCaml's PPX system and Template Haskell can be used to rewrite string literals, but these systems also do no adequately address any of these reasoning criteria.
+Hygienic macro systems, like those available in various Lisp-family languages and in Scala, are also unsuitable because the hygiene constraints prevent the macros from surfacing spliced expressions from literal bodies.
+The system of \emph{type-specific languages} (TSLs) developed in prior work by \citet{TSLs} has made perhaps the most headway toward these ideals, but it too does not adequately maintain these reasoning criteria.
+Moreover, it is not clear how to integrate this approach, which was developed for a simple monomorphic language with nominal types and local type inference, into ML, which supports structural types, parameterized types, abstract types, modules, pattern matching and non-local type inference.
+We will say more about these existing approaches in Sec.
+\ref{sec:existing-approaches}.
+For now, it suffices to say that after evaluating existing approaches, we found that none of them were suitable for integration into Reason.
+
+An approach that has made some headway toward this ideal is Omar et al.'s \emph{type-specific languages} (TSLs) \cite{TSLs}.
+A language that supports TSLs delegates control over the parsing and expansion of certain \emph{generalized literal forms} to a parser associated with the type that the form is being checked against.
+So if we add TSLs to our language, our example then requires a type annotation on \li{y}:
+\begin{lstlisting}[numbers=none,basicstyle=\ttfamily\fontsize{9pt}{1em}\selectfont]
+  let y : KQuery.t = `((!R)@&{&/x!/:2_!x}'!R)`
+\end{lstlisting}
+The parser associated with the type \li{KQuery.t} is statically invoked to lex, parse and expand the literal body, i.e.
+the sequence of characters between the outer delimiters, here \li{`(} and \li{)`}.
+The literal body is constrained by the context-free grammar of the language only in that nested delimiters must be matched (like comments in OCaml).
+TSLs are closely related to the mechanism that we will propose, so let us analyze TSLs from the perspective of the six reasoning criteria just outlined.
+\begin{enumerate}[leftmargin=12px]
+  \item[\LEFTcircle] \textbf{Responsibility}: The type-directed dispatch mechanism allows the client to easily determine which parser is responsible for each literal form: the parser that was associated with the type when it was defined.
+Clients do not need to worry about different TSLs conflicting syntactically because the context-free grammar of the language remains fixed and composition is mediated by splicing.
+The main limitation here is that it is impossible to define literal notation after a type has been defined, and it is also impossible to define multiple notations at a single type.
+  \item[\LEFTcircle] \textbf{Segmentation}: The parser can splice expressions out of the literal body, but there is no clear way to associate each spliced term with some particular segment (i.e.
+subsequence) of the literal body, nor is there a guarantee that spliced terms are non-overlapping.
+As such, there is no way to indicate to the programmer where base language expressions are located within a literal form.
+  \item[\LEFTcircle] \textbf{Capture}: The TSL mechanism as described enforces complete capture avoidance (though the formal specification given in the paper did not adequately enforce this constraint).
+  \item[\LEFTcircle] \textbf{Context Dependence}: The TSL mechanism as described requires that the expansion be completely closed, so it is trivially context independent.
+However, this comes at a significant expressive cost: the generated expansions cannot make use of any libraries whatsoever.
+  \item[\LEFTcircle] \textbf{Typing}: The expansion must necessarily be of the associated type, so abstract reasoning about the type of the expansion is straightforward.
+However, there is no easy way to determine the type expected for each spliced expression.
+In addition, the prior work considered only a monomorphic, nominally-typed language with local type inference, leaving open a number of problems that came up as we considered integrating TSLs into Reason/OCaml:
+\begin{enumerate}[leftmargin=15px,nolistsep,noitemsep]
+  \item \textbf{Structural Types}: There is no way to define literal notation at structural types, e.g. tuple and arrow types, because there is no ``definition site'' for such types.
+  \item \textbf{Parameterized \& Abstract Types}: There is no way to define literal notation over a type-parameterized family of types, e.g.
+at all types \li{list('a)}.
+Similarly, there is no way to define literal notation over a module-parameterized family of abstract types, e.g.
+at every abstract type defined by a module implementing the \li{QUEUE} signature.
+Parameterized and abstract type families are ubiquitous in ML-family languages.% Programmers in ML-family languages commonly define various implementations of abstract data types in this way.
+  \item \textbf{Pattern Literals}: Pattern matching is ubiquitous in ML-family languages but the prior work on TSLs considered only expression literals.
+%Pattern literals are dual to expression literals in that expression literals support value construction, while pattern literals support value deconstruction.
+  \item \textbf{ML-Style Type Inference}: It is not clear that a type-directed dispatch scheme can be cleanly  reconciled with ML-style type inference.
+\end{enumerate}
+\end{enumerate}
+
 \subsection{Syntax Definition Systems}\label{sec:syntax-dialects-intro}
@@
   \item $
-    \segof{\ce} = \sseq{\acesplicedk{m_i}{n_i}}{\nkind} \cup \sseq{\acesplicedc{m'_i}{n'_i}{\cekappa'_i}}{\ncon} \cup $ \\
+    \segofexisitng-{\ce} = \sseq{\acesplicedk{m_i}{n_i}}{\nkind} \cup \sseq{\acesplicedc{m'_i}{n'_i}{\cekappa'_i}}{\ncon} \cup $ \\
      ~~~~$          \sseq{\acesplicede{m''_i}{n''_i}{\ctau_i}}{\nexp}

### 2018-03-07 icfp18/syntax-icfp18.tex
more tweaks to introgit push!

@@
 Only existing composite forms can therefore be repurposed.
-\item An approach that came close was Omar et al.'s \emph{type-specific languages} (TSLs) \cite{TSLs}, but it fell short with regard to the six reasoning principles just discussed in ways that we will return to and in any case, the mechanism was designed for a simple monomorphic, nominally-typed language and relied critically on a particular local type inference scheme.
-It is not suitable for a language with an ML-like semantics, meaning a language with support for structural types (like tuple types and arrow types in ML), parameterized types, abstract types, modules, pattern matching and non-local type inference.
+\item \emph{Type-specific languages} (TSLs) \cite{TSLs} come closer to our goals in that they explicitly support splicing terms out of literal bodies, but the mechanism falls short with regard to the six reasoning principles just discussed in ways that we will return to in Sec.
+\ref{sec:existing-approaches}.
+In any case, this approach was designed for simple nominally-typed languages and relies critically on a particular local type inference scheme.
+It is not suitable for a language with an ML-like semantics, meaning a language with support for structural types (like tuple and function types in ML), parameterized types, abstract types, modules, pattern matching and non-local type inference.
 \end{itemize}
@@
 
-\paragraph{Brief Overview} To finish up our discussion of the K query example above, let us give a brief overview of the TLM-based solution outlined in Figure \ref{fig:K-tsm-example}.
-On lines 1-5, we define a TLM named \li{$kq}.
-On Line 9, we apply this TLM.
+\paragraph{Brief Overview} For a brief overview of our proposed mechanism, consider Figure \ref{fig:K-tsm-example}.
+On Lines 1-5, we define a TLM named \li{$kq} that provides K query literal notation.
+On Line 9, we apply this TLM to express the example from Figure \ref{fig:K-dialect}.
 We can reason abstractly about this program as follows:
-\begin{enumerate}[leftmargin=15pt]
+\begin{enumerate}[leftmargin=12pt]
 \item \textbf{Responsibility}: The lexer and parser specified by the applied TLM on Lines 2-3 are exclusively responsible for lexing, parsing and expanding the body of the generalized literal form, i.e.
 the characters between \li{`(} and \li{)`}.
-We will give more details on generalized literal forms, and on constructing a TLM lexer and parser in the next section.
-For now, it suffices to say that our design goal is to provide a mechanism where the programmer does not normally need to look up the definitions of \li{KQueryLexer} and \li{KQueryParser} to reason about types and binding.
+We will give more details on generalized literal forms and on constructing a TLM lexer and parser in the next section.
+For now, let us simply reiterate that our design goal is to provide a system where the programmer does not normally need to look up the definitions of \li{KQueryLexer} and \li{KQueryParser} to reason about types and binding.
 % To briefly summarize, our system uses Menhir, which is the most popular parser generator in the OCaml ecosystem and integrated into its standard build tools.
@@
 \item \textbf{Expansion Typing}: The type annotation on the definition of \li{$kq} (Line 1) determines the type that the expansion must have, here \li{KQuery.t}.
-\item \textbf{Context Dependence}: The external dependencies of the generated expansion are explicitly specified.
-Here, Lines 4-5 specify that expansions generated by \li{$kq} require the module \li{KQuery}.
+\item \textbf{Context Dependence}: Lines 4-5 specify that expansions generated by \li{$kq} are allowed to use the module \li{KQuery}, and no others.
 The system ensures that this dependency is  bound as specified even if the identifier \li{KQuery} has been shadowed at the application site.
-By making dependencies explicit, we can be sure that the expansion does not rely on the fact that, e.g., \li{w} is in scope at the application site.
+This completely relieves clients from needing to consider expansion-internal dependencies when naming variables.
+%For example, if the client renames \li{w}, there is not even a remote possibility that this could break or change the meaning of Line 9.
 \item \textbf{Segmentation}: The intermediate output that the TLM generates is structured so that the system can infer from it an accurate \emph{segmentation} of the literal body that distinguishes spliced terms, which appear directly in the expansion, from segments that are parsed in some other way by the TLM.
-The segmentation is all that needs to be communicated to editor services downstream of the expander, and ultimately to the programmer using secondary notation, e.g.
+The segmentation is all that needs to be communicated to language services downstream of the expander, e.g.
+editors and pretty printers, which can communicate this information to the programmer using secondary notation, e.g.
 colors in this document.
@@
 This, together with the context independence property, ensures that type inference at the TLM application site can be performed (by editor services or in the programmer's mind) abstractly, i.e.
-by requesting just the type annotations on the spliced segments.
+by reference only to the type annotations on the spliced segments, not the full expansion.
 %This information is usually not necessary to reason about typing, but it can be conveyed to the programmer upon request by the program editor if desired.
@@
 
-Consider the recursive datatype \li{Regex.t} defined by the module in Fig.~\ref{fig:Regex-module-def}, which encodes \emph{regular expressions} (regexes) into Reason \cite{Thompson:1968:PTR:363347.363387}.
-Regular expressions are  common in fields like bioinformatics, where they are used to express patterns in DNA sequences.
+Consider the recursive datatype \li{Regex.t} defined by the module in Fig.~\ref{fig:Regex-module-def}, which encodes {regular expressions} (regexes) into Reason \cite{Thompson:1968:PTR:363347.363387}.
+Regexes are  common in, for example, bioinformatics, where they are used to express patterns in DNA sequences.
 For example, we can construct a regex that matches the strings \li{"A"}, \li{"T"}, \li{"G"} or \li{"C"}, which represent the four bases in DNA, as follows:
@@
 \end{lstlisting}
-In Reason, the notation \li{Regex.(}$e$\li{)} locally opens the module \li{Regex} for use within the expression $e$, so we do not need to qualify the constructors \li{Regex.Or} and \li{Regex.Str}.
-Even allowing for this shorthand, however, this notation for constructing regexes is rather unwieldy.
+The notation \li{Regex.(}$e$\li{)} locally opens the module \li{Regex} for use within the expression $e$, so we do not need to qualify each application of \li{Regex.Or} and \li{Regex.Str}.
+Even with this shorthand, however, this notation for constructing regexes is rather unwieldy.
 Instead, we would like to be able to use the common POSIX-style notation for constructing values of type \li{Regex.t} \cite{STD95954}.

### 2018-03-07 icfp18/syntax-icfp18.tex
revisions to page 5

@@
 Consider the recursive datatype \li{Regex.t} defined by the module in Fig.~\ref{fig:Regex-module-def}, which encodes {regular expressions} (regexes) into Reason \cite{Thompson:1968:PTR:363347.363387}.
-Regexes are  common in, for example, bioinformatics, where they are used to express patterns in DNA sequences.
+Regexes are  common in, for instance, bioinformatics, where they are used to express patterns in DNA sequences.
 For example, we can construct a regex that matches the strings \li{"A"}, \li{"T"}, \li{"G"} or \li{"C"}, which represent the four bases in DNA, as follows:
@@
 Even with this shorthand, however, this notation for constructing regexes is rather unwieldy.
-Instead, we would like to be able to use the common POSIX-style notation for constructing values of type \li{Regex.t} \cite{STD95954}.
+Instead, we would like to be able to use the common POSIX-style notation  \cite{STD95954} for constructing values of type \li{Regex.t}, including values constructed compositionally from previously constructed regexes and strings.
 
@@
 \subsection{TLM Definition and Application}
-Figure \ref{fig:regex-tlm-def} defines a TLM named \li{$regex} for values of type \li{Regex.t} that supports a POSIX-style literal notation extended with splice forms for regexes and strings.
+We solve this problem in Figure \ref{fig:regex-tlm-def} by defining a TLM named \li{$regex} (pronounced ``lit regex'') that supports POSIX-style regex notation extended with splice forms for regexes and strings.
+We will consider the provider perspective, i.e.
+the details of \li{RegexLexer} and \li{RegexParser}, in Sec.
+\ref{sec:provider-perspective}, but the focus of this paper is on the client perspective, with an emphasis on holding the details of the lexer and parser abstract, so let us start instead with some examples of \li{$regex} being applied.
+
 Line 1 of Fig.
-\ref{fig:first-tlm-example} applies this TLM to construct the regex \li{any_base} just described.
+\ref{fig:first-tlm-example} applies \li{$regex} to construct the regex \li{any_base} just described.
 Line 2 of Fig.
 \ref{fig:first-tlm-example} then applies \li{$regex} again, using its regex splice form to construct a regex \li{bisA} matching the DNA sequences recognized by the BisA restriction enzyme, where the middle base can be any base.
-Finally, Lines 3-4 of Fig.
-\ref{fig:first-tlm-example} define a function that constructs a more complex regex from these two regexes and a given gene sequence represented as a string, which is spliced in using a string splice form delimited by \li{$$(} and \li{)}.
+Finally, Lines~3-4 of Fig.
+\ref{fig:first-tlm-example} define a function that constructs a more complex regex from these two regexes and a given gene sequence represented as a string, which is spliced in using the string splice form.
+Notice that the TLM provider has chosen to use \li{$(}$e$\li{)} for regex splicing and \li{$$(}$e$\li{)} for string splicing, where $e$ is an arbitrary Reason expression.
 
@@
 \end{lstlisting}
-According to the context-free grammar of the language, this form is simply a leaf of the unexpanded syntax tree, like a string literal would be. TLM names are prefixed by \li{$} 
+According to the context-free grammar of (this paper's dialect of) Reason, this form is simply a leaf of the unexpanded syntax tree, like a string literal would be. TLM names are prefixed by \li{$} 
  to distinguish them from variables.
-There are no lexical constraints on the body of the generalized literal form except that any occurrences of \li{`(} must be balanced by \li{)`}, much like nested comments in OCaml, Reason and other languages.
-Generalized literal forms, which first arose in the prior work on TSLs \cite{TSLs}, therefore syntactically subsume other literal forms.
+We call \li{`(GC$(any_base)GC)`} a generalized literal form, following the terminology introduced in the prior work on type-specific languages \cite{TSLs}.
+There are no lexical constraints on the body of a generalized literal form except that any occurrences of \li{`(} must be balanced by \li{)`}, much like nested comments in Reason/OCaml.
+Generalized literal forms therefore syntactically subsume other literal forms.
+The nesting constraint is to support TLM applications that appear inside spliced expressions (discussed below).
 The prior work specified several other choices of outer delimitation, including layout-sensitive delimitation, but for the purposes of this work, we will use \li{`(} and \li{)`} exclusively.
-The fact that the context-free grammar of the base language is fixed ensures that notation providers can reason modularly about syntactic ambiguity in their own grammars.
 
@@
 
+\subsection{Provider Perspective}
+\label{sec:provider-perspective}
+
+The fact that the context-free grammar of the base language is fixed ensures that notation providers can reason modularly about syntactic ambiguity in their own grammars.
 % \section{Simple Expression TLMs By Example}\label{sec:tsms-by-example}

### 2018-03-08 icfp18/syntax-icfp18.tex
near-final revisions up through context dependence section

--- 2018-03-07 icfp18/syntax-icfp18.tex
@@
 Instead, there has been persistent interest in mechanisms that allow library providers to define new literal notation on their own.
-For example, direct grammar extension systems like Camlp4 \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, macro-based systems like Template Haskell~\cite{SheardPeytonJones:Haskell-02,mainland2007s} and the \emph{type-specific languages} (TSLs) of \citet{TSLs}, and other systems that we will discuss below can all be used to define new literal notation (and, in some cases, other forms of new notation, such as new infix expression forms, control flow operations or type declaration forms, which we leave beyond the scope of this paper).
+For example, direct grammar extension systems like Camlp4 \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, macro-based systems like Template Haskell~\cite{SheardPeytonJones:Haskell-02,mainland2007s}, the system of \emph{type-specific languages} (TSLs) described by \citet{TSLs}, and other systems that we will discuss below can all be used to define new literal notation (and, in some cases, other forms of new notation, such as new infix expression forms, control flow operations or type declaration forms, which we leave beyond the scope of this paper).
 
@@
 \item \textbf{Context Dependence}: Which bindings does the expansion of Line~9 invisibly depend on?
-If I shadow or remove a module or other binding, could that break or change the meaning of Line~9 because its expansion depended invisibly on the original binding?
+If I shadow or remove a module or other binding, could that break or change the meaning of Line~9 because its expansion depends invisibly on the original binding?
 \item \textbf{Segmentation}: Are the characters \li{x}, \li{R} and \li{2} on Line~9 parsed as spliced terms, meaning that they appear directly in the underlying expansion, or are they parsed in some other way peculiar to this literal notation (e.g.
@@
 \begin{itemize}[leftmargin=15px]
-\item Unhygienic approaches, like OCaml's preprocessor extension point (PPX) rewriters \cite{ocaml-manual} and Template Haskell \cite{SheardPeytonJones:Haskell-02,mainland2007s} allow us to define new literal notation with support for splicing by repurposing existing string literal forms.
+\item Unhygienic approaches like OCaml's preprocessor extension point (PPX) rewriters \cite{ocaml-manual} and Template Haskell \cite{SheardPeytonJones:Haskell-02,mainland2007s} allow us to define new literal notation with support for splicing by repurposing existing string literal forms.
 They also partially or completely solve the problem of reasoning about \textbf{Responsibility} by using explicit annotations on terms that are being rewritten.
@@
 the rewriting is parametric in the arguments, so they are not suitable for the problem of defining new literal notation.
-\item Hygienic term rewriting macro systems, like those in various Lisp-family languages \cite{DBLP:conf/esop/HermanW08,Herman10:Theory,Kohlbecker86a,DBLP:conf/popl/Adams15,DBLP:conf/popl/ClingerR91,DBLP:journals/lisp/DybvigHB92,mccarthy1978history,Flatt:2012:CLR:2063176.2063195} and Scala \cite{ScalaMacros2013}, do not allow us to repurpose string literal forms to define literal forms at other types because the hygiene discipline does not allow us to splice terms out of literal bodies via parsing.
+\item Hygienic term rewriting macro systems, like those in various Lisp-family languages \cite{DBLP:conf/esop/HermanW08,Herman10:Theory,Kohlbecker86a,DBLP:conf/popl/Adams15,DBLP:conf/popl/ClingerR91,DBLP:journals/lisp/DybvigHB92,mccarthy1978history,Flatt:2012:CLR:2063176.2063195} (e.g.
+Racket \cite{Flatt:2012:CLR:2063176.2063195}) and Scala \cite{ScalaMacros2013}, do not allow us to repurpose string literal forms to define literal forms at composite types because the hygiene discipline does not allow us to splice terms out of literal bodies via parsing.
 Only existing composite forms can therefore be repurposed.
@@
 
-\paragraph{Contributions} This paper introduces \emph{typed literal macros} (TLMs): the first system for defining new literal notation, of nearly arbitrary design, that (1) provides the ability to reason abstractly about all six of the topics just outlined; and (2) is semantically expressive enough for integration into Reason/OCaml and other full-scale statically typed functional languages.
-We evaluate these claims with (1) a series of increasingly ML-like calculi equipped with proofs of the claimed abstract reasoning principles; and (2) a number of non-trivial examples that appear throughout the paper and involve the language features mentioned above.
-In describing these examples, we also demonstrate that literal parsing logic can be defined using standard, unmodified parser generators, so the burden on notation providers is comparable to that of existing systems despite these stronger client-side reasoning principles.
+\paragraph{Contributions} This paper introduces \emph{typed literal macros} (TLMs): the first system for defining new literal notation that (1) provides the ability to reason abstractly about all six of the topics just outlined; and (2) is semantically expressive enough for integration into Reason/OCaml and other full-scale statically typed functional languages.
+We evaluate these claims with (1) a series of increasingly ML-like calculi equipped with proofs of the claimed abstract reasoning principles; and (2) a number of non-trivial examples appearing throughout the paper that involve the advanced language features mentioned above.
+In describing these examples, we also demonstrate that literal parsing logic can be defined using standard, unmodified parser generators, so the burden on notation providers is comparable to that of existing systems despite these stronger reasoning principles.
 % From the client's perspective, which is the focus of our contributions, TLMs are uniquely reasonable.
@@
 On Line 9, we apply this TLM to express the example from Figure \ref{fig:K-dialect}.
-We can reason abstractly about this program as follows:
+We can reason abstractly about this program as follows.
 \begin{enumerate}[leftmargin=12pt]
@@
 \vspace{-3px}
-\newcommand{\seTLMsSec}{Expression TLMs}
+\newcommand{\seTLMsSec}{Expression TLMs in Reason}
 \section{\protect\seTLMsSec}
@@
 
-Consider the recursive datatype \li{Regex.t} defined by the module in Fig.~\ref{fig:Regex-module-def}, which encodes {regular expressions} (regexes) into Reason \cite{Thompson:1968:PTR:363347.363387}.
+Consider the recursive datatype \li{Regex.t} defined by Fig.~\ref{fig:Regex-module-def}, which encodes {regular expressions} (regexes) into Reason \cite{Thompson:1968:PTR:363347.363387}.
 Regexes are  common in, for example, bioinformatics, where they are used to express patterns in DNA sequences.
-For example, we can construct a regex that matches the strings \li{"A"}, \li{"T"}, \li{"G"} or \li{"C"}, which represent the four bases in DNA, as follows:
+We can construct a regex that matches the strings \li{"A"}, \li{"T"}, \li{"G"} or \li{"C"}, which represent the four bases in DNA, as follows:
 \begin{lstlisting}[numbers=none]
@@
 \end{lstlisting}
-The notation \li{Regex.(}$e$\li{)} locally opens the module \li{Regex} for use within the expression $e$, so we do not need to qualify each application of \li{Regex.Or} and \li{Regex.Str}.
-Even with this shorthand, however, this notation for constructing regexes is rather unwieldy.
-Instead, we would like to be able to use the common POSIX-style notation for constructing values of type \li{Regex.t} \cite{STD95954}.
-
+In Reason, the notation \li{Regex.(}$e$\li{)} locally opens the module \li{Regex} within the expression $e$, so we did not need to qualify each application of \li{Regex.Or} and \li{Regex.Str} above.
+Even with this shorthand, however, constructing regexes in this way is syntactically costly.
+Instead, we would like to have the option to use the common POSIX-style notation  \cite{STD95954} when constructing values of type \li{Regex.t}, including values constructed compositionally from previously constructed regexes and strings.
+We solve this problem in Fig.
+\ref{fig:regex-tlm-def} by defining a TLM named \li{$regex} (pronounced ``lit regex'') that supports POSIX-style regex notation extended with splice forms for regexes and strings.
 % One approach would be to simply define a function \li{parse_rx : string -> Regex.t} that parses a string at run-time to produce a regex.
@@
 \begin{lstlisting}
-  let any_base = $regex `(SURLA|T|G|CEURL)`;
-  let bisA = $regex `(SURLGC$(EURLany_baseSURL)GCEURL)`;
-  let gene_restriction_template(gene) => 
-    $regex `(SURL$(EURLbisASURL)$(EURLany_baseSURL)*$$(EURLgeneSURL)$(EURLany_baseSURL)*$(EURLbisASURL)EURL)`
+  module DNA = { let any_base = $regex `(SURLA|T|G|CEURL)`; };
+  let bisA = $regex `(SURLGC$(EURLDNA.any_baseSURL)GCEURL)`;
+  let restriction_template(gene) => 
+    $regex `(SURL$(EURLbisASURL)$(EURLDNA.any_baseSURL)*$$(EURLgeneSURL)$(EURLDNA.any_baseSURL)*$(EURLbisASURL)EURL)`;
 \end{lstlisting}
-\vspace{-5px}
+\vspace{-2px}
 \caption{Examples of the \li{$regex} TLM being applied in a bioinformatics application.
@@
 \vspace{-3px}
-\subsection{TLM Definition and Application}
-Figure \ref{fig:regex-tlm-def} defines a TLM named \li{$regex} for values of type \li{Regex.t} that supports a POSIX-style literal notation extended with splice forms for regexes and strings.
+\subsection{Client Perspective}
+The focus of this paper is on the client programmer's perspective, so we start in this section with some examples of \li{$regex} being applied in Fig.
+\ref{fig:first-tlm-example}, then use these examples to detail the client-side abstract reasoning principles that the mechanism provides, all before finally considering the details of the lexer and parser in Sec.
+\ref{sec:provider-perspective}.
+% The \li{open notation} directive on Line 2 of Fig.
+\ref{fig:first-tlm-example} allow qualifying each TLM application in the file.
+%reasoning about this program from the perspective of a client who wants to avoid looking at the lexer, parser and the underlying expansion itself.
+
 Line 1 of Fig.
-\ref{fig:first-tlm-example} applies this TLM to construct the regex \li{any_base} just described.
-Line 2 of Fig.
-\ref{fig:first-tlm-example} then applies \li{$regex} again, using its regex splice form to construct a regex \li{bisA} matching the DNA sequences recognized by the BisA restriction enzyme, where the middle base can be any base.
-Finally, Lines 3-4 of Fig.
-\ref{fig:first-tlm-example} define a function that constructs a more complex regex from these two regexes and a given gene sequence represented as a string, which is spliced in using a string splice form delimited by \li{$$(} and \li{)}.
+\ref{fig:first-tlm-example} applies \li{$regex} to construct the regex \li{any_base} that was  described above, this time using POSIX regex notation.
+Line~2 applies \li{$regex} again, using its regex splice form to construct a regex matching DNA sequences recognized by the BisA restriction enzyme, where the middle base can be any base.
+Finally, Lines~3-4 of Fig.
+\ref{fig:first-tlm-example} define a function that constructs a more complex regex from these two regexes and a given gene sequence represented as a string, which is spliced in using the string splice form.
+Notice that the TLM provider has chosen to use \li{$(}$e$\li{)} for regex splicing and \li{$$(}$e$\li{)} for string splicing, where $e$ is an unexpanded Reason expression of arbitrary form.
 
@@
 \begin{lstlisting}[numbers=none]
-    $regex `(GC$(any_base)GC)`
+    $regex `(GC$(DNA.any_base)GC)`
 \end{lstlisting}
-According to the context-free grammar of the language, this form is simply a leaf of the unexpanded syntax tree, like a string literal would be. TLM names are prefixed by \li{$} 
+According to the context-free grammar of (our extended dialect of) Reason, this form is simply a leaf of the unexpanded syntax tree, like a string literal would be. TLM names are prefixed by \li{$} 
  to distinguish them from variables.
-There are no lexical constraints on the body of the generalized literal form except that any occurrences of \li{`(} must be balanced by \li{)`}, much like nested comments in OCaml, Reason and other languages.
-Generalized literal forms, which first arose in the prior work on TSLs \cite{TSLs}, therefore syntactically subsume other literal forms.
-The prior work specified several other choices of outer delimitation, including layout-sensitive delimitation, but for the purposes of this work, we will use \li{`(} and \li{)`} exclusively.
-The fact that the context-free grammar of the base language is fixed ensures that notation providers can reason modularly about syntactic ambiguity in their own grammars.
-
-Notice that in the unexpanded expression above, we did not distinguish the spliced expression from the rest of the literal.
+We call the TLM argument, \li{`(GC$(DNA.any_base)GC)`}, a \emph{generalized literal form}, following the terminology introduced in the prior work on type-specific languages \cite{TSLs}.
+The only lexical constraint on the body of a generalized literal form, i.e.
+the characters between \li{`(} and \li{)`}, is that any further occurrences of \li{`(} must be balanced by \li{)`}, much like nested comments in Reason/OCaml.
+Generalized literal forms therefore lexically subsume other literal forms.
+The nesting constraint is to allow TLM applications to appear inside spliced expressions (discussed below).
+The prior work specified several other choices of outer delimitation, including layout-sensitive delimitation, but for the purposes of this work, we will use the outer delimiters \li{`(} and \li{)`} exclusively.
+Choices about how spliced expressions are recognized inside the literal body are entirely at the discretion of each TLM, as we will discuss below.
+
+% Notice that in the unexpanded expression, we have not yet distinguished the spliced expression \li{any_base} from the rest of the literal.
 This is to emphasize that it is only during the subsequent \emph{expansion} phase that the body of the generalized literal form is lexed, parsed and expanded to produce an expanded syntax tree where spliced expressions, like \li{any_base}, have been revealed.
-Responsibility for lexing, parsing and expanding the literal body is delegated to the lexer and parser specified by the applied TLM by the clauses \li{lexer} and \li{parser}, respectively.
-
-\vspace{-4px}
-\subsection{Abstract Reasoning Principles} 
-Rather than looking immediately at Fig.
-\ref{fig:lexer-and-parser}, which defines the lexer and parser specified by \li{$regex}, let us first consider the perspective of a client programmer who does not want to delve into the details of lexing and parsing for each piece of literal notation that they encounter.
-What can this programmer deduce just by reasoning from the information presented in Figures \ref{fig:Regex-module-def} through \ref{fig:first-tlm-example}?
-
+
+% \vspace{-4px}
+% \subsection{Client Perspective: Abstract Reasoning Principles} 
+% The question that this paper posits is central when evaluating a notation definition system is: what must be revealed about the lexer, parser and the expansion itself for the client programmer to be able to answer the sorts of questions about types and binding that they would normally ask?
+
+\paragraph{Responsibility} 
+Responsibility for lexing, parsing and expanding each literal body is delegated uniquely to the lexer and parser specified in the definition of the applied TLM by the clauses \li{lexer} and \li{parser}, respectively.
+We will return to consider the details of lexer and parser definition, and show the definitions of the lexer and parser specified by \li{$regex}, in Sec.
+\ref{sec:provider-perspective}.
+For now, let us purposefully hold these details abstract.% and take the perspective of a client that does not want to delve into these details.
+
+In Reason, TLM definitions like the one in Fig.
+\ref{fig:regex-tlm-def} are separately defined in files with the suffix \li{.relit}, which are processed and brought into scope by the Reason build system, \li{rebuild}.
+This is consistent with the protocol that other syntax extension systems use, e.g.
+the syntax libraries in Sugar* \cite{erdweg2011sugarj,erdweg2013framework,erdweg2012layout}.
+In our theoretical treatment in Sec.~\ref{sec:setsms-formally}, however, TLM definitions are integrated directly into the scoping structure of the language.
+This is not possible in Reason because this would require extending OCaml's semantics, i.e.
+adding TLM definitions to module signatures.
+While we were careful to choose a design where this modification would be feasible in principle---TLM definition equality requires only decidable type and module path equality---modifying the semantics of OCaml is intentionally beyond the scope of the Reason project.
+In any case, what is fundamentally important is that there is a well-defined protocol for finding the definition of the TLM uniquely responsible for each application site.
+% via a ``go to definition'' service.
+
+% nitions (Fig.
+\ref{fig:regex-tlm-def}) appear in \li{.relit} files, and they are brought into scope by the build system, \li{rebuild}, according to the usual protocols for library discovery.
+The fully qualified name of the \li{--}% In the theory that we develop in Sections \ref{sec:s-UL}, TLM definitions occur directly within code.
+However, in a language with a module system, this would require adding TLM definitions to signatures as well.
 % TLM definitions follow well-defined scoping rules, so the client programmer can follow the binding structure of the language in the usual manner (manually or assisted by a ``go to definition'' editor service) to determine which TLM is uniquely responsible for each literal form.
@@
 
-From the definition of the \li{$regex} TLM, the client can immediately determine the type of the expansion being generated at each application site because it is specified explicitly by the clause \li{at Regex.t} on Line 1 of Fig.
+\paragraph{Expansion Typing}
+Having found the definition of the \li{$regex} TLM, the client can immediately determine the type of the expansion being generated at each application site because it is specified explicitly by the clause \li{at Regex.t} on Line 1 of Fig.
 \ref{fig:regex-tlm-def}.
-This is analagous to the return type of a function.
-The identity of \li{Regex.t} is determined relative to the TLM definition site, not at each application site (so the module \li{Regex} need not be in scope at the application site, or it can be shadowed by a different module).
-
-The TLM mechanism enforces an similarly strong context independence property on generated expansions by requiring that the TLM explicitly specify the modules that the expansions it generates might internally depend on.
+The expansion type of a TLM is analagous to the return type of a function.
+The identity of \li{Regex.t} is determined relative to the TLM definition site, not at each application site, so the module \li{Regex} need not be in scope at the application site, or it can have been shadowed by a different module.
+In Reason, TLM definitions can refer to exactly those modules that a Reason module at the top level of the defining package would have access to, i.e.
+other top-level modules in the package as well as the modules exported by other packages that have been imported by the build system.
+
+\paragraph{Context Dependence}
+The TLM mechanism enforces a strong context independence condition on generated expansions by requiring that the TLM definition explicitly specify all modules that the expansions it generates might internally depend on.
 In this case, Lines 4-5 of Fig.
-\ref{fig:regex-tlm-def} specify that generated expansion might use the module \li{Regex}, again as it is bound at the definition site, using the module variable \li{Regex} internally.
-These happen to coincide in this case, but in general, the dependency can be an arbitrary module path while internally it must be bound to a module variable, for example:
+\ref{fig:regex-tlm-def} specify that generated expansions might use the module \li{Regex}, again as it is bound at the TLM definition site, using the module variable \li{Regex} internally.
+The expansion-internal module variable happens to coincide with the module path in this case, but in general, the dependency can be an arbitrary module path while internally it must be bound to a module variable.
+For example, consider the following dependency clause:
 \begin{lstlisting}[numbers=none]
@@
     module Regex as Regex
-    module Core.List as L
+    module Core.List as List
 \end{lstlisting}
-The module variable \li{L} will be bound internally to \li{Core.List} in all expansions generated by this TLM, including those where \li{L} or \li{Core.List} are not bound, or otherwise bound.
+The module variable \li{List} will be bound internally to \li{Core.List} in all expansions generated by this TLM, including expansions generated at application sites where \li{List} or \li{Core.List} are not bound, or bound to different modules.
 All other bindings, whether at the TLM definition site or the TLM application site, are not internally available to the expansion.
-This allows client programmers to freely rename and shadow variables as they are used to, without needing to inspect each parser to determine whether it requires certain modules be bound to particular names at each use site (as is common with, e.g., syntax extensions based on \li{camlp4} or Sugar*, or when using unhygienic term rewriting approaches like OCaml's PPX system or Template Haskell \cite{SheardPeytonJones:Haskell-02}, as further discussed in Sec.
+This allows programmers to freely rename and shadow variables as they are used to, without needing to inspect each parser to determine whether the expansions it generates require certain variables be bound in particular ways at each application site (as is common with, e.g., syntax extensions based on \li{camlp4} or Sugar*, or when using unhygienic term rewriting approaches like OCaml's PPX system or Template Haskell \cite{SheardPeytonJones:Haskell-02}, which we discuss further in Sec.
 \ref{sec:existing-approaches}).
@@
 Enforcing this strong context independence condition is quite subtle because TLM parsers need to be able to extract spliced expressions from the literal body and place them in the final expansion.
-Na\"ively checking that the final expansion is closed except for the explicitly named dependencies would inappropriately enforce context independence on application site spliced expressions, which should certainly not be prevented from referring to variables in the application site context.
+Na\"ively checking that the final expansion be closed except for the explicitly named dependencies would inappropriately enforce context independence on application site spliced expressions, which should certainly not be prevented from referring to variables in the application site context.
 For example, consider the final expansion of the example from Line 2 of Fig.
@@
 \begin{lstlisting}[numbers=none]
-  Regex.Seq(Regex.Str "GC", Regex.Seq(any_base, Regex.Str "GC"))
+  Regex.Seq(Regex.Str "GC", Regex.Seq(DNA.any_base, Regex.Str "GC"))
 \end{lstlisting}
-In this term, both \li{Regex} and \li{any_base} are free variables. There is nothing to indicate which free variables came from the literal body via an intentional act of splicing by the TLM. 
-%\footnote{An exhaustive search would not be feasible because final expansions are abstract syntax trees.
+In this term, both \li{Regex} and \li{DNA} are free module variables.
+There is nothing to distinguish references to \li{DNA} that arose from a spliced sub-expression parsed out of the literal body from those that would be a violation of the context independence condition.
+%\footnote{An exhaustive search for "DNA" in the literal body would not be  because final expansions are abstract syntax trees.
 There is not a unique mapping back to concrete syntax, e.g.
 because whitespace can be inserted arbitrarily.}
-As we will discuss more extensively in Sec.
-\ref{sec:existing-approaches}, this is why traditional hygienic term-rewriting macro systems, like those available in various Lisp-family languages \cite{mccarthy1978history} and in Scala \cite{ScalaMacros2013}, cannot be used to repurpose string literals for literal notation at other types.
-
-To address this problem, the TLM parser does not generate the final expansion directly.
-Instead, the parser generates a \emph{proto-expansion} that refers to spliced expressions indirectly by location relative to the start of the provided literal body.
-For example, the proto-expansion generated by \li{$regex} for the example above can be expressed as follows:
+
+As we will discuss more extensively in Sec.~\ref{sec:existing-approaches}, this issue is why hygienic term-rewriting macro systems, like those available in various Lisp-family languages \cite{mccarthy1978history} and in Scala \cite{ScalaMacros2013}, cannot be used to repurpose string literals for composite literal notation at other types.
+These systems would find that the appearance of \li{DNA.any_base} in the generated expansion violates the usual notion of context independence, because \li{DNA.any_base} is not, from the perspective of the context-free syntax, an argument to the macro nor a sub-term of an argument for which a tree path can be assigned \cite{DBLP:conf/esop/HermanW08,Herman10:Theory}.
+Instead, it arises as the result of performing a complex operation---parsing---on a sub-string of the string literal provided as the macro argument.
+
+To address this problem, TLM parsers are not tasked with generating the final expansion directly.
+Instead, the parser generates a \emph{proto-expansion} that refers to spliced terms indirectly by their location relative to the start of the provided literal body.
+For example, the proto-expansion generated by \li{$regex} for the example above can be pretty printed as follows:
 \begin{lstlisting}[numbers=none]
-  Regex.Seq(Regex.Str "GC", Regex.Seq(spliced<4; 11; Regex.t>, Regex.Str "GC"))
+  Regex.Seq(Regex.Str "GC", Regex.Seq(spliced<4; 15; Regex.t>, Regex.Str "GC"))
 \end{lstlisting}
-
-Here, \li{spliced<4; 11; Regex.t>} is a reference to the spliced expression \li{any_base} by its location relative to the start of the literal body being expanded.
-The context independence condition can be enforced straightforwardly on the proto-expansion -- the only free variable in the proto-expansion is \li{Regex}, which is an explicitly listed dependency.
+Here, \li{spliced<4; 15; Regex.t>} is a reference to the spliced expression \li{DNA.any_base} by its location relative to the start of the literal body being expanded---characters $4$ through $15$ (zero-indexed and inclusive) of the literal body, \li{GC$(DNA.any_base)GC}, are \li{DNA.any_base}.
+We say more about the type annotation on the splice reference when we discuss \textbf{Segment Typing} below.
+The context independence condition can be enforced straightforwardly on the proto-expansion -- the only free variable in the proto-expansion is \li{Regex}, which is an explicitly listed dependency, so all is well.
+Had a reference to \li{DNA} appeared in the proto-expansion, then it would be clear that the corresponding final expansion could only be valid in certain contexts (where \li{DNA} is bound) and not others, and therefore this would be a violation of the context independence condition.
 % Requiring that TLMs refer to spliced expressions indirectly in this manner ensures that a TLM cannot ``forge'' spliced terms, i.e.
@@
 
+\paragraph{Segmentation}
 The set of splice references in the proto-expansion generated for a literal body is called the \emph{segmentation} of that literal body.
@@
 \begin{lstlisting}[numbers=none]
-    $html `(SURL<div>Chemical structure of sucrose: {EURL
+    $html `(
+      SURL<div>Chemical structure of sucrose: {EURL
         Smiles.to_svg
           $smiles `(SCOLOR{ECOLORmono_glucoseSCOLOR}-O-{ECOLORmono_fructoseSCOLOR}ECOLOR)` 
-    SURL}</div>EURL)`
+      SURL}</div>EURL
+    )`
 \end{lstlisting}
 
-Each splice reference in the segmentation gives not just the location of the spliced segment but also a type annotation.
+\paragraph{Segment Typing} Each splice reference in the segmentation gives not just the location of the spliced segment but also a type annotation.
 Returning to the regex example above, the specified type annotation is \li{Regex.t}.
@@
 Consequently, type inference can be performed using only the segmentation together with the type annotation on the applied TLM's definition.
-For example, consider the function \li{gene_restriction_template} on Lines 3-4 of Fig.
+For example, consider the function \li{restriction_template} on Lines 3-4 of Fig.
 \ref{fig:first-tlm-example}.
@@
 
+\paragraph{Capture}
+The final expansion is generated by recursively expanding each spliced expression and then inserting it into the final expansion using capture-avoiding substitution.  
 ---------------------- {\color{red} stuff below is notes / not yet revised}

### 2018-03-09 icfp18/syntax-icfp18.tex
rewrite of sec 2.1 -- client perspective

@@
 More generally, evidence suggests that programmers frequently resort to ``stringly-typed programming'', i.e.
-they choose strings instead of composite data structures largely for reasons of notational convenience.
+they choose strings instead of composite data structures, largely for reasons of notational convenience.
 In particular, \citet{TSLs} sampled strings from open source projects and found that at least 15\% of them could be parsed by some readily apparent type-specific grammar, e.g.
@@
 
-Of course, it would not be scalable to ask general-purpose language designers to build in support for all known notations \emph{a priori}.
+Of course, it would not scale to ask general-purpose language designers to build in support for all known notations \emph{a priori}.
 Instead, there has been persistent interest in mechanisms that allow library providers to define new literal notation on their own.
-For example, direct grammar extension systems like Camlp4 \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, macro-based systems like Template Haskell~\cite{SheardPeytonJones:Haskell-02,mainland2007s}, the system of \emph{type-specific languages} (TSLs) described by \citet{TSLs}, and other systems that we will discuss below can all be used to define new literal notation (and, in some cases, other forms of new notation, such as new infix expression forms, control flow operations or type declaration forms, which we leave beyond the scope of this paper).
-
-\paragraph{Problem} The problem that specifically motivates this paper is that these existing systems  make it difficult or impossible to reason abstractly about such fundamental issues as types and variable binding.
+For example, direct grammar extension systems like Camlp4 \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, term rewriting systems like Template Haskell~\cite{SheardPeytonJones:Haskell-02,mainland2007s}, the system of \emph{type-specific languages} (TSLs) described by \citet{TSLs}, and other systems that we will discuss below can all be used to define new literal notation (and, in some cases, other forms of new notation, such as new infix expression forms, control flow operations or type declaration forms, which we leave beyond the scope of this paper).
+
+\paragraph{Problem} The problem that specifically motivates this paper is that these existing systems  make it difficult or impossible to reason abstractly about such fundamental issues as types and variable binding when presented with a program using user-defined literal forms.
 Instead, programmers and editor services can only reason transparently, i.e.
@@
 
-Consider, for example, the perspective of a programmer attempting to comprehend the program text in Figure \ref{fig:K-dialect}, which is written in a dialect of OCaml's surface syntax called Reason \cite{reason-what} that has, hypothetically, been extended with some number of new literal forms by a grammar extension system --- Lines 1-5 show the grammar extension syntax from Camlp4 \cite{ocaml-manual}, but Sugar*/SugarHaskell is  similar \cite{erdweg2011sugarj,erdweg2013framework,erdweg2012layout}.
-Line 9 uses one of the active syntax extensions to construct an encoding of a query in the niche database query language K, using its intentionally terse notation~\cite{Whitney:2001:LOR:376284.375783}.
+Consider, for instance, the perspective of a programmer trying to comprehend the program text in Fig.
+\ref{fig:K-dialect}, which is written in a  popular dialect of OCaml's surface syntax called Reason \cite{reason-what}\todo{up-to-date?}{} that has, hypothetically, been extended with some number of new literal forms by a grammar extension system --- Lines 1-5 show the grammar extension syntax from Camlp4 \cite{ocaml-manual}, but Sugar*/SugarHaskell is  similar \cite{erdweg2011sugarj,erdweg2013framework,erdweg2012layout}.
+Line 9 uses one of these active syntax extensions to construct an encoding of a query in the niche database query language K, using its intentionally terse notation~\cite{Whitney:2001:LOR:376284.375783}.
 The problem is that a programmer looking at the program as presented, and unfamiliar with (i.e.
@@
 If I shadow or remove a module or other binding, could that break or change the meaning of Line~9 because its expansion depends invisibly on the original binding?
-\item \textbf{Segmentation}: Are the characters \li{x}, \li{R} and \li{2} on Line~9 parsed as spliced terms, meaning that they appear directly in the underlying expansion, or are they parsed in some other way peculiar to this literal notation (e.g.
-as operators in the K query language)?
+\item \textbf{Segmentation}: Are the characters \li{x}, \li{R} and \li{2} on Line~9 parsed as spliced terms, meaning that they appear directly in the underlying expansion, or are they parsed in some other way peculiar to this literal notation, e.g.
+as operators in the K query language?
 \item \textbf{Segment Typing}: What type is each spliced term expected to have?
@@
 
-Given these issues, we concluded that direct grammar extension systems were not ideally suited for integration into Reason.
+Given these issues, we concluded that direct grammar extension systems were not ideally suited for integration into the Reason platform, which seeks to develop a clear and modern surface syntax for the OCaml programming language.
 We also evaluated various approaches that are based not on direct grammar extension but on term rewriting over a fixed grammar.
-We give a full account of this evaluation in Sec.~\ref{sec:existing-approaches}.
-Briefly, we found that:
+We give a full account of this evaluation in Sec.~\ref{sec:existing-approaches}, but briefly, we found that:
 \begin{itemize}[leftmargin=15px]
@@
 the rewriting is parametric in the arguments, so they are not suitable for the problem of defining new literal notation.
-\item Hygienic term rewriting macro systems, like those in various Lisp-family languages \cite{DBLP:conf/esop/HermanW08,Herman10:Theory,Kohlbecker86a,DBLP:conf/popl/Adams15,DBLP:conf/popl/ClingerR91,DBLP:journals/lisp/DybvigHB92,mccarthy1978history,Flatt:2012:CLR:2063176.2063195} (e.g.
-Racket \cite{Flatt:2012:CLR:2063176.2063195}) and Scala \cite{ScalaMacros2013}, do not allow us to repurpose string literal forms to define literal forms at composite types because the hygiene discipline does not allow us to splice terms out of literal bodies via parsing.
-Only existing composite forms can therefore be repurposed.
+\item Hygienic term rewriting macro systems, like those in various Lisp-family languages \cite{DBLP:conf/esop/HermanW08,Herman10:Theory,Kohlbecker86a,DBLP:conf/popl/Adams15,DBLP:conf/popl/ClingerR91,DBLP:journals/lisp/DybvigHB92,mccarthy1978history,Flatt:2012:CLR:2063176.2063195}, e.g.
+Racket \cite{Flatt:2012:CLR:2063176.2063195}, as well as Scala \cite{ScalaMacros2013}, do not allow us to repurpose string literal forms to define composite literal forms because the hygiene discipline does not allow us to splice terms out of literal bodies via parsing (the system cannot identify these terms as sub-terms of the macro arguments \cite{DBLP:conf/esop/HermanW08,Herman10:Theory}).% Only existing composite forms can therefore be repurposed.
 \item \emph{Type-specific languages} (TSLs) \cite{TSLs} come closer to our goals in that they explicitly support splicing terms out of literal bodies, but the mechanism falls short with regard to the six reasoning principles just discussed in ways that we will return to in Sec.
@@
 In any case, this approach was designed for simple nominally-typed languages and relies critically on a particular local type inference scheme.
-It is not suitable for a language with an ML-like semantics, meaning a language with support for structural types (like tuple and function types in ML), parameterized types, abstract types, modules, pattern matching and non-local type inference.
+It is not suitable for a language with an ML-like semantics, meaning a language with support for structural types (like tuple and function types in ML), parameterized type families, pattern matching and non-local type inference.
 \end{itemize}
@@
 
-\paragraph{Brief Overview} For a brief overview of our proposed mechanism, consider Figure \ref{fig:K-tsm-example}.
-On Lines 1-5, we define a TLM named \li{$kq} that provides K query literal notation.
+\paragraph{Brief Overview} For a brief overview of our proposed mechanism, consider Fig.~\ref{fig:K-tsm-example}.
+On Lines~1-5, we define a TLM named \li{$kq} that provides K query literal notation.
 On Line 9, we apply this TLM to express the example from Figure \ref{fig:K-dialect}.
@@
 \begin{enumerate}[leftmargin=12pt]
-\item \textbf{Responsibility}: The lexer and parser specified by the applied TLM on Lines 2-3 are exclusively responsible for lexing, parsing and expanding the body of the generalized literal form, i.e.
+\item \textbf{Responsibility}: The lexer and parser specified by the applied TLM on Lines 2-3 are together exclusively responsible for lexing, parsing and expanding the body of the generalized literal form, i.e.
 the characters between \li{`(} and \li{)`}.
@@
 \item \textbf{Context Dependence}: Lines 4-5 specify that expansions generated by \li{$kq} are allowed to use the module \li{KQuery}, and no others.
-The system ensures that this dependency is  bound as specified even if the identifier \li{KQuery} has been shadowed at the application site.
+The system ensures that this dependency is  bound as specified even if the variable \li{KQuery} has been shadowed at the application site.
 This completely relieves clients from needing to consider expansion-internal dependencies when naming variables.
 %For example, if the client renames \li{w}, there is not even a remote possibility that this could break or change the meaning of Line 9.
-\item \textbf{Segmentation}: The intermediate output that the TLM generates is structured so that the system can infer from it an accurate \emph{segmentation} of the literal body that distinguishes spliced terms, which appear directly in the expansion, from segments that are parsed in some other way by the TLM.
+\item \textbf{Segmentation}: The intermediate output that the TLM generates is structured so that the system can infer from it an accurate \emph{segmentation} of the literal body that distinguishes spliced terms, i.e.
+those that appear in the expansion, from segments  parsed in some other way.
 The segmentation is all that needs to be communicated to language services downstream of the expander, e.g.
-editors and pretty printers, which can communicate this information to the programmer using secondary notation, e.g.
+editors and pretty printers, which can pass it on to the programmer using secondary notation, e.g.
 colors in this document.
-So by examining Fig.
-\ref{fig:K-tsm-example}, the programmer knows that the two instances of \li{x} on Line 3 are parsed as spliced expressions (because they are black), whereas the \li{R}'s must be parsed in some other way, e.g.
-as operators of the K language (because they are green).
+So by examining Line 9, the programmer knows that the two instances of \li{x} are spliced expressions (because they are in black), whereas the \li{R}'s must be parsed in some other way, e.g.
+as operators of the K language (because they are in green).
+This also implies that errors in spliced terms can always be reported in terms of their original location.
 \item \textbf{Segment Typing}: Each spliced segment in the inferred segmentation also has  a type annotation.
-This, together with the context independence property, ensures that type inference at the TLM application site can be performed (by editor services or in the programmer's mind) abstractly, i.e.
+This, together with the context independence condition, ensures that type inference at the TLM application site can be performed (by editor services or in the programmer's mind) abstractly, i.e.
 by reference only to the type annotations on the spliced segments, not the full expansion.
@@
 % \item 
-Sec.~\ref{sec:setsms} introduces expression TLMs with a substantially more detailed case study. 
+Sec.~\ref{sec:setsms} details expression TLMs in Reason with a more detailed case study. 
 % \item 
@@
 % \item 
-Sec.~\ref{sec:ptsms} introduces the more general {parametric TLMs}, i.e.
-TLMs that can take type and module parameters, which allow us to define literal notation over a type- or module-parameterized family of types.
+Sec.~\ref{sec:ptsms} introduces the more general {parametric TLMs}, which allow us to define literal notation at a type- or module-parameterized family of types.
 This section also reveals that the \li{expansions require} clause can be understood as parameterization followed immediately by partial parameter application.
-Having introduced the basic machinery, we continue in Sec.~\ref{sec:more-examples} with examples that further demonstrate the expressive power of this approach and various parser implementation strategies that notation providers can consider.
-Notably, this section gives our take on the example from  Fig.~\ref{fig:urweb} of Ur/Web-style HTML literals as well as a TLM that implements quasiquotation for Reason language terms, which is useful for writing other TLMs.
-Next, Sec.~\ref{sec:setsms-formally} defines a type-theoretic calculus of simple expression and pattern TLMs and formally establishes the reasoning principles implied above in their simplest form. 
+Having introduced the basic machinery, we continue in Sec.~\ref{sec:more-examples} with examples that further demonstrate the expressive power of this approach and various parser implementation strategies.
+Notably, this section gives our take on the example from  Fig.~\ref{fig:urweb} of Ur/Web-style HTML literals as well as a TLM that implements quasiquotation for Reason language terms, which is useful for implementing other TLMs.
+Sec.~\ref{sec:setsms-formally} defines a type-theoretic calculus of simple expression and pattern TLMs and formally establishes the reasoning principles implied above in their simplest form. 
 Sec.~\ref{sec:ptlms-formally} adds type functions and an ML-style module system to this calculus, and gives a more general variant of the reasoning principles theorem. 
+Sec.~\ref{sec:implementation} provides a brief overview of how we are implementing TLMs  without modifying OCaml's type system. 
 % \item 
@@
 Sec.~\ref{sec:discussion} concludes with a summary of the contributions of this paper and a discussion of limitations and future work.
-Certain technical details, proofs and extensions to the calculi presented in this paper are given in the supplement.
+Certain technical details, proofs and extensions to the presented calculi are available in the supplement.
 % Also available is an additional series of case studies: notation for an encoding of regular expressions as an abstract data type, implemented using TLMs for quasiquotation and grammar-based parser generators.
@@
 Regexes are  common in, for example, bioinformatics, where they are used to express patterns in DNA sequences.
-We can construct a regex that matches the strings \li{"A"}, \li{"T"}, \li{"G"} or \li{"C"}, which represent the four bases in DNA, as follows:
+We might construct a regex that matches the strings \li{"A"}, \li{"T"}, \li{"G"} or \li{"C"}, which represent the four bases in DNA, as follows:
 \begin{lstlisting}[numbers=none]
@@
 In Reason, the notation \li{Regex.(}$e$\li{)} locally opens the module \li{Regex} within the expression $e$, so we did not need to qualify each application of \li{Regex.Or} and \li{Regex.Str} above.
-Even with this shorthand, however, constructing regexes in this way is syntactically costly.
-Instead, we would like to have the option to use the common POSIX-style notation  \cite{STD95954} when constructing values of type \li{Regex.t}, including values constructed compositionally from previously constructed regexes and strings.
+Even with the aid of this shorthand, however, constructing regexes in this way is syntactically costly.
+Instead, we would like to have the option to use the common POSIX-style notation  \cite{STD95954} when constructing values of type \li{Regex.t}, including values constructed compositionally from other regexes and strings.
 We solve this problem in Fig.
 \ref{fig:regex-tlm-def} by defining a TLM named \li{$regex} (pronounced ``lit regex'') that supports POSIX-style regex notation extended with splice forms for regexes and strings.
+Fig.
+\ref{fig:first-tlm-example} shows three examples of \li{$regex} being applied.
 % One approach would be to simply define a function \li{parse_rx : string -> Regex.t} that parses a string at run-time to produce a regex.
@@
   module Regex = {
-    type t = AnyChar | Str(string)
-           | Seq(t, t) | Or(t, t) 
+    type t = AnyChar 
+           | Str(string)
+           | Seq(t, t) 
+           | Or(t, t) 
            | Star(t);
@@
 \begin{lstlisting}[mathescape=|]
+  module RegexNotation {
     notation $regex at Regex.t
@@
         module Regex as Regex;
+  };
 \end{lstlisting}
@@
 \begin{lstlisting}
+  open RegexNotation;
   module DNA = { let any_base = $regex `(SURLA|T|G|CEURL)`; };
@@
 \end{lstlisting}
-\vspace{-2px}
+\vspace{-4px}
 \caption{Examples of the \li{$regex} TLM being applied in a bioinformatics application.
@@
 \end{subfigure}
-\vspace{3px}
+\vspace{0px}
 \caption{Case Study: POSIX-style regex literal notation, with support for string and regex splicing.}
-\vspace{-5px}
+\vspace{-6px}
+\label{fig:regex-case-study}
 \end{figure}
@@
 \subsection{Client Perspective}
-The focus of this paper is on the client programmer's perspective, so we start in this section with some examples of \li{$regex} being applied in Fig.
-\ref{fig:first-tlm-example}, then use these examples to detail the client-side abstract reasoning principles that the mechanism provides, all before finally considering the details of the lexer and parser in Sec.
+Let us start from the perspective of a client programmer examining Fig.
+\ref{fig:regex-case-study} but holding the underlying expansion, as well as the details of the lexer and parser, \li{RegexLexer} and \li{RegexParser}, abstract.
+We will return to describe the lexer and parser from the provider's perspective in Sec.
 \ref{sec:provider-perspective}.
+
+% Fig.
+\ref{fig:first-tlm-example}, then use these examples to detail the client-side abstract reasoning principles that the mechanism provides, all before finally .
 % The \li{open notation} directive on Line 2 of Fig.
@@
 
-Line 1 of Fig.
-\ref{fig:first-tlm-example} applies \li{$regex} to construct the regex \li{any_base} that was  described above, this time using POSIX regex notation.
-Line~2 applies \li{$regex} again, using its regex splice form to construct a regex matching DNA sequences recognized by the BisA restriction enzyme, where the middle base can be any base.
-Finally, Lines~3-4 of Fig.
-\ref{fig:first-tlm-example} define a function that constructs a more complex regex from these two regexes and a given gene sequence represented as a string, which is spliced in using the string splice form.
-Notice that the TLM provider has chosen to use \li{$(}$e$\li{)} for regex splicing and \li{$$(}$e$\li{)} for string splicing, where $e$ is an unexpanded Reason expression of arbitrary form.
+Line 2 of Fig.
+\ref{fig:first-tlm-example} applies \li{$regex} to construct the regex \li{DNA.any_base} that was  described above, this time using the more clear and familiar POSIX regex notation.
+Line~3 applies \li{$regex} again, using its regex splice form to compositionally construct a regex matching DNA sequences recognized by the BisA restriction enzyme, where the middle base can be any base.
+Finally, Lines~4-5 of Fig.~\ref{fig:first-tlm-example} define a function, \li{restriction_template}, that constructs a more complex regex from these first two regexes and a given gene sequence represented as a string.
 
@@
 \end{lstlisting}
-According to the context-free grammar of (our extended dialect of) Reason, this form is simply a leaf of the unexpanded syntax tree, like a string literal would be. TLM names are prefixed by \li{$} 
+According to the context-free grammar of (this paper's branch of) Reason, this form is simply a leaf of the unexpanded syntax tree, like a string literal would be. TLM names are prefixed by \li{$} 
  to distinguish them from variables.
 We call the TLM argument, \li{`(GC$(DNA.any_base)GC)`}, a \emph{generalized literal form}, following the terminology introduced in the prior work on type-specific languages \cite{TSLs}.
-The only lexical constraint on the body of a generalized literal form, i.e.
-the characters between \li{`(} and \li{)`}, is that any further occurrences of \li{`(} must be balanced by \li{)`}, much like nested comments in Reason/OCaml.
+The only lexical constraint on the \emph{body} of a generalized literal form, i.e.
+the characters between \li{`(} and \li{)`}, is that any nested occurrences of \li{`(} must be balanced by \li{)`}, much like nested comments in Reason/OCaml.
 Generalized literal forms therefore lexically subsume other literal forms.
 The nesting constraint is to allow TLM applications to appear inside spliced expressions (discussed below).
-The prior work specified several other choices of outer delimitation, including layout-sensitive delimitation, but for the purposes of this work, we will use the outer delimiters \li{`(} and \li{)`} exclusively.
-Choices about how spliced expressions are recognized inside the literal body are entirely at the discretion of each TLM, as we will discuss below.
+Prior work by \citet{TSLs} specified several other choices of outer delimitation, including layout-sensitive delimitation, but for the purposes of this work, we will use the outer delimiters \li{`(} and \li{)`} exclusively.
+Choices about how spliced expressions are recognized inside the literal body are entirely at the discretion of each TLM.
+In this case, notice that the TLM provider has chosen to use \li{$(}$e$\li{)} for regex splicing and \li{$$(}$e$\li{)} for string splicing, where $e$ is an unexpanded Reason expression of arbitrary form.
+We return to the topic of splicing below.
 
@@
 Responsibility for lexing, parsing and expanding each literal body is delegated uniquely to the lexer and parser specified in the definition of the applied TLM by the clauses \li{lexer} and \li{parser}, respectively.
-We will return to consider the details of lexer and parser definition, and show the definitions of the lexer and parser specified by \li{$regex}, in Sec.
-\ref{sec:provider-perspective}.
-For now, let us purposefully hold these details abstract.% and take the perspective of a client that does not want to delve into these details.
-
-In Reason, TLM definitions like the one in Fig.
+For now, let us purposefully continue to hold these details abstract.% and take the perspective of a client that does not want to delve into these details.
+
+TLM definitions follow the scoping structure of Reason, i.e.
+they can be defined in modules alongside other definitions.
+When a TLM definition appears inside a module, it must also appear in the signature with the same specification, up to the usual notions of type and module path equivalence in the type annotation and module dependencies, which are discussed below.
+This is much like the situation with datatype definitions in ML.
+In Sec.
+\ref{sec:implementation}, we will discuss how we use an encoding trick involving nested modules and temporary exceptions internally in the implementation to avoid having to primitively add TLM definitions to OCaml's type and module system.
+
+By convention, we define TLMs in a module suffixed with \li{Notation} so that client programmers can \li{open} just the TLM definition without bringing other definitions into scope, as shown on Line 1 of Fig.
+\ref{fig:first-tlm-example}.
+We could also have fully qualified each TLM application, \li{RegexNotation.$regex}, or abbreviated the TLM definition:
+\begin{lstlisting}[numbers=none]
+    let $regex = RegexNotation.$regex;
+\end{lstlisting}
+
+What is fundamental about this design is that there is a well-defined and familiar protocol for finding the definition of the TLM uniquely responsible for each generalized literal form in a program, following the usual scoping rules of the language.
+An editor service or documentation tool can use this protocol to integrate TLM definition lookup into a ``go to definition'' command.
+
+% In Reason, TLM definitions like the one in Fig.
 \ref{fig:regex-tlm-def} are separately defined in files with the suffix \li{.relit}, which are processed and brought into scope by the Reason build system, \li{rebuild}.
@@
 the syntax libraries in Sugar* \cite{erdweg2011sugarj,erdweg2013framework,erdweg2012layout}.
-In our theoretical treatment in Sec.~\ref{sec:setsms-formally}, however, TLM definitions are integrated directly into the scoping structure of the language.
-This is not possible in Reason because this would require extending OCaml's semantics, i.e.
-adding TLM definitions to module signatures.
-While we were careful to choose a design where this modification would be feasible in principle---TLM definition equality requires only decidable type and module path equality---modifying the semantics of OCaml is intentionally beyond the scope of the Reason project.
+In our theoretical treatment in Sec.~\ref{sec:setsms-formally}, however, TLM definitions will be integrated directly into the scoping structure of the language.
+This is not possible in Reason because this would require extending OCaml's semantics with TLM definitions in module signatures.
+While we were careful to choose a design where this modification would be feasible in principle---TLM definition equality requires only decidable type and module path equality---modifying the semantics of OCaml has intentionally been placed beyond the scope of the Reason project.
 In any case, what is fundamentally important is that there is a well-defined protocol for finding the definition of the TLM uniquely responsible for each application site.
@@
 \paragraph{Expansion Typing}
-Having found the definition of the \li{$regex} TLM, the client can immediately determine the type of the expansion being generated at each application site because it is specified explicitly by the clause \li{at Regex.t} on Line 1 of Fig.
+Having found the definition of the \li{$regex} TLM, the client can immediately determine the type of the expansion being generated at each application site because it is specified explicitly by the clause \li{at Regex.t} on Line~2 of Fig.
 \ref{fig:regex-tlm-def}.
@@
 The identity of \li{Regex.t} is determined relative to the TLM definition site, not at each application site, so the module \li{Regex} need not be in scope at the application site, or it can have been shadowed by a different module.
-In Reason, TLM definitions can refer to exactly those modules that a Reason module at the top level of the defining package would have access to, i.e.
-other top-level modules in the package as well as the modules exported by other packages that have been imported by the build system.
+% In Reason, TLM definitions can refer to exactly those modules that a Reason module at the top level of the defining package would have access to, i.e.
+other top-level modules in the package as well as modules exported by other packages that have been imported by the build system.
 
@@
 The TLM mechanism enforces a strong context independence condition on generated expansions by requiring that the TLM definition explicitly specify all modules that the expansions it generates might internally depend on.
-In this case, Lines 4-5 of Fig.
+In this case, Lines 5-6 of Fig.
 \ref{fig:regex-tlm-def} specify that generated expansions might use the module \li{Regex}, again as it is bound at the TLM definition site, using the module variable \li{Regex} internally.
-The expansion-internal module variable happens to coincide with the module path in this case, but in general, the dependency can be an arbitrary module path while internally it must be bound to a module variable.
+In this case, the expansion-internal module variable happens to coincide with the module path, but in general, the dependency can be an arbitrary module path.
 For example, consider the following dependency clause:
@@
 All other bindings, whether at the TLM definition site or the TLM application site, are not internally available to the expansion.
-This allows programmers to freely rename and shadow variables as they are used to, without needing to inspect each parser to determine whether the expansions it generates require certain variables be bound in particular ways at each application site (as is common with, e.g., syntax extensions based on \li{camlp4} or Sugar*, or when using unhygienic term rewriting approaches like OCaml's PPX system or Template Haskell \cite{SheardPeytonJones:Haskell-02}, which we discuss further in Sec.
+This allows both client programmers and the provider of the package where the TLM is defined to freely rename and shadow variables in the usual way, i.e.
+without needing to inspect each expansion or each TLM definition's parser to determine whether the syntax trees that it generates require certain variables be bound in particular ways at each application site (as is common with, e.g., syntax extensions based on \li{camlp4} or Sugar*, or when using unhygienic term rewriting approaches like OCaml's PPX system or Template Haskell, which we discuss further in Sec.
 \ref{sec:existing-approaches}).
 
-Enforcing this strong context independence condition is quite subtle because TLM parsers need to be able to extract spliced expressions from the literal body and place them in the final expansion.
-Na\"ively checking that the final expansion be closed except for the explicitly named dependencies would inappropriately enforce context independence on application site spliced expressions, which should certainly not be prevented from referring to variables in the application site context.
-For example, consider the final expansion of the example from Line 2 of Fig.
-\ref{fig:first-tlm-example} above:
+Enforcing this strong context independence condition is technically quite subtle because TLM parsers need to be able to perform splicing, i.e.
+they need to be able to parse  terms out of the literal body for placement in the expansion.
+Na\"ively checking that only the explicitly named dependencies are free in the expansion would inappropriately enforce context independence on application site spliced expressions, which should certainly not be prevented from referring to variables in scope at the application site.
+For example, consider the final expansion of our example from Line 2 of Fig.
+\ref{fig:first-tlm-example}:
 \begin{lstlisting}[numbers=none]
@@
 In this term, both \li{Regex} and \li{DNA} are free module variables.
-There is nothing to distinguish references to \li{DNA} that arose from a spliced sub-expression parsed out of the literal body from those that would be a violation of the context independence condition.
+There is nothing to distinguish references to \li{DNA} that arose from a spliced sub-expression parsed out of the literal body from those that would indicate that the context independence condition has been violated.
 %\footnote{An exhaustive search for "DNA" in the literal body would not be  because final expansions are abstract syntax trees.
@@
 As we will discuss more extensively in Sec.~\ref{sec:existing-approaches}, this issue is why hygienic term-rewriting macro systems, like those available in various Lisp-family languages \cite{mccarthy1978history} and in Scala \cite{ScalaMacros2013}, cannot be used to repurpose string literals for composite literal notation at other types.
-These systems would find that the appearance of \li{DNA.any_base} in the generated expansion violates the usual notion of context independence, because \li{DNA.any_base} is not, from the perspective of the context-free syntax, an argument to the macro nor a sub-term of an argument for which a tree path can be assigned \cite{DBLP:conf/esop/HermanW08,Herman10:Theory}.
+These systems would find that the appearance of \li{DNA.any_base} in the generated expansion violates their context independence condition, because \li{DNA.any_base} is not, from the perspective of the context-free syntax, an argument to the macro nor even a sub-term of an argument for which a tree path can be assigned \cite{DBLP:conf/esop/HermanW08,Herman10:Theory}.
 Instead, it arises as the result of performing a complex operation---parsing---on a sub-string of the string literal provided as the macro argument.
@@
 \end{lstlisting}
-Here, \li{spliced<4; 15; Regex.t>} is a reference to the spliced expression \li{DNA.any_base} by its location relative to the start of the literal body being expanded---characters $4$ through $15$ (zero-indexed and inclusive) of the literal body, \li{GC$(DNA.any_base)GC}, are \li{DNA.any_base}.
+Here, \li{spliced<4; 15; Regex.t>} is a reference to the spliced expression \li{DNA.any_base} by its location relative to the start of the literal body being expanded---characters $4$ through $15$ (zero-indexed and inclusive) of the literal body \li{GC$(DNA.any_base)GC} are \li{DNA.any_base}.
 We say more about the type annotation on the splice reference when we discuss \textbf{Segment Typing} below.
@@
 Had a reference to \li{DNA} appeared in the proto-expansion, then it would be clear that the corresponding final expansion could only be valid in certain contexts (where \li{DNA} is bound) and not others, and therefore this would be a violation of the context independence condition.
+\todo{why not spliced[e] somewhere}
 % Requiring that TLMs refer to spliced expressions indirectly in this manner ensures that a TLM cannot ``forge'' spliced terms, i.e.
@@
 \paragraph{Segmentation}
-The set of splice references in the proto-expansion generated for a literal body is called the \emph{segmentation} of that literal body.
-The system checks that the segmentation does indeed segment the literal body, i.e.
-that the segments are non-overlapping.
-For the purposes of abstract reasoning, only the segmentation needs to be reported to downstream editor services, and ultimately to the programmer.
-The rest of the proto-expansion can be held abstract.
-In this paper, spliced segment locations are reported using colors, with segments that are not part of spliced segment shown in color (green, so far) while spliced segments start out black.
+The finite set of splice references in the proto-expansion generated for a literal body is called the \emph{segmentation} of that literal body.
+The segmentation of the example above is the finite set containing one element, \li{spliced<4; 15; Regex.t>}.
+For the more complex example from Line 4 of Fig.
+\ref{fig:first-tlm-example}, the segmentation contains multiple spliced segments:
+\begin{lstlisting}[numbers=none]
+  { spliced<2; 5; Regex.t>, spliced<9; 20; Regex.t>; spliced<26; 29; string>; 
+    spliced<33; 44; Regex.t>; spliced<49; 52; Regex.t> } 
+\end{lstlisting}
+The system checks that the segmentation does in fact segment the literal body, i.e.
+that the segments are in-bounds, of positive extent and non-overlapping.
+Adjacent spliced segments must also be separated by at least one character.
+The spliced segment locations can therefore be communicated unambiguously to the programmer by tools downstream of the expander, e.g.
+program editors and pretty printers, using secondary notation.
+In this paper, non-spliced segments are shown in color and spliced segments start in black.
+
 When TLM applications are nested, a distinct color can be used at each depth.
@@
 \ref{sec:more-examples}, we will describe a TLM \li{$html} for HTML notation in the style of Ur/Web (see Fig.~\ref{fig:urweb}).
-We might apply it, together with another TLM \li{$smiles} for chemical structures described using the SMILES standard \cite{anderson1987smiles}, as follows:
+We can also define a TLM, \li{$smiles}, not shown, that allows for the use of the standard SMILES notation for chemical structures \cite{anderson1987smiles}, extended with splicing notation.
+Assuming that the expansion type of \li{$smiles} is \li{Smiles.t} and that \li{Smiles.to_svg : Smiles.t -> Html.t}, we can nicely transform encodings of chemical structures into a vector graphic embedded into an HTML page as follows:
 \begin{lstlisting}[numbers=none]
@@
 
-\paragraph{Segment Typing} Each splice reference in the segmentation gives not just the location of the spliced segment but also a type annotation.
-Returning to the regex example above, the specified type annotation is \li{Regex.t}.
-This type annotation, which must also be context independent, constrains the spliced expression.
-Consequently, type inference can be performed using only the segmentation together with the type annotation on the applied TLM's definition.
+Another benefit of explicitly tracking the locations of spliced segments is that errors that originate in spliced terms can be reported in terms of their original source location \cite{DBLP:journals/jsc/DeursenKT93}.
+
+\paragraph{Segment Typing} Each splice reference in the segmentation specifies not just the location of a spliced expression but also its expected type.
+The identity of this type is resolved in a context-independent manner, assuming only the dependencies explicitly specified by the TLM.
+
+By associating a type with each spliced segment, we guarantee that type inference can be performed abstractly, meaning that only the segment types, together with the expansion types specified by the applied TLMs, are necessary to infer types for variables appearing in a client-side function.
 For example, consider the function \li{restriction_template} on Lines 3-4 of Fig.
 \ref{fig:first-tlm-example}.
-The return type of this function can be determined to be \li{Regex.t} because of the type annotation on the \li{$regex} TLM.
-The type of the argument, \li{gene}, can be inferred to be \li{string} because the segmentation specifies the type \li{string} for the spliced segment where it appears.
-Context independence implies that \li{gene} cannot appear elsewhere in the expansion, and so no further typing constraints could possibly be collected from examining the portions of the expansion held abstract.
-Spliced segment types can be communicated directly to the programmer upon request by an editor service, e.g.
-\li{merlin} for OCaml/Reason \cite{Merlin}.
+The return type of this function can be inferred to be \li{Regex.t} from the type annotation on the \li{$regex} TLM, as previously discussed.
+The type of the argument, \li{gene}, can be inferred to be \li{string} because the segmentation specifies the type \li{string} for the spliced segment where it appears (cf.
+the segmentation shown under \textbf{Segmentation} above).
+The context independence condition implies that \li{gene} cannot appear elsewhere in the expansion, and so no further typing constraints could possibly be collected from examining the portions of the (proto-)expansion being held abstract.
+
+Segment types can be communicated directly to the programmer upon request by an editor service.
+For Reason, we are extending the {Merlin} editor service \cite{Merlin}, which is used by various Reason editor extensions (e.g.
+for Emacs, Vim, and so on), with a new editor command for reporting the expected type of the innermost spliced segment containing the cursor.
+Note that because the type is specified explictly in the output of the parser, this information can be reported even when there is a parse error or type error in a spliced expression.
+% An alternative design that simply reported the type assigned to each spliced expression based on the final expansion (which already exists in Merlin) would be less useful in situations where there is a type error in a spliced expression, or the programmer is in the process of editing the program and wants to know what type of expression to construct.
+
+Segment types are somewhat analagous to the argument types of a function.
+The difference is that the argument signature of a function is the same every time the function is applied, i.e.
+it is associated with the function itself, whereas the inferred segmentation can differ at each TLM application site.
+This is, of course, what gives TLMs their notational flexibility.
 
 \paragraph{Capture}
-The final expansion is generated by recursively expanding each spliced expression and then inserting it into the final expansion using capture-avoiding substitution.  
+In discussing the question of inferring a type for \li{gene} above, we neglected to consider one critical question: are we sure that the variable \li{gene} in the third spliced segment on Line 4 of Fig.
+\ref{fig:first-tlm-example} is, in fact, a reference to the argument \li{gene} of the \li{restriction_template} function?
+After all, if we hold the expansion abstract then it may well be that the third spliced segment appears under (i.e.
+captures) a different binding of the identifier \li{gene}.
+For more common identifiers, e.g.
+\li{x} and \li{tmp}, inadvertent capture is perhaps inevitable.
+
+For example, consider the following application site:
+\begin{lstlisting}[numbers=none]
+    let tmp = /* ... application site temporary ... */;
+    $html `(SURL<h1>{EURLf(tmp)SURL}</h1>EURL)`;
+\end{lstlisting}
+Now consider the scenario where the proto-expansion generated by \li{$html} has the following form:
+\begin{lstlisting}[numbers=none]
+    let tmp = /* ... expansion-internal temporary ... */;
+    Html.H1Element(tmp, spliced<5; 10; Html.t>);
+\end{lstlisting}
+If the final expansion was produced na\"ively, by syntactically replacing the spliced segment reference by the final expansion recursively determined for the corresponding spliced expression, then the  variable \li{tmp} in the spliced expression would capture the binding of \li{tmp} in the proto-expansion.
+The result when the types of the two bindings differed would be a type error that exposes the internal details of the expansion.
+If the types of the two bindings of \li{tmp} coincided, then there could be subtle changes in run-time behavior that would be impossible to diagnose without looking at the expansion itself.
+Capture of invisible bindings is, due to these hazards, clearly unreasonable.
+
+To address this problem in our design, splicing is guaranteed to be capture-avoiding.
+The final expansion is generated by recursively expanding each spliced expression and then inserting it into the final expansion via capture-avoiding substitution.
+In other words, the system automatically alpha-varies the bindings in the proto-expansion to ensure capture avoidance.
+There is no need for TLM providers to manually deploy a mechanism that generates fresh variables (as in, e.g., Racket's \emph{reader macro} system \cite{Flatt:2012:CLR:2063176.2063195}, which is further discussed in Sec.
+\ref{sec:existing-approaches}).
+For example, the final expansion of the example above would be alpha-equivalent to the following:
+\begin{lstlisting}[numbers=none]
+  let tmp = /* ... application site temporary ... */; 
+  let tmp' = /* ... expansion-internal temporary (automatically renamed) ... */;
+  Html.H1Element(tmp', f(tmp));
+\end{lstlisting}
+%Notice that the expansion-internal binding of \li{tmp} has been alpha-varied to \li{tmp'}.
+The reference to \li{tmp} in the spliced expression then refers, as intended, to the application site binding.
+
+% For TLM providers, the benefit of this mechanism is that they can name the variables used internally within expansions freely. TLM clients can, in turn, can reason abstractly about \textbf{Capture}.
+
+Although this capture avoidance discipline implies that TLMs cannot intentinoally expose bindings internal to the expansion directly to spliced expressions, this does not imply that values cannot flow from the expansion into a spliced expression.
+It simply means that when this is intended, the segment type must be a function type.
+Reason has concise lambda notation, \li{(x) => e}, which is useful in many of these use cases.
+For example, the capture avoidance discipline means that we cannot define numeric list comprehension notation\footnote{A polymorphic list comprehension TLM requires type parameters, introduced in Sec.
+\ref{sec:ptsms}.} like the following:
+\begin{lstlisting}[numbers=none,deletekeywords={in}]
+  $listcomp `(x + 1SURL | x in EURLloSURL to EURLhi)` (* NO! cannot reason abstractly *)
+\end{lstlisting}
+However, the following is permitted, because \li{x} is bound by the spliced lambda expression:
+\begin{lstlisting}[numbers=none]
+  $listcomp `((x) => x + 1 SURL|EURL lo SURLtoEURL hi)`  (* OK! *)
+\end{lstlisting}
+
+Our contention is that small syntactic costs like these are more than justified by the peace of mind of knowing that unfamiliar literal notation is absolutely prevented from obscuring the type and binding structure of the language.
+We say more about the future prospect of a  mechanism designed specifically for specialized binding forms, e.g.
+Haskell-style \li{do} notation \cite{jones2003haskell}, in Sec.
+\ref{sec:discussion}.\todo{do this}% For now, the existing infix operator support in Reason and OCaml suffices for working with monadic values.
+
+% There are various more relaxed capture avoidance conditions that one might consider, e.g.
+where the segmentation explicitly specifies which spliced identifiers in the literal body are to be captured in each spliced expression.
+However, in Reason
+
+% For example, Haskell's notation for values equipped with monadic structure, i.e.
+\li{do}-notation, cannot be expressed under this strong capture avoidance condition because it introduces bindings into sub-terms .
+Even without
+
+% An alternative capture avoidance discipline is one where each spliced segment is annotated with a list of spliced identifiers that will be bound in the corresponding spliced expression, together with their types.
+Spliced identifiers, like other spliced segments, must appear at an identified location within the literal body, and cannot overlap with other spliced segments.
+
+% In our initial designs, we were able to express this example because we distinguished spliced identifiers, much as we do spliced types and expressions.
+These could be explicitly bound inside spliced expressions by extending the \li{SplicedE} and \li{SplicedT} constructors to take finite sets of spliced identifiers.
+This is conceptually similar to the approach taken by \citet{DBLP:conf/esop/HermanW08} for giving (already parsed) sub-terms access to internal variables.
+However, we ultimately decided to remove this feature because it would disproportionately increase the reasoning burden on clients when they encounter an unfamiliar literal: \emph{might this obscure literal also be introducing bindings?} It is difficult for a program editor to display a set of ``hidden bindings'' to the programmer.
+Haskell-style infix notation for monadic values, where bind is \li{e >>= f}, can be expressed using TLMs without support for spliced identifiers.
+Reason 3.0 is adopting more concise function syntax, which will decrease the cost of this syntax.
+
+\paragraph{Summary}
+
+Let us briefly reiterate the key ideas of this section.
+The focus was on providing simple protocols to client programmers when they have questions about the syntactic structure, type structure or binding structure of a program.
+At no point in answering these sorts of questions is the client made to look at the generated expansion itself, nor inspect the parser implementation.
+Instead, the programmer need only be given knowledge of the expansion type from the TLM definition and the segmentation inferred by the expander at each application site, which carries a small volume of information that can be communicated straightforwardly by standard editor services.
+Certain questions related to binding structure simply do not need to be asked due to the strict context independence and capture avoidance discipline of the system, enabled by the explicit tracking of spliced segments by location.
+Despite these semantic constraints, the system is able to express a number of interesting examples with few compromises because there is only one simple lexical constraint on literal bodies.
+
 ---------------------- {\color{red} stuff below is notes / not yet revised}
 
-When a TLM definition appears inside a module, it must also appear in the signature with the same specification, up to the usual notions of type and module path equivalence in the type annotation and module dependencies (like datatype definitions in ML).
-Note that even when a TLM needs only a single helper function, it must be placed into a named module.
-This is to avoid needing to reason about the equivalence of arbitrary expressions when determining whether two signatures are compatible.
-
-Notice that regexes are spliced in using \li{$(}$e$\li{)}, where $e$ is any expression of type \li{Regex.t}, while strings are spliced in using \li{$$(}$e${\li)}, where $e$ is any expression of type \li{string}.
-These choices are made entire
-
- Every TLM definition also includes a \emph{type annotation}, here \li{at html}, and a \emph{parse function} between \li{by static \{} and \li{\}}.
-% The TLM definition follows the same scoping rules as type definitions.
-We will describe how TLM definitions are packaged into libraries in Sec \ref{sec:static-eval}.
-
-Using this TLM, we can express the Ur/Web example from Figure \ref{fig:urweb} as shown in Figure \ref{fig:first-tsm-example}. On both Lines 1 and 2, we apply 
-\li{$html} 
- to a \emph{generalized literal form} delimited by \li{[|} and \li{|]}.
+% Using this TLM, we can express the Ur/Web example from Figure \ref{fig:urweb} as shown in Figure \ref{fig:first-tsm-example}. On both Lines 1 and 2, we apply 
+% \li{$html} 
+%  to a \emph{generalized literal form} delimited by \li{[|} and \li{|]}.
 Generalized literal forms, which first arose in the prior work on TSLs \cite{TSLs}, syntactically subsume other literal forms because the context-free syntax of the language only specifies the outer delimiters.
@@
 
- The system delegates responsibility over the parsing and expansion of each literal body to the applied TLM's parse function during a semantic phase called \emph{expansion}, which starts just before and continues into the typing phase.
-
- Because the parse function is applied during this phase, rather than at run-time, we call it a \emph{static function}.
+%  The system delegates responsibility over the parsing and expansion of each literal body to the applied TLM's parse function during a semantic phase called \emph{expansion}, which starts just before and continues into the typing phase.
+
+%  Because the parse function is applied during this phase, rather than at run-time, we call it a \emph{static function}.
 Static functions  cannot refer to the surrounding variable bindings because those variables stand for run-time values.
@@
 \else \fi
-  % return to in Sec. \ref{sec:static-eval}.
-
- The input type of the parse function, \li{body}, classifies encodings of literal {bodies}.
+%   % return to in Sec. \ref{sec:static-eval}.
+
+%  The input type of the parse function, \li{body}, classifies encodings of literal {bodies}.
 Literal bodies are sequences of characters, so we define \li{body} as a synonym of \li{string} in \autoref{fig:indexrange-and-parseresult}.
@@
 Let us consider these two possibilities in turn.
+
+\subsection{Provider Perspective}
+\label{sec:provider-perspective}
 
@@
 These are discussed next.
-
-\subsection{Provider Perspective}
-\label{sec:provider-perspective}
 
@@
 Proto-expansion validation ensures that spliced terms have access \emph{only} to the bindings at the application site---spliced terms cannot capture bindings internal to the proto-expansion. 
-For example, consider the following application site:
-\begin{lstlisting}[numbers=none]
-    let tmp = /* ... application site temp ... */;
-    $html [|SURL<h1>{EURLf(tmp)SURL}</h1>EURL|];
-\end{lstlisting}
-Now consider the scenario where the proto-expansion generated by \li{$html} has the following form:
-\begin{lstlisting}[numbers=none]
-    let tmp = /* ... expansion-internal temp ... */;
-    H1Element(tmp, spliced<5; 10; html>);
-\end{lstlisting}
-Na\"ively, the binding of the variable \li{tmp} in the proto-expansion could shadow the application-site binding of \li{tmp} in the final expansion. 
-To address this problem, splicing is guaranteed to be capture-avoiding.
-When generating the final expansion, the system discharges the requirement that capture not occur by implicitly alpha-varying the bindings in the proto-expansion as needed.
-There is no need for TLM providers to explicitly deploy a mechanism that generates fresh variables (as in, e.g., prior \emph{reader macro} systems \cite{Flatt:2012:CLR:2063176.2063195}.
-% For example, the final expansion of the example above might take the following form:
-%\begin{lstlisting}[numbers=none]
-%  let tmp = /* ... application site temporary ... */; 
-%  let tmp' = /* ... expansion-internal temporary ... */;
-%  H1Element(tmp', f(tmp));
-%\end{lstlisting}
-%Notice that the expansion-internal binding of \li{tmp} has been alpha-varied to \li{tmp'}.
-The reference to \li{tmp} in the spliced expression then refers, as intended, to the application site binding.
-
-% For TLM providers, the benefit of this mechanism is that they can name the variables used internally within expansions freely. TLM clients can, in turn, can reason abstractly about \textbf{Capture}.
-
-Capture avoidance does prevent library providers from intentionally introducing bindings into spliced terms.
-For example, Haskell's literal notation for monadic values, i.e.
-\li{do}-notation, cannot be expressed because it introduces a new binding syntax \cite{jones2003haskell}.
-
-In our initial designs, we were able to express this example because we distinguished spliced identifiers, much as we do spliced types and expressions.
-These could be explicitly bound inside spliced expressions by extending the \li{SplicedE} and \li{SplicedT} constructors to take finite sets of spliced identifiers.
-This is conceptually similar to the approach taken by \citet{DBLP:conf/esop/HermanW08} for giving (already parsed) sub-terms access to internal variables.
-However, we ultimately decided to remove this feature because it would disproportionately increase the reasoning burden on clients when they encounter an unfamiliar literal: \emph{might this obscure literal also be introducing bindings?} It is difficult for a program editor to display a set of ``hidden bindings'' to the programmer.
-Haskell-style infix notation for monadic values, where bind is \li{e >>= f}, can be expressed using TLMs without support for spliced identifiers.
-Reason 3.0 is adopting more concise function syntax, which will decrease the cost of this syntax.
 %We will show an alternative formulation of Haskell's syntax for monadic commands that uses Reason's anonymous function syntax to bind variables in Sec. \ref{sec:application-monadic-commands}.
@@
 % \endgroup
+\newcommand{\implSec}{Implementation}
+\section{\protect\implSec}
+\label{sec:implementation}
+
+Note that even when a TLM needs only a single helper function, it must be placed into a named module.
+This is to avoid needing to reason about the equivalence of arbitrary expressions when determining whether two signatures are compatible.
 

### 2018-03-19 icfp18/syntax-icfp18.tex
ICFP18 submission

@@
 %% For double-blind review submission
-\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
+\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=false,printccs=false,printacmref=false}
 % \documentclass[sigplan,review,10pt,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
@@
   \@startsection{paragraph}{4}%
-  {\z@}{0.25ex \@plus 1ex \@minus .2ex}{-1em}%
+  {\z@}{1ex \@plus .2ex \@minus .2ex}{-1em}%
   {\normalfont\normalsize\bfseries}%
@@
 \makeatother
+
+% \usepackage{titlesec}
+% \titleformat{\subsubsection}[runin]{\large\bfseries}{\thesubsubsection}{1em}{}
+
+\makeatletter
+\renewcommand{\subsubsection}{%
+  \@startsection{subsubsection}{3}%
+  {\z@}{1ex \@plus .2ex \@minus .2ex}{-1em}%
+  {\normalfont\normalsize\bfseries}%
+}
+\makeatother
+
+%   \usepackage{titlesec}
+
+% \titleformat*{\subsubsection}{\large\bfseries}
 
@@
 mathescape=false,escapeinside={(~}{~)},
-numbers=left, numberstyle=\scriptsize\color{mygray}, language=ML,moredelim=[il][\sffamily]{?},showspaces=false,showstringspaces=false,xleftmargin=0pt, numbersep=-6pt, morekeywords=[1]{try,spliced,tyfam,opfam,let,fn,val,def,casetype,objtype,metadata,of,*,string,lexer,parser,datatype,new,toast,notation,module,switch,where,expansions,require,import,for,ana,syn,opcon,tycon,metasignature,metamodule,metasig,metamod,static,at,by,tycase,mod,macro,match,pattern,in,patterns,expressions,implicit,forall,exptsm,pattsm},deletekeywords={double,structure,of},classoffset=0, xleftmargin=0pt,
+numbers=left, numberstyle=\scriptsize\color{mygray}, language=ML,showspaces=false,showstringspaces=false,xleftmargin=0pt, numbersep=-6pt, morekeywords=[1]{try,spliced,tyfam,opfam,let,fn,val,def,casetype,objtype,metadata,of,*,string,lexer,parser,datatype,new,toast,notation,module,switch,where,expansions,require,import,for,ana,syn,opcon,tycon,metasignature,metamodule,metasig,metamod,static,at,by,tycase,mod,macro,match,pattern,in,patterns,expressions,implicit,forall,int,exptsm,pattsm},deletekeywords={double,structure,of},classoffset=0, xleftmargin=0pt,
 aboveskip=3pt,belowskip=2pt,
 moredelim=**[is][\color{red}]{SSTR}{ESTR},
-moredelim=**[is][\color{OliveGreen}]{SHTML}{EHTML},
+moredelim=**[is][\color{Periwinkle}]{SHTML}{EHTML},
 moredelim=**[is][\color{purple}]{SCSS}{ECSS},
@@
 moredelim=**[is][\color{gray}]{SNAT}{ENAT}, 
-moredelim=**[is][\color{OliveGreen}]{SURL}{EURL},
-moredelim=**[is][\color{SeaGreen}]{SQT}{EQT},
+moredelim=**[is][\color{Periwinkle}]{SURL}{EURL},
+moredelim=**[is][\color{Sealavender}]{SQT}{EQT},
 moredelim=**[is][\color{Periwinkle}]{SGRM}{EGRM},
-moredelim=**[is][\color{YellowGreen}]{SID}{EID},
+moredelim=**[is][\color{Yellowlavender}]{SID}{EID},
 moredelim=**[is][\color{Sepia}]{SUS}{EUS},
+moredelim=**[is][\color{black}]{SOK}{EOK},
 % deletestring=[d]{"},
@@
 \newcommand{\li}[1]{\lstinline[basicstyle=\ttfamily\fontsize{9pt}{1em}\selectfont]{#1}}
+\newcommand{\lismall}[1]{\lstinline[basicstyle=\ttfamily\fontsize{9pt}{1em}\selectfont]{#1}}
+\newcommand{\lifootnote}[1]{\lstinline[basicstyle=\ttfamily\fontsize{7pt}{1em}\selectfont]{#1}}
 
@@
 
-\input{macros}
+\input{thesis/macros}
 
@@
 This paper questions this common practice: why should an otherwise ordinary data structure be privileged in the definition of a general-purpose programming language?
-General-purpose programming languages typically define literal notation for only a small number of common data structures, e.g.
-lists.
+General-purpose programming languages typically define literal notation for only a small number of common data structures, like lists.
 This is unsatisfying because there are many other data structures for which literal notation might be useful, e.g.
-finite maps, regular expressions, HTML data, SQL queries, syntax trees and chemical structures.
-There may also be different implementations of each of these data structures, perhaps with different performance characteristics, that could all benefit from common literal notation.
+finite maps, regular expressions, HTML elements, SQL queries, syntax trees for various languages and chemical structures.
+There may also be different implementations of each of these data structures behind a common signature that could all benefit from common literal notation.
 %A less \emph{ad hoc} design would be one that allows library providers to define new literal forms in a decentralized manner.
@@
 % Partial parameter application lowers the syntactic cost of this strict style. 
-Compared to existing approaches, TLMs are uniquely \emph{reasonable}: TLM providers can reason modularly about syntactic ambiguity, and TLM clients can reason abstractly, i.e.
-without examining the underlying expansion, about types and binding.
-The system only needs to convey to clients, via secondary notation, the inferred \emph{segmentation} of each literal, which gives the locations and types of spliced subterms.
+Compared to existing approaches, TLMs are uniquely \emph{reasonable}.
+TLM clients can reason abstractly, i.e.
+without examining grammars or generated expansions, about types and binding.
+The system only needs to convey to clients, via secondary notation, the inferred \emph{segmentation} of each literal body, which gives the locations and types of spliced subterms.
+TLM providers can reason modularly about syntactic ambiguity and expansion correctness according to clear criteria.
+This paper integrates TLMs into Reason, an emerging alternative front-end for OCaml, and demonstrates, through several non-trivial case studies, how TLMs integrate with the advanced features of OCaml, including pattern matching and the module system.
+We discuss optional integration with MetaOCaml, which allows TLM providers to be more confident about type correctness.
+Finally, we establish these abstract reasoning principles formally with a detailed type-theoretic account of expression and pattern TLMs for ``core ML''.
+%This is a hygienic term-rewriting macro system, of any design, for a language with a ML-like type structure and pattern matching. 
 %Maintaining information about segmentation is necessary for the novel TLM hygiene mechanism, i.e.
@@
 % This mechanism captures many common syntactic idioms while avoiding the problem of syntactic conflicts by construction and supplying clients with clear abstract reasoning principles. 
-This paper establishes these abstract reasoning principles formally with a calculus of typed expressions, pattern matching and ML-style modules. 
-This calculus is the first detailed type-theoretic account of a hygienic macro system, of any design, for a language with these essential features of ML. 
-We are integrating TLMs into Reason, an emerging alternative front-end for OCaml.% This calculus is the first such formalization of a typed macro system.
+% This calculus is the first such formalization of a typed macro system.
 
@@
 When designing the surface syntax of a general-purpose programming language, it is common practice to define shorthand \emph{literal notation}, i.e.
-notation that decreases the syntactic cost of constructing and pattern matching over values of a particular data structure or parameterized family of data structures.
+notation that decreases the syntactic cost of constructing and pattern matching over values of some particular data structure or parameterized family of data structures.
 For example, many languages in the ML family support list literals like \li{[x1, x2, x3]} in both expression and pattern position \cite{harper1997programming,mthm97-for-dart}.
@@
 % We will return to this concept again later.} 
-While lists are common across problem domains, other literal notation is more specialized.
+Lists are common across problem domains, but other literal notation is more specialized.
 For example, Ur/Web extends the surface syntax of Ur (an ML-like language \cite{conf/pldi/Chlipala10}) with expression and pattern literals for encodings of XML and HTML data~\cite{conf/popl/Chlipala15}.
-The example in Fig.
-\ref{fig:urweb} shows two HTML literals, one that ``splices in'' a string expression delimited by \li{\{[} and \li{]\}} and the other an HTML expression delimited by \li{\{} and \li{\}}.
+The example in Fig.~\ref{fig:urweb} shows two HTML literals, one that ``splices in'' a string expression delimited by \li{\{[} and \li{]\}} and the other an HTML expression delimited by \li{\{} and \li{\}}.
 %The desugarings of these literal expressions, not shown, are substantially more verbose and less readable to programmers familiar with XHTML.
@@
 This design practice, where the language designer privileges certain library constructs with built-in literal notation, is \emph{ad hoc} in that it is easy to come up with other examples of data structures for which mathematicians, scientists or programmers have invented specialized notation \cite{DBLP:journals/cacm/Iverson80,cajori1928history,TSLs}.
-For example, 1) clients of a ``collections'' library might want not just list literals, but also literal notation for matrices, finite sets, maps and so on; 2) clients of a ``web programming'' library might want CSS literals (which Ur/Web lacks); 3) a compiler author might want ``quotation'' literals for the terms of the object language and various intermediate languages of interest; and 4) clients of a ``chemistry'' library might want chemical structure literals based on the SMILES standard \cite{anderson1987smiles}.
+For example, 1) clients of a ``collections'' library might want not just list literals, but also literal notation for matrices, finite maps, and so on; 2) clients of a ``web programming'' library might want CSS literals (which Ur/Web lacks); 3) a compiler author might want ``quotation'' literals for the terms of the object language and various intermediate languages of interest; and 4) clients of a ``chemistry'' library might want chemical structure literals based on the SMILES standard \cite{anderson1987smiles}.
 
@@
 Instead, there has been persistent interest in mechanisms that allow library providers to define new literal notation on their own.
-For example, direct grammar extension systems like Camlp4 \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, term rewriting systems like Template Haskell~\cite{SheardPeytonJones:Haskell-02,mainland2007s}, the system of \emph{type-specific languages} (TSLs) described by \citet{TSLs}, and other systems that we will discuss below can all be used to define new literal notation (and, in some cases, other forms of new notation, such as new infix expression forms, control flow operations or type declaration forms, which we leave beyond the scope of this paper).
+For example, direct grammar extension systems like Camlp4 \cite{de2003camlp4} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, term rewriting systems like Template Haskell~\cite{SheardPeytonJones:Haskell-02,mainland2007s}, the system of \emph{type-specific languages} (TSLs) described by \citet{TSLs}, and other systems that we will discuss below can all be used to define new literal notation (and, in some cases, other forms of new notation, such as new infix expression forms, control flow operations or type declaration forms, which we leave beyond the scope of this paper).
 
@@
 \begin{figure}[t!]
-\begin{subfigure}[t]{0.48\textwidth}
+\vspace{4px}
+\begin{subfigure}[t]{0.46\textwidth}
 \fcolorbox{gray!20}{gray!20}{\makebox[0.97\textwidth][l]{\textsc{existing grammar extension systems}}}
-\vspace{-12px}
 \begin{lstlisting}[xleftmargin=-2pt, morekeywords={EXTEND}]
@@
   /* ...more extensions defined... */
-  let w = compute_w();
-  let x = compute_x(w);
+  let x = compute_x();
   let y = `((!R)@&{&/x!/:2_!x}'!R)`;
@@
 \vspace{-5px}
-\caption{It is difficult to reason abstractly about programs that use unfamiliar grammar extensions (see the six reasoning criteria in the text).}
+\caption{It is difficult to reason abstractly given program text that uses a variety of grammar extensions (see the six reasoning criteria in the text).}
 \label{fig:K-dialect}
@@
 \fcolorbox{gray!20}{gray!20}{\makebox[\textwidth][l]{\textsc{\textbf{this paper}: typed literal macros}}}
-\vspace{-12px}
 \begin{lstlisting}[xleftmargin=-2pt]
-  notation $kq at KQuery.t 
-    lexer  KQueryLexer
-    parser KQueryParser.start
+  notation $kq at KQuery.t {
+    lexer  KQueryLexer;
+    parser KQueryParser.start;
     expansions require
-      module KQuery as KQuery;
-  /* ...more notations defined... */
-  let w = compute_w();
-  let x = compute_x(w);
+      { module KQuery as KQuery }
+  }; /* ...more notations defined... */
+  let x = compute_x();
   let y = $kq `(SURL(!R)@&{&/EURLxSURL!/:2_!EURLxSURL}'!REURL)`;
@@
 \vspace{-5px}
-\caption{TLMs make examples like these more reasonable by leaving the base grammar fixed and making the type and binding structure and the segmentation explicit.}
+\caption{TLMs make examples like these more reasonable by leaving the base grammar fixed and strictly enforcing a simple type, binding and segmentation discipline.}
 \label{fig:K-tsm-example}
@@
 
-Consider, for instance, the perspective of a programmer trying to comprehend the program text in Fig.
-\ref{fig:K-dialect}, which is written in a  popular dialect of OCaml's surface syntax called Reason \cite{reason-what}\todo{up-to-date?}{} that has, hypothetically, been extended with some number of new literal forms by a grammar extension system --- Lines 1-5 show the grammar extension syntax from Camlp4 \cite{ocaml-manual}, but Sugar*/SugarHaskell is  similar \cite{erdweg2011sugarj,erdweg2013framework,erdweg2012layout}.
-Line 9 uses one of these active syntax extensions to construct an encoding of a query in the niche database query language K, using its intentionally terse notation~\cite{Whitney:2001:LOR:376284.375783}.
+Consider, for instance, the perspective of a programmer trying to comprehend the program text in Fig.~\ref{fig:K-dialect}, which is written in an emerging dialect of OCaml's surface syntax called Reason \cite{reason-what} that has, hypothetically, been extended with some number of new literal forms by a grammar extension system --- Lines 1-5 show the grammar extension syntax from Camlp4 \cite{de2003camlp4}, but Sugar*/SugarHaskell is  similar \cite{erdweg2011sugarj,erdweg2013framework,erdweg2012layout}.
+Line 9 uses one of these active syntax extensions to construct an encoding of a query in the rather obscure database query language K, using its intentionally terse notation~\cite{Whitney:2001:LOR:376284.375783}.
 The problem is that a programmer looking at the program as presented, and unfamiliar with (i.e.
@@
 \item Unhygienic approaches like OCaml's preprocessor extension point (PPX) rewriters \cite{ocaml-manual} and Template Haskell \cite{SheardPeytonJones:Haskell-02,mainland2007s} allow us to define new literal notation with support for splicing by repurposing existing string literal forms.
-They also partially or completely solve the problem of reasoning about \textbf{Responsibility} by using explicit annotations on terms that are being rewritten.
-However, they do not satisfy the remaining five reasoning criteria.
-\item Hygienic staging macro systems, like MetaOCaml and related systems \cite{DBLP:conf/flops/Kiselyov14,ganz2001macros,Sheard:1999:UMS} do not give the expander access to the syntax trees of arguments, i.e.
-the rewriting is parametric in the arguments, so they are not suitable for the problem of defining new literal notation.
+They also partially or completely solve the problem of reasoning about \textbf{Responsibility} but they do not satisfy the remaining five reasoning criteria.
+\item Hygienic staging macro systems, like MetaOCaml and related systems \cite{DBLP:conf/flops/Kiselyov14,ganz2001macros,Sheard:1999:UMS} are unsuitable for defining literal notation because they do not give the expander access to syntax trees.%, i.e.
+the rewriting is parametric in the arguments, so they are unsuitable for defining new notation.
 \item Hygienic term rewriting macro systems, like those in various Lisp-family languages \cite{DBLP:conf/esop/HermanW08,Herman10:Theory,Kohlbecker86a,DBLP:conf/popl/Adams15,DBLP:conf/popl/ClingerR91,DBLP:journals/lisp/DybvigHB92,mccarthy1978history,Flatt:2012:CLR:2063176.2063195}, e.g.
-Racket \cite{Flatt:2012:CLR:2063176.2063195}, as well as Scala \cite{ScalaMacros2013}, do not allow us to repurpose string literal forms to define composite literal forms because the hygiene discipline does not allow us to splice terms out of literal bodies via parsing (the system cannot identify these terms as sub-terms of the macro arguments \cite{DBLP:conf/esop/HermanW08,Herman10:Theory}).% Only existing composite forms can therefore be repurposed.
+Racket \cite{Flatt:2012:CLR:2063176.2063195}, as well as Scala \cite{ScalaMacros2013}, do not allow us to flexibly repurpose string literal forms to define composite literal forms because the hygiene discipline cannot account for terms spliced out of literal bodies via parsing.%(the system cannot identify these terms as sub-terms of the macro arguments \cite{DBLP:conf/esop/HermanW08,Herman10:Theory}).% Only existing composite forms can therefore be repurposed.
 \item \emph{Type-specific languages} (TSLs) \cite{TSLs} come closer to our goals in that they explicitly support splicing terms out of literal bodies, but the mechanism falls short with regard to the six reasoning principles just discussed in ways that we will return to in Sec.
@@
 \paragraph{Contributions} This paper introduces \emph{typed literal macros} (TLMs): the first system for defining new literal notation that (1) provides the ability to reason abstractly about all six of the topics just outlined; and (2) is semantically expressive enough for integration into Reason/OCaml and other full-scale statically typed functional languages.
-We evaluate these claims with (1) a series of increasingly ML-like calculi equipped with proofs of the claimed abstract reasoning principles; and (2) a number of non-trivial examples appearing throughout the paper that involve the advanced language features mentioned above.
-In describing these examples, we also demonstrate that literal parsing logic can be defined using standard, unmodified parser generators, so the burden on notation providers is comparable to that of existing systems despite these stronger reasoning principles.
+We evaluate these claims with a number of non-trivial examples appearing throughout the paper that involve the advanced language features mentioned above.
+In describing these examples, we demonstrate that literal parsing logic can be defined using standard, unmodified parser generators, so the burden on notation providers is comparable to that of existing systems despite these stronger reasoning principles.
+Finally, we give a type-theoretic account of TLMs where we formally establish these abstract reasoning principles.
 % From the client's perspective, which is the focus of our contributions, TLMs are uniquely reasonable.
@@
 
-\paragraph{Brief Overview} For a brief overview of our proposed mechanism, consider Fig.~\ref{fig:K-tsm-example}.
-On Lines~1-5, we define a TLM named \li{$kq} that provides K query literal notation.
-On Line 9, we apply this TLM to express the example from Figure \ref{fig:K-dialect}.
+\paragraph{A Brief Overview} For a brief overview of the proposed mechanism, consider Fig.~\ref{fig:K-tsm-example}.
+Lines~1-5 define a TLM named \li{$kq} that provides K query literal notation.
+Line 9 applies this TLM to express the example from Figure \ref{fig:K-dialect}.
 We can reason abstractly about this program as follows.
@@
 So by examining Line 9, the programmer knows that the two instances of \li{x} are spliced expressions (because they are in black), whereas the \li{R}'s must be parsed in some other way, e.g.
-as operators of the K language (because they are in green).
+as operators of the K language (because they are in lavender).
 This also implies that errors in spliced terms can always be reported in terms of their original location.
@@
 % \item 
-Sec.~\ref{sec:setsms} details expression TLMs in Reason with a more detailed case study. 
+Sec.~\ref{sec:setlms} details expression TLMs in Reason with several more case studies of varying detail, notably including TLMs for regular expressions, Ur/Web style HTML literals, chemical structure literals and quasiquotation for Reason language terms, which can be used for implementing other TLMs.
+It also describes experimental integration with MetaOCaml, which can help providers reason about type correctness.
 % \item 
-Sec.~\ref{sec:sptsms} then introduces {pattern TLMs} and describes the special reasoning conditions in pattern position. %The hygiene condition for pattern TLMs is interesting.
+Sec.~\ref{sec:sptsms} then briefly introduces {pattern TLMs} and describes the special reasoning conditions in pattern position. %The hygiene condition for pattern TLMs is interesting.
 % \item 
 Sec.~\ref{sec:ptsms} introduces the more general {parametric TLMs}, which allow us to define literal notation at a type- or module-parameterized family of types.
-This section also reveals that the \li{expansions require} clause can be understood as parameterization followed immediately by partial parameter application.
-Having introduced the basic machinery, we continue in Sec.~\ref{sec:more-examples} with examples that further demonstrate the expressive power of this approach and various parser implementation strategies.
+This section also reveals that the \li{expansions require} clause can be understood as parameterization followed immediately by partial parameter application, or, alternatively, that parameterization can be understood in terms of the \li{expansions require} clause together with ML's module functions (functors).
+Having introduced the basic machinery by example, we proceed in
+%Sec.~\ref{sec:more-examples} with examples that further demonstrate the expressive power of this approach and various parser implementation strategies.
 Notably, this section gives our take on the example from  Fig.~\ref{fig:urweb} of Ur/Web-style HTML literals as well as a TLM that implements quasiquotation for Reason language terms, which is useful for implementing other TLMs.
-Sec.~\ref{sec:setsms-formally} defines a type-theoretic calculus of simple expression and pattern TLMs and formally establishes the reasoning principles implied above in their simplest form. 
-Sec.~\ref{sec:ptlms-formally} adds type functions and an ML-style module system to this calculus, and gives a more general variant of the reasoning principles theorem. 
-Sec.~\ref{sec:implementation} provides a brief overview of how we are implementing TLMs  without modifying OCaml's type system. 
+Sec.~\ref{sec:setlms-formally} to define a type-theoretic account of simple expression and pattern TLMs and formally establish the reasoning principles implied above in their essential form. 
+% Sec.~\ref{sec:ptlms-formally} adds type functions and an ML-style module system to this calculus, and gives a more general variant of the reasoning principles theorem. 
+Sec.~\ref{sec:implementation} provides a brief overview of how we are implementing TLMs for Reason without modifying OCaml's type system.
+% We also give a brief overview of optional integration with MetaOCaml, which can be used to guarantee that all expansion generated by a TLM are type correct.
 % \item 
@@
 Sec.~\ref{sec:discussion} concludes with a summary of the contributions of this paper and a discussion of limitations and future work.
-Certain technical details, proofs and extensions to the presented calculi are available in the supplement.
+Certain technical details, proofs and additional details on the case studies are available in the supplement.
 % Also available is an additional series of case studies: notation for an encoding of regular expressions as an abstract data type, implemented using TLMs for quasiquotation and grammar-based parser generators.
@@
 
-\vspace{-3px}
 \newcommand{\seTLMsSec}{Expression TLMs in Reason}
 \section{\protect\seTLMsSec}
-\label{sec:setsms}
-\vspace{-2px}
+\label{sec:setlms}
 
@@
 Regexes are  common in, for example, bioinformatics, where they are used to express patterns in DNA sequences.
-We might construct a regex that matches the strings \li{"A"}, \li{"T"}, \li{"G"} or \li{"C"}, which represent the four bases in DNA, as follows:
+We can construct a regex that matches the strings \li{"A"}, \li{"T"}, \li{"G"} or \li{"C"}, which represent the four bases in DNA, as follows:
 \begin{lstlisting}[numbers=none]
@@
 \end{lstlisting}
-In Reason, the notation \li{Regex.(}$e$\li{)} locally opens the module \li{Regex} within the expression $e$, so we did not need to qualify each application of \li{Regex.Or} and \li{Regex.Str} above.
+Note that in Reason, the notation \li{Regex.(}$e$\li{)} locally opens the module \li{Regex} within the expression $e$, so we do not need to qualify each application of \li{Regex.Or} and \li{Regex.Str}.
 Even with the aid of this shorthand, however, constructing regexes in this way is syntactically costly.
 Instead, we would like to have the option to use the common POSIX-style notation  \cite{STD95954} when constructing values of type \li{Regex.t}, including values constructed compositionally from other regexes and strings.
-We solve this problem in Fig.
-\ref{fig:regex-tlm-def} by defining a TLM named \li{$regex} (pronounced ``lit regex'') that supports POSIX-style regex notation extended with splice forms for regexes and strings.
-Fig.
-\ref{fig:first-tlm-example} shows three examples of \li{$regex} being applied.
+We solve this problem in Fig.~\ref{fig:regex-tlm-def} by defining a TLM named \li{$regex} (pronounced ``lit regex'') that supports POSIX-style regex notation extended with splice forms for regexes and strings.
+Fig.~\ref{fig:first-tlm-example} shows three examples of \li{$regex} being applied.
 % One approach would be to simply define a function \li{parse_rx : string -> Regex.t} that parses a string at run-time to produce a regex.
@@
   module Regex = {
-    type t = AnyChar 
+    type t = Empty
+           | AnyChar 
            | Str(string)
@@
            | Star(t);
-  };
+  }
 \end{lstlisting}
@@
 \begin{lstlisting}[mathescape=|]
-  module RegexNotation {
-    notation $regex at Regex.t
-      lexer   RegexLexer
-      parser  RegexParser.start
+  module RegexNotation = {
+    notation $regex at Regex.t {
+      lexer  RegexLexer;
+      parser RegexParser.start;
       expansions require
-        module Regex as Regex;
-  };
+        { module Regex = Regex; }
+    }
+  }
 \end{lstlisting}
@@
 \begin{subfigure}[t]{\textwidth}
-\vspace{4px}
+\vspace{5px}
 \begin{lstlisting}
-  open RegexNotation;
+  notation $regex = RegexNotation.$regex; /* or open RegexNotation */
   module DNA = { let any_base = $regex `(SURLA|T|G|CEURL)`; };
   let bisA = $regex `(SURLGC$(EURLDNA.any_baseSURL)GCEURL)`;
-  let restriction_template(gene) => 
+  let restriction_template = (gene) => 
     $regex `(SURL$(EURLbisASURL)$(EURLDNA.any_baseSURL)*$$(EURLgeneSURL)$(EURLDNA.any_baseSURL)*$(EURLbisASURL)EURL)`;
@@
 \caption{Case Study: POSIX-style regex literal notation, with support for string and regex splicing.}
-\vspace{-6px}
+% \vspace{-6px}
 \label{fig:regex-case-study}
@@
 
-\vspace{-3px}
-\subsection{Client Perspective}
-Let us start from the perspective of a client programmer examining Fig.
-\ref{fig:regex-case-study} but holding the underlying expansion, as well as the details of the lexer and parser, \li{RegexLexer} and \li{RegexParser}, abstract.
+\subsection{Client Perspective}\label{sec:client-perspective}
+Let us start from the perspective of a client programmer examining Fig.~\ref{fig:regex-case-study} but holding the underlying expansion of Fig.~\ref{fig:first-tlm-example}, as well as the details of the lexer and parser, \li{RegexLexer} and \li{RegexParser}, abstract.
 We will return to describe the lexer and parser from the provider's perspective in Sec.
@@
 
-% Fig.
-\ref{fig:first-tlm-example}, then use these examples to detail the client-side abstract reasoning principles that the mechanism provides, all before finally .
-% The \li{open notation} directive on Line 2 of Fig.
-\ref{fig:first-tlm-example} allow qualifying each TLM application in the file.
+% Fig.~\ref{fig:first-tlm-example}, then use these examples to detail the client-side abstract reasoning principles that the mechanism provides, all before finally .
+% The \li{open notation} directive on Line 2 of Fig.~\ref{fig:first-tlm-example} allow qualifying each TLM application in the file.
 %reasoning about this program from the perspective of a client who wants to avoid looking at the lexer, parser and the underlying expansion itself.
 
-Line 2 of Fig.
-\ref{fig:first-tlm-example} applies \li{$regex} to construct the regex \li{DNA.any_base} that was  described above, this time using the more clear and familiar POSIX regex notation.
+Line 2 of Fig.~\ref{fig:first-tlm-example} applies \li{$regex} to construct the regex \li{DNA.any_base} that was  described above, this time using the more concise and common POSIX regex notation.
 Line~3 applies \li{$regex} again, using its regex splice form to compositionally construct a regex matching DNA sequences recognized by the BisA restriction enzyme, where the middle base can be any base.
@@
 \end{lstlisting}
-According to the context-free grammar of (this paper's branch of) Reason, this form is simply a leaf of the unexpanded syntax tree, like a string literal would be. TLM names are prefixed by \li{$} 
+According to the context-free grammar of (this paper's version of) Reason, this form is simply a leaf of the unexpanded parse tree, like a string literal would be. TLM names are prefixed by \li{$} 
  to distinguish them from variables.
-We call the TLM argument, \li{`(GC$(DNA.any_base)GC)`}, a \emph{generalized literal form}, following the terminology introduced in the prior work on type-specific languages \cite{TSLs}.
-The only lexical constraint on the \emph{body} of a generalized literal form, i.e.
+We call the TLM argument, \li{`(GC$(DNA.any_base)GC)`}, a \emph{generalized literal form}, following the terminology introduced in  prior work on type-specific languages~\cite{TSLs}.
+The only lexical constraint imposed on the {literal body}, i.e.
 the characters between \li{`(} and \li{)`}, is that any nested occurrences of \li{`(} must be balanced by \li{)`}, much like nested comments in Reason/OCaml.
-Generalized literal forms therefore lexically subsume other literal forms.
-The nesting constraint is to allow TLM applications to appear inside spliced expressions (discussed below).
-Prior work by \citet{TSLs} specified several other choices of outer delimitation, including layout-sensitive delimitation, but for the purposes of this work, we will use the outer delimiters \li{`(} and \li{)`} exclusively.
-Choices about how spliced expressions are recognized inside the literal body are entirely at the discretion of each TLM.
-In this case, notice that the TLM provider has chosen to use \li{$(}$e$\li{)} for regex splicing and \li{$$(}$e$\li{)} for string splicing, where $e$ is an unexpanded Reason expression of arbitrary form.
-We return to the topic of splicing below.
-
+Generalized literal forms therefore lexically subsume many other literal forms.
+This nesting constraint is to allow TLM applications to appear inside spliced expressions.
+An example of nested TLM application is shown in Fig.~\ref{fig:nesting-example}, discussed later in this section.
+Related work by \citet{TSLs}, discussed in Sec.~\ref{sec:existing-approaches}, specified several other choices of outer delimitation, including layout-sensitive delimitation, but for the purposes of this paper, \li{`(} and \li{)`} suffice.
 % Notice that in the unexpanded expression, we have not yet distinguished the spliced expression \li{any_base} from the rest of the literal.
@@
 
-\paragraph{Responsibility} 
-Responsibility for lexing, parsing and expanding each literal body is delegated uniquely to the lexer and parser specified in the definition of the applied TLM by the clauses \li{lexer} and \li{parser}, respectively.
-For now, let us purposefully continue to hold these details abstract.% and take the perspective of a client that does not want to delve into these details.
-
-TLM definitions follow the scoping structure of Reason, i.e.
-they can be defined in modules alongside other definitions.
-When a TLM definition appears inside a module, it must also appear in the signature with the same specification, up to the usual notions of type and module path equivalence in the type annotation and module dependencies, which are discussed below.
+\subsubsection{Responsibility} 
+Responsibility for lexing, parsing and expanding each literal body is delegated uniquely to the applied TLM.
+% Choices about how spliced expressions are recognized inside the literal body are entirely at the discretion of each TLM.
+In this case, the TLM provider has chosen to use \li{$(}$e$\li{)} for regex splicing and \li{$$(}$e$\li{)} for string splicing, where $e$ is an unexpanded Reason expression of arbitrary form.
+% We return to these topics below.% reasoning about splicing works when encountering an unfamiliar literal form below.
+% and take the perspective of a client that does not want to delve into these details.
+TLM definitions follow the same scoping rules as Reason modules, i.e.
+they can be defined in modules alongside other definitions and accessed through module paths.
+When a TLM definition appears inside a module with an explicitly specified module type, it must also appear in the module type with the same specification, up to the usual notions of type and module path equivalence in the type annotation and module dependencies, which are discussed below.
 This is much like the situation with datatype definitions in ML.
 In Sec.
-\ref{sec:implementation}, we will discuss how we use an encoding trick involving nested modules and temporary exceptions internally in the implementation to avoid having to primitively add TLM definitions to OCaml's type and module system.
-
-By convention, we define TLMs in a module suffixed with \li{Notation} so that client programmers can \li{open} just the TLM definition without bringing other definitions into scope, as shown on Line 1 of Fig.
-\ref{fig:first-tlm-example}.
-We could also have fully qualified each TLM application, \li{RegexNotation.$regex}, or abbreviated the TLM definition:
-\begin{lstlisting}[numbers=none]
-    let $regex = RegexNotation.$regex;
-\end{lstlisting}
-
-What is fundamental about this design is that there is a well-defined and familiar protocol for finding the definition of the TLM uniquely responsible for each generalized literal form in a program, following the usual scoping rules of the language.
+\ref{sec:implementation}, we will describe how we use an encoding of TLM definitions as modules with singleton signatures to avoid having to primitively extend OCaml's module system.
+
+What is fundamental about this design is that there is a well-defined protocol for finding the definition of the TLM uniquely responsible for each generalized literal form in a program, following the usual scoping rules of the language.
 An editor service or documentation tool can use this protocol to integrate TLM definition lookup into a ``go to definition'' command.
-
-% In Reason, TLM definitions like the one in Fig.
-\ref{fig:regex-tlm-def} are separately defined in files with the suffix \li{.relit}, which are processed and brought into scope by the Reason build system, \li{rebuild}.
+Moreover, clients can define scoped abbreviations as shown on Line 1 of Fig.~\ref{fig:first-tlm-example}, again following the usual scoping rules.
+By convention, we define TLMs in a module suffixed with \li{Notation} so that client programmers can \li{open} just the relevant TLM definition(s) without bringing other definitions into scope.
+We will demonstrate a simple lexically scoped implicit application mechanism for situations where the same TLM is being repeatedly applied  in Sec.
+\ref{sec:regexparser}.
+
+% In Reason, TLM definitions like the one in Fig.~\ref{fig:regex-tlm-def} are separately defined in files with the suffix \li{.relit}, which are processed and brought into scope by the Reason build system, \li{rebuild}.
 This is consistent with the protocol that other syntax extension systems use, e.g.
 the syntax libraries in Sugar* \cite{erdweg2011sugarj,erdweg2013framework,erdweg2012layout}.
-In our theoretical treatment in Sec.~\ref{sec:setsms-formally}, however, TLM definitions will be integrated directly into the scoping structure of the language.
+In our theoretical treatment in Sec.~\ref{sec:setlms-formally}, however, TLM definitions will be integrated directly into the scoping structure of the language.
 This is not possible in Reason because this would require extending OCaml's semantics with TLM definitions in module signatures.
@@
 
-% nitions (Fig.
-\ref{fig:regex-tlm-def}) appear in \li{.relit} files, and they are brought into scope by the build system, \li{rebuild}, according to the usual protocols for library discovery.
+% nitions (Fig.~\ref{fig:regex-tlm-def}) appear in \li{.relit} files, and they are brought into scope by the build system, \li{rebuild}, according to the usual protocols for library discovery.
 The fully qualified name of the \li{--}% In the theory that we develop in Sections \ref{sec:s-UL}, TLM definitions occur directly within code.
@@
 
-\paragraph{Expansion Typing}
-Having found the definition of the \li{$regex} TLM, the client can immediately determine the type of the expansion being generated at each application site because it is specified explicitly by the clause \li{at Regex.t} on Line~2 of Fig.
-\ref{fig:regex-tlm-def}.
+\subsubsection{Expansion Typing}
+Having found the definition of the \li{$regex} TLM, the client can immediately determine the type of the expansion being generated at each application site because it is specified explicitly by the clause \li{at Regex.t} on Line~2 of Fig.~\ref{fig:regex-tlm-def}.
 The expansion type of a TLM is analagous to the return type of a function.
@@
 
-\paragraph{Context Dependence}
-The TLM mechanism enforces a strong context independence condition on generated expansions by requiring that the TLM definition explicitly specify all modules that the expansions it generates might internally depend on.
-In this case, Lines 5-6 of Fig.
-\ref{fig:regex-tlm-def} specify that generated expansions might use the module \li{Regex}, again as it is bound at the TLM definition site, using the module variable \li{Regex} internally.
-In this case, the expansion-internal module variable happens to coincide with the module path, but in general, the dependency can be an arbitrary module path.
-For example, consider the following dependency clause:
-\begin{lstlisting}[numbers=none]
-  expansions require 
-    module Regex as Regex
-    module Core.List as List
-\end{lstlisting}
-The module variable \li{List} will be bound internally to \li{Core.List} in all expansions generated by this TLM, including expansions generated at application sites where \li{List} or \li{Core.List} are not bound, or bound to different modules.
-All other bindings, whether at the TLM definition site or the TLM application site, are not internally available to the expansion.
-This allows both client programmers and the provider of the package where the TLM is defined to freely rename and shadow variables in the usual way, i.e.
-without needing to inspect each expansion or each TLM definition's parser to determine whether the syntax trees that it generates require certain variables be bound in particular ways at each application site (as is common with, e.g., syntax extensions based on \li{camlp4} or Sugar*, or when using unhygienic term rewriting approaches like OCaml's PPX system or Template Haskell, which we discuss further in Sec.
-\ref{sec:existing-approaches}).
-
-Enforcing this strong context independence condition is technically quite subtle because TLM parsers need to be able to perform splicing, i.e.
+\subsubsection{Context Dependence}
+\label{sec:context-dependence}
+The system enforces a strong context independence condition on generated expansions by requiring that the TLM definition explicitly specify all modules that the expansions it generates might internally depend on.
+In this case, Lines 5-7 of Fig.~\ref{fig:regex-tlm-def} specify that generated expansions might use the module \li{Regex}, again as it is bound at the TLM definition site, using the module variable \li{Regex} internally.
+In  general, the dependency can be an arbitrary module path, e.g.
+\li{module List = Core.Std.List}.
+The \li{Pervasives} module is implicitly opened in each expansion.
+All other bindings, both at the TLM definition site and the TLM application site, are not internally available to the expansion.
+
+% For example, consider the following dependency clause:
+% \begin{lstlisting}[numbers=none]
+%   expansions require { 
+%     module Regex = Regex;
+%     module List = Core.List;
+%   }
+% \end{lstlisting}
+% The module variable \li{List} will be bound internally to \li{Core.List} in all expansions generated by this TLM, including expansions generated at application sites where \li{List} or \li{Core.List} are not bound, or bound to different modules.
+
+From the client programmer's perspective, the benefit of an application site context independence discipline is clear---clients do not need to give any thought to which bindings the expansion might invisibly be assuming are in scope.
+This is a common problem with various unhygienic approaches, e.g., \li{camlp4} \cite{de2003camlp4}, Sugar* \cite{erdweg2011sugarj,erdweg2013framework,erdweg2012layout}, PPX rewriters \cite{ocaml-manual} or Template Haskell \cite{SheardPeytonJones:Haskell-02,mainland2007s}.
+
+The benefits of making the macro definition site dependencies explicit, rather than implicitly allowing expansions to access all bindings at the definition site as in many existing macro systems, are more subtle.
+One technical benefit is that this exposes the TLM's dependencies in the signature of the module where the TLM is defined.
+This is useful for build tools that extract dependencies by source code analysis.
+Moreover, if an ``internal'' module is being used by expansions, then this will be manifest in the signature and can be corrected if this was not intended.
+% (Similar concerns about unintended capture drives recent work on explicit closures for distributed computing \cite{DBLP:conf/ecoop/MillerHO14}.) 
+Implicit access to the  definition site would require the implementation to carefully ``smuggle'' internal values to each application site, thereby skirting the abstraction discipline of ML's module system \cite{culpepper2005syntactic}.% (and making the implementation substantially more complex!)\todo{I think there is some work on this...}  %This, in turn, ensures that the expansion does not depend on bindings inaccessible outside the module, i.e.
+the implementation does not need to ``smuggle'' references to these internal definitions to each application site.
+Consequently, the abstraction discipline of the ML module system is maintained.
+
+Another benefit of this explicit approach is that renamings can be propagated without needing to modify AST encodings in parser definitions.
+For example, if we decided to rename the module \li{Regex} to \li{Regexp}, this renaming would need only to propagate to the module paths in the TLM definition in the usual manner as long as the names of the internal module variables remain the same.
+(The TLM provider may wish, for the sake of aesthetics, to propagate the change into the parser definition as well, but this is not semantically necessary).
+This allows us to encode expansions using ordinary OCaml values (see Sec.
+\ref{sec:provider-perspective}).
+
+Enforcing this strong context independence condition is technically subtle because TLM parsers need to be able to implement splicing, i.e.
 they need to be able to parse  terms out of the literal body for placement in the expansion.
-Na\"ively checking that only the explicitly named dependencies are free in the expansion would inappropriately enforce context independence on application site spliced expressions, which should certainly not be prevented from referring to variables in scope at the application site.
-For example, consider the final expansion of our example from Line 2 of Fig.
-\ref{fig:first-tlm-example}:
+Na\"ively checking that only the explicitly named dependencies are free in the expansion would inappropriately constrain application site spliced expressions, which should certainly not be prevented from referring to variables in scope at the application site.
+For example, consider the final expansion of our example from Line 2 of Fig.~\ref{fig:first-tlm-example}:
 \begin{lstlisting}[numbers=none]
@@
 
-As we will discuss more extensively in Sec.~\ref{sec:existing-approaches}, this issue is why hygienic term-rewriting macro systems, like those available in various Lisp-family languages \cite{mccarthy1978history} and in Scala \cite{ScalaMacros2013}, cannot be used to repurpose string literals for composite literal notation at other types.
-These systems would find that the appearance of \li{DNA.any_base} in the generated expansion violates their context independence condition, because \li{DNA.any_base} is not, from the perspective of the context-free syntax, an argument to the macro nor even a sub-term of an argument for which a tree path can be assigned \cite{DBLP:conf/esop/HermanW08,Herman10:Theory}.
-Instead, it arises as the result of performing a complex operation---parsing---on a sub-string of the string literal provided as the macro argument.
+Hygienic term-rewriting macro systems, like those available in various Lisp-family languages \cite{mccarthy1978history} and in Scala \cite{ScalaMacros2013}, cannot be used to repurpose string literals for composite literal notation at other types for exactly this reason.
+These systems would find that the appearance of \li{DNA.any_base}  violates their context independence condition (which is one aspect of {hygiene}), because \li{DNA.any_base} is not, from the perspective of the context-free syntax, an argument to the macro nor even a sub-term of an argument for which a tree path can be assigned \cite{DBLP:conf/esop/HermanW08,Herman10:Theory,gorn1965explicit}.
+Instead, it arises as the result of performing a complex operation---parsing---on an arbitrary sub-sequence of the string literal passed into the macro for rewriting.
+%From the macro expander's perspective, it has appeared by magic''.
 
 To address this problem, TLM parsers are not tasked with generating the final expansion directly.
-Instead, the parser generates a \emph{proto-expansion} that refers to spliced terms indirectly by their location relative to the start of the provided literal body.
+Instead, the parser generates a \emph{proto-expansion} that refers to spliced terms indirectly by location relative to the start of the provided literal body.
 For example, the proto-expansion generated by \li{$regex} for the example above can be pretty printed as follows:
 \begin{lstlisting}[numbers=none]
-  Regex.Seq(Regex.Str "GC", Regex.Seq(spliced<4; 15; Regex.t>, Regex.Str "GC"))
+  Regex.Seq(Regex.Str "GC", Regex.Seq(spliced<4..15 : Regex.t>, Regex.Str "GC"))
 \end{lstlisting}
-Here, \li{spliced<4; 15; Regex.t>} is a reference to the spliced expression \li{DNA.any_base} by its location relative to the start of the literal body being expanded---characters $4$ through $15$ (zero-indexed and inclusive) of the literal body \li{GC$(DNA.any_base)GC} are \li{DNA.any_base}.
-We say more about the type annotation on the splice reference when we discuss \textbf{Segment Typing} below.
-The context independence condition can be enforced straightforwardly on the proto-expansion -- the only free variable in the proto-expansion is \li{Regex}, which is an explicitly listed dependency, so all is well.
-Had a reference to \li{DNA} appeared in the proto-expansion, then it would be clear that the corresponding final expansion could only be valid in certain contexts (where \li{DNA} is bound) and not others, and therefore this would be a violation of the context independence condition.
-\todo{why not spliced[e] somewhere}
+Here, \lismall{spliced<4..15 : Regex.t>} is a reference to the spliced expression \li{DNA.any_base} because characters $4$ through $15$, zero-indexed and inclusive, of \lismall{GC$(DNA.any_base)GC} are \lismall{DNA.any_base}.
+We return to the type annotation on the splice reference, \li{Regex.t}, in Sec.
+\ref{sec:segment-typing} below.
+The context independence condition can be enforced straightforwardly on the proto-expansion---the only free variable in the proto-expansion is \li{Regex}, which is an explicitly listed dependency, so all is well.
+
+% Had a reference to \li{DNA} appeared in the proto-expansion, it would be clear that it did not originate in a spliced expression.
+Leaving it in the final expansion would mean that the expansion would be valid only in contexts where \li{DNA} is appropriately bound, i.e.
+\li{DNA} would be an invisible assumption that clients could not reason abstractly about.
+More subtly, using the binding of \li{DNA} from the TLM definition site would also be problematic because if \li{DNA} is renamed, it would not be obvious that this renaming needs to be propagated into the OCaml values that represent proto-expansions inside the parser (see Sec.
+\ref{sec:provider-perspective} for further discussion of AST representations).
+We avoid both of these issues by making dependencies explicit.
+%  can propagate into the TLM definition without changing the internal module variable being used.% Similar approaches have worked well in other situations where there are subtleties %Recent work on closures in distributed programming took a similar approach of making the variables  explicit, albeit for different reasons \cite{DBLP:conf/ecoop/MillerHO14}\todo{cite heather's paper}\todo{talk somewhere about how you could use implicit dependencies if you defined parser in-line using MetaOCaml, but this would have other trade-offs}
 % Requiring that TLMs refer to spliced expressions indirectly in this manner ensures that a TLM cannot ``forge'' spliced terms, i.e.
@@
 
-\paragraph{Segmentation}
+\subsubsection{Segmentation}\label{sec:segmentation}
 The finite set of splice references in the proto-expansion generated for a literal body is called the \emph{segmentation} of that literal body.
-The segmentation of the example above is the finite set containing one element, \li{spliced<4; 15; Regex.t>}.
-For the more complex example from Line 4 of Fig.
-\ref{fig:first-tlm-example}, the segmentation contains multiple spliced segments:
+The segmentation of the example above is the finite set containing one splice reference, \lismall{spliced<4..15 : Regex.t>}.
+For the more complex example from Line 5 of Fig.~\ref{fig:first-tlm-example}, the segmentation contains five splice references:
 \begin{lstlisting}[numbers=none]
-  { spliced<2; 5; Regex.t>, spliced<9; 20; Regex.t>; spliced<26; 29; string>; 
-    spliced<33; 44; Regex.t>; spliced<49; 52; Regex.t> } 
+  { spliced<2..5 : Regex.t>, spliced<9..20 : Regex.t>, spliced<26..29 : string>, 
+    spliced<33..44 : Regex.t>, spliced<49..52 : Regex.t> } 
 \end{lstlisting}
@@
 When TLM applications are nested, a distinct color can be used at each depth.
-For example, in Sec.
-\ref{sec:more-examples}, we will describe a TLM \li{$html} for HTML notation in the style of Ur/Web (see Fig.~\ref{fig:urweb}).
-We can also define a TLM, \li{$smiles}, not shown, that allows for the use of the standard SMILES notation for chemical structures \cite{anderson1987smiles}, extended with splicing notation.
-Assuming that the expansion type of \li{$smiles} is \li{Smiles.t} and that \li{Smiles.to_svg : Smiles.t -> Html.t}, we can nicely transform encodings of chemical structures into a vector graphic embedded into an HTML page as follows:
+For example, Fig.~\ref{fig:nesting-example} shows a program fragment where we transform an encoding of a chemical structure expressed using the standard SMILES notation for chemical structures \cite{anderson1987smiles}, extended with splicing notation, into a vector graphic, then embed this directly into a fragment of a webpage.
+We detail the TLM \li{$html at Html.t}, which implements HTML notation similar to that found in Ur/Web, in the supplement, and assume a TLM, \li{$smiles at Smiles.t} and a function, \li{Smiles.to_svg : Smiles.t => Html.t}, not shown.
+In Reason, \li{|>} is reverse function application.
+\begin{figure}[t]
 \begin{lstlisting}[numbers=none]
-    $html `(
-      SURL<div>Chemical structure of sucrose: {EURL
-        Smiles.to_svg
-          $smiles `(SCOLOR{ECOLORmono_glucoseSCOLOR}-O-{ECOLORmono_fructoseSCOLOR}ECOLOR)` 
-      SURL}</div>EURL
-    )`
+  $html `( SURL<div>
+             <h3>Chemical Structure of Sucrose</h3>
+             <$>EURL$smiles `(SCOLOR{ECOLORmono_glucoseSCOLOR}-O-{ECOLORmono_fructoseSCOLOR}ECOLOR)` |> Smiles.to_svgSURL</$>
+           </div>EURL )`
 \end{lstlisting}
-
-Another benefit of explicitly tracking the locations of spliced segments is that errors that originate in spliced terms can be reported in terms of their original source location \cite{DBLP:journals/jsc/DeursenKT93}.
-
-\paragraph{Segment Typing} Each splice reference in the segmentation specifies not just the location of a spliced expression but also its expected type.
+\caption{A practical demonstration of nested TLM application. The colors communicate the segmentation.}
+\vspace{-3px}
+\label{fig:nesting-example}
+\end{figure}
+
+\subsubsection{Segment Typing}\label{sec:segment-typing} Each splice reference in the segmentation carries not just the location of a spliced expression but also its expected type.
 The identity of this type is resolved in a context-independent manner, assuming only the dependencies explicitly specified by the TLM.
 
-By associating a type with each spliced segment, we guarantee that type inference can be performed abstractly, meaning that only the segment types, together with the expansion types specified by the applied TLMs, are necessary to infer types for variables appearing in a client-side function.
-For example, consider the function \li{restriction_template} on Lines 3-4 of Fig.
-\ref{fig:first-tlm-example}.
-The return type of this function can be inferred to be \li{Regex.t} from the type annotation on the \li{$regex} TLM, as previously discussed.
+By associating a type with each spliced segment, type inference can be performed abstractly, meaning that only the segment types, together with the expansion types specified by the applied TLMs, are necessary to infer types for variables appearing in a client-side function.
+For example, consider the function \li{restriction_template} on Lines 4-5 of Fig.~\ref{fig:first-tlm-example}.
+The return type of this function can be inferred to be \li{Regex.t} from the expansion type annotation on the \li{$regex} TLM, as previously discussed.
 The type of the argument, \li{gene}, can be inferred to be \li{string} because the segmentation specifies the type \li{string} for the spliced segment where it appears (cf.
-the segmentation shown under \textbf{Segmentation} above).
+the segmentation shown in Sec.
+\ref{sec:segmentation} above).
 The context independence condition implies that \li{gene} cannot appear elsewhere in the expansion, and so no further typing constraints could possibly be collected from examining the portions of the (proto-)expansion being held abstract.
+Another important benefit of explicitly tracking the locations of spliced segments is that errors that originate in spliced terms can be reported in terms of their original source location \cite{DBLP:journals/jsc/DeursenKT93}.
 
 Segment types can be communicated directly to the programmer upon request by an editor service.
-For Reason, we are extending the {Merlin} editor service \cite{Merlin}, which is used by various Reason editor extensions (e.g.
-for Emacs, Vim, and so on), with a new editor command for reporting the expected type of the innermost spliced segment containing the cursor.
-Note that because the type is specified explictly in the output of the parser, this information can be reported even when there is a parse error or type error in a spliced expression.
+For Reason, we are equipping the {Merlin} tool \cite{Merlin}, which is used by various Reason editor modes (e.g.
+for Emacs, Vim, and so on), with a new editor command that reports the expected type of the innermost spliced segment containing the cursor.
+Note that because the type is explicit, this information can be reported even when there is a parse error or type error in a spliced expression.
 % An alternative design that simply reported the type assigned to each spliced expression based on the final expansion (which already exists in Merlin) would be less useful in situations where there is a type error in a spliced expression, or the programmer is in the process of editing the program and wants to know what type of expression to construct.
@@
 The difference is that the argument signature of a function is the same every time the function is applied, i.e.
-it is associated with the function itself, whereas the inferred segmentation can differ at each TLM application site.
-This is, of course, what gives TLMs their notational flexibility.
-
-\paragraph{Capture}
-In discussing the question of inferring a type for \li{gene} above, we neglected to consider one critical question: are we sure that the variable \li{gene} in the third spliced segment on Line 4 of Fig.
-\ref{fig:first-tlm-example} is, in fact, a reference to the argument \li{gene} of the \li{restriction_template} function?
+it is associated with the function itself, whereas the inferred segmentation can differ for each choice of literal body.
+This, of course, gives TLMs substantially more notational flexibility, even relative to infix or mixfix function notation (where the number of subexpressions is fixed \cite{wieland2009parsing}), as intended.% (even infix or mixfix functions \cite{wieland2009parsing}).
+
+\subsubsection{Capture}
+In discussing the question of inferring a type for \li{gene} above, we neglected to consider one critical question: are we sure that the variable \li{gene} in the third spliced segment on Line 5 of Fig.~\ref{fig:first-tlm-example} is, in fact, a reference to the argument \li{gene} of the \li{restriction_template} function?
 After all, if we hold the expansion abstract then it may well be that the third spliced segment appears under (i.e.
@@
 For more common identifiers, e.g.
-\li{x} and \li{tmp}, inadvertent capture is perhaps inevitable.
-
-For example, consider the following application site:
+\li{tmp}, inadvertent capture is not difficult to imagine.
+For example, consider this application site:
 \begin{lstlisting}[numbers=none]
   let tmp = /* ... application site temporary ... */;
-    $html `(SURL<h1>{EURLf(tmp)SURL}</h1>EURL)`;
+  $html `(SURL<h1><$>EURLf(tmp)SURL</$></h1>EURL)`;
 \end{lstlisting}
@@
   let tmp = /* ... expansion-internal temporary ... */;
-    Html.H1Element(tmp, spliced<5; 10; Html.t>);
+  Html.H1Element(tmp, spliced<7..12 : Html.t>);
 \end{lstlisting}
-If the final expansion was produced na\"ively, by syntactically replacing the spliced segment reference by the final expansion recursively determined for the corresponding spliced expression, then the  variable \li{tmp} in the spliced expression would capture the binding of \li{tmp} in the proto-expansion.
-The result when the types of the two bindings differed would be a type error that exposes the internal details of the expansion.
-If the types of the two bindings of \li{tmp} coincided, then there could be subtle changes in run-time behavior that would be impossible to diagnose without looking at the expansion itself.
-Capture of invisible bindings is, due to these hazards, clearly unreasonable.
-
-To address this problem in our design, splicing is guaranteed to be capture-avoiding.
-The final expansion is generated by recursively expanding each spliced expression and then inserting it into the final expansion via capture-avoiding substitution.
-In other words, the system automatically alpha-varies the bindings in the proto-expansion to ensure capture avoidance.
-There is no need for TLM providers to manually deploy a mechanism that generates fresh variables (as in, e.g., Racket's \emph{reader macro} system \cite{Flatt:2012:CLR:2063176.2063195}, which is further discussed in Sec.
+If the final expansion was produced na\"ively, by syntactically replacing the splice reference with the final expansion recursively determined for the corresponding spliced expression, then the  variable \li{tmp} in the spliced expression would capture the expansion-internal binding of \li{tmp}.
+The result if the types of the two bindings differed would be a type error exposing the internal details of the expansion.
+If the types of the two bindings of \li{tmp} coincided, then there would be no static indication of the problem but there could be subtle and mysterious changes in run-time behavior.
+% Capture of invisible bindings is, due to these hazards, clearly unreasonable.
+
+To address this problem, splicing is guaranteed to be capture-avoiding.
+The final expansion is generated by recursively expanding each spliced expression and then inserting it into the final expansion via capture-avoiding substitution, which automatically alpha-varies the internal bindings of the proto-expansion as necessary.
+There is no need for TLM providers to manually deploy a mechanism that generates fresh variables (as in, e.g., Racket's reader macros \cite{Flatt:2012:CLR:2063176.2063195}, further discussed in Sec.
 \ref{sec:existing-approaches}).
-For example, the final expansion of the example above would be alpha-equivalent to the following:
+For example, the final expansion of the example above is alpha-equivalent to the following:
 \begin{lstlisting}[numbers=none]
   let tmp = /* ... application site temporary ... */; 
-  let tmp' = /* ... expansion-internal temporary (automatically renamed) ... */;
-  Html.H1Element(tmp', f(tmp));
+  let tmp_fresh = /* ... expansion-internal temporary ... */;
+  Html.H1Element(tmp_fresh, f(tmp));
 \end{lstlisting}
@@
 
-Although this capture avoidance discipline implies that TLMs cannot intentinoally expose bindings internal to the expansion directly to spliced expressions, this does not imply that values cannot flow from the expansion into a spliced expression.
-It simply means that when this is intended, the segment type must be a function type.
-Reason has concise lambda notation, \li{(x) => e}, which is useful in many of these use cases.
-For example, the capture avoidance discipline means that we cannot define numeric list comprehension notation\footnote{A polymorphic list comprehension TLM requires type parameters, introduced in Sec.
-\ref{sec:ptsms}.} like the following:
+Although this strict capture avoidance discipline implies that TLMs cannot intentionally introduce  bindings directly into spliced expressions, this does not imply that values cannot flow from the expansion into a spliced expression.
+It simply means that when this is intended, the segment type must be a function type, which serves to make this interface explicit.
+Reason's concise lambda notation, \li{(x) => e}, decreases the syntactic cost of this approach.
+For example, we cannot define list comprehension notation like the following:
 \begin{lstlisting}[numbers=none,deletekeywords={in}]
-  $listcomp `(x + 1SURL | x in EURLloSURL to EURLhi)` (* NO! cannot reason abstractly *)
+  $listcomp `(x + 1SURL | x in EURLloSURL .. EURLhi)` /* NO! cannot reason abstractly */
 \end{lstlisting}
-However, the following is permitted, because \li{x} is bound by the spliced lambda expression:
+However, the following is permitted, because \li{x} is bound by the spliced lambda expression and the corresponding segment type makes the type of the interface explicit:
 \begin{lstlisting}[numbers=none]
-  $listcomp `((x) => x + 1 SURL|EURL lo SURLtoEURL hi)`  (* OK! *)
+  $listcomp `((x) => x + 1 SURL|EURL lo SURL..EURL hi)`  /* OK! */
 \end{lstlisting}
 
-Our contention is that small syntactic costs like these are more than justified by the peace of mind of knowing that unfamiliar literal notation is absolutely prevented from obscuring the type and binding structure of the language.
-We say more about the future prospect of a  mechanism designed specifically for specialized binding forms, e.g.
+Our contention is that small syntactic costs like these are more than justified by the peace of mind of knowing that unfamiliar literal notation cannot possibly be performing ``magic'' on the type and binding structure of the language.
+We say more about the future prospect of a  mechanism designed specifically for shorthand binding forms, e.g.
 Haskell-style \li{do} notation \cite{jones2003haskell}, in Sec.
-\ref{sec:discussion}.\todo{do this}% For now, the existing infix operator support in Reason and OCaml suffices for working with monadic values.
+\ref{sec:discussion}.% For now, the existing infix operator support in Reason and OCaml suffices for working with monadic values.
+
+%A polymorphic list comprehension TLM requires type parameters, introduced in Sec. \ref{sec:ptsms}.
 
@@
 Reason 3.0 is adopting more concise function syntax, which will decrease the cost of this syntax.
-
-\paragraph{Summary}
-
-Let us briefly reiterate the key ideas of this section.
-The focus was on providing simple protocols to client programmers when they have questions about the syntactic structure, type structure or binding structure of a program.
-At no point in answering these sorts of questions is the client made to look at the generated expansion itself, nor inspect the parser implementation.
-Instead, the programmer need only be given knowledge of the expansion type from the TLM definition and the segmentation inferred by the expander at each application site, which carries a small volume of information that can be communicated straightforwardly by standard editor services.
-Certain questions related to binding structure simply do not need to be asked due to the strict context independence and capture avoidance discipline of the system, enabled by the explicit tracking of spliced segments by location.
-Despite these semantic constraints, the system is able to express a number of interesting examples with few compromises because there is only one simple lexical constraint on literal bodies.
-
----------------------- {\color{red} stuff below is notes / not yet revised}
 
@@
 
+\vspace{-5px}
 \subsection{Provider Perspective}
 \label{sec:provider-perspective}
-
-% \vspace{-3px}\paragraph{Parse Errors} 
-If the parse function determines that the literal body is not well-formed according to the syntax that it implements, it must return \li{ParseError \{msg=}$e_\text{msg}$\li{, loc=}$e_\text{loc}$\li{\}}
-where $e_\text{msg}$ is a custom error message and $e_\text{loc}$ is a value of type \li{segment}, defined in Figure \ref{fig:indexrange-and-parseresult}, that designates a segment of the literal body as the origin of the error \cite{DBLP:journals/jsc/DeursenKT93}.
-
-% \vspace{-3px}\paragraph{Success} 
-If instead parsing succeeds, the parse function returns \li{Success} ~$\ecand$, 
-where $\ecand$ is called the \emph{encoding of the proto-expansion}.
+\begin{figure}
+\begin{subfigure}[t]{0.30\textwidth}
+\begin{lstlisting}[morekeywords={rule,parse,eof}]
+  { 
+  open RegexParser;
+  let readsplice = 
+    Relit.Segment.read_to(")");
+  let unescape = (s) => 
+    String.sub(s, 1, 1); 
+  }
+  let special = 
+    ['SCSS\\ECSS' 'SCSS.ECSS' 'SCSS|ECSS' 'SCSS*ECSS' 'SCSS+ECSS' 
+     'SCSS?ECSS' 'SCSS(ECSS' 'SCSS)ECSS' 'SCSS$ECSS']
+  let not_special = _#special
+  let escape = 'SCSS\\ECSS' special
+  rule read = 
+    parse
+    | "." { DOT }
+    | "|" { BAR }
+    | "*" { STAR }
+    | "+" { PLUS }
+    | "?" { QMARK }
+    | "(" { LPAREN }
+    | ")" { RPAREN }
+    | not_special+ as s
+      { STR(s) }
+    | escape as s
+      { STR(unescape(s)) }
+    | "$(" 
+      { SPLICED_REGEX
+        (readsplice(lexbuf)) }
+    | "$$("  
+      { SPLICED_STRING
+        (readsplice(lexbuf)) }
+    | eof { EOF }
+\end{lstlisting}
+\vspace{-4px}
+\caption{\li{RegexLexer.mll}}
+\label{fig:regex-lexer}
+\end{subfigure}
+\hfill
+\begin{subfigure}[t]{0.61\textwidth}
+\begin{lstlisting}[morekeywords={token,left,start},stringstyle=\ttfamily\color{black}]
+  %{ 
+  open notation Relit.$proto_expr; 
+  %}
+  %token DOT BAR STAR PLUS QMARK LPAREN RPAREN EOF
+  %token <string> STR
+  %token <Relit.Segment.t> SPLICED_REGEX 
+  %token <Relit.Segment.t> SPLICED_STRING
+  %left BAR
+  %start <Relit.ProtoExpr.t> staSOKEOKrt
+  %%
+  staSURLEURLrt:
+    | e = regex; EOF { e }
+    | EOF { `(SURL Regex.Empty EURL)` }
+  regex:
+    | DOT { `(SURL Regex.AnyChar EURL)` }
+    | s = STR { `(SURL Regex.Str($s `(EURLsSURL)`) EURL)` }
+    | r1 = regex; r2 = regex 
+      { `(SURL Regex.Seq(`(EURLr1SURL)`, `(EURLr2SURL)`) EURL)` }
+    | r1 = regex; BAR; r2 = regex 
+      { `(SURL Regex.Or(`(EURLr1SURL)`, `(EURLr2SURL)`)EURL )` }
+    | r = regex; STAR { `(SURL Regex.Star(`(EURLrSURL)`) EURL)` }
+    | r = regex; PLUS
+      { `(SURL leEURLSURLt r = `(EURLrSURL)`;
+           Regex.Seq(r, Regex.Star(r)) EURL)` }
+    | r = regex; QMARK 
+      { `(SURL Regex.Or(Regex.Empty, `(EURLrSURL)`) EURL)` }
+    | LPAREN; r = regex; RPAREN { r }
+    | seg = SPLICED_REGEX 
+      { `(SURL $spliced `(EURLsegSURL : Regex.t)` EURL)` }
+    | seg = SPLICED_STRING 
+      { `(SURL Regex.Str
+           ($spliced `(EURLsegSURL : strEURLSURLing)`) EURL)` }
+\end{lstlisting}
+\vspace{-4px}
+\caption{\li{RegexParser.mly}}
+\label{fig:regex-parser}
+\end{subfigure}
+\vspace{2px}
+\caption{The lexer and parser for the \li{$regex} TLM from Fig.~\ref{fig:regex-tlm-def}. See Fig.~\ref{fig:relit-util} for relevant definitions from \li{Relit}.}
+\label{fig:lexer-and-parser}
+\vspace{-5px}
+\end{figure}
+
+Let us turn now to the perspective of the TLM provider, whose principal task is to define the lexer and parser named in the TLM definition.
+In our implementation of TLMs for Reason, lexers are generated using \li{ocamllex} \cite{ocaml-manual} and parsers using Menhir \cite{pottier2016menhir}, which is a modern derivative of the parser generator Yacc~\cite{johnson1975yacc,TarditiDR:mly} with support for LR(1) grammars \cite{jourdan2012validating}.
+Reason itself is implemented using these same tools.
+We will say more about how lexer and parser definitions are resolved by our implementation in Sec.
+\ref{sec:implementation}.
+What is fundamentally important is that there is some clear protocol by which the lexer and parser names and the starting non-terminal name specified by the TLM definition, in this case \li{RegexLexer} and \li{RegexParser.start} on Lines 3-4 of Fig.~\ref{fig:regex-tlm-def}, are resolved to lexer and parser implementations that can be evaluated at expansion-time, i.e.
+during compilation.
+
+The \li{ocamllex} and Menhir definitions used to implement \li{RegexLexer} and \li{RegexParser} are given in Fig.~\ref{fig:lexer-and-parser}.
+A fragment of the signature of the \li{Relit} library, which provides various necessary definitions and utilities to TLM providers, is shown in Fig.~\ref{fig:relit-util}.
+After describing this example in detail, we will outline some alternative parsing strategies.
+
+\begin{figure}
+\begin{lstlisting}[deletekeywords={spliced}]
+  module Segment : { 
+    type t = {start_pos: int, end_pos: int};
+    let read_to : string => Lexing.lexbuf => Segment.t;
+  };
+  exception ExpansionError({msg: string, loc: option(Segment.t)});
+  module ProtoExpr : { 
+    type t = Parsetree.expression; /* from the OCaml compiler library */
+    let spliced : Segment.t => Parsetree.core_type => t;
+  };
+  notation $re_expr at Parsetree.expression { /* ... (see text) ... */ };
+  notation $re_type at Parsetree.core_type { /* ... (see text) ... */ };
+  notation $proto_expr at ProtoExpr.t = $re_expr;
+\end{lstlisting}
+\caption{A fragment of the \li{Relit} module signature relevant to expression TLM providers.}
+\label{fig:relit-util}
+\vspace{-4px}
+\end{figure}
+    % let match_spliced : t => option(Segment.t, ProtoTyp.t);
+
+\subsubsection{\li{RegexLexer}}
+\label{sec:regex-lexer}
+In most respects, the definition of the lexer in Fig.~\ref{fig:regex-lexer} is conventional.
+It reads various token patterns on Lines 6-23, emitting the corresponding tokens specified by the \lstinline[morekeywords={token}]{%token} declarations in Fig.~\ref{fig:regex-parser}, which have been brought into scope by Line 2 of Fig.~\ref{fig:regex-lexer}.
+Lines 20-21 combine sequences of non-special characters (e.g.
+alphanumeric characters) into a single string token.
+Lines 22-23 implement backslash-prefixed escape sequences for characters that have special meaning in this notation for regexes, emitting a string token in each instance.\footnote{A better approach would be to define a second lexing rule that greedily combines sequences consisting of non-special characters and escape characters into a single \li{STR} token, but for the sake of exposition, we stick to this simpler approach.}
+
+Lines 24-31 of the lexer, which recognize the notation for regex splicing \li{$(}$e$\li{)}, and string splicing, \li{$$(}$e$\li{)}, are more unusual.
+Recall from Sec.~\ref{sec:context-dependence} that the parser must represent spliced terms abstractly, by their location within the literal body.
+This implies that the TLM does not need to parse spliced expressions itself, as long as it can independently decide where the spliced expressions start and end.
+A paired start and end position is a segment,  represented by a value of the type \li{Segment.t} defined on Line 2 of Fig.~\ref{fig:relit-util}.
+Lines 26-27 and Lines 30-31 of the lexer produce a segment by calling a helper function, \li{Relit.Segment.read_to}, that internally invokes the Reason lexer on the provided lexing buffer until it sees an instance of the provided token outside of a Reason comment, string literal or generalized literal.
+If the provided token is a right delimiter also used by Reason, like \li{")"} on Line 4 of Fig.~\ref{fig:regex-lexer}, then \li{read_to} looks for the first unmatched instance.
+For example, if the remaining lexing buffer contains {\small \texttt{f("))"))A|G}} then \li{read_to} will consume up through the final instance of \li{)} and emit the start and end position, excluding the closing delimiter, relative to the original buffer.
+
+This approach has two major benefits.
+First, it supports splicing even within literal notations that have a very different lexical structure from Reason itself, as in this example.
+Second, it allows the parser to avoiding needing to link to Reason's expression grammar (though this is also supported by Menhir).
+Spliced expressions arrive as single, opaque tokens as shown on Lines 6-7 of Fig.~\ref{fig:regex-parser}, and do not need to be parsed by the TLM at all.
+The system will recursively process spliced expressions after the parser has finished generating the proto-expansion.
+
+Similar helper functions, not shown, are available for reading out single identifiers, simple paths like \li{X1.X2.xy} and a few other useful fragments of the Reason syntax, so that TLM providers can implement lower-cost splicing notation that clients can choose in situations where the spliced expression is of a restricted form that makes the end position of the spliced expression unambiguous without matching delimitation.
+As a matter of practice, we recommend that TLM providers start with a fully delimited form before adding optional alternative shorthand in order to avoid artificially restricting the client programmer's ability to directly convey the compositional structure of a value.
+Although \li{Segment.t} is not held abstract in order to support certain advanced scenarios, discussed below, we recommend that TLM providers generally treat it as such.
+
+\subsubsection{\li{RegexParser} and Quasiquotation}\label{sec:regexparser}The job of a TLM parser is to generate a proto-expansion given the tokens generated by the lexer, or if this is not possible, to indicate an expansion error by raising either a Menhir parse error or \li{Relit.ExpansionError} (Line 5 of Fig.~\ref{fig:relit-util}) with an appropriate error message and, if possible, an error location.
+In the case of expression TLMs, the proto-expansion must be a proto-expression (we consider proto-patterns in the next section).
+The Menhir directive on Line 9 of Fig~\ref{fig:regex-parser}, which specifies the starting non-terminal of the grammar, reveals that proto-expressions are values of type \li{Relit.ProtoExpr.t}.
+This type is defined on Line 7 of Fig.~\ref{fig:relit-util} as a synonym for \li{Parsetree.expression}, which is the standard representation of expression parse trees exposed by the OCaml compiler \cite{ocaml-manual}.% by many external tools, including the Reason toolchain.
+
+To be able to repurpose this existing parse tree representation for proto-expansions, there must be some way to unambiguously represent splice references, notated earlier in the paper as \li{spliced<m..n : ty>} where \li{m..n} specifies a segment and \li{ty} is the segment's expected type.
+The \li{ProtoExpr} module solves this problem by providing a function \li{spSURLEURLliced} that takes a segment, of type \li{Segment.t}, and a type representation of type \li{Parsetree.core_type}, which also comes from OCaml's compiler library, and produces a corresponding value of type \li{ProtoExpr.t} that uniquely represents the corresponding splice references.
+Internally, we use OCaml's parse tree annotations.
+
+The datatypes in OCaml's \li{Parsetree} module are necessarily intricate, given the sophistication of the OCaml system, so constructing expression encodings manually is generally too unwieldy.
+Addressing this class of problem is, of course, the motivation for this very paper, so it is only natural that we define TLMs  for working with OCaml parse trees using familiar surface syntax, extended with splice forms.
+In the literature, literal notation for the host language's own parse trees is called \emph{quasiquotation}, and splicing is referred to as \emph{unquotation} or \emph{antiquotation} \cite{Bawd99a,shabalin2013quasiquotes,mainland2007s}.
+The TLMs \li{$re_expr} and \li{$re_type} specified on Lines 10-11 of Fig.~\ref{fig:relit-util} implement quasiquotation for OCaml syntax trees using Reason's surface syntax.
+The supplement describes their implementation.
+For clarity, the \li{$proto_expr} TLM is defined as a synonym for \li{$re_expr} on Line~12 of Fig.~\ref{fig:relit-util} (the two TLMs may diverge in the future).
+We apply this TLM many times in the semantic actions in Fig.~\ref{fig:regex-parser}.
+Notice, however, that the generalized literal forms in Fig.~\ref{fig:regex-parser} are not each prefixed by \li{$proto_expr}.
+Instead, we use the \li{open notation} directive on Line 2, which implicitly applies \li{$proto_expr} to all unadorned generalized literal forms following the usual lexical scoping rules for \li{open}.
+
+To support antiquotation, the \li{$re_expr} TLM repurposes TLM application and generalized literal forms, as can be observed throughout Fig.~\ref{fig:regex-parser}.
+In particular, parse tree splicing is supported by unadorned generalized literal forms, e.g.
+on Line~23 where we implement the regex notation \li{r+} in terms of sequencing and star, taking care to bind \li{r} to an expansion-internal variable---also \li{r} but distinguished by the segmentation---to avoid double evaluation.
+String splicing (which converts the spliced string expression into a parse tree of string constant form) is performed by repurposing the TLM name \li{$s} (Line~16).
+There is also notation for references to spliced expressions that repurposes the TLM name \li{$spliced}, as seen on Lines 28-32, where the regex splicing logic is implemented.
+
+TLM-related forms are available to be repurposed in this way by \li{$proto_expr} because there is no reasonable way for \li{$re_expr} to support antiquotation across notational boundaries.
+For example, if \li{$re_expr} had instead used some other antiquotation notation, e.g.
+\li{^(}$e$\li{)}, then consider this example:
+\begin{lstlisting}[numbers=none]
+  $proto_expr `( SURLleEURLSURLt y = ^(EURLr1SURL); $m `(...{^(r2)}...)`EURL )`
+\end{lstlisting}
+The sequence \li{^(r1)} is in expression position, so antiquotation occurs.
+However, there is no way for \li{$re_expr} to know how the TLM that \li{$m} will eventually refer to will parse the sequence \li{^(r2)}---it may end up in a spliced expression, or it may have some other meaning---so it cannot perform antiquotation.
+In other words, TLM expansion is fundamentally ``outside-in''.
+Notice that the segmentation does communicate the fact that \li{r2} has not been antiquoted (it remains lavender).
+However, to avoid confusion, we decided not to support the generation of TLM application forms via \li{$re_expr}, and instead repurposed that notation for antiquotation.
+
+This issue does \textbf{not} come up when using TLMs to define quotation literals for languages other than Reason itself (an example is given in Sec.~\ref{sec:sptsms}), so if this design proves too limiting, it may be reasonable to build ``inside-out'' quasiquotation primitively into Reason as a special case.
+In fact, this is already possible using MetaOCaml, discussed below.
+%(Nested TLM application via splicing is, of course, useful and fully supported, e.g.
+in Fig.~\ref{fig:nesting-example}).
+ % Typed quotations with an explicit representation of contexts would need to be built in to the language  necessary to even think about expanding a
+
+\subsubsection{Correctness Criteria}
+In order to maintain the abstract reasoning principles discussed in Sec. \ref{sec:client-perspective}, the system \emph{validates} each proto-expansion generated by the parser as follows:
+\begin{enumerate}
+  \item First, the segmentation is computed and validated. This involves checking two criteria:
+    \begin{enumerate}
+      \item The segments must be in-bounds, of positive extent, non-overlapping and separated by at least one character.
+      \item The segment types must encode valid types assuming only the TLM dependencies.
+    \end{enumerate}
+  \item Second, the proto-expansion is typechecked assuming only the TLM dependencies (plus opened \li{Pervasives}), treating each splice reference as a variable of the specified segment type.
+\end{enumerate}
+If proto-expansion validation fails, the client is notified that the applied TLM is incorrectly implemented.
+The proto-expansion validation criteria are exactly the correctness criteria that parser writers must consider, quantifying over all possible literal bodies for which a proto-expansion is generated, in order to avoid exposing this class of compile-time error to clients.
+
+\subsubsection{MetaOCaml}\label{sec:metaocaml} In ``vanilla'' OCaml, it is not possible to express a datatype that classifies only valid proto-expansions according to the correctness criteria above.
+The MetaOCaml type system comes closer by extending the semantics of OCaml with values of type \li{'a code}, which are constructed using a primitive typed quasiquotation operation \cite{DBLP:conf/flops/Kiselyov14,taha2004gentle}.
+In MetaOCaml-BER, which is an active implementation of (an extension of) MetaOCaml as a fork of the compiler, and \li{ppx_stage}, which implements a subset of MetaOCaml as a compiler plugin \cite{ppxstage}, a value of type \li{'a code} is represented as, and can be coerced safely to, an OCaml parse tree for the corresponding expression of type \li{'a} \cite{DBLP:conf/flops/Kiselyov14}.
+It is therefore possible to implement a TLM parser using \li{'a code} values internally, doing the coercion only at the end.
+However, the typing guarantee is relative to the context where the quotation was constructed, i.e.
+the parser's implementation, so a validation step is still needed to ensure that the assumptions are consistent with those specified by the \li{expansions require} clause.
+
+The segmentation criteria cannot be statically enforced and, in both mentioned implementations of MetaOCaml, it is impossible to express splice references directly because they do not yet support terms that contain explicit type annotations.
+There are various workarounds, e.g.
+by defining for each possible segment type \li{t} a dummy value of type \li{t code} and then post-processing the parse tree after the coercion step.
+Overall, however, this approach can substantially increase a TLM provider's confidence in the type correctness of the expansion logic, with the caveats just mentioned.
+With \li{ppx_stage}, it is possible to compile the parser separately without requiring that clients use the plugin.
+TLMs can be applied inside quoted code as long as the applied TLM is accessed through a dependency of the TLM being defined (so that internal dependency references correctly resolve).
+%MetaOCaml-BER does not change the parse tree or bytecode representation of the OCaml system, and values of type \li{'a code} are not part of the public interface of a TLM parser, so it should, in principle, be possible to compile a parser using MetaOCaml-BER for use in  ``vanilla'' OCaml (and therefore Reason) programs.
+However, MetaOCaml-BER does not suitably package the necessary pieces of its run-time system, so we have not verified this conjecture.
+
+\subsubsection{Other Implementation Strategies}
+Although \li{ocamllex} and Menhir are powerful tools, they are not right for every parsing job.
+For example, we might want to use a different parser generator, a parser combinator library \cite{Hutton1992d}, or post-process the result of calling an existing parser to produce a corresponding proto-expansion.
+Fortunately, it is easy to bypass \li{ocamllex} and Menhir:
+%The most obvious method is to define a trivial lexer that reads the entire literal body into a single token, and a trivial parser with one semantic action.
+In fact, this is not actually necessary because the system does not interface with \li{ocamllex} and Menhir directly.
+Rather, it is the job of the user's build system to make sure that by the time the compiler is invoked, these tools have already been invoked as necessary to produce ordinary OCaml modules satisfying the expected interface.
+It is just as well to implement this interface manually.
+the \li{Relit} library includes a trivial lexer, \li{Relit.TrivLexer}, and a few functors that take care of the necessary boilerplate in situations where we simply want to define the parser as a function of type \li{string => Relit.ProtoExpr.t}.
+For example, the supplement briefly describes how \li{$re_expr} can be expressed in terms of the Reason parser, a parse tree serializer and a tree transformation.
+It also details how the \li{$html} TLM shown previously is implemented using an existing, ``production grade'' HTML parsing library, \li{Markup.ml} \cite{markupml}, together with a simple post-processing step.
+It is also worth noting that although there is no mechanism for applying TLMs at the module level, this can be emulated by using OCaml's first class modules.
+
+  % These names are resolved as if they were at the top level of the compilation unit, not as module paths at the TLM definition site (because the resulting module must not refer to bound module variables, e.g.
+as bound by enclosing functors, or else it would be impossible to resolve the lexer and parser implementations at compile-time).
+
+% Lexer and parser names are interpreted relative to the top level of the compilation unit. The lexer must be a module with a signature compatible with
+
+% ---------------------- {\color{red} stuff below is notes / not yet revised}
+% % \vspace{-3px}\paragraph{Parse Errors} 
+% If the parse function determines that the literal body is not well-formed according to the syntax that it implements, it must return \li{ParseError \{msg=}$e_\text{msg}$\li{, loc=}$e_\text{loc}$\li{\}}
+% where $e_\text{msg}$ is a custom error message and $e_\text{loc}$ is a value of type \li{segment}, defined in Figure \ref{fig:indexrange-and-parseresult}, that designates a segment of the literal body as the origin of the error \cite{DBLP:journals/jsc/DeursenKT93}.
+
+% % \vspace{-3px}\paragraph{Success} 
+% If instead parsing succeeds, the parse function returns \li{Success} ~$\ecand$, 
+% where $\ecand$ is called the \emph{encoding of the proto-expansion}.
 For expression TLMs, the proto-expansion is a \emph{proto-expression} and it is encoded as a value of the recursive datatype \li{proto_expr} that is outlined in Figure \ref{fig:candidate-exp-Reason}.
-Most of the constructors of \li{proto_expr} are individually uninteresting -- they encode OCaml's various expression forms. 
-Expressions can mention types, so we also need the type \li{proto_typ} also outlined in Figure \ref{fig:candidate-exp-Reason}.
+% Most of the constructors of \li{proto_expr} are individually uninteresting -- they encode OCaml's various expression forms. 
+% Expressions can mention types, so we also need the type \li{proto_typ} also outlined in Figure \ref{fig:candidate-exp-Reason}.
 It is only the \li{SplicedE} and \li{SplicedT} constructors that are novel.
@@
 
-The fact that the context-free grammar of the base language is fixed ensures that notation providers can reason modularly about syntactic ambiguity in their own grammars.
-% \section{Simple Expression TLMs By Example}\label{sec:tsms-by-example}
-% We begin in this section with a ``tutorial-style'' introduction to seTLMs in Reason.
+% The fact that the context-free grammar of the base language is fixed ensures that notation providers can reason modularly about syntactic ambiguity in their own grammars.
+% % \section{Simple Expression TLMs By Example}\label{sec:tsms-by-example}
+% % We begin in this section with a ``tutorial-style'' introduction to seTLMs in Reason.
 %In particular, we will define an seTLM for constructing values of the recursive labeled sum type \li{rx} that was defined in Figure \ref{fig:datatype-rx}.
-% Sec.
+% % Sec.
 \ref{sec:tsms-minimal-formalism} then formally defines a reduced dialect of Reason called $\miniVerseUE$.
@@
 
-% \subsection{TLM Application}\label{sec:uetsms-usage}
-% The following Reason expression, drawn textually, is of \emph{TLM application} form. Here, a TLM named \li{$rx} is applied to the \emph{generalized literal form} \li{/SURLA|T|G|CEURL/}:
-% \begin{lstlisting}[numbers=none,mathescape=|]
-% $rx /SURLA|T|G|CEURL/
-% \end{lstlisting}
-% Generalized literal forms are left unparsed according to the context-free syntax of Reason.
+% % \subsection{TLM Application}\label{sec:uetsms-usage}
+% % The following Reason expression, drawn textually, is of \emph{TLM application} form. Here, a TLM named \li{$rx} is applied to the \emph{generalized literal form} \li{/SURLA|T|G|CEURL/}:
+% % \begin{lstlisting}[numbers=none,mathescape=|]
+% % $rx /SURLA|T|G|CEURL/
+% % \end{lstlisting}
+% % Generalized literal forms are left unparsed according to the context-free syntax of Reason.
 Several other outer delimiters are also available, as summarized in Figure \ref{fig:literal-forms}.
-The client is free to choose any of these for use with any TLM, as long as the \emph{literal body} (shown in green above) satisfies the requirements stated in Figure \ref{fig:literal-forms}.
+The client is free to choose any of these for use with any TLM, as long as the \emph{literal body} (shown in lavender above) satisfies the requirements stated in Figure \ref{fig:literal-forms}.
 For example, we could have equivalently written the example above as \li{$rx `SURLA|T|G|CEURL`}.
@@
 
-% It is only during the subsequent \emph{typed expansion} phase that the applied TLM parses the {body} of the literal form to generate a \emph{proto-expansion}.
+% % It is only during the subsequent \emph{typed expansion} phase that the applied TLM parses the {body} of the literal form to generate a \emph{proto-expansion}.
 The language then \emph{validates} this proto-expansion according to criteria that we will describe in Sec.
@@
 
-% For example, the expansion of the TLM application above is equivalent to the following expression when the regex value constructors \li{Or} and \li{Str} are in scope:
-% \begin{lstlisting}[numbers=none]
-% Or(Str "SSTRAESTR", Or(Str "SSTRTESTR", Or(Str "SSTRGESTR", Str "SSTRCESTR")))
-% \end{lstlisting}
-% To avoid the assumption that the variables \li{Or} and \li{Str} are in scope at the TLM application site, the expansion actually uses the explicit \li{fold} and \li{inj} operators, as described in Sec.
+% % For example, the expansion of the TLM application above is equivalent to the following expression when the regex value constructors \li{Or} and \li{Str} are in scope:
+% % \begin{lstlisting}[numbers=none]
+% % Or(Str "SSTRAESTR", Or(Str "SSTRTESTR", Or(Str "SSTRGESTR", Str "SSTRCESTR")))
+% % \end{lstlisting}
+% % To avoid the assumption that the variables \li{Or} and \li{Str} are in scope at the TLM application site, the expansion actually uses the explicit \li{fold} and \li{inj} operators, as described in Sec.
 \ref{sec:lists}.
@@
 (We will show how TLM parameters can reduce the awkwardness of this requirement in Chapter \ref{chap:ptsms}.)
-% %The constructors above are those of the type \li{Rx} that was defined in Figure \ref{fig:datatype-rx}.
-
-% % A number of literal forms, ,  are available in Reason's concrete syntax.
+% % %The constructors above are those of the type \li{Rx} that was defined in Figure \ref{fig:datatype-rx}.
+
+% % % A number of literal forms, ,  are available in Reason's concrete syntax.
 Any literal form can be used with any TLM,  TLMs have access only to the literal bodies.
@@
 
-%  %The form does not directly determine the expansion.
-
-% \begin{figure}
-% \begin{lstlisting}
-% 'SURLbody cannot contain an apostropheEURL'
-% `SURLbody cannot contain a backtickEURL`
-% [SURLbody cannot contain unmatched square bracketsEURL]
-% {|SURLbody cannot contain unmatched barred curly bracesEURL|}
-% /SURLbody cannot contain a forward slashEURL/
-% \SURLbody cannot contain a backslashEURL\
-% \end{lstlisting}
-% %SURL<tag>body includes enclosing tags</tag>EURL
-% \caption[Available Generalized Literal Forms]{Generalized literal forms available for use in Reason's textual syntax.
-The characters in green indicate the literal bodies and describe how the literal body is constrained by the form shown on that line.
+% %  %The form does not directly determine the expansion.
+
+% % \begin{figure}
+% % \begin{lstlisting}
+% % 'SURLbody cannot contain an apostropheEURL'
+% % `SURLbody cannot contain a backtickEURL`
+% % [SURLbody cannot contain unmatched square bracketsEURL]
+% % {|SURLbody cannot contain unmatched barred curly bracesEURL|}
+% % /SURLbody cannot contain a forward slashEURL/
+% % \SURLbody cannot contain a backslashEURL\
+% % \end{lstlisting}
+% % %SURL<tag>body includes enclosing tags</tag>EURL
+% % \caption[Available Generalized Literal Forms]{Generalized literal forms available for use in Reason's textual syntax.
+The characters in lavender indicate the literal bodies and describe how the literal body is constrained by the form shown on that line.
 The Wyvern language defines additional forms, including whitespace-delimited forms \cite{TSLs} and multipart forms \cite{sac15}, but for simplicity we leave these out of Reason.}
-% \label{fig:literal-forms}
-% \end{figure}
-
-\subsection{Splicing}\label{sec:splicing-and-hygiene}
-When the parse function determines that some segment of the literal body is a spliced expression, according to whatever syntactic criteria it deems suitable, it can indirectly refer to it in the encoding it produces using the \li{SplicedE} constructor of \li{proto_expr}, which takes a value of type \li{segment} that indicates the zero-indexed location of the spliced expression relative to the start of the provided literal body.
+% % \label{fig:literal-forms}
+% % \end{figure}
+
+% \subsection{Splicing}\label{sec:splicing-and-hygiene}
+% When the parse function determines that some segment of the literal body is a spliced expression, according to whatever syntactic criteria it deems suitable, it can indirectly refer to it in the encoding it produces using the \li{SplicedE} constructor of \li{proto_expr}, which takes a value of type \li{segment} that indicates the zero-indexed location of the spliced expression relative to the start of the provided literal body.
 The \li{SplicedE} constructor also requires a value of type \li{proto_typ}, which indicates the type that the spliced expression is expected to have.
@@
 
-\begin{figure}
-\begin{lstlisting}[numbers=none,xleftmargin=0pt]
-type body = string;
-type segment = {startIdx: int, endIdx: int};
-type parse_result('a) 
-  = ParseError {msg: string, loc: segment}
-  | Success('a);
-type proto_typ = Arrow(proto_typ, proto_typ)
-               | StringTy
-               | /* ... */ 
-               | SplicedT(segment);
-type proto_expr = Tuple(list(proto_expr))
-                | /* ... */
-                | SplicedE(segment, proto_typ);
-\end{lstlisting}
-\caption[Definitions of various types used by TLM definitions.]{Definitions of various types available ambiently to TLM definitions.}
-\label{fig:indexrange-and-parseresult}
-\label{fig:candidate-exp-Reason}
-\end{figure}
-
-For example, consider again the two TLM applications in Figure \ref{fig:first-tsm-example}.
+% \begin{figure}
+% \begin{lstlisting}[numbers=none,xleftmargin=0pt]
+% type body = string;
+% type segment = {startIdx: int, endIdx: int};
+% type parse_result('a) 
+%   = ParseError {msg: string, loc: segment}
+%   | Success('a);
+% type proto_typ = Arrow(proto_typ, proto_typ)
+%                | StringTy
+%                | /* ... */ 
+%                | SplicedT(segment);
+% type proto_expr = Tuple(list(proto_expr))
+%                 | /* ... */
+%                 | SplicedE(segment, proto_typ);
+% \end{lstlisting}
+% \caption[Definitions of various types used by TLM definitions.]{Definitions of various types available ambiently to TLM definitions.}
+% \label{fig:indexrange-and-parseresult}
+% \label{fig:candidate-exp-Reason}
+% \end{figure}
+
+% For example, consider again the two TLM applications in Figure \ref{fig:first-tsm-example}.
 In each case, the parse function of the \li{$html} TLM (Figure \ref{fig:html-tlm-def}, Lines 2-4) first sends the literal body through an off-the-shelf HTML parser, \li{parse_html}.
@@
 For instance,  the proto-expansion generated for first TLM application in \ref{fig:first-tsm-example}, pretty-printed, is:
-\begin{lstlisting}[numbers=none]
-    H1Element(Nil, Cons(TextNode "Hello, ", Cons(
-      TextNode spliced<13; 22; string>, Nil)))
-\end{lstlisting}
-Here, \li{spliced<13; 22; string>} is a reference to the spliced string expression \li{first_name} by its location relative to the start of the literal body being expanded (the off-the-shelf HTML parser provides the necessary baseline location information for use by \li{html_to_ast}).
+% \begin{lstlisting}[numbers=none]
+%     H1Element(Nil, Cons(TextNode "Hello, ", Cons(
+%       TextNode spliced<13..22; string>, Nil)))
+% \end{lstlisting}
+% Here, \li{spliced<13; 22; string>} is a reference to the spliced string expression \li{first_name} by its location relative to the start of the literal body being expanded (the off-the-shelf HTML parser provides the necessary baseline location information for use by \li{html_to_ast}).
 It corresponds to the encoding \li{SplicedE(\{startIdx=13, endIdx=22\}, StringTy)}.
@@
 
-% The parse function can similarly extract \emph{spliced types} from a literal body using the \li{SplicedT} variant of \li{proto_typ}.
+% % The parse function can similarly extract \emph{spliced types} from a literal body using the \li{SplicedT} variant of \li{proto_typ}.
 %In particular, the parse function must provide the index range of spliced subexpressions to the \li{Spliced} constructor of the type \li{MarkedExp}.
@@
 
-%For example, had the  would not be a valid expansion, because the  that are not inside spliced subexpressions:
-%\begin{lstlisting}[numbers=none]
-%Q.Seq(Q.Str(name), Q.Seq(Q.Str ": ", ssn))
-%\end{lstlisting}
-
-% \subsection{Splice Summaries and Segmentations}
-The \emph{segmentation} inferred from a proto-expansion is the finite set of references to spliced terms contained within.
+% %For example, had the  would not be a valid expansion, because the  that are not inside spliced subexpressions:
+% %\begin{lstlisting}[numbers=none]
+% %Q.Seq(Q.Str(name), Q.Seq(Q.Str ": ", ssn))
+% %\end{lstlisting}
+
+% % \subsection{Splice Summaries and Segmentations}
+% The \emph{segmentation} inferred from a proto-expansion is the finite set of references to spliced terms contained within.
 For example, the segmentation inferred from the proto-expression above contains only \li{spliced<13; 22; string>}.
 % Notice that no information about  the spliced terms appear is communicated by the splice summary.
-The system checks that all of the locations in the segmentation are 1) in bounds relative to the literal body; and 2) non-overlapping. 
-This resolves the problem of \textbf{Segmentation} described in Sec.
+% The system checks that all of the locations in the segmentation are 1) in bounds relative to the literal body; and 2) non-overlapping. 
+% This resolves the problem of \textbf{Segmentation} described in Sec.
 \ref{sec:intro}, i.e.
@@
 A program editor or pretty-printer can communicate this segmentation information to the programmer, e.g.
-by coloring non-spliced segments green as is our convention in this document.
+by coloring non-spliced segments lavender as is our convention in this document.
 In general, spliced expressions might themselves apply TLMs, in which case the convention is to use a distinct color for unspliced segments at each depth.
 For example, consider a TLM \li{$smiles} for chemical structures \cite{anderson1987smiles} with support for splicing using curly braces:
-\begin{lstlisting}[numbers=none]
-    $html [|SURLChemical structure of sucrose: {EURL
-        $smiles [|SCSS{ECSSm_glucoseSCSS}-O-{ECSSm_fructoseSCSS}ECSS|] 
-        |> SMILES.to_svg SURL}EURL|]
-\end{lstlisting}
-%For example, if strings were not primitive but rather defined as sequences of characters, we might define a TLM \li{$str} to recover string literal notation and write \li{$html [|SURL<body>\{EURLheading $str [|SSTRWorld!ESTR|]SURL\} ...</body>EURL|]}.
-
-A program editor or pretty-printer can communicate the type of each spliced expression, also specified abstractly by the segmentation, upon request (for Reason, via Merlin \cite{Merlin}.) %When there is a type mismatch, the type annotation on the spliced segment also allows the error message to report the problem without revealing the full expansion.
-
-\subsection{Proto-Expansion Validation}\label{sec:uetsms-validation}
-Three important concerns described in Sec.
+% \begin{lstlisting}[numbers=none]
+%     $html [|SURLChemical structure of sucrose: {EURL
+%         $smiles [|SCSS{ECSSm_glucoseSCSS}-O-{ECSSm_fructoseSCSS}ECSS|] 
+%         |> SMILES.to_svg SURL}EURL|]
+% \end{lstlisting}
+% %For example, if strings were not primitive but rather defined as sequences of characters, we might define a TLM \li{$str} to recover string literal notation and write \li{$html [|SURL<body>\{EURLheading $str [|SSTRWorld!ESTR|]SURL\} ...</body>EURL|]}.
+
+% A program editor or pretty-printer can communicate the type of each spliced expression, also specified abstractly by the segmentation, upon request (for Reason, via Merlin \cite{Merlin}.) %When there is a type mismatch, the type annotation on the spliced segment also allows the error message to report the problem without revealing the full expansion.
+
+% \subsection{Proto-Expansion Validation}\label{sec:uetsms-validation}
+% Three important concerns described in Sec.
 \ref{sec:intro} remain: those related to reasoning abstractly about the \emph{hygiene properties}, i.e.
@@
 Addressing these concerns is the purpose of the \emph{proto-expansion validation} process, which occurs during the typing phase.
-Proto-expansion validation results in the \emph{final expansion}, which is simply the proto-expansion with the references to spliced segments replaced with their own final expansions.
-
-\subsubsection{Capture}\label{sec:capture}
-
-Proto-expansion validation ensures that spliced terms have access \emph{only} to the bindings at the application site---spliced terms cannot capture bindings internal to the proto-expansion. 
-%We will show an alternative formulation of Haskell's syntax for monadic commands that uses Reason's anonymous function syntax to bind variables in Sec. \ref{sec:application-monadic-commands}.
-
-\subsubsection{Context Dependence}\label{sec:context-dependence}
-%The prohibition on shadowing ensures only that variables that appear in spliced terms do not refer to bindings that appear in the surrounding expansion. 
-The proto-expansion validation process also ensures that variables that appear in the proto-expansion do not refer to bindings that appear either at the TLM definition or the application site.
+% Proto-expansion validation results in the \emph{final expansion}, which is simply the proto-expansion with the references to spliced segments replaced with their own final expansions.
+
+% \subsubsection{Capture}\label{sec:capture}
+
+% Proto-expansion validation ensures that spliced terms have access \emph{only} to the bindings at the application site---spliced terms cannot capture bindings internal to the proto-expansion. 
+% %We will show an alternative formulation of Haskell's syntax for monadic commands that uses Reason's anonymous function syntax to bind variables in Sec. \ref{sec:application-monadic-commands}.
+
+% \subsubsection{Context Dependence}\label{sec:context-dependence}
+% %The prohibition on shadowing ensures only that variables that appear in spliced terms do not refer to bindings that appear in the surrounding expansion. 
+% The proto-expansion validation process also ensures that variables that appear in the proto-expansion do not refer to bindings that appear either at the TLM definition or the application site.
 In other words, expansions must be completely \emph{context independent} -- they can make no assumptions about the surrounding context whatsoever.
 
-A minimal example of a ``broken'' TLM that does not generate context-independent proto-expansions is below:
-\begin{lstlisting}[numbers=none]
-    syntax $broken at t by static { 
-      fun(_) => Success (Var "SSTRxESTR") };
-\end{lstlisting}
-The proto-expansion that this TLM generates (for any literal body) refers to a variable \li{x} that it does not itself bind.
+% A minimal example of a ``broken'' TLM that does not generate context-independent proto-expansions is below:
+% \begin{lstlisting}[numbers=none]
+%     syntax $broken at t by static { 
+%       fun(_) => Success (Var "SSTRxESTR") };
+% \end{lstlisting}
+% The proto-expansion that this TLM generates (for any literal body) refers to a variable \li{x} that it does not itself bind.
 If proto-expansion validation permitted such a proto-expansion, it would be well-typed only under those application site typing contexts where \li{x} is bound.
@@
 
-Of course, this prohibition does not extend into the spliced terms in a proto-expansion -- spliced terms appear at the application site, so they can justifiably refer to application site bindings.
-(like \li{first_name} in Fig.
-\ref{fig:first-tsm-example}.) Because proto-expansions refer to spliced terms indirectly, enforcing context independence is straightforward -- we need only that the proto-expansion itself be closed.% In the next section, we will formalize this intuition.
+% Of course, this prohibition does not extend into the spliced terms in a proto-expansion -- spliced terms appear at the application site, so they can justifiably refer to application site bindings.
+(like \li{first_name} in Fig.~\ref{fig:first-tsm-example}.) Because proto-expansions refer to spliced terms indirectly, enforcing context independence is straightforward -- we need only that the proto-expansion itself be closed.% In the next section, we will formalize this intuition.
 % The TLM provider can only refer to them opaquely.
 
-Na\"ively, this restriction, also present in the prior work on TSLs \cite{TSLs}, is quite restrictive -- expansions cannot access any library functions.
+% Na\"ively, this restriction, also present in the prior work on TSLs \cite{TSLs}, is quite restrictive -- expansions cannot access any library functions.
 At best, they can require the client to ``pass in'' required library functions via splicing at every application.
@@
 
-% This prohibition on context dependence explains why the expansion generated by the TLM application in Sec.
+% % This prohibition on context dependence explains why the expansion generated by the TLM application in Sec.
 \ref{sec:uetsms-usage} cannot make use of the regex value constructors, e.g.
@@
 
-% Collectively, we refer to the prohibition on capture and the prohibition on context dependence as \emph{hygiene properties}, by conceptual analogy to corresponding properties in term-rewriting macro systems (see Sec.
+% % Collectively, we refer to the prohibition on capture and the prohibition on context dependence as \emph{hygiene properties}, by conceptual analogy to corresponding properties in term-rewriting macro systems (see Sec.
 \ref{sec:macro-systems}.) The novelty here comes from the fact that spliced terms are being extracted from an initially unparsed sequence of characters.
-% In the examples in Sec.
+% % In the examples in Sec.
 \ref{sec:uetsms-usage} and Sec.
@@
 
-\subsubsection{Typing}\label{sec:typing-e}
-Finally, proto-expansion validation maintains a reasonable {typing discipline} by (1) checking that the expansion is of the type specified by the TLM's type annotation; (2) checking that each spliced type is valid; (3) checking that the type annotation on each spliced expression is valid; and (4) checking each spliced expression against the specified type annotation.
+% \subsubsection{Typing}\label{sec:typing-e}
+% Finally, proto-expansion validation maintains a reasonable {typing discipline} by (1) checking that the expansion is of the type specified by the TLM's type annotation; (2) checking that each spliced type is valid; (3) checking that the type annotation on each spliced expression is valid; and (4) checking each spliced expression against the specified type annotation.
 Context independence implies that ML-style type inference can be performed using only the segmentation (because the remainder of an expansion cannot mention the very variables whose types are being inferred).
 In the prior work on TSLs, spliced terms did not have type annotations
-%The OCaml type system is not strong enough to allow us to express only contextually well-typed syntax trees. In The details are discussed as future work in Sec. \ref{sec:discussion}.
- % This addresses the problem of reasoning abstractly about \textbf{Typing} described in Sec. \ref{sec:intro}, i.e.:
- % \begin{enumerate}
- %   \item determining the type of an expansion requires examining only the type annotation on the TLM definition (much as determining the type of a function application requires examining only the function's type); and
- % \item determining the type that a spliced expression must have requires only the information in the splice summary (rather than complete knowledge of the proto-expansion).
- % \end{enumerate}
-
-% The language \emph{validates} proto-expansions before a final expansion is generated.
+% %The OCaml type system is not strong enough to allow us to express only contextually well-typed syntax trees. In The details are discussed as future work in Sec. \ref{sec:discussion}.
+%  % This addresses the problem of reasoning abstractly about \textbf{Typing} described in Sec. \ref{sec:intro}, i.e.:
+%  % \begin{enumerate}
+%  %   \item determining the type of an expansion requires examining only the type annotation on the TLM definition (much as determining the type of a function application requires examining only the function's type); and
+%  % \item determining the type that a spliced expression must have requires only the information in the splice summary (rather than complete knowledge of the proto-expansion).
+%  % \end{enumerate}
+
+% % The language \emph{validates} proto-expansions before a final expansion is generated.
 One aspect of proto-expansion validation is checking  the proto-expansion against the type annotation specified by the TLM, e.g.
@@
 
-% \subsection{Hygiene}
-% The spliced subexpressions that the proto-expansion refers to (by their position within the literal body, cf.
+% % \subsection{Hygiene}
+% % The spliced subexpressions that the proto-expansion refers to (by their position within the literal body, cf.
 above) must be parsed, typed and expanded during the proto-expansion validation process (otherwise, the language would not be able to check the type of the proto-expansion).
@@
 
-% \paragraph{Context Independent Expansion}
-
-% \paragraph{Expansion Independent Splicing} 
-% %These properties suffice to ensure that programmers and tools can freely rename a variable without changing the meaning of the program.
+% % \paragraph{Context Independent Expansion}
+
+% % \paragraph{Expansion Independent Splicing} 
+% % %These properties suffice to ensure that programmers and tools can freely rename a variable without changing the meaning of the program.
 The only information that is necessary to perform such a \emph{rename refactoring} is the locations of spliced subexpressions within all the literal forms for which the variable being renamed is in scope; the expansions need not otherwise be examined.
@@
 
-% \subsubsection{Final Expansion}
-
-% For example, the final expansion of the body of \li{lookup_rx} is equivalent to the following, under an environment where the regex value constructors are available:
-% \begin{lstlisting}[numbers=none]
-% Seq(Str(name), Seq(Str "SSTR: ESTR", ssn))
-% \end{lstlisting}
-% (Again, due to the prohibition on context dependent expansions, the final expansion actually involves explicit \li{fold} and \li{inj} operators.)
+% % \subsubsection{Final Expansion}
+
+% % For example, the final expansion of the body of \li{lookup_rx} is equivalent to the following, under an environment where the regex value constructors are available:
+% % \begin{lstlisting}[numbers=none]
+% % Seq(Str(name), Seq(Str "SSTR: ESTR", ssn))
+% % \end{lstlisting}
+% % (Again, due to the prohibition on context dependent expansions, the final expansion actually involves explicit \li{fold} and \li{inj} operators.)
 
@@
 
-Let us now briefly consider the topic of TLMs that generate \emph{patterns}, rather than expressions.
-Pattern literals are the dual to expression literals in that expression literals support construction whereas pattern literals support deconstruction \cite{moriconi2008inversion}.
-For example, we can pattern match on a value \li{x : html} by applying a pattern TLM \li{$html} as follows:
-\begin{lstlisting}[numbers=none]
-    switch x {
-    | $html [|SURL<h1>{EURLcsSURL}</h1>EURL|] => /* cs:list(html) */
-    | _ -> None
+\begin{figure}[t]
+\begin{subfigure}[t]{0.55\textwidth}
+\begin{lstlisting}[morekeywords={expression}]
+  module Lambda = {/* a typical encoding */};
+  module LambdaNotation {
+    notation $term at Lambda.term { 
+      lexer SimpleLangLexer;
+      expression parser LambdaParser.term_e;
+      pattern parser LambdaParser.term_p;
+      expansions require 
+        { module Lambda = Lambda } };
+    notation $v at Lambda.v {/* analagous */}
   };
 \end{lstlisting}
-Any list pattern, including one generated by another TLM application, can appear where \li{cs} appears in the example pattern above.
-For longer \li{switch} expressions, the shorthand \li{switch x using $html} applies \li{$html} to every rule where the outermost pattern is of generalized literal form.
-
-Notice that we can use the same name, \li{$html}, for this pattern TLM as for the expression TLM defined in the previous section.
-It does not make sense to apply an expression TLM in pattern position  (many expression-level constructs, e.g.
-lambdas, do not correspond even syntactically to patterns), and \emph{vice versa}, so this is unambiguous.
-
-Pattern TLM definitions differ from expression TLM definitions in two ways: (1) a \emph{sort qualifier}, \li{for patterns}, distinguishes them from expression TLM definitions; (2) the return type of the parse function is \li{parse_result(proto_pat)}, rather than \li{parse_result(proto_expr)}.
+\vspace{-6px}
+\caption{Expression and pattern TLMs for \lismall{Lambda} terms and values}
+\label{fig:lam}
+\vspace{2px}
+\end{subfigure}
+\hfill
+\begin{subfigure}[t]{0.42\textwidth}
+\begin{lstlisting}
+  open LambdaNotation;
+  exception Unbound(Lambda.var);
+  let rec eval = $term.(fun 
+  | `(x)` => raise Unbound(x)
+  | `(SURLlamEURL xSURL.EURLe)` => $v `(SURLlamEURL xSURL.EURLe)`
+  | `(e1SURL(EURLe2SURL)EURL)` => {
+    let $v `(SURLlamEURL xSURL.EURLe)` = eval(e1);
+    let v2 = eval(e2);
+    eval(`(SURL[EURLv2SURL/EURLxSURL]EURLe)`) 
+  });
+\end{lstlisting}
+\vspace{-6px}
+\caption{A reasonably elegant \lismall{Lambda} evaluator}
+\label{fig:lam-eval}
+\end{subfigure}
+\vspace{2px}
+\caption{Expression and pattern literal notation for lambda terms and values}
+\vspace{-8px}
+\label{fig:lam-example}
+\end{figure}
+
+The previous section introduced expression TLMs, which support value construction.
+This section introduces pattern TLMs, which support value deconstruction via the structural pattern matching facilities common to ML-like languages.
+
+For example, the definitions outlined in Fig.
+\ref{fig:lam} allow us to elegantly express an evaluator for the lambda calculus as shown in Fig.
+\ref{fig:lam-eval}.
+We assume that the terms of \li{Lambda} are represented differently from the values to avoid needing additional ``impossible'' cases, but we can use the same notation for both without conflict by defining separate TLMs, \li{$term} and \li{$v}, respectively.
+Each of these TLMs defines both expression and pattern literal notation, distinguished by qualifiers in the TLM definition as shown on Lines 5-6 of Fig.
+\ref{fig:lam}, and we see examples of all four of these notations in use in Fig.
+\ref{fig:lam-eval}.
+These TLMs were designed to be used for operating on \li{Lambda} terms, so we chose a splicing convention inspired by the typical convention ``on paper'', where identifiers appearing in the literal are parsed as spliced (OCaml) variables, as indicated by the colors in Fig.
+\ref{fig:lam-eval}.
+For example, on Line 5, \li{x} and \li{e} are bound by the literal pattern for \li{Lambda} terms on the left, then spliced into the expression literal on the right to construct the corresponding \li{Lambda} value.
+% A different pair of TLMs at the same types could make the opposite choice, parsing identifiers as \li{Lambda} language variables and defining a separate notation for splicing.
+%These would be less elegant for writing the evaluator but more useful when, for example, writing its tests.
+This sort of flexibility is difficult to achieve in settings where the grammar of the host language is itself extended (e.g.
+with the more limited mixfix systems of Coq \cite{Coq:manual} or Agda \cite{norell2007towards}) because syntactic conflicts ensue.
+
+From a client programmer's perspective, reasoning principles analagous to those described in Sec.
+\ref{sec:client-perspective} for expression TLMs are available.
+\textbf{Responsibility} is assigned by the same protocol, and the type annotation on the responsible TLM governs the type of the generated pattern, satisfying the \textbf{Expansion Typing} condition.
+Patterns can contain module paths and type annotations, so pattern TLMs are governed by the same \textbf{Context Independence} condition as expression TLMs.
+The same protocol around \textbf{Segmentation} is also enforced.
+
+The main novelty has to do with \textbf{Segment Typing} and \textbf{Capture}.
+Variables in patterns do not refer to existing bindings, as in expressions, but rather introduce bindings into other expressions, e.g.
+the corresponding branch of a case analysis or the body of a function, so the critical question is this: given only the segmentation of a pattern literal, can we determine exactly which variables the expansion binds, and what the types of these variables are?
+To answer in the affirmative, we need to preclude the possibility of ``invisible bindings'', so the system ensures that pattern literals bind only those variables that appear inside spliced patterns.
+This implies that \emph{proto-patterns simply cannot contain pattern variables}.
+The segmentation assigns a type to each spliced segment, so the question just asked can be answered given only the segmentation.
+
+From the provider's perspective, the parser specified by the \li{pattern parser} clause must generate a proto-pattern, rather than a proto-expression.
+In Reason, this is a value of type \li{Relit.ProtoPat.t}.
+Much as with \li{Relit.ProtoExpr.t} from Fig.
+\ref{fig:relit-util}, this type is defined as a synonym for \li{Parsetree.patteSURLEURLrn} from OCaml's compiler library equipped with an additional function, \li{Relit.ProtoPat.spSURLEURLliced}, that constructs splice references.
+In principle, it might be better to define \li{Relit.ProtoPat.t} to exclude the representation of pattern variables, and we do just that in the theoretical development in Sec.
+\ref{sec:setlms-formally}, but \li{Relit} checks this straightforward condition during proto-expansion validation.
+Note that in OCaml, boolean guards can be associated with rules of a \li{match} (in Reason, \li{switch}) expression, but the guard is not part of the syntax of patterns.
+(If it were, we would need to validate it as in Sec.~\ref{sec:setlms}).
+%In order to ensure that a pattern TLM can be applied anywhere a pattern can appear, e.g.
+in a \li{let} expression as on Line 7 of Fig.
+\ref{fig:lam-eval}, a pattern TLM cannot generate a guard.
+% In a language where boolean guards are included in the syntax of patterns, we would need to enforce context independence, capture avoidance and segmentation-related conditions on guard expressions as well.%  In languages where boolean guards can appear at any depth in a pattern, the same
+
+% \begin{figure}
+% \begin{subfigure}[t]{0.51\textwidth}
+% \begin{lstlisting}
+%   module Lambda = {
+%     type var = int;
+%     type tm = Var(var) | Lam(var, tm) | Ap(tm, tm);
+%     type v  = VLam(var, tm);
+%     let subst : v -> var -> tm -> tm = (* ... *)
+%   };
+% \end{lstlisting}
+% \caption{Lambda calculus terms}
+% \end{subfigure}
+% \hfill
+% \begin{subfigure}[t]{0.49\textwidth}
+% \begin{lstlisting}
+%   module Lam = { /* a typical encoding */ };
+%   module LamNotation {
+%     notation $term at Lam.term { 
+%       lexer SimpleLangLexer;
+%       expression parser LamParser.tm_expr;
+%       pattern parser Lam.tm_pat;
+%       expansions require { module Lam }
+%     }
+%     notation $v at Lam.v { /* analagous */ } 
+%   };
+% \end{lstlisting}
+% \vspace{-4px}
+% \caption{Lambda TLM definitions}
+% \vspace{2px}
+% \end{subfigure}
+% \caption{Lambda calculus example}
+% \end{figure}
+
+% From the client's perspective, pattern TLMs work much like expression TLMs. Let us briefly review the six reasoning principles from Sec. \ref{sec:setlms}.
+
+% \subsubsection{Responsibility} With respect to responsibility, client programmers can reason as in Sec.~\ref{sec:setlms}, following the usual scoping structure of the language.
+Notice that a single TLM definition, in this case \li{HtmlNotation.$html}, can define both expression and pattern literal notation, distinguishing the two parsers by qualifiers as shown on Lines 4-5 of Fig.
+\ref{fig:html-tlm-def}.
+
+% \subsubsection{Expansion Typing} When applying a pattern TLM, the expansion type specified by the TLM definition, again identified relative to the definition site, also classifies the generated pattern.
+
+% \subsubsection{Context Dependence} Patterns can contain module paths and type annotations, so pattern TLMs must be governed by the same context independence condition as expression TLMs.
+
+% \subsubsection{Segmentation} A pattern TLM can splice other patterns out from the provided literal body.
+The same guarantees with respect to segmentation apply.
+So, from the colors in Fig.
+\ref{fig:pattern-tlm-example}, we know where the spliced patterns are.
+Spliced patterns can be of any valid Reason pattern form.
+For example, we see Reason's built in list pattern notation in the outermost splice of Fig.
+\ref{fig:pattern-tlm-example}.
+
+% \subsubsection{Segment Typing}
+
+% \subsubsection{Capture}
+
+% capture --- here's the rub!!!
+
+%  In languages where guards can be part of patterns and refer to the variables that the pattern binds, the capture avoidance considerations from before apply.
+
+% \subsection{Provider Perspective}
+% Pattern TLM definitions differ from expression TLM definitions in two ways: (1) a \emph{sort qualifier}, \li{for patterns}, distinguishes them from expression TLM definitions; (2) the return type of the parse function is \li{parse_result(proto_pat)}, rather than \li{parse_result(proto_expr)}.
 The type \li{proto_pat}, outlined below, classifies encodings of \emph{proto-patterns}.
-\begin{lstlisting}[numbers=none]
-   type proto_pat = /* no variable pattern form! */
-                  | Wild
-                  | /* ... */
-                  | SplicedP(segment, proto_typ);
-\end{lstlisting}
-
-% \subsection{Splicing}
-The constructor \li{SplicedP} operates much like \li{SplicedE} to allow a proto-pattern to refer indirectly to spliced patterns.% by their location within the literal body.
-
-%  For example, the proto-pattern generated for the example at the top of this section would be written concretely as follows:
 % \begin{lstlisting}[numbers=none]
-%   H1Element (_, spliced<6; 7; list(html)>)
+%    type proto_pat = /* no variable pattern form! */
+%                   | Wild
+%                   | /* ... */
+%                   | SplicedP(segment, proto_typ);
 % \end{lstlisting}
 
-% \subsection{Proto-Pattern Validation}
-% Proto-pattern validation serves, like proto-expression validation, to maintain the ability to reason abstractly about binding and typing.
-
-To maintain the abstract binding discipline, variable patterns can appear only within spliced patterns.
+% % \subsection{Splicing}
+% The constructor \li{SplicedP} operates much like \li{SplicedE} to allow a proto-pattern to refer indirectly to spliced patterns.% by their location within the literal body.
+
+% %  For example, the proto-pattern generated for the example at the top of this section would be written concretely as follows:
+% % \begin{lstlisting}[numbers=none]
+% %   H1Element (_, spliced<6; 7; list(html)>)
+% % \end{lstlisting}
+
+% % \subsection{Proto-Pattern Validation}
+% % Proto-pattern validation serves, like proto-expression validation, to maintain the ability to reason abstractly about binding and typing.
+
+% To maintain the abstract binding discipline, variable patterns can appear only within spliced patterns.
 Enforcing this restriction is straightforward: we simply have not defined a variant of the \li{proto_pat} type that encodes variable patterns (wildcards are allowed.)  This restriction ensures that only variables visible to the client in a spliced pattern are bound in the corresponding branch expression.
@@
 
-Type annotations on references to spliced patterns could refer to type variables, so we also need to enforce context independence in the manner discussed in the previous section.
+% Type annotations on references to spliced patterns could refer to type variables, so we also need to enforce context independence in the manner discussed in the previous section.
 %Pattern guards are part of pattern matching rules, not patterns themselves, in Reason/OCaml, but were they part of patterns, context independence would also need to be enforced for guard expressions.% In languages like OCaml, which support arbitrary boolean guard expressions, we would also need to enforce both context independence and capture avoidance as discussed in the previous section for spliced expressions that end up in a guard expression.
@@
 
-% \subsubsection{Typing} 
-To maintain an abstract typing discipline, proto-pattern validation checks type annotations much as in Sec.
+% % \subsubsection{Typing} 
+% To maintain an abstract typing discipline, proto-pattern validation checks type annotations much as in Sec.
 \ref{sec:typing-e}.% (1) that the final expansion is a pattern that matches values of the type specified by the TLM's type annotation; (2) that each spliced pattern matches values of the type indicated in the segmentation; and (3) that each of these types are themselves well-formed types.
@@
 
-The simple TLMs in the previous sections operate only at one specified type, as did the TSLs in the prior work \cite{TSLs}.
-This is rather limiting.
-This section introduces \emph{parametric TLMs}, which can operate over a type- and module-parameterized family of types.
-They also neatly solve the problem discussed in Sec.
-\ref{sec:context-dependence} of giving expansions access to helper functions.
+All of the examples that we have discussed so far operate at a single specified type.
+This section introduces \emph{parametric TLMs}, which can take type and module parameters and operate over a type- and module-parameterized family of types.
+%Parametric TLMs also allow an alternative interpretation of TLM expansion dependencies as partially applied module parameters.
 %Moreover, as discussed at the end of Sec.
@@
 
-Consider the following Reason/OCaml module type (a.k.a. signature), which specifies an abstract data type \cite{liskov1974programming,harper1997programming} of string-keyed polymorphic dictionaries:
-\begin{lstlisting}[numbers=none]
-    module type DICT = {
+\begin{figure}
+\begin{subfigure}[t]{0.4\textwidth}
+\begin{lstlisting}
+  module Map = { 
+    module type S = {
+      type key;
       type t('a);
       let empty : t('a);
-      let extend : t('a) -> (string, 'a) -> t('a); 
-      /* ... */ 
-    };
+      let add : 
+        key -> 'a -> t('a) -> t('a);
+    /* ... */ }
+  }
 \end{lstlisting}
-We can define a TLM that is parametric over implementations of this signature, \li{D : DICT}, and over choices of the codomain type, \li{'a} as follows:
-\begin{lstlisting}[numbers=none]
-    syntax $dict (D : DICT) (type 'a) at D.t('a) 
-    by static { fun(b) => /* ... */ };
+\vspace{-7px}
+\caption{The \li{Map.S} signature}
+\label{fig:map-sig}
+\end{subfigure}
+\hfill
+\begin{subfigure}[t]{0.55\textwidth}
+\begin{subfigure}[t]{\textwidth}
+\begin{lstlisting}
+  notation $map(M : Map.S, type a) at M.t(a) {
+    lexer  MapLexer; parser MapParser; 
+  }
 \end{lstlisting}
-For example, given some module that implements this signature, \li{HashDict : DICT}, we can apply \li{$dict} as follows:
-\begin{lstlisting}[numbers=none]
-    $dict HashDict int [|"key1"SURL=>EURL10SURL; EURL"key2"SURL=>EURL15|]
+\vspace{-6px}
+\caption{The parametric TLM \li{$map}}
+\vspace{5px}
+\label{fig:map-tlm}
+\end{subfigure}
+\begin{subfigure}[b]{\textwidth}
+\begin{lstlisting}
+  notation $stringmap = $map(Misc.StringMap);
+  let make_request id req v = $stringmap
+    `( "session_id" SURL->EURL string_of_int(id)SURL,EURL
+                req SURL->EURL v )`;
 \end{lstlisting}
-Notice that the segmentation immediately reveals which punctuation is particular to this TLM and where the spliced key and value expressions appear.
+\vspace{-4px}
+\caption{Applying \li{$map}}
+\label{fig:map-ap}
+\end{subfigure}
+\end{subfigure}
+\vspace{4px}
+\caption{Literal notation for all finite map implementations using parametric TLMs.}
+\vspace{-5px}
+\label{fig:map-fig}
+\end{figure}
+
+Fig.
+\ref{fig:map-sig} shows a portion of the signature \li{Map.S} from the OCaml standard library, which specifies a polymorphic abstract data type \cite{liskov1974programming,harper1997programming} \li{t('a)} of finite maps with keys of type \li{key}.
+There are many ways to implement this signature.
+For example, the \li{Map.Make} functor in the standard library implements this signature using balanced binary trees given a module that specifies a \li{key} type equipped with comparison functions.
+The \li{Misc.StringMap} module in the standard library also provides a specialized implementation of this signature with type \li{key = string}.
+
+We can define a TLM, \li{$map}, that is parametric over implementations of this signature, \li{M : Map.S}, and over choices of the co-domain type, \li{a}, as shown in Fig.
+\ref{fig:map-tlm}.
+We can then partially apply \li{$map} to \li{Misc.StringMap} as shown on Line 1 of Fig.~\ref{fig:map-ap}, producing a TLM, \li{$stringmap},  parameterized only over the type \li{a}, with expansion type \li{Misc.StringMap.t(a)}.
+We can apply \li{$stringmap} to a type specialize it further, e.g.
+\li{$stringmap string} has expansion type \li{Misc.StringMap.t(string)}.
+For TLMs where all remaining parameters are types mentioned in the expansion type, we can also immediately apply the TLM to a generalized literal form as shown on Lines~2-3 of Fig.~\ref{fig:map-ap}.
+The type parameters are determined by unification using the types inferred for the spliced expressions together with the segment types specified in the segmentation, and any surrounding type constraints.
+In this example, we can infer \li{a} to be \li{string} because \li{string_of_int(id) : string} and the corresponding segment type is \li{a}.
+TLM abbreviations can themselves be parameterized to support partial application of parameters other than the last.
+
+% For example, given some module that implements this signature, \li{HashDict : DICT}, we can apply \li{$dict} as follows:
+% \begin{lstlisting}[numbers=none]
+%     $dict HashDict int [|"key1"SURL=>EURL10SURL; EURL"key2"SURL=>EURL15|]
+% \end{lstlisting}
+% Notice that the segmentation immediately reveals which punctuation is particular to this TLM and where the spliced key and value expressions appear.
 Because the context-free syntax of unexpanded terms is never modified, it is possible to reason modularly about syntactic determinism, i.e.
@@
 
-The proto-expansion generated for the TLM application above, shown below, can refer to the parameters:
+The proto-expansion generated by a parametric TLM can refer to its parameters.
+Validation checks that the proto-expansion is truly parametric---in this case, the proto-expansion must be valid for all modules \li{M : Map.S} and types \li{a}.
+It is only after validation that we substitute the actual parameters, here \li{Misc.StringMap} for \li{M} and \li{string} for \li{a}, into the final expansion.
+
+%Only the type annotations on references to spliced terms are subject to early parameter substitution (because they classify application site terms.)
+
+% If we will use the \li{HashDict} implementation ubiquitously, we can abbreviate the partial application of \li{$dict} to \li{HashDict}, resulting in a TLM that is parametric over only the type \li{'a}:
+% \begin{lstlisting}[numbers=none]
+%     syntax $dict' = $dict HashDict; 
+% \end{lstlisting}
+
+The \li{expansions require} clause shown in previous examples can be understood in terms of parameterization over the dependencies, using the given module variables, followed immediately by partial application of the corresponding module paths.
+Note, however, that module parameters need an explicit signature, like functor (module function) arguments in ML.
+
+An alternative point-of-view in ML is to treat the \li{expansions require} clause as fundamental and achieve parameterization using module functions (functors), e.g.
+we could express \li{$map} as follows:
 \begin{lstlisting}[numbers=none]
-    D.extend(D.extend D.empty (spliced<1;6;string>, 
-      spliced<11;12;'a>)) (spliced<15;20;string>, 
-      spliced<25;26;'a>)
+  module MapNotation = (M : Map.S, A : { type a; }) => {
+    notation $map at M.t(A.a) { 
+      lexer MapLexer; parser MapParser; 
+      expansions require { module M = M; type a = A.a; } } }
 \end{lstlisting}
-Validation checks that the proto-expansion is truly parametric, i.e.
-it must be valid for all modules \li{D : DICT} and types \li{'a}.
-It is only after validation that we substitute the actual parameters, here \li{HashDict} for \li{D} and \li{int} for \li{'a}, into the final expansion.
-Only the type annotations on references to spliced terms are subject to early parameter substitution (because they classify application site terms.)
-
-If we will use the \li{HashDict} implementation ubiquitously, we can abbreviate the partial application of \li{$dict} to \li{HashDict}, resulting in a TLM that is parametric over only the type \li{'a}:
-\begin{lstlisting}[numbers=none]
-    syntax $dict' = $dict HashDict; 
-\end{lstlisting}
-TLM abbreviations can themselves be parameterized to support partial application of parameters other than the last.
-
-In Sec.
+% The conveniences related to type parameter inference are not available when parameterization is encoded in this way.
+
+% In Sec.
 \ref{sec:context-dependence} we discussed the problem of the strict context independence discipline being too restrictive, in that it would seem to restrict expansions from referring to useful helpers bound at the TLM definition site.
@@
 Because this will be common in practice, we provide the following shorthand:
-\begin{lstlisting}[numbers=none]
-    syntax $a at t using X~${}_1$~=M~${}_1$~,...,X~${}_n$~=M~${}_n$~ by ...
-\end{lstlisting} 
-% This is the same as a definition that specifies $n$ type, expression or module parameters and immediately applies the given arguments, binding the result to the same TSM name.
-
-This explicit parameter passing discipline is reminiscent of work on explicit capture for distributed functions \cite{DBLP:conf/ecoop/MillerHO14}.
+% \begin{lstlisting}[numbers=none]
+%     syntax $a at t using X~${}_1$~=M~${}_1$~,...,X~${}_n$~=M~${}_n$~ by ...
+% \end{lstlisting} 
+% % This is the same as a definition that specifies $n$ type, expression or module parameters and immediately applies the given arguments, binding the result to the same TSM name.
+
+% This explicit parameter passing discipline is reminiscent of work on explicit capture for distributed functions \cite{DBLP:conf/ecoop/MillerHO14}.
 By not implicitly giving expansions access to all definition-site bindings, we need not examine parse functions to reason about, e.g., renaming.
@@
 
-\newcommand{\moreExamplesSec}{Additional Examples}
-\section{\protect\moreExamplesSec}
-\label{sec:more-examples}
-
-\subsection{HTML}
-We start in this section by describing the TLM mechanism in detail by way of a substantial case study: literal notation, like that built in to Ur/Web (see Figure \ref{fig:urweb}) and also, presently, Reason, for expressions of the type  \li{html} in Figure \ref{fig:html-type-def}.
-
-\begin{figure*}[t]
-
-\begin{subfigure}[t]{0.48\textwidth}
-\begin{lstlisting}
-  type html_attrs = list(string, string)
-  type html = BodyElement(html_attrs, list(html))
-            | H1Element(html_attrs, list(html))
-            | TextNode(string) 
-            | /* ... */;
-\end{lstlisting}
-\caption{The \li{html} type, which classifies encodings of HTML data.}
-\label{fig:html-type-def}
-\end{subfigure}
-\hfill
-\begin{subfigure}[t]{0.48\textwidth}
-\begin{lstlisting}[mathescape=|]
-  syntax $html at html by static {
-    fun(b : body) : parse_result(proto_expr) => 
-      try (parse_html b |> html_to_ast |> Success) {
-      | InvalidHTML msg loc => ParseError msg loc }
-  };
-\end{lstlisting}
-\caption{The \li{$html} TLM definition. Figure \ref{fig:candidate-exp-Reason} defines   TLM-related types.}
-\label{fig:html-tlm-def}
-\end{subfigure}
-
-\begin{subfigure}[t]{\textwidth}
-\begin{lstlisting}
-  let heading first_name = $html [|SURL<h1>Hello, {[EURLfirst_nameSURL]}!</h1>EURL|]
-  let body = $html [|SURL<body>{EURLheading "World!"SURL} ...</body>EURL|]
-\end{lstlisting}
-\caption{Two examples of the \li{$html} TLM being applied.
-Compare to Ur/Web's built in HTML literal notation from Figure \ref{fig:urweb}.
-%In each case, literal body, between backticks, is initially left unparsed according to the language's context-free syntax.
-%The applied TLM determines a segmentation and expansion during the typed expansion phase, which generalizes the usual typing phase.
-}
-\label{fig:first-tsm-example}
-\end{subfigure}
-\caption{Case Study: HTML literals}
-\end{figure*}
-
-\subsection{Quotation}
-
-\subsection{More Examples, Briefly}
-%
-
-\begin{figure}[t!]
-\begin{minipage}{\textwidth}
-\small
-$\arraycolsep=2pt\begin{array}{llcl}
-\mathsf{UTyp} & \utau & ::= & 
-\ut ~\vert~ 
-\parr{\utau}{\utau} ~\vert~
-\forallt{\ut}{\utau} ~\vert~
-\rect{\ut}{\utau}\\
-& & \vert & 
-\prodt{\mapschema{\utau}{i}{\labelset}} ~\vert~
-\sumt{\mapschema{\utau}{i}{\labelset}}\\
-\mathsf{UExp} & \ue & ::= & 
-\ux ~\vert~
-% \asc{\ue}{\utau} ~\vert~
-% \letsyn{\ux}{\ue}{\ue} & \text{value binding}\\
-\lam{\ux}{\utau}{\ue} ~\vert~
-\ap{\ue}{\ue} ~\vert~
-\Lam{\ut}{\ue} ~\vert~
-\App{\ue}{\utau} ~\vert~
-\fold{\ue} \\
-% \unfold{\ue} ~\vert~
-&& \vert & \tpl{\mapschema{\ue}{i}{\labelset}} ~\vert~
-\inj{\ell}{\ue} ~\vert~
-\matchwith{\ue}{\seqschemaX{\urv}}
- \\
-& & \vert & \uesyntaxq{\tsmv}{\utau}{e}{\ue} \\   
-& & \vert & \usyntaxup{\tsmv}{\utau}{e}{\ue} \\
-&&\vert&  \utsmap{\tsmv}{b} \\
-\mathsf{URule} & \urv & ::= & 
-%& \aumatchrule{\upv}{\ue} 
-\matchrule{\upv}{\ue} \\
-\mathsf{UPat} & \upv & ::= & 
-%& \ux 
-\ux ~\vert~
-\wildp ~\vert~ 
-\foldp{\upv} ~\vert~
-\tplp{\mapschema{\upv}{i}{\labelset}} ~\vert~
-\injp{\ell}{\upv} ~\vert~
-\utsmap{\tsmv}{b}
-
-% \LCC  &  & 
-% %& \lightgray 
-% & \color{Yellow} & \color{Yellow} \\
-% &&
-% %& \audefuetsm{\utau}{e}{\tsmv}{\ue} 
-% & \uesyntax{\tsmv}{\utau}{e}{\ue} & \text{seTLM definition}\\ 
-% &&
-% %& \autsmap{b}{\tsmv} 
-% & \utsmap{\tsmv}{b} & \text{seTLM application}\ECC
-\end{array}$
-\end{minipage}
-\caption[Syntax of the $\miniVersePat$ unexpanded language (UL)]{Syntax of the $\miniVersePat$ unexpanded language (UL).
-Metavariable $\ut$ ranges over type identifiers, $\ux$ over expression identifiers, $\ell$ over labels, $\labelset$ over finite sets of labels, $\tsmv$ over TLM names and $b$ over literal bodies.
-We write $\mapschema{\utau}{i}{\labelset}$ for a finite mapping of each label $i$ in $\labelset$ to some unexpanded type $\utau_i$, and similarly for other sorts.
-We write $\seqschemaX{\urv}$ for a finite sequence of $n$ unexpanded rules.
-}
-\label{fig:U-unexpanded-terms}
-\end{figure}
+% \newcommand{\moreExamplesSec}{Additional Examples}
+% \section{\protect\moreExamplesSec}
+% \label{sec:more-examples}
+
 \begin{comment}
@@
 \section{\protect\seTLMsFormallySec}
-\label{sec:setsms-formally}
-
-This section will present a calculus of simple expression and pattern TLMs called $\miniVersePat$.
-By the end of this section, we will have a theorem that encodes the six reasoning principles that were outlined informally in the previous sections.
-
-Because our focus is on these reasoning principles, and for reasons of space, we leave our full calculus of parametric TLMs to the supplement.
+\label{sec:setlms-formally}
+
+This section will present a calculus of simple expression and pattern TLMs based on ``core ML'' called $\miniVersePat$.
+By the end of this section, we will have a theorem that formalizes the six reasoning principles that were outlined informally in the previous sections.
+$\miniVersePat$ consists of an \emph{unexpanded language}, or \emph{UL}, defined by typed expansion to the core language, named for the purposes of this paper $\miniVerseBase$.
+
+% Because our focus is on these reasoning principles, and for reasons of space, we leave our full calculus of parametric TLMs to the supplement.
 The full calculus extends the simple calculus of this section with an ML module calculus based closely on the system defined by \citet{pfple1}, which in turn is based on early work by \citet{MacQueen:1984:MSM:800055.802036,DBLP:conf/popl/MacQueen86}, subsequent work on the phase splitting interpretation of modules \cite{harper1989higher} and on using dependent singleton kinds to track type identity \cite{stone2006extensional,DBLP:conf/lfmtp/Crary09}, and finally on formal developments by \citet{dreyer2005understanding} and \citet{conf/popl/LeeCH07}.
@@
 
-Both the simple and full calculus consist of an \emph{unexpanded language}, or \emph{UL}, defined by typed expansion to an \emph{expanded language}, or \emph{XL}.
-Figs.
-\ref{fig:U-unexpanded-terms}  and \ref{fig:U-expanded-terms} summarize the syntax of the UL and  the XL, respectively.
 % Programs are written as unexpanded expressions but evaluate as well-typed expanded expressions.
@@
 
-\subsection{Expanded Language (XL)}\label{sec:s-XL}
-The {XL} of $\miniVersePat$ forms a standard pure functional language with partial function types, quantification over types, recursive types, labeled product types and labeled sum types and support for pattern matching.
- The reader is directed to \emph{PFPL} \cite{pfple1} for a detailed introductory account of these constructs.
-We will only tersely summarize the statics and dynamics of the XL below because the particularities are not critical.
-
-% \subsubsection{Statics of the Expanded Language}
-The \emph{statics of the XL} is organized around the type formation judgement, $\istypeU{\Delta}{\tau}$, the expression typing judgement, $\hastypeU{\Delta}{\Gamma}{e}{\tau}$, and the pattern typing judgement, $\patType{\pctx}{p}{\tau}$.
-In the latter, $\pctx$ is a typing context that tracks the typing hypotheses generated by $p$.
+\subsection{Core Language}\label{sec:s-XL}
+Fig.
+\ref{fig:U-expanded-terms} gives the syntax of the core language, $\miniVerseBase$.
+This language forms a standard pure functional language with partial function types, quantification over types, recursive types, labeled product types, labeled sum types and support for pattern matching.\footnote{The supplement shows how to remove pattern matching  (and pattern TLMs) from the calculus by adding the usual elimination forms, e.g.
+$\unfold{e}$ rather than the $\aefoldp{p}$ pattern for values of recursive type.} Formally, core language terms are {abstract binding trees} (ABTs) identified up to alpha-equivalence, so we follow the syntactic conventions of \citet{pfple1}.
+This notational convention also clearly distinguishes core terms from unexpanded terms, which formally behave quite differently as we will describe below.
+
+ The reader is directed to \emph{PFPL} \cite{pfple1} for a detailed introductory account of all of these language constructs.
+We will only tersely summarize the static and dynamic semantics of the core language because the particularities are not critical to our ideas.
+The static semantics is organized around the type formation judgement, $\istypeU{\Delta}{\tau}$, the expression typing judgement, $\hastypeU{\Delta}{\Gamma}{e}{\tau}$, and the pattern typing judgement, $\patType{\pctx}{p}{\tau}$.
+Type formation contexts, $\Delta$, track hypothesis of the form $\Dhyp{t}$, and typing contexts, $\Gamma$, track hypotheses of the form $x : \tau$.
+In the pattern typing judgement, $\pctx$ collects the typing hypotheses generated by $p$.
 These judgements are inductively defined in the supplemental material along with necessary auxiliary structures and  standard lemmas.
-%\emph{Type formation contexts}, $\Delta$, and \emph{typing contexts}, $\Gamma$ are defined in the supplement in the standard way \cite{pfpl},
+The dynamic semantics of $\miniVerseBase$ is organized around the judgements $\isvalU{e}$, which says that $e$ is a value, and $\evalU{e}{e'}$, which says that $e$ evaluates to the value $e'$.
+As in ML, evaluation can diverge (general recursion is possible via recursive types \cite{pfple1}).
+It can also result in match failure.%\emph{Type formation contexts}, $\Delta$, and \emph{typing contexts}, $\Gamma$ are defined in the supplement in the standard way \cite{pfpl},
 % \[\begin{array}{ll}
@@
 
+\begin{figure}
+%\hspace{-5px}
+\begin{minipage}{\textwidth}
+\small
+$\arraycolsep=2pt\begin{array}{llcl}
+\mathsf{Typ} & \tau & ::= & t ~\vert~ \aparr{\tau}{\tau} ~\vert~ \aall{t}{\tau} ~\vert~ \arec{t}{\tau} ~\vert~ \aprod{\labelset}{\mapschema{\tau}{i}{\labelset}} ~\vert~ \asum{\labelset}{\mapschema{\tau}{i}{\labelset}}\\
+\mathsf{Exp} & e & ::= & x ~\vert~ \aelam{\tau}{x}{e} ~\vert~ \aeap{e}{e} ~\vert~ \aetlam{t}{e} ~\vert~ \aetap{e}{\tau} ~\vert~ \aefold{e} ~\vert~ \aetpl{\labelset}{\mapschema{e}{i}{\labelset}} ~\vert~  \aein{\ell}{e} \\
+& & \vert & \aematchwith{n}{e}{\seqschemaX{r}}\\
+\mathsf{Rule} & r & ::= & \aematchrule{p}{e}\\
+\mathsf{Pat} & p & ::= & x  ~\vert~ \aewildp ~\vert~ \aefoldp{p} ~\vert~ \aetplp{\labelset}{\mapschema{p}{i}{\labelset}} ~\vert~ \aeinjp{\ell}{p}
+\end{array}$
+\end{minipage}
+\caption[Syntax of the core language of $\miniVersePat$]{Syntax of the core language, $\miniVerseBase$, which is an entirely standard typed lambda calculus.
+Metavariable $x$ ranges over variables, $t$ over type variables, $\ell$ over labels and $L$ over finite sets of labels.
+We write $\mapschema{\tau}{i}{\labelset}$ for a finite mapping of each label $i$ in $\labelset$ to some type $\tau_i$, and similarly for other sorts of terms.
+We write $\seqschemaX{r}$ for a finite sequence of $n > 0$  rules.
+%When using stylized forms, the label set is omitted when it can be inferred, e.g.
+the labeled product type $\prodt{\finmap{\mapitem{\ell_1}{e_1}, \mapitem{\ell_2}{e_2}}}$ leaves the label set $\{\ell_1, \ell_2\}$ implicit.
+% When we use the stylized forms, we assume that the reader can infer suppressed indices and arguments from the surrounding context.
+}
+\label{fig:U-expanded-terms}
+\end{figure}
+
+\begin{figure}[t!]
+\begin{minipage}{\textwidth}
+\small
+$\arraycolsep=2pt\begin{array}{llcl}
+\mathsf{UTyp} & \utau & ::= & 
+\ut ~\vert~ 
+\parr{\utau}{\utau} ~\vert~
+\forallt{\ut}{\utau} ~\vert~
+\rect{\ut}{\utau} ~\vert~
+\prodt{\mapschema{\utau}{i}{\labelset}} ~\vert~
+\sumt{\mapschema{\utau}{i}{\labelset}}\\
+\mathsf{UExp} & \ue & ::= & 
+\ux ~\vert~
+% \asc{\ue}{\utau} ~\vert~
+% \letsyn{\ux}{\ue}{\ue} & \text{value binding}\\
+\lam{\ux}{\utau}{\ue} ~\vert~
+\ap{\ue}{\ue} ~\vert~
+\Lam{\ut}{\ue} ~\vert~
+\App{\ue}{\utau} ~\vert~
+\fold{\ue} ~\vert~
+\tpl{\mapschema{\ue}{i}{\labelset}} ~\vert~
+\inj{\ell}{\ue} \\
+& & \vert & 
+\matchwith{\ue}{\seqschemaX{\urv}}
+ \\
+& & \vert & \uesyntaxq{\tsmv}{\utau}{e}{\ue} \\   
+& & \vert & \usyntaxup{\tsmv}{\utau}{e}{\ue} \\
+&&\vert&  \utsmap{\tsmv}{b} \\
+\mathsf{URule} & \urv & ::= & 
+%& \aumatchrule{\upv}{\ue} 
+\matchrule{\upv}{\ue} \\
+\mathsf{UPat} & \upv & ::= & 
+%& \ux 
+\ux ~\vert~
+\wildp ~\vert~ 
+\foldp{\upv} ~\vert~
+\tplp{\mapschema{\upv}{i}{\labelset}} ~\vert~
+\injp{\ell}{\upv} ~\vert~
+\utsmap{\tsmv}{b}
+
+% \LCC  &  & 
+% %& \lightgray 
+% & \color{Yellow} & \color{Yellow} \\
+% &&
+% %& \audefuetsm{\utau}{e}{\tsmv}{\ue} 
+% & \uesyntax{\tsmv}{\utau}{e}{\ue} & \text{seTLM definition}\\ 
+% &&
+% %& \autsmap{b}{\tsmv} 
+% & \utsmap{\tsmv}{b} & \text{seTLM application}\ECC
+\end{array}$
+\end{minipage}
+\caption[Syntax of the $\miniVersePat$ unexpanded language (UL)]{Syntax of the $\miniVersePat$ unexpanded language (UL).
+Metavariable $\ut$ ranges over type identifiers, $\ux$ over expression identifiers, $\tsmv$ over TLM identifiers and $b$ over literal bodies.
+}
+\label{fig:U-unexpanded-terms}
+\end{figure}
 % \subsubsection{Evaluation Semantics}\label{sec:dynamics-U}
-The \emph{evaluation semantics} of $\miniVersePat$ is organized around the judgements $\isvalU{e}$, which says that $e$ is a value, and $\evalU{e}{e'}$, which says that $e$ evaluates to the value $e'$.% Additional  judgements, not shown, are needed to define the dynamics of pattern matching, but they do not appear directly in our subsequent developments, so we omit them.
+%The \emph{evaluation semantics} of $\miniVersePat$ is organized around the judgements $\isvalU{e}$, which says that $e$ is a value, and $\evalU{e}{e'}$, which says that $e$ evaluates to the value $e'$.% Additional  judgements, not shown, are needed to define the dynamics of pattern matching, but they do not appear directly in our subsequent developments, so we omit them.
 % We assume an eager dynamics.
@@
 \subsection{Syntax of the Unexpanded Language}\label{sec:syntax-U}\label{sec:s-UL}
-Unexpanded types and expressions are simple inductive structures.
-Unlike expanded types and expressions, they are \textbf{not} abstract binding trees -- we do \textbf{not} define the standard notions of renaming, alpha-equivalence or substitution for unexpanded terms.
-This is because unexpanded expressions remain ``partially parsed'' due to the presence of literal bodies, $b$, from which spliced terms might be extracted during typed expansion.
-In fact, unexpanded types and expressions do not involve variables at all, but rather \emph{type identifiers}, $\ut$, and \emph{expression identifiers}, $\ux$.
-Identifiers are given meaning by expansion to variables during typed expansion, as we will see.
-This distinction between identifiers and variables is technically crucial to our developments.
+Fig.
+\ref{fig:U-unexpanded-terms} defines the syntax of the unexpanded language.
+Unlike core language types and expressions, unexpanded expressiona and types are \textbf{not} abstract binding trees -- we do \textbf{not} assume the standard notions of renaming, alpha-equivalence or substitution.
+Instead, they are simple inductive  structures.
+This is because unexpanded expressions remain ``partially parsed'' due to the presence of literal bodies, $b$, from which spliced terms might be extracted during expansion.
+In fact, unexpanded types and expressions do not involve variables at all, but rather \emph{identifiers}, $\ut$ and $\ux$.
+% This distinction between identifiers and variables is technically crucial to our developments.
 %We \textbf{cannot} adopt the usual definitions of $\alpha$-renaming of identifiers, because unexpanded types and expressions are still in a ``partially parsed'' state -- the literal bodies, $b$, within an unexpanded expression might contain spliced subterms that are ``surfaced'' by a TLM only during typed expansion, as we will detail below.
@@
 %\] and so on for the remaining expanded expression forms.
-Most of the unexpanded forms in Figure \ref{fig:U-unexpanded-terms}  mirror the expanded forms.
-We refer to these as the \emph{common forms}.
-% The mapping from expanded forms to common unexpanded forms is defined explicitly in the supplement.
 
 There is also a corresponding context-free textual syntax for the UL. 
-Giving a complete definition of the context-free textual syntax as, e.g., a context-free grammar, is not critical to our purposes here. 
+Giving a complete definition of the context-free textual syntax with, for example, a context-free grammar is not critical to our purpose. 
 %Our paper on Wyvern defines a textual syntax for a similar system \cite{TSLs}. 
-Instead, we only posit partial metafunctions $\parseUTypF{b}$, $\parseUExpF{b}$ and $\parseUPatF{b}$  that go from character sequences, $b$, to unexpanded types, expressions and patterns (the supplement states the full condition.)
+Instead, we only posit partial metafunctions $\parseUTypF{b}$, $\parseUExpF{b}$ and $\parseUPatF{b}$  that go from character sequences, $b$, to unexpanded types, expressions and patterns, respectively.
 % \begingroup
@@
 
-\begin{figure}
-%\hspace{-5px}
-\begin{minipage}{\textwidth}
-\small
-$\arraycolsep=2pt\begin{array}{llcl}
-\mathsf{Typ} & \tau & ::= & t ~\vert~ \aparr{\tau}{\tau} ~\vert~ \aall{t}{\tau} ~\vert~ \arec{t}{\tau} \\
-& & \vert &  \aprod{\labelset}{\mapschema{\tau}{i}{\labelset}} ~\vert~ \asum{\labelset}{\mapschema{\tau}{i}{\labelset}}\\
-\mathsf{Exp} & e & ::= & x ~\vert~ \aelam{\tau}{x}{e} ~\vert~ \aeap{e}{e} ~\vert~ \aetlam{t}{e} ~\vert~ \aetap{e}{\tau} \\
-& & \vert & \aefold{e} ~\vert~ \aetpl{\labelset}{\mapschema{e}{i}{\labelset}} ~\vert~  \aein{\ell}{e} \\
-& & \vert & \aematchwith{n}{e}{\seqschemaX{r}}\\
-\mathsf{Rule} & r & ::= & \aematchrule{p}{e}\\
-\mathsf{Pat} & p & ::= & x  ~\vert~ \aewildp ~\vert~ \aefoldp{p} ~\vert~ \aetplp{\labelset}{\mapschema{p}{i}{\labelset}}\\
-& & \vert & \aeinjp{\ell}{p}
-\end{array}$
-\end{minipage}
-\caption[Syntax of the XL of $\miniVersePat$]{Syntax of the expanded language (XL).
-XL terms are \emph{abstract binding trees} (ABTs) identified up to alpha-equivalence, so we follow the syntactic conventions of \citet{pfple1}.
-Metavariable $x$ and $t$ ranges over variables.
-%When using stylized forms, the label set is omitted when it can be inferred, e.g.
-the labeled product type $\prodt{\finmap{\mapitem{\ell_1}{e_1}, \mapitem{\ell_2}{e_2}}}$ leaves the label set $\{\ell_1, \ell_2\}$ implicit.
-% When we use the stylized forms, we assume that the reader can infer suppressed indices and arguments from the surrounding context.
-}
-\label{fig:U-expanded-terms}
-\end{figure}
+\vspace{-5px}
+\subsection{Typed Expansion}\label{sec:typed-expansion-U}\label{sec:s-TE}
+The \emph{typed expansion judgements} below specify the expansion process, which in the setting of $\miniVersePat$ occurs simultaneously with typing (see Sec.
+\ref{sec:implementation} for certain nuances in our implementation).
+\[\begin{array}{ll}
+% \textbf{Judgement Form} & \textbf{Description}\\
+\expandsTU{\uDelta}{\utau}{\tau} & \text{$\utau$ has well-formed expansion $\tau$}\\
+\expandsUPX{\ue}{e}{\tau} & \text{$\ue$ has expansion $e$ of type $\tau$}\\
+% \ruleExpands{\uDelta}{\uGamma}{\uPsi}{\uPhi}{\urv}{r}{\tau}{\tau'} & \text{$\urv$ has expansion $r$ taking values of type $\tau$ to values of type $\tau'$}\\
+\patExpands{\upctx}{\uPhi}{\upv}{p}{\tau} & \text{$\upv$ has expansion $p$ matching $\tau$}
+\end{array}\]
+%\newcommand{\gray}[1]{{\color{gray} #1}}
+
+% These judgements are inductively defined in the supplement. 
+% \begingroup 
+% \def\thetheorem{\ref{thm:typed-expansion-short-U}}
+
+%These rules validate the following theorem, which establishes that typed expansion produces an expansion of the assigned type. 
+%\begin{theorem}[Typed Expression Expansion] If $\expandsU{\uDD{\uD}{\Delta}}{\uGG{\uG}{\Gamma}}{\uPsi}{\ue}{e}{\tau}$ and $\uetsmenv{\Delta}{\uPsi}$ then $\hastypeU{\Delta}{\Gamma}{e}{\tau}$.\end{theorem}
+%\begin{proof} This is the first part of Theorem \ref{thm:typed-expansion-U}, defined and proven below.\end{proof}
+
+Most of the unexpanded forms in Figure \ref{fig:U-unexpanded-terms}  mirror the expanded forms.
+We refer to these as the \emph{common forms}.
+% The mapping from expanded forms to common unexpanded forms is defined explicitly in the supplement.
+The typed expansion rules that handle common forms mirror the corresponding typing rules.
+The \emph{expression TLM context}, $\uPsi$, and the \emph{pattern TLM context}, $\uPhi$, detailed below pass through these rules opaquely.
+For example, the rules for variables and lambdas are shown being applied in the example derivation in Fig.~\ref{fig:expansion-exmpl}, discussed below.
+The full set of rules is in the supplement.
+%Each of these rules is based on the corresponding typing rule, i.e.
+Rules (\ref{rule:hastypeU-var}) through (\ref{rule:hastypeU-case}), respectively.
+For example, the following typed expansion rules are based on the typing rules (\ref{rule:hastypeU-var}), (\ref{rule:hastypeU-lam}) and (\ref{rule:hastypeU-ap}), respectively:% for unexpanded expressions of variable, function and application form, respectively:
+% {\small\begin{mathpar}
+%   \inferrule[ee-id]{ }{\expandsUP{\uDelta}{\uGamma, \uGhyp{\ux}{x}{\tau}}{\uPsi}{\uPhi}{\ux}{x}{\tau}}
+
+%   \inferrule[ee-lam]{
+%     \expandsTU{\uDelta}{\utau}{\tau}\\
+%     \expandsUP{\uDelta}{\uGamma, \uGhyp{\ux}{x}{\tau}}{\uPsi}{\uPhi}{\ue}{e}{\tau'}
+%   }{\expandsUPX{\lam{\ux}{\utau}{\ue}}{\aelam{\tau}{x}{e}}{\aparr{\tau}{\tau'}}}
+% \end{mathpar}}
 
 \begin{figure*}
+\vspace{-4px}
 {\small\begin{mathpar}
@@
 \end{mathpar}}
-\caption{An example expansion derviation demonstrating how identifiers and variables are separately tracked.}
+\vspace{-6px}
+\caption{An example expansion derivation demonstrating how identifiers and variables are separately tracked.}
 \label{fig:expansion-exmpl}
+\vspace{-8px}
 \end{figure*}
-
-\subsection{Typed Expansion}\label{sec:typed-expansion-U}\label{sec:s-TE}
-Unexpanded terms are checked and expanded simultaneously according to the central \emph{typed expansion judgements}:
-\[\begin{array}{ll}
-% \textbf{Judgement Form} & \textbf{Description}\\
-\expandsTU{\uDelta}{\utau}{\tau} & \text{$\utau$ has well-formed expansion $\tau$}\\
-\expandsUPX{\ue}{e}{\tau} & \text{$\ue$ has expansion $e$ of type $\tau$}\\
-% \ruleExpands{\uDelta}{\uGamma}{\uPsi}{\uPhi}{\urv}{r}{\tau}{\tau'} & \text{$\urv$ has expansion $r$ taking values of type $\tau$ to values of type $\tau'$}\\
-\patExpands{\upctx}{\uPhi}{\upv}{p}{\tau} & \text{$\upv$ has expansion $p$ matching $\tau$}
-\end{array}\]
-%\newcommand{\gray}[1]{{\color{gray} #1}}
-
-% These judgements are inductively defined in the supplement. 
-% \begingroup 
-% \def\thetheorem{\ref{thm:typed-expansion-short-U}}
-
-%These rules validate the following theorem, which establishes that typed expansion produces an expansion of the assigned type. 
-%\begin{theorem}[Typed Expression Expansion] If $\expandsU{\uDD{\uD}{\Delta}}{\uGG{\uG}{\Gamma}}{\uPsi}{\ue}{e}{\tau}$ and $\uetsmenv{\Delta}{\uPsi}$ then $\hastypeU{\Delta}{\Gamma}{e}{\tau}$.\end{theorem}
-%\begin{proof} This is the first part of Theorem \ref{thm:typed-expansion-U}, defined and proven below.\end{proof}
-
-The typed expansion rules that handle common forms mirror the corresponding typing rules.
-The \emph{expression TLM context}, $\uPsi$, and the \emph{pattern TLM context}, $\uPhi$, pass through these rules opaquely.
-For example, the rules for variables and lambdas are shown being applied in Fig.
-\ref{fig:expansion-exmpl}, discussed below.
-%Each of these rules is based on the corresponding typing rule, i.e.
-Rules (\ref{rule:hastypeU-var}) through (\ref{rule:hastypeU-case}), respectively.
-For example, the following typed expansion rules are based on the typing rules (\ref{rule:hastypeU-var}), (\ref{rule:hastypeU-lam}) and (\ref{rule:hastypeU-ap}), respectively:% for unexpanded expressions of variable, function and application form, respectively:
-% {\small\begin{mathpar}
-%   \inferrule[ee-id]{ }{\expandsUP{\uDelta}{\uGamma, \uGhyp{\ux}{x}{\tau}}{\uPsi}{\uPhi}{\ux}{x}{\tau}}
-
-%   \inferrule[ee-lam]{
-%     \expandsTU{\uDelta}{\utau}{\tau}\\
-%     \expandsUP{\uDelta}{\uGamma, \uGhyp{\ux}{x}{\tau}}{\uPsi}{\uPhi}{\ue}{e}{\tau'}
-%   }{\expandsUPX{\lam{\ux}{\utau}{\ue}}{\aelam{\tau}{x}{e}}{\aparr{\tau}{\tau'}}}
-% \end{mathpar}}
-
-The only subtlety related to common forms has to do with the relationship between identifiers, $\ux$, in the UL and variables, $x$, in the XL.
-To understand this, we must first describe in detail how unexpanded contexts work.
-\emph{Unexpanded typing contexts}, $\uGamma$, are pairs of the form $\uGG{\uG}{\Gamma}$, where $\uG$ maps each expression identifier $\ux \in \domof{\uG}$ to the hypothesis $\vExpands{\ux}{x}$, for some expression variable, $x$, called its expansion.
-The standard typing context, $\Gamma$, then tracks the type of $x$.
-We write $\ctxUpdate{\uG}{\ux}{x}$ for the expression identifier expansion context that maps $\ux$ to $\vExpands{\ux}{x}$ and defers to $\uG$ for all other expression identifiers (i.e.
-the previous mapping is \textbf{updated}.) Note the distinction between update and extension (which requires that the new identifier is not already in the domain.) %We write $\uGammaOK{\uGamma}$ when $\uGamma=\uGG{\uG}{\Gamma}$ and each expression variable in $\uG$ is assigned a type by $\Gamma$.
+The only technical subtlety related to common forms has to do with the relationship between identifiers, $\ux$, in the UL and variables, $x$, in the core language.
+We might hope to identify identifiers in the UL with variables in the core language and track the bindings in unexpanded terms using typing contexts as defined in  the core language, but we cannot because the only operation for producing a new typing context from an existing typing context is context extension, written $\Gamma, x : \tau$, which is defined only when $x \notin \domof{\Gamma}$.
+When working with abstract binding trees, i.e.
+terms identified up to variable renaming, we typically need to give no thought to this condition because it is always possible to implicitly rename the term under consideration when shadowing occurs to discharge this requirement.
+However, we cannot implicitly rename unexpanded terms.
+Changing the definition of typing contexts would have significant implications throughout the metatheory of the core language, which we seek to avoid touching.
+
+Instead, we define \emph{unexpanded typing contexts}, $\uGamma$, as pairs of the form $\uGG{\uG}{\Gamma}$, where $\uG$ maps each expression identifier $\ux \in \domof{\uG}$ to a variable, $x$, written $\vExpands{\ux}{x}$.
+The typing context, $\Gamma$, only tracks the type of these variables.
+We define the identifier update operation $\ctxUpdate{\uG}{\ux}{x}$ as producing the expression identifier expansion context that maps $\ux$ to $x$, written $\vExpands{\ux}{x}$, and defers to $\uG$ for all other expression identifiers, with no requirement that $\ux$ be apart from $\domof{\uG}$.
+%Note the distinction between update and extension (which requires that the new identifier is not already in the domain.) %We write $\uGammaOK{\uGamma}$ when $\uGamma=\uGG{\uG}{\Gamma}$ and each expression variable in $\uG$ is assigned a type by $\Gamma$.
 %\begin{definition} $\uGammaOK{\uGG{\uG}{\Gamma}}$ iff for each $\vExpands{\ux}{x} \in \uG$, we have $\Ghyp{x}{\tau} \in \Gamma$ for some $\tau$.\end{definition}
@@
 We define $\uGamma, \uGhyp{\ux}{x}{\tau}$ when $\uGamma = \uGG{\uG}{\Gamma}$ as an abbreviation of $\uGG{\ctxUpdate{\uG}{\ux}{x}}{\Gamma, \Ghyp{x}{\tau}}$.
-
-To develop an intuition for why the update operation is necessary, it is instructive to inspect in Fig.
-\ref{fig:expansion-exmpl} the derivation of the expansion of the unexpanded expression $\lam{\ux}{\utau}{\lam{\ux}{\utau}{\ux}}$ to $\aelam{\tau}{x_1}{\aelam{\tau}{x_2}{x_2}}$ assuming $\expandsTU{\uDelta}{\utau}{\tau}$.
-Notice that when Rule \textsc{ee-lam} is applied, the type identifier expansion context is updated but the typing context is extended with a (necessarily fresh) variable, first $x_1$ then $x_2$.
-Without this mechanism, expansions for unexpanded terms with shadowing, like this minimal example, would not exist, because we cannot implicitly alpha-vary the unexpanded term to sidestep this problem in the usual manner.
+Unexpanded type formation contexts, $\uDelta$, are analagous.
+
+To develop an intuition for how this formulation solves the problem, it is instructive to inspect the derivation in Fig.~\ref{fig:expansion-exmpl} of the expansion of the unexpanded expression $\lam{\ux}{\utau}{\lam{\ux}{\utau}{\ux}}$ assuming $\expandsTU{\uDelta}{\utau}{\tau}$.
+Notice that each time Rule \textsc{ee-lam} is applied, the type identifier expansion context is updated but the typing context is extended with a fresh variable, first $x_1$ then $x_2$, which is possible by alpha varying only the expansion, leaving the unexpanded term unchanged.
+% Without this mechanism, expansions for unexpanded terms with shadowing would not be derivable.\todo{add let notation to show how this is actually relevant?}
 
@@
 % \end{theorem}
-% For the typed expansion rules governing common forms, like the two example rules applied in Fig.
-\ref{fig:expansion-exmpl}, it is easy to see that this invariant holds.
+% For the typed expansion rules governing common forms, like the two example rules applied in Fig.~\ref{fig:expansion-exmpl}, it is easy to see that this invariant holds.
 The  rules of particular interest are the rules governing TLM definitions and TLM application, which are the topic of the remainder of this section.
@@
 
-\subsection{TLM Definitions}\label{sec:U-uetsm-definition}\label{sec:s-TLM-def}
+\subsubsection{TLM Definitions}\label{sec:U-uetsm-definition}\label{sec:s-TLM-def}
 % An unexpanded expression of seTLM definition form, $\uesyntaxq{\tsmv}{\utau}{\eparse}{\ue}$, 
@@
  % defines an {seTLM} identified as $\tsmv$ with \emph{unexpanded type annotation} $\utau$ and \emph{parse function} $\eparse$ for use within $\ue$. 
-Rule \textsc{ee-def-setsm} in Fig.
-\ref{fig:ee-def-setsm} governs simple expression TLM (seTLM) definitions.
-The first premise expands the unexpanded type annotation.
-The second premise checks that $\eparse$ is a closed expanded function of the given function type.
-(In the supplement, we add the machinery necessary for parse functions that are neither closed nor yet expanded.)
+ There are four unexpanded forms that are not common forms -- the two TLM definition forms and the two TLM application forms.
+Let us start with TLM definitions.
+Rule \textsc{ee-def-setlm} in Fig.~\ref{fig:ee-def-setlm} governs simple expression TLM (seTLM) definitions.
+% Each expression TLM definition specifies a TLM identifier, $\tsmv$, an unexpanded expansion type, $\utau$, a parse function, $\eparse$ and an unexpanded dependency, $\uedep$.
+TLM definitions are scoped to the specified unexpanded expression, $\ue$.
+The premises of Rule~\textsc{ee-def-setlm} operate as follows.
+
+ The first premise expands the unexpanded expansion type, $\utau$, producing the expansion type, $\tau$.
+
+ The second premise checks that the parse function, $\eparse$, is a closed expanded function with input type $\tBody$ and return type $\tParseResultExp$.\footnote{Think of $\eparse$ as the result of parser name resolution, which produces a ``compiled''---i.e.
+closed and expanded---term.}
+ %An alternative design would be to index the typed expansion judgements by a mapping of parser names to these parse functions, but we chose this design to avoid passing this index through every judgement.}
+ The type abbreviated $\tBody$ classifies encodings of literal bodies, $b$.
+Rather than defining $\tBody$ explicitly it suffices to take as a condition that there is an isomorphism between literal bodies and values of type $\tBody$ mediated in one direction by a judgement $\encodeBody{b}{\ebody}$ that is used in the rule for TLM application discussed below.
+The return type, $\tParseResultExp$, abbreviates a labeled sum type,
+{$
+% L_\mathtt{SE} & \defeq \lbltxt{ParseError}, \lbltxt{SuccessE}\\ \asumNL{
+  \asum{~}{\mapitem{\lbltxt{Error}}{\prodt{}}, 
+  \mapitem{\lbltxt{SuccessE}}{\tCEExp}}
+$}, that allows the TLM's parser to distinguish parse errors from successful parses.\footnote{In \lifootnote{Relit}, we used an exception, \lifootnote{Relit.ExpansionError}, rather than a sum type for the same purpose.}
 
@@
 {\small\begin{mathpar}
-\inferrule[ee-def-setsm]{
-  \expandsTU{\uDelta}{\utau}{\tau}\\
-  \hastypeU{\emptyset}{\emptyset}{\eparse}{\aparr{\tBody}{\tParseResultExp}}\\\\
-  \evalU{\eparse}{\eparse'}\\
-  \expandsUP{\uDelta}{\uGamma}{\uPsi, \uShyp{\tsmv}{a}{\tau}{\eparse'}}{\uPhi}{\ue}{e}{\tau'}
+\inferrule[ee-def-setlm]{
+  \expandsTU{\uDelta}{\utau}{\tau}\\\\
+  \hastypeU{\emptyset}{\emptyset}{\eparse}{\aparr{\tBody}{\tParseResultExp}}\\
+  \expandsUP{\uDelta}{\uGamma}{\uPsi}{\uPhi}{\uedep}{\edep}{\taudep}\\\\
+  \uGamma = \uGG{\uG}{\Gamma}\\
+  \expandsUP{\uDelta}{\uGG{\uG}{\Gamma, x : \taudep}}{\uPsi, \uShyp{\tsmv}{x}{\tau}{\eparse}}{\uPhi}{\ue}{e}{\tau'}\\\\
+  e_\text{defn} = \aeap{\aelam{\taudep}{x}{e}}{\edep}
 }{
-  \expandsUPX{\uesyntaxq{\tsmv}{\utau}{\eparse}{\ue}}{e}{\tau'}
-}
-~~~~~~~~~~~~\hspace{15px}
-\inferrule[ee-ap-setsm]{
-  \encodeBody{b}{\ebody}\\
-  \evalU{\ap{\eparse}{\ebody}}{\aein{\mathtt{SuccessE}}{\ecand}}\\\\
-  \decodeCondE{\ecand}{\ce}\\
-  \segOK{\segof{\ce}}{b}\\\\
-  \cvalidE{\emptyset}{\emptyset}{\esceneUP{\uDelta}{\uGamma}{\uPsi}{\uPhi}{b}}{\ce}{e}{\tau}
-}{
-  \expandsUP{\uDelta}{\uGamma}{\uPsi', \uShyp{\tsmv}{a}{\tau}{\eparse}}{\uPhi}{\utsmap{\tsmv}{b}}{e}{\tau}
+  \expandsUPX{\uesyntaxqr{\tsmv}{\utau}{\eparse}{\ue}}{e_\text{defn}}{\tau'}
 }
 \end{mathpar}}
-\caption{The typed expansion rules for expression TLM definition and application.}
-\label{fig:ee-def-setsm}
+{\small\begin{mathpar}
+\inferrule[ee-def-sptlm]{
+  \expandsTU{\uDelta}{\utau}{\tau}\\
+  \hastypeU{\emptyset}{\emptyset}{\eparse}{\aparr{\tBody}{\tParseResultPat}}\\\\
+  \expandsUP{\uDelta}{\uGamma}{\uPsi}{\uPhi, \uPhyp{\tsmv}{\_}{\tau}{\eparse}}{\ue}{e}{\tau'}
+}{
+  \expandsUPX{\usyntaxup{\tsmv}{\utau}{\eparse}{\ue}}{e}{\tau'}
+}
+\end{mathpar}}
+\caption{The typed expansion rules for expression and pattern TLM definitions.}
+\label{fig:ee-def-setlm}
+\vspace{-5px}
 \end{figure*}
@@
 
-The type abbreviated $\tBody$ classifies encodings of literal bodies, $b$.
-Rather than defining $\tBody$ explicitly it suffices to take as a condition that there is an isomorphism between literal bodies and values of type $\tBody$ mediated in one direction by a judgement $\encodeBody{b}{\ebody}$ that will come up below.
-
-The return type, $\tParseResultExp$, abbreviates a labeled sum type that distinguishes parse errors from successful parses:
-{$
-% L_\mathtt{SE} & \defeq \lbltxt{ParseError}, \lbltxt{SuccessE}\\ \asumNL{
-  \mapitem{\lbltxt{ParseError}}{\prodt{}}, 
-  \mapitem{\lbltxt{SuccessE}}{\tCEExp}
-$}.
 % \end{align*}} %[\mapitem{\lbltxt{ParseError}}{\prodt{}}, \mapitem{\lbltxt{SuccessE}}{\tCEExp}]
@@
 
-The type abbreviated $\tCEExp$ classifies encodings of \emph{proto-expressions}, $\ce$ (pronounced ``grave $e$''.) The syntax of proto-expressions, defined in Figure \ref{fig:U-candidate-terms}, will be described when we describe proto-expansion validation in Sec.
+The type abbreviated $\tCEExp$ classifies encodings of \emph{proto-expressions}, $\ce$ (pronounced ``grave $e$'').
+The syntax of proto-expressions, defined in Fig.~\ref{fig:U-candidate-terms}, will be described when we describe proto-expansion validation in Sec.
 \ref{sec:ce-syntax-U}.
-The mapping from proto-expressions to values of type $\tCEExp$ is defined by the \emph{proto-expression encoding judgement}, $\encodeCondE{\ce}{e}$.
-An inverse mapping is defined by the \emph{proto-expression decoding judgement}, $\decodeCondE{e}{\ce}$.
-Again, we need only take as a condition that there is an isomorphism between values of type $\tCEExp$ and closed proto-expressions mediated by these judgements (see supplement.)
-
-The third premise of Rule \textsc{ee-def-setsm} evaluates the parse function to a value. This is not necessary, but it is the choice one would expect to make in an eager language.
-
-The final premise of Rule \textsc{ee-def-setsm} extends the expression TLM context, $\uPsi$, with the newly determined {seTLM definition}, and proceeds to assign a type, $\tau'$, and expansion, $e$, to $\ue$.
+As with $\tBody$, we need only take as a condition that there is an isomorphism between values of type $\tCEExp$ and closed proto-expressions, which is mediated in one direction by the \emph{proto-expression decoding judgement}, $\decodeCondE{e}{\ce}$ that we will return to when we describe TLM application below (see supplement for the full condition).
+
+In $\miniVersePat$, we model the \lismall{expansions require} clause as specifying a single value dependency.
+%\footnote{In \lifootnote{Relit}, this is further restricted to a module expression with a singleton signature, for reasons discussed in Sec.~\ref{sec:implementation}.} 
+The third premise of Rule~\textsc{ee-def-setlm} generates the expanded dependency, $\edep$ of type $\taudep$, from the unexpanded dependency, $\uedep$.
+Note that we do not need to explicitly allow for type dependencies (which would be formally cumbersome because we cannot ``tuple together'' types like we can values in this setting).
+Instead, type dependencies can be expressed by dependency on a value of existential type to be unpacked by the expansion.
+This existential type can in turn be expressed in terms of universal types by the well-known encoding \cite{B304,pfple1}.
+This workaround is perhaps unsurprising given that existentials relate closely to modules, which package both types and values \cite{pfple1,mitchell1988abstract}.
+
+Having processed the TLM definition, we are ready to continue into the unexpanded expression $\ue$ where the TLM is bound.
+To activate the TLM definition for use by $\ue$, the third row of premises in Rule~\textsc{ee-def-setlm} first generates a fresh variable, \li{x}, to stand for the value dependency.
+We will use this variable to instantiate the dependency in each generated expansion when we discuss TLM application below.
+Note that there is no corresponding expression identifier in $\uG$.
+Second, the rule extends the expression TLM context, $\uPsi$, to associate the TLM identifier $\tsmv$ with $x$, as well as with the given expansion type and parse function.
+If $\tsmv$ was already defined, the previous definition is shadowed (the supplement gives some additional details on how TLM contexts are structured).
+
+The final premise of Rule~\textsc{ee-def-setlm}, together with the conclusion of the rule, specifies the expansion of the TLM definition as being of function application form---it wraps $e$, where the variable $x$ stands free for the dependency, with a lambda binding $x$, and then immediately applies it to pass down the actual value dependency, $\edep$.
+In other words, it \li{let}-binds the dependency.
+This defers to the core language with regard to whether $\edep$ is evaluated eagerly or lazily.
+
+% The final premise of Rule \textsc{ee-def-setlm} extends the expression TLM context, $\uPsi$, with the newly determined {seTLM definition}, and proceeds to assign a type, $\tau'$, and expansion, $e$, to $\ue$.
 The conclusion of the rule then assigns this type and expansion to the seTLM definition as a whole.
@@
 
-{Expression TLM contexts}, $\uPsi$, are of the form $\uAS{\uA}{\Psi}$, where $\uA$ is a \emph{TLM identifier expansion context} and $\Psi$ is an \emph{expression TLM definition context}.
+% {Expression TLM contexts}, $\uPsi$, are of the form $\uAS{\uA}{\Psi}$, where $\uA$ is a \emph{TLM identifier expansion context} and $\Psi$ is an \emph{expression TLM definition context}.
 We distinguish TLM identifiers, $\tsmv$, from TLM names, $a$, for much the same reason that we distinguish type and expression identifiers from type and expression variables: in order to allow a TLM definition to shadow a previously defined TLM definition without relying on an implicit identification convention.
@@
 
-An {expression TLM definition context}, $\Psi$, is a finite function mapping each TLM name $a \in \domof{\Psi}$ to an \emph{expanded seTLM definition}, $\xuetsmbnd{a}{\tau}{\eparse}$, where $\tau$ is the seTLM's type annotation, and $\eparse$ is its parse function.
-We define $\uPsi, \uShyp{\tsmv}{a}{\tau}{\eparse}$, when $\uPsi=\uAS{\uA}{\Psi}$, as an abbreviation of $\uAS{\ctxUpdate{\uA}{\tsmv}{a}}{\Psi, \xuetsmbnd{a}{\tau}{\eparse}}$.
+% An {expression TLM definition context}, $\Psi$, is a finite function mapping each TLM name $a \in \domof{\Psi}$ to an \emph{expanded seTLM definition}, $\xuetsmbnd{a}{\tau}{\eparse}$, where $\tau$ is the seTLM's type annotation, and $\eparse$ is its parse function.
+% We define $\uPsi, \uShyp{\tsmv}{a}{\tau}{\eparse}$, when $\uPsi=\uAS{\uA}{\Psi}$, as an abbreviation of $\uAS{\ctxUpdate{\uA}{\tsmv}{a}}{\Psi, \xuetsmbnd{a}{\tau}{\eparse}}$.
 
@@
 
-The simple pattern TLM (spTLM) definition form operates analagously (see supplement), with the spTLM context, $\uPhi$, rather than the $\uPsi$ updated.
-This allows expression and pattern TLMs to use the same identifiers.
+Rule \textsc{ee-ap-sptlm} for pattern TLM definitions shown in Fig.
+\ref{fig:ee-def-setlm} is analagous but simpler, because in $\miniVersePat$ patterns are entirely structural (there are no module paths that they might depend on).
+%As such, we do not need the machinery for dependency passing.
 % {\vspace{-3px}\small\begin{mathpar}
@@
 
-\subsection{TLM Application}\label{sec:U-uetsm-application}\label{sec:s-TLM-ap}
-The unexpanded expression form for applying an seTLM named $\tsmv$ to a literal form with literal body $b$ is $\utsmap{\tsmv}{b}$. 
+\subsubsection{TLM Application}\label{sec:U-uetsm-application}\label{sec:s-TLM-ap}
+The unexpanded expression form for applying an seTLM identified as $\tsmv$ to a literal form with literal body $b$ is $\utsmap{\tsmv}{b}$. 
 % This form uses forward slashes\todo{use backticks} to delimit the literal body, but other generalized literal forms could also be included as derived forms in the textual syntax.
@@
 $\mathsf{str}[s]$ and $\mathsf{num}[n]$.} This operator is indexed by the TLM name $\tsmv$ and takes no arguments.
-Rule \textsc{ee-ap-setsm} governing this form is shown in Fig. \ref{fig:ee-def-setsm}.
+Rule \textsc{ee-ap-setlm} governing this form is in Fig.~\ref{fig:ee-ap-setlm}.
 
@@
 % {\small\begin{mathpar}
-% \inferrule[ee-ap-setsm]{
+% \inferrule[ee-ap-setlm]{
 %   \encodeBody{b}{\ebody}\\
@@
 % \end{mathpar}}
-
-The first premise encodes the literal body, $\ebody$, which, as described above, is a value of type $\tBody$.
-
-The second premise applies the parse function $\eparse$ to the encoding of the literal body.
+The first two premises serve simply to ``look up'' $\tsmv$ in the expression TLM context, $\uPsi$, and to look up the correspondency dependency variable, $x$, in $\uGamma$.
+
+The third premise encodes the literal body, $\ebody$, producing a value $\ebody$ of type $\tBody$ according to the body encoding judgement $\encodeBody{b}{\ebody}$ described in Sec.~\ref{sec:s-TLM-def} above.
+
+The fourth premise applies the parse function $\eparse$ to the encoding of the literal body.
 If parsing succeeds, i.e.
-a value of the form $\aein{\mathtt{SuccessE}}{\ecand}$ results from evaluation, then $\ecand$ will be a value of type $\tCEExp$ (assuming a well-formed expression TLM context, by application of Type Safety.) We call $\ecand$ the \emph{encoding of the proto-expansion}.
-If the parse function produces a value labeled $\lbltxt{ParseError}$, then typed expansion fails and formally, no rule is necessary.
-
-The third premise decodes the encoding of the proto-expansion using the judgement described in Sec. \ref{sec:U-uetsm-definition}.
-
-The fourth premise of Rule \textsc{ee-ap-setsm} determines the segmentation of the proto-expansion, $\segof{\ce}$, and ensures that it is valid with respect to $b$ via the predicate $\segOK{\psi}{b}$, which  checks that each segment in the finite set of segments $\psi$ has non-negative length and is within bounds of $b$, and that the segments in $\psi$ do not overlap.
-
-The final premise \emph{validates} the proto-expansion and simultaneously generates the \emph{final expansion}, $e$, which appears in the conclusion of the rule.
-The proto-expression validation judgement is defined in the next subsection.
-
-The typed pattern expansion rule governing pattern TLM application is analagous (see supplement).
+a value of the form $\aein{\mathtt{SuccessE}}{\ecand}$ results from evaluation, then $\ecand$ will be a value of type $\tCEExp$ (by type safety and the canonical forms lemma, assuming a well-formed expression TLM context).
+We call $\ecand$ the \emph{encoding of the proto-expansion}.
+If the parse function produces a value labeled $\lbltxt{Error}$, then typed expansion fails and formally, no rule is necessary.
+
+\begin{figure*}
+{\small\begin{mathpar}
+\inferrule[ee-ap-setlm]{
+  \uPsi = \uPsi', \uShyp{\tsmv}{x}{\tau}{\eparse}\\
+  \uGamma = \uGG{\uG}{\Gamma, x : \taudep}\\\\
+  \encodeBody{b}{\ebody}\\
+  \evalU{\ap{\eparse}{\ebody}}{\aein{\mathtt{SuccessE}}{\ecand}}\\
+  \decodeCondE{\ecand}{\ce}\\\\
+  \segOK{\segof{\ce}}{b}\\
+  \cvalidE{\emptyset}{\emptyset}{\esceneUP{\uDelta}{\uGamma}{\uPsi}{\uPhi}{b}}{\ce}{e}{\aparr{\taudep}{\tau}}
+}{
+  \expandsUP{\uDelta}{\uGamma}{\uPsi}{\uPhi}{\utsmap{\tsmv}{b}}{\aeap{e}{x}}{\tau}
+}
+\end{mathpar}}
+{\small\begin{mathpar}
+\inferrule[pe-ap-sptlm]{
+  \uPhi = \uPhi', \uPhyp{\tsmv}{\_}{\tau}{\eparse}\\\\
+  \encodeBody{b}{\ebody}\\
+  \evalU{\ap{\eparse}{\ebody}}{\aein{\mathtt{SuccessP}}{\ecand}}\\
+  \decodeCEPat{\ecand}{\cpv}\\\\
+    \segOK{\segof{\cpv}}{b}\\
+  \cvalidP{\upctx}{\pscene{\uDelta}{\uPhi}{b}}{\cpv}{p}{\tau}
+}{
+  \patExpands{\upctx}{\uPhi}{\utsmap{\tsmv}{b}}{p}{\tau}
+}\end{mathpar}}
+\caption{The typed expansion rules for expression and pattern TLM application.}
+\label{fig:ee-ap-setlm}
+\end{figure*}
+
+The fifth premise decodes the encoding of the proto-expansion using the judgement described in Sec. \ref{sec:U-uetsm-definition}, producing the proto-expansion itself, $\ce$.
+
+The final two premises validate the proto-expansion.
+Proto-expansion validation is described in Sec.~\ref{sec:s-PEV} below.
+In $\miniVersePat$, the proto-expansion does not expand directly to the final expansion but to a function, here $e$, from the dependency type, $\taudep$, to the expansion type, $\tau$, as suggested by the right-hand side of the final premise.
+In the conclusion of the rule, we apply the dependency variable, $x$, to produce the final expansion.\footnote{In \lifootnote{Relit}, the dependency is always on the singleton module induced by the \lifootnote{expansions require} clause, which is \lifootnote{open}ed immediately, so the necessary boilerplate is inserted automatically.% If this were changed to require that the proto-expansion be a curried functor over dependencies listed in lexical order, the upside would be that the module identifiers used internally to a proto-expansion would not be exposed in the TLM definition, but the downside is that this would require the signature of each dependency be reproduced in each proto-expansion.
+We chose the more concise approach.
+}
+
+The typed pattern expansion rule governing pattern TLM application, Rule~\textsc{pe-ap-sptlm} in Fig.~\ref{fig:ee-ap-setlm}, is analagous but again simpler because there is no need to pass dependencies into patterns in $\miniVersePat$.\footnote{If there were a need to pass dependencies into patterns, we would need either pattern functions or to use an encoding trick, e.g.
+by requiring the result be a function containing a ``dummy'' match with one pattern.} We describe proto-pattern validation in Sec.~\ref{sec:s-PEV} below.
 % {\small\begin{mathpar}
@@
 \subsection{Proto-Expansion Validation}\label{sec:ce-validation-U}\label{sec:ce-syntax-U}\label{sec:s-PEV}
-
+Proto-expansion validation occurs in two steps, corresponding to the final two premises of the TLM application rules in Fig.~\ref{fig:ee-ap-setlm}.
+
+The first of these two premises determines the segmentation of the proto-expansion by pulling out the splice references and ensures that it is valid via the predicate $\segOK{\psi}{b}$, where $\psi$ is the segmentation.
+This checks that each segment  has positive length and is within bounds of $b$, and that the segments do not overlap and are separated by at least one character (see supplement).
+
+The second of these two premises {typechecks} the proto-expression or proto-pattern, and simultaneously generates a corresponding core language expression or pattern, from which the final expansion is constructed as described above.
+This step is specified by the following judgements:% are types and expanded expressions, respectively.
+\[\begin{array}{ll}
+\cvalidT{\Delta}{\tscenev}{\ctau}{\tau} & \text{$\ctau$ has well-formed expansion $\tau$}\\
+\cvalidE{\Delta}{\Gamma}{\escenev}{\ce}{e}{\tau} & \text{$\ce$ has expansion $e$ of type $\tau$}\\
+% \cvalidR{\Delta}{\Gamma}{\escenev}{\crv}{r}{\tau}{\tau'} & \text{$\crv$ has expansion $r$ taking values of type $\tau$ to values of type $\tau'$}\\
+\cvalidP{\upctx}{\pscenev}{\cpv}{p}{\tau} & \text{$\cpv$ has expansion $p$ matching $\tau$}
+\end{array}\]
+
+The purpose of the \emph{splicing scenes} $\tscenev$, $\escenev$ and $\pscenev$ is to ``remember'' the contexts and literal body from the TLM application site (cf.
+the final premise of Rule \textsc{ee-ap-setlm} in Fig.~\ref{fig:ee-def-setlm}) for when validation encounters spliced terms.
+For example, \emph{expression splicing scenes}, $\escenev$, are of the form $\esceneUP{\uDelta}{\uGamma}{\uPsi}{\uPhi}{b}$.
+
+\paragraph{Common Forms} Most of the proto-expansion forms, including all of those elided in Fig.~\ref{fig:U-candidate-terms} mirror corresponding expanded forms.
+The rules governing proto-expansion validation for these common forms  in the supplement correspondingly mirror the typing rules.
+Splicing scenes---$\escenev$, $\tscenev$ and $\pscenev$--pass opaquely through these rules, i.e.
+none of these rules can access the application site contexts.
+Notice that the initial expansion-interal typing contexts, $\Delta$ and $\Gamma$, start out empty in the rules in Fig.~\ref{fig:ee-ap-setlm}.
+Together, this maintains context independence (defined formally below).
+
+Notice that proto-rules, $\crv$, involve expanded patterns, $p$, not proto-patterns, $\cpv$, because proto-rules appear in proto-expressions, which are generated by expression TLMs.
+Proto-patterns arise only from pattern TLMs.
+There is no variable proto-pattern form, for the reasons described in Sec.
+\ref{sec:sptsms}.
+
+\paragraph{Splice References} 
 \begin{figure}
@@
 \aceap{\ce}{\ce} ~\vert~
-\cdots \\
+\cdots ~\vert~ 
 % \acetlam{t}{\ce} ~\vert~
@@
 % \acein{\ell}{\ce} ~\vert~
-& & \vert & 
 \acematchwith{n}{\ce}{\seqschemaX{\crv}} ~\vert~
@@
 \end{figure}
-
-Finally, we arrive at the crucial \emph{proto-expansion validation judgements}, which validate the proto-expansions generated by TLMs and simultaneously generate their final expansions:% are types and expanded expressions, respectively.
-\[\begin{array}{ll}
-\cvalidT{\Delta}{\tscenev}{\ctau}{\tau} & \text{$\ctau$ has well-formed expansion $\tau$}\\
-\cvalidE{\Delta}{\Gamma}{\escenev}{\ce}{e}{\tau} & \text{$\ce$ has expansion $e$ of type $\tau$}\\
-% \cvalidR{\Delta}{\Gamma}{\escenev}{\crv}{r}{\tau}{\tau'} & \text{$\crv$ has expansion $r$ taking values of type $\tau$ to values of type $\tau'$}\\
-\cvalidP{\upctx}{\pscenev}{\cpv}{p}{\tau} & \text{$\cpv$ has expansion $p$ matching $\tau$}
-\end{array}\]
-
-The purpose of the \emph{splicing scenes} $\tscenev$, $\escenev$ and $\pscenev$ is to ``remember'' the contexts and literal body from the TLM application site (cf.
-the final premise of Rule \textsc{ee-ap-setsm} in Fig.
-\ref{fig:ee-def-setsm}) for when validation encounters spliced terms.
-For example, \emph{expression splicing scenes}, $\escenev$, are of the form $\esceneUP{\uDelta}{\uGamma}{\uPsi}{\uPhi}{b}$.
-
-\paragraph{Common Forms} Most of the proto-expansion forms, including all of those elided in Fig.
-\ref{fig:U-candidate-terms} mirror corresponding expanded forms.
-The rules governing proto-expansion validation for these common forms  in the supplement correspondingly mirror the typing rules.
-Splicing scenes---$\escenev$, $\tscenev$ and $\pscenev$--pass opaquely through these rules, i.e.
-none of these rules can access the application site contexts.
-This maintains context independence (defined formally below.)
-
-Notice that proto-rules, $\crv$, involve expanded patterns, $p$, not proto-patterns, $\cpv$.
-The reason is that proto-rules appear in proto-expressions, which are generated by expression TLMs.
-Proto-patterns, in contrast, arise only from pattern TLMs.
-There is not a variable proto-pattern form, for the reasons described in Sec.
-\ref{sec:sptsms}.
-
-\paragraph{References to Spliced Terms} The only interesting forms are the references to spliced unexpanded types, expressions and patterns.
-Let us consider the rule for references to spliced unexpanded expressions:
+The only interesting forms in Fig.~\ref{fig:U-candidate-terms} are the references to spliced types, expressions and patterns. Let us consider the rule for references to spliced expressions:
+
 {\small\begin{mathpar}
@@
   \parseUExp{\bsubseq{b}{m}{n}}{\ue}\\
-  \cvalidT{\emptyset}{\tsceneUP{\uDelta}{b}}{\ctau}{\tau}\\
+  \cvalidT{\emptyset}{\tsceneUP{\uDD{\uD}{\Delta_\text{app}}}{b}}{\ctau}{\tau}\\
   \expandsUP{\uDD{\uD}{\Delta_\text{app}}}{\uGG{\uG}{\Gamma_\text{app}}}{\uPsi}{\uPhi}{\ue}{e}{\tau}\\\\
@@
 The second premise performs proto-type expansion on the given type annotation, $\ctau$, producing a type, $\tau$.
-The third premise then invokes type expansion on $\ue$ \emph{under the application site contexts}, $\uDD{\uD}{\Delta_\text{app}}$ and $\uGG{\uG}{\Gamma_\text{app}}$, but \emph{not} the expansion-local contexts, $\Delta$ and $\Gamma$.
+The expansion-internal type variables in $\Delta$ are not available to $\tau$.
+The third premise then invokes type expansion on $\ue$ \emph{under the application site contexts}, $\uDD{\uD}{\Delta_\text{app}}$ and $\uGG{\uG}{\Gamma_\text{app}}$, but \emph{not} the expansion-internal contexts, $\Delta$ and $\Gamma$.
 The final premise requires that the application site contexts are disjoint from the expansion-local type formation context.
@@
 This serves to enforce capture avoidance.
-%(we will formally state this property in the next section.)
-
-The rule for references to spliced unexpanded types and patterns are fundamentally analagous (see supplement). 
+Note that the purpose of $\miniVersePat$ is to specify the necessary conditions, not to specify a particular implementation strategy.
+There are various ways to implement this capture avoidance condition.
+We will formally state the capture avoidance property in terms of capture avoiding substitution below, which is one strategy.
+Another is to pro-actively generate fresh variables for all internal bindings \emph{a priori}.
+
+The rule for references to spliced unexpanded types and patterns are analagous (see supplement).
+Note that we did not give examples of type splicing in the previous sections, but it is occasionally useful.
+For example, a TLM that implements a parser generator in the style of Menhir could need type splicing for reading in the non-terminal types, e.g.
+as on Line 9 of Fig.~\ref{fig:regex-parser}.
 % \noindent
@@
 \subsection{Metatheory}\label{sec:s-metatheory}
+Let us now sketch the main metatheoretic properties of $\miniVersePat$.
 
@@
 \begin{enumerate}[nolistsep]
-\item If $\expandsU{\uDD{\uD}{\Delta}}{\uGG{\uG}{\Gamma}}{\uAS{\uA}{\Psi}}{\ue}{e}{\tau}$ then $\hastypeU{\Delta}{\Gamma}{e}{\tau}$.
-\item If $\cvalidE{\Delta}{\Gamma}{\esceneU{\uDD{\uD}{\Delta_\text{app}}}{\uGG{\uG}{\Gamma_\text{app}}}{\uAS{\uA}{\Psi}}{b}}{\ce}{e}{\tau}$ and $\Delta \cap \Delta_\text{app} = \emptyset$ and $\domof{\Gamma} \cap \domof{\Gamma_\text{app}} = \emptyset$ then $\hastypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\Gcons{\Gamma}{\Gamma_\text{app}}}{e}{\tau}$.
+\item If $\expandsUP{\uDD{\uD}{\Delta}}{\uGG{\uG}{\Gamma}}{\uPsi}{\uPhi}{\ue}{e}{\tau}$ then $\hastypeU{\Delta}{\Gamma}{e}{\tau}$.
+\item If $\cvalidE{\Delta}{\Gamma}{\esceneUP{\uDD{\uD}{\Delta_\text{app}}}{\uGG{\uG}{\Gamma_\text{app}}}{\uPsi}{\uPhi}{b}}{\ce}{e}{\tau}$ and $\Delta \cap \Delta_\text{app} = \emptyset$ and $\domof{\Gamma} \cap \domof{\Gamma_\text{app}} = \emptyset$ then $\hastypeU{(\Dcons{\Delta}{\Delta_\text{app}})}{(\Gcons{\Gamma}{\Gamma_\text{app}})}{e}{\tau}$.
 \end{enumerate}
 \end{theorem}
+\vspace{-4px}
 
 % \endgroup
-The additional second clause simply states that the final expansion produced by proto-expression validation is well-typed under the combined application site and expansion-internal context (because spliced terms are distinguished only in the proto-expansion, not in the final expansion.) The combined context can only be formed if these are disjoint.
-
-The proof proceeds by mutual rule induction and appeal to simple lemmas about type expansion and proto-type validation (see supplement).
-The proof is straightforward but for one issue: it is not immediately clear that the mutual induction is well-founded, because the case in the proof of part 2 for Rule \textsc{pev-spliced} invokes part 1 of the induction hypothesis on a term that is not a sub-term of the conclusion, but rather parsed out of the literal body, $b$.
+The additional second clause simply states that the final expansion produced by proto-expression validation is well-typed under the combined application site and expansion-internal context (because spliced terms are distinguished only in the proto-expansion, but not in the final expansion).
+% The combined context can only be formed if these are disjoint.
+The proof proceeds by mutual rule induction and appeal to the analagous typed pattern expansion theorem and simple lemmas about type expansion   and proto-type validation.
+The only issue is that it is not immediately clear that the mutual induction is well-founded, because the case in the proof of part 2 for Rule \textsc{pev-spliced} invokes part 1 of the induction hypothesis on a term that is not a sub-term of the conclusion, but rather parsed out of the literal body, $b$.
 To establish that the mutual induction is well-founded, then, we need to explicitly establish a decreasing metric.
@@
 % More specifically, {the sum of the lengths of the literal bodies that appear in the term strictly decreases each time you perform a nested TLM application} because some portion of the term has to be consumed by the TLM name and the delimiters.
-The details are given in the supplemental material.
+The details are given in the supplement.
 
@@
 
-\subsubsection{seTLM Reasoning Principles}
-
-The following theorem summarizes the abstract reasoning principles that programmers can rely on when applying a simple expression TLM.
+\subsubsection{TLM Reasoning Principles}
+
+The following theorem summarizes the abstract reasoning principles that client programmers can rely on when applying a simple expression TLM.
 Informal descriptions of the labeled clauses are given inline, in gray boxes.
@@
 % \def\thetheorem{\ref{thm:tsc-SES}}
-
-\begin{theorem}[seTLM Reasoning Principles]\label{thm:setlm-reasoning} ~\\
-If $\expandsU{\uDD{\uD}{\Delta}}{\uGG{\uG}{\Gamma}}{\uPsi}{\utsmap{\tsmv}{b}}{e}{\tau}$ then:
-\begin{enumerate}[nolistsep]
-\item (\textbf{Typing 1}) $\uPsi = \uPsi', \uShyp{\tsmv}{a}{\tau}{\eparse}$ and $\hastypeU{\Delta}{\Gamma}{e}{\tau}$
-  \begin{quote}
+\noindent
+\begin{theorem}[seTLM Reasoning Principles]\label{thm:setlm-reasoning} If $\expandsU{\uDD{\uD}{\Delta}}{\uGG{\uG}{\Gamma}}{\uPsi}{\utsmap{\tsmv}{b}}{e}{\tau}$ then
+\begin{enumerate}[nolistsep,leftmargin=10pt,label={\arabic*.}]
+\item (\textbf{Expansion Typing}) $\uPsi = \uPsi', \uShyp{\tsmv}{\_}{\tau}{\eparse}$ and $\hastypeU{\Delta}{\Gamma}{e}{\tau}$ 
   \begin{grayparbox}
@@
   \end{grayparbox}
-  \end{quote}
-\item $\encodeBody{b}{\ebody}$
-\item (\textbf{Responsibility}) $\evalU{\ap{\eparse}{\ebody}}{\aein{\lbltxt{SuccessE}}{\ecand}}$
-  \begin{quote}
+\item (\textbf{Responsibility}) $\encodeBody{b}{\ebody}$ and $\evalU{\ap{\eparse}{\ebody}}{\aein{\lbltxt{SuccessE}}{\ecand}}$ and $\decodeCondE{\ecand}{\ce}$
   \begin{grayparbox}
-  The parse function of the invoked TLM is responsible for the expansion.
+  The parse function of the applied TLM is responsible for generating the proto-expansion.
   \end{grayparbox}
-  \end{quote}
-\item $\decodeCondE{\ecand}{\ce}$
 \item (\textbf{Segmentation}) $\segOK{\segof{\ce}}{b}$ 
-        \begin{quote}
           \begin{grayparbox}
-        The segmentation determined by the proto-expansion actually segments the literal body (i.e. each segment is in-bounds and the segments are non-overlapping.)
+        The segmentation determined by the proto-expansion segments the literal body.
           \end{grayparbox}
-\end{quote}
-\item $\segof{\ce} = \sseq{\acesplicedt{m'_i}{n'_i}}{\nty} \cup$\\ $\quad\sseq{\acesplicede{m_i}{n_i}{\ctau_i}}{\nexp}$
-\item \textbf{(Typing 2)} $\sseq{
+\item (\textbf{Segment Typing})  $\segof{\ce} = \sseq{\acesplicedt{m'_i}{n'_i}}{\nty} \cup \sseq{\acesplicede{m_i}{n_i}{\ctau_i}}{\nexp}$ and
+  \begin{enumerate}
+  \item $\sseq{
         \expandsTU{\uDD{\uD}{\Delta}}
@@
       }{\nty}$ and $\sseq{\istypeU{\Delta}{\tau'_i}}{\nty}$
-      \begin{quote}
           \begin{grayparbox}
@@
           \end{grayparbox}
-\end{quote}
-\item \textbf{(Typing 3)} $\sseq{
+  \item $\sseq{
     \cvalidT{\emptyset}{
@@
   }{\nexp}$ and $\sseq{\istypeU{\Delta}{\tau_i}}{\nexp}$
-    \begin{quote}
         \begin{grayparbox}
-      Each type annotation on a reference to a spliced expression has a well-formed expansion.
+        Each segment type has a well-formed expansion.
         \end{grayparbox}
-\end{quote}
-\item \textbf{(Typing 4)} ~\\ $\sseq{
+  \item $\sseq{
     \expandsU
@@
       {\tau_i}
-}{\nexp}$ and $\sseq{\hastypeU{\Delta}{\Gamma}{e_i}{\tau_i}}{\nexp}$
-    \begin{quote}
+  }{\nexp}$ and $\sseq{\hastypeU{\Delta}{\Gamma}{e_i}{\tau_i}}{\nexp}$\vspace{-5px}
         \begin{grayparbox}
-      Each spliced expression has a well-typed expansion consistent with the type annotation in the segmentation.
+        Each spliced expression has a well-typed expansion consistent with the segment type.
         \end{grayparbox}
-\end{quote}
-\item (\textbf{Capture Avoidance}) ~\\$e = [\sseq{\tau'_i/t_i}{\nty}, \sseq{e_i/x_i}{\nexp}]e'$ for some variables $\sseq{t_i}{\nty}$ and $\sseq{x_i}{\nexp}$, and $e'$
-    \begin{quote}
+\end{enumerate}
+\item (\textbf{Capture Avoidance}) $e = [\sseq{\tau'_i/t_i}{\nty}, \sseq{e_i/x_i}{\nexp}]e'$ for fresh variables $\sseq{x_i}{\nexp}$ and $\sseq{t_i}{\nty}$ and some $e'$
     \begin{grayparbox}
-      The final expansion can be decomposed into a  term with variables in place of each spliced type or expression.
-The expansions of these spliced types and expressions can be substituted into this term in the standard capture avoiding manner.
+      We can decompose the final expansion, $e$, into an ``internal expression'', $e'$, with fresh variables in place of each spliced type or expression.
+The expansion can be produced by substituting in the expansions of these spliced types and expressions in the standard capture avoiding manner.
     \end{grayparbox}
-    \end{quote}
-\item (\textbf{Context Independence}) ~\\$\mathsf{fv}(e') \subset \sseq{t_i}{\nty} \cup \sseq{x_i}{\nexp}$
-    \begin{quote}
+\item (\textbf{Context Independence}) $\mathsf{fv}(e') \subset \sseq{t_i}{\nty} \cup \sseq{x_i}{\nexp}$
       \begin{grayparbox}
-
-      The decomposed term makes no mention of bindings in the application site context, i.e. the only free variables are those standing for spliced terms.
+      The internal expression makes no mention of bindings in the application site context, i.e. the only free variables remaining are those standing for spliced terms.
       \end{grayparbox}
-\end{quote}
   % $\hastypeU
@@
 Instead, we used standard notions of capture avoiding substitution and free variables combined with the context disjointness conditions in the rules above.
-This is possible only because we keep track of spliced terms explicitly in the proto-expansion, rather than going straight to the final expansion.
+This is possible only because we keep track of spliced terms explicitly in the proto-expansion.
 %In fact, doing so is critical for TLMs -- there is no notion of alpha-conversion for partially parsed terms, so any notion of hygiene that relies on this notion would be inapplicable.
 
-The reasoning principles theorem for pattern TLMs is in the supplement.
-The key clause establishes that the hypotheses generated by the TLM application form are exactly the union of the hypothesis generated by the spliced patterns.
+The reasoning principles theorem for pattern TLMs is below.
+The \textbf{Visibility} clause establishes that the hypotheses generated by a pattern of TLM application form are exactly the union of the hypothesis generated by the spliced patterns---there are no invisible bindings (see Sec.~\ref{sec:sptsms}).
 % No additional hypotheses can be produced by the TLM.
 
-In the full calculus, which supports parametric TLMs, the context independence clause allows reference to the variables standing for parameters.
-
-\newcommand{\pTLMsFormallySec}{Parametric TLMs, Formally}
-\section{\protect\pTLMsFormallySec}
-\label{sec:ptlms-formally}
-
-% \begin{figure}[p]
-% \[\begin{array}{llcl}
-% \mathsf{UMType} & \urho & ::= 
-% %& \autype{\utau} 
-% & \utau & \text{type annotation}\\
+\begin{theorem}[spTLM Abstract Reasoning Principles]
+\label{thm:spTLM-Typing-Segmentation}
+If $\patExpands{\upctx}{\uPhi}{\utsmap{\tsmv}{b}}{p}{\tau}$ where $\uDelta=\uDD{\uD}{\Delta}$ and $\uGamma=\uGG{\uG}{\Gamma}$ then all of the following hold:
+\begin{enumerate}[nolistsep,leftmargin=10pt,label={\arabic*.}]
+        \item (\textbf{Expansion Typing}) $\uPhi=\uPhi', \uPhyp{\tsmv}{\_}{\tau}{\eparse}$ and $\patType{\pctx}{p}{\tau}$
+        \item (\textbf{Responsibility}) $\encodeBody{b}{\ebody}$ and $\evalU{\eparse(\ebody)}{\aein{\mathtt{SuccessP}}{\ecand}}$ and $\decodeCEPat{\ecand}{\cpv}$
+        \item (\textbf{Segmentation}) $\segOK{\segof{\cpv}}{b}$
+        \item (\textbf{Segment Typing}) $\segof{\cpv} = \sseq{\acesplicedt{n'_i}{m'_i}}{\nty} \cup \sseq{\acesplicedp{m_i}{n_i}{\ctau_i}}{\npat}$ and
+          \begin{enumerate}
+          \item $\sseq{
+                \expandsTU{\uDelta}
+                {
+                  \parseUTypF{\bsubseq{b}{m'_i}{n'_i}}
+                }{\tau'_i}
+              }{\nty}$ and $\sseq{\istypeU{\Delta}{\tau'_i}}{\nty}$
+          \item $\sseq{
+            \cvalidT{\emptyset}{
+              \tsceneUP
+                {\uDelta}{b}
+            }{
+              \ctau_i
+            }{\tau_i}
+          }{\npat}$ and $\sseq{\istypeU{\Delta}{\tau_i}}{\npat}$
+          \item $\sseq{
+            \patExpands
+              {\uGG{\uG_i}{\pctx_i}}
+              {\uPhi}
+              {\parseUPatF{\bsubseq{b}{m_i}{n_i}}}
+              {p_i}
+              {\tau_i}
+          }{\npat}$  and $\sseq{\patType{\pctx_i}{p_i}{\tau_i}}{\npat}$
+          \end{enumerate}
+      \item (\textbf{Visibility}) $\uG = \biguplus_{0 \leq i < \npat} \uG_i$ and $\Gamma = \bigcup_{0 \leq i < \npat} \pctx_i$
+\end{enumerate}
+\end{theorem}
+
+% \newcommand{\pTLMsFormallySec}{Parametric TLMs, Formally}
+% \section{\protect\pTLMsFormallySec}
+% \label{sec:ptlms-formally}
+
+% % \begin{figure}[p]
+% % \[\begin{array}{llcl}
+% % \mathsf{UMType} & \urho & ::= 
+% % %& \autype{\utau} 
+% % & \utau & \text{type annotation}\\
+% % % &&
+% % %& \aualltypes{\ut}{\urho} 
+% % % & \alltypes{\ut}{\urho} & \text{type parameterization}\\
 % % &&
-% %& \aualltypes{\ut}{\urho} 
-% % & \alltypes{\ut}{\urho} & \text{type parameterization}\\
-% &&
-% %& \auallmods{\usigma}{\uX}{\urho} 
-% & \allmods{\uX}{\usigma}{\urho} & \text{module parameterization}\\
-% \mathsf{UMExp} & \uepsilon & ::= 
-% %& \abindref{\tsmv} 
-% & \tsmv & \text{TLM identifier reference}\\
+% % %& \auallmods{\usigma}{\uX}{\urho} 
+% % & \allmods{\uX}{\usigma}{\urho} & \text{module parameterization}\\
+% % \mathsf{UMExp} & \uepsilon & ::= 
+% % %& \abindref{\tsmv} 
+% % & \tsmv & \text{TLM identifier reference}\\
+% % % &&
+% % %& \auabstype{\ut}{\uepsilon} 
+% % % & \abstype{\ut}{\uepsilon} & \text{type abstraction}\\
 % % &&
-% %& \auabstype{\ut}{\uepsilon} 
-% % & \abstype{\ut}{\uepsilon} & \text{type abstraction}\\
-% &&
-% %& \auabsmod{\usigma}{\uX}{\uepsilon} 
-% & \absmod{\uX}{\usigma}{\uepsilon} & \text{module abstraction}\\
+% % %& \auabsmod{\usigma}{\uX}{\uepsilon} 
+% % & \absmod{\uX}{\usigma}{\uepsilon} & \text{module abstraction}\\
+% % % &&
+% % %& \auaptype{\utau}{\uepsilon} 
+% % % & \aptype{\uepsilon}{\utau} & \text{type application}\\
 % % &&
-% %& \auaptype{\utau}{\uepsilon} 
-% % & \aptype{\uepsilon}{\utau} & \text{type application}\\
-% &&
-% %& \auapmod{\uM}{\uepsilon} 
-% & \apmod{\uepsilon}{\uX} & \text{module application}\ECC
-% \end{array}
-% \]
-% \caption{Syntax of unexpanded TLM types and expressions.}
-% \label{fig:P-macro-expressions-types-u}
-% \end{figure}
-
-We will now outline $\miniVerseParam$, a calculus that extends $\miniVersePat$ with parametric TLMs.
+% % %& \auapmod{\uM}{\uepsilon} 
+% % & \apmod{\uepsilon}{\uX} & \text{module application}\ECC
+% % \end{array}
+% % \]
+% % \caption{Syntax of unexpanded TLM types and expressions.}
+% % \label{fig:P-macro-expressions-types-u}
+% % \end{figure}
+
+% We will now outline $\miniVerseParam$, a calculus that extends $\miniVersePat$ with parametric TLMs.
 This calculus is organized, like $\miniVersePat$, as an unexpanded language (UL) defined by typed expansion to an expanded language (XL).
 There is not enough space to describe $\miniVerseParam$ with the same level of detail as in Sec.
-\ref{sec:setsms-formally}, so we highlight only the most important concepts below.
+\ref{sec:setlms-formally}, so we highlight only the most important concepts below.
 The details are in the supplement.
 
-The XL consists of 1) module expressions, $M$, classified by signatures, $\sigma$; 2) constructions, $c$, classified by kinds, $\kappa$; and 3) expressions classified by types, which are constructions of kind $\akty$ (we use metavariables $\tau$ instead of $c$ for types by convention.) Metavariables $X$ ranges over module variables and $u$ or $t$ over construction variables.
+% The XL consists of 1) module expressions, $M$, classified by signatures, $\sigma$; 2) constructions, $c$, classified by kinds, $\kappa$; and 3) expressions classified by types, which are constructions of kind $\akty$ (we use metavariables $\tau$ instead of $c$ for types by convention.) Metavariables $X$ ranges over module variables and $u$ or $t$ over construction variables.
 The module and construction languages are based closely on those defined by \citet{pfple1}, which in turn are based on early work by \citet{MacQueen:1984:MSM:800055.802036,DBLP:conf/popl/MacQueen86}, subsequent work on the phase splitting interpretation of modules \cite{harper1989higher} and on using dependent singleton kinds to track type identity \cite{stone2006extensional,DBLP:conf/lfmtp/Crary09}, and finally on formal developments by \citet{dreyer2005understanding} and \citet{conf/popl/LeeCH07}.
@@
 
-The main conceptual difference between $\miniVersePat$ and $\miniVerseParam$ is that $\miniVerseParam$ introduces the notion of unexpanded and expanded TLM expressions and types, as shown in Fig.
-\ref{fig:P-macro-expressions-types}.
-
-\begin{figure}[h]
-\small
-\begin{minipage}{0.38\textwidth}
-$\arraycolsep=3pt\begin{array}{llcl}
-\mathsf{UMType} & \urho & ::= & \utau ~\vert~ \allmods{\uX}{\usigma}{\urho}\\
-\mathsf{UMExp} & \uepsilon & ::= & \tsmv ~\vert~ \absmod{\uX}{\usigma}{\uepsilon}~\vert~ \apmod{\uepsilon}{\uX}
-\end{array}$
-\end{minipage}
-\begin{minipage}{0.6\textwidth}
-$\arraycolsep=3pt\begin{array}{llcl}
-\mathsf{MType} & \rho & ::= & \aetype{\tau} ~\vert~ \aeallmods{\sigma}{X}{\rho} \\
-\mathsf{MExp} & \epsilon & ::= & \adefref{a} ~\vert~ \aeabsmod{\sigma}{X}{\epsilon} ~\vert~\aeapmod{X}{\epsilon} 
-\end{array}$
-\end{minipage}
-\caption{Syntax of unexpanded and expanded TLM types and expressions in $\miniVerseParam$}
-\label{fig:P-macro-expressions-types}
-\end{figure}
-The TLM type $\aeallmods{\sigma}{X}{\rho}$ classifies TLM expressions that have one module parameter matching $\sigma$.
+% The main conceptual difference between $\miniVersePat$ and $\miniVerseParam$ is that $\miniVerseParam$ introduces the notion of unexpanded and expanded TLM expressions and types, as shown in Fig.~\ref{fig:P-macro-expressions-types}.
+
+% \begin{figure}[h]
+% \small
+% \begin{minipage}{0.38\textwidth}
+% $\arraycolsep=3pt\begin{array}{llcl}
+% \mathsf{UMType} & \urho & ::= & \utau ~\vert~ \allmods{\uX}{\usigma}{\urho}\\
+% \mathsf{UMExp} & \uepsilon & ::= & \tsmv ~\vert~ \absmod{\uX}{\usigma}{\uepsilon}~\vert~ \apmod{\uepsilon}{\uX}
+% \end{array}$
+% \end{minipage}
+% \begin{minipage}{0.6\textwidth}
+% $\arraycolsep=3pt\begin{array}{llcl}
+% \mathsf{MType} & \rho & ::= & \aetype{\tau} ~\vert~ \aeallmods{\sigma}{X}{\rho} \\
+% \mathsf{MExp} & \epsilon & ::= & \adefref{a} ~\vert~ \aeabsmod{\sigma}{X}{\epsilon} ~\vert~\aeapmod{X}{\epsilon} 
+% \end{array}$
+% \end{minipage}
+% \caption{Syntax of unexpanded and expanded TLM types and expressions in $\miniVerseParam$}
+% \label{fig:P-macro-expressions-types}
+% \end{figure}
+% The TLM type $\aeallmods{\sigma}{X}{\rho}$ classifies TLM expressions that have one module parameter matching $\sigma$.
 For simplicity, we formalize only module parameters.
@@
 
-The rule governing expression TLM application, reproduced below, touches all of the main ideas in $\miniVerseParam$, so we will refer to it throughout the remainder of this section.%It might be useful to compare this rule to Rule \textsc{ee-ap-sptsm}.
-{\small\begin{mathpar}
-\inferrule[ee-ap-petsm]{
-  \uOmega = \uOmegaEx{\uD}{\uG}{\uMctx}{\Omega_\text{app}}\\
-  \uPsi=\uAS{\uA}{\Psi}\\\\
-  \tsmexpExpandsExp{\uOmega}{\uPsi}{\uepsilon}{\epsilon}{\aetype{\tau_\text{final}}}\\
-  \tsmexpEvalsExp{\Omega_\text{app}}{\Psi}{\epsilon}{\epsilon_\text{normal}}\\\\
-  \tsmdefof{\epsilon_\text{normal}}=a\\
-  \Psi = \Psi', \petsmdefn{a}{\rho}{\eparse}\\\\
-  \encodeBody{b}{\ebody}\\
-  \evalU{\ap{\eparse}{\ebody}}{\aein{\mathtt{SuccessE}}{e_\text{pproto}}}\\
-  \decodePCEExp{e_\text{pproto}}{\pce}\\\\
-  \prepce{\Omega_\text{app}}{\Psi}{\pce}{\ce}{\epsilon_\text{normal}}{\aetype{\tau_\text{proto}}}{\omega}{\Omega_\text{params}}\\\\
-  \segOK{\segof{\ce}}{b}\\
-  \cvalidEP{\Omega_\text{params}}{\esceneP{\omega : \OParams}{\uOmega}{\uPsi}{\uPhi}{b}}{\ce}{e}{\tau_\text{proto}}
-}{
-  \expandsP{\uOmega}{\uPsi}{\uPhi}{\utsmap{\uepsilon}{b}}{[\omega]e}{[\omega]\tau_\text{proto}}
-}
-\end{mathpar}}
-
-The first two premises simply deconstruct the (unified) unexpanded context $\uOmega$ (which tracks the expansion of expression, constructor and module identifiers, as $\uDelta$ and $\uGamma$ did in $\miniVersePat$) and peTLM context, $\uPsi$.
+% The rule governing expression TLM application, reproduced below, touches all of the main ideas in $\miniVerseParam$, so we will refer to it throughout the remainder of this section.%It might be useful to compare this rule to Rule \textsc{ee-ap-sptsm}.
+% {\small\begin{mathpar}
+% \inferrule[ee-ap-petsm]{
+%   \uOmega = \uOmegaEx{\uD}{\uG}{\uMctx}{\Omega_\text{app}}\\
+%   \uPsi=\uAS{\uA}{\Psi}\\\\
+%   \tsmexpExpandsExp{\uOmega}{\uPsi}{\uepsilon}{\epsilon}{\aetype{\tau_\text{final}}}\\
+%   \tsmexpEvalsExp{\Omega_\text{app}}{\Psi}{\epsilon}{\epsilon_\text{normal}}\\\\
+%   \tsmdefof{\epsilon_\text{normal}}=a\\
+%   \Psi = \Psi', \petsmdefn{a}{\rho}{\eparse}\\\\
+%   \encodeBody{b}{\ebody}\\
+%   \evalU{\ap{\eparse}{\ebody}}{\aein{\mathtt{SuccessE}}{e_\text{pproto}}}\\
+%   \decodePCEExp{e_\text{pproto}}{\pce}\\\\
+%   \prepce{\Omega_\text{app}}{\Psi}{\pce}{\ce}{\epsilon_\text{normal}}{\aetype{\tau_\text{proto}}}{\omega}{\Omega_\text{params}}\\\\
+%   \segOK{\segof{\ce}}{b}\\
+%   \cvalidEP{\Omega_\text{params}}{\esceneP{\omega : \OParams}{\uOmega}{\uPsi}{\uPhi}{b}}{\ce}{e}{\tau_\text{proto}}
+% }{
+%   \expandsP{\uOmega}{\uPsi}{\uPhi}{\utsmap{\uepsilon}{b}}{[\omega]e}{[\omega]\tau_\text{proto}}
+% }
+% \end{mathpar}}
+
+% The first two premises simply deconstruct the (unified) unexpanded context $\uOmega$ (which tracks the expansion of expression, constructor and module identifiers, as $\uDelta$ and $\uGamma$ did in $\miniVersePat$) and peTLM context, $\uPsi$.
 Next, we expand $\uepsilon$ according to straightforward unexpanded peTLM expression expansion rules.
@@
 
-The fourth premise performs \emph{peTLM expression normalization}, $\tsmexpEvalsExp{\Omega_\text{app}}{\Psi}{\epsilon}{\epsilon_\text{normal}}$.
+% The fourth premise performs \emph{peTLM expression normalization}, $\tsmexpEvalsExp{\Omega_\text{app}}{\Psi}{\epsilon}{\epsilon_\text{normal}}$.
 This is defined in terms of a structural operational semantics \cite{DBLP:journals/jlp/Plotkin04a} with two stepping rules:
-% \begin{equation*}\tag{\ref{rule:tsmexpEvalsExp}}
-% \inferrule{
-%   \tsmexpMultistepsExp{\Omega}{\Psi}{\epsilon}{\epsilon'}\\
-%   \tsmexpNormalExp{\Omega}{\Psi}{\epsilon'}
-% }{
-%   \tsmexpEvalsExp{\Omega}{\Psi}{\epsilon}{\epsilon'}
-% }
-% \end{equation*}
-% where the multistep judgement, $\tsmexpMultistepsExp{\Omega}{\Psi}{\epsilon}{\epsilon'}$, is defined as the reflexive, transitive closure of the stepping judgement defined by the following rules:
-% \begin{equation*}\tag{\ref{rule:tsmexpStepsExp-aptype-1}}
-% \inferrule{
+% % \begin{equation*}\tag{\ref{rule:tsmexpEvalsExp}}
+% % \inferrule{
+% %   \tsmexpMultistepsExp{\Omega}{\Psi}{\epsilon}{\epsilon'}\\
+% %   \tsmexpNormalExp{\Omega}{\Psi}{\epsilon'}
+% % }{
+% %   \tsmexpEvalsExp{\Omega}{\Psi}{\epsilon}{\epsilon'}
+% % }
+% % \end{equation*}
+% % where the multistep judgement, $\tsmexpMultistepsExp{\Omega}{\Psi}{\epsilon}{\epsilon'}$, is defined as the reflexive, transitive closure of the stepping judgement defined by the following rules:
+% % \begin{equation*}\tag{\ref{rule:tsmexpStepsExp-aptype-1}}
+% % \inferrule{
+% %   \tsmexpStepsExp{\Omega}{\Psi}{\epsilon}{\epsilon'}
+% % }{
+% %   \tsmexpStepsExp{\Omega}{\Psi}{\aeaptype{\tau}{\epsilon}}{\aeaptype{\tau}{\epsilon'}}
+% % }
+% % \end{equation*}
+% % \begin{equation*}\tag{\ref{rule:tsmexpStepsExp-aptype-2}}
+% % \inferrule{ }{
+% %   \tsmexpStepsExp{\Omega}{\Psi}{\aeaptype{\tau}{\aeabstype{t}{\epsilon}}}{[\tau/t]\epsilon}
+% % }
+% % \end{equation*}
+% {\small\begin{mathpar}
+% \inferrule[eps-dyn-apmod-subst-e]{ }{
+%   \tsmexpStepsExp{\Omega}{\Psi}{\aeapmod{X}{\aeabsmod{\sigma}{X'}{\epsilon}}}{[X/X']\epsilon}
+% }
+
+% \inferrule[eps-dyn-apmod-steps-e]{
 %   \tsmexpStepsExp{\Omega}{\Psi}{\epsilon}{\epsilon'}
 % }{
-%   \tsmexpStepsExp{\Omega}{\Psi}{\aeaptype{\tau}{\epsilon}}{\aeaptype{\tau}{\epsilon'}}
-% }
-% \end{equation*}
-% \begin{equation*}\tag{\ref{rule:tsmexpStepsExp-aptype-2}}
-% \inferrule{ }{
-%   \tsmexpStepsExp{\Omega}{\Psi}{\aeaptype{\tau}{\aeabstype{t}{\epsilon}}}{[\tau/t]\epsilon}
-% }
-% \end{equation*}
-{\small\begin{mathpar}
-\inferrule[eps-dyn-apmod-subst-e]{ }{
-  \tsmexpStepsExp{\Omega}{\Psi}{\aeapmod{X}{\aeabsmod{\sigma}{X'}{\epsilon}}}{[X/X']\epsilon}
-}
-
-\inferrule[eps-dyn-apmod-steps-e]{
-  \tsmexpStepsExp{\Omega}{\Psi}{\epsilon}{\epsilon'}
-}{
-  \tsmexpStepsExp{\Omega}{\Psi}{\aeapmod{X}{\epsilon}}{\aeapmod{X}{\epsilon'}}
-}
-\end{mathpar}}
-% The peTLM expression normal forms are defined as follows:
-% \begin{equation*}\tag{\ref{rule:tsmexpNormalExp-defref}}
-% \inferrule{ }{
-%   \tsmexpNormalExp{\Omega}{\Psi, \petsmdefn{a}{\rho}{\eparse}}{\adefref{a}}
-% }
-% \end{equation*}
-% % \begin{equation*}\tag{\ref{rule:tsmexpNormalExp-abstype}}
+%   \tsmexpStepsExp{\Omega}{\Psi}{\aeapmod{X}{\epsilon}}{\aeapmod{X}{\epsilon'}}
+% }
+% \end{mathpar}}
+% % The peTLM expression normal forms are defined as follows:
+% % \begin{equation*}\tag{\ref{rule:tsmexpNormalExp-defref}}
 % % \inferrule{ }{
-% %   \tsmexpNormalExp{\Omega}{\Psi}{\aeabstype{t}{\epsilon}}
+% %   \tsmexpNormalExp{\Omega}{\Psi, \petsmdefn{a}{\rho}{\eparse}}{\adefref{a}}
 % % }
 % % \end{equation*}
-% \begin{equation*}\tag{\ref{rule:tsmexpNormalExp-absmod}}
-% \inferrule{ }{
-%   \tsmexpNormalExp{\Omega}{\Psi}{\aeabsmod{\sigma}{X}{\epsilon}}
-% }
-% \end{equation*}
-% % \begin{equation*}\tag{\ref{rule:tsmexpNormalExp-aptype}}
+% % % \begin{equation*}\tag{\ref{rule:tsmexpNormalExp-abstype}}
+% % % \inferrule{ }{
+% % %   \tsmexpNormalExp{\Omega}{\Psi}{\aeabstype{t}{\epsilon}}
+% % % }
+% % % \end{equation*}
+% % \begin{equation*}\tag{\ref{rule:tsmexpNormalExp-absmod}}
+% % \inferrule{ }{
+% %   \tsmexpNormalExp{\Omega}{\Psi}{\aeabsmod{\sigma}{X}{\epsilon}}
+% % }
+% % \end{equation*}
+% % % \begin{equation*}\tag{\ref{rule:tsmexpNormalExp-aptype}}
+% % % \inferrule{
+% % %   \epsilon \neq \aeabstype{t}{\epsilon'}\\
+% % %   \tsmexpNormalExp{\Omega}{\Psi}{\epsilon}
+% % % }{
+% % %   \tsmexpNormalExp{\Omega}{\Psi}{\aeaptype{\tau}{\epsilon}}
+% % % }
+% % % \end{equation*}
+% % \begin{equation*}\tag{\ref{rule:tsmexpNormalExp-apmod}}
 % % \inferrule{
-% %   \epsilon \neq \aeabstype{t}{\epsilon'}\\
+% %   \epsilon \neq \aeabsmod{\sigma}{X'}{\epsilon'}\\
 % %   \tsmexpNormalExp{\Omega}{\Psi}{\epsilon}
 % % }{
-% %   \tsmexpNormalExp{\Omega}{\Psi}{\aeaptype{\tau}{\epsilon}}
+% %   \tsmexpNormalExp{\Omega}{\Psi}{\aeapmod{X}{\epsilon}}
 % % }
 % % \end{equation*}
-% \begin{equation*}\tag{\ref{rule:tsmexpNormalExp-apmod}}
-% \inferrule{
-%   \epsilon \neq \aeabsmod{\sigma}{X'}{\epsilon'}\\
-%   \tsmexpNormalExp{\Omega}{\Psi}{\epsilon}
-% }{
-%   \tsmexpNormalExp{\Omega}{\Psi}{\aeapmod{X}{\epsilon}}
-% }
-% \end{equation*}
-\vspace{-5px}
-
-Normalization eliminates parameters introduced in higher-order abbreviations, leaving only those parameter applications specified by the original TLM definition.
+% \vspace{-5px}
+
+% Normalization eliminates parameters introduced in higher-order abbreviations, leaving only those parameter applications specified by the original TLM definition.
 Normal forms and progress and preservation theorems are established in the supplement.
 
-The third row of premises looks up the applied TLM's definition by invoking a simple metafunction to extract its name, $a$, then looking up $a$ within the peTLM definition context, $\Psi$.
-% \begin{align}
-% \tsmdefof{\adefref{a}} & = a \tag{\ref{eqn:tsmdefof-adefref}}\\
-% % \tsmdefof{\aeabstype{t}{\epsilon}} & = \tsmdefof{\epsilon} \tag{\ref{eqn:tsmdefof-abstype}}\\
-% \tsmdefof{\aeabsmod{\sigma}{X}{\epsilon}} & = \tsmdefof{\epsilon} \tag{\ref{eqn:tsmdefof-absmod}}\\
-% % \tsmdefof{\aeaptype{\tau}{\epsilon}} & = \tsmdefof{\epsilon} \tag{\ref{eqn:tsmdefof-aptype}}\\
-% \tsmdefof{\aeapmod{X}{\epsilon}} & = \tsmdefof{\epsilon} \tag{\ref{eqn:tsmdefof-apmod}}
-% \end{align}
-
-The fourth row of premises 1) encodes the body as a value of the type $\tBody$; 2) applies the parse function; and 3) decodes the result, producing a \emph{parameterized proto-expression}, $\pce$.
+% The third row of premises looks up the applied TLM's definition by invoking a simple metafunction to extract its name, $a$, then looking up $a$ within the peTLM definition context, $\Psi$.
+% % \begin{align}
+% % \tsmdefof{\adefref{a}} & = a \tag{\ref{eqn:tsmdefof-adefref}}\\
+% % % \tsmdefof{\aeabstype{t}{\epsilon}} & = \tsmdefof{\epsilon} \tag{\ref{eqn:tsmdefof-abstype}}\\
+% % \tsmdefof{\aeabsmod{\sigma}{X}{\epsilon}} & = \tsmdefof{\epsilon} \tag{\ref{eqn:tsmdefof-absmod}}\\
+% % % \tsmdefof{\aeaptype{\tau}{\epsilon}} & = \tsmdefof{\epsilon} \tag{\ref{eqn:tsmdefof-aptype}}\\
+% % \tsmdefof{\aeapmod{X}{\epsilon}} & = \tsmdefof{\epsilon} \tag{\ref{eqn:tsmdefof-apmod}}
+% % \end{align}
+
+% The fourth row of premises 1) encodes the body as a value of the type $\tBody$; 2) applies the parse function; and 3) decodes the result, producing a \emph{parameterized proto-expression}, $\pce$.
 Parameterized proto-expressions, $\pce$, are ABTs that serve simply to introduce the parameter bindings into an underlying proto-expression, $\ce$.
@@
 
-\vspace{-4px}{\small\[\begin{array}{llcl}
-% \textbf{Sort} & & & \textbf{Operational Form} & \textbf{Stylized Form} & \textbf{Description}\\
-% \LCC \color{Yellow}&\color{Yellow}&\color{Yellow}& \color{Yellow} & \color{Yellow} & \color{Yellow}\\
-\mathsf{PPrExp} & \pce & ::= & \apceexp{\ce} ~\vert~ \apcebindmod{X}{\pce}
-\end{array}\]}%
-\vspace{-6px}%
-
-There must be one binder in $\pce$ for each TLM parameter specified by $a$. (In Reason, we can insert these binders automatically as a convenience.)
-
-The judgement on the fifth row of Rule \textsc{ee-ap-petsm} then \emph{deparameterizes} $\pce$ by peeling away these binders to produce 1) the underlying proto-expression, $\ce$, with the variables that stand for the parameters free; 2) a corresponding deparameterized type, $\tau_\text{proto}$, that uses the same free variables to stand for the parameters; 3) a \emph{substitution}, $\omega$, that pairs the applied parameters from $\epsilon_\text{normal}$ with the corresponding variables generated when peeling away the binders in $\pce$; and 4) a corresponding \emph{parameter context}, $\Omega_\text{params}$, that tracks the signatures of these variables.
+% \vspace{-4px}{\small\[\begin{array}{llcl}
+% % \textbf{Sort} & & & \textbf{Operational Form} & \textbf{Stylized Form} & \textbf{Description}\\
+% % \LCC \color{Yellow}&\color{Yellow}&\color{Yellow}& \color{Yellow} & \color{Yellow} & \color{Yellow}\\
+% \mathsf{PPrExp} & \pce & ::= & \apceexp{\ce} ~\vert~ \apcebindmod{X}{\pce}
+% \end{array}\]}%
+% \vspace{-6px}%
+
+% There must be one binder in $\pce$ for each TLM parameter specified by $a$. (In Reason, we can insert these binders automatically as a convenience.)
+
+% The judgement on the fifth row of Rule \textsc{ee-ap-petsm} then \emph{deparameterizes} $\pce$ by peeling away these binders to produce 1) the underlying proto-expression, $\ce$, with the variables that stand for the parameters free; 2) a corresponding deparameterized type, $\tau_\text{proto}$, that uses the same free variables to stand for the parameters; 3) a \emph{substitution}, $\omega$, that pairs the applied parameters from $\epsilon_\text{normal}$ with the corresponding variables generated when peeling away the binders in $\pce$; and 4) a corresponding \emph{parameter context}, $\Omega_\text{params}$, that tracks the signatures of these variables.
 The two rules governing the proto-expression deparameterization judgement are below:
-{\small\begin{mathpar}
-\inferrule{ }{
-  \prepce{\Omega_\text{app}}{\Psi, \petsmdefn{a}{\rho}{\eparse}}{\apceexp{\ce}}{\ce}{\adefref{a}}{\rho}{\emptyset}{\emptyset}
-}
-
-\vspace{-4px}\inferrule{
-  \prepce{\Omega_\text{app}}{\Psi}{\pce}{\ce}{\epsilon}{\aeallmods{\sigma}{X}{\rho}}{\omega}{\Omega}\\
-  X \notin \domof{\Omega_\text{app}}
-}{
-  \prepce{\Omega_\text{app}}{\Psi}{\apcebindmod{X}{\pce}}{\ce}{\aeapmod{X'}{\epsilon}}{\rho}{(\omega, X'/X)}{(\Omega, X : \sigma)}
-}
-\end{mathpar}}%
-This judgement can be pronounced ``when applying peTLM $\epsilon$, $\pce$ has deparameterization $\ce$ leaving $\rho$ with parameter substitution $\omega$''. 
-Notice based on the second rule that every module binding in $\pce$ must pair with a corresponding module parameter application.
+% {\small\begin{mathpar}
+% \inferrule{ }{
+%   \prepce{\Omega_\text{app}}{\Psi, \petsmdefn{a}{\rho}{\eparse}}{\apceexp{\ce}}{\ce}{\adefref{a}}{\rho}{\emptyset}{\emptyset}
+% }
+
+% \vspace{-4px}\inferrule{
+%   \prepce{\Omega_\text{app}}{\Psi}{\pce}{\ce}{\epsilon}{\aeallmods{\sigma}{X}{\rho}}{\omega}{\Omega}\\
+%   X \notin \domof{\Omega_\text{app}}
+% }{
+%   \prepce{\Omega_\text{app}}{\Psi}{\apcebindmod{X}{\pce}}{\ce}{\aeapmod{X'}{\epsilon}}{\rho}{(\omega, X'/X)}{(\Omega, X : \sigma)}
+% }
+% \end{mathpar}}%
+% This judgement can be pronounced ``when applying peTLM $\epsilon$, $\pce$ has deparameterization $\ce$ leaving $\rho$ with parameter substitution $\omega$''. 
+% Notice based on the second rule that every module binding in $\pce$ must pair with a corresponding module parameter application.
 Moreover, the variables standing for parameters must not appear in $\Omega_\text{app}$, i.e.
@@
 
-The final row of premises checks that the segmentation of $\ce$ is valid and  performs proto-expansion validation under the parameter context, $\Omega_\text{param}$ (rather than the empty context, as was the case in $\miniVersePat$.) The conclusion of the rule applies the parameter substitution, $\omega$, to the resulting expression and the deparameterized type.
-
-Proto-expansion validation operates conceptually as in $\miniVersePat$.
+% The final row of premises checks that the segmentation of $\ce$ is valid and  performs proto-expansion validation under the parameter context, $\Omega_\text{param}$ (rather than the empty context, as was the case in $\miniVersePat$.) The conclusion of the rule applies the parameter substitution, $\omega$, to the resulting expression and the deparameterized type.
+
+% Proto-expansion validation operates conceptually as in $\miniVersePat$.
 The only subtlety has to do with the type annotations on references to spliced terms.
@@
 However, the spliced term itself must treat parameters parametrically, so the substitution is not applied in the conclusion of the following rule:
-\begin{mathpar}\label{rule:cvalidE-P-splicede}
-\inferrule{
-  \parseUExp{\bsubseq{b}{m}{n}}{\ue}\\
-    \cvalidC{\OParams}{\csceneP{\omega : \OParams}{\uOmega}{b}}{\ctau}{\tau}{\akty}\\
-  \expandsP{\uOmega}{\uPsi}{\uPhi}{\ue}{e}{[\omega]\tau}\\\\
-  \uOmega=\uOmegaEx{\uD}{\uG}{\uMctx}{\Omega_\text{app}}\\
-  \domof{\Omega} \cap \domof{\Omega_\text{app}} = \emptyset
-}{
-  \cvalidEP{\Omega}{\esceneP{\omega : \OParams}{\uOmega}{\uPsi}{\uPhi}{b}}{\acesplicede{m}{n}{\ctau}}{e}{\tau}
-}
-\end{mathpar}
-(This is only sensible because we maintain the invariant that $\Omega$ is always an extension of $\Omega_\text{params}$.)
-
-The calculus enjoys metatheoretic properties analagous to those described in Sec.
+% \begin{mathpar}\label{rule:cvalidE-P-splicede}
+% \inferrule{
+%   \parseUExp{\bsubseq{b}{m}{n}}{\ue}\\
+%     \cvalidC{\OParams}{\csceneP{\omega : \OParams}{\uOmega}{b}}{\ctau}{\tau}{\akty}\\
+%   \expandsP{\uOmega}{\uPsi}{\uPhi}{\ue}{e}{[\omega]\tau}\\\\
+%   \uOmega=\uOmegaEx{\uD}{\uG}{\uMctx}{\Omega_\text{app}}\\
+%   \domof{\Omega} \cap \domof{\Omega_\text{app}} = \emptyset
+% }{
+%   \cvalidEP{\Omega}{\esceneP{\omega : \OParams}{\uOmega}{\uPsi}{\uPhi}{b}}{\acesplicede{m}{n}{\ctau}}{e}{\tau}
+% }
+% \end{mathpar}
+% (This is only sensible because we maintain the invariant that $\Omega$ is always an extension of $\Omega_\text{params}$.)
+
+% The calculus enjoys metatheoretic properties analagous to those described in Sec.
 \ref{sec:s-metatheory}, modified to account for the presence of modules, kinds and parameterization.
@@
 The details of parametric pattern TLMs (ppTLMs) are analagous (see supplement.)
-\vspace{-3px}
-\begin{theorem}[peTLM Reasoning Principles]
-If $\expandsP{\uOmega}{\uPsi}{\uPhi}{\utsmap{\uepsilon}{b}}{e}{\tau}$ then:
-\begin{enumerate}[nolistsep,noitemsep]
-  \item $\uOmega=\uOmegaEx{\uD}{\uG}{\uMctx}{\Omega_\text{app}}$
-  \item $\uPsi=\uAS{\uA}{\Psi}$
-  \item (\textbf{Typing 1}) $\tsmexpExpandsExp{\uOmega}{\uPsi}{\uepsilon}{\epsilon}{\aetype{\tau}}$ and $\hastypeP{\Omega_\text{app}}{e}{\tau}$
-  \item $\tsmexpEvalsExp{\Omega_\text{app}}{\Psi}{\epsilon}{\epsilon_\text{normal}}$
-  \item $\tsmdefof{\epsilon_\text{normal}}=a$
-  \item $\Psi = \Psi', \petsmdefn{a}{\rho}{\eparse}$
-  \item $\encodeBody{b}{\ebody}$
-    \item $\evalU{\ap{\eparse}{\ebody}}{\aein{\mathtt{SuccessE}}{e_\text{pproto}}}$
-  \item $\decodePCEExp{e_\text{pproto}}{\pce}$
-  \item $\prepce{\Omega_\text{app}}{\Psi}{\pce}{\ce}{\epsilon_\text{normal}}{\aetype{\tau_\text{proto}}}{\omega}{\Omega_\text{params}}$
-  \item (\textbf{Segmentation}) $\segOK{\segof{\ce}}{b}$
-  \item $\cvalidEP{\Omega_\text{params}}{\esceneP{\omega : \OParams}{\uOmega}{\uPsi}{\uPhi}{b}}{\ce}{e'}{\tau_\text{proto}}$
-  \item $e = [\omega]e'$
-  \item $\tau = [\omega]\tau_\text{proto}$
-  \item $
-    \segof{\ce} = \sseq{\acesplicedk{m_i}{n_i}}{\nkind} \cup \sseq{\acesplicedc{m'_i}{n'_i}{\cekappa'_i}}{\ncon} \cup $ \\
-     ~~~~$          \sseq{\acesplicede{m''_i}{n''_i}{\ctau_i}}{\nexp}
-    $
-  \item (\textbf{Kinding 1}) $\sseq{\kExpands{\uOmega}{\parseUKindF{\bsubseq{b}{m_i}{n_i}}}{\kappa_i}}{\nkind}$ and \\ ~~~~$\sseq{\iskind{\Omega_\text{app}}{\kappa_i}}{\nkind}$
-  \item (\textbf{Kinding 2}) $\sseq{\cvalidK{\OParams}{\csceneP{\omega : \OParams}{\uOmega}{b}}{\cekappa'_i}{\kappa'_i}}{\ncon}$ and $\sseq{\iskind{\Omega_\text{app}}{[\omega]\kappa'_i}}{\ncon}$
-  \item (\textbf{Kinding 3}) $\sseq{\cExpands{\uOmega}{\parseUConF{\bsubseq{b}{m'_i}{n'_i}}}{c_i}{[\omega]\kappa'_i}}{\ncon}$ and $\sseq{\haskind{\Omega_\text{app}}{c_i}{[\omega]\kappa'_i}}{\ncon}$
-  \item (\textbf{Kinding 4}) $\sseq{\cvalidC{\OParams}{\csceneP{\omega : \OParams}{\uOmega}{b}}{\ctau_i}{\tau_i}{\akty}}{\nexp}$ and $\sseq{\haskind{\Omega_\text{app}}{[\omega]\tau_i}{\akty}}{\nexp}$
-  \item (\textbf{Typing 2}) $\sseq{\expandsP{\uOmega}{\uPsi}{\uPhi}{\parseUExpF{\bsubseq{b}{m''_i}{n''_i}}}{e_i}{[\omega]\tau_i}}{\nexp}$ and $\sseq{\hastypeP{\Omega_\text{app}}{e_i}{[\omega]\tau_i}}{\nexp}$
-  \item (\textbf{Capture Avoidance}) $e = [\sseq{\kappa_i/k_i}{\nkind}, \sseq{c_i/u_i}{\ncon}, \sseq{e_i/x_i}{\nexp}, \omega]e''$ for some $e''$ and fresh $\sseq{k_i}{\nkind}$ and fresh $\sseq{u_i}{\ncon}$ and fresh $\sseq{x_i}{\nexp}$
-  \item (\textbf{Context Independence}) $\mathsf{fv}(e'') \subset \sseq{k_i}{\nkind} \cup \sseq{u_i}{\ncon} \cup \sseq{x_i}{\nexp} \cup \domof{\OParams}$
-  % $\hastypeP{\sseq{\Khyp{k_i}}{\nkind} \cup \sseq{u_i :: [\omega]\kappa'_i}{\ncon} \cup \sseq{x_i : [\omega]\tau_i}{\nexp}}{[\omega]e''}{\tau}$\todo{maybe restate this in terms of free variables of e'' here and elsewhere, because context isn't technically well-formed here?}
-\end{enumerate}
-\end{theorem}
-
-\vspace{-6px}
-
-% The following theorem summarizes the abstract reasoning principles available to programmers when applying a pattern TLM.
-Most of the labeled clauses are analagous to those described above, so we omit their descriptions.
-% % \begingroup
-% % \def\thetheorem{\ref{thm:spTLM-Typing-Segmentation}}
-% \begin{theorem}[spTLM Reasoning Principles]
-% % \label{thm:spTLM-Typing-Segmentation}
-% If $\patExpands{\upctx}{\uPhi}{\utsmap{\tsmv}{b}}{p}{\tau}$ where $\uDelta=\uDD{\uD}{\Delta}$ and $\uGamma=\uGG{\uG}{\Gamma}$ then all of the following hold:
-% \begin{enumerate}[noitemsep,nolistsep]
-%         \item (\textbf{Typing 1}) $\uPhi=\uPhi', \uPhyp{\tsmv}{a}{\tau}{\eparse}$ and $\patType{\pctx}{p}{\tau}$
+% \vspace{-3px}
+% \begin{theorem}[peTLM Reasoning Principles]
+% If $\expandsP{\uOmega}{\uPsi}{\uPhi}{\utsmap{\uepsilon}{b}}{e}{\tau}$ then:
+% \begin{enumerate}[nolistsep,noitemsep]
+%   \item $\uOmega=\uOmegaEx{\uD}{\uG}{\uMctx}{\Omega_\text{app}}$
+%   \item $\uPsi=\uAS{\uA}{\Psi}$
+%   \item (\textbf{Typing 1}) $\tsmexpExpandsExp{\uOmega}{\uPsi}{\uepsilon}{\epsilon}{\aetype{\tau}}$ and $\hastypeP{\Omega_\text{app}}{e}{\tau}$
+%   \item $\tsmexpEvalsExp{\Omega_\text{app}}{\Psi}{\epsilon}{\epsilon_\text{normal}}$
+%   \item $\tsmdefof{\epsilon_\text{normal}}=a$
+%   \item $\Psi = \Psi', \petsmdefn{a}{\rho}{\eparse}$
 %   \item $\encodeBody{b}{\ebody}$
-%         \item $\evalU{\eparse(\ebody)}{\aein{\mathtt{SuccessP}}{\ecand}}$
-%         \item $\decodeCEPat{\ecand}{\cpv}$
-%         \item (\textbf{Segmentation}) $\segOK{\segof{\cpv}}{b}$
-%         \item $\segof{\cpv} = \sseq{\acesplicedt{n'_i}{m'_i}}{\nty} \cup \sseq{\acesplicedp{m_i}{n_i}{\ctau_i}}{\npat}$
-%         \item (\textbf{Typing 2}) $\sseq{
-%               \expandsTU{\uDelta}
-%               {
-%                 \parseUTypF{\bsubseq{b}{m'_i}{n'_i}}
-%               }{\tau'_i}
-%             }{\nty}$ and $\sseq{\istypeU{\Delta}{\tau'_i}}{\nty}$
-%         \item (\textbf{Typing 3}) $\sseq{
-%           \cvalidT{\emptyset}{
-%             \tsceneUP
-%               {\uDelta}{b}
-%           }{
-%             \ctau_i
-%           }{\tau_i}
-%         }{\npat}$ and $\sseq{\istypeU{\Delta}{\tau_i}}{\npat}$
-%         \item (\textbf{Typing 4}) $\sseq{
-%           \patExpands
-%             {\uGG{\uG_i}{\pctx_i}}
-%             {\uPhi}
-%             {\parseUPatF{\bsubseq{b}{m_i}{n_i}}}
-%             {p_i}
-%             {\tau_i}
-%         }{\npat}$  and $\sseq{\patType{\pctx_i}{p_i}{\tau_i}}{\npat}$
-%       \item (\textbf{Visibility}) $\uG = \biguplus_{0 \leq i < \npat} \uG_i$ and $\Gamma = \bigcup_{0 \leq i < \npat} \pctx_i$
-%         \begin{quote}
-%           \begin{grayparbox}
-%           The hypotheses generated by the TLM application are exactly those generated by the spliced patterns.
-%           \end{grayparbox}
-%         \end{quote}
+%     \item $\evalU{\ap{\eparse}{\ebody}}{\aein{\mathtt{SuccessE}}{e_\text{pproto}}}$
+%   \item $\decodePCEExp{e_\text{pproto}}{\pce}$
+%   \item $\prepce{\Omega_\text{app}}{\Psi}{\pce}{\ce}{\epsilon_\text{normal}}{\aetype{\tau_\text{proto}}}{\omega}{\Omega_\text{params}}$
+%   \item (\textbf{Segmentation}) $\segOK{\segof{\ce}}{b}$
+%   \item $\cvalidEP{\Omega_\text{params}}{\esceneP{\omega : \OParams}{\uOmega}{\uPsi}{\uPhi}{b}}{\ce}{e'}{\tau_\text{proto}}$
+%   \item $e = [\omega]e'$
+%   \item $\tau = [\omega]\tau_\text{proto}$
+%   \item $
+%     \segof{\ce} = \sseq{\acesplicedk{m_i}{n_i}}{\nkind} \cup \sseq{\acesplicedc{m'_i}{n'_i}{\cekappa'_i}}{\ncon} \cup $ \\
+%      ~~~~$          \sseq{\acesplicede{m''_i}{n''_i}{\ctau_i}}{\nexp}
+%     $
+%   \item (\textbf{Kinding 1}) $\sseq{\kExpands{\uOmega}{\parseUKindF{\bsubseq{b}{m_i}{n_i}}}{\kappa_i}}{\nkind}$ and \\ ~~~~$\sseq{\iskind{\Omega_\text{app}}{\kappa_i}}{\nkind}$
+%   \item (\textbf{Kinding 2}) $\sseq{\cvalidK{\OParams}{\csceneP{\omega : \OParams}{\uOmega}{b}}{\cekappa'_i}{\kappa'_i}}{\ncon}$ and $\sseq{\iskind{\Omega_\text{app}}{[\omega]\kappa'_i}}{\ncon}$
+%   \item (\textbf{Kinding 3}) $\sseq{\cExpands{\uOmega}{\parseUConF{\bsubseq{b}{m'_i}{n'_i}}}{c_i}{[\omega]\kappa'_i}}{\ncon}$ and $\sseq{\haskind{\Omega_\text{app}}{c_i}{[\omega]\kappa'_i}}{\ncon}$
+%   \item (\textbf{Kinding 4}) $\sseq{\cvalidC{\OParams}{\csceneP{\omega : \OParams}{\uOmega}{b}}{\ctau_i}{\tau_i}{\akty}}{\nexp}$ and $\sseq{\haskind{\Omega_\text{app}}{[\omega]\tau_i}{\akty}}{\nexp}$
+%   \item (\textbf{Typing 2}) $\sseq{\expandsP{\uOmega}{\uPsi}{\uPhi}{\parseUExpF{\bsubseq{b}{m''_i}{n''_i}}}{e_i}{[\omega]\tau_i}}{\nexp}$ and $\sseq{\hastypeP{\Omega_\text{app}}{e_i}{[\omega]\tau_i}}{\nexp}$
+%   \item (\textbf{Capture Avoidance}) $e = [\sseq{\kappa_i/k_i}{\nkind}, \sseq{c_i/u_i}{\ncon}, \sseq{e_i/x_i}{\nexp}, \omega]e''$ for some $e''$ and fresh $\sseq{k_i}{\nkind}$ and fresh $\sseq{u_i}{\ncon}$ and fresh $\sseq{x_i}{\nexp}$
+%   \item (\textbf{Context Independence}) $\mathsf{fv}(e'') \subset \sseq{k_i}{\nkind} \cup \sseq{u_i}{\ncon} \cup \sseq{x_i}{\nexp} \cup \domof{\OParams}$
+%   % $\hastypeP{\sseq{\Khyp{k_i}}{\nkind} \cup \sseq{u_i :: [\omega]\kappa'_i}{\ncon} \cup \sseq{x_i : [\omega]\tau_i}{\nexp}}{[\omega]e''}{\tau}$\todo{maybe restate this in terms of free variables of e'' here and elsewhere, because context isn't technically well-formed here?}
 % \end{enumerate}
 % \end{theorem}
-% 
-% \begin{proof} 
-% The proof, in the supplement, relies on an auxiliary lemma about decomposing proto-patterns.
-
-% \subsection{Retrospective} Let us step back and briefly review what we have accomplished so far.
+
+% \vspace{-6px}
+
+% % The following theorem summarizes the abstract reasoning principles available to programmers when applying a pattern TLM.
+Most of the labeled clauses are analagous to those described above, so we omit their descriptions.
+% % % \begingroup
+% % % \def\thetheorem{\ref{thm:spTLM-Typing-Segmentation}}
+% % \begin{theorem}[spTLM Reasoning Principles]
+% % % \label{thm:spTLM-Typing-Segmentation}
+% % If $\patExpands{\upctx}{\uPhi}{\utsmap{\tsmv}{b}}{p}{\tau}$ where $\uDelta=\uDD{\uD}{\Delta}$ and $\uGamma=\uGG{\uG}{\Gamma}$ then all of the following hold:
+% % \begin{enumerate}[noitemsep,nolistsep]
+% %         \item (\textbf{Typing 1}) $\uPhi=\uPhi', \uPhyp{\tsmv}{a}{\tau}{\eparse}$ and $\patType{\pctx}{p}{\tau}$
+% %         \item $\encodeBody{b}{\ebody}$
+% %         \item $\evalU{\eparse(\ebody)}{\aein{\mathtt{SuccessP}}{\ecand}}$
+% %         \item $\decodeCEPat{\ecand}{\cpv}$
+% %         \item (\textbf{Segmentation}) $\segOK{\segof{\cpv}}{b}$
+% %         \item $\segof{\cpv} = \sseq{\acesplicedt{n'_i}{m'_i}}{\nty} \cup \sseq{\acesplicedp{m_i}{n_i}{\ctau_i}}{\npat}$
+% %         \item (\textbf{Typing 2}) $\sseq{
+% %               \expandsTU{\uDelta}
+% %               {
+% %                 \parseUTypF{\bsubseq{b}{m'_i}{n'_i}}
+% %               }{\tau'_i}
+% %             }{\nty}$ and $\sseq{\istypeU{\Delta}{\tau'_i}}{\nty}$
+% %         \item (\textbf{Typing 3}) $\sseq{
+% %           \cvalidT{\emptyset}{
+% %             \tsceneUP
+% %               {\uDelta}{b}
+% %           }{
+% %             \ctau_i
+% %           }{\tau_i}
+% %         }{\npat}$ and $\sseq{\istypeU{\Delta}{\tau_i}}{\npat}$
+% %         \item (\textbf{Typing 4}) $\sseq{
+% %           \patExpands
+% %             {\uGG{\uG_i}{\pctx_i}}
+% %             {\uPhi}
+% %             {\parseUPatF{\bsubseq{b}{m_i}{n_i}}}
+% %             {p_i}
+% %             {\tau_i}
+% %         }{\npat}$  and $\sseq{\patType{\pctx_i}{p_i}{\tau_i}}{\npat}$
+% %       \item (\textbf{Visibility}) $\uG = \biguplus_{0 \leq i < \npat} \uG_i$ and $\Gamma = \bigcup_{0 \leq i < \npat} \pctx_i$
+% %         \begin{quote}
+% %           \begin{grayparbox}
+% %           The hypotheses generated by the TLM application are exactly those generated by the spliced patterns.
+% %           \end{grayparbox}
+% %         \end{quote}
+% % \end{enumerate}
+% % \end{theorem}
+% % 
+% % \begin{proof} 
+% % The proof, in the supplement, relies on an auxiliary lemma about decomposing proto-patterns.
+
+% % \subsection{Retrospective} Let us step back and briefly review what we have accomplished so far.
 First, we defined an entirely standard expanded language (Sec.
@@
 \ref{sec:s-metatheory}).
-% We write $\tsfrom{\escenev}$ for the type splicing scene constructed by dropping unnecessary contexts from $\escenev$:
-% \[\tsfrom{\esceneUP{\uDelta}{\uGamma}{\uPsi}{\uPhi}{b}} = \tsceneUP{\uDelta}{b}\]
-
-% Figure \ref{fig:U-candidate-terms} defines the syntax of proto-types, $\ctau$, and proto-expressions, $\ce$. Proto-types and -expressions
-
-% Each expanded form maps onto a proto-expansion form. We refer to these as the \emph{common proto-expansion forms}. The mapping is given explicitly in Appendix \ref{appendix:proto-expansions-SES}.
-
-% There are two ``interesting'' proto-expansion forms, highlighted in yellow in Figure \ref{fig:U-candidate-terms}: a proto-type form for \emph{references to spliced unexpanded types}, $\acesplicedt{m}{n}$, and a proto-expression form for \emph{references to spliced unexpanded expressions}, $\acesplicede{m}{n}{\ctau}$, where $m$ and $n$ are natural numbers.%TLM utilize these to splice types and unexpanded expressions out of literal bodies.
-
-% \subsubsection{Proto-Type Validation}\label{sec:SE-proto-type-validation}
-% The \emph{proto-type validation judgement}, $\cvalidT{\Delta}{\tscenev}{\ctau}{\tau}$, is inductively defined by Rules (\ref{rules:cvalidT-U}).
-
-% \paragraph{Common Forms} Rules (\ref{rule:cvalidT-U-tvar}) through (\ref{rule:cvalidT-U-sum}) validate proto-types of common form. The first three of these are reproduced below.
-% %Each of these rules is defined based on the corresponding type formation rule, i.e.
+% % We write $\tsfrom{\escenev}$ for the type splicing scene constructed by dropping unnecessary contexts from $\escenev$:
+% % \[\tsfrom{\esceneUP{\uDelta}{\uGamma}{\uPsi}{\uPhi}{b}} = \tsceneUP{\uDelta}{b}\]
+
+% % Figure \ref{fig:U-candidate-terms} defines the syntax of proto-types, $\ctau$, and proto-expressions, $\ce$. Proto-types and -expressions
+
+% % Each expanded form maps onto a proto-expansion form. We refer to these as the \emph{common proto-expansion forms}. The mapping is given explicitly in Appendix \ref{appendix:proto-expansions-SES}.
+
+% % There are two ``interesting'' proto-expansion forms, highlighted in yellow in Figure \ref{fig:U-candidate-terms}: a proto-type form for \emph{references to spliced unexpanded types}, $\acesplicedt{m}{n}$, and a proto-expression form for \emph{references to spliced unexpanded expressions}, $\acesplicede{m}{n}{\ctau}$, where $m$ and $n$ are natural numbers.%TLM utilize these to splice types and unexpanded expressions out of literal bodies.
+
+% % \subsubsection{Proto-Type Validation}\label{sec:SE-proto-type-validation}
+% % The \emph{proto-type validation judgement}, $\cvalidT{\Delta}{\tscenev}{\ctau}{\tau}$, is inductively defined by Rules (\ref{rules:cvalidT-U}).
+
+% % \paragraph{Common Forms} Rules (\ref{rule:cvalidT-U-tvar}) through (\ref{rule:cvalidT-U-sum}) validate proto-types of common form. The first three of these are reproduced below.
+% % %Each of these rules is defined based on the corresponding type formation rule, i.e.
 Rules (\ref{rule:istypeU-var}) through (\ref{rule:istypeU-sum}), respectively.
 For example, the following proto-types validation rules are based on type formation rules (\ref{rule:istypeU-var}), (\ref{rule:istypeU-parr}) and (\ref{rule:istypeU-all}), respectively:
-% % \begin{subequations}%\label{rules:cvalidT-U}
-% \begin{equation*}\tag{\ref{rule:cvalidT-U-tvar}}
-% \inferrule{ }{
-%   \cvalidT{\Delta, \Dhyp{t}}{\tscenev}{t}{t}
-% }
-% \end{equation*}
-% \begin{equation*}\tag{\ref{rule:cvalidT-U-parr}}
-%   \inferrule{
-%     \cvalidT{\Delta}{\tscenev}{\ctau_1}{\tau_1}\\
-%     \cvalidT{\Delta}{\tscenev}{\ctau_2}{\tau_2}
-%   }{
-%     \cvalidT{\Delta}{\tscenev}{\aceparr{\ctau_1}{\ctau_2}}{\aparr{\tau_1}{\tau_2}}
-%   }
-% \end{equation*}
-% \begin{equation*}\tag{\ref{rule:cvalidT-U-all}}
-%   \inferrule {
-%     \cvalidT{\Delta, \Dhyp{t}}{\tscenev}{\ctau}{\tau}
-%   }{
-%     \cvalidT{\Delta}{\tscenev}{\aceall{t}{\ctau}}{\aall{t}{\tau}}
-%   }
-% \end{equation*}
-% % \begin{equation*}\label{rule:cvalidT-U-rec}
+% % % \begin{subequations}%\label{rules:cvalidT-U}
+% % \begin{equation*}\tag{\ref{rule:cvalidT-U-tvar}}
+% % \inferrule{ }{
+% %   \cvalidT{\Delta, \Dhyp{t}}{\tscenev}{t}{t}
+% % }
+% % \end{equation*}
+% % \begin{equation*}\tag{\ref{rule:cvalidT-U-parr}}
 % %   \inferrule{
+% %     \cvalidT{\Delta}{\tscenev}{\ctau_1}{\tau_1}\\
+% %     \cvalidT{\Delta}{\tscenev}{\ctau_2}{\tau_2}
+% %   }{
+% %     \cvalidT{\Delta}{\tscenev}{\aceparr{\ctau_1}{\ctau_2}}{\aparr{\tau_1}{\tau_2}}
+% %   }
+% % \end{equation*}
+% % \begin{equation*}\tag{\ref{rule:cvalidT-U-all}}
+% %   \inferrule {
 % %     \cvalidT{\Delta, \Dhyp{t}}{\tscenev}{\ctau}{\tau}
 % %   }{
-% %     \cvalidT{\Delta}{\tscenev}{\acerec{t}{\ctau}}{\arec{t}{\tau}}
+% %     \cvalidT{\Delta}{\tscenev}{\aceall{t}{\ctau}}{\aall{t}{\tau}}
 % %   }
 % % \end{equation*}
-% % \begin{equation*}\label{rule:cvalidT-U-prod}
-% %   \inferrule{
-% %     \{\cvalidT{\Delta}{\tscenev}{\ctau_i}{\tau_i}\}_{i \in \labelset}
-% %   }{
-% %     \cvalidT{\Delta}{\tscenev}{\aceprod{\labelset}{\mapschema{\ctau}{i}{\labelset}}}{\aprod{\labelset}{\mapschema{\tau}{i}{\labelset}}}
-% %   }
-% % \end{equation*}
-% % \begin{equation*}\label{rule:cvalidT-U-sum}
-% %   \inferrule{
-% %     \{\cvalidT{\Delta}{\tscenev}{\ctau_i}{\tau_i}\}_{i \in \labelset}
-% %   }{
-% %     \cvalidT{\Delta}{\tscenev}{\acesum{\labelset}{\mapschema{\ctau}{i}{\labelset}}}{\asum{\labelset}{\mapschema{\tau}{i}{\labelset}}}
-% %   }
-% % \end{equation*}
-
-% These rules, like the rules for common unexpanded type forms,  mirror the corresponding type formation rules, i.e.
+% % % \begin{equation*}\label{rule:cvalidT-U-rec}
+% % %   \inferrule{
+% % %     \cvalidT{\Delta, \Dhyp{t}}{\tscenev}{\ctau}{\tau}
+% % %   }{
+% % %     \cvalidT{\Delta}{\tscenev}{\acerec{t}{\ctau}}{\arec{t}{\tau}}
+% % %   }
+% % % \end{equation*}
+% % % \begin{equation*}\label{rule:cvalidT-U-prod}
+% % %   \inferrule{
+% % %     \{\cvalidT{\Delta}{\tscenev}{\ctau_i}{\tau_i}\}_{i \in \labelset}
+% % %   }{
+% % %     \cvalidT{\Delta}{\tscenev}{\aceprod{\labelset}{\mapschema{\ctau}{i}{\labelset}}}{\aprod{\labelset}{\mapschema{\tau}{i}{\labelset}}}
+% % %   }
+% % % \end{equation*}
+% % % \begin{equation*}\label{rule:cvalidT-U-sum}
+% % %   \inferrule{
+% % %     \{\cvalidT{\Delta}{\tscenev}{\ctau_i}{\tau_i}\}_{i \in \labelset}
+% % %   }{
+% % %     \cvalidT{\Delta}{\tscenev}{\acesum{\labelset}{\mapschema{\ctau}{i}{\labelset}}}{\asum{\labelset}{\mapschema{\tau}{i}{\labelset}}}
+% % %   }
+% % % \end{equation*}
+
+% % These rules, like the rules for common unexpanded type forms,  mirror the corresponding type formation rules, i.e.
 Rules (\ref{rules:istypeU}).
 The type splicing scene, $\tscenev$, passes opaquely through these rules.
-% % We can express this scheme more precisely with the following rule transformation. For each rule in Rules (\ref{rules:istypeU}), 
-% % \begin{mathpar}
-% % % \refstepcounter{equation}
-% % % \label{rule:cvalidT-U-rec}
-% % % \refstepcounter{equation}
-% % % \label{rule:cvalidT-U-prod}
-% % % \refstepcounter{equation}
-% % % \label{rule:cvalidT-U-sum}
-% % % \inferrule{J_1\\\cdots\\J_k}{J}
-% % \end{mathpar}
-% % the corresponding proto-types validation rule is
-% % \begin{mathpar}
-% % \inferrule{
-% %   \VTypof{J_1}\\
-% %   \cdots\\
-% %   \VTypof{J_k}
-% % }{
-% %   \VTypof{J}
-% % }
-% % \end{mathpar}
-% % where 
-% % \[\begin{split}
-% % \VTypof{\istypeU{\Delta}{\tau}} & = \cvalidT{\Delta}{\tscenev}{\VTypof{\tau}}{\tau}\\
-% % \VTypof{\{J_i\}_{i \in \labelset}} & = \{\VTypof{J_i}\}_{i \in \labelset}
-% % \end{split}\]
-% % and where $\VTypof{\tau}$, when $\tau$ is a metapattern of sort $\mathsf{Typ}$, is a metapattern of sort $\mathsf{CETyp}$ defined as follows:
-% % \begin{itemize}
-% % \item When $\tau$ is of definite form, $\VTypof{\tau}$ is defined as follows:
-% % \begin{align*}
-% % \VTypof{t} & = t\\
-% % \VTypof{\aparr{\tau_1}{\tau_2}} & = \aceparr{\VTypof{\tau_1}}{\VTypof{\tau_2}}\\
-% % \VTypof{\aall{t}{\tau}} & = \aceall{t}{\VTypof{\tau}}\\
-% % \VTypof{\arec{t}{\tau}} & = \acerec{t}{\VTypof{\tau}}\\
-% % \VTypof{\aprod{\labelset}{\mapschema{\tau}{i}{\labelset}}} & = \aceprod{\labelset}{\mapschemax{\VTypofv}{\tau}{i}{\labelset}}\\
-% % \VTypof{\asum{\labelset}{\mapschema{\tau}{i}{\labelset}}} & = \acesum{\labelset}{\mapschemax{\VTypofv}{\tau}{i}{\labelset}}
-% % \end{align*}
-% % \item When $\tau$ is of indefinite form, $\VTypof{\tau}$ is a uniquely corresponding metapattern also of indefinite form. For example, $\VTypof{\tau_1}=\ctau_1$ and $\VTypof{\tau_2}=\ctau_2$.
-% % \end{itemize}
-
-% % It is instructive to use this rule transformation to generate Rules (\ref{rule:cvalidT-U-tvar}) through (\ref{rule:cvalidT-U-all}) above.
+% % % We can express this scheme more precisely with the following rule transformation. For each rule in Rules (\ref{rules:istypeU}), 
+% % % \begin{mathpar}
+% % % % \refstepcounter{equation}
+% % % % \label{rule:cvalidT-U-rec}
+% % % % \refstepcounter{equation}
+% % % % \label{rule:cvalidT-U-prod}
+% % % % \refstepcounter{equation}
+% % % % \label{rule:cvalidT-U-sum}
+% % % % \inferrule{J_1\\\cdots\\J_k}{J}
+% % % \end{mathpar}
+% % % the corresponding proto-types validation rule is
+% % % \begin{mathpar}
+% % % \inferrule{
+% % %   \VTypof{J_1}\\
+% % %   \cdots\\
+% % %   \VTypof{J_k}
+% % % }{
+% % %   \VTypof{J}
+% % % }
+% % % \end{mathpar}
+% % % where 
+% % % \[\begin{split}
+% % % \VTypof{\istypeU{\Delta}{\tau}} & = \cvalidT{\Delta}{\tscenev}{\VTypof{\tau}}{\tau}\\
+% % % \VTypof{\{J_i\}_{i \in \labelset}} & = \{\VTypof{J_i}\}_{i \in \labelset}
+% % % \end{split}\]
+% % % and where $\VTypof{\tau}$, when $\tau$ is a metapattern of sort $\mathsf{Typ}$, is a metapattern of sort $\mathsf{CETyp}$ defined as follows:
+% % % \begin{itemize}
+% % % \item When $\tau$ is of definite form, $\VTypof{\tau}$ is defined as follows:
+% % % \begin{align*}
+% % % \VTypof{t} & = t\\
+% % % \VTypof{\aparr{\tau_1}{\tau_2}} & = \aceparr{\VTypof{\tau_1}}{\VTypof{\tau_2}}\\
+% % % \VTypof{\aall{t}{\tau}} & = \aceall{t}{\VTypof{\tau}}\\
+% % % \VTypof{\arec{t}{\tau}} & = \acerec{t}{\VTypof{\tau}}\\
+% % % \VTypof{\aprod{\labelset}{\mapschema{\tau}{i}{\labelset}}} & = \aceprod{\labelset}{\mapschemax{\VTypofv}{\tau}{i}{\labelset}}\\
+% % % \VTypof{\asum{\labelset}{\mapschema{\tau}{i}{\labelset}}} & = \acesum{\labelset}{\mapschemax{\VTypofv}{\tau}{i}{\labelset}}
+% % % \end{align*}
+% % % \item When $\tau$ is of indefinite form, $\VTypof{\tau}$ is a uniquely corresponding metapattern also of indefinite form. For example, $\VTypof{\tau_1}=\ctau_1$ and $\VTypof{\tau_2}=\ctau_2$.
+% % % \end{itemize}
+
+% % % It is instructive to use this rule transformation to generate Rules (\ref{rule:cvalidT-U-tvar}) through (\ref{rule:cvalidT-U-all}) above.
 We omit the remaining rules, i.e.
@@
 
-% Notice that in Rule (\ref{rule:cvalidT-U-tvar}), only type variables tracked by $\Delta$, the expansion's local type validation context, are well-formed.
+% % Notice that in Rule (\ref{rule:cvalidT-U-tvar}), only type variables tracked by $\Delta$, the expansion's local type validation context, are well-formed.
 Type variables tracked by the application site unexpanded type formation context, which is a component of the type splicing scene, $\tscenev$, are not validated.
@@
 
-% \paragraph{References to Spliced Types} The only proto-type form that does not correspond to a type form is $\acesplicedt{m}{n}$, which is a \emph{reference to a spliced unexpanded type}, i.e.
+% % \paragraph{References to Spliced Types} The only proto-type form that does not correspond to a type form is $\acesplicedt{m}{n}$, which is a \emph{reference to a spliced unexpanded type}, i.e.
 it indicates that an unexpanded type should be parsed out from the literal body, which appears in the type splicing scene $\tscenev$, beginning at position $m$ and ending at position $n$, where $m$ and $n$ are natural numbers.
 Rule (\ref{rule:cvalidT-U-splicedt}) governs this form:
-% \begin{equation*}\tag{\ref{rule:cvalidT-U-splicedt}}
-%   \inferrule{
-%     \parseUTyp{\bsubseq{b}{m}{n}}{\utau}\\
-%     \expandsTU{\uDD{\uD}{\Delta_\text{app}}}{\utau}{\tau}\\
-%     \Delta \cap \Delta_\text{app} = \emptyset
-%   }{
-%     \cvalidT{\Delta}{\tsceneU{\uDD{\uD}{\Delta_\text{app}}}{b}}{\acesplicedt{m}{n}}{\tau}
-%   }
-% \end{equation*}
-% The first premise of this rule extracts the indicated subsequence of $b$ using the partial metafunction $\bsubseq{b}{m}{n}$ and parses it using the partial metafunction $\mathsf{parseUTyp}(b)$, characterized in Sec.
+% % \begin{equation*}\tag{\ref{rule:cvalidT-U-splicedt}}
+% %   \inferrule{
+% %     \parseUTyp{\bsubseq{b}{m}{n}}{\utau}\\
+% %     \expandsTU{\uDD{\uD}{\Delta_\text{app}}}{\utau}{\tau}\\
+% %     \Delta \cap \Delta_\text{app} = \emptyset
+% %   }{
+% %     \cvalidT{\Delta}{\tsceneU{\uDD{\uD}{\Delta_\text{app}}}{b}}{\acesplicedt{m}{n}}{\tau}
+% %   }
+% % \end{equation*}
+% % The first premise of this rule extracts the indicated subsequence of $b$ using the partial metafunction $\bsubseq{b}{m}{n}$ and parses it using the partial metafunction $\mathsf{parseUTyp}(b)$, characterized in Sec.
 \ref{sec:syntax-U}, to produce the spliced unexpanded type itself, $\utau$.
 
-% The second premise of Rule (\ref{rule:cvalidT-U-splicedt}) performs type expansion of $\utau$ under the application site unexpanded type formation context, $\uDD{\uD}{\Delta_\text{app}}$, which is a component of the type splicing scene.
+% % The second premise of Rule (\ref{rule:cvalidT-U-splicedt}) performs type expansion of $\utau$ under the application site unexpanded type formation context, $\uDD{\uD}{\Delta_\text{app}}$, which is a component of the type splicing scene.
 The hypotheses in the expansion's local type formation context, $\Delta$, are not made available to $\tau$.
@@
 
-% The third premise of Rule (\ref{rule:cvalidT-U-splicedt}) imposes the constraint that the proto-expansion's type formation context, $\Delta$, be disjoint from the application site type formation context, $\Delta_\text{app}$.
+% % The third premise of Rule (\ref{rule:cvalidT-U-splicedt}) imposes the constraint that the proto-expansion's type formation context, $\Delta$, be disjoint from the application site type formation context, $\Delta_\text{app}$.
 This premise can always be discharged by $\alpha$-varying the proto-expansion that the reference to the spliced type appears within.
 
-% Together, these two premises enforce the injunction on type variable capture as described in Sec.
+% % Together, these two premises enforce the injunction on type variable capture as described in Sec.
 \ref{sec:uetsms-validation} -- the TLM provider can choose type variable names freely within a proto-expansion.
@@
 
-% Rules (\ref{rules:cvalidT-U}) validate the following lemma, which establishes that the final expansion of a valid proto-type is a well-formed type under the combined type formation context.
-% \begingroup
-% \def\thetheorem{\ref{lemma:candidate-expansion-type-validation}}
-% \begin{lemma}[Proto-Expansion Type Validation]
-% If $\cvalidT{\Delta}{\tsceneU{\uDD{\uD}{\Delta_\text{app}}}{b}}{\ctau}{\tau}$ and $\Delta \cap \Delta_\text{app}=\emptyset$ then $\istypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\tau}$.
-% \end{lemma}
-% \endgroup
-
-% \subsubsection{Proto-Expression Validation}
-% The \emph{proto-expression validation judgement}, $\cvalidE{\Delta}{\Gamma}{\escenev}{\ce}{e}{\tau}$, is defined mutually inductively with the typed expansion judgement by Rules (\ref{rules:cvalidE-U}) as follows.% This is necessary because a typed expansion judgement appears as a premise in Rule (\ref{rule:cvalidE-U-splicede}) below, and a proto-expression validation judgement appears as a premise in Rule (\ref{rule:expandsU-tsmap}) above.
-
-% \paragraph{Common Forms} Rules (\ref{rule:cvalidE-U-var}) through (\ref{rule:cvalidE-U-case}) validate proto-expressions of common form, as well as ascriptions and let binding.
+% % Rules (\ref{rules:cvalidT-U}) validate the following lemma, which establishes that the final expansion of a valid proto-type is a well-formed type under the combined type formation context.
+% % \begingroup
+% % \def\thetheorem{\ref{lemma:candidate-expansion-type-validation}}
+% % \begin{lemma}[Proto-Expansion Type Validation]
+% % If $\cvalidT{\Delta}{\tsceneU{\uDD{\uD}{\Delta_\text{app}}}{b}}{\ctau}{\tau}$ and $\Delta \cap \Delta_\text{app}=\emptyset$ then $\istypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\tau}$.
+% % \end{lemma}
+% % \endgroup
+
+% % \subsubsection{Proto-Expression Validation}
+% % The \emph{proto-expression validation judgement}, $\cvalidE{\Delta}{\Gamma}{\escenev}{\ce}{e}{\tau}$, is defined mutually inductively with the typed expansion judgement by Rules (\ref{rules:cvalidE-U}) as follows.% This is necessary because a typed expansion judgement appears as a premise in Rule (\ref{rule:cvalidE-U-splicede}) below, and a proto-expression validation judgement appears as a premise in Rule (\ref{rule:expandsU-tsmap}) above.
+
+% % \paragraph{Common Forms} Rules (\ref{rule:cvalidE-U-var}) through (\ref{rule:cvalidE-U-case}) validate proto-expressions of common form, as well as ascriptions and let binding.
 The first five of these rules are reproduced below:
-% %For each expanded expression form defined in Figure \ref{fig:U-expanded-terms}, Figure \ref{fig:U-candidate-terms} defines a corresponding proto-expression form.
+% % %For each expanded expression form defined in Figure \ref{fig:U-expanded-terms}, Figure \ref{fig:U-candidate-terms} defines a corresponding proto-expression form.
 The validation rules for proto-expressions of these forms are each based on the corresponding typing rule in Rules (\ref{rules:hastypeU}).
 For example, the validation rules for proto-expressions of variable, function and function application form  are based on Rules (\ref{rule:hastypeU-var}) through (\ref{rule:hastypeU-ap}), respectively:
-% %\begin{subequations}%\label{rules:cvalidE-U}
-% \begin{equation*}\tag{\ref{rule:cvalidE-U-var}}
-% \inferrule{ }{
-%   \cvalidE{\Delta}{\Gamma, \Ghyp{x}{\tau}}{\escenev}{x}{x}{\tau}
-% }
-% \end{equation*}
-% \begin{equation*}\tag{\ref{rule:cvalidE-U-asc}}
-% \inferrule{
-%   \cvalidT{\Delta}{\tsfrom{\escenev}}{\ctau}{\tau}\\
-%   \cvalidE{\Delta}{\Gamma}{\escenev}{\ce}{e}{\tau}
-% }{
-%   \cvalidE{\Delta}{\Gamma}{\escenev}{\aceasc{\ctau}{\ce}}{e}{\tau}
-% }
-% \end{equation*}
-% \begin{equation*}\tag{\ref{rule:cvalidE-U-letsyn}}
-%   \inferrule{
-%     \cvalidE{\Delta}{\Gamma}{\escenev}{\ce_1}{e_1}{\tau_1}\\
-%     \cvalidE{\Delta}{\Gamma, x : \tau_1}{\ce_2}{e_2}{\tau_2}
-%   }{
-%     \cvalidE{\Delta}{\Gamma}{\escenev}{\aceletsyn{x}{\ce_1}{\ce_2}}{
-%       \aeap{\aelam{\tau_1}{x}{e_2}}{e_1}
-%     }{\tau_2}
-%   }
-% \end{equation*}
-% \begin{equation*}\tag{\ref{rule:cvalidE-U-lam}}
-% \inferrule{
-%   \cvalidT{\Delta}{\tsfrom{\escenev}}{\ctau}{\tau}\\
-%   \cvalidE{\Delta}{\Gamma, \Ghyp{x}{\tau}}{\escenev}{\ce}{e}{\tau'}
-% }{
-%   \cvalidE{\Delta}{\Gamma}{\escenev}{\acelam{\ctau}{x}{\ce}}{\aelam{\tau}{x}{e}}{\aparr{\tau}{\tau'}}
-% }
-% \end{equation*}
-% \begin{equation*}\tag{\ref{rule:cvalidE-U-ap}}
-%   \inferrule{
-%     \cvalidE{\Delta}{\Gamma}{\escenev}{\ce_1}{e_1}{\aparr{\tau}{\tau'}}\\
-%     \cvalidE{\Delta}{\Gamma}{\escenev}{\ce_2}{e_2}{\tau}
-%   }{
-%     \cvalidE{\Delta}{\Gamma}{\escenev}{\aceap{\ce_1}{\ce_2}}{\aeap{e_1}{e_2}}{\tau'}
-%   }
-% \end{equation*}
-% Once again, the rules for common forms mirror the typing rules, i.e. Rules (\ref{rules:hastypeU}). The expression splicing scene, $\escenev$, passes opaquely through these rules.
-
-% Notice that in Rule (\ref{rule:cvalidE-U-var}), only variables tracked by the proto-expansion typing context, $\Gamma$, are validated.
-Variables  in the application site unexpanded typing context, which appears within the expression splicing scene $\escenev$, are not validated.
-This achieves \emph{context independence} as described in Sec.
-\ref{sec:uetsms-validation} -- seTLMs cannot impose ``hidden constraints'' on the application site unexpanded typing context, because the variable bindings at the application site are not directly available to proto-expansions.
-We will consider this formally in Sec.
-\ref{sec:SE-metatheory} below.
-
-% \paragraph{References to Spliced Unexpanded Expressions} The only proto-expression form that does not correspond to an expanded expression form is $\acesplicede{m}{n}{\ctau}$, which is a \emph{reference to a spliced unexpanded expression}, i.e.
-it indicates that an unexpanded expression should be parsed out from the literal body beginning at position $m$ and ending at position $n$.
-Rule (\ref{rule:cvalidE-U-splicede}) governs this form:
-% \begin{equation*}\tag{\ref{rule:cvalidE-U-splicede}}
-% \inferrule{
-%   \cvalidT{\emptyset}{\tsfrom{\escenev}}{\ctau}{\tau}\\
-%   \escenev=\esceneU{\uDD{\uD}{\Delta_\text{app}}}{\uGG{\uG}{\Gamma_\text{app}}}{\uPsi}{b}\\
-%   \parseUExp{\bsubseq{b}{m}{n}}{\ue}\\
-%   \expandsU{\uDD{\uD}{\Delta_\text{app}}}{\uGG{\uG}{\Gamma_\text{app}}}{\uPsi}{\ue}{e}{\tau}\\\\
-%   \Delta \cap \Delta_\text{app} = \emptyset\\
-%   \domof{\Gamma} \cap \domof{\Gamma_\text{app}} = \emptyset
-% }{
-%   \cvalidE{\Delta}{\Gamma}{\escenev}{\acesplicede{m}{n}{\ctau}}{e}{\tau}
-% }
-% \end{equation*}
-% % \begin{equation*}\label{rule:cvalidE-U-splicede}
-% % \inferrule{
-% %   \parseUExp{\bsubseq{b}{m}{n}}{\ue}\\\\
-% %   \expandsU{\Delta_\text{app}}{\Gamma_\text{app}}{\uPsi}{\ue}{e}{\tau}\\
-% %   \Delta \cap \Delta_\text{app} = \emptyset\\
-% %   \domof{\Gamma} \cap \domof{\Gamma_\text{app}} = \emptyset
-% % }{
-% %   \cvalidE{\Delta}{\Gamma}{\esceneU{\Delta_\text{app}}{\Gamma_\text{app}}{\uPsi}{b}}{\splicede{m}{n}}{e}{\tau}
-% % }
-% % \end{equation*}
-
-% The first premise of this rule validates and expands the type annotation. This type must be context independent.
-
-% The second premise of this rule serves simply to reveal the components of the expression splicing scene.
-
-% The third premise of this rule extracts the indicated subsequence of $b$ using the partial metafunction $\bsubseq{b}{m}{n}$ and parses it using the partial metafunction $\mathsf{parseUExp}(b)$, characterized in Sec.
-\ref{sec:syntax-U}, to produce the referenced spliced unexpanded expression, $\ue$.
-
-% The fourth premise of Rule (\ref{rule:cvalidE-U-splicede}) performs typed expansion of $\ue$ assuming the application site contexts that appear in the expression splicing scene.
-Notice that the hypotheses in $\Delta$ and $\Gamma$ are not made available to $\ue$.
-
-% The fifth premise of Rule (\ref{rule:cvalidE-U-splicede}) imposes the constraint that the proto-expansion's type formation context, $\Delta$, be disjoint from the application site type formation context, $\Delta_\text{app}$.
-Similarly, the sixth premise requires that the proto-expansion's typing context, $\Gamma$, be disjoint from the application site typing context, $\Gamma_\text{app}$.
-These two premises can always be discharged by $\alpha$-varying the proto-expression that the reference to the spliced unexpanded expression appears within.
-% Together, these premises enforce the prohibition on capture as described in Sec.
-\ref{sec:uetsms-validation} -- the TLM provider can choose variable names freely within a proto-expansion, because the language prevents them from shadowing those at the application site.
-% %\end{subequations}
-% % \begin{subequations}\label{rules:cvalidE-U}
-% % \begin{equation*}\label{rule:cvalidE-U-var}
+% % %\begin{subequations}%\label{rules:cvalidE-U}
+% % \begin{equation*}\tag{\ref{rule:cvalidE-U-var}}
 % % \inferrule{ }{
@@
 % % \end{equation*}
-% % \begin{equation*}\label{rule:cvalidE-U-lam}
+% % \begin{equation*}\tag{\ref{rule:cvalidE-U-asc}}
+% % \inferrule{
+% %   \cvalidT{\Delta}{\tsfrom{\escenev}}{\ctau}{\tau}\\
+% %   \cvalidE{\Delta}{\Gamma}{\escenev}{\ce}{e}{\tau}
+% % }{
+% %   \cvalidE{\Delta}{\Gamma}{\escenev}{\aceasc{\ctau}{\ce}}{e}{\tau}
+% % }
+% % \end{equation*}
+% % \begin{equation*}\tag{\ref{rule:cvalidE-U-letsyn}}
+% %   \inferrule{
+% %     \cvalidE{\Delta}{\Gamma}{\escenev}{\ce_1}{e_1}{\tau_1}\\
+% %     \cvalidE{\Delta}{\Gamma, x : \tau_1}{\ce_2}{e_2}{\tau_2}
+% %   }{
+% %     \cvalidE{\Delta}{\Gamma}{\escenev}{\aceletsyn{x}{\ce_1}{\ce_2}}{
+% %       \aeap{\aelam{\tau_1}{x}{e_2}}{e_1}
+% %     }{\tau_2}
+% %   }
+% % \end{equation*}
+% % \begin{equation*}\tag{\ref{rule:cvalidE-U-lam}}
 % % \inferrule{
@@
 % % \end{equation*}
-% % \begin{equation*}\label{rule:cvalidE-U-ap}
+% % \begin{equation*}\tag{\ref{rule:cvalidE-U-ap}}
 % %   \inferrule{
@@
 % % \end{equation*}
-% % \begin{equation*}\label{rule:cvalidE-U-tlam}
+% % Once again, the rules for common forms mirror the typing rules, i.e. Rules (\ref{rules:hastypeU}). The expression splicing scene, $\escenev$, passes opaquely through these rules.
+
+% % Notice that in Rule (\ref{rule:cvalidE-U-var}), only variables tracked by the proto-expansion typing context, $\Gamma$, are validated.
+Variables  in the application site unexpanded typing context, which appears within the expression splicing scene $\escenev$, are not validated.
+This achieves \emph{context independence} as described in Sec.
+\ref{sec:uetsms-validation} -- seTLMs cannot impose ``hidden constraints'' on the application site unexpanded typing context, because the variable bindings at the application site are not directly available to proto-expansions.
+We will consider this formally in Sec.
+\ref{sec:SE-metatheory} below.
+
+% % \paragraph{References to Spliced Unexpanded Expressions} The only proto-expression form that does not correspond to an expanded expression form is $\acesplicede{m}{n}{\ctau}$, which is a \emph{reference to a spliced unexpanded expression}, i.e.
+it indicates that an unexpanded expression should be parsed out from the literal body beginning at position $m$ and ending at position $n$.
+Rule (\ref{rule:cvalidE-U-splicede}) governs this form:
+% % \begin{equation*}\tag{\ref{rule:cvalidE-U-splicede}}
 % % \inferrule{
-% %     \cvalidE{\Delta, \Dhyp{t}}{\Gamma}{\escenev}{\ce}{e}{\tau}
+% %   \cvalidT{\emptyset}{\tsfrom{\escenev}}{\ctau}{\tau}\\
+% %   \escenev=\esceneU{\uDD{\uD}{\Delta_\text{app}}}{\uGG{\uG}{\Gamma_\text{app}}}{\uPsi}{b}\\
+% %   \parseUExp{\bsubseq{b}{m}{n}}{\ue}\\
+% %   \expandsU{\uDD{\uD}{\Delta_\text{app}}}{\uGG{\uG}{\Gamma_\text{app}}}{\uPsi}{\ue}{e}{\tau}\\\\
+% %   \Delta \cap \Delta_\text{app} = \emptyset\\
+% %   \domof{\Gamma} \cap \domof{\Gamma_\text{app}} = \emptyset
 % % }{
-% %     \cvalidEX{\acetlam{t}{\ce}}{\aetlam{t}{e}}{\aall{t}{\tau}}
+% %   \cvalidE{\Delta}{\Gamma}{\escenev}{\acesplicede{m}{n}{\ctau}}{e}{\tau}
 % % }
 % % \end{equation*}
-% % \begin{equation*}\label{rule:cvalidE-U-tap}
-% %   \inferrule{
-% %     \cvalidEX{\ce}{e}{\aall{t}{\tau}}\\
-% %     \cvalidT{\Delta}{\tsfrom{\escenev}}{\ctau'}{\tau'}
-% %   }{
-% %     \cvalidEX{\acetap{\ce}{\ctau'}}{\aetap{e}{\tau'}}{[\tau'/t]\tau}
-% %   }
-% % \end{equation*}
-% % \begin{equation*}\label{rule:cvalidE-U-fold}
-% %   \inferrule{
-% %     \cvalidT{\Delta, \Dhyp{t}}{\escenev}{\ctau}{\tau}\\
-% %     \cvalidEX{\ce}{e}{[\arec{t}{\tau}/t]\tau}
-% %   }{
-% %     \cvalidEX{\acefold{t}{\ctau}{\ce}}{\aefold{e}}{\arec{t}{\tau}}
-% %   }
-% % \end{equation*}
-% % \begin{equation*}\label{rule:cvalidE-U-unfold}
-% %   \inferrule{
-% %     \cvalidEX{\ce}{e}{\arec{t}{\tau}}
-% %   }{
-% %     \cvalidEX{\aceunfold{\ce}}{\aeunfold{e}}{[\arec{t}{\tau}/t]\tau}
-% %   }
-% % \end{equation*}
-% % \begin{equation*}\label{rule:cvalidE-U-tpl}
-% %   \inferrule{
-% %     \{\cvalidEX{\ce_i}{e_i}{\tau_i}\}_{i \in \labelset}
-% %   }{
-% %     \cvalidEX{\acetpl{\labelset}{\mapschema{\ce}{i}{\labelset}}}{\aetpl{\labelset}{\mapschema{e}{i}{\labelset}}}{\aprod{\labelset}{\mapschema{\tau}{i}{\labelset}}}
-% %   }
-% % \end{equation*}
-% % \begin{equation*}\label{rule:cvalidE-U-pr}
-% %   \inferrule{
-% %     \cvalidEX{\ce}{e}{\aprod{\labelset, \ell}{\mapschema{\tau}{i}{\labelset}; \mapitem{\ell}{\tau}}}
-% %   }{
-% %     \cvalidEX{\acepr{\ell}{\ce}}{\aepr{\ell}{e}}{\tau}
-% %   }
-% % \end{equation*}
-% % \begin{equation*}\label{rule:cvalidE-U-in}
-% %   \inferrule{
-% %     \{\cvalidT{\Delta}{\tsfrom{\escenev}}{\ctau_i}{\tau_i}\}_{i \in \labelset}\\
-% %     \cvalidT{\Delta}{\tsfrom{\escenev}}{\ctau}{\tau}\\
-% %     \cvalidEX{\ce}{e}{\tau}
-% %   }{
-% %     \left\{\shortstack{$\Delta~\Gamma \vdash_\uPsi \acein{\labelset, \ell}{\ell}{\mapschema{\ctau}{i}{\labelset}; \mapitem{\ell}{\ctau}}{\ce}$\\$\leadsto$\\$\aein{\labelset, \ell}{\ell}{\mapschema{\tau}{i}{\labelset}; \mapitem{\ell}{\tau}}{e} : \asum{\labelset, \ell}{\mapschema{\tau}{i}{\labelset}; \mapitem{\ell}{\tau}}$\vspace{-1.2em}}\right\}
-% %   }
-% % \end{equation*}
-% % \begin{equation*}\label{rule:cvalidE-U-case}
-% %   \inferrule{
-% %     \cvalidEX{\ce}{e}{\asum{\labelset}{\mapschema{\tau}{i}{\labelset}}}\\
-% %     \{\cvalidE{\Delta}{\Gamma, \Ghyp{x_i}{\tau_i}}{\escenev}{\ue_i}{e_i}{\tau}\}_{i \in \labelset}
-% %   }{
-% %     \cvalidEX{\acecase{\labelset}{\ce}{\mapschemab{x}{\ce}{i}{\labelset}}}{\aecase{\labelset}{e}{\mapschemab{x}{e}{i}{\labelset}}}{\tau}
-% %   }
-% % \end{equation*}
-% % \begin{equation*}\label{rule:cvalidE-U-splicede}
-% % \inferrule{
-% %   \parseUExp{\bsubseq{b}{m}{n}}{\ue}\\\\
-% %   \Delta \cap \Delta_\text{app} = \emptyset\\
-% %   \domof{\Gamma} \cap \domof{\Gamma_\text{app}} = \emptyset\\
-% %   \expandsU{\Delta_\text{app}}{\Gamma_\text{app}}{\uPsi}{\ue}{e}{\tau}
-% % }{
-% %   \cvalidE{\Delta}{\Gamma}{\esceneU{\Delta_\text{app}}{\Gamma_\text{app}}{\uPsi}{b}}{\acesplicede{m}{n}}{e}{\tau}
-% % }
-% % \end{equation*}
-% % \end{subequations}
-
-% % Each form of expanded expression, $e$, corresponds to a form of proto-expression, $\ce$ (compare Figure \ref{fig:U-expanded-terms} and Figure \ref{fig:U-candidate-terms}).
+% % % \begin{equation*}\label{rule:cvalidE-U-splicede}
+% % % \inferrule{
+% % %   \parseUExp{\bsubseq{b}{m}{n}}{\ue}\\\\
+% % %   \expandsU{\Delta_\text{app}}{\Gamma_\text{app}}{\uPsi}{\ue}{e}{\tau}\\
+% % %   \Delta \cap \Delta_\text{app} = \emptyset\\
+% % %   \domof{\Gamma} \cap \domof{\Gamma_\text{app}} = \emptyset
+% % % }{
+% % %   \cvalidE{\Delta}{\Gamma}{\esceneU{\Delta_\text{app}}{\Gamma_\text{app}}{\uPsi}{b}}{\splicede{m}{n}}{e}{\tau}
+% % % }
+% % % \end{equation*}
+
+% % The first premise of this rule validates and expands the type annotation. This type must be context independent.
+
+% % The second premise of this rule serves simply to reveal the components of the expression splicing scene.
+
+% % The third premise of this rule extracts the indicated subsequence of $b$ using the partial metafunction $\bsubseq{b}{m}{n}$ and parses it using the partial metafunction $\mathsf{parseUExp}(b)$, characterized in Sec.
+\ref{sec:syntax-U}, to produce the referenced spliced unexpanded expression, $\ue$.
+
+% % The fourth premise of Rule (\ref{rule:cvalidE-U-splicede}) performs typed expansion of $\ue$ assuming the application site contexts that appear in the expression splicing scene.
+Notice that the hypotheses in $\Delta$ and $\Gamma$ are not made available to $\ue$.
+
+% % The fifth premise of Rule (\ref{rule:cvalidE-U-splicede}) imposes the constraint that the proto-expansion's type formation context, $\Delta$, be disjoint from the application site type formation context, $\Delta_\text{app}$.
+Similarly, the sixth premise requires that the proto-expansion's typing context, $\Gamma$, be disjoint from the application site typing context, $\Gamma_\text{app}$.
+These two premises can always be discharged by $\alpha$-varying the proto-expression that the reference to the spliced unexpanded expression appears within.
+% % Together, these premises enforce the prohibition on capture as described in Sec.
+\ref{sec:uetsms-validation} -- the TLM provider can choose variable names freely within a proto-expansion, because the language prevents them from shadowing those at the application site.
+% % %\end{subequations}
+% % % \begin{subequations}\label{rules:cvalidE-U}
+% % % \begin{equation*}\label{rule:cvalidE-U-var}
+% % % \inferrule{ }{
+% % %   \cvalidE{\Delta}{\Gamma, \Ghyp{x}{\tau}}{\escenev}{x}{x}{\tau}
+% % % }
+% % % \end{equation*}
+% % % \begin{equation*}\label{rule:cvalidE-U-lam}
+% % % \inferrule{
+% % %   \cvalidT{\Delta}{\tsfrom{\escenev}}{\ctau}{\tau}\\
+% % %   \cvalidE{\Delta}{\Gamma, \Ghyp{x}{\tau}}{\escenev}{\ce}{e}{\tau'}
+% % % }{
+% % %   \cvalidE{\Delta}{\Gamma}{\escenev}{\acelam{\ctau}{x}{\ce}}{\aelam{\tau}{x}{e}}{\aparr{\tau}{\tau'}}
+% % % }
+% % % \end{equation*}
+% % % \begin{equation*}\label{rule:cvalidE-U-ap}
+% % %   \inferrule{
+% % %     \cvalidE{\Delta}{\Gamma}{\escenev}{\ce_1}{e_1}{\aparr{\tau}{\tau'}}\\
+% % %     \cvalidE{\Delta}{\Gamma}{\escenev}{\ce_2}{e_2}{\tau}
+% % %   }{
+% % %     \cvalidE{\Delta}{\Gamma}{\escenev}{\aceap{\ce_1}{\ce_2}}{\aeap{e_1}{e_2}}{\tau'}
+% % %   }
+% % % \end{equation*}
+% % % \begin{equation*}\label{rule:cvalidE-U-tlam}
+% % %   \inferrule{
+% % %     \cvalidE{\Delta, \Dhyp{t}}{\Gamma}{\escenev}{\ce}{e}{\tau}
+% % %   }{
+% % %     \cvalidEX{\acetlam{t}{\ce}}{\aetlam{t}{e}}{\aall{t}{\tau}}
+% % %   }
+% % % \end{equation*}
+% % % \begin{equation*}\label{rule:cvalidE-U-tap}
+% % %   \inferrule{
+% % %     \cvalidEX{\ce}{e}{\aall{t}{\tau}}\\
+% % %     \cvalidT{\Delta}{\tsfrom{\escenev}}{\ctau'}{\tau'}
+% % %   }{
+% % %     \cvalidEX{\acetap{\ce}{\ctau'}}{\aetap{e}{\tau'}}{[\tau'/t]\tau}
+% % %   }
+% % % \end{equation*}
+% % % \begin{equation*}\label{rule:cvalidE-U-fold}
+% % %   \inferrule{
+% % %     \cvalidT{\Delta, \Dhyp{t}}{\escenev}{\ctau}{\tau}\\
+% % %     \cvalidEX{\ce}{e}{[\arec{t}{\tau}/t]\tau}
+% % %   }{
+% % %     \cvalidEX{\acefold{t}{\ctau}{\ce}}{\aefold{e}}{\arec{t}{\tau}}
+% % %   }
+% % % \end{equation*}
+% % % \begin{equation*}\label{rule:cvalidE-U-unfold}
+% % %   \inferrule{
+% % %     \cvalidEX{\ce}{e}{\arec{t}{\tau}}
+% % %   }{
+% % %     \cvalidEX{\aceunfold{\ce}}{\aeunfold{e}}{[\arec{t}{\tau}/t]\tau}
+% % %   }
+% % % \end{equation*}
+% % % \begin{equation*}\label{rule:cvalidE-U-tpl}
+% % %   \inferrule{
+% % %     \{\cvalidEX{\ce_i}{e_i}{\tau_i}\}_{i \in \labelset}
+% % %   }{
+% % %     \cvalidEX{\acetpl{\labelset}{\mapschema{\ce}{i}{\labelset}}}{\aetpl{\labelset}{\mapschema{e}{i}{\labelset}}}{\aprod{\labelset}{\mapschema{\tau}{i}{\labelset}}}
+% % %   }
+% % % \end{equation*}
+% % % \begin{equation*}\label{rule:cvalidE-U-pr}
+% % %   \inferrule{
+% % %     \cvalidEX{\ce}{e}{\aprod{\labelset, \ell}{\mapschema{\tau}{i}{\labelset}; \mapitem{\ell}{\tau}}}
+% % %   }{
+% % %     \cvalidEX{\acepr{\ell}{\ce}}{\aepr{\ell}{e}}{\tau}
+% % %   }
+% % % \end{equation*}
+% % % \begin{equation*}\label{rule:cvalidE-U-in}
+% % %   \inferrule{
+% % %     \{\cvalidT{\Delta}{\tsfrom{\escenev}}{\ctau_i}{\tau_i}\}_{i \in \labelset}\\
+% % %     \cvalidT{\Delta}{\tsfrom{\escenev}}{\ctau}{\tau}\\
+% % %     \cvalidEX{\ce}{e}{\tau}
+% % %   }{
+% % %     \left\{\shortstack{$\Delta~\Gamma \vdash_\uPsi \acein{\labelset, \ell}{\ell}{\mapschema{\ctau}{i}{\labelset}; \mapitem{\ell}{\ctau}}{\ce}$\\$\leadsto$\\$\aein{\labelset, \ell}{\ell}{\mapschema{\tau}{i}{\labelset}; \mapitem{\ell}{\tau}}{e} : \asum{\labelset, \ell}{\mapschema{\tau}{i}{\labelset}; \mapitem{\ell}{\tau}}$\vspace{-1.2em}}\right\}
+% % %   }
+% % % \end{equation*}
+% % % \begin{equation*}\label{rule:cvalidE-U-case}
+% % %   \inferrule{
+% % %     \cvalidEX{\ce}{e}{\asum{\labelset}{\mapschema{\tau}{i}{\labelset}}}\\
+% % %     \{\cvalidE{\Delta}{\Gamma, \Ghyp{x_i}{\tau_i}}{\escenev}{\ue_i}{e_i}{\tau}\}_{i \in \labelset}
+% % %   }{
+% % %     \cvalidEX{\acecase{\labelset}{\ce}{\mapschemab{x}{\ce}{i}{\labelset}}}{\aecase{\labelset}{e}{\mapschemab{x}{e}{i}{\labelset}}}{\tau}
+% % %   }
+% % % \end{equation*}
+% % % \begin{equation*}\label{rule:cvalidE-U-splicede}
+% % % \inferrule{
+% % %   \parseUExp{\bsubseq{b}{m}{n}}{\ue}\\\\
+% % %   \Delta \cap \Delta_\text{app} = \emptyset\\
+% % %   \domof{\Gamma} \cap \domof{\Gamma_\text{app}} = \emptyset\\
+% % %   \expandsU{\Delta_\text{app}}{\Gamma_\text{app}}{\uPsi}{\ue}{e}{\tau}
+% % % }{
+% % %   \cvalidE{\Delta}{\Gamma}{\esceneU{\Delta_\text{app}}{\Gamma_\text{app}}{\uPsi}{b}}{\acesplicede{m}{n}}{e}{\tau}
+% % % }
+% % % \end{equation*}
+% % % \end{subequations}
+
+% % % Each form of expanded expression, $e$, corresponds to a form of proto-expression, $\ce$ (compare Figure \ref{fig:U-expanded-terms} and Figure \ref{fig:U-candidate-terms}).
 For each typing rule in Rules \ref{rules:hastypeU}, there is a corresponding proto-expression validation rule -- Rules (\ref{rule:cvalidE-U-var}) to (\ref{rule:cvalidE-U-case}) -- where the proto-expression and expanded expression correspond.
@@
 
-% %Candidate expansions cannot themselves define or apply TLMs.
+% % %Candidate expansions cannot themselves define or apply TLMs.
 This simplifies our metatheory, though it can be inconvenient at times for TLM providers.
@@
 
-% \subsection{Metatheory}\label{sec:SE-metatheory}
-% \subsubsection{Typed Expansion}
-% Let us now consider Theorem \ref{thm:typed-expansion-short-U}, which was mentioned at the beginnning of Sec. \ref{sec:typed-expansion-U} and is reproduced below:
-% \begingroup
-% \def\thetheorem{\ref{thm:typed-expansion-short-U}}
-% \begin{theorem}[Typed Expression Expansion] \hspace{-3px}If $\expandsU{\uDD{\uD}{\Delta}\hspace{-3px}}{\uGG{\uG}{\Gamma}\hspace{-3px}}{\uPsi}{\ue}{e}{\tau}$ then $\hastypeU{\Delta}{\Gamma}{e}{\tau}$.
-% \end{theorem}
-% \endgroup
-
-%  To prove this theorem, we must  prove the following stronger theorem, because the proto-expression validation judgement is defined mutually inductively with the typed expansion judgement:
-
-% \begingroup
-% \def\thetheorem{\ref{thm:typed-expansion-full-U}}
-% \begin{theorem}[Typed Expansion (Full)] ~
-% \begin{enumerate}
-% \item If $\expandsU{\uDD{\uD}{\Delta}}{\uGG{\uG}{\Gamma}}{\uAS{\uA}{\Psi}}{\ue}{e}{\tau}$ then $\hastypeU{\Delta}{\Gamma}{e}{\tau}$.
-% \item If $\cvalidE{\Delta}{\Gamma}{\esceneU{\uDD{\uD}{\Delta_\text{app}}}{\uGG{\uG}{\Gamma_\text{app}}}{\uAS{\uA}{\Psi}}{b}}{\ce}{e}{\tau}$ and $\Delta \cap \Delta_\text{app} = \emptyset$ and $\domof{\Gamma} \cap \domof{\Gamma_\text{app}} = \emptyset$ then $\hastypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\Gcons{\Gamma}{\Gamma_\text{app}}}{e}{\tau}$.
-% \end{enumerate}
-% \end{theorem}
-% \endgroup
-% \begin{proof}
-% By mutual rule induction over Rules (\ref{rules:expandsU}) and Rules (\ref{rules:cvalidE-U}).
+% % \subsection{Metatheory}\label{sec:SE-metatheory}
+% % \subsubsection{Typed Expansion}
+% % Let us now consider Theorem \ref{thm:typed-expansion-short-U}, which was mentioned at the beginnning of Sec. \ref{sec:typed-expansion-U} and is reproduced below:
+% % \begingroup
+% % \def\thetheorem{\ref{thm:typed-expansion-short-U}}
+% % \begin{theorem}[Typed Expression Expansion] \hspace{-3px}If $\expandsU{\uDD{\uD}{\Delta}\hspace{-3px}}{\uGG{\uG}{\Gamma}\hspace{-3px}}{\uPsi}{\ue}{e}{\tau}$ then $\hastypeU{\Delta}{\Gamma}{e}{\tau}$.
+% % \end{theorem}
+% % \endgroup
+
+% %  To prove this theorem, we must  prove the following stronger theorem, because the proto-expression validation judgement is defined mutually inductively with the typed expansion judgement:
+
+% % \begingroup
+% % \def\thetheorem{\ref{thm:typed-expansion-full-U}}
+% % \begin{theorem}[Typed Expansion (Full)] ~
+% % \begin{enumerate}
+% % \item If $\expandsU{\uDD{\uD}{\Delta}}{\uGG{\uG}{\Gamma}}{\uAS{\uA}{\Psi}}{\ue}{e}{\tau}$ then $\hastypeU{\Delta}{\Gamma}{e}{\tau}$.
+% % \item If $\cvalidE{\Delta}{\Gamma}{\esceneU{\uDD{\uD}{\Delta_\text{app}}}{\uGG{\uG}{\Gamma_\text{app}}}{\uAS{\uA}{\Psi}}{b}}{\ce}{e}{\tau}$ and $\Delta \cap \Delta_\text{app} = \emptyset$ and $\domof{\Gamma} \cap \domof{\Gamma_\text{app}} = \emptyset$ then $\hastypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\Gcons{\Gamma}{\Gamma_\text{app}}}{e}{\tau}$.
+% % \end{enumerate}
+% % \end{theorem}
+% % \endgroup
+% % \begin{proof}
+% % By mutual rule induction over Rules (\ref{rules:expandsU}) and Rules (\ref{rules:cvalidE-U}).
 The full proof is given in Appendix \ref{appendix:SES-typed-expression-expansion-metatheory}.
@@
 
-% The proof of part 1 proceeds by inducting over the typed expansion assumption.
+% % The proof of part 1 proceeds by inducting over the typed expansion assumption.
 The only interesting cases are those related to seTLM definition and application, reproduced below.
@@
 
-% \begin{byCases}
-% \item[\text{(\ref{rule:expandsU-syntax})}] We have 
-% \begin{pfsteps}
-%   \item \ue=\uesyntax{\tsmv}{\utau'}{\eparse}{\ue'} \BY{assumption}
-%   \item \expandsTU{\uDelta}{\utau'}{\tau'} \BY{assumption} \pflabel{expandsTU}
-%  \item \hastypeU{\emptyset}{\emptyset}{\eparse}{\aparr{\tBody}{\tParseResultExp}} \BY{assumption}\pflabel{eparse}
-%   \item \expandsU{\uDelta}{\uGamma}{\uPsi, \uShyp{\tsmv}{a}{\tau'}{\eparse}}{\ue'}{e}{\tau} \BY{assumption}\pflabel{expandsU}
-% %  \item \uetsmenv{\Delta}{\Psi} \BY{assumption}\pflabel{uetsmenv1}
-%  \item \istypeU{\Delta}{\tau'} \BY{Lemma \ref{lemma:type-expansion-U} to \pfref{expandsTU}} \pflabel{istype}
-% %  \item \uetsmenv{\Delta}{\Psi, \xuetsmbnd{\tsmv}{\tau'}{\eparse}} \BY{Definition \ref{def:seTLM-def-ctx-formation} on \pfref{uetsmenv1}, \pfref{istype} and \pfref{eparse}}\pflabel{uetsmenv3}
-%   \item \hastypeU{\Delta}{\Gamma}{e}{\tau} \BY{IH, part 1(a) on \pfref{expandsU}}
-% \end{pfsteps}
-% \resetpfcounter
-
-% \item[\text{(\ref{rule:expandsU-tsmap})}] We have 
-% \begin{pfsteps}
-%   \item \ue=\utsmap{\tsmv}{b} \BY{assumption}
-%   \item \uA = \uA', \vExpands{\tsmv}{a} \BY{assumption}
-%   \item \Psi=\Psi', \xuetsmbnd{a}{\tau}{\eparse} \BY{assumption}
-%   \item \encodeBody{b}{\ebody} \BY{assumption}
-%   \item \evalU{\eparse(\ebody)}{\aein{\lbltxt{SuccessE}}{\ecand}} \BY{assumption}
-%   \item \decodeCondE{\ecand}{\ce} \BY{assumption}
-%   \item \cvalidE{\emptyset}{\emptyset}{\esceneU{\uDelta}{\uGamma}{\uPsi}{b}}{\ce}{e}{\tau} \BY{assumption}\pflabel{cvalidE}
-% %  \item \uetsmenv{\Delta}{\Psi} \BY{assumption} \pflabel{uetsmenv}
-%   \item \emptyset \cap \Delta = \emptyset \BY{finite set intersection} \pflabel{delta-cap}
-%   \item {\emptyset} \cap \domof{\Gamma} = \emptyset \BY{finite set intersection} \pflabel{gamma-cap}
-%   \item \hastypeU{\emptyset \cup \Delta}{\emptyset \cup \Gamma}{e}{\tau} \BY{IH, part 2 on \pfref{cvalidE}, \pfref{delta-cap}, and \pfref{gamma-cap}} \pflabel{penultimate}
-%   \item \hastypeU{\Delta}{\Gamma}{e}{\tau} \BY{finite set and finite function identity over \pfref{penultimate}}
-% \end{pfsteps}
-% \resetpfcounter
-% \end{byCases}
-
-% The proof of part 2 proceeds by induction over the proto-expression validation assumption.
+% % \begin{byCases}
+% % \item[\text{(\ref{rule:expandsU-syntax})}] We have 
+% % \begin{pfsteps}
+% %   \item \ue=\uesyntax{\tsmv}{\utau'}{\eparse}{\ue'} \BY{assumption}
+% %   \item \expandsTU{\uDelta}{\utau'}{\tau'} \BY{assumption} \pflabel{expandsTU}
+% %  \item \hastypeU{\emptyset}{\emptyset}{\eparse}{\aparr{\tBody}{\tParseResultExp}} \BY{assumption}\pflabel{eparse}
+% %   \item \expandsU{\uDelta}{\uGamma}{\uPsi, \uShyp{\tsmv}{a}{\tau'}{\eparse}}{\ue'}{e}{\tau} \BY{assumption}\pflabel{expandsU}
+% % %  \item \uetsmenv{\Delta}{\Psi} \BY{assumption}\pflabel{uetsmenv1}
+% %  \item \istypeU{\Delta}{\tau'} \BY{Lemma \ref{lemma:type-expansion-U} to \pfref{expandsTU}} \pflabel{istype}
+% % %  \item \uetsmenv{\Delta}{\Psi, \xuetsmbnd{\tsmv}{\tau'}{\eparse}} \BY{Definition \ref{def:seTLM-def-ctx-formation} on \pfref{uetsmenv1}, \pfref{istype} and \pfref{eparse}}\pflabel{uetsmenv3}
+% %   \item \hastypeU{\Delta}{\Gamma}{e}{\tau} \BY{IH, part 1(a) on \pfref{expandsU}}
+% % \end{pfsteps}
+% % \resetpfcounter
+
+% % \item[\text{(\ref{rule:expandsU-tsmap})}] We have 
+% % \begin{pfsteps}
+% %   \item \ue=\utsmap{\tsmv}{b} \BY{assumption}
+% %   \item \uA = \uA', \vExpands{\tsmv}{a} \BY{assumption}
+% %   \item \Psi=\Psi', \xuetsmbnd{a}{\tau}{\eparse} \BY{assumption}
+% %   \item \encodeBody{b}{\ebody} \BY{assumption}
+% %   \item \evalU{\eparse(\ebody)}{\aein{\lbltxt{SuccessE}}{\ecand}} \BY{assumption}
+% %   \item \decodeCondE{\ecand}{\ce} \BY{assumption}
+% %   \item \cvalidE{\emptyset}{\emptyset}{\esceneU{\uDelta}{\uGamma}{\uPsi}{b}}{\ce}{e}{\tau} \BY{assumption}\pflabel{cvalidE}
+% % %  \item \uetsmenv{\Delta}{\Psi} \BY{assumption} \pflabel{uetsmenv}
+% %   \item \emptyset \cap \Delta = \emptyset \BY{finite set intersection} \pflabel{delta-cap}
+% %   \item {\emptyset} \cap \domof{\Gamma} = \emptyset \BY{finite set intersection} \pflabel{gamma-cap}
+% %   \item \hastypeU{\emptyset \cup \Delta}{\emptyset \cup \Gamma}{e}{\tau} \BY{IH, part 2 on \pfref{cvalidE}, \pfref{delta-cap}, and \pfref{gamma-cap}} \pflabel{penultimate}
+% %   \item \hastypeU{\Delta}{\Gamma}{e}{\tau} \BY{finite set and finite function identity over \pfref{penultimate}}
+% % \end{pfsteps}
+% % \resetpfcounter
+% % \end{byCases}
+
+% % The proof of part 2 proceeds by induction over the proto-expression validation assumption.
 The only interesting case governs references to spliced expressions.
 In the following cases, let $\uDelta_\text{app}=\uDD{\uD}{\Delta_\text{app}}$ and $\uGamma_\text{app}=\uGG{\uG}{\Gamma_\text{app}}$ and $\uPsi = \uAS{\uA}{\Psi}$.
-% \begin{byCases}
-% % \item[\text{(\ref{rule:cvalidE-U-var})}] ~
+% % \begin{byCases}
+% % % \item[\text{(\ref{rule:cvalidE-U-var})}] ~
+% % % \begin{pfsteps*}
+% % %   \item $\ce=x$ \BY{assumption}
+% % %   \item $e=x$ \BY{assumption}
+% % %   \item $\Gamma=\Gamma', \Ghyp{x}{\tau}$ \BY{assumption}
+% % %   \item $\hastypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\Gamma', \Ghyp{x}{\tau}}{x}{\tau}$ \BY{Rule (\ref{rule:hastypeU-var})} \pflabel{hastypeU}
+% % %   \item $\hastypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\Gcons{\Gamma', \Ghyp{x}{\tau}}{\Gamma_\text{app}}}{x}{\tau}$ \BY{Lemma \ref{lemma:weakening-U} over $\Gamma_\text{app}$ to \pfref{hastypeU}}
+% % % \end{pfsteps*}
+% % % \resetpfcounter
+
+% % % \item[\text{(\ref{rule:cvalidE-U-lam})}] ~
+% % % \begin{pfsteps*}
+% % %   \item $\ce=\acelam{\ctau_1}{x}{\ce'}$ \BY{assumption}
+% % %   \item $e=\aelam{\tau_1}{x}{e'}$ \BY{assumption}
+% % %   \item $\tau=\aparr{\tau_1}{\tau_2}$ \BY{assumption}
+% % %   \item $\cvalidT{\Delta}{\tsceneU{\uDelta_\text{app}}{b}}{\ctau_1}{\tau_1}$ \BY{assumption} \pflabel{cvalidT}
+% % %   \item $\cvalidE{\Delta}{\Gamma, \Ghyp{x}{\tau_1}}{\esceneU{\uDelta_\text{app}}{\uGamma_\text{app}}{\uPsi}{b}}{\ce'}{e'}{\tau_2}$ \BY{assumption} \pflabel{cvalidE}
+% % % %  \item $\uetsmenv{\Delta_\text{app}}{\Psi}$ \BY{assumption} \pflabel{uetsmenv}
+% % %   \item $\Delta \cap \Delta_\text{app}=\emptyset$ \BY{assumption} \pflabel{delta-disjoint}
+% % %   \item $\domof{\Gamma} \cap \domof{\Gamma_\text{app}}=\emptyset$ \BY{assumption} \pflabel{gamma-disjoint}
+% % %   \item $x \notin \domof{\Gamma_\text{app}}$ \BY{identification convention} \pflabel{x-fresh}
+% % %   \item $\domof{\Gamma, x : \tau_1} \cap \domof{\Gamma_\text{app}}=\emptyset$ \BY{\pfref{gamma-disjoint} and \pfref{x-fresh}} \pflabel{gamma-disjoint2}
+% % %   \item $\istypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\tau_1}$ \BY{Lemma \ref{lemma:candidate-expansion-type-validation} on \pfref{cvalidT} and \pfref{delta-disjoint}} \pflabel{istype}
+% % %   \item $\hastypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\Gcons{\Gamma, \Ghyp{x}{\tau_1}}{\Gamma_\text{app}}}{e'}{\tau_2}$ \BY{IH, part 2 on \pfref{cvalidE}, \pfref{delta-disjoint} and \pfref{gamma-disjoint2}} \pflabel{hastype1}
+% % %   \item $\hastypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\Gcons{\Gamma}{\Gamma_\text{app}}, \Ghyp{x}{\tau_1}}{e'}{\tau_2}$ \BY{exchange over $\Gamma_\text{app}$ on \pfref{hastype1}} \pflabel{hastype2}
+% % %   \item $\hastypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\Gcons{\Gamma}{\Gamma_\text{app}}}{\aelam{\tau_1}{x}{e'}}{\aparr{\tau_1}{\tau_2}}$ \BY{Rule (\ref{rule:hastypeU-lam}) on \pfref{istype} and \pfref{hastype2}}
+% % % \end{pfsteps*}
+% % % \resetpfcounter
+
+% % % \item[\text{(\ref{rule:cvalidE-U-ap})}] ~
+% % % \begin{pfsteps*}
+% % %   \item $\ce=\aceap{\ce_1}{\ce_2}$ \BY{assumption}
+% % %   \item $e=\aeap{e_1}{e_2}$ \BY{assumption}
+% % %   \item $\cvalidE{\Delta}{\Gamma}{\esceneU{\uDelta_\text{app}}{\uGamma_\text{app}}{\uPsi}{b}}{\ce_1}{e_1}{\aparr{\tau_2}{\tau}}$ \BY{assumption} \pflabel{cvalidE1}
+% % %   \item $\cvalidE{\Delta}{\Gamma}{\esceneU{\uDelta_\text{app}}{\uGamma_\text{app}}{\uPsi}{b}}{\ce_2}{e_2}{\tau_2}$ \BY{assumption} \pflabel{cvalidE2}
+% % % %  \item $\uetsmenv{\Delta_\text{app}}{\Psi}$ \BY{assumption} \pflabel{uetsmenv}
+% % %   \item $\Delta \cap \Delta_\text{app}=\emptyset$ \BY{assumption} \pflabel{delta-disjoint}
+% % %   \item $\domof{\Gamma} \cap \domof{\Gamma_\text{app}}=\emptyset$ \BY{assumption} \pflabel{gamma-disjoint}
+% % %   \item $\hastypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\Gcons{\Gamma}{\Gamma_\text{app}}}{e_1}{\aparr{\tau_2}{\tau}}$ \BY{IH, part 2 on \pfref{cvalidE1}, \pfref{delta-disjoint} and \pfref{gamma-disjoint}} \pflabel{hastypeU1}
+% % %   \item $\hastypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\Gcons{\Gamma}{\Gamma_\text{app}}}{e_2}{\tau_2}$ \BY{IH, part 2 on \pfref{cvalidE2}, \pfref{delta-disjoint} and \pfref{gamma-disjoint}} \pflabel{hastypeU2}
+% % %   \item $\hastypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\Gcons{\Gamma}{\Gamma_\text{app}}}{\aeap{e_1}{e_2}}{\tau}$ \BY{Rule (\ref{rule:hastypeU-ap}) on \pfref{hastypeU1} and \pfref{hastypeU2}}
+% % % \end{pfsteps*}
+% % % \resetpfcounter
+
+% % % \item[\text{(\ref{rule:cvalidE-U-tlam})}] ~
+% % % \begin{pfsteps}
+% % %   \item \ce=\acetlam{t}{\ce'} \BY{assumption}
+% % %   \item e = \aetlam{t}{e'} \BY{assumption}
+% % %   \item \tau = \aall{t}{\tau'}\BY{assumption}
+% % %   \item \cvalidE{\Delta, \Dhyp{t}}{\Gamma}{\esceneU{\uDelta_\text{app}}{\uGamma_\text{app}}{\uPsi}{b}}{\ce'}{e'}{\tau'} \BY{assumption} \pflabel{cvalidE}
+% % % %  \item \uetsmenv{\Delta_\text{app}}{\Psi} \BY{assumption} \pflabel{uetsmenv}
+% % %   \item \Delta \cap \Delta_\text{app}=\emptyset \BY{assumption} \pflabel{delta-disjoint}
+% % %   \item \domof{\Gamma} \cap \domof{\Gamma_\text{app}}=\emptyset \BY{assumption} \pflabel{gamma-disjoint}
+% % %   \item \Dhyp{t} \notin \Delta_\text{app} \BY{identification convention}\pflabel{t-fresh}
+% % %   \item \Delta, \Dhyp{t} \cap \Delta_\text{app} = \emptyset \BY{\pfref{delta-disjoint} and \pfref{t-fresh}}\pflabel{delta-disjoint2}
+% % %   \item \hastypeU{\Dcons{\Delta, \Dhyp{t}}{\Delta_\text{app}}}{\Gcons{\Gamma}{\Gamma_\text{app}}}{e'}{\tau'} \BY{IH, part 2 on \pfref{cvalidE}, \pfref{delta-disjoint2} and \pfref{gamma-disjoint}}\pflabel{hastype1}
+% % %   \item \hastypeU{\Dcons{\Delta}{\Delta_\text{app}, \Dhyp{t}}}{\Gcons{\Gamma}{\Gamma_\text{app}}}{e'}{\tau'} \BY{exchange over $\Delta_\text{app}$ on \pfref{hastype1}}\pflabel{hastype2}
+% % %   \item \hastypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\Gcons{\Gamma}{\Gamma_\text{app}}}{\aetlam{t}{e'}}{\aall{t}{\tau'}} \BY{Rule (\ref{rule:hastypeU-tlam}) on \pfref{hastype2}}
+% % % \end{pfsteps}
+% % % \resetpfcounter
+
+% % % \item[{\text{(\ref{rule:cvalidE-U-tap})}}~\textbf{through}~{\text{(\ref{rule:cvalidE-U-case})}}] These cases follow analagously, i.e.
+we apply the IH, part 2 to all proto-expression validation judgements, Lemma \ref{lemma:candidate-expansion-type-validation} to all proto-type validation judgements, the identification convention to ensure that extended contexts remain disjoint, weakening and exchange as needed, and the corresponding typing rule in Rules (\ref{rule:hastypeU-tap}) through (\ref{rule:hastypeU-case}).
+% % % \\
+
+% % \item[\text{(\ref{rule:cvalidE-U-splicede})}] ~
 % % \begin{pfsteps*}
-% %   \item $\ce=x$ \BY{assumption}
-% %   \item $e=x$ \BY{assumption}
-% %   \item $\Gamma=\Gamma', \Ghyp{x}{\tau}$ \BY{assumption}
-% %   \item $\hastypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\Gamma', \Ghyp{x}{\tau}}{x}{\tau}$ \BY{Rule (\ref{rule:hastypeU-var})} \pflabel{hastypeU}
-% %   \item $\hastypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\Gcons{\Gamma', \Ghyp{x}{\tau}}{\Gamma_\text{app}}}{x}{\tau}$ \BY{Lemma \ref{lemma:weakening-U} over $\Gamma_\text{app}$ to \pfref{hastypeU}}
-% % \end{pfsteps*}
-% % \resetpfcounter
-
-% % \item[\text{(\ref{rule:cvalidE-U-lam})}] ~
-% % \begin{pfsteps*}
-% %   \item $\ce=\acelam{\ctau_1}{x}{\ce'}$ \BY{assumption}
-% %   \item $e=\aelam{\tau_1}{x}{e'}$ \BY{assumption}
-% %   \item $\tau=\aparr{\tau_1}{\tau_2}$ \BY{assumption}
-% %   \item $\cvalidT{\Delta}{\tsceneU{\uDelta_\text{app}}{b}}{\ctau_1}{\tau_1}$ \BY{assumption} \pflabel{cvalidT}
-% %   \item $\cvalidE{\Delta}{\Gamma, \Ghyp{x}{\tau_1}}{\esceneU{\uDelta_\text{app}}{\uGamma_\text{app}}{\uPsi}{b}}{\ce'}{e'}{\tau_2}$ \BY{assumption} \pflabel{cvalidE}
+% %   \item $\ce=\acesplicede{m}{n}{\ctau}$ \BY{assumption}
+% %   \item $  \escenev=\esceneU{\uDD{\uD}{\Delta_\text{app}}}{\uGG{\uG}{\Gamma_\text{app}}}{\uPsi}{b}$ \BY{assumption}
+% %   \item   $\cvalidT{\emptyset}{\tsfrom{\escenev}}{\ctau}{\tau}$ \BY{assumption}
+% %   \item $\parseUExp{\bsubseq{b}{m}{n}}{\ue}$ \BY{assumption}
+% %   \item $\expandsU{\uDelta_\text{app}}{\uGamma_\text{app}}{\uPsi}{\ue}{e}{\tau}$ \BY{assumption} \pflabel{expands}
 % % %  \item $\uetsmenv{\Delta_\text{app}}{\Psi}$ \BY{assumption} \pflabel{uetsmenv}
@@
 % %   \item $\domof{\Gamma} \cap \domof{\Gamma_\text{app}}=\emptyset$ \BY{assumption} \pflabel{gamma-disjoint}
-% %   \item $x \notin \domof{\Gamma_\text{app}}$ \BY{identification convention} \pflabel{x-fresh}
-% %   \item $\domof{\Gamma, x : \tau_1} \cap \domof{\Gamma_\text{app}}=\emptyset$ \BY{\pfref{gamma-disjoint} and \pfref{x-fresh}} \pflabel{gamma-disjoint2}
-% %   \item $\istypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\tau_1}$ \BY{Lemma \ref{lemma:candidate-expansion-type-validation} on \pfref{cvalidT} and \pfref{delta-disjoint}} \pflabel{istype}
-% %   \item $\hastypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\Gcons{\Gamma, \Ghyp{x}{\tau_1}}{\Gamma_\text{app}}}{e'}{\tau_2}$ \BY{IH, part 2 on \pfref{cvalidE}, \pfref{delta-disjoint} and \pfref{gamma-disjoint2}} \pflabel{hastype1}
-% %   \item $\hastypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\Gcons{\Gamma}{\Gamma_\text{app}}, \Ghyp{x}{\tau_1}}{e'}{\tau_2}$ \BY{exchange over $\Gamma_\text{app}$ on \pfref{hastype1}} \pflabel{hastype2}
-% %   \item $\hastypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\Gcons{\Gamma}{\Gamma_\text{app}}}{\aelam{\tau_1}{x}{e'}}{\aparr{\tau_1}{\tau_2}}$ \BY{Rule (\ref{rule:hastypeU-lam}) on \pfref{istype} and \pfref{hastype2}}
-% % \end{pfsteps*}
-% % \resetpfcounter
-
-% % \item[\text{(\ref{rule:cvalidE-U-ap})}] ~
-% % \begin{pfsteps*}
-% %   \item $\ce=\aceap{\ce_1}{\ce_2}$ \BY{assumption}
-% %   \item $e=\aeap{e_1}{e_2}$ \BY{assumption}
-% %   \item $\cvalidE{\Delta}{\Gamma}{\esceneU{\uDelta_\text{app}}{\uGamma_\text{app}}{\uPsi}{b}}{\ce_1}{e_1}{\aparr{\tau_2}{\tau}}$ \BY{assumption} \pflabel{cvalidE1}
-% %   \item $\cvalidE{\Delta}{\Gamma}{\esceneU{\uDelta_\text{app}}{\uGamma_\text{app}}{\uPsi}{b}}{\ce_2}{e_2}{\tau_2}$ \BY{assumption} \pflabel{cvalidE2}
-% % %  \item $\uetsmenv{\Delta_\text{app}}{\Psi}$ \BY{assumption} \pflabel{uetsmenv}
-% %   \item $\Delta \cap \Delta_\text{app}=\emptyset$ \BY{assumption} \pflabel{delta-disjoint}
-% %   \item $\domof{\Gamma} \cap \domof{\Gamma_\text{app}}=\emptyset$ \BY{assumption} \pflabel{gamma-disjoint}
-% %   \item $\hastypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\Gcons{\Gamma}{\Gamma_\text{app}}}{e_1}{\aparr{\tau_2}{\tau}}$ \BY{IH, part 2 on \pfref{cvalidE1}, \pfref{delta-disjoint} and \pfref{gamma-disjoint}} \pflabel{hastypeU1}
-% %   \item $\hastypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\Gcons{\Gamma}{\Gamma_\text{app}}}{e_2}{\tau_2}$ \BY{IH, part 2 on \pfref{cvalidE2}, \pfref{delta-disjoint} and \pfref{gamma-disjoint}} \pflabel{hastypeU2}
-% %   \item $\hastypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\Gcons{\Gamma}{\Gamma_\text{app}}}{\aeap{e_1}{e_2}}{\tau}$ \BY{Rule (\ref{rule:hastypeU-ap}) on \pfref{hastypeU1} and \pfref{hastypeU2}}
-% % \end{pfsteps*}
-% % \resetpfcounter
-
-% % \item[\text{(\ref{rule:cvalidE-U-tlam})}] ~
-% % \begin{pfsteps}
-% %   \item \ce=\acetlam{t}{\ce'} \BY{assumption}
-% %   \item e = \aetlam{t}{e'} \BY{assumption}
-% %   \item \tau = \aall{t}{\tau'}\BY{assumption}
-% %   \item \cvalidE{\Delta, \Dhyp{t}}{\Gamma}{\esceneU{\uDelta_\text{app}}{\uGamma_\text{app}}{\uPsi}{b}}{\ce'}{e'}{\tau'} \BY{assumption} \pflabel{cvalidE}
-% % %  \item \uetsmenv{\Delta_\text{app}}{\Psi} \BY{assumption} \pflabel{uetsmenv}
-% %   \item \Delta \cap \Delta_\text{app}=\emptyset \BY{assumption} \pflabel{delta-disjoint}
-% %   \item \domof{\Gamma} \cap \domof{\Gamma_\text{app}}=\emptyset \BY{assumption} \pflabel{gamma-disjoint}
-% %   \item \Dhyp{t} \notin \Delta_\text{app} \BY{identification convention}\pflabel{t-fresh}
-% %   \item \Delta, \Dhyp{t} \cap \Delta_\text{app} = \emptyset \BY{\pfref{delta-disjoint} and \pfref{t-fresh}}\pflabel{delta-disjoint2}
-% %   \item \hastypeU{\Dcons{\Delta, \Dhyp{t}}{\Delta_\text{app}}}{\Gcons{\Gamma}{\Gamma_\text{app}}}{e'}{\tau'} \BY{IH, part 2 on \pfref{cvalidE}, \pfref{delta-disjoint2} and \pfref{gamma-disjoint}}\pflabel{hastype1}
-% %   \item \hastypeU{\Dcons{\Delta}{\Delta_\text{app}, \Dhyp{t}}}{\Gcons{\Gamma}{\Gamma_\text{app}}}{e'}{\tau'} \BY{exchange over $\Delta_\text{app}$ on \pfref{hastype1}}\pflabel{hastype2}
-% %   \item \hastypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\Gcons{\Gamma}{\Gamma_\text{app}}}{\aetlam{t}{e'}}{\aall{t}{\tau'}} \BY{Rule (\ref{rule:hastypeU-tlam}) on \pfref{hastype2}}
-% % \end{pfsteps}
-% % \resetpfcounter
-
-% % \item[{\text{(\ref{rule:cvalidE-U-tap})}}~\textbf{through}~{\text{(\ref{rule:cvalidE-U-case})}}] These cases follow analagously, i.e.
-we apply the IH, part 2 to all proto-expression validation judgements, Lemma \ref{lemma:candidate-expansion-type-validation} to all proto-type validation judgements, the identification convention to ensure that extended contexts remain disjoint, weakening and exchange as needed, and the corresponding typing rule in Rules (\ref{rule:hastypeU-tap}) through (\ref{rule:hastypeU-case}).
-% % \\
-
-% \item[\text{(\ref{rule:cvalidE-U-splicede})}] ~
-% \begin{pfsteps*}
-%   \item $\ce=\acesplicede{m}{n}{\ctau}$ \BY{assumption}
-%   \item $  \escenev=\esceneU{\uDD{\uD}{\Delta_\text{app}}}{\uGG{\uG}{\Gamma_\text{app}}}{\uPsi}{b}$ \BY{assumption}
-%   \item   $\cvalidT{\emptyset}{\tsfrom{\escenev}}{\ctau}{\tau}$ \BY{assumption}
-%   \item $\parseUExp{\bsubseq{b}{m}{n}}{\ue}$ \BY{assumption}
-%   \item $\expandsU{\uDelta_\text{app}}{\uGamma_\text{app}}{\uPsi}{\ue}{e}{\tau}$ \BY{assumption} \pflabel{expands}
-% %  \item $\uetsmenv{\Delta_\text{app}}{\Psi}$ \BY{assumption} \pflabel{uetsmenv}
-%   \item $\Delta \cap \Delta_\text{app}=\emptyset$ \BY{assumption} \pflabel{delta-disjoint}
-%   \item $\domof{\Gamma} \cap \domof{\Gamma_\text{app}}=\emptyset$ \BY{assumption} \pflabel{gamma-disjoint}
-%   \item $\hastypeU{\Delta_\text{app}}{\Gamma_\text{app}}{e}{\tau}$ \BY{IH, part 1 on \pfref{expands}} \pflabel{hastype}
-%   \item $\hastypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\Gcons{\Gamma}{\Gamma_\text{app}}}{e}{\tau}$ \BY{Lemma \ref{lemma:weakening-U} over $\Delta$ and $\Gamma$ and exchange on \pfref{hastype}}
-% \end{pfsteps*}
-% \resetpfcounter
-% \end{byCases}
-
-% The mutual induction can be shown to be well-founded by showing that the following numeric metric on the judgements that we induct over is decreasing:
-% \begin{align*}
-% \sizeof{\expandsU{\uDelta}{\uGamma}{\uPsi}{\ue}{e}{\tau}} & = \sizeof{\ue}\\
-% \sizeof{\cvalidE{\Delta}{\Gamma}{\esceneU{\uDelta_\text{app}}{\uGamma_\text{app}}{\uPsi}{b}}{\ce}{e}{\tau}} & = \sizeof{b}
-% \end{align*}
-% where $\sizeof{b}$ is the length of $b$ and $\sizeof{\ue}$ is the sum of the lengths of the literal bodies in $\ue$ (see Appendix \ref{appendix:SES-body-lengths}.)
-
-% The only case in the proof of part 1 that invokes part 2 is Case (\ref{rule:expandsU-tsmap}). There, we have that the metric remains stable: \begin{align*}
-%  & \sizeof{\expandsU{\uDelta}{\uGamma}{\uPsi}{\utsmap{\tsmv}{b}}{e}{\tau}}\\
-% =& \sizeof{\cvalidE{\emptyset}{\emptyset}{\esceneU{\uDelta}{\uGamma}{\uPsi}{b}}{\ce}{e}{\tau}}\\
-% =&\sizeof{b}\end{align*}
-
-% The only case in the proof of part 2 that invokes part 1 is Case (\ref{rule:cvalidE-U-splicede}).
-There, we have that $\parseUExp{\bsubseq{b}{m}{n}}{\ue}$ and the IH is applied to the judgement $\expandsU{\uDelta_\text{app}}{\uGamma_\text{app}}{\uPsi}{\ue}{e}{\tau}$ where $\uDelta_\text{app}=\uDD{\uD}{\Delta_\text{app}}$ and $\uGamma_\text{app}=\uGG{\uG}{\Gamma_\text{app}}$ and $\uPsi=\uAS{\uA}{\Psi}$.
-Because the metric is stable when passing from part 1 to part 2, we must have that it is strictly decreasing in the other direction:
-% \[\sizeof{\expandsU{\uDelta_\text{app}}{\uGamma_\text{app}}{\uPsi}{\ue}{e}{\tau}} < \sizeof{\cvalidE{\Delta}{\Gamma}{\esceneU{\uDelta_\text{app}}{\uGamma_\text{app}}{\uPsi}{b}}{\acesplicede{m}{n}{\ctau}}{e}{\tau}}\]
-% i.e. by the definitions above, 
-% \[\sizeof{\ue} < \sizeof{b}\]
-
-% This is established by appeal to the following two conditions.
-The first condition states that an unexpanded expression constructed by parsing a textual sequence $b$ is strictly smaller, as measured by the metric defined above, than the length of $b$, because some characters must necessarily be used to invoke a TLM and delimit each literal body.
-% \begingroup
-% \def\thetheorem{\ref{condition:body-parsing}}
-% \begin{condition}[Expression Parsing Monotonicity] If $\parseUExp{b}{\ue}$ then $\sizeof{\ue} < \sizeof{b}$.\end{condition}
-% \endgroup
-% The second condition simply states that subsequences of $b$ are no longer than $b$.
-% \begingroup
-% \def\thetheorem{\ref{condition:body-subsequences}}
-% \begin{condition}[Body Subsequencing] If $\bsubseq{b}{m}{n}=b'$ then $\sizeof{b'} \leq \sizeof{b}$. \end{condition}
-% \endgroup
-
-% Combining these two conditions, we have that $\sizeof{\ue} < \sizeof{b}$ as needed.
-% \end{proof}
-
-% % We need to define the following theorem about proto-expression validation mutually with Theorem \ref{thm:typed-expansion-U}. 
-% % \begin{theorem}[Proto-Expansion Expression Validation]\label{thm:candidate-expansion-validation-U}
-% % If $\cvalidE{\Delta}{\Gamma}{\esceneU{\Delta_\text{app}}{\Gamma_\text{app}}{\uPsi}{b}}{\ce}{e}{\tau}$ and $\uetsmenv{\Delta_\text{app}}{\uPsi}$ then $\hastypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\Gcons{\Gamma}{\Gamma_\text{app}}}{e}{\tau}$.
-% % \end{theorem}
-% % \begin{proof} By rule induction over Rules (\ref{rules:cvalidE-U}).
-% % \begin{byCases}
-% % \item[\text{(\ref{rule:cvalidE-U-var})}] ~
-% % \begin{pfsteps*}
-% %   \item $\ce=x$ \BY{assumption}
-% %   \item $e=x$ \BY{assumption}
-% %   \item $\Gamma=\Gamma', \Ghyp{x}{\tau}$ \BY{assumption}
-% %   \item $\hastypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\Gamma', \Ghyp{x}{\tau}}{x}{\tau}$ \BY{Rule (\ref{rule:hastypeU-var})} \pflabel{hastypeU}
-% %   \item $\hastypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\Gcons{\Gamma', \Ghyp{x}{\tau}}{\Gamma_\text{app}}}{x}{\tau}$ \BY{Lemma \ref{lemma:weakening-U} over $\Gamma_\text{app}$ to \pfref{hastypeU}}
-% % \end{pfsteps*}
-% % \resetpfcounter
-
-% % \item[\text{(\ref{rule:cvalidE-U-lam})}] ~
-% % \begin{pfsteps*}
-% %   \item $\ce=\acelam{\ctau_1}{x}{\ce'}$ \BY{assumption}
-% %   \item $e=\aelam{\tau_1}{x}{e'}$ \BY{assumption}
-% %   \item $\tau=\aparr{\tau_1}{\tau_2}$ \BY{assumption}
-% %   \item $\cvalidT{\Delta}{\esceneU{\Delta_\text{app}}{\Gamma_\text{app}}{\uPsi}{b}}{\ctau_1}{\tau_1}$ \BY{assumption} \pflabel{cvalidT}
-% %   \item $\cvalidE{\Delta}{\Gamma, \Ghyp{x}{\tau_1}}{\esceneU{\Delta_\text{app}}{\Gamma_\text{app}}{\uPsi}{b}}{\ce'}{e'}{\tau_2}$ \BY{assumption} \pflabel{cvalidE}
-% %   \item $\uetsmenv{\Delta_\text{app}}{\uPsi}$ \BY{assumption} \pflabel{uetsmenv}
-% %   \item $\istypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\tau_1}$ \BY{Lemma \ref{lemma:candidate-expansion-type-validation} on \pfref{cvalidT}} \pflabel{istype}
-% %   \item $\hastypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\Gcons{\Gamma, \Ghyp{x}{\tau_1}}{\Gamma_\text{app}}}{e'}{\tau_2}$ \BY{IH on \pfref{cvalidE} and \pfref{uetsmenv}} \pflabel{hastype1}
-% %   \item $\hastypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\Gcons{\Gamma}{\Gamma_\text{app}}, \Ghyp{x}{\tau_1}}{e'}{\tau_2}$ \BY{exchange over $\Gamma_\text{app}$ on \pfref{hastype1}} \pflabel{hastype2}
-% %   \item $\hastypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\Gcons{\Gamma}{\Gamma_\text{app}}}{\aelam{\tau_1}{x}{e'}}{\aparr{\tau_1}{\tau_2}}$ \BY{Rule (\ref{rule:hastypeU-lam}) on \pfref{istype} and \pfref{hastype2}}
-% % \end{pfsteps*}
-% % \resetpfcounter
-
-% % \item[\text{(\ref{rule:cvalidE-U-ap})}] ~
-% % \begin{pfsteps*}
-% %   \item $\ce=\aceap{\ce_1}{\ce_2}$ \BY{assumption}
-% %   \item $e=\aeap{e_1}{e_2}$ \BY{assumption}
-% %   \item $\cvalidE{\Delta}{\Gamma}{\esceneU{\Delta_\text{app}}{\Gamma_\text{app}}{\uPsi}{b}}{\ce_1}{e_1}{\aparr{\tau_1}{\tau}}$ \BY{assumption} \pflabel{cvalidE1}
-% %   \item $\cvalidE{\Delta}{\Gamma}{\esceneU{\Delta_\text{app}}{\Gamma_\text{app}}{\uPsi}{b}}{\ce_2}{e_2}{\tau_1}$ \BY{assumption} \pflabel{cvalidE2}
-% %   \item $\uetsmenv{\Delta_\text{app}}{\uPsi}$ \BY{assumption} \pflabel{uetsmenv}
-% %   \item $\hastypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\Gcons{\Gamma}{\Gamma_\text{app}}}{e_1}{\aparr{\tau_1}{\tau}}$ \BY{IH on \pfref{cvalidE1} and \pfref{uetsmenv}} \pflabel{hastypeU1}
-% %   \item $\hastypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\Gcons{\Gamma}{\Gamma_\text{app}}}{e_2}{\tau_1}$ \BY{IH on \pfref{cvalidE2} and \pfref{uetsmenv}} \pflabel{hastypeU2}
-% %   \item $\hastypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\Gcons{\Gamma}{\Gamma_\text{app}}}{\aeap{e_1}{e_2}}{\tau}$ \BY{Rule (\ref{rule:hastypeU-ap}) on \pfref{hastypeU1} and \pfref{hastypeU2}}
-% % \end{pfsteps*}
-% % \resetpfcounter
-
-% % \item[\VExpof{\text{\ref{rule:hastypeU-tlam}}}~\text{through}~\VExpof{\text{\ref{rule:hastypeU-case}}}] These cases follow analagously, i.e.
-we apply the IH to all proto-expression validation premises, Lemma \ref{lemma:candidate-expansion-type-validation} to all proto-types validation premises, weakening and exchange as needed, and then apply the corresponding typing rule.
-% % \\
-
-% % \item[\text{(\ref{rule:cvalidE-U-splicede})}] ~
-% % \begin{pfsteps*}
-% %   \item $\ce=\acesplicede{m}{n}$ \BY{assumption}
-% %   \item $\parseUExp{\bsubseq{b}{m}{n}}{\ue}$ \BY{assumption}
-% %   \item $\expandsU{\Delta_\text{app}}{\Gamma_\text{app}}{\uPsi}{\ue}{e}{\tau}$ \BY{assumption} \pflabel{expands}
-% %   \item $\uetsmenv{\Delta_\text{app}}{\uPsi}$ \BY{assumption} \pflabel{uetsmenv}
-% %   \item $\hastypeU{\Delta_\text{app}}{\Gamma_\text{app}}{e}{\tau}$ \BY{Theorem \ref{thm:typed-expansion-U} on \pfref{expands} and \pfref{uetsmenv}} \pflabel{hastype}
-% %   \item $\hastypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\Gcons{\Gamma}{\Gamma_\text{app}}}{e}{\tau}$ \BY{Lemma \ref{lemma:weakening-U} on \pfref{hastype}}
+% %   \item $\hastypeU{\Delta_\text{app}}{\Gamma_\text{app}}{e}{\tau}$ \BY{IH, part 1 on \pfref{expands}} \pflabel{hastype}
+% %   \item $\hastypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\Gcons{\Gamma}{\Gamma_\text{app}}}{e}{\tau}$ \BY{Lemma \ref{lemma:weakening-U} over $\Delta$ and $\Gamma$ and exchange on \pfref{hastype}}
 % % \end{pfsteps*}
@@
 % % \end{byCases}
+
+% % The mutual induction can be shown to be well-founded by showing that the following numeric metric on the judgements that we induct over is decreasing:
+% % \begin{align*}
+% % \sizeof{\expandsU{\uDelta}{\uGamma}{\uPsi}{\ue}{e}{\tau}} & = \sizeof{\ue}\\
+% % \sizeof{\cvalidE{\Delta}{\Gamma}{\esceneU{\uDelta_\text{app}}{\uGamma_\text{app}}{\uPsi}{b}}{\ce}{e}{\tau}} & = \sizeof{b}
+% % \end{align*}
+% % where $\sizeof{b}$ is the length of $b$ and $\sizeof{\ue}$ is the sum of the lengths of the literal bodies in $\ue$ (see Appendix \ref{appendix:SES-body-lengths}.)
+
+% % The only case in the proof of part 1 that invokes part 2 is Case (\ref{rule:expandsU-tsmap}). There, we have that the metric remains stable: \begin{align*}
+% %  & \sizeof{\expandsU{\uDelta}{\uGamma}{\uPsi}{\utsmap{\tsmv}{b}}{e}{\tau}}\\
+% % =& \sizeof{\cvalidE{\emptyset}{\emptyset}{\esceneU{\uDelta}{\uGamma}{\uPsi}{b}}{\ce}{e}{\tau}}\\
+% % =&\sizeof{b}\end{align*}
+
+% % The only case in the proof of part 2 that invokes part 1 is Case (\ref{rule:cvalidE-U-splicede}).
+There, we have that $\parseUExp{\bsubseq{b}{m}{n}}{\ue}$ and the IH is applied to the judgement $\expandsU{\uDelta_\text{app}}{\uGamma_\text{app}}{\uPsi}{\ue}{e}{\tau}$ where $\uDelta_\text{app}=\uDD{\uD}{\Delta_\text{app}}$ and $\uGamma_\text{app}=\uGG{\uG}{\Gamma_\text{app}}$ and $\uPsi=\uAS{\uA}{\Psi}$.
+Because the metric is stable when passing from part 1 to part 2, we must have that it is strictly decreasing in the other direction:
+% % \[\sizeof{\expandsU{\uDelta_\text{app}}{\uGamma_\text{app}}{\uPsi}{\ue}{e}{\tau}} < \sizeof{\cvalidE{\Delta}{\Gamma}{\esceneU{\uDelta_\text{app}}{\uGamma_\text{app}}{\uPsi}{b}}{\acesplicede{m}{n}{\ctau}}{e}{\tau}}\]
+% % i.e. by the definitions above, 
+% % \[\sizeof{\ue} < \sizeof{b}\]
+
+% % This is established by appeal to the following two conditions.
+The first condition states that an unexpanded expression constructed by parsing a textual sequence $b$ is strictly smaller, as measured by the metric defined above, than the length of $b$, because some characters must necessarily be used to invoke a TLM and delimit each literal body.
+% % \begingroup
+% % \def\thetheorem{\ref{condition:body-parsing}}
+% % \begin{condition}[Expression Parsing Monotonicity] If $\parseUExp{b}{\ue}$ then $\sizeof{\ue} < \sizeof{b}$.\end{condition}
+% % \endgroup
+% % The second condition simply states that subsequences of $b$ are no longer than $b$.
+% % \begingroup
+% % \def\thetheorem{\ref{condition:body-subsequences}}
+% % \begin{condition}[Body Subsequencing] If $\bsubseq{b}{m}{n}=b'$ then $\sizeof{b'} \leq \sizeof{b}$. \end{condition}
+% % \endgroup
+
+% % Combining these two conditions, we have that $\sizeof{\ue} < \sizeof{b}$ as needed.
 % % \end{proof}
 
-% %\qed
-
-% \subsubsection{Abstract Reasoning Principles}\label{sec:uetsms-reasoning-principles}
-% The following theorem summarizes the abstract reasoning principles that programmers can rely on when applying an seTLM. In particular, the programmer can be sure that:
-% \begin{enumerate} 
-% \item \textbf{Segmentation}: The segmentation determined by the proto-expansion actually segments the literal body (i.e. each segment is in-bounds and the segments are non-overlapping.)
-% \item \textbf{Typing 1}: The type of the expansion is consistent with the type annotation on the seTLM definition.
-% \item \textbf{Typing 2}: Each spliced type has a well-formed expansion at the application site.
-% \item \textbf{Typing 3}: Each type annotation on a reference to a spliced expression has a well-formed expansion at the application site.
-% \item \textbf{Typing 4}: Each spliced expression has a well-typed expansion consistent with its type annotation.
-% \item \textbf{Capture Avoidance}: The final expansion can be decomposed into a  term with variables in place of each spliced type or expression.
+% % % We need to define the following theorem about proto-expression validation mutually with Theorem \ref{thm:typed-expansion-U}. 
+% % % \begin{theorem}[Proto-Expansion Expression Validation]\label{thm:candidate-expansion-validation-U}
+% % % If $\cvalidE{\Delta}{\Gamma}{\esceneU{\Delta_\text{app}}{\Gamma_\text{app}}{\uPsi}{b}}{\ce}{e}{\tau}$ and $\uetsmenv{\Delta_\text{app}}{\uPsi}$ then $\hastypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\Gcons{\Gamma}{\Gamma_\text{app}}}{e}{\tau}$.
+% % % \end{theorem}
+% % % \begin{proof} By rule induction over Rules (\ref{rules:cvalidE-U}).
+% % % \begin{byCases}
+% % % \item[\text{(\ref{rule:cvalidE-U-var})}] ~
+% % % \begin{pfsteps*}
+% % %   \item $\ce=x$ \BY{assumption}
+% % %   \item $e=x$ \BY{assumption}
+% % %   \item $\Gamma=\Gamma', \Ghyp{x}{\tau}$ \BY{assumption}
+% % %   \item $\hastypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\Gamma', \Ghyp{x}{\tau}}{x}{\tau}$ \BY{Rule (\ref{rule:hastypeU-var})} \pflabel{hastypeU}
+% % %   \item $\hastypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\Gcons{\Gamma', \Ghyp{x}{\tau}}{\Gamma_\text{app}}}{x}{\tau}$ \BY{Lemma \ref{lemma:weakening-U} over $\Gamma_\text{app}$ to \pfref{hastypeU}}
+% % % \end{pfsteps*}
+% % % \resetpfcounter
+
+% % % \item[\text{(\ref{rule:cvalidE-U-lam})}] ~
+% % % \begin{pfsteps*}
+% % %   \item $\ce=\acelam{\ctau_1}{x}{\ce'}$ \BY{assumption}
+% % %   \item $e=\aelam{\tau_1}{x}{e'}$ \BY{assumption}
+% % %   \item $\tau=\aparr{\tau_1}{\tau_2}$ \BY{assumption}
+% % %   \item $\cvalidT{\Delta}{\esceneU{\Delta_\text{app}}{\Gamma_\text{app}}{\uPsi}{b}}{\ctau_1}{\tau_1}$ \BY{assumption} \pflabel{cvalidT}
+% % %   \item $\cvalidE{\Delta}{\Gamma, \Ghyp{x}{\tau_1}}{\esceneU{\Delta_\text{app}}{\Gamma_\text{app}}{\uPsi}{b}}{\ce'}{e'}{\tau_2}$ \BY{assumption} \pflabel{cvalidE}
+% % %   \item $\uetsmenv{\Delta_\text{app}}{\uPsi}$ \BY{assumption} \pflabel{uetsmenv}
+% % %   \item $\istypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\tau_1}$ \BY{Lemma \ref{lemma:candidate-expansion-type-validation} on \pfref{cvalidT}} \pflabel{istype}
+% % %   \item $\hastypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\Gcons{\Gamma, \Ghyp{x}{\tau_1}}{\Gamma_\text{app}}}{e'}{\tau_2}$ \BY{IH on \pfref{cvalidE} and \pfref{uetsmenv}} \pflabel{hastype1}
+% % %   \item $\hastypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\Gcons{\Gamma}{\Gamma_\text{app}}, \Ghyp{x}{\tau_1}}{e'}{\tau_2}$ \BY{exchange over $\Gamma_\text{app}$ on \pfref{hastype1}} \pflabel{hastype2}
+% % %   \item $\hastypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\Gcons{\Gamma}{\Gamma_\text{app}}}{\aelam{\tau_1}{x}{e'}}{\aparr{\tau_1}{\tau_2}}$ \BY{Rule (\ref{rule:hastypeU-lam}) on \pfref{istype} and \pfref{hastype2}}
+% % % \end{pfsteps*}
+% % % \resetpfcounter
+
+% % % \item[\text{(\ref{rule:cvalidE-U-ap})}] ~
+% % % \begin{pfsteps*}
+% % %   \item $\ce=\aceap{\ce_1}{\ce_2}$ \BY{assumption}
+% % %   \item $e=\aeap{e_1}{e_2}$ \BY{assumption}
+% % %   \item $\cvalidE{\Delta}{\Gamma}{\esceneU{\Delta_\text{app}}{\Gamma_\text{app}}{\uPsi}{b}}{\ce_1}{e_1}{\aparr{\tau_1}{\tau}}$ \BY{assumption} \pflabel{cvalidE1}
+% % %   \item $\cvalidE{\Delta}{\Gamma}{\esceneU{\Delta_\text{app}}{\Gamma_\text{app}}{\uPsi}{b}}{\ce_2}{e_2}{\tau_1}$ \BY{assumption} \pflabel{cvalidE2}
+% % %   \item $\uetsmenv{\Delta_\text{app}}{\uPsi}$ \BY{assumption} \pflabel{uetsmenv}
+% % %   \item $\hastypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\Gcons{\Gamma}{\Gamma_\text{app}}}{e_1}{\aparr{\tau_1}{\tau}}$ \BY{IH on \pfref{cvalidE1} and \pfref{uetsmenv}} \pflabel{hastypeU1}
+% % %   \item $\hastypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\Gcons{\Gamma}{\Gamma_\text{app}}}{e_2}{\tau_1}$ \BY{IH on \pfref{cvalidE2} and \pfref{uetsmenv}} \pflabel{hastypeU2}
+% % %   \item $\hastypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\Gcons{\Gamma}{\Gamma_\text{app}}}{\aeap{e_1}{e_2}}{\tau}$ \BY{Rule (\ref{rule:hastypeU-ap}) on \pfref{hastypeU1} and \pfref{hastypeU2}}
+% % % \end{pfsteps*}
+% % % \resetpfcounter
+
+% % % \item[\VExpof{\text{\ref{rule:hastypeU-tlam}}}~\text{through}~\VExpof{\text{\ref{rule:hastypeU-case}}}] These cases follow analagously, i.e.
+we apply the IH to all proto-expression validation premises, Lemma \ref{lemma:candidate-expansion-type-validation} to all proto-types validation premises, weakening and exchange as needed, and then apply the corresponding typing rule.
+% % % \\
+
+% % % \item[\text{(\ref{rule:cvalidE-U-splicede})}] ~
+% % % \begin{pfsteps*}
+% % %   \item $\ce=\acesplicede{m}{n}$ \BY{assumption}
+% % %   \item $\parseUExp{\bsubseq{b}{m}{n}}{\ue}$ \BY{assumption}
+% % %   \item $\expandsU{\Delta_\text{app}}{\Gamma_\text{app}}{\uPsi}{\ue}{e}{\tau}$ \BY{assumption} \pflabel{expands}
+% % %   \item $\uetsmenv{\Delta_\text{app}}{\uPsi}$ \BY{assumption} \pflabel{uetsmenv}
+% % %   \item $\hastypeU{\Delta_\text{app}}{\Gamma_\text{app}}{e}{\tau}$ \BY{Theorem \ref{thm:typed-expansion-U} on \pfref{expands} and \pfref{uetsmenv}} \pflabel{hastype}
+% % %   \item $\hastypeU{\Dcons{\Delta}{\Delta_\text{app}}}{\Gcons{\Gamma}{\Gamma_\text{app}}}{e}{\tau}$ \BY{Lemma \ref{lemma:weakening-U} on \pfref{hastype}}
+% % % \end{pfsteps*}
+% % % \resetpfcounter
+% % % \end{byCases}
+% % % \end{proof}
+
+% % %\qed
+
+% % \subsubsection{Abstract Reasoning Principles}\label{sec:uetsms-reasoning-principles}
+% % The following theorem summarizes the abstract reasoning principles that programmers can rely on when applying an seTLM. In particular, the programmer can be sure that:
+% % \begin{enumerate} 
+% % \item \textbf{Segmentation}: The segmentation determined by the proto-expansion actually segments the literal body (i.e. each segment is in-bounds and the segments are non-overlapping.)
+% % \item \textbf{Typing 1}: The type of the expansion is consistent with the type annotation on the seTLM definition.
+% % \item \textbf{Typing 2}: Each spliced type has a well-formed expansion at the application site.
+% % \item \textbf{Typing 3}: Each type annotation on a reference to a spliced expression has a well-formed expansion at the application site.
+% % \item \textbf{Typing 4}: Each spliced expression has a well-typed expansion consistent with its type annotation.
+% % \item \textbf{Capture Avoidance}: The final expansion can be decomposed into a  term with variables in place of each spliced type or expression.
 The expansions of these spliced types and expressions can be substituted into this term in the standard capture avoiding manner.
-% \item \textbf{Context Independence}: The decomposed term is indeed well-typed independent of the application site contexts.
-
-% \end{enumerate}
-
-% \begingroup
-% \def\thetheorem{\ref{thm:tsc-SES}}
-% \begin{theorem}[seTLM Abstract Reasoning Principles]
-% If $\expandsU{\uDD{\uD}{\Delta}}{\uGG{\uG}{\Gamma}}{\uPsi}{\utsmap{\tsmv}{b}}{e}{\tau}$ then:
-% \begin{enumerate}
-% \item (\textbf{Typing 1}) $\uPsi = \uPsi', \uShyp{\tsmv}{a}{\tau}{\eparse}$ and $\hastypeU{\Delta}{\Gamma}{e}{\tau}$
-% \item $\encodeBody{b}{\ebody}$
-% \item $\evalU{\ap{\eparse}{\ebody}}{\aein{\lbltxt{SuccessE}}{\ecand}}$
-% \item $\decodeCondE{\ecand}{\ce}$
-% \item (\textbf{Segmentation}) $\segOK{\segof{\ce}}{b}$
-% \item $\segof{\ce} = \sseq{\acesplicedt{m'_i}{n'_i}}{\nty} \cup \sseq{\acesplicede{m_i}{n_i}{\ctau_i}}{\nexp}$
-% \item \textbf{(Typing 2)} $\sseq{
-%       \expandsTU{\uDD{\uD}{\Delta}}
-%       {
-%         \parseUTypF{\bsubseq{b}{m'_i}{n'_i}}
-%       }{\tau'_i}
-%     }{\nty}$ and $\sseq{\istypeU{\Delta}{\tau'_i}}{\nty}$
-% \item \textbf{(Typing 3)} $\sseq{
-%   \cvalidT{\emptyset}{
-%     \tsceneUP
-%       {\uDD
-%         {\uD}{\Delta}
-%       }{b}
-%   }{
-%     \ctau_i
-%   }{\tau_i}
-% }{\nexp}$ and $\sseq{\istypeU{\Delta}{\tau_i}}{\nexp}$
-% \item \textbf{(Typing 4)} $\sseq{
-%   \expandsU
-%     {\uDD{\uD}{\Delta}}
-%     {\uGG{\uG}{\Gamma}}
-%     {\uPsi}
-%     {\parseUExpF{\bsubseq{b}{m_i}{n_i}}}
-%     {e_i}
-%     {\tau_i}
-% }{\nexp}$ and $\sseq{\hastypeU{\Delta}{\Gamma}{e_i}{\tau_i}}{\nexp}$
-% \item (\textbf{Capture Avoidance}) $e = [\sseq{\tau'_i/t_i}{\nty}, \sseq{e_i/x_i}{\nexp}]e'$ for some $\sseq{t_i}{\nty}$ and $\sseq{x_i}{\nexp}$ and $e'$
-% \item (\textbf{Context Independence}) $\mathsf{fv}(e') \subset \sseq{t_i}{\nty} \cup \sseq{x_i}{\nexp}$
-%   % $\hastypeU
-%   % {\sseq{\Dhyp{t_i}}{\nty}}
-%   % {\sseq{x_i : \tau_i}{\nexp}}
-%   % {e'}{\tau}$
-% \end{enumerate}
-% \end{theorem}
-% \begin{proof} The proof, which involves auxiliary lemmas about the decomposition of proto-types and proto-expressions, is given in Appendix \ref{appendix:SES-reasoning-principles}.
-% \end{proof}
-% \endgroup
+% % \item \textbf{Context Independence}: The decomposed term is indeed well-typed independent of the application site contexts.
+
+% % \end{enumerate}
+
+% % \begingroup
+% % \def\thetheorem{\ref{thm:tsc-SES}}
+% % \begin{theorem}[seTLM Abstract Reasoning Principles]
+% % If $\expandsU{\uDD{\uD}{\Delta}}{\uGG{\uG}{\Gamma}}{\uPsi}{\utsmap{\tsmv}{b}}{e}{\tau}$ then:
+% % \begin{enumerate}
+% % \item (\textbf{Typing 1}) $\uPsi = \uPsi', \uShyp{\tsmv}{a}{\tau}{\eparse}$ and $\hastypeU{\Delta}{\Gamma}{e}{\tau}$
+% % \item $\encodeBody{b}{\ebody}$
+% % \item $\evalU{\ap{\eparse}{\ebody}}{\aein{\lbltxt{SuccessE}}{\ecand}}$
+% % \item $\decodeCondE{\ecand}{\ce}$
+% % \item (\textbf{Segmentation}) $\segOK{\segof{\ce}}{b}$
+% % \item $\segof{\ce} = \sseq{\acesplicedt{m'_i}{n'_i}}{\nty} \cup \sseq{\acesplicede{m_i}{n_i}{\ctau_i}}{\nexp}$
+% % \item \textbf{(Typing 2)} $\sseq{
+% %       \expandsTU{\uDD{\uD}{\Delta}}
+% %       {
+% %         \parseUTypF{\bsubseq{b}{m'_i}{n'_i}}
+% %       }{\tau'_i}
+% %     }{\nty}$ and $\sseq{\istypeU{\Delta}{\tau'_i}}{\nty}$
+% % \item \textbf{(Typing 3)} $\sseq{
+% %   \cvalidT{\emptyset}{
+% %     \tsceneUP
+% %       {\uDD
+% %         {\uD}{\Delta}
+% %       }{b}
+% %   }{
+% %     \ctau_i
+% %   }{\tau_i}
+% % }{\nexp}$ and $\sseq{\istypeU{\Delta}{\tau_i}}{\nexp}$
+% % \item \textbf{(Typing 4)} $\sseq{
+% %   \expandsU
+% %     {\uDD{\uD}{\Delta}}
+% %     {\uGG{\uG}{\Gamma}}
+% %     {\uPsi}
+% %     {\parseUExpF{\bsubseq{b}{m_i}{n_i}}}
+% %     {e_i}
+% %     {\tau_i}
+% % }{\nexp}$ and $\sseq{\hastypeU{\Delta}{\Gamma}{e_i}{\tau_i}}{\nexp}$
+% % \item (\textbf{Capture Avoidance}) $e = [\sseq{\tau'_i/t_i}{\nty}, \sseq{e_i/x_i}{\nexp}]e'$ for some $\sseq{t_i}{\nty}$ and $\sseq{x_i}{\nexp}$ and $e'$
+% % \item (\textbf{Context Independence}) $\mathsf{fv}(e') \subset \sseq{t_i}{\nty} \cup \sseq{x_i}{\nexp}$
+% %   % $\hastypeU
+% %   % {\sseq{\Dhyp{t_i}}{\nty}}
+% %   % {\sseq{x_i : \tau_i}{\nexp}}
+% %   % {e'}{\tau}$
+% % \end{enumerate}
+% % \end{theorem}
+% % \begin{proof} The proof, which involves auxiliary lemmas about the decomposition of proto-types and proto-expressions, is given in Appendix \ref{appendix:SES-reasoning-principles}.
+% % \end{proof}
+% % \endgroup
 \newcommand{\implSec}{Implementation}
@@
 \label{sec:implementation}
-
-Note that even when a TLM needs only a single helper function, it must be placed into a named module.
+The implementation, \li{Relit}, which is an ongoing effort based on the finished design described in this paper, consists of a few changes to the context-free grammar of Reason together with a parse tree pre-processor for the OCaml compiler, currently using its PPX system (discussed in Sec.~\ref{sec:existing-approaches}).
+
+\paragraph{TLM Definitions} The Reason grammar extension turns TLM definitions into module definitions of a stereotyped form.
+For example, the module \li{RegexNotation} defining \li{$regex} from Fig.~\ref{fig:regex-tlm-def} is, eliding a few minor details, shown below.
+The comments describe how the components correspond.
+\begin{lstlisting}
+  module RegexNotation = {
+    module Notation_regex___relit /* = $regex */ = { 
+      type t = Regex.t; /* expansion type */
+      module Lexer = RegexLexer; /* lexer module must match ocamllex sig */
+      module Parser = RegexParser; /* parser module must match menhir sig */
+      module Dependencies = { module Regex = Regex; } /* see text */
+      exception Apply(string); /* used in TLM applications, see below */
+    } 
+  }
+\end{lstlisting}
+
+TLM definitions are constrained so that the signature of the generated module is a \emph{singleton}, i.e.
+that it uniquely determines the module.
+This implies that the dependencies cannot list a value---only module paths and concrete types can be listed.
+As such, TLM definition equality is decidable and TLM definitions can be tracked in module types (signatures) without any special effort.
+Note that this is more restrictive than in the formal system in the previous section, where parse functions were included inline, rather than by path, and dependencies were arbitrary values.
+It must be possible to resolve the implementation of the lexer and parser modules, not just their signatures, at expansion time, so lexer and module paths are lifted to the top level of the compilation unit.
+
+\paragraph{TLM Application} For a TLM application, the Reason grammar extension produces an expression annotated with \li{relit} that raises the corresponding \li{Apply} exception, supplying a string containing the literal body.
+For example, a fragment of the example from Fig.~\ref{fig:first-tlm-example} is below:
+\begin{lstlisting}
+  open RegexNotation; /* ... */
+  let restriction_template = (gene) => [%relit raise(Notation_regex___relit.Apply(
+      "$(bisA)$(DNA.any_base)*$$(gene)$(DNA.any_base)*$(bisA)"))]
+\end{lstlisting}
+
+The Relit preprocessor rewrites TLM applications by following the specification in the previous sections differing only in that the ``dependency variable'' is a ``dependency path'', here to \li{Notation_regex___relit.Dependencies}.
+The only difficult is in finding the TLM definition at all, which due to our integration into the module system may have been accessed indirectly, e.g.
+here via the \li{open} directive and in other situations via module synonyms, functors and so on.
+The trick is in the singleton restriction just described---if we can determine the signature for the module path that \li{Apply} is accessed from, we have the full definition in hand.
+To do so, we run the OCaml typechecker on this unexpanded representation.
+Because the TLM application raises an exception, its type can be generalized arbitrarily.
+As long as unrelated type errors at a particular depth are resolved (treating the TLM applications at that depth as ``holes'') expansion at that depth can proceed.
+This occasionally causes types that incidentally appear in error messages to be more general than the programmer might expect (e.g.
+here \li{restriction_template} will have type \li{'a -> 'b} during this ``pre-typing'' phase).
+%If this type ends up in an error message incidentally, e.g. due to a signature mismatch elsewhere, it may be somewhat confusing. 
+We are working to improve error reporting and, more generally, improving integration with various other tools in the Reason / OCaml ecosystem.
+
+\paragraph{Proto-Expansion Validation} The typing phase of proto-expansion validation also defers to the OCaml typechecker, with spliced expressions  represented as variables of the corresponding segment type, resolved relative to the typing context determined at the call site.
+The recursively generated substitutions are, after validation is complete, substituted in the standard capture-avoiding manner.
+
+\paragraph{OCaml Support} Although our efforts are focused on Reason, it would not be difficult to add TLM-related  forms to OCaml's standard grammar, and no changes to the preprocessor just described should be needed.
+The only caveat is that TLMs that support splicing would then, ideally, also be agnostic to the base language grammar in use.
+We are exploring various ways for TLM providers to opt in to this, e.g.
+by defining a version of \li{Relit.read_to} from Fig.~\ref{fig:relit-util} for both grammars.
+
+% The system interprets the 
+% lexer and parser names that appear in TLM definitions, e.g.
+\li{RegexLexer} and \li{RegexParser} in Fig.~\ref{fig:regex-tlm-def}, as if they were module paths at the top level of the compilation unit.
+The modules must have already been compiled to OCaml bytecode by the time the compiler reaches the TLM definition (because ``compile time'' for a program that uses TLMs is ``run-time'' for the lexer and parser).
+Concretely, the system will look for \li{RegexLexer.cmo} and \li{RegexParser.cmo} on the build path.
+The lexer module must have the same interface as modules generated by the \li{ocamllex} tool, and the parser must have the same interface as modules generated by Menhir \cite{menhir}\todo{cite menhir}{}, which is a mature parser generator based on Yacc \cite{johnson1975yacc,TarditiDR:mly}.
+Reason is itself implemented using Menhir.
+The most straightforward way to satisfy these requirements is, of course, to use these tools.
+However, the TLM mechanism never looks at the original lexer specification, i.e.
+the \li{.mll} file, nor at the grammar itself, i.e.
+the \li{.mly} file, nor does it invoke these tools directly.
+This is the job of the build system of the user's choice.
+There are many popular build systems in the OCaml ecosystem.
+For the build system bundled with the Reason platform, \li{rebuild}, which is based on \li{ocamlbuild}, the \li{-use-menhir} flag will invoke \li{ocamllex} and Menhir as necessary.
+
+% Note that even when a TLM needs only a single helper function, it must be placed into a named module.
 This is to avoid needing to reason about the equivalence of arbitrary expressions when determining whether two signatures are compatible.
+
+% More specifically, the internal representation can be pretty printed as follows, where \li{start_pos} and \li{end_pos} are integer literals corresponding to the start and end position of the provided segment, and \li{ann_ty} is the provided type annotation after the references to the dependencies specified by the TLM have been appropriately resolved.
+% \begin{lstlisting}[numbers=none]
+%   ((raise ((start_pos, end_pos); failwith "<final expansion here>")) : ann_ty) 
+%     [@relit.spliced]
+% \end{lstlisting}
+% An exception with a type annotation will always have the specified type, so this internal representation allows the system to check the proto-expansion against the expansion type specified by the TLM before moving on to recursively generate the final expansion.
 
@@
 
-Unhygienic term rewriting systems like OCaml's PPX system and Template Haskell can be used to rewrite string literals, but these systems also do no adequately address any of these reasoning criteria.
-Hygienic macro systems, like those available in various Lisp-family languages and in Scala, are also unsuitable because the hygiene constraints prevent the macros from surfacing spliced expressions from literal bodies.
-The system of \emph{type-specific languages} (TSLs) developed in prior work by \citet{TSLs} has made perhaps the most headway toward these ideals, but it too does not adequately maintain these reasoning criteria.
-Moreover, it is not clear how to integrate this approach, which was developed for a simple monomorphic language with nominal types and local type inference, into ML, which supports structural types, parameterized types, abstract types, modules, pattern matching and non-local type inference.
-We will say more about these existing approaches in Sec.
-\ref{sec:existing-approaches}.
-For now, it suffices to say that after evaluating existing approaches, we found that none of them were suitable for integration into Reason.
-
-An approach that has made some headway toward this ideal is Omar et al.'s \emph{type-specific languages} (TSLs) \cite{TSLs}.
-A language that supports TSLs delegates control over the parsing and expansion of certain \emph{generalized literal forms} to a parser associated with the type that the form is being checked against.
-So if we add TSLs to our language, our example then requires a type annotation on \li{y}:
-\begin{lstlisting}[numbers=none,basicstyle=\ttfamily\fontsize{9pt}{1em}\selectfont]
-  let y : KQuery.t = `((!R)@&{&/x!/:2_!x}'!R)`
-\end{lstlisting}
-The parser associated with the type \li{KQuery.t} is statically invoked to lex, parse and expand the literal body, i.e.
-the sequence of characters between the outer delimiters, here \li{`(} and \li{)`}.
-The literal body is constrained by the context-free grammar of the language only in that nested delimiters must be matched (like comments in OCaml).
-TSLs are closely related to the mechanism that we will propose, so let us analyze TSLs from the perspective of the six reasoning criteria just outlined.
-\begin{enumerate}[leftmargin=12px]
-  \item[\LEFTcircle] \textbf{Responsibility}: The type-directed dispatch mechanism allows the client to easily determine which parser is responsible for each literal form: the parser that was associated with the type when it was defined.
-Clients do not need to worry about different TSLs conflicting syntactically because the context-free grammar of the language remains fixed and composition is mediated by splicing.
-The main limitation here is that it is impossible to define literal notation after a type has been defined, and it is also impossible to define multiple notations at a single type.
-  \item[\LEFTcircle] \textbf{Segmentation}: The parser can splice expressions out of the literal body, but there is no clear way to associate each spliced term with some particular segment (i.e.
-subsequence) of the literal body, nor is there a guarantee that spliced terms are non-overlapping.
-As such, there is no way to indicate to the programmer where base language expressions are located within a literal form.
-  \item[\LEFTcircle] \textbf{Capture}: The TSL mechanism as described enforces complete capture avoidance (though the formal specification given in the paper did not adequately enforce this constraint).
-  \item[\LEFTcircle] \textbf{Context Dependence}: The TSL mechanism as described requires that the expansion be completely closed, so it is trivially context independent.
-However, this comes at a significant expressive cost: the generated expansions cannot make use of any libraries whatsoever.
-  \item[\LEFTcircle] \textbf{Typing}: The expansion must necessarily be of the associated type, so abstract reasoning about the type of the expansion is straightforward.
-However, there is no easy way to determine the type expected for each spliced expression.
-In addition, the prior work considered only a monomorphic, nominally-typed language with local type inference, leaving open a number of problems that came up as we considered integrating TSLs into Reason/OCaml:
-\begin{enumerate}[leftmargin=15px,nolistsep,noitemsep]
-  \item \textbf{Structural Types}: There is no way to define literal notation at structural types, e.g. tuple and arrow types, because there is no ``definition site'' for such types.
-  \item \textbf{Parameterized \& Abstract Types}: There is no way to define literal notation over a type-parameterized family of types, e.g.
-at all types \li{list('a)}.
-Similarly, there is no way to define literal notation over a module-parameterized family of abstract types, e.g.
-at every abstract type defined by a module implementing the \li{QUEUE} signature.
-Parameterized and abstract type families are ubiquitous in ML-family languages.% Programmers in ML-family languages commonly define various implementations of abstract data types in this way.
-  \item \textbf{Pattern Literals}: Pattern matching is ubiquitous in ML-family languages but the prior work on TSLs considered only expression literals.
-%Pattern literals are dual to expression literals in that expression literals support value construction, while pattern literals support value deconstruction.
-  \item \textbf{ML-Style Type Inference}: It is not clear that a type-directed dispatch scheme can be cleanly  reconciled with ML-style type inference.
-\end{enumerate}
-\end{enumerate}
-
-\subsection{Syntax Definition Systems}\label{sec:syntax-dialects-intro}
-One approach available to library providers seeking to introduce new literal forms is to use a syntax definition system to construct a library-specific \emph{syntax dialect}: a new syntax definition that extends the syntax definition given in the language definition with new forms, including literal forms.% For example, Ur/Web's syntax (Figure 1) is a library-specific dialect of Ur's syntax \cite{conf/pldi/Chlipala10,conf/popl/Chlipala15}.
-
+In this section, we will give an overview of the many existing mechanisms that library providers might use to define new notation, and more specifically, new literal notation.
+Rather than evaluating these mechanisms by asking ``how much syntactic power do they give the library provider?'', however, our approach is to ask ``what do these mechanisms quite reasonably \textbf{not} allow a library provider to express?'', using the rubric of six reasoning principles that this paper has developed.
+
+\paragraph{Syntax Definition Systems}\label{sec:syntax-dialects-intro}
+One approach available to library providers seeking to introduce new literal forms is to use a syntax definition system to extend the syntax of an existing language directly.
+% For example, Ur/Web's syntax (Figure 1) is a library-specific dialect of Ur's syntax \cite{conf/pldi/Chlipala10,conf/popl/Chlipala15}.
  % Such dialects are sometimes qualitatively taxonomized as amongst the ``domain-specific language'' for this reason \cite{fowler2010domain}.
-%Syntactic cost is often assessed qualitatively \cite{green1996usability}, though quantitative metrics can be defined.
+%Syntactic cost is often assessed qualitatively \cite{lavender1996usability}, though quantitative metrics can be defined.
 % Syntax definition systems , which we will discuss in Sec.
@@
 ``domain-specific'') syntax dialects like Ur/Web, and have thereby contributed to their ongoing proliferation.
-
 % Many have argued that a proliferation of syntax dialects constructed using these tools is harmless or even desirable, because programmers can simply choose the right syntax dialect for each job at hand \cite{journals/stp/Ward94}.
 However, we argue that in fact
-There are hundreds of syntax definition systems of various design.
-Notable examples include grammar-oriented systems like Camlp4 \cite{ocaml-manual}, Copper \cite{conf/gpce/WykS07,conf/pldi/SchwerdfegerW09} and Sugar*/SoundExt \cite{erdweg2011sugarj,erdweg2013framework,conf/icfp/LorenzenE13,conf/popl/LorenzenE16}, as well as parser combinator systems \cite{Hutton1992d}.
-The parsers generated by these systems can be invoked to preprocess program text in various ways, e.g.
-by invoking them from within a build script, by using a preprocessor-aware build system (e.g.
-\li{ocamlbuild}), or via language-integrated preprocessing directives, e.g.
-Racket's \li{#lang} directive or its reader macros \cite{Flatt:2012:CLR:2063176.2063195}, or the import mechanism of SugarJ \cite{erdweg2011sugarj}.% the directives of language-integrated ``mixfix'' systems \cite{wieland2009parsing,missura1997higher,5134} like Coq \cite{Coq:manual}.)
+There are hundreds of syntax definition systems of various design, and the parsers generated by these systems can be invoked to preprocess program text in various ways, e.g.
+by invoking them from within a build script, by using a preprocessor-aware compiler (e.g.
+\li{ocamlc}), or via language-integrated preprocessing directives, e.g.
+the import mechanism of SugarJ \cite{erdweg2011sugarj}, or Racket's \li{#lang} directive \cite{Flatt:2012:CLR:2063176.2063195}.
+%or its reader macros\todo{citation} \cite{Flatt:2012:CLR:2063176.2063195}.% the directives of language-integrated ``mixfix'' systems \cite{wieland2009parsing,missura1997higher,5134} like Coq \cite{Coq:manual}.)
 %library-specific (a.k.a. ``domain-specific'')
 
-The problem was described by example in Sec.
-\ref{sec:intro}: these systems make it difficult to reason abstractly.
-Let us reitierate more generally, before discussing a few systems that are exceptional along some strict subset of these dimensions:
-\begin{enumerate}[noitemsep,nolistsep,leftmargin=10pt]
-\item \textbf{Responsibility}: Clients using a combined syntax dialect cannot easily determine which constituent extension is responsible for a given form, whereas TLMs have explicit names which follow the usual scoping conventions.
-
-Moreover, there can be syntactic conflicts because multiple extensions can claim responsibility for the same form.
-TLMs sidestep these complexities entirely because the context-free syntax of the language is fixed, and composition is via splicing rather than direct combination.
-\item \textbf{Segmentation}: Clients of a syntax dialect cannot accurately determine which segments of the program text appear directly in the desugaring.
-In contrast, TLM clients can inspect the inferred segmentation (communicated via secondary notation.)
-\item \textbf{Capture}: Unlike TLM clients, clients of a syntax dialect cannot be sure that spliced terms are capture avoiding. 
-\item \textbf{Context Dependence}: Similarly, clients cannot be sure that the desugaring is context independent.
-Indeed, without a method to pass in parameters (Sec.
-\ref{sec:ptsms}), achieving strict context independence would be impractical.
-\item \textbf{Typing}: Clients of a syntax dialect cannot reason abstractly about what type a desugaring has.
-In contrast, TLM clients can determine the type of any expansion by referring to the parameter and type declarations on the TLM definition, and nothing else.
-The inferred segmentation also gives types for each spliced expression or pattern.
-%This information can be communicated upon request by the type inspection service of a program editor.
-\end{enumerate}
-
-% The first problem with using a syntax definition systems to directly extend the context-free syntax of a language is that clients cannot always deterministically combine the resulting extended syntax dialects when they want to use the new forms that they define together, i.e.
-there can be syntactic conflicts.
-This is a significant problem because large programs use many separately developed libraries \cite{DBLP:conf/sac/LammelPS11}.
-
-An extensible syntax definition system that has confronted the problem of \textbf{Responsibility} (but not the other problems) is Copper \cite{conf/pldi/SchwerdfegerW09}.
-Copper integrates a modular grammar analysis that guarantees that determinism is conserved when extensions of a certain restricted class are combined.
+These systems give a large amount of syntactic power to the notation provider, but this comes at a cost: with few exceptions, these systems make it difficult to reason abstractly about the six topics from Sec.~\ref{sec:intro}.
+Rather than reiterating the points made there, let us focus on the exceptional cases.
+
+A grammar extension system that has confronted the problem of \textbf{Responsibility} (but not the other problems) is Copper \cite{conf/pldi/SchwerdfegerW09,schwerdfeger2010context}.
+Copper performs a modular grammar analysis that guarantees that determinism is conserved (i.e.
+ambiguities are not possible) when extensions of a certain restricted class are combined.
 The caveat is that the constituent extensions must prefix all newly introduced forms with marking tokens drawn from disjoint sets.
-To be confident that the marking tokens used are disjoint, providers must base them on the domain name system or some other coordinating entity.
+To be confident that the marking tokens used are disjoint, providers must base them on, for example, the domain name system.
 Because the mechanism operates at the level of the context-free grammar, it is difficult for the client to define scoped abbreviations for these verbose marking tokens.
-TLMs can be abbreviated (Sec.
-\ref{sec:ptsms}).
+TLMs can, in contrast, be abbreviated and distributed within modules following the usual scoping structure of the language.
+Composition is via splicing, rather than direct grammar composition.
 
 Some programming languages, notably including theorem provers like Coq \cite{Coq:manual} and Agda \cite{norell2007towards}, support ``mixfix'' notation directives \cite{wieland2009parsing,missura1997higher,5134}.
-Many of these systems enforce capture avoidance and application-site context independence \cite{5134,DBLP:conf/gpce/TahaJ03,Coq:manual,DBLP:conf/ifl/DanielssonN08}.
+Many of these systems enforce \textbf{Capture Avoidance} and application-site \textbf{Context Independence} \cite{5134,DBLP:conf/gpce/TahaJ03,Coq:manual,DBLP:conf/ifl/DanielssonN08}.
 The problem is that mixfix notation requires a fixed number of sub-trees, e.g.
 \li{if _ then _ else _}.
-Coq has some limited extensions for list-like literals \cite{Coq:manual}.
+Coq has some \emph{ad hoc} extensions for list-like literals \cite{Coq:manual}.
 These systems cannot express the example literal forms from this paper, because they can have any number of spliced terms.
 
-The work of \citet{conf/icfp/LorenzenE13,conf/popl/LorenzenE16} introduces SoundExt, a grammar-based syntax extension system where extension providers can equip their new forms with derived typing rules.
-The system then attempts to automatically verify that the expansion logic (expressed using a rewrite system, rather than an arbitrary function) is sound relative to these derived rules.
-TLMs differ in several ways.
-First, as already discussed, we leave the context-free syntax fixed, so different TLMs cannot conflict.
-Second, SoundExt does not enforce hygiene, i.e.
+Racket allows \li{#lang} definitions and reader macros to gain complete control over lexing and parsing the remainder of a file \cite{Flatt:2012:CLR:2063176.2063195}.
+However, this flexibility makes it difficult to reason about segmentation and even responsibility (''is a reader macro active?'') Providers can optionally generate fresh variables to avoid capture, but this is not enforced.
+We say more about macros below.
+
+\citet{conf/icfp/LorenzenE13,conf/popl/LorenzenE16}'s SoundExt is a grammar-based syntax extension system where extension providers can equip their new forms with derived typing rules.
+The system then attempts to automatically verify that the expansion logic, expressed using a rewrite system, rather than an arbitrary function, is sound relative to these derived rules, so it is possible to reason about \textbf{Expansion Typing}.
+SoundExt does not enforce hygiene, i.e.
 expansions might depend on the context and intentionally induce capture.
-Similarly, there is no abstract segmentation discipline.
-A client can only indirectly reason about binding (but not segmentation) by inspecting the derived typing rules.
+A client can only indirectly reason about binding by inspecting the derived typing rules.
+There is no abstract segmentation discipline.
 Unlike TLMs, SoundExt supports type-dependent expansions \cite{conf/popl/LorenzenE16}.
-The trade-off is that TLMs can generate expansions, and therefore segmentations, even when the program is ill-typed.
-Another important distinction is that TLMs rely on proto-expansion validation, rather than verification as in SoundExt.
-The trade-off is that TLMs do not require that the expansion logic be written using a restricted rewriting system, nor does the system require a fully mechanized language definition.
-Finally, there is no clear notion of ``partial application'' in SoundExt or other syntax definition systems.
+The trade-off is that TLMs can generate proto-expansions, and therefore segmentations, even when the spliced expressions are ill-typed.
+Another important distinction is that TLMs rely on proto-expansion validation, rather than verification as in SoundExt (though providers can use MetaOCaml and related systems to reason about typing, as described in Sec.
+\ref{sec:metaocaml}).
+The trade-off is that TLMs do not require a fully mechanized host language definition.
+Finally, there is no clear notion of parameterization or partial application in SoundExt or other syntax definition systems, so it would be difficult to define notation for, for example, all finite map implementations as we demonstrated in Sec.~\ref{sec:ptsms}.
 
@@
 
-\subsection{Term Rewriting Systems}
+\paragraph{Term Rewriting Systems}
 Another approach -- and the approach that TLMs are rooted in -- is to leave the context-free syntax of the language fixed, and instead contextually rewrite existing literal forms.
-
-OCaml's textual syntax now includes \emph{preprocessor extension (ppx) points} used to identify terms that some external term rewriting preprocessor must rewrite \cite{ocaml-manual}.
-We could  mark a string literal as follows:
+For example, OCaml's textual syntax now includes \emph{preprocessor extension (PPX) points} used to identify terms that some external term rewriter will rewrite \cite{ocaml-manual}.
+For example, we could  mark a string literal as follows:
 \begin{lstlisting}[numbers=none]
-    [%xml "SSTR<h1>Hello, {[first_name]}!</h1>ESTR"]
+    [%xml "<h1>Hello, {[first_name]}!</h1>"]
 \end{lstlisting}
-More than one applied preprocessor might recognize this annotation (there are, in practice, many XML/HTML libraries), so the problems of \textbf{Responsibility} comes up.
-It is also impossible to reason abstractly about \textbf{Segmentation}, \textbf{Capture}, \textbf{Context Dependence} and \textbf{Typing} because the code that the preprocessor generates is unconstrained.% For these reasons, users of this mechanism warn that they should be used sparingly.\todo{citation}
-
-Term-rewriting macro systems are language-integrated local term rewriting systems that require that the client explicitly apply the intended rewriting, implemented by a macro, to the term that is to be rewritten.
-This addresses the issue of \textbf{Responsibility}.
-However, unhygienic, untyped macro systems, like the earliest variants of the Lisp macro system \cite{Hart63a}, Template Haskell \cite{SheardPeytonJones:Haskell-02} and GHC's quasiquotation system \cite{mainland2007s} (which is based on Template Haskell), do not allow clients to reason abstractly about the remaining issues, again because the expansion that they produce is unconstrained.
-(It is not enough that with Template Haskell / GHC quasiquotation, the generated expansion is typechecked -- to satisfy the \textbf{Typing} criterion, it must be possible to reason abstractly about \emph{what the type of the generated expansion is}.)
+This does help with reasoning about responsibility, but technically more than one applied preprocessor might recognize this annotation (there are, in practice, many XML/HTML libraries), and annotations do not follow scoping rules.
+It is impossible to reason abstractly about the other issues because the code that the preprocessor generates is unconstrained.% For these reasons, users of this mechanism warn that they should be used sparingly.\todo{citation}
+
+Term-rewriting macro systems require that the client explicitly apply the intended rewriting, implemented by a scoped macro, to the term that is to be rewritten.
+This addresses the issue of \textbf{Responsibility} more thoroughly.
+However, unhygienic, untyped macro systems, like the earliest variants of the Lisp macro system \cite{Hart63a}, and, more recently, Template Haskell \cite{SheardPeytonJones:Haskell-02} and GHC's quotation system \cite{mainland2007s} (which is based on Template Haskell), do not allow clients to reason abstractly about the remaining issues, again because the expansion that they produce is unconstrained.
+There is typically some way for library providers to opt in to a capture avoidance discipline by asking for fresh variables, but this is not enforced.
+Note that it is not enough that with Template Haskell / GHC quotation, the generated expansion is typechecked---to satisfy the \textbf{Expansion Typing} criterion, it must be possible to reason abstractly about \emph{what the type of the generated expansion is}.
 
 \emph{Hygienic} macro systems prevent, or abstractly account for \cite{DBLP:conf/esop/HermanW08,Herman10:Theory}, \textbf{Capture}, and they enforce application-site \textbf{Context Independence} \cite{Kohlbecker86a,DBLP:conf/popl/Adams15,DBLP:conf/popl/ClingerR91,DBLP:journals/lisp/DybvigHB92}.
-The critical problem is that a standard hygiene discipline makes it impossible to repurpose string literal forms to introduce compositional literal forms at other types.
-Consider again our running XHTML example, which we might try to realize by applying a hygienic macro, \li{xml!}, to a string literal that the macro parses:
-\begin{lstlisting}[numbers=none]
-    (xml! "SSTR<h1>Hello, {[first_name]}!</h1>ESTR")
-\end{lstlisting}
-The expansion of this macro will fail the check for application-site context independence because \li{first_name} will appear as a free variable, in no way different from any other.
-The hygiene mechanism for TLMs addresses this problem by explicitly distinguishing spliced segments of the literal body in the proto-expansion.
-This also addresses the problem of \textbf{Segmentation} -- the segmentation abstractly communicates the fact that (only) \li{first_name} is a spliced \li{string} expression.
+The critical problem, detailed in Sec.~\ref{sec:context-dependence}, is that the standard context independence discipline makes it impossible to repurpose string literal forms to introduce compositional literal forms at other types.
+
+TLMs also consider the problem of definition-site context independence by specifying expansion dependencies explicitly.
+This sidesteps problems faced in prior attempts to integrate macros into module systems \cite{culpepper2005syntactic} related to ``smuggling'' definition site values to application sites, without violating the abstraction discipline of the module system.
+It also prevents issues related to cross-stage persistence, since macro definitions do not refer to definition-site values directly but rather program against the dependency signature.
 
 Much of the research on macro systems has been for languages in the LISP tradition \cite{mccarthy1978history} that do not have rich static type structure.
-The formal macro calculus studied by \citet{DBLP:conf/esop/HermanW08} (which is not capable of expressing new literal forms, for the reasons just discussed) uses types only to encode the binding structure of the generated expansion (as discussed in Sec.
-\ref{sec:capture}).
-Research on typed \emph{staging macro systems} like MetaML \cite{Sheard:1999:UMS}, MetaOCaml \cite{DBLP:conf/flops/Kiselyov14} and MacroML \cite{ganz2001macros} is also not directly applicable to the problem of defining new literal forms -- the syntax tree of the arguments cannot be inspected at all (staging macros are used mainly for partial evaluation and performance-related reasons.)
-
-The Scala macro system is a hygienic macro system.
-Its ``black box'' macros support reasoning abstractly about \textbf{Typing} because type annotations constrain the macro arguments and the generated expansions, though the precise reasoning principles available are unclear because the Scala macro system has not been formally specified.
-The full calculus we have defined is the first detailed type-theoretic accounts of a typed, hygienic macro system of any design for an ML-like language, i.e.
-one with a rich static type system, support for pattern matching, type functions and ML-like modules.
+The formal macro calculus studied by \citet{DBLP:conf/esop/HermanW08} (which is not capable of expressing new literal forms, for the reason just discussed) uses types only as a technical advice in reasoning about the binding structure of the generated expansion.
+The Scala macro system does support reasoning abstractly about \textbf{Expansion Typing} due to the return type annotations.
+The full calculus we have defined is the first detailed type-theoretic account of a typed, hygienic macro system of any design for an ML- or Scala-like language, i.e.
+one with a rich static type structure and support for pattern matching.
+Many of the basic mechanisms would be relevant even if support for parsing literal bodies was removed and replaced with term rewriting.
+Segmentations can be considered a refinement of the tree paths in prior work \cite{Herman10:Theory,gorn1965explicit}.
+
+Research on typed \emph{staging macro systems} \cite{DBLP:conf/popl/DaviesP96} like MetaML \cite{Sheard:1999:UMS}, MetaOCaml \cite{DBLP:conf/flops/Kiselyov14}, MacroML \cite{ganz2001macros,DBLP:conf/gpce/TahaJ03} and Typed Template Haskell \cite{tth} is not directly applicable to the problem of defining new literal forms because the syntax tree of the arguments cannot be inspected at all (staging macros are used mainly for optimization).
+Sec.~\ref{sec:metaocaml} discussed how the typed quotations from these systems can help TLM providers reason about the type-correctness of their parsers.
+Future work could help address some of the mentioned limitations.
 
@@
 \emph{string interpolation}) forms, or similar but more general \emph{fragmentary quotation forms} \cite{conf/icfp/Slind91}, e.g.
-SML/NJ.
+SML/NJ \cite{SML/Quote}.
 These designate a particular delimiter to escape out into the expression language.
 The problem with using these together with macros as vehicles to introduce literal forms at various other types is 1) there is no ``one-size-fits-all'' escape delimiter, and 2) typing is problematic because every escaped term is checked against the same type.
-In the HTML example, we have splicing at two different types using two different delimiters.
-These forms also cannot appear in patterns.%In general, e.g.
+For example, in Fig.~\ref{fig:urweb}, we have splicing at two different types.
+%These forms also cannot appear in patterns.
+Scala's pattern matching machinery incorporates a dynamic metaobject protocol that can approximate pattern literal notation in some cases, though this makes it difficult to reason about exhaustiveness and redundancy.
+%In general, e.g.
 when defining syntax for a programming language with many sorts of terms, the most appropriate choice of delimiters might depend on where each spliced term appears.
 
-This brings us back to the most closely related work, that of \citet{TSLs} on \emph{type-specific languages} (TSLs).
+This brings us to the most closely related work, that of \citet{TSLs} on \emph{type-specific languages} (TSLs).
 Like simple expression TLMs (Sec.
-\ref{sec:setsms}), TSLs allow library providers to programmatically control the parsing of expressions of generalized literal form.
+\ref{sec:setlms}), TSLs allow library providers to programmatically control the parsing of expressions of generalized literal forms.
 With TSLs, parse functions are associated directly with nominal types and invoked according to a bidirectionally typed protocol.
 In contrast, TLMs are separately defined and explicitly applied.
-Accordingly, different TLMs can operate at the same type, and can operate at any type, including structural types.
+Accordingly, different TLMs can operate at the same type, and they can operate at any type, including structural types.
 In a subsequent short paper, \citet{sac15} suggested explicit application of simple expression TLMs  also in a bidirectional typed setting \cite{Pierce:2000:LTI:345099.345100}, but this paper did not have any formal content.
 With TLMs, it is not necessary for the language to be bidirectionally typed (see Sec.
-\ref{sec:typing-e} on type inference).% It should be possible to implicitly invoke TLMs based on the expected type in situations where the inference engine had enough information, but we leave this as future work.
-
-Perhaps most importantly, the metatheory presented by \citet{TSLs} establishes only that generated expansions are of the expected type (i.e.
+\ref{sec:segment-typing} on ML-style type inference).
+% It should be possible to implicitly invoke TLMs based on the expected type in situations where the inference engine had enough information, but we leave this as future work.
+The metatheory presented by \citet{TSLs} establishes only that generated expansions are of the expected type (i.e.
 a variant of the Typed Expression Expansion theorem from Sec.
-\ref{sec:s-metatheory}.) It does not establish the remaining abstract reasoning principles that have been the major focus of this paper.
-In particular, there is no formal hygiene theorem and indeed the formal system in the paper does not correctly handle substitution or capture avoidance, issues we emphasized because they were non-obvious in Sec.
+\ref{sec:s-metatheory}), though the induction principle used in the proof is not clear.
+It does not establish the remaining abstract reasoning principles that have been the major focus of this paper.
+There appears to be a context independence condition but it does not allow for any dependencies whatsoever, which is unreasonably restrictive.
+There is no formal hygiene theorem and indeed the formal system in the paper does not correctly handle substitution or capture avoidance, issues we emphasized because they were non-obvious in Sec.
 5.
-Moreover, the TLM does not guarantee that a valid segmentation will exist, nor associate types with segments.
-
-Finally, the prior work did not consider pattern matching, type functions, ML-style modules, parameters or static evaluation. This paper addresses all of these.
+Moreover, the TLM does not guarantee that a valid segmentation will exist, nor associate types with spliced segments.
+Finally, this prior work did not consider pattern matching, module system integration or type- or module-parameterized type families.
+This paper addresses all of these, resulting in a design suitable for integration into Reason/OCaml.
 
@@
 This class includes all of the examples enumerated in Sec.
-\ref{sec:intro} (up to the choice of outermost delimiter), the case studies detailed in this paper and in the supplement, and the examples collected from the empirical study by \citet{TSLs}.
+\ref{sec:intro} (up to the choice of outermost delimiter), the varied and non-trivial case studies outlined in this paper and in the supplement, and the examples collected from the empirical study by \citet{TSLs}.
+We anticipate many more interesting examples will emerge as the Reason community explores the mechanism.
 
@@
 The reasoning principles that TLMs provide, which are the primary contributions of this paper, allow clients to "reason around" poor literal designs, using principles analagous to those already familiar to programmers in languages like ML and Scala.
+Although we emphasized integration with modules, our formal system demonstrates that modules are not necessary to capture the fundamental ideas in this paper.
+We intend this paper to be useful to the designers of languages far from the ML tradition.
+More generally, we intend for the ``reasoning principles first'' approach advocated in this paper to be a useful intellectual framework for language designers considering the merits of other ``conveniences''.
 % We must leave a detailed empirical evaluation of the impact of particular TLMs on various quality attributes, like programmer productivity and program comprehensibility, as future work.
@@
 
-% Haskell's \li{do}-notation at types with monadic structure cannot be expressed directly using TLMs because it needs to purposefully induce capture in spliced sub-terms, and TLMs prevent capture.
-This limitation could be resolved by allowing identifiers that appear in literal bodies to be explicitly designated as bound within spliced expressions in the segmentation.
+% his limitation could be resolved by allowing identifiers that appear in literal bodies to be explicitly designated as bound within spliced expressions in the segmentation.
 The prior work by \citet{DBLP:conf/esop/HermanW08} considered a similar approach in a setting where the locations of sub-terms is known ahead of time (building on the \emph{tree locations} of \citet{gorn1965explicit}, which are conceptually similar to our spliced segment locations.)  We leave as future work the details of this proposal and a consideration of whether this additional expressive power would justify the increased reasoning complexity for clients (it is presently the author's opinion that it would not.)
 
-A correct parse function never returns an encoding of a proto-expansion that fails validation given well-typed splices, but this invariant cannot be enforced by the ML type system.
-% Using a proof assistant, it would be possible to verify that
-% a parse function generates only encodings of valid proto-expansions. 
-Under a
-richer type system, the return type of the parse function itself could be refined so as to
-enforce this invariant intrinsically.
+\paragraph{Limitations} Not all interesting properties of a program will, in general, be apparent from its type, particularly in a language like OCaml.
+As usual, programmers will sometimes need to peek behind the abstraction boundaries or rely on informal documentation to understand, in detail, what a literal notation is doing (e.g.
+with respect its equational properties, its side effects and so on).
+TLMs in languages with more expressive type systems would be commensurately more reasonable.
+
+Notation that intentionally introduces bindings into spliced terms, like Haskell's \li{do}-notation at types equipped with monadic structure, cannot be expressed using TLMs as defined in this paper, because splicing is capture avoiding.
+Although we considered weakening the capture avoidance condition in various somewhat reasonable ways, e.g.
+by communicating which identifiers are explicitly captured by each spliced segment, it is quite difficult to communicate this structure to client programmers even with tool support.
+Given that \li{do}-notation is already general and comes equipped with well-understood reasoning principles, the most reasonable approach is perhaps to  build it in primitively, suitably parameterized over the implementation of the \li{MONAD} signature.
+
+% A correct parse function never returns an encoding of a proto-expansion that fails validation given well-typed splices, but this invariant cannot be enforced by the ML type system.
+% % Using a proof assistant, it would be possible to verify that
+% % a parse function generates only encodings of valid proto-expansions. 
+% Under a
+% richer type system, the return type of the parse function itself could be refined so as to
+% enforce this invariant intrinsically.
 This problem -- of typed first-class typed term representations -- has been studied in a variety of settings, e.g.
@@
 % We leave integration of these approaches into the TLM mechanism as future work.
- Our present efforts aim to leave the semantics of OCaml unchanged.
+%  Our present efforts aim to leave the semantics of OCaml unchanged.
 
 %Another topic of interest has to do with intentional capture. 
-Another direction has to do with automated refactoring.
+Another future direction has to do with automated refactoring.
 The unexpanded language does not come with context-free notions of renaming and substitution.
@@
 At several points in the paper, we allude to editor integration.
-However, several important questions having to do with TLM-specific syntax highlighting, incremental parsing and error recovery \cite{graham1979practical} remain to be considered, and indeed these are our biggest remaining implementation challenges.
-% Another interesting direction would be to generalize TLMs to support non-textual notation in the setting of a structure editor.
-The semantic foundations for structure editors proposed recently by \citet{DBLP:conf/popl/OmarVHAH17} may guide such an effort.
+However, several important questions having to do with TLM-specific syntax highlighting, incremental parsing and error recovery \cite{graham1979practical} remain to be considered.
+Another interesting direction would be to generalize TLMs to support non-textual notation in the setting of a structure editor \cite{DBLP:conf/popl/OmarVHAH17}.
 
@@
 %This paper serves to guide those who hope to incorporate TLMs into their future language designs.
+
+% \newcommand{\concSec}{Conclusion}
+% \section{\protect\concSec}
+% \label{sec:conc}
+
+\paragraph{Concluding Remarks} 
+To conclude, let us briefly reiterate the key ideas of this paper.
+The focus was on ensuring that client programmers can follow simple protocols when they have questions about the syntactic structure, type structure or binding structure of a program.
+In answering these sorts of questions,  the client is not made to look at the generated expansion itself, nor inspect the parser implementation.
+Instead, the programmer need only be given knowledge of the expansion type from the TLM definition and the segmentation inferred by the expander at each application site, which carries a small volume of information that can be communicated straightforwardly by standard editor services.
+Certain questions related to the binding structure simply do not need to be asked due to the strict context independence and capture avoidance discipline of the system, enabled by the explicit tracking of dependencies, and of spliced segments.
+
+Despite these semantic constraints, the system is able to express a number of non-trivial examples with few compromises because there is only one simple lexical constraint on literal bodies.
+The mechanism integrates cleanly into Reason/OCaml, with full support for its type system, module system and pattern matching system.
+TLM providers can use existing, mature parser generators and parse tree representations, and can opt-in to a stronger typing discipline by using various implementations of MetaOCaml.
+Overall, we believe that TLMs represent a distinctly \emph{reasonable} and \emph{expressive} new point in the design space of literal notation definition systems.
+
+% Unhygienic term rewriting systems like OCaml's PPX system and Template Haskell can be used to rewrite string literals, but these systems also do no adequately address any of these reasoning criteria.
+Hygienic macro systems, like those available in various Lisp-family languages and in Scala, are also unsuitable because the hygiene constraints prevent the macros from surfacing spliced expressions from literal bodies.
+The system of \emph{type-specific languages} (TSLs) developed in prior work by \citet{TSLs} has made perhaps the most headway toward these ideals, but it too does not adequately maintain these reasoning criteria.
+Moreover, it is not clear how to integrate this approach, which was developed for a simple monomorphic language with nominal types and local type inference, into ML, which supports structural types, parameterized types, abstract types, modules, pattern matching and non-local type inference.
+We will say more about these existing approaches in Sec.
+\ref{sec:existing-approaches}.
+For now, it suffices to say that after evaluating existing approaches, we found that none of them were suitable for integration into Reason.
+
+% An approach that has made some headway toward this ideal is Omar et al.'s \emph{type-specific languages} (TSLs) \cite{TSLs}.
+A language that supports TSLs delegates control over the parsing and expansion of certain \emph{generalized literal forms} to a parser associated with the type that the form is being checked against.
+So if we add TSLs to our language, our example then requires a type annotation on \li{y}:
+% \begin{lstlisting}[numbers=none,basicstyle=\ttfamily\fontsize{9pt}{1em}\selectfont]
+%   let y : KQuery.t = `((!R)@&{&/x!/:2_!x}'!R)`
+% \end{lstlisting}
+% The parser associated with the type \li{KQuery.t} is statically invoked to lex, parse and expand the literal body, i.e.
+the sequence of characters between the outer delimiters, here \li{`(} and \li{)`}.
+The literal body is constrained by the context-free grammar of the language only in that nested delimiters must be matched (like comments in OCaml).
+TSLs are closely related to the mechanism that we will propose, so let us analyze TSLs from the perspective of the six reasoning criteria just outlined.
+% \begin{enumerate}[leftmargin=12px]
+%   \item[\LEFTcircle] \textbf{Responsibility}: The type-directed dispatch mechanism allows the client to easily determine which parser is responsible for each literal form: the parser that was associated with the type when it was defined.
+Clients do not need to worry about different TSLs conflicting syntactically because the context-free grammar of the language remains fixed and composition is mediated by splicing.
+The main limitation here is that it is impossible to define literal notation after a type has been defined, and it is also impossible to define multiple notations at a single type.
+%   \item[\LEFTcircle] \textbf{Segmentation}: The parser can splice expressions out of the literal body, but there is no clear way to associate each spliced term with some particular segment (i.e.
+subsequence) of the literal body, nor is there a guarantee that spliced terms are non-overlapping.
+As such, there is no way to indicate to the programmer where base language expressions are located within a literal form.
+%   \item[\LEFTcircle] \textbf{Capture}: The TSL mechanism as described enforces complete capture avoidance (though the formal specification given in the paper did not adequately enforce this constraint).
+%   \item[\LEFTcircle] \textbf{Context Dependence}: The TSL mechanism as described requires that the expansion be completely closed, so it is trivially context independent.
+However, this comes at a significant expressive cost: the generated expansions cannot make use of any libraries whatsoever.
+%   \item[\LEFTcircle] \textbf{Typing}: The expansion must necessarily be of the associated type, so abstract reasoning about the type of the expansion is straightforward.
+However, there is no easy way to determine the type expected for each spliced expression.
+In addition, the prior work considered only a monomorphic, nominally-typed language with local type inference, leaving open a number of problems that came up as we considered integrating TSLs into Reason/OCaml:
+% \begin{enumerate}[leftmargin=15px,nolistsep,noitemsep]
+%   \item \textbf{Structural Types}: There is no way to define literal notation at structural types, e.g. tuple and arrow types, because there is no ``definition site'' for such types.
+%   \item \textbf{Parameterized \& Abstract Types}: There is no way to define literal notation over a type-parameterized family of types, e.g.
+at all types \li{list('a)}.
+Similarly, there is no way to define literal notation over a module-parameterized family of abstract types, e.g.
+at every abstract type defined by a module implementing the \li{QUEUE} signature.
+Parameterized and abstract type families are ubiquitous in ML-family languages.% Programmers in ML-family languages commonly define various implementations of abstract data types in this way.
+%   \item \textbf{Pattern Literals}: Pattern matching is ubiquitous in ML-family languages but the prior work on TSLs considered only expression literals.
+%Pattern literals are dual to expression literals in that expression literals support value construction, while pattern literals support value deconstruction.
+%   \item \textbf{ML-Style Type Inference}: It is not clear that a type-directed dispatch scheme can be cleanly  reconciled with ML-style type inference.
+% \end{enumerate}
+% \end{enumerate}
 
@@
 There is not enough space to describe $\miniVerseParam$ with the same level of detail as in Sec.
-\ref{sec:setsms-formally}, so we highlight only the most important concepts below.
+\ref{sec:setlms-formally}, so we highlight only the most important concepts below.
 The details are in the supplement.

### 2018-03-19 icfp18/supplement/icfp18-supplement.tex
added supplement

\PassOptionsToPackage{svgnames,dvipsnames,svgnames}{xcolor}

\documentclass[12pt]{report}
\newif\ificfp
\icfptrue
\newcommand{\todolater}[1]{{\color{magenta} TODO (Later): #1}}
\newcommand{\todo}[1]{{\color{red} TODO: #1}}

% some useful packages
\usepackage{times}     % use times font for document
%\usepackage{lmodern}
\usepackage{newtxtt}
\usepackage{bbm}
%\renewcommand{\ttdefault}{txtt} % use txtt for typewriter font
\usepackage{mathpazo}
\usepackage{mathpartir} % use package for inference rules
\usepackage{upgreek} % package for alternative greek letters (\uppi)
\usepackage{fullpage}
\usepackage{colortab}
%\usepackage{graphicx}
\usepackage[labelfont=bf]{caption}

% \usepackage{cleveref}
%\usepackage{MnSymbol}
\usepackage{fancyvrb}
\usepackage{microtype}
\usepackage[framemethod=tikz]{mdframed}

% Get just llangle and rrangle from MnSymbol
\makeatletter
\DeclareFontFamily{OMX}{MnSymbolE}{}
\DeclareSymbolFont{MnLargeSymbols}{OMX}{MnSymbolE}{m}{n}
\SetSymbolFont{MnLargeSymbols}{bold}{OMX}{MnSymbolE}{b}{n}
\DeclareFontShape{OMX}{MnSymbolE}{m}{n}{
    <-6>  MnSymbolE5
   <6-7>  MnSymbolE6
   <7-8>  MnSymbolE7
   <8-9>  MnSymbolE8
   <9-10> MnSymbolE9
  <10-12> MnSymbolE10
  <12->   MnSymbolE12
}{}
\DeclareFontShape{OMX}{MnSymbolE}{b}{n}{
    <-6>  MnSymbolE-Bold5
   <6-7>  MnSymbolE-Bold6
   <7-8>  MnSymbolE-Bold7
   <8-9>  MnSymbolE-Bold8
   <9-10> MnSymbolE-Bold9
  <10-12> MnSymbolE-Bold10
  <12->   MnSymbolE-Bold12
}{}

\let\llangle\@undefined
\let\rrangle\@undefined
\DeclareMathDelimiter{\llangle}{\mathopen}%
                     {MnLargeSymbols}{'164}{MnLargeSymbols}{'164}
\DeclareMathDelimiter{\rrangle}{\mathclose}%
                     {MnLargeSymbols}{'171}{MnLargeSymbols}{'171}
\makeatother

% \usepackage{rotating}
% \usepackage{pdflscape}

\usepackage[colorlinks=true,allcolors=Blue,backref,pageanchor=true,plainpages=false, pdfpagelabels, bookmarks,bookmarksnumbered,
pdfborder={0 0 0},  %removes outlines around hyper links in online display
]{hyperref}

\usepackage{amsmath,amssymb, amsthm}

\allowdisplaybreaks[1]
\newtheorem{theorem}{Theorem}[chapter]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{assumption}[theorem]{Assumption}
\newtheorem{condition}[theorem]{Condition}

\usepackage{pfsteps}

\usepackage[noabbrev]{cleveref}

\newenvironment{proof-sketch}{\noindent{\emph{Proof Sketch.}}}{\qed}

% \makeatletter
% \renewenvironment{proof}[1][\proofname]{\par
%   \vspace{-\topsep}% remove the space after the theorem
%   \pushQED{\qed}%
%   \normalfont
%   \topsep0pt \partopsep0pt % no space before
%   \trivlist
%   \item[\hskip\labelsep
%         \itshape
%     #1\@addpunct{.}]\ignorespaces
% }{%
%   \popQED\endtrivlist\@endpefalse
%   \addvspace{6pt plus 6pt} % some space after
% }
% \makeatother
% \makeatletter
% \renewenvironment{proof-sketch}[1][\proofname]{\par
%   \vspace{-\topsep}% remove the space after the theorem
%   \pushQED{\qed}%
%   \normalfont
%   \topsep0pt \partopsep0pt % no space before
%   \trivlist
%   \item[\hskip\labelsep
%         \itshape
%     Proof Sketch\@addpunct{.}]\ignorespaces
% }{%
%   \popQED\endtrivlist\@endpefalse
%   \addvspace{6pt plus 6pt} % some space after
% }
% \makeatother
\usepackage{subcaption} %% For complex figures with subfigures/subcaptions

\usepackage{mathtools}
\usepackage{ stmaryrd }
\usepackage[numbers,sort]{natbib}

% \usepackage{subfigure}

% Approximately 1" margins, more space on binding side
%\usepackage[letterpaper,twoside,vscale=.8,hscale=.75,nomarginpar]{geometry}
%for general printing (not binding)
\usepackage[letterpaper,twoside,vscale=.8,hscale=.75,nomarginpar,hmarginratio=1:1]{geometry}

%% Listings
\definecolor{mygray}{rgb}{0.75,0.75,0.75}
\usepackage{listings}
\lstset{tabsize=2, 
basicstyle=\ttfamily\fontsize{8pt}{1em}\selectfont, 
commentstyle=\itshape\ttfamily\color{gray},
stringstyle=\ttfamily\color{purple},
mathescape=false,escapeinside={(~}{~)},
numbers=left, numberstyle=\scriptsize\color{mygray}, language=ML,showspaces=false,showstringspaces=false,xleftmargin=0pt, numbersep=-6pt, morekeywords=[1]{try,spliced,tyfam,opfam,let,fn,val,def,casetype,objtype,metadata,of,*,string,lexer,parser,datatype,new,toast,notation,module,switch,where,expansions,require,import,for,ana,syn,opcon,tycon,metasignature,metamodule,metasig,metamod,static,at,by,tycase,mod,macro,match,pattern,in,patterns,expressions,implicit,forall,int,exptsm,pattsm},deletekeywords={double,structure,of},classoffset=0, xleftmargin=0pt, 
aboveskip=3pt,belowskip=2pt,
moredelim=**[is][\color{red}]{SSTR}{ESTR},
moredelim=**[is][\color{Periwinkle}]{SHTML}{EHTML},
moredelim=**[is][\color{purple}]{SCSS}{ECSS},
moredelim=**[is][\color{brown}]{SSQL}{ESQL},
moredelim=**[is][\color{orange}]{SCOLOR}{ECOLOR},
moredelim=**[is][\color{magenta}]{SPCT}{EPCT}, 
moredelim=**[is][\color{gray}]{SNAT}{ENAT}, 
moredelim=**[is][\color{Periwinkle}]{SURL}{EURL},
moredelim=**[is][\color{Sealavender}]{SQT}{EQT},
moredelim=**[is][\color{Periwinkle}]{SGRM}{EGRM},
moredelim=**[is][\color{Yellowlavender}]{SID}{EID},
moredelim=**[is][\color{Sepia}]{SUS}{EUS},
moredelim=**[is][\color{black}]{SOK}{EOK},
% deletestring=[d]{"},
}
\lstset{morecomment=[n]{/*}{*/}}

\newcommand{\liv}[1]{\lstinline{#1}}
\newcommand{\li}[1]{\lstinline[basicstyle=\ttfamily\fontsize{10pt}{1em}\selectfont]{#1}}
\newcommand{\lismall}[1]{\lstinline[basicstyle=\ttfamily\fontsize{9pt}{1em}\selectfont]{#1}}
\newcommand{\lifootnote}[1]{\lstinline[basicstyle=\ttfamily\fontsize{7pt}{1em}\selectfont]{#1}}

% http://tex.stackexchange.com/q/43526
% fix the apparently deliberate but undocumented behaviour of disabling escapes other than mathescape in TextStyle (used by \lstinline)
% there may be a good reason why this is disabled by default, so beware in case it causes any problems
\usepackage{etoolbox}
\makeatletter
\patchcmd{\lsthk@TextStyle}{\let\lst@DefEsc\@empty}{}{}{\errmessage{failed to patch}}
\makeatother

% Provides a draft mark at the bottom of the document. 
% \draftstamp{\today}{DRAFT}

% I hate hyphenation.
%\lefthyphenmin=5
\definecolor{light-gray}{gray}{0.95}

\input{macros}

% allow interrupted equation numbering
% taken from http://tex.stackexchange.com/questions/101002/interrupting-and-resuming-subequations
% \makeatletter
% \def\user@resume{resume}
% \def\user@intermezzo{intermezzo}
% %
% \newcounter{previousequation}
% \newcounter{lastsubequation}
% \newcounter{savedparentequation}
% \setcounter{savedparentequation}{1}
% % 
% \renewenvironment{subequations}[1][]{%
%       \def\user@decides{#1}%
%       \setcounter{previousequation}{\value{equation}}%
%       \ifx\user@decides\user@resume 
%            \setcounter{equation}{\value{savedparentequation}}%
%       \else  
%       \ifx\user@decides\user@intermezzo
%            \refstepcounter{equation}%
%       \else
%            \setcounter{lastsubequation}{0}%
%            \refstepcounter{equation}%
%       \fi\fi
%       \protected@edef\theHparentequation{%
%           \@ifundefined {theHequation}\theequation \theHequation}%
%       \protected@edef\theparentequation{\theequation}%
%       \setcounter{parentequation}{\value{equation}}%
%       \ifx\user@decides\user@resume 
%            \setcounter{equation}{\value{lastsubequation}}%
%          \else
%            \setcounter{equation}{0}%
%       \fi
%       \def\theequation  {\theparentequation  \alph{equation}}%
%       \def\theHequation {\theHparentequation \alph{equation}}%
%       \ignorespaces
% }{%
% %  \arabic{equation};\arabic{savedparentequation};\arabic{lastsubequation}
%   \ifx\user@decides\user@resume
%        \setcounter{lastsubequation}{\value{equation}}%
%        \setcounter{equation}{\value{previousequation}}%
%   \else
%   \ifx\user@decides\user@intermezzo
%        \setcounter{equation}{\value{parentequation}}%
%   \else
%        \setcounter{lastsubequation}{\value{equation}}%
%        \setcounter{savedparentequation}{\value{parentequation}}%
%        \setcounter{equation}{\value{parentequation}}%
%   \fi\fi
% %  \arabic{equation};\arabic{savedparentequation};\arabic{lastsubequation}
%   \ignorespacesafterend
% }
% \makeatother

\begin{document} 
\title{Reasonably Programmable Literal Notation\\ (ICFP 2018 Anonymous Supplemental Material)}
\date{March 16, 2018}
\maketitle
\tableofcontents
\appendix

\chapter{Additional Examples}

\begin{figure*}[t]
\begin{subfigure}[t]{0.50\textwidth}
\begin{lstlisting}
  module Html = {
    /* a simplified encoding of HTML */
    type tag = string;
    type attr = (string, string);
    type attrs = list(attr);
    type t = 
      | Text(string) 
      | Elem(tag, attrs, list(t));
  }
\end{lstlisting}
\vspace{-4px}
\caption{The \li{Html} module, which defines \li{Html.t}.}
\label{fig:html-type-def}
\end{subfigure}
\hfill
\begin{subfigure}[t]{0.49\textwidth}
\begin{lstlisting}[morekeywords={expression}]
  module HtmlNotation {
    notation $html at Html.t {
      lexer Relit.TrivLexer;
      expression parser HtmlParser.expr;
      pattern parser HtmlParser.pat;
      expansions require 
        { module Html = Html; }
    }
  }
\end{lstlisting}
\vspace{-4px}
\caption{The \li{$html} expression and pattern TLMs.}
\label{fig:html-tlm-def}
\end{subfigure}

\begin{subfigure}[t]{\textwidth}
\vspace{5px}
% let pElem : Html.tag -> Html.attrs -> list(ProtoExpr.t) -> ProtoExpr.t = (* ... *)
% let pTextElem : Html.tag -> Html.attrs, string) -> ProtoExpr.t = (* ... *)
\begin{lstlisting}
  module HtmlParser = Relit.HandRolledExprPatParser({
    module ProtoHtml = {
      type t = PText(string)
             | PElem(Html.tag, Html.attrs, list(t))
             | PSplicedChildren(Html.tag, Html.attrs, Relit.Segment.t)
             | PSplicedElem(Relit.Segment.t) 
             | PSplicedText(Relit.Segment.t) 
      let parse : string => t = 
        /* ... via Markup.ml's stream combinators (see text) ... */;
    };
    module ExprExpander = {
      open notation Relit.$proto_expr; open ProtoHtml;
      let expand_attr = (x1, x2) => `(SURL ($s `(EURLx1SURL)`, $s `(EURLx2SURL)`)EURL )`;
      let expand_attrs = (attrs) => `(SURL $list `(EURLList.map(expand_attr, attrs)SURL)`EURL )`;
      let expand_proto = fun 
      | PText(text) => `(SURL Html.Text($s `(EURLtextSURL)`)EURL )` 
      | PElem(tag, attrs, children) => 
        `(SURL Html.Elem($s `(EURLtagSURL)`, `(EURLexpand_attrs(attrs)SURL)`, 
            $list `(EURLList.map(expand_parsed, children)SURL)`)EURL )`
      | PSplicedChildren(tag, attrs, seg) => 
        `(SURL Html.Elem($s `(EURLtagSURL)`, `(EURLexpand_attrs(attrs)SURL)`, 
            $spliced `(EURLsegSURL : list(Html.t))`)EURL )`
      | PSplicedHtml(seg) => `(SURL $spliced `(EURLsegSURL : Html.t) EURL)`);
      | PSplicedText(seg) => `(SURL Html.Text($spliced `(EURLsegSURL : striSURLEURLng)`) EURL)`
    };
    let expr = (body) => body |> ProtoHtml.parse |> ExprExpander.expand_proto;
    let pat  = (body) => expr(body) |> ProtoPat.from_expr;
  })
\end{lstlisting}
\vspace{-6px}
\caption{The implementation of \li{HtmlParser}, which defers the parsing step to the \li{Markup.ml} library \cite{markupml}.}
\label{fig:html-parser}
\end{subfigure}
\begin{subfigure}[t]{\textwidth}
\lstset{deletestring=[d]{"}}
\begin{lstlisting}
  open HtmlNotation; 
  exception UnexpectedFormat;
  let scrape_data = $html.(fun 
  | `(SURL<div... class="result">EURL
      [ `(SCOLOR<h1><$$>ECOLORtitleSCOLOR</$$></h1>ECOLOR)`, author, summary, ..._ ]SURL</div>EURL)` => 
    (title, author, summary)
  | _ => raise UnexpectedFormat);
\end{lstlisting}
\vspace{-4px}
\caption{Using a pattern TLM to deconstruct an HTML value.}
\label{fig:pattern-tlm-example}
\end{subfigure}
\vspace{2px}
\caption{Case Study: HTML literals as a library}
\label{fig:html-case-study}
\end{figure*}

\section{HTML Literals} 
Reason is seeing increasing adoption in the domain of client-side web programming, where generating and manipulating HTML data is central. 
There are many ways to represent HTML data in Reason, but one simple and useful representation is specified by \li{Html.t}, defined in Fig.~\ref{fig:html-type-def}. However, manually applying the constructors of this type is tedious and, for many web programmers, the induced notation is unfamiliar. It also makes it difficult to copy-and-paste from, for example, existing HTML files, as might be useful when refactoring an existing project to use Reason. Instead, we would like HTML literal notation. 
The \li{$html} TLM defined in Fig.~\ref{fig:html-tlm-def} implements the standard HTML notation extended with the tags \li{<$>} and \li{<$$>}, which support \li{Html.t} splicing and \li{string} splicing, respectively. It also supports the suffix \li{...} on a tag, which causes the body of the tag to be a splice of type \li{list(Html.t)}.\footnote{We would also in practice have splicing for  attribute values, but for simplicity we omit these.} The example in the paper demonstrated the first of these. The example in Fig.~\ref{fig:pattern-tlm-example} shows the list splicing form in pattern position.% HTML splicing form, \li{<$>}$e$\li{</$>}.

% The HTML standard is complex. In addition, good error handling is important in practice. As such, most practical HTML parsing libraries are ``hand-rolled'', i.e. they are not generated from a context-free grammar by a general parser generator like Menhir. For example, the robust \li{Markup.ml} library for OCaml provides a simple one-pass stream-based interface that can be used to generate an HTML representation of the user's choice \cite{ocaml}. Let us outline how we can use this library to implement HTML literals, with support for splicing as just specified. 
In implementing HTML literals, using a parser generator may not be ideal. Instead, ``post-processing'' the result of an existing parser is a good approach. Fig.~\ref{fig:html-parser} shows much of the implementation of \li{HtmlParser},  which is constructed by applying a functor that constructs a module that has the same essential interface as the modules that Menhir generates, with the functions \li{expr : string -> ProtoExpr.t} and \li{pat : string -> ProtoPat.t} of the input module exported as the ``non-terminals''. The companion lexer is \li{Relit.TrivLexer}. The implementation of \li{expr} on Line 23 of Fig.~\ref{fig:html-parser} reveals that the literal body is parsed by first calling \li{ProtoHtml.parse}, then passing the result on to \li{ExprExpander.expand_proto}.

The \li{ProtoHtml.parse} function, outlined on Lines 8-9, generates a value of the type \li{ProtoHtml.t} defined on Lines 2-6 and so named because it is similar to \li{Html.t} but distinguishes spliced segments, like a proto-expression. We omit the definition of \li{ProtoHtml.parse} but conceptually, it starts by creating a character stream from the literal body, then from that creates a \li{Markup.ml} HTML content stream, then transforms that to a stream that emits the signals from the HTML content stream until it sees a splice start tag (which is reported as a syntax error relative to the HTML standard), at which point it calls \li{Segment.read_to}, which was described in the paper, to advance the original character stream until it recognizes the closing tag, emitting the generated segment and then returning control to the HTML content stream. The final step is to fold over this stream to produce the desired result of type \li{ProtoHtml.t}.

The \li{ExprExpander.expand_proto} function defined, with some helper functions, on Lines~9-20 finishes the job of the TLM parser by mapping the parsed proto-HTML to a proto-expression using the \li{$proto_expr} TLM previously described (in this case, using the \li{$list} antiquotation form that produces a parse tree of OCaml's built in list literal form from a list of parse trees). It would be straightforward to reuse the \li{ProtoHtml} module for different variations on HTML encodings, implementing only the expander anew. Overall, the amount of code needed to implement HTML literals.

Reason currently builds in ``JSX'' literals, which also support an HTML-like notation with support for splicing in a somewhat idiosynchratic manner motivated by certain JavaScript libraries (in particular, React). The expansion of this notation must be interpretted by a PPX rewriter. The problem is that this makes it difficult to use different HTML-related libraries at once. For example, ReasonReact and an XML library might both want to use this notation. This approach solves this problem (as well as other problems with the PPX-based approach as discussed in the paper). % For example, another encoding is one where each tag in the HTML standard correspond to a distinct datatype constructor.

\section{Implementing Quasiquotation} The Reason quasiquotation TLMs, e.g. \li{$re_expr}, can similarly be implemented by the functional transformations outlined below, with an example from each step (grossly simplified from the actual \li{Parsetree} representation) on the right. In words, \li{$re_expr} first programmatically invokes the Reason parser on the literal body. It next serializes the generated parse tree to Reason source code, then parses that. This produces a parse tree that, if evaluated in the appropriate environment, will produce the original parse tree. The final step is to implement antiquotation as described above by repurposing the generalized literal forms in the body, using the source locations from the first parse tree, which have  been carried into the second parse tree as constants.
\begin{lstlisting}[numbers=none,basicstyle=\ttfamily\fontsize{8pt}{1em}\selectfont]
  body             "2 + `(xyz)`" 
  |> parse_re      Plus(Num(2, Loc(0)), GenLit("xyz", Loc(6)))
  |> serialize_re  "Plus(Num(2, Loc(0)), GenLit(\"SCSSxyz\ECSS", Loc(6)))"
  |> parse_re      Ap(V("Plus"), /*...*/Ap(V("GenLit"), Pair(Str("xyz"), /*...*/Num(6)))
  |> genlit_to_sp  Ap(V("Plus"), /*...*/Spliced(6,8,TyPath(["ProtoExpr","t"])))
\end{lstlisting}

% \section{Regular Expressions, More Abstractly}
% A regular expression, or \emph{regex}, is a description of a \emph{regular language} \cite{Thompson:1968:PTR:363347.363387}. Regexes arise with some frequency in fields like natural language processing and bioinformatics.

% \paragraph{Recursive Sums}
% One way to encode regular expressions in Reason is as values of the recursive datatype abbreviated \li{rx} in Figure \ref{fig:datatype-rx}.

% \begin{figure}[h]
% \begin{lstlisting}[numbers=none]
% type rx = Empty | Str(string) | Seq(rx, rx) 
%         | Or(rx, rx) | Star(rx);
% \end{lstlisting}
% \caption{Definition of the recursive labeled sum type \li{rx}}
% \label{fig:datatype-rx}
% \end{figure}
% We can construct a regex that matches the strings \li{"SSTRAESTR"}, \li{"SSTRTESTR"}, \li{"SSTRGESTR"} or \li{"SSTRCESTR"} (i.e. DNA bases) as follows:
% \begin{lstlisting}[numbers=none]
% Or(Str "SSTRAESTR", Or(Str "SSTRTESTR", Or(Str "SSTRGESTR", Str "SSTRCESTR")))
% \end{lstlisting}

% Given a value of type \li{rx}, we can deconstruct it by pattern matching. For example, the function \li{is_dna_rx} defined in Figure \ref{fig:is_dna_rx} detects regular expressions that match DNA sequences.

% \begin{figure}[h]
% \begin{lstlisting}[numbers=none]
% fun is_dna_rx(r : rx) : boolean => 
%   switch r { 
%   | Str "SSTRAESTR" => True;
%   | Str "SSTRTESTR" => True;
%   | Str "SSTRGESTR" => True;
%   | Str "SSTRCESTR" => True;
%   | Seq (r1, r2) => (is_dna_rx r1) andalso (is_dna_rx r2);
%   | Or  (r1, r2) => (is_dna_rx r1) andalso (is_dna_rx r2);
%   | Star(r') => is_dna_rx r';
%   | _ => False;
%   };
% \end{lstlisting}
% \caption{Pattern matching over regexes in Reason}
% \label{fig:is_dna_rx}
% \end{figure}

% \paragraph{Abstract Types} Encoding regexes as values of type \li{rx} is straightforward, but there are reasons why one might not wish to expose this encoding to clients directly.

% First, regexes are usually identified up to their reduction to a normal form. For example, \li{Seq(Empty, Str "SSTRAESTR")} has normal form \li{Str("SSTRAESTR")}. It can be useful for regexes with the same normal form to be  indistinguishable from the perspective of client code. (The details of regex normalization are not important for our purposes, so we omit them.)

% Second, it can be useful for performance reasons to maintain additional data alongside each regex (e.g. a corresponding finite automaton.) In fact, there may be many ways to represent regexes, each with different performance trade-offs, so we would like to provide a choice of representations behind a common interface.

% To achieve these goals, we turn to the OCaml module system, which is based closely on the SML module system \cite{mthm97-for-dart,dreyer2005understanding} (which originates in early work by MacQueen \cite{MacQueen:1984:MSM:800055.802036}.) In particular, let us define the {signature} abbreviated \li{RX} in Figure \ref{fig:signature-RX}.
% %Notice that it exposes an interface otherwise  to the one available using a case type:

% \begin{figure}[ht]
% \begin{lstlisting}[deletekeywords={case}]
% /* abstract regex unfoldings */
% type u('a) = UEmpty | UStr string | USeq ('a, 'a) 
%            | UOr ('a, 'a) | UStar of 'a;

% module type RX = {
%   type t (* abstract *);

%   (* constructors *)
%   let Empty : t;
%   let Str : string -> t;
%   let Seq : t * t -> t;
%   let Or : t * t -> t;
%   let Star : t -> t;

%   (* produces the normal unfolding *)
%   let unfold_norm : t -> u(t);
% end

% module R1 : RX = { (* ... *) };
% module R2 : RX = { (* ... *) };
% \end{lstlisting}
% \vspace{-5px}
% \caption{The \lstinline{RX} signature and two example implementations}
% \label{fig:signature-RX}
% \end{figure}

% The clients of any module \lstinline{R} that has been sealed by \lstinline{RX}, e.g. \li{R1} or \li{R2}  in Figure \ref{fig:signature-RX}, manipulate regexes as values of type \li{R.t} using the interface specified by \li{RX}. For example, a client can construct a regex matching DNA bases by projecting the value constructors out of \li{R} and applying them as follows:
% \begin{lstlisting}[numbers=none]
% R.Or(R.Str "SSTRAESTR", R.Or(R.Str "SSTRTESTR", R.Or (R.Str "SSTRGESTR", R.Str "SSTRCESTR")))
% \end{lstlisting}

% Because the identity of the representation type \lstinline{R.t} is held abstract by the signature, the only way for a client to construct a value of this type is through the values that \li{RX} specifies (i.e. we have defined an \emph{abstract data type (ADT)}  \cite{liskov1974programming}.) Consequently, representation invariants need only be established locally within each module.

% Similarly, clients cannot interrogate the structure of a value \li{r : R.t} directly. Instead, the signature specifies a function \li{R.unfold_norm} that produces the \emph{normal unfolding} of a given regex, i.e. a value of type \li{u(R.t)} that exposes only the outermost form of the regex in normal form (this normal form invariant is specified only as an unenforced side condition that implementations are expected to obey, as is common practice in languages like ML.) Clients can pattern match over the {normal unfolding} in the familiar manner, as shown in Figure \ref{fig:is_dna_rx_prime}. 
% \begin{figure}
% \begin{lstlisting}[numbers=none]
% fun is_dna_rx'(r : R.t) : boolean => 
%   switch R.unfold_norm r { 
%   | UStr "SSTRAESTR" => True;
%   | UStr "SSTRTESTR" => True;
%   | UStr "SSTRGESTR" => True;
%   | UStr "SSTRCESTR" => True;
%   | USeq (r1, r2) => (is_dna_rx' r1) && (is_dna_rx' r2);
%   | UOr (r1, r2) => (is_dna_rx' r1) && (is_dna_rx' r2);
%   | UStar r' => is_dna_rx' r';
%   | _ => False;
%   };
% \end{lstlisting}
% \vspace{-5px}
% \caption{Pattern matching over normal unfoldings of regexes}
% \label{fig:is_dna_rx_prime}
% \end{figure}

% The normal unfolding suffices in situations where a client needs to examine only the outermost structure of a regex. However, in general, a client may want to pattern match more deeply into a regex. There are various ways to approach this problem.

% One approach is to define auxiliary functions that construct $n$-deep unfoldings of \li{r}, where $n$ is the deepest level at which the client wishes to expose the normal structure of the regex. For example, it is easy to define a function \li{unfold_norm2 : R.t -> u(u(R.t))} in terms of \li{R.unfold_norm} that allows pattern matching to depth $2$.\footnote{Defining an unfolding \emph{generic} in $n$ is a more subtle problem that is beyond the scope of this work.}

% Another approach is to \emph{completely unfold} a value of type \li{t} by applying a function \li{view : R.t -> rx} that recursively applies \li{R.unfold_norm} to exhaustion. The type \li{rx} was defined in Figure \ref{fig:datatype-rx}.  Computing the complete unfolding (also called the \emph{view}) can have higher dynamic cost than computing an incomplete unfolding of appropriate depth, but it is also a simpler approach (i.e.   lower cognitive cost can justify higher dynamic cost.)

% \begin{figure}[t]
% \begin{lstlisting}[numbers=none]
% module RXUtil(R : RX) = {
%   let unfold_norm2(r : R.t) : u(u(R.t)) => /* ... */;

%   let view(r : R.t) : rx => 
%     switch R.unfold_norm r { 
%     | UEmpty => Empty;
%     | UStr s => Str s;
%     | USeq (r1, r2) => Seq (view r1, view r2);
%     | UOr (r1, r2) => Or (view r1, view r2);
%     | UStar r => Star (view r);
%     };

%   let parse_str(x : string) : R.t => /* ... */
% };
% \end{lstlisting}
% \vspace{-5px}
% \caption{The definition of \li{RXUtil}}
% \vspace{-5px}
% \label{fig:RXUtil}
% \end{figure}
% Typically, utility functions like \li{unfold_norm2} and \li{view} are defined in a \emph{functor} (i.e. a function at the level of modules) like \li{RXUtil} in Figure \ref{fig:RXUtil}, so that they need only be defined once, rather than separately for each module \li{R : RX}. The client can instantiate the functor by applying it to their choice of module as follows:
% \begin{lstlisting}[numbers=none]
% module RU = RXUtil(R)
% \end{lstlisting}

% \section{Using A Parser Generator to Define a TLM}
% Using these definitions, an implementation of a parser generator \li{P : PARSEGEN} and the static parametric TLMs \li{$grammar}, \li{$proto_typ} and \li{$proto_expr} as described in Sec. 6 of the paper, we might define a TLM for regexes (implementing a subset of the POSIX regex syntax for simplicity) as shown in Figure \ref{fig:rx-grammar-based}.

% The value \li{spliced_uexp} has type \li{proto_typ -> P.grammar(proto_expr)} and operates as described in Sec. 6 -- it recognizes the syntax of unexpanded expressions and produces the corresponding reference to a spliced expression, with the given type annotation.

% \begin{figure}[h!]
% \vspace{-5px}
% \begin{lstlisting}[deletekeywords={as}]
% (* we use regexes themselves to define lexemes in the grammar,
%  * so we need some static phase helpers. *)
% static module RS : RX = /* ... */;
% static module RU = RXUtil(RS);

% (* a TLM parametric over all modules R : RX *)
% syntax $rx(R : RX) at R.t by static {
%   P.generate ($grammar P proto_expr [|SHTML #\label{line:rx_parse_fn_start}#
%     start <- ""
%       EHTMLfun () => $proto_expr `SCSSR.EmptyECSS`SHTML
%     start <- "(" start ")"
%       EHTMLfun e => eSHTML
%     token str_tok #\label{line:str_tok_start}#
%       EHTML/* cannot use $rx within its own defn */
%       RU.parse_str "SSTR[^(@$]+ESTR" SHTML #\label{line:str_tok_end}#
%     start <- str_tok
%       EHTMLfun s => $proto_expr `SCSSR.Str %(ECSSstr_to_proto_str_lit sSCSS)ECSS`SHTML
%     start <- start start
%       EHTMLfun e1 e2 => $proto_expr `SCSSR.Seq (%ECSSe1SCSS, %ECSSe2SCSS)ECSS`SHTML
%     start <- start "|" start 
%       EHTMLfun e1 e2 => $proto_expr `SCSSR.Or (%ECSSe1SCSS, %ECSSe2SCSS)ECSS`SHTML
%     start <- start "*"
%       EHTMLfun e => $proto_expr `SCSSR.Star %ECSSe`SHTML
%     start <- "%{" ^(EHTMLspliced_uexp ($proto_typ `SCSSR.tECSS`)SHTML) "}" #\label{line:splicing-start}#
%       EHTMLfun e => eSHTML
%     start <- "${" ^(EHTMLspliced_uexp ($proto_typ `SCSSstringECSS`)SHTML) "}"
%       EHTMLfun e => $proto_expr `SCSSR.Str %(ECSSeSCSS)ECSS`SHTML #\label{line:splicing-end}#
%   EHTML|])
% }; #\label{line:rx_parse_fn_end}#
% \end{lstlisting}
% \vspace{-12px}
% \caption{A grammar-based definition of \texttt{\$rx}}
% \vspace{-15px}
% \label{fig:rx-grammar-based}
% \end{figure}

\input{appendix}

% \input{static-language}

\bibliographystyle{plainnat}
\bibliography{../../../../papers/research} %your bib file

\end{document}

### 2018-03-19 icfp18/supplement/popl18-supplement.tex
added supplement

\PassOptionsToPackage{svgnames,dvipsnames,svgnames}{xcolor}

\documentclass[12pt]{report}
\newif\ificfp
\icfptrue
\newcommand{\todolater}[1]{{\color{magenta} TODO (Later): #1}}
\newcommand{\todo}[1]{{\color{red} TODO: #1}}

% some useful packages
\usepackage{times}     % use times font for document
%\usepackage{lmodern}
\usepackage{newtxtt}
\usepackage{bbm}
%\renewcommand{\ttdefault}{txtt} % use txtt for typewriter font
\usepackage{mathpazo}
\usepackage{mathpartir} % use package for inference rules
\usepackage{upgreek} % package for alternative greek letters (\uppi)
\usepackage{fullpage}
\usepackage{colortab}
%\usepackage{graphicx}
\usepackage[labelfont=bf]{caption}

% \usepackage{cleveref}
%\usepackage{MnSymbol}
\usepackage{fancyvrb}
\usepackage{microtype}
\usepackage[framemethod=tikz]{mdframed}

% Get just llangle and rrangle from MnSymbol
\makeatletter
\DeclareFontFamily{OMX}{MnSymbolE}{}
\DeclareSymbolFont{MnLargeSymbols}{OMX}{MnSymbolE}{m}{n}
\SetSymbolFont{MnLargeSymbols}{bold}{OMX}{MnSymbolE}{b}{n}
\DeclareFontShape{OMX}{MnSymbolE}{m}{n}{
    <-6>  MnSymbolE5
   <6-7>  MnSymbolE6
   <7-8>  MnSymbolE7
   <8-9>  MnSymbolE8
   <9-10> MnSymbolE9
  <10-12> MnSymbolE10
  <12->   MnSymbolE12
}{}
\DeclareFontShape{OMX}{MnSymbolE}{b}{n}{
    <-6>  MnSymbolE-Bold5
   <6-7>  MnSymbolE-Bold6
   <7-8>  MnSymbolE-Bold7
   <8-9>  MnSymbolE-Bold8
   <9-10> MnSymbolE-Bold9
  <10-12> MnSymbolE-Bold10
  <12->   MnSymbolE-Bold12
}{}

\let\llangle\@undefined
\let\rrangle\@undefined
\DeclareMathDelimiter{\llangle}{\mathopen}%
                     {MnLargeSymbols}{'164}{MnLargeSymbols}{'164}
\DeclareMathDelimiter{\rrangle}{\mathclose}%
                     {MnLargeSymbols}{'171}{MnLargeSymbols}{'171}
\makeatother

% \usepackage{rotating}
% \usepackage{pdflscape}

\usepackage[colorlinks=true,allcolors=Blue,backref,pageanchor=true,plainpages=false, pdfpagelabels, bookmarks,bookmarksnumbered,
pdfborder={0 0 0},  %removes outlines around hyper links in online display
]{hyperref}

\usepackage{amsmath,amssymb, amsthm}

\allowdisplaybreaks[1]
\newtheorem{theorem}{Theorem}[chapter]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{assumption}[theorem]{Assumption}
\newtheorem{condition}[theorem]{Condition}

\usepackage{pfsteps}

\usepackage[noabbrev]{cleveref}

\newenvironment{proof-sketch}{\noindent{\emph{Proof Sketch.}}}{\qed}

% \makeatletter
% \renewenvironment{proof}[1][\proofname]{\par
%   \vspace{-\topsep}% remove the space after the theorem
%   \pushQED{\qed}%
%   \normalfont
%   \topsep0pt \partopsep0pt % no space before
%   \trivlist
%   \item[\hskip\labelsep
%         \itshape
%     #1\@addpunct{.}]\ignorespaces
% }{%
%   \popQED\endtrivlist\@endpefalse
%   \addvspace{6pt plus 6pt} % some space after
% }
% \makeatother
% \makeatletter
% \renewenvironment{proof-sketch}[1][\proofname]{\par
%   \vspace{-\topsep}% remove the space after the theorem
%   \pushQED{\qed}%
%   \normalfont
%   \topsep0pt \partopsep0pt % no space before
%   \trivlist
%   \item[\hskip\labelsep
%         \itshape
%     Proof Sketch\@addpunct{.}]\ignorespaces
% }{%
%   \popQED\endtrivlist\@endpefalse
%   \addvspace{6pt plus 6pt} % some space after
% }
% \makeatother

\usepackage{mathtools}
\usepackage{ stmaryrd }
\usepackage[numbers,sort]{natbib}

\usepackage{subfigure}

% Approximately 1" margins, more space on binding side
%\usepackage[letterpaper,twoside,vscale=.8,hscale=.75,nomarginpar]{geometry}
%for general printing (not binding)
\usepackage[letterpaper,twoside,vscale=.8,hscale=.75,nomarginpar,hmarginratio=1:1]{geometry}

\usepackage{listings}
\lstset{tabsize=2, 
basicstyle=\ttfamily\fontsize{11pt}{1em}\selectfont, 
commentstyle=\itshape\ttfamily\color{gray}, 
stringstyle=\ttfamily\color{red},
mathescape=false,escapechar=\#,
numbers=left, numberstyle=\scriptsize\color{gray}\ttfamily, language=ML,moredelim=[il][\sffamily]{?},showspaces=false,showstringspaces=false,xleftmargin=15pt, morekeywords=[1]{tyfam,opfam,let,fn,val,def,casetype,objtype,metadata,of,*,datatype,new,toast,syntax,module,where,import,switch,for,ana,syn,opcon,tycon,metasignature,metamodule,metasig,metamod,static,at,tycase,mod,macro,match,pattern,in,patterns,expressions,implicit,forall,rectype,fold,unfold,inj,by,spliced},deletekeywords={double},classoffset=0,belowskip=\smallskipamount,
moredelim=**[is][\color{red}]{SSTR}{ESTR},
moredelim=**[is][\color{Green}]{SHTML}{EHTML},
moredelim=**[is][\color{purple}]{SCSS}{ECSS},
moredelim=**[is][\color{brown}]{SSQL}{ESQL},
moredelim=**[is][\color{orange}]{SCOLOR}{ECOLOR},
moredelim=**[is][\color{magenta}]{SPCT}{EPCT}, 
moredelim=**[is][\color{gray}]{SNAT}{ENAT}, 
moredelim=**[is][\color{Green}]{SURL}{EURL},
moredelim=**[is][\color{SeaGreen}]{SQT}{EQT},
moredelim=**[is][\color{Periwinkle}]{SGRM}{EGRM},
moredelim=**[is][\color{YellowGreen}]{SID}{EID},
moredelim=**[is][\color{Sepia}]{SUS}{EUS},
deletestring=[d]{"},
}
\lstloadlanguages{Java,VBScript,XML,HTML,ML}
\let\li\lstinline

% http://tex.stackexchange.com/q/43526
% fix the apparently deliberate but undocumented behaviour of disabling escapes other than mathescape in TextStyle (used by \lstinline)
% there may be a good reason why this is disabled by default, so beware in case it causes any problems
\usepackage{etoolbox}
\makeatletter
\patchcmd{\lsthk@TextStyle}{\let\lst@DefEsc\@empty}{}{}{\errmessage{failed to patch}}
\makeatother

% Provides a draft mark at the bottom of the document. 
% \draftstamp{\today}{DRAFT}

% I hate hyphenation.
%\lefthyphenmin=5
\definecolor{light-gray}{gray}{0.95}

\input{macros}

% allow interrupted equation numbering
% taken from http://tex.stackexchange.com/questions/101002/interrupting-and-resuming-subequations
% \makeatletter
% \def\user@resume{resume}
% \def\user@intermezzo{intermezzo}
% %
% \newcounter{previousequation}
% \newcounter{lastsubequation}
% \newcounter{savedparentequation}
% \setcounter{savedparentequation}{1}
% % 
% \renewenvironment{subequations}[1][]{%
%       \def\user@decides{#1}%
%       \setcounter{previousequation}{\value{equation}}%
%       \ifx\user@decides\user@resume 
%            \setcounter{equation}{\value{savedparentequation}}%
%       \else  
%       \ifx\user@decides\user@intermezzo
%            \refstepcounter{equation}%
%       \else
%            \setcounter{lastsubequation}{0}%
%            \refstepcounter{equation}%
%       \fi\fi
%       \protected@edef\theHparentequation{%
%           \@ifundefined {theHequation}\theequation \theHequation}%
%       \protected@edef\theparentequation{\theequation}%
%       \setcounter{parentequation}{\value{equation}}%
%       \ifx\user@decides\user@resume 
%            \setcounter{equation}{\value{lastsubequation}}%
%          \else
%            \setcounter{equation}{0}%
%       \fi
%       \def\theequation  {\theparentequation  \alph{equation}}%
%       \def\theHequation {\theHparentequation \alph{equation}}%
%       \ignorespaces
% }{%
% %  \arabic{equation};\arabic{savedparentequation};\arabic{lastsubequation}
%   \ifx\user@decides\user@resume
%        \setcounter{lastsubequation}{\value{equation}}%
%        \setcounter{equation}{\value{previousequation}}%
%   \else
%   \ifx\user@decides\user@intermezzo
%        \setcounter{equation}{\value{parentequation}}%
%   \else
%        \setcounter{lastsubequation}{\value{equation}}%
%        \setcounter{savedparentequation}{\value{parentequation}}%
%        \setcounter{equation}{\value{parentequation}}%
%   \fi\fi
% %  \arabic{equation};\arabic{savedparentequation};\arabic{lastsubequation}
%   \ignorespacesafterend
% }
% \makeatother

\begin{document} 
\title{Reasonably Programmable Literal Forms\\ (POPL 2018 Anonymous Supplemental Material)}
\date{July 7, 2017}
\maketitle
\tableofcontents

\input{appendix}

\input{static-language}

\chapter{A Detailed, Non-Trivial Example: Regular Expression Literals}
\section{Basic Definitions}
A regular expression, or \emph{regex}, is a description of a \emph{regular language} \cite{Thompson:1968:PTR:363347.363387}. Regexes arise with some frequency in fields like natural language processing and bioinformatics.

\paragraph{Recursive Sums}
One way to encode regular expressions in Reason is as values of the recursive datatype abbreviated \li{rx} in Figure \ref{fig:datatype-rx}.

\begin{figure}[h]
\begin{lstlisting}[numbers=none]
type rx = Empty | Str string | Seq (rx, rx) 
        | Or (rx, rx) | Star of rx;
\end{lstlisting}
\caption{Definition of the recursive labeled sum type \li{rx}}
\label{fig:datatype-rx}
\end{figure}
We can construct a regex that matches the strings \li{"SSTRAESTR"}, \li{"SSTRTESTR"}, \li{"SSTRGESTR"} or \li{"SSTRCESTR"} (i.e. DNA bases) as follows:
\begin{lstlisting}[numbers=none]
Or(Str "SSTRAESTR", Or(Str "SSTRTESTR", Or(Str "SSTRGESTR", Str "SSTRCESTR")))
\end{lstlisting}

Given a value of type \li{rx}, we can deconstruct it by pattern matching. For example, the function \li{is_dna_rx} defined in Figure \ref{fig:is_dna_rx} detects regular expressions that match DNA sequences.

\begin{figure}[h]
\begin{lstlisting}[numbers=none]
fun is_dna_rx(r : rx) : boolean => 
  switch r { 
  | Str "SSTRAESTR" => True;
  | Str "SSTRTESTR" => True;
  | Str "SSTRGESTR" => True;
  | Str "SSTRCESTR" => True;
  | Seq (r1, r2) => (is_dna_rx r1) andalso (is_dna_rx r2);
  | Or  (r1, r2) => (is_dna_rx r1) andalso (is_dna_rx r2);
  | Star(r') => is_dna_rx r';
  | _ => False;
  };
\end{lstlisting}
\caption{Pattern matching over regexes in Reason}
\label{fig:is_dna_rx}
\end{figure}

\paragraph{Abstract Types} Encoding regexes as values of type \li{rx} is straightforward, but there are reasons why one might not wish to expose this encoding to clients directly.

First, regexes are usually identified up to their reduction to a normal form. For example, \li{Seq(Empty, Str "SSTRAESTR")} has normal form \li{Str("SSTRAESTR")}. It can be useful for regexes with the same normal form to be  indistinguishable from the perspective of client code. (The details of regex normalization are not important for our purposes, so we omit them.)

Second, it can be useful for performance reasons to maintain additional data alongside each regex (e.g. a corresponding finite automaton.) In fact, there may be many ways to represent regexes, each with different performance trade-offs, so we would like to provide a choice of representations behind a common interface.

To achieve these goals, we turn to the OCaml module system, which is based closely on the SML module system \cite{mthm97-for-dart,dreyer2005understanding} (which originates in early work by MacQueen \cite{MacQueen:1984:MSM:800055.802036}.) In particular, let us define the {signature} abbreviated \li{RX} in Figure \ref{fig:signature-RX}.
%Notice that it exposes an interface otherwise  to the one available using a case type:

\begin{figure}[ht]
\begin{lstlisting}[deletekeywords={case}]
/* abstract regex unfoldings */
type u('a) = UEmpty | UStr string | USeq ('a, 'a) 
           | UOr ('a, 'a) | UStar of 'a;

module type RX = {
  type t (* abstract *);

  (* constructors *)
  let Empty : t;
  let Str : string -> t;
  let Seq : t * t -> t;
  let Or : t * t -> t;
  let Star : t -> t;

  (* produces the normal unfolding *)
  let unfold_norm : t -> u(t);
end

module R1 : RX = { (* ... *) };
module R2 : RX = { (* ... *) };
\end{lstlisting}
\vspace{-5px}
\caption{The \lstinline{RX} signature and two example implementations}
\label{fig:signature-RX}
\end{figure}

The clients of any module \lstinline{R} that has been sealed by \lstinline{RX}, e.g. \li{R1} or \li{R2}  in Figure \ref{fig:signature-RX}, manipulate regexes as values of type \li{R.t} using the interface specified by \li{RX}. For example, a client can construct a regex matching DNA bases by projecting the value constructors out of \li{R} and applying them as follows:
\begin{lstlisting}[numbers=none]
R.Or(R.Str "SSTRAESTR", R.Or(R.Str "SSTRTESTR", R.Or (R.Str "SSTRGESTR", R.Str "SSTRCESTR")))
\end{lstlisting}

Because the identity of the representation type \lstinline{R.t} is held abstract by the signature, the only way for a client to construct a value of this type is through the values that \li{RX} specifies (i.e. we have defined an \emph{abstract data type (ADT)}  \cite{liskov1974programming}.) Consequently, representation invariants need only be established locally within each module.

Similarly, clients cannot interrogate the structure of a value \li{r : R.t} directly. Instead, the signature specifies a function \li{R.unfold_norm} that produces the \emph{normal unfolding} of a given regex, i.e. a value of type \li{u(R.t)} that exposes only the outermost form of the regex in normal form (this normal form invariant is specified only as an unenforced side condition that implementations are expected to obey, as is common practice in languages like ML.) Clients can pattern match over the {normal unfolding} in the familiar manner, as shown in Figure \ref{fig:is_dna_rx_prime}. 
\begin{figure}
\begin{lstlisting}[numbers=none]
fun is_dna_rx'(r : R.t) : boolean => 
  switch R.unfold_norm r { 
  | UStr "SSTRAESTR" => True;
  | UStr "SSTRTESTR" => True;
  | UStr "SSTRGESTR" => True;
  | UStr "SSTRCESTR" => True;
  | USeq (r1, r2) => (is_dna_rx' r1) && (is_dna_rx' r2);
  | UOr (r1, r2) => (is_dna_rx' r1) && (is_dna_rx' r2);
  | UStar r' => is_dna_rx' r';
  | _ => False;
  };
\end{lstlisting}
\vspace{-5px}
\caption{Pattern matching over normal unfoldings of regexes}
\label{fig:is_dna_rx_prime}
\end{figure}

The normal unfolding suffices in situations where a client needs to examine only the outermost structure of a regex. However, in general, a client may want to pattern match more deeply into a regex. There are various ways to approach this problem.

One approach is to define auxiliary functions that construct $n$-deep unfoldings of \li{r}, where $n$ is the deepest level at which the client wishes to expose the normal structure of the regex. For example, it is easy to define a function \li{unfold_norm2 : R.t -> u(u(R.t))} in terms of \li{R.unfold_norm} that allows pattern matching to depth $2$.\footnote{Defining an unfolding \emph{generic} in $n$ is a more subtle problem that is beyond the scope of this work.}

Another approach is to \emph{completely unfold} a value of type \li{t} by applying a function \li{view : R.t -> rx} that recursively applies \li{R.unfold_norm} to exhaustion. The type \li{rx} was defined in Figure \ref{fig:datatype-rx}.  Computing the complete unfolding (also called the \emph{view}) can have higher dynamic cost than computing an incomplete unfolding of appropriate depth, but it is also a simpler approach (i.e.   lower cognitive cost can justify higher dynamic cost.)

\begin{figure}[t]
\begin{lstlisting}[numbers=none]
module RXUtil(R : RX) = {
  let unfold_norm2(r : R.t) : u(u(R.t)) => /* ... */;

  let view(r : R.t) : rx => 
    switch R.unfold_norm r { 
    | UEmpty => Empty;
    | UStr s => Str s;
    | USeq (r1, r2) => Seq (view r1, view r2);
    | UOr (r1, r2) => Or (view r1, view r2);
    | UStar r => Star (view r);
    };

  let parse_str(x : string) : R.t => /* ... */
};
\end{lstlisting}
\vspace{-5px}
\caption{The definition of \li{RXUtil}}
\vspace{-5px}
\label{fig:RXUtil}
\end{figure}
Typically, utility functions like \li{unfold_norm2} and \li{view} are defined in a \emph{functor} (i.e. a function at the level of modules) like \li{RXUtil} in Figure \ref{fig:RXUtil}, so that they need only be defined once, rather than separately for each module \li{R : RX}. The client can instantiate the functor by applying it to their choice of module as follows:
\begin{lstlisting}[numbers=none]
module RU = RXUtil(R)
\end{lstlisting}

\section{Using A Parser Generator to Define a TLM}
Using these definitions, an implementation of a parser generator \li{P : PARSEGEN} and the static parametric TLMs \li{$grammar}, \li{$proto_typ} and \li{$proto_expr} as described in Sec. 6 of the paper, we might define a TLM for regexes (implementing a subset of the POSIX regex syntax for simplicity) as shown in Figure \ref{fig:rx-grammar-based}.

The value \li{spliced_uexp} has type \li{proto_typ -> P.grammar(proto_expr)} and operates as described in Sec. 6 -- it recognizes the syntax of unexpanded expressions and produces the corresponding reference to a spliced expression, with the given type annotation.

\begin{figure}[h!]
\vspace{-5px}
\begin{lstlisting}[deletekeywords={as}]
(* we use regexes themselves to define lexemes in the grammar,
 * so we need some static phase helpers. *)
static module RS : RX = /* ... */;
static module RU = RXUtil(RS);

(* a TLM parametric over all modules R : RX *)
syntax $rx(R : RX) at R.t by static {
  P.generate ($grammar P proto_expr [|SHTML #\label{line:rx_parse_fn_start}#
    start <- ""
      EHTMLfun () => $proto_expr `SCSSR.EmptyECSS`SHTML
    start <- "(" start ")"
      EHTMLfun e => eSHTML
    token str_tok #\label{line:str_tok_start}#
      EHTML/* cannot use $rx within its own defn */
      RU.parse_str "SSTR[^(@$]+ESTR" SHTML #\label{line:str_tok_end}#
    start <- str_tok
      EHTMLfun s => $proto_expr `SCSSR.Str %(ECSSstr_to_proto_str_lit sSCSS)ECSS`SHTML
    start <- start start
      EHTMLfun e1 e2 => $proto_expr `SCSSR.Seq (%ECSSe1SCSS, %ECSSe2SCSS)ECSS`SHTML
    start <- start "|" start 
      EHTMLfun e1 e2 => $proto_expr `SCSSR.Or (%ECSSe1SCSS, %ECSSe2SCSS)ECSS`SHTML
    start <- start "*"
      EHTMLfun e => $proto_expr `SCSSR.Star %ECSSe`SHTML
    start <- "%{" ^(EHTMLspliced_uexp ($proto_typ `SCSSR.tECSS`)SHTML) "}" #\label{line:splicing-start}#
      EHTMLfun e => eSHTML
    start <- "${" ^(EHTMLspliced_uexp ($proto_typ `SCSSstringECSS`)SHTML) "}"
      EHTMLfun e => $proto_expr `SCSSR.Str %(ECSSeSCSS)ECSS`SHTML #\label{line:splicing-end}#
  EHTML|])
}; #\label{line:rx_parse_fn_end}#
\end{lstlisting}
\vspace{-12px}
\caption{A grammar-based definition of \texttt{\$rx}}
\vspace{-15px}
\label{fig:rx-grammar-based}
\end{figure}

\bibliographystyle{plainnat}
\bibliography{../papers/research} %your bib file

\end{document}

### 2018-05-04 icfp18/rebuttal.txt
reviews and rebuttal

We thank the referees for their thoughtful reviews!

# Review A

Racket certainly does offer granular escape hatches out of its hygiene 
discipline. The problem is that the client does not know, abstractly, whether these 
escape hatches have been used strictly to support splicing from 
a string literal body.

To address this problem, it would be possible to use the escape hatches in Racket to implement, as a layer atop the Racket macro system, the literal macro system described in this paper. This point should also address the critique that the paper is "very closely tied to ML" --
in fact, the contributions are relevant even in dynamic languages like Racket 
(by simply ignoring the typing aspects, or by inserting a corresponding 
contract on each expansion). We will make these points clear in the final related work section.

With the additional page allowance in the final paper, we can include the HTML example from the supplement. It should not be necessary to remove the metatheory, which 
is a valuable contribution in that it makes the claimed reasoning 
principles precise (Theorems 5.2 and 5.3).

We would also be happy to provide the additional small clarifications suggested 
in the "Details" section of the review.

# Review B

> it does not appear that TLMs allow to easily splice in an arbitrary experssion of the host language

Spliced expressions can be any arbitrary expression of the host language that 
does not contain unbalanced \`( and )\`. Sec. 2.1 suggests this and many of the examples demonstrate this, but we would be happy to state this more directly.

The SQL example can certainly be expressed. If the "WHERE" expression appears directly in the expansion (i.e. it has a type like `DBRow -> bool`), this is straightforward. 
Alternatively, the TLM might internally extract its parse tree (like the 
`$re_expr` TLM in 2.2.2) and transform it. Any free 
variables in the transformed expression would need to be either 
(1) marked as spliced expressions, or (2) appear in the TLM's dependencies.

> I would have liked more details about the implementation

We would be happy to include the requested implementation details 
in the final version.

> would it be possible to rewrite the
second branch (with proper syntax support) :
```
| `( lam x . e )` as l => $v l
```

This is not possible -- the meaning of variables in OCaml is unchanged, 
so `l` stands for the matched value, not the generalized literal form 
appearing in the pattern. It is possible, of course, to instead apply a 
partial function `term_to_val : Lambda.term -> Lambda.v`.

# Review C

> the resulting system is presented only through a few partially-omitted 
examples

Fig. 3 together with Fig. 5 actually presents a complete example with no omissions, including the full 
lexer and parser. We would be happy to also include the second complete 
example from the supplement (see above).

There are in total 8 examples in the paper ($kq, $regex, $html, $smiles, 
$listcomp, $re_expr, \$term/\$v, $map). These span a variety
of application domains and demonstrate all of the features of the 
proposed system.

### 2018-06-12 icfp18/syntax-icfp18.tex
final submission
deanonymized + changed bib format + small edits to abstract

@@
 %% For double-blind review submission
-\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=false,printccs=false,printacmref=false}
+% \documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=false,printccs=false,printacmref=false}
+\documentclass[acmsmall]{acmart}
 % \documentclass[sigplan,review,10pt,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
@@
 \acmArticle{1}
-\acmYear{2017}
+\acmYear{2018}
 \acmMonth{1}
@@
 %% Supplied to authors by publisher for camera-ready submission
-\acmConference[PL'17]{ACM SIGPLAN Conference on Programming Languages}{January 01--03, 2017}{New York, NY, USA}
-\acmYear{2017}
+\acmConference[ICFP'18]{International Conference on Functional Programming}{September 23--29, 2018}{St. Louis, MO, USA}
+\acmYear{2018}
 \acmISBN{978-x-xxxx-xxxx-x/YY/MM}
@@
 %% issue of PACMPL.
-% \citestyle{acmauthoryear}   %% For author/year citations
+\citestyle{acmauthoryear}   %% For author/year citations
 
@@
 %% Author with single affiliation.
-\author{First1 Last1}
-\authornote{with author1 note}          %% \authornote is optional;
+\author{Cyrus Omar}
+\authornote{The majority of this research was performed while the first author was at Carnegie Mellon University.}          %% \authornote is optional;
                                         %% can be repeated if necessary
-\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
 \affiliation{
-  \position{Position1}
-  \department{Department1}              %% \department is recommended
-  \institution{Institution1}            %% \institution is required
-  \streetaddress{Street1 Address1}
-  \city{City1}
-  \state{State1}
-  \postcode{Post-Code1}
-  \country{Country1}
+  \institution{University of Chicago}            %% \institution is required
+  \city{Chicago}
+  \state{IL}
+  % \country{USA}
 }
-\email{first1.last1@inst1.edu}          %% \email is recommended
+\email{comar@cs.uchicago.edu}          %% \email is recommended
+\affiliation{
+  \institution{Carnegie Mellon University}           %% \institution is required
+  \city{Pittsburgh}
+  \state{PA}
+  % \country{USA}
+}
 
 %% Author with two affiliations and emails.
-\author{First2 Last2}
-\authornote{with author2 note}          %% \authornote is optional;
-                                        %% can be repeated if necessary
-\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
+\author{Jonathan Aldrich}
 \affiliation{
-  \position{Position2a}
-  \department{Department2a}             %% \department is recommended
-  \institution{Institution2a}           %% \institution is required
-  \streetaddress{Street2a Address2a}
-  \city{City2a}
-  \state{State2a}
-  \postcode{Post-Code2a}
-  \country{Country2a}
+  \institution{Carnegie Mellon University}           %% \institution is required
+  \city{Pittsburgh}
+  \state{PA}
+  % \country{USA}
 }
-\email{first2.last2@inst2a.com}         %% \email is recommended
-\affiliation{
-  \position{Position2b}
-  \department{Department2b}             %% \department is recommended
-  \institution{Institution2b}           %% \institution is required
-  \streetaddress{Street3b Address2b}
-  \city{City2b}
-  \state{State2b}
-  \postcode{Post-Code2b}
-  \country{Country2b}
-}
-\email{first2.last2@inst2b.org}         %% \email is recommended
+\email{jonathan.aldrich@cs.cmu.edu}         %% \email is recommended
 
@@
 TLM providers can reason modularly about syntactic ambiguity and expansion correctness according to clear criteria.
-This paper integrates TLMs into Reason, an emerging alternative front-end for OCaml, and demonstrates, through several non-trivial case studies, how TLMs integrate with the advanced features of OCaml, including pattern matching and the module system.
+This paper incorporates TLMs into Reason, an emerging alternative front-end for OCaml, and demonstrates, through several non-trivial case studies, how TLMs integrate with the advanced features of OCaml, including pattern matching and the module system.
 We discuss optional integration with MetaOCaml, which allows TLM providers to be more confident about type correctness.
@@
 \vspace{-4px}
-\caption{HTML literals with support for splicing at two different types are built primitively into Ur/Web \cite{conf/popl/Chlipala15}.}
+\caption{HTML literals with support for splicing at two types are built primitively into Ur/Web \cite{conf/popl/Chlipala15}.}
 \label{fig:urweb}

### 2018-06-12 icfp18/syntax-icfp18.tex
ACM copyright stuff, CCS stuff, made section titles all caps (blagh)

@@
 %% see authors.acm.org) by publisher for camera-ready submission
-\setcopyright{none}             %% For review submission
+% \setcopyright{none}             %% For review submission
 %\setcopyright{acmcopyright}
 %\setcopyright{acmlicensed}
-%\setcopyright{rightsretained}
+\setcopyright{rightsretained}
 %\copyrightyear{2017}           %% If different from \acmYear
@@
 \begin{abstract}
-\vspace{-5px}
 %Many languages define convenient list literal forms.
@@
 This paper incorporates TLMs into Reason, an emerging alternative front-end for OCaml, and demonstrates, through several non-trivial case studies, how TLMs integrate with the advanced features of OCaml, including pattern matching and the module system.
-We discuss optional integration with MetaOCaml, which allows TLM providers to be more confident about type correctness.
+We also discuss optional integration with MetaOCaml, which allows TLM providers to be more confident about type correctness.
 Finally, we establish these abstract reasoning principles formally with a detailed type-theoretic account of expression and pattern TLMs for ``core ML''.
@@
 %% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
-% \begin{CCSXML}
-% <ccs2012>
-% <concept>
-% <concept_id>10011007.10011006.10011008</concept_id>
-% <concept_desc>Software and its engineering~General programming languages</concept_desc>
-% <concept_significance>500</concept_significance>
-% </concept>
-% <concept>
-% <concept_id>10003456.10003457.10003521.10003525</concept_id>
-% <concept_desc>Social and professional topics~History of programming languages</concept_desc>
-% <concept_significance>300</concept_significance>
-% </concept>
-% </ccs2012>
-% \end{CCSXML}
-
-% \ccsdesc[500]{Software and its engineering~General programming languages}
-% \ccsdesc[300]{Social and professional topics~History of programming languages}
-%% End of generated code
+\begin{CCSXML}
+<ccs2012>
+<concept>
+<concept_id>10011007.10011006.10011008.10011009.10011019</concept_id>
+<concept_desc>Software and its engineering~Extensible languages</concept_desc>
+<concept_significance>500</concept_significance>
+</concept>
+<concept>
+<concept_id>10011007.10011006.10011050.10011055</concept_id>
+<concept_desc>Software and its engineering~Macro languages</concept_desc>
+<concept_significance>500</concept_significance>
+</concept>
+</ccs2012>
+\end{CCSXML}
+
+\ccsdesc[500]{Software and its engineering~Extensible languages}
+\ccsdesc[500]{Software and its engineering~Macro languages}
 
@@
 
-\vspace{-5px}
-\newcommand{\introSec}{Introduction}
-\section{\protect\introSec}
+\section{Introduction}
 \label{sec:intro}
-\vspace{-2px}
-% The surface syntax of a programming language is its user interface. 
 When designing the surface syntax of a general-purpose programming language, it is common practice to define shorthand \emph{literal notation}, i.e.
@@
 Lists are common across problem domains, but other literal notation is more specialized.
-For example, Ur/Web extends the surface syntax of Ur (an ML-like language \cite{conf/pldi/Chlipala10}) with expression and pattern literals for encodings of XML and HTML data~\cite{conf/popl/Chlipala15}.
-The example in Fig.~\ref{fig:urweb} shows two HTML literals, one that ``splices in'' a string expression delimited by \li{\{[} and \li{]\}} and the other an HTML expression delimited by \li{\{} and \li{\}}.
+For instance, Ur/Web extends the surface syntax of Ur (an ML-like language \cite{conf/pldi/Chlipala10}) with expression and pattern literals for encodings of XML and HTML data~\cite{conf/popl/Chlipala15}.
+For example, Fig.~\ref{fig:urweb} shows two Ur/Web HTML literals, one that ``splices in'' a string expression delimited by \li{\{[} and \li{]\}} and the other an HTML expression delimited by \li{\{} and \li{\}}.
 %The desugarings of these literal expressions, not shown, are substantially more verbose and less readable to programmers familiar with XHTML.
@@
 \end{lstlisting}
-\vspace{-4px}
 \caption{HTML literals with support for splicing at two types are built primitively into Ur/Web \cite{conf/popl/Chlipala15}.}
@@
 
-\newcommand{\seTLMsSec}{Expression TLMs in Reason}
-\section{\protect\seTLMsSec}
+\section{Expression Literals}
 \label{sec:setlms}
@@
 
-\newcommand{\spTLMsSec}{Pattern TLMs}
-\section{\protect\spTLMsSec}
+\section{Pattern Literals}
 \label{sec:sptsms}
@@
 
-\newcommand{\pTLMsSec}{Parametric TLMs}
-\section{\protect\pTLMsSec}
+\section{Parametric TLMs}
 \label{sec:ptsms}
@@
 
-\newcommand{\seTLMsFormallySec}{Simple TLMs, Formally}
-\section{\protect\seTLMsFormallySec}
+\section{Typed Literal Macros, Formally}
 \label{sec:setlms-formally}
@@
 % % \endgroup
-\newcommand{\implSec}{Implementation}
-\section{\protect\implSec}
+\section{Implementation}
 \label{sec:implementation}
@@
 
-\newcommand{\rwSec}{Related Work}
-\section{\protect\rwSec}
+\section{Related Work}
 \label{sec:existing-approaches}
@@
 
-\newcommand{\discussionSec}{Discussion}
-\section{\protect\discussionSec}
+\section{Discussion}
 \label{sec:discussion}

### 2018-06-16 icfp18/syntax-icfp18.tex
affiliation update per conversation with JA
some edits to intro

@@
 mathescape=false,escapeinside={(~}{~)},
-numbers=left, numberstyle=\scriptsize\color{mygray}, language=ML,showspaces=false,showstringspaces=false,xleftmargin=0pt, numbersep=-6pt, morekeywords=[1]{try,spliced,tyfam,opfam,let,fn,val,def,casetype,objtype,metadata,of,*,string,lexer,parser,datatype,new,toast,notation,module,switch,where,expansions,require,import,for,ana,syn,opcon,tycon,metasignature,metamodule,metasig,metamod,static,at,by,tycase,mod,macro,match,pattern,in,patterns,expressions,implicit,forall,int,exptsm,pattsm},deletekeywords={double,structure,of},classoffset=0, xleftmargin=0pt,
+numbers=left, numberstyle=\scriptsize\color{mygray}, language=ML,showspaces=false,showstringspaces=false,xleftmargin=0pt, numbersep=-6pt, morekeywords=[1]{try,spliced,tyfam,opfam,let,fn,val,def,casetype,objtype,metadata,of,*,string,lexer,parser,package,datatype,new,toast,notation,module,switch,where,dependencies,import,for,ana,syn,opcon,tycon,metasignature,metamodule,metasig,metamod,static,at,by,tycase,mod,macro,match,pattern,in,patterns,expressions,implicit,forall,int,exptsm,pattsm},deletekeywords={double,structure,of},classoffset=0, xleftmargin=0pt,
 aboveskip=3pt,belowskip=2pt,
@@
 \acmYear{2018}
-\acmMonth{1}
+\acmMonth{9}
 \acmDOI{10.1145/nnnnnnn.nnnnnnn}
@@
 \email{comar@cs.uchicago.edu}          %% \email is recommended
-\affiliation{
-  \institution{Carnegie Mellon University}           %% \institution is required
-  \city{Pittsburgh}
-  \state{PA}
-  % \country{USA}
-}
+% \affiliation{
+%   \institution{Carnegie Mellon University}           %% \institution is required
+%   \city{Pittsburgh}
+%   \state{PA}
+%   % \country{USA}
+% }
 
@@
 finite maps, regular expressions, HTML elements, SQL queries, syntax trees for various languages and chemical structures.
-There may also be different implementations of each of these data structures behind a common signature that could all benefit from common literal notation.
+There may also be different implementations of each of these data structures behind a common interface that could all benefit from common literal notation.
 %A less \emph{ad hoc} design would be one that allows library providers to define new literal forms in a decentralized manner.
@@
 \end{lstlisting}
+\vspace{-3px}
 \caption{HTML literals with support for splicing at two types are built primitively into Ur/Web \cite{conf/popl/Chlipala15}.}
 \label{fig:urweb}
-\vspace{-7px}
+\vspace{-10px}
 \end{figure}
@@
 This design practice, where the language designer privileges certain library constructs with built-in literal notation, is \emph{ad hoc} in that it is easy to come up with other examples of data structures for which mathematicians, scientists or programmers have invented specialized notation \cite{DBLP:journals/cacm/Iverson80,cajori1928history,TSLs}.
-For example, 1) clients of a ``collections'' library might want not just list literals, but also literal notation for matrices, finite maps, and so on; 2) clients of a ``web programming'' library might want CSS literals (which Ur/Web lacks); 3) a compiler author might want ``quotation'' literals for the terms of the object language and various intermediate languages of interest; and 4) clients of a ``chemistry'' library might want chemical structure literals based on the SMILES standard \cite{anderson1987smiles}.
+For example, (1) clients of a ``collections'' library might want not just list literals, but also literal notation for matrices, finite maps, and so on; (2) clients of a ``web programming'' library might want CSS literals (which Ur/Web lacks); (3) a compiler author might want ``quotation'' literals for the terms of the object language and various intermediate languages of interest; and (4) clients of a ``chemistry'' library might want chemical structure literals based on the SMILES standard \cite{anderson1987smiles}.
 
@@
 for URLs, paths, regular expressions and many others.
-Literal notation, with support for splicing, would decrease the syntactic cost of composite encodings, which are more amenable to programmatic manipulation and compositional reasoning.
+Literal notation, with support for splicing, would decrease the syntactic cost of composite encodings, which are more amenable to programmatic manipulation and compositional reasoning than string encodings.
 %Literal forms would  preferable to string encodings because they support compositional construction and decomposition by pattern matching via splicing.
@@
 Instead, there has been persistent interest in mechanisms that allow library providers to define new literal notation on their own.
-For example, direct grammar extension systems like Camlp4 \cite{de2003camlp4} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, term rewriting systems like Template Haskell~\cite{SheardPeytonJones:Haskell-02,mainland2007s}, the system of \emph{type-specific languages} (TSLs) described by \citet{TSLs}, and other systems that we will discuss below can all be used to define new literal notation (and, in some cases, other forms of new notation, such as new infix expression forms, control flow operations or type declaration forms, which we leave beyond the scope of this paper).
+For example, direct grammar extension systems like Camlp4 \cite{de2003camlp4} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, term rewriting systems like Template Haskell~\cite{SheardPeytonJones:Haskell-02,mainland2007s}, the system of \emph{type-specific languages} (TSLs) described by \citet{TSLs}, and other systems that we will discuss below can all be used to define new literal notation (and, in some cases, other forms of new notation, such as new infix operator forms, control flow operations or type declaration forms, which we leave beyond the scope of this paper).
 
@@
   notation $kq at KQuery.t {
-    lexer  KQueryLexer;
-    parser KQueryParser.start;
-    expansions require
-      { module KQuery as KQuery }
+    lexer  KQueryLexer
+    parser KQueryParser.start
+    in package kquery_parser;
+    dependencies = {module KQuery = KQuery}
   }; /* ...more notations defined... */
@@
 \end{subfigure}
-\vspace{4px}
+\vspace{2px}
 \caption{Two of the possible ways to introduce literal notation for encodings of K queries}
-\vspace{-10px}
+\vspace{-4px}
 \end{figure}
 
-Consider, for instance, the perspective of a programmer trying to comprehend the program text in Fig.~\ref{fig:K-dialect}, which is written in an emerging dialect of OCaml's surface syntax called Reason \cite{reason-what} that has, hypothetically, been extended with some number of new literal forms by a grammar extension system --- Lines 1-5 show the grammar extension syntax from Camlp4 \cite{de2003camlp4}, but Sugar*/SugarHaskell is  similar \cite{erdweg2011sugarj,erdweg2013framework,erdweg2012layout}.
-Line 9 uses one of these active syntax extensions to construct an encoding of a query in the rather obscure database query language K, using its intentionally terse notation~\cite{Whitney:2001:LOR:376284.375783}.
-The problem is that a programmer looking at the program as presented, and unfamiliar with (i.e.
-holding abstract) the details elided on Lines 5-6, cannot easily answer questions like the following:
+Consider, for instance, the perspective of a programmer trying to comprehend the program text in Fig.~\ref{fig:K-dialect}, which is written in an emerging dialect of OCaml's surface syntax called Reason \cite{reason-what} that has, hypothetically, been extended with some number of new literal forms by a grammar extension system --- Lines 1-6 outline the Camlp4 mechanism \cite{de2003camlp4}, but Sugar*/SugarHaskell is similar \cite{erdweg2011sugarj,erdweg2012layout}.
+Line 8 uses one of these active syntax extensions to construct an encoding of a query in the rather obscure database query language K, using its intentionally terse notation~\cite{Whitney:2001:LOR:376284.375783}.
+The problem is that a programmer examining the program as presented, and unfamiliar with (i.e.
+holding abstract) the details elided on Lines 1-6, cannot easily answer questions like the following:
 
 \begin{enumerate}[leftmargin=15px]
-\item \textbf{Responsibility}: Which syntax extension determined the expansion of the literal on Line~9? Might activating a new extension generate a conflicting expansion for the same literal?
-\item \textbf{Expansion Typing}: What type does \li{y} have?
-\item \textbf{Context Dependence}: Which bindings does the expansion of Line~9 invisibly depend on?
-If I shadow or remove a module or other binding, could that break or change the meaning of Line~9 because its expansion depends invisibly on the original binding?
-\item \textbf{Segmentation}: Are the characters \li{x}, \li{R} and \li{2} on Line~9 parsed as spliced terms, meaning that they appear directly in the underlying expansion, or are they parsed in some other way peculiar to this literal notation, e.g.
+\item \textbf{Responsibility}: Which syntax extension determined the expansion of the literal on Line~8? Might activating a new extension generate a conflicting expansion for the same literal?
+\item \textbf{Expansion Typing}: What type does the expansion, and thus the variable \li{y} on Line 8, have?
+\item \textbf{Context Dependence}: Which bindings does the expansion of Line~8 invisibly depend on?
+If we shadow or remove a module or other binding, could that break or change the meaning of Line~8 because its expansion depends invisibly on the original binding?
+\item \textbf{Segmentation}: Are the characters \li{x}, \li{R} and \li{2} on Line~8 parsed as spliced expressions, meaning that they appear directly in the underlying expansion, or are they parsed in some other way peculiar to this literal notation, e.g.
 as operators in the K query language?
 \item \textbf{Segment Typing}: What type is each spliced term expected to have?
-How can I infer a type for a variable that appears in a spliced term without looking at where it ends up in the expansion?
-\item \textbf{Capture}: If \li{x} is in fact a spliced term, does it refer to the binding of \li{x} on Line 5, or might it capture an invisible binding of the same identifier in the expansion of Line 9?
+How can we infer a type for a variable that appears in a spliced term without looking at where it ends up in the expansion?
+\item \textbf{Capture}: If \li{x} is in fact a spliced term, does it refer to the binding of \li{x} on Line 7, or might it capture an invisible binding of the same identifier in the expansion of Line 8?
 \end{enumerate}
@@
 
-Given these issues, we concluded that direct grammar extension systems were not ideally suited for integration into the Reason platform, which seeks to develop a clear and modern surface syntax for the OCaml programming language.
+Given these issues, we concluded that direct grammar extension systems like \li{camlp4} were not ideally suited for integration into the Reason platform, which seeks to develop a clear and modern surface syntax for the OCaml programming language \cite{reason-what}.
 We also evaluated various approaches that are based not on direct grammar extension but on term rewriting over a fixed grammar.
@@
 They also partially or completely solve the problem of reasoning about \textbf{Responsibility} but they do not satisfy the remaining five reasoning criteria.
-\item Hygienic staging macro systems, like MetaOCaml and related systems \cite{DBLP:conf/flops/Kiselyov14,ganz2001macros,Sheard:1999:UMS} are unsuitable for defining literal notation because they do not give the expander access to syntax trees.%, i.e.
+% \item Hygienic staging macro systems, like MetaOCaml and related systems \cite{DBLP:conf/flops/Kiselyov14,ganz2001macros,Sheard:1999:UMS} are unsuitable for defining literal notation because they do not give the expander access to syntax trees.%, i.e.
 the rewriting is parametric in the arguments, so they are unsuitable for defining new notation.
-\item Hygienic term rewriting macro systems, like those in various Lisp-family languages \cite{DBLP:conf/esop/HermanW08,Herman10:Theory,Kohlbecker86a,DBLP:conf/popl/Adams15,DBLP:conf/popl/ClingerR91,DBLP:journals/lisp/DybvigHB92,mccarthy1978history,Flatt:2012:CLR:2063176.2063195}, e.g.
-Racket \cite{Flatt:2012:CLR:2063176.2063195}, as well as Scala \cite{ScalaMacros2013}, do not allow us to flexibly repurpose string literal forms to define composite literal forms because the hygiene discipline cannot account for terms spliced out of literal bodies via parsing.%(the system cannot identify these terms as sub-terms of the macro arguments \cite{DBLP:conf/esop/HermanW08,Herman10:Theory}).% Only existing composite forms can therefore be repurposed.
-\item \emph{Type-specific languages} (TSLs) \cite{TSLs} come closer to our goals in that they explicitly support splicing terms out of literal bodies, but the mechanism falls short with regard to the six reasoning principles just discussed in ways that we will return to in Sec.
+\item Hygienic term rewriting macro systems, like those in various Lisp-family languages, 
+% \cite{DBLP:conf/esop/HermanW08,Herman10:Theory,Kohlbecker86a,DBLP:conf/popl/Adams15,DBLP:conf/popl/ClingerR91,DBLP:journals/lisp/DybvigHB92,mccarthy1978history,Flatt:2012:CLR:2063176.2063195}
+e.g.
+Racket \cite{Flatt:2012:CLR:2063176.2063195}, as well as Scala \cite{ScalaMacros2013}, do not allow us to flexibly repurpose string literal forms to define composite literal forms because the hygiene discipline cannot account for base language terms spliced out of string literal bodies via parsing (see Sec.~\ref{sec:existing-approaches} for more details).
+%(the system cannot identify these terms as sub-terms of the macro arguments \cite{DBLP:conf/esop/HermanW08,Herman10:Theory}).% Only existing composite forms can therefore be repurposed.
+\item \emph{Type-specific languages (TSLs)} \cite{TSLs} come closer to our goals in that they explicitly support splicing terms out of literal bodies, but the mechanism falls subtly short with regard to the six reasoning principles just discussed in ways that we detail in Sec.
 \ref{sec:existing-approaches}.
 In any case, this approach was designed for simple nominally-typed languages and relies critically on a particular local type inference scheme.
-It is not suitable for a language with an ML-like semantics, meaning a language with support for structural types (like tuple and function types in ML), parameterized type families, pattern matching and non-local type inference.
+It is not immediately suitable for a language with an ML-like semantics, meaning a language with support for structural types (like tuple and function types in ML), parameterized type families, pattern matching and non-local type inference.
 \end{itemize}
@@
 \paragraph{A Brief Overview} For a brief overview of the proposed mechanism, consider Fig.~\ref{fig:K-tsm-example}.
-Lines~1-5 define a TLM named \li{$kq} that provides K query literal notation.
-Line 9 applies this TLM to express the example from Figure \ref{fig:K-dialect}.
+Lines~1-6 define a TLM named \li{$kq} that provides K query literal notation.
+Line 8 applies this TLM to express the example from Fig.
+\ref{fig:K-dialect}.
 We can reason abstractly about this program as follows.
@@
 Sec.~\ref{sec:ptsms} introduces the more general {parametric TLMs}, which allow us to define literal notation at a type- or module-parameterized family of types.
-This section also reveals that the \li{expansions require} clause can be understood as parameterization followed immediately by partial parameter application, or, alternatively, that parameterization can be understood in terms of the \li{expansions require} clause together with ML's module functions (functors).
+This section also reveals that the \li{dependencies} clause can be understood as parameterization followed immediately by partial parameter application, or, alternatively, that parameterization can be understood in terms of the \li{dependencies} clause together with ML's module functions (functors).
 Having introduced the basic machinery by example, we proceed in
@@
       parser RegexParser.start;
-      expansions require
+      dependencies
         { module Regex = Regex; }
@@
 % \begin{lstlisting}[numbers=none]
-%   expansions require { 
+%   dependencies { 
 %     module Regex = Regex;
@@
 However, the typing guarantee is relative to the context where the quotation was constructed, i.e.
-the parser's implementation, so a validation step is still needed to ensure that the assumptions are consistent with those specified by the \li{expansions require} clause.
+the parser's implementation, so a validation step is still needed to ensure that the assumptions are consistent with those specified by the \li{dependencies} clause.
 
@@
       pattern parser LambdaParser.term_p;
-      expansions require 
+      dependencies 
         { module Lambda = Lambda } };
@@
 %       pattern parser Lam.tm_pat;
-%       expansions require { module Lam }
+%       dependencies { module Lam }
 %     }
@@
 
-The \li{expansions require} clause shown in previous examples can be understood in terms of parameterization over the dependencies, using the given module variables, followed immediately by partial application of the corresponding module paths.
+The \li{dependencies} clause shown in previous examples can be understood in terms of parameterization over the dependencies, using the given module variables, followed immediately by partial application of the corresponding module paths.
 Note, however, that module parameters need an explicit signature, like functor (module function) arguments in ML.
 
-An alternative point-of-view in ML is to treat the \li{expansions require} clause as fundamental and achieve parameterization using module functions (functors), e.g.
-we could express \li{$map} as follows:
+An alternative point-of-view in ML is to treat the \li{dependencies} clause as fundamental and achieve parameterization using module functions (functors), e.g. we could express \li{$map} as follows:
 \begin{lstlisting}[numbers=none]
@@
       lexer MapLexer; parser MapParser; 
-      expansions require { module M = M; type a = A.a; } } }
+      dependencies { module M = M; type a = A.a; } } }
 \end{lstlisting}
@@
 
-In $\miniVersePat$, we model the \lismall{expansions require} clause as specifying a single value dependency.
+In $\miniVersePat$, we model the \lismall{dependencies} clause as specifying a single value dependency.
 %\footnote{In \lifootnote{Relit}, this is further restricted to a module expression with a singleton signature, for reasons discussed in Sec.~\ref{sec:implementation}.} 
@@
 In $\miniVersePat$, the proto-expansion does not expand directly to the final expansion but to a function, here $e$, from the dependency type, $\taudep$, to the expansion type, $\tau$, as suggested by the right-hand side of the final premise.
-In the conclusion of the rule, we apply the dependency variable, $x$, to produce the final expansion.\footnote{In \lifootnote{Relit}, the dependency is always on the singleton module induced by the \lifootnote{expansions require} clause, which is \lifootnote{open}ed immediately, so the necessary boilerplate is inserted automatically.% If this were changed to require that the proto-expansion be a curried functor over dependencies listed in lexical order, the upside would be that the module identifiers used internally to a proto-expansion would not be exposed in the TLM definition, but the downside is that this would require the signature of each dependency be reproduced in each proto-expansion.
+In the conclusion of the rule, we apply the dependency variable, $x$, to produce the final expansion.\footnote{In \lifootnote{Relit}, the dependency is always on the singleton module induced by the \lifootnote{dependencies} clause, which is \lifootnote{open}ed immediately, so the necessary boilerplate is inserted automatically.% If this were changed to require that the proto-expansion be a curried functor over dependencies listed in lexical order, the upside would be that the module identifiers used internally to a proto-expansion would not be exposed in the TLM definition, but the downside is that this would require the signature of each dependency be reproduced in each proto-expansion.
 We chose the more concise approach.

### 2018-06-16 icfp18/syntax-icfp18.tex
acknowledgments

@@
   %% extraction tools.
-  This material is based upon work supported by the
-  \grantsponsor{GS100000001}{National Science
-    Foundation}{http://dx.doi.org/10.13039/100000001} under Grant
-  No.~\grantnum{GS100000001}{nnnnnnn} and Grant
-  No.~\grantnum{GS100000001}{mmmmmmm}.  Any opinions, findings, and
-  conclusions or recommendations expressed in this material are those
-  of the author and do not necessarily reflect the views of the
-  National Science Foundation.
+   This work was
+supported in part by AFRL and DARPA under agreement \#FA8750-16-2-0042;
+ by the NSA under lablet contract \#H98230-14-C-0140; and by Ravi Chugh
+ at the University of Chicago. The authors
+would like to thank Robert Harper, Karl Crary and Eric Van Wyk for 
+valuable feedback on the first author's doctoral dissertation, which forms the basis of this work.
+The authors would also like to thank Charles Chamberlain
+for his crucial contributions to the implementation and for valuable feedback on this
+paper. Finally, the authors would like to thank the anonymous referees
+at ICFP and at prior conferences, and the members of the PL Reading Group at University of Chicago, 
+for their valuable feedback, which substantially improved the paper.
+
+  % This material is based upon work supported by the
+  % \grantsponsor{GS100000001}{National Science
+  %   Foundation}{http://dx.doi.org/10.13039/100000001} under Grant
+  % No.~\grantnum{GS100000001}{nnnnnnn} and Grant
+  % No.~\grantnum{GS100000001}{mmmmmmm}.  Any opinions, findings, and
+  % conclusions or recommendations expressed in this material are those
+  % of the author and do not necessarily reflect the views of the
+  % National Science Foundation.
 \end{acks}

### 2018-06-16 icfp18/syntax-icfp18.tex
tightened acks

@@
 would like to thank Robert Harper, Karl Crary and Eric Van Wyk for 
-valuable feedback on the first author's doctoral dissertation, which forms the basis of this work.
-The authors would also like to thank Charles Chamberlain
-for his crucial contributions to the implementation and for valuable feedback on this
-paper. Finally, the authors would like to thank the anonymous referees
-at ICFP and at prior conferences, and the members of the PL Reading Group at University of Chicago, 
-for their valuable feedback, which substantially improved the paper.
+valuable feedback on the first author's doctoral dissertation, which forms the basis of this work; 
+Charles Chamberlain, 
+for his crucial contributions to the implementation and for valuable feedback on the
+paper; and members of the PL Reading Group at University of Chicago, and the anonymous referees
+at ICFP and at prior conferences, for their valuable critiques, which substantially improved the paper.
 

### 2018-06-16 icfp18/syntax-icfp18.tex
more tightening of acks

@@
 paper; and members of the PL Reading Group at University of Chicago, and the anonymous referees
-at ICFP and at prior conferences, for their valuable critiques, which substantially improved the paper.
+at ICFP and prior conferences, for their valuable critiques, which substantially improved the paper.
 

### 2018-06-16 icfp18/syntax-icfp18.tex
finished? revising intro text

@@
 %the context-free grammar of the language is fixed, so the TLM provider can modularly establish that the notation that the TLM implements is unambiguous (even in situations where a spliced term itself applies another TLM, which we will show in a later example).
-\item \textbf{Expansion Typing}: The type annotation on the definition of \li{$kq} (Line 1) determines the type that the expansion must have, here \li{KQuery.t}.
+\item \textbf{Expansion Typing}: The type annotation on the definition of \li{$kq} (Line 1) specifies the type that the expansion must have, here \li{KQuery.t}.
 \item \textbf{Context Dependence}: Lines 4-5 specify that expansions generated by \li{$kq} are allowed to use the module \li{KQuery}, and no others.
@@
 as operators of the K language (because they are in lavender).
-This also implies that errors in spliced terms can always be reported in terms of their original location.
+Errors in spliced terms can always be reported in terms of their original location.
 \item \textbf{Segment Typing}: Each spliced segment in the inferred segmentation also has  a type annotation.
@@
 % \item 
-Sec.~\ref{sec:setlms} details expression TLMs in Reason with several more case studies of varying detail, notably including TLMs for regular expressions, Ur/Web style HTML literals, chemical structure literals and quasiquotation for Reason language terms, which can be used for implementing other TLMs.
+Sec.~\ref{sec:setlms} details expression TLMs in Reason with several more case studies of varying detail, notably including TLMs for regular expressions, HTML literals, chemical structure literals and quasiquotation for Reason language terms, which can be used for implementing other TLMs.
 It also describes experimental integration with MetaOCaml, which can help providers reason about type correctness.
@@
 Sec.~\ref{sec:ptsms} introduces the more general {parametric TLMs}, which allow us to define literal notation at a type- or module-parameterized family of types.
-This section also reveals that the \li{dependencies} clause can be understood as parameterization followed immediately by partial parameter application, or, alternatively, that parameterization can be understood in terms of the \li{dependencies} clause together with ML's module functions (functors).
+ % This section also reveals that the \li{dependencies} clause can be understood as parameterization followed immediately by partial parameter application, or, alternatively, that parameterization can be understood in terms of the \li{dependencies} clause together with ML's module functions (functors).
  Having introduced the basic machinery by example, we proceed in 
@@
 Notably, this section gives our take on the example from  Fig.~\ref{fig:urweb} of Ur/Web-style HTML literals as well as a TLM that implements quasiquotation for Reason language terms, which is useful for implementing other TLMs.
-Sec.~\ref{sec:setlms-formally} to define a type-theoretic account of simple expression and pattern TLMs and formally establish the reasoning principles implied above in their essential form. 
+Sec.~\ref{sec:setlms-formally} to describe a type-theoretic account of simple expression and pattern TLMs and formally establish the reasoning principles implied above in their essential form.
+The full technical details and proofs are in the accompanying technical report \cite{icfp18tr}.\todo{fix .bib when we have the TR number}{}
 % Sec.~\ref{sec:ptlms-formally} adds type functions and an ML-style module system to this calculus, and gives a more general variant of the reasoning principles theorem. 
@@
 % We also give a brief overview of optional integration with MetaOCaml, which can be used to guarantee that all expansion generated by a TLM are type correct.
+This implementation, called \li{Relit}, and additional implementation details, documentation and examples are available from the project page\todo{get a github.io url?}{}:
+
+\begin{center}
+\url{https://github.com/cyrus-/relit}
+\end{center}
+
+\noindent
 % \item 
@@
 % \item 
-Sec.~\ref{sec:discussion} concludes with a summary of the contributions of this paper and a discussion of limitations and future work.
-Certain technical details, proofs and additional details on the case studies are available in the supplement.
+Sec.~\ref{sec:discussion} concludes with a discussion of contributions, limitations and future work.
 % Also available is an additional series of case studies: notation for an encoding of regular expressions as an abstract data type, implemented using TLMs for quasiquotation and grammar-based parser generators.

### 2018-06-16 icfp18/syntax-icfp18.tex
fixed line numbers and light tweaks to bullet points in Brief Overview section

@@
 \begin{enumerate}[leftmargin=12pt]
-\item \textbf{Responsibility}: The lexer and parser specified by the applied TLM on Lines 2-3 are together exclusively responsible for lexing, parsing and expanding the body of the generalized literal form, i.e.
+\item \textbf{Responsibility}: The lexer and parser specified by the applied TLM on Lines 2-4 are together exclusively responsible for lexing, parsing and expanding the body of the generalized literal form, i.e.
 the characters between \li{`(} and \li{)`}.
@@
 %the context-free grammar of the language is fixed, so the TLM provider can modularly establish that the notation that the TLM implements is unambiguous (even in situations where a spliced term itself applies another TLM, which we will show in a later example).
-\item \textbf{Expansion Typing}: The type annotation on the definition of \li{$kq} (Line 1) specifies the type that the expansion must have, here \li{KQuery.t}.
-\item \textbf{Context Dependence}: Lines 4-5 specify that expansions generated by \li{$kq} are allowed to use the module \li{KQuery}, and no others.
+\item \textbf{Expansion Typing}: The type annotation on Line 1 specifies the type that every expansion generated by \li{$kq} must have, here \li{KQuery.t}.
+\item \textbf{Context Dependence}: Line 5 specifies that expansions generated by \li{$kq} are allowed to use the module \li{KQuery}, and no others.
 The system ensures that this dependency is  bound as specified even if the variable \li{KQuery} has been shadowed at the application site.
@@
 colors in this document.
-So by examining Line 9, the programmer knows that the two instances of \li{x} are spliced expressions (because they are in black), whereas the \li{R}'s must be parsed in some other way, e.g.
+So by examining Line 8, the programmer knows that the two instances of \li{x} are spliced expressions (because they are in black), whereas the \li{R}'s must be parsed in some other way, e.g.
 as operators of the K language (because they are in lavender).
@@
 %TLM definitions follow the usual scoping rules, so it is easy to ``jump to the definition'' of \li{$kq}.
-\item \textbf{Capture}: Splicing is guaranteed to be capture-avoiding, so the spliced expression \li{x} must refer to the binding of \li{x} on Line 2.
+\item \textbf{Capture}: Splicing is guaranteed to be capture-avoiding, so the spliced expression \li{x} must refer to the binding of \li{x} on Line 7.
 It cannot have captured a coincidental binding of \li{x} in the expansion.

### 2018-06-18 icfp18/syntax-icfp18.tex
TR number and removed todo about github URL

@@
 \vspace{-5px}
-\caption{It is difficult to reason abstractly given program text that uses a variety of grammar extensions (see the six reasoning criteria in the text).}
+\caption{It is difficult to reason abstractly given program text that uses a variety of grammar extensions (see the six reasoning criteria in the paper text).}
 \label{fig:K-dialect}
@@
 Sec.~\ref{sec:setlms-formally} to describe a type-theoretic account of simple expression and pattern TLMs and formally establish the reasoning principles implied above in their essential form.
-The full technical details and proofs are in the accompanying technical report \cite{icfp18tr}.\todo{fix .bib when we have the TR number}{}
+The full technical details and proofs are in the accompanying technical report \cite{icfp18tr}.
 % Sec.~\ref{sec:ptlms-formally} adds type functions and an ML-style module system to this calculus, and gives a more general variant of the reasoning principles theorem. 
@@
 % We also give a brief overview of optional integration with MetaOCaml, which can be used to guarantee that all expansion generated by a TLM are type correct.
-This implementation, called \li{Relit}, and additional implementation details, documentation and examples are available from the project page\todo{get a github.io url?}{}:
+This implementation, called \li{Relit}, and additional implementation details, documentation and examples are available from the project page:
 

### 2018-06-18 icfp18/syntax-icfp18.tex
edits to 2.0

@@
 \author{Cyrus Omar}
-\authornote{The majority of this research was performed while the first author was at Carnegie Mellon University.}          %% \authornote is optional;
+\authornote{The majority of this research was performed while the first author attended Carnegie Mellon University.}          %% \authornote is optional;
                                         %% can be repeated if necessary
@@
 
-Consider, for instance, the perspective of a programmer trying to comprehend the program text in Fig.~\ref{fig:K-dialect}, which is written in an emerging dialect of OCaml's surface syntax called Reason \cite{reason-what} that has, hypothetically, been extended with some number of new literal forms by a grammar extension system --- Lines 1-6 outline the Camlp4 mechanism \cite{de2003camlp4}, but Sugar*/SugarHaskell is similar \cite{erdweg2011sugarj,erdweg2012layout}.
+Consider, for instance, the perspective of a programmer trying to comprehend the program text in Fig.~\ref{fig:K-dialect}, which is written in an emerging dialect of OCaml's surface syntax called Reason \cite{reason-what} that has, hypothetically, been extended with some number of new literal forms by a grammar extension system --- Lines 1-6 outline the Camlp4 mechanism \cite{de2003camlp4}; Sugar*/SugarHaskell is similar \cite{erdweg2011sugarj,erdweg2012layout}.
 Line 8 uses one of these active syntax extensions to construct an encoding of a query in the rather obscure database query language K, using its intentionally terse notation~\cite{Whitney:2001:LOR:376284.375783}.
@@
 % We also give a brief overview of optional integration with MetaOCaml, which can be used to guarantee that all expansion generated by a TLM are type correct.
-This implementation, called \li{Relit}, and additional implementation details, documentation and examples are available from the project page:
+This implementation, called \li{Relit}, and additional implementation details, documentation and examples are available from the \li{Relit} project page:
 
@@
 
-Consider the recursive datatype \li{Regex.t} defined by Fig.~\ref{fig:Regex-module-def}, which encodes {regular expressions} (regexes) into Reason \cite{Thompson:1968:PTR:363347.363387}.
+Consider the recursive datatype \li{Regex.t} defined in Fig.~\ref{fig:Regex-module-def}, which encodes {regular expressions} (regexes) into Reason \cite{Thompson:1968:PTR:363347.363387}.
 Regexes are  common in, for example, bioinformatics, where they are used to express patterns in DNA sequences.
-We can construct a regex that matches the strings \li{"A"}, \li{"T"}, \li{"G"} or \li{"C"}, which represent the four bases in DNA, as follows:
+For example, we can construct a regex that matches the strings \li{"A"}, \li{"T"}, \li{"G"} or \li{"C"}, which represent the four bases in DNA, as follows:
 \begin{lstlisting}[numbers=none]
@@
 Instead, we would like to have the option to use the common POSIX-style notation  \cite{STD95954} when constructing values of type \li{Regex.t}, including values constructed compositionally from other regexes and strings.
-We solve this problem in Fig.~\ref{fig:regex-tlm-def} by defining a TLM named \li{$regex} (pronounced ``lit regex'') that supports POSIX-style regex notation extended with splice forms for regexes and strings.
+We solve this problem in Fig.~\ref{fig:regex-tlm-def} by defining a TLM named \li{$regex} (pronounced ``lit regex'') that supports POSIX-style regex notation extended with splice forms for regexes (delimited by \li{$(} and \li{)}) and strings (delimited by \li{$$(} and \li{)}).
 Fig.~\ref{fig:first-tlm-example} shows three examples of \li{$regex} being applied.
@@
     notation $regex at Regex.t {
-      lexer  RegexLexer;
-      parser RegexParser.start;
-      dependencies
-        { module Regex = Regex; }
+      lexer  RegexLexer
+      parser RegexParser.start
+      in package regex_parser;
+      dependencies = {module Regex = Regex}
     }

### 2018-06-18 icfp18/syntax-icfp18.tex
edits to 2.1.0 and fig 3 formatting

@@
 \end{lstlisting}
-Note that in Reason, the notation \li{Regex.(}$e$\li{)} locally opens the module \li{Regex} within the expression $e$, so we do not need to qualify each application of \li{Regex.Or} and \li{Regex.Str}.
-Even with the aid of this shorthand, however, constructing regexes in this way is syntactically costly.
+Note that in Reason, the notation \li{Regex.(}$e$\li{)} locally opens the module \li{Regex} within $e$, so we do not need to qualify each constructor application.
+Even with this shorthand, however, constructing regexes in this way is syntactically costly.
 Instead, we would like to have the option to use the common POSIX-style notation  \cite{STD95954} when constructing values of type \li{Regex.t}, including values constructed compositionally from other regexes and strings.
 We solve this problem in Fig.~\ref{fig:regex-tlm-def} by defining a TLM named \li{$regex} (pronounced ``lit regex'') that supports POSIX-style regex notation extended with splice forms for regexes (delimited by \li{$(} and \li{)}) and strings (delimited by \li{$$(} and \li{)}).
-Fig.~\ref{fig:first-tlm-example} shows three examples of \li{$regex} being applied.
 % One approach would be to simply define a function \li{parse_rx : string -> Regex.t} that parses a string at run-time to produce a regex.
@@
 
+Fig.~\ref{fig:first-tlm-example} shows three examples of \li{$regex} being applied.
+Line 2 applies \li{$regex} to construct the regex \li{DNA.any_base} that was  described above, this time using the more concise and common POSIX regex notation.
+Line~3 applies \li{$regex} again, using its regex splice form to compositionally construct a regex matching DNA sequences recognized by the BisA restriction enzyme, where the middle base can be any base.
+Finally, Lines~4-5 define a function, \li{restriction_template}, that constructs a more complex regex from these first two regexes and a given gene sequence represented as a string.
+
 % Moreover, there already exists a broadly adopted shorthand notational standard -- the POSIX standard for regular expressions \cite{STD95954}.
@@
 \begin{subfigure}[t]{0.45\textwidth}
-\vspace{-1px}
+\vspace{-8px}
 \begin{lstlisting}[mathescape=~]
@@
 \begin{subfigure}[t]{0.53\textwidth}
-\vspace{-1px}
+\vspace{-8px}
 \begin{lstlisting}[mathescape=|]
@@
 \begin{subfigure}[t]{\textwidth}
-\vspace{5px}
+\vspace{4px}
 \begin{lstlisting}
@@
 \end{lstlisting}
-\vspace{-4px}
+\vspace{-6px}
 \caption{Examples of the \li{$regex} TLM being applied in a bioinformatics application.
@@
 \end{subfigure}
-\vspace{0px}
+\vspace{1px}
 \caption{Case Study: POSIX-style regex literal notation, with support for string and regex splicing.}
-% \vspace{-6px}
+% \vspace{-4px}
 \label{fig:regex-case-study}
@@
 
-Line 2 of Fig.~\ref{fig:first-tlm-example} applies \li{$regex} to construct the regex \li{DNA.any_base} that was  described above, this time using the more concise and common POSIX regex notation.
-Line~3 applies \li{$regex} again, using its regex splice form to compositionally construct a regex matching DNA sequences recognized by the BisA restriction enzyme, where the middle base can be any base.
-Finally, Lines~4-5 of Fig.~\ref{fig:first-tlm-example} define a function, \li{restriction_template}, that constructs a more complex regex from these first two regexes and a given gene sequence represented as a string.
-
 Let us consider the second of these three TLM applications more closely:
@@
 \end{lstlisting}
-According to the context-free grammar of (this paper's version of) Reason, this form is simply a leaf of the unexpanded parse tree, like a string literal would be. TLM names are prefixed by \li{$} 
+According to the context-free syntax of (this paper's extension to) Reason, this form is a leaf of the unexpanded parse tree, like a string literal would be. TLM names are prefixed by \li{$} 
  to distinguish them from variables.
-We call the TLM argument, \li{`(GC$(DNA.any_base)GC)`}, a \emph{generalized literal form}, following the terminology introduced in  prior work on type-specific languages~\cite{TSLs}.
+We call the TLM argument, \li{`(GC$(DNA.any_base)GC)`}, a \emph{generalized literal form}, following our prior work on type-specific languages~\cite{TSLs}.
 The only lexical constraint imposed on the {literal body}, i.e.
@@
 This nesting constraint is to allow TLM applications to appear inside spliced expressions.
-An example of nested TLM application is shown in Fig.~\ref{fig:nesting-example}, discussed later in this section.
-Related work by \citet{TSLs}, discussed in Sec.~\ref{sec:existing-approaches}, specified several other choices of outer delimitation, including layout-sensitive delimitation, but for the purposes of this paper, \li{`(} and \li{)`} suffice.
+An example of nested TLM applications is shown in Fig.~\ref{fig:nesting-example}, discussed later in this section.
+Our prior work \cite{TSLs} specified other choices of outer delimitation, including layout-sensitive delimitation, but for  this paper \li{`(} and \li{)`} suffice.
 % Notice that in the unexpanded expression, we have not yet distinguished the spliced expression \li{any_base} from the rest of the literal.

### 2018-06-18 icfp18/syntax-icfp18.tex
edits to 2.1.1 and tweaks to spacing

@@
 \begin{subfigure}[t]{0.45\textwidth}
-\vspace{-8px}
+\vspace{-4px}
 \begin{lstlisting}[mathescape=~]
@@
 \begin{subfigure}[t]{0.53\textwidth}
-\vspace{-8px}
+\vspace{-4px}
 \begin{lstlisting}[mathescape=|]
@@
 \vspace{-5px}
-\caption{The definition of the \li{$regex} TLM. Figure \ref{fig:lexer-and-parser} defines \li{RegexLexer} and \li{RegexParser}.}
+\caption{The definition of \li{$regex}. Fig. \ref{fig:lexer-and-parser} defines \li{RegexLexer} and \li{RegexParser}, which are detailed in Sec.~\ref{sec:provider-perspective}.}
 \label{fig:regex-tlm-def}
@@
 \begin{subfigure}[t]{\textwidth}
-\vspace{4px}
+\vspace{7px}
 \begin{lstlisting}
@@
 % and take the perspective of a client that does not want to delve into these details.
-TLM definitions follow the same scoping rules as Reason modules, i.e.
-they can be defined in modules alongside other definitions and accessed through module paths.
-When a TLM definition appears inside a module with an explicitly specified module type, it must also appear in the module type with the same specification, up to the usual notions of type and module path equivalence in the type annotation and module dependencies, which are discussed below.
+TLM definitions and abbreviations (like the abbreviation on Line 1 of Fig.~\ref{fig:first-tlm-example}) follow the same scoping rules as Reason modules, i.e.
+they can appear in modules and be accessed through module paths.
+When a TLM definition appears inside a module with an explicitly specified module type (a.k.a.
+signature), it must also appear in the module type with the same specification, up to the usual notions of type and module path equivalence in the type annotation and dependencies, which are discussed below.
 This is much like the situation with datatype definitions in ML.
-In Sec.
-\ref{sec:implementation}, we will describe how we use an encoding of TLM definitions as modules with singleton signatures to avoid having to primitively extend OCaml's module system.
-
-What is fundamental about this design is that there is a well-defined protocol for finding the definition of the TLM uniquely responsible for each generalized literal form in a program, following the usual scoping rules of the language.
-An editor service or documentation tool can use this protocol to integrate TLM definition lookup into a ``go to definition'' command.
-Moreover, clients can define scoped abbreviations as shown on Line 1 of Fig.~\ref{fig:first-tlm-example}, again following the usual scoping rules.
 By convention, we define TLMs in a module suffixed with \li{Notation} so that client programmers can \li{open} just the relevant TLM definition(s) without bringing other definitions into scope.
@@
 \ref{sec:regexparser}.
+
+What is fundamental about this design is that there is a well-defined protocol that follows the usual scoping rules of the language for finding the definition of the TLM uniquely responsible for each generalized literal form in a program.
+An editor service or documentation tool could use this protocol to integrate TLM definition lookup into a ``go to definition'' command.
+In Sec.
+\ref{sec:implementation}, we will describe how we use an encoding of TLM definitions as modules with singleton signatures to avoid having to primitively extend OCaml.
 
@@
 valuable feedback on the first author's doctoral dissertation, which forms the basis of this work; 
-Charles Chamberlain, 
+Charles Chamberlain 
 for his crucial contributions to the implementation and for valuable feedback on the
-paper; and members of the PL Reading Group at University of Chicago, and the anonymous referees
-at ICFP and prior conferences, for their valuable critiques, which substantially improved the paper.
+paper; and members of the PL Reading Group at University of Chicago and the anonymous referees
+at ICFP and prior conferences for their valuable critiques, which substantially improved the paper.
 

### 2018-06-18 icfp18/syntax-icfp18.tex
fixed spelling of analogous

@@
 Forcing the programmer to reason transparently to answer basic questions like these defeats the ultimate purpose of syntactic sugar: decreasing cognitive cost \cite{Green89}.
-Analagous problems do not arise when programming without syntax extensions in languages like ML --- programmers can reason lexically about where variables and other symbols are bound, and types mediate abstraction over function and module implementations \cite{B304}.
-Ideally, the programmer would be able to abstract in some analagous manner over the implementation of an unfamiliar notation.
+analogous problems do not arise when programming without syntax extensions in languages like ML --- programmers can reason lexically about where variables and other symbols are bound, and types mediate abstraction over function and module implementations \cite{B304}.
+Ideally, the programmer would be able to abstract in some analogous manner over the implementation of an unfamiliar notation.
 % From this perspective, it is unsurprising that the use of Camlp4 has been deprecated in the OCaml system.
@@
 Having found the definition of the \li{$regex} TLM, the client can immediately determine the type of the expansion being generated at each application site because it is specified explicitly by the clause \li{at Regex.t} on Line~2 of Fig.~\ref{fig:regex-tlm-def}.
-The expansion type of a TLM is analagous to the return type of a function.
+The expansion type of a TLM is analogous to the return type of a function.
 The identity of \li{Regex.t} is determined relative to the TLM definition site, not at each application site, so the module \li{Regex} need not be in scope at the application site, or it can have been shadowed by a different module.
@@
 
-Segment types are somewhat analagous to the argument types of a function.
+Segment types are somewhat analogous to the argument types of a function.
 The difference is that the argument signature of a function is the same every time the function is applied, i.e.
@@
 The \li{SplicedE} constructor also requires a value of type \li{proto_typ}, which indicates the type that the spliced expression is expected to have.
-Types can be spliced out by using the \li{SplicedT} constructor of \li{proto_typ} analagously.
+Types can be spliced out by using the \li{SplicedT} constructor of \li{proto_typ} analogously.
 
@@
         { module Lambda = Lambda } };
-    notation $v at Lambda.v {/* analagous */}
+    notation $v at Lambda.v {/* analogous */}
   };
@@
 
-From a client programmer's perspective, reasoning principles analagous to those described in Sec.
+From a client programmer's perspective, reasoning principles analogous to those described in Sec.
 \ref{sec:client-perspective} for expression TLMs are available.
@@
 %     }
-%     notation $v at Lam.v { /* analagous */ } 
+%     notation $v at Lam.v { /* analogous */ } 
 %   };
@@
 Enforcing this restriction is straightforward: we simply have not defined a variant of the \li{proto_pat} type that encodes variable patterns (wildcards are allowed.)  This restriction ensures that only variables visible to the client in a spliced pattern are bound in the corresponding branch expression.
-This is analagous to the capture avoidance principle for expression TLMs.
+This is analogous to the capture avoidance principle for expression TLMs.
 
@@
 We define $\uGamma, \uGhyp{\ux}{x}{\tau}$ when $\uGamma = \uGG{\uG}{\Gamma}$ as an abbreviation of $\uGG{\ctxUpdate{\uG}{\ux}{x}}{\Gamma, \Ghyp{x}{\tau}}$.
-Unexpanded type formation contexts, $\uDelta$, are analagous.
+Unexpanded type formation contexts, $\uDelta$, are analogous.
 
@@
 
-% \emph{Unexpanded type formation contexts}, $\uDelta$, consist of a \emph{type identifier expansion context}, $\uD$, paired with a standard type formation context, $\Delta$, and operate analagously (see supplement.)
+% \emph{Unexpanded type formation contexts}, $\uDelta$, consist of a \emph{type identifier expansion context}, $\uD$, paired with a standard type formation context, $\Delta$, and operate analogously (see supplement.)
 % of the form $\uDD{\uD}{\Delta}$, i.e.
@@
 Rule \textsc{ee-ap-sptlm} for pattern TLM definitions shown in Fig.
-\ref{fig:ee-def-setlm} is analagous but simpler, because in $\miniVersePat$ patterns are entirely structural (there are no module paths that they might depend on).
+\ref{fig:ee-def-setlm} is analogous but simpler, because in $\miniVersePat$ patterns are entirely structural (there are no module paths that they might depend on).
 %As such, we do not need the machinery for dependency passing.
@@
 
-The typed pattern expansion rule governing pattern TLM application, Rule~\textsc{pe-ap-sptlm} in Fig.~\ref{fig:ee-ap-setlm}, is analagous but again simpler because there is no need to pass dependencies into patterns in $\miniVersePat$.\footnote{If there were a need to pass dependencies into patterns, we would need either pattern functions or to use an encoding trick, e.g.
+The typed pattern expansion rule governing pattern TLM application, Rule~\textsc{pe-ap-sptlm} in Fig.~\ref{fig:ee-ap-setlm}, is analogous but again simpler because there is no need to pass dependencies into patterns in $\miniVersePat$.\footnote{If there were a need to pass dependencies into patterns, we would need either pattern functions or to use an encoding trick, e.g.
 by requiring the result be a function containing a ``dummy'' match with one pattern.} We describe proto-pattern validation in Sec.~\ref{sec:s-PEV} below.
@@
 
-The rule for references to spliced unexpanded types and patterns are analagous (see supplement).
+The rule for references to spliced unexpanded types and patterns are analogous (see supplement).
 Note that we did not give examples of type splicing in the previous sections, but it is occasionally useful.
@@
 
-% The rule for references to spliced unexpanded patterns is entirely analagous (see supplement).
+% The rule for references to spliced unexpanded patterns is entirely analogous (see supplement).
 % {\small\begin{mathpar}
@@
 % The combined context can only be formed if these are disjoint.
-The proof proceeds by mutual rule induction and appeal to the analagous typed pattern expansion theorem and simple lemmas about type expansion   and proto-type validation.
+The proof proceeds by mutual rule induction and appeal to the analogous typed pattern expansion theorem and simple lemmas about type expansion   and proto-type validation.
 The only issue is that it is not immediately clear that the mutual induction is well-founded, because the case in the proof of part 2 for Rule \textsc{pev-spliced} invokes part 1 of the induction hypothesis on a term that is not a sub-term of the conclusion, but rather parsed out of the literal body, $b$.
@@
 
-% The calculus enjoys metatheoretic properties analagous to those described in Sec.
+% The calculus enjoys metatheoretic properties analogous to those described in Sec.
 \ref{sec:s-metatheory}, modified to account for the presence of modules, kinds and parameterization.
 The following theorem establishes the abstract reasoning principles available when applying a parametric expression TLM.
-The clauses are directly analagous to those of Theorem \ref{thm:setlm-reasoning}, so for reasons of space we do not repeat the inline descriptions.
+The clauses are directly analogous to those of Theorem \ref{thm:setlm-reasoning}, so for reasons of space we do not repeat the inline descriptions.
 The \textbf{Kinding} clauses can be understood by analogy to the \textbf{Typing} clauses.
-The details of parametric pattern TLMs (ppTLMs) are analagous (see supplement.)
+The details of parametric pattern TLMs (ppTLMs) are analogous (see supplement.)
 % \vspace{-3px}
@@
 % % The following theorem summarizes the abstract reasoning principles available to programmers when applying a pattern TLM.
-Most of the labeled clauses are analagous to those described above, so we omit their descriptions.
+Most of the labeled clauses are analogous to those described above, so we omit their descriptions.
 % % % \begingroup
@@
 
-% % % \item[{\text{(\ref{rule:cvalidE-U-tap})}}~\textbf{through}~{\text{(\ref{rule:cvalidE-U-case})}}] These cases follow analagously, i.e.
+% % % \item[{\text{(\ref{rule:cvalidE-U-tap})}}~\textbf{through}~{\text{(\ref{rule:cvalidE-U-case})}}] These cases follow analogously, i.e.
 we apply the IH, part 2 to all proto-expression validation judgements, Lemma \ref{lemma:candidate-expansion-type-validation} to all proto-type validation judgements, the identification convention to ensure that extended contexts remain disjoint, weakening and exchange as needed, and the corresponding typing rule in Rules (\ref{rule:hastypeU-tap}) through (\ref{rule:hastypeU-case}).
@@
 
-% % % \item[\VExpof{\text{\ref{rule:hastypeU-tlam}}}~\text{through}~\VExpof{\text{\ref{rule:hastypeU-case}}}] These cases follow analagously, i.e.
+% % % \item[\VExpof{\text{\ref{rule:hastypeU-tlam}}}~\text{through}~\VExpof{\text{\ref{rule:hastypeU-case}}}] These cases follow analogously, i.e.
 we apply the IH to all proto-expression validation premises, Lemma \ref{lemma:candidate-expansion-type-validation} to all proto-types validation premises, weakening and exchange as needed, and then apply the corresponding typing rule.
@@
 
-% In contrast, when a programmer encounters, for example, a function call like the call to \li{read_data} on Line 3, the analagous questions can be answered by following clear protocols that become ``cognitive reflexes'' after sufficient experience with the language, even if the programmer has no experience with the library defining \li{read_data}:
+% In contrast, when a programmer encounters, for example, a function call like the call to \li{read_data} on Line 3, the analogous questions can be answered by following clear protocols that become ``cognitive reflexes'' after sufficient experience with the language, even if the programmer has no experience with the library defining \li{read_data}:
 % \begin{enumerate}
@@
 %TLMs leave the burden of establishing the value of any particular literal notation to individual library providers, rather than to the language designer.
-The reasoning principles that TLMs provide, which are the primary contributions of this paper, allow clients to "reason around" poor literal designs, using principles analagous to those already familiar to programmers in languages like ML and Scala.
+The reasoning principles that TLMs provide, which are the primary contributions of this paper, allow clients to "reason around" poor literal designs, using principles analogous to those already familiar to programmers in languages like ML and Scala.
 Although we emphasized integration with modules, our formal system demonstrates that modules are not necessary to capture the fundamental ideas in this paper.
@@
 % We must leave a detailed empirical evaluation of the impact of particular TLMs on various quality attributes, like programmer productivity and program comprehensibility, as future work.
-% Our own implementation efforts have focused on an emerging alternative front-end for OCaml called Reason, but Scala supports analagous features \cite{conf/oopsla/AminRO14} and would also be a suitable host for the TLM mechanism.
+% Our own implementation efforts have focused on an emerging alternative front-end for OCaml called Reason, but Scala supports analogous features \cite{conf/oopsla/AminRO14} and would also be a suitable host for the TLM mechanism.
 The intended audience for this paper is language designers seeking a reasonable solution to the problem of user-defined literal notation.% We also plan to implement TLMs into \li{typy}, a typed functional language embedded into Python as a library \cite{gpce/Omar16}.%We have evaluated this claim by stating the available abstract reasoning principles formally and proving that they hold for the calculi that we have described.
@@
 
-The calculus enjoys metatheoretic properties analagous to those described in Sec.
+The calculus enjoys metatheoretic properties analogous to those described in Sec.
 \ref{sec:s-metatheory}, modified to account for the presence of modules, kinds and parameterization.
 The following theorem establishes the abstract reasoning principles available when applying a parametric expression TLM.
-The clauses are directly analagous to those of Theorem \ref{thm:setlm-reasoning}, so for reasons of space we do not repeat the inline descriptions.
+The clauses are directly analogous to those of Theorem \ref{thm:setlm-reasoning}, so for reasons of space we do not repeat the inline descriptions.
 The \textbf{Kinding} clauses can be understood by analogy to the \textbf{Typing} clauses.
-The details of parametric pattern TLMs (ppTLMs) are analagous (see supplement.)
+The details of parametric pattern TLMs (ppTLMs) are analogous (see supplement.)
 \vspace{-3px}

### 2018-06-18 icfp18/syntax-icfp18.tex
edits to 2.1.2

@@
 The expansion type of a TLM is analogous to the return type of a function.
-The identity of \li{Regex.t} is determined relative to the TLM definition site, not at each application site, so the module \li{Regex} need not be in scope at the application site, or it can have been shadowed by a different module.
+The identity of \li{Regex.t} is determined relative to the TLM definition site, not at the application site, so the module \li{Regex} need not be in scope at the application site, or it can have been shadowed by a different module.
 % In Reason, TLM definitions can refer to exactly those modules that a Reason module at the top level of the defining package would have access to, i.e.

### 2018-06-19 icfp18/syntax-icfp18.tex
edits to 2.1.3-2.1.6 and new diagram figure

@@
 \usepackage{bbm}
+\usepackage[all]{xy}
 \usepackage{wasysym}
@@
 \label{sec:context-dependence}
+
 The system enforces a strong context independence condition on generated expansions by requiring that the TLM definition explicitly specify all modules that the expansions it generates might internally depend on.
-In this case, Lines 5-7 of Fig.~\ref{fig:regex-tlm-def} specify that generated expansions might use the module \li{Regex}, again as it is bound at the TLM definition site, using the module variable \li{Regex} internally.
+In this case, Line 6 of Fig.~\ref{fig:regex-tlm-def} specifies that generated expansions might use the module \li{Regex}, again as it is bound at the TLM definition site, using the module variable \li{Regex} internally.
 In  general, the dependency can be an arbitrary module path, e.g.
@@
 The \li{Pervasives} module is implicitly opened in each expansion.
-All other bindings, both at the TLM definition site and the TLM application site, are not internally available to the expansion.
+All other bindings, both at the TLM application site and the TLM definition site, are not internally available to the expansion.
 
@@
 
-From the client programmer's perspective, the benefit of an application site context independence discipline is clear---clients do not need to give any thought to which bindings the expansion might invisibly be assuming are in scope.
-This is a common problem with various unhygienic approaches, e.g., \li{camlp4} \cite{de2003camlp4}, Sugar* \cite{erdweg2011sugarj,erdweg2013framework,erdweg2012layout}, PPX rewriters \cite{ocaml-manual} or Template Haskell \cite{SheardPeytonJones:Haskell-02,mainland2007s}.
-
-The benefits of making the macro definition site dependencies explicit, rather than implicitly allowing expansions to access all bindings at the definition site as in many existing macro systems, are more subtle.
-One technical benefit is that this exposes the TLM's dependencies in the signature of the module where the TLM is defined.
+From the client programmer's perspective, the benefit of this application site context independence discipline is clear---clients do not need to give any thought to which bindings the expansion might invisibly be assuming are in scope, as they do when using unhygienic approaches (see Sec.~\ref{sec:existing-approaches}).%, e.g., \li{camlp4} \cite{de2003camlp4}, Sugar* \cite{erdweg2011sugarj}, PPX rewriters \cite{ocaml-manual} or Template Haskell \cite{SheardPeytonJones:Haskell-02,mainland2007s}.
+
+The benefits of making the macro definition site dependencies explicit, rather than implicitly allowing expansions to access all bindings at the definition site as in many existing macro systems, arise from the fact that this exposes the TLM's dependencies in the signature of the module where the TLM is defined.
 This is useful for build tools that extract dependencies by source code analysis.
-Moreover, if an ``internal'' module is being used by expansions, then this will be manifest in the signature and can be corrected if this was not intended.
+Moreover, if an ``internal'' module is being used by expansions, then this will be manifest in the signature and can be corrected if this was unintended.
 % (Similar concerns about unintended capture drives recent work on explicit closures for distributed computing \cite{DBLP:conf/ecoop/MillerHO14}.) 
-Implicit access to the  definition site would require the implementation to carefully ``smuggle'' internal values to each application site, thereby skirting the abstraction discipline of ML's module system \cite{culpepper2005syntactic}.% (and making the implementation substantially more complex!)\todo{I think there is some work on this...}  %This, in turn, ensures that the expansion does not depend on bindings inaccessible outside the module, i.e.
+Implicit access to the  definition site would require the implementation to carefully ``smuggle out'' otherwise internal values to each application site, thereby skirting the abstraction discipline of ML's module system \cite{culpepper2005syntactic}.% (and making the implementation substantially more complex!)\todo{I think there is some work on this...}  %This, in turn, ensures that the expansion does not depend on bindings inaccessible outside the module, i.e.
 the implementation does not need to ``smuggle'' references to these internal definitions to each application site.
@@
 
-Another benefit of this explicit approach is that renamings can be propagated without needing to modify AST encodings in parser definitions.
-For example, if we decided to rename the module \li{Regex} to \li{Regexp}, this renaming would need only to propagate to the module paths in the TLM definition in the usual manner as long as the names of the internal module variables remain the same.
-(The TLM provider may wish, for the sake of aesthetics, to propagate the change into the parser definition as well, but this is not semantically necessary).
-This allows us to encode expansions using ordinary OCaml values (see Sec.
-\ref{sec:provider-perspective}).
+\begin{figure}
+\vspace{-5px}
+\begin{displaymath}
+\begin{array}{l}
+  \text{\li{$regex `(GC$(DNA.any_base)GC)`}}\\[-10px]
+  \xymatrix{
+  ~ \ar[d]^{\text{The applied TLM lexes and parses the literal body to generate the \textbf{proto-expansion}.}} \\ ~
+  % \text{\li{Regex.Seq(Regex.Str "GC", Regex.Seq(spliced<4..15 : Regex.t>, Regex.Str "GC"))}} \ar[d]^f\\ 
+  }\\[-5px]
+  \text{\li{Regex.Seq(Regex.Str "GC", Regex.Seq(spliced<4..15 : Regex.t>, Regex.Str "GC"))}}\\[-10px]
+  \xymatrix{
+  ~ \ar[d]^{\text{The system performs proto-expansion validation and generates the \textbf{final expansion}.}} \\ ~
+  % \text{\li{Regex.Seq(Regex.Str "GC", Regex.Seq(spliced<4..15 : Regex.t>, Regex.Str "GC"))}} \ar[d]^f\\ 
+  }\\[-5px]
+  \text{\li{Regex.Seq(Regex.Str "GC", Regex.Seq(DNA.any_base, Regex.Str "GC"))}}
+\end{array}
+\end{displaymath}
+\caption{Expansion of TLM applications occurs in two steps.}
+\vspace{-5px}
+\label{fig:expansion}
+\end{figure}
 
@@
 Na\"ively checking that only the explicitly named dependencies are free in the expansion would inappropriately constrain application site spliced expressions, which should certainly not be prevented from referring to variables in scope at the application site.
-For example, consider the final expansion of our example from Line 2 of Fig.~\ref{fig:first-tlm-example}:
-\begin{lstlisting}[numbers=none]
-  Regex.Seq(Regex.Str "GC", Regex.Seq(DNA.any_base, Regex.Str "GC"))
-\end{lstlisting}
+For example, consider the  bottom of Fig.~\ref{fig:expansion}, which shows the final expansion of the example from Line 2 of Fig.~\ref{fig:first-tlm-example}.
+% \begin{lstlisting}[numbers=none]
+%   Regex.Seq(Regex.Str "GC", Regex.Seq(DNA.any_base, Regex.Str "GC"))
+% \end{lstlisting}
 In this term, both \li{Regex} and \li{DNA} are free module variables.
@@
 
-Hygienic term-rewriting macro systems, like those available in various Lisp-family languages \cite{mccarthy1978history} and in Scala \cite{ScalaMacros2013}, cannot be used to repurpose string literals for composite literal notation at other types for exactly this reason.
-These systems would find that the appearance of \li{DNA.any_base}  violates their context independence condition (which is one aspect of {hygiene}), because \li{DNA.any_base} is not, from the perspective of the context-free syntax, an argument to the macro nor even a sub-term of an argument for which a tree path can be assigned \cite{DBLP:conf/esop/HermanW08,Herman10:Theory,gorn1965explicit}.
-Instead, it arises as the result of performing a complex operation---parsing---on an arbitrary sub-sequence of the string literal passed into the macro for rewriting.
-%From the macro expander's perspective, it has appeared by magic''.
-
-To address this problem, TLM parsers are not tasked with generating the final expansion directly.
-Instead, the parser generates a \emph{proto-expansion} that refers to spliced terms indirectly by location relative to the start of the provided literal body.
-For example, the proto-expansion generated by \li{$regex} for the example above can be pretty printed as follows:
-\begin{lstlisting}[numbers=none]
-  Regex.Seq(Regex.Str "GC", Regex.Seq(spliced<4..15 : Regex.t>, Regex.Str "GC"))
-\end{lstlisting}
+To address this problem, TLM parsers do not generate the final expansion directly, but rather a \emph{proto-expansion} that refers to spliced terms indirectly by location relative to the start of the provided literal body.
+For example, the proto-expansion generated by \li{$regex} for the example above can be pretty printed as shown in the middle of Fig.~\ref{fig:expansion}.
+% \begin{lstlisting}[numbers=none]
+%   Regex.Seq(Regex.Str "GC", Regex.Seq(spliced<4..15 : Regex.t>, Regex.Str "GC"))
+% \end{lstlisting}
 Here, \lismall{spliced<4..15 : Regex.t>} is a reference to the spliced expression \li{DNA.any_base} because characters $4$ through $15$, zero-indexed and inclusive, of \lismall{GC$(DNA.any_base)GC} are \lismall{DNA.any_base}.
@@
 \ref{sec:segment-typing} below.
-The context independence condition can be enforced straightforwardly on the proto-expansion---the only free variable in the proto-expansion is \li{Regex}, which is an explicitly listed dependency, so all is well.
+The context independence condition can be enforced straightforwardly on the proto-expansion---the only free variable in the proto-expansion is \li{Regex}, which is an explicitly listed dependency in Fig.~\ref{fig:regex-tlm-def}, so all is well.
 
@@
 \caption{A practical demonstration of nested TLM application. The colors communicate the segmentation.}
-\vspace{-3px}
 \label{fig:nesting-example}
@@
 
-Segment types can be communicated directly to the programmer upon request by an editor service.
-For Reason, we are equipping the {Merlin} tool \cite{Merlin}, which is used by various Reason editor modes (e.g.
-for Emacs, Vim, and so on), with a new editor command that reports the expected type of the innermost spliced segment containing the cursor.
-Note that because the type is explicit, this information can be reported even when there is a parse error or type error in a spliced expression.
+Segment types can be communicated directly to the programmer upon request by an editor service For Reason, we are equipping the {Merlin} tool \cite{Merlin}, which is used by various Reason editor modes (e.g.
+for Emacs and Vim), with a new editor command that reports the expected type of the innermost spliced segment containing the cursor.
+Note that because the type is explicit, this information can be reported even when there is a parse or type error in a spliced expression.
 % An alternative design that simply reported the type assigned to each spliced expression based on the final expansion (which already exists in Merlin) would be less useful in situations where there is a type error in a spliced expression, or the programmer is in the process of editing the program and wants to know what type of expression to construct.
@@
 it is associated with the function itself, whereas the inferred segmentation can differ for each choice of literal body.
-This, of course, gives TLMs substantially more notational flexibility, even relative to infix or mixfix function notation (where the number of subexpressions is fixed \cite{wieland2009parsing}), as intended.% (even infix or mixfix functions \cite{wieland2009parsing}).
+This, of course, gives TLMs substantially more notational flexibility, even relative to infix or mixfix function notation (where the number of subexpressions is fixed \cite{wieland2009parsing}).% (even infix or mixfix functions \cite{wieland2009parsing}).
 
@@
 \end{lstlisting}
-%Notice that the expansion-internal binding of \li{tmp} has been alpha-varied to \li{tmp'}.
+Notice that the expansion-internal binding of \li{tmp} has been alpha-varied to \li{tmp_fresh}.
 The reference to \li{tmp} in the spliced expression then refers, as intended, to the application site binding.
@@
 Reason's concise lambda notation, \li{(x) => e}, decreases the syntactic cost of this approach.
-For example, we cannot define list comprehension notation like the following:
+For example, we cannot define list comprehension notation like the following because the binding site of \li{x} is not clear:
 \begin{lstlisting}[numbers=none,deletekeywords={in}]
@@
 
-Our contention is that small syntactic costs like these are more than justified by the peace of mind of knowing that unfamiliar literal notation cannot possibly be performing ``magic'' on the type and binding structure of the language.
+Our contention is that small syntactic costs like these are more than justified by the peace of mind of knowing that unfamiliar literal notation cannot possibly be performing ``magic'' with the type and binding structure.
 We say more about the future prospect of a  mechanism designed specifically for shorthand binding forms, e.g.
@@
 
+ \todo{integrate me}{}Another benefit of this explicit approach is that renamings can be performed without needing to reason about or modify parse tree encodings in parser definitions.
+For example, if we decided to rename the module \li{Regex} to \li{Regexp}, this renaming would need only to be performed on the module paths in the TLM definition, as long as the name of the internal module variable in the dependencies list remains \li{Regex}.
+%(The TLM provider may wish, for the sake of aesthetics, to propagate the change into the parser definition as well, but this is not semantically necessary). 
+This allows us to encode expansions using ordinary OCaml values (in fact, using the OCaml compiler's own parse tree representation), rather than requiring the introduction of primitive quasiquotation into OCaml (see Sec.
+\ref{sec:provider-perspective}).
+
 \subsubsection{Correctness Criteria}
@@
 \cvalidE{\Delta}{\Gamma}{\escenev}{\ce}{e}{\tau} & \text{$\ce$ has expansion $e$ of type $\tau$}\\
-% \cvalidR{\Delta}{\Gamma}{\escenev}{\crv}{r}{\tau}{\tau'} & \text{$\crv$ has expansion $r$ taking values of type $\tau$ to values of type $\tau'$}\\
+% \cvalidR{\Delta}{\Gamma}{\escenev}{\crulev}{r}{\tau}{\tau'} & \text{$\crulev$ has expansion $r$ taking values of type $\tau$ to values of type $\tau'$}\\
 \cvalidP{\upctx}{\pscenev}{\cpv}{p}{\tau} & \text{$\cpv$ has expansion $p$ matching $\tau$}
@@
 
-Notice that proto-rules, $\crv$, involve expanded patterns, $p$, not proto-patterns, $\cpv$, because proto-rules appear in proto-expressions, which are generated by expression TLMs.
+Notice that proto-rules, $\crulev$, involve expanded patterns, $p$, not proto-patterns, $\cpv$, because proto-rules appear in proto-expressions, which are generated by expression TLMs.
 Proto-patterns arise only from pattern TLMs.
@@
 % \acein{\ell}{\ce} ~\vert~
-\acematchwith{n}{\ce}{\seqschemaX{\crv}} ~\vert~
+\acematchwith{n}{\ce}{\seqschemaX{\crulev}} ~\vert~
 \acesplicede{m}{n}{\ctau}\\
-\mathsf{PrRule} & \crv & ::= & \acematchrule{p}{\ce}\\
+\mathsf{PrRule} & \crulev & ::= & \acematchrule{p}{\ce}\\
 \mathsf{PrPat} & \cpv & ::= & \acewildp ~\vert~
@@
 
+\todo{integrate me}{}Hygienic term-rewriting macro systems, like those available in various Lisp-family languages \cite{mccarthy1978history} and in Scala \cite{ScalaMacros2013}, cannot be used to repurpose string literals for composite literal notation at other types for exactly this reason.
+These systems would find that the appearance of \li{DNA.any_base}  violates their context independence condition (which is one aspect of {hygiene}), because \li{DNA.any_base} is not, from the perspective of the context-free syntax, an argument to the macro nor even a sub-term of an argument for which a tree path can be assigned \cite{DBLP:conf/esop/HermanW08,Herman10:Theory,gorn1965explicit}.
+Instead, it arises as the result of performing a complex operation---parsing---on an arbitrary sub-sequence of the string literal passed into the macro.
+(It is  possible for a macro to selectively opt out of the hygiene discipline---see Sec.~\ref{sec:existing-approaches} for more on why relying on this sort of ``escape hatch'' is an unreasonable solution.) %From the macro expander's perspective, it has appeared by magic''.
+
 TLMs also consider the problem of definition-site context independence by specifying expansion dependencies explicitly.

### 2018-06-19 icfp18/syntax-icfp18.tex
moved note about first class modules to 2.1.2

@@
 
+It is also worth noting here that although there is no direct mechanism for applying TLMs at the module level, this can be achieved by using OCaml's first class modules \cite{ocaml-manual}.
+
 \subsubsection{Context Dependence}
@@
 \begin{figure}
-\vspace{-5px}
+\vspace{-10px}
 \begin{displaymath}
@@
 \end{displaymath}
+\vspace{-4px}
 \caption{Expansion of TLM applications occurs in two steps.}
-\vspace{-5px}
+\vspace{-6px}
 \label{fig:expansion}
@@
 It also details how the \li{$html} TLM shown previously is implemented using an existing, ``production grade'' HTML parsing library, \li{Markup.ml} \cite{markupml}, together with a simple post-processing step.
-It is also worth noting that although there is no mechanism for applying TLMs at the module level, this can be emulated by using OCaml's first class modules.
 

### 2018-06-21 icfp18/syntax-icfp18.tex
more edits to 2.1 and changed indexing notation to be consistent with relit

@@
 % Also available is an additional series of case studies: notation for an encoding of regular expressions as an abstract data type, implemented using TLMs for quasiquotation and grammar-based parser generators.
-%are also available in the supplement.
+%are also available in the accompanying technical report.
 Sec.~\ref{sec:static-eval} considers the topic of term evaluation during the  expansion phase, i.e.
@@
 %  We say more about LISP macros in Sec.
-\ref{sec:existing-approaches}.% For the present purposes, we can consider Reason as essentially the calculus defined in the supplement extended with various conveniences that are commonly found in other ML-like languages and, notionally, orthogonal to TLMs.
+\ref{sec:existing-approaches}.% For the present purposes, we can consider Reason as essentially the calculus defined in the accompanying technical report extended with various conveniences that are commonly found in other ML-like languages and, notionally, orthogonal to TLMs.
 %Reason is, as its name suggests, a conceptual descendent of ML.
@@
            | Star(t);
-  }
+  };
 \end{lstlisting}
@@
       dependencies = {module Regex = Regex}
-    }
+    };
   }
@@
 \end{lstlisting}
-\vspace{-6px}
+\vspace{-4px}
 \caption{Examples of the \li{$regex} TLM being applied in a bioinformatics application.
@@
 The \li{Pervasives} module is implicitly opened in each expansion.
-All other bindings, both at the TLM application site and the TLM definition site, are not internally available to the expansion.
+All other bindings, whether at the TLM application site or the TLM definition site, are not internally available to the expansion.
 
@@
 
-From the client programmer's perspective, the benefit of this application site context independence discipline is clear---clients do not need to give any thought to which bindings the expansion might invisibly be assuming are in scope, as they do when using unhygienic approaches (see Sec.~\ref{sec:existing-approaches}).%, e.g., \li{camlp4} \cite{de2003camlp4}, Sugar* \cite{erdweg2011sugarj}, PPX rewriters \cite{ocaml-manual} or Template Haskell \cite{SheardPeytonJones:Haskell-02,mainland2007s}.
+From the client's perspective, the benefit of this application site context independence discipline is clear---clients do not need to give any thought to which bindings the expansion might invisibly be assuming are in scope, as they do when using unhygienic approaches (see Sec.~\ref{sec:existing-approaches}).%, e.g., \li{camlp4} \cite{de2003camlp4}, Sugar* \cite{erdweg2011sugarj}, PPX rewriters \cite{ocaml-manual} or Template Haskell \cite{SheardPeytonJones:Haskell-02,mainland2007s}.
 
@@
   ~ \ar[d]^{\text{The applied TLM lexes and parses the literal body to generate the \textbf{proto-expansion}.}} \\ ~
-  % \text{\li{Regex.Seq(Regex.Str "GC", Regex.Seq(spliced<4..15 : Regex.t>, Regex.Str "GC"))}} \ar[d]^f\\ 
   }\\[-5px]
-  \text{\li{Regex.Seq(Regex.Str "GC", Regex.Seq(spliced<4..15 : Regex.t>, Regex.Str "GC"))}}\\[-10px]
+  \text{\li{Regex.Seq(Regex.Str "GC", Regex.Seq(spliced<4..16 : Regex.t>, Regex.Str "GC"))}}\\[-10px]
   \xymatrix{
-  ~ \ar[d]^{\text{The system performs proto-expansion validation and generates the \textbf{final expansion}.}} \\ ~
-  % \text{\li{Regex.Seq(Regex.Str "GC", Regex.Seq(spliced<4..15 : Regex.t>, Regex.Str "GC"))}} \ar[d]^f\\ 
+  ~ \ar[d]^{\text{The system performs proto-expansion validation and recursively expands the splices to generate the final \textbf{expansion}.}} \\ ~
   }\\[-5px]
@@
 \vspace{-4px}
-\caption{Expansion of TLM applications occurs in two steps.}
+\caption{TLM applications expand in two steps.}
 \vspace{-6px}
@@
 % \end{lstlisting}
-Here, \lismall{spliced<4..15 : Regex.t>} is a reference to the spliced expression \li{DNA.any_base} because characters $4$ through $15$, zero-indexed and inclusive, of \lismall{GC$(DNA.any_base)GC} are \lismall{DNA.any_base}.
-We return to the type annotation on the splice reference, \li{Regex.t}, in Sec.
+Here, \lismall{spliced<4..16 : Regex.t>} is a reference to the spliced expression \li{DNA.any_base} because the zero-indexed subsequence $[4, 16)$ of \lismall{GC$(DNA.any_base)GC} is \lismall{DNA.any_base}.
+We return to the type annotation on the splice reference, \li{Regex.t}, when we discuss \textbf{Segment Typing} in Sec.
 \ref{sec:segment-typing} below.
-The context independence condition can be enforced straightforwardly on the proto-expansion---the only free variable in the proto-expansion is \li{Regex}, which is an explicitly listed dependency in Fig.~\ref{fig:regex-tlm-def}, so all is well.
+The context independence condition can be enforced directly on the proto-expansion---the only free variable in the proto-expansion is \li{Regex}, which is an explicitly listed dependency in Fig.~\ref{fig:regex-tlm-def}, so all is well.
 
@@
 The finite set of splice references in the proto-expansion generated for a literal body is called the \emph{segmentation} of that literal body.
-The segmentation of the example above is the finite set containing one splice reference, \lismall{spliced<4..15 : Regex.t>}.
+The segmentation of the example above is the finite set containing one splice reference, \lismall{spliced<4..16 : Regex.t>}.
 For the more complex example from Line 5 of Fig.~\ref{fig:first-tlm-example}, the segmentation contains five splice references:
 \begin{lstlisting}[numbers=none]
-  { spliced<2..5 : Regex.t>, spliced<9..20 : Regex.t>, spliced<26..29 : string>, 
-    spliced<33..44 : Regex.t>, spliced<49..52 : Regex.t> } 
+  { spliced<2..6 : Regex.t>, spliced<9..21 : Regex.t>, spliced<26..30 : string>, 
+    spliced<33..45 : Regex.t>, spliced<49..53 : Regex.t> } 
 \end{lstlisting}
@@
 For example, Fig.~\ref{fig:nesting-example} shows a program fragment where we transform an encoding of a chemical structure expressed using the standard SMILES notation for chemical structures \cite{anderson1987smiles}, extended with splicing notation, into a vector graphic, then embed this directly into a fragment of a webpage.
-We detail the TLM \li{$html at Html.t}, which implements HTML notation similar to that found in Ur/Web, in the supplement, and assume a TLM, \li{$smiles at Smiles.t} and a function, \li{Smiles.to_svg : Smiles.t => Html.t}, not shown.
+We detail the TLM \li{$html at Html.t}, which implements HTML notation similar to that found in Ur/Web, in the accompanying technical report, and assume a TLM, \li{$smiles at Smiles.t} and a function, \li{Smiles.to_svg : Smiles.t => Html.t}, not shown.
 In Reason, \li{|>} is reverse function application.
@@
 \end{lstlisting}
+\vspace{-2px}
 \caption{A practical demonstration of nested TLM application. The colors communicate the segmentation.}
@@
 
-Segment types can be communicated directly to the programmer upon request by an editor service For Reason, we are equipping the {Merlin} tool \cite{Merlin}, which is used by various Reason editor modes (e.g.
+Segment types can be communicated directly to the programmer upon request by an editor service For Reason, we are equipping the {Merlin} tool\todo{cite ICFP merlin paper}{} \cite{Merlin}, which is used by various Reason editor modes (e.g.
 for Emacs and Vim), with a new editor command that reports the expected type of the innermost spliced segment containing the cursor.
@@
   let tmp = /* ... expansion-internal temporary ... */;
-  Html.H1Element(tmp, spliced<7..12 : Html.t>);
+  Html.H1Element(tmp, spliced<7..13 : Html.t>);
 \end{lstlisting}
@@
 Instead, there is are separate \emph{static bindings} marked by the \li{static} keyword that populate a \emph{static environment} that is discarded after expansion finishes.
-A more detailed account of static evaluation, both informal and informal, is given in the supplement.
+A more detailed account of static evaluation, both informal and informal, is given in the accompanying technical report.
 An alternative design that allows for the explicit lowering of standard-phase modules to the static phase has also been proposed for OCaml \cite{Ocaml/macros}.
@@
   notation $re_expr at Parsetree.expression { /* ... (see text) ... */ };
-  notation $re_type at Parsetree.core_type { /* ... (see text) ... */ };
   notation $proto_expr at ProtoExpr.t = $re_expr;
@@
 In the literature, literal notation for the host language's own parse trees is called \emph{quasiquotation}, and splicing is referred to as \emph{unquotation} or \emph{antiquotation} \cite{Bawd99a,shabalin2013quasiquotes,mainland2007s}.
-The TLMs \li{$re_expr} and \li{$re_type} specified on Lines 10-11 of Fig.~\ref{fig:relit-util} implement quasiquotation for OCaml syntax trees using Reason's surface syntax.
-The supplement describes their implementation.
+The TLM \li{$re_expr}  specified on Line 10 of Fig.~\ref{fig:relit-util} provides quasiquotation for OCaml parse trees using Reason's surface syntax.
+The accompanying technical report describes its implementation.
 For clarity, the \li{$proto_expr} TLM is defined as a synonym for \li{$re_expr} on Line~12 of Fig.~\ref{fig:relit-util} (the two TLMs may diverge in the future).
@@
 the \li{Relit} library includes a trivial lexer, \li{Relit.TrivLexer}, and a few functors that take care of the necessary boilerplate in situations where we simply want to define the parser as a function of type \li{string => Relit.ProtoExpr.t}.
-For example, the supplement briefly describes how \li{$re_expr} can be expressed in terms of the Reason parser, a parse tree serializer and a tree transformation.
+For example, the accompanying technical report briefly describes how \li{$re_expr} can be expressed in terms of the Reason parser, a parse tree serializer and a tree transformation.
 It also details how the \li{$html} TLM shown previously is implemented using an existing, ``production grade'' HTML parsing library, \li{Markup.ml} \cite{markupml}, together with a simple post-processing step.
@@
 Consequently, encodings of proto-terms can be values of standard datatypes (e.g.
-\li{proto_expr}) with variables represented simply as, e.g., strings, and quasiquotation notation can be expressing using TLMs (see supplement).
+\li{proto_expr}) with variables represented simply as, e.g., strings, and quasiquotation notation can be expressing using TLMs (see accompanying technical report).
 A central design goal of Reason is to leave the OCaml semantics unchanged, so building quasiquotation in primitively, to integrate the free variables in term encodings into the overall binding discipline \cite{Bawd99a}, was in any case infeasible.
@@
 To support splicing, we need non-terminals that recognize unexpanded terms and produce the corresponding splice references, rather than the AST itself.
-This requires that the parser generator keep track of location information (as most production-grade parser generators already do for error reporting.) A more detailed example of this mechanism being used to define a TLM for a modular encoding of regular expressions is given in the supplement.% For spliced expressions, this non-terminal would need to be a family of non-terminals indexed by a value of type \li{proto_typ}.
+This requires that the parser generator keep track of location information (as most production-grade parser generators already do for error reporting.) A more detailed example of this mechanism being used to define a TLM for a modular encoding of regular expressions is given in the accompanying technical report.% For spliced expressions, this non-terminal would need to be a family of non-terminals indexed by a value of type \li{proto_typ}.
 
@@
 The substitution $\omega$ is applied to the parse function after it is expanded.
-The full details are defined as a small patch of $\miniVerseParam$ called $\miniVersePH$ in the supplement.
+The full details are defined as a small patch of $\miniVerseParam$ called $\miniVersePH$ in the accompanying technical report.
 
@@
 
-% Because our focus is on these reasoning principles, and for reasons of space, we leave our full calculus of parametric TLMs to the supplement.
+% Because our focus is on these reasoning principles, and for reasons of space, we leave our full calculus of parametric TLMs to the accompanying technical report.
 The full calculus extends the simple calculus of this section with an ML module calculus based closely on the system defined by \citet{pfple1}, which in turn is based on early work by \citet{MacQueen:1984:MSM:800055.802036,DBLP:conf/popl/MacQueen86}, subsequent work on the phase splitting interpretation of modules \cite{harper1989higher} and on using dependent singleton kinds to track type identity \cite{stone2006extensional,DBLP:conf/lfmtp/Crary09}, and finally on formal developments by \citet{dreyer2005understanding} and \citet{conf/popl/LeeCH07}.
@@
 \ref{sec:ptsms}.
-Proofs are in the supplement for both the simple and full calculus.
+Proofs are in the accompanying technical report for both the simple and full calculus.
 
@@
 \ref{fig:U-expanded-terms} gives the syntax of the core language, $\miniVerseBase$.
-This language forms a standard pure functional language with partial function types, quantification over types, recursive types, labeled product types, labeled sum types and support for pattern matching.\footnote{The supplement shows how to remove pattern matching  (and pattern TLMs) from the calculus by adding the usual elimination forms, e.g.
+This language forms a standard pure functional language with partial function types, quantification over types, recursive types, labeled product types, labeled sum types and support for pattern matching.\footnote{The accompanying technical report shows how to remove pattern matching  (and pattern TLMs) from the calculus by adding the usual elimination forms, e.g.
 $\unfold{e}$ rather than the $\aefoldp{p}$ pattern for values of recursive type.} Formally, core language terms are {abstract binding trees} (ABTs) identified up to alpha-equivalence, so we follow the syntactic conventions of \citet{pfple1}.
@@
 In the pattern typing judgement, $\pctx$ collects the typing hypotheses generated by $p$.
-These judgements are inductively defined in the supplemental material along with necessary auxiliary structures and  standard lemmas.
+These judgements are inductively defined in the accompanying technical reportal material along with necessary auxiliary structures and  standard lemmas.
 The dynamic semantics of $\miniVerseBase$ is organized around the judgements $\isvalU{e}$, which says that $e$ is a value, and $\evalU{e}{e'}$, which says that $e$ evaluates to the value $e'$.
 As in ML, evaluation can diverge (general recursion is possible via recursive types \cite{pfple1}).
-It can also result in match failure.%\emph{Type formation contexts}, $\Delta$, and \emph{typing contexts}, $\Gamma$ are defined in the supplement in the standard way \cite{pfpl},
+It can also result in match failure.%\emph{Type formation contexts}, $\Delta$, and \emph{typing contexts}, $\Gamma$ are defined in the accompanying technical report in the standard way \cite{pfpl},
 % \[\begin{array}{ll}
@@
 % \emph{Typing contexts}, $\Gamma$, are finite functions that map each variable $x \in \domof{\Gamma}$, where $\domof{\Gamma}$ is a finite set of variables, to the hypothesis $\Ghyp{x}{\tau}$, for some $\tau$.
-The judgements above are inductively defined in the supplemental material and validate standard lemmas, also given in the supplement.
+The judgements above are inductively defined in the accompanying technical reportal material and validate standard lemmas, also given in the accompanying technical report.
 %Empty typing contexts are written $\emptyset$, or omitted entirely within judgements, and non-empty typing contexts are written as finite sequences of hypotheses identified up to exchange and contraction.
@@
 
-% These judgements are inductively defined in the supplement. 
+% These judgements are inductively defined in the accompanying technical report. 
 % \begingroup 
@@
 We refer to these as the \emph{common forms}.
-% The mapping from expanded forms to common unexpanded forms is defined explicitly in the supplement.
+% The mapping from expanded forms to common unexpanded forms is defined explicitly in the accompanying technical report.
 The typed expansion rules that handle common forms mirror the corresponding typing rules.
@@
 For example, the rules for variables and lambdas are shown being applied in the example derivation in Fig.~\ref{fig:expansion-exmpl}, discussed below.
-The full set of rules is in the supplement.
+The full set of rules is in the accompanying technical report.
 %Each of these rules is based on the corresponding typing rule, i.e.
@@
 
-% \emph{Unexpanded type formation contexts}, $\uDelta$, consist of a \emph{type identifier expansion context}, $\uD$, paired with a standard type formation context, $\Delta$, and operate analogously (see supplement.)
+% \emph{Unexpanded type formation contexts}, $\uDelta$, consist of a \emph{type identifier expansion context}, $\uD$, paired with a standard type formation context, $\Delta$, and operate analogously (see accompanying technical report.)
 % of the form $\uDD{\uD}{\Delta}$, i.e.
@@
 
-% The \emph{type expansion judgement}, $\expandsTU{\uDelta}{\utau}{\tau}$, is inductively defined by the rules given in the supplement. The first three of these rules are reproduced below:
+% The \emph{type expansion judgement}, $\expandsTU{\uDelta}{\utau}{\tau}$, is inductively defined by the rules given in the accompanying technical report.
+The first three of these rules are reproduced below:
 % % \begin{subequations}%\label{rules:expandsTU}
@@
 \ref{sec:ce-syntax-U}.
-As with $\tBody$, we need only take as a condition that there is an isomorphism between values of type $\tCEExp$ and closed proto-expressions, which is mediated in one direction by the \emph{proto-expression decoding judgement}, $\decodeCondE{e}{\ce}$ that we will return to when we describe TLM application below (see supplement for the full condition).
+As with $\tBody$, we need only take as a condition that there is an isomorphism between values of type $\tCEExp$ and closed proto-expressions, which is mediated in one direction by the \emph{proto-expression decoding judgement}, $\decodeCondE{e}{\ce}$ that we will return to when we describe TLM application below (see accompanying technical report for the full condition).
 
@@
 Second, the rule extends the expression TLM context, $\uPsi$, to associate the TLM identifier $\tsmv$ with $x$, as well as with the given expansion type and parse function.
-If $\tsmv$ was already defined, the previous definition is shadowed (the supplement gives some additional details on how TLM contexts are structured).
+If $\tsmv$ was already defined, the previous definition is shadowed (the accompanying technical report gives some additional details on how TLM contexts are structured).
 
@@
 The first of these two premises determines the segmentation of the proto-expansion by pulling out the splice references and ensures that it is valid via the predicate $\segOK{\psi}{b}$, where $\psi$ is the segmentation.
-This checks that each segment  has positive length and is within bounds of $b$, and that the segments do not overlap and are separated by at least one character (see supplement).
+This checks that each segment  has positive length and is within bounds of $b$, and that the segments do not overlap and are separated by at least one character (see accompanying technical report).
 
@@
 \paragraph{Common Forms} Most of the proto-expansion forms, including all of those elided in Fig.~\ref{fig:U-candidate-terms} mirror corresponding expanded forms.
-The rules governing proto-expansion validation for these common forms  in the supplement correspondingly mirror the typing rules.
+The rules governing proto-expansion validation for these common forms  in the accompanying technical report correspondingly mirror the typing rules.
 Splicing scenes---$\escenev$, $\tscenev$ and $\pscenev$--pass opaquely through these rules, i.e.
@@
 
-The rule for references to spliced unexpanded types and patterns are analogous (see supplement).
+The rule for references to spliced unexpanded types and patterns are analogous (see accompanying technical report).
 Note that we did not give examples of type splicing in the previous sections, but it is occasionally useful.
@@
 
-% The rule for references to spliced unexpanded patterns is entirely analogous (see supplement).
+% The rule for references to spliced unexpanded patterns is entirely analogous (see accompanying technical report).
 % {\small\begin{mathpar}
@@
 % More specifically, {the sum of the lengths of the literal bodies that appear in the term strictly decreases each time you perform a nested TLM application} because some portion of the term has to be consumed by the TLM name and the delimiters.
-The details are given in the supplement.
+The details are given in the accompanying technical report.
 
@@
 
-% The proof, which involves auxiliary lemmas about the decomposition of proto-types and proto-expressions, is given in the supplement.
+% The proof, which involves auxiliary lemmas about the decomposition of proto-types and proto-expressions, is given in the accompanying technical report.
 
@@
 \ref{sec:setlms-formally}, so we highlight only the most important concepts below.
-The details are in the supplement.
+The details are in the accompanying technical report.
 
@@
 % Normalization eliminates parameters introduced in higher-order abbreviations, leaving only those parameter applications specified by the original TLM definition.
-Normal forms and progress and preservation theorems are established in the supplement.
+Normal forms and progress and preservation theorems are established in the accompanying technical report.
 
@@
 The \textbf{Kinding} clauses can be understood by analogy to the \textbf{Typing} clauses.
-The details of parametric pattern TLMs (ppTLMs) are analogous (see supplement.)
+The details of parametric pattern TLMs (ppTLMs) are analogous (see accompanying technical report.)
 % \vspace{-3px}
@@
 % % \begin{proof} 
-% % The proof, in the supplement, relies on an auxiliary lemma about decomposing proto-patterns.
+% % The proof, in the accompanying technical report, relies on an auxiliary lemma about decomposing proto-patterns.
 
@@
 This class includes all of the examples enumerated in Sec.
-\ref{sec:intro} (up to the choice of outermost delimiter), the varied and non-trivial case studies outlined in this paper and in the supplement, and the examples collected from the empirical study by \citet{TSLs}.
+\ref{sec:intro} (up to the choice of outermost delimiter), the varied and non-trivial case studies outlined in this paper and in the accompanying technical report, and the examples collected from the empirical study by \citet{TSLs}.
 We anticipate many more interesting examples will emerge as the Reason community explores the mechanism.
@@
 
-% Other directions for future work are given in the supplement\todo{do this?}.
+% Other directions for future work are given in the accompanying technical report\todo{do this?}.
 
@@
 would like to thank Robert Harper, Karl Crary and Eric Van Wyk for 
-valuable feedback on the first author's doctoral dissertation, which forms the basis of this work; 
+valuable feedback on the first author's doctoral dissertation \cite{omar-thesis}, which forms the basis of this work; 
 Charles Chamberlain 
@@
 paper; and members of the PL Reading Group at University of Chicago and the anonymous referees
-at ICFP and prior conferences for their valuable critiques, which substantially improved the paper.
+ for their thoughtful critiques, which substantially improved the paper.
 
@@
 Normalization eliminates parameters introduced in higher-order abbreviations, leaving only those parameter applications specified by the original TLM definition.
-Normal forms and progress and preservation theorems are established in the supplement.
+Normal forms and progress and preservation theorems are established in the accompanying technical report.
 
@@
 The \textbf{Kinding} clauses can be understood by analogy to the \textbf{Typing} clauses.
-The details of parametric pattern TLMs (ppTLMs) are analogous (see supplement.)
+The details of parametric pattern TLMs (ppTLMs) are analogous (see accompanying technical report.)
 \vspace{-3px}

### 2018-06-22 icfp18/syntax-icfp18.tex
finished edits through 2.2.0

@@
 \begin{figure}[t]
+\vspace{-8px}
 \begin{lstlisting}[numbers=none]
@@
 \label{fig:nesting-example}
+\vspace{-4px}
 \end{figure}
@@
 
-Segment types can be communicated directly to the programmer upon request by an editor service For Reason, we are equipping the {Merlin} tool\todo{cite ICFP merlin paper}{} \cite{Merlin}, which is used by various Reason editor modes (e.g.
+Segment types can be communicated directly to the programmer upon request by an editor service For Reason, we plan to equip the {Merlin} tool \cite{Merlin}, which is used by various Reason editor extensions (e.g.
 for Emacs and Vim), with a new editor command that reports the expected type of the innermost spliced segment containing the cursor.
-Note that because the type is explicit, this information can be reported even when there is a parse or type error in a spliced expression.
+Note that because the type is explicitly stated, this information can be reported even when there is a parse or type error in a spliced expression.
 % An alternative design that simply reported the type assigned to each spliced expression based on the final expansion (which already exists in Merlin) would be less useful in situations where there is a type error in a spliced expression, or the programmer is in the process of editing the program and wants to know what type of expression to construct.
@@
 The difference is that the argument signature of a function is the same every time the function is applied, i.e.
-it is associated with the function itself, whereas the inferred segmentation can differ for each choice of literal body.
-This, of course, gives TLMs substantially more notational flexibility, even relative to infix or mixfix function notation (where the number of subexpressions is fixed \cite{wieland2009parsing}).% (even infix or mixfix functions \cite{wieland2009parsing}).
+it is associated with the function itself, whereas the segmentation can differ for each choice of literal body.
+This, of course, is what gives TLMs substantially more notational flexibility, even relative to infix or mixfix function notation (where the number of subexpressions is fixed \cite{wieland2009parsing}).% (even infix or mixfix functions \cite{wieland2009parsing}).
 
@@
 Let us turn now to the perspective of the TLM provider, whose principal task is to define the lexer and parser named in the TLM definition.
-In our implementation of TLMs for Reason, lexers are generated using \li{ocamllex} \cite{ocaml-manual} and parsers using Menhir \cite{pottier2016menhir}, which is a modern derivative of the parser generator Yacc~\cite{johnson1975yacc,TarditiDR:mly} with support for LR(1) grammars \cite{jourdan2012validating}.
+Our implementation of TLMs for Reason, called \li{Relit}, assumes that the TLM provider is using \li{ocamllex} to generate the lexer \cite{ocaml-manual} and Menhir to generate the parser \cite{pottier2016menhir}.
+Menhir is a modern derivative of Yacc~\cite{johnson1975yacc,TarditiDR:mly} with support for LR(1) grammars \cite{jourdan2012validating}.
 Reason itself is implemented using these same tools.
-We will say more about how lexer and parser definitions are resolved by our implementation in Sec.
-\ref{sec:implementation}.
-What is fundamentally important is that there is some clear protocol by which the lexer and parser names and the starting non-terminal name specified by the TLM definition, in this case \li{RegexLexer} and \li{RegexParser.start} on Lines 3-4 of Fig.~\ref{fig:regex-tlm-def}, are resolved to lexer and parser implementations that can be evaluated at expansion-time, i.e.
+\li{Relit} requires that the TLM provider package the lexer and parser into an \li{ocamlfind} package named in the TLM definition---in the case of Fig.~\ref{fig:regex-tlm-def}, \li{regex_parser}---so that \li{Relit} can load them, together with their dependencies, at expansion-time.
+The TLM also specifies the name of the starting non-terminal---here, \li{start} on Line 4 of Fig.~\ref{fig:regex-tlm-def}.
+% What is fundamentally important is that there is some clear protocol by which the lexer and parser names and the starting non-terminal name specified by the TLM definition, in this case \li{RegexLexer} and \li{RegexParser.start} on Lines 3-4 of Fig.~\ref{fig:regex-tlm-def}, are resolved to lexer and parser implementations that can be evaluated at expansion-time, i.e.
 during compilation.
-
-The \li{ocamllex} and Menhir definitions used to implement \li{RegexLexer} and \li{RegexParser} are given in Fig.~\ref{fig:lexer-and-parser}.
-A fragment of the signature of the \li{Relit} library, which provides various necessary definitions and utilities to TLM providers, is shown in Fig.~\ref{fig:relit-util}.
-After describing this example in detail, we will outline some alternative parsing strategies.
+The \li{ocamllex} and Menhir definitions used to implement \li{RegexLexer} and \li{RegexParser} appear in Fig.~\ref{fig:lexer-and-parser}.
+A fragment of the signature of the \li{Relit_helper} library, which provides various necessary definitions and utilities to TLM providers, appears in Fig.~\ref{fig:relit-util}.
+After describing this example in detail, we outline some alternative parsing strategies that do not directly require the use of \li{ocamllex} and Menhir in Sec.~\ref{sec:html-example}.
+In particular, we describe the \li{$html} TLM from Fig.~\ref{fig:nesting-example}.
 
@@
 \subsubsection{Other Implementation Strategies}
+\label{sec:html-example}
 Although \li{ocamllex} and Menhir are powerful tools, they are not right for every parsing job.
@@
       let add : 
-        key -> 'a -> t('a) -> t('a);
+        key => 'a => t('a) => t('a);
     /* ... */ }
@@
 In the pattern typing judgement, $\pctx$ collects the typing hypotheses generated by $p$.
-These judgements are inductively defined in the accompanying technical reportal material along with necessary auxiliary structures and  standard lemmas.
+These judgements are inductively defined in the accompanying technical report along with necessary auxiliary structures and standard lemmas.
 The dynamic semantics of $\miniVerseBase$ is organized around the judgements $\isvalU{e}$, which says that $e$ is a value, and $\evalU{e}{e'}$, which says that $e$ evaluates to the value $e'$.
@@
 This occasionally causes types that incidentally appear in error messages to be more general than the programmer might expect (e.g.
-here \li{restriction_template} will have type \li{'a -> 'b} during this ``pre-typing'' phase).
+here \li{restriction_template} will have type \li{'a => 'b} during this ``pre-typing'' phase).
 %If this type ends up in an error message incidentally, e.g. due to a signature mismatch elsewhere, it may be somewhat confusing. 

### 2018-06-22 icfp18/syntax-icfp18.tex
tightened up 2.2.0

@@
 For example, Fig.~\ref{fig:nesting-example} shows a program fragment where we transform an encoding of a chemical structure expressed using the standard SMILES notation for chemical structures \cite{anderson1987smiles}, extended with splicing notation, into a vector graphic, then embed this directly into a fragment of a webpage.
-We detail the TLM \li{$html at Html.t}, which implements HTML notation similar to that found in Ur/Web, in the accompanying technical report, and assume a TLM, \li{$smiles at Smiles.t} and a function, \li{Smiles.to_svg : Smiles.t => Html.t}, not shown.
+We will detail the TLM \li{$html at Html.t}, which implements HTML notation similar to that found in Ur/Web, in Sec.~\ref{sec:html-example}, and we assume a TLM \li{$smiles at Smiles.t}, and a function, \li{Smiles.to_svg : Smiles.t => Html.t}, not shown.
 In Reason, \li{|>} is reverse function application.
@@
 \label{sec:provider-perspective}
+\vspace{-1px}
 \begin{figure}
 \begin{subfigure}[t]{0.30\textwidth}
+\vspace{-8px}
 \begin{lstlisting}[morekeywords={rule,parse,eof}]
@@
 \end{lstlisting}
-\vspace{-4px}
+\vspace{-5px}
 \caption{\li{RegexLexer.mll}}
@@
 \begin{subfigure}[t]{0.61\textwidth}
+\vspace{-8px}
 \begin{lstlisting}[morekeywords={token,left,start},stringstyle=\ttfamily\color{black}]
@@
 \end{lstlisting}
-\vspace{-4px}
+\vspace{-5px}
 \caption{\li{RegexParser.mly}}
@@
 \end{subfigure}
-\vspace{2px}
+\vspace{1px}
 \caption{The lexer and parser for the \li{$regex} TLM from Fig.~\ref{fig:regex-tlm-def}. See Fig.~\ref{fig:relit-util} for relevant definitions from \li{Relit}.}
 \label{fig:lexer-and-parser}
-\vspace{-5px}
+\vspace{-7px}
 \end{figure}
 
-Let us turn now to the perspective of the TLM provider, whose principal task is to define the lexer and parser named in the TLM definition.
-Our implementation of TLMs for Reason, called \li{Relit}, assumes that the TLM provider is using \li{ocamllex} to generate the lexer \cite{ocaml-manual} and Menhir to generate the parser \cite{pottier2016menhir}.
+Let us turn now to the perspective of the TLM provider, whose principal task is to define the lexer and parser named in the TLM definition---\li{RegexLexer} and \li{RegexParser} in Fig.~\ref{fig:regex-tlm-def}.
+Our implementation of TLMs for Reason, called \li{Relit}, assumes that the provider is using \li{ocamllex} \cite{ocaml-manual} and Menhir  \cite{pottier2016menhir} to generate the lexer and parser, respectively.
 Menhir is a modern derivative of Yacc~\cite{johnson1975yacc,TarditiDR:mly} with support for LR(1) grammars \cite{jourdan2012validating}.
-Reason itself is implemented using these same tools.
+The Reason implementation also uses these tools.
 \li{Relit} requires that the TLM provider package the lexer and parser into an \li{ocamlfind} package named in the TLM definition---in the case of Fig.~\ref{fig:regex-tlm-def}, \li{regex_parser}---so that \li{Relit} can load them, together with their dependencies, at expansion-time.
-The TLM also specifies the name of the starting non-terminal---here, \li{start} on Line 4 of Fig.~\ref{fig:regex-tlm-def}.
+The TLM must also name the starting non-terminal---here, \li{start} on Line~4 of Fig.~\ref{fig:regex-tlm-def}.
 % What is fundamentally important is that there is some clear protocol by which the lexer and parser names and the starting non-terminal name specified by the TLM definition, in this case \li{RegexLexer} and \li{RegexParser.start} on Lines 3-4 of Fig.~\ref{fig:regex-tlm-def}, are resolved to lexer and parser implementations that can be evaluated at expansion-time, i.e.
 during compilation.
-The \li{ocamllex} and Menhir definitions used to implement \li{RegexLexer} and \li{RegexParser} appear in Fig.~\ref{fig:lexer-and-parser}.
-A fragment of the signature of the \li{Relit_helper} library, which provides various necessary definitions and utilities to TLM providers, appears in Fig.~\ref{fig:relit-util}.
-After describing this example in detail, we outline some alternative parsing strategies that do not directly require the use of \li{ocamllex} and Menhir in Sec.~\ref{sec:html-example}.
-In particular, we describe the \li{$html} TLM from Fig.~\ref{fig:nesting-example}.
+The \li{ocamllex} and Menhir definitions that implement \li{RegexLexer} and \li{RegexParser} appear in Fig.~\ref{fig:lexer-and-parser}. 
+After detailing this example in Sec.~\ref{sec:regex-lexer}-\ref{sec:regexparser}, we summarize the relevant parser correctness criteria in Sec.~\ref{sec:correctness-criteria}, describe how providers might use MetaOCaml to reason more precisely about type correctness in Sec.~\ref{sec:metaocaml}, and finally outline an alternative implementation strategy in Sec.~\ref{sec:html-example} that sidesteps \li{ocamllex} and Menhir to support the use of arbitrary parse functions.
+In particular, we implement the \li{$html} TLM from Fig.~\ref{fig:nesting-example} using an off-the-shelf HTML parser.
 
@@
 \label{sec:regex-lexer}
+% A fragment of the signature of the \li{Relit_helper} library, which provides various necessary definitions and utilities to TLM providers, appears in Fig.~\ref{fig:relit-util}. 
 In most respects, the definition of the lexer in Fig.~\ref{fig:regex-lexer} is conventional.
@@
 \subsubsection{Correctness Criteria}
+\label{sec:correctness-criteria}
 In order to maintain the abstract reasoning principles discussed in Sec. \ref{sec:client-perspective}, the system \emph{validates} each proto-expansion generated by the parser as follows:

### 2018-06-22 icfp18/syntax-icfp18.tex
tighten up 2.2.1-2

@@
     type t = {start_pos: int, end_pos: int};
-    let read_to : string => Lexing.lexbuf => Segment.t;
-  };
+    let read_to : string => Lexing.lexbuf => Segment.t; };
   exception ExpansionError({msg: string, loc: option(Segment.t)});
@@
     type t = Parsetree.expression; /* from the OCaml compiler library */
-    let spliced : Segment.t => Parsetree.core_type => t;
-  };
+    let spliced : Segment.t => Parsetree.core_type => t; };
   notation $re_expr at Parsetree.expression { /* ... (see text) ... */ };
@@
 \end{lstlisting}
+\vspace{-2px}
 \caption{A fragment of the \li{Relit} module signature relevant to expression TLM providers.}
 \label{fig:relit-util}
-\vspace{-4px}
+\vspace{-5px}
 \end{figure}
@@
 In most respects, the definition of the lexer in Fig.~\ref{fig:regex-lexer} is conventional.
-It reads various token patterns on Lines 6-23, emitting the corresponding tokens specified by the \lstinline[morekeywords={token}]{%token} declarations in Fig.~\ref{fig:regex-parser}, which have been brought into scope by Line 2 of Fig.~\ref{fig:regex-lexer}.
-Lines 20-21 combine sequences of non-special characters (e.g.
+It reads various token patterns on Lines 8-32, emitting the corresponding tokens specified by the \lstinline[morekeywords={token}]{%token} declarations in Fig.~\ref{fig:regex-parser}, which have been brought into scope by Line 2 of Fig.~\ref{fig:regex-lexer}.
+Lines 22-23 combine sequences of non-special characters (e.g.
 alphanumeric characters) into a single string token.
-Lines 22-23 implement backslash-prefixed escape sequences for characters that have special meaning in this notation for regexes, emitting a string token in each instance.\footnote{A better approach would be to define a second lexing rule that greedily combines sequences consisting of non-special characters and escape characters into a single \li{STR} token, but for the sake of exposition, we stick to this simpler approach.}
+Lines 24-25 implement backslash-prefixed escape sequences for characters that have special meaning in this notation for regexes, emitting a string token in each instance.\footnote{A better approach would be to define a second lexing rule that greedily combines sequences consisting of non-special characters and escape characters into a single \li{STR} token, but for the sake of exposition, we stick to this simpler approach.}
 
@@
 This implies that the TLM does not need to parse spliced expressions itself, as long as it can independently decide where the spliced expressions start and end.
-A paired start and end position is a segment,  represented by a value of the type \li{Segment.t} defined on Line 2 of Fig.~\ref{fig:relit-util}.
-Lines 26-27 and Lines 30-31 of the lexer produce a segment by calling a helper function, \li{Relit.Segment.read_to}, that internally invokes the Reason lexer on the provided lexing buffer until it sees an instance of the provided token outside of a Reason comment, string literal or generalized literal.
+A paired start and end position is a segment,  represented by a value of the type \li{Segment.t} defined in the \li{Relit} helper library as specified on Line 2 of Fig.~\ref{fig:relit-util}.
+Lines 26-28 and Lines 29-31 of the lexer produce a segment by calling a helper function, \li{Relit.Segment.read_to}, that internally invokes the Reason lexer on the provided lexing buffer until it sees an instance of the provided token outside of a Reason comment, string literal or generalized literal.
 If the provided token is a right delimiter also used by Reason, like \li{")"} on Line 4 of Fig.~\ref{fig:regex-lexer}, then \li{read_to} looks for the first unmatched instance.
-For example, if the remaining lexing buffer contains {\small \texttt{f("))"))A|G}} then \li{read_to} will consume up through the final instance of \li{)} and emit the start and end position, excluding the closing delimiter, relative to the original buffer.
+For example, if the remaining lexing buffer contains {\small \texttt{f("))"))A|G}} then \li{read_to} will consume up until just before the final instance of \li{)} and emit the appropriate segment.
+Similar helper functions, not shown, are available for reading out single identifiers, simple paths like \li{X1.X2.xy} and a few other unambiguously delimited fragments of the Reason syntax.
 
@@
 First, it supports splicing even within literal notations that have a very different lexical structure from Reason itself, as in this example.
-Second, it allows the parser to avoiding needing to link to Reason's expression grammar (though this is also supported by Menhir).
-Spliced expressions arrive as single, opaque tokens as shown on Lines 6-7 of Fig.~\ref{fig:regex-parser}, and do not need to be parsed by the TLM at all.
+Second, it allows the parser to avoiding needing to link to Reason's expression grammar (though this is supported by Menhir), which substantially simplifies reasoning about ambiguities.
+Spliced expressions arrive as single, opaque tokens as specified on Lines 6-7 of Fig.~\ref{fig:regex-parser}.
 The system will recursively process spliced expressions after the parser has finished generating the proto-expansion.
 
-Similar helper functions, not shown, are available for reading out single identifiers, simple paths like \li{X1.X2.xy} and a few other useful fragments of the Reason syntax, so that TLM providers can implement lower-cost splicing notation that clients can choose in situations where the spliced expression is of a restricted form that makes the end position of the spliced expression unambiguous without matching delimitation.
-As a matter of practice, we recommend that TLM providers start with a fully delimited form before adding optional alternative shorthand in order to avoid artificially restricting the client programmer's ability to directly convey the compositional structure of a value.
-Although \li{Segment.t} is not held abstract in order to support certain advanced scenarios, discussed below, we recommend that TLM providers generally treat it as such.
-
-\subsubsection{\li{RegexParser} and Quasiquotation}\label{sec:regexparser}The job of a TLM parser is to generate a proto-expansion given the tokens generated by the lexer, or if this is not possible, to indicate an expansion error by raising either a Menhir parse error or \li{Relit.ExpansionError} (Line 5 of Fig.~\ref{fig:relit-util}) with an appropriate error message and, if possible, an error location.
+%, so that TLM providers can implement lower-cost splicing notation that clients can choose in situations where the spliced expression is of a restricted form that makes the end position of the spliced expression unambiguous.
+
+%As a matter of practice, we recommend that TLM providers start with a fully delimited form before adding optional alternative shorthand in order to avoid artificially restricting the client programmer's ability to directly convey the compositional structure of a value.
+% Although \li{Segment.t} is not held abstract in order to support certain advanced scenarios, discussed below, we recommend that TLM providers generally treat it as such.
+
+\subsubsection{\li{RegexParser} and Quasiquotation}\label{sec:regexparser}The job of the parser is to generate a proto-expansion given the tokens generated by the lexer, or if this is not possible, to indicate an expansion error by raising either a Menhir parse error or \li{Relit.ExpansionError} (Line 4 of Fig.~\ref{fig:relit-util}) with an appropriate error message and, if possible, an error location.
 In the case of expression TLMs, the proto-expansion must be a proto-expression (we consider proto-patterns in the next section).
-The Menhir directive on Line 9 of Fig~\ref{fig:regex-parser}, which specifies the starting non-terminal of the grammar, reveals that proto-expressions are values of type \li{Relit.ProtoExpr.t}.
-This type is defined on Line 7 of Fig.~\ref{fig:relit-util} as a synonym for \li{Parsetree.expression}, which is the standard representation of expression parse trees exposed by the OCaml compiler \cite{ocaml-manual}.% by many external tools, including the Reason toolchain.
-
-To be able to repurpose this existing parse tree representation for proto-expansions, there must be some way to unambiguously represent splice references, notated earlier in the paper as \li{spliced<m..n : ty>} where \li{m..n} specifies a segment and \li{ty} is the segment's expected type.
-The \li{ProtoExpr} module solves this problem by providing a function \li{spSURLEURLliced} that takes a segment, of type \li{Segment.t}, and a type representation of type \li{Parsetree.core_type}, which also comes from OCaml's compiler library, and produces a corresponding value of type \li{ProtoExpr.t} that uniquely represents the corresponding splice references.
-Internally, we use OCaml's parse tree annotations.
+Proto-expressions are encoded as values of type \li{Relit.ProtoExpr.t}.
+This type is defined on Line 6 of Fig.~\ref{fig:relit-util} as a synonym for \li{Parsetree.expression}, which is the standard representation of expression parse trees exposed by the OCaml compiler library \cite{ocaml-manual}.% by many external tools, including the Reason toolchain.
+
+In order to repurpose this existing parse tree representation for proto-expansions, we must provide some way to unambiguously represent splice references, notated earlier in the paper as \li{spliced<m..n : ty>} where \li{m..n} specifies a segment and \li{ty} is the segment's expected type.
+The \li{ProtoExpr} module solves this problem by providing a function \li{spSURLEURLliced} that takes a segment, of type \li{Segment.t}, and a type representation of type \li{Parsetree.core_type}, which also comes from OCaml's compiler library, and produces a corresponding value of type \li{ProtoExpr.t} that uniquely represents the corresponding splice reference (see Sec.~\ref{sec:implementation} for more on the internal representation).\todo{say something about splice references in sec 6}{}
 

### 2018-06-22 icfp18/syntax-icfp18.tex
revised 2.2.2

@@
 In most respects, the definition of the lexer in Fig.~\ref{fig:regex-lexer} is conventional.
-It reads various token patterns on Lines 8-32, emitting the corresponding tokens specified by the \lstinline[morekeywords={token}]{%token} declarations in Fig.~\ref{fig:regex-parser}, which have been brought into scope by Line 2 of Fig.~\ref{fig:regex-lexer}.
+It reads various lexical patterns on Lines 8-32, emitting the corresponding tokens specified by the \lstinline[morekeywords={token}]{%token} declarations in Fig.~\ref{fig:regex-parser}, which have been brought into scope by Line 2 of Fig.~\ref{fig:regex-lexer}.
 Lines 22-23 combine sequences of non-special characters (e.g.
@@
 Recall from Sec.~\ref{sec:context-dependence} that the parser must represent spliced terms abstractly, by their location within the literal body.
-This implies that the TLM does not need to parse spliced expressions itself, as long as it can independently decide where the spliced expressions start and end.
+This implies that the TLM does not need to parse the spliced expression itself, as long as it can independently decide where the spliced expression starts and ends.
 A paired start and end position is a segment,  represented by a value of the type \li{Segment.t} defined in the \li{Relit} helper library as specified on Line 2 of Fig.~\ref{fig:relit-util}.
@@
 Second, it allows the parser to avoiding needing to link to Reason's expression grammar (though this is supported by Menhir), which substantially simplifies reasoning about ambiguities.
-Spliced expressions arrive as single, opaque tokens as specified on Lines 6-7 of Fig.~\ref{fig:regex-parser}.
-The system will recursively process spliced expressions after the parser has finished generating the proto-expansion.
+Spliced expressions arrive into the parser as single, opaque tokens as specified on Lines 6-7 of Fig.~\ref{fig:regex-parser}.
+The system will recursively expand spliced expressions after the parser has finished generating the proto-expansion.
 
@@
 In order to repurpose this existing parse tree representation for proto-expansions, we must provide some way to unambiguously represent splice references, notated earlier in the paper as \li{spliced<m..n : ty>} where \li{m..n} specifies a segment and \li{ty} is the segment's expected type.
-The \li{ProtoExpr} module solves this problem by providing a function \li{spSURLEURLliced} that takes a segment, of type \li{Segment.t}, and a type representation of type \li{Parsetree.core_type}, which also comes from OCaml's compiler library, and produces a corresponding value of type \li{ProtoExpr.t} that uniquely represents the corresponding splice reference (see Sec.~\ref{sec:implementation} for more on the internal representation).\todo{say something about splice references in sec 6}{}
+The \li{ProtoExpr} module solves this problem by providing a function \li{spSURLEURLliced} that takes a segment, of type \li{Segment.t}, and a type representation, of type \li{Parsetree.core_type}, which also comes from OCaml's compiler library, and produces a corresponding value of type \li{ProtoExpr.t} that uniquely represents the corresponding splice reference (see Sec.~\ref{sec:implementation} for more on the internal representation).\todo{say something about splice references in sec 6}{}
+
+Note that it is only because we explicitly name dependencies in TLM definitions that we can get away with using a simple, unprivileged representation of parse trees. In particular, 
+% Another benefit of this explicit approach is that 
+variable renamings can be performed without needing to reason about or modify parse tree encodings in parser definitions (where variables are represented using strings, and are thus not part of the binding structure).
+For example, if we decided to rename the module \li{Regex} to \li{Regexp}, this renaming would need only to be performed on the module paths in the definition of \li{$rx}, as long as the name of the internal module variable in the dependencies list remains \li{Regex}.
+%(The TLM provider may wish, for the sake of aesthetics, to propagate the change into the parser definition as well, but this is not semantically necessary). 
+% This allows us to encode expansions using ordinary OCaml values (in fact, using the OCaml compiler's own parse tree representation), rather than requiring the introduction of primitive quasiquotation into OCaml (see Sec.
+\ref{sec:provider-perspective}).
 
@@
 Notice, however, that the generalized literal forms in Fig.~\ref{fig:regex-parser} are not each prefixed by \li{$proto_expr}.
-Instead, we use the \li{open notation} directive on Line 2, which implicitly applies \li{$proto_expr} to all unadorned generalized literal forms following the usual lexical scoping rules for \li{open}.
+Instead, we use the \li{open notation} directive on Line 2, which implicitly applies \li{$proto_expr} to all unadorned generalized literal forms in its scope.
 
@@
 
-This issue does \textbf{not} come up when using TLMs to define quotation literals for languages other than Reason itself (an example is given in Sec.~\ref{sec:sptsms}), so if this design proves too limiting, it may be reasonable to build ``inside-out'' quasiquotation primitively into Reason as a special case.
-In fact, this is already possible using MetaOCaml, discussed below.
+This issue does \textbf{not} come up when using TLMs to define quotation literals for languages other than Reason itself (an example is given in Sec.~\ref{sec:sptsms}), so if this design proves too limiting, it may be reasonable to build ``inside-out'' quasiquotation primitively into Reason.
+In fact, this is already possible using a preprocessor, \li{ppx_metaquot}, from the \li{ppx_tools} library \cite{ppx-tools}, or with MetaOCaml and related systems, discussed below.
 %(Nested TLM application via splicing is, of course, useful and fully supported, e.g.
@@
  % Typed quotations with an explicit representation of contexts would need to be built in to the language  necessary to even think about expanding a
-
- \todo{integrate me}{}Another benefit of this explicit approach is that renamings can be performed without needing to reason about or modify parse tree encodings in parser definitions.
-For example, if we decided to rename the module \li{Regex} to \li{Regexp}, this renaming would need only to be performed on the module paths in the TLM definition, as long as the name of the internal module variable in the dependencies list remains \li{Regex}.
-%(The TLM provider may wish, for the sake of aesthetics, to propagate the change into the parser definition as well, but this is not semantically necessary). 
-This allows us to encode expansions using ordinary OCaml values (in fact, using the OCaml compiler's own parse tree representation), rather than requiring the introduction of primitive quasiquotation into OCaml (see Sec.
-\ref{sec:provider-perspective}).
 

### 2018-06-22 icfp18/syntax-icfp18.tex
edits through 2.2.4

@@
 \vspace{-2px}
-\caption{A fragment of the \li{Relit} module signature relevant to expression TLM providers.}
+\caption{A fragment of the \li{Relit} module's signature relevant to expression TLM providers.}
 \label{fig:relit-util}
@@
 \end{enumerate}
-If proto-expansion validation fails, the client is notified that the applied TLM is incorrectly implemented.
-The proto-expansion validation criteria are exactly the correctness criteria that parser writers must consider, quantifying over all possible literal bodies for which a proto-expansion is generated, in order to avoid exposing this class of compile-time error to clients.
-
-\subsubsection{MetaOCaml}\label{sec:metaocaml} In ``vanilla'' OCaml, it is not possible to express a datatype that classifies only valid proto-expansions according to the correctness criteria above.
+If validation fails, the client is notified that the applied TLM is incorrectly implemented.% The proto-expansion validation criteria are exactly the correctness criteria that parser writers must consider, quantifying over all possible literal bodies for which a proto-expansion is generated, in order to avoid exposing this class of compile-time error to clients.
+
+\subsubsection{MetaOCaml}\label{sec:metaocaml} In ``vanilla'' OCaml, it is not possible to define a datatype that classifies only valid proto-expansions according to the correctness criteria above.
 The MetaOCaml type system comes closer by extending the semantics of OCaml with values of type \li{'a code}, which are constructed using a primitive typed quasiquotation operation \cite{DBLP:conf/flops/Kiselyov14,taha2004gentle}.
-In MetaOCaml-BER, which is an active implementation of (an extension of) MetaOCaml as a fork of the compiler, and \li{ppx_stage}, which implements a subset of MetaOCaml as a compiler plugin \cite{ppxstage}, a value of type \li{'a code} is represented as, and can be coerced safely to, an OCaml parse tree for the corresponding expression of type \li{'a} \cite{DBLP:conf/flops/Kiselyov14}.
+In MetaOCaml-BER, which is an active implementation of (an extension of) MetaOCaml as a fork of the compiler, and \li{ppx_stage}, which implements a subset of MetaOCaml as an ordinary compiler plugin rather than a fork of the compiler \cite{ppxstage}, a value of type \li{'a code} is represented as, and can be coerced safely to, an OCaml parse tree for the corresponding expression of type \li{'a} \cite{DBLP:conf/flops/Kiselyov14}.
 It is therefore possible to implement a TLM parser using \li{'a code} values internally, doing the coercion only at the end.
@@
 by defining for each possible segment type \li{t} a dummy value of type \li{t code} and then post-processing the parse tree after the coercion step.
-Overall, however, this approach can substantially increase a TLM provider's confidence in the type correctness of the expansion logic, with the caveats just mentioned.
+
+Overall, this approach can substantially increase a TLM provider's confidence in the type correctness of the expansion logic, with the caveats just mentioned.
 With \li{ppx_stage}, it is possible to compile the parser separately without requiring that clients use the plugin.
@@
 one with a rich static type structure and support for pattern matching.
-Many of the basic mechanisms would be relevant even if support for parsing literal bodies was removed and replaced with term rewriting.
+Many of the mechanisms would be relevant even if support for parsing literal bodies was removed and replaced with term rewriting.
 Segmentations can be considered a refinement of the tree paths in prior work \cite{Herman10:Theory,gorn1965explicit}.
@@
 
-At several points in the paper, we allude to editor integration.
+At several points, we allude to editor integration.
 However, several important questions having to do with TLM-specific syntax highlighting, incremental parsing and error recovery \cite{graham1979practical} remain to be considered.

### 2018-06-22 icfp18/supplement/icfp18-supplement.tex
added HTML case study

@@
 \chapter{Additional Examples}
-
-\begin{figure*}[t]
-\begin{subfigure}[t]{0.50\textwidth}
-\begin{lstlisting}
-  module Html = {
-    /* a simplified encoding of HTML */
-    type tag = string;
-    type attr = (string, string);
-    type attrs = list(attr);
-    type t = 
-      | Text(string) 
-      | Elem(tag, attrs, list(t));
-  }
-\end{lstlisting}
-\vspace{-4px}
-\caption{The \li{Html} module, which defines \li{Html.t}.}
-\label{fig:html-type-def}
-\end{subfigure}
-\hfill
-\begin{subfigure}[t]{0.49\textwidth}
-\begin{lstlisting}[morekeywords={expression}]
-  module HtmlNotation {
-    notation $html at Html.t {
-      lexer Relit.TrivLexer;
-      expression parser HtmlParser.expr;
-      pattern parser HtmlParser.pat;
-      expansions require 
-        { module Html = Html; }
-    }
-  }
-\end{lstlisting}
-\vspace{-4px}
-\caption{The \li{$html} expression and pattern TLMs.}
-\label{fig:html-tlm-def}
-\end{subfigure}
-
-\begin{subfigure}[t]{\textwidth}
-\vspace{5px}
-% let pElem : Html.tag -> Html.attrs -> list(ProtoExpr.t) -> ProtoExpr.t = (* ... *)
-% let pTextElem : Html.tag -> Html.attrs, string) -> ProtoExpr.t = (* ... *)
-\begin{lstlisting}
-  module HtmlParser = Relit.HandRolledExprPatParser({
-    module ProtoHtml = {
-      type t = PText(string)
-             | PElem(Html.tag, Html.attrs, list(t))
-             | PSplicedChildren(Html.tag, Html.attrs, Relit.Segment.t)
-             | PSplicedElem(Relit.Segment.t) 
-             | PSplicedText(Relit.Segment.t) 
-      let parse : string => t = 
-        /* ... via Markup.ml's stream combinators (see text) ... */;
-    };
-    module ExprExpander = {
-      open notation Relit.$proto_expr; open ProtoHtml;
-      let expand_attr = (x1, x2) => `(SURL ($s `(EURLx1SURL)`, $s `(EURLx2SURL)`)EURL )`;
-      let expand_attrs = (attrs) => `(SURL $list `(EURLList.map(expand_attr, attrs)SURL)`EURL )`;
-      let expand_proto = fun 
-      | PText(text) => `(SURL Html.Text($s `(EURLtextSURL)`)EURL )` 
-      | PElem(tag, attrs, children) => 
-        `(SURL Html.Elem($s `(EURLtagSURL)`, `(EURLexpand_attrs(attrs)SURL)`, 
-            $list `(EURLList.map(expand_parsed, children)SURL)`)EURL )`
-      | PSplicedChildren(tag, attrs, seg) => 
-        `(SURL Html.Elem($s `(EURLtagSURL)`, `(EURLexpand_attrs(attrs)SURL)`, 
-            $spliced `(EURLsegSURL : list(Html.t))`)EURL )`
-      | PSplicedHtml(seg) => `(SURL $spliced `(EURLsegSURL : Html.t) EURL)`);
-      | PSplicedText(seg) => `(SURL Html.Text($spliced `(EURLsegSURL : striSURLEURLng)`) EURL)`
-    };
-    let expr = (body) => body |> ProtoHtml.parse |> ExprExpander.expand_proto;
-    let pat  = (body) => expr(body) |> ProtoPat.from_expr;
-  })
-\end{lstlisting}
-\vspace{-6px}
-\caption{The implementation of \li{HtmlParser}, which defers the parsing step to the \li{Markup.ml} library \cite{markupml}.}
-\label{fig:html-parser}
-\end{subfigure}
-\begin{subfigure}[t]{\textwidth}
-\lstset{deletestring=[d]{"}}
-\begin{lstlisting}
-  open HtmlNotation; 
-  exception UnexpectedFormat;
-  let scrape_data = $html.(fun 
-  | `(SURL<div... class="result">EURL
-      [ `(SCOLOR<h1><$$>ECOLORtitleSCOLOR</$$></h1>ECOLOR)`, author, summary, ..._ ]SURL</div>EURL)` => 
-    (title, author, summary)
-  | _ => raise UnexpectedFormat);
-\end{lstlisting}
-\vspace{-4px}
-\caption{Using a pattern TLM to deconstruct an HTML value.}
-\label{fig:pattern-tlm-example}
-\end{subfigure}
-\vspace{2px}
-\caption{Case Study: HTML literals as a library}
-\label{fig:html-case-study}
-\end{figure*}
 

### 2018-06-22 icfp18/syntax-icfp18.tex
added HTML case study

@@
     };
-  }
+  };
 \end{lstlisting}
@@
 In other words, TLM expansion is fundamentally ``outside-in''.
-Notice that the segmentation does communicate the fact that \li{r2} has not been antiquoted (it remains lavender).
-However, to avoid confusion, we decided not to support the generation of TLM application forms via \li{$re_expr}, and instead repurposed that notation for antiquotation.
+Notice that the segmentation does communicate the fact that \li{r2} has not been antiquoted (it remains lavender), but to avoid confusion, we decided not to support the generation of TLM applications via \li{$re_expr}.
 
 This issue does \textbf{not} come up when using TLMs to define quotation literals for languages other than Reason itself (an example is given in Sec.~\ref{sec:sptsms}), so if this design proves too limiting, it may be reasonable to build ``inside-out'' quasiquotation primitively into Reason.
-In fact, this is already possible using a preprocessor, \li{ppx_metaquot}, from the \li{ppx_tools} library \cite{ppx-tools}, or with MetaOCaml and related systems, discussed below.
+In fact, this is already available using a preprocessor, \li{ppx_metaquot}, from the \li{ppx_tools} library \cite{ppx-tools}, or with MetaOCaml and related systems, discussed below.
 %(Nested TLM application via splicing is, of course, useful and fully supported, e.g.
@@
     \begin{enumerate}
-      \item The segments must be in-bounds, of positive extent, non-overlapping and separated by at least one character.
+      \item The segments must be in-bounds, of positive extent, non-overlapping and separated.
       \item The segment types must encode valid types assuming only the TLM dependencies.
@@
 
-The segmentation criteria cannot be statically enforced and, in both mentioned implementations of MetaOCaml, it is impossible to express splice references directly because they do not yet support terms that contain explicit type annotations.
-There are various workarounds, e.g.
-by defining for each possible segment type \li{t} a dummy value of type \li{t code} and then post-processing the parse tree after the coercion step.
-
-Overall, this approach can substantially increase a TLM provider's confidence in the type correctness of the expansion logic, with the caveats just mentioned.
-With \li{ppx_stage}, it is possible to compile the parser separately without requiring that clients use the plugin.
+The segmentation criteria cannot be statically enforced and, in both mentioned implementations of MetaOCaml, it is impossible to express splice references directly because they do not yet support terms with type annotations.
+One workaround is to define for each  segment type \li{t} a dummy value of type \li{t code} that is replaced with a splice reference by post-processing after the coercion step.
+
+Overall, this approach can, at least, substantially increase a TLM provider's confidence in the type correctness of the expansion logic. 
+% With \li{ppx_stage}, it is possible to compile the parser separately without requiring that clients use the plugin. 
 TLMs can be applied inside quoted code as long as the applied TLM is accessed through a dependency of the TLM being defined (so that internal dependency references correctly resolve).
@@
 \label{sec:html-example}
+
+\begin{figure}[t]
+\begin{subfigure}[t]{0.46\textwidth}
+\begin{lstlisting}
+  module Html = {
+    /* a simplified encoding of HTML */
+    type tag = string;
+    type attr = (string, string);
+    type attrs = list(attr);
+    type t = 
+      | Text(string) 
+      | Elem(tag, attrs, list(t));
+  };
+\end{lstlisting}
+\vspace{-6px}
+\caption{The \li{Html} module, which defines \li{Html.t}.}
+\label{fig:html-type-def}
+\end{subfigure}
+\hfill
+\begin{subfigure}[t]{0.52\textwidth}
+\begin{lstlisting}[morekeywords={expression}]
+  module HtmlNotation = {
+    notation $html at Html.t {
+      lexer HtmlLexer
+      expression parser HtmlParser.expr
+      pattern parser HtmlParser.pat 
+      in package html_parser;
+      dependencies = {module Html = Html}
+    };
+  };
+\end{lstlisting}
+\vspace{-6px}
+\caption{The \li{$html} expression and pattern TLMs.}
+\label{fig:html-tlm-def}
+\end{subfigure}
+
+\begin{subfigure}[t]{\textwidth}
+\vspace{3px}
+% let pElem : Html.tag -> Html.attrs -> list(ProtoExpr.t) -> ProtoExpr.t = (* ... *)
+% let pTextElem : Html.tag -> Html.attrs, string) -> ProtoExpr.t = (* ... *)
+\begin{lstlisting}
+  module HtmlLexer = Relit.TrivLexer;
+  module HtmlParser = Relit.HandRolledExprPatParser({
+    module ProtoHtml = {
+      type t = PText(string)
+             | PElem(Html.tag, Html.attrs, list(t))
+             | PSplicedElem(Relit.Segment.t) 
+             | PSplicedText(Relit.Segment.t) 
+      let parse : string => t = 
+        /* ... via Markup.ml's stream combinators (see text) ... */;
+    };
+    module ExprExpander = {
+      open notation Relit.$proto_expr; open ProtoHtml;
+      let expand_attr = (x1, x2) => `(SURL ($s `(EURLx1SURL)`, $s `(EURLx2SURL)`)EURL )`;
+      let expand_attrs = (attrs) => `(SURL $list `(EURLList.map(expand_attr, attrs)SURL)`EURL )`;
+      let expand_proto = fun 
+      | PText(text) => `(SURL Html.Text($s `(EURLtextSURL)`)EURL )` 
+      | PElem(tag, attrs, children) => 
+        `(SURL Html.Elem($s `(EURLtagSURL)`, `(EURLexpand_attrs(attrs)SURL)`, 
+            $list `(EURLList.map(expand_parsed, children)SURL)`)EURL )`
+      | PSplicedHtml(seg) => `(SURL $spliced `(EURLsegSURL : Html.t) EURL)`);
+      | PSplicedText(seg) => `(SURL Html.Text($spliced `(EURLsegSURL : striSURLEURLng)`) EURL)`
+    };
+    let expr = (body) => body |> ProtoHtml.parse |> ExprExpander.expand_proto;
+    let pat  = (body) => expr(body) |> ProtoPat.from_expr;
+  })
+\end{lstlisting}
+\vspace{-6px}
+\caption{The implementation of \li{HtmlParser}, which defers the parsing step to the \li{Markup.ml} library \cite{markupml}.}
+\label{fig:html-parser}
+\vspace{-5px}
+\end{subfigure}
+% \begin{subfigure}[t]{\textwidth}
+% \lstset{deletestring=[d]{"}}
+% \begin{lstlisting}
+%   open HtmlNotation; 
+%   exception UnexpectedFormat;
+%   let scrape_data = $html.(fun 
+%   | `(SURL<div... class="result">EURL
+%       [ `(SCOLOR<h1><$$>ECOLORtitleSCOLOR</$$></h1>ECOLOR)`, author, summary, ..._ ]SURL</div>EURL)` => 
+%     (title, author, summary)
+%   | _ => raise UnexpectedFormat);
+% \end{lstlisting}
+% \vspace{-4px}
+% \caption{Using a pattern TLM to deconstruct an HTML value.}
+% \label{fig:pattern-tlm-example}
+% \end{subfigure}
+\vspace{5px}
+\caption{Case Study: HTML literals as a library}
+\vspace{-4px}
+\label{fig:html-case-study}
+\end{figure}
+
 Although \li{ocamllex} and Menhir are powerful tools, they are not right for every parsing job.
 For example, we might want to use a different parser generator, a parser combinator library \cite{Hutton1992d}, or post-process the result of calling an existing parser to produce a corresponding proto-expansion.
-Fortunately, it is easy to bypass \li{ocamllex} and Menhir:
+Fortunately, it is easy to bypass \li{ocamllex} and Menhir.
 %The most obvious method is to define a trivial lexer that reads the entire literal body into a single token, and a trivial parser with one semantic action.
@@
 It is just as well to implement this interface manually.
-the \li{Relit} library includes a trivial lexer, \li{Relit.TrivLexer}, and a few functors that take care of the necessary boilerplate in situations where we simply want to define the parser as a function of type \li{string => Relit.ProtoExpr.t}.
-For example, the accompanying technical report briefly describes how \li{$re_expr} can be expressed in terms of the Reason parser, a parse tree serializer and a tree transformation.
-It also details how the \li{$html} TLM shown previously is implemented using an existing, ``production grade'' HTML parsing library, \li{Markup.ml} \cite{markupml}, together with a simple post-processing step.
+% the \li{Relit} library includes a trivial lexer, \li{Relit.TrivLexer}, and a few functors that take care of the necessary boilerplate in situations where we simply want to define the parser as a function of type \li{string => Relit.ProtoExpr.t}.
+%For example, the accompanying technical report briefly describes how \li{$re_expr}, described above, can be expressed in terms of the Reason parser, a parse tree serializer and a simple tree transformation.
+Let us consider one substantial example of this approach here: the \li{$html} TLM, shown being applied in Fig.~\ref{fig:nesting-example}, is implemented using an existing, ``production grade'' HTML parsing library, \li{Markup.ml} \cite{markupml}, together with a simple post-processing step as outlined in Fig.~\ref{fig:html-case-study}.
+
+There are many ways to represent HTML data, but one simple representation is specified by \li{Html.t}, defined in Fig.~\ref{fig:html-type-def}.
+However, manually applying the constructors is tedious and, for many web programmers, the induced notation is unfamiliar.
+It also makes it difficult to copy-and-paste from, for example, existing HTML files, as might be useful when refactoring an existing project to use Reason.
+The \li{$html} TLM defined in Fig.~\ref{fig:html-tlm-def} implements the standard HTML notation extended with the tags \li{<$>} and \li{<$$>}, which support \li{Html.t} splicing and \li{string} splicing, respectively.
+%It also supports the suffix \li{...} on a tag, which causes the body of the tag to be a splice of type \li{list(Html.t)}. 
+{We would also in practice have splicing for HTML lists and attribute values, but we omit these here.}
+
+Fig.~\ref{fig:html-parser} shows much of the implementation of \li{HtmlParser},  which is constructed by applying a functor that constructs a module that has the same essential interface as the modules that Menhir generates, with the functions \li{expr : string -> ProtoExpr.t} and \li{pat : string -> ProtoPat.t} of the input module exported as the ``non-terminals''.
+The companion lexer is \li{Relit.TrivLexer}, which simply passes the whole literal body on as a single token.
+The implementation of \li{expr} on Line 23 of Fig.~\ref{fig:html-parser} reveals that the literal body is parsed by first calling \li{ProtoHtml.parse}, then passing the result on to \li{ExprExpander.expand_proto}.
+(We discuss \li{pat} in the next section\todo{do this}{}.)
+
+The \li{ProtoHtml.parse} function, outlined on Lines 8-9, generates a value of the type \li{ProtoHtml.t} defined on Lines 4-7 and so named because it is similar to \li{Html.t} but distinguishes spliced segments, like a proto-expression.
+We omit the definition of \li{ProtoHtml.parse} but conceptually, it starts by creating a character stream from the literal body, then from that creates a \li{Markup.ml} HTML content stream, then transforms that to a stream that emits the signals from the HTML content stream until it sees a splice start tag (which is reported as a syntax error relative to the HTML standard), at which point it calls \li{Segment.read_to}, described earlier, to advance the original character stream until it recognizes the closing tag, emitting the generated segment and then returning control to the HTML content stream.
+Finally, it folds over this stream to produce the result of type \li{ProtoHtml.t}.
+
+The \li{ExprExpander.expand_proto} function defined, with some helper functions, on Lines~12-21 finishes the job of the TLM parser by mapping the parsed proto-HTML to a proto-expression using the \li{$proto_expr} TLM previously described (in this case, using the \li{$list} antiquotation form that produces a parse tree of OCaml's built in list literal form from a list of parse trees).% It would be straightforward to reuse the \li{ProtoHtml} module for different variations on HTML encodings, implementing only the expander anew.
+Overall, the amount of code needed to implement HTML literals.
+
+Reason currently builds in ``JSX'' literals, which also support an HTML-like notation with support for splicing in a somewhat idiosynchratic manner motivated by certain JavaScript libraries (in particular, React).
+The expansion of this notation must be interpretted by a PPX rewriter.
+The problem is that this makes it difficult to use different HTML-related libraries at once.
+For example, ReasonReact and an XML library might both want to use this notation.
+This approach solves this problem (as well as other problems with the PPX-based approach as discussed in Sec.~\ref{sec:existing-approaches}).
 
@@
 \label{sec:sptsms}
+
+The previous section introduced expression TLMs, which support value construction.
+This section introduces pattern TLMs, which support value deconstruction via the structural pattern matching facilities common to ML-like languages.
+
+For example, the definitions outlined in Fig.
+\ref{fig:lam} allow us to elegantly express an evaluator for the lambda calculus as shown in Fig.
+\ref{fig:lam-eval}.
+We assume that the terms of \li{Lambda} are represented differently from the values to avoid needing additional ``impossible'' cases, but we can use the same notation for both without conflict by defining separate TLMs, \li{$term} and \li{$v}, respectively.
+Each of these TLMs defines both expression and pattern literal notation, distinguished by qualifiers in the TLM definition as shown on Lines 5-6 of Fig.
+\ref{fig:lam}, and we see examples of all four of these notations in use in Fig.
+\ref{fig:lam-eval}.
+These TLMs were designed to be used for operating on \li{Lambda} terms, so we chose a splicing convention inspired by the typical convention ``on paper'', where identifiers appearing in the literal are parsed as spliced (OCaml) variables, as indicated by the colors in Fig.
+\ref{fig:lam-eval}.
+For example, on Line 5, \li{x} and \li{e} are bound by the literal pattern for \li{Lambda} terms on the left, then spliced into the expression literal on the right to construct the corresponding \li{Lambda} value.
+% A different pair of TLMs at the same types could make the opposite choice, parsing identifiers as \li{Lambda} language variables and defining a separate notation for splicing.
+%These would be less elegant for writing the evaluator but more useful when, for example, writing its tests.
+This sort of flexibility is difficult to achieve in settings where the grammar of the host language is itself extended (e.g.
+with the more limited mixfix systems of Coq \cite{Coq:manual} or Agda \cite{norell2007towards}) because syntactic conflicts ensue.
+
+From a client programmer's perspective, reasoning principles analogous to those described in Sec.
+\ref{sec:client-perspective} for expression TLMs are available.
+\textbf{Responsibility} is assigned by the same protocol, and the type annotation on the responsible TLM governs the type of the generated pattern, satisfying the \textbf{Expansion Typing} condition.
+Patterns can contain module paths and type annotations, so pattern TLMs are governed by the same \textbf{Context Independence} condition as expression TLMs.
+The same protocol around \textbf{Segmentation} is also enforced.
 
@@
 
-The previous section introduced expression TLMs, which support value construction.
-This section introduces pattern TLMs, which support value deconstruction via the structural pattern matching facilities common to ML-like languages.
-
-For example, the definitions outlined in Fig.
-\ref{fig:lam} allow us to elegantly express an evaluator for the lambda calculus as shown in Fig.
-\ref{fig:lam-eval}.
-We assume that the terms of \li{Lambda} are represented differently from the values to avoid needing additional ``impossible'' cases, but we can use the same notation for both without conflict by defining separate TLMs, \li{$term} and \li{$v}, respectively.
-Each of these TLMs defines both expression and pattern literal notation, distinguished by qualifiers in the TLM definition as shown on Lines 5-6 of Fig.
-\ref{fig:lam}, and we see examples of all four of these notations in use in Fig.
-\ref{fig:lam-eval}.
-These TLMs were designed to be used for operating on \li{Lambda} terms, so we chose a splicing convention inspired by the typical convention ``on paper'', where identifiers appearing in the literal are parsed as spliced (OCaml) variables, as indicated by the colors in Fig.
-\ref{fig:lam-eval}.
-For example, on Line 5, \li{x} and \li{e} are bound by the literal pattern for \li{Lambda} terms on the left, then spliced into the expression literal on the right to construct the corresponding \li{Lambda} value.
-% A different pair of TLMs at the same types could make the opposite choice, parsing identifiers as \li{Lambda} language variables and defining a separate notation for splicing.
-%These would be less elegant for writing the evaluator but more useful when, for example, writing its tests.
-This sort of flexibility is difficult to achieve in settings where the grammar of the host language is itself extended (e.g.
-with the more limited mixfix systems of Coq \cite{Coq:manual} or Agda \cite{norell2007towards}) because syntactic conflicts ensue.
-
-From a client programmer's perspective, reasoning principles analogous to those described in Sec.
-\ref{sec:client-perspective} for expression TLMs are available.
-\textbf{Responsibility} is assigned by the same protocol, and the type annotation on the responsible TLM governs the type of the generated pattern, satisfying the \textbf{Expansion Typing} condition.
-Patterns can contain module paths and type annotations, so pattern TLMs are governed by the same \textbf{Context Independence} condition as expression TLMs.
-The same protocol around \textbf{Segmentation} is also enforced.
-
 The main novelty has to do with \textbf{Segment Typing} and \textbf{Capture}.
@@
 % \subsection{Parametric TLMs By Example}\label{sec:ptsms-by-example}
+
+Fig.
+\ref{fig:map-sig} shows a portion of the signature \li{Map.S} from the OCaml standard library, which specifies a polymorphic abstract data type \cite{liskov1974programming,harper1997programming} \li{t('a)} of finite maps with keys of type \li{key}.
+There are many ways to implement this signature.
+For example, the \li{Map.Make} functor in the standard library implements this signature using balanced binary trees given a module that specifies a \li{key} type equipped with comparison functions.
+The \li{Misc.StringMap} module in the standard library also provides a specialized implementation of this signature with type \li{key = string}.
+
+We can define a TLM, \li{$map}, that is parametric over implementations of this signature, \li{M : Map.S}, and over choices of the co-domain type, \li{a}, as shown in Fig.
+\ref{fig:map-tlm}.
+We can then partially apply \li{$map} to \li{Misc.StringMap} as shown on Line 1 of Fig.~\ref{fig:map-ap}, producing a TLM, \li{$stringmap},  parameterized only over the type \li{a}, with expansion type \li{Misc.StringMap.t(a)}.
+We can apply \li{$stringmap} to a type specialize it further, e.g.
+\li{$stringmap string} has expansion type \li{Misc.StringMap.t(string)}.
+For TLMs where all remaining parameters are types mentioned in the expansion type, we can also immediately apply the TLM to a generalized literal form as shown on Lines~2-3 of Fig.~\ref{fig:map-ap}.
+The type parameters are determined by unification using the types inferred for the spliced expressions together with the segment types specified in the segmentation, and any surrounding type constraints.
+In this example, we can infer \li{a} to be \li{string} because \li{string_of_int(id) : string} and the corresponding segment type is \li{a}.
+TLM abbreviations can themselves be parameterized to support partial application of parameters other than the last.
+
+% For example, given some module that implements this signature, \li{HashDict : DICT}, we can apply \li{$dict} as follows:
+% \begin{lstlisting}[numbers=none]
+%     $dict HashDict int [|"key1"SURL=>EURL10SURL; EURL"key2"SURL=>EURL15|]
+% \end{lstlisting}
+% Notice that the segmentation immediately reveals which punctuation is particular to this TLM and where the spliced key and value expressions appear.
+Because the context-free syntax of unexpanded terms is never modified, it is possible to reason modularly about syntactic determinism, i.e.
+we can reason above that \li{=>} does not appear in the follow set of unexpanded expressions \cite{conf/pldi/SchwerdfegerW09}, so there can never be an ambiguity about where a key expression ends.
+
+The proto-expansion generated by a parametric TLM can refer to its parameters.
+Validation checks that the proto-expansion is truly parametric---in this case, the proto-expansion must be valid for all modules \li{M : Map.S} and types \li{a}.
+It is only after validation that we substitute the actual parameters, here \li{Misc.StringMap} for \li{M} and \li{string} for \li{a}, into the final expansion.
+
+%Only the type annotations on references to spliced terms are subject to early parameter substitution (because they classify application site terms.)
+
+% If we will use the \li{HashDict} implementation ubiquitously, we can abbreviate the partial application of \li{$dict} to \li{HashDict}, resulting in a TLM that is parametric over only the type \li{'a}:
+% \begin{lstlisting}[numbers=none]
+%     syntax $dict' = $dict HashDict; 
+% \end{lstlisting}
 
@@
 \end{figure}
-
-Fig.
-\ref{fig:map-sig} shows a portion of the signature \li{Map.S} from the OCaml standard library, which specifies a polymorphic abstract data type \cite{liskov1974programming,harper1997programming} \li{t('a)} of finite maps with keys of type \li{key}.
-There are many ways to implement this signature.
-For example, the \li{Map.Make} functor in the standard library implements this signature using balanced binary trees given a module that specifies a \li{key} type equipped with comparison functions.
-The \li{Misc.StringMap} module in the standard library also provides a specialized implementation of this signature with type \li{key = string}.
-
-We can define a TLM, \li{$map}, that is parametric over implementations of this signature, \li{M : Map.S}, and over choices of the co-domain type, \li{a}, as shown in Fig.
-\ref{fig:map-tlm}.
-We can then partially apply \li{$map} to \li{Misc.StringMap} as shown on Line 1 of Fig.~\ref{fig:map-ap}, producing a TLM, \li{$stringmap},  parameterized only over the type \li{a}, with expansion type \li{Misc.StringMap.t(a)}.
-We can apply \li{$stringmap} to a type specialize it further, e.g.
-\li{$stringmap string} has expansion type \li{Misc.StringMap.t(string)}.
-For TLMs where all remaining parameters are types mentioned in the expansion type, we can also immediately apply the TLM to a generalized literal form as shown on Lines~2-3 of Fig.~\ref{fig:map-ap}.
-The type parameters are determined by unification using the types inferred for the spliced expressions together with the segment types specified in the segmentation, and any surrounding type constraints.
-In this example, we can infer \li{a} to be \li{string} because \li{string_of_int(id) : string} and the corresponding segment type is \li{a}.
-TLM abbreviations can themselves be parameterized to support partial application of parameters other than the last.
-
-% For example, given some module that implements this signature, \li{HashDict : DICT}, we can apply \li{$dict} as follows:
-% \begin{lstlisting}[numbers=none]
-%     $dict HashDict int [|"key1"SURL=>EURL10SURL; EURL"key2"SURL=>EURL15|]
-% \end{lstlisting}
-% Notice that the segmentation immediately reveals which punctuation is particular to this TLM and where the spliced key and value expressions appear.
-Because the context-free syntax of unexpanded terms is never modified, it is possible to reason modularly about syntactic determinism, i.e.
-we can reason above that \li{=>} does not appear in the follow set of unexpanded expressions \cite{conf/pldi/SchwerdfegerW09}, so there can never be an ambiguity about where a key expression ends.
-
-The proto-expansion generated by a parametric TLM can refer to its parameters.
-Validation checks that the proto-expansion is truly parametric---in this case, the proto-expansion must be valid for all modules \li{M : Map.S} and types \li{a}.
-It is only after validation that we substitute the actual parameters, here \li{Misc.StringMap} for \li{M} and \li{string} for \li{a}, into the final expansion.
-
-%Only the type annotations on references to spliced terms are subject to early parameter substitution (because they classify application site terms.)
-
-% If we will use the \li{HashDict} implementation ubiquitously, we can abbreviate the partial application of \li{$dict} to \li{HashDict}, resulting in a TLM that is parametric over only the type \li{'a}:
-% \begin{lstlisting}[numbers=none]
-%     syntax $dict' = $dict HashDict; 
-% \end{lstlisting}
 

### 2018-06-25 icfp18/syntax-icfp18.tex
second phase submission

@@
 Let us turn now to the perspective of the TLM provider, whose principal task is to define the lexer and parser named in the TLM definition---\li{RegexLexer} and \li{RegexParser} in Fig.~\ref{fig:regex-tlm-def}.
-Our implementation of TLMs for Reason, called \li{Relit}, assumes that the provider is using \li{ocamllex} \cite{ocaml-manual} and Menhir  \cite{pottier2016menhir} to generate the lexer and parser, respectively.
-Menhir is a modern derivative of Yacc~\cite{johnson1975yacc,TarditiDR:mly} with support for LR(1) grammars \cite{jourdan2012validating}.
+Our implementation of TLMs for Reason, called \li{Relit}, assumes that the provider is using \li{ocamllex} \cite{ocaml-manual} and Menhir  \cite{pottier2016menhir,DBLP:journals/entcs/PottierR06} to generate the lexer and parser, respectively.
+Menhir is a modern derivative of Yacc~\cite{johnson1975yacc} with support for LR(1) grammars \cite{jourdan2012validating,DBLP:journals/entcs/PottierR06}.
 The Reason implementation also uses these tools.
-\li{Relit} requires that the TLM provider package the lexer and parser into an \li{ocamlfind} package named in the TLM definition---in the case of Fig.~\ref{fig:regex-tlm-def}, \li{regex_parser}---so that \li{Relit} can load them, together with their dependencies, at expansion-time.
+\li{Relit} requires that the TLM provider package the lexer and parser into an \li{ocamlfind} package named in the TLM definition---in Fig.~\ref{fig:regex-tlm-def}, \li{regex_parser}---so that \li{Relit} can load them, together with their dependencies, at expansion-time.
 The TLM must also name the starting non-terminal---here, \li{start} on Line~4 of Fig.~\ref{fig:regex-tlm-def}.
@@
 The \li{ocamllex} and Menhir definitions that implement \li{RegexLexer} and \li{RegexParser} appear in Fig.~\ref{fig:lexer-and-parser}. 
-After detailing this example in Sec.~\ref{sec:regex-lexer}-\ref{sec:regexparser}, we summarize the relevant parser correctness criteria in Sec.~\ref{sec:correctness-criteria}, describe how providers might use MetaOCaml to reason more precisely about type correctness in Sec.~\ref{sec:metaocaml}, and finally outline an alternative implementation strategy in Sec.~\ref{sec:html-example} that sidesteps \li{ocamllex} and Menhir to support the use of arbitrary parse functions.
+After detailing this example in Sec.~\ref{sec:regex-lexer}-\ref{sec:regexparser}, we summarize the relevant correctness criteria in Sec.~\ref{sec:correctness-criteria}, describe how providers might use MetaOCaml to reason more precisely about type correctness in Sec.~\ref{sec:metaocaml}, and outline an alternative implementation strategy in Sec.~\ref{sec:html-example} that sidesteps \li{ocamllex} and Menhir to support the use of arbitrary parse functions.
 In particular, we implement the \li{$html} TLM from Fig.~\ref{fig:nesting-example} using an off-the-shelf HTML parser.
@@
 In order to repurpose this existing parse tree representation for proto-expansions, we must provide some way to unambiguously represent splice references, notated earlier in the paper as \li{spliced<m..n : ty>} where \li{m..n} specifies a segment and \li{ty} is the segment's expected type.
-The \li{ProtoExpr} module solves this problem by providing a function \li{spSURLEURLliced} that takes a segment, of type \li{Segment.t}, and a type representation, of type \li{Parsetree.core_type}, which also comes from OCaml's compiler library, and produces a corresponding value of type \li{ProtoExpr.t} that uniquely represents the corresponding splice reference (see Sec.~\ref{sec:implementation} for more on the internal representation).\todo{say something about splice references in sec 6}{}
+The \li{ProtoExpr} module solves this problem by providing a function \li{spSURLEURLliced} that takes a segment, of type \li{Segment.t}, and a type representation, of type \li{Parsetree.core_type}, which also comes from OCaml's compiler library, and produces a corresponding value of type \li{ProtoExpr.t} that uniquely represents the corresponding splice reference (see Sec.~\ref{sec:implementation} for more on the internal representation).
 
@@
 The implementation of \li{expr} on Line 23 of Fig.~\ref{fig:html-parser} reveals that the literal body is parsed by first calling \li{ProtoHtml.parse}, then passing the result on to \li{ExprExpander.expand_proto}.
-(We discuss \li{pat} in the next section\todo{do this}{}.)
+(We discuss \li{pat} in the next section.)
 
@@
 
-The \li{ExprExpander.expand_proto} function defined, with some helper functions, on Lines~12-21 finishes the job of the TLM parser by mapping the parsed proto-HTML to a proto-expression using the \li{$proto_expr} TLM previously described (in this case, using the \li{$list} antiquotation form that produces a parse tree of OCaml's built in list literal form from a list of parse trees).% It would be straightforward to reuse the \li{ProtoHtml} module for different variations on HTML encodings, implementing only the expander anew.
+The \li{ExprExpander.expand_proto} function defined, together with helper functions, on Lines~12-21 finishes the job of the TLM parser by mapping the parsed proto-HTML to a proto-expression using the \li{$proto_expr} TLM previously described (in this case, using the \li{$list} antiquotation form that produces a parse tree of OCaml's built in list literal form from a list of parse trees).% It would be straightforward to reuse the \li{ProtoHtml} module for different variations on HTML encodings, implementing only the expander anew.
 Overall, the amount of code needed to implement HTML literals.
 
-Reason currently builds in ``JSX'' literals, which also support an HTML-like notation with support for splicing in a somewhat idiosynchratic manner motivated by certain JavaScript libraries (in particular, React).
+Reason currently builds in ``JSX'' literals, which also support an HTML-like notation with support for splicing.
 The expansion of this notation must be interpretted by a PPX rewriter.
@@
 
-The previous section introduced expression TLMs, which support value construction.
-This section introduces pattern TLMs, which support value deconstruction via the structural pattern matching facilities common to ML-like languages.
-
-For example, the definitions outlined in Fig.
-\ref{fig:lam} allow us to elegantly express an evaluator for the lambda calculus as shown in Fig.
-\ref{fig:lam-eval}.
-We assume that the terms of \li{Lambda} are represented differently from the values to avoid needing additional ``impossible'' cases, but we can use the same notation for both without conflict by defining separate TLMs, \li{$term} and \li{$v}, respectively.
-Each of these TLMs defines both expression and pattern literal notation, distinguished by qualifiers in the TLM definition as shown on Lines 5-6 of Fig.
-\ref{fig:lam}, and we see examples of all four of these notations in use in Fig.
-\ref{fig:lam-eval}.
-These TLMs were designed to be used for operating on \li{Lambda} terms, so we chose a splicing convention inspired by the typical convention ``on paper'', where identifiers appearing in the literal are parsed as spliced (OCaml) variables, as indicated by the colors in Fig.
-\ref{fig:lam-eval}.
-For example, on Line 5, \li{x} and \li{e} are bound by the literal pattern for \li{Lambda} terms on the left, then spliced into the expression literal on the right to construct the corresponding \li{Lambda} value.
-% A different pair of TLMs at the same types could make the opposite choice, parsing identifiers as \li{Lambda} language variables and defining a separate notation for splicing.
-%These would be less elegant for writing the evaluator but more useful when, for example, writing its tests.
-This sort of flexibility is difficult to achieve in settings where the grammar of the host language is itself extended (e.g.
-with the more limited mixfix systems of Coq \cite{Coq:manual} or Agda \cite{norell2007towards}) because syntactic conflicts ensue.
-
-From a client programmer's perspective, reasoning principles analogous to those described in Sec.
-\ref{sec:client-perspective} for expression TLMs are available.
-\textbf{Responsibility} is assigned by the same protocol, and the type annotation on the responsible TLM governs the type of the generated pattern, satisfying the \textbf{Expansion Typing} condition.
-Patterns can contain module paths and type annotations, so pattern TLMs are governed by the same \textbf{Context Independence} condition as expression TLMs.
-The same protocol around \textbf{Segmentation} is also enforced.
-
 \begin{figure}[t]
@@
   module Lambda = {/* a typical encoding */};
-  module LambdaNotation {
+  module LambdaNotation = {
     notation $term at Lambda.term { 
-      lexer SimpleLangLexer;
-      expression parser LambdaParser.term_e;
-      pattern parser LambdaParser.term_p;
-      dependencies 
-        { module Lambda = Lambda } };
+      lexer LambdaLexer
+      expression parser LambdaParser.term_e
+      pattern parser LambdaParser.term_p
+      in package lambda_parser;
+      dependencies = {module Lambda=Lambda} };
     notation $v at Lambda.v {/* analogous */}
@@
 \end{subfigure}
-\vspace{2px}
 \caption{Expression and pattern literal notation for lambda terms and values}
@@
 
-The main novelty has to do with \textbf{Segment Typing} and \textbf{Capture}.
-Variables in patterns do not refer to existing bindings, as in expressions, but rather introduce bindings into other expressions, e.g.
-the corresponding branch of a case analysis or the body of a function, so the critical question is this: given only the segmentation of a pattern literal, can we determine exactly which variables the expansion binds, and what the types of these variables are?
-To answer in the affirmative, we need to preclude the possibility of ``invisible bindings'', so the system ensures that pattern literals bind only those variables that appear inside spliced patterns.
-This implies that \emph{proto-patterns simply cannot contain pattern variables}.
-The segmentation assigns a type to each spliced segment, so the question just asked can be answered given only the segmentation.
-
-From the provider's perspective, the parser specified by the \li{pattern parser} clause must generate a proto-pattern, rather than a proto-expression.
+The previous section introduced expression TLMs, which support value construction.
+This section introduces pattern TLMs, which support value deconstruction via the structural pattern matching facilities common to ML-like languages.
+
+For example, the definitions outlined in Fig.
+\ref{fig:lam} allow us to elegantly express an evaluator for the lambda calculus as shown in Fig.
+\ref{fig:lam-eval}.
+We assume that the terms of \li{Lambda} are of type \li{Lambda.term}, and the values have a different representation, of type \li{Lambda.v}, to avoid needing additional ``impossible'' cases.
+We can provide common notation for both without conflict by defining separate TLMs, \li{$term} and \li{$v}, respectively, in Fig.~\ref{fig:lam}.
+Each of these TLMs defines both expression and pattern literal notation, distinguished by qualifiers in the TLM definition as shown on Lines 5-6 of Fig.
+\ref{fig:lam}, and we see examples of all four of these notations in use in Fig.
+\ref{fig:lam-eval}.
+These TLMs were designed to be used for operating on \li{Lambda} terms, so we chose a splicing convention inspired by the typical convention ``on paper'', where identifiers appearing in the literal are parsed as spliced (OCaml) variables, as indicated by the colors in Fig.
+\ref{fig:lam-eval}.
+% For example, on Line 5, \li{x} and \li{e} are bound by the literal pattern for \li{Lambda} terms on the left, then spliced into the expression literal on the right.
+% A different pair of TLMs at the same types could make the opposite choice, parsing identifiers as \li{Lambda} language variables and defining a separate notation for splicing.
+%These would be less elegant for writing the evaluator but more useful when, for example, writing its tests.
+This sort of flexibility is difficult to achieve in settings where the grammar of the host language is itself extended (e.g.
+with the more limited mixfix systems of Coq \cite{Coq:manual} or Agda \cite{norell2007towards}) because syntactic conflicts ensue.
+
+\subsection{Client Perspective} From a client programmer's perspective, reasoning principles analogous to those described in Sec.~\ref{sec:client-perspective} for expression TLMs are available.
+\textbf{Responsibility} is assigned by the same protocol, and the type annotation on the responsible TLM governs the type of the generated pattern, satisfying the \textbf{Expansion Typing} condition.
+Patterns can contain module paths and type annotations, so pattern TLMs are governed by the same \textbf{Context Independence} condition as expression TLMs.
+The same protocol around \textbf{Segmentation} is also enforced.
+The segmentation assigns a type to each spliced segment, so we can also reason about \textbf{Segment Typing}.
+
+Variables in patterns do not refer to existing bindings, as in expressions, so we do not need to worry about capture avoidance.
+Instead, patterns introduce bindings into other expressions, e.g.
+the corresponding branch of a case analysis, so the critical question is this: given only the segmentation of a pattern literal, how can we determine exactly which variables the expansion binds?
+To answer, we need to preclude the possibility of ``invisible bindings'', so the system ensures that pattern literals bind only those variables that appear inside spliced patterns.
+We call this property \textbf{Visibility}.
+
+Note that in OCaml, boolean guards can be associated with rules of a \li{match} expression, but the guard is not part of the syntax of patterns.
+(If it were, we would need to validate it as in Sec.~\ref{sec:setlms}.) %In order to ensure that a pattern TLM can be applied anywhere a pattern can appear, e.g.
+in a \li{let} expression as on Line 7 of Fig.
+\ref{fig:lam-eval}, a pattern TLM cannot generate a guard.
+% In a language where boolean guards are included in the syntax of patterns, we would need to enforce context independence, capture avoidance and segmentation-related conditions on guard expressions as well.%  In languages where boolean guards can appear at any depth in a pattern, the same
+
+\subsection{Provider Perspective}
+From the provider's perspective, the parser specified by the \li{pattern parser} clause differs only in that it must generate a proto-pattern, rather than a proto-expression.
 In Reason, this is a value of type \li{Relit.ProtoPat.t}.
@@
 \ref{fig:relit-util}, this type is defined as a synonym for \li{Parsetree.patteSURLEURLrn} from OCaml's compiler library equipped with an additional function, \li{Relit.ProtoPat.spSURLEURLliced}, that constructs splice references.
-In principle, it might be better to define \li{Relit.ProtoPat.t} to exclude the representation of pattern variables, and we do just that in the theoretical development in Sec.
+The \textbf{Visibility} property is maintained by enforcing the rule that a \emph{proto-pattern simply cannot contain pattern variables}.
+
+We omit the straightforward details of the parsers in Fig.~\ref{fig:lam-example}.
+In cases where the output of an expression parser uses only constructors and constants, like the HTML example in Fig.~\ref{fig:html-parser}, we can simply call the expression parser and then convert the result to a proto-pattern (Line 24 of Fig.~\ref{fig:html-parser}).
+%In cases where an expression parser uses other constructs, e.g.
+the \li{let} bindings used in the regex parser in Fig.~\ref{fig:regex-parser}, we would need to define it separately.
+
+ % In principle, it might be better to define \li{Relit.ProtoPat.t} to exclude the representation of pattern variables, and we do just that in the theoretical development in Sec.
 \ref{sec:setlms-formally}, but \li{Relit} checks this straightforward condition during proto-expansion validation.
-Note that in OCaml, boolean guards can be associated with rules of a \li{match} (in Reason, \li{switch}) expression, but the guard is not part of the syntax of patterns.
-(If it were, we would need to validate it as in Sec.~\ref{sec:setlms}).
-%In order to ensure that a pattern TLM can be applied anywhere a pattern can appear, e.g.
-in a \li{let} expression as on Line 7 of Fig.
-\ref{fig:lam-eval}, a pattern TLM cannot generate a guard.
-% In a language where boolean guards are included in the syntax of patterns, we would need to enforce context independence, capture avoidance and segmentation-related conditions on guard expressions as well.%  In languages where boolean guards can appear at any depth in a pattern, the same
 
@@
 Fig.
-\ref{fig:map-sig} shows a portion of the signature \li{Map.S} from the OCaml standard library, which specifies a polymorphic abstract data type \cite{liskov1974programming,harper1997programming} \li{t('a)} of finite maps with keys of type \li{key}.
+\ref{fig:map-sig} shows a portion of the signature \li{Map.S} from the OCaml standard library, which specifies a polymorphic abstract data type \li{t('a)} of finite maps with keys of type \li{key}.
 There are many ways to implement this signature.
@@
 We can then partially apply \li{$map} to \li{Misc.StringMap} as shown on Line 1 of Fig.~\ref{fig:map-ap}, producing a TLM, \li{$stringmap},  parameterized only over the type \li{a}, with expansion type \li{Misc.StringMap.t(a)}.
-We can apply \li{$stringmap} to a type specialize it further, e.g.
+We might apply \li{$stringmap} to a type specialize it further, e.g.
 \li{$stringmap string} has expansion type \li{Misc.StringMap.t(string)}.
-For TLMs where all remaining parameters are types mentioned in the expansion type, we can also immediately apply the TLM to a generalized literal form as shown on Lines~2-3 of Fig.~\ref{fig:map-ap}.
+Alternatively, for TLMs where all remaining parameters are types mentioned in the expansion type, we can immediately apply the TLM to a generalized literal form as shown on Lines~2-3 of Fig.~\ref{fig:map-ap}.
 The type parameters are determined by unification using the types inferred for the spliced expressions together with the segment types specified in the segmentation, and any surrounding type constraints.
@@
         key => 'a => t('a) => t('a);
-    /* ... */ }
-  }
+      /* ... */
+    };
+  };
 \end{lstlisting}
@@
   notation $map(M : Map.S, type a) at M.t(a) {
-    lexer  MapLexer; parser MapParser; 
+    lexer MapLexer parser MapParser 
+    in package map_parser;
   }
 \end{lstlisting}
-\vspace{-6px}
+\vspace{-8px}
 \caption{The parametric TLM \li{$map}}
@@
     notation $map at M.t(A.a) { 
-      lexer MapLexer; parser MapParser; 
+      lexer MapLexer parser MapParser in package map_parser;
       dependencies { module M = M; type a = A.a; } } }
@@
 
-This section will present a calculus of simple expression and pattern TLMs based on ``core ML'' called $\miniVersePat$.
-By the end of this section, we will have a theorem that formalizes the six reasoning principles that were outlined informally in the previous sections.
-$\miniVersePat$ consists of an \emph{unexpanded language}, or \emph{UL}, defined by typed expansion to the core language, named for the purposes of this paper $\miniVerseBase$.
+This section presents  $\miniVersePat$, a calculus of simple expression and pattern TLMs based on ``core ML''.
+By the end of this section, we will have theorems that formalize the reasoning principles that were outlined informally in the previous sections.
+$\miniVersePat$ consists of an {unexpanded language} ({UL}) defined by typed expansion to the core language, named for the purposes of this paper $\miniVerseBase$.
 
@@
 \ref{fig:U-unexpanded-terms} defines the syntax of the unexpanded language.
-Unlike core language types and expressions, unexpanded expressiona and types are \textbf{not} abstract binding trees -- we do \textbf{not} assume the standard notions of renaming, alpha-equivalence or substitution.
+Unlike core language types and expressions, unexpanded expressions and types are \textbf{not} abstract binding trees -- we do \textbf{not} assume the standard notions of renaming, alpha-equivalence or substitution.
 Instead, they are simple inductive  structures.
@@
 For example, the rules for variables and lambdas are shown being applied in the example derivation in Fig.~\ref{fig:expansion-exmpl}, discussed below.
-The full set of rules is in the accompanying technical report.
+The full set of rules is in the accompanying technical report \cite{icfp18tr}.
 %Each of these rules is based on the corresponding typing rule, i.e.
@@
 The typing context, $\Gamma$, only tracks the type of these variables.
-We define the identifier update operation $\ctxUpdate{\uG}{\ux}{x}$ as producing the expression identifier expansion context that maps $\ux$ to $x$, written $\vExpands{\ux}{x}$, and defers to $\uG$ for all other expression identifiers, with no requirement that $\ux$ be apart from $\domof{\uG}$.
+We define the identifier update operation $\ctxUpdate{\uG}{\ux}{x}$ as mapping $\ux$ to $x$, written $\vExpands{\ux}{x}$, and deferring to $\uG$ for all other expression identifiers, with no requirement that $\ux$ be apart from $\domof{\uG}$.
 %Note the distinction between update and extension (which requires that the new identifier is not already in the domain.) %We write $\uGammaOK{\uGamma}$ when $\uGamma=\uGG{\uG}{\Gamma}$ and each expression variable in $\uG$ is assigned a type by $\Gamma$.
@@
 \end{mathpar}}
+\vspace{-4px}
 \caption{The typed expansion rules for expression and pattern TLM definitions.}
@@
 \ref{sec:ce-syntax-U}.
-As with $\tBody$, we need only take as a condition that there is an isomorphism between values of type $\tCEExp$ and closed proto-expressions, which is mediated in one direction by the \emph{proto-expression decoding judgement}, $\decodeCondE{e}{\ce}$ that we will return to when we describe TLM application below (see accompanying technical report for the full condition).
+As with $\tBody$, we need only take as a condition that there is an isomorphism between values of type $\tCEExp$ and closed proto-expressions, which is mediated in one direction by the \emph{proto-expression decoding judgement}, $\decodeCondE{e}{\ce}$ that we will return to when we describe TLM application below (see accompanying technical report for the full conditions).
 
@@
 Second, the rule extends the expression TLM context, $\uPsi$, to associate the TLM identifier $\tsmv$ with $x$, as well as with the given expansion type and parse function.
-If $\tsmv$ was already defined, the previous definition is shadowed (the accompanying technical report gives some additional details on how TLM contexts are structured).
+If $\tsmv$ was already defined, the previous definition is shadowed (the accompanying technical report gives some additional details on TLM contexts).
 
@@
 If parsing succeeds, i.e.
-a value of the form $\aein{\mathtt{SuccessE}}{\ecand}$ results from evaluation, then $\ecand$ will be a value of type $\tCEExp$ (by type safety and the canonical forms lemma, assuming a well-formed expression TLM context).
+a value of the form $\aein{\mathtt{SuccessE}}{\ecand}$ results from evaluation, then $\ecand$ will be a value of type $\tCEExp$ (by type safety).
 We call $\ecand$ the \emph{encoding of the proto-expansion}.
-If the parse function produces a value labeled $\lbltxt{Error}$, then typed expansion fails and formally, no rule is necessary.
+If the parse function produces a value labeled $\lbltxt{Error}$, then typed expansion fails and no rule is necessary.
 
@@
 
-The typed pattern expansion rule governing pattern TLM application, Rule~\textsc{pe-ap-sptlm} in Fig.~\ref{fig:ee-ap-setlm}, is analogous but again simpler because there is no need to pass dependencies into patterns in $\miniVersePat$.\footnote{If there were a need to pass dependencies into patterns, we would need either pattern functions or to use an encoding trick, e.g.
-by requiring the result be a function containing a ``dummy'' match with one pattern.} We describe proto-pattern validation in Sec.~\ref{sec:s-PEV} below.
+The typed pattern expansion rule governing pattern TLM application, Rule~\textsc{pe-ap-sptlm} in Fig.~\ref{fig:ee-ap-setlm}, is analogous but again simpler because there is no need to pass dependencies into patterns in $\miniVersePat$.
+%\footnote{If there were a need to pass dependencies into patterns, we would need either pattern functions or to use an encoding trick, e.g.
+by requiring the result be a function containing a ``dummy'' match with one pattern.}
+We describe proto-pattern validation in Sec.~\ref{sec:s-PEV} below.
 % {\small\begin{mathpar}
@@
 Notice that the initial expansion-interal typing contexts, $\Delta$ and $\Gamma$, start out empty in the rules in Fig.~\ref{fig:ee-ap-setlm}.
-Together, this maintains context independence (defined formally below).
+This maintains context independence (defined formally below).
 
@@
 \label{sec:implementation}
-The implementation, \li{Relit}, which is an ongoing effort based on the finished design described in this paper, consists of a few changes to the context-free grammar of Reason together with a parse tree pre-processor for the OCaml compiler, currently using its PPX system (discussed in Sec.~\ref{sec:existing-approaches}).
+The implementation, \li{Relit}, which is an ongoing effort based on the design described in this paper, consists of a few extensions to the context-free syntax of Reason together with a parse tree pre-processor for the OCaml compiler, currently using its PPX system (discussed in Sec.~\ref{sec:existing-approaches}).
 
 \paragraph{TLM Definitions} The Reason grammar extension turns TLM definitions into module definitions of a stereotyped form.
-For example, the module \li{RegexNotation} defining \li{$regex} from Fig.~\ref{fig:regex-tlm-def} is, eliding a few minor details, shown below.
+For example, definition of \li{$regex} from Fig.~\ref{fig:regex-tlm-def} is, eliding a few minor details, implemented as shown below.
 The comments describe how the components correspond.
+
 \begin{lstlisting}
-  module RegexNotation = {
-    module Notation_regex___relit /* = $regex */ = { 
-      type t = Regex.t; /* expansion type */
-      module Lexer = RegexLexer; /* lexer module must match ocamllex sig */
-      module Parser = RegexParser; /* parser module must match menhir sig */
-      module Dependencies = { module Regex = Regex; } /* see text */
-      exception Apply(string); /* used in TLM applications, see below */
-    } 
-  }
+    module RelitInternalDefn_regex = struct
+      type t = Regex.t (* corresponds to "at Regex.t" *)
+      module Lexer_RegexLexer = struct end (* ... "lexer RegexLexer" *)
+      module Parser_RegexParser = struct end (* ... "parser RegexParser" *)
+      module Nonterminal_start = struct end (* ... ".start" *)
+      module Package_regex_parser = struct end (* ... "in package regex_parser" *)
+      module Dependencies = struct (* ... "dependencies = {" *)
+        module Regex = Regex
+      end
+      exception Apply of string * string (* used internally, see below *)
+    end
 \end{lstlisting}
 
-TLM definitions are constrained so that the signature of the generated module is a \emph{singleton}, i.e.
+The name of the TLM is stored as the module name, with the internal prefix \li{RelitInternalDefn_}.
+By encoding TLM definitions as modules, TLMs can be packaged inside other modules (\li{RegexNotation} in Fig.~\ref{fig:regex-tlm-def}) and it is straightforward to support scoped TLM abbreviations, as on Line 1 of Fig.~\ref{fig:first-tlm-example}.
+By encoding the relevant information from the TLM definition in module names, e.g.
+\li{Lexer_RegexLexer} and \li{Nonterminal_start}, we ensure that the corresponding signature is a \emph{singleton signature}, i.e.
+it uniquely identifies the TLM definition, so TLM definitions can be included in module signatures, exported from functors and packaged using the standard tooling.
+
+% TLM definitions are constrained so that the signature of the generated module is a \emph{singleton}, i.e.
 that it uniquely determines the module.
@@
 As such, TLM definition equality is decidable and TLM definitions can be tracked in module types (signatures) without any special effort.
-Note that this is more restrictive than in the formal system in the previous section, where parse functions were included inline, rather than by path, and dependencies were arbitrary values.
-It must be possible to resolve the implementation of the lexer and parser modules, not just their signatures, at expansion time, so lexer and module paths are lifted to the top level of the compilation unit.
+%Note that this is more restrictive than in the formal system in the previous section, where parse functions were included inline, rather than by path, and dependencies were arbitrary values.
+It must be possible to resolve the implementation of the lexer and parser modules, not just their signatures, at expansion time, so lexer and module paths must be separately packaged as described in Sec.~\ref{sec:provider-perspective}.
 
@@
 \begin{lstlisting}
-  open RegexNotation; /* ... */
-  let restriction_template = (gene) => [%relit raise(Notation_regex___relit.Apply(
-      "$(bisA)$(DNA.any_base)*$$(gene)$(DNA.any_base)*$(bisA)"))]
+  raise (RelitInternalDefn_regex.Apply
+          ("You're using relit syntax without the relit ppx!", "A|T|G|C") [@relit])
 \end{lstlisting}
 
-The Relit preprocessor rewrites TLM applications by following the specification in the previous sections differing only in that the ``dependency variable'' is a ``dependency path'', here to \li{Notation_regex___relit.Dependencies}.
+The Relit preprocessor rewrites TLM applications by following the specification in the previous section, differing only in that the ``dependency variable'' is a ``dependency path'', here to \li{Notation_regex___relit.Dependencies}.
 The only difficult is in finding the TLM definition at all, which due to our integration into the module system may have been accessed indirectly, e.g.
@@
 
-\paragraph{Proto-Expansion Validation} The typing phase of proto-expansion validation also defers to the OCaml typechecker, with spliced expressions  represented as variables of the corresponding segment type, resolved relative to the typing context determined at the call site.
-The recursively generated substitutions are, after validation is complete, substituted in the standard capture-avoiding manner.
+\paragraph{Proto-Expansion Validation} The typing phase of proto-expansion validation also defers to the OCaml typechecker, with spliced expressions  represented as exceptions ascribed the corresponding segment type.
+The recursively generated substitutions are, after validation is complete, substituted in the standard capture-avoiding manner by simply replacing splice references with variables, and then wrapping the proto-expansion in a function that is immediately applied to the splices.
 
@@
 
-In this section, we will give an overview of the many existing mechanisms that library providers might use to define new notation, and more specifically, new literal notation.
+This section will give an overview of the many existing mechanisms that library providers might use to define new notation, and more specifically, new literal notation.
 Rather than evaluating these mechanisms by asking ``how much syntactic power do they give the library provider?'', however, our approach is to ask ``what do these mechanisms quite reasonably \textbf{not} allow a library provider to express?'', using the rubric of six reasoning principles that this paper has developed.
@@
 Because the mechanism operates at the level of the context-free grammar, it is difficult for the client to define scoped abbreviations for these verbose marking tokens.
-TLMs can, in contrast, be abbreviated and distributed within modules following the usual scoping structure of the language.
-Composition is via splicing, rather than direct grammar composition.
-
-Some programming languages, notably including theorem provers like Coq \cite{Coq:manual} and Agda \cite{norell2007towards}, support ``mixfix'' notation directives \cite{wieland2009parsing,missura1997higher,5134}.
+TLMs can, in contrast, be abbreviated and distributed within modules following the usual scoping structure.
+Composition is via splicing, rather than grammar composition.
+
+Some programming languages, notably including theorem provers like Agda \cite{norell2007towards} and Coq \cite{Coq:manual}, support ``mixfix'' notation directives \cite{wieland2009parsing,missura1997higher,5134}.
 Many of these systems enforce \textbf{Capture Avoidance} and application-site \textbf{Context Independence} \cite{5134,DBLP:conf/gpce/TahaJ03,Coq:manual,DBLP:conf/ifl/DanielssonN08}.
@@
 The trade-off is that TLMs do not require a fully mechanized host language definition.
-Finally, there is no clear notion of parameterization or partial application in SoundExt or other syntax definition systems, so it would be difficult to define notation for, for example, all finite map implementations as we demonstrated in Sec.~\ref{sec:ptsms}.
+Finally, there is no clear notion of parameterization or partial application in these systems, so it would be difficult to define notation for, for example, all finite map implementations as we demonstrated in Sec.~\ref{sec:ptsms}.
 
@@
 
-\emph{Hygienic} macro systems prevent, or abstractly account for \cite{DBLP:conf/esop/HermanW08,Herman10:Theory}, \textbf{Capture}, and they enforce application-site \textbf{Context Independence} \cite{Kohlbecker86a,DBLP:conf/popl/Adams15,DBLP:conf/popl/ClingerR91,DBLP:journals/lisp/DybvigHB92}.
+% \emph{Hygienic} macro systems prevent, or abstractly account for \cite{DBLP:conf/esop/HermanW08,Herman10:Theory}, \textbf{Capture}, and they enforce application-site \textbf{Context Independence} \cite{Kohlbecker86a,DBLP:conf/popl/Adams15,DBLP:conf/popl/ClingerR91,DBLP:journals/lisp/DybvigHB92}.
 The critical problem, detailed in Sec.~\ref{sec:context-dependence}, is that the standard context independence discipline makes it impossible to repurpose string literal forms to introduce compositional literal forms at other types.
 
-\todo{integrate me}{}Hygienic term-rewriting macro systems, like those available in various Lisp-family languages \cite{mccarthy1978history} and in Scala \cite{ScalaMacros2013}, cannot be used to repurpose string literals for composite literal notation at other types for exactly this reason.
+\emph{Hygienic} term-rewriting macro systems, like those available in various Lisp-family languages \cite{mccarthy1978history} and in Scala \cite{ScalaMacros2013}, prevent, or abstractly account for \cite{DBLP:conf/esop/HermanW08,Herman10:Theory}, \textbf{Capture}, and they enforce application-site \textbf{Context Independence} \cite{Kohlbecker86a,DBLP:conf/popl/Adams15,DBLP:conf/popl/ClingerR91,DBLP:journals/lisp/DybvigHB92}.
+However, a strictly hygienic term-rewriting macro system cannot be used to repurpose string literals for composite literal notation at other types because the system cannot distinguish  sub-terms in the generated expansion that came from spliced segments of the literal body.
+From the perspective of the context-free syntax, spliced sub-terms are neither an argument to the macro nor even a sub-term of an argument for which a tree path can be assigned \cite{DBLP:conf/esop/HermanW08,Herman10:Theory,gorn1965explicit}.
+Instead, they arise as the result of performing a complex operation---parsing---on an arbitrary sub-sequence of the string literal passed into the macro.
+TLMs address this problem by distinguishing spliced sub-terms explicitly in proto-expansions.
+
+% \todo{integrate me}{}Hygienic term-rewriting macro systems, like those available in various Lisp-family languages \cite{mccarthy1978history} and in Scala \cite{ScalaMacros2013}, cannot be used to repurpose string literals for composite literal notation at other types for exactly this reason.
 These systems would find that the appearance of \li{DNA.any_base}  violates their context independence condition (which is one aspect of {hygiene}), because \li{DNA.any_base} is not, from the perspective of the context-free syntax, an argument to the macro nor even a sub-term of an argument for which a tree path can be assigned \cite{DBLP:conf/esop/HermanW08,Herman10:Theory,gorn1965explicit}.
 Instead, it arises as the result of performing a complex operation---parsing---on an arbitrary sub-sequence of the string literal passed into the macro.
-(It is  possible for a macro to selectively opt out of the hygiene discipline---see Sec.~\ref{sec:existing-approaches} for more on why relying on this sort of ``escape hatch'' is an unreasonable solution.) %From the macro expander's perspective, it has appeared by magic''.
+
+It is  possible in many production-grade hygienic macro systems, e.g.
+in Racket \cite{Flatt:2012:CLR:2063176.2063195}, for a macro to granularly opt out of the hygiene discipline in parts of an expansion, but clients cannot be sure that this powerful ``escape hatch'' was used in a disciplined way.
+However, a typed literal macro system could be implemented as a layer on top of an existing hygienic macro system, e.g.
+in Racket, by internally deploying the escape hatch exactly as specified by this paper.
+%From the macro expander's perspective, it has appeared by magic''.
 
@@
 Segmentations can be considered a refinement of the tree paths in prior work \cite{Herman10:Theory,gorn1965explicit}.
+Note also that the contributions are relevant even in dynamic languages like Racket
+(by simply ignoring the typing aspects, or by inserting corresponding 
+contracts).
 
 Research on typed \emph{staging macro systems} \cite{DBLP:conf/popl/DaviesP96} like MetaML \cite{Sheard:1999:UMS}, MetaOCaml \cite{DBLP:conf/flops/Kiselyov14}, MacroML \cite{ganz2001macros,DBLP:conf/gpce/TahaJ03} and Typed Template Haskell \cite{tth} is not directly applicable to the problem of defining new literal forms because the syntax tree of the arguments cannot be inspected at all (staging macros are used mainly for optimization).
-Sec.~\ref{sec:metaocaml} discussed how the typed quotations from these systems can help TLM providers reason about the type-correctness of their parsers.
-Future work could help address some of the mentioned limitations.
+Sec.~\ref{sec:metaocaml} discussed how the typed quotations from these systems can help TLM providers reason about the type-correctness of expansions.% Future work could help address some of the mentioned limitations.
 
@@
 
-This brings us to the most closely related work, that of \citet{TSLs} on \emph{type-specific languages} (TSLs).
+This brings us to the most closely related work: our own prior work on \emph{type-specific languages (TSLs)} \cite{TSLs}.
 Like simple expression TLMs (Sec.
@@
 Accordingly, different TLMs can operate at the same type, and they can operate at any type, including structural types.
-In a subsequent short paper, \citet{sac15} suggested explicit application of simple expression TLMs  also in a bidirectional typed setting \cite{Pierce:2000:LTI:345099.345100}, but this paper did not have any formal content.
+In a subsequent short paper, \citet{sac15} suggested explicit application of simple expression TLMs  also in a bidirectional typed setting \cite{Pierce:2000:LTI:345099.345100}, but that paper did not have any formal content.
 With TLMs, it is not necessary for the language to be bidirectionally typed (see Sec.
@@
 a variant of the Typed Expression Expansion theorem from Sec.
-\ref{sec:s-metatheory}), though the induction principle used in the proof is not clear.
+\ref{sec:s-metatheory}).
 It does not establish the remaining abstract reasoning principles that have been the major focus of this paper.
-There appears to be a context independence condition but it does not allow for any dependencies whatsoever, which is unreasonably restrictive.
+There is a context independence condition but it does not allow for any dependencies whatsoever, which is unreasonably restrictive.
 There is no formal hygiene theorem and indeed the formal system in the paper does not correctly handle substitution or capture avoidance, issues we emphasized because they were non-obvious in Sec.
@@
 Finally, this prior work did not consider pattern matching, module system integration or type- or module-parameterized type families.
-This paper addresses all of these, resulting in a design suitable for integration into Reason/OCaml.
+This paper addresses all of these, resulting in a design suitable for integration into Reason/OCaml and other languages broadly in the ML tradition.
 
@@
 This class includes all of the examples enumerated in Sec.
-\ref{sec:intro} (up to the choice of outermost delimiter), the varied and non-trivial case studies outlined in this paper and in the accompanying technical report, and the examples collected from the empirical study by \citet{TSLs}.
+\ref{sec:intro} (up to the choice of outermost delimiter), the varied and non-trivial case studies outlined in this paper, and the examples collected from the empirical study by \citet{TSLs}.
 We anticipate many more interesting examples will emerge as the Reason community explores the mechanism.
@@
 Although we emphasized integration with modules, our formal system demonstrates that modules are not necessary to capture the fundamental ideas in this paper.
-We intend this paper to be useful to the designers of languages far from the ML tradition.
-More generally, we intend for the ``reasoning principles first'' approach advocated in this paper to be a useful intellectual framework for language designers considering the merits of other ``conveniences''.
-% We must leave a detailed empirical evaluation of the impact of particular TLMs on various quality attributes, like programmer productivity and program comprehensibility, as future work.
+We intend this paper to be useful to the designers of languages both near and distant from the ML tradition.
+More generally, we intend for the ``reasoning principles first'' design philosophy that we followed in this paper to be a useful example for language designers considering the merits of other ``conveniences'' --- syntactic cost alone is not the whole story.% We must leave a detailed empirical evaluation of the impact of particular TLMs on various quality attributes, like programmer productivity and program comprehensibility, as future work.
 % Our own implementation efforts have focused on an emerging alternative front-end for OCaml called Reason, but Scala supports analogous features \cite{conf/oopsla/AminRO14} and would also be a suitable host for the TLM mechanism.
@@
 However, several important questions having to do with TLM-specific syntax highlighting, incremental parsing and error recovery \cite{graham1979practical} remain to be considered.
-Another interesting direction would be to generalize TLMs to support non-textual notation in the setting of a structure editor \cite{DBLP:conf/popl/OmarVHAH17}.
+Another interesting direction would be to generalize TLMs to support non-textual notation in the setting of a structure editor \cite{DBLP:conf/popl/OmarVHAH17,snapl17}.
 
@@
 Instead, the programmer need only be given knowledge of the expansion type from the TLM definition and the segmentation inferred by the expander at each application site, which carries a small volume of information that can be communicated straightforwardly by standard editor services.
-Certain questions related to the binding structure simply do not need to be asked due to the strict context independence and capture avoidance discipline of the system, enabled by the explicit tracking of dependencies, and of spliced segments.
+Certain questions related to the binding structure simply do not need to be asked due to the strict context independence and capture avoidance discipline of the system, enabled by the explicit tracking of dependencies and of spliced segments.
 
@@
 Overall, we believe that TLMs represent a distinctly \emph{reasonable} and \emph{expressive} new point in the design space of literal notation definition systems.
+\vspace{-5px}
 

### 2018-06-25 icfp18/syntax-icfp18.tex
fixed a couple of line number omissions and dropped unnecessary distinction between re_expr and proto_expr

@@
     let spliced : Segment.t => Parsetree.core_type => t; };
-  notation $re_expr at Parsetree.expression { /* ... (see text) ... */ };
-  notation $proto_expr at ProtoExpr.t = $re_expr;
+  notation $proto_expr at ProtoExpr.t { /* ... (see text) ... */ };
 \end{lstlisting}
@@
 In the literature, literal notation for the host language's own parse trees is called \emph{quasiquotation}, and splicing is referred to as \emph{unquotation} or \emph{antiquotation} \cite{Bawd99a,shabalin2013quasiquotes,mainland2007s}.
-The TLM \li{$re_expr}  specified on Line 10 of Fig.~\ref{fig:relit-util} provides quasiquotation for OCaml parse trees using Reason's surface syntax.
+The TLM \li{$proto_expr}  specified on Line 8 of Fig.~\ref{fig:relit-util} provides quasiquotation for OCaml parse trees using Reason's surface syntax.
 The accompanying technical report describes its implementation.
-For clarity, the \li{$proto_expr} TLM is defined as a synonym for \li{$re_expr} on Line~12 of Fig.~\ref{fig:relit-util} (the two TLMs may diverge in the future).
 We apply this TLM many times in the semantic actions in Fig.~\ref{fig:regex-parser}.
@@
 
-To support antiquotation, the \li{$re_expr} TLM repurposes TLM application and generalized literal forms, as can be observed throughout Fig.~\ref{fig:regex-parser}.
+To support antiquotation, the \li{$proto_expr} TLM repurposes TLM application and generalized literal forms, as can be observed throughout Fig.~\ref{fig:regex-parser}.
 In particular, parse tree splicing is supported by unadorned generalized literal forms, e.g.
@@
 
-TLM-related forms are available to be repurposed in this way by \li{$proto_expr} because there is no reasonable way for \li{$re_expr} to support antiquotation across notational boundaries.
-For example, if \li{$re_expr} had instead used some other antiquotation notation, e.g.
-\li{^(}$e$\li{)}, then consider this example:
+TLM-related forms are available to be repurposed in this way by \li{$proto_expr} because there is no reasonable way for \li{$proto_expr} to support antiquotation across notational boundaries.
+For example, if \li{$proto_expr} had instead used \li{^(}$e$\li{)} for antiquotation, then consider this example:
 \begin{lstlisting}[numbers=none]
@@
 The sequence \li{^(r1)} is in expression position, so antiquotation occurs.
-However, there is no way for \li{$re_expr} to know how the TLM that \li{$m} will eventually refer to will parse the sequence \li{^(r2)}---it may end up in a spliced expression, or it may have some other meaning---so it cannot perform antiquotation.
+However, there is no way for \li{$proto_expr} to know how the TLM that \li{$m} will eventually refer to will parse the sequence \li{^(r2)}---it may end up in a spliced expression, or it may have some other meaning---so it cannot perform antiquotation.
 In other words, TLM expansion is fundamentally ``outside-in''.
-Notice that the segmentation does communicate the fact that \li{r2} has not been antiquoted (it remains lavender), but to avoid confusion, we decided not to support the generation of TLM applications via \li{$re_expr}.
+Notice that the segmentation does communicate the fact that \li{r2} has not been antiquoted (it remains lavender), but to avoid confusion, we decided not to support the generation of TLM applications via \li{$proto_expr}.
 
@@
 % the \li{Relit} library includes a trivial lexer, \li{Relit.TrivLexer}, and a few functors that take care of the necessary boilerplate in situations where we simply want to define the parser as a function of type \li{string => Relit.ProtoExpr.t}.
-%For example, the accompanying technical report briefly describes how \li{$re_expr}, described above, can be expressed in terms of the Reason parser, a parse tree serializer and a simple tree transformation.
+%For example, the accompanying technical report briefly describes how \li{$proto_expr}, described above, can be expressed in terms of the Reason parser, a parse tree serializer and a simple tree transformation.
 Let us consider one substantial example of this approach here: the \li{$html} TLM, shown being applied in Fig.~\ref{fig:nesting-example}, is implemented using an existing, ``production grade'' HTML parsing library, \li{Markup.ml} \cite{markupml}, together with a simple post-processing step as outlined in Fig.~\ref{fig:html-case-study}.
@@
 
-The \li{ProtoHtml.parse} function, outlined on Lines 8-9, generates a value of the type \li{ProtoHtml.t} defined on Lines 4-7 and so named because it is similar to \li{Html.t} but distinguishes spliced segments, like a proto-expression.
+The \li{ProtoHtml.parse} function, elided on Lines 8-9, generates a value of the type \li{ProtoHtml.t} defined on Lines 4-7 and so named because it is similar to \li{Html.t} but distinguishes spliced segments, like a proto-expression.
 We omit the definition of \li{ProtoHtml.parse} but conceptually, it starts by creating a character stream from the literal body, then from that creates a \li{Markup.ml} HTML content stream, then transforms that to a stream that emits the signals from the HTML content stream until it sees a splice start tag (which is reported as a syntax error relative to the HTML standard), at which point it calls \li{Segment.read_to}, described earlier, to advance the original character stream until it recognizes the closing tag, emitting the generated segment and then returning control to the HTML content stream.
@@
     in package map_parser;
-  }
+  };
 \end{lstlisting}
@@
       lexer MapLexer parser MapParser in package map_parser;
-      dependencies { module M = M; type a = A.a; } } }
+      dependencies { module M = M; type a = A.a; } 
+    };
+  };
 \end{lstlisting}
@@
 % \end{equation*}
-\begin{figure*}
+\begin{figure}
 {\small\begin{mathpar}
@@
 \end{mathpar}}
-\vspace{-4px}
 \caption{The typed expansion rules for expression and pattern TLM definitions.}
@@
 \vspace{-5px}
-\end{figure*}
+\end{figure}
 % \end{subequations}

### 2018-06-25 icfp18/syntax-icfp18.tex
Cleaned up bib
added brief mention of recent work by Pombrio and Krishnamurthy
added brief mention of Rust macros

@@
 
-Some programming languages, notably including theorem provers like Agda \cite{norell2007towards} and Coq \cite{Coq:manual}, support ``mixfix'' notation directives \cite{wieland2009parsing,missura1997higher,5134}.
+Some programming languages, notably including Rust \cite{Rust/Macros} and theorem provers like Agda \cite{norell2007towards} and Coq \cite{Coq:manual}, support ``mixfix'' notation directives \cite{wieland2009parsing,missura1997higher,5134}.
 Many of these systems enforce \textbf{Capture Avoidance} and application-site \textbf{Context Independence} \cite{5134,DBLP:conf/gpce/TahaJ03,Coq:manual,DBLP:conf/ifl/DanielssonN08}.
@@
 \li{if _ then _ else _}.
-Coq has some \emph{ad hoc} extensions for list-like literals \cite{Coq:manual}.
+Rust and Coq have some \emph{ad hoc} extensions for list-like literals \cite{Rust/Macros,Coq:manual}.
 These systems cannot express the example literal forms from this paper, because they can have any number of spliced terms.
@@
 The system then attempts to automatically verify that the expansion logic, expressed using a rewrite system, rather than an arbitrary function, is sound relative to these derived rules, so it is possible to reason about \textbf{Expansion Typing}.
-SoundExt does not enforce hygiene, i.e.
+Recent work by \citet{DBLP:conf/pldi/PombrioK18} instead infers the typing rules.
+These systems do not enforce hygiene, i.e.
 expansions might depend on the context and intentionally induce capture.

### 2018-06-25 icfp18/syntax-icfp18.tex
added brief mention of dirty macros paper

@@
 It is  possible in many production-grade hygienic macro systems, e.g.
-in Racket \cite{Flatt:2012:CLR:2063176.2063195}, for a macro to granularly opt out of the hygiene discipline in parts of an expansion, but clients cannot be sure that this powerful ``escape hatch'' was used in a disciplined way.
+in Racket \cite{Flatt:2012:CLR:2063176.2063195}, for a macro to granularly opt out of the hygiene discipline in parts of an expansion \cite{kiselyov2002write}, but clients cannot be sure that this powerful ``escape hatch'' was used in a disciplined way.
 However, a typed literal macro system could be implemented as a layer on top of an existing hygienic macro system, e.g.

### 2018-06-25 icfp18/syntax-icfp18.tex
added brief mention of debugging paper and active code completion paper

@@
 At several points, we allude to editor integration.
-However, several important questions having to do with TLM-specific syntax highlighting, incremental parsing and error recovery \cite{graham1979practical} remain to be considered.
-Another interesting direction would be to generalize TLMs to support non-textual notation in the setting of a structure editor \cite{DBLP:conf/popl/OmarVHAH17,snapl17}.
+However, several important questions having to do with TLM-specific syntax highlighting, incremental parsing and error recovery \cite{graham1979practical}, as well as debugging \cite{DBLP:journals/scp/CulpepperF10}, remain to be considered.
+Another interesting direction would be to generalize TLMs to support non-textual notation in a structure editor \cite{DBLP:conf/popl/OmarVHAH17,snapl17}, thereby internalizing active code completion \cite{Omar:2012:ACC:2337223.2337324}.
 
@@
 Instead, the programmer need only be given knowledge of the expansion type from the TLM definition and the segmentation inferred by the expander at each application site, which carries a small volume of information that can be communicated straightforwardly by standard editor services.
-Certain questions related to the binding structure simply do not need to be asked due to the strict context independence and capture avoidance discipline of the system, enabled by the explicit tracking of dependencies and of spliced segments.
+Certain questions related to the binding structure simply do not need to be asked due to the strict context independence and capture avoidance discipline of the system.%enabled by the explicit tracking of dependencies and of spliced segments.
 

### 2018-06-27 icfp18/supplement/icfp18-supplement.tex
formatting of supplement into CMU TR

@@
 \documentclass[12pt]{report}
+\usepackage{fullpage,cmu-titlepage2}
+
 \newif\ificfp
@@
 \allowdisplaybreaks[1]
-\newtheorem{theorem}{Theorem}[chapter]
+\newtheorem{theorem}{Theorem}[section]
 \newtheorem{lemma}[theorem]{Lemma}
@@
 % \makeatother
+\title{Reasonably Programmable Literal Notation: Supplemental Material}
+
+\author{Cyrus Omar \and Jonathan Aldrich}
+
+\date{July 8, 2018}
+
+\abstract{
+This report presents a complete technical account of the formal system that was described in the accompanying paper, as well as more details on the quasiquotation TLM.
+}
+
+\keywords{extensible syntax; macros; hygiene; type systems}
+
+\trnumber{CMU-ISR-18-104}
+
+%\citationinfo{}
+
+%\arpasupport{fox}
+\support{This work was
+supported in part by AFRL and DARPA under agreement \#FA8750-16-2-0042;
+ by the NSA under lablet contract \#H98230-14-C-0140; and by Ravi Chugh
+ at the University of Chicago.}
+% \authorsupport{The author holds a Froboz Gradual Fellowship.}
+
+% \otherdisclaimer{NSF}
 
 \begin{document}
-\title{Reasonably Programmable Literal Notation\\ (ICFP 2018 Anonymous Supplemental Material)}
-\date{March 16, 2018}
+\renewcommand*{\thepage}{title-\arabic{page}} 
 \maketitle
+\renewcommand*{\thepage}{\arabic{page}} 
+\appendix
+
 \tableofcontents
@@
 
-\chapter{Additional Examples}
-
-\section{HTML Literals} 
-Reason is seeing increasing adoption in the domain of client-side web programming, where generating and manipulating HTML data is central. 
-There are many ways to represent HTML data in Reason, but one simple and useful representation is specified by \li{Html.t}, defined in Fig.~\ref{fig:html-type-def}.
+% \section{HTML Literals} 
+% Reason is seeing increasing adoption in the domain of client-side web programming, where generating and manipulating HTML data is central. 
+% There are many ways to represent HTML data in Reason, but one simple and useful representation is specified by \li{Html.t}, defined in Fig.~\ref{fig:html-type-def}.
 However, manually applying the constructors of this type is tedious and, for many web programmers, the induced notation is unfamiliar.
@@
 Instead, we would like HTML literal notation.
-The \li{$html} TLM defined in Fig.~\ref{fig:html-tlm-def} implements the standard HTML notation extended with the tags \li{<$>} and \li{<$$>}, which support \li{Html.t} splicing and \li{string} splicing, respectively.
+% The \li{$html} TLM defined in Fig.~\ref{fig:html-tlm-def} implements the standard HTML notation extended with the tags \li{<$>} and \li{<$$>}, which support \li{Html.t} splicing and \li{string} splicing, respectively.
 It also supports the suffix \li{...} on a tag, which causes the body of the tag to be a splice of type \li{list(Html.t)}.\footnote{We would also in practice have splicing for  attribute values, but for simplicity we omit these.} The example in the paper demonstrated the first of these.
@@
 
-% The HTML standard is complex.
+% % The HTML standard is complex.
 In addition, good error handling is important in practice.
@@
 Let us outline how we can use this library to implement HTML literals, with support for splicing as just specified.
-In implementing HTML literals, using a parser generator may not be ideal.
+% In implementing HTML literals, using a parser generator may not be ideal.
 Instead, ``post-processing'' the result of an existing parser is a good approach.
@@
 
-The \li{ProtoHtml.parse} function, outlined on Lines 8-9, generates a value of the type \li{ProtoHtml.t} defined on Lines 2-6 and so named because it is similar to \li{Html.t} but distinguishes spliced segments, like a proto-expression.
+% The \li{ProtoHtml.parse} function, outlined on Lines 8-9, generates a value of the type \li{ProtoHtml.t} defined on Lines 2-6 and so named because it is similar to \li{Html.t} but distinguishes spliced segments, like a proto-expression.
 We omit the definition of \li{ProtoHtml.parse} but conceptually, it starts by creating a character stream from the literal body, then from that creates a \li{Markup.ml} HTML content stream, then transforms that to a stream that emits the signals from the HTML content stream until it sees a splice start tag (which is reported as a syntax error relative to the HTML standard), at which point it calls \li{Segment.read_to}, which was described in the paper, to advance the original character stream until it recognizes the closing tag, emitting the generated segment and then returning control to the HTML content stream.
@@
 
-The \li{ExprExpander.expand_proto} function defined, with some helper functions, on Lines~9-20 finishes the job of the TLM parser by mapping the parsed proto-HTML to a proto-expression using the \li{$proto_expr} TLM previously described (in this case, using the \li{$list} antiquotation form that produces a parse tree of OCaml's built in list literal form from a list of parse trees).
+% The \li{ExprExpander.expand_proto} function defined, with some helper functions, on Lines~9-20 finishes the job of the TLM parser by mapping the parsed proto-HTML to a proto-expression using the \li{$proto_expr} TLM previously described (in this case, using the \li{$list} antiquotation form that produces a parse tree of OCaml's built in list literal form from a list of parse trees).
 It would be straightforward to reuse the \li{ProtoHtml} module for different variations on HTML encodings, implementing only the expander anew.
@@
 
-Reason currently builds in ``JSX'' literals, which also support an HTML-like notation with support for splicing in a somewhat idiosynchratic manner motivated by certain JavaScript libraries (in particular, React).
+% Reason currently builds in ``JSX'' literals, which also support an HTML-like notation with support for splicing in a somewhat idiosynchratic manner motivated by certain JavaScript libraries (in particular, React).
 The expansion of this notation must be interpretted by a PPX rewriter.
@@
 
-\section{Implementing Quasiquotation} The Reason quasiquotation TLMs, e.g.
-\li{$re_expr}, can similarly be implemented by the functional transformations outlined below, with an example from each step (grossly simplified from the actual \li{Parsetree} representation) on the right.
+\chapter{Implementing Quasiquotation} The Reason quasiquotation TLMs, e.g.
+\li{$proto_expr}, can be implemented by the functional transformations outlined below, with an example from each step (grossly simplified from the actual \li{Parsetree} representation) on the right.
 In words, \li{$re_expr} first programmatically invokes the Reason parser on the literal body.
@@
 The final step is to implement antiquotation as described above by repurposing the generalized literal forms in the body, using the source locations from the first parse tree, which have  been carried into the second parse tree as constants.
-\begin{lstlisting}[numbers=none,basicstyle=\ttfamily\fontsize{8pt}{1em}\selectfont]
+\begin{lstlisting}[numbers=none,basicstyle=\ttfamily\fontsize{8.3pt}{1em}\selectfont]
 body             "2 + `(xyz)`" 

### 2018-06-27 icfp18/supplement/icfp18-supplement.tex
tweaks to description of typographic conventions

@@
 The final step is to implement antiquotation as described above by repurposing the generalized literal forms in the body, using the source locations from the first parse tree, which have  been carried into the second parse tree as constants.
-\begin{lstlisting}[numbers=none,basicstyle=\ttfamily\fontsize{8.3pt}{1em}\selectfont]
+\begin{lstlisting}[numbers=none,basicstyle=\ttfamily\fontsize{8.4pt}{1em}\selectfont]
 body             "2 + `(xyz)`" 

### 2018-06-28 icfp18/syntax-icfp18.tex
formatting tweaks required by ACM

@@
 % \documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=false,printccs=false,printacmref=false}
-\documentclass[acmsmall]{acmart}
+\documentclass[acmsmall,screen]{acmart}
 % \documentclass[sigplan,review,10pt,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
@@
   {\z@}{1ex \@plus .2ex \@minus .2ex}{-1em}%
-  {\normalfont\normalsize\bfseries}%
+  {\sffamily\normalsize\bfseries\maybe@addperiod}%
+}
+
+\newcommand{\maybe@addperiod}[1]{%
+  #1\@addpunct{.}%
 }
@@
 
-\makeatletter\if@ACM@journal\makeatother
-%% Journal information (used by PACMPL format)
-%% Supplied to authors by publisher for camera-ready submission
-\acmJournal{PACMPL}
-\acmVolume{1}
-\acmNumber{1}
-\acmArticle{1}
-\acmYear{2018}
-\acmMonth{9}
-\acmDOI{10.1145/nnnnnnn.nnnnnnn}
-\startPage{1}
-\else\makeatother
-%% Conference information (used by SIGPLAN proceedings format)
-%% Supplied to authors by publisher for camera-ready submission
-\acmConference[ICFP'18]{International Conference on Functional Programming}{September 23--29, 2018}{St. Louis, MO, USA}
-\acmYear{2018}
-\acmISBN{978-x-xxxx-xxxx-x/YY/MM}
-\acmDOI{10.1145/nnnnnnn.nnnnnnn}
-\startPage{1}
-\fi
+% \makeatletter\if@ACM@journal\makeatother
+% %% Journal information (used by PACMPL format)
+% %% Supplied to authors by publisher for camera-ready submission
+% \acmJournal{PACMPL}
+% \acmVolume{1}
+% \acmNumber{1}
+% \acmArticle{1}
+% \acmYear{2018}
+% \acmMonth{9}
+% \acmDOI{10.1145/nnnnnnn.nnnnnnn}
+% \startPage{1}
+% \else\makeatother
+% %% Conference information (used by SIGPLAN proceedings format)
+% %% Supplied to authors by publisher for camera-ready submission
+% \acmConference[ICFP'18]{International Conference on Functional Programming}{September 23--29, 2018}{St. Louis, MO, USA}
+% \acmYear{2018}
+% \acmISBN{978-x-xxxx-xxxx-x/YY/MM}
+% \acmDOI{10.1145/nnnnnnn.nnnnnnn}
+% \startPage{1}
+% \fi
 
@@
 %\setcopyright{acmlicensed}
+%%% The following is specific to ICFP'18 and the paper
+%%% 'Reasonably Programmable Literal Notation'
+%%% by Cyrus Omar and Jonathan Aldrich.
+%%%
 \setcopyright{rightsretained}
-%\copyrightyear{2017}           %% If different from \acmYear
+\acmPrice{}
+\acmDOI{10.1145/3234629}
+\acmYear{2018}
+\copyrightyear{2018}
+\acmJournal{PACMPL}
+\acmVolume{2}
+\acmNumber{ICFP}
+\acmArticle{40}
+\acmMonth{9}
 
@@
   \state{IL}
-  % \country{USA}
+  \country{USA}
 }
@@
   \state{PA}
-  % \country{USA}
+  \country{USA}
 }
@@
 %% comma separated list
-% \keywords{keyword1, keyword2, keyword3}  %% \keywords is optional
+\keywords{parsing, macro systems, hygiene, type systems, pattern matching}  %% \keywords is optional
 
@@
 \begin{figure}[h]
-\vspace{-9px}
+% \vspace{-9px}
 \begin{lstlisting}

### 2018-07-08 icfp18/supplement/icfp18-supplement.tex
final submitted version of paper and supplement!

@@
 supported in part by AFRL and DARPA under agreement \#FA8750-16-2-0042;
- by the NSA under lablet contract \#H98230-14-C-0140; and by Ravi Chugh
- at the University of Chicago.}
+ by the NSA under lablet contract \#H98230-14-C-0140.}
 % \authorsupport{The author holds a Froboz Gradual Fellowship.}

### 2018-07-08 icfp18/syntax-icfp18.tex
final submitted version of paper and supplement!

@@
 \acmPrice{}
-\acmDOI{10.1145/3234629}
+\acmDOI{10.1145/3236801}
 \acmYear{2018}
@@
 \acmNumber{ICFP}
-\acmArticle{40}
+\acmArticle{106}
 \acmMonth{9}
@@
 
-\input{thesis/macros}
+\input{macros}
 
@@
 \begin{figure}[h]
-% \vspace{-9px}
+\vspace{-4px}
 \begin{lstlisting}
@@
 \label{fig:urweb}
-\vspace{-10px}
+\vspace{-5px}
 \end{figure}
@@
 By the end of this section, we will have theorems that formalize the reasoning principles that were outlined informally in the previous sections.
-$\miniVersePat$ consists of an {unexpanded language} ({UL}) defined by typed expansion to the core language, named for the purposes of this paper $\miniVerseBase$.
+$\miniVersePat$ consists of an {unexpanded language} ({UL}) defined by typed expansion to the core language, named for the purposes of this paper $\miniVerseCore$.
 
@@
 Fig.
-\ref{fig:U-expanded-terms} gives the syntax of the core language, $\miniVerseBase$.
+\ref{fig:U-expanded-terms} gives the syntax of the core language, $\miniVerseCore$.
 This language forms a standard pure functional language with partial function types, quantification over types, recursive types, labeled product types, labeled sum types and support for pattern matching.\footnote{The accompanying technical report shows how to remove pattern matching  (and pattern TLMs) from the calculus by adding the usual elimination forms, e.g.
@@
 These judgements are inductively defined in the accompanying technical report along with necessary auxiliary structures and standard lemmas.
-The dynamic semantics of $\miniVerseBase$ is organized around the judgements $\isvalU{e}$, which says that $e$ is a value, and $\evalU{e}{e'}$, which says that $e$ evaluates to the value $e'$.
+The dynamic semantics of $\miniVerseCore$ is organized around the judgements $\isvalU{e}$, which says that $e$ is a value, and $\evalU{e}{e'}$, which says that $e$ evaluates to the value $e'$.
 As in ML, evaluation can diverge (general recursion is possible via recursive types \cite{pfple1}).
@@
 \end{minipage}
-\caption[Syntax of the core language of $\miniVersePat$]{Syntax of the core language, $\miniVerseBase$, which is an entirely standard typed lambda calculus.
+\caption[Syntax of the core language of $\miniVersePat$]{Syntax of the core language, $\miniVerseCore$, which is an entirely standard typed lambda calculus.
 Metavariable $x$ ranges over variables, $t$ over type variables, $\ell$ over labels and $L$ over finite sets of labels.
@@
 The rules governing proto-expansion validation for these common forms  in the accompanying technical report correspondingly mirror the typing rules.
-Splicing scenes---$\escenev$, $\tscenev$ and $\pscenev$--pass opaquely through these rules, i.e.
+Splicing scenes---$\tscenev$, $\escenev$ and $\pscenev$--pass opaquely through these rules, i.e.
 none of these rules can access the application site contexts.
@@
 \paragraph{TLM Definitions} The Reason grammar extension turns TLM definitions into module definitions of a stereotyped form.
-For example, definition of \li{$regex} from Fig.~\ref{fig:regex-tlm-def} is, eliding a few minor details, implemented as shown below.
+For example, the definition of \li{$regex} from Fig.~\ref{fig:regex-tlm-def} is, eliding a few minor details, implemented as shown below.
 The comments describe how the components correspond.
@@
 \begin{lstlisting}
-    module RelitInternalDefn_regex = struct
+    module RelitInternalDefn_regex = struct (* notation $regex *)
       type t = Regex.t (* corresponds to "at Regex.t" *)
@@
 
-The name of the TLM is stored as the module name, with the internal prefix \li{RelitInternalDefn_}.
-By encoding TLM definitions as modules, TLMs can be packaged inside other modules (\li{RegexNotation} in Fig.~\ref{fig:regex-tlm-def}) and it is straightforward to support scoped TLM abbreviations, as on Line 1 of Fig.~\ref{fig:first-tlm-example}.
-By encoding the relevant information from the TLM definition in module names, e.g.
+% The name of the TLM is stored as the module name, with the internal prefix \li{RelitInternalDefn_}. 
+By encoding TLM definitions as modules and encoding the information from the TLM definition in module names, e.g.
 \li{Lexer_RegexLexer} and \li{Nonterminal_start}, we ensure that the corresponding signature is a \emph{singleton signature}, i.e.
-it uniquely identifies the TLM definition, so TLM definitions can be included in module signatures, exported from functors and packaged using the standard tooling.
+it uniquely identifies the TLM definition, so TLM definitions can be included in module signatures, exported from within module definitions (including those generated by functors) and packaged using the standard tooling.
 
@@
 Because the TLM application raises an exception, its type can be generalized arbitrarily.
-As long as unrelated type errors at a particular depth are resolved (treating the TLM applications at that depth as ``holes'') expansion at that depth can proceed.
+As long as unrelated type errors at a particular depth are resolved (treating the TLM applications at that depth as ``holes''), expansion at that depth can proceed.
 This occasionally causes types that incidentally appear in error messages to be more general than the programmer might expect (e.g.
@@
 We are working to improve error reporting and, more generally, improving integration with various other tools in the Reason / OCaml ecosystem.
+
+Pattern TLMs are similar, but OCaml does not support exceptions in pattern position so we require a preprocessing step that turns any expression that the pattern binds over into an exception.
 
@@
 TLM providers can use existing, mature parser generators and parse tree representations, and can opt-in to a stronger typing discipline by using various implementations of MetaOCaml.
-Overall, we believe that TLMs represent a distinctly \emph{reasonable} and \emph{expressive} new point in the design space of literal notation definition systems.
+Overall, we believe that TLMs represent a distinctly \emph{reasonable} and \emph{expressive} new point in the design space of literal notation definition systems and macro systems.
 \vspace{-5px}
@@
 %% Bibliography
-\bibliography{../../../papers/research}
+\bibliography{research}
 

### 2018-07-16 icfp18/supplement/icfp18-supplement.tex
changed subsubsection formatting per publisher request

@@
 
-\usepackage[colorlinks=true,allcolors=Blue,backref,pageanchor=true,plainpages=false, pdfpagelabels, bookmarks,bookmarksnumbered,
+\usepackage[colorlinks=true,allcolors=Blue,pageanchor=true,plainpages=false, pdfpagelabels, bookmarks,bookmarksnumbered,
 pdfborder={0 0 0},  %removes outlines around hyper links in online display
@@
 \usepackage{ stmaryrd }
-\usepackage[numbers,sort]{natbib}
+\usepackage[numbers]{natbib}
 
@@
 \abstract{
-This report presents a complete technical account of the formal system that was described in the accompanying paper, as well as more details on the quasiquotation TLM.
+This report presents a complete technical account of the formal system that was described in the accompanying paper \cite{tlms-icfp18}, as well as more details on the quasiquotation TLM.
 }
@@
 \support{This work was
-supported in part by AFRL and DARPA under agreement \#FA8750-16-2-0042;
+supported in part by AFRL and DARPA under agreement \#FA8750-16-2-0042 and 
  by the NSA under lablet contract \#H98230-14-C-0140.}
@@
 \bibliographystyle{plainnat}
-\bibliography{../../../../papers/research} %your bib file
+\bibliography{../research} %your bib file
 
